### 2009-09-28 ISBMS09/paper.tex
ADD: first draft of ISBMS 09 paper
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@59 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\documentclass{llncs}

\usepackage[english]{babel}
\usepackage[tight]{subfigure}
% \usepackage{floatrow}
\usepackage{epsfig}
\usepackage{amsmath}

\begin{document}

\title{A simple shell model for real-time simulation of intra-ocular implant deployment}
\author{Olivier Comas, St\'ephane Cotin and Christian Duriez}
\institute{INRIA, Alcove team, Lille, France}

\maketitle

\begin{abstract}
With 30 million interventions a year worldwide, cataract surgery is one of the most frequently performed procedures. Yet, no tool currently allows teaching all steps of the procedure without putting patients at risk. A particularly challenging stage of this surgery deals with the injection and deployment of the intraocular lens implant. In this paper we propose to rely on thin plate theory to accurately describe the large bending of the implant. Our approach extends the co-rotational method used in finite element analysis to incorporate a bending energy. This results in a relatively simple and computationally efficient approach which was applied to the simulation of the lens' deployment. This simulation also accounts for the complex contacts that take place between the implant and the injection device. Particular care was given to the validation of the proposed model as well as its stability. 
\end{abstract}

\section{Introduction}
According to the latest assessment of the World Health Organization, age related cataract is responsible for $48\,\%$ of world blindness, which represents about 18 million people. Cataract is the clouding of the lens of the eye which impedes the passage of light. Its treatment is a surgical operation, which is very successful in restoring sight. The opaque lens is removed and replaced by an artificial intraocular lens. As people in the world live longer, the number of people with cataract is growing.  Current methods of teaching based on companionship not only expose patient to complications inherent to the inexperience of the operator, but they are also not very adequate with the increasing demands in training for this procedure. The development of a cataract surgery simulator could offer a new way of learning for young ophthalmologists. This work focuses on intraocular lens modelling to simulate the crucial step of the surgery dealing with the lens injection and its deployment within the lens capsule.

The implant is a three-dimensional structure, thin in one direction and long in the other two directions and may be characterised as a thin object. Thin objects are usually classified in two categories: plates and shells. The difference between these two kinds of structure is very well explained by Liu \cite{Liu03}. A plate structure carries only transversal loads that lead to bending deformation. As an example, consider the horizontal boards of a bookshelf that supports books. Those boards can be approximated as a plate structure and the transversal loads are of course the weight of the books. Conversely, a shell structure carries loads in all directions, and therefore undergoes bending, twisting and in-plane deformation, like the fuselage of an aircraft. Considering the folded position of the implant within the injection device (figure \ref{fig-surgery}), we need to model the intraocular implant as a shell. 
% + 2 EXAMPLES TO ILLUSTRATE THE DIFFERENCE BETWEEN PLATE AND SHELL

%\begin{figure}[h]
%\begin{center}
%\includegraphics[width=6cm]{images/deployment}
%\caption [Implant within injector] {Insertion of the lens implant through the injector}
%\label{fig-injector}
%\end{center}
%\end{figure}

\subsection{Related work}

Numerous models are available in the literature to describe physics of thin objects, from fairly simple and naive approaches to more complex and thourought representations. Continuum mechanics provides many formulations able to accurately describe stresses occuring within thin objects, all of them fall into two categories depending whether they are based on plate or shell theory. Those theories have been a subject of interest in the mechanical community for decades. Plate theory can be inferred from shell theory by considering the special case of planar midsurfaces \cite{Chapelle03}. Indeed it is often not necessary to use three-dimensional elasticity theory like in shell theory to model plates. Simple two-dimensional theories that account for bending deformation of thin bodies have been developped, and they are known as plate theories \cite{Reddy05}. However, formulations based on continuum mechanics are often quite complex and they are not always suitable for real-time applications. Therefore many researchers have tried to circumvent this limit. One of the most successful approach is the mass-spring model \cite{Provot95,Hammer08,Yu08}. Not only elasticity parameters included in mass-springs systems have no relation with the physical parameters but this type of model does not account for bending. Another type of modelling is position based approaches, which rely on projection as a complete simulation method \cite{Muller07}. Those approaches use impulses instead of forces to control the animation. Because impulses directly change velocities, one level of integration can be skipped. While this is an interesting approach for computer game industry, the main drawback of this method is that it is not physically accurate. So is the approach of Grinspun \emph{et al.} \cite{Grinspun03}, targeted at computer animation, which geometrically derives a discrete model for thin shells. Finite difference method \cite{Terzopoulos87} and finite element method have also been widely used, especially in cloth modelling, which is a very active field in computer graphics \cite{Volino00}. 
% TOO MANY REFERENCES OUTSIDE SHELLS?

\subsection{Our approach}

Our idea is to create a shell element by combining a classic two-dimensional element to handle the in-plane forces (membrane) and a plate element for off-plane forces (bending and twisting). However our real-time requirements demands a computationally inexpensive formulation and we would like to introduce the co-rotational framework to thin plate theory. Indeed co-rotational approaches have been successfully applied to real-time simulation of volumetric objects by the community over the last few years. They offer a good trade-off between computational efficiency and accuracy by allowing materially linear FEM to handle large deformations. We propose to improve and extend a plate model first introduced by Przemieniecki \cite{Przemieniecki68} to a co-rotational formulation. Once combined with a two-dimensional formulation, we obtain an accurate, yet computationally efficient, shell finite element method featuring both membrane and bending energies.

\section{Plate model's description}

\subsection{Thin plate theory}

Since the computation of the membrane stiffness matrix is classic and presents no difficulty, we will only detail the bending part of the formulation. To calculate the stiffness matrix for the transverse deflections and rotations shown on figure \ref{fig-triangle}, we start with the assumed deflection $u_z$ of the form

\begin{equation}
 u_z = c_1 + c_2x + c_3y + c_4x^2 + c_5xy + c_6y^2 + c_7x^3 + c_8xy^2 + c_9y^3
\label{eq-deflection}
\end{equation}

where $c_1$, \ldots , $c_9$ are constants. An issue of symmetry was observed with the deflection proposed by Przemieniecki, but was corrected by removing the term $x^2y$ in factor of the 8th constant. These constants can be evaluated in terms of the displacements and slopes at the three corners of the triangular plate using 
\begin{equation}
\textbf{u} = \textbf{Cc}
\label{eq-U}
\end{equation} 
where $\textbf{u} = \left\{u_1 u_2 \ldots u_9 \right\} $ and $\textbf{c} = \left\{c_1 c_2 \ldots c_9 \right\} $ while the matrix $\textbf{C}$ is given by
\begin{equation}
C = 
	\begin{bmatrix}
	1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 	0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
	0 & -1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
	1 & x_2 & 0 & x_2^2 & 0 & 0 & x_2^3 & 0 & 0 \\
	0 & 0 & 1 & 0 & x_2 & 0 & 0 & 0 & 0 \\
	0 & -1 & 0 & -2x_2 & 0 & 0 & -3x_2 & 0 & 0 \\
	1 & x_3 & y_3 & x_3^2 & x_3y_3 & y_3^2 & x_3^2 & x_3y_3^2& y_3^3 \\
	0 & 0 & 1 & 0 & x_3 & 2y_3 & 0 & 2x_3y_3 & 3y_3^2 \\
	0 & -1 & 0 & -2x_3 & -y_3 & 0 & -3x_3^2 & -y_3^2 & 0
	\end{bmatrix}
\end{equation} 
in deriving equation \ref{eq-U}. The notation

\begin{equation}
(u_z)_{x_1,y_1} = u_1 \hspace{1cm} \left(\frac{\partial u_z}{\partial y}\right)_{x_1,y_1} = u_2 \hspace{1cm} \left(\frac{\partial u_z}{\partial x}\right)_{x_1,y_1} = -u_3
\end{equation} 
and so on, has been used.

\begin{figure}
\begin{center}
\includegraphics[width=12cm]{images/triangle}
\caption {Triangular plate element in bending}
\label{fig-triangle}
\end{center}
\end{figure}

We can calculate the strains from the flat-plate theory using
\begin{equation}
e_{xx} = -z \frac{\partial^2u_z}{\partial x^2}
\end{equation} 
\begin{equation}
e_{yy} = -z \frac{\partial^2u_z}{\partial y^2}
\end{equation} 
\begin{equation}
e_{xy} = -2z \frac{\partial^2u_z}{\partial x \partial y}
\end{equation} 
Hence, using the above equations and equation \ref{eq-deflection}, we have
\begin{equation}
\begin{bmatrix}
e_{xx} \\
e_{yy} \\
e_{xy}
\end{bmatrix}
= 
-z
\begin{bmatrix}
0 & 0 & 0 & 2 & 0 & 0 & 6x & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 2 & 0 & 2x & 6y \\
0 & 0 & 0 & 0 & 2 & 0 & 0 & 4y & 0 \\
\end{bmatrix}
\textbf{c}
\label{eq-strains}
\end{equation} 
or symbolically $\textbf{e} = \textbf{Dc}$ where $\textbf{D}$ stands for the $3\times 9$ matrix in equation \ref{eq-strains}, including the premultiplying constant $-z$. Noting from equation \ref{eq-U} that $\textbf{c} = \textbf{C}^{-1}\textbf{u}$, we have 
\begin{equation}
\textbf{e} = \textbf{DC}^{-1}\textbf{u} = \textbf{bu}
\end{equation} 
where $\textbf{b} = \textbf{DC}^{-1}$. 
Knowing that the stiffness matrix $\textbf{K}_e$ for an element is obtained from
\begin{equation}
\textbf{K}_e = \int_v \textbf{b}^{T} \boldsymbol\chi \textbf{b} dV
\end{equation} 
where $\boldsymbol\chi$ is the material matrix. The substitution of $\textbf{b}$ into this expression yields
\begin{equation}
\textbf{K}_e = (\textbf{C}^{-1})^T \int_v \textbf{D}^{T} \boldsymbol\chi \textbf{D} dV \textbf{C}^{-1}
\end{equation} 
The integration is carried out numerically using Gauss points located at the centre of each edge.

\subsection{Implementation}

This plate formulation has been successfully implemented into the open-source framework SOFA \cite{SOFA}. In practice the formulation is very simple and for each triangle it consists of the following steps:

\begin{enumerate}
\item compute the rotation matrix from global to triangle (local) frame
\item measure the orientation of each vertex in the local frame and compare it to the local orientation obtained for the rest position to compute the local displacement vector $\textbf{u} = \left\{0 \, \Omega _{Ax} \, \Omega _{Ay}  \hspace{0.5cm} 0 \, \Omega _{Bx} \, \Omega _{By}  \hspace{0.5cm}  0 \, \Omega _{Cx} \, \Omega _{Cy} \right\}$. Indeed in a co-rotational framework the normal displacement of each vertex is locally null and only the rotation around the local $x$ and $y$ axes need to be measured at each of the three vertices A, B and C.
\item the matrix $\textbf{D}_i$ is expressed at each Gauss point $i$
\item the strain-displacement matrix at each Gauss point $i$ is calculated with $\textbf{J}_i = \textbf{D}_i \textbf{C}^{-1}$
\item the local stiffness matrix of the element $\textbf{K}_e$ is then obtained $\textbf{K}_e = \displaystyle{\sum^3_{i=1}} \textbf{J}_i \boldsymbol\chi \textbf{J}_i^T$
\item if a direct solver is used, the local element stiffness matrix is transformed into the global frame and added to the global stiffness matrix
\end{enumerate}

\section{Validation}

We compared the model with some theoretical results to assess its quality in modelling bending. The test that we carried out uses a square shape mesh clamped on the edges. A uniform pressure is then applied on the square and the maximum deflection $z_{max}$ at the centre can be calculated. In his paper Zhongnian \cite{Zhongnian86} uses the following parameters:
\begin{itemize}
 \item Young Modulus $E = 1.092 \times 10^6$
 \item Poisson's ratio $\nu = 0.3$
 \item edge of the square $L = 10\,m$
 \item thickness $h = 0.1\,m$
 \item pressure q is a uniformly distributed load per unit area
\end{itemize}

It can be shown that $z_{max} = 0.126\,q$. We experimentally measured the maximum deflection for different values of pressure $q$ (see table \ref{tab-results}). In average we found $z_{max} = 0.1194\,q $. Hence a $5\,\%$ error between our model and the theory on that test.

\begin{table}[h!]
	\begin{center}
		\begin{tabular}{|p{2cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
		\hline
		 \centering q & \centering 1 & \centering 2 & \centering 3 & \centering 8 \tabularnewline
		\hline
		\centering $z_{max}$ & \centering 0.11 & \centering 0.23 & \centering 0.48 & \centering 1.09 \tabularnewline
		\hline
		\end{tabular}
	\vspace{0.3cm}
	\caption{Experimental results}
	\label{tab-results}
	\end{center}
\end{table}

\section{Injection and deployment of the intraocular implant}
\subsection{Description of cataract surgery procedure}
Cataract surgery is the removal of the natural lens of the eye (also called \emph{crystalline lens}) that has developed an opacification, which is referred to as a cataract. Metabolic changes of the crystalline lens fibers over the time lead to the development of the cataract and loss of transparency, causing impairment or loss of vision. During cataract surgery, a patient's cloudy natural lens is removed and replaced with a synthetic lens to restore the lens' transparency (figure \ref{fig-surgery}). The last step in the simulation of this surgery is the injection of the implant and its deployment within the lens capsule.

\begin{figure}[h]
\begin{center}
\includegraphics[height=4.5cm]{images/suction}
\hspace{0.2cm}
\includegraphics[height=4.5cm]{images/deployment}
\caption [Cataract surgery] {Left: removal of the opacified lens. Right: insertion of the lens implant.}
\label{fig-surgery}
\end{center}
\end{figure}

%\begin{figure}[h]
%\begin{center}
%\includegraphics[height=4.5cm]{images/implant_injection}
%\caption [Cataract surgery] {Left: insertion of the lens implant. Right: deployment within the eye}
%\label{fig-surgery2}
%\end{center}
%\end{figure}

\subsection{Simulation}

A mesh of lens implant with the proper geometry has been designed with Blender (figure \ref{fig-implant}). The physical parameters have been provided by the manufacturer Alcon (table \ref{tab-parameters}).

\begin{table}[h!]
	\begin{center}
		\begin{tabular}{|p{3cm}|p{3cm}|p{3cm}|}
		\hline
		 \centering Young's modulus & \centering Poisson's ratio & \centering Mass density \tabularnewline
		\hline
		\centering $1 MPa$ & \centering $0.42$ & \centering $1.2 g/cm^3$ \tabularnewline
		\hline
		\end{tabular}
	\vspace{0.3cm}
	\caption{Physical parameters of the intraocular implant (source: Alcon)}
	\label{tab-parameters}
	\end{center}
\end{table}

We start by folding the haptics onto the implant body, then the body is bent while keeping the haptics inside to obtain the same position as the one chosen by the surgeon when he puts the implant inside the injection device (figure \ref{fig-implantFolded}). The bent implant is then placed into the injection device. The simulation eventually consists of pushing the intraocular implant within the injection device into the eye. We can notice the progressive deployment of the implant when it slowly gets out of the injector. 
% + SCREENSHOTS OF SIMULATION

\begin{figure}[!h]
\begin{center}
\includegraphics[height=6.5cm]{images/IOL}
\hspace{1cm}
\includegraphics[height=6.5cm]{images/mesh_implant}
\caption [Lens implant and its mesh] {The real implant and the mesh created with Blender (473 triangles)}
\label{fig-implant}
\end{center}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[height=5.5cm]{images/implant_haptics_in}
\includegraphics[height=5.5cm]{images/implant_folded}
\caption [Lens imlant] {Folding of the intraocular implant}
\label{fig-implantFolded}
\end{center}
\end{figure}

\section{Conclusion}

We improved a plate finite element formulation by extending it to a co-rotational framework in order to handle large deformations. By combining this formulation to a classic two-dimensional element we created a simple shell element able to efficiently handle both membrane and bending forces. The validity of our approach has been demonstrated though comparison with theoretical results and a non-trivial application in cataract surgery simulation.
% + NEXT STEPS

 \bibliographystyle{splncs}
 \bibliography{bibliography}

\end{document}

### 2009-10-16 ISBMS09/paper.tex
J'ai modifie les valeurs obtenues lors du test theorique et recalcule l'erreur
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@60 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
Ajout de testContatcs.scn pour tester les contacts d'une beam dans un cylindre controle par les shells + mise a jour du papier ISBMS
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@67 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \section{Plate model's description}
+
+\subsection{Triangular elastic membrane}
+The computation of the triangular elastic membrane stiffness matrix can be simply derived from previous works dealing with tetrahedral co-rotational elements (see Muller \emph{et al.} \cite{Muller04}) for instance.
+The element stiffness matrix $\textbf{K}_e$ can be computed as follows:
+
+\begin{equation}
+\textbf{K}_e = \int_v \textbf{J} \boldsymbol\chi \textbf{J}^{T} dV
+\end{equation}
+
+where $\textbf{J}$ is the strain-displacement matrix and $\boldsymbol\chi$ embodies the material's behaviour. In the simple case of Hooke's law we have:
+
+\begin{equation}
+\boldsymbol\chi = \frac{E}{12(1-\nu^2)}
+	\begin{bmatrix}
+	1 & \nu & 0 \\
+ 	\nu & 1 & 0 \\
+	0 & 0 & \frac{1}{2} (1-\nu)
+	\end{bmatrix}
+\end{equation}
+
+The stiffness matrix in the global frame is eventually obtained using the rotation matrix of the element: $\textbf{K} = \textbf{R} \textbf{K}_e \textbf{R}^{T}$.
 
@@
 We experimentally measured the maximum deflection for different values of pressure $q$ (see table \ref{tab-results}).
-In average we found $z_{max} = 0.1194\,q $.
-Hence a $5\,\%$ error between our model and the theory on that test.
+In average we found $z_{max} = 0.1248\,q $.
+Hence a $0.93\,\%$ error between our model and the theory on that test.
 
@@
 	\begin{center}
-		\begin{tabular}{|p{2cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
+		\begin{tabular}{|p{1.7cm}|p{1.7cm}|p{1.7cm}|p{1.7cm}|p{1.7cm}|p{1.7cm}|}
 		\hline
-		 \centering q & \centering 1 & \centering 2 & \centering 3 & \centering 8 \tabularnewline
+		 \centering q & \centering 1 & \centering 2 & \centering 3 & \centering 4 & \centering 5 \tabularnewline
 		\hline
-		\centering $z_{max}$ & \centering 0.11 & \centering 0.23 & \centering 0.48 & \centering 1.09 \tabularnewline
+		\centering $z_{max}$ & \centering 0.1218 & \centering 0.2475 & \centering 0.3747 & \centering 0.505 & \centering 0.637 \tabularnewline
 		\hline
@@
 \end{figure}
+
+% Ref suction : http://www.coastal-eye-care-maine.com/cataracts-surgery/cataract-surgery-maine.html
 

### 2009-11-10 ISBMS09/paper.tex
Updated version of ISBMS paper with first small changes for camera-ready paper
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@68 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2009-09-28 ISBMS09/paper.tex
@@
 \usepackage{amsmath}
+\usepackage{multirow}
+\renewcommand*\arraystretch{1.25} % extra vertical space in cells
 
@@
 
-\title{A simple shell model for real-time simulation of intra-ocular implant deployment}
+\title{A Shell Model for Real-Time Simulation of Intra-ocular Implant Deployment}
 \author{Olivier Comas, St\'ephane Cotin and Christian Duriez}
@@
 Yet, no tool currently allows teaching all steps of the procedure without putting patients at risk.
-A particularly challenging stage of this surgery deals with the injection and deployment of the intraocular lens implant.
-In this paper we propose to rely on thin plate theory to accurately describe the large bending of the implant.
-Our approach extends the co-rotational method used in finite element analysis to incorporate a bending energy.
-This results in a relatively simple and computationally efficient approach which was applied to the simulation of the lens' deployment.
-This simulation also accounts for the complex contacts that take place between the implant and the injection device.
-Particular care was given to the validation of the proposed model as well as its stability.
+A particularly challenging stage of this surgery deals with the injection and deployment of the intra-ocular lens implant.
+In this paper we propose to rely on shell theory to accurately describe the complex deformations of the implant.
+Our approach extends the co-rotational method used in finite element analysis of in-plane deformations to incorporate a bending energy.
+This results in a relatively simple and computationally efficient approach which was applied to the simulation of the lens deployment.
+This simulation also accounts for the complex contacts that take place during the injection phase.
+%Particular care was given to the validation of the proposed model, its computational efficiency and stability. 
 \end{abstract}
@@
 Cataract is the clouding of the lens of the eye which impedes the passage of light.
-Its treatment is a surgical operation, which is very successful in restoring sight.
-The opaque lens is removed and replaced by an artificial intraocular lens.
-As people in the world live longer, the number of people with cataract is growing.
-Current methods of teaching based on companionship not only expose patient to complications inherent to the inexperience of the operator, but they are also not very adequate with the increasing demands in training for this procedure.
-The development of a cataract surgery simulator could offer a new way of learning for young ophthalmologists.
-This work focuses on intraocular lens modelling to simulate the crucial step of the surgery dealing with the lens injection and its deployment within the lens capsule.
+Its treatment is a surgical operation, where the opaque lens is removed and replaced by an artificial intra-ocular lens.
+As people live longer, more surgical interventions are needed.
+However, current methods of teaching are not compatible with increasing demands of training for this procedure, and also expose patients to complications inherent to the inexperience of the operator.
+A cataract surgery simulator can offer an efficient alternative teaching method for ophthalmologists, as illustrated in recent reports such as \cite{Feudner09}.
+The work presented in this paper focuses on the simulation of a crucial step of cataract surgery: the injection and deployment of an artificial lens.
+To reach this objective, we propose an efficient yet simple formulation for triangular shells, as well as a physically accurate model of the complex (self-)contacts that occur during the injection and deployment of the artificial lens.
 
-The implant is a three-dimensional structure, thin in one direction and long in the other two directions and may be characterised as a thin object.
-Thin objects are usually classified in two categories: plates and shells.
-The difference between these two kinds of structure is very well explained by Liu \cite{Liu03}.
-A plate structure carries only transversal loads that lead to bending deformation.
-As an example, consider the horizontal boards of a bookshelf that supports books.
-Those boards can be approximated as a plate structure and the transversal loads are of course the weight of the books.
-Conversely, a shell structure carries loads in all directions, and therefore undergoes bending, twisting and in-plane deformation, like the fuselage of an aircraft.
-Considering the folded position of the implant within the injection device (figure \ref{fig-surgery}), we need to model the intraocular implant as a shell.
-% + 2 EXAMPLES TO ILLUSTRATE THE DIFFERENCE BETWEEN PLATE AND SHELL
+\subsection{Related work}
 
-%\begin{figure}[h]
+Although various shapes of intra-ocular lens implants exist they all share common properties: they are three-dimensional structures, usually made of acrylic
+%\footnote{acrylic is the common name for polymethylmethacrylate} 
+material, and are much thinner in one direction than in the two other ones.
+This last characteristic is what classifies these lenses as thin objects, i.e.
+three-dimensional objects that can be abstracted as two-dimensional structures.
+Numerous models are available in the literature to describe physics of thin objects, from fairly simple and naive approaches to more complex and thorough representations.
+Continuum mechanics provides many formulations able to accurately describe stresses occurring within thin objects.
+Most of them fall into one of the following two categories: plate theory or shell theory.
+Those theories have been a subject of interest in the mechanical community for decades.
+The difference between these two kinds of structures is very well explained by Liu \textit{et al.}~\cite{Liu03} and can be summarized by the fact that plate bending elements can only carry transversal loads while shells can undergo more complex deformations.
+%For instance, if we consider the horizontal board of a bookshelf, that board can be approximated as a plate structure and the transversal loads are the weight of the books.
+A typical deformation of the board is illustrated in figure~\ref{fig-boards}(a).
+Conversely, a shell structure can carry loads in all directions, and therefore undergoes bending, twisting and in-plane deformation (see Figure~\ref{fig-boards}(b)).
+Considering the folded shape of the implant within the injection device (see Figure \ref{fig-surgery}), a thin plate model would not correctly capture the deformation and therefore a shell formulation was retained to model the intra-ocular lens behaviour.
+
+%\begin{figure}[ht]
 %\begin{center}
-%\includegraphics[width=6cm]{images/deployment}
-%\caption [Implant within injector] {Insertion of the lens implant through the injector}
-%\label{fig-injector}
+%\includegraphics[width=4.5cm]{images/intraocular_implant_1}
+%\hspace{1.5cm}
+%\includegraphics[width=4.5cm]{images/intraocular_implant_2}
+%\caption{Different types of lens implants.
+All share similar properties: they are three-dimensional structures, usually made of acrylic material, and are much thinner in one direction than in the two other ones.}
+%\label{fig-implants}
 %\end{center}
@@
 
-\subsection{Related work}
+%\begin{figure}[ht]
+%\begin{center}
+%\includegraphics[width=5.5cm]{images/board_bending}
+%\hfill
+%\includegraphics[width=5.2cm]{images/board_twist}
+%\caption{Illustration of a key difference between different models of thin objects.
+While thin plate theory allows to describe bending (left), it cannot represent more complex deformations such as twist (right) which is captured by shell theory.}
+%\label{fig-boards}
+%\end{center}
+%\end{figure}
 
-Numerous models are available in the literature to describe physics of thin objects, from fairly simple and naive approaches to more complex and thourought representations.
-Continuum mechanics provides many formulations able to accurately describe stresses occuring within thin objects, all of them fall into two categories depending whether they are based on plate or shell theory.
-Those theories have been a subject of interest in the mechanical community for decades.
-Plate theory can be inferred from shell theory by considering the special case of planar midsurfaces \cite{Chapelle03}.
-Indeed it is often not necessary to use three-dimensional elasticity theory like in shell theory to model plates.
-Simple two-dimensional theories that account for bending deformation of thin bodies have been developped, and they are known as plate theories \cite{Reddy05}.
-However, formulations based on continuum mechanics are often quite complex and they are not always suitable for real-time applications.
-Therefore many researchers have tried to circumvent this limit.
-One of the most successful approach is the mass-spring model \cite{Provot95,Hammer08,Yu08}.
-Not only elasticity parameters included in mass-springs systems have no relation with the physical parameters but this type of model does not account for bending.
-Another type of modelling is position based approaches, which rely on projection as a complete simulation method \cite{Muller07}.
-Those approaches use impulses instead of forces to control the animation.
-Because impulses directly change velocities, one level of integration can be skipped.
-While this is an interesting approach for computer game industry, the main drawback of this method is that it is not physically accurate.
-So is the approach of Grinspun \emph{et al.} \cite{Grinspun03}, targeted at computer animation, which geometrically derives a discrete model for thin shells.
-Finite difference method \cite{Terzopoulos87} and finite element method have also been widely used, especially in cloth modelling, which is a very active field in computer graphics \cite{Volino00}.
-% TOO MANY REFERENCES OUTSIDE SHELLS?
+Development of a satisfactory physical model that runs in real-time but produces visually convincing animation of thin objects has been a challenge in Computer Graphics, particularly in the area of cloth modelling.
+Rather than resorting to shell theory which involves the most complex formulations in continuum mechanics, previous works have often relied on discrete formulations.
+Early approaches to cloth modelling only considered in-plane deformation, and often relied on mass-spring models (see \cite{Provot95}, \cite{Hammer08} or \cite{Yu08} for instance).
+More recent works have considered adding bending through angular springs.
+For instance Taskiran \emph{et al.} \cite{Taskiran05} have chosen to use a linear representation of angular springs to supply bending and torsion effects in hair modelling, which allowed them to simulate curly hair with good realism.
+Wang \emph{et al.} \cite{Wang07} successfully used a network of linear and angular springs to describe bending and twisting of catheters and guidewires in an interventional radiology simulator.
+Yet, such models are limited in their ability to describe certain behaviour, as they do not rely on continuum mechanics.
+Another limitation of such models is the difficulty to derive spring stiffness (in particular for angular springs) from elastic properties (Young's modulus and Poisson's ratio).
+For these reasons, other approaches have been proposed.
+Among the different models introduced recently we can mention the work of Choi \emph{et al.} \cite{Choi07} and Bridson \emph{et al} \cite{Bridson03}.
+Choi \emph{et al.} proposed a real-time simulation technique for thin shells undergoing large deformations.
+The authors adopt the energy functions from the discrete shells proposed by \cite{Grinspun03}.
+For real-time integration of the governing equation, they adapted a modal analysis technique, called modal warping.
+The resulting simulations run in real-time even for large meshes, and the model can handle large bending and/or twisting deformations with acceptable realism.
+Bridson \emph{et al.} followed the same energy approach to derive their bending model but improved the resolution of the equations by suggesting a novel mixed implicit/explicit integration scheme.
+They also presented a post-processing method for treating cloth-character collisions that preserves folds and wrinkles.
+Pabst \emph{et al.} improved the bending modelling used by Bridson \emph{et al.} to allow the integration of measured material data.
+A method to use thin shell dynamics with point sampled surfaces for efficient animation was recently proposed by Wicke \emph{et al.} \cite{Wicke05} where the curvature of the shell is measured through the use of fibers.
 
-\subsection{Our approach}
+In this paper we consider the problem of lens implant folding, unfolding and deployment within a very constrained space.
+An important difference with previous works is that the ratio stiffness / mass of artificial lenses is very high compared to many material (such as fabric), resulting in added difficulties to obtain a numerically stable simulation.
+The closest work to what we propose is the one of Thomaszewski \emph{et al.} \cite{Thomaszewski06} who derive their formulation from thin shell analysis in a co-rotational framework.
+However, the authors use very soft and deformable material (Young's modulus $E = 5000\,Pa$ and Poisson's ratio $\nu = 0.25$ while the stiffness of intra-ocular implants are three orders of magnitude higher).
 
-Our idea is to create a shell element by combining a classic two-dimensional element to handle the in-plane forces (membrane) and a plate element for off-plane forces (bending and twisting).
-However our real-time requirements demands a computationally inexpensive formulation and we would like to introduce the co-rotational framework to thin plate theory.
-Indeed co-rotational approaches have been successfully applied to real-time simulation of volumetric objects by the community over the last few years.
-They offer a good trade-off between computational efficiency and accuracy by allowing materially linear FEM to handle large deformations.
+\section{Co-rotational triangular shell model}
+
+We propose to define a triangular shell element by combining a two-dimensional in-plane membrane energy, with an off-plane energy for describing bending and twist (see figure \ref{fig-triangle}).
+To allow for real-time simulation, a computationally efficient formulation is needed.
+We therefore propose to extend the co-rotational idea introduced by Fellippa in \cite{Felippa00}.
+Indeed co-rotational approaches have been successfully applied to real-time simulation of volumetric objects over the last few years \cite{Muller04}.
+They offer a good trade-off between computational efficiency and accuracy by allowing small deformations but large displacements.
 We propose to improve and extend a plate model first introduced by Przemieniecki \cite{Przemieniecki68} to a co-rotational formulation.
-Once combined with a two-dimensional formulation, we obtain an accurate, yet computationally efficient, shell finite element method featuring both membrane and bending energies.
+Once combined with an in-plane membrane formulation we obtain an accurate, yet computationally efficient, shell finite element method featuring both membrane and bending energies.
 
-\section{Plate model's description}
+\begin{figure}
+\centering
+\includegraphics[height=3cm]{images/triangle_membrane}
+\hspace{1cm}
+\includegraphics[height=3.25cm]{images/triangle_bending}
+\caption {A triangular shell element can be defined as a combination of a triangular in-plane membrane element (left) and a triangular thin plate in bending (right).
+The different degrees of freedom $v$ and $u$ of both models are illustrated above.}
+\label{fig-triangle}
+\end{figure}
 
-\subsection{Thin plate theory}
+\subsection{Triangular elastic membrane}
 
-Since the computation of the membrane stiffness matrix is classic and presents no difficulty, we will only detail the bending part of the formulation.
+The computation of the triangular elastic membrane stiffness matrix can be derived from previous works dealing with tetrahedral co-rotational elements (see Muller \emph{et al.} \cite{Muller04} for instance).
+The element stiffness matrix $\textbf{K}_e$ can be computed as follows:
+
+\begin{equation}
+\textbf{K}_e = \int_v \textbf{J} \boldsymbol\chi \textbf{J}^{T} dV
+\end{equation}
+
+where $\textbf{J}$ is the strain-displacement matrix and $\boldsymbol\chi$ embodies the material's behaviour. In the simple case of Hooke's law we have:
+
+\begin{equation}
+\boldsymbol\chi = \frac{E}{12(1-\nu^2)}
+\begin{bmatrix}
+1 & \nu & 0 \\
+\nu & 1 & 0 \\
+0 & 0 & \frac{1}{2} (1-\nu)
+\end{bmatrix}
+\end{equation}
+
+The stiffness matrix in the global frame is eventually obtained using the rotation matrix of the element: $\textbf{K} = \textbf{R} \textbf{K}_e \textbf{R}^{T}$ where $\textbf{R}$ describes the rotation of the (triangular) element with respect to its initial configuration.
+
+\subsection{Triangular plate bending}
+
 To calculate the stiffness matrix for the transverse deflections and rotations shown on figure \ref{fig-triangle}, we start with the assumed deflection $u_z$ of the form
-
 \begin{equation}
@@
 where $c_1$, \ldots , $c_9$ are constants.
-An issue of symmetry was observed with the deflection proposed by Przemieniecki, but was corrected by removing the term $x^2y$ in factor of the 8th constant.
+Equation~\ref{eq-deflection} solves an issue of symmetry which was observed with the deflection function proposed by Przemieniecki in \cite{Przemieniecki68}.
 These constants can be evaluated in terms of the displacements and slopes at the three corners of the triangular plate using
@@
 \end{equation} 
-where $\textbf{u} = \left\{u_1 u_2 \ldots u_9 \right\} $ and $\textbf{c} = \left\{c_1 c_2 \ldots c_9 \right\} $ while the matrix $\textbf{C}$ is given by
+where $\textbf{u} = \left\{u_1 u_2 \ldots u_9 \right\} $ and $\textbf{c} = \left\{c_1 c_2 \ldots c_9 \right\} $. Matrix $\textbf{C}$ derives from equation~\ref{eq-U}:
 \begin{equation}
@@
 \end{equation} 
-in deriving equation \ref{eq-U}. The notation
+The notations 
+\begin{equation}
+u_1 = (u_z)_{x_1,y_1} \hspace{1cm} u_2 = \left(\frac{\partial u_z}{\partial y}\right)_{x_1,y_1} \hspace{1cm} u_3 = - \left(\frac{\partial u_z}{\partial x}\right)_{x_1,y_1}
+\end{equation} 
+and so on for the two other vertices, were used.
 
-\begin{equation}
-(u_z)_{x_1,y_1} = u_1 \hspace{1cm} \left(\frac{\partial u_z}{\partial y}\right)_{x_1,y_1} = u_2 \hspace{1cm} \left(\frac{\partial u_z}{\partial x}\right)_{x_1,y_1} = -u_3
-\end{equation} 
-and so on, has been used.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=12cm]{images/triangle}
-\caption {Triangular plate element in bending}
-\label{fig-triangle}
-\end{center}
-\end{figure}
-
-We can calculate the strains from the flat-plate theory using
-\begin{equation}
-e_{xx} = -z \frac{\partial^2u_z}{\partial x^2}
-\end{equation} 
-\begin{equation}
-e_{yy} = -z \frac{\partial^2u_z}{\partial y^2}
-\end{equation} 
-\begin{equation}
-e_{xy} = -2z \frac{\partial^2u_z}{\partial x \partial y}
-\end{equation} 
-Hence, using the above equations and equation \ref{eq-deflection}, we have
+We can calculate the strains from the flat-plate theory using $  e_{xx} = -z \frac{\partial^2u_z}{\partial x^2} $, $ e_{yy} = -z \frac{\partial^2u_z}{\partial y^2} $, and $ e_{xy} = -2z \frac{\partial^2u_z}{\partial x \partial y} $.
+Hence, using the above equations and equation (\ref{eq-deflection}), we have
 \begin{equation}
@@
 \end{equation} 
-or symbolically $\textbf{e} = \textbf{Dc}$ where $\textbf{D}$ stands for the $3\times 9$ matrix in equation \ref{eq-strains}, including the premultiplying constant $-z$.
-Noting from equation \ref{eq-U} that $\textbf{c} = \textbf{C}^{-1}\textbf{u}$, we have
+or symbolically $\textbf{e} = \textbf{Dc}$ where $\textbf{D}$ stands for the $3\times 9$ matrix in equation \ref{eq-strains}, including the pre-multiplying constant $-z$.
+Noting from equation (\ref{eq-U}) that $\textbf{c} = \textbf{C}^{-1}\textbf{u}$, we have:
 \begin{equation}
@@
 \end{equation} 
-where $\boldsymbol\chi$ is the material matrix. The substitution of $\textbf{b}$ into this expression yields
+where $\boldsymbol\chi$ is the material matrix, the substitution of $\textbf{b}$ into this expression yields
 \begin{equation}
@@
 \end{equation} 
-The integration is carried out numerically using Gauss points located at the centre of each edge.
+The integration is carried out numerically using Gauss points located at the middle of each edge of the triangle.
 
@@
 
-This plate formulation has been successfully implemented into the open-source framework SOFA \cite{SOFA}.
-In practice the formulation is very simple and for each triangle it consists of the following steps:
-
+In practical terms, the different computations associated with each triangular shell element can be described as follows:
+%
 \begin{enumerate}
-\item compute the rotation matrix from global to triangle (local) frame
-\item measure the orientation of each vertex in the local frame and compare it to the local orientation obtained for the rest position to compute the local displacement vector $\textbf{u} = \left\{0 \, \Omega _{Ax} \, \Omega _{Ay}  \hspace{0.5cm} 0 \, \Omega _{Bx} \, \Omega _{By}  \hspace{0.5cm}  0 \, \Omega _{Cx} \, \Omega _{Cy} \right\}$.
-Indeed in a co-rotational framework the normal displacement of each vertex is locally null and only the rotation around the local $x$ and $y$ axes need to be measured at each of the three vertices A, B and C.
-\item the matrix $\textbf{D}_i$ is expressed at each Gauss point $i$
-\item the strain-displacement matrix at each Gauss point $i$ is calculated with $\textbf{J}_i = \textbf{D}_i \textbf{C}^{-1}$
-\item the local stiffness matrix of the element $\textbf{K}_e$ is then obtained $\textbf{K}_e = \displaystyle{\sum^3_{i=1}} \textbf{J}_i \boldsymbol\chi \textbf{J}_i^T$
-\item if a direct solver is used, the local element stiffness matrix is transformed into the global frame and added to the global stiffness matrix
+\item Compute the rotation matrix $\textbf{R}$ from global to triangle (local) frame
+\item Compute the local displacement vector $\textbf{u} = \{v_1, v_2, 0, u_2, u_3, v_3, v_4, 0, u_5, u_6,$ $v_5, v_6, 0, u_8, u_9 \} $ for each of the 3 nodes of the triangle (the nodes have 6 degrees-of-freedom).
+As we are in a co-rotational framework the normal displacements $u_1, u_4, u_7$ of all 3 nodes are null in the local frame of the triangle.
+\item Compute matrix $\textbf{D}_i$  at each Gauss point $i$
+\item The strain-displacement matrix at each Gauss point $i$ is computed with $\textbf{J}_i = \textbf{D}_i \textbf{C}^{-1}$
+\item Compute the local stiffness matrix $\textbf{K}_e$ of the element as $\textbf{K}_e = \displaystyle{\sum^3_{i=1}} \textbf{J}_i \boldsymbol\chi \textbf{J}_i^T$
+\item Transform the local element stiffness matrix into the global frame and add it to a global stiffness matrix
 \end{enumerate}
 
-\section{Validation}
+The co-rotational shell formulation has been successfully implemented into the open-source framework SOFA \cite{SOFA}.
 
-We compared the model with some theoretical results to assess its quality in modelling bending.
-The test that we carried out uses a square shape mesh clamped on the edges.
-A uniform pressure is then applied on the square and the maximum deflection $z_{max}$ at the centre can be calculated.
-In his paper Zhongnian \cite{Zhongnian86} uses the following parameters:
-\begin{itemize}
- \item Young Modulus $E = 1.092 \times 10^6$
- \item Poisson's ratio $\nu = 0.3$
- \item edge of the square $L = 10\,m$
- \item thickness $h = 0.1\,m$
- \item pressure q is a uniformly distributed load per unit area
-\end{itemize}
+\subsection{Validation}
 
-It can be shown that $z_{max} = 0.126\,q$.
-We experimentally measured the maximum deflection for different values of pressure $q$ (see table \ref{tab-results}).
-In average we found $z_{max} = 0.1194\,q $.
-Hence a $5\,\%$ error between our model and the theory on that test.
+We compared our model with some theoretical results reported by Zhongnian \cite{Zhongnian86} to assess its quality in modelling bending.
+The test that we carried out uses a square shape mesh clamped on all four edges.
+A uniform load is then applied on the square and the maximum deflection $z_{max}$ at the centre can be calculated.
+Several simulations are performed with increasing load values $q$ (ranging from $1$ to $5 N/m^2$) and the following parameters were used: Young's Modulus $E = 1.092 \times 10^6 \,Pa$, Poisson's ratio $\nu = 0.3$, square edge length $L = 10\,m$, thickness $h = 0.1\,m$.
+Using these particular values it can be shown that $z_{max} = 0.126\,q$.
+The maximum deflection obtained in our simulations are reported in table \ref{tab-results}.
+In average we found $z_{max} = 0.1248\,q $, resulting in a $0.93\,\%$ error between our model and theoretical results on that test.
 
-\begin{table}[h!]
-	\begin{center}
-		\begin{tabular}{|p{2cm}|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
-		\hline
-		 \centering q & \centering 1 & \centering 2 & \centering 3 & \centering 8 \tabularnewline
-		\hline
-		\centering $z_{max}$ & \centering 0.11 & \centering 0.23 & \centering 0.48 & \centering 1.09 \tabularnewline
-		\hline
+\begin{table}[ht]
+	\centering
+	\begin{tabular}{p{8.5cm}|c|c|}
+	\cline{2-3}
+	\multirow{5}{*}{\includegraphics[height=2.75cm]{images/clamped_square}} & $q$ & $z_{max}$ \tabularnewline
+	\cline{2-3}
+	& $\,1\,$ & $\, 0.1218 \,$ \tabularnewline
+	& $\,2\,$ & $\, 0.2475 \,$ \tabularnewline		
+	& $\,3\,$ & $\, 0.3747 \,$ \tabularnewline	
+	& $\,4\,$ & $\, 0.5050 \,$ \tabularnewline		
+	& $\,5\,$ & $\, 0.6374 \,$ \tabularnewline
+	\cline{2-3}
 	\end{tabular}
-	\vspace{0.3cm}
-	\caption{Experimental results}
+	\caption{Comparison of our shell model with theoretical results on the bending of a square plate. An error of less than $1\,\%$ was found between our simulation and the theoretical results.}
 	\label{tab-results}
-	\end{center}
 \end{table}
 
-\section{Injection and deployment of the intraocular implant}
-\subsection{Description of cataract surgery procedure}
-Cataract surgery is the removal of the natural lens of the eye (also called \emph{crystalline lens}) that has developed an opacification, which is referred to as a cataract.
+\section{Simulation of the intra-ocular lens injection}
+
+%Cataract surgery is the removal of the natural lens of the eye (also called \emph{crystalline lens}) that has developed an opacification, which is referred to as a cataract.
 Metabolic changes of the crystalline lens fibers over the time lead to the development of the cataract and loss of transparency, causing impairment or loss of vision.
@@
 
+Cataract surgery consists in three main steps: capsulorhexis, phacoemulsification, and implantation of an intra-ocular lens.
+Capsulorhexis is the technique used to remove a part of the anterior lens capsule.
+Phacoemulsification consists in using a surgical device which tip vibrates at an ultrasonic frequency to emulsify the natural lens material and then aspirate the  fragments of the cortical material.
+After the removal of the diseased lens, an intra-ocular lens is implanted into the eye, through a small incision (about 2 mm) using a foldable intra-ocular lens (see figure \ref{fig-surgery}).
+The foldable lens, usually made of acrylic material, is then implanted within the lens capsule through the incision used during phacoemulsification.
+
 \begin{figure}[h]
 \begin{center}
-\includegraphics[height=4.5cm]{images/suction}
-\hspace{0.2cm}
-\includegraphics[height=4.5cm]{images/deployment}
-\caption [Cataract surgery] {Left: removal of the opacified lens. Right: insertion of the lens implant.}
+\includegraphics[height=3.2cm]{images/suction}
+\hfill
+\includegraphics[height=3.2cm]{images/implant_injection_step_1}
+\hfill
+\includegraphics[height=3.2cm]{images/implant_injection_step_2}
+\caption [Cataract surgery] {Left: removal of the opacified lens by phacoemulsification.
+Center: insertion of the lens implant which is folded inside the injection device and then deploys within the lens capsule.
+Right: the lens in place in the capsule.}
 \label{fig-surgery}
@@
 
-%\begin{figure}[h]
-%\begin{center}
-%\includegraphics[height=4.5cm]{images/implant_injection}
-%\caption [Cataract surgery] {Left: insertion of the lens implant. Right: deployment within the eye}
-%\label{fig-surgery2}
-%\end{center}
-%\end{figure}
-
-\subsection{Simulation}
-
-A mesh of lens implant with the proper geometry has been designed with Blender (figure \ref{fig-implant}).
-The physical parameters have been provided by the manufacturer Alcon (table \ref{tab-parameters}).
+To simulate the insertion and deployment of an intra-ocular lens, we first created a triangulation of the lens surface.
+Particular care was given to the mesh, to ensure that areas where large stresses occur contain a higher density of elements (figure \ref{fig-implant}).
+The lens mesh contains $743$ triangles and $473$ nodes.
+Models of the injection device and the entire eye anatomy were also created.
+Physical parameters of the lens implant have been provided by the manufacturer Alcon and they are presented into the table \ref{tab-parameters}.
 
@@
 	\vspace{0.3cm}
-	\caption{Physical parameters of the intraocular implant (source: Alcon)}
+	\caption{Physical parameters of the intra-ocular implant (source: Alcon)}
 	\label{tab-parameters}
@@
 
-We start by folding the haptics onto the implant body, then the body is bent while keeping the haptics inside to obtain the same position as the one chosen by the surgeon when he puts the implant inside the injection device (figure \ref{fig-implantFolded}).
-The bent implant is then placed into the injection device.
-The simulation eventually consists of pushing the intraocular implant within the injection device into the eye.
-We can notice the progressive deployment of the implant when it slowly gets out of the injector.
-% + SCREENSHOTS OF SIMULATION
+To obtain the folded geometry of the lens we simulated the folding process by first folding the hooks (also called \emph{haptics}) onto the implant body.
+The body was then bent while keeping the hooks inside to obtain the shape described in figure \ref{fig-implant}.
+The folded implant was then placed into the injection device.
+The simulation of the injection consists in pushing the intra-ocular implant within the injection device into the lens capsule.
+During these stages of the simulation, complex contacts occur and consist of self collisions of the lens as well as collisions between the lens and the injector and later with the capsule.
+To solve the contacts we use the contact warping method proposed Saupin \emph{et al.} \cite{Saupin08} as it offers an efficient way to compute physically correct contact responses in the case of co-rotational models.
 
 \begin{figure}[!h]
-\begin{center}
-\includegraphics[height=6.5cm]{images/IOL}
-\hspace{1cm}
-\includegraphics[height=6.5cm]{images/mesh_implant}
-\caption [Lens implant and its mesh] {The real implant and the mesh created with Blender (473 triangles)}
+\centering
+\includegraphics[height=3.5cm]{images/IOL}
+\hfill
+\includegraphics[height=3.5cm]{images/mesh_implant}
+\hfill
+\includegraphics[height=2.5cm]{images/implant_haptics_in}
+\hfill
+\includegraphics[height=3.0cm]{images/implant_folded}
+\caption [Lens implant and its mesh] {An actual intra-ocular implant and the triangular mesh used in our simulations.
+Notice the higher density of elements in areas where large deformations will take place.}
 \label{fig-implant}
-\end{center}
 \end{figure}
 
+Results of our simulation are illustrated in figure \ref{fig-simu-results}.
+We can notice the progressive deployment of the implant when it exits the injector.
+The shape of the intra-ocular lens remains very close to that of a real one during all stages of the simulation: within the injector, during the ejection phase, and when in place within the capsule.
+Due to the high stiffness and low mass of the lens, a direct solver was used at each time step rather than an iterative solver, resulting in a more accurate and more stable simulation, to the detriment of computation times (about 5 FPS for the complete simulation, and about 10 FPS for the deformation only, on a 2.4 GHz processor).
+
 \begin{figure}[!h]
-\begin{center}
-\includegraphics[height=5.5cm]{images/implant_haptics_in}
-\includegraphics[height=5.5cm]{images/implant_folded}
-\caption [Lens imlant] {Folding of the intraocular implant}
-\label{fig-implantFolded}
-\end{center}
+\centering
+\includegraphics[height=3.5cm]{images/simu_injection_step_1}
+\hfill
+\includegraphics[height=3.5cm]{images/simu_injection_step_2}
+\hfill
+\includegraphics[height=3.5cm]{images/simu_injection_step_3}
+\caption [Lens imlant] {Three steps of the simulation of the intra-ocular lens implant injection and its deployment within the lens capsule.}
+\label{fig-simu-results}
 \end{figure}
@@
 
-We improved a plate finite element formulation by extending it to a co-rotational framework in order to handle large deformations.
-By combining this formulation to a classic two-dimensional element we created a simple shell element able to efficiently handle both membrane and bending forces.
-The validity of our approach has been demonstrated though comparison with theoretical results and a non-trivial application in cataract surgery simulation.
-% + NEXT STEPS
+We proposed a co-rotational formulation for shell elements by extending a classical in-plane triangular finite element approach.
+This simple shell element can efficiently handle both membrane, bending and twist forces.
+The validity of our approach has been demonstrated though comparison with theoretical results.
+We then applied it to a rather complex application: the simulation of intra-ocular lens implant deployment in a cataract surgery simulation.
+These preliminary results are very encouraging, and show the potential of such models.
+Examples include the modelling of hollow anatomical structures (stomach, colon, etc.), the simulation of cardiac valve leaflets, and blood vessels.
+
+\section*{Acknowledgments}
+We are very grateful to J\'er\'emy Ringard for his invaluable help in improving the eye anatomical model, to J\'er\'emie Dequidt and Damien Marchal for understanding Blender, to Fr\'ederick Roy for his help with SOFA, and Nadia Boubchir for her persistence in motivating us to obtain these results in time...
 

### 2009-11-18 ISBMS09/paper.tex
Final version
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@69 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 %\footnote{acrylic is the common name for polymethylmethacrylate} 
-material, and are much thinner in one direction than in the two other ones.
+material, and are much thinner in one direction ($0.13$\,mm) than in the two other ones ($6$ and $13$\,mm).
 This last characteristic is what classifies these lenses as thin objects, i.e.
@@
 The difference between these two kinds of structures is very well explained by Liu \textit{et al.}~\cite{Liu03} and can be summarized by the fact that plate bending elements can only carry transversal loads while shells can undergo more complex deformations.
-%For instance, if we consider the horizontal board of a bookshelf, that board can be approximated as a plate structure and the transversal loads are the weight of the books.
-A typical deformation of the board is illustrated in figure~\ref{fig-boards}(a).
-Conversely, a shell structure can carry loads in all directions, and therefore undergoes bending, twisting and in-plane deformation (see Figure~\ref{fig-boards}(b)).
+For instance, if we consider the horizontal board of a bookshelf, that board can be approximated as a plate structure and the transversal loads are the weight of the books.
+A typical deformation of the board is illustrated in figure~\ref{fig-boards} left.
+Conversely, a shell structure can carry loads in all directions, and therefore can undergo bending, twisting and in-plane deformation (see Figure~\ref{fig-boards} right).
 Considering the folded shape of the implant within the injection device (see Figure \ref{fig-surgery}), a thin plate model would not correctly capture the deformation and therefore a shell formulation was retained to model the intra-ocular lens behaviour.
 
-%\begin{figure}[ht]
-%\begin{center}
-%\includegraphics[width=4.5cm]{images/intraocular_implant_1}
-%\hspace{1.5cm}
-%\includegraphics[width=4.5cm]{images/intraocular_implant_2}
-%\caption{Different types of lens implants.
-All share similar properties: they are three-dimensional structures, usually made of acrylic material, and are much thinner in one direction than in the two other ones.}
-%\label{fig-implants}
-%\end{center}
-%\end{figure}
-
-%\begin{figure}[ht]
-%\begin{center}
-%\includegraphics[width=5.5cm]{images/board_bending}
-%\hfill
-%\includegraphics[width=5.2cm]{images/board_twist}
-%\caption{Illustration of a key difference between different models of thin objects.
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=6cm]{images/board_bending}
+\hfill
+\includegraphics[width=6cm]{images/board_twist}
+\caption{Illustration of a key difference between various models of thin objects.
 While thin plate theory allows to describe bending (left), it cannot represent more complex deformations such as twist (right) which is captured by shell theory.}
-%\label{fig-boards}
-%\end{center}
-%\end{figure}
+\label{fig-boards}
+\end{center}
+\end{figure}
 
@@
 An important difference with previous works is that the ratio stiffness / mass of artificial lenses is very high compared to many material (such as fabric), resulting in added difficulties to obtain a numerically stable simulation.
+Indeed, if the mass density of cotton and acrylic is about the same, the stiffness of intra-ocular implants is three orders of magnitude higher.
 The closest work to what we propose is the one of Thomaszewski \emph{et al.} \cite{Thomaszewski06} who derive their formulation from thin shell analysis in a co-rotational framework.
-However, the authors use very soft and deformable material (Young's modulus $E = 5000\,Pa$ and Poisson's ratio $\nu = 0.25$ while the stiffness of intra-ocular implants are three orders of magnitude higher).
+However, the authors use very soft and deformable material (Young's modulus $E = 5000\,Pa$ and Poisson's ratio $\nu = 0.25$).
 
@@
 \includegraphics[height=3cm]{images/triangle_membrane}
-\hspace{1cm}
-\includegraphics[height=3.25cm]{images/triangle_bending}
+\includegraphics[height=3.5cm]{images/triangle_bending}
 \caption {A triangular shell element can be defined as a combination of a triangular in-plane membrane element (left) and a triangular thin plate in bending (right).
@@
 
-where $\textbf{J}$ is the strain-displacement matrix and $\boldsymbol\chi$ embodies the material's behaviour. In the simple case of Hooke's law we have:
+where $\textbf{J}$ is the strain-displacement matrix and $\boldsymbol\chi$ embodies the material's behaviour.
+The implant is very stiff and we therefore assume that the local deformations remain limited during the deployment and a linear constitutive law is sufficient.
+Thus in the simple case of Hooke's law we have:
 
@@
 
-We can calculate the strains from the flat-plate theory using $  e_{xx} = -z \frac{\partial^2u_z}{\partial x^2} $, $ e_{yy} = -z \frac{\partial^2u_z}{\partial y^2} $, and $ e_{xy} = -2z \frac{\partial^2u_z}{\partial x \partial y} $.
+We can calculate the strains from the flat-plate theory using:
+\begin{equation}
+e_{xx} = -z \frac{\partial^2u_z}{\partial x^2}
+\end{equation} 
+\begin{equation}
+e_{yy} = -z \frac{\partial^2u_z}{\partial y^2}
+\end{equation} 
+\begin{equation}
+e_{xy} = -2z \frac{\partial^2u_z}{\partial x \partial y}
+\end{equation}
+
 Hence, using the above equations and equation (\ref{eq-deflection}), we have
@@
 	\centering
-	\begin{tabular}{p{8.5cm}|c|c|}
+	\begin{tabular}{p{9cm}|c|c|}
 	\cline{2-3}
-	\multirow{5}{*}{\includegraphics[height=2.75cm]{images/clamped_square}} & $q$ & $z_{max}$ \tabularnewline
+	\multirow{5}{*}{\includegraphics[height=3.5cm]{images/clamped_square}} & $q$ & $z_{max}$ \tabularnewline
 	\cline{2-3}
@@
 	\end{tabular}
-	\caption{Comparison of our shell model with theoretical results on the bending of a square plate. An error of less than $1\,\%$ was found between our simulation and the theoretical results.}
+	\vspace{1cm}
+	\caption{Comparison of our shell model with theoretical results on the bending of a square plate. An error of less than $1\,\%$ was found between our simulation and theoretical results.}
 	\label{tab-results}
@@
 Cataract surgery consists in three main steps: capsulorhexis, phacoemulsification, and implantation of an intra-ocular lens.
+Prior to starting the surgery, a viscoelastic fluid is introduced into the capsule to facilitate capsulorhexis creation and provide protection during phacoemulsification.
+This fluid remains in the capsule for the duration of the surgery, including the injection of the implant.
 Capsulorhexis is the technique used to remove a part of the anterior lens capsule.
@@
 The foldable lens, usually made of acrylic material, is then implanted within the lens capsule through the incision used during phacoemulsification.
+In some cases the implant can be flipped or  the hooks (also called \emph{haptics}) can break.
+Therefore the simulation of the insertion and deployment of the implant is crucial for achieving a successful surgery.
 
@@
 To simulate the insertion and deployment of an intra-ocular lens, we first created a triangulation of the lens surface.
-Particular care was given to the mesh, to ensure that areas where large stresses occur contain a higher density of elements (figure \ref{fig-implant}).
+Particular care was given to the mesh, to ensure that areas where large stresses occur contain a higher density of elements (figure \ref{fig-mesh}).
+This was done by noting the constraints applied by the surgeon to the haptics while inserting the implant within the injection device.
+During this stage, the haptics are folded onto the implant body, leading to high stresses at the junctions.
 The lens mesh contains $743$ triangles and $473$ nodes.
 Models of the injection device and the entire eye anatomy were also created.
-Physical parameters of the lens implant have been provided by the manufacturer Alcon and they are presented into the table \ref{tab-parameters}.
+Physical parameters of the lens implant have been provided by the manufacturer Alcon and they are presented in table \ref{tab-parameters}.
 
@@
 
-To obtain the folded geometry of the lens we simulated the folding process by first folding the hooks (also called \emph{haptics}) onto the implant body.
-The body was then bent while keeping the hooks inside to obtain the shape described in figure \ref{fig-implant}.
+The first difficulty is to obtain the folded geometry of the lens within the injection device.
+This step is not important for the training process and does not need to be interactive.
+Indeed the surgeon does not always have to prepare the implant as some injection devices are readily available with a folded implant already in place.
+We simulate the folding process by first folding the haptics onto the implant body.
+The body was then bent while keeping the haptics inside to obtain the shape described in figure \ref{fig-implantFolding}.
+The whole process was carried out by applying the necessary forces and boundary conditions on the body and haptics of the implant.
 The folded implant was then placed into the injection device.
@@
 To solve the contacts we use the contact warping method proposed Saupin \emph{et al.} \cite{Saupin08} as it offers an efficient way to compute physically correct contact responses in the case of co-rotational models.
+As adhesions between the haptics and the body is often observed in surgery, friction is also taken into account in the contact response process.
 
@@
 \centering
-\includegraphics[height=3.5cm]{images/IOL}
-\hfill
-\includegraphics[height=3.5cm]{images/mesh_implant}
-\hfill
-\includegraphics[height=2.5cm]{images/implant_haptics_in}
-\hfill
-\includegraphics[height=3.0cm]{images/implant_folded}
+\includegraphics[height=5cm]{images/IOL}
+\includegraphics[height=5cm]{images/mesh_implant}
 \caption [Lens implant and its mesh] {An actual intra-ocular implant and the triangular mesh used in our simulations.
 Notice the higher density of elements in areas where large deformations will take place.}
-\label{fig-implant}
+\label{fig-mesh}
+\end{figure}
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=4cm]{images/implant_folding1}
+\includegraphics[width=3.5cm]{images/implant_folding2} \\
+\includegraphics[width=3cm]{images/implant_folding3}
+\includegraphics[width=5.5cm]{images/implant_folding4}
+\caption [Folding of intra-ocular implant] {Top: intermediary steps of the intra-ocular implant folding. Bottom: fully folded implant ready to be placed into the injection device.}
+\label{fig-implantFolding}
 \end{figure}
@@
 The shape of the intra-ocular lens remains very close to that of a real one during all stages of the simulation: within the injector, during the ejection phase, and when in place within the capsule.
-Due to the high stiffness and low mass of the lens, a direct solver was used at each time step rather than an iterative solver, resulting in a more accurate and more stable simulation, to the detriment of computation times (about 5 FPS for the complete simulation, and about 10 FPS for the deformation only, on a 2.4 GHz processor).
+Due to the high stiffness and low mass of the lens, a direct sparse solver was used at each time step (dt $= 0.01$\,s) rather than an iterative solver, resulting in a more accurate and more stable simulation, to the detriment of computation times (about 5 FPS for the complete simulation, and about 10 FPS for the deformation only, on a 2.4 GHz processor).
 
@@
 \centering
-\includegraphics[height=3.5cm]{images/simu_injection_step_1}
+\includegraphics[width=4cm]{images/surgery1}
 \hfill
-\includegraphics[height=3.5cm]{images/simu_injection_step_2}
+\includegraphics[width=4cm]{images/surgery2}
 \hfill
-\includegraphics[height=3.5cm]{images/simu_injection_step_3}
-\caption [Lens imlant] {Three steps of the simulation of the intra-ocular lens implant injection and its deployment within the lens capsule.}
+\includegraphics[width=4cm]{images/surgery3} \\
+\vspace{0.1cm}
+\includegraphics[width=4cm]{images/simu1}
+\hfill
+\includegraphics[width=4cm]{images/simu2}
+\hfill
+\includegraphics[width=4cm]{images/simu3}
+\caption [Lens imlant] {Three steps of the simulation of the intra-ocular lens implant injection and its deployment within the lens capsule.
+Top: images from a real cataract surgery, courtesy of Dr.
+Tarek Youssef, ophtalmologist in Canada.
+Below: our simulation of the implant's deployment.}
 \label{fig-simu-results}
@@
 The validity of our approach has been demonstrated though comparison with theoretical results.
-We then applied it to a rather complex application: the simulation of intra-ocular lens implant deployment in a cataract surgery simulation.
-These preliminary results are very encouraging, and show the potential of such models.
+It was then applied it to a rather complex application: the simulation of intra-ocular lens implant deployment in a cataract surgery simulation.
+These preliminary results are very encouraging and show the potential of such models.
 Examples include the modelling of hollow anatomical structures (stomach, colon, etc.), the simulation of cardiac valve leaflets, and blood vessels.
+Regarding improvements of our current model and simulation, we plan to account for the interaction with the viscoelastic fluid (currently approximated by a high damping).
+Computation times also need to be improved through the use of more optimised meshes and by porting our parallelisable algorithm to Graphics Processing Units.
 
 \section*{Acknowledgments}
-We are very grateful to J\'er\'emy Ringard for his invaluable help in improving the eye anatomical model, to J\'er\'emie Dequidt and Damien Marchal for understanding Blender, to Fr\'ederick Roy for his help with SOFA, and Nadia Boubchir for her persistence in motivating us to obtain these results in time...
+We are very grateful to J\'er\'emy Ringard for his invaluable help in improving the eye anatomical model, to J\'er\'emie Dequidt and Damien Marchal for understanding Blender, to Fr\'ederick Roy for his help with SOFA, and Nadia Boubchir for her persistence in motivating us to obtain these results in time.
+
+\bigskip
+
+We would also like to thank INRIA and the SOFA project, as well as CSIRO Preventative Health National Research Flagship, ICTC, The Australian e-Health Research Centre for their fundings.
 

### 2010-03-02 forcefield/TriangularBendingFEMForceField.h
FIX: issue with shader and mechanical mapping CHG: takes area of the triangle into account for the membrane energy - updates of scene files to take the changes into account
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@78 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@ -129,8 +129,8 @@ public:
                 // Stiffness matrix for bending K = Jt * M * J
                 StiffnessMatrixBending stiffnessMatrixBending;
                 
-                // Third of the surface
-                Real thirdSurface;
+                // Surface
+                Real area;
                 // Variables needed for drawing the shell
                 Vec<9, Real> u; // displacement vector
                 Mat<9, 9, Real> invC; // inverse of C (used in bending mode only)
@@ -176,13 +176,15 @@ public:
         Data<VecCoord> targetVertices;
         Data<SeqTriangles> targetTriangles;
 
+        unsigned int indexTop;
+
 protected :
 
         TriangleData<TriangleInformation> triangleInfo;
 
 	void computeDisplacement(Displacement &Disp, const VecCoord &x, const Index elementIndex);
         void computeDisplacementBending(DisplacementBending &Disp, const VecCoord &x, const Index elementIndex);
-	void computeStrainDisplacementMatrix(StrainDisplacement &J, const Vec3& b, const Vec3& c);
+	void computeStrainDisplacementMatrix(StrainDisplacement &J, const Index elementIndex, const Vec3& b, const Vec3& c);
         void computeStrainDisplacementMatrixBending(TriangleInformation *tinfo, const Vec3& b, const Vec3& c);
         void tensorFlatPlate(Mat<3, 9, Real>& D, const Vec3 &P);
         void computeStiffnessMatrix(StiffnessMatrix &K, const StrainDisplacement &J, const MaterialStiffness &M);

### 2010-03-02 mapping/BendingPlateMechanicalMapping.h

@@ -28,7 +28,8 @@
 #include <sofa/core/componentmodel/behavior/MechanicalMapping.h>
 #include <sofa/core/componentmodel/behavior/MechanicalState.h>
 #include <sofa/helper/vector.h>
-#include <sofa/core/VisualModel.h>
+#include <sofa/helper/gl/GLSLShader.h>
+//#include <sofa/core/VisualModel.h>
 
 #include "../forcefield/TriangularBendingFEMForceField.h"
 #include <sofa/component/topology/TriangleSubdivisionTopologicalMapping.h>
@@ -53,7 +54,8 @@ using namespace sofa::component::topology;
 using namespace sofa::helper::system::thread;
 
 template <class BasicMapping>
-class BendingPlateMechanicalMapping : public BasicMapping, public core::VisualModel, public virtual core::objectmodel::BaseObject
+class BendingPlateMechanicalMapping : public BasicMapping, public virtual core::objectmodel::BaseObject
+// public core::VisualModel,
 {
 public:
     SOFA_CLASS(SOFA_TEMPLATE(BendingPlateMechanicalMapping,BasicMapping), BasicMapping);
@@ -88,7 +90,8 @@ public:
 
     void init();
     void reinit();
-    virtual void drawVisual();
+    virtual void draw();
+//    virtual void drawVisual();
     
     virtual ~BendingPlateMechanicalMapping();
     
@@ -98,6 +101,8 @@ public:
     void applyJT( typename In::VecConst& out, const typename Out::VecConst& in );
 
 protected:
+
+        helper::gl::GLSLShader shader;
 
         Data<bool> measureError;
         Data<InVecCoord> targetVertices;

### 2010-03-02 MICCAI10/paper.tex
miccai paper v0.0
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@79 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2009-11-18 ISBMS09/paper.tex
@@
 
-\title{A Shell Model for Real-Time Simulation of Intra-ocular Implant Deployment}
+\title{A Shell Model for Real-Time Medical Simulation}
 \author{Olivier Comas, St\'ephane Cotin and Christian Duriez}
@@
 \begin{abstract}
-With 30 million interventions a year worldwide, cataract surgery is one of the most frequently performed procedures.
-Yet, no tool currently allows teaching all steps of the procedure without putting patients at risk.
-A particularly challenging stage of this surgery deals with the injection and deployment of the intra-ocular lens implant.
-In this paper we propose to rely on shell theory to accurately describe the complex deformations of the implant.
-Our approach extends the co-rotational method used in finite element analysis of in-plane deformations to incorporate a bending energy.
-This results in a relatively simple and computationally efficient approach which was applied to the simulation of the lens deployment.
-This simulation also accounts for the complex contacts that take place during the injection phase.
-%Particular care was given to the validation of the proposed model, its computational efficiency and stability. 
+Contributions: Shell model with real-time implementation + method for fitting medical data
+
 \end{abstract}
@@
 \section{Introduction}
-According to the latest assessment of the World Health Organization, age related cataract is responsible for $48\,\%$ of world blindness, which represents about 18 million people.
-Cataract is the clouding of the lens of the eye which impedes the passage of light.
-Its treatment is a surgical operation, where the opaque lens is removed and replaced by an artificial intra-ocular lens.
-As people live longer, more surgical interventions are needed.
-However, current methods of teaching are not compatible with increasing demands of training for this procedure, and also expose patients to complications inherent to the inexperience of the operator.
-A cataract surgery simulator can offer an efficient alternative teaching method for ophthalmologists, as illustrated in recent reports such as \cite{Feudner09}.
-The work presented in this paper focuses on the simulation of a crucial step of cataract surgery: the injection and deployment of an artificial lens.
-To reach this objective, we propose an efficient yet simple formulation for triangular shells, as well as a physically accurate model of the complex (self-)contacts that occur during the injection and deployment of the artificial lens.
 
@@
 
-Although various shapes of intra-ocular lens implants exist they all share common properties: they are three-dimensional structures, usually made of acrylic
-%\footnote{acrylic is the common name for polymethylmethacrylate} 
-material, and are much thinner in one direction ($0.13$\,mm) than in the two other ones ($6$ and $13$\,mm).
-This last characteristic is what classifies these lenses as thin objects, i.e.
-three-dimensional objects that can be abstracted as two-dimensional structures.
 Numerous models are available in the literature to describe physics of thin objects, from fairly simple and naive approaches to more complex and thorough representations.
@@
 
-\section{Simulation of the intra-ocular lens injection}
-
-%Cataract surgery is the removal of the natural lens of the eye (also called \emph{crystalline lens}) that has developed an opacification, which is referred to as a cataract.
-Metabolic changes of the crystalline lens fibers over the time lead to the development of the cataract and loss of transparency, causing impairment or loss of vision.
-During cataract surgery, a patient's cloudy natural lens is removed and replaced with a synthetic lens to restore the lens' transparency (figure \ref{fig-surgery}).
-The last step in the simulation of this surgery is the injection of the implant and its deployment within the lens capsule.
-
-Cataract surgery consists in three main steps: capsulorhexis, phacoemulsification, and implantation of an intra-ocular lens.
-Prior to starting the surgery, a viscoelastic fluid is introduced into the capsule to facilitate capsulorhexis creation and provide protection during phacoemulsification.
-This fluid remains in the capsule for the duration of the surgery, including the injection of the implant.
-Capsulorhexis is the technique used to remove a part of the anterior lens capsule.
-Phacoemulsification consists in using a surgical device which tip vibrates at an ultrasonic frequency to emulsify the natural lens material and then aspirate the  fragments of the cortical material.
-After the removal of the diseased lens, an intra-ocular lens is implanted into the eye, through a small incision (about 2 mm) using a foldable intra-ocular lens (see figure \ref{fig-surgery}).
-The foldable lens, usually made of acrylic material, is then implanted within the lens capsule through the incision used during phacoemulsification.
-In some cases the implant can be flipped or  the hooks (also called \emph{haptics}) can break.
-Therefore the simulation of the insertion and deployment of the implant is crucial for achieving a successful surgery.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[height=3.2cm]{images/suction}
-\hfill
-\includegraphics[height=3.2cm]{images/implant_injection_step_1}
-\hfill
-\includegraphics[height=3.2cm]{images/implant_injection_step_2}
-\caption [Cataract surgery] {Left: removal of the opacified lens by phacoemulsification.
-Center: insertion of the lens implant which is folded inside the injection device and then deploys within the lens capsule.
-Right: the lens in place in the capsule.}
-\label{fig-surgery}
-\end{center}
-\end{figure}
-
-To simulate the insertion and deployment of an intra-ocular lens, we first created a triangulation of the lens surface.
-Particular care was given to the mesh, to ensure that areas where large stresses occur contain a higher density of elements (figure \ref{fig-mesh}).
-This was done by noting the constraints applied by the surgeon to the haptics while inserting the implant within the injection device.
-During this stage, the haptics are folded onto the implant body, leading to high stresses at the junctions.
-The lens mesh contains $743$ triangles and $473$ nodes.
-Models of the injection device and the entire eye anatomy were also created.
-Physical parameters of the lens implant have been provided by the manufacturer Alcon and they are presented in table \ref{tab-parameters}.
-
-\begin{table}[h!]
-	\begin{center}
-		\begin{tabular}{|p{3cm}|p{3cm}|p{3cm}|}
-		\hline
-		 \centering Young's modulus & \centering Poisson's ratio & \centering Mass density \tabularnewline
-		\hline
-		\centering $1 MPa$ & \centering $0.42$ & \centering $1.2 g/cm^3$ \tabularnewline
-		\hline
-		\end{tabular}
-	\vspace{0.3cm}
-	\caption{Physical parameters of the intra-ocular implant (source: Alcon)}
-	\label{tab-parameters}
-	\end{center}
-\end{table}
-
-The first difficulty is to obtain the folded geometry of the lens within the injection device.
-This step is not important for the training process and does not need to be interactive.
-Indeed the surgeon does not always have to prepare the implant as some injection devices are readily available with a folded implant already in place.
-We simulate the folding process by first folding the haptics onto the implant body.
-The body was then bent while keeping the haptics inside to obtain the shape described in figure \ref{fig-implantFolding}.
-The whole process was carried out by applying the necessary forces and boundary conditions on the body and haptics of the implant.
-The folded implant was then placed into the injection device.
-The simulation of the injection consists in pushing the intra-ocular implant within the injection device into the lens capsule.
-During these stages of the simulation, complex contacts occur and consist of self collisions of the lens as well as collisions between the lens and the injector and later with the capsule.
-To solve the contacts we use the contact warping method proposed Saupin \emph{et al.} \cite{Saupin08} as it offers an efficient way to compute physically correct contact responses in the case of co-rotational models.
-As adhesions between the haptics and the body is often observed in surgery, friction is also taken into account in the contact response process.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[height=5cm]{images/IOL}
-\includegraphics[height=5cm]{images/mesh_implant}
-\caption [Lens implant and its mesh] {An actual intra-ocular implant and the triangular mesh used in our simulations.
-Notice the higher density of elements in areas where large deformations will take place.}
-\label{fig-mesh}
-\end{figure}
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=4cm]{images/implant_folding1}
-\includegraphics[width=3.5cm]{images/implant_folding2} \\
-\includegraphics[width=3cm]{images/implant_folding3}
-\includegraphics[width=5.5cm]{images/implant_folding4}
-\caption [Folding of intra-ocular implant] {Top: intermediary steps of the intra-ocular implant folding. Bottom: fully folded implant ready to be placed into the injection device.}
-\label{fig-implantFolding}
-\end{figure}
-
-Results of our simulation are illustrated in figure \ref{fig-simu-results}.
-We can notice the progressive deployment of the implant when it exits the injector.
-The shape of the intra-ocular lens remains very close to that of a real one during all stages of the simulation: within the injector, during the ejection phase, and when in place within the capsule.
-Due to the high stiffness and low mass of the lens, a direct sparse solver was used at each time step (dt $= 0.01$\,s) rather than an iterative solver, resulting in a more accurate and more stable simulation, to the detriment of computation times (about 5 FPS for the complete simulation, and about 10 FPS for the deformation only, on a 2.4 GHz processor).
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=4cm]{images/surgery1}
-\hfill
-\includegraphics[width=4cm]{images/surgery2}
-\hfill
-\includegraphics[width=4cm]{images/surgery3} \\
-\vspace{0.1cm}
-\includegraphics[width=4cm]{images/simu1}
-\hfill
-\includegraphics[width=4cm]{images/simu2}
-\hfill
-\includegraphics[width=4cm]{images/simu3}
-\caption [Lens imlant] {Three steps of the simulation of the intra-ocular lens implant injection and its deployment within the lens capsule.
-Top: images from a real cataract surgery, courtesy of Dr.
-Tarek Youssef, ophtalmologist in Canada.
-Below: our simulation of the implant's deployment.}
-\label{fig-simu-results}
-\end{figure}
-
 \section{Conclusion}
@@
 
-\section*{Acknowledgments}
-We are very grateful to J\'er\'emy Ringard for his invaluable help in improving the eye anatomical model, to J\'er\'emie Dequidt and Damien Marchal for understanding Blender, to Fr\'ederick Roy for his help with SOFA, and Nadia Boubchir for her persistence in motivating us to obtain these results in time.
-
-\bigskip
-
-We would also like to thank INRIA and the SOFA project, as well as CSIRO Preventative Health National Research Flagship, ICTC, The Australian e-Health Research Centre for their fundings.
-
 \bibliographystyle{splncs}

### 2010-03-07 forcefield/TriangularBendingFEMForceField.h
Removed MICCAI folder from Christian :p
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@81 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@ -96,7 +96,10 @@ protected:
         typedef Mat<18, 18, Real> StiffnessMatrixGlobalSpace;
 
 	sofa::core::componentmodel::topology::BaseMeshTopology* _topology;
+
         sofa::core::componentmodel::topology::BaseMeshTopology* _topologyHigh;
+        VecCoord verticesTarget;
+        SeqTriangles trianglesTarget;
 
         TriangularBendingFEMForceFieldInternalData<DataTypes> data;
         friend class TriangularBendingFEMForceFieldInternalData<DataTypes>;
@@ -173,8 +176,8 @@ public:
         Data <Real> f_membraneRatio;
         Data <Real> f_bendingRatio;
         Data<bool> refineMesh;
-        Data<VecCoord> targetVertices;
-        Data<SeqTriangles> targetTriangles;
+//        Data<VecCoord> targetVertices;
+//        Data<SeqTriangles> targetTriangles;
 
         unsigned int indexTop;
 

### 2010-03-07 mapping/BendingPlateMechanicalMapping.h

@@ -105,8 +105,13 @@ protected:
         helper::gl::GLSLShader shader;
 
         Data<bool> measureError;
-        Data<InVecCoord> targetVertices;
-        Data<SeqTriangles> targetTriangles;
+//        Data<InVecCoord> targetVertices;
+//        Data<SeqTriangles> targetTriangles;
+
+        sofa::core::componentmodel::topology::BaseMeshTopology* _topologyHigh;
+        InVecCoord verticesTarget;
+        SeqTriangles trianglesTarget;
+        helper::vector<Vec3> colourMapping;
 
         helper::vector<Real> vectorError;
         helper::vector<Vec3> normals;

### 2010-03-07 MICCAI2010/paper.tex
Changed name
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@83 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\documentclass{llncs}

\usepackage[english]{babel}
%\usepackage[tight]{subfigure}
%\usepackage{floatrow}
\usepackage{graphicx}
%\usepackage{amsmath}
%\usepackage{multirow}
%\renewcommand*\arraystretch{1.25} % extra vertical space in cells

\begin{document}

\title{A Framework for Modelling Thin Objects: \\ from Meshing to Shell Finite Element Method}
%\author{Olivier Comas, St\'ephane Cotin and Christian Duriez}
%\institute{INRIA, Alcove team, Lille, France}

\maketitle

\begin{abstract}

\end{abstract}

\section{Introduction}
Cadavers and live animal surgery are often used in the training of surgeons. But ethics concerns and high costs are progressively leading to the elimination of these methods in favor of more modern and effective alternatives. Medical simulators offer an efficient teaching method for physicians. Numerous surgical interventions involve manipulation and deformation of thin anatomical structures and hollow organs. Examples include the interaction between the catheter and the blood vessels during interventional radiology, the deformation of the colon in colonoscopy or the interaction with a beating heart. Simulators of those interventions must showcase a physically realistic behaviour of thin structures to be accepted by the medical community. This is even crucial in the case of colonoscopy for instance as the practitioner only guides himself through the force feedback that he receives from the interaction between the colonoscope and the colon. Although volumic organs are quite accurately simulated in recent simulators hollow organs are poorly modelled, which limits the type of medical procedures that a simulator can provide.

We aim to fill this gap by offering a framework able to handle all the steps from meshing to realistic simulation of hollow organs. Not only we propose an efficient yet simple co-rotational finite element formulation for triangular shells but we also present a method to generate a mesh of any hollow organ with shell-type curved elements in an attempt to optimise even more the computation of deformations by using fewer elements. % Finally we show the pertinence of our approach by coupling tetrahedral and shell elements in an advanced liver model.

\section{A physical model for thin structures}

\subsection{Related work}

Development of a satisfactory physical model that runs in real-time but produces visually convincing animation of thin objects is challenging. Rather than resorting to shell theory which involves the most complex formulations in continuum mechanics, previous works have often relied on discrete formulations. Early approaches only considered in-plane deformation, and often relied on mass-spring models. More recent works have considered adding bending through angular springs. For instance Wang et al.~\cite{Wang07} successfully used a network of linear and angular springs to describe bending and twisting of catheters and guidewires in an interventional radiology simulator. Yet, such models are limited in their ability to describe certain behaviour, as they do not rely on continuum mechanics. Another limitation of such models is the difficulty to derive spring stiffness (in particular for angular springs) from elastic properties (Young's modulus and Poisson's ratio). For these reasons, other approaches have been proposed.

Continuum mechanics provides many formulations able to accurately describe stresses occurring within thin objects. Most of them fall into one of the following two categories: plate theory or shell theory. Those theories have been a subject of interest in the mechanical community for decades. The difference between these two kinds of structures is very well explained by Liu et al.~\cite{Liu03} and can be summarized by the fact that plate bending elements can only carry transversal loads while shells can undergo more complex deformations. Considering the wide range of shapes and solicitations that can be applied to hollow organs, a thin plate model would not correctly capture their deformation and therefore a shell formulation was retained.

\subsection{Co-rotational triangular shell model}

A complete description of our co-rotational triangular shell finite element model is available in one of our previous publication \cite{Comas2010ISBMS}. Therefore we will only remind briefly the key points. We improved and extended a plate model first introduced by Przemieniecki \cite{Przemieniecki68} to a co-rotational formulation. Co-rotational approaches offer a good trade-off between computational efficiency and accuracy by allowing small deformations but large displacements. Once combined with an in-plane membrane formulation we obtain an accurate, yet computationally efficient, shell finite element method featuring both membrane and bending energies.

\subsubsection{Triangular elastic membrane}

The element stiffness matrix $\textbf{K}_e$ can be computed as follows:

\begin{equation}
\textbf{K}_e = \int_v \textbf{J} \boldsymbol\chi \textbf{J}^{T} dV
\end{equation}
where $\textbf{J}$ is the strain-displacement matrix and $\boldsymbol\chi$ embodies the material's behaviour. The stiffness matrix in the global frame is eventually obtained using the rotation matrix of the element: $\textbf{K} = \textbf{R}^{T} \textbf{K}_e \textbf{R} $ where $\textbf{R}$ describes the rotation of the (triangular) element with respect to its initial configuration.

\subsubsection{Triangular plate bending}

\begin{figure}
\centering
\includegraphics[height=5cm]{images/bending}
\caption {The different degrees of freedom $u$ of a triangular thin plate in bending are illustrated above.}
\label{fig-triangle}
\end{figure}

To calculate the stiffness matrix for the transverse deflections and rotations shown on figure \ref{fig-triangle}, we start with the assumed deflection $u_z$ of the form
\begin{equation}
 u_z = c_1 + c_2x + c_3y + c_4x^2 + c_5xy + c_6y^2 + c_7x^3 + c_8xy^2 + c_9y^3
\label{eq-deflection}
\end{equation}

where $c_1$, \ldots , $c_9$ are constants. Let us define the vector $\textbf{u} = \left\{u_1 u_2 \ldots u_9 \right\} $ of the displacements and slopes at the three corners of the triangular plate using the following notations:
\begin{equation}
u_1 = (u_z)_{x_1,y_1} \hspace{1cm} u_2 = \left(\frac{\partial u_z}{\partial y}\right)_{x_1,y_1} \hspace{1cm} u_3 = - \left(\frac{\partial u_z}{\partial x}\right)_{x_1,y_1}
\end{equation} 
and so on for the two other vertices.
We then can derive a matrix $C$ such as $\textbf{u} = \textbf{Cc}$ where $\textbf{c} = \left\{c_1 c_2 \ldots c_9 \right\} $.

We can calculate the strains from the flat-plate theory using:
\begin{equation}
e_{xx} = -z \frac{\partial^2u_z}{\partial x^2}
\end{equation} 
\begin{equation}
e_{yy} = -z \frac{\partial^2u_z}{\partial y^2}
\end{equation} 
\begin{equation}
e_{xy} = -2z \frac{\partial^2u_z}{\partial x \partial y}
\end{equation}

Symbolically this may be expressed as $\textbf{e} = \textbf{Dc}$ where $\textbf{D}$ derives from equation~\ref{eq-deflection} and the above relations. Noting that $\textbf{c} = \textbf{C}^{-1}\textbf{u}$, we have:
\begin{equation}
\textbf{e} = \textbf{DC}^{-1}\textbf{u} = \textbf{bu}
\end{equation} 
where the strain-displacement matrix $\textbf{b} = \textbf{DC}^{-1}$. 
The stiffness matrix $\textbf{K}_e$ for an element is then classically obtained from
\begin{equation}
\textbf{K}_e = \int_v \textbf{b}^{T} \boldsymbol\chi \textbf{b} dV
\end{equation} 
where $\boldsymbol\chi$ is the material matrix. The stiffness matrix in the global frame is eventually obtained using the rotation matrix of the element as in the membrane case.

\section{Meshing a curved surface}

\subsection{Related work}

A key component of finite element methods is the mesh of the simulated object being used. While a sufficient number of elements is needed to accurately describe the geometry, too many of them will yield a computationally expensive simulation and solving the system in real-time may no longer be possible. Literature about volumic mesh generation algorithms is fairly dense. However, there are only a few that are concerned about the generation of meshes over curved surfaces. One of the most widely used techniques for the creation of surface meshes is the plane to surface transformation method \cite{Zienkiewicz71}, mesh is first generated on a two-dimensional domain and then mapped onto the surface. If this method gives reasonably good meshes on smooth surfaces, the results are usually quite poor with more complex curved surfaces. The finite elements may be generated directly on the curved surfaces based on the advancing front technique \cite{Lo85,Lau96}. The main issue with this approach is that an analytical description of the geometry is needed, which is not the case in medical simulation. Another method consists of using an a priori error estimator to build an adaptive mesh generation \cite{Baumann97}. However, the tolerance of this indicator should be chosen depending on the desired accuracy of the finite element solution, and therefore requires some knowledge about the problem in order to choose an effective tolerance. In \cite{Bechet02} the authors start from an existing triangular mesh created with a CAD software and refine and smooth the mesh based on element quality and surface curvature. While all those methods allow their authors to get satisfactory meshes over curved surfaces according to their needs, they all make use of flat elements to mesh geometries and do not generate actual shells.

\subsection{A mechanically-driven method to mesh curved surfaces with shells}

Indeed, one main difference between shell and plate elements is that the rest shape of the former can describe a curved surface. We propose to take this crucial feature into account to optimise the meshing process. While many flat triangles are often required to describe highly curved surfaces, fewer triangular shell elements are needed to describe the given geometry with the same precision since they can be naturally curved. In the following we assume that we have a high resolution triangular mesh obtained from a binary segmented image of the organ we want to simulate (via a Marching Cube algorithm for instance). Our goal is to create a mesh featuring the optimal number of shell elements while staying as close as possible to our targeted geometry.

Therefore we need to insure that the distance between the surface of our shell-based mesh and the targeted high resolution mesh will be minimal. To our knowledge the only tool available to measure error between two surfaces is the Hausdorff distance \cite{Klein96,Cignoni98}. As a reminder the Hausdorff distance between two meshes is the maximum between the two so-called one-sided Hausdorff distances:

\begin{equation}
d_{\mathrm{H}}(X,Y) = \max \left\{ \sup_{x \in X} \inf_{y \in Y} d(x,y), \sup_{y \in Y} \inf_{x \in X} d(x,y) \right\} \enspace .
\end{equation}
where $d()$ is the Euclidian distance between two points. Since the curved surface of each shell must be taken into account in the computation of the Hausdorff distance, a finer mesh is created by adding vertices on the surface of each element. We first sample the flat surface of each element by recursively dividing each triangle into four smaller ones and the deflection of each new vertex is computed using (\ref{eq-deflection}). We then use this new mesh that samples the actual surface described by the shells to compute the Hausdorff distance with the targeted high resolution mesh. The same process of subdivision allows us to render each shell as a curved triangle (Fig.~\ref{fig-shell} ).

\begin{figure}
\centering
\includegraphics[height=4cm]{images/shell_curvature}
\caption {The triangle formed by the three vertices of the shell has been recursively subdivided 3 times and the deflection of each new vertex was computed according to the same deflection function used in our shell finite element formulation. Sampling the actual surface of the shell allows rendering and more accurate Hausdorff distance computation. }
\label{fig-shell}
\end{figure}

The first step in the process of generating a shell-based mesh is an aggressive decimation of the targeted high resolution mesh. We then apply a heuristic method derived from the work of Saupin et. al \cite{Saupin07} with tetrahedral meshes based on very simple geometrical rules. For each node of the coarse mesh, we find the three closest triangles on the high resolution mesh and we move the node to the barycenter of the three centres of mass of those triangles. This technique locally smooths the surface of the mesh while converging towards the desired high resolution mesh. At each iteration of this algorithm we measure the error with the target using the Hausdorff distance and the process is stopped when the required precision has been reached. A simple example is shown Fig. \ref{fig-cylinder} and \ref{fig-cylinderResults} to illustrate the method.

\begin{figure}
\centering
\includegraphics[height=4cm]{images/cylinderHighWireframe}
\includegraphics[height=4cm]{images/cylinderLow}
\caption {Illustration of our method on a simple example. The target is the high resolution cylinder mesh (left) and the start point is a very coarse mesh approximating the shape of a cylinder (right).}
\label{fig-cylinder}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=7cm]{images/cylinderResults}
\caption {(a) Coarse mesh rendered with shells. (b) Gaps between the coarse mesh and the targeted cylinder. (c) New mesh after one iteration of our algorithm. (d) Differences between the two meshes have been subtantially reduced.}
\label{fig-cylinderResults}
\end{figure}

\section{Results}
\subsection{Meshing of anatomical structures (error maps)}

\subsection{Deformations}
\subsubsection{Computation times on 3 examples}
\subsubsection{Coupling between tetrahedra and shells for more advanced modelling}

\section{Conclusion}
\section*{Acknowledgments}

\bibliographystyle{splncs}
\bibliography{bibliography}

\end{document}

### 2010-03-09 MICCAI2010/paper.tex
CHG: - changed computation of the error (it's now the actual two-sided Hausdorff distance that is computed) - changed colour gradient (conversion from HSL colours) - simplification of the vertex shader - updates of all screenshots for the paper (with white backgrounds) and better image to illustrate the results on cylinder - added meshes of liverHighRes, liverLowRes300 (coarse mesh) and liverLowRes300(coarse mesh after one iteration) - updates scenes to take the changes into account - loads high res mesh as Vec3d (instead of Rigids), changed made accordingly in ForceField and MechanicalMapping

@@
 \centering
-\includegraphics[height=7cm]{images/cylinderResults}
-\caption {(a) Coarse mesh rendered with shells.
-(b) Gaps between the coarse mesh and the targeted cylinder.
-(c) New mesh after one iteration of our algorithm.
-(d) Differences between the two meshes have been subtantially reduced.}
+\includegraphics[height=4cm]{images/cylinderResults2}
+\caption {(Left) Hausdorff distance colour map of the coarse mesh rendered with shells.
+(Right) Hausdorff distance colour map after one iteration of our algorithm.
+We can notice that differences with the target mesh have been subtantially reduced.}
 \label{fig-cylinderResults}

### 2010-03-09 forcefield/TriangularBendingFEMForceField.h

@@ -98,11 +98,11 @@ protected:
 	sofa::core::componentmodel::topology::BaseMeshTopology* _topology;
 
         sofa::core::componentmodel::topology::BaseMeshTopology* _topologyHigh;
-        VecCoord verticesTarget;
+        helper::vector<Vec3> verticesTarget;
         SeqTriangles trianglesTarget;
 
-        TriangularBendingFEMForceFieldInternalData<DataTypes> data;
-        friend class TriangularBendingFEMForceFieldInternalData<DataTypes>;
+//        TriangularBendingFEMForceFieldInternalData<DataTypes> data;
+//        friend class TriangularBendingFEMForceFieldInternalData<DataTypes>;
 
 public:
 

### 2010-03-09 mapping/BendingPlateMechanicalMapping.h

@@ -52,6 +52,7 @@ using namespace sofa::defaulttype;
 using namespace sofa::component::forcefield;
 using namespace sofa::component::topology;
 using namespace sofa::helper::system::thread;
+using namespace core::componentmodel::topology;
 
 template <class BasicMapping>
 class BendingPlateMechanicalMapping : public BasicMapping, public virtual core::objectmodel::BaseObject
@@ -65,8 +66,9 @@ public:
     typedef typename Out::VecCoord OutVecCoord;
     typedef typename Out::VecDeriv OutVecDeriv;
     typedef typename Out::Coord OutCoord;
-    typedef typename Out::Deriv OutDeriv;   
-    typedef typename std::map<unsigned int, OutDeriv>::const_iterator OutConstraintIterator;        
+    typedef typename Out::Deriv OutDeriv;
+    typedef typename Out::DataTypes OutDataTypes;
+//    typedef typename std::map<unsigned int, OutDeriv>::const_iterator OutConstraintIterator;
 
     typedef typename In::VecCoord InVecCoord;
     typedef typename In::VecDeriv InVecDeriv;
@@ -74,10 +76,10 @@ public:
     typedef typename In::Deriv InDeriv;
     typedef typename In::DataTypes InDataTypes;
 
-    typedef core::componentmodel::topology::BaseMeshTopology::Edge	Edge;
-    typedef core::componentmodel::topology::BaseMeshTopology::SeqEdges	SeqEdges;
-    typedef core::componentmodel::topology::BaseMeshTopology::Triangle	Triangle;
-    typedef core::componentmodel::topology::BaseMeshTopology::SeqTriangles SeqTriangles;
+    typedef BaseMeshTopology::Edge	Edge;
+    typedef BaseMeshTopology::SeqEdges	SeqEdges;
+    typedef BaseMeshTopology::Triangle	Triangle;
+    typedef BaseMeshTopology::SeqTriangles SeqTriangles;
 
     typedef typename Out::Real Real;
     typedef Vec<3, Real> Vec3;
@@ -108,16 +110,17 @@ protected:
 //        Data<InVecCoord> targetVertices;
 //        Data<SeqTriangles> targetTriangles;
 
-        sofa::core::componentmodel::topology::BaseMeshTopology* _topologyHigh;
-        InVecCoord verticesTarget;
+        BaseMeshTopology* _topologyHigh;
+        OutVecCoord verticesTarget;
         SeqTriangles trianglesTarget;
+        
         helper::vector<Vec3> colourMapping;
[…]

### 2010-03-09 MICCAI2010/paper.tex
intro + figures stent
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@85 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 %\renewcommand*\arraystretch{1.25} % extra vertical space in cells
+\usepackage{amsmath}
 
@@
 \section{Introduction}
-Cadavers and live animal surgery are often used in the training of surgeons.
-But ethics concerns and high costs are progressively leading to the elimination of these methods in favor of more modern and effective alternatives.
-Medical simulators offer an efficient teaching method for physicians.
-Numerous surgical interventions involve manipulation and deformation of thin anatomical structures and hollow organs.
-Examples include the interaction between the catheter and the blood vessels during interventional radiology, the deformation of the colon in colonoscopy or the interaction with a beating heart.
-Simulators of those interventions must showcase a physically realistic behaviour of thin structures to be accepted by the medical community.
-This is even crucial in the case of colonoscopy for instance as the practitioner only guides himself through the force feedback that he receives from the interaction between the colonoscope and the colon.
-Although volumic organs are quite accurately simulated in recent simulators hollow organs are poorly modelled, which limits the type of medical procedures that a simulator can provide.
 
-We aim to fill this gap by offering a framework able to handle all the steps from meshing to realistic simulation of hollow organs.
-Not only we propose an efficient yet simple co-rotational finite element formulation for triangular shells but we also present a method to generate a mesh of any hollow organ with shell-type curved elements in an attempt to optimise even more the computation of deformations by using fewer elements.
-% Finally we show the pertinence of our approach by coupling tetrahedral and shell elements in an advanced liver model.
+Human body is composed of various deformable anatomical structures. 
+Realistic modeling of organ deformations is a challenging research field that leads the way to numerous clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedure and per-operative guidance based on simulation.
+In all these cases, the clinician needs fast update of the deformation model to obtain a real-time display of the computed deformations.  
+To realize this kind of interactive systems, one face the trade-off problem between accuracy of the model and the need of fast computations.
+
+An other key challenge of real-time human soft-tissue modeling is the variousness of the mechanical behaviors. 
+It seems unrealistic to target a unique model for all tissues even if there are obvious advantages to make maximum use of optimized generic models.
+Yet most of previous work focus on volume models that are able to capture the behavior of volume organs like liver \cite{} or brain tissues \cite{}.
+In contrast, this paper seeks to propose a solution for simulating, in real-time, deformation of thin anatomical structures, whose volume is negligible compared to their surface area.
+Examples of such anatomical structures include hollow structures, like the wall of blood vessels or the colon but also the envelope of volume organs, like the capsule of Glisson.
+
+In the following of this paper, we present a framework for realistic and real-time simulation of thin organs.
+Our method, presented in section \ref{sec:model} is based on shell theory and computes the internal forces using Finite Element Modeling (FEM) on curved triangles.
+The design of the elements relies on a polynomial interpolation that is also use to capture the interactions with other models, including volume models.
+We also present an automatic process to obtain meshes from image based reconstruction in an attempt to optimize  the computation using fewer elements  (see section \ref{sec:mesh}).
+The benefits of our approach (meshing of a curved surface, fast computation and possible interactions with volume models) are illustrated using various examples presented in section \ref{sec:results}.
 
 \section{A physical model for thin structures}
-
+\label{sec:model}
 \subsection{Related work}
@@
 \begin{equation}
-\textbf{K}_e = \int_v \textbf{J} \boldsymbol\chi \textbf{J}^{T} dV
+\textbf{K}_e = \int_v \textbf{J} \boldsymbol{\chi} \textbf{J}^{T} dV
 \end{equation}
@@
 \section{Meshing a curved surface}
-
+\label{sec:mesh}
 \subsection{Related work}
@@
 \section{Results}
+\label{sec:results}
 \subsection{Meshing of anatomical structures (error maps)}
@@
 \subsubsection{Coupling between tetrahedra and shells for more advanced modelling}
+\begin{figure}
+\centering
+\includegraphics[height=4cm]{images/stent_begin.png}
+\includegraphics[height=4cm]{images/stent_end.png}
+\caption {Avant... Apres...}
+\label{fig-stent}
+\end{figure}
 

### 2010-03-09 MICCAI2010/paper.tex
comments
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@86 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \usepackage{amsmath}
+\usepackage[latin1]{inputenc} % pour les caracteres francais dans les remarques
+
+\usepackage[usenames]{color}
+\newcommand{\SC}[1]{{\color{red}{\textbf{SC: #1}}}}
+\newcommand{\OC}[1]{{\color{blue}{\textbf{OC: #1}}}}
+\newcommand{\CD}[1]{{\color{magenta}{\textbf{CD: #1}}}}
+
+\newcommand{\nomore}[1]{}
 
@@
 Human body is composed of various deformable anatomical structures. 
-Realistic modeling of organ deformations is a challenging research field that leads the way to numerous clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedure and per-operative guidance based on simulation.
+Realistic modeling of organ deformations is a challenging research field that opens the door to new clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedure and per-operative guidance based on simulation.
 In all these cases, the clinician needs fast update of the deformation model to obtain a real-time display of the computed deformations.  
@@
 It seems unrealistic to target a unique model for all tissues even if there are obvious advantages to make maximum use of optimized generic models.
-Yet most of previous work focus on volume models that are able to capture the behavior of volume organs like liver \cite{} or brain tissues \cite{}.
+Yet most of previous work focus on volume models that are able to capture the behavior of volume organs like liver \cite{} or brain tissues \cite{} \CD{TODO: biblio}.
 In contrast, this paper seeks to propose a solution for simulating, in real-time, deformation of thin anatomical structures, whose volume is negligible compared to their surface area.
@@
 
-Development of a satisfactory physical model that runs in real-time but produces visually convincing animation of thin objects is challenging.
+Development of a satisfactory physical model that runs in real-time but produces visually convincing animation of thin objects is challenging. \CD{utile?}
 Rather than resorting to shell theory which involves the most complex formulations in continuum mechanics, previous works have often relied on discrete formulations.
@@
 For instance Wang et al.~\cite{Wang07} successfully used a network of linear and angular springs to describe bending and twisting of catheters and guidewires in an interventional radiology simulator.
-Yet, such models are limited in their ability to describe certain behaviour, as they do not rely on continuum mechanics.
+\CD{C'est pas des shells si?} Yet, such models are limited in their ability to describe certain behaviour, as they do not rely on continuum mechanics.
 Another limitation of such models is the difficulty to derive spring stiffness (in particular for angular springs) from elastic properties (Young's modulus and Poisson's ratio).
@@
 The difference between these two kinds of structures is very well explained by Liu et al.~\cite{Liu03} and can be summarized by the fact that plate bending elements can only carry transversal loads while shells can undergo more complex deformations.
-Considering the wide range of shapes and solicitations that can be applied to hollow organs, a thin plate model would not correctly capture their deformation and therefore a shell formulation was retained.
+Considering the wide range of shapes and solicitations that can be applied to hollow organs, a thin plate model would not correctly capture their deformation and therefore a shell formulation was retained for this work.
 
@@
 
-A complete description of our co-rotational triangular shell finite element model is available in one of our previous publication \cite{Comas2010ISBMS}.
+A complete description of our co-rotational triangular shell finite element model is available in one of our previous publication \cite{Comas2010ISBMS} \CD{Attention, il faut faire une citation anonyme !!}.
 Therefore we will only remind briefly the key points.
@@
 \begin{equation}
+\label{eq-deformation}
 e_{xx} = -z \frac{\partial^2u_z}{\partial x^2}
-\end{equation} 
-\begin{equation}
+\hspace{1cm}
 e_{yy} = -z \frac{\partial^2u_z}{\partial y^2}
-\end{equation} 
-\begin{equation}
+\hspace{1cm}
 e_{xy} = -2z \frac{\partial^2u_z}{\partial x \partial y}
@@
 
-Symbolically this may be expressed as $\textbf{e} = \textbf{Dc}$ where $\textbf{D}$ derives from equation~\ref{eq-deflection} and the above relations.
+Symbolically this may be expressed as $\textbf{e} = \textbf{Dc}$ where $\textbf{D}$ derives from equations~\ref{eq-deflection} and~\ref{eq-deformation}.
 Noting that $\textbf{c} = \textbf{C}^{-1}\textbf{u}$, we have:
@@
 We first sample the flat surface of each element by recursively dividing each triangle into four smaller ones and the deflection of each new vertex is computed using (\ref{eq-deflection}).
+\CD{Lien avec la fonction d'interpolation a mettre plus en avant !}.
 We then use this new mesh that samples the actual surface described by the shells to compute the Hausdorff distance with the targeted high resolution mesh.
@@
 The first step in the process of generating a shell-based mesh is an aggressive decimation of the targeted high resolution mesh.
-We then apply a heuristic method derived from the work of Saupin et.
+We then apply a heuristi"c method derived from the work of Saupin et.
 al \cite{Saupin07} with tetrahedral meshes based on very simple geometrical rules.

### 2010-03-10 MICCAI2010/paper.tex
Last update MICCAI
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@87 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 Human body is composed of various deformable anatomical structures.
-Realistic modeling of organ deformations is a challenging research field that opens the door to new clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedure and per-operative guidance based on simulation.
-In all these cases, the clinician needs fast update of the deformation model to obtain a real-time display of the computed deformations.  
-To realize this kind of interactive systems, one face the trade-off problem between accuracy of the model and the need of fast computations.
+Realistic modelling of organ deformations is a challenging research field that opens the door to new clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedure and per-operative guidance based on simulation.
+In all these cases the clinician needs fast update of the deformation model to obtain a real-time display of the computed deformations.
+To realise this kind of interactive systems, one faces the trade-off problem between accuracy of the model and the need of fast computations.
 
-An other key challenge of real-time human soft-tissue modeling is the variousness of the mechanical behaviors. 
-It seems unrealistic to target a unique model for all tissues even if there are obvious advantages to make maximum use of optimized generic models.
-Yet most of previous work focus on volume models that are able to capture the behavior of volume organs like liver \cite{} or brain tissues \cite{} \CD{TODO: biblio}.
+Another key challenge of real-time human soft-tissue modelling is the variousness of the mechanical behaviours.
+It seems unrealistic to target a unique model for all tissues even if there are obvious advantages to make maximum use of optimised generic models.
+Yet most of previous work focus on volume models that are able to capture the behavior of solid organs like liver \cite{} or brain tissues \cite{} \CD{TODO: biblio}.
 In contrast, this paper seeks to propose a solution for simulating, in real-time, deformation of thin anatomical structures, whose volume is negligible compared to their surface area.
-Examples of such anatomical structures include hollow structures, like the wall of blood vessels or the colon but also the envelope of volume organs, like the capsule of Glisson.
+Examples of such anatomical structures include hollow structures, like the wall of blood vessels or the colon but also the envelope of solid organs, like the capsule of Glisson.
 
 In the following of this paper, we present a framework for realistic and real-time simulation of thin organs.
-Our method, presented in section \ref{sec:model} is based on shell theory and computes the internal forces using Finite Element Modeling (FEM) on curved triangles.
-The design of the elements relies on a polynomial interpolation that is also use to capture the interactions with other models, including volume models.
-We also present an automatic process to obtain meshes from image based reconstruction in an attempt to optimize  the computation using fewer elements  (see section \ref{sec:mesh}).
-The benefits of our approach (meshing of a curved surface, fast computation and possible interactions with volume models) are illustrated using various examples presented in section \ref{sec:results}.
+Our method, presented in section~\ref{sec:model} is based on shell theory and computes the internal forces using Finite Element Modelling (FEM) on curved triangles.
+The design of the elements relies on a polynomial interpolation that is also used to capture the interactions with other models, including solid models.
+We also present an automatic process to obtain meshes from image based reconstruction in an attempt to optimise the computation using fewer elements (see section~\ref{sec:mesh}).
+The benefits of our approach (meshing of a curved surface, fast computation and possible interactions with solid models) are illustrated using various examples presented in section~\ref{sec:results}.
 
@@
 
-Development of a satisfactory physical model that runs in real-time but produces visually convincing animation of thin objects is challenging. \CD{utile?}
 Rather than resorting to shell theory which involves the most complex formulations in continuum mechanics, previous works have often relied on discrete formulations.
@@
 
-A complete description of our co-rotational triangular shell finite element model is available in one of our previous publication \cite{Comas2010ISBMS} \CD{Attention, il faut faire une citation anonyme !!}.
+A complete description and validation of our co-rotational triangular shell finite element model is available in one of our previous publication \cite{Comas2010ISBMS}.
 Therefore we will only remind briefly the key points.
@@
 \end{equation} 
-
 where $c_1$, \ldots , $c_9$ are constants.
+Using a third-degree polynomial expression allows us to reach a greater precision for both the computation of the bending energy and the interpolation within the surface of the shell.
 Let us define the vector $\textbf{u} = \left\{u_1 u_2 \ldots u_9 \right\} $ of the displacements and slopes at the three corners of the triangular plate using the following notations:
@@
 and so on for the two other vertices.
-We then can derive a matrix $C$ such as $\textbf{u} = \textbf{Cc}$ where $\textbf{c} = \left\{c_1 c_2 \ldots c_9 \right\} $.
+We then can derive a matrix $\textbf{C}$ such as $\textbf{u} = \textbf{Cc}$ where $\textbf{c} = \left\{c_1 c_2 \ldots c_9 \right\} $.
 
@@
 \end{equation} 
-
 Symbolically this may be expressed as $\textbf{e} = \textbf{Dc}$ where $\textbf{D}$ derives from equations~\ref{eq-deflection} and~\ref{eq-deformation}.
@@
 
+\subsubsection{Mechanical interactions with the curved surface of shells}
+But the practical interest of modelling complex behaviours such as bending and twisting would remain fairly low for medical simulation if contacts were not handled properly.
+In our case the difficulty comes from different sources.
+First the collision detection must be carried out with the curved surface of shell elements as opposed to the classic detection on plane triangles.
+Then forces applied to a given triangle need to be distributed between linear forces and torques onto its three vertices.
+As we will see, the same polynomial interpolation function \ref{eq-deflection} chosen to compute the bending energy in our FEM formulation is also used to capture the interactions between the curved surface and other objects.
+
+In order to detect the collision with the bent surface, we have chosen the subdivision approach.
+Therefore a finer mesh is created by adding vertices on the surface of each element.
+We first sample the flat surface of each element by recursively dividing each triangle into four smaller ones and the deflection of each new vertex is computed using (\ref{eq-deflection}) and the displacements and slopes at the three vertices of the triangular element.
+This process of subdivision allows us to render each shell as a curved triangle (Fig.~\ref{fig-shell} ) and detect any collision with the curved surface of the shell using any of the classic collision detection algorithms working on flat triangles.
+\begin{figure}
+\centering
+\includegraphics[height=3cm]{images/shell_curvature}
+\caption {The triangle formed by the three vertices of the shell has been recursively subdivided 3 times and the deflection of each new vertex was computed according to the same deflection function used in our shell finite element formulation.
+Sampling the actual surface of the shell allows rendering and more accurate collision detection.}
+\label{fig-shell}
+\end{figure}
+
+Once a collision has been detected, it must be processed by distributing the linear force received on the bent surface between the three vertices of the triangle.
+First the linear part of the force is simply transmitted on each node using the barycentric coordinates of the contact point's projection onto the triangle.
+
+The main difficulty is to convert the normal component of the force applied to the bent surface into a torque at each of the three nodes.
+Our approach is the following: during force computation, we use the angle measured at each node to compute the local deflection of each subvertex within the triangle.
+Differentiating the formulation twice yields a relation between the torque applied at each node and the generated force in bending.
+We therefore need to invert the latter formulation to convert a bending force into torques at each vertex.
+We start by retrieving the normal component of the applied force vector $F_{\mathrm{z}}$.
+We project the application point of the force into the triangle's plane and compute its local coordinates $(x,y)$.
+We create the polynom $P = F_{\mathrm{z}}(1 \hspace{0.2cm} x \hspace{0.2cm} y \hspace{0.2cm} x^2 \hspace{0.2cm} xy \hspace{0.2cm} y^2 \hspace{0.2cm} x^3 \hspace{0.2cm} xy^2 \hspace{0.2cm} y^3)^T$.
+The moments at each vertex are then obtained with $\Omega = (\textbf{C}^{-1})^T P$.
+Thus we are able to transmit any force coming from elaborate interactions with the curved surface of shells to the mechanical vertices used in our FEM formulation.
+
 \section{Meshing a curved surface}
@@
 
-A key component of finite element methods is the mesh of the simulated object being used.
+Another key component of finite element methods is the mesh of the simulated object being used.
 While a sufficient number of elements is needed to accurately describe the geometry, too many of them will yield a computationally expensive simulation and solving the system in real-time may no longer be possible.
@@
 Therefore we need to insure that the distance between the surface of our shell-based mesh and the targeted high resolution mesh will be minimal.
-To our knowledge the only tool available to measure error between two surfaces is the Hausdorff distance \cite{Klein96,Cignoni98}.
+An efficient technique for measuring the error between two surfaces is the Hausdorff distance \cite{Klein96,Cignoni98}.
 As a reminder the Hausdorff distance between two meshes is the maximum between the two so-called one-sided Hausdorff distances:
-
 \begin{equation}
@@
 where $d()$ is the Euclidian distance between two points.
-Since the curved surface of each shell must be taken into account in the computation of the Hausdorff distance, a finer mesh is created by adding vertices on the surface of each element.
-We first sample the flat surface of each element by recursively dividing each triangle into four smaller ones and the deflection of each new vertex is computed using (\ref{eq-deflection}).
-\CD{Lien avec la fonction d'interpolation a mettre plus en avant !}.
-We then use this new mesh that samples the actual surface described by the shells to compute the Hausdorff distance with the targeted high resolution mesh.
-The same process of subdivision allows us to render each shell as a curved triangle (Fig.~\ref{fig-shell} ).
+The same technique of subdivision used for rendering allows us to sample the actual surface described by the shells to compute the Hausdorff distance with the targeted high resolution mesh.
 
-\begin{figure}
-\centering
-\includegraphics[height=4cm]{images/shell_curvature}
-\caption {The triangle formed by the three vertices of the shell has been recursively subdivided 3 times and the deflection of each new vertex was computed according to the same deflection function used in our shell finite element formulation.
-Sampling the actual surface of the shell allows rendering and more accurate Hausdorff distance computation.
-}
-\label{fig-shell}
-\end{figure}
-
-The first step in the process of generating a shell-based mesh is an aggressive decimation of the targeted high resolution mesh.
-We then apply a heuristi"c method derived from the work of Saupin et.
+The first step in the process of generating a shell-based mesh is an aggressive simplification of the targeted high resolution mesh by using the quadric edge collapse decimation technique implemented in Meshlab \cite{Meshlab}.
+The algorithm tries as much as possible to preserve mesh boundaries and generate high quality triangular elements.
+We then apply a heuristic method derived from the work of Saupin et.
 al \cite{Saupin07} with tetrahedral meshes based on very simple geometrical rules.
@@
 \end{figure}
-
 \begin{figure}
@@
 \includegraphics[height=4cm]{images/cylinderResults2}
-\caption {(Left) Hausdorff distance colour map of the coarse mesh rendered with shells.
-(Right) Hausdorff distance colour map after one iteration of our algorithm.
+\caption {(Left) One-sided Hausdorff distance colour map of the coarse mesh rendered with shells.
+(Right) One-sided Hausdorff distance colour map after only one iteration of our algorithm.
 We can notice that differences with the target mesh have been subtantially reduced.}
@@
 \subsection{Deformations}
+
 \subsubsection{Computation times on 3 examples}
-\subsubsection{Coupling between tetrahedra and shells for more advanced modelling}
+
+\subsubsection{Coupling between tetrahedra and shells for advanced modelling}
+
 \begin{figure}
@@
 \includegraphics[height=4cm]{images/stent_end.png}
-\caption {Avant... Apres...}
+\caption {Before... After...}
 \label{fig-stent}

### 2010-03-10 MICCAI2010/compile.sh

@@ -1,0 +1,12 @@
+# Compiles paper properly, deletes temporary files and open the generated pdf
+
+rm paper.aux paper.bbl paper.blg paper.log
+
+pdflatex paper.tex
+bibtex paper.aux
+pdflatex paper.tex
+pdflatex paper.tex
+
+rm paper.aux paper.bbl paper.blg paper.log
+
+open paper.pdf

### 2010-03-10 MICCAI2010/paper.tex
Update angioplasty
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@88 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \subsubsection{Mechanical interactions with the curved surface of shells}
+\label{sec:interactions}
 But the practical interest of modelling complex behaviours such as bending and twisting would remain fairly low for medical simulation if contacts were not handled properly.
@@
 
+Structures in human body can be either solid (brain, liver, prostrate etc.) or hollow (blood vessels, colon, stomach etc.).
+However knowing how to model the two kind of structures is not sufficient to reach a high degree of accuracy in medical simulation.
+Indeed real life situations are more complex.
+As an example, the external surface of the liver is covered by a single layer of cells called Glisson's capsule.
+Its interaction with the liver plays an important role into the overall structure's mechanical behaviour.
+Therefore considering the interaction between solid and hollow objects is as crucial as modelling the two structures separately.
+
+An example of medical procedure to illustrate this point even further is angioplasty.
+Angioplasty is the technique of mechanically widening a narrowed or obstructed blood vessel, typically as a result of atherosclerosis.
+An empty and collapsed balloon on a guide wire, known as a balloon catheter, is passed into the narrowed locations and then inflated to a fixed size using water pressures.
+The balloon crushes the fatty deposits, so opening up the blood vessel to improved flow.
+As a proof of concept we tried to simulate an angioplasty (Fig.~\ref{fig-stent}).
+The blood vessel is modelled using the shell FEM formulation described in this paper and the fatty deposits are simulated with a tetrahedral FEM method and are fixed to the interior wall of the blood vessel.
+When the balloon inflates it crushes the deposits and the deposits then apply a pressure onto the curved surfaces of shells modelling the interior wall.
+The forces are then distributed onto the mechanical nodes of the blood vessel mesh as detailled in section \ref{sec:interactions}, which widens the blood vessel as expected.
+
 \begin{figure}
@@
 \includegraphics[height=4cm]{images/stent_end.png}
-\caption {Before... After...}
+\caption {Simulation of an angioplasty procedure.
+(Left) The collapsed balloon was inserted into the blood vessel simulated with our shell FEM formulation.
+The fatty deposits (in yellow) are modelled with a tetrahedral FEM.
+(Right) Upon inflation the balloon is crushing the fatty deposits, which is applying a pressure onto the interior wall and widening the blood vessel.}
 \label{fig-stent}

### 2010-03-10 MICCAI2010/paper.tex
Last update for MICCAI (added screenshots for results section)
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@90 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \centering
-\includegraphics[height=4cm]{images/cylinderResults2}
+\includegraphics[height=4cm]{images/resultsCylinder}
 \caption {(Left) One-sided Hausdorff distance colour map of the coarse mesh rendered with shells.
@@
 \label{sec:results}
-\subsection{Meshing of anatomical structures (error maps)}
+\subsection{Meshing of anatomical structures}
+
+This approach has been applied to mesh more complex anatomical structures with curved shell elements. In each case the error is expressed as a percentage of the diagonal of the object's bounding box.
+
+\begin{figure}
+\centering
+\includegraphics[height=4cm]{images/resultsLiver}
+\caption {Left: the targeteted high resolution liver mesh (8,000 triangles).
+Right: the one-sided Hausdorff distance error map after applying only one iteration of our algorithm to the coarse mesh (1,200 shells)}.
+\label{fig-liver}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[height=4cm]{images/resultsAnevrism}
+\caption {Left: the targeteted high resolution anevrism mesh (28,368 triangles).
+Right: the one-sided Hausdorff distance error map after applying only one iteration of our algorithm to the coarse mesh (1,160 shells)}.
+\label{fig-anevrism}
+\end{figure}
 

### 2010-03-10 MICCAI2010/paper.tex
Added computation time (no sentence)
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@91 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
Modifs intro
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@92 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 To realise this kind of interactive systems, one faces the trade-off problem between accuracy of the model and the need of fast computations.
-
 Another key challenge of real-time human soft-tissue modelling is the variousness of the mechanical behaviours.
@@
 
-In the following of this paper, we present a framework for realistic and real-time simulation of thin organs.
+Shell theory allows the modelling of  structure  deformations when the thickness is small compared to its other dimensions~\cite{Liu03}.
+The key idea is to model the physical shell as if it were a surface of zero thickness, but endowed with mechanical properties in the form of elastic resistance to stretching, twisting, bending and shearing forces.
+Rather than resorting to shell theory, previous work on medical simulation often relies on mass-spring models (see \ref{Hammer08} \CD{voir fichier bib isbms} for instance).
+%Rather than resorting to shell theory which involves the most complex formulations in continuum mechanics, previous works have often relied on discrete formulations.
+Early approaches only considered in-plane deformation, and often relied on mass-spring models.
+More recent works have considered adding bending through angular springs.
+For instance Wang et al.~\cite{Wang07} successfully used a network of linear and angular springs to describe bending and twisting of catheters and guidewires in an interventional radiology simulator.
+\CD{C'est pas des shells si?}
+Yet, such models are limited in their ability to describe certain behaviour, as they do not rely on continuum mechanics: it is difficult to derive spring stiffness (in particular for angular springs) from elastic properties (Young's modulus and Poisson's ratio).
+The use of fast shell based model for computer graphics and interactive simulation was recently introduced with the work of Choi~\textit{et al.}~\cite{Choi}.
+The model relies on simplified energy functions and precomputed modal analysis.
+It allows for fast and visually realistic results.
+We propose to rely an a similar approach but with more accuracy in order to be applicable to medical simulation.
+Our model is not based on modal analysis but uses a corotational formulation and polynomial shape functions.
+In \cite{Comas2010ISBMS}, we present and apply this shell model for a flat deformable implant.
+
+In this paper, we propose to apply this technique for modelling the deformations of some anatomical structures.
+Therefore, one of the main problem is to geometrically model the surface of the bodies using the shape functions of the shell model.
+This problem is quite similar to the reconstruction of the surface of organs with continuous surface patches that has been extensively studied \CD{L'ideal serait un survey en fait !!}~\cite{} .
+\CD{Je te laisse completer}
+
+Our contribution is to unify these two approaches (shell based modeling and geometric reconstruction using shape functions) in a framework  for realistic and real-time simulation of thin organ deformations.
+In addition, we demonstrate the ability of the technique to simulate thin anatomical membrane which enfolds a solid organ without any requirement on mesh correspondence.
 Our method, presented in section~\ref{sec:model} is based on shell theory and computes the internal forces using Finite Element Modelling (FEM) on curved triangles.
@@
 
-\section{A physical model for thin structures}
+\section{Co-rotational triangular shell model for thin structures}
 \label{sec:model}
-\subsection{Related work}
+%\subsection{Related work}
 
-Rather than resorting to shell theory which involves the most complex formulations in continuum mechanics, previous works have often relied on discrete formulations.
-Early approaches only considered in-plane deformation, and often relied on mass-spring models.
-More recent works have considered adding bending through angular springs.
-For instance Wang et al.~\cite{Wang07} successfully used a network of linear and angular springs to describe bending and twisting of catheters and guidewires in an interventional radiology simulator.
-\CD{C'est pas des shells si?} Yet, such models are limited in their ability to describe certain behaviour, as they do not rely on continuum mechanics.
-Another limitation of such models is the difficulty to derive spring stiffness (in particular for angular springs) from elastic properties (Young's modulus and Poisson's ratio).
-For these reasons, other approaches have been proposed.
-
-Continuum mechanics provides many formulations able to accurately describe stresses occurring within thin objects.
-Most of them fall into one of the following two categories: plate theory or shell theory.
-Those theories have been a subject of interest in the mechanical community for decades.
-The difference between these two kinds of structures is very well explained by Liu et al.~\cite{Liu03} and can be summarized by the fact that plate bending elements can only carry transversal loads while shells can undergo more complex deformations.
-Considering the wide range of shapes and solicitations that can be applied to hollow organs, a thin plate model would not correctly capture their deformation and therefore a shell formulation was retained for this work.
-
-\subsection{Co-rotational triangular shell model}
+%\subsection{}
 
@@
 where the strain-displacement matrix $\textbf{b} = \textbf{DC}^{-1}$. 
-The stiffness matrix $\textbf{K}_e$ for an element is then classically obtained from
+The stiffness matrix $\textbf{K}_e$ for an element is then obtained from:
 \begin{equation}
@@
 
-\subsubsection{Computation times on 3 examples}
+\subsubsection{Computation times}
+
+Anevrism: 7.5 FPS with a timestep of 0.02s so we would need to be about 7 times faster to be real-time (easily doable on GPU)
 

### 2010-03-10 MICCAI2010/paper.tex
abstract + qqs modifs
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@93 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{abstract}
-
+This paper presents a new modeling technique for the deformation of thin anatomical structures like membrane and hollow organs. 
+We show that the behavior of this type of tissue can be abstracted with a 2D modeling of their surface elastic resistance using shell theory. 
+In order to apply the shell theory in the particular context of medical simulation, our method propose to base the geometrical reconstruction of the organ on the shape functions of the shell element. 
+Moreover, we also use these continuous shape functions to handle the contacts and the interactions with other types of deformable tissues.
+The technique is illustrated using several examples including the simulation of angioplasty procedure.
 \end{abstract}
@@
 Shell theory allows the modelling of  structure  deformations when the thickness is small compared to its other dimensions~\cite{Liu03}.
-The key idea is to model the physical shell as if it were a surface of zero thickness, but endowed with mechanical properties in the form of elastic resistance to stretching, twisting, bending and shearing forces.
+The key idea is to model the physical shell as if it were a surface of zero thickness, but endowed with mechanical properties in the form of elastic resistance to stretching, bending and shearing forces.
 Rather than resorting to shell theory, previous work on medical simulation often relies on mass-spring models (see \ref{Hammer08} \CD{voir fichier bib isbms} for instance).
@@
 \label{sec:interactions}
-But the practical interest of modelling complex behaviours such as bending and twisting would remain fairly low for medical simulation if contacts were not handled properly.
+But the practical interest of modelling complex behaviours such as bending and twisting would remain fairly low for medical simulation if contacts and constraints were not handled properly.
 In our case the difficulty comes from different sources.

### 2010-03-10 MICCAI2010/paper.tex
figures
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@94 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \centering
-\includegraphics[height=3cm]{images/shell_curvature}
+\includegraphics[height=4cm]{images/shell_curvature2.png}
 \caption {The triangle formed by the three vertices of the shell has been recursively subdivided 3 times and the deflection of each new vertex was computed according to the same deflection function used in our shell finite element formulation.
@@
 A simple example is shown Fig.
-\ref{fig-cylinder} and \ref{fig-cylinderResults} to illustrate the method.
+\ref{fig-cylinder}
+%and \ref{fig-cylinderResults} 
+to illustrate the method.
 
@@
 \centering
-\includegraphics[height=4cm]{images/cylinderHighWireframe}
-\includegraphics[height=4cm]{images/cylinderLow}
-\caption {Illustration of our method on a simple example.
-The target is the high resolution cylinder mesh (left) and the start point is a very coarse mesh approximating the shape of a cylinder (right).}
-\label{fig-cylinder}
-\end{figure}
-\begin{figure}
-\centering
-\includegraphics[height=4cm]{images/resultsCylinder}
-\caption {(Left) One-sided Hausdorff distance colour map of the coarse mesh rendered with shells.
+\includegraphics[height=2.6cm]{images/cylinderHighWireframe}
+\includegraphics[height=2.6cm]{images/cylinderLow}
+\includegraphics[height=2.6cm]{images/resultsCylinder}
+\caption {\CD{Ca serait bien d'ajouter des (a) - (b) - (c) - (d) en dessous des figures}Illustration of our method on a simple example.
+The target is the high resolution cylinder mesh (left) and the start point is a very coarse mesh approximating the shape of a cylinder.
+(Left) One-sided Hausdorff distance colour map of the coarse mesh rendered with shells.
 (Right) One-sided Hausdorff distance colour map after only one iteration of our algorithm.
 We can notice that differences with the target mesh have been subtantially reduced.}
-\label{fig-cylinderResults}
+\label{fig-cylinder}
 \end{figure}
@@
 \centering
-\includegraphics[height=4cm]{images/resultsLiver}
+\includegraphics[height=3cm]{images/resultsLiver}
 \caption {Left: the targeteted high resolution liver mesh (8,000 triangles).
@@
 \centering
-\includegraphics[height=4cm]{images/resultsAnevrism}
+\includegraphics[height=3cm]{images/resultsAnevrism}
 \caption {Left: the targeteted high resolution anevrism mesh (28,368 triangles).

### 2010-03-11 MICCAI2010/paper.tex
Last update (introduction+review reconstruction)
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@95 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
+%\newcommand{\SC}[1]{}
+%\newcommand{\OC}[1]{}
+%\newcommand{\CD}[1]{}
+
 \newcommand{\nomore}[1]{}
@@
 
-\title{A Framework for Modelling Thin Objects: \\ from Meshing to Shell Finite Element Method}
+\title{Shell Model for Reconstruction and Simulation of Thin Anatomical Structures}
 %\author{Olivier Comas, St\'ephane Cotin and Christian Duriez}
@@
 \begin{abstract}
-This paper presents a new modeling technique for the deformation of thin anatomical structures like membrane and hollow organs. 
-We show that the behavior of this type of tissue can be abstracted with a 2D modeling of their surface elastic resistance using shell theory. 
+This paper presents a new modelling technique for the deformation of thin anatomical structures like membranes and hollow organs.
+We show that the behaviour of this type of tissue can be abstracted with a 2D modelling of their surface elastic resistance using shell theory.
 In order to apply the shell theory in the particular context of medical simulation, our method propose to base the geometrical reconstruction of the organ on the shape functions of the shell element.
 Moreover, we also use these continuous shape functions to handle the contacts and the interactions with other types of deformable tissues.
-The technique is illustrated using several examples including the simulation of angioplasty procedure.
+The technique is illustrated using several examples including the simulation of an angioplasty procedure.
 \end{abstract}
@@
 To realise this kind of interactive systems, one faces the trade-off problem between accuracy of the model and the need of fast computations.
-Another key challenge of real-time human soft-tissue modelling is the variousness of the mechanical behaviours.
+Another key challenge of soft-tissue modelling is the variousness of the mechanical behaviours.
 It seems unrealistic to target a unique model for all tissues even if there are obvious advantages to make maximum use of optimised generic models.
-Yet most of previous work focus on volume models that are able to capture the behavior of solid organs like liver \cite{} or brain tissues \cite{} \CD{TODO: biblio}.
-In contrast, this paper seeks to propose a solution for simulating, in real-time, deformation of thin anatomical structures, whose volume is negligible compared to their surface area.
+Yet most of previous work focus on volumetric models that are able to capture the behaviour of solid organs like liver \cite{} or brain tissues \cite{} \CD{TODO: biblio}.
+In contrast, this paper seeks to propose a solution for simulating, close to real-time, deformation of thin anatomical structures whose volume is negligible compared to their surface area.
 Examples of such anatomical structures include hollow structures, like the wall of blood vessels or the colon but also the envelope of solid organs, like the capsule of Glisson.
@@
 Shell theory allows the modelling of structure deformations when the thickness is small compared to its other dimensions~\cite{Liu03}.
-The key idea is to model the physical shell as if it were a surface of zero thickness, but endowed with mechanical properties in the form of elastic resistance to stretching, bending and shearing forces.
-Rather than resorting to shell theory, previous work on medical simulation often relies on mass-spring models (see \ref{Hammer08} \CD{voir fichier bib isbms} for instance).
-%Rather than resorting to shell theory which involves the most complex formulations in continuum mechanics, previous works have often relied on discrete formulations.
-Early approaches only considered in-plane deformation, and often relied on mass-spring models.
-More recent works have considered adding bending through angular springs.
-For instance Wang et al.~\cite{Wang07} successfully used a network of linear and angular springs to describe bending and twisting of catheters and guidewires in an interventional radiology simulator.
-\CD{C'est pas des shells si?}
+The key idea is to model the physical shell as if it were a surface of zero thickness, but endowed with mechanical properties in the form of elastic resistance to stretching and bending forces.
+\OC{Pas de shearing il me semble.
+Et il y a bien une thickness dans la formulation, ta phrase laisse sous-entendre qu'on la prend pas en compte.}
+Rather than resorting to shell theory, previous work on medical simulation often relies on linear or angular mass-spring models as in \cite{Nedel98,Hammer08}.
 Yet, such models are limited in their ability to describe certain behaviour, as they do not rely on continuum mechanics: it is difficult to derive spring stiffness (in particular for angular springs) from elastic properties (Young's modulus and Poisson's ratio).
-The use of fast shell based model for computer graphics and interactive simulation was recently introduced with the work of Choi~\textit{et al.}~\cite{Choi}.
+The use of fast shell based model for computer graphics and interactive simulation was recently introduced with the work of Choi et al.~\cite{Choi07}.
 The model relies on simplified energy functions and precomputed modal analysis.
-It allows for fast and visually realistic results.
+It allows fast and visually realistic results.
 We propose to rely an a similar approach but with more accuracy in order to be applicable to medical simulation.
-Our model is not based on modal analysis but uses a corotational formulation and polynomial shape functions.
-In \cite{Comas2010ISBMS}, we present and apply this shell model for a flat deformable implant.
+Our model is not based on modal analysis but uses a co-rotational formulation and polynomial shape functions.
+In~\cite{Comas2010ISBMS}, we present and apply this shell model to a flat deformable implant.
+\OC{ptet un peu trop specifique pour preserver l'anonymat non ?}
 
 In this paper, we propose to apply this technique for modelling the deformations of some anatomical structures.
-Therefore, one of the main problem is to geometrically model the surface of the bodies using the shape functions of the shell model.
-This problem is quite similar to the reconstruction of the surface of organs with continuous surface patches that has been extensively studied \CD{L'ideal serait un survey en fait !!}~\cite{} .
-\CD{Je te laisse completer}
+To be able to do this, one of the main problem is to geometrically model the surface of the structures using curved triangular shell elements.
+This problem is quite similar to the reconstruction of the surface of objects in computer vision.
+Indeed calculating curvature maps of 3D surfaces represented as digitised data (point clouds or triangulated meshes) has been extensively studied in this field.
+One of the most common approach is to use continuous surface patches~\cite{Kolb95,Douros02}.
+The surface is locally approximated by an analytic representation called surface patch, usually chosen to be an implicit surface.
+In~\cite{Tang99} an elegant and non noise-sensitive approach that infers sign and direction of principal curvatures directly from the input is presented and the authors use this information for coherent surface and curve extraction from 3D data.
+However, our situation is substantially different as extracting the surface is not our ultimate goal.
+We want to model the deformation of the structure and in that regard the curvature of the surface has a physical meaning: it represents the rest shape of our anatomical structure, that is the configuration with zero energy.
+Therefore we need to match the shape (polynomial) functions used in our shell FEM formulation to describe the surface of the shells with the surface of the anatomical structures that we wish to simulate.
 
-Our contribution is to unify these two approaches (shell based modeling and geometric reconstruction using shape functions) in a framework  for realistic and real-time simulation of thin organ deformations.
+Our contribution is to unify these two approaches (shell based modelling and geometric surface reconstruction) separately used in mechanical engineering and computer vision in a framework for realistic simulation of thin organ deformations.
 In addition, we demonstrate the ability of the technique to simulate thin anatomical membrane which enfolds a solid organ without any requirement on mesh correspondence.
-Our method, presented in section~\ref{sec:model} is based on shell theory and computes the internal forces using Finite Element Modelling (FEM) on curved triangles.
-The design of the elements relies on a polynomial interpolation that is also used to capture the interactions with other models, including solid models.
-We also present an automatic process to obtain meshes from image based reconstruction in an attempt to optimise the computation using fewer elements (see section~\ref{sec:mesh}).
-The benefits of our approach (meshing of a curved surface, fast computation and possible interactions with solid models) are illustrated using various examples presented in section~\ref{sec:results}.
+\OC{j'enleverais bien cette derniere phrase moi, on en fait pas la demonstration au final, on evoque juste que ca serait possible} Our method, presented in section~\ref{sec:model}, is based on shell theory and computes the internal forces using Finite Element Modelling (FEM) on curved triangles.
+The design of the elements relies on a polynomial interpolation function that is also used to capture and process the interactions with other models, including solid models.
+In section~\ref{sec:mesh} we also present an automatic process to obtain meshes from image based reconstruction that relies on the exact same polynomial function.
+% in an attempt to optimise the computation using fewer elements. 
+Finally the benefits of our approach (meshing of a curved surface, fast computation and possible interactions with solid models) are illustrated using various examples presented in section~\ref{sec:results}.
 
@@
 \label{sec:model}
-%\subsection{Related work}
-
-%\subsection{}
 
@@
 
-\subsubsection{Triangular elastic membrane}
+\subsection{Triangular elastic membrane}
 
@@
 
-\subsubsection{Triangular plate bending}
+\subsection{Triangular plate bending}
 
@@
 \centering
-\includegraphics[height=5cm]{images/bending}
+\includegraphics[height=4cm]{images/bending}
 \caption {The different degrees of freedom $u$ of a triangular thin plate in bending are illustrated above.}
@@
 
-\subsubsection{Mechanical interactions with the curved surface of shells}
+\subsection{Mechanical interactions with the curved surface of shells}
 \label{sec:interactions}
@@
 Then forces applied to a given triangle need to be distributed between linear forces and torques onto its three vertices.
-As we will see, the same polynomial interpolation function \ref{eq-deflection} chosen to compute the bending energy in our FEM formulation is also used to capture the interactions between the curved surface and other objects.
+As we will see, the same polynomial interpolation function (\ref{eq-deflection}) chosen to compute the bending energy in our FEM formulation is also used to capture the interactions between the curved surface and other objects.
 
@@
 
-\section{Meshing a curved surface}
+\section{A mechanically-driven method to mesh curved surfaces with shells}
 \label{sec:mesh}
-\subsection{Related work}
+%\subsection{Related work}
 
-Another key component of finite element methods is the mesh of the simulated object being used.
+%Another key component of finite element methods is the mesh of the simulated object being used.
 While a sufficient number of elements is needed to accurately describe the geometry, too many of them will yield a computationally expensive simulation and solving the system in real-time may no longer be possible.
@@
 
-\subsection{A mechanically-driven method to mesh curved surfaces with shells}
-
-Indeed, one main difference between shell and plate elements is that the rest shape of the former can describe a curved surface.
-We propose to take this crucial feature into account to optimise the meshing process.
-While many flat triangles are often required to describe highly curved surfaces, fewer triangular shell elements are needed to describe the given geometry with the same precision since they can be naturally curved.
+Because the surface of an anatomical structure has a physical meaning, we cannot use any type of function to reconstruct this surface.
+The surface must be patched with triangular elements whose interpolation makes use of the same shape function designed for our shell FEM formulation.
+By suggesting a method to reconstruct the surface of any structures with curved shell, the meshing process is optimised.
+Indeed while many flat triangles are often required to describe highly curved surfaces, fewer triangular shell elements are needed to describe the given geometry with the same precision since they can be naturally curved.
 In the following we assume that we have a high resolution triangular mesh obtained from a binary segmented image of the organ we want to simulate (via a Marching Cube algorithm for instance).
@@
 This technique locally smooths the surface of the mesh while converging towards the desired high resolution mesh.
-At each iteration of this algorithm we measure the error with the target using the Hausdorff distance and the process is stopped when the required precision has been reached.
-A simple example is shown Fig.
-\ref{fig-cylinder}
-%and \ref{fig-cylinderResults} 
-to illustrate the method.
+At each iteration of this algorithm we measure the error between the curved surface of shells and the target using the Hausdorff distance and the process is stopped when the required precision has been reached.
+A simple example is shown Fig.~\ref{fig-cylinder} to illustrate the method.
+We can notice that differences with the target mesh have been subtantially reduced.
 
@@
 \centering
-\includegraphics[height=2.6cm]{images/cylinderHighWireframe}
-\includegraphics[height=2.6cm]{images/cylinderLow}
-\includegraphics[height=2.6cm]{images/resultsCylinder}
-\caption {\CD{Ca serait bien d'ajouter des (a) - (b) - (c) - (d) en dessous des figures}Illustration of our method on a simple example.
-The target is the high resolution cylinder mesh (left) and the start point is a very coarse mesh approximating the shape of a cylinder.
-(Left) One-sided Hausdorff distance colour map of the coarse mesh rendered with shells.
-(Right) One-sided Hausdorff distance colour map after only one iteration of our algorithm.
-We can notice that differences with the target mesh have been subtantially reduced.}
+\includegraphics[height=2.5cm]{images/exampleCylinder}
+\caption {Illustration of our method on a simple example.
+The target is a high resolution cylinder mesh of 16,384 triangles (a) and we start from a very coarse mesh (12 triangles) approximating the shape of a cylinder, rendered with flat triangles here (b).
+In (c) the same coarse mesh is rendered with shells and a one-sided Hausdorff distance colour map is applied to show the initial error with the high resolution mesh.
+(d) One-sided Hausdorff distance colour map after only one iteration of our algorithm (48 shells).}
 \label{fig-cylinder}
@@
 
-\subsection{Deformations}
-
-\subsubsection{Computation times}
+\subsection{Computation times}
 
@@
 
-\subsubsection{Coupling between tetrahedra and shells for advanced modelling}
+\subsection{Coupling between tetrahedra and shells for advanced modelling}
 
@@
 \centering
-\includegraphics[height=4cm]{images/stent_begin.png}
-\includegraphics[height=4cm]{images/stent_end.png}
+\includegraphics[height=3.8cm]{images/angioplasty.png}
 \caption {Simulation of an angioplasty procedure.
-(Left) The collapsed balloon was inserted into the blood vessel simulated with our shell FEM formulation.
+(a) The collapsed balloon was inserted into the blood vessel simulated with our shell FEM formulation.
 The fatty deposits (in yellow) are modelled with a tetrahedral FEM.
-(Right) Upon inflation the balloon is crushing the fatty deposits, which is applying a pressure onto the interior wall and widening the blood vessel.}
+(b) Upon inflation the balloon is crushing the fatty deposits, which is applying a pressure onto the interior wall and widening the blood vessel.}
 \label{fig-stent}

### 2010-03-11 examples/conv.sh
demos for paper MICCAI
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@97 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@ -1,1 +1,12 @@
-awk 'BEGIN { print "<Mesh position=\""; section=0 } /Coordinates/ { section=1 } (NF==4&&section==1) { print $2 " " $3 " " $4 } /Elements/ { section+=1; if (section==2) { print "\"\n tetrahedra=\"" } } (NF==5&&section==2) { print $2-1 " " $3-1 " " $4-1 " " $5-1 } END { print "\" />" }'
+awk 'BEGIN { print "<Mesh position=\""; section=0 } 
+/Coordinates/ { section=section+1 } 
+/Triangle/ {section=2}
+(NF==4&&section==1) { print $2 " " $3 " " $4 } 
+/Elements/ {  
+	     if (section==1) { print "\"\n tetrahedra=\""  } 
+	     if (section==2)  {print "\"\n triangles=\"" } 
+	    } 
+
+(NF==5&&section==1) { print $2-1 " " $3-1 " " $4-1 " " $5-1 } 
+(NF==4&&section==2) {print $2-1 " " $4-1 " " $3-1 }
+END { print "\" />" }'

### 2010-03-11 MICCAI2010/paper.tex
Last update (added conclusion + few changes in intro and images)
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@103 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\title{Shell Model for Reconstruction and Simulation of Thin Anatomical Structures}
+\title{Shell Model for Reconstruction and Real-Time Simulation of Thin Anatomical Structures}
 %\author{Olivier Comas, St\'ephane Cotin and Christian Duriez}
@@
 The key idea is to model the physical shell as if it were a surface of zero thickness, but endowed with mechanical properties in the form of elastic resistance to stretching and bending forces.
-\OC{Pas de shearing il me semble.
-Et il y a bien une thickness dans la formulation, ta phrase laisse sous-entendre qu'on la prend pas en compte.}
+\OC{TODO: reformuler cette phrase pour enlever zero thickness}
 Rather than resorting to shell theory, previous work on medical simulation often relies on linear or angular mass-spring models as in \cite{Nedel98,Hammer08}.
@@
 We propose to rely an a similar approach but with more accuracy in order to be applicable to medical simulation.
-Our model is not based on modal analysis but uses a co-rotational formulation and polynomial shape functions.
-In~\cite{Comas2010ISBMS}, we present and apply this shell model to a flat deformable implant.
-\OC{ptet un peu trop specifique pour preserver l'anonymat non ?}
+Our model is not based on modal analysis but uses a co-rotational formulation and polynomial shape functions presented in~\cite{Comas2010ISBMS}.
 
@@
 Our contribution is to unify these two approaches (shell based modelling and geometric surface reconstruction) separately used in mechanical engineering and computer vision in a framework for realistic simulation of thin organ deformations.
-In addition, we demonstrate the ability of the technique to simulate thin anatomical membrane which enfolds a solid organ without any requirement on mesh correspondence.
-\OC{j'enleverais bien cette derniere phrase moi, on en fait pas la demonstration au final, on evoque juste que ca serait possible} Our method, presented in section~\ref{sec:model}, is based on shell theory and computes the internal forces using Finite Element Modelling (FEM) on curved triangles.
+Our method, presented in section~\ref{sec:model}, is based on shell theory and computes the internal forces using Finite Element Modelling (FEM) on curved triangles.
 The design of the elements relies on a polynomial interpolation function that is also used to capture and process the interactions with other models, including solid models.
@@
 \includegraphics[height=3cm]{images/resultsLiver}
-\caption {Left: the targeteted high resolution liver mesh (8,000 triangles).
+\caption {Left: the targeteted high resolution Glisson's capsule mesh (8,000 triangles).
 Right: the one-sided Hausdorff distance error map after applying only one iteration of our algorithm to the coarse mesh (1,200 shells)}.
@@
 \centering
-\includegraphics[height=3cm]{images/resultsAnevrism}
-\caption {Left: the targeteted high resolution anevrism mesh (28,368 triangles).
-Right: the one-sided Hausdorff distance error map after applying only one iteration of our algorithm to the coarse mesh (1,160 shells)}.
-\label{fig-anevrism}
+\includegraphics[height=3cm]{images/resultsAneurysm}
+\caption {(a) the targeteted high resolution aneurysm mesh (28,368 triangles). (b): the one-sided Hausdorff distance error map on a mesh of 1,160 shells generated with our method}.
+\label{fig-aneurysm}
 \end{figure}
@@
 \subsection{Computation times}
-
-Anevrism: 7.5 FPS with a timestep of 0.02s so we would need to be about 7 times faster to be real-time (easily doable on GPU)
 
@@
 \section{Conclusion}
-\section*{Acknowledgments}
+We proposed a complete framework for real-time modelling of thin anatomical structures.
+The novelty of our method relies on the combination of a shell finite element formulation and a geometric surface reconstruction both based on the same polynomial interpolation function used to describe the surface of shells.
+We also showed how contacts and interactions with the curved surfaces of shells could be handled using the same function.
+The efficiency of the whole approach was demonstrated through shell reconstruction and real-time modelling of a blood vessel showing signs of an aneurysm.
+Finally these results are very encouraging towards fine modelling of more complex situations like interactions existing between the liver and the Glisson's capsule or the simulation of an angioplasty procedure.
+
+%\section*{Acknowledgments}
 

### 2010-03-11 MICCAI2010/paper.tex
h-11
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@105 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 This paper presents a new modelling technique for the deformation of thin anatomical structures like membranes and hollow organs.
-We show that the behaviour of this type of tissue can be abstracted with a 2D modelling of their surface elastic resistance using shell theory.
+We show that the behaviour of this type of tissue can be abstracted with a modelling of their surface elastic resistance using shell theory.
 In order to apply the shell theory in the particular context of medical simulation, our method propose to base the geometrical reconstruction of the organ on the shape functions of the shell element.
@@
 
-Human body is composed of various deformable anatomical structures.
+The human body is composed of various deformable anatomical structures.
 Realistic modelling of organ deformations is a challenging research field that opens the door to new clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedure and per-operative guidance based on simulation.
 In all these cases the clinician needs fast update of the deformation model to obtain a real-time display of the computed deformations.
-To realise this kind of interactive systems, one faces the trade-off problem between accuracy of the model and the need of fast computations.
+To develop this kind of interactive systems, one faces the trade-off problem between accuracy of the model and the need for fast computations.
 Another key challenge of soft-tissue modelling is the variousness of the mechanical behaviours.
 It seems unrealistic to target a unique model for all tissues even if there are obvious advantages to make maximum use of optimised generic models.
-Yet most of previous work focus on volumetric models that are able to capture the behaviour of solid organs like liver \cite{} or brain tissues \cite{} \CD{TODO: biblio}.
-In contrast, this paper seeks to propose a solution for simulating, close to real-time, deformation of thin anatomical structures whose volume is negligible compared to their surface area.
-Examples of such anatomical structures include hollow structures, like the wall of blood vessels or the colon but also the envelope of solid organs, like the capsule of Glisson.
+Yet most of previous work focus on volumetric models that are able to capture the behaviour of solid organs like the liver or the brain (see for instance \cite{Miller07} \cite{Delingette08}).
+In contrast, this paper seeks to propose a solution for simulating, in real-time, the deformation of thin anatomical structures whose volume is negligible compared to their surface area.
+Examples of such anatomical structures include hollow structures, such as the wall of blood vessels, or membranes,  such the capsule of Glisson.
 
 Shell theory allows the modelling of structure deformations when the thickness is small compared to its other dimensions~\cite{Liu03}.
-The key idea is to model the physical shell as if it were a surface of zero thickness, but endowed with mechanical properties in the form of elastic resistance to stretching and bending forces.
-\OC{TODO: reformuler cette phrase pour enlever zero thickness}
+The key idea is to model the physical shell as a surface but endowed with mechanical properties in the form of elastic resistance to stretching and bending forces.
 Rather than resorting to shell theory, previous work on medical simulation often relies on linear or angular mass-spring models as in \cite{Nedel98,Hammer08}.
@@
 The model relies on simplified energy functions and precomputed modal analysis.
-It allows fast and visually realistic results.
-We propose to rely an a similar approach but with more accuracy in order to be applicable to medical simulation.
+It allows for fast and visually realistic results.
+We propose to rely on a similar approach but with more accuracy in order to be applicable to medical simulation.
 Our model is not based on modal analysis but uses a co-rotational formulation and polynomial shape functions presented in~\cite{Comas2010ISBMS}.
 
-In this paper, we propose to apply this technique for modelling the deformations of some anatomical structures.
-To be able to do this, one of the main problem is to geometrically model the surface of the structures using curved triangular shell elements.
-This problem is quite similar to the reconstruction of the surface of objects in computer vision.
-Indeed calculating curvature maps of 3D surfaces represented as digitised data (point clouds or triangulated meshes) has been extensively studied in this field.
+In this paper, we propose for the first time to apply this technique for modelling the deformations of some anatomical structures.
+The first step is to geometrically model the surface of the structures using curved triangular shell elements.
+This process is quite similar to the reconstruction of the surface of objects in computer vision.
+Indeed calculating curvature maps of 3D surfaces represented as digitized data (point clouds or triangulated meshes) has been extensively studied.
 One of the most common approach is to use continuous surface patches~\cite{Kolb95,Douros02}.
 The surface is locally approximated by an analytic representation called surface patch, usually chosen to be an implicit surface.
-In~\cite{Tang99} an elegant and non noise-sensitive approach that infers sign and direction of principal curvatures directly from the input is presented and the authors use this information for coherent surface and curve extraction from 3D data.
-However, our situation is substantially different as extracting the surface is not our ultimate goal.
-We want to model the deformation of the structure and in that regard the curvature of the surface has a physical meaning: it represents the rest shape of our anatomical structure, that is the configuration with zero energy.
-Therefore we need to match the shape (polynomial) functions used in our shell FEM formulation to describe the surface of the shells with the surface of the anatomical structures that we wish to simulate.
+These works target non noise-sensitive approaches and coherent surface and curve extraction from 3D data~\cite{Tang99}.
+%In~\cite{Tang99} an elegant and non noise-sensitive approach that infers sign and direction of principal curvatures directly from the input is presented and the authors use this information for coherent surface and curve extraction from 3D data.
+However, our situation is substantially different as %extracting the surface is not our ultimate goal. 
+we want to model the deformation of the structure. 
+%In that regard the curvature of the surface has a physical meaning: it represents the rest shape of our anatomical structure, that is the configuration with zero energy. 
+%\CD{Non !!! Rien ne te dit que l'objet ne te dit que l'objet n'est pas déformé au moment où tu prend l'image !!!}
+In that regard the curvature of the surface has a physical meaning: it represents the neutral fiber surface of the shell modelling on which the plane stress hypothesis~\cite{Liu03} applies. 
+Therefore we propose to match the shape (polynomial) functions used in our shell FEM formulation to describe the surface of the shells with the surface of the anatomical structures that we wish to simulate.
 
-Our contribution is to unify these two approaches (shell based modelling and geometric surface reconstruction) separately used in mechanical engineering and computer vision in a framework for realistic simulation of thin organ deformations.
-Our method, presented in section~\ref{sec:model}, is based on shell theory and computes the internal forces using Finite Element Modelling (FEM) on curved triangles.
-The design of the elements relies on a polynomial interpolation function that is also used to capture and process the interactions with other models, including solid models.
-In section~\ref{sec:mesh} we also present an automatic process to obtain meshes from image based reconstruction that relies on the exact same polynomial function.
+One originality of this work is that the polynomial shape functions, defined on  curved triangles, are used in three different ways to optimize the computation and reach real-time computations.
+The shell model, presented in section~\ref{sec:model}, uses them to computes the internal forces using a Finite Element Modelling (FEM).
+The interactions with other models, including deformable solid models, are captured and processed using the same polynomial interpolations.
+Moreover, in section~\ref{sec:mesh} we also present an automatic process to obtain meshes from image based reconstruction that relies on the exact same polynomial functions.
 % in an attempt to optimise the computation using fewer elements. 
@@
 Once combined with an in-plane membrane formulation we obtain an accurate, yet computationally efficient, shell finite element method featuring both membrane and bending energies.
+In the following, we detail the bending stiffness computation in order to present the polynomial shape functions that are used in the shell model.
 
-\subsection{Triangular elastic membrane}
+%\paragraph{Triangular elastic membrane}
 
-The element stiffness matrix $\textbf{K}_e$ can be computed as follows:
+%The element stiffness matrix $\textbf{K}_e$ can be computed as follows:
 
-\begin{equation}
-\textbf{K}_e = \int_v \textbf{J} \boldsymbol{\chi} \textbf{J}^{T} dV
-\end{equation}
-where $\textbf{J}$ is the strain-displacement matrix and $\boldsymbol\chi$ embodies the material's behaviour.
+%\begin{equation}
+%\textbf{K}_e = \int_v \textbf{J} \boldsymbol{\chi} \textbf{J}^{T} dV
+%\end{equation}
+%where $\textbf{J}$ is the strain-displacement matrix and $\boldsymbol\chi$ embodies the material's behaviour.
 The stiffness matrix in the global frame is eventually obtained using the rotation matrix of the element: $\textbf{K} = \textbf{R}^{T} \textbf{K}_e \textbf{R} $ where $\textbf{R}$ describes the rotation of the (triangular) element with respect to its initial configuration.
 
-\subsection{Triangular plate bending}
-
+\paragraph{Polynomial shape function}
+%
 \begin{figure}
+\vspace{-1.0cm}
 \centering
 \includegraphics[height=4cm]{images/bending}
-\caption {The different degrees of freedom $u$ of a triangular thin plate in bending are illustrated above.}
+\caption {The different degrees of freedom $u$ of a triangular thin plate in bending.}
 \label{fig-triangle}
+\vspace{-0.5cm}
 \end{figure}
-
-To calculate the stiffness matrix for the transverse deflections and rotations shown on figure \ref{fig-triangle}, we start with the assumed deflection $u_z$ of the form
+%
+To calculate the stiffness matrix for the transverse deflections and rotations shown on figure \ref{fig-triangle}, the deflection $u_z$ is computed using a polynomial interpolation:
 \begin{equation}
@@
 We then can derive a matrix $\textbf{C}$ such as $\textbf{u} = \textbf{Cc}$ where $\textbf{c} = \left\{c_1 c_2 \ldots c_9 \right\} $.
-
 We can calculate the strains from the flat-plate theory using:
@@
 \end{equation}
+
 Symbolically this may be expressed as $\textbf{e} = \textbf{Dc}$ where $\textbf{D}$ derives from equations~\ref{eq-deflection} and~\ref{eq-deformation}.
-Noting that $\textbf{c} = \textbf{C}^{-1}\textbf{u}$, we have:
-\begin{equation}
-\textbf{e} = \textbf{DC}^{-1}\textbf{u} = \textbf{bu}
-\end{equation} 
-where the strain-displacement matrix $\textbf{b} = \textbf{DC}^{-1}$. 
+Noting that $\textbf{c} = \textbf{C}^{-1}\textbf{u}$, we have $\textbf{e} = \textbf{DC}^{-1}\textbf{u} = \textbf{bu}$ where the strain-displacement matrix $\textbf{b} = \textbf{DC}^{-1}$.
 The stiffness matrix $\textbf{K}_e$ for an element is then obtained from:
@@
 \end{equation} 
-where $\boldsymbol\chi$ is the material matrix. The stiffness matrix in the global frame is eventually obtained using the rotation matrix of the element as in the membrane case.
+where $\boldsymbol\chi$ is the material matrix. The stiffness matrix in the global frame is eventually obtained using the rotation matrix of the element. %as in the membrane case.
 
-\subsection{Mechanical interactions with the curved surface of shells}
+\paragraph{Mechanical interactions with the curved surface of shells}
 \label{sec:interactions}
-But the practical interest of modelling complex behaviours such as bending and twisting would remain fairly low for medical simulation if contacts and constraints were not handled properly.
+The practical interest of modelling complex behaviours such as bending and twisting would remain fairly low for medical simulation if contacts and constraints were not handled properly.
 In our case the difficulty comes from different sources.
@@
 We first sample the flat surface of each element by recursively dividing each triangle into four smaller ones and the deflection of each new vertex is computed using (\ref{eq-deflection}) and the displacements and slopes at the three vertices of the triangular element.
-This process of subdivision allows us to render each shell as a curved triangle (Fig.~\ref{fig-shell} ) and detect any collision with the curved surface of the shell using any of the classic collision detection algorithms working on flat triangles.
+This process of subdivision allows us to render each shell as a curved triangle (Fig.~\ref{fig-shell} (a) and (b)) and detect any collision with the curved surface of the shell using any of the classic collision detection algorithms working on flat triangles.
 \begin{figure}
 \centering
-\includegraphics[height=4cm]{images/shell_curvature2.png}
-\caption {The triangle formed by the three vertices of the shell has been recursively subdivided 3 times and the deflection of each new vertex was computed according to the same deflection function used in our shell finite element formulation.
-Sampling the actual surface of the shell allows rendering and more accurate collision detection.}
+\vspace{-0.5cm}
+\includegraphics[height=3.5cm]{images/shell_curvature2.png}
+\vspace{-0.3cm}
+\caption {(a)  The triangle formed by the three vertices of the shell has been recursively subdivided 3 times and the deflection of each new vertex was computed according to the same deflection function used in our shell finite element formulation.
+(b) Sampling the actual surface of the shell allows rendering and more accurate collision detection.
+(c) The shape function is used to distribute an external force $F$ on the triangle nodes}
+
 \label{fig-shell}
+\vspace{-0.4cm}
 \end{figure}
@@
 
-The main difficulty is to convert the normal component of the force applied to the bent surface into a torque at each of the three nodes.
+The main difficulty is to convert the normal component of the force applied to the bent surface into a torque at each of the three nodes (Fig.~\ref{fig-shell} (c)).
 Our approach is the following: during force computation, we use the angle measured at each node to compute the local deflection of each subvertex within the triangle.
@@
 \label{sec:mesh}
-%\subsection{Related work}
+%\paragraph{Related work}
 
@@
 
-Because the surface of an anatomical structure has a physical meaning, we cannot use any type of function to reconstruct this surface.
-The surface must be patched with triangular elements whose interpolation makes use of the same shape function designed for our shell FEM formulation.
+Because the surface of an anatomical structure has a physical meaning, %we cannot use any type of function to reconstruct this surface. The surface must be patched 
+we propose to patch the surface with triangular elements whose interpolation makes use of the same shape function designed for our shell FEM formulation.
 By suggesting a method to reconstruct the surface of any structures with curved shell, the meshing process is optimised.
@@
 A simple example is shown Fig.~\ref{fig-cylinder} to illustrate the method.
-We can notice that differences with the target mesh have been subtantially reduced.
+%We can notice that differences with the target mesh have been subtantially reduced.
+% Christian: je ne comprend pas cette phrase.. du coup, je l'enlève !
 
@@
 \label{sec:results}
-\subsection{Meshing of anatomical structures}
+\paragraph{Meshing of anatomical structures}
 
@@
 \includegraphics[height=3cm]{images/resultsAneurysm}
-\caption {(a) the targeteted high resolution aneurysm mesh (28,368 triangles). (b): the one-sided Hausdorff distance error map on a mesh of 1,160 shells generated with our method}.
+\caption {(a) the targeteted high resolution aneurysm mesh (28,368 triangles).
+(b): the one-sided Hausdorff distance error map on a mesh of 1,160 shells generated with our method \CD{Montre plutot celui qui part de 200 et fait environ 800 triangles a la fin !!!}}.
 \label{fig-aneurysm}
@@
 
-\subsection{Computation times}
+\paragraph{Computation times}
 
-\subsection{Coupling between tetrahedra and shells for advanced modelling}
+We perform several tests on the aneurysm model at different resolution. 
+The shells are resisting to a uniform pressure load and are solved using an Conjugate Gradient (CG) iterative solver.
+The computation times are reported in the Fig.\ref{fig-computation}. 
+Implicit integration allows for large time steps (40ms) and the computation is real-time for 800 shell elements and a reasonable error criterion ($5\%$).
+When the computation time must be bounded (critical real-time applications), one can fix the number of CG iterations to, for instance, $100$ and remain real-time for 1000 shell elements.
+However, in that case, the accuracy of the results are not checked.
+
+\begin{figure}
+\centering
+\includegraphics[height=3cm]{images/computation_time.png}
+\caption {Computation time on meshes of the Aneurysm model with 200, 400, 600, 800 and 1000 elements.}.
+\label{fig-computation}
+\end{figure}
+
+\paragraph{Coupling between tetrahedra and shells for advanced modelling}
 
@@
 \centering
-\includegraphics[height=3.8cm]{images/angioplasty.png}
+\includegraphics[height=3.8cm]{images/stenose_final.png}
 \caption {Simulation of an angioplasty procedure.

### 2010-03-11 MICCAI2010/paper.tex
modifs intro avec Stephane
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@106 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 We propose to rely on a similar approach but with more accuracy in order to be applicable to medical simulation.
-Our model is not based on modal analysis but uses a co-rotational formulation and polynomial shape functions presented in~\cite{Comas2010ISBMS}.
+Our model is not based on modal analysis but uses a co-rotational formulation, triangular elements with 18 degrees of freedom, and polynomial shape functions presented in~\cite{Comas2010ISBMS}.
 
-In this paper, we propose for the first time to apply this technique for modelling the deformations of some anatomical structures.
+To model the deformation of complex anatomical structures using shell elements, the first step is to describe its surface with curved patches.
+%In this paper, we propose for the first time to apply this technique for modelling the deformations of some anatomical structures.
 The first step is to geometrically model the surface of the structures using curved triangular shell elements.
@@
 In that regard the curvature of the surface has a physical meaning: it represents the neutral fiber surface of the shell modelling on which the plane stress hypothesis~\cite{Liu03} applies.
-Therefore we propose to match the shape (polynomial) functions used in our shell FEM formulation to describe the surface of the shells with the surface of the anatomical structures that we wish to simulate.
 
-One originality of this work is that the polynomial shape functions, defined on  curved triangles, are used in three different ways to optimize the computation and reach real-time computations.
-The shell model, presented in section~\ref{sec:model}, uses them to computes the internal forces using a Finite Element Modelling (FEM).
-The interactions with other models, including deformable solid models, are captured and processed using the same polynomial interpolations.
-Moreover, in section~\ref{sec:mesh} we also present an automatic process to obtain meshes from image based reconstruction that relies on the exact same polynomial functions.
+In this paper, we propose to determine the shape (polynomial) functions used in our shell formulation to approximate the surface of anatomical structures with a surface composed of shell elements.
+These polynomial shape functions are used in three different ways in our computational model: (a) to approximate complex geometrical shapes, (b) to compute internal forces, (c) to compute contact forces  inside a curved triangle.
+Section~\ref{sec:model} presents the  Finite Element Modelling (FEM) of the shell elements
+and the process of contacts and interactions with other models.
+In section~\ref{sec:mesh} we present the automatic process to obtain meshes from image based reconstruction.
 % in an attempt to optimise the computation using fewer elements. 

### 2010-03-11 MICCAI2010/paper.tex
tard...
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@108 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\section{A mechanically-driven method to mesh curved surfaces with shells}
+\section{Physics-based reconstruction using shell elements}
+%A mechanically-driven method to mesh curved surfaces with shells}
 \label{sec:mesh}
@@
 
-The first step in the process of generating a shell-based mesh is an aggressive simplification of the targeted high resolution mesh by using the quadric edge collapse decimation technique implemented in Meshlab \cite{Meshlab}.
+The first step in the process of generating a shell-based mesh is a important decimation of the high resolution mesh, using quadric edge collapse technique implemented in  Meshlab \cite{Meshlab}.
+%aggressive simplification of the targeted high resolution mesh by using the quadric edge collapse decimation technique implemented in Meshlab \cite{Meshlab}. 
 The algorithm tries as much as possible to preserve mesh boundaries and generate high quality triangular elements.
@@
 For each node of the coarse mesh, we find the three closest triangles on the high resolution mesh and we move the node to the barycenter of the three centres of mass of those triangles.
-This technique locally smooths the surface of the mesh while converging towards the desired high resolution mesh.
+This technique locally smoothes the surface of the mesh while converging towards the desired high resolution mesh.
 At each iteration of this algorithm we measure the error between the curved surface of shells and the target using the Hausdorff distance and the process is stopped when the required precision has been reached.
@@
 % Christian: je ne comprend pas cette phrase.. du coup, je l'enlève !
-
-\begin{figure}
+%
+\begin{figure}[htb]
 \centering
@@
 \end{figure}
+%
 
+\vspace{-0.7cm}
 \section{Results}
+\vspace{-0.2cm}
 \label{sec:results}
 \paragraph{Meshing of anatomical structures}
-
-This approach has been applied to mesh more complex anatomical structures with curved shell elements. In each case the error is expressed as a percentage of the diagonal of the object's bounding box.
-
-\begin{figure}
+This approach has been applied to approximate more complex anatomical geometries with curved shell elements.
+In each case the error is expressed as a percentage of the diagonal of the object's bounding box. 
+\begin{figure}[h]
+\vspace{-0.8cm}
 \centering
 \includegraphics[height=3cm]{images/resultsLiver}
+\vspace{-0.2cm}
 \caption {Left: the targeteted high resolution Glisson's capsule mesh (8,000 triangles).
@@
 \end{figure}
-
-\begin{figure}
+%
+\begin{figure}[h]
+\vspace{-1.5cm}
 \centering
 \includegraphics[height=3cm]{images/resultsAneurysm}
+\vspace{-0.2cm}
 \caption {(a) the targeteted high resolution aneurysm mesh (28,368 triangles).
@@
 \label{fig-aneurysm}
+\vspace{-1.4cm}
 \end{figure}
-
+%
+%
 \paragraph{Computation times}
-
 We perform several tests on the aneurysm model at different resolution. 
@@
 When the computation time must be bounded (critical real-time applications), one can fix the number of CG iterations to, for instance, $100$ and remain real-time for 1000 shell elements.
-However, in that case, the accuracy of the results are not checked.
-
+However, in that case, the accuracy of the results is not checked.
+%
 \begin{figure}
+\vspace{-0.5cm}
 \centering
 \includegraphics[height=3cm]{images/computation_time.png}
+\vspace{-0.5cm}
 \caption {Computation time on meshes of the Aneurysm model with 200, 400, 600, 800 and 1000 elements.}.
+\vspace{-1.5cm}
 \label{fig-computation}
 \end{figure}
-
+%
+%
 \paragraph{Coupling between tetrahedra and shells for advanced modelling}
-
+%
 Structures in human body can be either solid (brain, liver, prostrate etc.) or hollow (blood vessels, colon, stomach etc.).
@@
 The forces are then distributed onto the mechanical nodes of the blood vessel mesh as detailled in section \ref{sec:interactions}, which widens the blood vessel as expected.
-
+%
 \begin{figure}
@@
 \caption {Simulation of an angioplasty procedure.
-(a) The collapsed balloon was inserted into the blood vessel simulated with our shell FEM formulation.
-The fatty deposits (in yellow) are modelled with a tetrahedral FEM.
-(b) Upon inflation the balloon is crushing the fatty deposits, which is applying a pressure onto the interior wall and widening the blood vessel.}
+(a, c): A collapsed stent is inserted into the blood vessel simulated with our shell FEM formulation.
+The fatty deposits (in yellow) are modelled with tetrahedral FEM.
+(b, d): The stent is crushing the fatty deposits which create a pressure onto the interior wall and widen the blood vessel.}
 \label{fig-stent}
 \end{figure}
-
+%
 \section{Conclusion}
-We proposed a complete framework for real-time modelling of thin anatomical structures.
+In this paper, we propose a framework for real-time modelling of thin anatomical structures.
 The novelty of our method relies on the combination of a shell finite element formulation and a geometric surface reconstruction both based on the same polynomial interpolation function used to describe the surface of shells.
-We also showed how contacts and interactions with the curved surfaces of shells could be handled using the same function.
-The efficiency of the whole approach was demonstrated through shell reconstruction and real-time modelling of a blood vessel showing signs of an aneurysm.
-Finally these results are very encouraging towards fine modelling of more complex situations like interactions existing between the liver and the Glisson's capsule or the simulation of an angioplasty procedure.
+We also show how contacts and interactions with the curved surfaces of shells can be handled using the same function.
+The efficiency of the proposed method is illustrated through shell-based reconstruction and real-time simulation of the deformations of various anatomical structures.
+We also present preliminary results on an application of our work through the simulation of an angioplasty procedure.
+%
 

### 2010-03-12 MICCAI2010/paper.tex
Oups
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@112 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 This paper presents a new modelling technique for the deformation of thin anatomical structures like membranes and hollow organs.
-We show that the behaviour of this type of tissue can be abstracted with a modelling of their surface elastic resistance using shell theory.
+We show that the behaviour of this type of surface tissue can be abstracted with a modelling of their elastic resistance using shell theory.
 In order to apply the shell theory in the particular context of medical simulation, our method propose to base the geometrical reconstruction of the organ on the shape functions of the shell element.
@@
 In all these cases the clinician needs fast update of the deformation model to obtain a real-time display of the computed deformations.
-To develop this kind of interactive systems, one faces the trade-off problem between accuracy of the model and the need for fast computations.
-Another key challenge of soft-tissue modelling is the variousness of the mechanical behaviours.
-It seems unrealistic to target a unique model for all tissues even if there are obvious advantages to make maximum use of optimised generic models.
-Yet most of previous work focus on volumetric models that are able to capture the behaviour of solid organs like the liver or the brain (see for instance \cite{Miller07} \cite{Delingette08}).
+%To develop this kind of interactive systems, one faces the trade-off problem between accuracy of the model and the need for fast computations. 
+Another key challenge of soft-tissue modelling is the variousness of the mechanical behaviours. It seems unrealistic to target a unique model for all tissues.
+% even if there are obvious advantages to make maximum use of optimised generic models. 
+Yet most of previous works focus on volumetric models that are able to capture the behaviour of solid organs like the liver or the brain (see for instance \cite{Miller07,Delingette08}).
 In contrast, this paper seeks to propose a solution for simulating, in real-time, the deformation of thin anatomical structures whose volume is negligible compared to their surface area.
-Examples of such anatomical structures include hollow structures, such as the wall of blood vessels, or membranes,  such the capsule of Glisson.
+Examples of such anatomical structures include hollow structures, such as the wall of blood vessels, or membranes, such as the Glisson's capsule surrounding the liver.
 
@@
 The key idea is to model the physical shell as a surface but endowed with mechanical properties in the form of elastic resistance to stretching and bending forces.
-Rather than resorting to shell theory, previous work on medical simulation often relies on linear or angular mass-spring models as in \cite{Nedel98,Hammer08}.
+Rather than resorting to shell theory, previous works in medical simulation often rely on linear or angular mass-spring models as in \cite{Nedel98,Hammer08}.
 Yet, such models are limited in their ability to describe certain behaviour, as they do not rely on continuum mechanics: it is difficult to derive spring stiffness (in particular for angular springs) from elastic properties (Young's modulus and Poisson's ratio).
-The use of fast shell based model for computer graphics and interactive simulation was recently introduced with the work of Choi et al.~\cite{Choi07}.
-The model relies on simplified energy functions and precomputed modal analysis.
-It allows for fast and visually realistic results.
-We propose to rely on a similar approach but with more accuracy in order to be applicable to medical simulation.
-Our model is not based on modal analysis but uses a co-rotational formulation, triangular elements with 18 degrees of freedom, and polynomial shape functions presented in~\cite{Comas2010ISBMS}.
+The use of fast shell-based modelling for interactive simulation was recently introduced in computer graphics with the work of Choi et al.~\cite{Choi07}.
+Their model relies on simplified energy functions and precomputed modal analysis for fast and visually realistic results.
+We propose to rely on a similar approach but with more accuracy to be applicable to medical simulation.
+Our model is not based on modal analysis but uses a co-rotational formulation and polynomial shape functions presented in~\cite{Comas10}.
+% , triangular elements with 18 degrees of freedom, => pas vraiment, on s'interesse seulement a x,y en translation dans le plan et la rotation autour de x et y pour le bending...
 
 To model the deformation of complex anatomical structures using shell elements, the first step is to describe its surface with curved patches.
-%In this paper, we propose for the first time to apply this technique for modelling the deformations of some anatomical structures.
-The first step is to geometrically model the surface of the structures using curved triangular shell elements.
 This process is quite similar to the reconstruction of the surface of objects in computer vision.
-Indeed calculating curvature maps of 3D surfaces represented as digitized data (point clouds or triangulated meshes) has been extensively studied.
+Indeed calculating curvature maps of 3D surfaces represented as digitised data (point clouds or triangulated meshes) has been extensively studied.
 One of the most common approach is to use continuous surface patches~\cite{Kolb95,Douros02}.
@@
 %In~\cite{Tang99} an elegant and non noise-sensitive approach that infers sign and direction of principal curvatures directly from the input is presented and the authors use this information for coherent surface and curve extraction from 3D data.
-However, our situation is substantially different as %extracting the surface is not our ultimate goal. 
-we want to model the deformation of the structure. 
-%In that regard the curvature of the surface has a physical meaning: it represents the rest shape of our anatomical structure, that is the configuration with zero energy. 
-%\CD{Non !!! Rien ne te dit que l'objet ne te dit que l'objet n'est pas déformé au moment où tu prend l'image !!!}
-In that regard the curvature of the surface has a physical meaning: it represents the neutral fiber surface of the shell modelling on which the plane stress hypothesis~\cite{Liu03} applies.
-
-In this paper, we propose to determine the shape (polynomial) functions used in our shell formulation to approximate the surface of anatomical structures with a surface composed of shell elements.
-These polynomial shape functions are used in three different ways in our computational model: (a) to approximate complex geometrical shapes, (b) to compute internal forces, (c) to compute contact forces  inside a curved triangle.
-Section~\ref{sec:model} presents the  Finite Element Modelling (FEM) of the shell elements
-and the process of contacts and interactions with other models.
-In section~\ref{sec:mesh} we present the automatic process to obtain meshes from image based reconstruction.
-% in an attempt to optimise the computation using fewer elements. 
-Finally the benefits of our approach (meshing of a curved surface, fast computation and possible interactions with solid models) are illustrated using various examples presented in section~\ref{sec:results}.
+However, our situation is substantially different as we want to model the deformation of the structure.
+In that regard the curvature of the surface has a physical meaning: it represents the mid-surface of the shell on which the plane stress hypothesis~\cite{Liu03} applies.
+We propose to approximate the surface of anatomical structures with shell elements whose each surface is described by the shape function used in our shell formulation.
+%We propose to determine the shape (polynomial) functions used in our shell formulation to approximate the surface of anatomical structures with a surface composed of shell elements.
+=> pas vraiment...
+pour un maillage donne elles sont fixees pour un triangle par les normales de chaque sommet.
+On joue sur la densite d'elements pour bien matcher la surface plutot...
+These polynomial shape functions are used in three different ways in our computational model: (a) to approximate complex geometrical shapes, (b) to compute internal forces, (c) to compute contact forces onto a curved triangle.
+Section~\ref{sec:model} presents our Finite Element Modelling (FEM) for shell elements and how we process contacts and interactions with other models.
+In section~\ref{sec:mesh} we introduce an automatic process to obtain meshes from image based reconstruction.
+Finally the benefits of our approach (meshing of a curved surface, fast computation and possible interactions with solid models) are illustrated using various examples showed in section~\ref{sec:results}.
 
@@
 \label{sec:model}
-
-A complete description and validation of our co-rotational triangular shell finite element model is available in one of our previous publication \cite{Comas2010ISBMS}.
+%
+A complete description and validation of our co-rotational triangular shell finite element model is available in one of our previous publication \cite{Comas10}.
 Therefore we will only remind briefly the key points.
@@
 Once combined with an in-plane membrane formulation we obtain an accurate, yet computationally efficient, shell finite element method featuring both membrane and bending energies.
-In the following, we detail the bending stiffness computation in order to present the polynomial shape functions that are used in the shell model.
+In the following we detail the bending stiffness computation in order to present the polynomial shape functions that are used in the shell model.
 
@@
 
-\paragraph{Polynomial shape function}
+\subsubsection{Polynomial shape function}
 %
 \begin{figure}
-\vspace{-1.0cm}
+\vspace{-1.3cm}
 \centering
-\includegraphics[height=4cm]{images/bending}
+\includegraphics[height=3.8cm]{images/bending}
+\vspace{-0.2cm}
 \caption {The different degrees of freedom $u$ of a triangular thin plate in bending.}
@@
 %
-To calculate the stiffness matrix for the transverse deflections and rotations shown on figure \ref{fig-triangle}, the deflection $u_z$ is computed using a polynomial interpolation:
+To calculate the stiffness matrix for the transverse deflections and rotations shown on Fig.~\ref{fig-triangle}, the deflection $u_z$ is computed using a polynomial interpolation:
 \begin{equation}
@@
 \end{equation} 
-and so on for the two other vertices.
-We then can derive a matrix $\textbf{C}$ such as $\textbf{u} = \textbf{Cc}$ where $\textbf{c} = \left\{c_1 c_2 \ldots c_9 \right\} $.
-We can calculate the strains from the flat-plate theory using:
+and so on for the two other vertices and we can derive a matrix $\textbf{C}$ such as $\textbf{u} = \textbf{Cc}$ where $\textbf{c} = \left\{c_1 c_2 \ldots c_9 \right\} $.
+We can then calculate the strains from the flat-plate theory using:
 \begin{equation}
@@
 \end{equation} 
-
-Symbolically this may be expressed as $\textbf{e} = \textbf{Dc}$ where $\textbf{D}$ derives from equations~\ref{eq-deflection} and~\ref{eq-deformation}.
+Symbolically this may be expressed as $\textbf{e} = \textbf{Dc}$ where $\textbf{D}$ derives from (\ref{eq-deflection}) and (\ref{eq-deformation}).
 Noting that $\textbf{c} = \textbf{C}^{-1}\textbf{u}$, we have $\textbf{e} = \textbf{DC}^{-1}\textbf{u} = \textbf{bu}$ where the strain-displacement matrix $\textbf{b} = \textbf{DC}^{-1}$.
@@
 \begin{equation}
-\textbf{K}_e = \int_v \textbf{b}^{T} \boldsymbol\chi \textbf{b} dV
+\textbf{K}_e = \int_v \textbf{b}^{T} \boldsymbol\chi \textbf{b} dV \quad
+\mbox{where } \boldsymbol\chi \mbox{ is the material matrix} \enspace .
 \end{equation} 
-where $\boldsymbol\chi$ is the material matrix. The stiffness matrix in the global frame is eventually obtained using the rotation matrix of the element. %as in the membrane case.
+%The stiffness matrix in the global frame is eventually obtained using the rotation matrix of the element.
 
-\paragraph{Mechanical interactions with the curved surface of shells}
+\subsubsection{Mechanical interactions with the curved surface of shells}
 \label{sec:interactions}
@@
 Then forces applied to a given triangle need to be distributed between linear forces and torques onto its three vertices.
-As we will see, the same polynomial interpolation function (\ref{eq-deflection}) chosen to compute the bending energy in our FEM formulation is also used to capture the interactions between the curved surface and other objects.
+As we will see, the same polynomial interpolation function chosen to compute the bending energy in our FEM formulation is also used to capture the interactions between the curved surface and other objects.
 
@@
 Therefore a finer mesh is created by adding vertices on the surface of each element.
-We first sample the flat surface of each element by recursively dividing each triangle into four smaller ones and the deflection of each new vertex is computed using (\ref{eq-deflection}) and the displacements and slopes at the three vertices of the triangular element.
+We first sample the flat surface of each element by recursively dividing each triangle into four smaller ones and the deflection of each new vertex is computed using (\ref{eq-deflection}) according to the displacements and slopes at the three vertices of the triangular element.
 This process of subdivision allows us to render each shell as a curved triangle (Fig.~\ref{fig-shell} (a) and (b)) and detect any collision with the curved surface of the shell using any of the classic collision detection algorithms working on flat triangles.
@@
 \vspace{-0.5cm}
-\includegraphics[height=3.5cm]{images/shell_curvature2.png}
-\vspace{-0.3cm}
-\caption {(a)  The triangle formed by the three vertices of the shell has been recursively subdivided 3 times and the deflection of each new vertex was computed according to the same deflection function used in our shell finite element formulation.
+\includegraphics[height=3.3cm]{images/shell_curvature.pdf}
+\vspace{-0.4cm}
+\caption {(a)  The triangle formed by the three vertices of the shell has been recursively subdivided 3 times and the deflection of each new vertex was computed according to the same deflection function used in our shell FEM.
 (b) Sampling the actual surface of the shell allows rendering and more accurate collision detection.
-(c) The shape function is used to distribute an external force $F$ on the triangle nodes}
-
+(c) The shape function is used to distribute an external force $F$ onto the triangle nodes.}
 \label{fig-shell}
@@
 The main difficulty is to convert the normal component of the force applied to the bent surface into a torque at each of the three nodes (Fig.~\ref{fig-shell} (c)).
-Our approach is the following: during force computation, we use the angle measured at each node to compute the local deflection of each subvertex within the triangle.
+Our approach is the following: during force computation, we use the change in orientation measured at each node to compute the local deflection of each subvertex within the triangle.
 Differentiating the formulation twice yields a relation between the torque applied at each node and the generated force in bending.
@@
 The moments at each vertex are then obtained with $\Omega = (\textbf{C}^{-1})^T P$.
-Thus we are able to transmit any force coming from elaborate interactions with the curved surface of shells to the mechanical vertices used in our FEM formulation.
+Thus we are able to transmit any force coming from interactions with the curved surface of shells to the mechanical vertices used in our FEM formulation.
 
 \section{Physics-based reconstruction using shell elements}
-%A mechanically-driven method to mesh curved surfaces with shells}
 \label{sec:mesh}
+
 %\paragraph{Related work}
-
 %Another key component of finite element methods is the mesh of the simulated object being used.
@@
 
-Because the surface of an anatomical structure has a physical meaning, %we cannot use any type of function to reconstruct this surface. The surface must be patched 
-we propose to patch the surface with triangular elements whose interpolation makes use of the same shape function designed for our shell FEM formulation.
+Because the surface of an anatomical structure has a physical meaning, we propose to patch the surface with triangular elements whose interpolation makes use of the same shape function designed for our shell FEM formulation.
 By suggesting a method to reconstruct the surface of any structures with curved shell, the meshing process is optimised.
-Indeed while many flat triangles are often required to describe highly curved surfaces, fewer triangular shell elements are needed to describe the given geometry with the same precision since they can be naturally curved.
+Indeed while many flat triangles are required to describe highly curved surfaces, fewer triangular shell elements are needed to describe the given geometry with the same precision since they can be curved.
 In the following we assume that we have a high resolution triangular mesh obtained from a binary segmented image of the organ we want to simulate (via a Marching Cube algorithm for instance).
@@
 The first step in the process of generating a shell-based mesh is a important decimation of the high resolution mesh, using quadric edge collapse technique implemented in Meshlab \cite{Meshlab}.
-%aggressive simplification of the targeted high resolution mesh by using the quadric edge collapse decimation technique implemented in Meshlab \cite{Meshlab}. 
-The algorithm tries as much as possible to preserve mesh boundaries and generate high quality triangular elements.
+The algorithm tries as much as possible to preserve mesh boundaries and generates high quality triangular elements.
 We then apply a heuristic method derived from the work of Saupin et.
-al \cite{Saupin07} with tetrahedral meshes based on very simple geometrical rules.
+al \cite{Saupin07} with tetrahedral meshes based on simple geometrical rules.
 For each node of the coarse mesh, we find the three closest triangles on the high resolution mesh and we move the node to the barycenter of the three centres of mass of those triangles.
@@
 A simple example is shown Fig.~\ref{fig-cylinder} to illustrate the method.
-%We can notice that differences with the target mesh have been subtantially reduced.
-% Christian: je ne comprend pas cette phrase.. du coup, je l'enlève !
 %
@@
 \includegraphics[height=2.5cm]{images/exampleCylinder}
-\caption {Illustration of our method on a simple example.
-The target is a high resolution cylinder mesh of 16,384 triangles (a) and we start from a very coarse mesh (12 triangles) approximating the shape of a cylinder, rendered with flat triangles here (b).
+\caption {The target (a) is a high resolution cylinder mesh of 16,384 triangles and we start from a very coarse mesh (12 triangles) approximating the shape of a cylinder, rendered with flat triangles here (b).
 In (c) the same coarse mesh is rendered with shells and a one-sided Hausdorff distance colour map is applied to show the initial error with the high resolution mesh.
@@
 %
+\section{Results}
+\label{sec:results}
 
-\vspace{-0.7cm}
-\section{Results}
-\vspace{-0.2cm}
-\label{sec:results}
-\paragraph{Meshing of anatomical structures}
+\subsubsection{Meshing of anatomical structures}
+%
 This approach has been applied to approximate more complex anatomical geometries with curved shell elements.
@@
 \begin{figure}[h]
-\vspace{-0.8cm}
 \centering
-\includegraphics[height=3cm]{images/resultsLiver}
-\vspace{-0.2cm}
-\caption {Left: the targeteted high resolution Glisson's capsule mesh (8,000 triangles).
-Right: the one-sided Hausdorff distance error map after applying only one iteration of our algorithm to the coarse mesh (1,200 shells)}.
+\includegraphics[height=3.2cm]{images/resultsLiver}
+\caption {(a) the targeteted high resolution Glisson's capsule mesh (8,000 triangles).
+(b) the one-sided Hausdorff distance error map after applying only one iteration of our algorithm to the coarse mesh (1,200 shells).}
 \label{fig-liver}
+\vspace{-0.5cm}
 \end{figure}
@@
 \begin{figure}[h]
-\vspace{-1.5cm}
 \centering
-\includegraphics[height=3cm]{images/resultsAneurysm}
-\vspace{-0.2cm}
-\caption {(a) the targeteted high resolution aneurysm mesh (28,368 triangles).
-(b): the one-sided Hausdorff distance error map on a mesh of 1,160 shells generated with our method \CD{Montre plutot celui qui part de 200 et fait environ 800 triangles a la fin !!!}}.
+\includegraphics[height=3.2cm]{images/resultsAneurysm}
+\caption {(a) the targeteted high resolution aneurysm mesh (28,368 triangles). (b): the one-sided Hausdorff distance error map on a mesh of 772 shells generated with our method.}
 \label{fig-aneurysm}
-\vspace{-1.4cm}
 \end{figure}
 %
+\subsubsection{Computation times}
 %
-\paragraph{Computation times}
-We perform several tests on the aneurysm model at different resolution. 
-The shells are resisting to a uniform pressure load and are solved using an Conjugate Gradient (CG) iterative solver.
-The computation times are reported in the Fig.\ref{fig-computation}. 
+We perform several tests on the aneurysm model at different resolutions. 
+The shells are resisting to a uniform pressure load and are solved using a Conjugate Gradient (CG) iterative solver.
+The computation times are reported in Fig.~\ref{fig-computation}. 
 Implicit integration allows for large time steps (40ms) and the computation is real-time for 800 shell elements and a reasonable error criterion ($5\%$).
-When the computation time must be bounded (critical real-time applications), one can fix the number of CG iterations to, for instance, $100$ and remain real-time for 1000 shell elements.
-However, in that case, the accuracy of the results is not checked.
+When the computation time must be bounded (critical real-time applications), one can fix the number of CG iterations to, for instance, $100$ and remains real-time for 1000 shell elements.
+However, in that case the accuracy of the results is not checked.
 %
 \begin{figure}
-\vspace{-0.5cm}
 \centering
-\includegraphics[height=3cm]{images/computation_time.png}
-\vspace{-0.5cm}
-\caption {Computation time on meshes of the Aneurysm model with 200, 400, 600, 800 and 1000 elements.}.
-\vspace{-1.5cm}
+\includegraphics[height=3.6cm]{images/computation_time.pdf}
+\vspace{-0.2cm}
+\caption {Computation time on meshes of the aneurysm model with 200, 400, 600, 800 and 1000 elements.}.
+\vspace{-0.6cm}
 \label{fig-computation}
@@
 %
-\paragraph{Coupling between tetrahedra and shells for advanced modelling}
+\subsubsection{Coupling between tetrahedra and shells for advanced modelling}
 %
-Structures in human body can be either solid (brain, liver, prostrate etc.) or hollow (blood vessels, colon, stomach etc.).
+Structures in human body can be either solid (brain, liver, prostate etc.) or hollow (blood vessels, colon, stomach etc.).
 However knowing how to model the two kind of structures is not sufficient to reach a high degree of accuracy in medical simulation.
-Indeed real life situations are more complex.
-As an example, the external surface of the liver is covered by a single layer of cells called Glisson's capsule.
+Real life situations are more complex.
+As an example, the external surface of the liver is covered by a layer of cells called Glisson's capsule.
 Its interaction with the liver plays an important role into the overall structure's mechanical behaviour.
@@
 Angioplasty is the technique of mechanically widening a narrowed or obstructed blood vessel, typically as a result of atherosclerosis.
-An empty and collapsed balloon on a guide wire, known as a balloon catheter, is passed into the narrowed locations and then inflated to a fixed size using water pressures.
+An empty and collapsed balloon on a guide wire is passed into the narrowed locations and then inflated to a fixed size.
 The balloon crushes the fatty deposits, so opening up the blood vessel to improved flow.
@@
 The blood vessel is modelled using the shell FEM formulation described in this paper and the fatty deposits are simulated with a tetrahedral FEM method and are fixed to the interior wall of the blood vessel.
-When the balloon inflates it crushes the deposits and the deposits then apply a pressure onto the curved surfaces of shells modelling the interior wall.
-The forces are then distributed onto the mechanical nodes of the blood vessel mesh as detailled in section \ref{sec:interactions}, which widens the blood vessel as expected.
+When the balloon inflates it crushes the deposits and they then apply a pressure onto the curved surfaces of shells modelling the interior wall.
+The forces are then distributed onto the mechanical nodes of the blood vessel mesh as detailled in section~\ref{sec:interactions}, which widens the blood vessel as expected.
 %
@@
 \centering
-\includegraphics[height=3.8cm]{images/stenose_final.png}
+\includegraphics[height=4.3cm]{images/stenose_final.pdf}
 \caption {Simulation of an angioplasty procedure.
@@
 The fatty deposits (in yellow) are modelled with tetrahedral FEM.
-(b, d): The stent is crushing the fatty deposits which create a pressure onto the interior wall and widen the blood vessel.}
+(b, d): The stent is crushing the fatty deposits which creates a pressure onto the interior wall and widens the blood vessel.}
 \label{fig-stent}
@@
 %
+\vspace{-0.1cm}
 \section{Conclusion}
-In this paper, we propose a framework for real-time modelling of thin anatomical structures.
+\vspace{-0.1cm}
+We propose a framework for real-time modelling of thin anatomical structures.
 The novelty of our method relies on the combination of a shell finite element formulation and a geometric surface reconstruction both based on the same polynomial interpolation function used to describe the surface of shells.
 We also show how contacts and interactions with the curved surfaces of shells can be handled using the same function.
-The efficiency of the proposed method is illustrated through shell-based reconstruction and real-time simulation of the deformations of various anatomical structures.
-We also present preliminary results on an application of our work through the simulation of an angioplasty procedure.
+The efficiency of the method is illustrated through shell-based reconstruction and real-time simulation of the deformations of various anatomical structures.
+We also present preliminary results on the simulation of an angioplasty procedure.
 %
-
 %\section*{Acknowledgments}
-
+%
 \bibliographystyle{splncs}

### 2010-05-28 MICCAI2010/paper.tex
Update for camera-ready version (fits in 8 pages without cheating :p)
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@118 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \usepackage{amsmath}
-\usepackage[latin1]{inputenc} % pour les caracteres francais dans les remarques
+%\usepackage[latin1]{inputenc} % pour les caracteres francais dans les remarques
 
@@
 
-%\newcommand{\SC}[1]{}
-%\newcommand{\OC}[1]{}
-%\newcommand{\CD}[1]{}
-
 \newcommand{\nomore}[1]{}
@@
 \title{Shell Model for Reconstruction and Real-Time Simulation of Thin Anatomical Structures}
-%\author{Olivier Comas, St\'ephane Cotin and Christian Duriez}
-%\institute{INRIA, Alcove team, Lille, France}
+\author{Olivier Comas\inst{1,2}, Christian Duriez\inst{1} and St\'ephane Cotin\inst{1}}
+\institute{INRIA, Shaman team, Lille, France
+\and The Australian e-Health Research Centre, CSIRO, Brisbane, Australia}
 
@@
 We show that the behaviour of this type of surface tissue can be abstracted with a modelling of their elastic resistance using shell theory.
-In order to apply the shell theory in the particular context of medical simulation, our method propose to base the geometrical reconstruction of the organ on the shape functions of the shell element.
+In order to apply the shell theory in the context of medical simulation, our method propose to base the geometrical reconstruction of the organ on the shape functions of the shell element.
 Moreover, we also use these continuous shape functions to handle the contacts and the interactions with other types of deformable tissues.
@@
 The human body is composed of various deformable anatomical structures. 
-Realistic modelling of organ deformations is a challenging research field that opens the door to new clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedure and per-operative guidance based on simulation.
+%Realistic modelling of organ deformations is a challenging research field that opens the door to new clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedure and per-operative guidance based on simulation.
 In all these cases the clinician needs fast update of the deformation model to obtain a real-time display of the computed deformations.
-%To develop this kind of interactive systems, one faces the trade-off problem between accuracy of the model and the need for fast computations. 
-Another key challenge of soft-tissue modelling is the variousness of the mechanical behaviours. It seems unrealistic to target a unique model for all tissues.
-% even if there are obvious advantages to make maximum use of optimised generic models. 
+%Another 
+A key challenge of soft-tissue modelling is the variousness of the mechanical behaviours. It seems unrealistic to use a unique model for all tissues.
 Yet most of previous works focus on volumetric models that are able to capture the behaviour of solid organs like the liver or the brain (see for instance \cite{Miller07,Delingette08}).
 In contrast, this paper seeks to propose a solution for simulating, in real-time, the deformation of thin anatomical structures whose volume is negligible compared to their surface area.
-Examples of such anatomical structures include hollow structures, such as the wall of blood vessels, or membranes, such as the Glisson's capsule surrounding the liver.
+Examples include hollow structures, such as the wall of blood vessels, or membranes, such as the Glisson's capsule surrounding the liver.
+%of such anatomical structures
 
@@
 Yet, such models are limited in their ability to describe certain behaviour, as they do not rely on continuum mechanics: it is difficult to derive spring stiffness (in particular for angular springs) from elastic properties (Young's modulus and Poisson's ratio).
-The use of fast shell-based modelling for interactive simulation was recently introduced in computer graphics with the work of Choi et al.~\cite{Choi07}.
+The work of Choi et al.~\cite{Choi07} in computer graphics refers to a fast shell-based modelling for interactive simulation.
 Their model relies on simplified energy functions and precomputed modal analysis for fast and visually realistic results.
@@
 Our model is not based on modal analysis but uses a co-rotational formulation and polynomial shape functions presented in~\cite{Comas10}.
-% , triangular elements with 18 degrees of freedom, => pas vraiment, on s'interesse seulement a x,y en translation dans le plan et la rotation autour de x et y pour le bending...
 
@@
 Indeed calculating curvature maps of 3D surfaces represented as digitised data (point clouds or triangulated meshes) has been extensively studied.
-One of the most common approach is to use continuous surface patches~\cite{Kolb95,Douros02}.
+One of the most common approach is to use continuous surface patches~\cite{Kolb95}.
 The surface is locally approximated by an analytic representation called surface patch, usually chosen to be an implicit surface.
-These works target non noise-sensitive approaches and coherent surface and curve extraction from 3D data~\cite{Tang99}.
+These works target approaches that are not noise-sensitive and coherent surface and curve extraction from 3D data~\cite{Tang99}.
+% removed citation Douros02 which was with Kolb95.
 %In~\cite{Tang99} an elegant and non noise-sensitive approach that infers sign and direction of principal curvatures directly from the input is presented and the authors use this information for coherent surface and curve extraction from 3D data.
@@
 We propose to approximate the surface of anatomical structures with shell elements whose each surface is described by the shape function used in our shell formulation.
-%We propose to determine the shape (polynomial) functions used in our shell formulation to approximate the surface of anatomical structures with a surface composed of shell elements.
-=> pas vraiment...
-pour un maillage donne elles sont fixees pour un triangle par les normales de chaque sommet.
-On joue sur la densite d'elements pour bien matcher la surface plutot...
+
 These polynomial shape functions are used in three different ways in our computational model: (a) to approximate complex geometrical shapes, (b) to compute internal forces, (c) to compute contact forces onto a curved triangle.
@@
 Finally the benefits of our approach (meshing of a curved surface, fast computation and possible interactions with solid models) are illustrated using various examples showed in section~\ref{sec:results}.
+Implementations were carried out within the open source framework SOFA \cite{SOFA}.
 
@@
 %
-A complete description and validation of our co-rotational triangular shell finite element model is available in one of our previous publication \cite{Comas10}.
-Therefore we will only remind briefly the key points.
+A complete description and validation of our co-rotational triangular shell finite element model is available in one of our previous publication~\cite{Comas10}. 
+%one of our previous publication 
+Therefore we will only remind the key points.
 We improved and extended a plate model first introduced by Przemieniecki \cite{Przemieniecki68} to a co-rotational formulation.
@@
 
-%\paragraph{Triangular elastic membrane}
-
-%The element stiffness matrix $\textbf{K}_e$ can be computed as follows:
-
-%\begin{equation}
-%\textbf{K}_e = \int_v \textbf{J} \boldsymbol{\chi} \textbf{J}^{T} dV
-%\end{equation}
-%where $\textbf{J}$ is the strain-displacement matrix and $\boldsymbol\chi$ embodies the material's behaviour.
-The stiffness matrix in the global frame is eventually obtained using the rotation matrix of the element: $\textbf{K} = \textbf{R}^{T} \textbf{K}_e \textbf{R} $ where $\textbf{R}$ describes the rotation of the (triangular) element with respect to its initial configuration.
-
 \subsubsection{Polynomial shape function}
@@
 \begin{figure}
-\vspace{-1.3cm}
 \centering
-\includegraphics[height=3.8cm]{images/bending}
-\vspace{-0.2cm}
+\includegraphics[height=3.2cm]{images/bending}
 \caption {The different degrees of freedom $u$ of a triangular thin plate in bending.}
 \label{fig-triangle}
-\vspace{-0.5cm}
 \end{figure}
@@
 In order to detect the collision with the bent surface, we have chosen the subdivision approach. 
-Therefore a finer mesh is created by adding vertices on the surface of each element.
+% Therefore a finer mesh is created by adding vertices on the surface of each element. 
 We first sample the flat surface of each element by recursively dividing each triangle into four smaller ones and the deflection of each new vertex is computed using (\ref{eq-deflection}) according to the displacements and slopes at the three vertices of the triangular element.
 This process of subdivision allows us to render each shell as a curved triangle (Fig.~\ref{fig-shell} (a) and (b)) and detect any collision with the curved surface of the shell using any of the classic collision detection algorithms working on flat triangles.
-\begin{figure}
-\centering
-\vspace{-0.5cm}
-\includegraphics[height=3.3cm]{images/shell_curvature.pdf}
-\vspace{-0.4cm}
-\caption {(a)  The triangle formed by the three vertices of the shell has been recursively subdivided 3 times and the deflection of each new vertex was computed according to the same deflection function used in our shell FEM.
-(b) Sampling the actual surface of the shell allows rendering and more accurate collision detection.
-(c) The shape function is used to distribute an external force $F$ onto the triangle nodes.}
-\label{fig-shell}
-\vspace{-0.4cm}
-\end{figure}
 
@@
 Thus we are able to transmit any force coming from interactions with the curved surface of shells to the mechanical vertices used in our FEM formulation.
+
+\begin{figure}
+\centering
+\includegraphics[height=3.6cm]{images/shell_curvature.pdf}
+\caption {(a,b)  The triangle formed by the three vertices of the shell has been recursively subdivided 3 times to allow more accurate rendering and collision detection.
+(c) The shape function is used to distribute an external force $F$ onto the triangle nodes.}
+\label{fig-shell}
+\end{figure}
+% \caption {(a)  The triangle formed by the three vertices of the shell has been recursively subdivided 3 times and the deflection of each new vertex was computed according to the same deflection function used in our shell FEM.
+(b) Sampling the actual surface of the shell allows more accurate rendering and collision detection.
+(c) The shape function is used to distribute an external force $F$ onto the triangle nodes.}
 
@@
 Because the surface of an anatomical structure has a physical meaning, we propose to patch the surface with triangular elements whose interpolation makes use of the same shape function designed for our shell FEM formulation.
-By suggesting a method to reconstruct the surface of any structures with curved shell, the meshing process is optimised.
-Indeed while many flat triangles are required to describe highly curved surfaces, fewer triangular shell elements are needed to describe the given geometry with the same precision since they can be curved.
+% By suggesting a method to reconstruct the surface of any structures with curved shell, the meshing process is optimised. Indeed 
+Moreover, while many flat triangles are required to describe highly curved surfaces, fewer triangular shell elements are needed to describe the given geometry with the same precision since they can be curved.
 In the following we assume that we have a high resolution triangular mesh obtained from a binary segmented image of the organ we want to simulate (via a Marching Cube algorithm for instance).
@@
 
-Therefore we need to insure that the distance between the surface of our shell-based mesh and the targeted high resolution mesh will be minimal.
-An efficient technique for measuring the error between two surfaces is the Hausdorff distance \cite{Klein96,Cignoni98}.
+Therefore we need to ensure that the distance between the surface of our shell-based mesh and the targeted high resolution mesh will be minimal.
+An efficient technique for measuring the error between two surfaces is the Hausdorff distance \cite{Klein96}.
 As a reminder the Hausdorff distance between two meshes is the maximum between the two so-called one-sided Hausdorff distances:
+% removed citration to Cignoni98
 \begin{equation}
@@
 
-The first step in the process of generating a shell-based mesh is a important decimation of the high resolution mesh, using quadric edge collapse technique implemented in Meshlab \cite{Meshlab}.
+The first step in the process of generating a shell-based mesh is an important decimation of the high resolution mesh, using quadric edge collapse technique implemented in Meshlab \cite{Meshlab}.
 The algorithm tries as much as possible to preserve mesh boundaries and generates high quality triangular elements.
@@
 A simple example is shown Fig.~\ref{fig-cylinder} to illustrate the method.
-%
-\begin{figure}[htb]
+\begin{figure}
 \centering
 \includegraphics[height=2.5cm]{images/exampleCylinder}
-\caption {The target (a) is a high resolution cylinder mesh of 16,384 triangles and we start from a very coarse mesh (12 triangles) approximating the shape of a cylinder, rendered with flat triangles here (b).
-In (c) the same coarse mesh is rendered with shells and a one-sided Hausdorff distance colour map is applied to show the initial error with the high resolution mesh.
-(d) One-sided Hausdorff distance colour map after only one iteration of our algorithm (48 shells).}
+\caption {The target (a) is a high resolution cylinder mesh of 16,384 triangles and we start from a very coarse mesh (12 triangles), rendered with flat triangles here (b).
+In (c) the coarse mesh is rendered with shells and a one-sided Hausdorff distance colour map is applied to show the initial error with the high resolution mesh.
+(d) One-sided Hausdorff distance colour map after one iteration of our algorithm (48 shells).}
 \label{fig-cylinder}
 \end{figure}
-%
+
 \section{Results}
@@
 
-\subsubsection{Meshing of anatomical structures}
+\subsubsection{Meshing of anatomical structures.}
 %
@@
 In each case the error is expressed as a percentage of the diagonal of the object's bounding box.
-\begin{figure}[h]
+\begin{figure}
 \centering
-\includegraphics[height=3.2cm]{images/resultsLiver}
+\includegraphics[height=3.3cm]{images/resultsLiver}
 \caption {(a) the targeteted high resolution Glisson's capsule mesh (8,000 triangles).
@@
 \label{fig-liver}
-\vspace{-0.5cm}
 \end{figure}
-%
-\begin{figure}[h]
+
+\begin{figure}
 \centering
-\includegraphics[height=3.2cm]{images/resultsAneurysm}
+\includegraphics[height=3.3cm]{images/resultsAneurysm}
 \caption {(a) the targeteted high resolution aneurysm mesh (28,368 triangles). (b): the one-sided Hausdorff distance error map on a mesh of 772 shells generated with our method.}
@@
 \end{figure}
+
+\subsubsection{Computation times.}
 %
-\subsubsection{Computation times}
-%
-We perform several tests on the aneurysm model at different resolutions. 
-The shells are resisting to a uniform pressure load and are solved using a Conjugate Gradient (CG) iterative solver.
-The computation times are reported in Fig.~\ref{fig-computation}. 
+We perform several tests on the aneurysm model at different resolutions to measure computation times (Fig.~\ref{fig-computation}).
+The shells are resisting to a uniform pressure load and solved using a Conjugate Gradient (CG) iterative solver.
+% The computation times are reported in Fig.~\ref{fig-computation}. 
 Implicit integration allows for large time steps (40ms) and the computation is real-time for 800 shell elements and a reasonable error criterion ($5\%$).
@@
 However, in that case the accuracy of the results is not checked.
-%
+
 \begin{figure}
 \centering
-\includegraphics[height=3.6cm]{images/computation_time.pdf}
-\vspace{-0.2cm}
-\caption {Computation time on meshes of the aneurysm model with 200, 400, 600, 800 and 1000 elements.}.
-\vspace{-0.6cm}
+\includegraphics[height=3.5cm]{images/computation_time.pdf}
+\caption {Computation time on meshes of 200, 400, 600, 800 and 1000 elements.}.
 \label{fig-computation}
 \end{figure}
+
+\subsubsection{Coupling between tetrahedra and shells for advanced modelling.}
 %
-%
-\subsubsection{Coupling between tetrahedra and shells for advanced modelling}
-%
-Structures in human body can be either solid (brain, liver, prostate etc.) or hollow (blood vessels, colon, stomach etc.).
-However knowing how to model the two kind of structures is not sufficient to reach a high degree of accuracy in medical simulation.
-Real life situations are more complex.
+%Structures in human body can be either solid (brain, liver, prostate etc.) or hollow (blood vessels, colon, stomach etc.). 
+Structures in human body can be either solid or hollow.
+However knowing how to model the two kind of structures is not sufficient to reach a high degree of accuracy, real life situations are more complex.
 As an example, the external surface of the liver is covered by a layer of cells called Glisson's capsule.
@@
 When the balloon inflates it crushes the deposits and they then apply a pressure onto the curved surfaces of shells modelling the interior wall.
-The forces are then distributed onto the mechanical nodes of the blood vessel mesh as detailled in section~\ref{sec:interactions}, which widens the blood vessel as expected.
-%
+The forces are then distributed onto the mechanical nodes of the blood vessel mesh as detailed in section~\ref{sec:interactions}, which widens the blood vessel as expected.
+
 \begin{figure}
 \centering
-\includegraphics[height=4.3cm]{images/stenose_final.pdf}
+\includegraphics[height=3.1cm]{images/stenose_final.pdf}
 \caption {Simulation of an angioplasty procedure.
+(a, c): A collapsed stent is inserted into the blood vessel.
+(b, d): The stent is crushing the fatty deposits which creates a pressure onto the interior wall and widens the blood vessel.}
+\label{fig-stent}
+\end{figure}
+%\caption {Simulation of an angioplasty procedure.
 (a, c): A collapsed stent is inserted into the blood vessel simulated with our shell FEM formulation.
@@
 (b, d): The stent is crushing the fatty deposits which creates a pressure onto the interior wall and widens the blood vessel.}
-\label{fig-stent}
-\end{figure}
-%
-\vspace{-0.1cm}
+
 \section{Conclusion}
-\vspace{-0.1cm}
 We propose a framework for real-time modelling of thin anatomical structures.
@@
 We also present preliminary results on the simulation of an angioplasty procedure.
-%
-%\section*{Acknowledgments}
-%
+
+\section*{Acknowledgments}
+We would like to thank INRIA and the SOFA project, as well as CSIRO Preventative Health National Research Flagship, ICTC, The Australian e-Health Research Centre for their fundings.
+
 \bibliographystyle{splncs}

### 2010-06-28 MICCAI2010/paper.tex
- Latest modifications for camera-ready MICCAI 2010 - fixed cataract scene (threshold angle to return a null quaternion was changed in the qDiffZ method) - scenes updated
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@121 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \institute{INRIA, Shaman team, Lille, France
-\and The Australian e-Health Research Centre, CSIRO, Brisbane, Australia}
+%\and Preventative Health National Research Flagship, CSIRO ICT, The Australian e-Health Research Centre, Brisbane, Australia}
+\and Preventative Health Flagship, CSIRO ICT, AEHRC, Brisbane, Australia}
 
@@
 Examples include hollow structures, such as the wall of blood vessels, or membranes, such as the Glisson's capsule surrounding the liver.
+It is also of particular interest to us for modelling the colon in our colonoscopy simulator \cite{deVisser10}.
 %of such anatomical structures
@@
 However, our situation is substantially different as we want to model the deformation of the structure.
-In that regard the curvature of the surface has a physical meaning: it represents the mid-surface of the shell on which the plane stress hypothesis~\cite{Liu03} applies.
+In that regard the curvature of the surface has a physical meaning: it represents the mid-surface of the shell.
+% on which the plane stress hypothesis~\cite{Liu03} applies. 
 We propose to approximate the surface of anatomical structures with shell elements whose each surface is described by the shape function used in our shell formulation.
@@
 %Structures in human body can be either solid (brain, liver, prostate etc.) or hollow (blood vessels, colon, stomach etc.). 
-Structures in human body can be either solid or hollow.
+Structures in human body can be either solid (brain, liver, prostate etc.) or hollow (colon, blood vessels, stomach etc.).
 However knowing how to model the two kind of structures is not sufficient to reach a high degree of accuracy, real life situations are more complex.
@@
 
-\section*{Acknowledgments}
-We would like to thank INRIA and the SOFA project, as well as CSIRO Preventative Health National Research Flagship, ICTC, The Australian e-Health Research Centre for their fundings.
+%\section*{Acknowledgments}
+%We would like to thank INRIA and the SOFA project, as well as CSIRO Preventative Health National Research Flagship, ICTC, The Australian e-Health Research Centre for their fundings.
 

### 2010-06-28 mapping/BendingPlateMechanicalMapping.h

@@ -31,6 +31,8 @@
 #include <sofa/helper/gl/GLSLShader.h>
 //#include <sofa/core/VisualModel.h>
 
+#include <sofa/component/topology/TriangleSetTopologyContainer.h>
+
 #include "../forcefield/TriangularBendingFEMForceField.h"
 #include <sofa/component/topology/TriangleSubdivisionTopologicalMapping.h>
 
@@ -110,10 +112,9 @@ protected:
 	BaseMeshTopology* outputTopo;
         
         Data<bool> measureError;
-//        Data<OutVecCoord> targetVertices;
-//        Data<SeqTriangles> targetTriangles;
+        Data<std::string> nameTargetTopology;
 
-        BaseMeshTopology* _topologyHigh;
+        TriangleSetTopologyContainer* topologyTarget;
         OutVecCoord verticesTarget;
         SeqTriangles trianglesTarget;
         

### 2010-07-09 Thesis/Abstract.tex
Added directory for PhD thesis
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@122 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\cleardoublepage

\begin{vcenterpage}
\noindent\rule[2pt]{\textwidth}{0.5pt}
\begin{center}
{\large\textbf{Real-time soft tissue modelling on GPU for medical simulation\\}}
\end{center}
{\large\textbf{Abstract:}}
Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla. Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.
\\
Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla. \\

Blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla. \\

Blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla.
\\
{\large\textbf{Keywords:}}
Medical simulation, soft tissue modelling, finite element method, GPU
\\
\noindent\rule[2pt]{\textwidth}{0.5pt}
\end{vcenterpage}

### 2010-07-09 Thesis/Appendix1.tex
Added directory for PhD thesis
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@122 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{Appendix Example}
\label{chap:appendix1}

\section{Appendix example section}

And I cite myself to show by bibtex style file (two authors) \citep{Commowick_MICCAI_2007}.

This for other bibtex stye file : only one author \cite{Oakes_RStat_1999} and many authors \cite{Guimond_CVIU_2000}.

### 2010-07-09 Thesis/Chapter1.tex
Added directory for PhD thesis
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@122 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{Introduction}
\label{chap:intro}
\begin{shortAbstract}
A short abstract for the upcoming chapter
\end{shortAbstract}

\section{Illustration Example}

\subsection{A subsection just for fun}

Sorry I won't write your PhD here ;) This small text just to mention that this style supports writing with accents such as in french words (thse, dfinir, ...). Also I put here a simple way to include an image. This is standard latex. For pdflatex compilation, the extension of the images is jpg. For latex compilation, this is ps or eps. The base folder containing images is set in formatAndDefs.tex, as well as the default extensions added to the image names.

\begin{figure}[!htbp]
  \begin{center}
    \includegraphics[width=0.9\textwidth]{Chapter1/arctic_control}
  \end{center}
  \caption{A nice image...}
  \label{fig:jolieImage}
\end{figure}

\section{An equation}

Just to show argmin and partial derivative commands.
First, we want to obtain a registration method which is as independent as possible w.r.t. the setting of its parameters. This setting, done by the clinician, indeed needs to be minimal while guaranteeing a robust result. We therefore propose registration methods allowing to better control the obtained transformation, using outlier rejection techniques or locally affine transformations.First, we want to obtain a registration method which is as independent as possible w.r.t. the setting of its parameters. This setting, done by the clinician, indeed needs to be minimal while guaranteeing a robust result. We therefore propose registration methods allowing to better control the obtained transformation, using outlier rejection techniques or locally affine transformations.First, we want to obtain a registration method which is as independent as possible w.r.t. the setting of its parameters. This setting, done by the clinician, indeed needs to be minimal while guaranteeing a robust result. We therefore propose registration methods allowing to better control the obtained transformation, using outlier rejection techniques or locally affine transformations.First, we want to obtain a registration method which is as independent as possible w.r.t. the setting of its parameters. This setting, done by the clinician, indeed needs to be minimal while guaranteeing a robust result. We therefore propose registration methods allowing to better control the obtained transformation, using outlier rejection techniques or locally affine transformations.
First, we want to obtain a registration method which is as independent as possible w.r.t. the setting of its parameters. This setting, done by the clinician, indeed needs to be minimal while guaranteeing a robust result. We therefore propose registration methods allowing to better control the obtained transformation, using outlier rejection techniques or locally affine transformations.First, we want to obtain a registration method which is as independent as possible w.r.t. the setting of its parameters. This setting, done by the clinician, indeed needs to be minimal while guaranteeing a robust result. We therefore propose registration methods allowing to better control the obtained transformation, using outlier rejection techniques or locally affine transformations.
Regularization:

\begin{equation}
  \pd{T}{t} = \Delta T
\end{equation}

\section{An other section}

Showing a great bullet list environme

### 2010-07-09 Thesis/Resume.tex
Added directory for PhD thesis
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@122 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2010-07-09 Thesis/Abstract.tex
@@
-\cleardoublepage
+\chapter*{Résumé}
 
-\begin{vcenterpage}
-\noindent\rule[2pt]{\textwidth}{0.5pt}
-\begin{center}
-{\large\textbf{Real-time soft tissue modelling on GPU for medical simulation\\}}
-\end{center}
-{\large\textbf{Abstract:}}
+\large\textbf{Simulation de tissus mous en temps réel sur GPU pour la simulation médicale} \\
+
 Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla. Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.
@@
 Blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla.
-\\
-{\large\textbf{Keywords:}}
-Medical simulation, soft tissue modelling, finite element method, GPU
-\\
-\noindent\rule[2pt]{\textwidth}{0.5pt}
-\end{vcenterpage}

### 2010-07-09 Thesis/TitlePage.tex
Added directory for PhD thesis
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@122 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\begin{titlepage}
\begin{center}
\noindent {\LARGE \textbf{UNIVERSITY OF LILLE 1}} \\
\vspace*{0.3cm}
\noindent {\LARGE \textbf{DOCTORAL SCHOOL OF ENGINEERING SCHOOL}} \\
\vspace*{0.5cm}
\noindent \Huge \textbf{P H D\ \ T H E S I S} \\
\vspace*{0.3cm}
\noindent \large {to obtain the title of} \\
\vspace*{0.3cm}
\noindent \LARGE \textbf{PhD of Science} \\
\vspace*{0.3cm}
\noindent \Large of the University of Lille 1 \\
\noindent \Large \textbf{Specialty: \textsc{Computer Science}}\\
\vspace*{0.4cm}
\noindent \large {Defended by\\}
\noindent \LARGE Olivier \textsc{Comas} \\
\vspace*{0.8cm}
\noindent {\Huge \textbf{Real-time soft tissue modelling on GPU for medical simulation}} \\
\vspace*{0.8cm}
\noindent \Large Thesis Advisor: Stéphane \textsc{Cotin} \\
\vspace*{0.2cm}
\noindent \Large prepared at \\ 
INRIA Lille, \textsc{Shaman} Team and CSIRO ICT Brisbane, EAHRC \\
\vspace*{0.7cm}
\noindent \large defended on the $00^{th}$ of December 2010 \\
\end{center}
\noindent \large \textbf{Jury:} \\
\begin{center}
\noindent \large 
\begin{tabular}{llcl}
      \textit{Reviewers :}	& Bernard \textsc{Name}		& - & Origin\\
				& Bernard \textsc{Name}		& - & Origin\\
      \textit{Advisor :}	& Stéphane \textsc{Cotin}		& - & INRIA (Shaman)\\
      \textit{President :}	& Bernard \textsc{Name}		& - & Origin\\
      \textit{Examinators :}   & Bernard \textsc{Name}          & - & Origin\\
      				& Bernard \textsc{Name}			& - & Origin\\
      				& Bernard \textsc{Name}		& - & Origin\\
      \textit{Invited :}		& Bernard \textsc{Name}		& - & Origin
\end{tabular}
\end{center}
\end{titlepage}
\sloppy

\titlepage

### 2010-07-09 Thesis/compile.sh

@@ -1,0 +1,13 @@
+# Compiles paper properly, deletes temporary files and open the generated pdf (tested on Mac OS)
+
+rm -f *.{blg,bbl,toc,out,log,aux,pdf,maf,brf,mtc,mtc0,mtc1,mtc2,mtc3,nlo,lot,lof,nls,ilg}
+
+pdflatex Thesis
+bibtex Thesis
+makeindex Thesis.nlo -s nomencl.ist -o Thesis.nls
+pdflatex Thesis
+pdflatex Thesis
+
+rm -f *.{blg,bbl,toc,out,log,aux,maf,brf,mtc,mtc0,mtc1,mtc2,mtc3,nlo,lot,lof,nls,ilg}
+
+open Thesis.pdf

### 2010-07-12 Thesis/Chapter1.tex
Delete most of dummy contents
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@123 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \section{Illustration Example}
-
-\subsection{A subsection just for fun}
-
-Sorry I won't write your PhD here ;) This small text just to mention that this style supports writing with accents such as in french words (thse, dfinir, ...).
-Also I put here a simple way to include an image.
-This is standard latex.
-For pdflatex compilation, the extension of the images is jpg.
-For latex compilation, this is ps or eps.
-The base folder containing images is set in formatAndDefs.tex, as well as the default extensions added to the image names.
-
-\begin{figure}[!htbp]
-  \begin{center}
-    \includegraphics[width=0.9\textwidth]{Chapter1/arctic_control}
-  \end{center}
-  \caption{A nice image...}
-  \label{fig:jolieImage}
-\end{figure}
-
-\section{An equation}
-
-Just to show argmin and partial derivative commands.
-First, we want to obtain a registration method which is as independent as possible w.r.t.
-the setting of its parameters.
-This setting, done by the clinician, indeed needs to be minimal while guaranteeing a robust result.
-We therefore propose registration methods allowing to better control the obtained transformation, using outlier rejection techniques or locally affine transformations.First, we want to obtain a registration method which is as independent as possible w.r.t.
-the setting of its parameters.
-This setting, done by the clinician, indeed needs to be minimal while guaranteeing a robust result.
-We therefore propose registration methods allowing to better control the obtained transformation, using outlier rejection techniques or locally affine transformations.First, we want to obtain a registration method which is as independent as possible w.r.t.
-the setting of its parameters.
-This setting, done by the clinician, indeed needs to be minimal while guaranteeing a robust result.
-We therefore propose registration methods allowing to better control the obtained transformation, using outlier rejection techniques or locally affine transformations.First, we want to obtain a registration method which is as independent as possible w.r.t.
-the setting of its parameters.
-This setting, done by the clinician, indeed needs to be minimal while guaranteeing a robust result.
-We therefore propose registration methods allowing to better control the obtained transformation, using outlier rejection techniques or locally affine transformations.
-First, we want to obtain a registration method which is as independent as possible w.r.t.
-the setting of its parameters.
-This setting, done by the clinician, indeed needs to be minimal while guaranteeing a robust result.
-We therefore propose registration methods allowing to better control the obtained transformation, using outlier rejection techniques or locally affine transformations.First, we want to obtain a registration method which is as independent as possible w.r.t.
-the setting of its parameters.
-This setting, done by the clinician, indeed needs to be minimal while guaranteeing a robust result.
-We therefore propose registration methods allowing to better control the obtained transformation, using outlier rejection techniques or locally affine transformations.
-Regularization:
-
-\begin{equation}
-  \pd{T}{t} = \Delta T
-\end{equation}
-
-\section{An other section}
-
-Showing a great bullet list environme

### 2010-07-12 Thesis/Chapter2.tex
Added directory for PhD thesis
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@122 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
Delete most of dummy contents
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@123 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2010-07-09 Thesis/Chapter1.tex
@@
-\chapter{Introduction}
+\chapter{My first real chapter}
 \label{chap:intro}
@@
 \end{shortAbstract}
-
-\section{Illustration Example}
-
-\subsection{A subsection just for fun}
-
-Sorry I won't write your PhD here ;) This small text just to mention that this style supports writing with accents such as in french words (thse, dfinir, ...).
-Also I put here a simple way to include an image.
-This is standard latex.
-For pdflatex compilation, the extension of the images is jpg.
-For latex compilation, this is ps or eps.
-The base folder containing images is set in formatAndDefs.tex, as well as the default extensions added to the image names.
-
-\begin{figure}[!htbp]
-  \begin{center}
-    \includegraphics[width=0.9\textwidth]{Chapter1/arctic_control}
-  \end{center}
-  \caption{A nice image...}
-  \label{fig:jolieImage}
-\end{figure}
-
-\section{An equation}
-
-Just to show argmin and partial derivative commands.
-First, we want to obtain a registration method which is as independent as possible w.r.t.
-the setting of its parameters.
-This setting, done by the clinician, indeed needs to be minimal while guaranteeing a robust result.
-We therefore propose registration methods allowing to better control the obtained transformation, using outlier rejection techniques or locally affine transformations.First, we want to obtain a registration method which is as independent as possible w.r.t.
-the setting of its parameters.
-This setting, done by the clinician, indeed needs to be minimal while guaranteeing a robust result.
-We therefore propose registration methods allowing to better control the obtained transformation, using outlier rejection techniques or locally affine transformations.First, we want to obtain a registration method which is as independent as possible w.r.t.
-the setting of its parameters.
-This setting, done by the clinician, indeed needs to be minimal while guaranteeing a robust result.
-We therefore propose registration methods allowing to better control the obtained transformation, using outlier rejection techniques or locally affine transformations.First, we want to obtain a registration method which is as independent as possible w.r.t.
-the setting of its parameters.
-This setting, done by the clinician, indeed needs to be minimal while guaranteeing a robust result.
-We therefore propose registration methods allowing to better control the obtained transformation, using outlier rejection techniques or locally affine transformations.
-First, we want to obtain a registration method which is as independent as possible w.r.t.
-the setting of its parameters.
-This setting, done by the clinician, indeed needs to be minimal while guaranteeing a robust result.
-We therefore propose registration methods allowing to better control the obtained transformation, using outlier rejection techniques or locally affine transformations.First, we want to obtain a registration method which is as independent as possible w.r.t.
-the setting of its parameters.
-This setting, done by the clinician, indeed needs to be minimal while guaranteeing a robust result.
-We therefore propose registration methods allowing to better control the obtained transformation, using outlier rejection techniques or locally affine transformations.
-Regularization:
-
-\begin{equation}
-  \pd{T}{t} = \Delta T
-\end{equation}
-
-\section{An other section}
-
-Showing a great bullet list environme

### 2010-07-14 Thesis/TitlePage.tex
First table of contents + update of bibliography
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@125 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
Added draft features, watermark, time stamp
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@126 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \vspace*{0.3cm}
-\noindent {\LARGE \textbf{DOCTORAL SCHOOL OF ENGINEERING SCHOOL}} \\
+\noindent {\LARGE \textbf{DOCTORAL SCHOOL OF ENGINEERING SCIENCE}} \\
 \vspace*{0.5cm}
@@
 \vspace*{0.8cm}
-\noindent {\Huge \textbf{Real-time soft tissue modelling on GPU for medical simulation}} \\
+\noindent {\Huge \textbf{Real-time Soft Tissue Modelling on GPU for Medical Simulation}} \\
 \vspace*{0.8cm}
@@
 \vspace*{0.7cm}
-\noindent \large defended on the $00^{th}$ of December 2010 \\
+%\noindent \large defended on the $00^{th}$ of December 2010 \\
+\noindent \large \color{red}{DRAFT - \today ~at \currenttime} \\
 \end{center}
@@
 \begin{tabular}{llcl}
-      \textit{Reviewers :}	& Bernard \textsc{Name}		& - & Origin\\
-				& Bernard \textsc{Name}		& - & Origin\\
-      \textit{Advisor :}	& Stéphane \textsc{Cotin}		& - & INRIA (Shaman)\\
-      \textit{President :}	& Bernard \textsc{Name}		& - & Origin\\
-      \textit{Examinators :}   & Bernard \textsc{Name}          & - & Origin\\
-      				& Bernard \textsc{Name}			& - & Origin\\
-      				& Bernard \textsc{Name}		& - & Origin\\
-      \textit{Invited :}		& Bernard \textsc{Name}		& - & Origin
+      \textit{Reviewers:}	& Bernard \textsc{}		& - & INRIA (Shaman)\\
+				& Bernard \textsc{}		& - & INRIA (Shaman)\\
+      \textit{Advisor:}	& Stéphane \textsc{Cotin}		& - & INRIA (Shaman)\\
+      \textit{President:}	& Bernard \textsc{}		& - & INRIA (Shaman)\\
+      \textit{Examinators:}   & Bernard \textsc{}          & - & INRIA (Shaman)\\
+      				& Bernard \textsc{}			& - & INRIA (Shaman)\\
+      				& Bernard \textsc{}		& - & INRIA (Shaman)\\
 \end{tabular}

### 2010-07-15 Thesis/tableofcontents.tex
2 new designs for the title page + add table of contents
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@127 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\part{Introduction}

	\chapter{Medical simulation}
		\section{General context and goal: medical training, patient-specific planning and per-operative guidance}
		\section{Challenges (trade-off between accuracy and real-time)}

	\chapter{One key point in medical simulation: soft-tissue modelling}
		\section{Necessary background in continuum mechanics}
			\subsection{Deformation tensor and strain tensor}
			\subsection{Stress and constitutive laws}
		\section{Tissue characterisation}
			\subsection{Material models for organs (non-linear, visco-elastic and anisotropic)}
			\subsection{Measure/estimation of model parameters}

	\chapter{Main principles of Finite Element Method (or how to solve equations of continuum mechanics from previous section)}
		\section{Discretisation}
		\section{Derivation of element equations}
		\section{Assembly of element equations}
		\section{Solution of global problem}

\part{Solid organs modelling}

	\chapter{State of art: FEM}

	\chapter{Linear not accurate => Non-linear FEM => Introduction of TLED}
		\section{Differences with classic FEM and reasons of its efficiency}
		\section{Visco-elasticity and anisotropy added \OC{MICCAI 2008; MedIA 2009}}

	\chapter{GPU implementation of TLED}
		\section{What is GPGPU}
		\section{Re-formulation of the algorithm for its Cg implementation}
		\section{CUDA implementation/optimisations \OC{ISBMS 2008a}}

	\chapter{Implementation in SOFA}
		\section{Presentation of SOFA project and architecture}
		\section{Implementation in SOFA and TLED released in open-source}

\part{Hollow organs modelling}

	\chapter{State of art: hollow structures}
		\section{Non-physic approaches (computer graphics stuff)}
		\section{Physically accurate approches (plates/shells)}

	\chapter{Colonoscopy simulator project}
		\section{Project introduction}
		\section{Mass-spring model for colon implemented on GPU \OC{ISBMS 2008b}}

	\chapter{More accurate: a co-rotational triangular shell model \OC{ISBMS 2010}}
		\section{Model description}
		\section{Validation}
		\section{Application to implant deployment simulation in cataract surgery}

	\chapter{'Shell meshing' technique \OC{MICCAI 2010}}
		\section{State of art: reconstruction/simplification}
		\section{Our method}

	\chapter{Applications to medical simulation}
		\section{Nice medical stuff to show}
		\section{Interaction solid/hollow organs}

\part{Conclusion}

### 2010-07-16 Thesis/TitlePage.tex
Change tableofcontents.tex to outline.tex + changed titlepage to reflect a draft
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@128 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{titlepage}
+
 \begin{center}
-\noindent {\LARGE \textbf{UNIVERSITY OF LILLE 1}} \\
-\vspace*{0.3cm}
-\noindent {\LARGE \textbf{DOCTORAL SCHOOL OF ENGINEERING SCIENCE}} \\
+
+\noindent
+\begin{tabular*}{\textwidth}{p{5cm}p{8.2cm}}
+\multirow{5}{*}{\includegraphics[width=5cm]{images/univ-lille}} & \hspace{\stretch{1}} \\
+& \hspace{\stretch{1}} \\
+& \hspace{\stretch{1}} \\
+& \hspace{\stretch{1}} Thesis submitted to obtain the title of\\
+& \hspace{\stretch{1}} Doctor of Philosophy\\ 
+\end{tabular*}
+
 \vspace*{0.5cm}
-\noindent \Huge \textbf{P H D\ \ T H E S I S} \\
-\vspace*{0.3cm}
-\noindent \large {to obtain the title of} \\
-\vspace*{0.3cm}
-\noindent \LARGE \textbf{PhD of Science} \\
-\vspace*{0.3cm}
-\noindent \Large of the University of Lille 1 \\
-\noindent \Large \textbf{Specialty: \textsc{Computer Science}}\\
-\vspace*{0.4cm}
-\noindent \large {Defended by\\}
-\noindent \LARGE Olivier \textsc{Comas} \\
-\vspace*{0.8cm}
-\noindent {\Huge \textbf{Real-time Soft Tissue Modelling on GPU for Medical Simulation}} \\
-\vspace*{0.8cm}
-\noindent \Large Thesis Advisor: Stéphane \textsc{Cotin} \\
-\vspace*{0.2cm}
-\noindent \Large prepared at \\ 
+\hspace{\stretch{1}} Doctoral School of Engineering Science\\
+\hspace{\stretch{1}} Field: Computer Science \\
+
+\vspace*{2.5cm}
+\Huge \textbf{Real-time Soft Tissue Modelling on GPU for Medical Simulation} \\
+\vspace*{2.5cm}
+\noindent \large Prepared by Olivier \textsc{Comas} at \\ 
 INRIA Lille, \textsc{Shaman} Team and CSIRO ICT Brisbane, EAHRC \\
-\vspace*{0.7cm}
-%\noindent \large defended on the $00^{th}$ of December 2010 \\
+\vspace*{0.5cm}
 \noindent \large \color{red}{DRAFT - \today ~at \currenttime} \\
+%\noindent \large Defended on the $00^{th}$ of December 2010 \\
 \end{center}
+\vspace*{1cm}
 \noindent \large \textbf{Jury:} \\
@@
       				& Bernard \textsc{}			& - & INRIA (Shaman)\\
-      				& Bernard \textsc{}		& - & INRIA (Shaman)\\
 \end{tabular}

### 2010-07-16 Thesis/Chapter2.tex
Update 15 July - Started chapter "soft-tissue modelling"
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@129 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\chapter{My first real chapter}
-\label{chap:intro}
+\chapter{Soft-tissue modelling}
+\label{chap:softtissue}
 \begin{shortAbstract}
-A short abstract for the upcoming chapter
+As seen in the previous chapter, realistic modelling of organs' deformation is a challenging research field that opens the door to new clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedure and per-operative guidance based on simulation.
+In all these cases the clinician needs fast updates of the deformation model to obtain a real-time display of the computed deformations.
+If for medical training devices the haptic feedback from touching organs merely needs to feel real, the accuracy of the information provided to the clinician in the cases of planning or per-operative guidance is crucial.
+Therefore a substantial comprehension of the mechanics involved and a knowledge of physical properties of anatomical structures are both mandatory in our quest to realistically model organs' deformation.
+This chapter will first introduce a few necessary basic concepts of continuum mechanics.
+It will then present the different theoretical models able to describe organs' mechanical behaviours.
 \end{shortAbstract}
+
+\section{Necessary background in continuum mechanics}
+
+\subsection{What is continuum mechanics}
+In our everyday life matter appears smooth and continuous: from the wood used to build your desk to the water you drink.
+But this is just illusion.
+The concept that matter is composed of discrete units has been around for millennia.
+In fact we know with certainty that our world is composed of microscopic atoms and molecules separated by empty space since the beginning of the twentieth century~\cite{Lautrup05}.
+However, certain physical phenomena can be predicted with theories that pay no attention to the molecular structure of materials.
+Consider for instance the deformation of the horizontal board of a bookshelf under the weight of the books.
+The bending of the shelf can be modelled without considering its molecular composition.
+The branch of physics in which materials are treated as continuous is known as continuum mechanics.
+In this theory matter is assumed to exist as a continuum, meaning that the matter in the body is continuously distributed and fills the entire region of space it occupies~\cite{Lai96}.
+Whether the approximation of continuum mechanics is justified in a given situation is a matter of experimental test.
+
+Continuum mechanics studies the response of materials to different loading conditions and can be divided into two main parts: general principles common to all media and constitutive equations defining idealised materials.
+
+\subsection{General principles}
+\subsubsection{Deformation tensor and strain tensor}
+\subsubsection{Stress}
+
+\subsection{Idealised materials}
+So far we have described the state of stress and strain and these concepts are valid for every continuum.
+However these equations are not sufficient to describe the response of a specific material due to a given loading.
+\subsubsection{Constitutive laws}
+
+\section{Tissue characterisation}
+	\subsection{Material models for organs (non-linear, visco-elastic and anisotropic)}
+	\subsection{Measure/estimation of model parameters?}

### 2010-07-19 Thesis/Chapter2.tex
- Update 19 July - Added detailled outline for chapter soft-tissue modelling and changed introduction to this chapter + .bib updated
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@130 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\chapter{Soft-tissue modelling}
+\chapter{Background in continuum mechanics for soft-tissue modelling}
 \label{chap:softtissue}
@@
 Therefore a substantial comprehension of the mechanics involved and a knowledge of physical properties of anatomical structures are both mandatory in our quest to realistically model organs' deformation.
-This chapter will first introduce a few necessary basic concepts of continuum mechanics.
+This chapter will introduce a few necessary concepts of continuum mechanics.
 It will then present the different theoretical models able to describe organs' mechanical behaviours.
@@
 
-\section{Necessary background in continuum mechanics}
-
-\subsection{What is continuum mechanics}
-In our everyday life matter appears smooth and continuous: from the wood used to build your desk to the water you drink.
+\section{Introduction}
+In our everyday life, matter appears smooth and continuous: from the wood used to build your desk to the water you drink.
 But this is just illusion.
 The concept that matter is composed of discrete units has been around for millennia.
-In fact we know with certainty that our world is composed of microscopic atoms and molecules separated by empty space since the beginning of the twentieth century~\cite{Lautrup05}.
+In fact we know with certainty that our world is composed of microscopic atoms and molecules separated by empty space since the beginning of the twentieth century~\citep{Lautrup05}.
 However, certain physical phenomena can be predicted with theories that pay no attention to the molecular structure of materials.
@@
 The branch of physics in which materials are treated as continuous is known as continuum mechanics.
-In this theory matter is assumed to exist as a continuum, meaning that the matter in the body is continuously distributed and fills the entire region of space it occupies~\cite{Lai96}.
+Continuum mechanics studies the response of materials to different loading conditions.
+In this theory, matter is assumed to exist as a continuum, meaning that the matter in the body is continuously distributed and fills the entire region of space it occupies~\citep{Lai96}.
 Whether the approximation of continuum mechanics is justified in a given situation is a matter of experimental test.
 
-Continuum mechanics studies the response of materials to different loading conditions and can be divided into two main parts: general principles common to all media and constitutive equations defining idealised materials.
+Modelling anatomical structures requires an understanding of the deformation and stresses caused by the different sollicitations that occur during medical procedures.
+A sufficient knowledge of continuum mechanics is therefore essential to follow the rest of this manuscript.
+Continuum mechanics can be divided into two main parts: general principles common to all media (analysis of deformation, strain and stress concepts) and constitutive equations defining idealised materials.
+This chapter will not only deal with those two aspects but will also introduce experiments carried out on organs in order to assess the physical parameters used in the theoretical models.
+This chapter will follow the notation used by~\cite{Reddy07} and the interested reader may refer to this book for more details.
 
-\subsection{General principles}
-\subsubsection{Deformation tensor and strain tensor}
-\subsubsection{Stress}
+\section{Description of motion}
+	\subsection{Lagrangian description}
+	\subsection{Eulerian description}
+	\subsection{Displacement field}
 
-\subsection{Idealised materials}
-So far we have described the state of stress and strain and these concepts are valid for every continuum.
-However these equations are not sufficient to describe the response of a specific material due to a given loading.
-\subsubsection{Constitutive laws}
+\section{Analysis of deformation}
+	\subsection{Deformation gradient tensor}
+	\subsection{Change of volume}
+
+\section{Strain measures}
+	\subsection{Cauchy-Green deformation tensors}
+	\subsection{Green strain tensor}
+	\subsection{Cauchy and Euler tensor}
+	\subsection{Infinitesimal strain tensor}
+
+\section{Stress}
+	\subsection{Cauchy stress}
+	\subsection{First Piola-Kirchhoff stress tensor}
+	\subsection{Second Piola-Kirchhoff stress tensor}
+
+\section{Constitutive equations}
+	\subsection{Elastic solids}	
+	\subsection{Generalised Hooke's law}
+	\subsection{Orthotropic materials}
+	\subsection{Isotropic materials}
 
@@
 	\subsection{Material models for organs (non-linear, visco-elastic and anisotropic)}
-	\subsection{Measure/estimation of model parameters?}
+	\subsection{Measure/estimation of model parameters}

### 2010-07-20 Thesis/Chapter2.tex
- Update 20 July - Updates on chapter soft tissue modelling + biblio update + add packages for mathfont and sets of reals, complex etc.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@131 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 If for medical training devices the haptic feedback from touching organs merely needs to feel real, the accuracy of the information provided to the clinician in the cases of planning or per-operative guidance is crucial.
-Therefore a substantial comprehension of the mechanics involved and a knowledge of physical properties of anatomical structures are both mandatory in our quest to realistically model organs' deformation.
-This chapter will introduce a few necessary concepts of continuum mechanics.
-It will then present the different theoretical models able to describe organs' mechanical behaviours.
+Therefore a substantial comprehension of the mechanics involved and a knowledge of physical properties of anatomical structures are both mandatory in our quest to realistically model the deformation of organs.
+This chapter will start by introducing the main concepts of continuum mechanics that are fundamental to study the mechanical response of organs.
+It will then present mathematical models able to describe the different mechanical aspects of materials and briefly expose their experimental validation.
 \end{shortAbstract}
@@
 The concept that matter is composed of discrete units has been around for millennia.
-In fact we know with certainty that our world is composed of microscopic atoms and molecules separated by empty space since the beginning of the twentieth century~\citep{Lautrup05}.
+In fact we know with certainty that our world is composed of microscopic atoms and molecules separated by empty space since the beginning of the twentieth century \citep{Lautrup05}.
 However, certain physical phenomena can be predicted with theories that pay no attention to the molecular structure of materials.
@@
 The bending of the shelf can be modelled without considering its molecular composition.
-The branch of physics in which materials are treated as continuous is known as continuum mechanics.
+The branch of physics in which materials are treated as continuous is known as \emph{continuum mechanics}.
 Continuum mechanics studies the response of materials to different loading conditions.
-In this theory, matter is assumed to exist as a continuum, meaning that the matter in the body is continuously distributed and fills the entire region of space it occupies~\citep{Lai96}.
-Whether the approximation of continuum mechanics is justified in a given situation is a matter of experimental test.
+In this theory, matter is assumed to exist as a continuum, meaning that the matter in the body is continuously distributed and fills the entire region of space it occupies \citep{Lai96}.
+This assumption is generally valid if the length scales of interest are large compared with the length scales of discrete molecular structure, but whether the approximation of continuum mechanics is actually justified in a given situation is merely a matter of experimental test.
 
@@
 Continuum mechanics can be divided into two main parts: general principles common to all media (analysis of deformation, strain and stress concepts) and constitutive equations defining idealised materials.
-This chapter will not only deal with those two aspects but will also introduce experiments carried out on organs in order to assess the physical parameters used in the theoretical models.
-This chapter will follow the notation used by~\cite{Reddy07} and the interested reader may refer to this book for more details.
+This chapter will not only deal with those two aspects but will also introduce experiments carried out on organs in order to assess the physical parameters used in the mathematical models.
+This chapter will follow the notation used by \cite{Reddy07} and the interested reader may refer to this book for more details.
 
 \section{Description of motion}
+Let us consider a body $\boldsymbol \beta$ of known geometry in a three-dimensional Euclidian space $\field{R}^3$.
+For a given geometry and loading, $\boldsymbol \beta$ will undergo a set of macroscopic changes which is called \emph{deformation}.
+The region of space occupied by the body at a given time $t$ is termed a \emph{configuration}.
+A change in the configuration of a continuum body results in a displacement.
+The displacement of a body has two components: a rigid-body displacement and a deformation.
+A rigid-body displacement consists of a simultaneous translation and rotation of the body without changing its shape or size.
+Deformation implies the change in shape and/or size of the body from an initial configuration noted $\kappa_{0}$ to a new configuration $\kappa$ called the \emph{current} or \emph{deformed configuration}.
+
+Let us now consider a given particule of the body that we call $X$.
+What we will call particule in the following is in fact an infinitesimal volume of material.
+We denote the position it occupies in the initial configuration $\textbf{X}$ and note $\textbf{x}$ its position in the deformed configuration, both expressed in the chosen frame of reference.
+The mapping $\chi : \boldsymbol \beta_{\kappa_{0}} \rightarrow \boldsymbol \beta_{\kappa}$ is called the \emph{deforming mapping} of the body $\boldsymbol \beta$ from $\kappa_{0}$ to $\kappa$.
+When analysing the deformation of a continuous body, it is necessary to describe the evolution of configurations through time.
+Its mathematical description follows one of the two approaches: the material description or the spatial description.
+The material description is known as \emph{Lagrangian description} whereas the spatial description is also called \emph{Eulerian description}.
+These two approaches are detailled next.
+
 	\subsection{Lagrangian description}
+In the Lagrangian description, the position and physical properties of the particles are referred to a reference configuration $\kappa_{R}$, often chosen to be the undeformed configuration $\kappa_{0}$.
+Thus, the current coordinates ($\textbf{x} \in \kappa$) are expressed in terms of the reference coordinates ($\textbf{X} \in \kappa_{0}$):
+\begin{equation}
+\textbf{x} = \chi(\textbf{X}, t), \quad \chi(\textbf{X}, 0) = \textbf{X},
+\end{equation}
+and the variation of a typical variable $\phi$ over the body is described with respect to the coordinates $\textbf{X}$ and time $t$:
+\begin{equation}
+\phi = \phi(\textbf{X}, t) \enspace .
+\end{equation}
+For a fixed value of $\textbf{X} \in \kappa_{0}$, $\phi(\textbf{X},t)$ gives the value of $\phi$ at time $t$ associated with the fixed particule $X$ whose position in the reference configuration is $\textbf{X}$.
+We note that a change in time $t$ implies that the same particule $X$ has a different value $\phi$.
+Thus, the Lagrangian description focuses its attention on the particules of the continuous body and it is usually used in solid mechanics.
+
 	\subsection{Eulerian description}
+Rather than following the particules, the Eulerian description observes the changes at fixed locations.
+	In the Eulerian description, the motion is this time referred to the current configuration $\kappa$ and $\phi$ is described with respect to the current position ($\textbf{x} \in \kappa$):
+\begin{equation}
+\phi = \phi(\textbf{x}, t), \quad \textbf{X} = \textbf{X}(\textbf{x}, t) \enspace .
+\end{equation}
+For a fixed value of $\textbf{x} \in \kappa$, $\phi(\textbf{x},t)$ gives the value of $\phi$ associated with different material particules at different times, because different particules occupy the position $\textbf{x} \in \kappa$ at different times.
+A change in time $t$ implies that a different value $\phi$ is observed at the same spatial location $\textbf{x} \in \kappa$, now probably occupied by a different particule.
+Hence, the Eulerian description is focused on a spatial position.
+	This approach is convenient for the study of fluid flow where the kinematic property of greatest interest is the rate at which change is taking place rather than the shape of the body of fluid at a reference time \citep{Spencer80}.
+Because we are only interested in the study of solid bodies, the Lagrangian description will be used in the rest of the text.
+
 	\subsection{Displacement field}
+The deplacement of a particule $X$ is called \emph{displacement vector} and its expression in the Lagrangian description is the following:
+\begin{equation}
+\textbf{u}(\textbf{X}, t) = \textbf{x}(\textbf{X}, t) - \textbf{X} \enspace .
+\end{equation}
+A \emph{displacement field} is a vector field of all displacement vectors for all particles in the body, which relates the deformed configuration $\kappa$ with the undeformed configuration $\kappa_{0}$.
+Indeed if the displacement field is known, we can construct the current configuration $\kappa$ from the undeformed configuration $\kappa_{0}$: $\chi(\textbf{X}, t) = \textbf{X} + \textbf{u}(\textbf{X})$.
 
 \section{Analysis of deformation}
+
 	\subsection{Deformation gradient tensor}
+The displacement field tells us how a point displaces from the reference to the deformed configuration.
+However, we would also like to know how a piece of material is stretched and rotated as the body moves from the reference to the deformed configuration.
+Since the length of a material line $d\textbf{X}$ can change when going to the deformed configuration as well as its orientation, we can say that $d\textbf{X}$ deforms into $d\textbf{x}$.
+The question then becomes how to relate $d\textbf{x}$ in the deformed configuration into $d\textbf{X}$ of the reference configuration.
+The \emph{deformation gradient} gives this relationship and is defined as:
+\begin{equation}
+d\textbf{x} = \textbf{F}  d\textbf{X}
+\end{equation}
+
 	\subsection{Change of volume}
+
+	\newpage
 

### 2010-07-21 Thesis/Chapter2.tex
- Update 21 July - Wrote Analysis of deformation and Strain measures sections + biblio updated
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@132 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 This chapter will not only deal with those two aspects but will also introduce experiments carried out on organs in order to assess the physical parameters used in the mathematical models.
-This chapter will follow the notation used by \cite{Reddy07} and the interested reader may refer to this book for more details.
+This chapter will follow the notation used by \cite{Bonnet97} and \cite{Reddy07}.
+The interested reader may refer to these books for more details.
 
@@
 A rigid-body displacement consists of a simultaneous translation and rotation of the body without changing its shape or size.
-Deformation implies the change in shape and/or size of the body from an initial configuration noted $\kappa_{0}$ to a new configuration $\kappa$ called the \emph{current} or \emph{deformed configuration}.
+Deformation implies the change in shape and\slash or size of the body from an initial configuration noted $\kappa_{0}$ to a new configuration $\kappa$ called the \emph{current} or \emph{deformed configuration}.
 
@@
 What we will call particule in the following is in fact an infinitesimal volume of material.
-We denote the position it occupies in the initial configuration $\textbf{X}$ and note $\textbf{x}$ its position in the deformed configuration, both expressed in the chosen frame of reference.
-The mapping $\chi : \boldsymbol \beta_{\kappa_{0}} \rightarrow \boldsymbol \beta_{\kappa}$ is called the \emph{deforming mapping} of the body $\boldsymbol \beta$ from $\kappa_{0}$ to $\kappa$.
+We denote the position it occupies in the initial configuration $\mathbf{X}$ and note $\mathbf{x}$ its position in the deformed configuration, both expressed in the chosen frame of reference.
+The mapping $\chi$ defined as the following:
+\begin{equation}
+\label{chap2:mapping}
+\chi : \boldsymbol \beta_{\kappa_{0}} \rightarrow \boldsymbol \beta_{\kappa}
+\end{equation} 
+is called the \emph{deforming mapping} of the body $\boldsymbol \beta$ from $\kappa_{0}$ to $\kappa$.
 When analysing the deformation of a continuous body, it is necessary to describe the evolution of configurations through time.
@@
 In the Lagrangian description, the position and physical properties of the particles are referred to a reference configuration $\kappa_{R}$, often chosen to be the undeformed configuration $\kappa_{0}$.
-Thus, the current coordinates ($\textbf{x} \in \kappa$) are expressed in terms of the reference coordinates ($\textbf{X} \in \kappa_{0}$):
+Thus, the current coordinates ($\mathbf{x} \in \kappa$) are expressed in terms of the reference coordinates ($\mathbf{X} \in \kappa_{0}$):
 \begin{equation}
@@
 \end{equation}
-and the variation of a typical variable $\phi$ over the body is described with respect to the coordinates $\textbf{X}$ and time $t$:
+and the variation of a typical variable $\phi$ over the body is described with respect to the coordinates $\mathbf{X}$ and time $t$:
 \begin{equation}
-\phi = \phi(\textbf{X}, t) \enspace .
+\phi = \phi(\textbf{X}, t).
 \end{equation}
-For a fixed value of $\textbf{X} \in \kappa_{0}$, $\phi(\textbf{X},t)$ gives the value of $\phi$ at time $t$ associated with the fixed particule $X$ whose position in the reference configuration is $\textbf{X}$.
+For a fixed value of $\mathbf{X} \in \kappa_{0}$, $\phi(\mathbf{X},t)$ gives the value of $\phi$ at time $t$ associated with the fixed particule $X$ whose position in the reference configuration is $\mathbf{X}$.
 We note that a change in time $t$ implies that the same particule $X$ has a different value $\phi$.
@@
 Rather than following the particules, the Eulerian description observes the changes at fixed locations.
-	In the Eulerian description, the motion is this time referred to the current configuration $\kappa$ and $\phi$ is described with respect to the current position ($\textbf{x} \in \kappa$):
+	In the Eulerian description, the motion is this time referred to the current configuration $\kappa$ and $\phi$ is described with respect to the current position ($\mathbf{x} \in \kappa$):
 \begin{equation}
-\phi = \phi(\textbf{x}, t), \quad \textbf{X} = \textbf{X}(\textbf{x}, t) \enspace .
+\phi = \phi(\textbf{x}, t), \quad \textbf{X} = \textbf{X}(\textbf{x}, t).
 \end{equation}
-For a fixed value of $\textbf{x} \in \kappa$, $\phi(\textbf{x},t)$ gives the value of $\phi$ associated with different material particules at different times, because different particules occupy the position $\textbf{x} \in \kappa$ at different times.
-A change in time $t$ implies that a different value $\phi$ is observed at the same spatial location $\textbf{x} \in \kappa$, now probably occupied by a different particule.
+For a fixed value of $\mathbf{x} \in \kappa$, $\phi(\mathbf{x},t)$ gives the value of $\phi$ associated with different material particules at different times, because different particules occupy the position $\textbf{x} \in \kappa$ at different times.
+A change in time $t$ implies that a different value $\phi$ is observed at the same spatial location $\mathbf{x} \in \kappa$, now probably occupied by a different particule.
 Hence, the Eulerian description is focused on a spatial position.
@@
 \begin{equation}
-\textbf{u}(\textbf{X}, t) = \textbf{x}(\textbf{X}, t) - \textbf{X} \enspace .
+\textbf{u}(\textbf{X}, t) = \textbf{x}(\textbf{X}, t) - \textbf{X}.
 \end{equation}
 A \emph{displacement field} is a vector field of all displacement vectors for all particles in the body, which relates the deformed configuration $\kappa$ with the undeformed configuration $\kappa_{0}$.
-Indeed if the displacement field is known, we can construct the current configuration $\kappa$ from the undeformed configuration $\kappa_{0}$: $\chi(\textbf{X}, t) = \textbf{X} + \textbf{u}(\textbf{X})$.
+Indeed if the displacement field is known, we can construct the current configuration $\kappa$ from the undeformed configuration $\kappa_{0}$: $\chi(\mathbf{X}, t) = \mathbf{X} + \mathbf{u}(\mathbf{X})$.
 
@@
 However, we would also like to know how a piece of material is stretched and rotated as the body moves from the reference to the deformed configuration.
-Since the length of a material line $d\textbf{X}$ can change when going to the deformed configuration as well as its orientation, we can say that $d\textbf{X}$ deforms into $d\textbf{x}$.
-The question then becomes how to relate $d\textbf{x}$ in the deformed configuration into $d\textbf{X}$ of the reference configuration.
-The \emph{deformation gradient} gives this relationship and is defined as:
+Since the length of a material line $d\mathbf{X}$ can change when going to the deformed configuration as well as its orientation, we can say that $d\mathbf{X}$ deforms into $d\mathbf{x}$.
+The question then becomes how to relate $d\mathbf{x}$ in the deformed configuration with $d\mathbf{X}$ of the reference configuration.
+
+Consider two particules $P_{1}$ and $P_{2}$ in a continuous body separated by an infinitesimal distance $d\mathbf{X}$:
 \begin{equation}
-d\textbf{x} = \textbf{F}  d\textbf{X}
+\label{chap2:dX}
+d\mathbf{X} = \mathbf{X}_{P_{2}} - \mathbf{X}_{P_{1}}.
+\end{equation}
+After deformation the two particules have deformed to their current positions given by the mapping $\chi$~\eqref{chap2:mapping} as:
+\begin{equation}
+\mathbf{x}_{P_1} = \chi(\mathbf{X}_{P_{1}}, t) \quad \mbox{and} \quad \mathbf{x}_{P_2} = \chi(\mathbf{X}_{P_{2}}, t).
+\end{equation}
+Using \eqref{chap2:dX} the distance $d\textbf{x}$ between $P_{1}$ and $P_{2}$ can then be expressed as:
+\begin{equation}
+d\mathbf{x} = \mathbf{x}_{P_{2}} - \mathbf{x}_{P_{1}} = \chi(\mathbf{X}_{P_{1}} + d\mathbf{X}, t) - \chi(\mathbf{X}_{P_{1}}, t).
+\end{equation}
+The \emph{deformation gradient} $\textbf{F}$ can be defined as:
+\begin{equation}
+\mathbf{F} = \pd{\chi}{\mathbf{X}}
+\end{equation}
+and the vector $d\textbf{x}$ can then be obtained in terms of $d\mathbf{X}$ as:
+\begin{equation}
+d\textbf{x} = \textbf{F} \, d\textbf{X}.
+\end{equation}
+Note that $\textbf{F}$ transforms vectors from the reference configuration into vectors in the current configuration and is therefore a second-order tensor.
+
+Knowing that $\chi(\mathbf{X}, t)$ is of course $\mathbf{x}$, the deformation gradient may also be written as:
+\begin{equation}
+\label{chap2:gradient}
+\mathbf{F} = \pd{\mathbf{x}}{\mathbf{X}} = \nabla_{0} \mathbf{x} = \nabla_{0} \mathbf{u} + \mathbf{I},
+\end{equation}
+where $\nabla_{0}$ is the gradient operator with respect to $\mathbf{X}$ and $\mathbf{u}$ the displacement vector.
+In indicial notation in a Cartesian coordinate system, \eqref{chap2:gradient} can be explicited as:
+\begin{equation}
+[F] = 
+	\begin{bmatrix}
+		\pd{\mathbf{x}_{1}}{\mathbf{X}_{1}} & \pd{\mathbf{x}_{1}}{\mathbf{X}_{2}} & \pd{\mathbf{x}_{1}}{\mathbf{X}_{3}} \\\\
+		\pd{\mathbf{x}_{2}}{\mathbf{X}_{1}} & \pd{\mathbf{x}_{2}}{\mathbf{X}_{2}} & \pd{\mathbf{x}_{2}}{\mathbf{X}_{3}} \\\\
+		\pd{\mathbf{x}_{3}}{\mathbf{X}_{1}} & \pd{\mathbf{x}_{3}}{\mathbf{X}_{2}} & \pd{\mathbf{x}_{3}}{\mathbf{X}_{3}}
+	\end{bmatrix}
+	.
 \end{equation}
@@
 	\subsection{Change of volume}
-
-	\newpage
+At this point, we have seen how a deformation can affect a vector.
+We will now look into its effect on volumes.
+Our motivation comes from the need to write global equilibrium statements that involve integrals over volumes.
+We can define volume elements in the reference and deformed configurations.
+Consider three non-coplanar line elements $d\mathbf{X}^{(1)}$, $d\mathbf{X}^{(2)}$ and $d\mathbf{X}^{(3)}$ forming a parallelepiped in the reference configuration.
+The three vectors after deformation $d\mathbf{x}^{(1)}$, $d\mathbf{x}^{(2)}$ and $d\mathbf{x}^{(3)}$ can be obtained with:
+\begin{equation}
+\label{chap2:dxdX}
+d\mathbf{x}^{(i)} = \mathbf{F} \, d\mathbf{X}^{(i)}, \quad i = 1, 2, 3.
+\end{equation}
+The volume of the parallelepiped that we will note $dV$  can be calculated using the triple product between the three vectors:
+\begin{align}
+dV &= d\mathbf{X}^{(1)} \cdot d\mathbf{X}^{(2)} \times d\mathbf{X}^{(3)} = (\mathbf{\hat{N}}_{1} \cdot \mathbf{\hat{N}}_{2} \times \mathbf{\hat{N}}_{3}) \, dX^{(1)} dX^{(2)} dX^{(3)} \notag \\
+&= dX^{(1)} dX^{(2)} dX^{(3)}, \label{chap2:dV}
+\end{align}
+where $\mathbf{\hat{N}}_{i}$ denote the unit vector along $d\mathbf{X}^{i}$. The corresponding volume in the deformed configuration is given by:
+\begin{align}
+dv &= d\mathbf{x}^{(1)} \cdot d\mathbf{x}^{(2)} \times d\mathbf{x}^{(3)} \notag \\
+&=  (\mathbf{F} \cdot \mathbf{\hat{N}}_{1}) \cdot (\mathbf{F} \cdot \mathbf{\hat{N}}_{2}) \times (\mathbf{F} \cdot \mathbf{\hat{N}}_{3}) \,  dX^{(1)} dX^{(2)} dX^{(3)} \quad \text{by \eqref{chap2:dxdX}} \notag \\
+&= \det \mathbf{F} \, dX^{(1)} dX^{(2)} dX^{(3)}. \label{chap2:dv}
+\end{align}
+The determinant of $\mathbf{F}$ is called the \emph{Jacobian} and it is denoted by $J = \det \mathbf{F}$. And we have:
+\begin{equation}
+dv = J dV.
+\end{equation}
+Thus, J has the physical meaning of being the local ratio of current to reference volume of a material volume element.
 
 \section{Strain measures}
+The length of a material curve from the reference configuration can change when displaced to a curve in the deformed configuration.
+If all the curves do not change length, it is said that a rigid body displacement occurred.
+The concept of strain is used to evaluate how much a given displacement differs locally from a rigid body displacement.
+Therefore, although we know how to transform vectors from the reference configuration into vectors in the current configuration using the deformation gradient, it is more useful to find a measure of the change in length of $d\mathbf{X}$.
+Many measures of strains can be defined and the most common ones are now introduced.
+
 	\subsection{Cauchy-Green deformation tensors}
+Consider two particules $P_{1}$ and $P_{2}$ in the neighbourhood of each other, separated by $d\mathbf{X}$ in the reference configuration.
+In the deformed configuration $P_{1}$ and $P_{2}$ occupy the positions $\tilde{P}_{1}$ and $\tilde{P}_{2}$ and they are separated by $d\mathbf{x}$.
+We are interested in the change of distance between the two points $P_{1}$ and $P_{2}$ as the body deforms.
+
+The squared distances between $P_{1}$ and $P_{2}$ and $\tilde{P}_{1}$ and $\tilde{P}_{2}$ are respectively given by:
+\begin{align}
+(dS)^2 &= d\mathbf{X} \cdot d\mathbf{X} \label{chap2:dS} \\
+(ds)^2 &= d\mathbf{x} \cdot d\mathbf{x} = (\mathbf{F} \, d\mathbf{X}) \cdot (\mathbf{F} \, d\mathbf{X}) \notag \\
+&= (\mathbf{F} \, d\mathbf{X})^T (\mathbf{F} \, d\mathbf{X}) = (d\mathbf{X}^T \, \mathbf{F}^T) (\mathbf{F} \, d\mathbf{X}) = d\mathbf{X}^T (\mathbf{F}^T \mathbf{F} \, d\mathbf{X}) \notag \\
+&= d\mathbf{X} \cdot (\mathbf{F}^T \mathbf{F} \, d\mathbf{X}).
+\end{align}
+using the property that the dot product between two vectors $a$ and $b$ can also be expressed as the simple product between the transpose of $a$ and $b$ ($a \cdot b = a^T b$).
+We define the \emph{right Cauchy-Green deformation tensor} $\mathbf{C}$ as:
+\begin{equation}
+\mathbf{C} = \mathbf{F}^T \mathbf{F}.
+\end{equation}
+Thus, the change of distance between the two points $P_{1}$ and $P_{2}$ after deformation of the continuous body may be written as:
+\begin{equation}
+\label{chap2:ds}
+(ds)^2 = d\mathbf{X} \cdot (\mathbf{C} \, d\mathbf{X}).
+\end{equation}
+By definition, $\mathbf{C}$ is a symmetric second-order tensor. The transpose of $\mathbf{C}$ is denoted $\mathbf{B}$ and is called the \emph{left Cauchy-Green deformation tensor}:
+\begin{equation}
+\mathbf{B} = \mathbf{F} \mathbf{F}^T.
+\end{equation}
+
 	\subsection{Green strain tensor}
+Using \eqref{chap2:dS} and \eqref{chap2:ds}, the change in the squared lengths due to the body deformation between the reference and the current configuration can be expressed as:
+\begin{align}
+(ds)^2 - (dS)^2 &= d\mathbf{X} \cdot (\mathbf{C} \, d\mathbf{X}) - d\mathbf{X} \cdot d\mathbf{X} \notag \\
+&= d\mathbf{X} \cdot  (\mathbf{C} - \mathbf{I}) \, d\mathbf{X}.
+\end{align}
+Let us define the \emph{Green-St. Venant strain tensor} $\mathbf{E}$ as:
+\begin{equation}
+\mathbf{E} = \frac{1}{2}	 (\mathbf{C} - \mathbf{I})
+\end{equation}
+so we can write:
+\begin{equation}
+(ds)^2 - (dS)^2 = 2 \, d\mathbf{X} \cdot \mathbf{E} \, d\mathbf{X}.
+\end{equation}
+By definition, the Green strain tensor is a symmetric second-order tensor.
+Also, the change in squared lengths is zero if and only if $\mathbf{E} = \mathbf{0}$.
+Using \eqref{chap2:gradient}, the Green strain tensor may be developped as the following:
+\begin{align}
+\mathbf{E} &= \frac{1}{2} (\mathbf{F}^T \mathbf{F} - \mathbf{I}) \notag \\
+&= \frac{1}{2} \left[ (\nabla_{0} \mathbf{u} + \mathbf{I})^T (\nabla_{0} \mathbf{u} + \mathbf{I}) - \mathbf{I} \right] \notag \\
+&= \frac{1}{2} \left[ (\nabla_{0} \mathbf{u})^T  + \nabla_{0} \mathbf{u}+ (\nabla_{0} \mathbf{u})^T(\nabla_{0} \mathbf{u}) \right] \label{chap2:GreenTensor}
+\end{align}
+
+The Green strain tensor can be expressed in terms of its components in any coordinate system.
+In particular, in the cartesian coordinate system $(X_1, X_2, X_3)$ the components $E_{ij}$ of $\mathbf{E}$ are the following:
+\begin{equation}
+E_{i,j} = \frac{1}{2} \left( \pd{u_i}{Xj} + \pd{u_j}{Xi} + \pd{u_k}{Xi} \pd{u_k}{Xj} \right), \quad i = 1, 2, 3.
+\end{equation}
+The components $E_{11}$, $E_{22}$ and $E_{33}$ are called \emph{normal strains} and it can be shown that they are in fact the ratio of the change in length to the original length along each of the three unit vectors.
+The components $E_{12}$, $E_{23}$ and $E_{13}$ are called \emph{shear strains} and they can be interpreted as a measure of the change in angle between line elements that were perpendicular to each other in the undeformed configuration.
+
 	\subsection{Cauchy and Euler tensor}
+The change in the squared lengths during the body deformation can also be expressed relative to the current length. The length $dS$ can be written in terms of $d\mathbf{x}$ as:
+\begin{equation}
+(dS)^2 = d\mathbf{X} \cdot d\mathbf{X} = d\mathbf{x} \cdot (\mathbf{F}^{-T} \mathbf{F}^{-1}) d\mathbf{x} = d\mathbf{x} \cdot \mathbf{\tilde{B}} d\mathbf{x}
+\end{equation}
+where $\mathbf{\tilde{B}} = \mathbf{F}^{-T} \mathbf{F}^{-1}$ is called the \emph{Cauchy strain tensor}.
+$\mathbf{\tilde{B}}$ is in fact the inverse of the left Cauchy-Green tensor $\mathbf{B}$ introduced previously.
+
+In a similar way we defined the Green strain tensor, we can write the change in the squared lengths but relative to the current length:
+\begin{equation}
+(ds)^2 - (dS)^2 = 2 \, d\mathbf{x} \cdot \mathbf{e} \, d\mathbf{x}.
+\end{equation}
+where $\mathbf{e}$ is called \emph{Almansi-Hamel strain tensor} or simply \emph{Euler strain tensor}.
+
+	\subsection{\TODO{Principal strains?}}
+
 	\subsection{Infinitesimal strain tensor}
+In some cases, it is possible to simplify the expression of the Green strain tensor defined in \eqref{chap2:GreenTensor}.
+Indeed, when the displacement gradients are small (that is, $|\nabla \mathbf{u}| \ll 1$) we can neglect the nonlinear terms in the definition.
+In the case of infinitesimal strains, the Green-Lagrange strain tensor and the Eulerian strain tensor are approximately the same and can be approximated by the infinitesimal strain tensor denoted $\boldsymbol \epsilon$ and is given by:
+\begin{equation}
+\boldsymbol \epsilon = \frac{1}{2} \left[ \nabla \mathbf{u} + (\nabla \mathbf{u})^T \right]
+\end{equation}
+Its Cartesian components $\epsilon_{ij}$ are the following:
+\begin{equation}
+\epsilon_{ij} = \frac{1}{2} (\pd{u_i}{X_j} + \pd{u_j}{X_i}).
+\end{equation}
+
+The strain quantities defined in the previous sections are nonlinear expressions in terms of the mapping $\chi$ and will lead to nonlinear governing equations.
+In solid mechanics, whenever the hypothesis is acceptable, it is common practice to assume that the displacements are small and the infinitesimal strain tensor is used as a measure of the deformation.
 

### 2010-07-22 Thesis/Chapter2.tex
- Update 22 July - Added section on stresses
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@133 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{equation}
+\label{chap2:dxdX}
 d\textbf{x} = \textbf{F} \, d\textbf{X}.
@@
 Thus, J has the physical meaning of being the local ratio of current to reference volume of a material volume element.
+
+	\subsection{\TODO{Change of surface?}}
 
@@
 
-	\subsection{Infinitesimal strain tensor}
+	\subsection{Infinitesimal strain tensor} \label{chap2:infinitesimalStrainTensor}
 In some cases, it is possible to simplify the expression of the Green strain tensor defined in \eqref{chap2:GreenTensor}.
@@
 \begin{equation}
-\boldsymbol \epsilon = \frac{1}{2} \left[ \nabla \mathbf{u} + (\nabla \mathbf{u})^T \right]
+\boldsymbol \epsilon = \frac{1}{2} \left[ \nabla \mathbf{u} + (\nabla \mathbf{u})^T \right].
 \end{equation}
@@
 The strain quantities defined in the previous sections are nonlinear expressions in terms of the mapping $\chi$ and will lead to nonlinear governing equations.
-In solid mechanics, whenever the hypothesis is acceptable, it is common practice to assume that the displacements are small and the infinitesimal strain tensor is used as a measure of the deformation.
+In solid mechanics, whenever the hypothesis is acceptable, it is common practice to assume that the displacements are small and the infinitesimal strain tensor is then used as a measure of the deformation.
 
 \section{Stress}
-	\subsection{Cauchy stress}
-	\subsection{First Piola-Kirchhoff stress tensor}
-	\subsection{Second Piola-Kirchhoff stress tensor}
+Stress is a measure of the intensity of the internal forces acting between particles of a deformable body across imaginary internal surfaces.
+These internal forces are produced between the particles in the body as a reaction to external forces applied on the body.
+The $SI$ unit for stress is pascal (symbol $Pa$), which is equivalent to one newton (force) per square meter (unit area).
+As with strain, different measures of stress can be defined.
+For instance, stress can be measured per unit of deformed area (section \ref{chap2:CauchyStress}) or undeformed area (sections \ref{chap2:PiolaStress1} and \ref{chap2:PiolaStress2}).
+In general, the stress is not uniformly distributed over the cross section of a material body, and consequently the stress at a point on a given area is different from the average stress over this entire area.
+Therefore, it is necessary to define the stress not over a given area but at a specific point in the body.
+Stress at a point in a three-dimensional continuum can be measured in terms of nine quantities, three per plane, on three mutually perpendicular planes at the point.
+These nine quantities may be viewed as the components of a second-order tensor that we call \emph{stress tensor}.
+
+	\subsection{Cauchy stress}\label{chap2:CauchyStress}
+The Euler-Cauchy's stress principle states that on any surface (real or imaginary) that divides the body, the action of one part of the body on the other is equivalent to the system of distributed forces and couples on the surface dividing the body.
+With that consideration in mind, let us consider a plane $S$ that passes through an arbitrary internal point $P$ which has a unit normal vector $\mathbf{n}$.
+This plane separates the body into two regions and we are interested in the forces applied by one region onto the other.
+First we will introduce the true stress, defined as the stress in the deformed configuration $\chi$ measured per unit area of the deformed configuration.
+The force $\Delta\mathbf{f}(\mathbf{n})$ acting on a small element of area $\Delta a$ in a continuous medium depends not only on the size of the area but also depends on the orientation of this area $\mathbf{n}$.
+The \emph{Cauchy stress vector} at $P$ on $S$ can be defined as:
+\begin{equation}
+\mathbf{t}(\mathbf{n}) = \lim_{\Delta a \to 0} \frac{\Delta \mathbf{f}(\mathbf{n})}{\Delta a}.
+\end{equation}
+This equation means that the stress vector depends on its location in the body and the orientation of the plane on which it is acting.
+In general, the stress vector is not perpendicular to that plane and it can be separated into two components: one component normal to the plane called \emph{normal stress} and one component tangent to the plane named \emph{shear stress}.
+Since $\mathbf{t}$ depends on $\mathbf{n}$ but is not in the direction of $\mathbf{n}$, it may be interesting to look into the relationship between $\mathbf{t}$ and $\mathbf{n}$.
+
+To establish this relationship, let us consider an infinitesimal tetrahedron in a Cartesian coordinate system.
+\TODO{Add figure of a tetrahedron with all notations}.
+We note $-\mathbf{t}_1$, $-\mathbf{t}_2$, $-\mathbf{t}_3$ and $\mathbf{t}$ the stress vectors in the outwards directions on the faces of the tetrahedron whose areas are respectively $\Delta a_1$, $\Delta a_2$, $\Delta a_3$ and $\Delta a$.
+For the mass $m$ inside the tetrahedron, we have by Newton's second law:
+\begin{equation}
+\label{chap2:Newton}
+\sum \mathbf{F} = \mathbf{t} \Delta a - \mathbf{t}_1 \Delta a_1 - \mathbf{t}_2 \Delta a_2 - \mathbf{t}_3 \Delta a_3 + \Delta v \mathbf{f} = m \mathbf{a},
+\end{equation}
+where $\mathbf{f}$ is the force per unit volume acting on the body, $\Delta v$ the volume of the tetrahedron and $\mathbf{a}$ the acceleration.
+The areas $\Delta a_1$, $\Delta a_2$ and $\Delta a_3$ being the projections of $\Delta a$, they are related to $\Delta a$ in the Cartesian coordinate system $(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3)$ by:
+\begin{equation}
+\label{chap2:DeltaAi}
+\Delta a_i = (\mathbf{n} \cdot \mathbf{e}_i) \Delta a, \quad i=1, 2, 3.
+\end{equation}
+Dividing \eqref{chap2:Newton} by $\Delta a$, using \eqref{chap2:DeltaAi} and noting that $\Delta v \slash \Delta a \rightarrow 0$ when the tetrahedron shrinks to a point yields:
+\begin{align}
+\mathbf{t} &= (\mathbf{n} \cdot \mathbf{e}_1) \mathbf{t}_1 + (\mathbf{n} \cdot \mathbf{e}_2) \mathbf{t}_2 + (\mathbf{n} \cdot \mathbf{e}_3) \mathbf{t}_3 \notag \\
+&= \mathbf{n} \cdot (\mathbf{e}_1 \mathbf{t}_1 + \mathbf{e}_2 \mathbf{t}_2 + \mathbf{e}_3 \mathbf{t}_3)
+\end{align}
+
+We define the \emph{Cauchy stress tensor} $\sigma = \mathbf{e}_1 \mathbf{t}_1 + \mathbf{e}_2 \mathbf{t}_2 + \mathbf{e}_3 \mathbf{t}_3$ and we have:
+\begin{equation}
+\label{chap2:CauchyFormula}
+\mathbf{t}(\mathbf{n}) = \mathbf{n} \cdot \sigma
+\end{equation}
+The stress vector $\mathbf{t}$ represents the vectorial stress on a plane whose normal is $\mathbf{n}$.
+As we demonstrated, the stress tensor is a property of the medium that is independent of $\mathbf{n}$.
+It represents the current force per unit of deformed area: $d\mathbf{f} = \mathbf{t} da = \sigma \cdot d\mathbf{a}$.
+
+The matrix form of \eqref{chap2:CauchyFormula} in a Cartesian coordinate system is given by:
+\begin{equation}
+	\begin{bmatrix}
+		t_1 \\
+		t_2 \\
+		t_3
+	\end{bmatrix}
+= 
+	\begin{bmatrix}
+		\sigma_{11} && \sigma_{21} && \sigma_{31} \\
+		\sigma_{12} && \sigma_{22} && \sigma_{32} \\
+		\sigma_{13} && \sigma_{23} && \sigma_{33} \\
+	\end{bmatrix}
+	\begin{bmatrix}
+		n_1 \\
+		n_2 \\
+		n_3
+	\end{bmatrix}
+	.
+\end{equation}
+
+The Cauchy stress tensor is used for stress analysis of material bodies experiencing small deformations.
+This stress is basically defined as force/(unit deformed area).
+The strain measure that is appropriate to use with the Cauchy stress tensor is the infinitesimal strain tensor (see section~\ref{chap2:infinitesimalStrainTensor}).
+The problem with using the Cauchy stress tensor for analysing materials undergoing large deformation is that we generally do not know the area in the deformed configuration.
+Therefore, we need to develop alternative stress tensors that we can use in the reference configuration.
+We will introduce two new stress measures: the first Piola-Kirchhoff stress tensor and the second Piola-Kirchhoff stress tensor.
+
+	\subsection{First Piola-Kirchhoff stress tensor}\label{chap2:PiolaStress1}
+We know that the Cauchy stress tensor represents the current force per unit of deformed area: $d\mathbf{f} = \mathbf{t}(\mathbf{n}) da = \sigma \cdot d\mathbf{a}$.
+We define a stress vector $\mathbf{T}$ over the area element $dA$ with normal $\mathbf{N}$ in the underformed configuration such that it results in the same total force
+\begin{equation}
+\label{chap2:df}
+d\mathbf{f} = \mathbf{t}(\mathbf{n}) da = \mathbf{T}(\mathbf{N}) dA.
+\end{equation} 
+The two stress vectors have the obviously the same direction but different magnitudes since they are not applied to the same area.
+
+In a similar way we defined the Cauchy stress $\sigma$ with $\mathbf{t}(\mathbf{n}) = \sigma \cdot \mathbf{n}$, we introduce a stress tensor $\mathbf{P}$ called the \emph{first Piola-Kirchhoff stress tensor} such that $\mathbf{T}(\mathbf{n}) = \mathbf{P} \cdot \mathbf{N}$.
+The first Piola-Kirchhoff stress tensor gives the current force per unit undeformed area.
+
+The relation between the first Piola-Kirchhoff stress tensor and the Cauchy stress tensor can be obtained as follows. From \eqref{chap2:df} we have:
+\begin{equation}
+\mathbf{T} = \frac{da}{dA} \mathbf{t}
+\end{equation}
+Let us assume the \emph{Nanson's relation} \TODO{demonstration in section \emph{Change of surface}?} expressed as:
+\begin{equation}
+da \mathbf{n} = J \, dA \, \mathbf{F}^{-T} \, \mathbf{N},
+\end{equation}
+where $da$ is an area of the deformed configuration, $dA$ an area from the reference configuration, $\mathbf{F}$ the deformation gradient and $J = \det \mathbf{F}$ the Jacobian of the transformation.
+Using $d\mathbf{f} = \mathbf{t} da$ and Nanson's relation we have:
+\begin{equation}
+\mathbf{T} =  J \, \mathbf{t} \, \mathbf{F}^{-T}
+\end{equation}
+Finally we can express the first Piola-Kirchhoff stress tensor $\mathbf{P}$ as
+\begin{equation}
+\mathbf{P} = J \, \sigma \, \mathbf{F}^{-T}.
+\end{equation}
+
+	\subsection{Second Piola-Kirchhoff stress tensor}\label{chap2:PiolaStress2}
+The \emph{second Piola-Kirchhoff stress tensor} $\mathbf{S}$ is introduced as the stress tensor associated with the force $d\mathbf{F}$ in the undeformed area $d\mathbf{A}$ that corresponds to the force $d\mathbf{f}$ in the deformed area $d\mathbf{a}$:
+\begin{equation}
+d\mathbf{F} = \mathbf{S} \, d\mathbf{A}
+\end{equation}
+Thus, the second Piola-Kirchhoff stress tensor gives the transformed current force per unit undeformed area.
+Similar to the relationship \eqref{chap2:dxdX} between $d\mathbf{x}$ and $d\mathbf{X}$, the force $d\mathbf{f}$ is related to the force $d\mathbf{F}$ by:
+\begin{align}
+d\mathbf{F} &= \mathbf{F}^{-1} \, d\mathbf{f} \notag \\
+&= \mathbf{F}^{-1} \, (\mathbf{P} \, d\mathbf{A}) \notag \\
+&= \mathbf{S} \, d\mathbf{A}
+\end{align}
+Therefore, the second Piola-Kirchhoff stress tensor is related to the first Piola-Kirchhoff stress tensor as follows:
+\begin{equation}
+\mathbf{S} = \mathbf{F}^{-1} \mathbf{P} = J \, \mathbf{F}^{-1} \, \sigma \, \mathbf{F}^{-T}
+\end{equation}
 
 \section{Constitutive equations}
+%After the introduction of strain tensors, we are now in possesion of different measures to assess the state of deformation in a continuous solid.
+
 	\subsection{Elastic solids}	

### 2010-07-26 Thesis/Chapter2.tex
- Update 26 July - Added constitutive models
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@134 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
+	\subsection{\TODO{Isochoric component of the deformation gradient?}}
+
 	\subsection{Change of volume}
@@
 We can define volume elements in the reference and deformed configurations.
-Consider three non-coplanar line elements $d\mathbf{X}^{(1)}$, $d\mathbf{X}^{(2)}$ and $d\mathbf{X}^{(3)}$ forming a parallelepiped in the reference configuration.
+Consider three non-coplanar line elements $d\mathbf{X}^{(1)}$, $d\mathbf{X}^{(2)}$ and $d\mathbf{X}^{(3)}$ forming a parallelepiped in the reference configuration \TODO{Add figure with all notations}.
 The three vectors after deformation $d\mathbf{x}^{(1)}$, $d\mathbf{x}^{(2)}$ and $d\mathbf{x}^{(3)}$ can be obtained with:
@@
 \begin{equation}
+\label{chap2:dvdV}
 dv = J dV.
@@
 
-	\subsection{\TODO{Change of surface?}}
+	\subsection{Change of surface}
+For similar reasons, let us find the relation between an element of area in the reference configuration $d\mathbf{A}$ which becomes $d\mathbf{a}$ in the deformed configuration.
+Considering an arbitrary material line $d\mathbf{L}$ in the reference configuration and noting $d\mathbf{l}$ the same line after deformation \TODO{Add figure with all notations}, the corresponding reference and current volumes are given respectively by:
+\begin{align}
+d\mathbf{V} &= d\mathbf{L} \cdot d\mathbf{A} \\
+d\mathbf{v} &= d\mathbf{l} \cdot d\mathbf{a}
+\end{align}
+Using \eqref{chap2:dvdV} which relates the reference and deformed volumes and recalling that $d\mathbf{l} = \mathbf{F} \, d\mathbf{L}$, we have:
+\begin{equation}
+J \, d\mathbf{L} \cdot d\mathbf{A} = (\mathbf{F} \, d\mathbf{L}) \cdot d\mathbf{a}.
+\end{equation}
+The fact that this expression is valid for any vector $d\mathbf{L}$ and using the property $\mathbf{a} \cdot \mathbf{b} = \mathbf{a} ^{T} \mathbf{b}$ yields:
+\begin{equation}
+\label{chap2:dadA}
+d\mathbf{a} = J \, \mathbf{F}^{-T} \, d\mathbf{A}.
+\end{equation}
 
@@
 Indeed, when the displacement gradients are small (that is, $|\nabla \mathbf{u}| \ll 1$) we can neglect the nonlinear terms in the definition.
-In the case of infinitesimal strains, the Green-Lagrange strain tensor and the Eulerian strain tensor are approximately the same and can be approximated by the infinitesimal strain tensor denoted $\boldsymbol \epsilon$ and is given by:
+In the case of infinitesimal strains, the Green-Lagrange strain tensor and the Eulerian strain tensor are approximately the same and can be approximated by the infinitesimal strain tensor denoted $\boldsymbol \varepsilon$ and is given by:
 \begin{equation}
-\boldsymbol \epsilon = \frac{1}{2} \left[ \nabla \mathbf{u} + (\nabla \mathbf{u})^T \right].
+\boldsymbol \varepsilon = \frac{1}{2} \left[ \nabla \mathbf{u} + (\nabla \mathbf{u})^T \right].
 \end{equation}
-Its Cartesian components $\epsilon_{ij}$ are the following:
+Its Cartesian components $\varepsilon_{ij}$ are the following:
 \begin{equation}
-\epsilon_{ij} = \frac{1}{2} (\pd{u_i}{X_j} + \pd{u_j}{X_i}).
+\varepsilon_{ij} = \frac{1}{2} (\pd{u_i}{X_j} + \pd{u_j}{X_i}).
 \end{equation}
@@
 	\begin{bmatrix}
-		\sigma_{11} && \sigma_{21} && \sigma_{31} \\
-		\sigma_{12} && \sigma_{22} && \sigma_{32} \\
-		\sigma_{13} && \sigma_{23} && \sigma_{33} \\
+		\sigma_{11} & \sigma_{21} & \sigma_{31} \\
+		\sigma_{12} & \sigma_{22} & \sigma_{32} \\
+		\sigma_{13} & \sigma_{23} & \sigma_{33} \\
 	\end{bmatrix}
@@
 \end{equation}
-Let us assume the \emph{Nanson's relation} \TODO{demonstration in section \emph{Change of surface}?} expressed as:
-\begin{equation}
-da \mathbf{n} = J \, dA \, \mathbf{F}^{-T} \, \mathbf{N},
-\end{equation}
-where $da$ is an area of the deformed configuration, $dA$ an area from the reference configuration, $\mathbf{F}$ the deformation gradient and $J = \det \mathbf{F}$ the Jacobian of the transformation.
-Using $d\mathbf{f} = \mathbf{t} da$ and Nanson's relation we have:
+Using $d\mathbf{f} = \mathbf{t} da$ and the relation \eqref{chap2:dadA} between $d\mathbf{A}$ and $d\mathbf{a}$, we have:
 \begin{equation}
@@
 \section{Constitutive equations}
-%After the introduction of strain tensors, we are now in possesion of different measures to assess the state of deformation in a continuous solid.
+So far, all the relations that we have established are valid for every continuum.
+Indeed, no mention was made of any material in the derivations.
+Therefore these relations are not sufficient to describe the response of a material to a given loading.
+We know that the stresses result from the deformation of the material and it is now necessary to express them in terms of some measure of this deformation such as, for instance, the strain.
+These relationships, known as \emph{constitutive equations}, obviously depend on the type of material under consideration and may be dependent on time or not.
+Constitutive equations are not derived from any physical principles, although they are subject to obeying certain rules and the entropy inequality.
+In essence, constitutive equations are mathematical models of the behaviour of materials that are validated against experimental results.
 
-	\subsection{Elastic solids}	
-	\subsection{Generalised Hooke's law}
+The relationship between stress and strain depends on the nature, or constitution of the particular material that we are studying.
+Of course, any material mechanics, and especially tissue mechanics, can be quite complex, and many assumptions must be made when deriving a constitutive equation.
+It is basically impossible to derive a particular constitutive equation that would accurately model all aspects of tissue behaviour under any type of loading.
+This is because tissues may behave very differently under small loads as compared to large loads.
+For instance, we may expect damage under large loads.
+Also, under different loading rates different tissues may behave differently.
+Therefore, whenever we develop a constitutive equation to model a tissue, we need to balance the need to accurately model tissue behaviour under the range of loadings with the need to have a constitutive equation that is simple enough to:
+\begin{enumerate}
+\item be computed in a numerical model 
+\item experimentally measure all the material constants used in the constitutive equation
+\end{enumerate}
+This section will now introduce the most common constitutive models relevant to soft-tissue modelling.
+
+	\subsection{Elasticity}	
+If under applied loads a material stores but does not dissipate energy, and it returns to its orginal shape when the loads are removed, we call that material elastic.
+The work done by the stress is generally dependent on the deformation path.
+However, for elastic materials the stress does not depend on the path of deformation and therefore the state of stress in the deformed configuration is determined only by the state of deformation.
+It means that the stress does not depend on the time taken to achieve the deformation or the rate at which the state of deformation is reached.
+
+One special case of elastic materials are \emph{hyperelastic} materials. A material is said to be hyperelastic if there exists a \emph{strain energy density} function $U_0(\varepsilon)$ such that
+\begin{equation}
+\label{chap2:energy}
+\boldsymbol \sigma = \pd{U_0}{\varepsilon}.
+\end{equation}
+In order to develop a mathematical model of an hyperelastic material, $U_0$ is expanded in Taylor's series about $\varepsilon = 0$:
+\begin{equation}
+U_0 = C_0 + C_{ij}\varepsilon_{ij} + \frac{1}{2!}\hat{C}_{ijkl}\varepsilon{ij}\varepsilon_{kl} + \frac{1}{3!}\hat{C}_{ijklmn}\varepsilon_{ij}\varepsilon_{kl}\varepsilon_{mn} + \ldots , 
+\end{equation}
+where $C_0$, $C_{ij}$ and $\hat{C}$ are material stiffnesses.
+The form of $U_0$ may varied for different materials.
+A material is said to be \emph{linear} if the relationship between strain and stress is also linear.
+Based on this definition, $U_0$ is a cubic or higher-order function of the strains for nonlinear materials whereas $U_0$ is a quadratic function of strain for linear materials.
+We will start by discussing linear materials for the case of infinitesimal deformations (sections \ref{chap2:Hooke} to \ref{chap2:isotropic}).
+Consequently, we will not make any distinction between the various measures of stress and strain and use $\boldsymbol \sigma$ and $\boldsymbol \varepsilon$ for stress and strain tensors respectively.
+
+	\subsection{Linear materials: generalised Hooke's law} \label{chap2:Hooke}
+The linear constitutive model for infinitesimal deformations is called the generalised Hooke's law.
+To derive the relation between stress and strain for a linear material, we assume the quadratic form of $U_0$:
+\begin{equation}
+U_0 = C_0 + C_{ij}\varepsilon_{ij} + \frac{1}{2!}\hat{C}_{ijkl}\varepsilon{ij}\varepsilon_{kl}
+\end{equation}
+From there and using \eqref{chap2:energy} it can be shown that there exists some coefficients $C_{ijkl}$ such as:
+\begin{equation}
+\sigma_{ij} = C_{ijkl} \varepsilon_{kl}
+\end{equation}
+The $C_{ijkl}$ are the coefficients of the fourth-order tensor $\mathbf{C}$ that relates the second-order tensors of strain and stress.
+The $81$ scalar components of this fourth-order tensor ($=3^4$) can be reduced to $21$ independent coefficients using existing symmetries.
+Recalling that both strain and stress tensors are symmetric yields:
+\begin{equation}
+	\begin{bmatrix}
+		\sigma_{11} \\
+		\sigma_{22} \\
+		\sigma_{33} \\
+		\sigma_{23} \\
+		\sigma_{13} \\
+		\sigma_{12}
+	\end{bmatrix}
+= 
+	\begin{bmatrix}
+		C_{1111} & C_{1122} & C_{1133} & C_{1123} & C_{1113} & C_{1112} \\
+		C_{1122} & C_{2222} & C_{2233} & C_{2223} & C_{2213} & C_{2212} \\
+		C_{1133} & C_{2233} & C_{3333} & C_{3323} & C_{3313} & C_{3312} \\
+		C_{1123} & C_{2223} & C_{3323} & C_{2323} & C_{2313} & C_{2312} \\
+		C_{1113} & C_{2213} & C_{3313} & C_{2313} & C_{1313} & C_{1312} \\
+		C_{1112} & C_{2212} & C_{3312} & C_{2312} & C_{1312} & C_{1212}
+	\end{bmatrix}
+	\begin{bmatrix}
+		\varepsilon_{11} \\
+		\varepsilon_{22} \\
+		\varepsilon_{33} \\
+		\varepsilon_{23} \\
+		\varepsilon_{13} \\
+		\varepsilon_{12}
+	\end{bmatrix}
+	.
+\end{equation}
+
 	\subsection{Orthotropic materials}
-	\subsection{Isotropic materials}
+Further reduction in the number of independent stiffness parameters can be achieved using the so-called material symmetry.
+This symmetry is due to the internal structure of the material, that is the arrangement of molecules (crystallographic form).
+We note that the symmetry under discussion is a directional property and not a positional property.
+Thus, a material may have certain elastic symmetry at every point of a material body and the properties may vary from point to point.
+Positional dependence of material properties is what we called the inhomogeneity of the material.
+
+\emph{Orthotropic} materials are characterised by three mutually orthogonal planes of material symmetry, that is, three mutually orthogonal preferred directions.
+In this case, the number of coefficients is reduced to 9 and the stress-strain relation for an orthotropic material is the following:
+\begin{equation}
+\label{chap2:orthotropic}
+	\begin{bmatrix}
+		\sigma_{11} \\
+		\sigma_{22} \\
+		\sigma_{33} \\
+		\sigma_{23} \\
+		\sigma_{13} \\
+		\sigma_{12}
+	\end{bmatrix}
+= 
+	\begin{bmatrix}
+		C_{1111} & C_{1122} & C_{1133} &       0       &       0       &       0 \\
+		C_{1122} & C_{2222} & C_{2233} &       0       &       0       &       0 \\
+		C_{1133} & C_{2233} & C_{3333} &       0       &       0       &       0 \\
+		      0       &        0     &        0       & C_{2323} &       0       &       0 \\
+		      0       &        0     &        0       &       0       & C_{1313} &       0 \\
+		      0       &        0     &        0       &       0       &       0       & C_{1212}
+	\end{bmatrix}
+	\begin{bmatrix}
+		\varepsilon_{11} \\
+		\varepsilon_{22} \\
+		\varepsilon_{33} \\
+		\varepsilon_{23} \\
+		\varepsilon_{13} \\
+		\varepsilon_{12}
+	\end{bmatrix}
+	.
+\end{equation}
+
+Most often, the material properties are determined in a laboratory in terms of the engineering constants such as Young?s modulus, shear modulus, and so on.
+These constants are measured using simple tests like uniaxial tension test or pure shear test (see section \ref{chap2:estimationParameters}).
+Because of their direct and obvious physical meaning, engineering constants are used in place of the more abstract stiffness coefficients $C_{ijkl}$.
+Therefore, we will now reformulate \eqref{chap2:orthotropic} with respect to those engineering constants.
+Nothing $E_i$ the Young's modulus in the direction $x_i$ and $G_{ij}$ the shear moduli in the $x_i-x_j$ plane, it can be shown that:
+
+%Since we consider infinitesimal deformations and we also linearised the constitutive material law between stress and strain, the principle of superpositions applies.
+It means that the sum of the displacements (and hence strains) produced by two sets of loads is equal to the displacements (and strains) produced by the sum of the two sets of loads.
+For example, the extensional strain $\varepsilon_{11}^{(1)}$ in the  coordinate direction $x_1$ due to the stress $\sigma_{11}$ in the same direction is $\sigma_{11}/E_1$, where $E_1$ denotes Young?s modulus of the material in the $x_1$ direction.
+The extensional strain $\varepsilon_{11}^{(2)}$ due to the stress $\sigma_{22}$ applied in the direction $x_2$ (as a result of the Poisson's effect) is $-\nu_{21} (\sigma_{22}/E_2)$, where $\nu_{21}$ is the Poisson's ratio:
+%\begin{equation}
+%\nu_{21} = - \frac{\varepsilon_{11}}{\varepsilon_{22}}
+%\end{equation} 
+%and  $E_2$ denotes Young?s modulus of the material in the $x_2$ direction.
+Similarly, $\sigma_{33}$ produces a strain $\varepsilon_{11}^{(3)}$ valued at $-\nu_{31} (\sigma_{33}/E_3)$.
+The total strain $\varepsilon_{11}$ due to the simulatneous application of all three normal stress components is:
+%\begin{equation}
+%\varepsilon_{11} = \varepsilon_{11}^{(1)} + \varepsilon_{11}^{(2)} + \varepsilon_{11}^{(3)} = \frac{\sigma_{11}}{E_1} - \nu_{21} \, \frac{\sigma_{22}}{E_2}  - \nu_{31} \, \frac{\sigma_{33}}{E_3}.
+%\end{equation}
+
+\begin{equation}
+\label{chap2:orthotropic}
+	\begin{bmatrix}
+		\varepsilon_{11} \\
+		\varepsilon_{22} \\
+		\varepsilon_{33} \\
+		\varepsilon_{23} \\
+		\varepsilon_{13} \\
+		\varepsilon_{12}
+	\end{bmatrix}
+= 
+	\begin{bmatrix}
+		\frac{1}{E_1} & -\frac{\nu_{21}}{E_2} & -\frac{\nu_{31}}{E_3} &       0       &       0       &       0 \\
+		-\frac{\nu_{12}}{E_1} & \frac{1}{E_2} & -\frac{\nu_{32}}{E_3} &       0       &       0       &       0 \\
+		-\frac{\nu_{13}}{E_1} & -\frac{\nu_{23}}{E_2} & \frac{1}{E_3} &       0       &       0       &       0 \\
+		      0       &        0     &        0       & \frac{1}{G_{23}} &       0       &       0 \\
+		      0       &        0     &        0       &       0       & \frac{1}{G_{13}}&       0 \\
+		      0       &        0     &        0       &       0       &       0       & \frac{1}{G_{12}}
+	\end{bmatrix}
+	\begin{bmatrix}
+		\sigma_{11} \\
+		\sigma_{22} \\
+		\sigma_{33} \\
+		\sigma_{23} \\
+		\sigma_{13} \\
+		\sigma_{12}
+	\end{bmatrix}
+	.
+\end{equation}
+
+	\subsection{Isotropic materials} \label{chap2:isotropic}
+A material is said \emph{isotropic} if its properties are independent of the directions. Therefore, we have the following relations:
+\begin{equation}
+E_1 = E_2 = E_3 = E, \quad G_{12} = G_{13} = G_{23} = G, \quad \nu_{12} = \nu_{23} = \nu_{13} = \nu.
+\end{equation}
+The stress-strain relations may be written:
+\begin{equation}
+\sigma_{ij} = \frac{E}{1+\nu} \varepsilon_{ij} + \frac{\nu E}{(1+\nu)(1-2\nu)} \varepsilon_{kk}\delta_{ij}.
+\end{equation}	
+where the summation of repeated indices is implied (Einstein's notation). The inverse relations are:
+\begin{equation}
+\varepsilon_{ij} = \frac{1+\nu}{E} \sigma_{ij} - \frac{\nu}{E} \sigma_{kk} \delta_{ij},
+\end{equation}
+recalling that $G = E/[2(1+\nu)]$.
+
+	\subsection{Non-linear materials}
 
@@
 	\subsection{Material models for organs (non-linear, visco-elastic and anisotropic)}
-	\subsection{Measure/estimation of model parameters}
+	\subsection{Measure/estimation of model parameters} \label{chap2:estimationParameters}

### 2010-07-27 Thesis/Chapter2.tex
- Update 27 July - Added 'volumetric and isochoric components', 'Principal strains and invariants', 'Principal stresses and invariants', 'Non-linear materials', 'Viscoelastic materials' and 'Tissue characterisation'
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@135 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-	\subsection{\TODO{Isochoric component of the deformation gradient?}}
-
 	\subsection{Change of volume}
@@
 d\mathbf{a} = J \, \mathbf{F}^{-T} \, d\mathbf{A}.
+\end{equation}
+
+	\subsection{Volumetric and isochoric components}
+We recall that the Jacobian $J$ has the physical meaning of being the local ratio of current to reference volume of a material volume element.
+Therefore, if $J = 1$ the volume does not change locally during the deformation and the latter is qualified as \emph{isochoric} at this given particule $P$.
+If $J = 1$ everywhere in the body, the deformation of the body is isochoric.
+
+When dealing with incompressible and nearly incompressible materials it is necessary to separate the volumetric from the isochoric components of the deformation.
+Such a separation must ensure that the isochoric component $\mathbf{\hat{F}}$ does not imply any change in volume.
+This condition can be achieved by choosing $\mathbf{\hat{F}}$ as:
+\begin{equation}
+\mathbf{\hat{F}} = J^{-1/3} \mathbf{F}.
+\end{equation}
+Indeed,
+\begin{align}
+\det \mathbf{\hat{F}} &= \det (J^{-1/3} \mathbf{F}) \notag \\
+&= (J^{-1/3})^3 \det \mathbf{F} \notag \\
+&= J^{-1} J \notag \\
+&= 1
+\end{align}
+The deformation gradient $\mathbf{F}$ can now be expressed in terms of the volumetric and isochoric component $J$ and $\mathbf{\hat{F}}$ as:
+\begin{equation}
+\mathbf{F} = J^{1/3} \mathbf{\hat{F}}
 \end{equation}
@@
 
-	\subsection{\TODO{Principal strains?}}
+	\subsection{Principal strains and invariants}	\label{chap2:strainInvariants}
+The components $\varepsilon_{ij}$ of the strain tensor depend on of the coordinate system at the point under consideration.
+However, the strain tensor itself is a physical quantity and as such, it is independent of the coordinate system chosen to represent it.
+Certain operations on strain tensors give the same result independently of the coordinate system chosen to represent the components of strain.
+As an example, a vector is a simple tensor of rank one.
+In three dimensions, it has three components.
+The value of these components will depend on the coordinate system chosen to represent the vector, but the length of the vector is a physical quantity (a scalar) and is independent of the coordinate system chosen to represent the vector.
+Similarly, every second rank tensor (such as the strain tensors) has three independent invariant quantities associated with it.
+One set of such \emph{invariants} are the principal strains of the strain tensor, which are just the eigenvalues of the strain tensor.
+Because they are independent of any coordinate system, they are very convenient to define strain energy density functions (see section \ref{chap2:non-linear}).
+The most commonly used invariants are:
+\begin{align}
+I &= tr(\varepsilon) \notag \\
+II &= \frac{1}{2} \left\lbrace tr(\varepsilon^{2}) - \left[tr(\varepsilon)^{2} \right] \right\rbrace \\
+III &= det(\varepsilon). \notag
+\end{align}
+
+It can be shown that in the coordinate system $(\mathbf{p}_1, \mathbf{p}_2, \mathbf{p}_3)$ formed by the three eigenvectors, the expression of the strain tensor is:
+\begin{equation}
+\boldsymbol \varepsilon =
+	\begin{bmatrix}
+	\varepsilon_1 &          0           &           0           \\
+	         0           & \varepsilon_2 &           0           \\
+	         0           &           0           & \varepsilon_3
+	\end{bmatrix}	
+\end{equation}
+Since there are no shear strain components in this particular coordinate system, the principal strains represent the maximum and minimum stretches of an elemental volume.
+Because of the obvious simplicity of the strain tensor's expression, this coordinate system is often used in mechanics to derive and solve equations.
 
@@
 In some cases, it is possible to simplify the expression of the Green strain tensor defined in \eqref{chap2:GreenTensor}.
-Indeed, when the displacement gradients are small (that is, $|\nabla \mathbf{u}| \ll 1$) we can neglect the nonlinear terms in the definition.
+Indeed, when the displacement gradients are small (that is, $|\nabla \mathbf{u}| \ll 1$) we can neglect the non-linear terms in the definition.
 In the case of infinitesimal strains, the Green-Lagrange strain tensor and the Eulerian strain tensor are approximately the same and can be approximated by the infinitesimal strain tensor denoted $\boldsymbol \varepsilon$ and is given by:
@@
 
-The strain quantities defined in the previous sections are nonlinear expressions in terms of the mapping $\chi$ and will lead to nonlinear governing equations.
+The strain quantities defined in the previous sections are non-linear expressions in terms of the mapping $\chi$ and will lead to non-linear governing equations.
 In solid mechanics, whenever the hypothesis is acceptable, it is common practice to assume that the displacements are small and the infinitesimal strain tensor is then used as a measure of the deformation.
@@
 
+		\subsection{Principal stresses and invariants}
+In the same way that we defined principal strains in section \ref{chap2:strainInvariants} and like any second-rank tensor, the stress tensor has three invariants and we can find a coordinate system $(\mathbf{n}_1, \mathbf{n}_2, \mathbf{n}_3)$ in which the expression of the stress tensor is:
+\begin{equation}
+\boldsymbol \sigma =
+	\begin{bmatrix}
+	\sigma_1 &         0      &        0       \\
+	        0      & \sigma_2 &        0       \\
+	        0      &          0     & \sigma_3
+	\end{bmatrix}	
+\end{equation}
+This result means that at every point in a stressed body there are at least three planes called \emph{principal planes}, with normal vectors $\mathbf{n}_1$, $\mathbf{n}_2$ and $\mathbf{n}_3$ called \emph{principal directions} (in fact, the eigenvectors), where the corresponding stress vector is perpendicular to the plane and where there are no normal shear stresses.
+The three stresses normal $\sigma_1$, $\sigma_2$ and $\sigma_3$ to these principal planes are called \emph{principal stresses}.
+
 \section{Constitutive equations}
@@
 \end{equation}
-In order to develop a mathematical model of an hyperelastic material, $U_0$ is expanded in Taylor's series about $\varepsilon = 0$:
+However, for \emph{imcompressible} materials the stress tensor is not completely determined by the deformation.
+The hydrostatic pressure\footnote{pressure exerted by a fluid at equilibrium due to the force of gravity} affects the stress and the above equation is then written as:
+\begin{equation}
+\label{chap2:energy2}
+\boldsymbol \sigma = -p \mathbf{I} + \pd{U_0}{\varepsilon},
+\end{equation}
+where $p$ is the hydrostatic pressure. In order to develop a mathematical model of an hyperelastic material, $U_0$ is expanded in Taylor's series about $\varepsilon = 0$:
 \begin{equation}
@@
 A material is said to be \emph{linear} if the relationship between strain and stress is also linear.
-Based on this definition, $U_0$ is a cubic or higher-order function of the strains for nonlinear materials whereas $U_0$ is a quadratic function of strain for linear materials.
+Based on this definition, $U_0$ is a cubic or higher-order function of the strains for non-linear materials whereas $U_0$ is a quadratic function of strain for linear materials.
 We will start by discussing linear materials for the case of infinitesimal deformations (sections \ref{chap2:Hooke} to \ref{chap2:isotropic}).
@@
 
-	\subsection{Orthotropic materials}
+	\subsection{Orthotropic materials} \label{chap2:orthotropic}
 Further reduction in the number of independent stiffness parameters can be achieved using the so-called material symmetry.
@@
 
-Most often, the material properties are determined in a laboratory in terms of the engineering constants such as Young?s modulus, shear modulus, and so on.
-These constants are measured using simple tests like uniaxial tension test or pure shear test (see section \ref{chap2:estimationParameters}).
+Most often, the material properties are determined in a laboratory in terms of the engineering constants such as Young's modulus, shear modulus, and so on.
+These constants are measured using simple tests like uniaxial tension test or pure shear test.
 Because of their direct and obvious physical meaning, engineering constants are used in place of the more abstract stiffness coefficients $C_{ijkl}$.
@@
 
-	\subsection{Non-linear materials}
+	\subsection{Non-linear materials} \label{chap2:non-linear}
+For most materials, the linear relation between stress and strain is only valid for small deformation.
+Beyond a certain threshold, Hooke's law is no longer valid.
+Note that it does mean that the material is not elastic, it may well recover all its deformation after removal of the loads.
+However, materials often exibit an elastic threshold in the non-linear range after which permanent deformation occurs.
+This range of deformation is qualified of \emph{plastic}.
+
+The strain energy density function takes different forms for different materials.
+It is often expressed as a linar combination of unknown coefficients (determined experimentally) and principal invariants of Green strain tensor $\mathbf{E}$, left Cauchy-Green strain tensor $\mathbf{B}$ and deformation gradient tensor $\mathbf{F}$ (see section \ref{chap2:strainInvariants} for more information on invariants).
+The constitutive equations of  two well-known non-linear elastic materials (Mooney-Rivlin and neo-Hookean) will now be introduced.
+
+	\subsubsection*{Mooney-Rivlin}
+The Mooney-Rivlin model was proposed by Melvin Mooney and Ronald Rivlin in two independent papers in 1952.
+For an incompressible Mooney-Rivlin material, the strain energy density function $U_0$ is taken as a linear function of the principal invariants of the left Cauchy-Green strain tensor $\mathbf{B}$ and is written as:
+\begin{equation}
+U_0 = C_1 (I_B-3) + C_2(II_B-3),
+\end{equation}
+where $C_1$ and $C_2$ are two constants and $I_B$ and $II_B$ are the two principal invariants of $\mathbf{B}$. For those materials, it can be shown that the stress tensor has the form
+\begin{equation}
+\sigma = -p \mathbf{I} + \alpha \mathbf{B} + \beta \mathbf{B}^{-1},
+\end{equation}
+where $\alpha$ and $\beta$ are given by
+\begin{equation}
+\alpha = 2 \rho \pd{U_0}{I_B} = 2 \rho C_1, \quad \beta = -2\rho \pd{U_0}{II_B} = -2\rho C_2 \quad \mbox{with } \rho \, \mbox{the material density.}
+\end{equation}
+The Mooney-Rivlin incompressible model is often used to represent the behaviour of rubber-like materials.
+
+	\subsubsection*{Neo-Hookean}			
+The neo-Hookean model was proposed by Ronald Rivlin in 1948.
+It can be derived from the Mooney-Rivlin model in the special case where $C_2 = 0$ in the strain energy density function.
+Therefore, we have:
+\begin{equation}
+U_0 = C_1 (I_B-3)
+\end{equation}
+and the constitutive equation for the stress may be written as:
+\begin{equation}
+\sigma = -p \mathbf{I} + 2 \rho C_1 \mathbf{B}
+\end{equation}
+The neo-Hookean model provides a reasonable approximation of rubber-like material for moderate strains. It is typically accurate for strains less than $20\, \%$ \citep{Gent01}.
+
+	\subsection{Viscoelastic materials}
+Viscoelasticity is the property of materials that exhibit both viscous and elastic characteristics when undergoing deformation.
+Viscous materials, like honey, resist shear flow and strain linearly with time when a stress is applied.
+Elastic materials strain instantaneously when stretched and just as quickly return to their original state once the stress is removed.
+Viscoelastic materials have elements of both of these properties and, as such, exhibit time dependent strain.
+Depending on the change of strain rate versus stress inside a material the viscosity can be categorised as having a linear or non-linear response.
+When a material exhibits a linear response it is categorized as a Newtonian material.
+In this case the stress is linearly proportional to the strain rate.
+If the material exhibits a non-linear response to the strain rate, it is categorised as a Non-Newtonian material.
+
+Viscoelastic materials present a few characteristics: 
+\begin{enumerate}
+\item if the stress is held constant, the strain increases with time (creep)
+\item if the strain is held constant, the stress decreases with time (relaxation)
+\item the effective stiffness depends on the rate of application of the load
+\item if cyclic loading is applied, hysteresis occurs, leading to a dissipation of mechanical energy
+\end{enumerate}
+
+\TODO{Add details on constitutive equations? Level of details to be determined later (depending on TLED needs)}
 
 \section{Tissue characterisation}
-	\subsection{Material models for organs (non-linear, visco-elastic and anisotropic)}
-	\subsection{Measure/estimation of model parameters} \label{chap2:estimationParameters}
+One of the main obstacles in developing realistic organ models is the lack of data on the material properties of live organ tissues.
+Measuring and characterising in vivo organ properties is a highly challenging task, but is a requirement for realistic organ modelling.
+Indeed, as we have seen in sections \ref{chap2:orthotropic} and \ref{chap2:non-linear}, constitutive equations require the experimental determination of material constants.
+Organ models with incorrect material properties will affect training in surgical simulator systems \citep{Sedef06}.
+
+The constitutive equations are used to define idealised materials, which represent different aspects of the mechanical behaviour of natural materials.
+However, no real material is known to behave exactly as one of these mathematical models.
+In particular, few biological tissues obey Hooke's law.
+If because of their high stiffness (and therefore small deformations) cortical bones are accurately described by a linear model, Hooke's law becomes insufficient to describe the mechanical behaviour of the liver with precision for instance.
+In fact, in vivo experiments conducted by \cite{Melvin73} suggest that mechanical behaviour of liver is non-linear and viscoelastic.
+The brain also appears to be non-linear and viscoelastic \citep{Miller97}.
+The liver has the additional property of anisotropy, more precisely, it is transversely isotropic according to \cite{Chui07}.
+
+\TODO{Transition with next chapter (complexity of models => FEM to solve them)}

### 2010-07-28 Thesis/Chapter2.tex
- Update 28 July - Read chapter 2 and corrected spellings, transitions etc. + bib updated
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@136 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{shortAbstract}
-As seen in the previous chapter, realistic modelling of organs' deformation is a challenging research field that opens the door to new clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedure and per-operative guidance based on simulation.
+As seen in the previous chapter, realistic modelling of organ deformation is a challenging research field that opens the door to new clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedure and per-operative guidance based on simulation.
 In all these cases the clinician needs fast updates of the deformation model to obtain a real-time display of the computed deformations.
 If for medical training devices the haptic feedback from touching organs merely needs to feel real, the accuracy of the information provided to the clinician in the cases of planning or per-operative guidance is crucial.
-Therefore a substantial comprehension of the mechanics involved and a knowledge of physical properties of anatomical structures are both mandatory in our quest to realistically model the deformation of organs.
+Therefore, a substantial comprehension of the mechanics involved and a knowledge of the physical properties of anatomical structures are both mandatory in our quest to realistically model the deformation of organs.
 This chapter will start by introducing the main concepts of continuum mechanics that are fundamental to study the mechanical response of organs.
-It will then present mathematical models able to describe the different mechanical aspects of materials and briefly expose their experimental validation.
+It will then present mathematical models able to describe the different mechanical aspects of materials and briefly discuss the mechanical characterisation of tissues.
 \end{shortAbstract}
@@
 The concept that matter is composed of discrete units has been around for millennia.
-In fact we know with certainty that our world is composed of microscopic atoms and molecules separated by empty space since the beginning of the twentieth century \citep{Lautrup05}.
+In fact, we know with certainty that our world is composed of microscopic atoms and molecules separated by empty space since the beginning of the twentieth century \citep{Lautrup05}.
 However, certain physical phenomena can be predicted with theories that pay no attention to the molecular structure of materials.
-Consider for instance the deformation of the horizontal board of a bookshelf under the weight of the books.
+Consider for instance the deformation of the horizontal board of a bookshelf under the weight of books.
 The bending of the shelf can be modelled without considering its molecular composition.
@@
 Continuum mechanics can be divided into two main parts: general principles common to all media (analysis of deformation, strain and stress concepts) and constitutive equations defining idealised materials.
-This chapter will not only deal with those two aspects but will also introduce experiments carried out on organs in order to assess the physical parameters used in the mathematical models.
-This chapter will follow the notation used by \cite{Bonnet97} and \cite{Reddy07}.
+This chapter will not only deal with those two aspects but will also discuss the different aspects of the mechanical behaviour that we find in anatomical structures.
+Note that this chapter will mostly follow the notation used by \cite{Bonnet97} and \cite{Reddy07}.
 The interested reader may refer to these books for more details.
+%This chapter will not only deal with those two aspects but will also introduce experiments carried out on organs in order to assess the physical parameters used in the mathematical models.
 
 \section{Description of motion}
-Let us consider a body $\boldsymbol \beta$ of known geometry in a three-dimensional Euclidian space $\field{R}^3$.
-For a given geometry and loading, $\boldsymbol \beta$ will undergo a set of macroscopic changes which is called \emph{deformation}.
+Let us consider a body $\textbf{\textit{B}}$ of known geometry in a three-dimensional Euclidian space $\field{R}^3$.
+For a given geometry and loading, $\textbf{\textit{B}}$ will undergo a set of macroscopic changes which is called \emph{deformation}.
 The region of space occupied by the body at a given time $t$ is termed a \emph{configuration}.
@@
 \label{chap2:mapping}
-\chi : \boldsymbol \beta_{\kappa_{0}} \rightarrow \boldsymbol \beta_{\kappa}
+\chi : \textbf{\textit{B}}_{\kappa_{0}} \rightarrow \textbf{\textit{B}}_{\kappa}
 \end{equation} 
-is called the \emph{deforming mapping} of the body $\boldsymbol \beta$ from $\kappa_{0}$ to $\kappa$.
+is called the \emph{deforming mapping} of the body $\textbf{\textit{B}}$ from $\kappa_{0}$ to $\kappa$.
 When analysing the deformation of a continuous body, it is necessary to describe the evolution of configurations through time.
@@
 \end{equation}
-For a fixed value of $\mathbf{x} \in \kappa$, $\phi(\mathbf{x},t)$ gives the value of $\phi$ associated with different material particules at different times, because different particules occupy the position $\textbf{x} \in \kappa$ at different times.
-A change in time $t$ implies that a different value $\phi$ is observed at the same spatial location $\mathbf{x} \in \kappa$, now probably occupied by a different particule.
-Hence, the Eulerian description is focused on a spatial position.
+For a fixed value of $\mathbf{x} \in \kappa$, $\phi(\mathbf{x},t)$ gives the value of $\phi$ associated with the particule occupying the position $\textbf{x} \in \kappa$, which may very well be a different particule for each new time $ t $.
+Because a change in time $t$ implies that a different value $\phi$ is observed at the same spatial location $\mathbf{x} \in \kappa$, now probably occupied by a different particule, the Eulerian description is focused on a spatial position.
 This approach is convenient for the study of fluid flow where the kinematic property of greatest interest is the rate at which change is taking place rather than the shape of the body of fluid at a reference time \citep{Spencer80}.
@@
 	\subsection{Displacement field}
-The deplacement of a particule $X$ is called \emph{displacement vector} and its expression in the Lagrangian description is the following:
+The displacement $ \mathbf{u} $ of a particule $X$ is called \emph{displacement vector} and its expression in the Lagrangian description is given by the following:
 \begin{equation}
-\textbf{u}(\textbf{X}, t) = \textbf{x}(\textbf{X}, t) - \textbf{X}.
+\mathbf{u}(\textbf{X}, t) = \mathbf{x}(\mathbf{X}, t) - \mathbf{X}.
 \end{equation}
@@
 \end{equation}
-After deformation the two particules have deformed to their current positions given by the mapping $\chi$~\eqref{chap2:mapping} as:
+After deformation, the two particules have deformed to their current positions given by the mapping $\chi$~\eqref{chap2:mapping} as:
 \begin{equation}
@@
 Therefore, although we know how to transform vectors from the reference configuration into vectors in the current configuration using the deformation gradient, it is more useful to find a measure of the change in length of $d\mathbf{X}$.
-Many measures of strains can be defined and the most common ones are now introduced.
+Many measures of strains can be defined and the most common ones will now be introduced.
 
 	\subsection{Cauchy-Green deformation tensors}
-Consider two particules $P_{1}$ and $P_{2}$ in the neighbourhood of each other, separated by $d\mathbf{X}$ in the reference configuration.
+Let us consider two particules $P_{1}$ and $P_{2}$ in the neighbourhood of each other, separated by $d\mathbf{X}$ in the reference configuration.
 In the deformed configuration $P_{1}$ and $P_{2}$ occupy the positions $\tilde{P}_{1}$ and $\tilde{P}_{2}$ and they are separated by $d\mathbf{x}$.
@@
 \end{align}
-Let us define the \emph{Green-St. Venant strain tensor} $\mathbf{E}$ as:
+Let us define the \emph{Green-St.\,Venant strain tensor} $\mathbf{E}$ as:
 \begin{equation}
@@
 	\subsection{Principal strains and invariants}	\label{chap2:strainInvariants}
-The components $\varepsilon_{ij}$ of the strain tensor depend on of the coordinate system at the point under consideration.
+The components $\varepsilon_{ij}$ of the strain tensor depend on the coordinate system at the point under consideration.
 However, the strain tensor itself is a physical quantity and as such, it is independent of the coordinate system chosen to represent it.
@@
 \begin{align}
-I &= tr(\varepsilon) \notag \\
-II &= \frac{1}{2} \left\lbrace tr(\varepsilon^{2}) - \left[tr(\varepsilon)^{2} \right] \right\rbrace \\
-III &= det(\varepsilon). \notag
+I_1 &= tr(\varepsilon) \notag \\
+I_2 &= \frac{1}{2} \left\lbrace tr(\varepsilon^{2}) - \left[tr(\varepsilon)\right]^{2} \right\rbrace \\
+I_3 &= det(\varepsilon). \notag
 \end{align}
@@
 		\subsection{Principal stresses and invariants}
-In the same way that we defined principal strains in section \ref{chap2:strainInvariants} and like any second-rank tensor, the stress tensor has three invariants and we can find a coordinate system $(\mathbf{n}_1, \mathbf{n}_2, \mathbf{n}_3)$ in which the expression of the stress tensor is:
+In the same way that we defined principal strains in section \ref{chap2:strainInvariants} and like any second-rank tensor, the stress tensor has at least three invariants and we can find a coordinate system $(\mathbf{n}_1, \mathbf{n}_2, \mathbf{n}_3)$ in which the expression of the stress tensor is:
 \begin{equation}
@@
 Indeed, no mention was made of any material in the derivations.
-Therefore these relations are not sufficient to describe the response of a material to a given loading.
+Therefore, these relations are not sufficient to describe the response of a material to a given loading.
 We know that the stresses result from the deformation of the material and it is now necessary to express them in terms of some measure of this deformation such as, for instance, the strain.
@@
 	\subsection{Elasticity}	
-If under applied loads a material stores but does not dissipate energy, and it returns to its orginal shape when the loads are removed, we call that material elastic.
+If under applied loads a material stores but does not dissipate energy, and it returns to its orginal shape when the loads are removed, we call that material \emph{elastic}.
 The work done by the stress is generally dependent on the deformation path.
@@
 Therefore, we will now reformulate \eqref{chap2:orthotropic} with respect to those engineering constants.
-Nothing $E_i$ the Young's modulus in the direction $x_i$ and $G_{ij}$ the shear moduli in the $x_i-x_j$ plane, it can be shown that:
+Noting $E_i$ the Young's modulus in the direction $x_i$ and $G_{ij}$ the shear moduli in the $x_i-x_j$ plane, it can be shown that:
 
@@
 \begin{equation}
-\label{chap2:orthotropic}
+\label{chap2:orthotropic2}
 	\begin{bmatrix}
@@
 	\begin{bmatrix}
-		\frac{1}{E_1} & -\frac{\nu_{21}}{E_2} & -\frac{\nu_{31}}{E_3} &       0       &       0       &       0 \\
-		-\frac{\nu_{12}}{E_1} & \frac{1}{E_2} & -\frac{\nu_{32}}{E_3} &       0       &       0       &       0 \\
-		-\frac{\nu_{13}}{E_1} & -\frac{\nu_{23}}{E_2} & \frac{1}{E_3} &       0       &       0       &       0 \\
-		      0       &        0     &        0       & \frac{1}{G_{23}} &       0       &       0 \\
-		      0       &        0     &        0       &       0       & \frac{1}{G_{13}}&       0 \\
-		      0       &        0     &        0       &       0       &       0       & \frac{1}{G_{12}}
+		\dfrac{1}{E_1} & -\dfrac{\nu_{21}}{E_2} & -\dfrac{\nu_{31}}{E_3} &       0       &       0       &       0 \\
+		-\dfrac{\nu_{12}}{E_1} & \dfrac{1}{E_2} & -\dfrac{\nu_{32}}{E_3} &       0       &       0       &       0 \\
+		-\dfrac{\nu_{13}}{E_1} & -\dfrac{\nu_{23}}{E_2} & \dfrac{1}{E_3} &       0       &       0       &       0 \\
+		      0       &        0     &        0       & \dfrac{1}{G_{23}} &       0       &       0 \\
+		      0       &        0     &        0       &       0       & \dfrac{1}{G_{13}}&       0 \\
+		      0       &        0     &        0       &       0       &       0       & \dfrac{1}{G_{12}}
 	\end{bmatrix}
@@
 	\subsection{Isotropic materials} \label{chap2:isotropic}
-A material is said \emph{isotropic} if its properties are independent of the directions. Therefore, we have the following relations:
+A material is said \emph{isotropic} if its properties are independent of directions. Therefore, we have the following relations:
 \begin{equation}
@@
 \end{equation}
-where $\alpha$ and $\beta$ are given by
+where $\rho$ is the material density and $\alpha$ and $\beta$ are given by
 \begin{equation}
-\alpha = 2 \rho \pd{U_0}{I_B} = 2 \rho C_1, \quad \beta = -2\rho \pd{U_0}{II_B} = -2\rho C_2 \quad \mbox{with } \rho \, \mbox{the material density.}
+\alpha = 2 \rho \pd{U_0}{I_B} = 2 \rho C_1, \quad \beta = -2\rho \pd{U_0}{II_B} = -2\rho C_2
 \end{equation}
@@
 Depending on the change of strain rate versus stress inside a material the viscosity can be categorised as having a linear or non-linear response.
-When a material exhibits a linear response it is categorized as a Newtonian material.
+When a material exhibits a linear response it is categorised as a Newtonian material.
 In this case the stress is linearly proportional to the strain rate.

### 2010-07-29 Thesis/Chapter2.tex
- Update 29 July - Added Josh's corrections for first half of chapter 2
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@137 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{shortAbstract}
-As seen in the previous chapter, realistic modelling of organ deformation is a challenging research field that opens the door to new clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedure and per-operative guidance based on simulation.
+As seen in the previous chapter, realistic modelling of organ deformation is a challenging research field that opens the door to new clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedures and per-operative guidance based on simulation.
 In all these cases the clinician needs fast updates of the deformation model to obtain a real-time display of the computed deformations.
@@
 
-Modelling anatomical structures requires an understanding of the deformation and stresses caused by the different sollicitations that occur during medical procedures.
+Modelling anatomical structures requires an understanding of the deformation and stresses caused by the different interactions that occur during medical procedures.
 A sufficient knowledge of continuum mechanics is therefore essential to follow the rest of this manuscript.
@@
 
-Let us now consider a given particule of the body that we call $X$.
-What we will call particule in the following is in fact an infinitesimal volume of material.
-We denote the position it occupies in the initial configuration $\mathbf{X}$ and note $\mathbf{x}$ its position in the deformed configuration, both expressed in the chosen frame of reference.
+Let us now consider a given particle of the body that we call $X$.
+What we will call particle in the following is in fact an infinitesimal volume of material.
+We denote the position it occupies in the initial configuration $\mathbf{X}$ and note its position in the deformed configuration $\mathbf{x}$, both expressed in the chosen frame of reference.
 The mapping $\chi$ defined as the following:
@@
 Its mathematical description follows one of the two approaches: the material description or the spatial description.
-The material description is known as \emph{Lagrangian description} whereas the spatial description is also called \emph{Eulerian description}.
-These two approaches are detailled next.
+The material description is known as \emph{Lagrangian description} whereas the spatial description is called \emph{Eulerian description}.
+These two approaches are detailed next.
 
@@
 \end{equation}
-For a fixed value of $\mathbf{X} \in \kappa_{0}$, $\phi(\mathbf{X},t)$ gives the value of $\phi$ at time $t$ associated with the fixed particule $X$ whose position in the reference configuration is $\mathbf{X}$.
-We note that a change in time $t$ implies that the same particule $X$ has a different value $\phi$.
-Thus, the Lagrangian description focuses its attention on the particules of the continuous body and it is usually used in solid mechanics.
+For a fixed value of $\mathbf{X} \in \kappa_{0}$, $\phi(\mathbf{X},t)$ gives the value of $\phi$ at time $t$ associated with the fixed particle $X$ whose position in the reference configuration is $\mathbf{X}$.
+We note that a change in time $t$ implies that the same particle $X$ has a different value $\phi$.
+Thus, the Lagrangian description focuses its attention on the particles of the continuous body and it is usually used in solid mechanics.
 
 	\subsection{Eulerian description}
-Rather than following the particules, the Eulerian description observes the changes at fixed locations.
+Rather than following the particles, the Eulerian description observes the changes at fixed locations.
 	In the Eulerian description, the motion is this time referred to the current configuration $\kappa$ and $\phi$ is described with respect to the current position ($\mathbf{x} \in \kappa$):
@@
 \end{equation}
-For a fixed value of $\mathbf{x} \in \kappa$, $\phi(\mathbf{x},t)$ gives the value of $\phi$ associated with the particule occupying the position $\textbf{x} \in \kappa$, which may very well be a different particule for each new time $ t $.
-Because a change in time $t$ implies that a different value $\phi$ is observed at the same spatial location $\mathbf{x} \in \kappa$, now probably occupied by a different particule, the Eulerian description is focused on a spatial position.
+For a fixed value of $\mathbf{x} \in \kappa$, $\phi(\mathbf{x},t)$ gives the value of $\phi$ associated with the particle occupying the position $\textbf{x} \in \kappa$, which may very well be a different particle for each new time $ t $.
+Because a change in time $t$ implies that a different value $\phi$ is observed at the same spatial location $\mathbf{x} \in \kappa$, now probably occupied by a different particle, the Eulerian description is focused on a spatial position.
 This approach is convenient for the study of fluid flow where the kinematic property of greatest interest is the rate at which change is taking place rather than the shape of the body of fluid at a reference time \citep{Spencer80}.
@@
 	\subsection{Displacement field}
-The displacement $ \mathbf{u} $ of a particule $X$ is called \emph{displacement vector} and its expression in the Lagrangian description is given by the following:
+The displacement $ \mathbf{u} $ of a particle $X$ is called \emph{displacement vector} and its expression in the Lagrangian description is given by the following:
 \begin{equation}
@@
 
-Consider two particules $P_{1}$ and $P_{2}$ in a continuous body separated by an infinitesimal distance $d\mathbf{X}$:
+Consider two particles $P_{1}$ and $P_{2}$ in a continuous body separated by an infinitesimal distance $d\mathbf{X}$:
 \begin{equation}
@@
 \end{equation}
-After deformation, the two particules have deformed to their current positions given by the mapping $\chi$~\eqref{chap2:mapping} as:
+After deformation, the two particles have deformed to their current positions given by the mapping $\chi$~\eqref{chap2:mapping} as:
 \begin{equation}
@@
 d\mathbf{V} &= d\mathbf{L} \cdot d\mathbf{A} \\
-d\mathbf{v} &= d\mathbf{l} \cdot d\mathbf{a}
+\mbox{and } \quad d\mathbf{v} &= d\mathbf{l} \cdot d\mathbf{a}.
 \end{align}
@@
 We recall that the Jacobian $J$ has the physical meaning of being the local ratio of current to reference volume of a material volume element.
-Therefore, if $J = 1$ the volume does not change locally during the deformation and the latter is qualified as \emph{isochoric} at this given particule $P$.
+Therefore, if $J = 1$ the volume does not change locally during the deformation and the latter is qualified as \emph{isochoric} at this given particle $P$.
 If $J = 1$ everywhere in the body, the deformation of the body is isochoric.
@@
 &= J^{-1} J \notag \\
-&= 1
+&= 1.
 \end{align}
@@
 \begin{equation}
-\mathbf{F} = J^{1/3} \mathbf{\hat{F}}
+\mathbf{F} = J^{1/3} \mathbf{\hat{F}}.
 \end{equation}
@@
 	\subsection{Cauchy-Green deformation tensors}
-Let us consider two particules $P_{1}$ and $P_{2}$ in the neighbourhood of each other, separated by $d\mathbf{X}$ in the reference configuration.
+Let us consider two particles $P_{1}$ and $P_{2}$ in the neighbourhood of each other, separated by $d\mathbf{X}$ in the reference configuration.
 In the deformed configuration $P_{1}$ and $P_{2}$ occupy the positions $\tilde{P}_{1}$ and $\tilde{P}_{2}$ and they are separated by $d\mathbf{x}$.
@@
 Also, the change in squared lengths is zero if and only if $\mathbf{E} = \mathbf{0}$.
-Using \eqref{chap2:gradient}, the Green strain tensor may be developped as the following:
+Using \eqref{chap2:gradient}, the Green strain tensor may be expanded as the following:
 \begin{align}
@@
 &= \frac{1}{2} \left[ (\nabla_{0} \mathbf{u} + \mathbf{I})^T (\nabla_{0} \mathbf{u} + \mathbf{I}) - \mathbf{I} \right] \notag \\
-&= \frac{1}{2} \left[ (\nabla_{0} \mathbf{u})^T  + \nabla_{0} \mathbf{u}+ (\nabla_{0} \mathbf{u})^T(\nabla_{0} \mathbf{u}) \right] \label{chap2:GreenTensor}
+&= \frac{1}{2} \left[ (\nabla_{0} \mathbf{u})^T  + \nabla_{0} \mathbf{u}+ (\nabla_{0} \mathbf{u})^T(\nabla_{0} \mathbf{u}) \right]. \label{chap2:GreenTensor}
 \end{align}
@@
 The Green strain tensor can be expressed in terms of its components in any coordinate system.
-In particular, in the cartesian coordinate system $(X_1, X_2, X_3)$ the components $E_{ij}$ of $\mathbf{E}$ are the following:
+In particular, in the Cartesian coordinate system $(X_1, X_2, X_3)$ the components $E_{ij}$ of $\mathbf{E}$ are the following:
 \begin{equation}
@@
 	\end{bmatrix}	
+	.
 \end{equation}
@@
 These internal forces are produced between the particles in the body as a reaction to external forces applied on the body.
-The $SI$ unit for stress is pascal (symbol $Pa$), which is equivalent to one newton (force) per square meter (unit area).
+The $SI$ unit for stress is pascal (symbol $Pa$), which is equivalent to one newton (force) per square metre (unit area).
 As with strain, different measures of stress can be defined.
@@
 \mathbf{t} &= (\mathbf{n} \cdot \mathbf{e}_1) \mathbf{t}_1 + (\mathbf{n} \cdot \mathbf{e}_2) \mathbf{t}_2 + (\mathbf{n} \cdot \mathbf{e}_3) \mathbf{t}_3 \notag \\
-&= \mathbf{n} \cdot (\mathbf{e}_1 \mathbf{t}_1 + \mathbf{e}_2 \mathbf{t}_2 + \mathbf{e}_3 \mathbf{t}_3)
+&= \mathbf{n} \cdot (\mathbf{e}_1 \mathbf{t}_1 + \mathbf{e}_2 \mathbf{t}_2 + \mathbf{e}_3 \mathbf{t}_3).
 \end{align}
@@
 \label{chap2:CauchyFormula}
-\mathbf{t}(\mathbf{n}) = \mathbf{n} \cdot \sigma
+\mathbf{t}(\mathbf{n}) = \mathbf{n} \cdot \sigma .
 \end{equation}
@@
 \begin{equation}
-\mathbf{T} = \frac{da}{dA} \mathbf{t}
+\mathbf{T} = \frac{da}{dA} \mathbf{t}.
 \end{equation}
@@
 \begin{equation}
-\mathbf{T} =  J \, \mathbf{t} \, \mathbf{F}^{-T}
+\mathbf{T} =  J \, \mathbf{t} \, \mathbf{F}^{-T}.
 \end{equation}
@@
 \begin{equation}
-d\mathbf{F} = \mathbf{S} \, d\mathbf{A}
+d\mathbf{F} = \mathbf{S} \, d\mathbf{A}.
 \end{equation}
@@
 &= \mathbf{F}^{-1} \, (\mathbf{P} \, d\mathbf{A}) \notag \\
-&= \mathbf{S} \, d\mathbf{A}
+&= \mathbf{S} \, d\mathbf{A}.
 \end{align}
@@
 \begin{equation}
-\mathbf{S} = \mathbf{F}^{-1} \mathbf{P} = J \, \mathbf{F}^{-1} \, \sigma \, \mathbf{F}^{-T}
+\mathbf{S} = \mathbf{F}^{-1} \mathbf{P} = J \, \mathbf{F}^{-1} \, \sigma \, \mathbf{F}^{-T}.
 \end{equation}
@@
 \begin{equation}
-U_0 = C_0 + C_{ij}\varepsilon_{ij} + \frac{1}{2!}\hat{C}_{ijkl}\varepsilon{ij}\varepsilon_{kl}
+U_0 = C_0 + C_{ij}\varepsilon_{ij} + \frac{1}{2!}\hat{C}_{ijkl}\varepsilon{ij}\varepsilon_{kl}.
 \end{equation}
@@
 \begin{equation}
-\sigma_{ij} = C_{ijkl} \varepsilon_{kl}
+\sigma_{ij} = C_{ijkl} \varepsilon_{kl}.
 \end{equation}
@@
 \begin{equation}
-\alpha = 2 \rho \pd{U_0}{I_B} = 2 \rho C_1, \quad \beta = -2\rho \pd{U_0}{II_B} = -2\rho C_2
+\alpha = 2 \rho \pd{U_0}{I_B} = 2 \rho C_1, \quad \beta = -2\rho \pd{U_0}{II_B} = -2\rho C_2.
 \end{equation}
@@
 \begin{equation}
-U_0 = C_1 (I_B-3)
+U_0 = C_1 (I_B-3).
 \end{equation}
@@
 \begin{equation}
-\sigma = -p \mathbf{I} + 2 \rho C_1 \mathbf{B}
+\sigma = -p \mathbf{I} + 2 \rho C_1 \mathbf{B}.
 \end{equation}

### 2010-07-30 Thesis/Chapter3.tex
- Update 30 July - Changed outline according to what we discussed with Stephane + increased level of details + bib updated + creation of chapter 3
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@138 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2010-07-15 Thesis/tableofcontents.tex
@@
-\part{Introduction}
+\chapter{Main principles of Finite Element Method (most used technique to solve equations of continuum mechanics)}
+\label{chap3}
+\begin{shortAbstract}
+Blabla
+\end{shortAbstract}
 
-	\chapter{Medical simulation}
-		\section{General context and goal: medical training, patient-specific planning and per-operative guidance}
-		\section{Challenges (trade-off between accuracy and real-time)}
+\section{Introduction}
 
-	\chapter{One key point in medical simulation: soft-tissue modelling}
-		\section{Necessary background in continuum mechanics}
-			\subsection{Deformation tensor and strain tensor}
-			\subsection{Stress and constitutive laws}
-		\section{Tissue characterisation}
-			\subsection{Material models for organs (non-linear, visco-elastic and anisotropic)}
-			\subsection{Measure/estimation of model parameters}
+\section{Discretisation}
 
-	\chapter{Main principles of Finite Element Method (or how to solve equations of continuum mechanics from previous section)}
-		\section{Discretisation}
 \section{Derivation of element equations}
+
 \section{Assembly of element equations}
+
 \section{Solution of global problem}
-
-\part{Solid organs modelling}
-
-	\chapter{State of art: FEM}
-
-	\chapter{Linear not accurate => Non-linear FEM => Introduction of TLED}
-		\section{Differences with classic FEM and reasons of its efficiency}
-		\section{Visco-elasticity and anisotropy added \OC{MICCAI 2008; MedIA 2009}}
-
-	\chapter{GPU implementation of TLED}
-		\section{What is GPGPU}
-		\section{Re-formulation of the algorithm for its Cg implementation}
-		\section{CUDA implementation/optimisations \OC{ISBMS 2008a}}
-
-	\chapter{Implementation in SOFA}
-		\section{Presentation of SOFA project and architecture}
-		\section{Implementation in SOFA and TLED released in open-source}
-
-\part{Hollow organs modelling}
-
-	\chapter{State of art: hollow structures}
-		\section{Non-physic approaches (computer graphics stuff)}
-		\section{Physically accurate approches (plates/shells)}
-
-	\chapter{Colonoscopy simulator project}
-		\section{Project introduction}
-		\section{Mass-spring model for colon implemented on GPU \OC{ISBMS 2008b}}
-
-	\chapter{More accurate: a co-rotational triangular shell model \OC{ISBMS 2010}}
-		\section{Model description}
-		\section{Validation}
-		\section{Application to implant deployment simulation in cataract surgery}
-
-	\chapter{'Shell meshing' technique \OC{MICCAI 2010}}
-		\section{State of art: reconstruction/simplification}
-		\section{Our method}
-
-	\chapter{Applications to medical simulation}
-		\section{Nice medical stuff to show}
-		\section{Interaction solid/hollow organs}
-
-\part{Conclusion}

### 2010-07-30 Thesis/outline.tex
Update 15 July - Started chapter "soft-tissue modelling"
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@129 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 30 July - Changed outline according to what we discussed with Stephane + increased level of details + bib updated + creation of chapter 3
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@138 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2010-07-15 Thesis/tableofcontents.tex
@@
 	\chapter{Medical simulation}
-		\section{General context and goal: medical training, patient-specific planning and per-operative guidance}
-		\section{Challenges (trade-off between accuracy and real-time)}
+		\section{General context}
+			\subsection{Simulators for everything: we started by simulating what is simple}
+			\subsection{As complexity increases, applications to medical training, patient-specific planning and per-operative guidance}
+		\section{Challenges}					
+			\subsection{Photorealistic rendering}
+			\subsection{Accurate haptic feedback}
+			\subsection{Physically realist organ modelling in real-time}
 
-	\chapter{One key point in medical simulation: soft-tissue modelling}
-		\section{Necessary background in continuum mechanics}
-			\subsection{Deformation tensor and strain tensor}
-			\subsection{Stress and constitutive laws}
-		\section{Tissue characterisation}
-			\subsection{Material models for organs (non-linear, visco-elastic and anisotropic)}
-			\subsection{Measure/estimation of model parameters}
+	\input{chapter2}
 
-	\chapter{Main principles of Finite Element Method (or how to solve equations of continuum mechanics from previous section)}
-		\section{Discretisation}
-		\section{Derivation of element equations}
-		\section{Assembly of element equations}
-		\section{Solution of global problem}
+	\input{chapter3}
 
@@
 	\chapter{State of art: FEM}
+		\section{Expose the problem: how to model solid organs?}
+		\section{Techniques using geometrical kind of approaches}
+			\subsection{Shape matching}
+		\section{Techniques relying on physics}
+			\subsection{Mass-spring}
+		\section{Techniques based on continuum mechanics}	
+			\subsection{FEM with mesh (highlight evolution: linear, topological changes, co-rotational, non-linear, GPU)}
+			\subsection{Meshless (supposedly good for handling topological changes but other issues}
 
-	\chapter{Linear not accurate => Non-linear FEM => Introduction of TLED}
-		\section{Differences with classic FEM and reasons of its efficiency}
+	\chapter{Introduction of TLED}
+		\section{Description of the algorithm (neo-Hookean, explicit integration, independent elements, parallelisable on GPU)}
 		\section{Visco-elasticity and anisotropy added \OC{MICCAI 2008; MedIA 2009}}
@@
 		\section{CUDA implementation/optimisations \OC{ISBMS 2008a}}
-
-	\chapter{Implementation in SOFA}
-		\section{Presentation of SOFA project and architecture}
-		\section{Implementation in SOFA and TLED released in open-source}
+		\section{Implementation into SOFA}
+			\subsection{Presentation of SOFA project and architecture}
+			\subsection{TLED in SOFA and release in open-source}
 
@@
 	\chapter{State of art: hollow structures}
+		\section{Expose the problem: models presented for solid organs are insufficient. How to model hollow organs?}
+			\subsection{Colonoscopy simulator project: needs for colon}
+			\subsection{Cataract surgery, stenosis: other needs for implants/blood vessels}
 		\section{Non-physic approaches (computer graphics stuff)}
@@
 
-	\chapter{Colonoscopy simulator project}
-		\section{Project introduction}
-		\section{Mass-spring model for colon implemented on GPU \OC{ISBMS 2008b}}
+	\chapter{Mass-spring model for colon implemented on GPU \OC{ISBMS 2008b}}
+		\section{Colon is big, we need a fast model}
+		\section{Limitations (difficult to link it to actual physical properties)}
 
-	\chapter{More accurate: a co-rotational triangular shell model \OC{ISBMS 2010}}
+	\chapter{A co-rotational triangular shell model \OC{ISBMS 2010}}
 		\section{Model description}
@@
 
-	\chapter{Applications to medical simulation}
-		\section{Nice medical stuff to show}
-		\section{Interaction solid/hollow organs}
-
 \part{Conclusion}
+	\chapter{Summary}
+	\chapter{Discussion and perspectives (Interaction solid/hollow organs)}

### 2010-08-02 Thesis/Appendix2.tex
- Update 2 August - Started chapter3 + minor changes on chapter2 + bib updated
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@139 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{The weighted residual method}
\label{appendix2}

The \emph{method of weighted residuals} is an approximation technique for solving differential equations. A subclass of this method, the \emph{Galerkin method of weighted residuals}, is often used to derive the element equations for the finite element method.

Let us suppose that we have a linear differential operator $ D $ acting on a function $ u $ to produce a function $ p $:
\begin{equation}
D(u(x)) = p(x).
\end{equation}
We wish to approximate $ u $ by a function $ \tilde{u} $, which is a linear combination of basis functions chosen from a linearly independent set, That is: 
\begin{equation}
u \approx \tilde{u} = \sum_{i=1}^n a_i \phi_i .
\end{equation}
Now, when substituted into the differential operator $ D $, the result of the operation is not, in general, $ p(x) $. Hence, an error or \emph{residual} will exist:
\begin{equation}
R(x) = D(\tilde{u}(x)-p(x)) \neq 0.
\end{equation}
The idea behind the weighted residual method is to force the residual to zero in some average sense over the domain:
\begin{equation}
\int_X R(x) W_i(x) dx = 0, \quad i = 1, 2, \ldots, n
\end{equation}
where the number of \emph{weight functions} $ W_i $ is exactly equal to the number of unknown constants $ a_i $ in $ \tilde{u} $. The result is a set of $ n $ algebraic equations for the unknown constants $ a_i $. Depending on the choice for $ W_i $, several sub-methods can be derived: collocation, sub-domain, least squares, Ritz, Galerkin etc. For instance, the weight function used by the Galerkin method \TODO{need to check this: Wi are the shape functions?} is:
\begin{equation}
W_i = \pd{\tilde{u}}{a_i}.
\end{equation}

### 2010-08-02 Thesis/Chapter3.tex
- Update 2 August - Started chapter3 + minor changes on chapter2 + bib updated
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@139 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\chapter{Main principles of Finite Element Method (most used technique to solve equations of continuum mechanics)}
+\chapter{Practical approach of the Finite Element Method}
 \label{chap3}
@@
 
+	\subsection{A numerical method}
+One of the most important things engineers and scientists do is to model physical phenomena.
+Using assumptions concerning how the phenomena works and using the appropriate laws of physics governing the process, they can derive a mathematical model, often characterised by complex algebraic, differential or integral equations relating various quantities of interest.
+However, because of the complexities in the geometry and complex boundary conditions found in real life problems, we cannot analytically solve these equations.
+A few decades ago, the only possible approach was to drastically simplify them, which was not always sufficient to find an approximate solution.
+Nowadays, in practice, most of the problems are solved using numerical methods.
+Indeed, with suitable mathematical models and numerical methods, computers can help solving many practical problems of engineering.
+Numerical methods typically transform differential equations governing a continuum to a set of algebraic equations of a discrete model of the continuum that are to be solved using computers \citep{Reddy93}.
+The \emph{Finite Element Method} (FEM) is the most popular numerical procedure that is used to approximately solve differential equations, especially in continuum mechanics.
+
+	\subsection{The basic ideas of FEM}
+The finite element method begins by dividing the structure into small pieces, manageable regions, called \emph{elements}.
+The collection of all these elements makes up a \emph{mesh} which approximates the problem geometry.
+Why is this a good idea?
+If we can expect the solution for an engineering problem to be very complex, it is mostly due to the complex geometry, on which a global solution is difficult to figure out.
+If the problem domain can be divided (\emph{meshed}) into small elements, the solution within an element is easier to approximate \citep{MacDonald07}.
+Over each finite element, the unknown variables are approximated using known functions.
+These functions can be linear or higher-order polynomial expressions that depend on the geometrical locations (\emph{nodes}) used to define the finite element shape.
+The governing equations are integrated over each finite element using linear algebra techniques and the solution over the entire domain problem is obtained by summing (\emph{assembling}) the solution of each element.
+Thus, the finite element method transforms an infinite number of differential equations (one can be defined at any point of the continuum) into a finite number of algebraic equations (depending on the chosen number of elements).
+
+This approach may be compared to trying to find the area under a curve.
+We know that we can find the exact solution for the area under the curve by integration.
+However, sometimes the function describing the curve is not known, or is difficult to integrate.
+One method to obtain an approximate solution is to break up the area into a series of rectangles and add the areas of all rectangles \TODO{Add figure with curve and rectangles}.
+It is worth nothing that the solution accuracy can be increased by reducing the width of the rectangles to better follow the curve.
+
+\bigskip
+
+It is crucial to keep in mind that approximations occur at different stages during finite element analysis.
+The division of the whole domain into finite elements may not be exact \TODO{add figure from a domain divided into elements}, introducing error in the domain being modelled.
+The second stage is when element equations are derived.
+As mentionned earlier, the unknowns of the problem are approximated using the idea that any continuous function can be representated by a linear combination of known functions and undetermined coefficients.
+Algebraic relations between the undetermined coefficients are then obtained by satisfying the governing equations over each element.
+There are a few types of approach for establishing these equations but, without going into details, the mathematical foundation of all these approaches is the \emph{weighted residual method} (see Appendix~\ref{appendix2}), which leads to integrals during the process.
+Therefore, the second stage creates two sources of error: the representation of the solution by a linear combination of functions and the evaluation of the integrals.
+Finally, errors are introduced when solving the assembled system of algebraic equations.
+
 \section{Discretisation}
 
+	\subsection{Meshing process}
+The first step in the finite element method is to create a mesh of the domain to study.
+Mesh generation is a very important task and can be very time consuming.
+The domain has to be meshed properly into elements of specific shapes.
+All the elements together form the entire domain of the problem without any gap or overlapping.
+For example, triangles or quadrilaterals can be used in two dimensions, and tetrahedra and hexahedra in three dimensions.
+Information, such as \emph{element connectivity}, must be created during the meshing process for later use in the formulation of the FEM equations.
+The number of elements into which the domain is divided in a problem depends mainly on the geometry of the domain and on the desired accuracy of the solution.
+Usually, the number of elements increases with the complexity of the geometry.
+For instance, if one part of the domain is thiner \TODO{add figure to illustrate this point}, the size of the elements must be reduced in order to tile this particular part, which increases the total number of elements, and hence  the number of algebraic equations to solve.
+However, adding elements is sometimes desirable.
+Indeed, increasing the number of elements tends to get an approximate solution closer to the exact analytical solution (as reducing the width of the rectangles allows for a better approximation of the area under the curve).
+A trade-off between accuracy and computational time must be found.
+This is why the meshing process on the problem domain must be carried out carefully.
+One needs to create a mesh which gives an accurate enough solution for the desired application while restraining the computational time according to the problem time constraints.
+As an example, finding the maximum load that can sustain a bridge in structural mechanics demands a high degree of accuracy, no matter how much time it is needed to compute the solution.
+Conversely, organ deformation in medical training simulators must be computed at an interactive rate so that no apparent delay can be observed between the manipulation of a given organ and its deformation.
+It does not mean that precision is not required, simply that the time constraints will necessarily limit the accuracy of the solution.
+
+	\subsection{Elements and shape functions}
+As we have seen, the finite element method is based on finding an approximate solution over each element rather than the whole domain.
+To connect the approximate solution from each element to form a continuous over the whole domain, we require the solution to be the same at points common to the elements.
+Therefore, the element nodes (vertices defining each element's shape) needs to play the role of interpolation points in constructing the approximation functions over an element.
+Depending on the degree of polynomial approximation used to represent the solution, additional nodes may be identified inside the element.
+It is worth noting that the type of interpolation directly affects the accuracy of the solution.
+
 \section{Derivation of element equations}	
+	\subsection{Numerical integration}
+	\subsection{Generation of element matrices}
 

### 2010-08-04 mapping/BendingPlateMechanicalMapping.h
FIX: new version of constraints data, fix applyJT declaration (VecConst does not exist anymore)
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@140 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@ -102,7 +102,7 @@ public:
     void apply( typename Out::VecCoord& out, const typename In::VecCoord& in );    
     void applyJ( typename Out::VecDeriv& out, const typename In::VecDeriv& in );    
     void applyJT( typename In::VecDeriv& out, const typename Out::VecDeriv& in );
-    void applyJT( typename In::VecConst& out, const typename Out::VecConst& in );
+    void applyJT( typename In::MatrixDeriv& out, const typename Out::MatrixDeriv& in );
 
 protected:
 

### 2010-08-04 Thesis/Appendix2.tex
- Update 4 August - Added changes on chapter 3 + bib updated
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@141 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \end{equation}
-We wish to approximate $ u $ by a function $ \tilde{u} $, which is a linear combination of basis functions chosen from a linearly independent set, That is: 
+We wish to approximate $ u $ by a function $ \tilde{u} $, which is a linear combination of basis functions chosen from a linearly independent set, that is: 
 \begin{equation}

### 2010-08-04 Thesis/Chapter3.tex
- Update 4 August - Added changes on chapter 3 + bib updated
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@141 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 One of the most important things engineers and scientists do is to model physical phenomena.
-Using assumptions concerning how the phenomena works and using the appropriate laws of physics governing the process, they can derive a mathematical model, often characterised by complex algebraic, differential or integral equations relating various quantities of interest.
+Using assumptions concerning how the phenomena works and using the appropriate laws of physics governing the process, they can derive a mathematical model, often characterised by complex differential or integral equations relating various quantities of interest.
 However, because of the complexities in the geometry and complex boundary conditions found in real life problems, we cannot analytically solve these equations.
@@
 
-\bigskip
+\medskip
 
@@
 Algebraic relations between the undetermined coefficients are then obtained by satisfying the governing equations over each element.
-There are a few types of approach for establishing these equations but, without going into details, the mathematical foundation of all these approaches is the \emph{weighted residual method} (see Appendix~\ref{appendix2}), which leads to integrals during the process.
+There are a few types of approach for establishing these equations but, without going into details, the mathematical foundations of all these approaches are energy principles or the \emph{weighted residual methods} (see Appendix~\ref{appendix2}), which both lead to integrals during the process.
 Therefore, the second stage creates two sources of error: the representation of the solution by a linear combination of functions and the evaluation of the integrals.
@@
 Indeed, increasing the number of elements tends to get an approximate solution closer to the exact analytical solution (as reducing the width of the rectangles allows for a better approximation of the area under the curve).
+Elements are classically added in the particular regions of interest for a given problem and where substantial deformation may occur if this information is known a priori.
 A trade-off between accuracy and computational time must be found.
@@
 
-	\subsection{Elements and shape functions}
-As we have seen, the finite element method is based on finding an approximate solution over each element rather than the whole domain.
-To connect the approximate solution from each element to form a continuous over the whole domain, we require the solution to be the same at points common to the elements.
-Therefore, the element nodes (vertices defining each element's shape) needs to play the role of interpolation points in constructing the approximation functions over an element.
+	\subsection{Solution interpolation}	\label{chap3:solutionInterpolation}
+As we have seen, the finite element method is based on finding an approximate solution over each simple element rather than the whole domain.
+Any continuous function $ f $ may be approximated by a linear combination of known functions $ \phi_i $ and undetermined coefficients $ c_i $:
+\begin{equation}
+f \approx \tilde{f} = \sum_{i=1}^n c_i \phi_i .
+\end{equation}
+Moreover, the approximation solution $ u^e $within the element $\Omega_e $ must fullfill certain conditions of continuity in order to be convergent to the actual solution $ u $ as the number of elements is increased.
+The finite element method approximates the solution by the following polynomial expression:
+\begin{equation}
+\label{chap3:polynom}
+u^e(x) = \sum_{i=1}^n u^e_i \psi^e_i(x) 
+\end{equation}
+where $ \Omega_e $ is a one-dimensional element\footnote{For the sake of simplicity, all derivations are carried out in 1D but remain valid for each component in higher dimensions.}, $ x $ a position within this element, $ u^e_i $ are the values of the solution $ u $ at the nodes and $ \psi^e_i(x) $ the approximation functions over the element.
+This particular form will be assumed for brevity but the interested reader may refer to \cite{Reddy93} for demonstration.
+Note that $ u^e_i  $ plays the role of the undetermined coefficients and $ \psi^e_i $ the role of approximation functions.
+Writing the approximation in terms of the nodal values of the solution is necessitated by the fact that we require the solution to be the same at points common to the elements in order to connect the approximate solution from each element and form a continuous solution over the whole domain.
+The approximation $ \psi^e_i $ can be linear or higher-order polynomial expressions and are called interpolation functions.
+%They are also commonly named \emph{shape functions} because they define the `shape' that a variable can take within an element (linear, quadratic etc.). 
 Depending on the degree of polynomial approximation used to represent the solution, additional nodes may be identified inside the element.
 It is worth noting that the type of interpolation directly affects the accuracy of the solution.
+In other words, finding the solution to the problem consists of merely figuring out the values of the sought variable at every node of the mesh.
+Its value at any other point of the domain may then be deducted by interpolation within elements.
+
+\medskip
+
+The next logical question is how to derive the interpolation functions $ \psi^e_i  $ for a given element.
+Their derivation depends only on the geometry of the element and the number and location of the nodes.
+The number of nodes must be equal to the number of terms in the polynomial expansion.
+Therefore each element contains a single interpolation function for each of its nodes.
+As stated by interpolation theory, each interpolation function is required to be equal to $1$ at its corresponding node, and $0$ at all other nodes:
+\begin{equation} 
+	\psi^e_i(x_j) = \delta_{ij} = 
+	\begin{cases} 
+		1 & \text{if } i = j \\ 
+		0 & \text{if } i \neq j
+	\end{cases}
+\end{equation}
+where $ x_j $ is the position of node $j$. From there, we can state that the interpolation function at node $ i $ may be written as:
+\begin{equation}
+\psi^e_i(x) = C_i \prod_{i \neq j}^{n-1} (x-x_j)
+\end{equation}
+where $ j $ are the indices of the other nodes and $ C_i $ is a constant to be determined such as:
+\begin{equation}
+\psi^e_i(x_i) = 1.
+\end{equation}
+This function is zero at all nodes except the $i^{th}$ node. It is worth noting that such a definition for the interpolation functions $ \psi^e_i $ used in \eqref{chap3:polynom} yields:
+\begin{equation}
+u^e(x_j) = \sum_{i=1}^n u^e_i \psi^e_i(x_j) = u^e_j
+\end{equation}
+as expected.
+
+Since elements over the whole domain are generally not identical (different shapes), this process of determining all interpolation functions can become fastidious.
+We will now introduce a general method for defining interpolation functions which allows for arbitrary element shapes (subject to a given topology).
+To this end, we discuss the concept of element \emph{natural coordinates}.
+
+	\subsection{Natural coordinates}
+The natural coordinate system allows us to map every element into a typical and simpler element.
+As an example, an 8-node hexahedron element is shown in Fig.
+\TODO{add figure of hexahedron in global and natural coordinates} as it appears in the global $(x_1, x_2, x_3)$ and natural $(\xi_1,\xi_2,\xi_3)$ coordinate systems.
+While in its natural coordinate system the element is a regular aligned cube, in the global coordinate system the element may assume any admissible arbitrary form.
+Essentially, admissible means that the element must not be too distorted, and must certainly not fold back on itself.
+Another example is given Fig.
+\TODO{add figure of tetrahedron in global and natural coordinates} with a tetrahedral element.
+The components of each node has a simple expression in natural coordinates, usually $0, 1$ or $-1$ and the centre of the coordinate system is taken at one of the nodes or at the centre of the element.
+
+The use of the natural coordinate system has several advantages \citep{Biswas76}.
+Not only the interpolation functions can be derivated only once per type of element in the mesh, but their expression is also much simpler, regardless of the actual element shape.
+Consequently, the element equations and the derived element matrices get to be simplified too.
+However, we now need to switch between the natural and global coordinate systems before solving the equations on the whole domain.
+Indeed, it is necessary to account for the diversity of element shapes within the mesh.
+Solving the element equations expressed into the natural coordinate system would not yield the expected result for the whole domain.
+
+	\subsection{Geometry interpolation}
+Let us assume a relation between the global coordinate $ x $ and the natural coordinate $ \xi $ in the following form:
+\begin{equation}
+\label{chap3:f}
+x = f(\xi)
+\end{equation}
+where $ f $ is assumed to be a one-to-one correspondence (that is, a bijective function).
+This function may be seen as a transformation between the natural shape of the element and its arbitrary shape within the actual mesh, it describes a change in geometry.
+It is natural to think of approximating the geometry in a similar way that we approximated the solution in section \ref{chap3:solutionInterpolation}.
+Hence, the transformation defined by \eqref{chap3:f} may also be written as
+\begin{equation}
+x = \sum_{i=1}^m x_i^e \hat{\psi}_i^e(\xi)
+\end{equation}
+where $ x_i^e $ is the global coordinate of the $ i^{th} $ node of the element $ \Omega_e $ and $ \hat{\psi}_i^e(\xi) $ are the interpolation functions of degree $ m-1 $.
+Thus, we have a linear transformation when $ m = 2 $ and the relation betweeen $ x $ and $ \xi $ is quadratic when $ m=3 $.
+The interpolation functions $ \hat{\psi}_i^e(\xi) $ are called \emph{shape functions} because they are used to express the geometry or shape of the element.
+% Assuming the numbering of nodes in the global system follows the one of the natural system, the shape functions act as a mapping between the two representations.
+
+	\subsection{A particular case: isoparametric elements}
+% \noindent \textbf{Some remarks.}
+Most of the time, we choose to interpolate the solution and the element geometry with the same interpolation functions.
+In this case, the elements are said to be \emph{isoparametric}.
+Under this configuration, natural coordinates appear as parameters that define the shape functions.
+Note that the shape functions describe the geometry and the solution with the same degree of interpolation.
+Because isoparametric elements are very common in practice, the use of the phrase \emph{shape functions} is often extented to denote the interpolation functions employed to approximate the solution.
 
 \section{Derivation of element equations}
-	\subsection{Numerical integration}
-	\subsection{Generation of element matrices}
+
+Let us sum up where we are into the finite element analysis process so far.
+The whole domain has been divided into sub-domains, smaller, that we call elements, for which the solution to be sought will be simpler to find.
+We know how to approximate the solution within each of these elements using a linear combination of shape functions.
+And we even have a method to simplify the expression of these shape functions using natural coordinates.
+
+	\subsection{Strong and weak forms}
+
+The next step is therefore to derive the element equations themselves.
+Obviously these equations depend on the type of problem that we seek to solve.
+From now on, we will only consider the context of continuum mechanics since our overall goal is the modelling of organ deformation.
+However, a similar approach can be used in the other fields of physics.
+In continuum mechanics, applying Newton's second law yields a partial differential system of equations.
+Such equations are called \emph{strong forms}.
+The strong form, in contrast to a \emph{weak form}, requires strong continuity on the dependent field variables, such as displacements \citep{Liu03}.
+The functions defining these field variables have to be differentiable up to the order of the partial differential equations that exist in the strong from.
+Obtaining the exact solution for a strong form is usually very difficult for practical engineering problems.
+We know that the finite element method can be used to find an approximated solution but the method usually works well only with regular geometry and boundary conditions.
+
+A weak form is often created using energy principles or weighted residual methods.
+The weak form is often an integral form and requires a weaker continuity on the field variables.
+Due to the weaker requirement on the field variables, and the integral operation, a formulation based on a weak form usually produces a set of discretised system equations that give much more accurate results, especially for problems of complex geometry.
+Hence, the weak form is preferred by many for obtaining an approximate solution.
+Using the weak form usually leads to a set of well-behaved algebraic system equations, if the problem domain is discretised properly into elements.
+
+	\subsection{Time dependence}
+In all the equations of the finite element method that we derived so far, we have not taken time into account.
+For some problem, this is not an issue, the motion is sufficiently slow to assume that the dynamic effects (such as damping effect) are neglectable.
+However, this assumption is not always valid, particulary in our case where we want to model organ deformation.
+The deformation obviously depends on time.
+Finite element models of time-dependent problems can be developed in two alternatives ways \citep{Reddy93}:
+\begin{description}
+\item[(a)] coupled formulation in which the time $ t $ is treated as an additional coordinate along with the spatial coordinate $ x $
+\item[(b)] decoupled formulation where time and spatial variations are assumede to be separable.
+\end{description}
+Thus, the approximation of the solution $ u $ takes one of these two forms:
+\begin{align}
+u(x, t) \approx u^e(x, t) &= \sum_{i=1}^n \hat{u}^e_i \hat{\psi}^e_i(x, t)  \quad \text{(coupled formulation)} \label{chap3:approxTime1}\\
+u(x, t) \approx u^e(x, t) &= \sum_{i=1}^n u^e_i(t) \psi^e_i(x) \quad \text{(decoupled formulation)} \label{chap3:approxTime2}
+\end{align}
+where $ \hat{\psi}^e_i(x, t) $ are time-space interpolation functions, $ \hat{u}^e_i $ are the nodal values that are independent of $ x $ and $ t $, $ \psi^e_i(x) $ are the usual interpolation functions in spatial coordinate $ x $ only and the nodal values $ u^e_i(t) $ are functions of time $ t $ only.
+Space-time coupled finite element formulations are not common and they have not been adequately studied.
+Hence, we consider the decoupled formulation only.
+Of course, the assumption that the time and spatial variations are separable is not valid in general.
+However, with sufficiently small time steps, it is possible to obtain accurate solutions nevertheless.
+
+	\subsection{Dynamic system of equations}	
+The space-time decoupled finite element formulation of time-dependent problems involves two steps:
+\begin{description}
+\item[1. Spatial approximation.]
+The solution $ u $ is first approximated by expressions of the form \eqref{chap3:approxTime2} while keeping the time-dependent term during the finite element model derivation.
+This step leads to a set of ordinary differential equations in time.
+\item[2. Temporal approximation.]
+The system of ordinary differential equations is then approximated in time where different schemes may be used to assess the time derivatives.
+This step yields a set of algebraic equations for $ u^e_i $ at discrete time $ t_{k+1} = (k+1) \Delta t$ where $ \Delta t $ is the time increment and $ k > 0$ an integer.
+\end{description}
+At the end of this two-step approximation, we have a continuous spatial solution at discrete intervals in time:
+\begin{equation}
+\label{chap3:approxTime}
+u(x, t_k) \approx u^e(x, t_k) = \sum_{i=1}^n u^e_i(t_k) \psi^e_i(x) \quad \text{with } k = 0, 1, \ldots
+\end{equation}
+
+The construction of the weak form using either energy principles or weighted residual methods will not be detailled here.
+The reader may easily find it in dozens of textbooks.
+By substituting \eqref{chap3:approxTime} in the weak form, we obtain the finite element equations of equilibrium in matrix form:
+\begin{equation}
+\label{chap3:eqDynamic}
+\mathbf{M} \mathbf{\ddot U} + \mathbf{D} \mathbf{ \dot U} + \mathbf{K}(\mathbf{U}) \cdot \mathbf{U} = \mathbf{R},
+\end{equation}
+where $ \mathbf{M} $ is a constant mass matrix, $\mathbf{D}$ is a constant damping matrix, $ \mathbf{K}(\mathbf{U}) $ is the stiffness matrix, which is a function of global nodal displacements $\mathbf{U}$, and $\mathbf{R}$ are externally applied loads.
+
+	\subsection{Static system of equations}	
+Under some assumption, \eqref{chap3:eqDynamic} may be simplified.
+For instance, let us consider the problem of finding the maximum load that a desk can sustain.
+It is fair to say that strain and stress within the structure does not fluctuate over time.
+Therefore, we can assume that the equations describing the structure do not depend on time: the problem is said to be \emph{static}.
+In this case, the static system of equations can be easily obtained by merely dropping out the dynamics terms in \eqref{chap3:eqDynamic}.
+Indeed, the static case may just be seen as a special case of the dynamic equations and the static equations are:
+\begin{equation}
+\label{chap3:eqStatic}
+\mathbf{K}(\mathbf{U}) \cdot \mathbf{U} = \mathbf{R}.
+\end{equation}
+
+	\subsection{Determination of the matrices involved}
+
+%	\subsection{Generation of element matrices}
+%The basic postulate is that a continuum must satisfy Newton's second law of motion.
+This law can be stated as: the time rate of change of momentum of a collection of particules equals the resultant force exerted on the collection:
+%\begin{equation}
+%\dfrac{d}{dt}(m \mathbf{v}) = \mathbf{F},
+%\end{equation}
+%where $ m $ is the total mass, $ \mathbf{v} $ the velocity and $ \mathbf{F} $ the resultant force on the collection of particules.
 
 \section{Assembly of element equations}
+%Since the element is physically connected to its neighbours, the resulting algebraic equations will contain more unknowns than the number of algebraic equations.
+Thus, it becomes necessary to put the elements together to eliminate the extra unknowns.
+This process is described in the next section.
 
 \section{Solution of global problem}
+
+	\subsection{Static problems}
+
+	\subsection{Dynamic problems}

### 2010-08-05 Thesis/Chapter3.tex
- Update 5 August -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@142 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\medskip
+\bigskip
 
@@
 \begin{equation}
+\label{chap3:shapeFunctions}
 x = \sum_{i=1}^m x_i^e \hat{\psi}_i^e(\xi)
 \end{equation}
-where $ x_i^e $ is the global coordinate of the $ i^{th} $ node of the element $ \Omega_e $ and $ \hat{\psi}_i^e(\xi) $ are the interpolation functions of degree $ m-1 $.
+where $ x_i^e $ is the global coordinate of the $ i^{th} $ node of the element $ \Omega_e $, $ m $ the number of nodes for the element and $ \hat{\psi}_i^e(\xi) $ are the interpolation functions of degree $ m-1 $.
 Thus, we have a linear transformation when $ m = 2 $ and the relation betweeen $ x $ and $ \xi $ is quadratic when $ m=3 $.
@@
 % Assuming the numbering of nodes in the global system follows the one of the natural system, the shape functions act as a mapping between the two representations.
+
+\bigskip
+
+\noindent \textbf{Remark.}
+It is worth noting that \eqref{chap3:shapeFunctions} can be easily extented to three-dimensional problems.
+Let us consider a Cartesian coordinate system and we denote the position of node $ i $ $ \mathbf{x}_i = [x_{i1}, x_{i2}, x_{i3}]^T$.
+The position of a point $ \mathbf{x} = [x_{1}, x_{2}, x_{3}]^T$ within the element may then be interpolated by applying \eqref{chap3:shapeFunctions} to each component of the position:
+\begin{align}
+x_1 &= \sum_{i=1}^m x_{i1} \hat{\psi}_i^e(\xi) \notag \\
+x_2 &= \sum_{i=1}^m x_{i2} \hat{\psi}_i^e(\xi) \\
+x_3 &= \sum_{i=1}^m x_{i3} \hat{\psi}_i^e(\xi) \notag
+\end{align}
+If we construct the vector $ \hat{\mathbf{x}}^e $ of all nodal positions $ \mathbf{x}_i $ of the element, we can define a shape function matrix $ \mathbf{H} $ such as:
+\begin{equation}
+\label{chap3:H}
+\mathbf{x} = \mathbf{H} \hat{\mathbf{x}}^e
+\end{equation}
+where $ \mathbf{H} $ is a $ 3 \times m $ matrix built from submatrices $ \mathbf{H}_i $
+\begin{equation}
+\mathbf{H}_i = 
+	\begin{bmatrix}
+		\hat{\psi}_i^e & 0 & 0 \\
+		0 & \hat{\psi}_i^e & 0 \\
+		0 & 0 & \hat{\psi}_i^e
+	\end{bmatrix}
+\quad i = 1, 2, \ldots, m.
+\end{equation}
 
@@
 \begin{align}
-u(x, t) \approx u^e(x, t) &= \sum_{i=1}^n \hat{u}^e_i \hat{\psi}^e_i(x, t)  \quad \text{(coupled formulation)} \label{chap3:approxTime1}\\
+u(x, t) \approx u^e(x, t) &= \sum_{i=1}^n \hat{u}^e_i \hat{\psi}^e_i(x, t)  \quad \text{ (coupled formulation)} \label{chap3:approxTime1}\\
 u(x, t) \approx u^e(x, t) &= \sum_{i=1}^n u^e_i(t) \psi^e_i(x) \quad \text{(decoupled formulation)} \label{chap3:approxTime2}
@@
 
-	\subsection{Determination of the matrices involved}
+	\subsection{A few words on the matrices involved}
 
-%	\subsection{Generation of element matrices}
-%The basic postulate is that a continuum must satisfy Newton's second law of motion.
-This law can be stated as: the time rate of change of momentum of a collection of particules equals the resultant force exerted on the collection:
-%\begin{equation}
-%\dfrac{d}{dt}(m \mathbf{v}) = \mathbf{F},
-%\end{equation}
-%where $ m $ is the total mass, $ \mathbf{v} $ the velocity and $ \mathbf{F} $ the resultant force on the collection of particules.
+		\subsubsection{Mass matrix}
+The mass matrix $ \mathbf{M} $ for the whole domain may be computed by summing the mass contributions from every elements:
+\begin{equation}
+\mathbf{M} = \sum_e \mathbf{M}^e,
+\end{equation}
+where $ \mathbf{M}^e $ is the mass of element $ \Omega_e $ given by:
+\begin{equation}
+\mathbf{M}^e = \int_{V^e} \rho^e \mathbf{H}^T \mathbf{H} dV
+\end{equation}
+where $ \rho $ the mass density, $ V^e $ the volume of the element and $ \mathbf{H} $ is the shape function matrix defined by \eqref{chap3:H}.
+This integral may be evaluated using a procedure of numerical integration and generally results in a band mass matrix (non-zero entries are confined to a diagonal band, comprising the main diagonal and more diagonals on either side).
+This method for computing the mass matrix is called \emph{variational mass lumping}.
+
+However, it is very common in finite element analysis to use a diagonal mass matrix.
+The most common way of achieving this is to define the diagonal entries with the sum of the corresponding rows and then set all other values to 0.
+The physical interpretation of this is lumping all of the mass in the volume surrounding a node at the node itself.
+A diagonal mass matrix may offer significant computational and storage advantages.
+First, it is easily inverted, since the inverse of a diagonal matrix is also diagonal.
+And a diagonal mass matrix can also be stored simply as a vector.
+Therefore, such a matrix may be built by directly computing the mass of each element and assigning an equal proportion of this to each of the element's nodes.
+This method is named \emph{direct mass lumping}.
+The mass $ m^e $ of element $ \Omega_e $ is given by:
+\begin{equation}
+m^e = \rho^e \int_{V^e} dV,
+\end{equation}
+which is integrated over the element using a numerical method. Each diagonal component of the mass matrix then receives a contribution of $ m^e/m $.
+
+		\subsubsection{Damping matrix}
+In a similar way to the mass matrix, a damping matrix may be defined by summing the damping contributions from every elements:
+\begin{equation}
+\mathbf{D} = \sum_e \mathbf{D}^e,
+\end{equation}
+where $ \mathbf{D}^e $ is the damping of element $ \Omega_e $ given by:
+\begin{equation}
+\mathbf{D}^e = \int_{V^e} \kappa^e \mathbf{H}^T \mathbf{H} dV
+\end{equation}
+where $ \kappa^e $ is a damping parameter.
+In practice, the components of damping matrices are difficult to determine experimentally.
+Therefore, it is common to employ the \emph{Rayleigh damping}.
+The Rayleigh damping assumes that the damping may be written in the form of:
+\begin{equation}
+\mathbf{D}^e = \alpha \mathbf{M}^e + \beta \mathbf{K}^e,
+\end{equation}
+where $ \alpha $ and $ \beta $ are coefficients defining mass and stiffness proportional damping components.
+However, for systems with large degrees of freedom, it is difficult to guess meaningful values of $ \alpha $ and $ \beta $.
+\cite{Chowdhury03} have proposed a method to calculate these coefficients.
+In practice, further assumptions may only lead to include the mass proportional component ($ \beta = 0 $), which yields a diagonal damping matrix if a diagonal mass matrix is used.
+
+		\subsubsection{Stiffness matrix}
+The expression of the stiffness matrix depends on the chosen strain tensor as well as the appropriate constitutive model for the material.
+Its general form cannot be explicited.
+However, it can be demonstrated that $ \mathbf{K}^e $ may be written as
+\begin{equation}
+\mathbf{K}^e = \int_{V^e} \mathbf{B}^T \mathbf{M} \, \mathbf{B} \, dV,
+\end{equation}
+where $ \mathbf{M} $ is a material matrix, which only depends on the constitutive model chosen, and $ \mathbf{B} $ is the \emph{strain-displacement matrix} defined as the derivatives of the shape functions (see \cite{Liu03,Belytschko00} for details).
 
 \section{Assembly of element equations}
-%Since the element is physically connected to its neighbours, the resulting algebraic equations will contain more unknowns than the number of algebraic equations.
+We have now derived the equations for each element.
+However, since the element is physically connected to its neighbours, the resulting algebraic equations will contain more unknowns than the number of algebraic equations.
 Thus, it becomes necessary to put the elements together to eliminate the extra unknowns.
-This process is described in the next section.
+The assembly should be carried out as soon as the element matrices are computed, rather than waiting until the element coefficients of all elements are computed.
+This would require storage of the coefficients for each element.
+Instead, we can perform the assembly while we calculate the element matrices.
+
+The assembly operation is actually quite simple.
+Let us consider an example with stiffness matrices for one-dimensional problem.
+We assume that we have three two-node elements.
+And we suppose the element connectivity information given by Table \ref{chap3:connectivity}, which provides the global node number for each local node number within an element.
+\begin{table}
+	\begin{center}
+	\begin{tabular}{|c|c|c|}
+		\hline Elements & 1 & 2 \\ 
+		\hline 1 & 1 & 2 \\ 
+		\hline 2 & 2 & 3 \\ 
+		\hline 3 & 3 & 4 \\ 
+		\hline 
+		\end{tabular} \caption{Element connectivity information}
+		\label{chap3:connectivity}
+	\end{center}
+\end{table}
+
+% The element matrices are expressed with respect to local node numbers.
+\TODO{add figure with global stiffness matrix and a few arrows issued from the element stiffness matrices}
 

### 2010-08-06 Thesis/Chapter3.tex
- Update 6 August -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@143 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-	\subsection{Dynamic system of equations}	
+	\subsection{Dynamic system of equations}	 \label{chap3:dynamic}
 The space-time decoupled finite element formulation of time-dependent problems involves two steps:
@@
 
-	\subsection{Static system of equations}	
+	\subsection{Static system of equations}	\label{chap3:static}
 Under some assumption, \eqref{chap3:eqDynamic} may be simplified.
 For instance, let us consider the problem of finding the maximum load that a desk can sustain.
-It is fair to say that strain and stress within the structure does not fluctuate over time.
+It is fair to say that strain and stress within the structure does not fluctuate over time: the system is in an equilibrium state.
+At all times, the sum of all forces applied on the system is equal to $ 0 $.
 Therefore, we can assume that the equations describing the structure do not depend on time: the problem is said to be \emph{static}.
@@
 
-		\subsubsection{Mass matrix}
+		\subsubsection{Mass matrix} \label{chap3:massMatrix}
 The mass matrix $ \mathbf{M} $ for the whole domain may be computed by summing the mass contributions from every elements:
@@
 
-	\subsection{Static problems}
+\cite{Belytschko00}
 
-	\subsection{Dynamic problems}
+	\subsection{Explicit time integration}
+
+	\subsection{Static solutions and implicit time integration}
+
+%The complete and global system of equations has now been established.
+In this section, we will examine the techniques commonly used for solving these equations.
+Note that this is only the most classical approaches and the list is obviously not exhaustive.
+The choice of a resolution method depends on the problem (linear or non-linear) and the properties of the stiffness matrix.
+The system will first be solved in static before reviewing different methods of time integration to find the solutions in dynamic.
+%
+%	\subsection{Static problems}
+%First, we recall the general system of equations in static introduced in section \ref{chap3:static}, which is also valid for the whole system: 
+%\begin{equation}
+%\mathbf{K}(\mathbf{U}) \cdot \mathbf{U} = \mathbf{R}.
+%\end{equation}	
+%In this case, the displacement $ \mathbf{U} $ at each node may be found using an appropriate solver.
+If the problem is linear, the stiffness matrix does not depend on the displacement and the resolution is facilitated.
+The unknown displacements may be computed by
+%\begin{equation}
+%\mathbf{U} = \mathbf{K}^{-1} \, \mathbf{R}.
+%\end{equation}
+%Many techniques are available in the literature to invert a matrix and a few of them are introduced in section \ref{chap3:solvers}. 
+%	
+%	\subsection{Dynamic problems}
+%Let us recall the general system of equations in dynamic introduced in section \ref{chap3:dynamic}, which is also valid for the whole system: 
+%\begin{equation}
+%\label{chap3:dynamic2}
+%\mathbf{M} \mathbf{\ddot U} + \mathbf{D} \mathbf{ \dot U} + \mathbf{K}(\mathbf{U}) \cdot \mathbf{U} = \mathbf{R}.
+%\end{equation}	
+%In the general case of a dynamic system of equations, the equations are not yet algebraic and still depends on derivatives of the unknown displacements with respect to time.
+Therefore, an additional step needs to be performed before using a solver by integrating the equations over time.
+The different methods of time integration are classified into two main schemes: \emph{explicit} or \emph{implicit} integration.
+If positions and velocities are extrapolated from their values obtained at the previous time step, then this approach is qualified of explicit.
+Conversely, if positions and velocities are considered as unknowns for the current time step, the scheme is called implicit.
+The choice of one scheme over the other impacts the stability and the computational time.
+%
+%		\subsubsection*{Explicit time integration}
+%Explicit schemes make use	 of data from the previous time step to assess the ones from current time step. From \eqref{chap3:dynamic2} we have:
+%\begin{align}
+%\mathbf{M} \mathbf{\ddot U} &= \mathbf{R} - (\mathbf{D} \mathbf{ \dot U} + \mathbf{K}(\mathbf{U}) \cdot \mathbf{U}) \notag \\
+%\mathbf{M} \mathbf{\ddot U} &= \mathbf{F}_{\textbf{ext}} - \mathbf{F}_{\textbf{int}}(\mathbf{U}, \mathbf{\dot U})
+%\end{align}
+%where $ \mathbf{F}_{\textbf{int}} $ and $ \mathbf{F}_{\textbf{ext}} $ are the internal and external forces, respectively.
+We denote $ \mathbf{U}_i - \mathbf{U}_{i-1} = \Delta U = h \mathbf{\dot U}_{i-1} $ and $ \mathbf{\dot U}_i - \mathbf{\dot U}_{i-1} = \Delta \dot U = h \mathbf{\ddot U}_{i-1} $ and we may write:
+%\begin{equation}
+%	\begin{pmatrix}
+%		\Delta U \\
+%		\Delta \dot U
+%	\end{pmatrix}
+%	= 
+%	h 
+%	\begin{pmatrix}
+%		\mathbf{\dot U}_{i-1} \\
+%		\mathbf{M}^{-1} \, (\mathbf{F}_{\textbf{ext}} - \mathbf{F}_{\textbf{int}}(\mathbf{U}, \mathbf{\dot U}))
+%	\end{pmatrix}
+%\end{equation}
+%We note that this scheme does not require any resolution of system if the mass matrix was diagonalised by mass lumping as explained in section \ref{chap3:massMatrix}.
+It is therefore a particulary efficient algorithm.
+However, this kind of scheme is only conditionally stable and requires small time steps.
+Thus, the gain in computation must be balanced against the need to calculating more steps.
+This lack of stability increases with smaller elements and stiffer materials.
+
+%		\subsubsection*{Implicit time integration}
+
+	\subsection{Solvers}	\label{chap3:solvers}
+
+		\subsubsection*{Direct solvers}
+		\subsubsection*{Iterative solvers}

### 2010-08-07 Thesis/Chapter3.tex
- Update 7 August -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@144 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 where $ \mathbf{M} $ is a constant mass matrix, $\mathbf{D}$ is a constant damping matrix, $ \mathbf{K}(\mathbf{U}) $ is the stiffness matrix, which is a function of global nodal displacements $\mathbf{U}$, and $\mathbf{R}$ are externally applied loads.
+This equation is applicable to individual elements as well as the entire element assemblage.
 
@@
 Therefore, we can assume that the equations describing the structure do not depend on time: the problem is said to be \emph{static}.
+Accelerations can be neglected.
 In this case, the static system of equations can be easily obtained by merely dropping out the dynamics terms in \eqref{chap3:eqDynamic}.
@@
 
-	\subsection{A few words on the matrices involved}
+	\subsection{A few words on the matrices involved} \label{chap3:wordOnMatrices}
 
-		\subsubsection{Mass matrix} \label{chap3:massMatrix}
+		\subsubsection*{Mass matrix}
 The mass matrix $ \mathbf{M} $ for the whole domain may be computed by summing the mass contributions from every elements:
@@
 
-		\subsubsection{Damping matrix}
+		\subsubsection*{Damping matrix}
 In a similar way to the mass matrix, a damping matrix may be defined by summing the damping contributions from every elements:
@@
 
-		\subsubsection{Stiffness matrix}
+		\subsubsection*{Stiffness matrix}
 The expression of the stiffness matrix depends on the chosen strain tensor as well as the appropriate constitutive model for the material.
@@
 \begin{equation}
-\mathbf{K}^e = \int_{V^e} \mathbf{B}^T \mathbf{M} \, \mathbf{B} \, dV,
+\mathbf{K}^e = \int_{V^e} \mathbf{B}^T \mathbf{E} \, \mathbf{B} \, dV,
 \end{equation}
-where $ \mathbf{M} $ is a material matrix, which only depends on the constitutive model chosen, and $ \mathbf{B} $ is the \emph{strain-displacement matrix} defined as the derivatives of the shape functions (see \cite{Liu03,Belytschko00} for details).
+where $ \mathbf{E} $ is a material matrix, which only depends on the constitutive model chosen, and $ \mathbf{B} $ is the \emph{strain-displacement matrix} defined as the derivatives of the shape functions (see \cite{Liu03,Belytschko00} for details).
 
@@
 \section{Solution of global problem}
-
-\cite{Belytschko00}
+The complete and global system of equations has now been established.
+Let us recall the general system of equations in dynamic introduced in section \ref{chap3:dynamic}, which is also valid for the whole system:
+\begin{equation}
+\label{chap3:dynamic2}
+\mathbf{M} \mathbf{\ddot U} + \mathbf{D} \mathbf{ \dot U} + \mathbf{K}(\mathbf{U}) \cdot \mathbf{U} = \mathbf{R}.
+\end{equation}	
+In this section, we will examine the techniques commonly used for solving these equations.
+In the general case of a dynamic system of equations, the equations are not yet algebraic and still depends on derivatives of the unknown displacements with respect to time.
+Therefore, an additional step needs to be performed before using a solver by integrating the equations over time.
+As we will see shortly, the different methods of time integration are classified into two main schemes: \emph{explicit} or \emph{implicit} integration.
+We begin by introducing the simplest method, the Euler method to explain the concept of an explicit method.
+Then the central difference method will be detailled.
+We then describe a few implicit methods, allowing us to highlight the relative advantages of the two types of methods.
+Finally, the static case will be solved by emphasizing the similarities with implicit systems through the critical step of linearisation of the governing equations \citep{Belytschko00}.
 
@@
 
-	\subsection{Static solutions and implicit time integration}
+		\subsubsection*{The Euler method}	
+We want to approximate the solution of the following differential equation:
+\begin{equation}
+\label{chap3:eqToSolve}
+y^{'}(t) = f(t, y(t))
+\end{equation}
+Note that we restrict ourselves to first-order differential equations (meaning that only the first derivative of $y$ appears in the equation, and no higher derivatives).
+However, a higher-order equation can easily be converted to a system of first-order equations by introducing extra variables.
+For example, the second-order equation $y^{''} = -y$ can be rewritten as two first-order equations: $y^{'} = z$ and $z^{'} = -y$.
+By solving equations like \eqref{chap3:eqToSolve}, we will therefore be able to integrate the global system of equations over time.
 
-%The complete and global system of equations has now been established.
-In this section, we will examine the techniques commonly used for solving these equations.
-Note that this is only the most classical approaches and the list is obviously not exhaustive.
-The choice of a resolution method depends on the problem (linear or non-linear) and the properties of the stiffness matrix.
-The system will first be solved in static before reviewing different methods of time integration to find the solutions in dynamic.
-%
-%	\subsection{Static problems}
-%First, we recall the general system of equations in static introduced in section \ref{chap3:static}, which is also valid for the whole system: 
+We start by replacing the derivative $y^{'} $ by the finite difference approximation:
+\begin{equation}
+\label{chap3:approxExplicit}
+y^{'}(t) \approx \dfrac{y(t+\Delta t) - y(t)}{\Delta t},
+\end{equation}
+which gives:
+\begin{equation}
+y(t+\Delta t) \approx y(t) + \Delta t y^{'}(t)
+\end{equation}
+And substituting the last relation in \eqref{chap3:eqToSolve} yields:
+\begin{equation}
+\label{chap3:updateExplicit}
+y(t+\Delta t) \approx y(t) + \Delta t f(t, y(t))
+\end{equation}
+This formula is usually applied in a recursive scheme.
+We choose a time step $ \Delta t $ and we construct the sequence of time $ t^n = n \Delta t$.
+Using \eqref{chap3:updateExplicit}, we can approximate the exact solution $ y(t^n) $ at each time step by $ y^n $:
+\begin{equation}
+y^{n+1} = y^n + \Delta t f(t^n, y^n).
+\end{equation}
+This method is called \emph{Euler method} and is said to be explicit because the new value $ y^{n+1} $ is defined in terms of variables from the previous time step ($ y^n $).
+
+		\subsubsection*{The central difference method}	
+The central difference method is one of the most popular explicit methods in computational mechanics.
+This method is based on central difference formulas for the velocity and acceleration.
+Let us the time $ t $ of the simulation be subdivided into time steps $ \Delta t $.
+Let us denote by $ \mathbf{U}^t $, $ \mathbf{\dot U}^t $ and $ \mathbf{\ddot U }^t $ the matrices of nodal displacements, velocities and accelerations, respectively, at time $ t $.
+The central difference method makes use of the following finite differences approximations for acceleration and velocity:
+\begin{align}
+\mathbf{\ddot U}^t &= \dfrac{1}{(\Delta t)^2} \left( \mathbf{U}^{t-\Delta t} - 2 \mathbf{U}^{t} +\mathbf{U}^{t+\Delta t} \right) \label{chap3:accelerationApprox}\\
+\mathbf{\dot U}^t &= \dfrac{1}{2 \Delta t} \left( \mathbf{U}^{t+\Delta t} - \mathbf{U}^{t-\Delta t} \right). \label{chap3:velocityApprox}
+\end{align}
+One of the key advantage of the method is that the stifness term in \eqref{chap3:dynamic2} may be computed from
+\begin{equation}
+\mathbf{K}(\mathbf{U}) \cdot \mathbf{U} = \mathbf{F}(\mathbf{U}) = \sum_e \tilde{\mathbf{F}}^e,
+\end{equation}
+where $ \tilde{\mathbf{F}}^e $ are the global nodal force contributions due to stresses in element $ \Omega_e $.
+$ \tilde{\mathbf{F}} $ may be computed using strain-displacement equations, the constitutive equation and the element connectivity.
+Therefore, let us assume that all nodal forces $\mathbf{F}^t $  at time $ t $ are known.
+
+By subtituting \eqref{chap3:accelerationApprox} and \eqref{chap3:velocityApprox} into \eqref{chap3:dynamic2} we obtain:
+\begin{equation}
+\left( \dfrac{\mathbf{M}}{(\Delta t)^2} + \dfrac{\mathbf{D}}{2 \Delta t} \right) \mathbf{U}^{t+\Delta t} = \mathbf{R}^t - \mathbf{F}^t + \dfrac{2 \mathbf{M}}{(\Delta t)^2} \mathbf{U}^t + \left( \dfrac{\mathbf{D}}{2 \Delta t} - \dfrac{\mathbf{M}}{(\Delta t)^2} \right) \mathbf{U}^{t-\Delta t}.
+\end{equation}
+Both the right-hand side of the equation and the factor in front $ \mathbf{U}^{t+\Delta t} $ may be evaluated, which gives the nodal displacements for the next time step.
+The entire update can be accomplished without solving any system of equations provided that the mass matrix $ \mathbf{M} $ and the damping matrix $ \mathbf{D} $ are diagonal.
+As we have seen in section \ref{chap3:wordOnMatrices}, these approximations are often used in practice.
+And this is the characteristic of an explicit method: the time integration of the general system of equations for finite element model does not require the solution of any equations.
+This is a key point because it gives explicit methods a tremendous computational advantage.
+However, explicit methods are only conditionally stable.
+If the time step exceeds a critical value, the solution may be erroneous.
+This critical time step depends on the mesh and the material properties.
+It will decrease with mesh refinement and increasing stiffness of the material.
+
 %\begin{equation}
-%\mathbf{K}(\mathbf{U}) \cdot \mathbf{U} = \mathbf{R}.
+%\label{chap3:diffV}
+%\mathbf{V} ^{n+\frac{1}{2}} = \dfrac{1}{\Delta t^{n+\frac{1}{2}}} \left(\mathbf{U} ^{n+1} - \mathbf{U} ^{n} \right).
 %\end{equation}
-%In this case, the displacement $ \mathbf{U} $ at each node may be found using an appropriate solver.
-If the problem is linear, the stiffness matrix does not depend on the displacement and the resolution is facilitated.
-The unknown displacements may be computed by
+%By rearranging \eqref{chap3:diffV}, we may obtain:
 %\begin{equation}
-%\mathbf{U} = \mathbf{K}^{-1} \, \mathbf{R}.
+%\mathbf{U} ^{n+1} = \mathbf{U} ^{n} + \Delta t^{n+\frac{1}{2}} \mathbf{V}^{n+\frac{1}{2}}
 %\end{equation}
-%Many techniques are available in the literature to invert a matrix and a few of them are introduced in section \ref{chap3:solvers}. 
-%	
+%The acceleration is given by:
+%\begin{align}
+%\label{chap3:diffA}
+%\mathbf{A} ^{n} &= \dfrac{1}{ \Delta t^{n}} \left(\mathbf{V}^{n+\frac{1}{2}} - \mathbf{V}^{n-\frac{1}{2}} \right), \\
+%\mathbf{V} ^{n+\frac{1}{2}} &= \mathbf{V}^{n-\frac{1}{2}} + \Delta t^{n} \mathbf{A}^{n}.
+%\end{align}
+%By substituting \eqref{chap3:diffV} and its counterpart for the previous time step into \eqref{chap3:diffA}, the acceleration can be expressed directly in terms of the displacements:
+%\begin{equation}
+%\mathbf{A}^n = \dfrac{ \Delta t^{n-\frac{1}{2}} \left( \mathbf{U}^{n+1} - \mathbf{U}^{n} \right) - \Delta t^{n+\frac{1}{2}} \left( \mathbf{U}^{n} - \mathbf{U}^{n-1} \right) }{ \Delta t^{n} \Delta t^{n-\frac{1}{2}} \Delta t^{n+\frac{1}{2}} }
+%\end{equation}
+%If the time steps are all equals, the latter expression can be reduced to:
+%\begin{equation}
+%\mathbf{A}^n = \dfrac{ \left( \mathbf{U}^{n+1} -2 \mathbf{U}^{n} + \mathbf{U}^{n-1} \right) }{ \left( \Delta t^n \right)^2 }.
+%\end{equation}
+%This is the well-known formula for the second derivative of a function.
+
+%We now consider the time integration of the dynamic system of equations. Equations \eqref{chap3:dynamic2} may be written as:
+%\begin{equation}
+%\mathbf{M} \mathbf{A}^n = \mathbf{F}_{\text{ext}}(\mathbf{U}, t^n) - \mathbf{F}_{\text{int}}(\mathbf{U}, \mathbf{V}, t^n).
+%\end{equation}
+%where $ \mathbf{F}_{\text{int}} $ and $ \mathbf{F}_{\text{ext}} $ are the internal and external forces, respectively.
+The external loads are usually provided as functions of time and they may depends on nodal displacements (as when a pressure is applied).
+The internal forces depend on displacements and velocities through the stiffness and damping matrices.
+
+	\subsection{Implicit time integration}
+
+		\subsubsection*{The backward Euler method}
+The problem that we wish to solve is the same than the one we introduced with the Euler method, which we reproduce here for convenience:
+\begin{equation}
+y^{'}(t) = f(t, y(t))
+\end{equation}
+This time, instead of using \eqref{chap3:approxExplicit}, we approximate the derivative $ y^{'}(t) $ by:
+\begin{equation}
+y^{'}(t) \approx \dfrac{y(t) - y(t-\Delta t)}{\Delta t}.
+\end{equation}
+The update expression for $ y^n $ then becomes:
+\begin{equation}
+y^{n+1} = y^n + \Delta t f(t^{n+1}, y^{n+1}).
+\end{equation}
+The \emph{backward Euler method} is an implicit method, meaning that we have to solve an equation to find $ y^{n+1} $.
+Of course, it costs time to solve this equation and this cost must be taken into consideration when we select the time integration method to use.
+The advantage of implicit methods is that they are usually more stable for solving a stiff equation, meaning that larger steps can be used.
+
+	\subsubsection*{The Runge-Kutta method}
+The accuracy of backward Euler method can be improved by making the solution depends on more values. The $ 4^{th} $ order Runge-Kutta method gives the following update equation:
+\begin{equation}
+y^{n+1} = y^{n} + \frac{1}{6} \Delta t \left( k_1 + 2 k_2 + 2 k_3 +k_4 \right),
+\end{equation}
+where :
+\begin{align}
+k_1 &= f(t^n, y^n) \notag \\
+k_2 &= f(t^n + \frac{1}{2} \Delta t, y^n + \frac{1}{2} \Delta t \, k_1) \notag \\
+k_3 &= f(t^n + \frac{1}{2} \Delta t, y^n + \frac{1}{2} \Delta t \, k_2) \notag \\
+k_4 &= f(t^n + \Delta t, y^n + \Delta t \, k_3) \notag
+\end{align}
+This method is a fourth-order method because the total accumulated error has order $ \Delta t^4 $.
+It may be seen as a generalisation of the backward Euler method \citep{Press92}.
+Again, we have to solve an equation to find $ y^{n+1} $.
+
+	\subsection{Static solutions}
+First, we recall the general system of equations in static introduced in section \ref{chap3:static}, which is also valid for the whole system: 
+\begin{equation}
+\mathbf{K}(\mathbf{U}) \cdot \mathbf{U} = \mathbf{R}.
+\end{equation}	
+The situation is similar to implicit time integration methods for dynamic systems by the fact that we also have to solve a set of algebraic equations to find the nodal displacements.
+In the general case, the equations are non-linear and therefore quite difficult to solve.
+A suitable solver for this kind of system is, for instance, the \emph{Newton-Raphson} method.
+In essence, this method consists of obtaining linear equations instead.
+This process is called \emph{linearisation}.
+One may refer to \citep{Belytschko00} for more details.
+
+However, if the problem is linear, the stiffness matrix does not depend on the displacement and the resolution is facilitated. The unknown displacements may be computed by:
+\begin{equation}
+\mathbf{U} = \mathbf{K}^{-1} \, \mathbf{R}.
+\end{equation}
+Many techniques are available in the literature to solve this type of system, and a few of them are introduced in the next section.
+
 %	\subsection{Dynamic problems}
@@
 %\label{chap3:dynamic2}
-%\mathbf{M} \mathbf{\ddot U} + \mathbf{D} \mathbf{ \dot U} + \mathbf{K}(\mathbf{U}) \cdot \mathbf{U} = \mathbf{R}.
+%\mathbf{M} \mathbf{\ddot U} + \mathbf{U} \mathbf{ \dot U} + \mathbf{K}(\mathbf{U}) \cdot \mathbf{U} = \mathbf{R}.
 %\end{equation}	
@@
 %\begin{align}
-%\mathbf{M} \mathbf{\ddot U} &= \mathbf{R} - (\mathbf{D} \mathbf{ \dot U} + \mathbf{K}(\mathbf{U}) \cdot \mathbf{U}) \notag \\
+%\mathbf{M} \mathbf{\ddot U} &= \mathbf{R} - (\mathbf{U} \mathbf{ \dot U} + \mathbf{K}(\mathbf{U}) \cdot \mathbf{U}) \notag \\
 %\mathbf{M} \mathbf{\ddot U} &= \mathbf{F}_{\textbf{ext}} - \mathbf{F}_{\textbf{int}}(\mathbf{U}, \mathbf{\dot U})
@@
 
-%		\subsubsection*{Implicit time integration}
+	\subsection{Solvers}	\label{chap3:solvers}
+We will now examine the techniques commonly used for solving linear systems of algebraic equations.
+Note that this is only the most classical approaches and the list is obviously not exhaustive.
+Two categories of techniques are in opposition: \emph{direct} or \emph{iterative} solvers \citep{Press02}.
+In any case, we seek the solution for the unknown solution $ \mathbf{x}$ of the following system of linear equations:
+\begin{equation}
+\label{chap3:linearEquation}
+\mathbf{A} \mathbf{x} = \mathbf{b}.
+\end{equation}
+The system may already be linear by nature, or been linearised using techniques such as Newton-Raphson methods.
+At this point in the discussion, we are therefore only interested in solving a linear system of equations.
+If the matrix $ \mathbf{A} $ is squared, which is the case in finite element methods, the problem may be written as follows:
+\begin{equation}
+\mathbf{x} = \mathbf{A}^{-1} \, \mathbf{b}.
+\end{equation}
 
-	\subsection{Solvers}	\label{chap3:solvers}
+However, only solvers relevant to the context of finite element methods will be discussed.
+Indeed the matrix $ \mathbf{A} $ takes a specific form in problems solved by finite element methods.
+The matrix $ \mathbf{A} $ is \emph{sparse}.
+A sparse matrix is merely a matrix populated mainly with zeros.
+An example is given by Fig.~\ref{chap3:fig-sparse} for a finite element problem in two dimensions.
+\begin{figure}
+\begin{center}
+\includegraphics[width=10cm]{chapter3/sparse_matrix.pdf}
+\end{center}
+\caption{Illustration of a sparse matrix for a two-dimensional finite element problem. Each non-zero elements of the matrix is represented in black. Image courtesy of Oleg Alexandrov.}
+\label{chap3:fig-sparse}
+\end{figure}
+Because of the peculiar structure of the matrix involved in finite element computations, methods have been specifically designed to take advantage of this structure.
+Indeed, it is wasteful to use general methods of linear algebra on such problems, because most of the arithmetic operations devoted to solving the set of equations or inverting the matrix involve zero operands.
+Furthermore, it is wasteful to reserve storage for zero elements.
+
+%The choice of a resolution method depends on the problem (linear or non-linear) and the properties of the stiffness matrix.
 
 		\subsubsection*{Direct solvers}
+As their name is suggesting, direct solvers make an attempt into finding a solution directly by inverting the matrix $ \mathbf{A} $.
+They execute in a predictable number of operations.
+We will now introduce a few direct methods in the context of finite element methods.
+% Methods such as Gauss-Jordan elimination or LU decomposition will not be discussed in this section for instance for their lack of efficiency with sparse matrices.
+
+\paragraph*{LU decomposition.}
+The LU decomposition is a matrix decomposition which writes a matrix as the product of a lower triangular matrix and an upper triangular matrix. In other words, $ \mathbf{A} $ may be written as:
+\begin{equation}
+\label{chap3:LU}
+\mathbf{A} = \mathbf{L} \mathbf{U},
+\end{equation}
+where $ \mathbf{L} $ is lower triangular (has elements only on the diagonal and below) and $ \mathbf{U} $ is upper triangular (has elements only on the diagonal and above).
+As an example, for the case of a $ 4 \times 4 $ matrix $ \mathbf{A} $, we have:
+\begin{equation}
+	\begin{bmatrix}
+	a_{11} & a_{12} & a_{13} & a_{14} \\
+	a_{21} & a_{22} & a_{23} & a_{24} \\
+	a_{31} & a_{32} & a_{33} & a_{34} \\
+	a_{41} & a_{42} & a_{43} & a_{44}
+	\end{bmatrix}
+=
+	\begin{bmatrix}
+	\alpha_{11} & 0 & 0 & 0 \\
+	\alpha_{21} & \alpha_{22} & 0 & 0 \\
+	\alpha_{31} & \alpha_{32} & \alpha_{33} & 0 \\
+	\alpha_{41} & \alpha_{42} & \alpha_{43} & \alpha_{44}
+	\end{bmatrix}
+	\begin{bmatrix}
+	\beta_{11} & \beta_{12} & \beta_{13} & \beta_{14} \\
+	0 & \beta_{22} & \beta_{23} & \beta_{24} \\
+	0 & 0 & \beta_{33} & \beta_{34} \\
+	0 & 0 & 0 & \beta_{44}
+	\end{bmatrix}
+\end{equation}
+By subtituting \eqref{chap3:LU} into \eqref{chap3:linearEquation} we obtain:
+\begin{equation}
+\mathbf{A} \mathbf{x} = (\mathbf{L} \mathbf{U}) \mathbf{x} = \mathbf{L} (\mathbf{U} \mathbf{x}) = \mathbf{b}.
+\end{equation}
+Therefore, we can solve \eqref{chap3:linearEquation} by first solving the vector $ \mathbf{y} $ such that:
+\begin{equation}
+\label{chap3:eqLyb}
+\mathbf{L} \mathbf{y} = \mathbf{b}
+\end{equation}
+and then solving 
+\begin{equation}
+\label{chap3:eqUxy}
+\mathbf{U} \mathbf{x} = \mathbf{y}.
+\end{equation}
+The advantage of breaking up one linear set of equations into two successive ones is that the solution of a triangular set of equations is quite trivial. Indeed, \eqref{chap3:eqLyb} can be solved by:
+\begin{equation} 
+\left\lbrace
+	\begin{aligned}
+		y_1 &= \dfrac{b_1}{\alpha_{11}}  \\	
+		y_i &= \dfrac{1}{\alpha_{ii}} \left[ b_i - \sum_{j=1}^{i-1} \alpha_{ij} y_j \right] \quad i = 2, 3, \ldots, N.
+	\end{aligned} 
+\right.
+\end{equation}
+for a $ N \times N $ matrix $ \mathbf{A} $. We then solve \eqref{chap3:eqUxy} in a similar way:
+\begin{equation} 
+\left\lbrace
+	\begin{aligned}
+		x_N &= \dfrac{y_N}{\beta_{NN}} \\	
+		x_i &= \dfrac{1}{\beta_{ii}} \left[ y_i - \sum_{j=i+1}^{N} \beta_{ij} x_j \right] \quad i = N-1, N-2, \ldots, 1. 
+	\end{aligned} 
+\right.
+\end{equation}
+
 		\subsubsection*{Iterative solvers}
+An iterative method attempts to solve the system of equations by finding successive approximations to the solution starting from an initial guess.
+In the case of a system of linear equations, the two main classes of iterative methods are the stationary iterative methods, and the more general Krylov subspace methods.
+Stationary iterative methods solve a linear system with an operator approximating the original one and based on a measurement of the error in the result (called the residual), form an equation of correction for which this process is repeated.
+While these methods are simple to derive, convergence is only guaranteed for a limited class of matrices.
+Examples of stationary iterative methods are the Jacobi method and Gauss?Seidel method.
+Krylov subspace methods work by forming an orthogonal basis of the sequence of successive matrix powers times the initial residual (the Krylov sequence).
+The approximations to the solution are then formed by minimising the residual over the subspace formed.
+The typical method in this class is the conjugate gradient method (CG).
+In practice, for large system of equations (as often encountered in finite element analysis), direct methods would take too much time and an iterative solvers is commonly used.
+
+\paragraph*{The Gauss-Seidel method.}
+For this method, convergence is only guaranteed if the matrix is either diagonally dominant, or symmetric and positive-definite.
+In a similar way than the LU decomposition, we start by decomposing the matrix $ \mathbf{A} $ into a lower triangular component $ \mathbf{L}_ {*} $ and a strictly upper triangular component $ \mathbf{U} $:
+\begin{equation}
+\mathbf{A} = \mathbf{L}_{*} + \mathbf{U}.
+\end{equation}
+The system of linear equations \eqref{chap3:linearEquation} may be rewritten as:
+\begin{equation}
+\mathbf{L}_{*} \mathbf{x} = \mathbf{b} - \mathbf{U} \mathbf{x}
+\end{equation}
+The Gauss-Seidel method is an iterative technique that solves the left-hand side of the expression for $ \mathbf{x} $ using previous values for $ \mathbf{x} $ on the right-hand side. That is:
+\begin{equation}
+\mathbf{x}^{n+1} = \mathbf{L}_{*}^{-1} (\mathbf{b} - \mathbf{U} \mathbf{x}^n).
+\end{equation}
+Similarly to the LU decomposition method, we take advantage of the triangular form of $ \mathbf{L}_{*} $. The components of $ \mathbf{x}^{n+1} $ can be computed sequentially in the following manner:
+\begin{equation}
+x_i^{n+1} = \dfrac{1}{a_{ii}} \left[ b_i - \sum_{j>i} a_{ij} x_j^n -\sum_{j<i} a_{ij} x_j^{n+1} \right] \quad i = 1, 2, \ldots, N.
+\end{equation}
+The procedure is usually continued until the changes made by an interation are below a given threshold.
+
+\paragraph*{The conjugate gradient method.}		
+The simplest class of conjugate gradient method is based on the idea of minimising the quadratic function:
+\begin{equation}
+f(\mathbf{x}) = \frac{1}{2} \mathbf{x}^T \, \mathbf{A} \, \mathbf{x} - b^T \, \mathbf{x} + c.
+\end{equation}
+Indeed, if the matrix $ \mathbf{A} $ is symmetric and positive-definite, $ f(\mathbf{x}) $ is minimised by the solution to $ \mathbf{A} \mathbf{x} = \mathbf{b} $\citep{Shewchuk94}.
+The minimisation is carried out by generating a succession of search directions $ \mathbf{p} _k $ and improved minimisers $ \mathbf{x} _k $.
+At each stage a quantity $ \alpha_k $ is found that minimises $f(\mathbf{x}_k  + \alpha_k \mathbf{p}_k)$, and $\mathbf{x}_{k+1}$ is set equal to the new point $\mathbf{x}_k + \alpha_k \mathbf{p}_k$.
+The $ \mathbf{p}_k $  and $ \mathbf{x}_k $  are built up in such a way that $ \mathbf{x}_{k+1} $  is also the minimizer of $ f $ over the whole vector space of directions already taken, that is the space formed by the basis of vector ${\mathbf{p}_1, \mathbf{p}_2, \ldots, \mathbf{p}_k }$.
+After $N$ iterations, you arrive at the minimiser over the entire vector space, that is, the solution to our linear system of equations.
+
+Later, the \emph{biconjugate gradient method} was introduced as a generalisation of ordinary conjugate gradient to solve systems where the matrix $ \mathbf{A} $ was not necessarily symmetric and positive-definite (see \cite{Belytschko00} for details).

### 2010-08-09 Thesis/Chapter10.tex
- Update 9 August -
Added all chapters files + transition chapter2-chapter3 + summary chapter3 + removed outline



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@146 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2010-07-15 Thesis/tableofcontents.tex
@@
-\part{Introduction}
+\chapter{'Shell meshing' technique \OC{MICCAI 2010}}
+\label{chap10}
+\begin{shortAbstract}
+A short abstract for the upcoming chapter
+\end{shortAbstract}
 
-	\chapter{Medical simulation}
-		\section{General context and goal: medical training, patient-specific planning and per-operative guidance}
-		\section{Challenges (trade-off between accuracy and real-time)}
+\section{State of art: reconstruction/simplification}
 
-	\chapter{One key point in medical simulation: soft-tissue modelling}
-		\section{Necessary background in continuum mechanics}
-			\subsection{Deformation tensor and strain tensor}
-			\subsection{Stress and constitutive laws}
-		\section{Tissue characterisation}
-			\subsection{Material models for organs (non-linear, visco-elastic and anisotropic)}
-			\subsection{Measure/estimation of model parameters}
-
-	\chapter{Main principles of Finite Element Method (or how to solve equations of continuum mechanics from previous section)}
-		\section{Discretisation}
-		\section{Derivation of element equations}
-		\section{Assembly of element equations}
-		\section{Solution of global problem}
-
-\part{Solid organs modelling}
-
-	\chapter{State of art: FEM}
-
-	\chapter{Linear not accurate => Non-linear FEM => Introduction of TLED}
-		\section{Differences with classic FEM and reasons of its efficiency}
-		\section{Visco-elasticity and anisotropy added \OC{MICCAI 2008; MedIA 2009}}
-
-	\chapter{GPU implementation of TLED}
-		\section{What is GPGPU}
-		\section{Re-formulation of the algorithm for its Cg implementation}
-		\section{CUDA implementation/optimisations \OC{ISBMS 2008a}}
-
-	\chapter{Implementation in SOFA}
-		\section{Presentation of SOFA project and architecture}
-		\section{Implementation in SOFA and TLED released in open-source}
-
-\part{Hollow organs modelling}
-
-	\chapter{State of art: hollow structures}
-		\section{Non-physic approaches (computer graphics stuff)}
-		\section{Physically accurate approches (plates/shells)}
-
-	\chapter{Colonoscopy simulator project}
-		\section{Project introduction}
-		\section{Mass-spring model for colon implemented on GPU \OC{ISBMS 2008b}}
-
-	\chapter{More accurate: a co-rotational triangular shell model \OC{ISBMS 2010}}
-		\section{Model description}
-		\section{Validation}
-		\section{Application to implant deployment simulation in cataract surgery}
-
-	\chapter{'Shell meshing' technique \OC{MICCAI 2010}}
-		\section{State of art: reconstruction/simplification}
 \section{Our method}
-
-	\chapter{Applications to medical simulation}
-		\section{Nice medical stuff to show}
-		\section{Interaction solid/hollow organs}
-
-\part{Conclusion}

### 2010-08-09 Thesis/Chapter2.tex
- Update 30 July - Changed outline according to what we discussed with Stephane + increased level of details + bib updated + creation of chapter 3
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@138 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 2 August - Started chapter3 + minor changes on chapter2 + bib updated
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@139 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 9 August -
Added all chapters files + transition chapter2-chapter3 + summary chapter3 + removed outline



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/so

@@
 \chapter{Background in continuum mechanics for soft-tissue modelling}
-\label{chap:softtissue}
+\label{chap2}
 \begin{shortAbstract}
@@
 \end{equation}	
-where the summation of repeated indices is implied (Einstein's notation). The inverse relations are:
+where the summation of repeated indices is implied (Einstein's notation) and $\delta_{ij}$ is the Kronecker delta\footnote{$ \delta_{ij}$ is equal to $1$ if $i = j$ and to $0$ if $i \neq j$}.
+The inverse relations are:
 \begin{equation}
@@
 
+In other words, the resistance of a viscoelastic material to a given force depends on the velocity to which this force is applied.
+
 \TODO{Add details on constitutive equations? Level of details to be determined later (depending on TLED needs)}
@@
 
-\TODO{Transition with next chapter (complexity of models => FEM to solve them)}
+These various mechanical aspects of materials make realistic modelling of anatomical structures fairly challenging.
+In medical simulation, where modelling must occur in real-time or close to real-time, very often simplifications have to be made.
+As we will see in chapter \ref{chap4} and \ref{chap7}, numerous models are available in the literature to describe physics of objects, from fairly simple and naive approaches to more complex and thorough representations.
+However, the most common technique to solve equations of continuum mechanics is the Finite Element Method.
+Moreover, since the two main contributions of this thesis deal with finite element methods formulations, the general principles of this method will be detailled in the next section.

### 2010-08-09 Thesis/Chapter3.tex
- Update 9 August -
Added all chapters files + transition chapter2-chapter3 + summary chapter3 + removed outline



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@146 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{shortAbstract}
-Blabla
+The previous chapter explained in details how physics of anatomical structures could be described by continuun mechanics.
+When an organ is submitted to a force, the first effect of this application is a change in the organ's shape.
+This deformation can be measured using different strain tensors.
+By modelling the mechanical aspects of the material with an appropriate constitutive law, we can then deduce the stress tensor (or internal forces) at each point within the organ.
+The next step is to apply Newton's second law (law of conservation of momentum), which yields a partial differential equation at every point of the volume.
+The Finite Element Method is a numerical procedure that allows to turn an infinite number of partial differential equations into a finite number of algebraic equations.
+The method follows different steps.
+We will first begin by describing the subdivision of the domain into smaller sub-domains into which local simpler equations are derived.
+We will then discuss how all the local equations are summed over the whole domain which eventually leads to the global solution.
 \end{shortAbstract}
@@
 \item[(a)] coupled formulation in which the time $ t $ is treated as an additional coordinate along with the spatial coordinate $ x $
-\item[(b)] decoupled formulation where time and spatial variations are assumede to be separable.
+\item[(b)] decoupled formulation where time and spatial variations are assumed to be separable.
 \end{description}
@@
 where $ \mathbf{M} $ is a constant mass matrix, $\mathbf{D}$ is a constant damping matrix, $ \mathbf{K}(\mathbf{U}) $ is the stiffness matrix, which is a function of global nodal displacements $\mathbf{U}$, and $\mathbf{R}$ are externally applied loads.
-This equation is applicable to individual elements as well as the entire element assemblage.
+This equation is applicable to individual elements as well as the entire element assemblage (this is why the subscripts $ ^e $ were dropped).
 

### 2010-08-09 Thesis/Chapter4.tex
- Update 9 August -
Added all chapters files + transition chapter2-chapter3 + summary chapter3 + removed outline



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@146 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2010-07-30 Thesis/outline.tex
@@
-\part{Introduction}
+\chapter{State of art: FEM}
+\label{chap4}
+\begin{shortAbstract}
+A short abstract for the upcoming chapter
+\end{shortAbstract}
 
-	\chapter{Medical simulation}
-		\section{General context}
-			\subsection{Simulators for everything: we started by simulating what is simple}
-			\subsection{As complexity increases, applications to medical training, patient-specific planning and per-operative guidance}
-		\section{Challenges}					
-			\subsection{Photorealistic rendering}
-			\subsection{Accurate haptic feedback}
-			\subsection{Physically realist organ modelling in real-time}
+\section{Expose the problem: how to model solid organs?}
 
-	\input{chapter2}
-
-	\input{chapter3}
-
-\part{Solid organs modelling}
-
-	\chapter{State of art: FEM}
-		\section{Expose the problem: how to model solid organs?}
 \section{Techniques using geometrical kind of approaches}
@@
 	\subsection{Mass-spring}
+
 \section{Techniques based on continuum mechanics}	
@@
 	\subsection{Meshless (supposedly good for handling topological changes but other issues}
-
-	\chapter{Introduction of TLED}
-		\section{Description of the algorithm (neo-Hookean, explicit integration, independent elements, parallelisable on GPU)}
-		\section{Visco-elasticity and anisotropy added \OC{MICCAI 2008; MedIA 2009}}
-
-	\chapter{GPU implementation of TLED}
-		\section{What is GPGPU}
-		\section{Re-formulation of the algorithm for its Cg implementation}
-		\section{CUDA implementation/optimisations \OC{ISBMS 2008a}}
-		\section{Implementation into SOFA}
-			\subsection{Presentation of SOFA project and architecture}
-			\subsection{TLED in SOFA and release in open-source}
-
-\part{Hollow organs modelling}
-
-	\chapter{State of art: hollow structures}
-		\section{Expose the problem: models presented for solid organs are insufficient. How to model hollow organs?}
-			\subsection{Colonoscopy simulator project: needs for colon}
-			\subsection{Cataract surgery, stenosis: other needs for implants/blood vessels}
-		\section{Non-physic approaches (computer graphics stuff)}
-		\section{Physically accurate approches (plates/shells)}
-
-	\chapter{Mass-spring model for colon implemented on GPU \OC{ISBMS 2008b}}
-		\section{Colon is big, we need a fast model}
-		\section{Limitations (difficult to link it to actual physical properties)}
-
-	\chapter{A co-rotational triangular shell model \OC{ISBMS 2010}}
-		\section{Model description}
-		\section{Validation}
-		\section{Application to implant deployment simulation in cataract surgery}
-
-	\chapter{'Shell meshing' technique \OC{MICCAI 2010}}
-		\section{State of art: reconstruction/simplification}
-		\section{Our method}
-
-\part{Conclusion}
-	\chapter{Summary}
-	\chapter{Discussion and perspectives (Interaction solid/hollow organs)}

### 2010-08-09 Thesis/Chapter5.tex
- Update 9 August -
Added all chapters files + transition chapter2-chapter3 + summary chapter3 + removed outline



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@146 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2010-07-15 Thesis/tableofcontents.tex
@@
-\part{Introduction}
+\chapter{Introduction of TLED}
+\label{chap5}
+\begin{shortAbstract}
+A short abstract for the upcoming chapter
+\end{shortAbstract}
 
-	\chapter{Medical simulation}
-		\section{General context and goal: medical training, patient-specific planning and per-operative guidance}
-		\section{Challenges (trade-off between accuracy and real-time)}
+\section{Description of the algorithm (neo-Hookean, explicit integration, independent elements, parallelisable on GPU)}
 
-	\chapter{One key point in medical simulation: soft-tissue modelling}
-		\section{Necessary background in continuum mechanics}
-			\subsection{Deformation tensor and strain tensor}
-			\subsection{Stress and constitutive laws}
-		\section{Tissue characterisation}
-			\subsection{Material models for organs (non-linear, visco-elastic and anisotropic)}
-			\subsection{Measure/estimation of model parameters}
-
-	\chapter{Main principles of Finite Element Method (or how to solve equations of continuum mechanics from previous section)}
-		\section{Discretisation}
-		\section{Derivation of element equations}
-		\section{Assembly of element equations}
-		\section{Solution of global problem}
-
-\part{Solid organs modelling}
-
-	\chapter{State of art: FEM}
-
-	\chapter{Linear not accurate => Non-linear FEM => Introduction of TLED}
-		\section{Differences with classic FEM and reasons of its efficiency}
 \section{Visco-elasticity and anisotropy added \OC{MICCAI 2008; MedIA 2009}}
-
-	\chapter{GPU implementation of TLED}
-		\section{What is GPGPU}
-		\section{Re-formulation of the algorithm for its Cg implementation}
-		\section{CUDA implementation/optimisations \OC{ISBMS 2008a}}
-
-	\chapter{Implementation in SOFA}
-		\section{Presentation of SOFA project and architecture}
-		\section{Implementation in SOFA and TLED released in open-source}
-
-\part{Hollow organs modelling}
-
-	\chapter{State of art: hollow structures}
-		\section{Non-physic approaches (computer graphics stuff)}
-		\section{Physically accurate approches (plates/shells)}
-
-	\chapter{Colonoscopy simulator project}
-		\section{Project introduction}
-		\section{Mass-spring model for colon implemented on GPU \OC{ISBMS 2008b}}
-
-	\chapter{More accurate: a co-rotational triangular shell model \OC{ISBMS 2010}}
-		\section{Model description}
-		\section{Validation}
-		\section{Application to implant deployment simulation in cataract surgery}
-
-	\chapter{'Shell meshing' technique \OC{MICCAI 2010}}
-		\section{State of art: reconstruction/simplification}
-		\section{Our method}
-
-	\chapter{Applications to medical simulation}
-		\section{Nice medical stuff to show}
-		\section{Interaction solid/hollow organs}
-
-\part{Conclusion}

### 2010-08-09 Thesis/Chapter6.tex
- Update 9 August -
Added all chapters files + transition chapter2-chapter3 + summary chapter3 + removed outline



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@146 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2010-07-15 Thesis/tableofcontents.tex
@@
-\part{Introduction}
+\chapter{GPU implementation of TLED}
+\label{chap6}
+\begin{shortAbstract}
+A short abstract for the upcoming chapter
+\end{shortAbstract}
 
-	\chapter{Medical simulation}
-		\section{General context and goal: medical training, patient-specific planning and per-operative guidance}
-		\section{Challenges (trade-off between accuracy and real-time)}
+\section{What is GPGPU}
 
-	\chapter{One key point in medical simulation: soft-tissue modelling}
-		\section{Necessary background in continuum mechanics}
-			\subsection{Deformation tensor and strain tensor}
-			\subsection{Stress and constitutive laws}
-		\section{Tissue characterisation}
-			\subsection{Material models for organs (non-linear, visco-elastic and anisotropic)}
-			\subsection{Measure/estimation of model parameters}
+\section{Re-formulation of the algorithm for its Cg implementation}
 
-	\chapter{Main principles of Finite Element Method (or how to solve equations of continuum mechanics from previous section)}
-		\section{Discretisation}
-		\section{Derivation of element equations}
-		\section{Assembly of element equations}
-		\section{Solution of global problem}
-
-\part{Solid organs modelling}
-
-	\chapter{State of art: FEM}
-
-	\chapter{Linear not accurate => Non-linear FEM => Introduction of TLED}
-		\section{Differences with classic FEM and reasons of its efficiency}
-		\section{Visco-elasticity and anisotropy added \OC{MICCAI 2008; MedIA 2009}}
-
-	\chapter{GPU implementation of TLED}
-		\section{What is GPGPU}
-		\section{Re-formulation of the algorithm for its Cg implementation}
 \section{CUDA implementation/optimisations \OC{ISBMS 2008a}}
 
-	\chapter{Implementation in SOFA}
-		\section{Presentation of SOFA project and architecture}
-		\section{Implementation in SOFA and TLED released in open-source}
-
-\part{Hollow organs modelling}
-
-	\chapter{State of art: hollow structures}
-		\section{Non-physic approaches (computer graphics stuff)}
-		\section{Physically accurate approches (plates/shells)}
-
-	\chapter{Colonoscopy simulator project}
-		\section{Project introduction}
-		\section{Mass-spring model for colon implemented on GPU \OC{ISBMS 2008b}}
-
-	\chapter{More accurate: a co-rotational triangular shell model \OC{ISBMS 2010}}
-		\section{Model description}
-		\section{Validation}
-		\section{Application to implant deployment simulation in cataract surgery}
-
-	\chapter{'Shell meshing' technique \OC{MICCAI 2010}}
-		\section{State of art: reconstruction/simplification}
-		\section{Our method}
-
-	\chapter{Applications to medical simulation}
-		\section{Nice medical stuff to show}
-		\section{Interaction solid/hollow organs}
-
-\part{Conclusion}
+\section{Implementation into SOFA}
+	\subsection{Presentation of SOFA project and architecture}
+	\subsection{TLED in SOFA and release in open-source}

### 2010-08-09 Thesis/Chapter7.tex
- Update 9 August -
Added all chapters files + transition chapter2-chapter3 + summary chapter3 + removed outline



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@146 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2010-07-30 Thesis/outline.tex
@@
-\part{Introduction}
+\chapter{State of art: hollow structures}
+\label{chap7}
+\begin{shortAbstract}
+A short abstract for the upcoming chapter
+\end{shortAbstract}
 
-	\chapter{Medical simulation}
-		\section{General context}
-			\subsection{Simulators for everything: we started by simulating what is simple}
-			\subsection{As complexity increases, applications to medical training, patient-specific planning and per-operative guidance}
-		\section{Challenges}					
-			\subsection{Photorealistic rendering}
-			\subsection{Accurate haptic feedback}
-			\subsection{Physically realist organ modelling in real-time}
-
-	\input{chapter2}
-
-	\input{chapter3}
-
-\part{Solid organs modelling}
-
-	\chapter{State of art: FEM}
-		\section{Expose the problem: how to model solid organs?}
-		\section{Techniques using geometrical kind of approaches}
-			\subsection{Shape matching}
-		\section{Techniques relying on physics}
-			\subsection{Mass-spring}
-		\section{Techniques based on continuum mechanics}	
-			\subsection{FEM with mesh (highlight evolution: linear, topological changes, co-rotational, non-linear, GPU)}
-			\subsection{Meshless (supposedly good for handling topological changes but other issues}
-
-	\chapter{Introduction of TLED}
-		\section{Description of the algorithm (neo-Hookean, explicit integration, independent elements, parallelisable on GPU)}
-		\section{Visco-elasticity and anisotropy added \OC{MICCAI 2008; MedIA 2009}}
-
-	\chapter{GPU implementation of TLED}
-		\section{What is GPGPU}
-		\section{Re-formulation of the algorithm for its Cg implementation}
-		\section{CUDA implementation/optimisations \OC{ISBMS 2008a}}
-		\section{Implementation into SOFA}
-			\subsection{Presentation of SOFA project and architecture}
-			\subsection{TLED in SOFA and release in open-source}
-
-\part{Hollow organs modelling}
-
-	\chapter{State of art: hollow structures}
 \section{Expose the problem: models presented for solid organs are insufficient. How to model hollow organs?}
@@
 	\subsection{Cataract surgery, stenosis: other needs for implants/blood vessels}
+
 \section{Non-physic approaches (computer graphics stuff)}
+
 \section{Physically accurate approches (plates/shells)}
-
-	\chapter{Mass-spring model for colon implemented on GPU \OC{ISBMS 2008b}}
-		\section{Colon is big, we need a fast model}
-		\section{Limitations (difficult to link it to actual physical properties)}
-
-	\chapter{A co-rotational triangular shell model \OC{ISBMS 2010}}
-		\section{Model description}
-		\section{Validation}
-		\section{Application to implant deployment simulation in cataract surgery}
-
-	\chapter{'Shell meshing' technique \OC{MICCAI 2010}}
-		\section{State of art: reconstruction/simplification}
-		\section{Our method}
-
-\part{Conclusion}
-	\chapter{Summary}
-	\chapter{Discussion and perspectives (Interaction solid/hollow organs)}

### 2010-08-09 Thesis/Chapter8.tex
- Update 9 August -
Added all chapters files + transition chapter2-chapter3 + summary chapter3 + removed outline



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@146 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2010-07-30 Thesis/outline.tex
@@
-\part{Introduction}
+\chapter{Mass-spring model for colon implemented on GPU \OC{ISBMS 2008b}}
+\label{chap8}
+\begin{shortAbstract}
+A short abstract for the upcoming chapter
+\end{shortAbstract}
 
-	\chapter{Medical simulation}
-		\section{General context}
-			\subsection{Simulators for everything: we started by simulating what is simple}
-			\subsection{As complexity increases, applications to medical training, patient-specific planning and per-operative guidance}
-		\section{Challenges}					
-			\subsection{Photorealistic rendering}
-			\subsection{Accurate haptic feedback}
-			\subsection{Physically realist organ modelling in real-time}
+\section{Colon is big, we need a fast model}
 
-	\input{chapter2}
-
-	\input{chapter3}
-
-\part{Solid organs modelling}
-
-	\chapter{State of art: FEM}
-		\section{Expose the problem: how to model solid organs?}
-		\section{Techniques using geometrical kind of approaches}
-			\subsection{Shape matching}
-		\section{Techniques relying on physics}
-			\subsection{Mass-spring}
-		\section{Techniques based on continuum mechanics}	
-			\subsection{FEM with mesh (highlight evolution: linear, topological changes, co-rotational, non-linear, GPU)}
-			\subsection{Meshless (supposedly good for handling topological changes but other issues}
-
-	\chapter{Introduction of TLED}
-		\section{Description of the algorithm (neo-Hookean, explicit integration, independent elements, parallelisable on GPU)}
-		\section{Visco-elasticity and anisotropy added \OC{MICCAI 2008; MedIA 2009}}
-
-	\chapter{GPU implementation of TLED}
-		\section{What is GPGPU}
-		\section{Re-formulation of the algorithm for its Cg implementation}
-		\section{CUDA implementation/optimisations \OC{ISBMS 2008a}}
-		\section{Implementation into SOFA}
-			\subsection{Presentation of SOFA project and architecture}
-			\subsection{TLED in SOFA and release in open-source}
-
-\part{Hollow organs modelling}
-
-	\chapter{State of art: hollow structures}
-		\section{Expose the problem: models presented for solid organs are insufficient. How to model hollow organs?}
-			\subsection{Colonoscopy simulator project: needs for colon}
-			\subsection{Cataract surgery, stenosis: other needs for implants/blood vessels}
-		\section{Non-physic approaches (computer graphics stuff)}
-		\section{Physically accurate approches (plates/shells)}
-
-	\chapter{Mass-spring model for colon implemented on GPU \OC{ISBMS 2008b}}
-		\section{Colon is big, we need a fast model}
 \section{Limitations (difficult to link it to actual physical properties)}
-
-	\chapter{A co-rotational triangular shell model \OC{ISBMS 2010}}
-		\section{Model description}
-		\section{Validation}
-		\section{Application to implant deployment simulation in cataract surgery}
-
-	\chapter{'Shell meshing' technique \OC{MICCAI 2010}}
-		\section{State of art: reconstruction/simplification}
-		\section{Our method}
-
-\part{Conclusion}
-	\chapter{Summary}
-	\chapter{Discussion and perspectives (Interaction solid/hollow organs)}

### 2010-08-09 Thesis/Chapter9.tex
- Update 9 August -
Added all chapters files + transition chapter2-chapter3 + summary chapter3 + removed outline



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@146 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2010-07-15 Thesis/tableofcontents.tex
@@
-\part{Introduction}
+\chapter{A co-rotational triangular shell model \OC{ISBMS 2010}}
+\label{chap9}
+\begin{shortAbstract}
+A short abstract for the upcoming chapter
+\end{shortAbstract}
 
-	\chapter{Medical simulation}
-		\section{General context and goal: medical training, patient-specific planning and per-operative guidance}
-		\section{Challenges (trade-off between accuracy and real-time)}
+\section{Model description}
 
-	\chapter{One key point in medical simulation: soft-tissue modelling}
-		\section{Necessary background in continuum mechanics}
-			\subsection{Deformation tensor and strain tensor}
-			\subsection{Stress and constitutive laws}
-		\section{Tissue characterisation}
-			\subsection{Material models for organs (non-linear, visco-elastic and anisotropic)}
-			\subsection{Measure/estimation of model parameters}
+\section{Validation}
 
-	\chapter{Main principles of Finite Element Method (or how to solve equations of continuum mechanics from previous section)}
-		\section{Discretisation}
-		\section{Derivation of element equations}
-		\section{Assembly of element equations}
-		\section{Solution of global problem}
-
-\part{Solid organs modelling}
-
-	\chapter{State of art: FEM}
-
-	\chapter{Linear not accurate => Non-linear FEM => Introduction of TLED}
-		\section{Differences with classic FEM and reasons of its efficiency}
-		\section{Visco-elasticity and anisotropy added \OC{MICCAI 2008; MedIA 2009}}
-
-	\chapter{GPU implementation of TLED}
-		\section{What is GPGPU}
-		\section{Re-formulation of the algorithm for its Cg implementation}
-		\section{CUDA implementation/optimisations \OC{ISBMS 2008a}}
-
-	\chapter{Implementation in SOFA}
-		\section{Presentation of SOFA project and architecture}
-		\section{Implementation in SOFA and TLED released in open-source}
-
-\part{Hollow organs modelling}
-
-	\chapter{State of art: hollow structures}
-		\section{Non-physic approaches (computer graphics stuff)}
-		\section{Physically accurate approches (plates/shells)}
-
-	\chapter{Colonoscopy simulator project}
-		\section{Project introduction}
-		\section{Mass-spring model for colon implemented on GPU \OC{ISBMS 2008b}}
-
-	\chapter{More accurate: a co-rotational triangular shell model \OC{ISBMS 2010}}
-		\section{Model description}
-		\section{Validation}
 \section{Application to implant deployment simulation in cataract surgery}
-
-	\chapter{'Shell meshing' technique \OC{MICCAI 2010}}
-		\section{State of art: reconstruction/simplification}
-		\section{Our method}
-
-	\chapter{Applications to medical simulation}
-		\section{Nice medical stuff to show}
-		\section{Interaction solid/hollow organs}
-
-\part{Conclusion}

### 2010-08-09 Thesis/Chapter2.tex
- Update 9 August -
Added images for chapter 2 and 3



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@147 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 We can define volume elements in the reference and deformed configurations.
-Consider three non-coplanar line elements $d\mathbf{X}^{(1)}$, $d\mathbf{X}^{(2)}$ and $d\mathbf{X}^{(3)}$ forming a parallelepiped in the reference configuration \TODO{Add figure with all notations}.
+Consider three non-coplanar line elements $d\mathbf{X}^{(1)}$, $d\mathbf{X}^{(2)}$ and $d\mathbf{X}^{(3)}$ forming a parallelepiped in the reference configuration as can be seen on Fig.~\ref{chap2:fig-changeVolume}.
+% 
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=13cm]{chapter2/changeVolume.pdf}
+\end{center}
+\caption{Transformation of a volume element under a deformation mapping. \TODO{CHANGE TO A BETTER FIGURE.}}
+\label{chap2:fig-changeVolume}
+\end{figure}
+%
+
 The three vectors after deformation $d\mathbf{x}^{(1)}$, $d\mathbf{x}^{(2)}$ and $d\mathbf{x}^{(3)}$ can be obtained with:
@@
 For similar reasons, let us find the relation between an element of area in the reference configuration $d\mathbf{A}$ which becomes $d\mathbf{a}$ in the deformed configuration.
-Considering an arbitrary material line $d\mathbf{L}$ in the reference configuration and noting $d\mathbf{l}$ the same line after deformation \TODO{Add figure with all notations}, the corresponding reference and current volumes are given respectively by:
+Considering an arbitrary material line $d\mathbf{L}$ in the reference configuration and noting $d\mathbf{l}$ the same line after deformation (see Fig.~\ref{chap2:fig-changeSurface}, the corresponding reference and current volumes are given respectively by:
 \begin{align}
@@
 \end{equation}
+% 
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=13cm]{chapter2/changeSurface.pdf}
+\end{center}
+\caption{Transformation of a surface element under a deformation mapping. \TODO{CHANGE TO A BETTER FIGURE.}}
+\label{chap2:fig-changeSurface}
+\end{figure}
 
@@
 
-To establish this relationship, let us consider an infinitesimal tetrahedron in a Cartesian coordinate system.
-\TODO{Add figure of a tetrahedron with all notations}.
+To establish this relationship, let us consider an infinitesimal tetrahedron in a Cartesian coordinate system as represented Fig.~\ref{chap2:fig-CauchyStress}. 
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=13cm]{chapter2/CauchyStress.pdf}
+\end{center}
+\caption{Tetrahedral element in Cartesian coordinates. \TODO{CHANGE TO A BETTER FIGURE.}}
+\label{chap2:fig-CauchyStress}
+\end{figure}
+%
 We note $-\mathbf{t}_1$, $-\mathbf{t}_2$, $-\mathbf{t}_3$ and $\mathbf{t}$ the stress vectors in the outwards directions on the faces of the tetrahedron whose areas are respectively $\Delta a_1$, $\Delta a_2$, $\Delta a_3$ and $\Delta a$.

### 2010-08-09 Thesis/Chapter3.tex
- Update 9 August -
Added images for chapter 2 and 3



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@147 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\chapter{Practical approach of the Finite Element Method}
+\chapter{A practical approach of the Finite Element Method}
 \label{chap3}
@@
 Using assumptions concerning how the phenomena works and using the appropriate laws of physics governing the process, they can derive a mathematical model, often characterised by complex differential or integral equations relating various quantities of interest.
-However, because of the complexities in the geometry and complex boundary conditions found in real life problems, we cannot analytically solve these equations.
+However, because of the complexities in the geometry and complex boundary conditions found in real life problems, they cannot analytically solve these equations.
 A few decades ago, the only possible approach was to drastically simplify them, which was not always sufficient to find an approximate solution.
@@
 However, sometimes the function describing the curve is not known, or is difficult to integrate.
-One method to obtain an approximate solution is to break up the area into a series of rectangles and add the areas of all rectangles \TODO{Add figure with curve and rectangles}.
+One method to obtain an approximate solution is to break up the area into a series of rectangles and add the areas of all rectangles (see Fig.~\ref{chap3:fig-areaUnderCurve}).
 It is worth nothing that the solution accuracy can be increased by reducing the width of the rectangles to better follow the curve.
-
-\bigskip
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=8cm]{chapter3/areaUnderCurve.pdf}
+\end{center}
+\caption[Approximation of the area under a curve]{Approximation of the area under a curve by adding the areas of all rectangles (left Rieman sum).}
+\label{chap3:fig-areaUnderCurve}
+\end{figure}
 
 It is crucial to keep in mind that approximations occur at different stages during finite element analysis.
-The division of the whole domain into finite elements may not be exact \TODO{add figure from a domain divided into elements}, introducing error in the domain being modelled.
+The division of the whole domain into finite elements may not be exact (see Fig.~\ref{chap3:fig-errorWithSubdivision}), introducing error in the domain being modelled.
 The second stage is when element equations are derived.
@@
 Finally, errors are introduced when solving the assembled system of algebraic equations.
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=8cm]{chapter3/errorWithSubdivision.pdf}
+\end{center}
+\caption[Error introduced by the division into elements]{The division of the domain into elements is not always perfect.}
+\label{chap3:fig-errorWithSubdivision}
+\end{figure}
 
@@
 Usually, the number of elements increases with the complexity of the geometry.
-For instance, if one part of the domain is thiner \TODO{add figure to illustrate this point}, the size of the elements must be reduced in order to tile this particular part, which increases the total number of elements, and hence  the number of algebraic equations to solve.
+For instance, if one part of the domain is thiner (see Fig.~\ref{chap3:fig-sizeOfElementMeshing}), the size of the elements must be reduced in order to tile this particular part, which increases the total number of elements, and hence  the number of algebraic equations to solve.
 However, adding elements is sometimes desirable.
 Indeed, increasing the number of elements tends to get an approximate solution closer to the exact analytical solution (as reducing the width of the rectangles allows for a better approximation of the area under the curve).
-Elements are classically added in the particular regions of interest for a given problem and where substantial deformation may occur if this information is known a priori.
+Elements are classically added in the particular regions of interest for a given problem, if this information is known a priori.
 A trade-off between accuracy and computational time must be found.
 This is why the meshing process on the problem domain must be carried out carefully.
-One needs to create a mesh which gives an accurate enough solution for the desired application while restraining the computational time according to the problem time constraints.
+One needs to create a mesh which gives an accurate enough solution for the desired application while restraining the computational time according to the time constraints of the problem.
 As an example, finding the maximum load that can sustain a bridge in structural mechanics demands a high degree of accuracy, no matter how much time it is needed to compute the solution.
@@
 It does not mean that precision is not required, simply that the time constraints will necessarily limit the accuracy of the solution.
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=13cm]{chapter3/sizeOfElementMeshing.pdf}
+\end{center}
+\caption[Finer mesh required for complex geometry]{By adding details on the tree's geometry (right), the surface delimited by the tree becomes more difficult to mesh and would require smaller elements to tile all additional small branches.}
+\label{chap3:fig-sizeOfElementMeshing}
+\end{figure}
 
@@
 The natural coordinate system allows us to map every element into a typical and simpler element.
-As an example, an 8-node hexahedron element is shown in Fig.
-\TODO{add figure of hexahedron in global and natural coordinates} as it appears in the global $(x_1, x_2, x_3)$ and natural $(\xi_1,\xi_2,\xi_3)$ coordinate systems.
+As an example, an 8-node hexahedron element is shown in Fig.~\ref{chap3:fig-naturalCoordinatesHexa} as it appears in the global $(x_1, x_2, x_3)$ and natural $(\xi_1,\xi_2,\xi_3)$ coordinate systems.
 While in its natural coordinate system the element is a regular aligned cube, in the global coordinate system the element may assume any admissible arbitrary form.
 Essentially, admissible means that the element must not be too distorted, and must certainly not fold back on itself.
-Another example is given Fig.
-\TODO{add figure of tetrahedron in global and natural coordinates} with a tetrahedral element.
+Another example is given Fig.~\ref{chap3:fig-naturalCoordinatesTetra} with a tetrahedral element.
 The components of each node has a simple expression in natural coordinates, usually $0, 1$ or $-1$ and the centre of the coordinate system is taken at one of the nodes or at the centre of the element.
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=13cm]{chapter3/naturalCoordinatesHexa.pdf}
+\end{center}
+\caption[Natural coordinates of a hexaedron]{Hexahedral element as it appears in the global coordinate system (a) and its natural coordinate system (b).}
+\label{chap3:fig-naturalCoordinatesHexa}
+\end{figure}
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=13cm]{chapter3/naturalCoordinatesTetra.pdf}
+\end{center}
+\caption[Natural coordinates of a tetrahedron]{Tetrahedral element as it appears in the global coordinate system (a) and its natural coordinate system (b).}
+\label{chap3:fig-naturalCoordinatesTetra}
+\end{figure}
 
@@
 Due to the weaker requirement on the field variables, and the integral operation, a formulation based on a weak form usually produces a set of discretised system equations that give much more accurate results, especially for problems of complex geometry.
-Hence, the weak form is preferred by many for obtaining an approximate solution.
+Hence, the weak form is usually preferred for obtaining an approximate solution.
 Using the weak form usually leads to a set of well-behaved algebraic system equations, if the problem domain is discretised properly into elements.
@@
 \label{chap3:eqDynamic}
-\mathbf{M} \mathbf{\ddot U} + \mathbf{D} \mathbf{ \dot U} + \mathbf{K}(\mathbf{U}) \cdot \mathbf{U} = \mathbf{R},
+\mathbf{M}^e \mathbf{\ddot U}^e + \mathbf{D}^e \mathbf{ \dot U}^e + \mathbf{K}^e(\mathbf{U}^e) \cdot \mathbf{U}^e = \mathbf{R}^e,
 \end{equation}
-where $ \mathbf{M} $ is a constant mass matrix, $\mathbf{D}$ is a constant damping matrix, $ \mathbf{K}(\mathbf{U}) $ is the stiffness matrix, which is a function of global nodal displacements $\mathbf{U}$, and $\mathbf{R}$ are externally applied loads.
-This equation is applicable to individual elements as well as the entire element assemblage (this is why the subscripts $ ^e $ were dropped).
+where $ \mathbf{M}^e $ is a constant mass matrix, $\mathbf{D}^e$ is a constant damping matrix, $ \mathbf{K}^e(\mathbf{U}^e) $ is the stiffness matrix, which is a function of nodal displacements $\mathbf{U}^e$, and $\mathbf{R}^e$ are externally applied loads.
+All matrices are expressed at the level of individual elements (which is the reason for the superscript $ ^e $).
 
@@
 \label{chap3:eqStatic}
-\mathbf{K}(\mathbf{U}) \cdot \mathbf{U} = \mathbf{R}.
+\mathbf{K}^e (\mathbf{U}^e ) \cdot \mathbf{U}^e  = \mathbf{R}^e .
 \end{equation}
@@
 Thus, it becomes necessary to put the elements together to eliminate the extra unknowns.
-The assembly should be carried out as soon as the element matrices are computed, rather than waiting until the element coefficients of all elements are computed.
+The assembly of the global stiffness matrix should be carried out as soon as the element matrices are computed, rather than waiting until the element coefficients of all elements are computed.
 This would require storage of the coefficients for each element.
@@
 \end{table}
-
-% The element matrices are expressed with respect to local node numbers.
-\TODO{add figure with global stiffness matrix and a few arrows issued from the element stiffness matrices}
-
-\section{Solution of global problem}
-The complete and global system of equations has now been established.
-Let us recall the general system of equations in dynamic introduced in section \ref{chap3:dynamic}, which is also valid for the whole system:
+We also assume the following stiffness matrices for the three elements:
+\begin{equation}
+\mathbf{K}^1 = 
+\color{blue}
+	\begin{bmatrix}
+	-1 & 2 \\
+	3 & 1
+	\end{bmatrix}
+\color{black}
+\text{, } \quad 
+\mathbf{K}^2 = 
+\color{green}
+	\begin{bmatrix}
+	5 & 2 \\
+	-2 & 1
+	\end{bmatrix}
+\color{black}
+\text{ and } \quad 
+\mathbf{K}^3 = 
+\color{red}
+	\begin{bmatrix}
+	3 & 1 \\
+	2 & -1
+	\end{bmatrix}
+\color{black}
+\end{equation}
+The assembled matrix is a square matrix of dimension number of nodes $\times$ number of degrees of freedom.
+In our case, the global matrix is therefore a $ 4 \times 4 $ and is assembled in the following manner:
+\begin{equation}
+\mathbf{K} = 
+	\begin{bmatrix}
+	\color{blue}{-1} & \color{blue}{2} & \color{black}{0} & 0 \\
+	\color{blue}{3} & \color{blue}{1}\color{black}{+}\color{green}{5} & \color{green}{2} & \color{black}{0} \\
+	0 & \color{green}{-2} & \color{green}{1} \color{black}{+} \color{red}{3} & \color{red}{1} \\
+	\color{black}{0} & 0 & \color{red}{2} & \color{red}{-1}
+	\end{bmatrix}
+\color{black}
+\end{equation}
+In other words, the global matrix is merely a rewritting for the whole domain of the element stiffness matrices, which are expressed with respect to local node numbers, in terms of global node numbers.
+In addition to constructing the global mass and damping matrices (as shown in section \ref{chap3:wordOnMatrices}), this process leads to the following global system of equations:
 \begin{equation}
@@
 \end{equation}	
+The same set of equations was established at the element level in section \ref{chap3:dynamic} but we may note that the same system is also valid for the whole domain.
+
+\section{Solution of global problem}
+The complete and global system of equations has now been established.
 In this section, we will examine the techniques commonly used for solving these equations.
@@
 \label{chap3:eqToSolve}
-y^{'}(t) = f(t, y(t))
+\mathbf{\dot U}^t = f(t, \mathbf{U}^t).
 \end{equation}
-Note that we restrict ourselves to first-order differential equations (meaning that only the first derivative of $y$ appears in the equation, and no higher derivatives).
+Note that we restrict ourselves to first-order differential equations (meaning that only the first derivative of $\mathbf{U}^t$ appears in the equation, and no higher derivatives).
 However, a higher-order equation can easily be converted to a system of first-order equations by introducing extra variables.
-For example, the second-order equation $y^{''} = -y$ can be rewritten as two first-order equations: $y^{'} = z$ and $z^{'} = -y$.
+For example, the second-order equation $\mathbf{\ddot U}^t = -\mathbf{U}^t$ can be rewritten as two first-order equations: $\mathbf{\dot U}^t = \mathbf{V}^t$ and $\mathbf{\dot V}^t = -\mathbf{U}^t$.
 By solving equations like \eqref{chap3:eqToSolve}, we will therefore be able to integrate the global system of equations over time.
 
-We start by replacing the derivative $y^{'} $ by the finite difference approximation:
+We start by replacing the derivative $\mathbf{\dot U}^t $ by the finite difference approximation:
 \begin{equation}
 \label{chap3:approxExplicit}
-y^{'}(t) \approx \dfrac{y(t+\Delta t) - y(t)}{\Delta t},
+\mathbf{\dot U}^t \approx \dfrac{\mathbf{U}^{t+\Delta t} - \mathbf{U}^t}{\Delta t},
 \end{equation}
@@
 \begin{equation}
-y(t+\Delta t) \approx y(t) + \Delta t y^{'}(t)
+\mathbf{U}^{t+\Delta t} \approx \mathbf{U}^t + \Delta t \, \mathbf{\dot U}^t
 \end{equation}
@@
 \label{chap3:updateExplicit}
-y(t+\Delta t) \approx y(t) + \Delta t f(t, y(t))
+\mathbf{U}^{t+\Delta t} \approx \mathbf{U}^t + \Delta t \, f(t, \mathbf{U}^t)
 \end{equation}
@@
 We choose a time step $ \Delta t $ and we construct the sequence of time $ t^n = n \Delta t$.
-Using \eqref{chap3:updateExplicit}, we can approximate the exact solution $ y(t^n) $ at each time step by $ y^n $:
+Using \eqref{chap3:updateExplicit}, we can approximate the exact solution $ \mathbf{U}^{t^n} $ at each time step by $ \mathbf{U}^n $:
 \begin{equation}
-y^{n+1} = y^n + \Delta t f(t^n, y^n).
+\mathbf{U}^{n+1} = \mathbf{U}^n + \Delta t \, f(t^n, \mathbf{U}^n).
 \end{equation}
-This method is called \emph{Euler method} and is said to be explicit because the new value $ y^{n+1} $ is defined in terms of variables from the previous time step ($ y^n $).
+This method is called \emph{Euler method} and is said to be explicit because the new value $ \mathbf{U}^{n+1} $ is defined in terms of variables from the previous time step $ \mathbf{U}^n $.
 
@@
 \begin{equation}
-y^{'}(t) = f(t, y(t))
+\mathbf{\dot U}^t = f(t, \mathbf{U}^t).
 \end{equation}
-This time, instead of using \eqref{chap3:approxExplicit}, we approximate the derivative $ y^{'}(t) $ by:
+This time, instead of using \eqref{chap3:approxExplicit}, we approximate the derivative $ \mathbf{\dot U}^t $ by:
 \begin{equation}
-y^{'}(t) \approx \dfrac{y(t) - y(t-\Delta t)}{\Delta t}.
+\mathbf{\dot U}^t \approx \dfrac{\mathbf{U}^t - \mathbf{U}^{t-\Delta t}}{\Delta t}.
 \end{equation}
@@
 \begin{equation}
-y^{n+1} = y^n + \Delta t f(t^{n+1}, y^{n+1}).
+\mathbf{U}^{n+1} = \mathbf{U}^n + \Delta t \, f(t^{n+1}, \mathbf{U}^{n+1}).
 \end{equation}
-The \emph{backward Euler method} is an implicit method, meaning that we have to solve an equation to find $ y^{n+1} $.
+The \emph{backward Euler method} is an implicit method, meaning that we have to solve an equation to find $ \mathbf{U}^{n+1} $.
 Of course, it costs time to solve this equation and this cost must be taken into consideration when we select the time integration method to use.
@@
 \begin{equation}
-y^{n+1} = y^{n} + \frac{1}{6} \Delta t \left( k_1 + 2 k_2 + 2 k_3 +k_4 \right),
+\mathbf{U}^{n+1} = \mathbf{U}^{n} + \frac{1}{6} \Delta t \left( k_1 + 2 k_2 + 2 k_3 +k_4 \right),
 \end{equation}
@@
 \begin{align}
-k_1 &= f(t^n, y^n) \notag \\
-k_2 &= f(t^n + \frac{1}{2} \Delta t, y^n + \frac{1}{2} \Delta t \, k_1) \notag \\
-k_3 &= f(t^n + \frac{1}{2} \Delta t, y^n + \frac{1}{2} \Delta t \, k_2) \notag \\
-k_4 &= f(t^n + \Delta t, y^n + \Delta t \, k_3) \notag
+k_1 &= f(t^n, \mathbf{U}^n) \notag \\
+k_2 &= f(t^n + \frac{1}{2} \Delta t, \mathbf{U}^n + \frac{1}{2} \Delta t \, k_1) \notag \\
+k_3 &= f(t^n + \frac{1}{2} \Delta t, \mathbf{U}^n + \frac{1}{2} \Delta t \, k_2) \notag \\
+k_4 &= f(t^n + \Delta t, \mathbf{U}^n + \Delta t \, k_3) \notag
 \end{align}
@@
 It may be seen as a generalisation of the backward Euler method \citep{Press92}.
-Again, we have to solve an equation to find $ y^{n+1} $.
+Again, we have to solve an equation to find $ \mathbf{U}^{n+1} $.
 
@@
 \begin{center}
-\includegraphics[width=10cm]{chapter3/sparse_matrix.pdf}
+\includegraphics[width=13cm]{chapter3/sparse_matrix.pdf}
 \end{center}
-\caption{Illustration of a sparse matrix for a two-dimensional finite element problem. Each non-zero elements of the matrix is represented in black. Image courtesy of Oleg Alexandrov.}
+\caption[Illustration of a sparse matrix for a 2D finite element problem]{Illustration of a sparse matrix for a two-dimensional finite element problem.
+Each non-zero elements of the matrix is represented in black.
+Image courtesy of Oleg Alexandrov.}
 \label{chap3:fig-sparse}

### 2010-08-10 Thesis/Chapter3.tex
- Update 10 August -
Started chapter4 + few fixes on chapter3 + bib updated



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@148 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 However, since the element is physically connected to its neighbours, the resulting algebraic equations will contain more unknowns than the number of algebraic equations.
-Thus, it becomes necessary to put the elements together to eliminate the extra unknowns.
+In other words, elements share vertices.
+The deformation of an element induces the deformation of its neighbours.
+Thus, it becomes necessary to put the elements together to eliminate the extra unknowns, to model this interaction between elements.
 The assembly of the global stiffness matrix should be carried out as soon as the element matrices are computed, rather than waiting until the element coefficients of all elements are computed.
@@
 \end{equation}
-In other words, the global matrix is merely a rewritting for the whole domain of the element stiffness matrices, which are expressed with respect to local node numbers, in terms of global node numbers.
+In other words, the global matrix is merely a rewritting on the whole domain of the element stiffness matrices, which are expressed with respect to local node numbers, in terms of global node numbers.
 In addition to constructing the global mass and damping matrices (as shown in section \ref{chap3:wordOnMatrices}), this process leads to the following global system of equations:
@@
 
-Later, the \emph{biconjugate gradient method} was introduced as a generalisation of ordinary conjugate gradient to solve systems where the matrix $ \mathbf{A} $ was not necessarily symmetric and positive-definite (see \cite{Belytschko00} for details).
+Later, the \emph{biconjugate gradient method} was introduced as a generalisation of the ordinary conjugate gradient to solve systems where the matrix $ \mathbf{A} $ was not necessarily symmetric and positive-definite (see \cite{Belytschko00} for details).

### 2010-08-10 Thesis/Chapter4.tex
- Update 10 August -
Started chapter4 + few fixes on chapter3 + bib updated



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@148 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\chapter{State of art: FEM}
+\chapter{Modelling the deformation of solid objects in real-time}
 \label{chap4}
@@
 
-\section{Expose the problem: how to model solid organs?}
+\section{Introduction: the problematic}
 
-\section{Techniques using geometrical kind of approaches}
+At the beginning of computer simulation, users were immersed into environments that were merely a decor.
+If it could virtually take the user somewhere else, a static scene quickly showed its limits.
+Indeed, in a real-life environment many objects are in motion, interacting with each others.
+Therefore, in order to increase the fidelity of simulations, elements of physics were progressively added.
+It could be as simple as adding motion to clouds in a flight simulator for instance.
+But by improving the dynamism of the scene, the simulation appears more natural to users.
+Over time, with the increase in computational power of computers, objects stopped being all rigid and simulators started taking laws of physics into account.
+In other words, they became deformable.
+If this took the realism a step further, demands in computational power were multiplied.
+Simulators eventually faced a challenge: they had to weight the desired degree of realism against the computational power at their disposal.
+
+In the field of medical simulation, the correctness of deformation is often crucial.
+This is particulary true in the cases of per-operative guidance.
+As an example, removing a brain tumor relies heavily on knowing the spatial relation between the patient's brain and the images acquired prior to surgery.
+Unfortunately, drilling a hole into the patient's skull releases pressure and the brain deforms.
+This phenomenon is called brain shift.
+The overall deformation is quite complex and the displacements depend on local elasticity of the brain, the size and the location of the tumor.
+A large brain shift, if not corrected, will result in inaccuracies in the surgical procedure and has the potential to cause damage to normal tissue.
+A solution is to accurately model the mechanical response of the patient's brain to predict the actual deformation and hence keep the knowledge of the tumor's location.
+
+Conversely, when medical simulation is applied to training, such a precision in terms of distance is not required.
+The physician would not even visually notice reasonable errors as soon as it looks realistic enough.
+However, in some medical procedures, the sense of touch and the forces that the physician can feel are a fey factor for the success of the operation.
+For instance, during a colonoscopy procedure, force feedback gives precious information about what type of loop is forming and the physician can then react accordingly to prevent the loop formation.
+Consequently, a good colonoscopy simulator is required to provide accurate force feedbacks.
+If the contact forces were not reproduced to the operator, he would never be able to learn how to detect loop forming.
+Obviously, the computation of a realistic force to be returned demands an appropriate mechanical modelling of all structures.
+
+If the reasons differ, the need for modelling the mechanical response of organs with precision remains.
+However, strong time constraints often limit the complexity of the modelling and affects the overall accuracy.
+Because of this trade-off, various approaches were proposed over the years to fit into real-time constraints and may be grouped into three main categories: (1) geometrically based techniques, (2) approaches physically motivated usually relying on Newton's second law and (3) methods actually based on the equations of continuum mechanics.
+
+\section{Techniques based on geometry}
+
 	\subsection{Shape matching}
+One field particulary interested in modelling deformable objects in a very efficient way is the game industry.
+Of course, for games we are more attracted by computational efficiency and extreme stability features than accuracy to physics laws.
+Most of the time, we can tolerate a degradation of realism as long as the result looks realistic.
+\cite{Muller05} developed a technique with this idea in mind called \emph{shape matching}.
+We start with a set of particules with masses $ m_i $ in an initial configuration where we denote by $ \mathbf{x}_i^0 $ the initial positions of the particules.
+No connectivity information is required.
+The particles are simulated as a simple particle system without particle-particle interactions, but including response to collisions with the environment and including external forces such as gravity.
+The positions in the deformed configuration are noted $ \mathbf{x}_i $.
+The method consists of finding a set of points $ \mathbf{g}_i $ which minimises the difference between the two sets $ \mathbf{x}_i^0 $ and $ \mathbf{x}_i $.
+The first step is to find the rotation matrix $ \mathbf{R} $ and the translation vectors $ \mathbf{t} $ and $ \mathbf{t}_0 $ such that
+\begin{equation}
+\sum_i m_i (\mathbf{R}(\mathbf{x}_i^0 - \mathbf{t}_0) + \mathbf{t} - \mathbf{x}_i)^2
+\end{equation}
+is minimal.
+The optimal translation vectors $ \mathbf{t} $ and $ \mathbf{t}_0 $ turn out to be the centre of mass of the initial shape and actual shape, that we will note $ \mathbf{x}_{c}^0 $ and $ \mathbf{x}_c $, respectively.
+The optimal rotation matrix is found by first finding the optimal linear transformation $ \mathbf{A} $.
+The optimal rotation is eventually obtained through the rotational part of $ \mathbf{A} $ after a polar decomposition.
+Finally, the goal positions $ \mathbf{g}_i $ can be computed as follows:
+\begin{equation}
+\mathbf{g}_i = \mathbf{R}(\mathbf{x}_i^0 - \mathbf{x}_{c}^0) + \mathbf{x}_{c}.
+\end{equation}
+Once the goal positions are known, they are used to integrate the positions of the particules:
+\begin{equation}  
+	\begin{cases} 
+		\mathbf{v}_i(t+h) = \mathbf{v}_i(t) + \alpha \dfrac{\mathbf{g}_i(t) - \mathbf{x}_i(t)}{h}  + h \dfrac{f_{\text{ext}}(t)}{m_i} \\\\
+		\mathbf{x}_i(t+h) = \mathbf{x}_i(t) + h \mathbf{v}_i(t+h)
+	\end{cases}
+\end{equation}
+where $ \alpha $ is a parameter between $ 0 $ and $ 1 $ which simulates stiffness.
+
 \section{Techniques relying on physics}
+
 	\subsection{Mass-spring}
@@
 \section{Techniques based on continuum mechanics}
+
 	\subsection{FEM with mesh (highlight evolution: linear, topological changes, co-rotational, non-linear, GPU)}
+
 	\subsection{Meshless (supposedly good for handling topological changes but other issues}

### 2010-08-11 Thesis/Chapter4.tex
- Update 11 August -
Added free-form deformation, shape matching and mass-spring systems + big update for bib



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@149 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 However, in some medical procedures, the sense of touch and the forces that the physician can feel are a fey factor for the success of the operation.
-For instance, during a colonoscopy procedure, force feedback gives precious information about what type of loop is forming and the physician can then react accordingly to prevent the loop formation.
+For instance, during a colonoscopy procedure, force feedback gives precious information about what type of loop is forming and the physician can then react accordingly to prevent loop formation.
 Consequently, a good colonoscopy simulator is required to provide accurate force feedbacks.
@@
 \section{Techniques based on geometry}
+
+	\subsection{Free-form deformation}
+The first needs to deforming solid objects came from for the field of solid modelling.
+The goal was to allow the designer to shape an object into the form he wishes in a similar way that clay is manipulated by a sculptor's hands.
+\cite{Sederberg86} introduced a technique called \emph{free-form deformation} based on the paper of \cite{Barr84}.
+It involves a mapping from $ \field{R}^3 $ to $ \field{R}^3 $ through a tensor product trivariate Bernstein polynomial.
+The deformed shape of the object is interpolated using control points as in the case of Bézier curves and surfaces patches (see Fig.~{chap4:fig-cubeFFD}).
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=6.5cm]{chapter4/cubeFFD1.pdf}
+\includegraphics[width=6.6cm]{chapter4/cubeFFD2.pdf}
+\end{center}
+\caption[Deformation of objects by free-form deformation]{Let us imagine that the objects that we want to deform (spheres and cubes) are placed into a transparent parallelepiped.
+Control points are indicated by small white diamonds.
+The free-form deformation technique offers to compute the deformation of the objects from the positions of the control points (right).}
+\label{chap4:fig-cubeFFD}
+\end{figure}
+The capability of locally applying deformations makes the technique strongly analogous to sculpting with clay, which was the sought target.
+Although the authors do not give details on computational efficiency, their work constitutes the beginnings of deformable solid objects.
+
+Coquillard later developed the idea of Extented free-form deformation \citeyearpar{Coquillart90} and Animated free-form technique \citeyearpar{Coquillart91}.
+The idea of free-form deformation with lattices of arbitrary topology was introduced by \cite{MacCracken96}.
+Finally, \cite{Schein04} presented a technique to facilitate the incorporation of discontinuities in a model while deforming it.
 
@@
 \begin{equation}
+\label{chap4:gi}
 \mathbf{g}_i = \mathbf{R}(\mathbf{x}_i^0 - \mathbf{x}_{c}^0) + \mathbf{x}_{c}.
@@
 where $ \alpha $ is a parameter between $ 0 $ and $ 1 $ which simulates stiffness.
+Because the method in this basic form is not suitable for objects undergoing large deformations, it is then extended to linear and quadratic deformations by replacing $ \mathbf{R} $ in \eqref{chap4:gi} with combinations of the form $ \beta \mathbf{A} + (1-\beta) \mathbf{R}$ where $ \beta $ is an additional parameter.
+Linear transformation can represent shear and stretch while quadratic transformation add twist and bending.
+Of course, this technique is not physically realistic.
+But this is beyond the point.
+This method allows the authors to simulate hundreds of deformable objects at an interactive rate with an unconditional stability (see Fig.~\ref{chap4:fig-shoes}).
+Another illustration of this stability is shown Fig.~\ref{chap4:fig-duck}.
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=13.9cm]{chapter4/shoes.pdf}
+\end{center}
+\caption[Shape matching technique]{Shape matching allows to simulate hundreds of deformable objects in real-time.}
+\label{chap4:fig-shoes}
+\end{figure}
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=11cm]{chapter4/duck1.pdf} \\
+\vspace{0.1cm}
+\includegraphics[width=11cm]{chapter4/duck2.pdf}
+\end{center}
+\caption[Squeezing of a duck using Shape Matching]{A duck model is squeezed to demonstrate the stability of the shape matching technique.
+Although the flat shape of the duck is not physically plausible, this example exposes the impressive ability of the approach to recover from highly deformed or inverted configurations.}
+\label{chap4:fig-duck}
+\end{figure}
+
+Later, \cite{Rivers07} extented the technique to regular lattices with embedded geometry.
+From a surface mesh to be deformed, they start by voxelising the model to construct a lattice of cubic cells containing the mesh.
+The embedded mesh is then deformed using trilinear interpolation of lattice vertex positions.
+A particule is placed at each vertex of the lattice and associated with a shape matching region.
+Their technique of \emph{Fast Lattice Shape Matching} allows them to compute per-region transformations in an efficient manner (see Fig.~\ref{chap4:fig-penguins}).
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=13.9cm]{chapter4/penguins.pdf}
+\end{center}
+\caption[Penguins animation with Fast Lattice Shape Matching]{(Left) $150$ cartoon penguins deforming dynamically using Fast Lattice Shape Matching.
+(Right) Deformed lattices consisting of $150$ particles per penguin ($22\,500$ particles).
+The penguins can be deformed at $25$ frames per second on a Pentium4 $3.4 \,$GHz.}
+\label{chap4:fig-penguins}
+\end{figure}
+
+One limitation of fast lattice shape matching is that small features of a geometry yield an explosion of the runtime cost.
+Indeed, a small surface feature may require fine sampling in order to be deformed independently from non-adjacent material, but this fine sampling must be applied to the whole object.
+Therefore, \cite{Steinemann08} further improved the technique by allowing dynamic adaptive selection of levels of detail through the use of an octree-based hierarchical sampling.
+This adaptative sampling is illustrate with Fig.~\ref{chap4:fig-flowers}.
+Their technique also handles efficient topological changes.
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=13.9cm]{chapter4/flowers.pdf}
+\end{center}
+\caption[Flowers in the wind]{Complex scene with $40$ deforming flowers.
+A coarse sampling is employed when the flowers are moved by the wind, for a total of $5\,680$ nodes in the scene.
+However, the models are dynamically refined when touched by the user, as shown in the left image.}
+\label{chap4:fig-flowers}
+\end{figure}
 
@@
 
-	\subsection{Mass-spring}
+	\subsection{Mass-spring model}
+A classic technique for modelling a deformable object is the mass-spring system.
+In this method, the geometry is described by a network of masses connected together by springs.
+The force $ F $ exerted onto each mass by the spring may be computed with:
+\begin{equation}
+\label{chap4:massSpring1}
+F_{\text{spring}} = - k x,
+\end{equation}
+where $ k $ is the spring stiffness constant, $ x $ represents the stretch of the spring from its rest position.
+From our experience, we know that a mass attached to a single spring has a tendency to oscillate.
+If no appropriate measure is taken, a network of springs will a forciori oscillate as well.
+To prevent the system from oscillating, a damper is also added between each couple of masses.
+The force created by the damper acts to reduce the velocity and may be expressed as:
+\begin{equation}
+\label{chap4:massSpring2}
+F_{\text{damper}} = - d v
+\end{equation}
+where $ d $ is the damping constant and $ v $ the derivative of $ x $ with respect to time.
+The damper added between each couple of masses allows to model friction and reduce the amplitude of the oscillations.
+Applying Newton's second law $ \sum \mathbf{F} = m \mathbf{a} $ yields the standard differential equation for a mass-spring system:
+\begin{equation}
+\label{chap4:massSpring3}
+m \ddot x = - d \dot x - k x.
+\end{equation}
+For complex systems with more than one spring attached to each mass, the right-hand side of the equation above becomes more complex as it needs to be repeated and adjusted for each spring.
+The easiest way to solve a complex mass-spring system iteratively, is to first calculate the total force currently acting on each mass by summing all forces currently exerted by all springs attached to this mass (using \eqref{chap4:massSpring1} and \eqref{chap4:massSpring2}).
+Dividing the total force by the mass gives the current acceleration of each mass (by \eqref{chap4:massSpring3}).
+This equation must be numerically solved for each mass, either by explicit or implicit methods.
+A discussion of integration methods for mass-spring models can be found in \cite{Shinya05}.
+
+Mass-spring systems have been extensively used and are still very popular.
+And this is not surprising because they are easy to implement and computationally efficient.
+Of course, they have drawbacks.
+But as we will see shortly, various recent works attempt to circumvent them.
+One limitation often mentioned is the difficulty to derive spring stiffnesses from elastic properties (Young's modulus and Poisson's ratio).
+To overcome this deficiency, \cite{Lloyd07} mention two ways of obtaining the parameters of a mass-spring model.
+The first one consists in varying the parameters until the behaviour of the system is similar to the one obtained through the experiments or the one obtained with the Finite Element Method \citep{Bianchi04}.
+The second way consists in establishing an analytical reasoning for calculating the constants of the mass-spring model.
+For instance, starting from the definition of Young's modulus, Poisson's ration, shear and bulk modulus, \cite{Baudet07} apply simple tests to their mass-spring model to find the most appropriate stiffnesses.
+\cite{Lloyd07} proposes a linearisation of the mass-spring equations with the aim of equating the linearised stiffness matrix to the stiffness matrix of a linear finite element method.
+By following the same method, \cite{SanVicente09} derived a mass-spring model equivalent to a linear finite element model for maxillofacial surgery simulation.
+
+Because they do not derive from the equations of continuum mechanics, mass-spring models have a limited capacity to model the various aspects of a material like anisotropy, viscoelasticity etc.
+Even worse, according to \cite{Bourguignon00}, if all springs are set to the same stiffness, the mesh geometry may generate undesired anisotropy as shown in Fig.~\ref{chap4:fig-undesiredAnisotropy}.
+However, some works tried to take advantage of this feature by giving directions of interest to their model by specifically designing the mesh to align springs on specific directions.
+For instance, it was used by \cite{Miller88} to animate the motion of snakes and worms.
+It was also used by \cite{Ng-Thow-Hing97} in their muscle model where some of the springs were aligned with the muscle fibers and more recently by \cite{Magjarevic09} for modelling the mechanical behavior of the myocardial tissue.
+More details on controlling anisotropy in mass-spring systems may be found in \cite{Bourguignon00}.
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=8cm]{chapter4/undesiredAnisotropy.pdf}
+\end{center}
+\caption[Undesired anisotropy with mass-spring models]{The two meshes are undergoing a downward pull at their bottom.
+While the tetrahedral mass-spring system (left) shows signs of anisotropy, the hexahedral mesh (right) with springs aligned along the gravity does not.}
+\label{chap4:fig-undesiredAnisotropy}
+\end{figure}
+
+A mass-spring model may also be enhanced to feature viscoelasticity.
+Indeed, \cite{Stiles72} proposed a viscoelastic mass-spring system to test hypotheses on the mechanical response of muscles.
+Much more recently, \cite{Tamura05} introduced a method to simulate viscoelastic material using a spring-mass system.
+They use a large number of particules to create a randomly connected mesh to mimic the structure of polymeric materials and hence their viscoelascity characteristics.
+\cite{Basafa10} went even further and extended a mass-spring system to simulate non-linear viscoelastic deformations of soft tissue for laparoscopic surgery.
+They tune their parameters by a simple optimisation procedure to fit the mechanical response obtained on a set of experimental data.
+
+	\subsection{Modal analysis}
 
 \section{Techniques based on continuum mechanics}
+
+	\subsection{Point-based}
 

### 2010-08-12 Thesis/Chapter4.tex
- Update 12 August -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@150 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
+Although realism is not a crucial feature for game-like environments, other models based on laws of physics were created for other applications.
+
 \section{Techniques relying on physics}
+
+	\subsection{3D Chainmail algorithm}
+The \emph{3D Chainmail} was introduced by \cite{Gibson97} as a fast algorithm for deforming solid objects.
+Rather than carrying out complex computations on a low resolution geometry (as opposed to FEM techniques), the main idea is to take advantage of the original data resolution (of a CT scan for instance) but perform simple deformation calculation for each element.
+In this algorithm, each volume element is linked to its six nearest neighbours.
+If a link between two nodes is stretched to its limit, displacements are transferred to its neighbouring links.
+It works in a very similar way to a chain and is illustrated by Fig.~\ref{chap4:fig-chainmail}.
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=12cm]{chapter4/chainmail.pdf}
+\end{center}
+\caption[Deformation of a 2D chain]{Deformation of a 2D chain mail when the link indicated by an arrow is moved.}
+\label{chap4:fig-chainmail}
+\end{figure}	
+%
+The chain mail algorithm is then followed by an elastic relaxation, which relaxes the shape of the deformed object approximated by the chain mail algorithm.
+This step consists of minimising an elastic energy measure within the object to insure that the calculated deformation is a minimum energy configuration.
+
+This method was extented by \cite{Schill98} to integrate inhomogeneous and anisotropic behaviour.
+This extension of the chain mail algorithm allows the authors to simulate the vitrous humor in the eye during a vitrectomy procedure.
+However, the main drawback of this algorithm is that it does not model volume preservation, an important characteristic of many tissues in the human body.
+
+	\subsection{Modal analysis}
+\emph{Modal analysis} was first introduced to the graphics community by \cite{Pentland89} as a fast method for approximating deformation.
+Modal analysis is the process of taking the non-linear description of a system, finding a good linear approximation, and then finding a coordinate system that diagonalises the linear approximation.
+This process transforms a complicated system of non-linear equations into a simple set of decoupled linear equations that may be individually solved analytically \citep{Hauser03}.
+Moreover, because each decoupled equation may be solved analytically, stability concerns due to the choice of a time integration procedure are eliminated.
+The linearised system of equations may be written as:
+\begin{equation}
+\label{chap4:linearisedEquation}
+\mathbf{M} \mathbf{\ddot U} + \mathbf{D} \mathbf{ \dot U} + \mathbf{K} \mathbf{U} = \mathbf{R}.
+\end{equation}	
+where $ \mathbf{M} $ is the mass matrix, $\mathbf{D}$ is the damping matrix, $ \mathbf{K} $ is the stiffness matrix and $\mathbf{R}$ are externally applied loads.
+Modal decomposition refers to the process of diagonalising equation \eqref{chap4:linearisedEquation}.
+By using Rayleigh damping (see discussion on damping in section \ref{chap3:wordOnMatrices}), \eqref{chap4:linearisedEquation} may be rewritten as:
+\begin{equation}
+\label{chap4:eqRayleigh}
+\mathbf{K} (\mathbf{U} + \alpha_1  \mathbf{ \dot U}) + \mathbf{M} (\alpha_2 \mathbf{ \dot U} + \mathbf{\ddot U})  = \mathbf{R},
+\end{equation}	
+where $ \alpha_1 $ and $ \alpha_2 $ are the Rayleigh coefficients.
+We denote $ \mathbf{W} $ the solution to the generalised eigenproblem $ \mathbf{K} \mathbf{x} + \lambda \mathbf{M} \mathbf{x} = 0 $ and $\boldsymbol \Lambda$ the diagonal matrix of eigenvalues, then \eqref{chap4:eqRayleigh} may be expressed as:
+\begin{equation}
+\boldsymbol \Lambda (\mathbf{Z} + \alpha_1  \mathbf{ \dot Z}) + (\alpha_2 \mathbf{ \dot Z} + \mathbf{\ddot Z})  = \mathbf{G},
+\end{equation}
+where $ \mathbf{Z} = \mathbf{W}^{-1} \mathbf{U} $ is the vector of modal coordinates and $ \mathbf{G} = \mathbf{W}^T \mathbf{U} $ is the external force vector in the modal coordinate system.
+Each row of this equation now corresponds to a single scalar differential equation:
+\begin{equation}
+\ddot z_i + (\alpha_1 \lambda_i +\alpha_2) \dot z_i + \lambda_i z_i = g_i.
+\end{equation}
+The analytical solutions to each equations are well known and are the following:
+\begin{equation}
+z_i = c_1 e^{t \omega_i^+} + c_2 e^{t \omega_i^-}
+\end{equation}
+where $ c_1 $ and $ c_2 $ are complex constants and $ \omega_i $ is the complex frequency given by:
+\begin{equation}
+\omega_i^{\pm} = \dfrac{-(\alpha_1 \lambda_i + \alpha_2) \pm \sqrt{(\alpha_1 \lambda_i +\alpha_2)^2 - 4 \lambda_i}}{2}.
+\end{equation}
+The main advantage of modal analysis is the possibility to model only the necessary modes.
+Indeed if the eigenvalue $ \lambda_i $ associated with a particular mode is large, then the force required to cause a discernible displacement of that mode will also be large.
+Similarly, some displacements may be too small to be detected.
+Removing those modes from the computation will not change the appearance of the resulting simulation while substantially reducing the computational cost.
+
+\cite{James02} went even further by using graphics hardware to reduce CPU costs to a minimum.
+In particular, they use modal analysis to model human skin and secondary tissues in a laparascopic surgical simulation to allow the CPU to focus on more complex tissue models and user contact interactions (see Fig.~\ref{chap4:fig-modalAnalysisLaparo}).
+The technique is fairly efficient as it allows precomputations of the vibration modes.
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=13cm]{chapter4/modalAnalysisLaparo.pdf}
+\end{center}
+\caption[Applications of modal analysis]{(Left) A jumping motion that leads to significant thigh and belly vibrations of human skin when rendered with modal analysis.
+(Right) The same hardware accelerated technique is applied to secondary tissues in a laparoscopic simulation to ease the CPU of computations.
+}
+\label{chap4:fig-modalAnalysisLaparo}
+\end{figure}
+
+Although modal analysis significantly accelerates the simulation, it generates noticeable artifacts when applied to large deformations due to the step of linearisation.
+It can produce unnatural results with deformations of large amplitude.
+\cite{Choi05} proposed to overcome this limitation by taking the rotational component of the deformation into account at each node.
+\cite{Barbic05} make use of modal analysis to build a deformation basis from modal derivatives.
+It allows them to simulate large deformations of solid objects with non-linear material.
 
@@
 
-	\subsection{Modal analysis}
+The most complete mathematical formalism available to describe the mechanical behaviour of a solid is provided by continuum mechanics.
+Consequently, it feels natural to derive computational models from the equations of continuum mechanics and a few techniques based on these equations will now be introduced.
 
@@
 
-	\subsection{Point-based}
-
 	\subsection{FEM with mesh (highlight evolution: linear, topological changes, co-rotational, non-linear, GPU)}
 
-	\subsection{Meshless (supposedly good for handling topological changes but other issues}
+	\subsection{Meshless, position-based dynamics (supposedly good for handling topological changes but other issues)}

### 2010-08-14 Thesis/Chapter4.tex
- Update 14 August -
Essentially added references



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@151 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 	\subsection{FEM with mesh (highlight evolution: linear, topological changes, co-rotational, non-linear, GPU)}
+The finite element method brings accuracy a step further.
+The method will not be explained here since it was already discussed with great details in chapter~\ref{chap3}.
+This is the method of first choice for modelling deformable objects with precision as it is directly derived from the most complete theory available for describing the mechanical response of a continuum.
+However, because of its complexity, this method is substantially more computationally expensive than the ones introduced so far.
+And this has been a problem in computer graphics and interactive simulation for many years.
+So researchers came up with simplifying assumptions and ideas to make the computations more efficient.
+Over time, with the increase in computational power, the simplifications were progressively relaxed and the observed tendency for real-time simulation is now towards more and more complete finite element methods.
+In this section, we will describe this evolution in the use of finite element methods for interactive simulation.
+
+ The use of the finite element method in computer graphics was pioneered by \cite{Terzopoulost87} for simulations of elastic deformations.
 

### 2010-08-16 Thesis/Chapter4.tex
- Update 16 August -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@152 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-The most complete mathematical formalism available to describe the mechanical behaviour of a solid is provided by continuum mechanics.
+\cite{Terzopoulos91} even designed a mass-spring system that experience transitions from solid to liquid.
+Nodes are connected by springs in a hexahedral lattice and each of them has an associated temperature.
+Spring stiffnesses decreased with the increase in temperature.
+The diffusion of heat through the material is computed by using a discretised form of the heat equation.
+When the melting point is reached, the stiffness is set to zero and the node is detached from this particular spring.
+Once a node is freed from all his neighbouring springs, it becomes an independent glop of fluid and its behaviour is then modelled using a discrete fluid model.
+
+If mass-spring systems are quite versatile and easy to implement, the most complete mathematical formalism available to describe the mechanical behaviour of a solid is provided by continuum mechanics.
 Consequently, it feels natural to derive computational models from the equations of continuum mechanics and a few techniques based on these equations will now be introduced.

### 2010-08-17 Thesis/Chapter4.tex
- Update 17 August -
Major update on FEM state of the art + bib updated



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@153 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-	\subsection{FEM with mesh (highlight evolution: linear, topological changes, co-rotational, non-linear, GPU)}
+	\subsection{The Finite Element Method}
 The finite element method brings accuracy a step further.
@@
 This is the method of first choice for modelling deformable objects with precision as it is directly derived from the most complete theory available for describing the mechanical response of a continuum.
-However, because of its complexity, this method is substantially more computationally expensive than the ones introduced so far.
+Different kind of analysis may be carried out with the finite element method.
+As seen in section~\ref{chap3:derivationEquations}, an analysis may be either static (that is, the transient response is ignored) or dynamic (more general).
+In all analyses, the deformation of the body at a point is described by some measure of strain.
+If this measure is chosen to be linear (see section~\ref{chap2:infinitesimalStrainTensor} for details), the analysis is said to be \emph{geometrically} linear.
+In general, we consider this assumption to be valid for strain less than $ 10 \,$\%.
+In contrast, geometrically non-linear analyses make use of a non-linear strain tensor to measure the deformation and thus may handle large deformations.
+Another characteristic of an analysis is whether the relationship between strain and stress (the constitutive model) is assumed to be linear or non-linear.
+The analysis is qualified as \emph{materially} linear or non-linear, respectively.
+
+Because of its complexity, the finite element method is substantially more computationally expensive than the ones introduced so far.
 And this has been a problem in computer graphics and interactive simulation for many years.
@@
 
- The use of the finite element method in computer graphics was pioneered by \cite{Terzopoulost87} for simulations of elastic deformations.
+\bigskip
+
+The use of the finite element method in computer graphics was pioneered by \cite{Terzopoulos87} for simulations of elastic deformations.
+They derived a geometrically non-linear strain tensor from differential geometry theory.
+Obviously, the computations were not carried out in real-time but this technique allows \citeauthor{Terzopoulos87} to calculate the interaction between solid and deformable models in a physically realistic manner (see Fig.~\ref{chap4:fig-ball}).
+ %
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=7cm]{chapter4/ball.pdf}
+\end{center}
+\caption[Ball on a deformable model]{A ball resting on a supporting elastic solid as computed by \citeauthor{Terzopoulos87}.} 
+\label{chap4:fig-ball}
+\end{figure}	
+In order to speed up the computations, linear strain measures started to be used (see section~\ref{chap2:infinitesimalStrainTensor}).
+For instance, \cite{Gourret89} make use of a linear finite element method to model the deformation of grasped objects.
+The equations are however solved in static for simplicity.
+An approach to solve the system of equations in dynamic in a reasonable amount of time is to decrease the number of elements.
+This reduction must be compensated by the use of high-order elements, that is the degree of the shape functions used in the interpolation of the solution is higher to maintain the overall accuracy.
+This approach was applied to an elaborate model of a muscle by \cite{Chen92} and to modelling skin in a simulator of plastic surgery by \cite{Pieper92} where they both use 20-node brick elements.
+Yet, finite element modelling remained very computationally expensive and the interactivity required for medical simulators was still to be reached.
+
+In 1996, \citeauthor{Bro-Nielsen96} developed a real-time finite element method formulation by introducing three improvements.
+The first one is to compress the stiffness matrix of the volumetric system into a matrix with the same complexity as a surface model of the same object.
+This technique is called \emph{condensation}.
+The deformation is solved only at surface points while still taking the volumetric nature of the object into account.
+The second improvement concerns the fact that they explicitly precompute the inverse of the stiffness matrix and use matrix vector multiplication with this matrix to achieve a low calculation time.
+Finally, they exploit the sparse structure of the force vector.
+They showed the efficiency of their method by modelling a leg with 700 nodes at 20 frames per second.
+
+Later, \cite{James99} had another idea to improve the speed of the computation.
+They based their model on boundary integrals and the \emph{boundary element method}, used for the first time in computer graphics.
+The method consists of calculating only boundary displacements and forces.
+Unfortunately, the authors do not provide the performance of their framework.
+
+\cite{Cotin99} took advantage of the superposition principle which applies to linear problems.
+Their idea was to approximate the deformation of a solid object with a linear combination of pre-computed states.
+In addition, they enhanced the formulation by adding a non-linear radial component.
+This method was applied to compute the deformation of liver in a hepatic surgery simulator.
+Their approach substantially improved the computational time for a finite element model as it allowed them to compute the positions and the forces of a $1\,500$ node liver at $300\,$Hz.
+An important drawback emphasised by the authors themselves is the impossibility to simulate tissue cutting.
+Indeed, the large precomputations associated with this method are dependent of the geometry of the mesh and therefore cannot be updated in real-time.
+
+\bigskip
+
+As we know, the assumption of linearity is usually only valid for very small deformations and strains.
+If this is often an acceptable trade-off between the visual deformation result and the speed of the system, research was carried out to handle large deformations.
+However, since geometrically non-linear analyses are highly computationally demanding, another approach was found which allows large deformation while still relying on a linear strain measure: the \emph{co-rotational} methods.
+A very complete description of the co-rotational approach is given by \cite{Felippa00}.
+The problem with geometrically linear analyses is the error brought by the approximated (linear) strain measure, which increases with the rotation of the element.
+The idea behind co-rotational methods is the decomposition of the motion into a rigid-body and deformational components.
+In other words, the deformation of each element may be seen as measured in a coordinate system which follows the rigid-body motion of the element, hence cancelling the part of the error due to the rotation in the strain measure.
+Consequently, co-rotational formulations allow large displacements and rotations (since the rigid-body motion is not considered) as long as the strain remains small (the strain measure is still linear).
+
+Among the firsts to apply this technique to static analyses are \cite{Argyris64} and \cite{Wempner69}.
+Although the birth of co-rotational methods is not very clear \citep{Felippa05}, \cite{Belytschko73} had a similar thinking and introduced the so-called \emph{convected coordinates} for solving dynamic equations.
+However, while the spirit is the same than co-rotational methods, convected coordinates forms a curvilinear system that fits the change of metric as the body deforms, therefore the convected metric necessarily emcompasses deformations.
+In other words, the decomposition in rigid-body and deformational components is not exact.
+During the last decade, the co-rotational approach has regained interest in the computer graphics community.
+
+\cite{Hauth04} made use of the polar decomposition of the deformation gradient to extract the rotation.
+They also applied this formulation in a hierarchical finite element setting where the coarse level is computed with a fully non-linear strain tensor and the finer levels are computed with a co-rotational method.
+In addition, they examined the error that is arising from using the co-rotational approximation instead of a non-linear strain tensor.
+They measured that the co-rotational approximation induces an error of about $5-20\,$\% but can give a speedup of about a factor seven or more.
+
+\cite{Pouliquen05} addressed the simulation of human fingers with haptic feedback.
+During the grasping task, the deformable finger pads undergo large rotations while their deformation remains small.
+Using a co-rotational approach for modelling the finger pads allows them to efficiently compute the contact forces between the deformable pads and the rigid object to be grasped.
+
+\bigskip
+
+Another major concern in medical simulators is the ability to cutting through soft tissue models.
+\cite{Mor00} proposed a technique to generate a minimal set of new elements to replace intersected tetrahedra along the path formed by the cut.
+The method do not wait for the cut to be completed for splitting an element, instead, a minimal subdivision of a partially cute tetrahedron is generated.
+The authors used a very small time step to insure the stability of the soft tissue model and their simulations were not real-time.
+In addition, their method can create very small elements and the model can become unstable.
+This occurs because the stiffness matrices for the small elements are quite large, and even small displacements can generate very large forces.
+
+\cite{Cotin00} proposed a hybrid model allowing real-time cutting.
+They combined a quasi-static precomputed model, which is very efficient but does not allow topology changes, with a tensor-mass model which requires more computation but authorises cutting.
+The key idea is to separate a same structure into two parts: one susceptible to be cut during the surgery procedure and one that will not be submitted to any topological change.
+While the former will be simulated with the tensor-mass model, the deformation of the latter will be computed using the precomputed model.
+Since both linearly elastic models follow the same physical law, the combination of these two models should behave exactly as a global, linearly elastic model.
+To achieve this goal, \citeauthor{Cotin00} describe how to impose the additional boundary conditions at the connection nodes for the global model to be consistent in terms of forces and displacements.
+The efficiency of this hybrid model was demonstrated by simulating an hepatectomy (removing of one of the eight anatomical segments of the liver).
+They reported the simulation (allowing deformation and cutting) of a mesh with more than $ 8\,000 $ tetrahedra in real-time (see Fig.~\ref{chap4:fig-hybridModel}).
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[height=4.5cm]{chapter4/hybrid1.pdf}
+\hspace{0.2cm}
+\includegraphics[height=4.5cm]{chapter4/hybrid2.pdf}
+\end{center}
+\caption[Simulation of a hepatectomy using a hybrid model]{Simulation of a hepatectomy using a hybrid model.
+$ 18 $\% of the mesh was modelled with the tensor-mass model (green) and therefore could be cut out.
+The rest of the mesh was simulated with the precomputed finite element model.
+}
+\label{chap4:fig-hybridModel}
+\end{figure}
+
+\cite{Serby01} described a new approach to cutting in finite element models.
+The central idea is not to introduce new nodes/elements but to displace the existing ones to account for the topological changes introduced by a cut.
+They also added a step of homogenisation of the mesh to avoid tiny elements.
+Thus, the problem of decreasing element size is minimised and consequently the stability of the solution of the equations of motion is increased.
+
+Cutting is a problem for all techniques precomputing the inverse of the stiffness matrix.
+Indeed, topological changes require updates of this matrix, which is too computationally heavy to be done in real-time.
+\cite{Nienhuys01} avoid the precomputations by approximating the inverse at runtime with an iterative algorithm (Conjugate Gradient).
+Moreover, instead of subdividing each tetrahedron to make the cut appear where the user performed it, they adapt the mesh locally so that there are always triangles on the scalpel path, and perform the cut along those triangles.
+If the approach is interesting, the authors reported a lag between the scalpel and the realised cut.
+In addition, repositioning the nodes within the mesh can easily generate degeneracies, which have to be eliminated.
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=13cm]{chapter4/cutUnderScalpel.pdf}
+\end{center}
+\caption[The three steps in performing a cut]{The three steps in performing a cut (represented here in 2D).} 
+\label{chap4:fig-cutUnderScalpel}
+\end{figure}
+
+\bigskip
+
+viscoelastic + non-linear
+
+\bigskip
+
+GPU stuff
 

### 2010-08-18 Thesis/Chapter2.tex
- Update 18 August -
State of the art on FEM finished + bib updated



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@154 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\section{Strain measures}
+\section{Strain measures}	\label{chap2:strainMeasures}
 The length of a material curve from the reference configuration can change when displaced to a curve in the deformed configuration.

### 2010-08-18 Thesis/Chapter4.tex
- Update 18 August -
State of the art on FEM finished + bib updated



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@154 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \end{center}
-\caption[Penguins animation with Fast Lattice Shape Matching]{(Left) $150$ cartoon penguins deforming dynamically using Fast Lattice Shape Matching.
-(Right) Deformed lattices consisting of $150$ particles per penguin ($22\,500$ particles).
+\caption[Penguins animation with Fast Lattice Shape Matching]{(\textbf{Left}) $150$ cartoon penguins deforming dynamically using Fast Lattice Shape Matching.
+(\textbf{Right}) Deformed lattices consisting of $150$ particles per penguin ($22\,500$ particles).
 The penguins can be deformed at $25$ frames per second on a Pentium4 $3.4 \,$GHz.}
@@
 \end{center}
-\caption[Applications of modal analysis]{(Left) A jumping motion that leads to significant thigh and belly vibrations of human skin when rendered with modal analysis.
-(Right) The same hardware accelerated technique is applied to secondary tissues in a laparoscopic simulation to ease the CPU of computations.
+\caption[Applications of modal analysis]{(\textbf{Left}) A jumping motion that leads to significant thigh and belly vibrations of human skin when rendered with modal analysis.
+(\textbf{Right}) The same hardware accelerated technique is applied to secondary tissues in a laparoscopic simulation to ease the CPU of computations.
 }
@@
 
-As we know, the assumption of linearity is usually only valid for very small deformations and strains.
-If this is often an acceptable trade-off between the visual deformation result and the speed of the system, research was carried out to handle large deformations.
-However, since geometrically non-linear analyses are highly computationally demanding, another approach was found which allows large deformation while still relying on a linear strain measure: the \emph{co-rotational} methods.
-A very complete description of the co-rotational approach is given by \cite{Felippa00}.
-The problem with geometrically linear analyses is the error brought by the approximated (linear) strain measure, which increases with the rotation of the element.
-The idea behind co-rotational methods is the decomposition of the motion into a rigid-body and deformational components.
-In other words, the deformation of each element may be seen as measured in a coordinate system which follows the rigid-body motion of the element, hence cancelling the part of the error due to the rotation in the strain measure.
-Consequently, co-rotational formulations allow large displacements and rotations (since the rigid-body motion is not considered) as long as the strain remains small (the strain measure is still linear).
-
-Among the firsts to apply this technique to static analyses are \cite{Argyris64} and \cite{Wempner69}.
-Although the birth of co-rotational methods is not very clear \citep{Felippa05}, \cite{Belytschko73} had a similar thinking and introduced the so-called \emph{convected coordinates} for solving dynamic equations.
-However, while the spirit is the same than co-rotational methods, convected coordinates forms a curvilinear system that fits the change of metric as the body deforms, therefore the convected metric necessarily emcompasses deformations.
-In other words, the decomposition in rigid-body and deformational components is not exact.
-During the last decade, the co-rotational approach has regained interest in the computer graphics community.
-
-\cite{Hauth04} made use of the polar decomposition of the deformation gradient to extract the rotation.
-They also applied this formulation in a hierarchical finite element setting where the coarse level is computed with a fully non-linear strain tensor and the finer levels are computed with a co-rotational method.
-In addition, they examined the error that is arising from using the co-rotational approximation instead of a non-linear strain tensor.
-They measured that the co-rotational approximation induces an error of about $5-20\,$\% but can give a speedup of about a factor seven or more.
-
-\cite{Pouliquen05} addressed the simulation of human fingers with haptic feedback.
-During the grasping task, the deformable finger pads undergo large rotations while their deformation remains small.
-Using a co-rotational approach for modelling the finger pads allows them to efficiently compute the contact forces between the deformable pads and the rigid object to be grasped.
-
-\bigskip
-
-Another major concern in medical simulators is the ability to cutting through soft tissue models.
+Besides real-time computation of soft tissue, another major concern for medical simulators is the ability to cutting through soft tissue models.
 \cite{Mor00} proposed a technique to generate a minimal set of new elements to replace intersected tetrahedra along the path formed by the cut.
@@
 
-viscoelastic + non-linear
+As we know, the assumption of linearity is usually only valid for very small deformations and strains.
+If this is often an acceptable trade-off between the visual deformation result and the speed of the system, research was carried out to handle large deformations.
+However, since geometrically non-linear analyses are highly computationally demanding, another approach was found which allows large deformation while still relying on a linear strain measure: the \emph{co-rotational} methods.
+A very complete description of the co-rotational approach is given by \cite{Felippa00}.
+As explained in section~\ref{chap2:strainMeasures}, an appropriate strain measure is a tensor which only measures the deformation applied to the object.
+This is the case of the Cauchy-Green or Green tensors for instance.
+Those strain metrics only measures deformation, they are invariant with respect to rigid-body transformations applied.
+However, geometrically linear analyses make use of the infinitesimal strain tensor, which is only an approximation of the Green tensor.
+Consequently, this linearised strain measure is not invariant with respect to rigid-body transformations.
+In particular, the error brought by the approximated (linear) strain measure increases with the rotation of the element.
+This error in the measure creates ghost forces which overdeforms the simulated object.
+The idea behind co-rotational methods is the decomposition of the motion into a rigid-body and deformational components.
+In other words, the deformation of each element may be seen as measured in a coordinate system which follows the rigid-body motion of the element, hence cancelling the part of the error due to the rotation in the strain measure.
+Consequently, co-rotational formulations allow large displacements and rotations (since the rigid-body motion is not considered) as long as the strain remains small (the strain measure is still linear).
+
+Among the firsts to apply this technique to static analyses are \cite{Argyris64} and \cite{Wempner69}.
+Although the birth of co-rotational methods is not very clear \citep{Felippa05}, \cite{Belytschko73} had a similar thinking and introduced the so-called \emph{convected coordinates} for solving dynamic equations.
+However, while the spirit is the same than co-rotational methods, convected coordinates forms a curvilinear system that fits the change of metric as the body deforms, therefore the convected metric necessarily emcompasses deformations.
+In other words, the decomposition in rigid-body and deformational components is not exact.
+During the last decade, the co-rotational approach has regained interest in the computer graphics community.
+
+\cite{Hauth04} made use of the polar decomposition of the deformation gradient to extract the rotation.
+They also applied this formulation in a hierarchical finite element setting where the coarse level is computed with a fully non-linear strain tensor and the finer levels are computed with a co-rotational method.
+In addition, they examined the error that is arising from using the co-rotational approximation instead of a non-linear strain tensor.
+They measured that the co-rotational approximation induces an error of about $5-20\,$\% but can give a speedup of about a factor seven or more.
+
+\cite{Pouliquen05} addressed the simulation of human fingers with haptic feedback.
+During the grasping task, the deformable finger pads undergo large rotations while their deformation remains small.
+Using a co-rotational approach for modelling the finger pads allows them to efficiently compute the contact forces between the deformable pads and the rigid object to be grasped.
 
@@
 
-GPU stuff
+If co-rotational approaches increase accuracy in the strain measure, all the examples we have seen so far make use of a linear constitutive law.
+Yet, the mechanical behaviour of soft tissue is known to be non-linear, viscoelastic and often anisotropic \citep{Fung93}.
+Moreover, \cite{Misra07} demonstrated a noticeable difference between the force feedbacks provided by linear and non-linear tissue models.
+One of the first materially non-linear finite element formulation used in medical simulation was probably the work of \cite{Sagar94}.
+While developing a teleoperated micro-surgical robot for eye surgery, they wanted to provide a virtual environment has part of the system for training.
+Since the cornea is anisotropic and non-linearly elastic and undergoes large deformations, they have used large deformation elasticity theory with orthotopic and non-linear material properties in the finite element model.
+Coupled with a detailed visual model of the eye, the simulation produces a fair level of realism with a performance greater than $ 10 \,$Hz (see Fig.~\ref{chap4:fig-eyeSurgery}).
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[height=5.5cm]{chapter4/eyeSurgery1.pdf}
+\hspace{0.2cm}
+\includegraphics[height=5.5cm]{chapter4/eyeSurgery2.pdf}
+\end{center}
+\caption[Simulation of eye surgery]{(\textbf{Left}) Exterior view of the model showing the eye and the eyelashes.
+(\textbf{Right}) Surgical virtual environment including micro-tools and guidance information.}
+\label{chap4:fig-eyeSurgery}
+\end{figure}
+
+\cite{Keeve98} compared a fully (geometrically and materially) non-linear finite element model with a mass-spring model to compute the deformation of skin after bone realignment in a craniofacial surgery simulator.
+The efficiency of the mass-spring model gives the ability to the surgeon to realise interactive simulation of the resulting tissue changes and to improve his planning process.
+The best precision was given by the finite element model which could be used off-line to verify the chosen surgical procedure.
+
+The first geometrically non-linear finite element model running in real-time was reported by \cite{Debunne01}.
+They provided an adaptative method for animating viscoelastic deformable objects in a guaranteed frame rate.
+As the object moves and deforms, the sampling is refined to concentrate the computational load into the regions that deform the most (see Fig.~\ref{chap4:fig-gummyBear}).
+They reported animating a few hundred points in real-time at a frequency higher than $ 300\,$Hz.
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=13cm]{chapter4/gummyBear.pdf}
+\end{center}
+\caption[Simulation of a gummy bear]{The approach of \cite{Debunne01} makes use of a local refinement technique to ensure high physical fidelity while bounding the global computation load to guarantee real-time simulation with a geometrically non-linear and visco-elastic finite element model.}
+\label{chap4:fig-gummyBear}
+\end{figure}
+
+In the same spirit, \cite{Wu01} applied an adaptative meshing technique to provide sufficient detail where required while minimising unnecessary computation.
+They included both Mooney-Rivlin and Neo-Hookean material models and used the non-linear Cauchy-Green tensor.
+In addition to the local mesh refinement technique, they used mass lumping and an explicit time integration to deform a liver model of $ 1\,210 $ tetrahedra in real-time.
+
+Still in 2001, \citeauthor{Picinbono01} proposed to use a non-linear model only where the displacements are larger than a given threshold, the remaining part of the object still uses linear elasticity.
+
+Recently, research on non-linear finite element increased significantly.
+For instance, \cite{Pathmanathan04} used a fully non-linear model to predict deformation of breast under compression during mammography.
+However, they were not interested in real-time.
+\cite{Zhong05} claimed achieving real-time with their non-linear finite element method using an interpolation approach but do not give any figure.
+\cite{Sedef06} proposed a viscoelastic model to simulate the deformation of a liver.
+They calculate the deformation in real-time by carrying out the computations in static and using the superposition principle.
+However, their approach used a linear strain measure, which limits their system to small deformations.
+\cite{Yan07} claimed to compute their non-linear finite element model in real-time using a technique of \emph{graded mesh} but do not give any information on their performance.
+
+Finally, \cite{Miller07} proposed the total Lagrangian explicit dynamics (TLED) algorithm, a very efficient fully non-linear formulation.
+The algorithm is based on the finite element method using the total Lagrangian formulation, where stresses and strains are measured with respect to the original configuration.
+This choice allows for precomputing of most spatial derivatives before the commencement of the time-stepping procedure.
+The authors reported that the average number of floating-point operations per element per time step is $35\,$\% lower than for the similar implementation of the algorithm based on updated Lagrangian formulation.
+They were able to compute the deformation of a hexahedral mesh of $ 6\,000 $ elements in about $ 16\,$ms for a time step of $ 1\,$ms.
+If they could not achieve real-time computations, their work constituted a step towards the simulation of entire organs in real-time.
+
+\bigskip
+
+The finite element method is very computationally demanding.
+However, by nature, very similar computations are carried out for each element.
+Consequently, the finite element method is a good candidate for being implemented on parallel architectures.
+Thus, \cite{Szekely00} implemented a fully non-linear formulation for an 8-processor machine.
+However, the dedicated hardware was not yet assembled and functional at the time so they did not provide any measure of performance.
+
+Early 2000s, the rapid increase in the performance of graphics hardware, coupled with recent improvements in its programmability, have made graphics hardware a compelling platform for computationally demanding tasks in a wide variety of application domains.
+Researchers and developers have become interested in harnessing this power for general-purpose computing, an effort known collectively as GPGPU (for General-Purpose computing on the GPU).
+A survey of GPGPU algorithms and techniques was written by \cite{Owens07}.
+
+To the best of our knowledge, \cite{Rumpf01} were the firsts to leverage the power of Graphics Processing Units (GPU) to solve partial differential equations.
+They established a correspondence between common mathematical operations and OpenGL operations and structures.
+Thus, as an example, a vector is represented as a RGBA color vector in a color image on the graphics card.
+They decided to implement an iterative solver for a linear system of equations, which we find at the core of most finite element codes.
+Using this hardware accelerated solver, they solved the linear heat equations and presented an anisotropic diffusion model for image processing.
+They obtained promising results but noted the hardware restrictions that forced them to approximate all non-linear functions by linear ones in the implementation of the anisotropic diffusion.
+Although a complete hardware accelerated finite element method implementation was yet to be done, \citeauthor{Rumpf01} definitely stepped towards the right direction.
+
+\cite{Wu04} proposed the first application of GPU in medical simulation.
+They were able to achieve an interactive frame rate for the topolgy change in surgical simulation.
+Not only they used the condensation technique for merely calculating the surface nodes in the non-operation parts, but they implemented the conjugate gradient solver to GPU.
+Indeed, entering 2003, the full IEEE floating point precision data type supported in shaders and texture stages allows the GPU to be applied into more general applications.
+The conjugate gradient solver includes three primary operations: multiplication of the sparse matrix and the vector, addition of two vectors and the sum-reduction operation.
+The core component is the multiplication of the sparse matrix and the vector.
+Therefore, they migrated this part of calculation from the CPU into the fragment processor of the GPU, to take advantage of the fragment processor in its efficient manipulation of the local texture memory on the mathematical calculation.
+The performance of their method on the GPU can reach $ 2.15 $ times faster than the CPU implementation.
+
+In 2007, \citeauthor{Taylor07b} introduced the first GPU implementation of a fully non-linear finite element method (more details in \cite{Taylor08}).
+They reformulated the TLED algorithm presented by \cite{Miller07} to accomodate with the limited hardware possibilities of GPUs.
+In particular, the main restriction when using graphics-based GPU execution is the inability to scatter, which necessitated reformulation of force summation as a gather in their implementation.
+It is worth noting that the entire finite element method implementation was ported to GPU.
+They reported significant solution speed gains up to $ 16.8 \times$ over the CPU implementation.
+Thus, they were able to compute a simple cube model with up to $ 16\,000 $ tetrahedral elements in real-time.
 

### 2010-08-19 Thesis/Chapter4.tex
- Update 19 August -
Chapter 4 finished (well, I'd like to believe it is)



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@156 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-	\subsection{Meshless, position-based dynamics (supposedly good for handling topological changes but other issues)}
+	\subsection{Meshless methods (supposedly good for handling topological changes but other issues)}
+
+If the finite element method is a robust method and widely used in engineering, the method has a few shortcomings \citep{Liu05}.
+First, the creation of a mesh is compulsory and producing a good quality mesh is both difficult and time consuming.
+Second, the stresses obtained in FEM are often discontinuous at the interfaces of the elements.
+Special techniques are required in a post-processing stage to recover accurate stresses.
+In addition, under large deformations, considerable loss in accuracy in FEM results can arise from the element distortions.
+The origin of these problems is the use of elements.
+Consequently, the idea of getting rid of the elements and the mesh has emerged and the concept of \emph{meshless} methods was born.
+Instead, meshless methods use a set of nodes scattered within the problem domain.
+They do not form a mesh, meaning it does not require any a priori information on the relationship between the nodes for the interpolation of the unknown variables.
+In theory, this should facilitate topological changes, a feature fairly desirable in surgery simulation.
+It is worth noting though that meshless methods are still in their developing stage and are not as mature as finite element formulations.
+We will now briefly review the research carried out in soft tissue modelling for medical simulation that makes use of meshless method.
+A complete review and details on meshless methods are beyond the scope of this thesis.
+
+\bigskip
+
+A meshless method was introduced for the first time for modelling soft tissue in medical simulation by \cite{De01}.
+They use the so-called method of finite spheres (MFS).
+Nodal points are only sprinkled locally around the surgical tool tip and not over the whole domain.
+The interpolation is performed by functions that are nonzero only on spheres surrounding the nodes.
+A point collocation technique is used to generate the discrete equations.
+The authors qualitatively compared the method of finite spheres with the finite element technique from a commercial package and the displacement profil was similar in both cases.
+They were able to achieved a computational rate of about $ 100\, $Hz when 34 spheres where used.
+Real-time haptic rates of about $ 1\, $kHz was then obtained using a force extrapolation technique.
+
+\cite{Lim04} used the same technique of finite spheres introduced by \citeauthor{De01} to model soft tissue but they added progressive cutting, without the generation of new primitives.
+As the cut progresses, the nearest vertex to the intersection point of the tool is snapped to the tool path.
+Then, they observed that the prescribed boundary condition changes on only a very small portion as the tool interacts with the organ.
+Rather than solving the entire problem over and over again, they proposed instead to make incremental corrections, which results in an accelerated solution procedure.
+
+Meshless methods were then introduced to computer graphics by \cite{Muller04}.
+In each step, they compute the spatial derivatives of the displacement field using a moving least squares (MLS) procedure.
+From these derivatives, they obtain strains, stresses and elastic forces at each simulated point.
+In addition, they proposed techniques for modeling and animating a point-sampled surface that dynamically adapts to deformations of the underlying volumetric model.
+They used a non-linear measure of strain with a linear constitutive model.
+They reported real-time simulation of models exibiting elastic, plastic, melting and flowing effects.
+
+\cite{Horton06} tested meshless methods for surgical simulation.
+They proposed a total Lagrangian explicit dynamics algorithm using a non-linear material formulation.
+They applied the method on an example of brain shift and compared their results with a commercial finite element code.
+The shape functions are created from a cloud of unconnected nodes using the element free Galerkin method.
+They were interested in the displacement of the centreof mass for the brain and found the absolute displacement differences between the two algorithm are less than $ 0.85\,$mm which is the resolution of the MRIs in their study.
+The same team kept investigating and testing the technique.
+However, they used moving least squares procedure for creating the shape functions instead of the element free Galerkin method.
+They also realised experimental indentations \citep{Horton07} and provided more extensive comparisons with a finite element method \citep{Horton10}.
+
+Another example of application is the used of the technique called moving particule semi-implicit method (MPS) by \cite{Chhatkuli09}.
+They applied this method to simulate the deformation of lung and a tumor inside the left lung during inspiration.
+The lung tissues were considered to be homogeneous, isotropic and viscoelastic.
+They compared their results with the experimental CT taken at the end of inspiration and foundout that the deformation predicted by numerical simulation matches reasonably well with the experimental results.
+
+\bigskip
+
+\TODO{add drawbacks of meshless methods}

### 2010-08-22 Thesis/Chapter2.tex
- Update 23 August -
Auto-review of chapters 3 and 4 + summary of chapter 4 added



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@157 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 We can define volume elements in the reference and deformed configurations.
-Consider three non-coplanar line elements $d\mathbf{X}^{(1)}$, $d\mathbf{X}^{(2)}$ and $d\mathbf{X}^{(3)}$ forming a parallelepiped in the reference configuration as can be seen on Fig.~\ref{chap2:fig-changeVolume}.
+Consider three non-coplanar line elements $d\mathbf{X}^{(1)}$, $d\mathbf{X}^{(2)}$ and $d\mathbf{X}^{(3)}$ forming a parallelepiped in the reference configuration as can be seen on \fig{chap2:fig-changeVolume}.
 % 
@@
 For similar reasons, let us find the relation between an element of area in the reference configuration $d\mathbf{A}$ which becomes $d\mathbf{a}$ in the deformed configuration.
-Considering an arbitrary material line $d\mathbf{L}$ in the reference configuration and noting $d\mathbf{l}$ the same line after deformation (see Fig.~\ref{chap2:fig-changeSurface}, the corresponding reference and current volumes are given respectively by:
+Considering an arbitrary material line $d\mathbf{L}$ in the reference configuration and noting $d\mathbf{l}$ the same line after deformation (see \fig{chap2:fig-changeSurface}, the corresponding reference and current volumes are given respectively by:
 \begin{align}
@@
 
-To establish this relationship, let us consider an infinitesimal tetrahedron in a Cartesian coordinate system as represented Fig.~\ref{chap2:fig-CauchyStress}. 
+To establish this relationship, let us consider an infinitesimal tetrahedron in a Cartesian coordinate system as represented \fig{chap2:fig-CauchyStress}. 
 %

### 2010-08-22 Thesis/Chapter3.tex
- Update 17 August -
Major update on FEM state of the art + bib updated



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@153 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 23 August -
Auto-review of chapters 3 and 4 + summary of chapter 4 added



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@157 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\chapter{A practical approach of the Finite Element Method}
+\chapter{A practical approach of the finite element method}
 \label{chap3}
@@
 The next step is to apply Newton's second law (law of conservation of momentum), which yields a partial differential equation at every point of the volume.
-The Finite Element Method is a numerical procedure that allows to turn an infinite number of partial differential equations into a finite number of algebraic equations.
+The finite element method is a numerical procedure that allows to turn an infinite number of partial differential equations into a finite number of algebraic equations.
 The method follows different steps.
@@
 Numerical methods typically transform differential equations governing a continuum to a set of algebraic equations of a discrete model of the continuum that are to be solved using computers \citep{Reddy93}.
-The \emph{Finite Element Method} (FEM) is the most popular numerical procedure that is used to approximately solve differential equations, especially in continuum mechanics.
+The \emph{finite element method} (FEM) is the most popular numerical procedure that is used to approximately solve differential equations, especially in continuum mechanics.
 
@@
 However, sometimes the function describing the curve is not known, or is difficult to integrate.
-One method to obtain an approximate solution is to break up the area into a series of rectangles and add the areas of all rectangles (see Fig.~\ref{chap3:fig-areaUnderCurve}).
+One method to obtain an approximate solution is to break up the area into a series of rectangles and add the areas of all rectangles (see \fig{chap3:fig-areaUnderCurve}).
 It is worth nothing that the solution accuracy can be increased by reducing the width of the rectangles to better follow the curve.
@@
 \end{center}
-\caption[Approximation of the area under a curve]{Approximation of the area under a curve by adding the areas of all rectangles (left Rieman sum).}
+\caption[Approximation of the area under a curve]{Approximation of the area under a curve by adding the areas of all rectangles (left Rieman sum for this example).}
 \label{chap3:fig-areaUnderCurve}
@@
 It is crucial to keep in mind that approximations occur at different stages during finite element analysis.
-The division of the whole domain into finite elements may not be exact (see Fig.~\ref{chap3:fig-errorWithSubdivision}), introducing error in the domain being modelled.
+The division of the whole domain into finite elements may not be exact (see \fig{chap3:fig-errorWithSubdivision}), introducing error in the domain being modelled.
 The second stage is when element equations are derived.
@@
 Usually, the number of elements increases with the complexity of the geometry.
-For instance, if one part of the domain is thiner (see Fig.~\ref{chap3:fig-sizeOfElementMeshing}), the size of the elements must be reduced in order to tile this particular part, which increases the total number of elements, and hence  the number of algebraic equations to solve.
+For instance, if one part of the domain is thiner (see \fig{chap3:fig-sizeOfElementMeshing}), the size of the elements must be reduced in order to tile this particular part, which increases the total number of elements, and hence  the number of algebraic equations to solve.
 However, adding elements is sometimes desirable.
@@
 \end{center}
-\caption[Finer mesh required for complex geometry]{By adding details on the tree's geometry (right), the surface delimited by the tree becomes more difficult to mesh and would require smaller elements to tile all additional small branches.}
+\caption[Finer mesh required for complex geometry]{By adding details on the tree's geometry \textbf{(right)}, the surface delimited by the tree becomes more difficult to mesh and would require smaller elements to tile all additional small branches.}
 \label{chap3:fig-sizeOfElementMeshing}
@@
 The natural coordinate system allows us to map every element into a typical and simpler element.
-As an example, an 8-node hexahedron element is shown in Fig.~\ref{chap3:fig-naturalCoordinatesHexa} as it appears in the global $(x_1, x_2, x_3)$ and natural $(\xi_1,\xi_2,\xi_3)$ coordinate systems.
+As an example, an 8-node hexahedron element is shown in \fig{chap3:fig-naturalCoordinatesHexa} as it appears in the global $(x_1, x_2, x_3)$ and natural $(\xi_1,\xi_2,\xi_3)$ coordinate systems.
 While in its natural coordinate system the element is a regular aligned cube, in the global coordinate system the element may assume any admissible arbitrary form.
 Essentially, admissible means that the element must not be too distorted, and must certainly not fold back on itself.
-Another example is given Fig.~\ref{chap3:fig-naturalCoordinatesTetra} with a tetrahedral element.
+Another example is given \fig{chap3:fig-naturalCoordinatesTetra} with a tetrahedral element.
 The components of each node has a simple expression in natural coordinates, usually $0, 1$ or $-1$ and the centre of the coordinate system is taken at one of the nodes or at the centre of the element.
@@
 
-\section{Derivation of element equations}
+\section{Derivation of element equations}	\label{chap3:derivationEquations}
 
@@
 
-The construction of the weak form using either energy principles or weighted residual methods will not be detailled here.
+The construction of the weak form using either energy principles or weighted residual methods will not be detailed here.
 The reader may easily find it in dozens of textbooks.
@@
 It is fair to say that strain and stress within the structure does not fluctuate over time: the system is in an equilibrium state.
-At all times, the sum of all forces applied on the system is equal to $ 0 $.
+At all times, the sum of all forces applied on the system is equal to zero.
 Therefore, we can assume that the equations describing the structure do not depend on time: the problem is said to be \emph{static}.
-Accelerations can be neglected.
+Static analyses assume that loading of a body occurs slowly enough that inertial and damping terms may be neglected.
 In this case, the static system of equations can be easily obtained by merely dropping out the dynamics terms in \eqref{chap3:eqDynamic}.
@@
 The assembly operation is actually quite simple.
-Let us consider an example with stiffness matrices for one-dimensional problem.
+Let us consider an example with stiffness matrices for an one-dimensional problem.
 We assume that we have three two-node elements.
@@
 A sparse matrix is merely a matrix populated mainly with zeros.
-An example is given by Fig.~\ref{chap3:fig-sparse} for a finite element problem in two dimensions.
+An example is given by \fig{chap3:fig-sparse} for a finite element problem in two dimensions.
 \begin{figure}
@@
 While these methods are simple to derive, convergence is only guaranteed for a limited class of matrices.
-Examples of stationary iterative methods are the Jacobi method and Gauss?Seidel method.
+Examples of stationary iterative methods are the Jacobi method and Gauss-Seidel method.
 Krylov subspace methods work by forming an orthogonal basis of the sequence of successive matrix powers times the initial residual (the Krylov sequence).
@@
 \end{equation}
-Indeed, if the matrix $ \mathbf{A} $ is symmetric and positive-definite, $ f(\mathbf{x}) $ is minimised by the solution to $ \mathbf{A} \mathbf{x} = \mathbf{b} $\citep{Shewchuk94}.
+Indeed, if the matrix $ \mathbf{A} $ is symmetric and positive-definite, $ f(\mathbf{x}) $ is minimised by the solution to $ \mathbf{A} \mathbf{x} = \mathbf{b} $ \citep{Shewchuk94}.
 The minimisation is carried out by generating a succession of search directions $ \mathbf{p} _k $ and improved minimisers $ \mathbf{x} _k $.
 At each stage a quantity $ \alpha_k $ is found that minimises $f(\mathbf{x}_k  + \alpha_k \mathbf{p}_k)$, and $\mathbf{x}_{k+1}$ is set equal to the new point $\mathbf{x}_k + \alpha_k \mathbf{p}_k$.
-The $ \mathbf{p}_k $  and $ \mathbf{x}_k $  are built up in such a way that $ \mathbf{x}_{k+1} $  is also the minimizer of $ f $ over the whole vector space of directions already taken, that is the space formed by the basis of vector ${\mathbf{p}_1, \mathbf{p}_2, \ldots, \mathbf{p}_k }$.
+The $ \mathbf{p}_k $  and $ \mathbf{x}_k $  are built up in such a way that $ \mathbf{x}_{k+1} $  is also the minimiser of $ f $ over the whole vector space of directions already taken, that is the space formed by the basis of vector ${\mathbf{p}_1, \mathbf{p}_2, \ldots, \mathbf{p}_k }$.
 After $N$ iterations, you arrive at the minimiser over the entire vector space, that is, the solution to our linear system of equations.

### 2010-08-22 Thesis/Chapter4.tex
- Update 23 August -
Auto-review of chapters 3 and 4 + summary of chapter 4 added



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@157 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{shortAbstract}
-A short abstract for the upcoming chapter
+Modelling the deformation of solid organs is a crucial problem in medical simulation.
+If the finite element method is classicaly used in mechanical engineering for its ability to solve the equations of continuum mechanics fairly accurately, its development in medical simulation was more progressive.
+The main reason for this is that the finite element method is computationally demanding and not easily compatible with time constraints required in interactive simulation.
+Therefore, other methods were developed to allow faster computations, at the cost of larger approximations of course.
+Over time, along with the increase in computational power, methods became more elaborate and less approximative.
+These different approaches may be grouped into three main categories: (1) geometrically based techniques, (2) approaches physically motivated and (3) methods actually based on the equations of continuum mechanics.
+In this chapter, we will review the most classic methods applied to modelling of solid objects with a particular emphasis on medical simulation.
 \end{shortAbstract}
@@
 But by improving the dynamism of the scene, the simulation appears more natural to users.
-Over time, with the increase in computational power of computers, objects stopped being all rigid and simulators started taking laws of physics into account.
-In other words, they became deformable.
-If this took the realism a step further, demands in computational power were multiplied.
+Over time, with the increase of computers' computational power, objects stopped being all rigid and simulators started taking laws of physics into account.
+In other words, objects became deformable.
+If this took the realism a step further, the demands in computational power were multiplied.
 Simulators eventually faced a challenge: they had to weight the desired degree of realism against the computational power at their disposal.
@@
 However, strong time constraints often limit the complexity of the modelling and affects the overall accuracy.
-Because of this trade-off, various approaches were proposed over the years to fit into real-time constraints and may be grouped into three main categories: (1) geometrically based techniques, (2) approaches physically motivated usually relying on Newton's second law and (3) methods actually based on the equations of continuum mechanics.
+Because of this trade-off, various approaches were proposed over the years to fit into real-time constraints and may be grouped into three main categories: (1) geometrically based techniques, (2) approaches physically motivated (usually relying on Newton's second law) and (3) methods actually based on the equations of continuum mechanics.
 
@@
 It involves a mapping from $ \field{R}^3 $ to $ \field{R}^3 $ through a tensor product trivariate Bernstein polynomial.
-The deformed shape of the object is interpolated using control points as in the case of Bézier curves and surfaces patches (see Fig.~{chap4:fig-cubeFFD}).
+The deformed shape of the object is interpolated using control points as in the case of Bézier curves and surface patches (see \fig{chap4:fig-cubeFFD}).
 %
@@
 \end{center}
-\caption[Deformation of objects by free-form deformation]{Let us imagine that the objects that we want to deform (spheres and cubes) are placed into a transparent parallelepiped.
+\caption[Deformation of objects by free-form deformation]{Let us imagine that the objects that we want to deform (spheres and cubes) are placed into a transparent parallelepiped \textbf{(left)}.
 Control points are indicated by small white diamonds.
-The free-form deformation technique offers to compute the deformation of the objects from the positions of the control points (right).}
+The free-form deformation technique offers to compute the deformation of the objects from the positions of the control points \textbf{(right)}.}
 \label{chap4:fig-cubeFFD}
@@
 But this is beyond the point.
-This method allows the authors to simulate hundreds of deformable objects at an interactive rate with an unconditional stability (see Fig.~\ref{chap4:fig-shoes}).
-Another illustration of this stability is shown Fig.~\ref{chap4:fig-duck}.
+This method allows the authors to simulate hundreds of deformable objects at an interactive rate with an unconditional stability (see \fig{chap4:fig-shoes}).
+Another illustration of this stability is shown \fig{chap4:fig-duck}.
 %
@@
 \end{center}
-\caption[Squeezing of a duck using Shape Matching]{A duck model is squeezed to demonstrate the stability of the shape matching technique.
+\caption[Squeezing of a duck using shape matching]{A duck model is squeezed to demonstrate the stability of the shape matching technique.
 Although the flat shape of the duck is not physically plausible, this example exposes the impressive ability of the approach to recover from highly deformed or inverted configurations.}
@@
 A particule is placed at each vertex of the lattice and associated with a shape matching region.
-Their technique of \emph{Fast Lattice Shape Matching} allows them to compute per-region transformations in an efficient manner (see Fig.~\ref{chap4:fig-penguins}).
+Their technique of \emph{fast lattice shape matching} allows them to compute per-region transformations in an efficient manner (see \fig{chap4:fig-penguins}).
 %
@@
 \end{center}
-\caption[Penguins animation with Fast Lattice Shape Matching]{(\textbf{Left}) $150$ cartoon penguins deforming dynamically using Fast Lattice Shape Matching.
-(\textbf{Right}) Deformed lattices consisting of $150$ particles per penguin ($22\,500$ particles).
+\caption[Penguins animation with fast lattice shape matching]{\textbf{(Left)} $150$ cartoon penguins deforming dynamically using fast lattice shape matching\textbf{(Right)}.
+Deformed lattices consisting of $150$ particles per penguin ($22\,500$ particles).
 The penguins can be deformed at $25$ frames per second on a Pentium4 $3.4 \,$GHz.}
@@
 Therefore, \cite{Steinemann08} further improved the technique by allowing dynamic adaptive selection of levels of detail through the use of an octree-based hierarchical sampling.
-This adaptative sampling is illustrate with Fig.~\ref{chap4:fig-flowers}.
+This adaptative sampling is illustrate with \fig{chap4:fig-flowers}.
 Their technique also handles efficient topological changes.
@@
 A coarse sampling is employed when the flowers are moved by the wind, for a total of $5\,680$ nodes in the scene.
-However, the models are dynamically refined when touched by the user, as shown in the left image.}
+However, the models are dynamically refined when touched by the user, as shown in the right image.}
 \label{chap4:fig-flowers}
@@
 If a link between two nodes is stretched to its limit, displacements are transferred to its neighbouring links.
-It works in a very similar way to a chain and is illustrated by Fig.~\ref{chap4:fig-chainmail}.
+It works in a very similar way to a chain and is illustrated by \fig{chap4:fig-chainmail}.
 %
@@
 \end{equation}	
-where $ \mathbf{M} $ is the mass matrix, $\mathbf{D}$ is the damping matrix, $ \mathbf{K} $ is the stiffness matrix and $\mathbf{R}$ are externally applied loads.
+where $ \mathbf{M} $ is the mass matrix, $\mathbf{D}$ is the damping matrix, $ \mathbf{K} $ is the stiffness matrix and $\mathbf{R}$ are the externally applied loads.
 Modal decomposition refers to the process of diagonalising equation \eqref{chap4:linearisedEquation}.
@@
 \cite{James02} went even further by using graphics hardware to reduce CPU costs to a minimum.
-In particular, they use modal analysis to model human skin and secondary tissues in a laparascopic surgical simulation to allow the CPU to focus on more complex tissue models and user contact interactions (see Fig.~\ref{chap4:fig-modalAnalysisLaparo}).
+In particular, they use modal analysis to model human skin and secondary tissues in a laparascopic surgical simulation to allow the CPU to focus on more complex tissue models and user contact interactions (see \fig{chap4:fig-modalAnalysisLaparo}).
 The technique is fairly efficient as it allows precomputations of the vibration modes.
@@
 \end{equation}
-For complex systems with more than one spring attached to each mass, the right-hand side of the equation above becomes more complex as it needs to be repeated and adjusted for each spring.
+For complex systems with more than one spring attached to each mass, the right-hand side of the equation \eqref{chap4:massSpring3} becomes more complex as it needs to be repeated and adjusted for each spring.
 The easiest way to solve a complex mass-spring system iteratively, is to first calculate the total force currently acting on each mass by summing all forces currently exerted by all springs attached to this mass (using \eqref{chap4:massSpring1} and \eqref{chap4:massSpring2}).
@@
 Because they do not derive from the equations of continuum mechanics, mass-spring models have a limited capacity to model the various aspects of a material like anisotropy, viscoelasticity etc.
-Even worse, according to \cite{Bourguignon00}, if all springs are set to the same stiffness, the mesh geometry may generate undesired anisotropy as shown in Fig.~\ref{chap4:fig-undesiredAnisotropy}.
+Even worse, according to \cite{Bourguignon00}, if all springs are set to the same stiffness, the mesh geometry may generate undesired anisotropy as shown in \fig{chap4:fig-undesiredAnisotropy}.
 However, some works tried to take advantage of this feature by giving directions of interest to their model by specifically designing the mesh to align springs on specific directions.
@@
 \caption[Undesired anisotropy with mass-spring models]{The two meshes are undergoing a downward pull at their bottom.
-While the tetrahedral mass-spring system (left) shows signs of anisotropy, the hexahedral mesh (right) with springs aligned along the gravity does not.}
+While the tetrahedral mass-spring system \textbf{(left)} shows signs of anisotropy, the hexahedral mesh \textbf{(right)} with springs aligned along the gravity does not.}
 \label{chap4:fig-undesiredAnisotropy}
@@
 Indeed, \cite{Stiles72} proposed a viscoelastic mass-spring system to test hypotheses on the mechanical response of muscles.
-Much more recently, \cite{Tamura05} introduced a method to simulate viscoelastic material using a spring-mass system.
+Much more recently, \cite{Tamura05} introduced a method to simulate viscoelastic material using a mass-spring system.
 They use a large number of particules to create a randomly connected mesh to mimic the structure of polymeric materials and hence their viscoelascity characteristics.
@@
 
-	\subsection{The Finite Element Method}
+	\subsection{The finite element method}
 The finite element method brings accuracy a step further.
@@
 They derived a geometrically non-linear strain tensor from differential geometry theory.
-Obviously, the computations were not carried out in real-time but this technique allows \citeauthor{Terzopoulos87} to calculate the interaction between solid and deformable models in a physically realistic manner (see Fig.~\ref{chap4:fig-ball}).
+Obviously, the computations were not carried out in real-time but this technique allows \citeauthor{Terzopoulos87} to calculate the interaction between solid and deformable models in a physically realistic manner (see \fig{chap4:fig-ball}).
  %
@@
 The efficiency of this hybrid model was demonstrated by simulating an hepatectomy (removing of one of the eight anatomical segments of the liver).
-They reported the simulation (allowing deformation and cutting) of a mesh with more than $ 8\,000 $ tetrahedra in real-time (see Fig.~\ref{chap4:fig-hybridModel}).
+They reported the simulation (allowing deformation and cutting) of a mesh with more than $ 8\,000 $ tetrahedra in real-time (see \fig{chap4:fig-hybridModel}).
 %
@@
 
-\cite{Serby01} described a new approach to cutting in finite element models.
+\cite{Serby01} described a new approach to cutting with finite element models.
 The central idea is not to introduce new nodes/elements but to displace the existing ones to account for the topological changes introduced by a cut.
@@
 \cite{Hauth04} made use of the polar decomposition of the deformation gradient to extract the rotation.
-They also applied this formulation in a hierarchical finite element setting where the coarse level is computed with a fully non-linear strain tensor and the finer levels are computed with a co-rotational method.
+They also applied this formulation in a hierarchical finite element setting where the coarse level is computed with a non-linear strain tensor and the finer levels are computed with a co-rotational method.
 In addition, they examined the error that is arising from using the co-rotational approximation instead of a non-linear strain tensor.
@@
 While developing a teleoperated micro-surgical robot for eye surgery, they wanted to provide a virtual environment has part of the system for training.
-Since the cornea is anisotropic and non-linearly elastic and undergoes large deformations, they have used large deformation elasticity theory with orthotopic and non-linear material properties in the finite element model.
-Coupled with a detailed visual model of the eye, the simulation produces a fair level of realism with a performance greater than $ 10 \,$Hz (see Fig.~\ref{chap4:fig-eyeSurgery}).
+Since the cornea is anisotropic and non-linearly elastic and undergoes large deformations, they have used large deformation elasticity theory with orthotropic and non-linear material properties in the finite element model.
+Coupled with a detailed visual model of the eye, the simulation produces a fair level of realism with a performance greater than $ 10 \,$Hz (see \fig{chap4:fig-eyeSurgery}).
 %
@@
 They provided an adaptative method for animating viscoelastic deformable objects in a guaranteed frame rate.
-As the object moves and deforms, the sampling is refined to concentrate the computational load into the regions that deform the most (see Fig.~\ref{chap4:fig-gummyBear}).
+As the object moves and deforms, the sampling is refined to concentrate the computational load into the regions that deform the most (see \fig{chap4:fig-gummyBear}).
 They reported animating a few hundred points in real-time at a frequency higher than $ 300\,$Hz.
@@
 
-Recently, research on non-linear finite element increased significantly.
+Recently, research on non-linear finite element and more elaborate constitutive models increased significantly.
 For instance, \cite{Pathmanathan04} used a fully non-linear model to predict deformation of breast under compression during mammography.
@@
 Early 2000s, the rapid increase in the performance of graphics hardware, coupled with recent improvements in its programmability, have made graphics hardware a compelling platform for computationally demanding tasks in a wide variety of application domains.
-Researchers and developers have become interested in harnessing this power for general-purpose computing, an effort known collectively as GPGPU (for General-Purpose computing on the GPU).
+Researchers and developers have become interested in harnessing this power for general-purpose computing, an effort known collectively as GPGPU (for General-Purpose computing on Graphics Processing Units).
 A survey of GPGPU algorithms and techniques was written by \cite{Owens07}.
 
-To the best of our knowledge, \cite{Rumpf01} were the firsts to leverage the power of Graphics Processing Units (GPU) to solve partial differential equations.
+To the best of our knowledge, \cite{Rumpf01} were the firsts to leverage the power of GPUs to solve partial differential equations.
 They established a correspondence between common mathematical operations and OpenGL operations and structures.
@@
 \cite{Wu04} proposed the first application of GPU in medical simulation.
-They were able to achieve an interactive frame rate for the topolgy change in surgical simulation.
-Not only they used the condensation technique for merely calculating the surface nodes in the non-operation parts, but they implemented the conjugate gradient solver to GPU.
+They were able to achieve an interactive frame rate with topology changes in surgical simulation.
+Not only they used the condensation technique for merely calculating the surface nodes in the non-operation parts, but they implemented the conjugate gradient solver onto GPU.
 Indeed, entering 2003, the full IEEE floating point precision data type supported in shaders and texture stages allows the GPU to be applied into more general applications.
@@
 Therefore, they migrated this part of calculation from the CPU into the fragment processor of the GPU, to take advantage of the fragment processor in its efficient manipulation of the local texture memory on the mathematical calculation.
-The performance of their method on the GPU can reach $ 2.15 $ times faster than the CPU implementation.
+The performance of their method on the GPU is up to $ 2.15 $ times faster than the CPU implementation.
 
@@
 
-	\subsection{Meshless methods (supposedly good for handling topological changes but other issues)}
+	\subsection{Meshless methods}
 
@@
 The authors qualitatively compared the method of finite spheres with the finite element technique from a commercial package and the displacement profil was similar in both cases.
-They were able to achieved a computational rate of about $ 100\, $Hz when 34 spheres where used.
+They were able to achieve a computational rate of about $ 100\, $Hz when 34 spheres where used.
 Real-time haptic rates of about $ 1\, $kHz was then obtained using a force extrapolation technique.
@@
 From these derivatives, they obtain strains, stresses and elastic forces at each simulated point.
-In addition, they proposed techniques for modeling and animating a point-sampled surface that dynamically adapts to deformations of the underlying volumetric model.
+In addition, they proposed techniques for modelling and animating a point-sampled surface that dynamically adapts to deformations of the underlying volumetric model.
 They used a non-linear measure of strain with a linear constitutive model.
@@
 The shape functions are created from a cloud of unconnected nodes using the element free Galerkin method.
-They were interested in the displacement of the centreof mass for the brain and found the absolute displacement differences between the two algorithm are less than $ 0.85\,$mm which is the resolution of the MRIs in their study.
+They were interested in the displacement of the centre of mass for the brain and found the absolute displacement differences between the two algorithm are less than $ 0.85\,$mm which is the resolution of the MRIs in their study.
 The same team kept investigating and testing the technique.
@@
 The lung tissues were considered to be homogeneous, isotropic and viscoelastic.
-They compared their results with the experimental CT taken at the end of inspiration and foundout that the deformation predicted by numerical simulation matches reasonably well with the experimental results.
+They compared their results with the experimental CT taken at the end of inspiration and found out that the deformation predicted by numerical simulation matches reasonably well with the experimental results.
 
@@
 
-\TODO{add drawbacks of meshless methods}
+\TODO{add drawbacks of meshless methods: supposedly good for handling topological changes but other issues. Couldn't really find drawbacks in meshless' literature.}

### 2010-08-25 Thesis/Chapter2.tex
- Update 25 August -
Started chapter5 (TLED description) + update of labels + package left indices + bib updated



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@165 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\section{Description of motion}
+\section{Description of motion} 	\label{chap2:descriptionMotion}
 Let us consider a body $\textbf{\textit{B}}$ of known geometry in a three-dimensional Euclidian space $\field{R}^3$.
@@
 
-	\subsection{Green strain tensor}
+	\subsection{Green-Lagrange strain tensor}
 Using \eqref{chap2:dS} and \eqref{chap2:ds}, the change in the squared lengths due to the body deformation between the reference and the current configuration can be expressed as:
@@
 \end{align}
-Let us define the \emph{Green-St.\,Venant strain tensor} $\mathbf{E}$ as:
+Let us define the \emph{Green-Lagrange strain tensor} $\mathbf{E}$ as:
 \begin{equation}
@@
 \end{equation}
-By definition, the Green strain tensor is a symmetric second-order tensor.
+By definition, the Green-Lagrange strain tensor is a symmetric second-order tensor.
 Also, the change in squared lengths is zero if and only if $\mathbf{E} = \mathbf{0}$.
@@
 
-	\subsection{Elasticity}	
+	\subsection{Elasticity}	\label{chap2:elasticity}
 If under applied loads a material stores but does not dissipate energy, and it returns to its orginal shape when the loads are removed, we call that material \emph{elastic}.
@@
 \end{equation}
-The neo-Hookean model provides a reasonable approximation of rubber-like material for moderate strains. It is typically accurate for strains less than $20\, \%$ \citep{Gent01}.
+The neo-Hookean model provides a reasonable approximation of rubber-like material for moderate strains.
+\TODO{Do I keep this range of validity, which seems somewhat limited...} It is typically accurate for strains less than $20\, \%$ \citep{Gent01}.
 

### 2010-08-25 Thesis/Chapter3.tex
- Update 25 August -
Started chapter5 (TLED description) + update of labels + package left indices + bib updated



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@165 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{equation}
-\mathbf{M}^e = \int_{V^e} \rho^e \mathbf{H}^T \mathbf{H} dV
+\mathbf{M}^e = \int_{V^e} \rho^e \mathbf{H}^T \mathbf{H} \, dV
 \end{equation}
@@
 \begin{equation}
-\mathbf{D}^e = \int_{V^e} \kappa^e \mathbf{H}^T \mathbf{H} dV
+\mathbf{D}^e = \int_{V^e} \kappa^e \mathbf{H}^T \mathbf{H} \, dV
 \end{equation}
@@
 
-		\subsubsection*{The central difference method}	
+		\subsubsection*{The central difference method}		\label{chap3:centralDifferenceMethod}
 The central difference method is one of the most popular explicit methods in computational mechanics.
@@
 \begin{equation}
+\label{chap3:relationCDM}
 \left( \dfrac{\mathbf{M}}{(\Delta t)^2} + \dfrac{\mathbf{D}}{2 \Delta t} \right) \mathbf{U}^{t+\Delta t} = \mathbf{R}^t - \mathbf{F}^t + \dfrac{2 \mathbf{M}}{(\Delta t)^2} \mathbf{U}^t + \left( \dfrac{\mathbf{D}}{2 \Delta t} - \dfrac{\mathbf{M}}{(\Delta t)^2} \right) \mathbf{U}^{t-\Delta t}.
@@
 
+\paragraph*{\TODO{Cholesky decomposition?}}
+
 		\subsubsection*{Iterative solvers}

### 2010-08-25 Thesis/Chapter5.tex
- Update 25 August -
Started chapter5 (TLED description) + update of labels + package left indices + bib updated



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@165 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\chapter{Introduction of TLED}
+\chapter{The total Lagrangian explicit dynamics (TLED) algorithm}
 \label{chap5}
@@
 
-\section{Description of the algorithm (neo-Hookean, explicit integration, independent elements, parallelisable on GPU)}
+\section{Description of the TLED algorithm}
 
-\section{Visco-elasticity and anisotropy added \OC{MICCAI 2008; MedIA 2009}}
+	\subsection{Key ideas}
+
+The total Lagrangian explicit dynamics (TLED) algorithm was first introduced to the field of medical simulation by \cite{Miller07}.
+More details on this finite element method formulation may also be found in \cite{Bathe95}.
+It is worth noting that the finite element analysis carried out by the TLED algorithm is dynamic and fully non-linear (geometrically and materially).
+Let us begin by exposing the key ideas of the approach.
+
+The deformation of a body may be described by two different kind of description: Lagrangian or Eulerian (see section~\ref{chap2:descriptionMotion}).
+Because the Lagrangian description focuses its attention on the particles of the continuous body, it is usually used in solid mechanics.
+In the Lagrangian description, the position and physical properties of the particles are referred to a reference configuration.
+For the great majority of commercial finite element programs, the reference configuration is the previous configuration, that is the one at the end of the previous time step.
+In this case, where all variables are referred to the previous configuration, the formulation is called \emph{updated Lagrangian}.
+The advantage of this approach is the simplicity of the incremental strain description.
+The disadvantage is that all derivatives with respect to spatial coordinates must be recomputed in each time step, because the reference configuration is changing.
+The reason for this choice is historical, at the time of solver development the memory was expensive and caused more problems than actual speed of computations \citep{Miller07}.
+
+The first key idea of the TLED algorithm is to refer all variables to the underformed configuration.
+This type of formulation is said to be \emph{total Lagrangian}.
+Because of this, the choice of the second Piola-Kirchhoff stress tensor as a measure of stress is required.
+The strain measure that is work-conjugate with the second Piola-Kirchhoff stress is the Green-St.Venant strain tensor.
+That is,  the work of stress increments on the strain increments gives an accurate expression for work \citep{Ji10}.
+We should note that all derivatives in the definition of the Green tensor are with respect to the initial and undeformed configuration.
+The decisive advantage is that all derivatives are calculated with respect to the undeformed configuration and therefore can be precomputed.
+This is at the cost to a more complex strain-displacement matrix due an initial displacement effect in the incremental strain \citep{Bathe95}.
+However, the TLED algorithm performs significantly fewer mathematical operations in each time step.
+
+The second crucial idea is to use the central difference method for an explicit time integration of dynamic equilibrium equations.
+The advantage is that the stiffness term $ \mathbf{K(\mathbf{U}).\mathbf{U}} $ of the system of equations may be computed from:
+\begin{equation}
+\mathbf{K(\mathbf{U}).\mathbf{U}} = \mathbf{F}(\mathbf{U}) = \sum_e \mathbf{\tilde{F}}^e,
+\end{equation}
+where $ \mathbf{\tilde{F}}^e $ are the global nodal force contributions due to stresses in element $ \Omega_e $.
+This implies that the stiffness matrix does not need to be assembled since element nodal force contributions can be computed at the element level instead, which is a decisive computational advantage.
+
+	\subsection{Computation of element nodal forces}
+
+We adopt the notation of \citeauthor{Bathe95} with respect to indication of the relevant configuration of the body: a left superscript indicates the configuration in which a quantity occurs and, when applicable, a left subscript indicates the configuration with respect to which the quantity is measured.
+We note the coordinates of a point at time t $ ^t \mathbf{x}$.
+
+\subsubsection*{The deformation gradient} 
+As we know, a fundamental measure of deformation is the deformation gradient tensor $ _0^t \mathbf{X} $ and we recall that it may be written as:
+\begin{equation}
+_0^t\mathbf{X} = \pd{^t\mathbf{x}}{^0\mathbf{x}}.
+\end{equation}
+This tensor describes the stretches and rotations that the material fibres have undergone from time $0$ to time $t$.
+In order to compute the element deformation gradients, we first compute derivatives of displacements $ u_{i,j} $ with respect to global coordinates.
+Since we use a total Lagrangian framework, the derivatives are referred to the undeformed configuration and we have:
+\begin{equation}
+_0^t u_{i,j} = \pd{^t u_i}{_0 x_j} = \sum_{a=1}^N \pd{h_a}{_0 x_j} \, ^t u_{ai},
+\end{equation}
+where $ h_a $ is the shape function associated with node $ a $ and $ N $ the number of nodes of the element. If we define a matrix $ _0^t \mathbf{\partial u_x} $ of displacements derivatives 
+\begin{equation}
+_0^t\mathbf{\partial u_x} = 
+	\begin{bmatrix}
+	_0^t u_{1,1} &         _0^t u_{1,2}      &        _0^t u_{1,3}       \\\\
+	_0^t u_{2,1} &         _0^t u_{2,2}      &        _0^t u_{2,3}       \\\\
+	_0^t u_{3,1} &         _0^t u_{3,2}      &        _0^t u_{3,3}       
+	\end{bmatrix}	
+\end{equation}
+The deformation gradient may be obtained from the displacement derivatives with:
+\begin{equation}
+_0^t\mathbf{X} = \leftidx{_0^t}{ \partial \mathbf{u}}{\mathbf{x}} + \mathbf{I}.
+\end{equation}
+
+\subsubsection*{The Green-Lagrange strain tensor} 
+From the deformation gradient we may obtain the right Cauchy-Green deformation tensor $ _0^t \mathbf{C}  $:
+\begin{equation}
+_0^t\mathbf{C} = \leftidx{_0^t}{ \mathbf{X} }{^T} \, _0^t\mathbf{X},
+\end{equation}
+and then yields the Green-Lagrange strain tensor $ _0^t \mathbf{E}  $:
+\begin{equation}
+\label{chap5:E}
+_0^t\mathbf{E}  = \dfrac{1}{2} (_0^t\mathbf{C} - \mathbf{I}), 
+\end{equation}
+where $ \mathbf{I} $ is the rank $ 2 $ identity tensor.
+
+		\subsubsection*{Constitutive equations and evaluation of stress}
+We know that the stresses result from the deformation of the material and they may be expressed in terms of some measure of this deformation such as the strain.
+The constitutive equations, which depends on the material under consideration, relate the stresses to the strain.
+As we have seen in section \ref{chap2:elasticity}, hyperelastic materials are a general class of materials in which the constitutive relationship is expressed in the form of a strain energy density function $ _0^t W $.
+In such materials, the stresses are obtained by differentiating this energy density function with respect to the appropriate strain measure (that is energically conjugate).
+In a total Lagrangian framework, the appropriate stress and strain measures are second Piola-Kirchhoff stress $ _0^t \mathbf{S} $ and Green-Lagrange strain $ _0^t \mathbf{E} $.
+Consequently, the stress may be computed from:
+\begin{equation}
+_0^t\mathbf{S} = \pd{ \leftidx{_0^t}{ \mathbf{W} }{} }{_0^t \mathbf{E}}.
+\end{equation}
+
+The form of the strain energy function depends on the chosen material.
+For the TLED algorithm we choose a neo-Hookean material already described section \ref{chap2:non-linear}.
+For a compressible neo-Hookean material, the strain energy density is given by:
+\begin{equation}
+_0^t W = \dfrac{1}{2} \mu (_0^t I_1 - 3 - 2 \ln ^t J) + \dfrac{1}{2} \lambda (^t J - 1)^2,
+\end{equation}
+where $_0^t I_1 $ is the first invariant of the right Cauchy-Green deformation tensor $ _0^t \mathbf{C} $ given by $ _0^t I_1 = tr(_0^t \mathbf{C}) $ and $ ^t J = det(_0^t \mathbf{X}) = det(_0^t \mathbf{C})^2 $ is the Jacobian.
+%		
+Then, using the chain rule we have:
+\begin{equation}
+\pd{_0^t W}{_0^t\mathbf{E}} = \pd{_0^t W}{_0^t\mathbf{C}} \pd{_0^t\mathbf{C}}{_0^t\mathbf{E}}.
+\end{equation}
+Noting that
+\begin{equation}
+\pd{_0^t\mathbf{C}}{_0^t\mathbf{E}} = 2 \quad \mbox{by \eqref{chap5:E}},
+\end{equation}
+Therefore, we may evaluate the stress from:
+\begin{equation}
+_0^t\mathbf{S}  = 2 \pd{_0^t W}{_0^t\mathbf{C}},
+\end{equation}
+which we can then expressed as the following:
+\begin{equation}
+_0^t\mathbf{S}  = 2 \left( \pd{_0^t W}{_0^t\mathbf{I_1}} \pd{_0^t\mathbf{I_1}}{_0^t\mathbf{E}} + \pd{_0^t W}{^t J} \pd{^t J}{_0^t\mathbf{E}} \right). 
+\end{equation}
+It may be shown that this yields:
+\begin{equation}
+_0^t S_{ij} = \mu (\delta_{ij} - _0^tC_{ij}^{-1}) + \lambda ^t J (^t J -1) _0^t C_{ij}^{-1}.
+\end{equation}
+
+		\subsubsection*{Element nodal forces}
+For a given element at time $ t $, the global nodal force contributions $ \mathbf{\tilde{F}} $ due to stresses in the element may be computed from:
+\begin{equation}
+\label{chap5:elementForce}
+\leftidx{^t}{ \mathbf{\tilde{F}} }{} = \int_{^0 V} \, \leftidx{_0^t}{ \mathbf{B} }{_L^t} \, \leftidx{_0^t}{ \mathbf{\hat{S}} }{} \, d\leftidx{^0}{ V }{},
+\end{equation}
+where $ _0^t\mathbf{\hat{S}} $ is the vector form of the second Piola-Kirchhoff stress given by:
+\begin{equation}
+_0^t\mathbf{\hat{S}} = \left[ _0^t S_{11} \quad _0^t S_{22} \quad _0^t S_{33} \quad _0^t S_{12} \quad _0^t S_{23} \quad _0^t S_{13} \right] ^T,
+\end{equation}
+$ _0^t\mathbf{B}^T_L $ is the strain-displacement matrix and $ d^0V $ is the initial (undeformed) volume of the element.
+The strain-displacement matrix $ \mathbf{B}_L $ relates the strains in an element to the element's nodal displacements.
+If an implicit analysis is performed, the strain-displacement matrix is used in the assembly of the stiffness matrix.
+In explicit analyses it is used to compute element nodal force contributions according to \eqref{chap5:elementForce}.
+In geometrically linear analyses, we assume that the displacements are infinitesimally small so that the geometry of an element does not change over time and the strain-displacement matrix is constant.
+Conversely, because geometrically non-linear analyses (such as carried out by the TLED algorithm) take the change of geometry of the elements into account, the strain-displacement matrix $ \mathbf{B}_L = _0^t\mathbf{B}_L $ varies.
+However, the strain-displacement matrix at time $ t $ may be computed by transforming a stationary matrix $ _0^t\mathbf{B}_{L0} $ using the deformation gradient.
+We begin by defining the stationary matrix as:
+\begin{equation}
+_0^t\mathbf{B}_{L0} = \left[ _0\mathbf{B}_{L0}^{(1)} \quad _0\mathbf{B}_{L0}^{(2)} \quad \ldots \quad _0\mathbf{B}_{L0}^{(N)} \right],
+\end{equation}
+where $ N $ is the number of nodes per element and the submatrix $ _0\mathbf{B}_{L0}^{(a)}  $ are given by:
+\begin{equation}
+_0\mathbf{B}_{L0}^{(a)} =
+	\begin{bmatrix}
+	_0 h_{a,1} & 0 & 0 \\
+	0 & _0 h_{a,2} & 0 \\
+	0 & 0 & _0 h_{a,3} \\
+	_0 h_{a,2} & _0 h_{a,1} & 0 \\
+	0 & _0 h_{a,3} & _0 h_{a,2} \\
+	_0 h_{a,3} & 0 & _0 h_{a,1}
+	\end{bmatrix}
+\quad a = 1, 2, \ldots, N.
+\end{equation}
+The subscripted comma denotes partial differentiation such as:
+\begin{equation}
+_0 h_{a,i} = \pd{h_a}{^0 x_i}.
+\end{equation}
+The full strain-displacement matrix (accounting for initial displacement effect) is then computed via
+\begin{equation}
+_0^t\mathbf{B}_{L}^{(a)} = \leftidx{_0^t}{ \mathbf{B} }{_{L0}^{(a)}} \, \leftidx{_0^t}{ \mathbf{X} }{^T}.
+\end{equation} 
+We note that the stationary strain-displacement matrix is composed of derivatives of shape functions with respect to the original undeformed coordinates of the body.
+These derivatives are therefore constant and may be precomputed, affording the total Lagrangian formulation a significant computational advantage.
+
+	\subsection{Computation of node displacements}
+We recall that the dynamic system of equations describing a deformable solid is the following (see section \ref{chap3:dynamic2}:
+\begin{equation}
+\label{chap5:eqDynamic}
+\mathbf{M} \mathbf{\ddot U} + \mathbf{D} \mathbf{ \dot U} + \mathbf{K}(\mathbf{U}) \cdot \mathbf{U} = \mathbf{R}.
+\end{equation}		
+where $ \mathbf{M} $ is a constant mass matrix, $\mathbf{D}$ is a constant damping matrix, $ \mathbf{K}^e(\mathbf{U}^e) $ is the stiffness matrix, which is a function of nodal displacements $\mathbf{U}$, and $\mathbf{R}$ are externally applied loads.
+And we seek the unknown displacements $ \mathbf{U} $.
+
+		\subsubsection*{Mass and damping matrices}					
+The mass matrix for the system may be computed by summing contributions from individual elements:
+\begin{equation}
+\mathbf{M} = \sum_e \mathbf{M}^e,
+\end{equation}		
+where $ \mathbf{M}^e $ is the mass contribution from element $ \Omega_e $ given by:
+\begin{equation}
+\mathbf{M}^e = \int_{^0 V_e} \rho_e \mathbf{H}^T \mathbf{H} \, d^0 V.
+\end{equation}
+Following the total Lagrangian framework, we observe that the expression is integrated over the undeformed volume and the mass density $ \rho_e $ of the undeformed configuration is also used.
+We then diagonalise the mass matrix by applying the technique of mass lumping as explained in section \ref{chap3:wordOnMatrices}.
+Therefore, the mass matrix may be built by directly computing the mass of each element and assigning an equal proportion of this to each of the element's nodes.
+In other words, if the mass of element $ \Omega_e $ is $ m_e $ then for every node $ a $ which is attached to this element, the $ a^{th} $ diagonal component $ M_{aa} $ of the global mass matrix will receive a contribution of $ m_e/N $ where $ N $ is the number of nodes per element.
+The complete matrix is compiled by summing contributions from all elements.
+
+We also employ a Rayleigh damping from which we only consider the mass-proportional component to obtain a diagonal damping matrix. Consequently, $ \mathbf{D} $ is computed from:
+\begin{equation}
+\mathbf{D} = \alpha \mathbf{M}.
+\end{equation}
+
+		\subsubsection*{Gathering of nodal forces}
+As stated early in this chapter, the stiffness matrix of \eqref{chap5:eqDynamic} may be obtained from
+\begin{equation}
+\mathbf{K(\mathbf{U}).\mathbf{U}} = \mathbf{F}(\mathbf{U}) = \sum_e \mathbf{\tilde{F}}^e.
+\end{equation}
+The previous section detailed how to calculate element nodal forces $ \mathbf{\tilde{F}}^e $.
+For each node we can now add the element force contributions up from all elements attached to this node.
+Gathering those nodal force contributions allows us to compute the stiffness term of the equilibrium equations $ ^t\mathbf{F} $.
+Therefore, \eqref{chap5:eqDynamic} may be re-written as the following:
+\begin{equation}
+\label{chap5:eqDynamic2}
+\mathbf{M} \leftidx{^t}{ \mathbf{\mathbf{\ddot U}} }{} + \mathbf{D} \leftidx{^t}{ \mathbf{\mathbf{\dot U}} }{} + \leftidx{^t}{\mathbf{F}}{} = \leftidx{^t}{\mathbf{R}}{}.
+\end{equation}
+
+		\subsubsection*{Explicit time integration}
+The remaining step is the time integration of \eqref{chap5:eqDynamic2} to find out the expression of $ ^t\mathbf{U} $.
+For this, we employ the explicit central difference method described page \pageref{chap3:centralDifferenceMethod}.
+We assume that all displacements, velocities and accelerations for the current time step is known and we seek a formula for computing displacements at time $ t+\Delta t $.
+Let us remind \eqref{chap3:relationCDM} for convenience:
+\begin{equation}
+\left( \dfrac{\mathbf{M}}{(\Delta t)^2} + \dfrac{\mathbf{D}}{2 \Delta t} \right) \mathbf{U}^{t+\Delta t} = \mathbf{R}^t - \mathbf{F}^t + \dfrac{2 \mathbf{M}}{(\Delta t)^2} \mathbf{U}^t + \left( \dfrac{\mathbf{D}}{2 \Delta t} - \dfrac{\mathbf{M}}{(\Delta t)^2} \right) \mathbf{U}^{t-\Delta t}.
+\end{equation}
+By using diagonalised mass and damping matrices, this expression may be rearranged to give a formula for updating displacements component-wise:
+\begin{equation}
+\label{chap5:CDMcomponents}
+\leftidx{^{t+\Delta t}}{ \mathbf{U} }{_i} = \dfrac{\leftidx{^t}{R}{_i} - \leftidx{^t}{F}{_i} + \frac{2 M_{ii}}{\Delta t^2} \leftidx{^t}{U}{_i}  + \left( \frac{D_{ii}}{2 \Delta t} -\frac{M_{ii}}{\Delta t^2} \right) \leftidx{^{t-\Delta t}}{U}{_i} }{ \frac{D_{ii}}{2 \Delta t} +\frac{M_{ii}}{\Delta t^2} }
+\end{equation}
+Let us define the following vectors:
+\begin{align}
+A_i &= \dfrac{1}{ \frac{D_{ii}}{2 \Delta t} +\frac{M_{ii}}{\Delta t^2} } \\
+B_i &= \dfrac{ \frac{2 M_{ii}}{\Delta t^2} }{ \frac{D_{ii}}{2 \Delta t} +\frac{M_{ii}}{\Delta t^2} } = \dfrac{2 M_{ii}}{\Delta t^2} A_i \\
+C_i &= \dfrac{ \frac{D_{ii}}{2 \Delta t} -\frac{M_{ii}}{\Delta t^2} }{ \frac{D_{ii}}{2 \Delta t} +\frac{M_{ii}}{\Delta t^2} } = \dfrac{D_{ii}}{2\Delta t} A_i - \dfrac{B_i}{2},
+\end{align}
+Equation \eqref{chap5:CDMcomponents} may be rewritten as:
+\begin{equation}
+\leftidx{^{t+\Delta t}}{ \mathbf{U} }{_i} = A_i \, (\leftidx{^t}{R}{_i} - \leftidx{^t}{F}{_i}) + B_i \, \leftidx{^t}{U}{_i} + C_i \, \leftidx{^{t-\Delta t}}{U}{_i}.
+\end{equation}
+It is worth noting that the coefficient vectors $ \mathbf{A} $, $ \mathbf{B} $ and $ \mathbf{C} $ can be precomputed.
+
+\section{Anisotropic and viscoelastic constitutive equation \OC{MICCAI 2008; MedIA 2009}}
+
+%\section{Summary of the procedure (algo type of thing)}

### 2010-08-26 Thesis/Chapter5.tex
- Update 26 August -
End chapter 5 + start of chapter 6 + bib updated + PhD comics added for later use :)



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@166 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{shortAbstract}
-A short abstract for the upcoming chapter
+TLED proposed by Miller almost real-time \\
+GPU implementation to reach real-time \\
+But first: TLED description \\
+\OC{MICCAI 2008; MedIA 2009}
 \end{shortAbstract}
@@
 \begin{equation}
-\mathbf{K(\mathbf{U}).\mathbf{U}} = \mathbf{F}(\mathbf{U}) = \sum_e \mathbf{\tilde{F}}^e,
+\mathbf{K(\mathbf{U}) \cdot \mathbf{U}} = \mathbf{F}(\mathbf{U}) = \sum_e \mathbf{\tilde{F}}^e,
 \end{equation}
@@
 This implies that the stiffness matrix does not need to be assembled since element nodal force contributions can be computed at the element level instead, which is a decisive computational advantage.
+
+We will first explain how to compute element nodal forces based on the deformation gradient and a neo-Hookean constitutive model.
+We will then show how to use these element nodal forces to calculate the unknown displacements for each node of the model \citep{Taylor06}.
 
@@
 \begin{equation}
+\label{chap5:strainEnergyFunction}
 _0^t W = \dfrac{1}{2} \mu (_0^t I_1 - 3 - 2 \ln ^t J) + \dfrac{1}{2} \lambda (^t J - 1)^2,
@@
 Conversely, because geometrically non-linear analyses (such as carried out by the TLED algorithm) take the change of geometry of the elements into account, the strain-displacement matrix $ \mathbf{B}_L = _0^t\mathbf{B}_L $ varies.
-However, the strain-displacement matrix at time $ t $ may be computed by transforming a stationary matrix $ _0^t\mathbf{B}_{L0} $ using the deformation gradient.
-We begin by defining the stationary matrix as:
+However, the strain-displacement matrix at time $ t $ may be computed by transforming a linear matrix $ _0^t\mathbf{B}_{L0} $ using the deformation gradient.
+We begin by defining the linear matrix as:
 \begin{equation}
@@
 \end{equation} 
-We note that the stationary strain-displacement matrix is composed of derivatives of shape functions with respect to the original undeformed coordinates of the body.
+We note that the linear strain-displacement matrix is composed of derivatives of shape functions with respect to the original undeformed coordinates of the body.
 These derivatives are therefore constant and may be precomputed, affording the total Lagrangian formulation a significant computational advantage.
@@
 
-\section{Anisotropic and viscoelastic constitutive equation \OC{MICCAI 2008; MedIA 2009}}
+\section{Anisotropic and viscoelastic constitutive equations}
 
-%\section{Summary of the procedure (algo type of thing)}
+	\subsection{An extension to the TLED algorithm}	
+An attractive feature of explicit analyses is the relative ease with which arbitrarily complicated constitutive models may be incorporated.
+This arises from the fact that element stresses are computed directly from strains in the course of the procedure.
+Additionally, there is no requirement for computation of tangent matrices as in implicit or quasi-static procedures, since there is no involvement of Newtonian iterations.
+While the developments above incorporated a hyperelastic constitutive formulation (neo-Hookean), thus accommodating nonlinearity of the strain-related tissue stress response, the formulation was intentionally simple nonetheless since the focus was on validation of the computational framework.
+Two further key features of the response of most biological tissues are time- (and rate-)dependence and anisotropy \citep{Fung93}.
+
+Time-dependence manifests itself in many aspects of the mechanical response.
+Soft tissues under constant load generally exhibit creep, while those under constant deformation exhibit stress relaxation.
+Additionally, most tissues appear stiffer at higher loading velocities.
+In particular, so-called visco-hyperelastic models based on strain energy functions with time-dependent parameters have been shown to reproduce both the time-dependent and large strain aspects of the response \citep{Miller97,Miller00,Nava08}.
+
+Anisotropic mechanical response may arise, for example, from the presence of a highly organised microstructure such as those of connective tissues.
+These are predominantly composed of collagen or elastin fibres embedded in an amorphous matrix \citep{Fung93}, and may be considered as fibre reinforced composites in some cases.
+Alternatively, the presence of vasculature and other functional components means even non-load bearing organs may exhibit directional dependence \citep{Picinbono01,Prange02}.
+Whereas isotropic constitutive models may be formulated in terms of the usual principal strain invariants (as used in \eqref{chap5:strainEnergyFunction} for a neo-Hookean material), directional dependence requires inclusion of so-called pseudo-invariants of strain and material direction.
+
+In order to model such soft-tissue features with the explicit TLED formulation without significant performance penalties, an efficient constitutive update procedure involving time integration of the relevant hereditary equations is required.
+We address this problem by presenting a procedure similar to that developed by \cite{Poon98} for analysis of anisotropic linear viscoelastic models.
+This was also adapted by \cite{Taylor07} for solution of their fibre composite-based microstructural model.
+In these cases the rate independent responses were based on linear elasticity, rather than a hyperelastic formulation suitable for large deformations.
+In the present work we begin with a class of anisotropic visco-hyperelastic models, and develop a constitutive update procedure for explicit analyses based on these.
+
+	\subsection{Visco-hyperelasticity}
+Large recoverable deformations and the time- and rate-dependence of the mechanical response of soft-tissues have led to the formulation of visco-hyperelastic constitutive models in which an underlying hyperelastic formulation is augmented by time dependent (viscoelastic) material parameters.
+Models of this type are well known in the continuum mechanics community, and provide a kinematically consistent basis for modelling non-linear materials at large deformations.
+
+For such materials the constitutive response is defined in terms of a time-dependent Helmholtz free energy (strain energy) function $ \hat{\Psi} $, expressed in the form of a convolution integral:
+\begin{equation}
+\hat{\Psi} (\Psi ,t) = \int_0^t \alpha (t-s) \pd{\Psi}{s} ds,
+\end{equation}
+where $ t $ is time and $ \Psi $ is the underlying hyperelastic strain energy function. The relaxation functions $ \alpha (t) $ commonly assume the form of a Prony series:
+\begin{equation}
+\label{chap5:PronySeries}
+\alpha (t) = \alpha_{\infty} + \sum_{i=1}^N \alpha_i e^{-t/\tau_i},
+\end{equation}
+where $ \alpha_{\infty} $, $ \alpha_i $ and $ \tau_i $are positive real constants.
+Such forms for the relaxation functions have a physical interpretation, namely that of a generalised Maxwell model \citep{Holzapfel96}.
+If we impose the condition
+\begin{equation}
+\left( \alpha_{\infty} + \sum_{i=1}^N \alpha_i \right) = 1
+\end{equation}
+we may rewrite\eqref{chap5:PronySeries} as
+\begin{equation}
+\label{chap5:PronySeries2}
+\alpha (t) = 1 - \sum_{i=1}^N \alpha_i (1-e^{-t/\tau_i}),
+\end{equation}
+which will be of use in subsequent sections.
+
+\bigskip
+
+\noindent The required stress $ \mathbf{S} $ may be obtained via differentiation with respect to strain:
+\begin{equation}
+\label{chap5:stress}
+\mathbf{S} = 2 \pd{\hat{\Psi}(\Psi , t)}{\mathbf{C}} = \int_0^t \alpha (t-s) \left( 2 \pd{}{s} \pd{\Psi}{\mathbf{C}} \right) ds = \int_0^t \alpha (t-s) \pd{\boldsymbol \Phi}{s} ds,
+\end{equation}
+where $ \mathbf{C} $ is the right Cauchy-Green deformation tensor, and we have introduced $ \boldsymbol \Phi \, {\buildrel\rm	def\over=} \, 2 \partial \Psi / \partial \mathbf{C} $ as the instantaneous hyperelastic stress response.
+
+\bigskip
+
+Models of this form have been presented by Miller and co-workers for analysis of brain tissue \citep{Miller97,Miller02} and of liver and kidney \citep{Miller00}.
+They were shown to model the tissue responses to compressive (and in the case of brain, tensile) loading at strain rates varying over two orders of magnitude very well.
+We next consider evaluation of the hyperelastic stress response $ \boldsymbol \Phi $.
+
+	\subsection{Hyperelastic response}
+We firstly consider the case of isotropic materials, from which anisotropic formulations follow.
+We then consider the standard cases of transverse isotropy and orthotropy, which may be viewed as arising from, for example, the presence of mutually orthogonal reinforcing fibre phases.
+In considering these cases we encompass constitutive equations which have been proposed for a wide variety of biological tissues.
+
+	\subsubsection*{Isotropic materials}
+For isotropic materials the strain energy is a function of strain only, hence $ \Psi = \Psi (\mathbf{C}) $.
+In this work we consider strain energy functions with separated isochoric (volume preserving) and volumetric components \citep{Holzapfel00}:
+\begin{equation}
+\label{chap5:IstrainEnergy}
+\Psi (\mathbf{C}) = \Psi ^{iso} (\mathbf{\bar{C}}) + \Psi ^{vol} (J) = \Psi ^{iso} (\bar{I}_1, \bar{I}_2) + \Psi ^{vol} (J),
+\end{equation}
+where $ J $ is the Jacobian determinant, $ \mathbf{\bar{C}} = J^{2/3} \mathbf{C} $ is the modified right Cauchy-Green deformation, and $ \bar{I}_1 = tr \mathbf{\bar{C}} $ and $ \bar{I}_2 = \left[ (tr \mathbf{\bar{C}})^2 - tr (\mathbf{\bar{C}}^2) \right] / 2 $ are invariants of $ \mathbf{\bar{C}} $.
+The hyperelastic stress $ \boldsymbol \Phi $ then also consists of isochoric and volumetric components:
+\begin{equation}
+\boldsymbol \Phi = \boldsymbol \Phi^{iso} +\boldsymbol \Phi^{vol} ,
+\end{equation}
+where
+\begin{align}
+\boldsymbol \Phi^{vol} &= 2 \pd{\Psi^{vol}(J)}{\mathbf{C}} = J \dfrac{d\Psi^{vol}(J)}{dJ} \mathbf{C}^{-1}, \label{chap5:Phi_vol} \\
+\intertext{and}
+\boldsymbol \Phi^{iso} &= 2 \pd{\Psi^{iso}(\mathbf{\bar{C}})}{\mathbf{C}} = J^{-2/3} \mbox{Dev}  \bar{\boldsymbol \Phi}, \label{chap5:Phi_iso}
+\end{align}
+where Dev$(\bullet) = (\bullet) - (1/3)[(\bullet) : \mathbf{C}] \mathbf{C}^{-1}$ is the referential configuration deviatoric operator for a second order tensor \citep{Holzapfel00}, and
+\begin{equation}
+\label{chap5:PhiBar}
+\bar{\boldsymbol \Phi} = 2 \pd{\Psi^{iso}(\mathbf{\bar{C}})}{\mathbf{\bar{C}}} = \bar{\gamma}_1 \mathbf{I} + \bar{\gamma}_2 \mathbf{\bar{C}},
+\end{equation}
+with 
+\begin{equation}
+\bar{\gamma}_1 = 2 \left( \pd{\Psi^{iso}}{\bar{I}_1} + \bar{I}_1 \pd{\Psi^{iso}}{\bar{I}_2} \right) \quad \mbox{and} \quad \bar{\gamma}_2 = -2 \pd{\Psi^{iso}}{\bar{I}_2}.
+\end{equation}
+
+	\subsubsection*{Transversely isotropic materials}
+Transversely isotropic materials are characterised by a single preferred direction $\mathbf{a}_0$ in the reference configuration; the mechanical response is isotropic in the plane orthogonal to this direction.
+The strain energy is then a function of both $  \mathbf{C} $ and a structure tensor $\mathbf{A}_0 \, {\buildrel\rm	def\over=} \,  \mathbf{a}_0 \otimes \mathbf{a}_0$, where $\otimes$ denotes a tensor product.
+Analogous to the isotropic case, we consider strain energy functions of the form
+\begin{equation}
+\label{chap5:TIstrainEnergy}
+\Psi (\mathbf{C}, \mathbf{A}_0) = \Psi^{iso}(\bar{I}_1, \bar{I}_2, \bar{I}_4, \bar{I}_5) + \Psi^{vol}(J),
+\end{equation}
+where $ \bar{I}_4 = \mathbf{a}_0 \cdot \mathbf{\bar{C}} \mathbf{a}_0$ and $ \bar{I}_5 = \mathbf{a}_0 \cdot \mathbf{\bar{C}}^2 \mathbf{a}_0 $ are pseudo-invariants of $ \mathbf{\bar{C}} $ and $ \mathbf{A}_0 $.
+
+\bigskip
+
+\noindent As can be seen transversely isotropic strain energy functions \eqref{chap5:TIstrainEnergy} differ from isotropic ones \eqref{chap5:IstrainEnergy} only in the form of the isochoric term.
+Therefore the volumetric and isochoric stresses $ \boldsymbol \Phi^{vol} $ and $ \boldsymbol \Phi^{iso} $ remain as in \eqref{chap5:Phi_vol} and \eqref{chap5:Phi_iso} but with $\bar{\boldsymbol \Phi} $ given by
+\begin{equation}
+\label{chap5:PhiBarTI}
+\bar{\boldsymbol \Phi} = \bar{\gamma}_1 \mathbf{I} + \bar{\gamma}_2 \mathbf{\bar{C}} + \bar{\gamma}_4 \mathbf{A}_0 + \bar{\gamma}_5 ( \mathbf{a}_0 \otimes \mathbf{\bar{C}} \mathbf{a}_0 +  \mathbf{\bar{C}} \mathbf{a}_0 \otimes \mathbf{a}_0 ),
+\end{equation}
+with 
+\begin{equation}
+\bar{\gamma}_a = 2 \pd{\Psi^{iso}}{\bar{I}_a}, \quad a = 4, 5.
+\end{equation}
+
+	\subsubsection*{Orthotropic materials}
+Orthotropic materials are characterised by three mutually orthogonal preferred directions, which we identify with unit vectors $\mathbf{a}_0$ and $\mathbf{b}_0$ (and corresponding structure tensors $\mathbf{A}_0$ and $\mathbf{B}_0$), in the reference configuration.
+We need specify only two vectors, since the direction orthogonal to these naturally emerges as a preferred direction also.
+Orthotropic strain energy functions are then of the form
+\begin{equation}
+\Psi (\mathbf{C}, \mathbf{A}_0, \mathbf{B}_0) = \Psi^{iso}(\bar{I}_1, \bar{I}_2, \bar{I}_4, \bar{I}_5, \bar{I}_6, \bar{I}_7) + \Psi^{vol}(J),
+\end{equation}
+where $ \bar{I}_6 = \mathbf{b}_0 \cdot \mathbf{\bar{C}} \mathbf{b}_0 $ and $ \bar{I}_7 =  \mathbf{b}_0 \cdot \mathbf{\bar{C}}^2 \mathbf{b}_0 $ are pseudo-invariants of $ \mathbf{\bar{C}} $ and $ \mathbf{B}_0 $.
+In a similar manner to the transversely isotropic case we obtain $ \boldsymbol \Phi^{vol} $ and $ \boldsymbol \Phi^{iso} $ from \eqref{chap5:Phi_vol} and \eqref{chap5:Phi_iso}, with $\bar{\boldsymbol \Phi} $ now given by
+\begin{equation}
+\label{chap5:PhiBarOrtho}
+\begin{split}
+\bar{\boldsymbol \Phi} = \bar{\gamma}_1 \mathbf{I} + \bar{\gamma}_2 \mathbf{\bar{C}} + \bar{\gamma}_4 \mathbf{A}_0 + \bar{\gamma}_5 ( \mathbf{a}_0 \otimes \mathbf{\bar{C}} \mathbf{a}_0 +  \mathbf{\bar{C}} \mathbf{a}_0 \otimes \mathbf{a}_0 ) \\
++ \bar{\gamma}_6 \mathbf{B}_0 + \bar{\gamma}_7 ( \mathbf{b}_0 \otimes \mathbf{\bar{C}} \mathbf{b}_0 +  \mathbf{\bar{C}} \mathbf{b}_0 \otimes \mathbf{b}_0 )
+\end{split}
+\end{equation}
+where
+\begin{equation}
+\bar{\gamma}_a = 2 \pd{\Psi^{iso}}{\bar{I}_a}, \quad a = 6, 7.
+\end{equation}
+
+	\subsection{Recapitulation}	
+For a visco-hyperelastic material stress may be obtained from \eqref{chap5:stress}.
+This form is general in the sense that any underlying hyperelastic response may be used, including the anisotropic formulations described \eqref{chap5:IstrainEnergy}, supplemented with \eqref{chap5:Phi_vol}, \eqref{chap5:Phi_iso}, and \eqref{chap5:PhiBar} is the general form of an isotropic hyperelastic stress response, defined in terms of invariants.
+Transversely isotropic or orthotropic models may be produced by substituting \eqref{chap5:PhiBar} for \eqref{chap5:PhiBarTI} or \eqref{chap5:PhiBarOrtho}, respectively.
+The specification of particular forms of  $ \Psi^{vol} $ and $ \Psi^{iso} $ would be motivated by the particular tissue/material under analysis, and may stem from phenomenological or microstructural considerations.
+Finally we note that for separated isochoric and volumetric hyperelastic functions as used here, viscoelastic terms may be applied to either or both independently.
+
+\section{Constitutive update procedure for explicit analyses}
+
+Use of the above visco-hyperelastic models within the TLED algorithm (or any other explicit dynamic finite element procedure) requires a constitutive update scheme involving time integration of \eqref{chap5:stress}.
+
+	\subsection{Stress update equations}	
+We proceed by restating \eqref{chap5:stress} and using \eqref{chap5:PronySeries2}:	
+\begin{equation}
+\mathbf{S} = \int_0^t \left[ 1 - \sum_{i=1}^N \alpha_i (1-e^{(s-t)/\tau_i}) \right] \pd{\boldsymbol \Phi}{s} ds .
+\end{equation}
+This may be separated into rate-dependent and -independent terms as
+\begin{equation}
+\mathbf{S} = \boldsymbol \Phi - \sum_{i=1}^N \boldsymbol \Upsilon_i,
+\end{equation}
+where
+\begin{equation}
+\label{chap5:Upsilon}
+\boldsymbol \Upsilon_i = \int_0^t \alpha_i \left( 1-e^{(s-t)/\tau_i} \right) \pd{\boldsymbol \Phi}{s} ds, \quad i \in [1, N]
+\end{equation}
+are rate-dependent terms associated with each term in the Prony series.
+
+\bigskip
+
+\noindent In an incremental analysis we require the stress at the current increment given the deformation state and history of the material.
+Adding superscripts to indicate time increments the stress may be updated using
+\begin{equation}
+\label{chap5:updateStress}
+\mathbf{S}^n = \boldsymbol \Phi^n - \sum_{i=1}^N \boldsymbol \Upsilon_i^n. 
+\end{equation}
+The instantaneous terms $ \boldsymbol \Phi^n $ may be computed directly from the (known) current deformation $ \mathbf{C}^n $.
+The main difficulty is then computation of the in- cremental rate-dependent terms $ \boldsymbol \Upsilon_i^n $ .
+Following \cite{Poon98}, our strategy is to maintain each $ \boldsymbol \Upsilon_i^n $ as a separate state variable to be updated at each increment also.
+
+	\subsection{State variable update equations}	
+Our approach is to convert the integral equation \eqref{chap5:Upsilon} into a rate form which may then be numerically integrated to produce an incremental update formula for $ \boldsymbol \Upsilon_i^n $.
+We note that \eqref{chap5:Upsilon} is of the form
+\begin{equation}
+y(t) = \int_0^t f(t,s) ds.
+\end{equation}
+\cite{Poon98} provide the following formula for differentiating such equations with respect to $ t $:
+\begin{equation}
+\dot y = f(t, t) + \int_0^t \dot f(t,s) ds.
+\end{equation}
+Applying this formula to \eqref{chap5:Upsilon} we obtain
+\begin{align}
+f(t, t) &= \alpha_i \left( 1-e^{(t-t)/\tau_i} \right) \pd{\boldsymbol \Phi}{s} \notag \\
+&= 0 \\
+\intertext{and}
+\dot f(t, s) &= \dfrac{d}{dt} \left[ \alpha_i \left( 1-e^{(s-t)/\tau_i} \right) \pd{\boldsymbol \Phi}{s} \right] \notag \\
+&= \alpha_i e^{(s-t)/\tau_i} \pd{\boldsymbol \Phi}{s} \dfrac{1}{\tau_i},
+\end{align}
+leading to the required rate form for $ \boldsymbol \Upsilon_i $:
+\begin{align}
+\dot{\boldsymbol \Upsilon_i}  &= \dfrac{1}{\tau_i} \int_0^t \alpha_i e^{(s-t)/\tau_i} \pd{\boldsymbol \Phi}{s} ds \notag \\
+&= \dfrac{1}{\tau_i} (\alpha_i \boldsymbol \Phi - \boldsymbol \Upsilon_i). \label{chap5:UpsilonRate}
+\end{align}
+
+\bigskip
+
+\noindent Equation \eqref{chap5:UpsilonRate} may be integrated using a convenient numerical method.
+In particular the unconditionally stable backward Euler method suggested by \cite{Poon98} yields the following formula for $ \boldsymbol \Upsilon_i^n $:
+\begin{align}
+\boldsymbol \Upsilon_i^n &= \left( \dfrac{\alpha_i \boldsymbol \Phi^n}{\tau_i} + \dfrac{\boldsymbol \Upsilon_i^{n-1}}{\Delta t} \right) / \left(\dfrac{1}{\Delta t} + \dfrac{1}{\tau_i} \right) \notag \\
+&= A \boldsymbol \Phi^n + B \boldsymbol \Upsilon_i^{n-1}, \label{chap5:updateStateVar}
+\end{align}
+where $ \Delta t $ is the time step size, and $ A = \Delta t \alpha_i / (\Delta t + \tau_i) $ and $ B = \tau_i / (\Delta t + \tau_i) $ are constant coefficients.
+
+	\subsection{Summary}	
+The constitutive update procedure consists of
+\begin{enumerate}
+\item Updating state variables (one for each Prony term) via \eqref{chap5:updateStateVar}.
+\item Updating stresses via \eqref{chap5:updateStress}.
+\end{enumerate}
+Assuming a constant time step size $ \Delta t $, the coefficients $ A $ and $ B $ may be precomputed.

### 2010-08-26 Thesis/Chapter6.tex
- Update 25 August -
Started chapter5 (TLED description) + update of labels + package left indices + bib updated



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@165 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 26 August -
End chapter 5 + start of chapter 6 + bib updated + PhD comics added for later use :)



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@166 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\section{What is GPGPU}
+\section{Summary of the TLED formulation}
+A complete description of the TLED algorithm for soft tissue simulation was given in the previous chapter.
+However, let us remind the main steps of the algorithm before tackling the implementation.
+Briefly, the algorithm consists of a precomputation phase in which element shape function derivatives $ \partial \mathbf{h} $ (and other quantities) and the system mass matrix $ \mathbf{M} $ are calculated, followed by a time-loop in which incremental solutions for the node displacements $ \mathbf{U} $ are found.
+During each step of the time-loop we:
+\begin{enumerate}
+\item Apply loads (displacements and/or forces) and boundary conditions to relevant nodal degrees of freedom
+\item For each element compute
+\begin{description}
+\item[(a)] deformation gradient $ \mathbf{X} $ and right Cauchy-Green deformation tensor $ \mathbf{C} $
+\item[(b)] linear strain-displacement matrix $ \mathbf{B}_L $
+\item[(c)] second Piola-Kirchhoff stress $ \mathbf{S} $
+\item[(d)] element nodal forces $ \mathbf{\tilde{F}} $, and add these forces to the total nodal forces $ \mathbf{F} $
+\end{description}
+\item For each node compute new displacements $ \mathbf{U} $ using the central difference method.
+\end{enumerate}
 
-\section{Re-formulation of the algorithm for its Cg implementation}
+\bigskip
 
-\section{CUDA implementation/optimisations \OC{ISBMS 2008a}}
+\noindent The nodal force contributions $ \mathbf{\tilde{F}} $ from each element are obtained 
+\begin{equation}
+\mathbf{\tilde{F}} = \int_{^0 V} \, \mathbf{B}_L^t \, \mathbf{\hat{S}} \, d\leftidx{^0}{ V }{},
+\end{equation}
+where $ ^0 V $ is the initial volume of the element and $ \mathbf{\hat{S}} $ is the vector form of the stress tensor $ \mathbf{S} $.
+This integral is generally evaluated numerically, for ex- ample using Gaussian quadrature.
+For reduced integration 8-node hexahedral elements we obtain
+\begin{equation}
+\mathbf{\tilde{F}} = 8 \, \mathbf{B}_L^t \, \mathbf{\hat{S}} \, J
+\end{equation}
+where $ J $ is the Jacobian determinant. For 4-node tetrahedral elements we obtain
+\begin{equation}
+\mathbf{\tilde{F}} = V \, \mathbf{B}_L^t \, \mathbf{\hat{S}}.
+\end{equation}
+The above equations make no assumption concerning the constitutive model employed.
+The deformation state $ \mathbf{F} $ in each element is known, allowing stresses $ \mathbf{S} $ to be computed from any valid constitutive equation.
+
+\section{General-purpose computation on graphics processing units}
+
+	\subsection{Goal and motivation}
+Graphics processing units (GPU) functionality has, traditionally, been very limited.
+In fact, for many years the GPU was only used to accelerate certain parts of the graphics pipeline.
+A GPU is essentially a special purpose hardware designed to accelerate each stage of the geometry pipeline, the process of matching image data or a computer model to the pixels on the computer's screen.
+Initially, GPU could only run two kinds of program: vertex and pixel shaders.
+Vertex shaders are run once for each vertex given to the graphics processor.
+The purpose is to transform each vertex's 3D position in virtual space to the 2D coordinate at which it appears on the screen.
+Vertex shaders can manipulate properties such as position, color, texture coordinate and normal vector.
+Pixel shaders are functions that compute color and other attributes of each pixel.
+They range from always outputting the same color, to applying a lighting value, to doing shadows, specular highlights or translucency for instance.
+
+Eventually, vertex and pixel shaders became programmable to enable game programmers to generate even more realistic effects.
+Programmable pixel shaders allow the programmer to substitute, for example, a lighting model other than those provided by default by the graphics card.
+Shaders have enabled graphics programmers to create lens effects, displacement mapping, and depth of field.
+This evolution of GPU's hardware and the increasing programmable capability naturally lead to use GPUs for non-graphics applications.
+The term GPGPU (General-purpose computation on graphics processing units) was coined by Mark Harris in 2002 when he recognised an early trend of using GPUs for non-graphics applications.
+However, capabilities of GPU's were still fairly limited for non-graphics applications.
+
+Things dramatically changed in 2007 when the two types of shaders were unified.
+While early shader models used very different instruction sets for vertex and pixel shaders, unified shader models have almost the same capabilities.
+An unified shader architecture allows more flexible use of the graphics rendering hardware.
+The computing units of the GPU can run vertex or pixel shaders according to work loads.
+Along with unified shader models, a new type of shader was created: geometry shaders.
+They are executed after vertex shaders and can generate new graphics primitives, such as points, lines and triangles.
+
+\bigskip
+
+GPUs may be seen as high-performance many-core processors that can be used to accelerate a wide range of applications.
+An example (given by Sanford Russell from NVIDIA) to think about to illustrate the difference between a traditional CPU and a GPU is this: if you were looking for a word in a book, and handed the task to a CPU, it would start at page $ 1 $ and read it all the way to the end, because it's a serial processor.
+It would be fast, but would take time because it has to go in order.
+A GPU, which is a parallel processor, would tear the book into a thousand pieces and read it all at the same time.
+Even if each individual word is read more slowly, the book may be read in its entirety quicker, because words are read simultaneously.
+
+In addition to an increasing programmability, it is easy to understand why the development of GPGPU is soaring.
+GPU is now used in imaging, finance, signal processing, simulation, video and many other fields.
+A recent review of GPGPU algorithms may be found in \cite{Owens07}.
+This is therefore not surprising that research has been carried out to leverage the power of GPUs to solve partial differential equations of continuum mechanics via the computationally expensive finite element method.
+Indeed, we have already insisted on the strong time constraints demanded by the field of medical simulation and the path towards GPGPU is natural.
+A non-exhaustive review of algorithms implemented on GPU in the context of medical simulation was given chapter \ref{chap4} (page~\pageref{chap4:GPUMedicalSimulation}).
+
+Before introducing our own GPU implementation of the TLED formulation, we will discuss the different GPU programming languages at our disposal.
+
+	\subsection{Programming languages for GPUs}
+
+		\subsubsection*{Graphics API}
+
+		\subsubsection*{Non-graphics API}
 
@@
 	\subsection{Presentation of SOFA project and architecture}
-	\subsection{TLED in SOFA and release in open-source}
+	\subsection{CUDA description (why CUDA over Cg?)}	
+	\subsection{GPU implementation of TLED in SOFA and release in open-source \OC{ISBMS 2008a}}
+
+\section{Results}	
+	\subsection{Pure shear of a cube}
+	\subsection{Compression of a cube}
+	\subsection{GPU performance}
+	\subsection{Simulation of liver deformation}
+	\subsection{A medical application: cataract surgery}
+
+\section{Discussion}
+	\subsection{Critical time step}
+	\subsection{Difficulties to handle contacts}

### 2010-08-27 Thesis/Chapter6.tex
- Update 27 August -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@167 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{shortAbstract}
-A short abstract for the upcoming chapter
+\OC{ISBMS 2008a}
 \end{shortAbstract}
@@
 Programmable pixel shaders allow the programmer to substitute, for example, a lighting model other than those provided by default by the graphics card.
-Shaders have enabled graphics programmers to create lens effects, displacement mapping, and depth of field.
+Shaders have enabled graphics programmers to create lens effects, displacement mapping and depth of field.
 This evolution of GPU's hardware and the increasing programmable capability naturally lead to use GPUs for non-graphics applications.
 The term GPGPU (General-purpose computation on graphics processing units) was coined by Mark Harris in 2002 when he recognised an early trend of using GPUs for non-graphics applications.
-However, capabilities of GPU's were still fairly limited for non-graphics applications.
+However, capabilities of GPU's were still fairly limited for non-graphics applications at this time.
 
@@
 GPUs may be seen as high-performance many-core processors that can be used to accelerate a wide range of applications.
-An example (given by Sanford Russell from NVIDIA) to think about to illustrate the difference between a traditional CPU and a GPU is this: if you were looking for a word in a book, and handed the task to a CPU, it would start at page $ 1 $ and read it all the way to the end, because it's a serial processor.
+An example (given by Sanford Russell from NVIDIA) to think about to illustrate the difference between a traditional CPU and a GPU is this: if you were looking for a word in a book, and handed the task to a CPU, it would start at page 1 and read it all the way to the end, because it's a serial processor.
 It would be fast, but would take time because it has to go in order.
@@
 In addition to an increasing programmability, it is easy to understand why the development of GPGPU is soaring.
-GPU is now used in imaging, finance, signal processing, simulation, video and many other fields.
-A recent review of GPGPU algorithms may be found in \cite{Owens07}.
+GPU is now used in imaging, finance, signal processing, simulation, audio and video processing, astronomy, weather forecasting, molecular modelling, cryptography, quantum mechanics and many other fields.
+A fairly recent review of GPGPU algorithms may be found in \cite{Owens07}.
 This is therefore not surprising that research has been carried out to leverage the power of GPUs to solve partial differential equations of continuum mechanics via the computationally expensive finite element method.
@@
 
+Interface with a GPU has traditionally been via a graphic application programming interface (API) such as \citetalias{Board05} or \citetalias{DirectX10}.
+OpenGL is an open standard API that provides a number of functions for the rendering of 2D and 3D graphics and is available on most modern operating systems including but not limited to Windows, Mac OS X and Linux.
+It was initially designed by Silicon Graphics Inc.
+and now managed by the technology consortium Khronos group.
+DirectX is a proprietary API that provides functions to render three dimensional graphics, and uses hardware acceleration if it is available on the graphics card.
+It was designed by Microsoft for use on the Windows platform.
+
+Because OpenGL and DirectX are both low-level APIs and include many irrelevant functionality from a GPGPU perspective, higher level languages were designed to give developers more direct control of the graphics pipeline without having to use assembly language or hardware-specific languages.
+
 		\subsubsection*{Graphics API}
+OpenGL (version 1.5 and newer) provides shader language based on the C programming language called OpenGL shading language (GLSL).
+In the DirectX API (DirectX 9 and newer), shaders are programmed with the high level shader language (HLSL).
+Cg is another high-level shading language developed by NVIDIA in close collaboration with Microsoft and it is very similar to Microsoft's HLSL.
+However, the Cg compiler is not specific to a graphics API and can output both OpenGL and DirectX shader programs.
+
+These languages all share basic data types (float, integer, bool etc.) and also feature vector and matrix data types that are based on the basic data types (like float3 or float4x4 in Cg and vec3 or mat4 in GLSL) and standard library functions are provided (vector and matrix multiplication, cross product etc.).
+In addition, they support loops and branching like if/else, while, and for.
+However, because these languages are just an abstraction on the top of graphics API, there are significant restrictions on what they can do.
+Specifically, shader programs are unable to determine the destination of their outputs.
+They merely operate on fragment data (color, texture coordinates, normal etc.) and return the modified fragment at the same location.
+Each shader is also executed independently of its neighbours since no communication is allowed between processors.
+Note that while this was made possible on recent hardware, these graphics API do not take advantage of this feature and therefore this statement remains valid.
+
+Memory storage is allowed through the use of textures, which essentially are structured collections of floating point values arranged as arrays.
+Shaders can read textyres from any location and as many time as they like.
+As such it is a concept very similar to regular arrays.
+However, the individual components of textures consist of up to four floating point values in the form of a vector.
+This comes from the fact that graphics processing units were initially designed to manipulate color images, so each component of a texture is in fact a vector holding the red, green, blue and alpha (transparency) values for the given pixel.
+Textures are therefore optimised for retrieving up to four floating point values in a single read.
+
+When it comes to writing to GPU memory though, things are a bit more limited.
+Indeed, in contrast to texture reads, shaders can only write once into textures (at the end of the set of instructions).
+Morevoer, as stated earlier, shaders can only write in predefined locations corresponding to their potential screen location.
+Therefore, graphics API does not allow writing to random memory locations (this feature is called \emph{scattering}).
+Nethertheless, shaders can have multiple render targets.
+The write locations within each texture are still fixed though.
 
 		\subsubsection*{Non-graphics API}
+Eventually, non-graphics API were released to support the development of GPGPU.
+The two major graphics card manufacturers first released their own API.
+While the objective of hiding graphics aspects is the same, the approaches of NVIDIA and ATI (now owned by AMD) are quite different.
+
+In November 2006, NVIDIA released CUDA which consists of C language extensions.
+CUDA is fairly high level.
+Programs (called kernels) may be coded as C functions and invoked directly without going through a render pass as with graphics APIs.
+When called, kernels are executed N times in parallel by N threads as opposed to only once like regular C functions.
+Memory management functions analogous to standard C are also provided (memcpy, free etc.).
+
+At the same time, ATI released CTM released a software development layer aimed for low-level access (assembly-style language).
+Memory management functions similar to graphics API are also provided.
+A year later, in December 2007, ATI released Stream SDK and added a new high-level language derived from C, called ATI Brook+ which later evolved to ATI CAL (Compute Abstraction Layer).
+In both cases, these APIs work solely on each manufacturer's devices, CUDA with NVIDIA cards and Stream SDK on ATI cards.
+
+OpenCL is a more general framework for writing programs that execute across heterogeneous platforms consisting of CPUs, GPUs, and other processors.
+It is not specific to a single type of hardware.
+OpenCL was initially developed by Apple and refined into an initial proposal in collaboration AMD (ATI), IBM, Intel, and NVIDIA.
+Apple submitted this initial proposal to the Khronos Group and the first release occurred in December 2008.
+AMD decided to support OpenCL instead of the now deprecated CTM in its Stream SDK.
+
+The latest API to be released was DirectCompute from Microsoft.
+DirectCompute was released as part of DirectX11 in October 2009 and allows access to general-purpose computing with a fairly low level language.
+
+	\subsubsection*{The choice of API}	
+The first GPU implementation	of the TLED algorithm was carried out by \cite{Taylor07b,Taylor08}.
+They used the Cg language and had to reformulate the TLED algorithm presented to accomodate with the limited possibilities of this graphics API.
+Indeed, the major difference between this parallel implementation and a serial one is the writing of element nodal forces to memory at the end of a first kernel and their subsequent retrieval and summation during a second kernel.
+In a serial implementation nodal force contributions would most likely be added to a global sum as they are computed, rather than stored and summed in a second loop.
+But as stated earlier, this graphics API does not allow scattering, that is writing to random memory locations.
+Because of this limitation, element nodal forces cannot be directly added to the total nodal forces.
+Instead, the authors had to reformulate force summation as a gather operation in their implementation.
+They reported significant solution speed gains up to $ 16.8 \times$ over the CPU implementation.
+Thus, they were able to compute a simple cube model with up to $ 16\,000 $ tetrahedral elements in real-time.
+
+A major improvement was introduced with the release of CUDA.
+In theory, CUDA allows scattered writes and this new feature could dramatically change the GPU implementation of the TLED as the reformulation of the scatter as a gather would no longer be useful.
+A single kernel would be sufficient and an increase in performance is expected.
+Along with a more readable and simple code, this is the main feature that made us choose the CUDA API to reimplement the TLED algorithm.
+
+In addition, we decided to develop our CUDA-based re-implementation of the TLED algorithm within the international and open source framework SOFA.
+We will show that this integration has a very limited cost in terms of performance by comparing the SOFA version with a standalone implementation.
+By providing an efficient and accurate non-linear FEM for soft tissue modelling to worldwide researchers, we thus hope to assist in enhancing the realism of medical simulators.
 
 \section{Implementation into SOFA}
-	\subsection{Presentation of SOFA project and architecture}
-	\subsection{CUDA description (why CUDA over Cg?)}	
-	\subsection{GPU implementation of TLED in SOFA and release in open-source \OC{ISBMS 2008a}}
+
+	\subsection{SOFA, an open source simulation framework}
+
+\subsubsection*{Objectives}
+The multi-disciplinary aspect of medical simulation requires the integration within a single environment of solutions in areas as diverse as visualisation, biomechanical modelling, haptic feedback and contact modelling.
+Although their interaction is essential to design a realistic simulator, only few teams have the sufficient resources to build such frameworks.
+This diversity of problems creates challenges for researchers to advance specific areas, and leads rather often to duplication of effort.
+The open source SOFA framework \citep{Allard07} was created to overcome this issue by providing researchers with an advanced software architecture that facilitates the development of new algorithms and simulators.
+SOFA has been mostly developed by INRIA (the French national institute for research in computer science and control) and CIMIT (Center for Integration of Medicine and Innovative Technology) and is primarily targeted at real-time simulation with an emphasis on medical simulation.
+SOFA is highly modular and flexible: it allows independently developed algorithms to interact together within a common simulation while minimising the development time required for integration.
+The overall goal is to develop a flexible framework while minimising the impact of this flexibility on the computation overhead.
+To achieve these objectives, SOFA proposes a new architecture that implements a series of concepts described below.
+
+\subsubsection*{SOFA architecture}
+
+\paragraph{High-level modularity.}
+The SOFA architecture relies on the innovative notion of multi-model representation where an object is explicitly decomposed into various representations: Behaviour Model, Collision Model, Collision Response Model, Visual Model and Haptic Model.
+Each representation can then be optimised for a particular task (biomechanics,  collision detection, visualisation, haptics) while at the same time improving interoperability by creating a clear separation between the functional aspects of the simulation components.
+These representations are then connected together via a mechanism called \emph{mapping}.
+Various mapping functions can be defined, and each mapping associates a set of primitives of a representation to a set of primitives in the other representation (\fig{chap6:fig-SOFA}).
+For instance, a mapping can connect degrees of freedom in a Behaviour Model to vertices in a Visual Model.
+
+\begin{figure}
+\includegraphics[width=13.9cm]{chapter6/sofa.pdf}
+\caption [Multi-model representation in SOFA] {Multi-model representation in SOFA.
+\textbf{Left:} a Behaviour Model controls the other representations via a series of mappings.
+\textbf{Right:} examples of representations with a liver model.}
+\label{chap6:fig-SOFA}
+\end{figure}
+
+\paragraph{Finer level modularity.}
+In order to easily compare algorithms within SOFA, more flexibility was added to the Behaviour Model by introducing an even finer level of granularity.
+A series of generic primitives common to most physics-based simulations have been defined: DoF, Mass, Force Field and Solver.
+The DoF component describes the degrees of freedom, and their derivatives, of the object.
+The Mass component represents its mass.
+The Force Field describes both internal and external forces that can be applied to this object.
+The Solver component handles the time step integration, i.e.
+advancing the state of the system from time $t$ to time $t+\Delta t$.
+
+\paragraph{Scene-graph.}
+Finally, another key aspect of SOFA is the use of a scene-graph to organise and process the elements of a simulation.
+Each component is attached to a node of a tree structure.
+This simple structure makes it easy to visit all or a subset of the components in a scene, and dependencies between components are handled by retrieving sibling components attached to the same node.
+During the simulation loop, most computations can be expressed as a traversal of the scene-graph.
+For instance, at each time step, the simulation state is updated by processing all Solver components, which will then forward requests to the appropriate components by recursively sending actions within its sub-tree.
+
+These different functionalities and levels of abstraction allow the user to switch from one component to another by simply editing an XML file, without having to recompile.
+In particular this permits testing of different computational models of soft tissue deformation, and to assess the pros and cons of various algorithms within the same context.
+
+	\subsection{CUDA description}
+GPUs achieve a high floating point capacity by distributing computation across a high number of parallel execution threads.
+They perform optimally as single instruction, multiple data devices.
+CUDA is a relatively new C API for compatible NVIDIA GPUs.
+CUDA organises threads in two hierarchical levels: blocks, which are groups of threads executed on one of the GPU's multiprocessors, and grids, which are groups of blocks launched concurrently on the device, and which all execute the same kernel.
+Figure~\ref{chap6:fig-cuda} represents this thread organisation.
+% As an example, the NVIDIA 8800 GTX used for the results presented in section \ref{results_section} has 16 multiprocessors, each containing eight processors.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=10cm]{chapter6/cuda_architecture.pdf} 
+\caption[CUDA architecture] {Each kernel is executed by CUDA as a group of threads within a grid. Image courtesy of NVIDIA. }
+\label{chap6:fig-cuda}
+\end{center}
+\end{figure}
+
+CUDA allows developers to specify the number of threads per block in each execution (the so-called execution configuration), effectively defining the distribution of computational load across all processors.
+For a given kernel the block dimensions are chosen to optimise the utilisation of the available computational resources.
+Care should be taken at the multiprocessor level in balancing the available memory required by the kernels with the ability to hide global memory latency.
+Since a finite amount of memory is available on a multiprocessor, the memory requirements of a kernel will determine how many threads can run concurrently on each.
+Importantly, CUDA's use of time slicing allows more than one block to be executed concurrently on a single multiprocessor, which has important implications for hiding memory latency.
+If more than one block is executing, the multiprocessor is able to switch processing between blocks while others are stalled on memory accesses, whereas it has no option but to wait for these if only one block is executing.
+Therefore for memory bandwidth bound kernels it may be preferable to launch several smaller blocks on each multiprocessor rather than a single larger one if both configurations make the same use of multiprocessor memory resources.
+While tools are available from NVIDIA for estimating the optimal execution configuration, it has proved necessary to fine tune the configuration experimentally for each kernel.
+
+	\subsection{First implementation: scatter as a gather}
+
+\subsubsection*{SOFA integration.}
+Implementing a biomechanical model in SOFA translates essentially into writing a new Force Field, i.e.
+describing the algorithm used to compute internal forces in the model.
+It merely comes down to creating a single C++ class and changing the position reads and force writes to integrate the algorithm into SOFA's design.
+The precomputation phase takes place in the initialisation method where relevant variables are computed and passed to an external C function that allocates memory on the GPU and binds textures to it.
+During the simulation loop, the Solver requests the computation of the forces by launching the appropriate kernels on the GPU.
+
+\subsubsection*{Kernel organisation.}
+Although CUDA allows scattered writes in theory, it offers no write conflict management between threads.
+Why is this a problem?
+The major steps of the TLED algorithm are the following: (1) we compute the element nodal forces, that is for each node of a given element, we calculate the force contribution that this element has on this node; (2) we obtain the global force for each node by summing up the force contributions of all elements to which this node is attached; (3) we compute the displacement for each node by integration using the central difference method.
+In a serial implementation (CPU) we would do a loop on all elements and for each element we compute all element nodal force contributions.
+We would then directly add the element force contributions computed for each node into a global node array at the corresponding location for each node of the current processed element.
+In a parallel implementation, many elements are processed concurrently, in parallel.
+Inevitably, two elements sharing a node may be processed in the same time and may therefore try to both write their computed nodal force contribution into the exact same location in GPU's memory.
+At the time of this first implementation (2007), the result of concurrent writes was undefined.
+In fact, only one write was guaranteed to happen and there was no way to know which one succeeded.
+Potential measures to address this have proved extremely detrimental to performance.
+For this reason, the kernel arrangement in our first CUDA implementation eventually ends up to be essentially the same as the one used by \cite{Taylor07b}.
+
+Consequently, the TLED CUDA implementation also relies on 2 kernels.
+The first kernel operates over elements in the model and computes the element stresses based on the current model configuration.
+It then converts these into nodal force contributions, which are written to global memory.
+The second kernel operates over nodes and reads the previously calculated element force contributions and sums them for each node.
+The SOFA central difference solver computes the displacements from the nodal forces.
+Therefore, due to the impracticability of scattered writes, the sum operation is reformulated as a gather and the second kernel is needed to sum the nodal forces.
+The use of the developed viscoelastic constitutive update scheme necessitates storage of an additional array of state variables $ \boldsymbol \Upsilon_i $.
+
+\subsubsection*{Memory usage.}
+One efficient method for reading global memory data within kernels is texture fetching.
+Textures may be bound to either \emph{cudaArrays} or regions of linear memory.
+CudaArrays have been designed to achieve optimal fetching when the access pattern has a high level of 2D locality.
+In the present application, the access pattern among threads is essentially random (since unstructured meshes are used) and our experiments have shown that texture fetching from linear memory is in fact fastest.
+Therefore all global memory precomputed variables were accessed using this method.
+
+In SOFA, the forces are stored on the GPU in global memory.
+Since this memory space is not cached, it is important to follow the appropriate access pattern to obtain maximum memory bandwidth, especially given how costly accesses to device memory are.
+A multiprocessor takes 4 clock cycles to issue one memory instruction for a set of threads.
+When accessing global memory, there are, in addition, 400 to 600 clock cycles of memory latency.
+A suboptimal access pattern would yield incoherent writes.
+The memory bandwidth would then be an order of magnitude lower.
+In order to prevent this, a key feature of CUDA has been used: \emph{shared memory}.
+This is a very fast memory shared by all the processors of a multiprocessor.
+Hence, the results of the second kernel are first copied to shared memory and then moved to global memory.
+If the copies are well organised, it is possible to re-order the access to fulfil all the memory requirements (for both shared and global) and thus reach the maximum bandwidth.
+
+\subsubsection*{CPU-GPU interaction.}
+CPU-GPU interaction is generally a significant bottleneck in General Purpose GPU applications due to the relatively low interface bandwidth and it is desirable to minimise such interaction.
+However, interaction cannot be entirely removed from the present implementation since, for example, the solver requires inputs in the form of loaded nodes (which may change due to the interaction with the user) and their displacements, and may need to provide outputs in the form of reaction forces for haptic feedback.
+CUDA alleviates the problem somewhat by allowing allocation of areas of page-locked host memory which are directly accessible by the GPU and therefore offer much higher bandwidth.
+In SOFA, all transfers between CPU and GPU are made via this mechanism.
+
+\subsubsection*{Element technology.}
+We used both reduced integration 8-node hexahedral and 4-node (linear) tetrahedral elements.
+Tetrahedral meshes are easily generated and therefore widely used in simulations.
+However, hexahedra are preferable both in terms of solution accuracy and computational efficiency.
+Indeed, 4-node linear tetrahedra are known to be susceptible to volumetric locking \citep{Hughes00}.
+Yet, a disadvantage of hexahedra is the difficulty in automatically generating hexahedral meshes over arbitrary domains.
+Construction of hexahedral meshes is time consuming and laborious, which fact is of even greater significance when patient-specific simulations are considered.
+For this reason tetrahedral meshes are widely used in simulations despite their short comings.
+
+Another drawback of using hexahedra is the existence of so-called Hourglass modes that have to be addressed to avoid deterioration of the solution \citep{Flanagan81}.
+Techniques for suppressing these modes exist, but naturally involve additional computations.
+However, for a given number of degrees of freedom (DOF), a hexahedral mesh can be built with far fewer elements than a tetrahedral one.
+Since the majority of the calculations in explicit dynamic analyses are performed per element, this results in reduced overall computation time.
+
+From a GPU perspective, hexahedral element computations are substantially heavier and demand more memory resources than linear tetrahedral elements.
+Most of the matrices (such as shape function derivatives and nodal displacements) are twice as large for hexahedra, which necessitates twice as many texture fetches per element, and use of twice as many registers per thread.
+Similarly, twice as many nodal forces per element are written to global memory.
+Additional variables associated with hourglass control are required also.
+Therefore, on a per element basis hexahedra are significantly less efficient than tetrahedra, especially for GPU execution where memory efficiency is crucial.
+Thus, we observe that the occupancy (GPU percentage usage) drops from $25\,\%$ to only $8\,\%$ when using hexahedral elements.
+However, from the point of view of an entire model the lower number of hexahedra required for a given number of degrees of freedom still outweighs this element-wise inefficiency.
+
+	\subsection{Second implementation: a better use of shared memory}
+
+\subsubsection*{Limitations of the previous implementation}
+The previous TLED algorithm has been extensively tested on regular meshes like cubes.
+The model was purposely simple to test our new CUDA implementation.
+However, the TLED algorithm was developed to compute the deformation of organs.
+Thus, we started to assess our TLED implementation on an irregular mesh: a liver.
+From a segmented liver obtained from IRCAD (Research institute against digestive system cancer, France), we meshed the surface using a marching cube technique \citep{Lorensen87}.
+After smoothing it with Blender, a tetrahedral mesh has been generated with Tetgen, a free mesh generator.
+From there, we realised that the TLED was highly non efficient with such a mesh.
+After an extensive analysis with the profiler provided by NVIDIA, we became able to explain it.
+The liver mesh has a higher valency, meaning that each node has more elements attached than within a regular mesh like a cube.
+And on GPU, nodes are processed in parallel by group of 32 threads (called a warp).
+But if the instructions are different within a warp, the warp is divergent and the computations are serialised.
+With its higher valency, the liver mesh involves much more divergent warps when the forces are added up.
+To give an idea of the difference, a cube ($343$ nodes and $1\,296$ elements) is computed in $0.16\,$ms when a liver mesh ($399$ nodes and $1\,250$ elements) is processed in $0.27\,$ms.
+Based on the limitations discovered when the TLED has been applied to this irregular mesh, we came up with a new design, that we implemented and tested.
+
+\subsubsection*{A single kernel approach}
+The main bottleneck of the previous implementation was adding up the forces on each node in a second kernel after the computation of element force contributions in a first kernel.
+The latter were stored in 4 different textures (8 for the hexahedral formulation) and in order to add up the forces the second kernel had to switch all the time between the correct textures within the warps, causing them to diverge and slowing down substantially the overall process.
+Our idea was therefore to use shared memory to avoid writing in global memory at the end of the first kernel and prevent the warps to diverge.
+As explained before, shared memory is a very fast memory shared by all the processors of a multiprocessor.
+And each block of threads is sent to a multiprocessor to be executed.
+Obviously the amount of shared memory is finite, which limits the validity of this new approach and the results in performance will depend on the hardware (which may have different amount of shared memory).
+
+In the precomputation phase, nodes and elements are sorted by block to leave enough space for storage in shared memory of the element force contributions initially calculated at the end of the first kernel.
+We add nodes to a block along with their unique elements until the shared memory is saturated.
+Of course, in this approach some elements will be computed several times (because being into different blocks).
+However, storing the force contributions in shared memory instead of global memory to read them later on via texture fetches widely outperforms the need of computing several elements several times.
+
+\TODO{If time, try this implementation on Geforce GTX 480 as it features 3 times the amount of shared memory.}
+
+	\subsection{Third implementation: atomic writes}
+In April 2010, NVIDIA released a family of graphics cards based on a new GPU architecture called Fermi.
+Among many new features and improvements over previous generation, these cards added the capability of so-called \emph{atomic writes} for floats.
+The write operation is said to be atomic in the sense that it is guaranteed to be performed without interference from other threads.
+In other words, no other thread can access this address until the operation is complete.
+The concurrent writes are simply serialised in hardware.
+This feature allows us to implement the TLED algorithm as we would on a CPU, using a single kernel for all computations, and without the need to re-order the nodes and elements organisation as in the second implementation.
+
+\TODO{If time, try simple implementation using atomic writes with floats on Geforce GTX 480}
 
@@
 	\subsection{Compression of a cube}
-	\subsection{GPU performance}
+	\subsection{GPU performance (+comparison of different implementations)}
 	\subsection{Simulation of liver deformation}

### 2010-08-30 Thesis/Chapter5.tex
- Update 30 August -
Almost finished chapter 5 and 6 (first pages yet to be written) + bib udpated + added reviewer Sorensen



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@168 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-	\subsection{Computation of node displacements}
+	\subsection{Computation of node displacements} \label{chap5:computeNodalDisp}
 We recall that the dynamic system of equations describing a deformable solid is the following (see section \ref{chap3:dynamic2}:
@@
 It is worth noting that the coefficient vectors $ \mathbf{A} $, $ \mathbf{B} $ and $ \mathbf{C} $ can be precomputed.
+
+\bigskip
+
+Of crucial importance in explicit analyses is the restriction on time step sizes imposed by stability constraints \label{chap5:discussionTimestep}.
+As mentioned, the central difference method is only stable if $ \Delta t <  \Delta t_{cr} $.
+Essentially, stability requires that the time step be not larger than the time required for a dilatational wave to propagate through the smallest element.
+In linear analyses this translates into the following formula for $ \Delta t_{cr}$:
+\begin{equation}
+\label{chap5:criticalTimestep}
+\Delta t_{cr} = \dfrac{L_e}{c},
+\end{equation}
+where $ L_e $ is the smallest characteristic element length (roughly interpreted as the smallest edge length) in the assembly, and $c$ is the dilatational wave speed of the material, given by
+\begin{equation}
+\label{chap5:waveSpeed}
+c = \sqrt{\dfrac{E (1 - \nu)}{\rho (1+\nu)(1-2\nu)}},
+\end{equation}
+where $E$ is Young's modulus and $\nu$ is Poisson's ratio.
+If nonlinear analyses (either geometric or material) are conducted, a somewhat smaller time step must be employed since $c$ will change with deformation, and in particular will generally increase.
+Additionally, the presence of damping necessitates a further decrease in the limit.
+Nonetheless, \eqref{chap5:waveSpeed} may be used to estimate $ \Delta t_{cr}$.
+
+Some important observations arise from \eqref{chap5:criticalTimestep} and \eqref{chap5:waveSpeed}.
+Firstly, since the stiffness of soft tissues is very much smaller than that of common engineering materials ($E \approx 3 \times 10^3\, $Pa for brain tissue versus $\approx 2 \times 10^{11}\, $Pa for steel), the allowable time step is much larger for analysis of the former.
+This is a key reason for the expediency of explicit time integration for analysis of soft tissues.
+Secondly, soft tissues are generally considered to be incompressible, leading to $\nu \approx 0.49$ commonly being employed.
+However, if this value can be relaxed even further, significant increases in $ \Delta t_{cr}$ are possible.
+As an example, for a given $ Le $ a material with $E = 3\,000\, $Pa and $\nu = 0.45$ allows a time step of more than double that of a material with $\nu = 0.49$.
+Of course, lowering the Poisson ratio introduces inaccuracies, but for some applications (interactive medical simulation for instance) it may be that this is acceptable.
+Finally, it must be highlighted that $ \Delta t_{cr}$ is proportional to $ L_e $.
+If the element size is decreased (by using a finer mesh, or by analysing very small objects), inexorably $ \Delta t_{cr}$ is also reduced.
 

### 2010-08-30 Thesis/Chapter6.tex
- Update 30 August -
Almost finished chapter 5 and 6 (first pages yet to be written) + bib udpated + added reviewer Sorensen



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@168 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 The concurrent writes are simply serialised in hardware.
-This feature allows us to implement the TLED algorithm as we would on a CPU, using a single kernel for all computations, and without the need to re-order the nodes and elements organisation as in the second implementation.
+This feature allows us to implement the TLED algorithm as we would on a CPU, using a single kernel for all computations, and without the need to re-order the nodes and elements organisation as with the second implementation.
 
@@
 \section{Results}
+
+We present a series of examples based on pure shear and compression of cube models to demonstrate the validity of the constitutive update procedure and the performance of the GPU-based implementation.
+We conclude with an example of simulation of liver deformation, since as mentioned there is experimental evidence that liver exhibits an anisotropic response \citep{Chui07}.
+In each case we used a transversely isotropic visco-hyperelastic model with elastic strain energy components defined by
+\begin{align}
+\Psi^{iso} &= \Psi^{I}(\bar{I}_1) + \Psi^{TI}(\bar{I}_4) \notag \\
+&= \dfrac{\mu}{2} (\bar{I}_1 - 3) + \dfrac{\eta}{2} (\bar{I}_4 - 1)^2, \\
+\Psi^{vol} &= \dfrac{\kappa}{2} (J - 1)^2,
+\end{align}
+where $ \Psi^{I} $ and $ \Psi^{TI} $ are isotropic and transversely isotropic components, respectively, $ \mu $ is the small strain shear modulus, $ \kappa $ is the bulk modulus, and $ \eta $ is a material parameter with units of Pa.
+The isotropic component $ \Psi^{I} $ represents the well known neo-Hookean model, while the anisotropic component $ \Psi^{TI} $ is the same form used by \cite{Picinbono01}.
+Here the dependence of $ \Psi^{TI} $ on $ \bar{I}_5 $ is omitted and we include only $ \bar{I}_4 $ terms.
+This was done foremost for simplicity, but is also a common practical feature of anisotropic models for soft tissues.
+The main reason is the clear physical interpretation of $ \bar{I}_4 $ as the square of the stretch $ \lambda $ \footnote{Strictly, $ \bar{I}_4 = J^{-2/3} \lambda^2 $} along the preferred direction $ \mathbf{a}_0 $ \cite{Holzapfel00}, whereas $ \bar{I}_5 $ has a less clear physical meaning.
+$ \bar{I}_4 $ terms are therefore more amenable to experimental investigation, and of more use in modelling structural elements (e.g.
+fibres) responsible for the anisotropy.
+It must be emphasised that this model is used as an example only, and we make no claim concerning its appropriateness for any particular tissue.
+
+\bigskip
+
+The instantaneous isochoric and volumetric responses are then given by
+\begin{align}
+\boldsymbol \Phi^{iso} &= J^{-2/3} \left[ \mu \left( \mathbf{I} + \dfrac{I_1}{3} \mathbf{C}^{-1} \right) + \eta (\bar{I}_4 - 1) \left( \mathbf{A}_0 + \dfrac{I_4}{3} \mathbf{C}^{-1} \right) \right], \\
+\boldsymbol \Phi^{vol} &= \kappa J (J - 1) \mathbf{C}^{-1}.
+\end{align}
+We used viscoelastic isochoric terms only, with a single Prony series term for simplicity. The complete stress response is then
+\begin{align}
+\mathbf{S} &= \boldsymbol \Phi^{vol} + \int_0^t \left[ 1 - \alpha_1 \left(1-e^{(s-t)/\tau_1} \right) \right] \pd{\boldsymbol \Phi^{iso}}{\tau} ds \notag \\
+&= \boldsymbol \Phi^{vol} + \boldsymbol \Phi^{iso}(1 - \alpha_1) + \alpha_1 \int_0^t e^{(s-t)/\tau_1} \pd{\boldsymbol \Phi^{iso}}{\tau} ds.
+\end{align}
+Note that in this case the developed constitutive update procedure need only to be employed for isochoric terms.
+
+\bigskip
+
+Material parameters were chosen based on recent results for the viscoelastic response of human liver in vivo.
+\cite{Nava08} used an aspiration device to obtain an estimate of $\mu = 19\,700\,$Pa for the combined liver/liver capsule.
+However, they noted that an earlier study suggested the presence of the liver capsule meant this value could overestimate liver parenchyma stiffness by up to $3 \times$.
+Since it is the parenchyma we are concerned with, we employed a value of $\mu = 19\,700/3 = 6\,567\,$Pa.
+Assuming near incompressibility we specified $\kappa = 326\,210\,$Pa (corresponding to a Poisson ratio of 0.49).
+\citeauthor{Nava08} employed two Prony terms, and we use parameters from the first of these: $\alpha_1 = 0.5, \tau_1 = 0.58$.
+The short time constant $ \tau_1 $ means this Prony term most affects faster loadings, which are of interest in surgical simulations.
+Finally, since \citeauthor{Nava08} assumed an isotropic response their data provide no basis for estimating $ \eta $.
+Therefore we chose a conservative value of $\eta = 2\mu$, conservative since \cite{Picinbono01} used a value closer to $\eta = 16\mu$.
+
 	\subsection{Pure shear of a cube}
+The first example involved shearing of a unit cube model.
+We imposed a ramped loading such that the displacements $u$ of the loaded nodes at time $t$ were given by $u = rt$, where $r$ is the loading speed, and we chose the $x$-direction to be the preferred material direction ($\mathbf{a}_0 = [1\, 0\, 0 ]$).
+Since the deformation is homogenous only a single element was required to reproduce the stresses exactly.
+Moreover an analytical solution is available with which numerical results may be compared \citep{Taylor2009}.
+Four independent nonzero stress terms are produced.
+
+In \fig{chap6:fig-stressCurvesCube1} we demonstrate the effects of material anisotropy independent of viscoelastic effects by plotting the limiting instantaneous (hyperelastic) stress curves, produced by setting the relaxation parameter $\alpha_1 = 0$.
+\fig{chap6:fig-stressCurvesCube1}(a) shows curves for the anisotropic model, while \fig{chap6:fig-stressCurvesCube1}(b) shows curves for an equivalent isotropic model (i.e.
+with $\eta = 0$).
+Firstly we confirm that the FE solutions (data marks) match the analytical solutions (solid curves) very accurately.
+Secondly we observe that the anisotropic term significantly affects both the magnitude and shape of the stress curves.
+In particular the $S_{11}$ curve which is coincident with the $S_{33}$ curve in the isotropic model actually changes sign when anisotropy is introduced.
+The other curves exhibit a more pronounced stiffening for the anisotropic case.
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=13.9cm]{chapter6/stressCurvesCube1.pdf}
+\end{center}
+\caption[Stress comparison with analytical solution for a pure shear of a cube]{Instantaneous stress curves for pure shear deformation of (a) the anisotropic model compared with those of (b) an isotropic model.
+Solid lines correspond to the analytical solution, while markers indicate the FE solution.}
+\label{chap6:fig-stressCurvesCube1}
+\end{figure}
+
+In \fig{chap6:fig-stressCurvesCube2} we demonstrate the effect of loading rate on the complete viscoelastic model by plotting the $S_{11}$ component for strain rates of $2.5\,s^{-1}, 0.25\,s^{-1}$, and $0.025\,s^{-1}$, plus the bounding instantaneous and equilibrium responses.
+The strain rate-dependence introduced by the model (and commonly observed in biological tissues) is clearly shown.
+Additionally the close match between the analytic and numerical solutions demonstrates the validity of the developed constitutive update scheme.
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=10cm]{chapter6/stressCurvesCube2.pdf}
+\end{center}
+\caption[Stress comparison with analytical solution for a pure shear of a cube at various strain rate]{$S_{11}$ curves for pure shear deformation of the anisotropic viscoelastic model at varying strain rates.
+Curves for strain rates of $2.5\,s^{-1}, 0.25\,s^{-1}$, and $0.025\,s^{-1}$ are given (as labelled), along with the bounding instantaneous and equilibrium responses.
+The latter two are labelled $ \infty $ and $0^+$, respectively, indicating that they correspond to strain rates approaching these values.
+Solid lines correspond to the analytical solution, while markers indicate the FE solution.}
+\label{chap6:fig-stressCurvesCube2}
+\end{figure}
+
 	\subsection{Compression of a cube}
-	\subsection{GPU performance (+comparison of different implementations)}
-	\subsection{Simulation of liver deformation}
-	\subsection{A medical application: cataract surgery}
+The second example involved compression of a cube of edge length $0.1\,$m (comparable to some human organs).
+The cube was imagined to be fixed to opposing load platens and free on the remaining four faces.
+Compression was applied along the $x$-axis.
+A stress-relaxation type test protocol was simulated, in which the cube was compressed by $30\, \%$ in 0.5 seconds and the compression was held for a further 4.5 seconds.
+We demonstrate the effects of anisotropy on the deformation response by comparing the deformed shapes of anisotropic models with $\mathbf{a}_0 = [0\, 1\, 0] \, {\buildrel\rm	def\over=} \, \mathbf{a}_0^y $ and $\mathbf{a}_0 = [0\, \frac{1}{\sqrt{2}} \, \frac{1}{\sqrt{2}}] \, {\buildrel\rm	def\over=} \, \mathbf{a}_0^{yz} $  with that of an isotropic model (with $\eta = 0$).
+
+\bigskip
+
+Figure \ref{chap6:fig-cubeAnisotropic}(a) shows the undeformed cube, while \fig{chap6:fig-cubeAnisotropic}(b) shows the deformed isotropic model.
+With no preferred direction the lateral expansion was uniform.
+In \fig{chap6:fig-cubeAnisotropic}(c) the increased $y$-direction stiffness of the first anisotropic model lead to a much reduced expansion in this direction (vertical in the image), and an accompanying increase in the orthogonal $z$-direction.
+For the second anisotropic model the direction defined by $y = z$ was stiffened, and \fig{chap6:fig-cubeAnisotropic}(d) shows the resulting reduced expansion along this axis and the increased orthogonal expansion.
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=13.9cm]{chapter6/cubeAnisotropic.pdf}
+\end{center}
+\caption[Deformation patterns of transversely isotropic models compared with that of an isotropic model]{Deformation patterns of transversely isotropic models compared with that of an isotropic model: (a) the undeformed cube, (b) the deformed isotropic model, (c) the deformed anisotropic model with $ \mathbf{a}_0 = \mathbf{a}_0^y $, (d) the deformed anisotropic model with $\mathbf{a}_0^{yz}$.
+Colour maps indicate relative magnitude of lateral displacement  $((u_y^2 + u_z^2)^{1/2})$.}
+\label{chap6:fig-cubeAnisotropic}
+\end{figure}
+
+\bigskip
+
+The end face reaction force history for each model is shown in \fig{chap6:fig-stressDecay}.
+A distinct decay curve resulting from stress relaxation, and similar to that commonly noted in biological tissues \citep{Fung93}, was observed.
+The increased stiffness afforded by the anisotropic models (albeit in an orthogonal direction to the loading) result in greater reaction forces than the isotropic model.
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=10cm]{chapter6/stressDecay.pdf}
+\end{center}
+\caption[FE computed reaction forces on the end faces of the cube models over time.]{FE computed reaction forces on the end faces of the cube models over time.
+Curves are labelled according to the constitutive model used, where \emph{Anisotropic 1} refers to the transversely isotropic viscoelastic model with $\mathbf{a}_0 = \mathbf{a}_0^y$ and \emph{Anisotropic 2} refers to the case of $\mathbf{a}_0 = \mathbf{a}_0^{yz}$.
+Note that in this figure both markers and solid lines correspond to FE solutions.}
+\label{chap6:fig-stressDecay}
+\end{figure}
+
+	\subsection{GPU performance}
+
+		\subsubsection*{General GPU performance}
+Next, we assessed the computational performance of the GPU implementation and efficiency of the constitutive update scheme by measuring computation times for a range of mesh densities.
+Using the cube geometry as above we generated models with mesh sizes ranging from $3\,993$ DOF ($1\,331$ nodes, $1\,000$ hexahedral elements) to $177\,957$ DOF ($59\,319$ nodes, $54\,872$ hexahedral elements).
+We compared the GPU solution times for a single time step with those of a CPU-based implementation developed using C++.
+The test machine included an Intel Core2Duo 2.4GHz CPU, 2GB RAM, and an NVIDIA GeForce 8800GTX GPU.
+Three constitutive models were considered: the transversely isotropic viscoelastic model (TIV), TIV minus the viscoelastic terms (TIE), and TIE minus the anisotropy terms (NHE).
+Comparison of times for TIV and TIE indicates the computational load introduced by the developed viscoelastic constitutive update scheme.
+Comparison of times for TIE and NHE indicates the computational load introduced by material anisotropy.
+Figure~\ref{chap6:fig-timingsGPU} shows the results of the experiments.
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=13.9cm]{chapter6/timingsGPU.pdf}
+\end{center}
+\caption[GPU solution times]{(a), (b) show CPU and GPU solution times, respectively, for a single time step for each constitutive model as a function of model size, (c) shows the ratio of CPU to GPU solution times for model TIV, and (d) shows ratios of GPU solution times for the constitutive models: TIV to TIE, TIE to NHE, and TIV to NHE.
+(b) also shows the critical time steps $ \Delta t_{cr} $ for each model size using liver-like and brain-like properties.}
+\label{chap6:fig-timingsGPU}
+\end{figure}
+
+In \fig{chap6:fig-timingsGPU}(a) and \fig{chap6:fig-timingsGPU}(b) it can be seen that GPU solution affords significant speed improvements over CPU solution for all constitutive models.
+Additionally it appears that solution times are little affected by the introduction of the more complex constitutive models, and importantly by the use of the developed constitutive update scheme.
+This is borne out in \fig{chap6:fig-timingsGPU}(d), where we observe that the maximum solution time ratio for model TIE to model NHE (anisotropy vs isotropy) was $1.013$, and that of model TIV to model TIE (viscoelastic vs elastic) was $1.043$.
+The largest total solution time increase for an anisotropic viscoelastic model compared with an isotropic hyperelastic one was $5.1\, \%$ We conclude that the key features of anisotropy and viscoelasticity may be included in simulations at very little additional computational cost.
+Referring to \fig{chap6:fig-timingsGPU}(c) we observe a maximum speed improvement for GPU solution over CPU solution of $56.3 \times$.
+
+\fig{chap6:fig-timingsGPU}(b) also shows the decreasing critical time steps $ \Delta t_{cr} $ with increasing model sizes (which results from the decreasing element sizes), using liver-like and brain-like materials.
+We define liver-like materials to be those described by model TIV with material parameters quoted at the beginning of the section, whereas brain-like materials are those described by model NHE with material parameters $\mu = 1006.71\, $Pa and Poisson?s ratio $\nu = 0.45$.
+The intersections of these curves with the solution time curves provide an estimate of the real-time capacity of the implementation.
+This is discussed further in section~\ref{chap6:criticalTimestep}.
+
+\TODO{Add comparison of different implementations}
+
+		\subsubsection*{Performance within SOFA}
+The efficiency and the side-effects of porting the algorithm into the flexible SOFA framework need to be measured.
+Therefore the performance has been assessed by comparing the computational time of the algorithm running within and outside SOFA.
+NVIDIA provides a tool to check the GPU implementation by evaluating many variables during the execution like for instance timings, counts of inconsistent reads and writes or GPU occupancy.
+We used this tool to carry out two measures:
+\begin{enumerate}
+\item GPU time only estimates the GPU computational time.
+\item CPU time allows the evaluation of the execution time with the additional overhead due to the framework.
+\end{enumerate}
+
+\begin{figure}
+\includegraphics[width=13.9cm]{chapter6/cube_and_curves.pdf}
+\caption [Computational timings within SOFA] {\textbf{Left:} FEM mesh of cube with $29\,791$ nodes deformed under a uniformed load.
+\textbf{Right:} Comparison of GPU computational timings and CPU overheads between the SOFA and standalone implementations for different mesh sizes.}
+\label{chap6:fig-cube_and_curves}
+\end{figure}
+
+The tests were performed on a simple test scene featuring a cube under gravity.
+Hexahedral meshes with different resolutions from $1\,331$ to $29\,791$ nodes were used and the results are presented in figure \ref{chap6:fig-cube_and_curves}.
+In our standalone implementation, the second kernel not only accumulates nodal forces but also adds gravity and updates positions based on the central difference integration scheme.
+These operations are split into separated components in SOFA, in order to introduce more flexibility (such as applying additional forces or changing the integration algorithm).
+While this introduces no noticeable difference on a CPU-based simulation, when using the GPU it is more costly due to overheads in the CUDA API for the additional kernel launches.
+Although it reduces the performance by $8.4\, \%$ for large meshes, this could be optimised away by adding a kernel specific to a given combination of components.
+
+	\subsection{Simulation of liver deformation} \label{chap6:liverSimu}
+Finally, we performed simulations of manipulation of a liver model in order to demonstrate the feasibility of the procedure with more realistic organ geometries.
+The model geometry was based on the liver mesh available in the SOFA framework, which we smoothed and remeshed to improve uniformity of element sizes and node valencies.
+A mesh of 748 nodes and $2\,428$ tetrahedral elements was solvable in real-time.
+For anisotropic models a preferred material direction $\mathbf{a}_0 = [0\, 1\, 0] $ was used, corresponding to the vertical direction in \fig{chap6:fig-liver}.
+We note that in a real organ the preferred direction likely varies depending on the local vasculature and orientation of lobules \citep{Chui07}, but in the absence of experimental data for the present specimen the mentioned direction was assumed throughout.
+Moreover, on this point we note that different preferred directions may be specified for each element in the mesh, if such data are available.
+
+\bigskip
+
+The Von Mises equivalent stress distributions resulting from stretching the model with and without inclusion of viscoelasticity are compared in \fig{chap6:fig-liver}.
+For the viscoelastic case (\fig{chap6:fig-liver}, right image) the model is depicted several seconds after the completion of the stretch, so that stresses had reached an approximately steady state.
+The stress dissipation introduced by the viscoelastic model is clearly evident, and for example would result in significantly altered virtual tool reaction forces.
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=13.9cm]{chapter6/liver.pdf}
+\end{center}
+\caption[GPU solution times]{Deformation of a liver model using the TI model (middle) and the TIV model (right).
+The undeformed mesh is shown on the left.
+Colour maps indicate the relative Von Mises equivalent stress magnitude.}
+\label{chap6:fig-liver}
+\end{figure}
+
+\TODO{Add movies online? If possible...}
 
 \section{Discussion}
-	\subsection{Critical time step}
-	\subsection{Difficulties to handle contacts}
+
+	\subsection{Critical time step}	\label{chap6:criticalTimestep}
+As already emphasised in section~\ref{chap5:computeNodalDisp} page \pageref{chap5:discussionTimestep}, the use of an explicit integration scheme imposes that the solution time remains below the critical time step.
+This critical time step depends on two kind of criterion: the size of the elements (due to the choice of the mesh) and the material's properties (Young's modulus and Poisson's ratio).
+
+The first factor leads to a strong constraint on the quality of the mesh.
+Indeed, meshing a complex shape with good and uniform elements is not a straightforward task and rather often bad and very small elements are created.
+In such a case, the critical time step for the whole simulation would be small even if there is only a single small element in the mesh.
+As an example, the liver mesh used in the section~\ref{chap6:liverSimu} had to be smoothed and remeshed properly to allow a decent simulation.
+
+Material's properties are also a very influent factor for the determination of a critical time step.
+If the critical time step may be fairly large for very soft tissue (like brain), it is already divided by 2 when modelling liver.
+As we have seen, the value of the Poisson's ratio may be relaxed from 0.49 to 0.45 for instance.
+If it allows a time step twice as large, this relaxation introduces inaccuracies.
+If this is certainly acceptable for medical training simulators, the demand in precision for surgical planning or per-operating assistance would probably prevent such an approximation.
+
+In a nutshell, additional care must be taken to produce quality meshes and the range of materials that may be used while maintaining interactive computations is limited when using an explicit time integration scheme.
+However, it allows a huge gain in speed for very soft tissue when the compromises on material's properties are acceptable for the desired application, as with medical simulators used for training.
+For the interested reader, a more thorough discussion on the critical time step of explicit analyses may be found in \cite{Taylor08}.
+
+	\subsection{Handling contacts}
+The present work was entirely concerned with numerical solution of constitutive models in explicit FE analyses, however an equally important issue from the point of view of interactive simulation is the modelling of interaction between organs and virtual tools, and its possible impact on $ \Delta t_{cr} $ and the stability of the simulation.
+Explicit time integration schemes are known to be more sensitive to contacts than implicit schemes.
+While this is beyond the scope of the present work, we note that a range of contact formulations are available in industry standard commercial explicit FE codes (e.g.
+LS-Dyna and ABAQUS).
+The methods are well tested and understood in the computational mechanics community, with two main types in common use: penalty methods and kinematic constraint methods.
+Penalty methods involve use of fictitious contact springs at interfaces, effectively to penalise interpenetration of opposing surfaces.
+If excessively stiff springs are used $ \Delta t_{cr} $ may indeed be reduced, however except in extreme conditions (e.g.
+from explosive or high speed impacts, or imposition of very large forces on highly constrained objects) spring stiffnesses of the same order of magnitude as the material stiffness are suitable.
+The effect on $ \Delta t_{cr} $ is therefore minimal \citep{Hallquist06}.
+Kinematic constraint methods involve (loosely) direct relocation of penetrating nodes back to the outside surface of the penetrated body, and have no effect on $ \Delta t_{cr} $.
+These may be especially useful for modelling interaction between soft tissues and stiff (effectively rigid) surgical tools.
+The computational cost of either variety has not been investigated here.

### 2010-08-30 Thesis/TitlePage.tex
- Update 29 July - Added Josh's corrections for first half of chapter 2
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@137 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 30 August -
Almost finished chapter 5 and 6 (first pages yet to be written) + bib udpated + added reviewer Sorensen



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@168 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \noindent \large Prepared by Olivier \textsc{Comas} at \\ 
-INRIA Lille, \textsc{Shaman} Team and CSIRO ICT Brisbane, EAHRC \\
+INRIA Lille, \textsc{Shaman} Team and CSIRO ICT Brisbane, AEHRC \\
 \vspace*{0.5cm}
@@
 \begin{tabular}{llcl}
-      \textit{Reviewers:}	& Bernard \textsc{}		& - & INRIA (Shaman)\\
+      \textit{Reviewers:}	& Thomas Sangild \textsc{S\o rensen}          & - & University of Aarhus, Denmark\\
 				& Bernard \textsc{}		& - & INRIA (Shaman)\\
-      \textit{Advisor:}	& Stéphane \textsc{Cotin}		& - & INRIA (Shaman)\\
+      \textit{Advisor:}	& Stéphane \textsc{Cotin}		& - & INRIA (Shaman), France\\
       \textit{President:}	& Bernard \textsc{}		& - & INRIA (Shaman)\\

### 2010-08-31 Thesis/Chapter4.tex
- Update 26 August -
End chapter 5 + start of chapter 6 + bib updated + PhD comics added for later use :)



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@166 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 31 August -
Finished chapter 5 and 6 and started added references for chapter 7



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@169 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
+\bigskip
+
+Early 2000s, the rapid increase in the performance of graphics hardware, coupled with recent improvements in its programmability, have made graphics hardware a compelling platform for computationally demanding tasks in a wide variety of application domains.
+Researchers and developers have become interested in harnessing this power for general-purpose computing, an effort known collectively as GPGPU (for General-Purpose computing on Graphics Processing Units).
+A survey of GPGPU algorithms and techniques was written by \cite{Owens07}.
+
+\cite{Mosegaard05a,Mosegaard05b} were probably the first to implement a mass-spring model on GPU in medical simulation.
+For the first time, the whole simulation of was performed on the GPU (physics, interaction and visualisation) and they used their framework in a congenital cardiac surgery simulation.
+They compared two implementations: the first one explicitly defined the connectivity between springs using a connectivity texture while the second one determines implicitly determines at runtime that springs are connected when they are neighbours.
+They were able to compute 188 iterations per second on a $42\,745$ particule model of a pig heart.
+Depending on the size of the model and the type of implementation, the authors achieved speed-ups between 10 and 30 $ \times $ over their CPU implementation.
+
+Taking advantage of the GPU to reach a substantial speed-up is usually not straightforward.
+One must understand the limitations inherent in its design and devise algorithms accordingly.
+With this idea in mind, \cite{Sorensen06} reviewed the most important concepts and data structures required to realise two popular deformable models on the GPU: the spring-mass model and the finite element model.
+
+\TODO{Other GPU-based mass-spring implementations?}
+
 If mass-spring systems are quite versatile and easy to implement, the most complete mathematical formalism available to describe the mechanical behaviour of a solid is provided by continuum mechanics.
@@
 
-Finally, \cite{Miller07} proposed the total Lagrangian explicit dynamics (TLED) algorithm, a very efficient fully non-linear formulation.
+In 2007, \citeauthor{Miller07} proposed the total Lagrangian explicit dynamics (TLED) algorithm, a very efficient fully non-linear formulation.
 The algorithm is based on the finite element method using the total Lagrangian formulation, where stresses and strains are measured with respect to the original configuration.
 This choice allows for precomputing of most spatial derivatives before the commencement of the time-stepping procedure.
-The authors reported that the average number of floating-point operations per element per time step is $35\,$\% lower than for the similar implementation of the algorithm based on updated Lagrangian formulation.
+The authors reported that the average number of floating-point operations per element per time step is $35\, $\% lower than for the similar implementation of the algorithm based on updated Lagrangian formulation.
 They were able to compute the deformation of a hexahedral mesh of $ 6\,000 $ elements in about $ 16\,$ms for a time step of $ 1\,$ms.
 If they could not achieve real-time computations, their work constituted a step towards the simulation of entire organs in real-time.
+
+Finally, \cite{Taylor07a} presented an efficient constitutive update procedure for viscoelastic models suitable for simulation of soft tissues at large strains and varying strain rates.
+The procedure is formulated for use in explicit dynamic finite element algorithms like the TLED algorithm.
+The procedure is based on a class of visco-hyperelastic constitutive models developed from purely hyperelastic strain energy functions by introducing relaxation terms and expressing in convolution integral form.
+The resulting constitutive equations are separated into rate-dependent and -independent terms, with the former being designated as state variables to be stored and updated at each time step also.
+These are converted to a differential form, which upon integration in time yields the required incremental update formula.
+They showed the validity of the procedure with a series of numerical examples.
 
@@
 
-\cite{Wu04} proposed the first application of GPU in medical simulation.
+\cite{Wu04} proposed the first application of GPU in medical simulation \label{chap4:GPUMedicalSimulation}.
 They were able to achieve an interactive frame rate with topology changes in surgical simulation.

### 2010-08-31 Thesis/Chapter5.tex
- Update 31 August -
Finished chapter 5 and 6 and started added references for chapter 7



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@169 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{shortAbstract}
-TLED proposed by Miller almost real-time \\
-GPU implementation to reach real-time \\
-But first: TLED description \\
-\OC{MICCAI 2008; MedIA 2009}
+In 2007, \citeauthor{Miller07} proposed the total Lagrangian explicit dynamics (TLED) algorithm, a very efficient fully non-linear formulation.
+The authors reported that the average number of floating-point operations per element per time step is $35\, $\% lower than for the similar implementation of the algorithm based on updated Lagrangian formulation.
+Their work constituted a step towards the simulation of entire organs in real-time.
+Later in the year, \citeauthor{Taylor07b} reformulated the TLED algorithm to propose the first GPU implementation of a fully non-linear finite element method.
+However, restrictions due to using a graphics-based API necessitated reformulation of the force summation in their implementation.
+\cite{Taylor07a} also presented an efficient constitutive update procedure for viscoelastic models formulated for use in explicit dynamic finite element algorithms like the TLED algorithm.
+With the release by NVIDIA of a new graphics card architecture along with a new and more flexible non-graphics API specially designed for general-purpose GPU, we decided to investigate the re-implementation of the TLED for this new architecture with the idea of a more straightforward and efficient implementation in mind.
+We also wanted to extend the simple non-linear TLED formulation by adding the constitutive update procedure for viscoelastic models introduced by \citeauthor{Taylor07a} and an anisotropic constitutive formulation as well.
+This work eventually yields a very efficient GPU-based non-linear, viscoelastic and anisotropic formulation for modelling solid organs in medical simulation and was the object of three publications \citep{Comas2008,Taylor2008,Taylor2009}.
+This chapter aims at describing the TLED algorithm in details, as well as introducing the viscoelasticity and anisotropy extensions.
+Their GPU implementations will be the object of the next chapter.
 \end{shortAbstract}
@@
 We adopt the notation of \citeauthor{Bathe95} with respect to indication of the relevant configuration of the body: a left superscript indicates the configuration in which a quantity occurs and, when applicable, a left subscript indicates the configuration with respect to which the quantity is measured.
-We note the coordinates of a point at time t $ ^t \mathbf{x}$.
+We note the coordinates of a point $ ^t \mathbf{x}$ at time t.
 
@@
 In geometrically linear analyses, we assume that the displacements are infinitesimally small so that the geometry of an element does not change over time and the strain-displacement matrix is constant.
-Conversely, because geometrically non-linear analyses (such as carried out by the TLED algorithm) take the change of geometry of the elements into account, the strain-displacement matrix $ \mathbf{B}_L = _0^t\mathbf{B}_L $ varies.
+Conversely, because geometrically non-linear analyses (such as carried out by the TLED algorithm) take the change of geometry of the elements into account, the strain-displacement matrix $ \mathbf{B}_L = \leftidx{_0^t}{\mathbf{B}}{_L} $ varies.
 However, the strain-displacement matrix at time $ t $ may be computed by transforming a linear matrix $ _0^t\mathbf{B}_{L0} $ using the deformation gradient.

### 2010-08-31 Thesis/Chapter6.tex
- Update 31 August -
Finished chapter 5 and 6 and started added references for chapter 7



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@169 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{shortAbstract}
-\OC{ISBMS 2008a}
+The previous chapter introduced the TLED algorithm in details as well as enhancements with a viscoelastic and anisotropic formulation.
+If the TLED algorithm had already been implemented on GPU by \cite{Taylor07b}, restrictions due to using a graphics-based API and the limited hardware possibilities of GPUs necessitated reformulation of the force summation in their implementation.
+With the release by NVIDIA of a new graphics card architecture along with a new and more flexible non-graphics API specially designed for general computation on GPU, we decided to investigate the re-implementation of the TLED for this new architecture with the aim to achieve a more straightforward and efficient implementation.
+The aim was also to enhance the GPU-based implementation of the TLED with a viscoelastic and anisotropic formulation.
+This chapter will describe the implementation of this non-linear, viscoelastic and anisotropic finite element method algorithm in the open source framework SOFA.
+By providing an efficient and accurate non-linear FEM for soft tissue modelling to worldwide researchers, we thus hope to assist in enhancing the realism of medical simulators.
+This work has led to three publications \citep{Comas2008,Taylor2008,Taylor2009}.
 \end{shortAbstract}
@@
 In the DirectX API (DirectX 9 and newer), shaders are programmed with the high level shader language (HLSL).
-Cg is another high-level shading language developed by NVIDIA in close collaboration with Microsoft and it is very similar to Microsoft's HLSL.
+Cg is another high-level shading language developed by NVIDIA in close collaboration with Microsoft and it is very similar to Microsoft's HLSL \citep{Mark03}.
 However, the Cg compiler is not specific to a graphics API and can output both OpenGL and DirectX shader programs.
@@
 \end{center}
-\caption[GPU solution times]{Deformation of a liver model using the TI model (middle) and the TIV model (right).
+\caption[GPU solution times]{Deformation of a liver model using the TI model \textbf{(middle)} and the TIV model \textbf{(right)}.
 The undeformed mesh is shown on the left.

### 2010-09-08 Thesis/Chapter2.tex
- Update 8 September -
Started writing of state of the art for shells + bib updated



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@170 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 For similar reasons, let us find the relation between an element of area in the reference configuration $d\mathbf{A}$ which becomes $d\mathbf{a}$ in the deformed configuration.
-Considering an arbitrary material line $d\mathbf{L}$ in the reference configuration and noting $d\mathbf{l}$ the same line after deformation (see \fig{chap2:fig-changeSurface}, the corresponding reference and current volumes are given respectively by:
+Considering an arbitrary material line $d\mathbf{L}$ in the reference configuration and noting $d\mathbf{l}$ the same line after deformation (see \fig{chap2:fig-changeSurface}), the corresponding reference and current volumes are given respectively by:
 \begin{align}
@@
 
-\TODO{Add details on constitutive equations? Level of details to be determined later (depending on TLED needs)}
-
 \section{Tissue characterisation}

### 2010-09-08 Thesis/Chapter4.tex
- Update 8 September -
Started writing of state of the art for shells + bib updated



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@170 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-	\subsection{Mass-spring model}
+	\subsection{Mass-spring model}	\label{chap4:massSprings}
 A classic technique for modelling a deformable object is the mass-spring system.
@@
 
-\cite{Hauth04} made use of the polar decomposition of the deformation gradient to extract the rotation.
+In constrast of \cite{Muller02} who discretised the rotation field at each vertex, \cite{Hauth04} determined the rotation at the level of the element.
+If \citeauthor{Muller02}'s technique of stiffness warping yields ghost forces due to inaccuracies in the determination of the rotation field, \citeauthor{Hauth04} made use of the polar decomposition of the deformation gradient to extract the exact rotation.
 They also applied this formulation in a hierarchical finite element setting where the coarse level is computed with a non-linear strain tensor and the finer levels are computed with a co-rotational method.

### 2010-09-08 Thesis/Chapter7.tex
- Update 12 August -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@150 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 17 August -
Major update on FEM state of the art + bib updated



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@153 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 31 August -
Finished chapter 5 and 6 and started added references for chapter 7



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@169 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 8 September -
Started writing of state of the art for shells + bib updated



git-svn-id: svn+

@@
-\chapter{State of art: hollow structures}
+\chapter{Modelling the deformation of hollow objects in real-time}
 \label{chap7}
@@
 
-\section{Expose the problem: models presented for solid organs are insufficient. How to model hollow organs?}
-	\subsection{Colonoscopy simulator project: needs for colon}
-	\subsection{Cataract surgery, stenosis: other needs for implants/blood vessels}
+\section{Introduction: the problematic}
+The human body is composed of various deformable anatomical structures.
+A key challenge of soft-tissue modelling is the variousness of the mechanical behaviours.
+The shape and the internal structure may be very different from one organ to another.
+If you take the liver for instance, it is a solid organ composed of four lobes, each one made up of units called lobules.
+Each lobule consists of a central vein surrounded by liver cells and constitutes the end of a dense network of blood vessels.
+This particular internal structure has a strong influence on the mechanical properties of the liver.
+In contrast, the colon is a long and hollow organ.
+It consists of four sections which all have different shapes and material properties.
+Consequently, it seems unrealistic to use a unique model for all tissues.
+Yet, most of previous works focus on volumetric models that are able to capture the behaviour of solid organs.
+In fact in the field of medical simulation, very few models have been proposed for simulating, in real-time, the deformation of thin anatomical structures whose volume is negligible compared to their surface area.
+Examples include hollow structures, such as the wall of blood vessels, or membranes, such as the Glisson?s capsule surrounding the liver.
+It is also of particular interest to us for modelling the colon in our colonoscopy simulator and for simulating the deployment of the implant in cataract surgery.
 
-\section{Non-physic approaches (computer graphics stuff)}
+Numerous models are available in the literature to describe physics of thin objects, from fairly simple and naive approaches to more complex and thorough representations.
+Continuum mechanics provides many formulations able to accurately describe stresses occurring within thin objects.
+Most of them fall into one of the following two categories: plate theory or shell theory.
+Those theories have been a subject of interest in the mechanical community for decades.
+The difference between these two kinds of structures is very well explained by \cite{Liu03} and can be summarised by the fact that plate bending elements can only carry transversal loads while shells can undergo more complex deformations.
+For instance, if we consider the horizontal board of a bookshelf, that board can be approximated as a plate structure and the transversal loads are the weight of the books.
+A typical deformation of the board is illustrated in \fig{chap7:fig-boards} (a).
+Conversely, a shell structure can carry loads in all directions, and therefore can undergo bending, twisting and in-plane deformation (see \fig{chap7:fig-boards} (b)).
+%
+\begin{figure}[ht]
+\centering 
+\subfloat[Plate theory]{\includegraphics[width=6.5cm]{chapter7/board_bending.pdf}}
+\hfill 
+\subfloat[Shell theory]{\includegraphics[width=6.5cm]{chapter7/board_twist.pdf}}
+\caption[Illustration of the key difference between plate and shell theory]{Illustration of a key difference between various models of thin objects.
+While thin plate theory allows to describe bending (a), it cannot represent more complex deformations such as twist (b) which is captured by shell theory.}
+\label{chap7:fig-boards}
+\end{figure}
 
-\section{Physically accurate approches (plates/shells)}
+Development of a satisfactory physical model that runs in real-time but produces visually convincing animation of thin objects has been a challenge in Computer Graphics, particularly in the area of cloth modelling.
+Rather than resorting to shell theory which involves the most complex formulations in continuum mechanics, most of works have relied on discrete formulations.
+We will first discuss early approaches based on mass-spring models.
+Then we will present techniques relying on the derivation of a bending energy from geometric considerations.
+At last, we will introduce approaches based on shell theory using the more computationaly demanding finite element method.
+
+%	\subsection{Colonoscopy simulator project: needs for colon}
+%	\subsection{Cataract surgery, stenosis: other needs for implants/blood vessels}
+
+\section{Mass-spring models}
+
+Early approaches to thin objects modelling only considered in-plane deformation, and often relied on mass-spring models.
+This technique was already presented in section \ref{chap4:massSprings} so we only present their use for modelling thin structures.
+\cite{Provot95} improved a mass-spring model to take into account the non-elastic properties of woven fabrics.
+Because of the high stiffness of textiles, mass-spring models are particulary unstable when used to simulate cloth.
+The author proposed a new method inspired from dynamic inverse procedure.
+Under certain conditions and because of the linear law employed in a mass-spring system, the fabrics may become super-elastic.
+\citeauthor{Provot95} suggested to set a threshold on the deformation rate to correct this super-elasticity (see \fig{chap7:fig-sheet}).
+%
+\begin{figure}[ht]
+\centering 
+\subfloat[Initial position]{\includegraphics[width=4.5cm]{chapter7/sheet1.pdf}}
+\hfill 
+\subfloat[Uncorrected]{\includegraphics[width=4.5cm]{chapter7/sheet2.pdf}}
+\hfill 
+\subfloat[Corrected]{\includegraphics[width=4.5cm]{chapter7/sheet3.pdf}}
+\caption{Deformation of a sheet hanging by two adjacent corners. Images courtesy of \cite{Provot95}.}
+\label{chap7:fig-sheet}
+\end{figure}
+
+To improve the computational efficiency, \cite{Hutchinson96} presented a mechanism for adaptively refining the network of mass-springs to concentrate efforts only where it is needed.
+When potential inaccuracies are detected, the object is locally refined in the affected region, and the simulation may carry on.
+\cite{Oshita01} used a sparse triangular mesh and an interpolation to generate a dense mesh.
+
+A limitation of mass-spring models that we often hear is the difficulty to derive spring stiffness from elastic properties (Young's modulus and Poisson's ratio).
+With this drawback in mind, \cite{Volino97} improved a mass-spring system by allowing the modelisation of Young's modulus and Poisson's ratio (see \fig{chap7:fig-fabricSquares}).
+They also added a correction technique relying on the control of damping and velocities to ensure the stability of their model.
+%
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=12cm]{chapter7/fabricSquares.pdf}
+\caption[A $400\,$\% stretched fabric square]{A $400\,$\% stretched fabric square with various Poisson's ratio.}
+\label{chap7:fig-fabricSquares}
+\end{center}
+\end{figure}
+
+Other works have considered adding bending through angular springs.
+For instance, \cite{Taskiran05} have chosen to use a linear representation of angular springs to supply bending and torsion effects in hair modelling, which allowed them to simulate curly hair with good realism.
+All the forces related to springs were implemented: gravity, repulsions from collision, absorption, friction, etc.
+They could simulate up to a few hundreds of hair stripes in real-time.
+\cite{Wang07} successfully used a network of linear and angular springs to describe bending and twisting of catheters and guidewires in an interventional radiology simulator.
+
+More recently and in the field of medical simulation, \cite{Hammer08} employed a mass-spring model to simulate the mitral valve closure for surgical planning.
+In their model, triangle sides are treated as linear springs and sides shared by two triangles are treated as bending springs.
+
+\section{Techniques relying on the derivation of a bending energy}
+
+Another kind of approach is to geometrically derive a bending energy.
+For instance, \cite{Baraff98} derived stretch, bending and shear energies geometrically.
+They also used an adaptative time stepping based on detection of large stretches.
+Using an implicit integration allows the authors to use large time steps and substantially increase the stiffness with a limited impact on performance.
+
+%Among the different models introduced recently we can mention the work of \cite{Choi07} and \cite{Bridson03}.
+\citeauthor{Choi07}  proposed a real-time simulation technique for thin shells undergoing large deformations.
+The authors adopt the energy functions from the discrete shells proposed by \cite{Grinspun03}.
+For real-time integration of the governing equation, they adapted a modal analysis technique, called modal warping.
+The resulting simulations run in real-time even for large meshes, and the model can handle large bending and/or twisting deformations with acceptable realism.
+\citeauthor{Bridson03} followed the same energy approach to derive their bending model but improved the resolution of the equations by suggesting a novel mixed implicit/explicit integration scheme.
+They also presented a post-processing method for treating cloth-character collisions that preserves folds and wrinkles.
+\cite{Pabst08} improved the bending modelling used by \citeauthor{Bridson03} to allow the integration of measured material data.
+%
+%A method to use thin shell dynamics with point sampled surfaces for efficient animation was recently proposed by \cite{Wicke05} where the curvature of the shell is measured through the use of fibers.
+
+\section{Techniques based on continuum mechanics}
+
+% Co-rotational methods
+% More recently, \cite{Allard09} modelled the membrane of the lens capsule with a co-rotational finite element formulation in a cataract surgery simulator.
+The goal was to simulate the anisotropic fracture progation

### 2010-09-09 Thesis/Chapter4.tex
- Update 9 September -
End of chapter 7



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@171 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 If this is often an acceptable trade-off between the visual deformation result and the speed of the system, research was carried out to handle large deformations.
-However, since geometrically non-linear analyses are highly computationally demanding, another approach was found which allows large deformation while still relying on a linear strain measure: the \emph{co-rotational} methods.
+However, since geometrically non-linear analyses are highly computationally demanding, another approach was found which allows large deformation while still relying on a linear strain measure: the \emph{co-rotational} methods \label{chap4:corotationalMethods}.
 A very complete description of the co-rotational approach is given by \cite{Felippa00}.

### 2010-09-09 Thesis/Chapter7.tex
- Update 9 September -
End of chapter 7



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@171 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-Numerous models are available in the literature to describe physics of thin objects, from fairly simple and naive approaches to more complex and thorough representations.
+The first idea that may come to mind for modelling thin structures is to apply the same methods than with solid objects.
+For instance, \cite{Holzapfel02} proposed a very complex 3-layer model of an aterial wall for simulating an angioplasty procedure.
+Their model is non-linear and anisotropic along fibers placed according to histology.
+They also model diseased part of the vessel through a hardening parameter which adds plasticity.
+However, they discretised their geometric model with eight-node brick elements.
+Another example is given by \cite{Aloisio04} who assume that the blood flow in the vessels may be considered as incompressible and the artery consequently becomes a solid body and can be modelled as a solid deformable object.
+\TODO{why is this not appropriate?
+High-order elements could do the trick to follow the curvature...} Hence, specific models for thin objects were developed.
+
+Indeed, numerous models are available in the literature to describe physics of thin objects, from fairly simple and naive approaches to more complex and thorough representations.
 Continuum mechanics provides many formulations able to accurately describe stresses occurring within thin objects.
@@
 A typical deformation of the board is illustrated in \fig{chap7:fig-boards} (a).
-Conversely, a shell structure can carry loads in all directions, and therefore can undergo bending, twisting and in-plane deformation (see \fig{chap7:fig-boards} (b)).
+Conversely, a shell structure can carry loads in all directions, and therefore can undergo bending, twisting and membrane (that is, in-plane) deformation (see \fig{chap7:fig-boards} (b)).
 %
@@
 \includegraphics[width=12cm]{chapter7/fabricSquares.pdf}
-\caption[A $400\,$\% stretched fabric square]{A $400\,$\% stretched fabric square with various Poisson's ratio.}
+\caption[A $400\,$\% stretched fabric square]{A $400\,$\% stretched fabric square with various Poisson's ratio. Image courtesy of \cite{Volino97}.}
 \label{chap7:fig-fabricSquares}
@@
 Another kind of approach is to geometrically derive a bending energy.
+Theses approaches are more computationally demanding than mass-spring models but allows for higher flexibility and accuracy.
 For instance, \cite{Baraff98} derived stretch, bending and shear energies geometrically.
@@
 
-%Among the different models introduced recently we can mention the work of \cite{Choi07} and \cite{Bridson03}.
+Among the geometry-based models, there is the noticeable work of \cite{Grinspun03} who introduced a discrete shell model.
+They designed a simple shell model for triangle meshes by geometrically deriving membrane and bending energies.
+They applied geometric operators over piecewise-linear surfaces to define a discrete constitutive model, which results in simpler expressions than tensorial derivations from shell theory.
+They applied their method to simulate the deformation of common objects like beams, sheets of paper or hats (see \fig{chap7:fig-hat}).
+%
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=8cm]{chapter7/hat.pdf}
+\caption[Comparison between real footage and simulation of a hat dropped on a table]{Comparison between real footage and simulation of a hat dropped on a table. Image courtesy of \cite{Grinspun03}.}
+\label{chap7:fig-hat}
+\end{center}
+\end{figure}
+
+Among the different models introduced recently we can also mention the work of \cite{Choi07} and \cite{Bridson03}.
 \citeauthor{Choi07}  proposed a real-time simulation technique for thin shells undergoing large deformations.
@@
 \citeauthor{Bridson03} followed the same energy approach to derive their bending model but improved the resolution of the equations by suggesting a novel mixed implicit/explicit integration scheme.
+In particular, the explicit update for positions allows the authors to modify velocities to enforce constraints and allows a strain limiting procedure which they used to create folds and wrinkles in their cloth simulation.
 They also presented a post-processing method for treating cloth-character collisions that preserves folds and wrinkles.
-\cite{Pabst08} improved the bending modelling used by \citeauthor{Bridson03} to allow the integration of measured material data.
+\cite{Pabst08} later improved the bending modelling used by \citeauthor{Bridson03} to allow the integration of measured material data.
+
+\cite{Volino06} suggested an alternative somewhere between mass-spring and geometrically derived approaches.
+They compute a bending vector that represents the bending of the surface through  a simple linear combination of particule positions and they distribute this vector as particule forces according to the bending stiffness.
+The authors claimed much better accuracy than mass-spring models (particulary when dealing with low curvature and high bending stiffness) with similar computation time (see \fig{chap7:fig-dress} for an illustration).
 %
-%A method to use thin shell dynamics with point sampled surfaces for efficient animation was recently proposed by \cite{Wicke05} where the curvature of the shell is measured through the use of fibers.
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=10cm]{chapter7/dress.pdf}
+\caption[Simulation of complete garments that have large bending stiffness]{The model of \citeauthor{Volino06} allows the simulation of complete garments with material that may have fairly large bending stiffness.
+Image courtesy of \cite{Volino06}.}
+\label{chap7:fig-dress}
+\end{center}
+\end{figure}
+
+A method to use thin shell dynamics with point sampled surfaces for efficient animation was recently proposed by \cite{Wicke05} where the curvature of the shell is measured through the use of fibers.
+The fibers are used to approximate the differential surface operators.
+Their method supports both elastic and plastic deformations as well as fracturing and tearing of the material.
 
@@
 
-% Co-rotational methods
-% More recently, \cite{Allard09} modelled the membrane of the lens capsule with a co-rotational finite element formulation in a cataract surgery simulator.
-The goal was to simulate the anisotropic fracture progation
+Following a similar train of thoughts than with solid objects, methods based on continuum mechanics provide the most accurate solutions but are computationally very demanding.
+The mathematical framework of plate and shell theories is also fairly complex and deriving finite element methods from these theories is challenging.
+Their implementation is also far from being straightforward.
+Nevertheless, we will now review a few attempts of applying shell theory to simulation.
+
+In 1991, \citeauthor{Black91} presented a materially non-linear shell finite element model to simulate the leaflets of a bicuspid bioprosthetic heart valve.
+The results indicate that modelling bending has a strong influence on the accuracy of the deformation.
+They concluded that models where only membrane stresses are taken into account were likely to produce significant errors in the stress distribution.
+
+Shell models are fairly common in the field of cloth modelling.
+For instance, \cite{Eischen96} applied fully non-linear shell theory for modelling fabrics.
+The interested reader may also refer to their fairly complete review on the use of shell finite element methods in cloth modelling.
+However, their attention was towards the accuracy provided by shell theory and were not really interested in fast solution times.
+
+Shell finite element modelling followed the same trend than solid objects and co-rotational formulations were introduced to simplify strain expressions while allowing large rotations (but small strain) (see page \pageref{chap4:corotationalMethods} for more details).
+For instance, \cite{Etzmuss03} used the linear Cauchy strain tensor but applied a polar decomposition on the deformation gradient to allow large deformations.
+The bending energy is separated from the membrane formulation and most of material parameters were measured from experiments on actual pieces of clothing.
+As an example, the computation of a shirt consisting of $ 8\,300 $ triangles took an average of less than $ 3\,$s per animation frame with a time step of $ 0.02\,$s (see an illustration \fig{chap7:fig-shirt}).
+While still being far from achieving real-time, the intention of the authors was clearly to propose an efficiency model for cloth simulation.
+%
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=12cm]{chapter7/shirt.pdf}
+\caption[A shirt shown with different material properties]{A shirt shown with different material properties: (a) wool/viscose, (b) wool and (c) polyester/polyacrylics/acetate.
+Image courtesy of \cite{Etzmuss03}.}
+\label{chap7:fig-shirt}
+\end{center}
+\end{figure}
+
+\cite{Thomaszewski06} based their model on the Kirchhoff-Love shell theory and built a subdivision-based finite element formulation similar to \cite{Cirak00}.
+They also employed a co-rotational framework to simulate wrinkles and folds of textiles.
+Although the rotation field is computed for each vertex, they found that using the rotation obtained for the barycentre of all vertices involved was sufficient.
+More recently, \cite{Allard09} also used a co-rotational finite element formulation to model the membrane of the lens capsule in a cataract surgery simulator.
+The goal was to simulate the anisotropic fracture progation during capsulorhexis, the technique used to create a circular opening in the lens capsule.
+The formulation did not feature bending however.
+%
+\begin{figure}[ht]
+\centering 
+\subfloat[Under gravity]{\includegraphics[width=4.5cm]{chapter7/sleeve1.pdf}}
+\hfill 
+\subfloat[Torsion]{\includegraphics[width=4.5cm]{chapter7/sleeve2.pdf}}
+\hfill 
+\subfloat[Compression]{\includegraphics[width=4.5cm]{chapter7/sleeve3.pdf}}
+\caption[Different types of folds on a garment's sleeve ]{Different types of folds on a garment's sleeve. Images courtesy of \cite{Thomaszewski06}.}
+\label{chap7:fig-sheet}
+\end{figure}
+
+Regarding medical applications, some attempts have been made to apply complex shell finite element model to simulate anatomical structures.
+For instance, \cite{Lim05} used an asymmetrical shell element model to simulate the deformation of the mitral valve.
+The computation of the finite element model was carried out with the commercial software ANSYS.
+\cite{Conti10} modelled the non-linear and anisotropic mechanical response of the aortic leaflets during the entire cardiac cycle using a shell finite element model in Abaqus FEA.
+However, these models are not designed for interactive simulation and may need up to dozens of hours to be computed.
+
+\section{Objectives}
+
+We seek to propose a solution for simulating, in real-time, the deformation of thin anatomical structures.
+Since \cite{Black91} showed that formulations with only membrane energy were likely to produce significant errors in the stress distribution, this model is required to feature bending.
+Ideally, this solution would be versatile and allow us to simulate thin objects with various shapes and material properties with good accuracy.
+In this regard, methods derived from continuum mechanics are the most appropriate.
+In some situations, a thin plate model would not correctly capture the deformation of some objects and therefore a shell formulation was retained for more flexibility.
+To the best of our knowledge, no real-time shell finite element method was presented in the field of medical simulation.
+\TODO{need to triple-check this...}
+
+In the next section, we will first introduce our co-rotational triangular shell finite element model and show how to efficiently apply this model to simulate the implant deployment in cataract surgery.
+In a second time, we will present a technique to mesh complex anatomical shapes with curved triangles, hence allowing us to model, in real-time, any anatomical structures with our co-rotational shell FEM in an optimal way.
+
+% The key idea is to model the physical shell as a surface but endowed with mechanical properties in the form of elastic resistance to stretching and bending forces.

### 2010-09-10 Thesis/Chapter10.tex
- Update 10 September -
Huge update with chapter9 and 10 added + substantial update of bib



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@172 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\chapter{'Shell meshing' technique \OC{MICCAI 2010}}
+\chapter{Physics-based reconstruction using shell elements}
 \label{chap10}
 \begin{shortAbstract}
-A short abstract for the upcoming chapter
+A short abstract for the upcoming chapter \\
+\OC{MICCAI 2010}
 \end{shortAbstract}
 
-\section{State of art: reconstruction/simplification}
+\section{Techniques of mesh simplification}
+
+\subsection{Introduction: our motivation}
+
+Our idea is the following: while many flat triangles are required to describe highly curved surfaces, fewer triangular shell elements are needed to describe the given geometry with the same precision since they can be curved.
+In the following, we assume that we have a high resolution triangular mesh obtained from a binary segmented image of the organ we want to simulate (via a Marching Cube algorithm for instance, \cite{Lorensen87}).
+Our goal is to create a mesh featuring the optimal number of shell elements while staying as close as possible to our targeted geometry.
+For this reason, we need a technique to simplify the high resolution mesh of the organ to be simulated.
+
+Simplification strategies may be grouped into two categories: local strategies that iteratively simplify the mesh by the repeated application of some local operator, and global strategies that are applied to the input mesh as a whole \citep{Cignoni97,Talton04}.
+
+\subsection{Local simplification strategies}
+
+Typically, they define some mesh operation $S$ that, when applied to a mesh $M$, acts on a small collection of its elements and produces a new mesh $\overline{M}$  with fewer elements.
+By repeated application of $S$, a mesh may be simplified arbitrarily.
+In order to determine the mesh elements to which $S$ should be applied on a given iteration, $S$ may be associated with an error function (or cost function) that measures the amount of error the operation will introduce into the approximation.
+By computing the error associated with every possible application of $S$ at a particular iteration, the algorithm can apply the one with minimal cost.
+This type of heuristic is quite reasonable for simplification problems, and in practice these methods work well.
+
+One nice property of local iterative algorithms is that they allow the user to specify the desired attributes of the target approximating mesh with a high degree of precision.
+For example, the user may allow the algorithm to run until the mesh contains $k$ faces, or until the error at a given vertex exceeds some threshold $\varepsilon$.
+
+\subsubsection*{Vertex decimation.}
+The decimation algorithm, first proposed by \cite{Schroeder92}, is simple.
+Multiple passes are made over all vertices in the mesh.
+During a pass, each vertex is a candidate for removal and, if it meets the specified decimation criteria, the vertex and all triangles that use the vertex are deleted.
+The resulting hole in the mesh is patched by forming a local triangulation.
+The vertex removal process repeats, with possible adjustment of the decimation criteria, until some termination condition is met.
+
+The error associated with a particular decimation typically depends on the classification of the vertex being decimated.
+For the general case of a manifold vertex $\lbrace i \rbrace$ not near a boundary, \citeauthor{Schroeder92} consider the set of planes formed by the adjacent triangles and compute a single approximating plane based on the area-weighted average of the triangle normals $\textbf{n}_k$, centers $x_k$, and areas $A_k$:
+%
+\begin{equation}
+\textbf{n} = \frac{\sum_k A_k\textbf{n}_k}{\sum_k A_k},  \hspace*{1cm} \widehat{\textbf{n}} = \frac{\textbf{n}}{\vert \textbf{n} \vert}, \hspace*{1cm}  x = \frac{\sum_k A_k\textbf{x}_k}{\sum_k A_k}
+\end{equation}
+
+Then they calculate the distance from vertex $\lbrace i \rbrace$ to this plane by  $\vert \widehat{\textbf{n}} \cdot (v_i - x) \vert$ and define this quantity to be the cost associated with the decimation.
+This metric ensures that vertices in smooth regions will be decimated before vertices that define sharp features.
+
+In practice, simplification schemes based on vertex decimation excel at eliminating un-necessary geometry, such as is typically found in meshes generated by the marching cubes algorithm where large planar regions may be subdivided into many redundant triangles.
+
+\subsubsection*{Edge contraction.}
+Edge contraction, originally proposed by \cite{Hoppe93}, is the most common simplification operation.
+An edge contraction operates on a single edge $\lbrace v_1,v_2 \rbrace$ and contracts that edge to a single vertex $\lbrace w \rbrace$, updating all edges previously incident on $\lbrace v_1 \rbrace$ and $\lbrace v_2 \rbrace$ to reference $\lbrace w \rbrace$.
+On a manifold mesh without boundary, each edge contraction will collapse precisely two faces (see \fig{chap10:fig-edgeCollapse}).
+Edge contractions can alter the topology of a mesh, since repeatedly contracting all the edges around a hole will eventually close it.
+%
+\begin{figure}
+\centering
+\includegraphics[height=3cm]{chapter10/edgeCollapse.pdf}
+\caption[The edge contraction operation]{The edge contraction operation. Image courtesy of \cite{Hoppe93}.}
+\label{chap10:fig-edgeCollapse}
+\end{figure}
+
+For a given contraction $\lbrace v_1,v_2 \rbrace \rightarrow \lbrace w \rbrace$, it is not immediately clear what value should be assigned to $v_h$, i.e.
+where the resulting vertex should be placed.
+Obvious choices such as $v_h = v_i$, $v_h = v_j$, or $v_h = (v_i + v_j)/2$ are convenient, but can easily be shown to be non-optimal.
+Rather than placing $v_h$ arbitrarily, it is better to consider the error function associated with the contraction operation and attempt to minimise its value over the space of possible vertex placements.
+Once the optimal target position has been computed for every possible contraction, the contraction with the smallest associated error can be selected and applied.
+
+\subsubsection*{Adaptative subdivision.}
+Coplanar or nearly coplanar faces are searched for in the mesh, merged into larger polygons, and then retriangulated into fewer faces than those originally required \citep{Dehaemer91,Hinker93}.
+Face merging is driven by a co-planarity test.
+
+The basic adaptive subdivision method as presented by \cite{Schmitt86} is sketched in \fig{chap10:fig-adaptativeSubdivision}.
+(a) shows several data points produced by the sampling process and sketch (b) adds a trial polygon that approximates the surface represented by the sample points.
+A bilinear interpolation is performed over the surface of the approximating polygon to locate the point on the surface 'directly below' a sample data point.
+The distance (or error) between the surface and the data point is then calculated, as shown in (c), and is compared to the user specified tolerance.
+If any of the errors exceeds the tolerance value then the polygon is divided into four smaller polygons, as shown in (d), and the process is repeated recursively.
+%
+\begin{figure}
+\centering
+\includegraphics[width=13.5cm]{chapter10/adaptativeSubdivision.pdf}
+\caption[Surface-fitting polygons with adaptive subdivision]{Surface-fitting polygons with adaptive subdivision. Image courtesy of \cite{Schmitt86}.}
+\label{chap10:fig-adaptativeSubdivision}
+\end{figure}
+
+When the approximation to the data is within the specified tolerance, the recursion is terminated and the polygon is saved in a linked list.
+Upon completion of the subdivision algorithm, the resulting polygons are displayed for the user to judge the results.
+Generally, the user runs the program repeatedly while varying the maximum tolerance until the results are suitable.
+
+\subsubsection*{Polygon growth.}
+Another method proposed by \cite{Dehaemer91} to systematically generate surface-fitting polygons is to use a polygon growth technique (see \fig{chap10:fig-polygonGrowth}).
+First a seed polygon is selected from the original set of data, as in (a).
+Then a neighbouring polygon is selected and the two are combined into a larger trial polygon ((b) and (c)).
+If the trial polygon passes the accuracy metric, 10 additional neighbors are selected in an attempt at further growth, as in (d) through (g).
+When all attempts at further growth fail, the polygon is added to the list of polygons describing the reduced object, as shown in (h).
+A new seed polygon is selected and the process repeated.
+When there are no more polygons from which to choose a new seed, the process is terminated.
+%
+\begin{figure}
+\centering
+\includegraphics[width=12cm]{chapter10/polygonGrowth.pdf}
+\caption[Polygon growth]{Polygon growth. Image courtesy of \cite{Dehaemer91}.}
+\label{chap10:fig-polygonGrowth}
+\end{figure}
+
+This method of fitting polygons to the sampled data is not as useful as the adaptive subdivision methods, for several reasons.
+First, it is an essentially brute force method of surface fitting and very repetitive in nature.
+Many trial polygons are created, only to be 'thrown out' because they exceed the error tolerance.
+This causes the procedure to execute quite slowly (orders of magnitude more slowly) than the adaptive subdivision method.
+Second, the data is not significantly reduced.
+Many of the generated polygons are of complicated shapes and use a large number of vertices.
+Thus many of the original sample data points are retained.
+
+\subsubsection*{Appearence preserving simplification.}
+A completely different approach to determining the error associated with a given simplification operation was described by \cite{Cohen98}.
+Their metric, which is appearance-based, measures the amount of deviation caused by the operation in the screen-space (pixelised) representation of the mesh.
+Their algorithm functions by decoupling surface position from color and curvature information and storing the latter two quantities in texture and normal maps.
+A traditional geometric simplification algorithm can then be employed to filter the surface position, while a hardware-based approach is used to filter the color and normal information, resulting in simplified representations that are nearly visually indistinguishable from the original.
+However, the algorithm requires a parameterisation of the mesh in order to function.
+
+\subsection{Global simplification strategies}
+
+\subsubsection*{Vertex clustering.}
+The method of vertex clustering was originally proposed by \cite{Rossignac93} to handle meshes of arbitrary topological structure.
+In their algorithm, each vertex in the input mesh is assigned a weight based on its perceptual importance: vertices adjacent to triangles with large faces and those in areas of high curvature are weighted more heavily than vertices in smooth regions adjacent to smaller triangles.
+Next, a bounding box is placed around the mesh and subdivided into a three-dimensional grid.
+Finally, all the vertices in a given grid cell are clustered to the position of the vertex with maximum weight.
+Degenerate geometry may then be removed, although the authors proposed a novel visualisation of certain degenerate faces and edges in an effort to enhance the recognizability of their simplified meshes.
+The algorithm is extremely efficient and simple to implement, and the level of simplification may be controlled (although with some difficulty) by choosing the resolution of the overlaid grid.
+However, vertex clustering can drastically alter the topology of the input mesh in an unpredictable manner, and in practice does not produce very faithful geometric approximations at low levels of detail.
+
+\subsubsection*{Re-tiling.}
+The re-tiling method described by \cite{Turk92} begins with a polygonal surface and creates a triangulation of this surface with a user-specified number of vertices.
+The points are placed uniformly over a given polygonal surface by distributing points at random over the surface and then having each point repel all of its neighbors.
+An alternative to placing the points randomly is using the variation in surface curvature.
+The features of the surface would be more accurately reflected in a re-tiling by increasing the density of vertices in regions of high curvature.
+
+The key notion in the re-tiling procedure is the creation of an intermediate model called the mutual tessellation of a surface that contains both the vertices from the original model and the new points that are to become vertices in the re-tiled surface.
+The new model is then created by removing each original vertex and locally re-triangulating the surface in a way that matches the local connectedness of the initial surface (see \fig{chap10:fig-retiling}).
+
+If this method produces nice meshes, an issue is finding measures of how closely matched a given re-tiling is to the original model. 
+%
+\begin{figure}
+\centering
+\includegraphics[width=12cm]{chapter10/retiling.pdf}
+\caption[Retiling technique]{Upper left: original surface.
+Upper right: candidate vertices after point-repulsion.
+Lower left: mutual tesselation.
+Lower right: final tesselation.
+Image courtesy of \cite{Turk92}.}
+\label{chap10:fig-retiling}
+\end{figure}
 
 \section{Our method}
+
+	\subsection{A simple algorithm}
+
+However, we need to keep in mind that our situation is substantially different than all previously mentioned geometrical approaches as we want to model the deformation of the structure.
+In that regard, the curvature of the surface has a physical meaning: it represents the mid-surface of the shell.
+We propose to approximate the surface of anatomical structures with shell elements whose each surface is described by the shape function used in our shell formulation.
+
+The first step in the process of generating a shell-based mesh is an important decimation of the high resolution mesh, using quadric edge collapse technique implemented in Meshlab \cite{Meshlab}.
+The algorithm tries as much as possible to preserve mesh boundaries and generates high quality triangular elements.
+We then apply a heuristic method derived from the work of \cite{Saupin07} with tetrahedral meshes based on simple geometrical rules.
+For each node of the coarse mesh, we find the three closest triangles on the high resolution mesh and we move the node to the barycentre of the three centres of mass of those triangles.
+This technique locally smoothes the surface of the mesh while converging towards the desired high resolution mesh.
+
+At each iteration of this algorithm, we want to measure the error between the curved surface of shells and the target.
+Indeed, we need to ensure that the distance between the surface of our shell-based mesh and the targeted high resolution mesh will be minimal.
+An efficient technique for measuring the error between two surfaces is the Hausdorff distance \citep{Klein96}.
+As a reminder the Hausdorff distance between two meshes is the maximum between the two so-called one-sided Hausdorff distances:
+\begin{equation}
+d_{\mathrm{H}}(X,Y) = \max \left\{ \sup_{x \in X} \inf_{y \in Y} d(x,y), \sup_{y \in Y} \inf_{x \in X} d(x,y) \right\},
+\end{equation}
+where $d()$ is the Euclidian distance between two points.
+The same technique of subdivision used for rendering allows us to sample the actual surface described by the shells to compute the Hausdorff distance with the targeted high resolution mesh.
+
+By using the Hausdorff distance between two iterations, the process may be stopped when the required precision has been reached.
+A simple example is shown \fig{chap10:fig-cylinder} to illustrate the method.
+\begin{figure}
+\centering
+\includegraphics[width=13.5cm]{chapter10/exampleCylinder.pdf}
+\caption {The target (a) is a high resolution cylinder mesh of 16,384 triangles and we start from a very coarse mesh (12 triangles), rendered with flat triangles here (b).
+In (c) the coarse mesh is rendered with shells and a one-sided Hausdorff distance colour map is applied to show the initial error with the high resolution mesh.
+(d) One-sided Hausdorff distance colour map after one iteration of our algorithm (48 shells).}
+\label{chap10:fig-cylinder}
+\end{figure}
+
+	\subsection{Results on complex anatomical structures}
+
+\subsubsection*{Meshing of anatomical structures.}
+This approach has been applied to approximate more complex anatomical geometries with curved shell elements.
+Two examples are given with the Glisson's capsule, which is the membrane surrounding the liver (\fig{chap10:fig-liver}) and an aneurysm (\fig{chap10:fig-aneurysm}).
+In each case, the error is computed through a Hausdorff distance and expressed as a percentage of the diagonal of the object's bounding box.
+%
+\begin{figure}
+\centering
+\includegraphics[width=13.5cm]{chapter10/resultsLiver.pdf}
+\caption {(a) the targeteted high resolution Glisson's capsule mesh ($8\,000$ triangles).
+(b) the one-sided Hausdorff distance error map after applying only one iteration of our algorithm to the coarse mesh ($1\,200$ shells).}
+\label{chap10:fig-liver}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=13.5cm]{chapter10/resultsAneurysm.pdf}
+\caption {(a) the targeteted high resolution aneurysm mesh ($28\,368$ triangles). (b): the one-sided Hausdorff distance error map on a mesh of 772 shells generated with our method.}
+\label{chap10:fig-aneurysm}
+\end{figure}
+
+\subsubsection*{Computation times.}
+We perform several tests on the aneurysm model at different resolutions to measure computation times.
+The shells are resisting to a uniform pressure load and solved using a Conjugate Gradient (CG) iterative solver.
+The computation times are reported in \fig{chap10:fig-computation}.
+Implicit integration allows for large time steps ($40\,ms$) and the computation is real-time for 800 shell elements and a reasonable error criterion ($5\,\%$).
+When the computation time must be bounded (critical real-time applications), one can fix the number of CG iterations to, for instance, $100$ and remains real-time for $1\,000$ shell elements.
+However, in that case the accuracy of the results is not checked.
+%
+\begin{figure}
+\centering
+\includegraphics[width=13.5cm]{chapter10/computation_times.pdf}
+\caption {Computation time on meshes of 200, 400, 600, 800 and $ 1\,000 $ elements.}.
+\label{chap10:fig-computation}
+\end{figure}
+
+\subsection{Coupling between tetrahedra and shells for advanced modelling}
+Structures in human body can be either solid (brain, liver, prostate etc.) or hollow (colon, blood vessels, stomach etc.).
+However, knowing how to model the two kind of structures is not sufficient to reach a high degree of accuracy, real life situations are more complex.
+As an example, the external surface of the liver is covered by a layer of cells called Glisson's capsule.
+Its interaction with the liver plays an important role into the overall structure's mechanical behaviour.
+Therefore considering the interaction between solid and hollow objects is as crucial as modelling the two structures separately.
+
+An example of medical procedure to illustrate this point even further is angioplasty.
+Angioplasty is the technique of mechanically widening a narrowed or obstructed blood vessel, typically as a result of atherosclerosis.
+An empty and collapsed balloon on a guide wire is passed into the narrowed locations and then inflated to a fixed size.
+The balloon crushes the fatty deposits, so opening up the blood vessel to improved flow.
+As a proof of concept, we tried to simulate an angioplasty.
+Our preliminarty results are presented on \fig{chap10:fig-stent}.
+The blood vessel is modelled using the shell FEM formulation described in this paper and the fatty deposits are simulated with a tetrahedral FEM method and are fixed to the interior wall of the blood vessel.
+When the balloon inflates it crushes the deposits and they then apply a pressure onto the curved surfaces of shells modelling the interior wall.
+The forces are then distributed onto the mechanical nodes of the blood vessel mesh as detailed in section~\ref{chap9:interactions}, which widens the blood vessel as expected.
+%
+\begin{figure}
+\centering
+\includegraphics[width=12cm]{chapter10/stenose_final.pdf}
+\caption {Simulation of an angioplasty procedure.
+(a, c): A collapsed stent is inserted into the blood vessel simulated with our shell FEM formulation.
+The fatty deposits (in yellow) are modelled with tetrahedral FEM.
+(b, d): The stent is crushing the fatty deposits which creates a pressure onto the interior wall and widens the blood vessel.}
+\label{chap10:fig-stent}
+\end{figure}
+
+\section{Discussion}
+
+We proposed a framework for real-time modelling of thin anatomical structures.
+The novelty of our method relies on the combination of a shell finite element formulation and a geometric surface reconstruction both based on the same polynomial interpolation function used to describe the surface of shells.
+We also show how contacts and interactions with the curved surfaces of shells can be handled using the same function.
+In summary, these polynomial shape functions are used in three different ways in our computational model:
+%
+\begin{description}
+\item[(a)] to approximate complex geometrical shapes
+\item[(b)] to compute internal forces via a co-rotational shell FEM
+\item[(c)] to compute contact forces onto a curved triangle
+\end{description}
+
+The efficiency of the method is illustrated through shell-based reconstruction and real-time simulation of the deformations of various anatomical structures.
+We also present preliminary results on the simulation of an angioplasty procedure.
+
+\bigskip
+
+\TODO{Did we answer our problematic? Limitations? }

### 2010-09-10 Thesis/Chapter9.tex
- Update 10 September -
Huge update with chapter9 and 10 added + substantial update of bib



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@172 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\chapter{A co-rotational triangular shell model \OC{ISBMS 2010}}
+\chapter{A co-rotational triangular shell FEM model}
 \label{chap9}
 \begin{shortAbstract}
-A short abstract for the upcoming chapter
+A short abstract for the upcoming chapter \\
+\OC{ISBMS 2010}
 \end{shortAbstract}
@@
 
+We propose to define a triangular shell element by combining a two-dimensional in-plane membrane energy, with an off-plane energy for describing bending and twist.
+To allow for real-time simulation, a computationally efficient formulation is needed.
+We therefore propose to extend the co-rotational idea introduced by \cite{Felippa00}.
+Indeed, as we have seen in the previous chapters, co-rotational approaches have been successfully applied to real-time simulation over the last few years.
+They offer a good trade-off between computational efficiency and accuracy by allowing small deformations but large displacements.
+We propose to improve and extend a plate model first introduced by \cite{Przemieniecki68} to a co-rotational formulation.
+Once combined with an in-plane membrane formulation we obtain an accurate, yet computationally efficient, shell finite element method featuring both membrane and bending energies.
+
+\subsection{Triangular elastic membrane}
+
+The computation of the triangular elastic membrane stiffness matrix can be derived from previous works dealing with tetrahedral co-rotational elements \citep[like][for instance]{Muller04}.
+The element stiffness matrix $\textbf{K}_e$ can be computed as follows:
+%
+\begin{equation}
+\textbf{K}_e = \int_v \textbf{J} \boldsymbol\chi \textbf{J}^{T} dV,
+\end{equation}
+where $\textbf{J}$ is the strain-displacement matrix and $\boldsymbol\chi$ embodies the material's behaviour.
+The implant is very stiff and we therefore assume that the local deformations remain limited during the deployment and a linear constitutive law is sufficient.
+Thus in the simple case of Hooke's law we have:
+%
+\begin{equation}
+\boldsymbol\chi = \frac{E}{12(1-\nu^2)}
+\begin{bmatrix}
+1 & \nu & 0 \\
+\nu & 1 & 0 \\
+0 & 0 & \frac{1}{2} (1-\nu)
+\end{bmatrix}
+\end{equation}
+
+The stiffness matrix in the global frame is eventually obtained using the rotation matrix of the element: $\textbf{K} = \textbf{R} \textbf{K}_e \textbf{R}^{T}$ where $\textbf{R}$ describes the rotation of the (triangular) element with respect to its initial configuration.
+
+\subsection{Triangular plate bending}
+
+\begin{figure}
+\centering
+\includegraphics[height=4cm]{chapter9/bending.pdf}
+\caption {The different degrees of freedom $u$ of a triangular thin plate in bending.}
+\label{chap9:fig-triangle}
+\end{figure}
+%
+To calculate the stiffness matrix for the transverse deflections and rotations shown on \fig{chap9:fig-triangle}, we start with the assumed deflection $u_z$ of the form
+\begin{equation}
+ u_z = c_1 + c_2x + c_3y + c_4x^2 + c_5xy + c_6y^2 + c_7x^3 + c_8xy^2 + c_9y^3,
+\label{chap9:eq-deflection}
+\end{equation} 
+where $c_1$, \ldots , $c_9$ are constants.
+Equation~\ref{chap9:eq-deflection} solves an issue of symmetry which was observed with the deflection function proposed by \cite{Przemieniecki68}.
+These constants can be evaluated in terms of the displacements and slopes at the three corners of the triangular plate using
+\begin{equation}
+\textbf{u} = \textbf{Cc}
+\label{chap9:eq-U}
+\end{equation} 
+where $\textbf{u} = \left\{u_1 u_2 \ldots u_9 \right\} $ and $\textbf{c} = \left\{c_1 c_2 \ldots c_9 \right\} $. Matrix $\textbf{C}$ derives from \eqref{chap9:eq-U}:
+\begin{equation}
+C = 
+	\begin{bmatrix}
+	1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+ 	0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
+	0 & -1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+	1 & x_2 & 0 & x_2^2 & 0 & 0 & x_2^3 & 0 & 0 \\
+	0 & 0 & 1 & 0 & x_2 & 0 & 0 & 0 & 0 \\
+	0 & -1 & 0 & -2x_2 & 0 & 0 & -3x_2 & 0 & 0 \\
+	1 & x_3 & y_3 & x_3^2 & x_3y_3 & y_3^2 & x_3^2 & x_3y_3^2& y_3^3 \\
+	0 & 0 & 1 & 0 & x_3 & 2y_3 & 0 & 2x_3y_3 & 3y_3^2 \\
+	0 & -1 & 0 & -2x_3 & -y_3 & 0 & -3x_3^2 & -y_3^2 & 0
+	\end{bmatrix}
+\end{equation}
+
+We used the following notations:
+\begin{align}
+u_1 &= (u_z)_{x_1,y_1}, & u_2 &= \left(\frac{\partial u_z}{\partial y}\right)_{x_1,y_1}, & u_3 &= - \left(\frac{\partial u_z}{\partial x}\right)_{x_1,y_1} \notag \\
+u_4 &= (u_z)_{x_2,y_2}, & u_5 &= \left(\frac{\partial u_z}{\partial y}\right)_{x_2,y_2}, & u_6 &= - \left(\frac{\partial u_z}{\partial x}\right)_{x_2,y_2} \label{chap9:ui} \\
+u_7 &= (u_z)_{x_3,y_3}, & u_8 &= \left(\frac{\partial u_z}{\partial y}\right)_{x_3,y_3}, & u_9 &= - \left(\frac{\partial u_z}{\partial x}\right)_{x_3,y_3} \notag \\
+\end{align}
+
+\noindent We can calculate the strains from the flat-plate theory using:
+\begin{equation}
+e_{xx} = -z \frac{\partial^2u_z}{\partial x^2}
+\end{equation} 
+\begin{equation}
+e_{yy} = -z \frac{\partial^2u_z}{\partial y^2}
+\end{equation} 
+\begin{equation}
+e_{xy} = -2z \frac{\partial^2u_z}{\partial x \partial y}
+\end{equation}
+
+\noindent Hence, using the above equations and \eqref{chap9:eq-deflection}, we have
+\begin{equation}
+\begin{bmatrix}
+e_{xx} \\
+e_{yy} \\
+e_{xy}
+\end{bmatrix}
+= 
+-z
+\begin{bmatrix}
+0 & 0 & 0 & 2 & 0 & 0 & 6x & 0 & 0 \\
+0 & 0 & 0 & 0 & 0 & 2 & 0 & 2x & 6y \\
+0 & 0 & 0 & 0 & 2 & 0 & 0 & 4y & 0 \\
+\end{bmatrix}
+\textbf{c}
+\label{chap9:eq-strains}
+\end{equation} 
+or symbolically $\textbf{e} = \textbf{Dc}$ where $\textbf{D}$ stands for the $3\times 9$ matrix in \eqref{chap9:eq-strains}, including the pre-multiplying constant $-z$.
+Noting from \eqref{chap9:eq-U} that $\textbf{c} = \textbf{C}^{-1}\textbf{u}$, we have:
+\begin{equation}
+\textbf{e} = \textbf{DC}^{-1}\textbf{u} = \textbf{bu},
+\end{equation} 
+where $\textbf{b} = \textbf{DC}^{-1}$.
+
+\noindent Knowing that the stiffness matrix $\textbf{K}_e$ for an element is obtained from
+\begin{equation}
+\textbf{K}_e = \int_v \textbf{b}^{T} \boldsymbol\chi \textbf{b} dV,
+\end{equation} 
+where $\boldsymbol\chi$ is the material matrix, the substitution of $\textbf{b}$ into this expression yields
+\begin{equation}
+\textbf{K}_e = (\textbf{C}^{-1})^T \int_v \textbf{D}^{T} \boldsymbol\chi \textbf{D} dV \textbf{C}^{-1}.
+\end{equation} 
+The integration is carried out numerically using Gauss points located at the middle of each edge of the triangle.
+
+\subsection{Implementation}
+
+In practical terms, the different computations associated with each triangular shell element can be described as follows:
+%
+\begin{enumerate}
+\item Compute the rotation matrix $\textbf{R}$ from global to triangle (local) frame
+\item Compute the local displacement vector $\textbf{u} = \{v_1, v_2, 0, u_2, u_3, v_3, v_4, 0, u_5, u_6,$ $v_5, v_6, 0, u_8, u_9 \} $ for each triangle, where $ (v_1, v_2), (v_3, v_4) and (v_5, v_6) $ are the in-plane displacements on x and y local axes and $ u_i $ were defined by \eqref{chap9:ui}.
+As we are in a co-rotational framework the normal displacements $u_1, u_4, u_7$ at each of the three nodes are null in the local frame of the triangle.
+\item Compute matrix $\textbf{D}_i$  at each Gauss point $i$
+\item The strain-displacement matrix at each Gauss point $i$ is computed with $\textbf{J}_i = \textbf{D}_i \textbf{C}^{-1}$
+\item Compute the local stiffness matrix $\textbf{K}_e$ of the element as $\textbf{K}_e = \displaystyle{\sum^3_{i=1}} \textbf{J}_i \boldsymbol\chi \textbf{J}_i^T$
+\item Transform the local element stiffness matrix into the global frame and add it to a global stiffness matrix
+\end{enumerate}
+
+The co-rotational shell formulation has been successfully implemented as a ForceField into the open-source framework SOFA \citep{Allard07}.
+
+\bigskip
+
+\TODO{add a few words about a GPU implementation.}
+
 \section{Validation}
 
+We compared our model with some theoretical results reported by \cite{Zhongnian86} to assess its quality in modelling bending.
+The test that we carried out uses a square shape mesh clamped on all four edges.
+A uniform load is then applied on the square and the maximum deflection $z_{max}$ at the centre can be calculated.
+Several simulations are performed with increasing load values $q$ (ranging from $1$ to $5 N/m^2$) and the following parameters were used: Young's Modulus $E = 1.092 \times 10^6 \,Pa$, Poisson's ratio $\nu = 0.3$, square edge length $L = 10\,m$, thickness $h = 0.1\,m$.
+Using these particular values it can be shown that $z_{max} = 0.126\,q$.
+The maximum deflection obtained in our simulations are reported in Table~\ref{chap9:tab-results}.
+In average we found $z_{max} = 0.1248\,q $, resulting in a $0.93\,\%$ error between our model and theoretical results on that test.
+%
+\begin{table}[ht]
+	\centering
+	\begin{tabular}{p{9cm}|c|c|}
+	\cline{2-3}
+	\multirow{5}{*}{\includegraphics[height=3.5cm]{chapter9/clamped_square.pdf}} & $q$ & $z_{max}$ \tabularnewline
+	\cline{2-3}
+	& $\,1\,$ & $\, 0.1218 \,$ \tabularnewline
+	& $\,2\,$ & $\, 0.2475 \,$ \tabularnewline		
+	& $\,3\,$ & $\, 0.3747 \,$ \tabularnewline	
+	& $\,4\,$ & $\, 0.5050 \,$ \tabularnewline		
+	& $\,5\,$ & $\, 0.6374 \,$ \tabularnewline
+	\cline{2-3}
+	\end{tabular}
+	\vspace{1cm}
+	\caption{Comparison of our shell model with theoretical results on the bending of a square plate. An error of less than $1\,\%$ was found between our simulation and theoretical results.}
+	\label{chap9:tab-results}
+\end{table}
+
+\section{Mechanical interactions with the curved surface of shells}	\label{chap9:interactions}
+
+The practical interest of modelling complex behaviours such as bending and twisting would remain fairly low for medical simulation if contacts and constraints were not handled properly.
+In our case the difficulty comes from different sources.
+First the collision detection must be carried out with the curved surface of shell elements as opposed to the classic detection on plane triangles.
+Then forces applied to a given triangle need to be distributed between linear forces and torques onto its three vertices.
+As we will see, the same polynomial interpolation function chosen to compute the bending energy in our FEM formulation is also used to capture the interactions between the curved surface and other objects.
+
+In order to detect the collision with the bent surface, we have chosen the subdivision approach.
+We first sample the flat surface of each element by recursively dividing each triangle into four smaller ones and the deflection of each new vertex is computed using \eqref{chap9:eq-deflection} according to the displacements and slopes at the three vertices of the triangular element.
+This process of subdivision allows us to render each shell as a curved triangle (\fig{chap9:fig-shell} (a) and (b)) and detect any collision with the curved surface of the shell using any of the classic collision detection algorithms working on flat triangles.
+
+Once a collision has been detected, it must be processed by distributing the linear force received on the bent surface between the three vertices of the triangle.
+First the linear part of the force is simply transmitted on each node using the barycentric coordinates of the contact point's projection onto the triangle.
+
+The main difficulty is to convert the normal component of the force applied to the bent surface into a torque at each of the three nodes (\fig{chap9:fig-shell} (c)).
+Our approach is the following: during force computation, we use the change in orientation measured at each node to compute the local deflection of each subvertex within the triangle.
+Differentiating the formulation twice yields a relation between the torque applied at each node and the generated force in bending.
+We therefore need to invert the latter formulation to convert a bending force into torques at each vertex.
+We start by retrieving the normal component of the applied force vector $F_{\mathrm{z}}$.
+We project the application point of the force into the triangle's plane and compute its local coordinates $(x,y)$.
+We create the polynom $P$ such as
+%
+\begin{equation}
+P = F_{\mathrm{z}}(1 \hspace{0.2cm} x \hspace{0.2cm} y \hspace{0.2cm} x^2 \hspace{0.2cm} xy \hspace{0.2cm} y^2 \hspace{0.2cm} x^3 \hspace{0.2cm} xy^2 \hspace{0.2cm} y^3)^T. 
+\end{equation}
+%
+The moments at each vertex are then obtained with 
+\begin{equation}
+\Omega = (\textbf{C}^{-1})^T P.
+\end{equation}
+%
+Thus, we are able to transmit any force coming from interactions with the curved surface of shells to the mechanical vertices used in our FEM formulation. 
+%
+\begin{figure}
+\centering
+\includegraphics[width=12cm]{chapter9/shell_curvature.pdf}
+\caption {(a)  The triangle formed by the three vertices of the shell has been recursively subdivided 3 times and the deflection of each new vertex was computed according to the same deflection function used in our shell FEM.
+(b) Sampling the actual surface of the shell allows more accurate rendering and collision detection.
+(c) The shape function is used to distribute an external force $F$ onto the triangle nodes.}
+\label{chap9:fig-shell}
+\end{figure}
+
 \section{Application to implant deployment simulation in cataract surgery}
+
+Cataract surgery consists in three main steps: capsulorhexis, phacoemulsification, and implantation of an intra-ocular lens.
+Prior to starting the surgery, a viscoelastic fluid is introduced into the capsule to facilitate capsulorhexis creation and provide protection during phacoemulsification.
+This fluid remains in the capsule for the duration of the surgery, including the injection of the implant.
+Capsulorhexis is the technique used to remove a part of the anterior lens capsule.
+Phacoemulsification consists in using a surgical device which tip vibrates at an ultrasonic frequency to emulsify the natural lens material and then aspirate the  fragments of the cortical material.
+After the removal of the diseased lens, an intra-ocular lens is implanted into the eye, through a small incision (about 2 mm) using a foldable intra-ocular lens (see \fig{chap9:fig-surgery}).
+The foldable lens, usually made of acrylic material, is then implanted within the lens capsule through the incision used during phacoemulsification.
+In some cases the implant can be flipped or  the hooks (also called \emph{haptics}) can break.
+Therefore the simulation of the insertion and deployment of the implant is crucial for achieving a successful surgery.
+%
+\begin{figure}[h]
+\centering 
+\subfloat[ ]{\includegraphics[height=3.2cm]{chapter9/suction.pdf}}
+\hspace{1cm} 
+\subfloat[ ]{\includegraphics[height=3.2cm]{chapter9/implant_injection_step_1.pdf}}
+\hspace{1cm} 
+\subfloat[ ]{\includegraphics[height=3.2cm]{chapter9/implant_injection_step_2.pdf}}
+\caption [Steps of cataract surgery] {(a) removal of the opacified lens by phacoemulsification.
+(b) insertion of the lens implant which is folded inside the injection device and then deploys within the lens capsule.
+(c) the lens in place in the capsule.}
+\label{chap9:fig-surgery}
+\end{figure}
+
+To simulate the insertion and deployment of an intra-ocular lens, we first created a triangulation of the lens surface.
+Particular care was given to the mesh, to ensure that areas where large stresses occur contain a higher density of elements (see \fig{chap9:fig-mesh}).
+This was done by noting the constraints applied by the surgeon to the haptics while inserting the implant within the injection device.
+During this stage, the haptics are folded onto the implant body, leading to high stresses at the junctions.
+The lens mesh contains $743$ triangles and $473$ nodes.
+Models of the injection device and the entire eye anatomy were also created.
+Physical parameters of the lens implant have been provided by the manufacturer Alcon and they are presented in Table~\ref{chap9:tab-parameters}.
+%
+\begin{table}[h!]
+	\begin{center}
+		\begin{tabular}{|p{3cm}|p{3cm}|p{3cm}|}
+		\hline
+		 \centering Young's modulus & \centering Poisson's ratio & \centering Mass density \tabularnewline
+		\hline
+		\centering $1 MPa$ & \centering $0.42$ & \centering $1.2 g/cm^3$ \tabularnewline
+		\hline
+		\end{tabular}
+	\vspace{0.3cm}
+	\caption{Physical parameters of the intra-ocular implant (source: Alcon)}
+	\label{chap9:tab-parameters}
+	\end{center}
+\end{table}
+
+The first difficulty is to obtain the folded geometry of the lens within the injection device.
+This step is not important for the training process and does not need to be interactive.
+Indeed the surgeon does not always have to prepare the implant as some injection devices are readily available with a folded implant already in place.
+We simulate the folding process by first folding the haptics onto the implant body.
+The body was then bent while keeping the haptics inside to obtain the shape described in \fig{chap9:fig-implantFolding}.
+The whole process was carried out by applying the necessary forces and boundary conditions on the body and haptics of the implant.
+The folded implant was then placed into the injection device.
+The simulation of the injection consists in pushing the intra-ocular implant within the injection device into the lens capsule.
+During these stages of the simulation, complex contacts occur and consist of self collisions of the lens as well as collisions between the lens and the injector and later with the capsule.
+To solve the contacts we use the contact warping method proposed by \cite{Saupin08} as it offers an efficient way to compute physically correct contact responses in the case of co-rotational models \TODO{is it still true?}.
+As adhesions between the haptics and the body is often observed in surgery, friction is also taken into account in the contact response process.
+%
+\begin{figure}[h]
+\centering 
+\subfloat[Real implant]{\includegraphics[height=5cm]{chapter9/IOL.pdf}}
+\hspace{1cm} 
+\subfloat[Meshed model]{\includegraphics[height=5cm]{chapter9/mesh_implant.pdf}}
+\caption [Lens implant and its mesh] {An actual intra-ocular implant (a) and the triangular (b) mesh used in our simulations.
+Notice the higher density of elements in areas where large deformations will take place.
+Image of implant courtesy of Alcon.}
+\label{chap9:fig-mesh}
+\end{figure}
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=6cm]{chapter9/implant_folding1.pdf}
+\includegraphics[width=5cm]{chapter9/implant_folding2.pdf} \\
+\includegraphics[width=4.5cm]{chapter9/implant_folding3.pdf}
+\includegraphics[width=7cm]{chapter9/implant_folding4.pdf}
+\caption [Folding of intra-ocular implant] {Top: intermediary steps of the intra-ocular implant folding. Bottom: fully folded implant ready to be placed into the injection device.}
+\label{chap9:fig-implantFolding}
+\end{figure}
+
+Results of our simulation are illustrated in \fig{chap9:fig-simu-results}.
+We can notice the progressive deployment of the implant when it exits the injector.
+The shape of the intra-ocular lens remains very close to that of a real one during all stages of the simulation: within the injector, during the ejection phase, and when in place within the capsule.
+Due to the high stiffness and low mass of the lens, a direct sparse solver was used at each time step (dt $= 0.01$\,s) rather than an iterative solver, resulting in a more accurate and more stable simulation, to the detriment of computation times (about 5 FPS for the complete simulation, and about 10 FPS for the deformation only, on a 2.4 GHz processor).
+
+% Link where images from surgery have been found: http://asianophthalmology.org/articles/?a=articles&p=6
+\begin{figure}[!h]
+\centering
+\includegraphics[width=4.5cm]{chapter9/surgery1.pdf}
+\hfill
+\includegraphics[width=4.5cm]{chapter9/surgery2.pdf}
+\hfill
+\includegraphics[width=4.5cm]{chapter9/surgery3.pdf} \\
+\vspace{0.1cm}
+\includegraphics[width=4.5cm]{chapter9/simu1.pdf}
+\hfill
+\includegraphics[width=4.5cm]{chapter9/simu2.pdf}
+\hfill
+\includegraphics[width=4.5cm]{chapter9/simu3.pdf}
+\caption [Lens imlant] {Three steps of the simulation of the intra-ocular lens implant injection and its deployment within the lens capsule.
+Top: images from a real cataract surgery, courtesy of Dr.
+Tarek Youssef, ophtalmologist in Canada.
+Below: our simulation of the implant's deployment.}
+\label{chap9:fig-simu-results}
+\end{figure}
+
+\section{Discussion}
+
+We proposed a co-rotational formulation for shell elements by extending a classical in-plane triangular finite element approach.
+This simple shell element can efficiently handle both membrane, bending and twist forces.
+The validity of our approach has been demonstrated though comparison with theoretical results.
+It was then applied it to a rather complex application: the simulation of intra-ocular lens implant deployment in a cataract surgery simulation.
+These preliminary results are very encouraging and show the potential of such models.
+Examples include the modelling of hollow anatomical structures (stomach, colon, etc.), the simulation of cardiac valve leaflets, and blood vessels.
+However, to model the deformation of complex anatomical structures using shell elements, the first step is to describe its surface with curved patches.
+In the next section, we propose to study how to approximate the surface of anatomical structures with shell elements.

### 2010-09-11 Thesis/Chapter6.tex
- Update 11 September -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@173 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-	\subsection{SOFA, an open source simulation framework}
+	\subsection{SOFA, an open source simulation framework}	\label{chap6:SOFA}
 
@@
 \subsubsection*{SOFA integration.}
-Implementing a biomechanical model in SOFA translates essentially into writing a new Force Field, i.e.
-describing the algorithm used to compute internal forces in the model.
+Implementing a biomechanical model in SOFA translates essentially into writing a new Force Field, that is describing the algorithm used to compute internal forces in the model.
 It merely comes down to creating a single C++ class and changing the position reads and force writes to integrate the algorithm into SOFA's design.

### 2010-09-11 Thesis/Chapter8.tex
- Update 11 August -
Added free-form deformation, shape matching and mass-spring systems + big update for bib



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@149 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 11 September -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@173 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\chapter{Mass-spring model for colon implemented on GPU \OC{ISBMS 2008b}}
+\chapter{\TODO{Mass-spring + mapping for colon on GPU?}}
 \label{chap8}
@@
 
-\section{Colon is big, we need a fast model}
+\section{Colon is big, we need a fast model: mass-spring on coarse mesh+ mapping for high res mesh}
 
-\section{Limitations (difficult to link it to actual physical properties)}
+\section{Limitations... But difficult since it was somewhat sufficient for the colonoscopy simulator.}

### 2010-09-11 Thesis/Chapter9.tex
- Update 11 September -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@173 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-The computation of the triangular elastic membrane stiffness matrix can be derived from previous works dealing with tetrahedral co-rotational elements \citep[like][for instance]{Muller04}.
+The computation of the triangular elastic membrane stiffness matrix can be derived from previous works dealing with tetrahedral co-rotational elements \citep[like][for instance]{Muller04b}.
 The element stiffness matrix $\textbf{K}_e$ can be computed as follows:
@@
 \begin{equation}
+\label{chap9:Ke}
 \textbf{K}_e = (\textbf{C}^{-1})^T \int_v \textbf{D}^{T} \boldsymbol\chi \textbf{D} dV \textbf{C}^{-1}.
 \end{equation}
-The integration is carried out numerically using Gauss points located at the middle of each edge of the triangle.
 
-\subsection{Implementation}
+\section{Mechanical interactions with the curved surface of shells}	\label{chap9:interactions}
 
+The practical interest of modelling complex behaviours such as bending and twisting would remain fairly low for medical simulation if contacts and constraints were not handled properly.
+In our case the difficulty comes from different sources.
+First the collision detection must be carried out with the curved surface of shell elements as opposed to the classic detection on plane triangles.
+Then forces applied to a given triangle need to be distributed between linear forces and torques onto its three vertices.
+As we will see, the same polynomial interpolation function chosen to compute the bending energy in our FEM formulation is also used to capture the interactions between the curved surface and other objects.
+
+In order to detect the collision with the bent surface, we have chosen the subdivision approach.
+We first sample the flat surface of each element by recursively dividing each triangle into four smaller ones and the deflection of each new vertex is computed using \eqref{chap9:eq-deflection} according to the displacements and slopes at the three vertices of the triangular element.
+This process of subdivision allows us to render each shell as a curved triangle (\fig{chap9:fig-shell} (a) and (b)) and detect any collision with the curved surface of the shell using any of the classic collision detection algorithms working on flat triangles.
+
+Once a collision has been detected, it must be processed by distributing the linear force received on the bent surface between the three vertices of the triangle.
+First the linear part of the force is simply transmitted on each node using the barycentric coordinates of the contact point's projection onto the triangle.
+
+The main difficulty is to convert the normal component of the force applied to the bent surface into a torque at each of the three nodes (\fig{chap9:fig-shell} (c)).
+Our approach is the following: during force computation, we use the change in orientation measured at each node to compute the local deflection of each subvertex within the triangle.
+Differentiating the formulation twice yields a relation between the torque applied at each node and the generated force in bending.
+We therefore need to invert the latter formulation to convert a bending force into torques at each vertex.
+Thus, we are able to transmit any force coming from interactions with the curved surface of shells to the mechanical vertices used in our FEM formulation.
+%
+\begin{figure}
+\centering
+\includegraphics[width=12cm]{chapter9/shell_curvature.pdf}
+\caption {(a)  The triangle formed by the three vertices of the shell has been recursively subdivided 3 times and the deflection of each new vertex was computed according to the same deflection function used in our shell FEM.
+(b) Sampling the actual surface of the shell allows more accurate rendering and collision detection.
+(c) The shape function is used to distribute an external force $F$ onto the triangle nodes.}
+\label{chap9:fig-shell}
+\end{figure}
+
+\section{Implementation}
+
+The co-rotational shell formulation has been successfully implemented as a Force Field into the open-source framework SOFA \citep{Allard07}.
+A description of SOFA's architecture was given in section \ref{chap6:SOFA}.
+We will now review the key points of the implementation.
+
+	\subsection{Evaluation of displacements}
+The computation of the stiffness matrix, and subsequently the internal forces, starts with the measure of the displacements at the three vertices of every element.
+As defined by \eqref{chap9:ui}, we need to measure the displacement along the z axis and the rotations around local x and y axes for each vertex.
+For \cite{Przemieniecki68}, the deflection z is measured in the global frame.
+But in a co-rotational formulation, a local frame is attached to each triangle and used to compute displacements and forces.
+In this local frame (formed from the 3 vertices) the deflection is therefore always null (hence, $ u_1, u_4, u_7 $ are null).
+To evaluate the remaining rotations, we first associate a quaternion at each vertex which embodies its frame (or its orientation).
+In mathematics, quaternions are an extension of complex numbers and may be written as follows:
+\begin{equation}
+q = w + x i + y j + z k \quad \mbox{with } i^2 = j^2 = k^2 = -1.
+\end{equation}
+For use in geometry, quaternions are often written as a sum of a scalar and a vector:
+\begin{equation}
+q = w + (x, y, z) = \cos (\frac{\alpha}{2}) + \mathbf{u} \sin (\frac{\alpha}{2}),
+\end{equation}
+where $ \mathbf{u} $ is a unit vector and $ \alpha $ a scalar.
+We denote by $ \mathbf{v} $ a vector in $\field{R}^3$ and it can be shown that the quaternion product $ q \mathbf{v} q^{-1} $ yields a vector which is the rotation of $ \mathbf{v} $ by an angle $ \alpha $ around the axis $ \mathbf{u} $.
+Quaternions require less storage than rotation matrices and are known to be more numerically stable.
+
+From there, our approach to measure the rotations around local x and y axes is simple.
+At runtime and for each vertex, we compute the rotation that we need to apply to transform the z axis of the triangle's frame into the z axis of the vertex's frame (see listing \ref{chap9:qDiffZ}).
+The differences with the same quantities computed for the rest position during the initialisation lead to the desired displacements.
+
+\begin{lstlisting}[caption={Method used to compute the rotation of the z axis between the frames embodied by two quaternions},label=chap9:qDiffZ]
+inline Quat qDiffZ(const Quat& Qvertex, const Quat& Qtriangle)
+{
+    // dQ is the quaternion that embodies the rotation between the z axis of the vertex and the z axis of the local triangle's frame (in local space)
+    Quat dQ;
+
+    // u = z axis of the triangle's frame (in local space)
+    Vec3d u(0,0,1);
+
+    // v = z axis of the vertex's frame is expressed into world space
+    Vec3d v = Qvertex.rotate(Vec3d(0.0, 0.0, 1.0));
+
+    // v0 = v expressed into local triangle's frame
+    Vec3d v0 = Qtriangle.rotate(v);
+
+    // Axis of rotation between the 2 vectors u and v lies into the plan of the 2 vectors
+    Vec3d axis = cross(u, v0);
+
+    // Shortest angle between the 2 vectors
+    double angle = acos(dot(u, v0));
+
+    // Quaternion associated to this axis and this angle
+    if (fabs(angle)>1e-6)
+    {
+        dQ.axisToQuat(axis,angle);
+    }
+    else
+    {
+        dQ = Quat(0,0,0,1);
+    }
+
+    return dQ;
+}
+\end{lstlisting}
+
+	\subsection{Inversion of matrix $ \mathbf{C} $}
+\cite{Przemieniecki68} described the condition of non-inversion for the matrix $\mathbf{C}$ as the following:
+\TODO{add condition of inversion... but need the book :)}
+
+Each triangle of a mesh obtained from a regular grid of points fits this condition of non-inversion (right triangles).
+We fixed this limitation by choosing a different local coordinate system that the one introduced by \citeauthor{Przemieniecki68}.
+
+\TODO{add figure with the two different coordinate system}
+
+	\subsection{Numerical integration}
+In practice, the integration of the integral in \eqref{chap9:Ke} must be carried out numerically using Gaussian quadrature, a numerical technique which allows to calculate the value of an integral.
+The Gaussian quadrature states that the integral of a function $ f $ over the area $ A $ of a triangular surface may be evaluated as a weighted sum of function values at specific points within the area of integration \citep{Cowper73}.
+In other words, we have
+\begin{equation}
+\int_{A} f dA = A \sum_{i=1}^n w_i f_i,
+\end{equation}
+where $ w_i $ is a weight factor and $ f_i $ is the evaluation of $ f $ at the Gauss point $ i $.
+Different numbers and locations may be chosen for the integration points depending on the order of accuracy desired.
+We classically choose 3 integration points located at the middle of each edge of the triangle, for which the weight factors $ w_i $ are $ 1/3 $.
+The required integration over the volume is achieved by multiplying the result by the thickness $ t $ of the shell.
+Therefore, \eqref{chap9:Ke} may be numerically evaluated as
+\begin{equation}
+\textbf{K}_e = \dfrac{A t}{3} (\textbf{C}^{-1})^T \sum_{i=1}^3 \textbf{D}_i^{T} \boldsymbol\chi \textbf{D}_i  \textbf{C}^{-1},
+\end{equation}
+where $ D_i $ is the evaluation of the matrix $ \mathbf{D} $ defined by \eqref{chap9:eq-strains} at each Gauss point $ i $.
+
+\bigskip
+
+\subsection{Computation of stiffness matrix: summary}	\label{chap9:summary}
 In practical terms, the different computations associated with each triangular shell element can be described as follows:
@@
 \item Compute the local displacement vector $\textbf{u} = \{v_1, v_2, 0, u_2, u_3, v_3, v_4, 0, u_5, u_6,$ $v_5, v_6, 0, u_8, u_9 \} $ for each triangle, where $ (v_1, v_2), (v_3, v_4) and (v_5, v_6) $ are the in-plane displacements on x and y local axes and $ u_i $ were defined by \eqref{chap9:ui}.
-As we are in a co-rotational framework the normal displacements $u_1, u_4, u_7$ at each of the three nodes are null in the local frame of the triangle.
+As we are in a co-rotational framework, we recall that the normal displacements $u_1, u_4, u_7$ at each of the three nodes are null in the local frame of the triangle.
 \item Compute matrix $\textbf{D}_i$  at each Gauss point $i$
@@
 
-The co-rotational shell formulation has been successfully implemented as a ForceField into the open-source framework SOFA \citep{Allard07}.
+	\subsection{Concept of mappings in SOFA}	\label{chap9:mappings}
+Let us explain the concept of \emph{mappings} and \emph{mechanical mappings} in SOFA.
+In SOFA, the state of a simulated system can be described by the values and time derivatives of its independent degrees of freedom (DOF) gathered in two vectors $x_0$ and $v_0$.
+Geometry can be attached to the DOF for visualisation or contact computation.
+We call it the \emph{shape}.
+It is typically defined by points, such as triangle vertices or sphere centers, and additional data such as triangle connectivity or sphere radii.
+We call these points the vertices.
+Their positions, velocities and associated forces are stored in vectors $x_1$ , $v_1$ and $f_1$, respectively.
+They are not independent variables, since the positions and velocities are bound to the DOF using kinematic operators which we call the \emph{mappings}:
+\begin{eqnarray*}
+x_1 &=&\JNL_1(x_0)\\ 
+v_1 &=& J_1 v_0
+\end{eqnarray*}
+When the vertices and the DOF are the same, the mapping is the identity.
+Matrix $J_1 = \frac{\partial x_1}{\partial x_0}$ encodes the linear relation between the DOF velocities and the shape velocities.
+Due to linearity, the same relation holds on elementary displacements $dx$.
+It also holds on accelerations, with an additional offset due to velocities when the position mapping \JNL is nonlinear.
+In most cases, operators \JNL~ and \J~ are the same, but in the case of rigid bodies, \JNL~ is nonlinear with respect to $x$ and it can not be written as a matrix.
 
-\bigskip
+When shapes collide, additional geometry can be necessary to model the contact.
+For instance, when an edge intersects another one, a contact force is applied to the intersection points.
+These points are defined by their barycentric coordinates with respect to their edge vertices.
+Other relations can be used, depending on the kind of geometrical primitives in contact.
+This additional geometry requires another geometrical layer connected to the shape by a mapping, as illustrated in \fig{chap9:fig-mappings}.
 
-\TODO{add a few words about a GPU implementation.}
+\begin{figure}
+	\begin{center}
+ 		\includegraphics[width=13cm]{chapter9/mappings}
+ 	\end{center}
+ \caption[Concept of mappings in SOFA]{Mappings from the DOFs to a contact point.
+Top: two simulated objects in contact (red point).
+Bottom: hierarchy of geometrical layers.
+Positions and velocities are propagated top-down.
+The contact force $f_c$ is accumulated in the contact layers.
+Forces are then propagated bottom-up.
+Image courtesy of SOFA's documentation.
+}
+\label{chap9:fig-mappings}
+\end{figure}
+
+Positions and velocities can be propagated top-down through our layer hierarchy using the relations presented.
+In order to take the contact forces into account in the dynamics equation, we have to convert the contact forces applied to the contact points to forces applied to the DOF, where our mechanical model is applied.
+This requires an extension to the position and velocity mappings presented: we call it \emph{mechanical mapping}.
+This is a general method to propagate forces bottom-up through the layers of the geometrical hierarchy.
+Given forces $f_n$ applied to a geometry layer $n$, we derive the equivalent force applied to its parent layer $n-1$.
+Equivalent forces must have the same power.
+Thus, we have to compute $f_{n-1}$ such that:
+$$
+v_{n-1}^T f_{n-1} = v_n^T f_n
+$$
+The relation $v_{n} = J_{n}v_{n-1}$ allows us to rewrite the previous equation as
+$$
+v_{n-1}^T f_{n-1} = v_{n-1}^T J_{n}^T f_n
+$$
+Since this relation must hold for any velocity $v_{n-1}$, we simplify it and get
+\begin{equation} \label{eq:mapF}
+f_{n-1} = J_n^T f_n
+\end{equation}
+This corresponds to the principle of virtual work.
+
+\subsection{Visualisation}
+Because shells have a curved surface, a specific rendering technique is required to visualise the deformation in bending.
+Without any consideration, the shells would appear flat (as simple triangles).
+Our idea is to make use of a higher resolution mesh whose deformation would be controlled by the underlying physical mesh of shells.
+In other words, a mapping was implemented to connect the DOFs of our mechanical model to a set of vertices useful for rendering.
+In SOFA, a mapping has only 2 methods: init(), which allows computations during the initialisation phase, and apply() called at runtime and in charge of updating the positions of the high resolution mesh according to the coarser mesh constituted by the shell elements.
+
+\begin{itemize}
+\item \textbf{init()} \\
+During the initialisation phase, each vertex from the high resolution mesh is projected onto the coarse mesh.
+Thus, for each vertex we find the closest primitive (DOF, edge or triangle) on the coarse mesh and we take the list of triangles attached to it.
+Then, the barycentric coordinates of the projected point is computed within each triangle of this list.
+%
+\item \textbf{apply()} deals with the relation between the DOFs and the vertices used for rendering (shape vertices).
+The position of each vertex from the high resolution mesh must be updated according to the underlying mesh.
+After the initialisation phase, we have associated each rendering vertex to a triangle and its barycentric coordinates within this triangle are known.
+Therefore, we first start by computing the new position of the rendering vertex by weighting the coordinates of each triangle's vertex with the associated barycentric weight.
+Hence, we obtain a new position in the plane of the triangle.
+If we were to stop the procedure here, the shell elements would still be rendered as flat triangles since all rendering vertex are located in the plane of the triangles.
+Consequently, the additional step of computing the deflection is required.
+We know that the deflection at any point of local coordinates $(x, y)$ may be computed from $u_z = c_1 + c_2x + c_3y + c_4x^2 + c_5xy + c_6y^2 + c_7x^3 + c_8xy^2 + c_9y^3$ as described by \eqref{chap9:eq-deflection}.
+Hence, we calculate the local coordinates $(x, y)$ of each rendering vertex within its associated triangle.
+Moreover, we notice that the coefficients $ \mathbf{c} = \left\{c_1 c_2 \ldots c_9 \right\} $ may be evaluated using \eqref{chap9:eq-U} from
+\begin{equation}
+\mathbf{c} = \mathbf{C}^{-1} \mathbf{u},
+\end{equation}
+The deflection may therefore be computed for each vertex and added to the in-plane position previously calculated. 
+\end{itemize}
+
+It is worth noting that a high resolution mesh may be obtained by recursively subdividing each triangle.
+However, the vertices are closely related to the structure of the mechanical mesh and rendering artifacts may appear.
+As we can notice on \fig{chap9:fig-rendering}, the quality of the visualisation may be improved by using a totally different mesh.
+%
+%\begin{figure}
+%	\begin{center}
+%		\begin{tabular}{cc}
+%			\includegraphics[width=6cm]{chapter9/sub_grid_red.png} & \includegraphics[width=6cm]{chapter9/irregular_grid_red.png} \\
+%			\includegraphics[width=6cm]{chapter9/sub_grid_mesh.png} & \includegraphics[width=6cm]{chapter9/irregular_grid_mesh.png} \\
+%			Subdivided triangles & Irregular grid			
+%		\end{tabular}
+%	\end{center}
+%	\label{chap9:fig-rendering}
+%	\caption{We can notice some rendering artifacts when we use subdivided triangles. Using an irregular mesh instead reduces the impact.}
+%\end{figure}
+%
+\begin{figure}[ht]
+\centering 
+\subfloat[Subdivided triangles]{\includegraphics[width=6.5cm]{chapter9/sub_grid_red.png}}
+\hfill 
+\subfloat[Irregular grid]{\includegraphics[width=6.5cm]{chapter9/irregular_grid_red.png}} \\
+\subfloat[Subdivided triangles]{\includegraphics[width=6.5cm]{chapter9/sub_grid_mesh.png}}
+\hfill
+\subfloat[Irregular grid]{\includegraphics[width=6.5cm]{chapter9/irregular_grid_mesh.png}}
+\caption[Visual artifacts using triangle subdivisions]{In some situations, some rendering artifacts may appear when we use subdivided triangles to render the shells (a,c).
+Using an irregular mesh which is not closely related to the structure of the mechanical mesh reduces the impact (b,d).}
+\label{chap9:fig-rendering}
+\end{figure}
+
+	\subsection{Contacts with the curved surface of shells}
+If a specific rendering technique was developed to visualise the curved shape of the shells, it is also essential to interact with it.
+Indeed, as explained section \ref{chap9:interactions}, the force applied onto the curved surface of the shell must be processed by distributing the linear force received on the bent surface between the three vertices of the triangle.
+Consequently, the mapping previously described is extented to a mechanical mapping and two additional methods need to be implemented: applyJ and applyJT.
+%
+\begin{itemize}
+\item \textbf{applyJ()} encodes the linear relation between the DOF velocities and the shape velocities as described in section \ref{chap9:mappings}.
+applyJ() basically embodies the same computations than apply() except that the vector of displacement $\mathbf{u}$ is different.
+Whereas the local displacement in rotation is measured from the positions to create the vector $\mathbf{u}$, this time the vector is directly obtained from the angular velocities we had at each vertex (once converted into the local frame).
+We therefore have a relation between the angular velocities at each vertex and the velocity of the deflection in the direction normal to each triangle.
+\item \textbf{applyJT()} must convert the contact forces applied to the contact points to forces applied to the DOF.
+If the linear forces are easily distributed between DOFs using barycentric coordinates, it also encodes the relation between the vertical force $ \mathbf{F} $ applied onto the bent surface of the plate and the torque at each DOF.
+We start by retrieving the normal component of the applied force vector $F_{\mathrm{z}}$.
+We project the application point of the force into the triangle's plane and compute its local coordinates $(x,y)$.
+We create the polynom $P$ such as
+%
+\begin{equation}
+P = F_{\mathrm{z}}(1 \hspace{0.2cm} x \hspace{0.2cm} y \hspace{0.2cm} x^2 \hspace{0.2cm} xy \hspace{0.2cm} y^2 \hspace{0.2cm} x^3 \hspace{0.2cm} xy^2 \hspace{0.2cm} y^3)^T. 
+\end{equation}
+%
+The moments at each vertex are then obtained with 
+\begin{equation}
+\Omega = (\textbf{C}^{-1})^T P.
+\end{equation}
+\end{itemize}
+
+\subsection{Allowing a curved rest configuration}
+In \cite{Przemieniecki68}, the rest and initial configurations are both assumed to be the flat one.
+The only way to have an initial deformed shape is to apply a torque on the vertices of the shells.
+However, in this configuration we would compute a displacement vector $ \mathbf{u} $ that is not null, which leads to a bending energy.
+In order to define a curved rest configuration for our shells, we propose to compute the bending energy between the initial and rest configuration during the initialisation phase, and add the energy between the current and initial configurations at runtime.
+Consequently, by separating initial and rest configurations, it is possible to have an initial deformed shape without any initial bending energy.
+In other words, we shifted the zero energy configuration.
+
+\subsection{Possible extension: parallel implementation on GPU}
+If we look closely all the steps required to compute the local stiffness matrix for each element (see section \ref{chap9:summary} for a summary), one may notice that the operations carried out for a given shell element are independent of the other elements.
+This feature makes our co-rotational shell FEM implementation highly parallelisable.
+One potential problem however, is the assembly of the global stiffness matrix.
+Indeed, in a parallel implementation, many elements are processed concurrently.
+Inevitably, two elements sharing a node may be processed in the same time and may therefore try to both write their element contribution into the exact same location in the global stiffness matrix, and therefore the same location in GPU's memory.
+As with the TLED implementation described in chapter \ref{chap6}, there is two solutions.
+The first one would be to re-order the elements organisation so that no concurrent writes will occur.
+Unfortunately, this re-organisation of the element reduces the number of elements that can be processed in parallel.
+The second solution would be to take advantage of a new GPU NVIDIA architecture called Fermi, which added the capability atomic writes for floats.
+We recall that the write operation is said to be atomic in the sense that it is guaranteed to be performed without interference from other threads.
+In other words, the concurrent writes are simply serialised in hardware.
+Obviously, an additional overhead is to be expected.
+
+To conclude on a possible parallelisation of the algorithm, an optimal GPU implementation is always a challenge to achieve.
+One needs to choose wisely the type of memory to be used for each variable and optimise the access patterns.
+However, it seems that there is no major obstacle to a GPU implementation.
+
+\TODO{enough details?}
 
@@
 A uniform load is then applied on the square and the maximum deflection $z_{max}$ at the centre can be calculated.
-Several simulations are performed with increasing load values $q$ (ranging from $1$ to $5 N/m^2$) and the following parameters were used: Young's Modulus $E = 1.092 \times 10^6 \,Pa$, Poisson's ratio $\nu = 0.3$, square edge length $L = 10\,m$, thickness $h = 0.1\,m$.
+Several simulations are performed with increasing load values $q$ (ranging from $1$ to $5 N/m^2$) and the following parameters were used:
+\begin{itemize}
+ \item Young Modulus $E = 1.092 \times 10^6$
+ \item Poisson's ratio $\nu = 0.3$
+ \item edge of the square $L = 10\,m$
+ \item thickness $h = 0.1\,m$
+ \item pressure q is a uniformly distributed load per unit area
+\end{itemize}
+
 Using these particular values it can be shown that $z_{max} = 0.126\,q$.
@@
 
-\section{Mechanical interactions with the curved surface of shells}	\label{chap9:interactions}
+\bigskip
 
-The practical interest of modelling complex behaviours such as bending and twisting would remain fairly low for medical simulation if contacts and constraints were not handled properly.
-In our case the difficulty comes from different sources.
-First the collision detection must be carried out with the curved surface of shell elements as opposed to the classic detection on plane triangles.
-Then forces applied to a given triangle need to be distributed between linear forces and torques onto its three vertices.
-As we will see, the same polynomial interpolation function chosen to compute the bending energy in our FEM formulation is also used to capture the interactions between the curved surface and other objects.
-
-In order to detect the collision with the bent surface, we have chosen the subdivision approach.
-We first sample the flat surface of each element by recursively dividing each triangle into four smaller ones and the deflection of each new vertex is computed using \eqref{chap9:eq-deflection} according to the displacements and slopes at the three vertices of the triangular element.
-This process of subdivision allows us to render each shell as a curved triangle (\fig{chap9:fig-shell} (a) and (b)) and detect any collision with the curved surface of the shell using any of the classic collision detection algorithms working on flat triangles.
-
-Once a collision has been detected, it must be processed by distributing the linear force received on the bent surface between the three vertices of the triangle.
-First the linear part of the force is simply transmitted on each node using the barycentric coordinates of the contact point's projection onto the triangle.
-
-The main difficulty is to convert the normal component of the force applied to the bent surface into a torque at each of the three nodes (\fig{chap9:fig-shell} (c)).
-Our approach is the following: during force computation, we use the change in orientation measured at each node to compute the local deflection of each subvertex within the triangle.
-Differentiating the formulation twice yields a relation between the torque applied at each node and the generated force in bending.
-We therefore need to invert the latter formulation to convert a bending force into torques at each vertex.
-We start by retrieving the normal component of the applied force vector $F_{\mathrm{z}}$.
-We project the application point of the force into the triangle's plane and compute its local coordinates $(x,y)$.
-We create the polynom $P$ such as
-%
-\begin{equation}
-P = F_{\mathrm{z}}(1 \hspace{0.2cm} x \hspace{0.2cm} y \hspace{0.2cm} x^2 \hspace{0.2cm} xy \hspace{0.2cm} y^2 \hspace{0.2cm} x^3 \hspace{0.2cm} xy^2 \hspace{0.2cm} y^3)^T. 
-\end{equation}
-%
-The moments at each vertex are then obtained with 
-\begin{equation}
-\Omega = (\textbf{C}^{-1})^T P.
-\end{equation}
-%
-Thus, we are able to transmit any force coming from interactions with the curved surface of shells to the mechanical vertices used in our FEM formulation. 
-%
-\begin{figure}
-\centering
-\includegraphics[width=12cm]{chapter9/shell_curvature.pdf}
-\caption {(a)  The triangle formed by the three vertices of the shell has been recursively subdivided 3 times and the deflection of each new vertex was computed according to the same deflection function used in our shell FEM.
-(b) Sampling the actual surface of the shell allows more accurate rendering and collision detection.
-(c) The shape function is used to distribute an external force $F$ onto the triangle nodes.}
-\label{chap9:fig-shell}
-\end{figure}
+\TODO{maybe a bit short for a so-called validation...}
 

### 2010-09-13 Thesis/Chapter10.tex
- Update 13 September -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@174 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2010-08-09 Thesis/Chapter10.tex
@@
-\chapter{'Shell meshing' technique \OC{MICCAI 2010}}
+\chapter{Physics-based reconstruction using shell elements}
 \label{chap10}
 \begin{shortAbstract}
-A short abstract for the upcoming chapter
+A short abstract for the upcoming chapter \\
+\OC{MICCAI 2010}
 \end{shortAbstract}
 
-\section{State of art: reconstruction/simplification}
+\section{Techniques of mesh simplification}
+
+\subsection{Introduction: our motivation}
+
+Our idea is the following: while many flat triangles are required to describe highly curved surfaces, fewer triangular shell elements are needed to describe the given geometry with the same precision since they can be curved.
+In the following, we assume that we have a high resolution triangular mesh obtained from a binary segmented image of the object we want to simulate (via a Marching Cube algorithm for instance, \cite{Lorensen87}).
+Our goal is to create a mesh featuring the optimal number of shell elements while staying as close as possible to our targeted geometry.
+For this reason, we need a technique to simplify the high resolution mesh of the object to be simulated.
+
+\bigskip
+
+\TODO{state of the art to be added... }
+
+%Simplification strategies may be grouped into two categories: local strategies that iteratively simplify the mesh by the repeated application of some local operator, and global strategies that are applied to the input mesh as a whole \citep{Cignoni97,Talton04}.
+%
+%\subsection{Local simplification strategies}
+%
+%Typically, they define some mesh operation $S$ that, when applied to a mesh $M$, acts on a small collection of its elements and produces a new mesh $\overline{M}$  with fewer elements.
+By repeated application of $S$, a mesh may be simplified arbitrarily.
+In order to determine the mesh elements to which $S$ should be applied on a given iteration, $S$ may be associated with an error function (or cost function) that measures the amount of error the operation will introduce into the approximation.
+By computing the error associated with every possible application of $S$ at a particular iteration, the algorithm can apply the one with minimal cost.
+This type of heuristic is quite reasonable for simplification problems, and in practice these methods work well.
+%
+%One nice property of local iterative algorithms is that they allow the user to specify the desired attributes of the target approximating mesh with a high degree of precision.
+For example, the user may allow the algorithm to run until the mesh contains $k$ faces, or until the error at a given vertex exceeds some threshold $\varepsilon$.
+%
+%\subsubsection*{Vertex decimation.}
+%The decimation algorithm, first proposed by \cite{Schroeder92}, is simple.
+Multiple passes are made over all vertices in the mesh.
+During a pass, each vertex is a candidate for removal and, if it meets the specified decimation criteria, the vertex and all triangles that use the vertex are deleted.
+The resulting hole in the mesh is patched by forming a local triangulation.
+The vertex removal process repeats, with possible adjustment of the decimation criteria, until some termination condition is met.
+%
+%The error associated with a particular decimation typically depends on the classification of the vertex being decimated.
+For the general case of a manifold vertex $\lbrace i \rbrace$ not near a boundary, \citeauthor{Schroeder92} consider the set of planes formed by the adjacent triangles and compute a single approximating plane based on the area-weighted average of the triangle normals $\textbf{n}_k$, centers $x_k$, and areas $A_k$:
+%%
+%\begin{equation}
+%\textbf{n} = \frac{\sum_k A_k\textbf{n}_k}{\sum_k A_k},  \hspace*{1cm} \widehat{\textbf{n}} = \frac{\textbf{n}}{\vert \textbf{n} \vert}, \hspace*{1cm}  x = \frac{\sum_k A_k\textbf{x}_k}{\sum_k A_k}
+%\end{equation}
+%
+%Then they calculate the distance from vertex $\lbrace i \rbrace$ to this plane by  $\vert \widehat{\textbf{n}} \cdot (v_i - x) \vert$ and define this quantity to be the cost associated with the decimation.
+This metric ensures that vertices in smooth regions will be decimated before vertices that define sharp features.
+%
+%In practice, simplification schemes based on vertex decimation excel at eliminating un-necessary geometry, such as is typically found in meshes generated by the marching cubes algorithm where large planar regions may be subdivided into many redundant triangles.
+%
+%\subsubsection*{Edge contraction.}
+%Edge contraction, originally proposed by \cite{Hoppe93}, is the most common simplification operation.
+An edge contraction operates on a single edge $\lbrace v_1,v_2 \rbrace$ and contracts that edge to a single vertex $\lbrace w \rbrace$, updating all edges previously incident on $\lbrace v_1 \rbrace$ and $\lbrace v_2 \rbrace$ to reference $\lbrace w \rbrace$.
+On a manifold mesh without boundary, each edge contraction will collapse precisely two faces (see \fig{chap10:fig-edgeCollapse}).
+Edge contractions can alter the topology of a mesh, since repeatedly contracting all the edges around a hole will eventually close it.
+%%
+%\begin{figure}
+%\centering
+%\includegraphics[height=3cm]{chapter10/edgeCollapse.pdf}
+%\caption[The edge contraction operation]{The edge contraction operation. Image courtesy of \cite{Hoppe93}.}
+%\label{chap10:fig-edgeCollapse}
+%\end{figure}
+%
+%For a given contraction $\lbrace v_1,v_2 \rbrace \rightarrow \lbrace w \rbrace$, it is not immediately clear what value should be assigned to $v_h$, i.e.
+where the resulting vertex should be placed.
+Obvious choices such as $v_h = v_i$, $v_h = v_j$, or $v_h = (v_i + v_j)/2$ are convenient, but can easily be shown to be non-optimal.
+Rather than placing $v_h$ arbitrarily, it is better to consider the error function associated with the contraction operation and attempt to minimise its value over the space of possible vertex placements.
+Once the optimal target position has been computed for every possible contraction, the contraction with the smallest associated error can be selected and applied.
+%
+%\subsubsection*{Adaptative subdivision.}
+%Coplanar or nearly coplanar faces are searched for in the mesh, merged into larger polygons, and then retriangulated into fewer faces than those originally required \citep{Dehaemer91,Hinker93}.
+Face merging is driven by a co-planarity test.
+%
+%The basic adaptive subdivision method as presented by \cite{Schmitt86} is sketched in \fig{chap10:fig-adaptativeSubdivision}.
+(a) shows several data points produced by the sampling process and sketch (b) adds a trial polygon that approximates the surface represented by the sample points.
+A bilinear interpolation is performed over the surface of the approximating polygon to locate the point on the surface 'directly below' a sample data point.
+The distance (or error) between the surface and the data point is then calculated, as shown in (c), and is compared to the user specified tolerance.
+If any of the errors exceeds the tolerance value then the polygon is divided into four smaller polygons, as shown in (d), and the process is repeated recursively.
+%%
+%\begin{figure}
+%\centering
+%\includegraphics[width=13.5cm]{chapter10/adaptativeSubdivision.pdf}
+%\caption[Surface-fitting polygons with adaptive subdivision]{Surface-fitting polygons with adaptive subdivision. Image courtesy of \cite{Schmitt86}.}
+%\label{chap10:fig-adaptativeSubdivision}
+%\end{figure}
+%
+%When the approximation to the data is within the specified tolerance, the recursion is terminated and the polygon is saved in a linked list.
+Upon completion of the subdivision algorithm, the resulting polygons are displayed for the user to judge the results.
+Generally, the user runs the program repeatedly while varying the maximum tolerance until the results are suitable.
+%
+%\subsubsection*{Polygon growth.}
+%Another method proposed by \cite{Dehaemer91} to systematically generate surface-fitting polygons is to use a polygon growth technique (see \fig{chap10:fig-polygonGrowth}).
+First a seed polygon is selected from the original set of data, as in (a).
+Then a neighbouring polygon is selected and the two are combined into a larger trial polygon ((b) and (c)).
+If the trial polygon passes the accuracy metric, 10 additional neighbors are selected in an attempt at further growth, as in (d) through (g).
+When all attempts at further growth fail, the polygon is added to the list of polygons describing the reduced object, as shown in (h).
+A new seed polygon is selected and the process repeated.
+When there are no more polygons from which to choose a new seed, the process is terminated.
+%%
+%\begin{figure}
+%\centering
+%\includegraphics[width=12cm]{chapter10/polygonGrowth.pdf}
+%\caption[Polygon growth]{Polygon growth. Image courtesy of \cite{Dehaemer91}.}
+%\label{chap10:fig-polygonGrowth}
+%\end{figure}
+%
+%This method of fitting polygons to the sampled data is not as useful as the adaptive subdivision methods, for several reasons.
+First, it is an essentially brute force method of surface fitting and very repetitive in nature.
+Many trial polygons are created, only to be 'thrown out' because they exceed the error tolerance.
+This causes the procedure to execute quite slowly (orders of magnitude more slowly) than the adaptive subdivision method.
+Second, the data is not significantly reduced.
+Many of the generated polygons are of complicated shapes and use a large number of vertices.
+Thus many of the original sample data points are retained.
+%
+%
+%\subsubsection*{Appearence preserving simplification.}
+%A completely different approach to determining the error associated with a given simplification operation was described by \cite{Cohen98}.
+Their metric, which is appearance-based, measures the amount of deviation caused by the operation in the screen-space (pixelised) representation of the mesh.
+Their algorithm functions by decoupling surface position from color and curvature information and storing the latter two quantities in texture and normal maps.
+A traditional geometric simplification algorithm can then be employed to filter the surface position, while a hardware-based approach is used to filter the color and normal information, resulting in simplified representations that are nearly visually indistinguishable from the original.
+However, the algorithm requires a parameterisation of the mesh in order to function.
+%
+%
+%\subsection{Global simplification strategies}
+%
+%\subsubsection*{Vertex clustering.}
+%The method of vertex clustering was originally proposed by \cite{Rossignac93} to handle meshes of arbitrary topological structure.
+In their algorithm, each vertex in the input mesh is assigned a weight based on its perceptual importance: vertices adjacent to triangles with large faces and those in areas of high curvature are weighted more heavily than vertices in smooth regions adjacent to smaller triangles.
+Next, a bounding box is placed around the mesh and subdivided into a three-dimensional grid.
+Finally, all the vertices in a given grid cell are clustered to the position of the vertex with maximum weight.
+Degenerate geometry may then be removed, although the authors proposed a novel visualisation of certain degenerate faces and edges in an effort to enhance the recognizability of their simplified meshes.
+The algorithm is extremely efficient and simple to implement, and the level of simplification may be controlled (although with some difficulty) by choosing the resolution of the overlaid grid.
+However, vertex clustering can drastically alter the topology of the input mesh in an unpredictable manner, and in practice does not produce very faithful geometric approximations at low levels of detail.
+%
+%\subsubsection*{Re-tiling.}
+%The re-tiling method described by \cite{Turk92} begins with a polygonal surface and creates a triangulation of this surface with a user-specified number of vertices.
+The points are placed uniformly over a given polygonal surface by distributing points at random over the surface and then having each point repel all of its neighbors.
+An alternative to placing the points randomly is using the variation in surface curvature.
+The features of the surface would be more accurately reflected in a re-tiling by increasing the density of vertices in regions of high curvature.
+%
+%The key notion in the re-tiling procedure is the creation of an intermediate model called the mutual tessellation of a surface that contains both the vertices from the original model and the new points that are to become vertices in the re-tiled surface.
+The new model is then created by removing each original vertex and locally re-triangulating the surface in a way that matches the local connectedness of the initial surface (see \fig{chap10:fig-retiling}).
+%
+%If this method produces nice meshes, an issue is finding measures of how closely matched a given re-tiling is to the original model. 
+%%
+%\begin{figure}
+%\centering
+%\includegraphics[width=12cm]{chapter10/retiling.pdf}
+%\caption[Retiling technique]{Upper left: original surface.
+Upper right: candidate vertices after point-repulsion.
+Lower left: mutual tesselation.
+Lower right: final tesselation.
+Image courtesy of \cite{Turk92}.}
+%\label{chap10:fig-retiling}
+%\end{figure}
 
 \section{Our method}
+
+	\subsection{A simple algorithm}
+
+However, we need to keep in mind that our situation is substantially different than all previously mentioned geometrical approaches as we want to model the deformation of the structure.
+In that regard, the curvature of the surface has a physical meaning: it represents the mid-surface of the shell.
+We propose to approximate the surface of anatomical structures with shell elements whose each surface is described by the shape function used in our shell formulation.
+
+The first step in the process of generating a shell-based mesh is an important decimation of the high resolution mesh, using quadric edge collapse technique implemented in \cite{Meshlab}.
+The algorithm tries as much as possible to preserve mesh boundaries and generates high quality triangular elements.
+We then apply a heuristic method derived from the work of \cite{Saupin07} with tetrahedral meshes based on simple geometrical rules.
+For each node of the coarse mesh, we find the three closest triangles on the high resolution mesh and we move the node to the barycentre of the three centres of mass of those triangles.
+This technique locally smoothes the surface of the mesh while converging towards the desired high resolution mesh.
+
+At each iteration of this algorithm, we want to measure the error between the curved surface of shells and the target.
+Indeed, we need to ensure that the distance between the surface of our shell-based mesh and the targeted high resolution mesh will be minimal.
+An efficient technique for measuring the error between two surfaces is the Hausdorff distance \citep{Klein96}.
+As a reminder the Hausdorff distance between two meshes is the maximum between the two so-called one-sided Hausdorff distances:
+\begin{equation}
+d_{\mathrm{H}}(X,Y) = \max \left\{ \sup_{x \in X} \inf_{y \in Y} d(x,y), \sup_{y \in Y} \inf_{x \in X} d(x,y) \right\},
+\end{equation}
+where $d()$ is the Euclidian distance between two points.
+The same technique of subdivision used for rendering allows us to sample the actual surface described by the shells to compute the Hausdorff distance with the targeted high resolution mesh.
+
+By using the Hausdorff distance between two iterations, the process may be stopped when the required precision has been reached.
+A simple example is shown \fig{chap10:fig-cylinder} to illustrate the method.
+\begin{figure}
+\centering
+\includegraphics[width=13.5cm]{chapter10/exampleCylinder.pdf}
+\caption {The target (a) is a high resolution cylinder mesh of 16,384 triangles and we start from a very coarse mesh (12 triangles), rendered with flat triangles here (b).
+In (c) the coarse mesh is rendered with shells and a one-sided Hausdorff distance colour map is applied to show the initial error with the high resolution mesh.
+(d) One-sided Hausdorff distance colour map after one iteration of our algorithm (48 shells).}
+\label{chap10:fig-cylinder}
+\end{figure}
+
+	\subsection{Results on complex anatomical structures}
+
+\subsubsection*{Meshing of anatomical structures.}
+This approach has been applied to approximate more complex anatomical geometries with curved shell elements.
+Two examples are given with the Glisson's capsule, which is the membrane surrounding the liver (\fig{chap10:fig-liver}) and an aneurysm (\fig{chap10:fig-aneurysm}).
+In each case, the error is computed through a Hausdorff distance and expressed as a percentage of the diagonal of the object's bounding box.
+%
+\begin{figure}
+\centering
+\includegraphics[width=13.5cm]{chapter10/resultsLiver.pdf}
+\caption {(a) the targeteted high resolution Glisson's capsule mesh ($8\,000$ triangles).
+(b) the one-sided Hausdorff distance error map after applying only one iteration of our algorithm to the coarse mesh ($1\,200$ shells).}
+\label{chap10:fig-liver}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=13.5cm]{chapter10/resultsAneurysm.pdf}
+\caption {(a) the targeteted high resolution aneurysm mesh ($28\,368$ triangles). (b): the one-sided Hausdorff distance error map on a mesh of 772 shells generated with our method.}
+\label{chap10:fig-aneurysm}
+\end{figure}
+
+\subsubsection*{Computation times.}
+We perform several tests on the aneurysm model at different resolutions to measure computation times.
+The shells are resisting to a uniform pressure load and solved using a Conjugate Gradient (CG) iterative solver.
+The computation times are reported in \fig{chap10:fig-computation}.
+Implicit integration allows for large time steps ($40\,ms$) and the computation is real-time for 800 shell elements and a reasonable error criterion ($5\,\%$).
+When the computation time must be bounded (critical real-time applications), one can fix the number of CG iterations to, for instance, $100$ and remains real-time for $1\,000$ shell elements.
+However, in that case the accuracy of the results is not checked.
+%
+\begin{figure}
+\centering
+\includegraphics[width=13.5cm]{chapter10/computation_times.pdf}
+\caption {Computation time on meshes of 200, 400, 600, 800 and $ 1\,000 $ elements.}.
+\label{chap10:fig-computation}
+\end{figure}
+
+\subsection{Coupling between tetrahedra and shells for advanced modelling}
+Structures in human body can be either solid (brain, liver, prostate etc.) or hollow (colon, blood vessels, stomach etc.).
+However, knowing how to model the two kind of structures is not sufficient to reach a high degree of accuracy, real life situations are more complex.
+As an example, the external surface of the liver is covered by a layer of cells called Glisson's capsule.
+Its interaction with the liver plays an important role into the overall structure's mechanical behaviour.
+Therefore considering the interaction between solid and hollow objects is as crucial as modelling the two structures separately.
+
+An example of medical procedure to illustrate this point even further is angioplasty.
+Angioplasty is the technique of mechanically widening a narrowed or obstructed blood vessel, typically as a result of atherosclerosis.
+An empty and collapsed balloon on a guide wire is passed into the narrowed locations and then inflated to a fixed size.
+The balloon crushes the fatty deposits, so opening up the blood vessel to improved flow.
+As a proof of concept, we tried to simulate an angioplasty.
+Our preliminarty results are presented on \fig{chap10:fig-stent}.
+The blood vessel is modelled using the shell FEM formulation described in this paper and the fatty deposits are simulated with a tetrahedral FEM method and are fixed to the interior wall of the blood vessel.
+When the balloon inflates it crushes the deposits and they then apply a pressure onto the curved surfaces of shells modelling the interior wall.
+The forces are then distributed onto the mechanical nodes of the blood vessel mesh as detailed in section~\ref{chap9:interactions}, which widens the blood vessel as expected.
+%
+\begin{figure}
+\centering
+\includegraphics[width=12cm]{chapter10/stenose_final.pdf}
+\caption {Simulation of an angioplasty procedure.
+(a, c): A collapsed stent is inserted into the blood vessel simulated with our shell FEM formulation.
+The fatty deposits (in yellow) are modelled with tetrahedral FEM.
+(b, d): The stent is crushing the fatty deposits which creates a pressure onto the interior wall and widens the blood vessel.}
+\label{chap10:fig-stent}
+\end{figure}
+
+\section{Discussion}
+
+We proposed a framework for real-time modelling of thin anatomical structures.
+The novelty of our method relies on the combination of a shell finite element formulation and a geometric surface reconstruction both based on the same polynomial interpolation function used to describe the surface of shells.
+We also show how contacts and interactions with the curved surfaces of shells can be handled using the same function.
+In summary, these polynomial shape functions are used in three different ways in our computational model:
+%
+\begin{description}
+\item[(a)] to approximate complex geometrical shapes
+\item[(b)] to compute internal forces via a co-rotational shell FEM
+\item[(c)] to compute contact forces onto a curved triangle
+\end{description}
+
+The efficiency of the method is illustrated through shell-based reconstruction and real-time simulation of the deformations of various anatomical structures.
+We also present preliminary results on the simulation of an angioplasty procedure.
+
+\bigskip
+
+\TODO{Did we answer our problematic? Limitations? }

### 2010-09-13 Thesis/Chapter9.tex
- Update 13 September -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@174 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 They offer a good trade-off between computational efficiency and accuracy by allowing small deformations but large displacements.
-We propose to improve and extend a plate model first introduced by \cite{Przemieniecki68} to a co-rotational formulation.
+We propose to improve and extend a plate model first introduced by \cite{Przemieniecki85} to a co-rotational formulation.
 Once combined with an in-plane membrane formulation we obtain an accurate, yet computationally efficient, shell finite element method featuring both membrane and bending energies.
@@
 where $c_1$, \ldots , $c_9$ are constants.
-Equation~\ref{chap9:eq-deflection} solves an issue of symmetry which was observed with the deflection function proposed by \cite{Przemieniecki68}.
+Equation~\ref{chap9:eq-deflection} solves an issue of symmetry which was observed with the deflection function proposed by \cite{Przemieniecki85}.
 These constants can be evaluated in terms of the displacements and slopes at the three corners of the triangular plate using
@@
 \label{chap9:Ke}
-\textbf{K}_e = (\textbf{C}^{-1})^T \int_v \textbf{D}^{T} \boldsymbol\chi \textbf{D} dV \textbf{C}^{-1}.
+\textbf{K}_e = (\textbf{C}^{-1})^T \int_v \textbf{D}^{T} \boldsymbol\chi \textbf{D} dV \, \textbf{C}^{-1}.
 \end{equation}
+
+Similarly to the membrane stiffness matrix, the bending stiffness matrix in the global frame is eventually obtained using the rotation matrix of the element: $\textbf{K} = \textbf{R} \textbf{K}_e \textbf{R}^{T}$.
 
@@
 We first sample the flat surface of each element by recursively dividing each triangle into four smaller ones and the deflection of each new vertex is computed using \eqref{chap9:eq-deflection} according to the displacements and slopes at the three vertices of the triangular element.
-This process of subdivision allows us to render each shell as a curved triangle (\fig{chap9:fig-shell} (a) and (b)) and detect any collision with the curved surface of the shell using any of the classic collision detection algorithms working on flat triangles.
+This process of subdivision may also allow us to render each shell as a curved triangle (\fig{chap9:fig-shell} (a) and (b)) and detect any collision with the curved surface of the shell using any of the classic collision detection algorithms working on flat triangles.
 
@@
 	\subsection{Evaluation of displacements}
-The computation of the stiffness matrix, and subsequently the internal forces, starts with the measure of the displacements at the three vertices of every element.
+The computation of the stiffness matrix, and subsequently the internal forces, starts with the measure of the displacements $\left\{u_1 u_2 \ldots u_9 \right\} $ for every element.
 As defined by \eqref{chap9:ui}, we need to measure the displacement along the z axis and the rotations around local x and y axes for each vertex.
-For \cite{Przemieniecki68}, the deflection z is measured in the global frame.
+For \cite{Przemieniecki85}, the deflection z is measured in the global frame.
 But in a co-rotational formulation, a local frame is attached to each triangle and used to compute displacements and forces.
@@
 where $ \mathbf{u} $ is a unit vector and $ \alpha $ a scalar.
-We denote by $ \mathbf{v} $ a vector in $\field{R}^3$ and it can be shown that the quaternion product $ q \mathbf{v} q^{-1} $ yields a vector which is the rotation of $ \mathbf{v} $ by an angle $ \alpha $ around the axis $ \mathbf{u} $.
+If we denote by $ \mathbf{v} $ a vector in $\field{R}^3$, it can be shown that the quaternion product $ q \mathbf{v} q^{-1} $ yields a vector which is the rotation of $ \mathbf{v} $ by an angle $ \alpha $ around the axis $ \mathbf{u} $.
 Quaternions require less storage than rotation matrices and are known to be more numerically stable.
@@
 From there, our approach to measure the rotations around local x and y axes is simple.
-At runtime and for each vertex, we compute the rotation that we need to apply to transform the z axis of the triangle's frame into the z axis of the vertex's frame (see listing \ref{chap9:qDiffZ}).
+At runtime and for each vertex, we compute the rotation that we need to transform the z axis of the triangle's frame into the z axis of the vertex's frame (see listing \ref{chap9:qDiffZ}).
 The differences with the same quantities computed for the rest position during the initialisation lead to the desired displacements.
@@
 	\subsection{Inversion of matrix $ \mathbf{C} $}
-\cite{Przemieniecki68} described the condition of non-inversion for the matrix $\mathbf{C}$ as the following:
+\cite{Przemieniecki85} described the condition of non-inversion for the matrix $\mathbf{C}$ as the following:
 \TODO{add condition of inversion... but need the book :)}
@@
 Each triangle of a mesh obtained from a regular grid of points fits this condition of non-inversion (right triangles).
-We fixed this limitation by choosing a different local coordinate system that the one introduced by \citeauthor{Przemieniecki68}.
+We fixed this limitation by choosing a different local coordinate system that the one introduced by \citeauthor{Przemieniecki85}.
 
@@
 \end{equation}
-where $ w_i $ is a weight factor and $ f_i $ is the evaluation of $ f $ at the Gauss point $ i $.
-Different numbers and locations may be chosen for the integration points depending on the order of accuracy desired.
+where $ w_i $ is a weight factor and $ f_i $ is the evaluation of $ f $ at the integration point $ i $ (also called Gauss points).
+The choice of the number and the locations of the integration points depends on the order of accuracy desired.
 We classically choose 3 integration points located at the middle of each edge of the triangle, for which the weight factors $ w_i $ are $ 1/3 $.
@@
 \begin{equation}
-\textbf{K}_e = \dfrac{A t}{3} (\textbf{C}^{-1})^T \sum_{i=1}^3 \textbf{D}_i^{T} \boldsymbol\chi \textbf{D}_i  \textbf{C}^{-1},
+\textbf{K}_e = \dfrac{A t}{3} (\textbf{C}^{-1})^T \sum_{i=1}^3 \textbf{D}_i^{T} \boldsymbol\chi \textbf{D}_i  \, \textbf{C}^{-1},
 \end{equation}
@@
 \item Compute the local stiffness matrix $\textbf{K}_e$ of the element as $\textbf{K}_e = \displaystyle{\sum^3_{i=1}} \textbf{J}_i \boldsymbol\chi \textbf{J}_i^T$
-\item Transform the local element stiffness matrix into the global frame and add it to a global stiffness matrix
+\item Transform the local element stiffness matrix into the global frame and add it to the global stiffness matrix
 \end{enumerate}
@@
 It is worth noting that a high resolution mesh may be obtained by recursively subdividing each triangle.
-However, the vertices are closely related to the structure of the mechanical mesh and rendering artifacts may appear.
+However, in this case the vertices are closely related to the structure of the mechanical mesh and rendering artifacts may appear.
 As we can notice on \fig{chap9:fig-rendering}, the quality of the visualisation may be improved by using a totally different mesh.
@@
 \subsection{Allowing a curved rest configuration}
-In \cite{Przemieniecki68}, the rest and initial configurations are both assumed to be the flat one.
+In \cite{Przemieniecki85}, the rest and initial configurations are both assumed to be the flat one.
 The only way to have an initial deformed shape is to apply a torque on the vertices of the shells.
@@
 \subsection{Possible extension: parallel implementation on GPU}
-If we look closely all the steps required to compute the local stiffness matrix for each element (see section \ref{chap9:summary} for a summary), one may notice that the operations carried out for a given shell element are independent of the other elements.
+If we look closely at all the steps required to compute the local stiffness matrix for each element (see section \ref{chap9:summary} for a summary), we can notice that the operations carried out for a given shell element are independent of the other elements.
 This feature makes our co-rotational shell FEM implementation highly parallelisable.
@@
 One needs to choose wisely the type of memory to be used for each variable and optimise the access patterns.
-However, it seems that there is no major obstacle to a GPU implementation.
+However, it seems that there is no major obstacle to a GPU implementation and our co-rotation shell FEM will eventually be ported to GPU.
 
@@
 \begin{itemize}
- \item Young Modulus $E = 1.092 \times 10^6$
+ \item Young Modulus $E = 1.092 \times 10^6\,$Pa
  \item Poisson's ratio $\nu = 0.3$
@@
 Phacoemulsification consists in using a surgical device which tip vibrates at an ultrasonic frequency to emulsify the natural lens material and then aspirate the fragments of the cortical material.
-After the removal of the diseased lens, an intra-ocular lens is implanted into the eye, through a small incision (about 2 mm) using a foldable intra-ocular lens (see \fig{chap9:fig-surgery}).
+After the removal of the diseased lens, an intra-ocular lens is implanted into the eye, through a small incision (about $2\,$mm) using a foldable intra-ocular lens (see \fig{chap9:fig-surgery}).
 The foldable lens, usually made of acrylic material, is then implanted within the lens capsule through the incision used during phacoemulsification.
@@
 		\hline
-		\centering $1 MPa$ & \centering $0.42$ & \centering $1.2 g/cm^3$ \tabularnewline
+		\centering $1\,$MPa & \centering $0.42$ & \centering $1.2\,$g/cm$^3$ \tabularnewline
 		\hline
@@
 During these stages of the simulation, complex contacts occur and consist of self collisions of the lens as well as collisions between the lens and the injector and later with the capsule.
-To solve the contacts we use the contact warping method proposed by \cite{Saupin08} as it offers an efficient way to compute physically correct contact responses in the case of co-rotational models \TODO{is it still true?}.
+To solve the contacts we use the contact warping method proposed by \cite{Saupin08} as it offers an efficient way to compute physically correct contact responses in the case of co-rotational models \TODO{is it still true?
+I'm not even sure we used this technique in the final submission to ISBMS}.
 As adhesions between the haptics and the body is often observed in surgery, friction is also taken into account in the contact response process.
@@
 The shape of the intra-ocular lens remains very close to that of a real one during all stages of the simulation: within the injector, during the ejection phase, and when in place within the capsule.
-Due to the high stiffness and low mass of the lens, a direct sparse solver was used at each time step (dt $= 0.01$\,s) rather than an iterative solver, resulting in a more accurate and more stable simulation, to the detriment of computation times (about 5 FPS for the complete simulation, and about 10 FPS for the deformation only, on a 2.4 GHz processor).
+Due to the high stiffness and low mass of the lens, a direct sparse solver was used at each time step (dt $= 0.01\,$s) rather than an iterative solver, resulting in a more accurate and more stable simulation, to the detriment of computation times (about 5 FPS for the complete simulation, and about 10 FPS for the deformation only, on a 2.4 GHz processor).
 

### 2010-09-13 Thesis/TitlePage.tex
- Update 13 September -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@174 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
       \textit{Reviewers:}	& Thomas Sangild \textsc{S\o rensen}          & - & University of Aarhus, Denmark\\
-				& Bernard \textsc{}		& - & INRIA (Shaman)\\
-      \textit{Advisor:}	& Stéphane \textsc{Cotin}		& - & INRIA (Shaman), France\\
-      \textit{President:}	& Bernard \textsc{}		& - & INRIA (Shaman)\\
-      \textit{Examinators:}   & Bernard \textsc{}          & - & INRIA (Shaman)\\
+				& Georges \textsc{Dumont}		& - & INRIA (Bunraku team), Rennes, France\\
+      \textit{Advisor:}	& Stéphane \textsc{Cotin}		& - & INRIA (Shaman team), Lille, France\\
+      \textit{President:}	& Laurent \textsc{Grisoni}		& - & INRIA (Mint team), Lille, France\\
+      \textit{Examinators:}   & Tamy \textsc{Boubekeur}          & - & Telecom ParisTech, Paris, France\\
       				& Bernard \textsc{}			& - & INRIA (Shaman)\\

### 2010-09-14 Thesis/Chapter10.tex
- Update 14 September -
Chapter 7, 9 and 10 finished and sent for review



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@175 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{shortAbstract}
-A short abstract for the upcoming chapter \\
-\OC{MICCAI 2010}
+In the previous chapter, we proposed a co-rotational formulation for shell elements by extending a classical in-plane triangular finite element approach.
+This simple shell element can efficiently handle both membrane and offplane forces (bending and twist).
+The validity of our approach has been demonstrated though comparison with theoretical results.
+It was then applied it to a rather complex application: the simulation of intra-ocular lens implant deployment in a cataract surgery simulation.
+These preliminary results are very encouraging and show the potential of such models.
+Examples include the modelling of hollow anatomical structures (stomach, colon, etc.), the simulation of cardiac valve leaflets, and blood vessels.
+However, to model the deformation of complex anatomical structures using shell elements, the first step is to describe its surface with curved patches.
+In the next section, we propose to study how to approximate the surface of anatomical structures with (curved) shell elements.
+This work was the object of one publication \citep{Comas2010b}.
 \end{shortAbstract}
@@
 Our idea is the following: while many flat triangles are required to describe highly curved surfaces, fewer triangular shell elements are needed to describe the given geometry with the same precision since they can be curved.
-In the following, we assume that we have a high resolution triangular mesh obtained from a binary segmented image of the object we want to simulate (via a Marching Cube algorithm for instance, \cite{Lorensen87}).
 Our goal is to create a mesh featuring the optimal number of shell elements while staying as close as possible to our targeted geometry.
-For this reason, we need a technique to simplify the high resolution mesh of the object to be simulated.
+Literature about volumic mesh generation algorithms is fairly dense.
+However, there are only a few that are concerned about the generation of meshes over curved surfaces.
+One of the most widely used techniques for the creation of surface meshes is the plane to surface transformation method \citep{Zienkiewicz71}, mesh is first generated on a two-dimensional domain and then mapped onto the surface.
+If this method gives reasonably good meshes on smooth surfaces, the results are usually quite poor with more complex curved surfaces.
+The finite elements may be generated directly on the curved surfaces based on the advancing front technique \citep{Lo85,Lau96}.
+The main issue with this approach is that an analytical description of the geometry is needed, which is not the case in medical simulation.
+Another method consists of using an a priori error estimator to build an adaptive mesh generation \citep{Baumann97}.
+However, the tolerance of this indicator should be chosen depending on the desired accuracy of the finite element solution, and therefore requires some knowledge about the problem in order to choose an effective tolerance.
+\cite{Bechet02} start from an existing triangular mesh created with a CAD (computer-aided design) software and refine and smooth the mesh based on element quality and surface curvature.
+While all those methods allow their authors to get satisfactory meshes over curved surfaces according to their needs, they all make use of flat elements to mesh geometries and do not generate actual shells.
 
-\bigskip
+In fact, it is interesting to note that this process of meshing a curved surface is similar to the reconstruction of the surface of objects in computer vision.
+Indeed, calculating curvature maps of 3D surfaces represented as digitised data (point clouds or triangulated meshes) has been extensively studied.
+One of the most common approach is to use continuous surface patches \citep{Kolb95,Douros02}.
+The surface is locally approximated by an analytic representation called surface patch, usually chosen to be an implicit surface.
+\cite{Tang99} presented an elegant and non noise-sensitive approach that infers sign and direction of principal curvatures directly from the input and the authors used this information for coherent surface and curve extraction from 3D data.
 
-\TODO{state of the art to be added... }
+However, we need to keep in mind that our situation is substantially different than all previously mentioned geometrical approaches as we want to model the deformation of the structure.
+In that regard, the curvature of the surface has a physical meaning: it represents the mid-surface of the shell.
+By keeping this in mind, we propose an approach to mesh a curved surface based on the same polynomial shape function \eqref{chap9:eq-deflection} used in our co-rotational shell FEM.
 
@@
 
-However, we need to keep in mind that our situation is substantially different than all previously mentioned geometrical approaches as we want to model the deformation of the structure.
-In that regard, the curvature of the surface has a physical meaning: it represents the mid-surface of the shell.
-We propose to approximate the surface of anatomical structures with shell elements whose each surface is described by the shape function used in our shell formulation.
-
-The first step in the process of generating a shell-based mesh is an important decimation of the high resolution mesh, using quadric edge collapse technique implemented in \cite{Meshlab}.
+In the following, we assume that we have a high resolution triangular mesh obtained from a binary segmented image of the object we want to simulate (via a Marching Cube algorithm for instance, \cite{Lorensen87}).
+We propose a technique that can simplify the high resolution mesh of the object to be simulated and approximate the surface of anatomical structures with shell elements whose each surface is described by the shape function used in our shell formulation.
+Thus, the first step in the process of generating a shell-based mesh is an important decimation of the high resolution mesh, using quadric edge collapse technique implemented in \cite{Meshlab}.
+MeshLab is an open source program for the processing and editing of unstructured 3D triangular meshes.
 The algorithm tries as much as possible to preserve mesh boundaries and generates high quality triangular elements.
@@
 
-The efficiency of the method is illustrated through shell-based reconstruction and real-time simulation of the deformations of various anatomical structures.
-We also present preliminary results on the simulation of an angioplasty procedure.
+We achieved our objectives to propose a versatile solution which can simulate, in real-time, thin objects with various shapes and material properties with good accuracy.
+The efficiency of the method was illustrated through shell-based reconstruction and real-time simulation of the deformations of various anatomical structures and other thin objects (like an intra-ocular implant).
+We also presented preliminary results on the coupling between solid (tetrahedra) and thin objects (shells) for the advanced modelling of anatomical structures via the simulation of an angioplasty procedure.
 
@@
 
-\TODO{Did we answer our problematic? Limitations? }
+The finite element procedure that we propose is largely developed based on physical understanding without the use of mathematical shell theories.
+Indeed, our shell FEM was obtained by simply superimposing a plane stress and bending energies.
+According to \cite{Chapelle03}, this type of approach has a limited level of accuracy.
+The effective study of shell structures clearly requires a deep physical understanding of shell structural behaviours.
+The complexity of the physical behaviours of shells requires advanced mathematical analyses from theories that are still in development.
+The interested reader may refer to the book of \cite{Chapelle03} for more details on shell theory.
+If the range of application of our shell formulation might be somewhat limited, we did not find out any inconsistencies through our tests.
+We also need to keep in mind that the accuracy required in medical simulation is not the same than the one demanded in structural analyses for instance.
+Moreover, the additional complexity of a shell finite element formulation based on true shell theory may also very well jeopardise our real-time constraints.
+Therefore, we do not think that the use of this simple shell formulation significantly diminishes our contribution to the field of medical simulation.
+Indeed, to our knowledge, our co-rotational shell FEM is the first description of a shell finite element model applied to simulating the deformation of thin anatomical structures.
+
+Nevertheless, a few improvements are conceivable.
+Large strain could be allowed by using a non-linear strain tensor and more complex material laws (non-linear, viscoelasticity, etc.) could also be added to our shell formulation.
+However, those changes are substantial and their implementations are not straightforward.
+In addition, it is not clear how the benefits would be the greatest: whether from enhancing our model to a non-linear formulation or completely changing to another model based on true shell theory (where membrane and bending energies are not separated).
+The answer may depend on the type of constraints applied on the object.
+Also (as already mentioned section \ref{chap9:GPU}), the real-time capacity of our algorithm could be further improved by implementing our shell FEM on GPU.
+Although an optimal parallel re-implementation is always challenging, we have already reviewed why there was no major obstacle to a GPU implementation.
+
+Although we were able to mesh complex anatomical structures with a fairly small number of shell elements, our approach suffers from two major drawbacks.
+In our current implementation, if the error measured with the Hausdorff distance is greater than a threshold (which is also difficult to choose), all triangular elements are subdivised into 4 new triangles and each of the new vertices is moved towards the targeted geometry based on our heuristic algorithm.
+Obviously, this is not optimal as the error between the two meshes may very well be local and therefore the use of a global mesh refinement may be very limited.
+The second drawbrack is that we also do not check the quality of the elements during our mesh process.
+If some elements are stretched from the heavy decimation, these elements are not corrected during the process of remeshing.
+Our algorithm could even worsen the situation by splitting those stretched elements into four smaller (and more stretched) elements.
+One solution for the latter issue would be to enhance our algorithm with a relaxation phase where a repulsion force is applied between each vertex (constrained to remain on the surface of the object).
+This physical approach would allow a more even distribution of the vertices across the surface and would contribute to obtain better quality elements overall.
+Yet, meshing a surface with curved elements is a very complex problem and still an active area of research.
+In contrast to all previous geometrical techniques, we suggested a more physical approach consistent with our shell FEM formulation and obtained good results in meshing complex anatomical structures.

### 2010-09-14 Thesis/Chapter7.tex
- Update 14 September -
Chapter 7, 9 and 10 finished and sent for review



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@175 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{shortAbstract}
-A short abstract for the upcoming chapter
+The human body is composed of various deformable anatomical structures and finding appropriate soft-tissue models for all these type of structures is challenging.
+In the previous two chapters, we described an efficient non-linear FEM formulation for simulating the deformation of solid organs (like brain, liver, prostate etc.).
+But there is also a need to simulate the deformation of hollow organs, whose volume is negligible compared to their surface area, such as colon, blood vessels, stomach etc.
+The solid models previously mentioned are not appropriate and specific approaches are required for these thin anatomical structures.
+In this chapter, we will review diverse techniques applied for the simulation of thin objects.
+These different approaches may be grouped into three main categories: (1) mass-spring models, (2) approaches which derive a bending energy from geometrical considerations and (3) methods based on the equations of continuum mechanics.
+Although very few models have been proposed for simulating the deformation of thin anatomical structures in the field of medical simulation, we will try to emphasize medical applications whenever possible.
 \end{shortAbstract}
@@
 In fact in the field of medical simulation, very few models have been proposed for simulating, in real-time, the deformation of thin anatomical structures whose volume is negligible compared to their surface area.
-Examples include hollow structures, such as the wall of blood vessels, or membranes, such as the Glisson?s capsule surrounding the liver.
+Examples include hollow structures, such as the wall of blood vessels, or membranes, such as the Glisson's capsule surrounding the liver.
 It is also of particular interest to us for modelling the colon in our colonoscopy simulator and for simulating the deployment of the implant in cataract surgery.
@@
 \TODO{why is this not appropriate?
-High-order elements could do the trick to follow the curvature...} Hence, specific models for thin objects were developed.
+High-order elements could do the trick of following the curvature...} Hence, specific models for thin objects were developed.
 

### 2010-09-14 Thesis/Chapter9.tex
- Update 14 September -
Chapter 7, 9 and 10 finished and sent for review



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@175 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{shortAbstract}
-A short abstract for the upcoming chapter \\
-\OC{ISBMS 2010}
+In the field of medical simulation, very few models have been proposed for simulating, in real-time, the deformation of thin anatomical structures.
+In fact, to our knowledge, no model based on the equations of continuum mechanics was described and applied to simulating the deformation of objects.
+In this chapter, we propose to rely on continuum mechanics to accurately describe the complex deformations of the implant.
+Our approach extends the co-rotational method used in finite element analysis of in-plane deformations to incorporate a bending energy.
+We will first describe the model in details before discussing its implementation in the open source framework SOFA.
+Lastly, this co-rotational shell FEM was applied to the simulation of the lens deployment during catarct surgery.
+This simulation also accounts for the complex contacts that take place during the injection phase.
+This work was the object of two publications \citep{Comas2010a,Comas2010b}.
 \end{shortAbstract}
@@
 
-\subsection{Possible extension: parallel implementation on GPU}
+\subsection{Possible extension: parallel implementation on GPU}	\label{chap9:GPU}
 If we look closely at all the steps required to compute the local stiffness matrix for each element (see section \ref{chap9:summary} for a summary), we can notice that the operations carried out for a given shell element are independent of the other elements.
@@
 A uniform load is then applied on the square and the maximum deflection $z_{max}$ at the centre can be calculated.
-Several simulations are performed with increasing load values $q$ (ranging from $1$ to $5 N/m^2$) and the following parameters were used:
+Several simulations are performed with increasing load values $q$ (ranging from $1$ to 5 N/m$^2$) and the following parameters were used:
 \begin{itemize}
@@
  \item Poisson's ratio $\nu = 0.3$
- \item edge of the square $L = 10\,m$
- \item thickness $h = 0.1\,m$
+ \item edge of the square $L = 10\,$m
+ \item thickness $h = 0.1\,$m
  \item pressure q is a uniformly distributed load per unit area
@@
 \end{figure}
-
-\section{Discussion}
-
-We proposed a co-rotational formulation for shell elements by extending a classical in-plane triangular finite element approach.
-This simple shell element can efficiently handle both membrane, bending and twist forces.
-The validity of our approach has been demonstrated though comparison with theoretical results.
-It was then applied it to a rather complex application: the simulation of intra-ocular lens implant deployment in a cataract surgery simulation.
-These preliminary results are very encouraging and show the potential of such models.
-Examples include the modelling of hollow anatomical structures (stomach, colon, etc.), the simulation of cardiac valve leaflets, and blood vessels.
-However, to model the deformation of complex anatomical structures using shell elements, the first step is to describe its surface with curved patches.
-In the next section, we propose to study how to approximate the surface of anatomical structures with shell elements.

### 2010-09-16 Thesis/Chapter2.tex
- Update 16 September -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@176 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \section{Tissue characterisation}
-One of the main obstacles in developing realistic organ models is the lack of data on the material properties of live organ tissues.
-Measuring and characterising in vivo organ properties is a highly challenging task, but is a requirement for realistic organ modelling.
-Indeed, as we have seen in sections \ref{chap2:orthotropic} and \ref{chap2:non-linear}, constitutive equations require the experimental determination of material constants.
-Organ models with incorrect material properties will affect training in surgical simulator systems \citep{Sedef06}.
 
-The constitutive equations are used to define idealised materials, which represent different aspects of the mechanical behaviour of natural materials.
-However, no real material is known to behave exactly as one of these mathematical models.
+\ON If continuum mechanics does teach us how to model idealised continuous structures, modelling soft tissues remains very complex.
+Indeed, in practice soft tissues are complex structures: they do not consist of a simple continuous piece of material, they are made of various constituents.
+Mechanical properties of soft tissues not only depends on the properties of these constituents, but how they are arranged relative to each other.
+This elaborate arrangement often leads to material properties for the global structure that are non-homogenous, non-linear, anisotropic and even viscoelastic.
+
+As we know, \OFF the constitutive equations are used to define idealised materials, which represent different aspects of the mechanical behaviour of natural materials.
+\ON However, in practice, \OFF no real material is known to behave exactly as one of these mathematical models.
 In particular, few biological tissues obey Hooke's law.
@@
 These various mechanical aspects of materials make realistic modelling of anatomical structures fairly challenging.
-In medical simulation, where modelling must occur in real-time or close to real-time, very often simplifications have to be made.
-As we will see in chapter \ref{chap4} and \ref{chap7}, numerous models are available in the literature to describe physics of objects, from fairly simple and naive approaches to more complex and thorough representations.
-However, the most common technique to solve equations of continuum mechanics is the Finite Element Method.
-Moreover, since the two main contributions of this thesis deal with finite element methods formulations, the general principles of this method will be detailled in the next section.
+\ON The development of algorithms that allow the interactive computation of the deformation of soft tissues is very difficult.
+Therefore, \OFF in medical simulation, where modelling must occur in real-time or close to real-time, very often simplifications have to be made \ON and only the key characteristics of the material that are relevant for the objective of the simulation are considered and simulated.
+Experiments have to be performed to check the relevancy of the chosen constitutive model and to determine its parameters.
+At last, an optimised implementation of the model is required to perform in real-time.
+
+\OFF One of the main obstacles in developing realistic organ models is the lack of data on the material properties of live organ tissues.
+Measuring and characterising in vivo organ properties is a highly challenging task, but is a requirement for realistic organ modelling.
+Indeed, as we have seen in sections \ref{chap2:orthotropic} and \ref{chap2:non-linear}, constitutive equations require the experimental determination of material constants.
+Organ models with incorrect material properties will affect training in surgical simulator systems \citep{Sedef06}.
+\ON Some devices exist for inert materials which allow the measure of Young's modulus, shear modulus and Poisson's ratio.
+However, the same measures are much more difficult to obtain for living tissues.
+In particular, experiments confirmed the important differences between in vitro and in vivo measures, as demonstrated by \cite{Kerdok06} with the viscoelastic properties of the liver for instance.
+Absence of blood perfusion and difference in temperature are two of most important factors to explain these differences.
+Obviously, in vivo constraints are very heavy and developing experiments and devices for in vivo measures of soft tissue is a very complex task.
+The preparation of samples and their testings (tensile, compression, thermal effects, rate of loading etc.) are not straightforward procedures.
+
+A non-invasive technique called \emph{elastography} was introduced recently to measure Young's modulus.
+The first device based on this technique was commercialised by Echosens in 2003 \cite{Bastard09}.
+The techniques of elastography may be classified in two categories: static and dynamic elastography.
+In static elastography, a quasi-static constraint is applied onto the surface of the tissue and its deformation is measured.
+If the constraint is known, the Young's modulus may be obtained via Hooke's law.
+Dynamic elastography relies on shear waves propagation in the tissue.
+Shear waves are generated by an acoustic driver and the displacement caused by the waves is determined by ultrasounds, magnetic resonance imaging (MRI) or optical tomography.
+As an example, magnetic resonance elastography is a quantitative technique which allow the generation of maps describing the shear modulus within the tissues (see \fig{chap2:fig-elastography_liver}).
+However, elastography is still in development and only a few techniques of elastography are actually used clinically.
+In the future, elastography might be able to provide a map of the material properties of tissues which may be used as input data for any computational model.
+This technique would be a step towards patient-specific simulation.
+As diseased organs are often much stiffer than healthy structures, the integration of these data would greatly increase the fidelity of the simulation.
+%
+\begin{figure}
+\centering
+\includegraphics[width=12cm]{chapter2/elastography_liver.pdf}
+\caption {\ON Magnetic resonance elastography of the liver in a healthy volunteer and a patient with cirrhosis. Image courtesy of \cite{Talwalkar07}.}
+\label{chap2:fig-elastography_liver}
+\end{figure}
+
+Yet, precised characterisation of soft tissue is not sufficient to achieve a good simulation.
+The development of a simplified or optimised version of the model suitable for real-time computation is also mandatory.
+\OFF As we will see in chapter \ref{chap4} and \ref{chap7}, numerous models are available in the literature to describe physics of objects, from fairly simple and naive approaches to more complex and thorough representations.
+However, the most common technique to solve equations of continuum mechanics is the finite element method.
+Moreover, since the two main contributions of this thesis deal with finite element methods formulations, the general principles of this method will be detailed in the next section.

### 2010-09-16 Thesis/Chapter4.tex
- Update 16 September -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@176 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\TODO{add drawbacks of meshless methods: supposedly good for handling topological changes but other issues. Couldn't really find drawbacks in meshless' literature.}
+\ON The main advantage of meshless methods is of course in the absence of a mesh.
+This advantage of not having to go through the complex process of meshing was often claimed by meshless methods partisans.
+It also allows much easier cutting as it saves the cost of a computationally demanding remeshing around the cut.
+Yet, in contrast of finite element methods, meshless techniques are young and still in development.
+Boundary conditions are an aspect of the numerical solution of partial equations where meshless methods have many difficulties for instance.
+According to \cite{Horton10}, if meshless methods are accurate in terms of overall reaction forces, they are not as good with individual displacements or forces.
+Therefore the authors suggest to use a meshless algorithm to fill the interior sections of an organ (where a local discrepancy has minor consequences) and use the finite element method for the boundary.
+As a conclusion, if meshless methods are promising, the technique does not seem as mature as the finite element method for modelling soft tissues.
+\OFF

### 2010-09-28 Thesis/Chapter3.tex
- Update 16 September -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@176 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 28 September -
Major update: lot of fixes on the first part that has been reviewed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@177 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 %
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=8cm]{chapter3/errorWithSubdivision.pdf}
-\end{center}
-\caption[Error introduced by the division into elements]{The division of the domain into elements is not always perfect.}
+\begin{figure}[ht]
+\centering 
+\subfloat[ ]{\includegraphics[width=4.5cm]{chapter3/errorWithSubdivision1.pdf}}
+\hfill 
+\subfloat[ ]{\includegraphics[width=4.5cm]{chapter3/errorWithSubdivision2.pdf}}
+\hfill 
+\subfloat[ ]{\includegraphics[width=4.5cm]{chapter3/errorWithSubdivision3.pdf}}
+\caption[Error introduced by the division into elements]{The division of the domain (a) into elements is not always perfect (b).
+The discretisation error may be reduced by using smaller elements (c).
+However, this improvement is obtained to the detriment of computational efficiency as the number of algebraic equations to solve will increase.}
 \label{chap3:fig-errorWithSubdivision}
@@
 Usually, the number of elements increases with the complexity of the geometry.
-For instance, if one part of the domain is thiner (see \fig{chap3:fig-sizeOfElementMeshing}), the size of the elements must be reduced in order to tile this particular part, which increases the total number of elements, and hence  the number of algebraic equations to solve.
+For instance, if one part of the domain is thiner (see \fig{chap3:fig-errorWithSubdivision} (c)), the size of the elements must be reduced in order to tile this particular part, which increases the total number of elements, and hence the number of algebraic equations to solve.
 However, adding elements is sometimes desirable.
@@
 It does not mean that precision is not required, simply that the time constraints will necessarily limit the accuracy of the solution.
-%
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=13cm]{chapter3/sizeOfElementMeshing.pdf}
-\end{center}
-\caption[Finer mesh required for complex geometry]{By adding details on the tree's geometry \textbf{(right)}, the surface delimited by the tree becomes more difficult to mesh and would require smaller elements to tile all additional small branches.}
-\label{chap3:fig-sizeOfElementMeshing}
-\end{figure}
 
@@
 We will now introduce a few direct methods in the context of finite element methods.
-% Methods such as Gauss-Jordan elimination or LU decomposition will not be discussed in this section for instance for their lack of efficiency with sparse matrices.
 
-\paragraph*{LU decomposition.}
+\paragraph*{LU decomposition.}	\label{chap3:LUdecompo}
 The LU decomposition is a matrix decomposition which writes a matrix as the product of a lower triangular matrix and an upper triangular matrix. In other words, $ \mathbf{A} $ may be written as:
@@
 
-\paragraph*{\TODO{Cholesky decomposition?}}
+\ON
+\paragraph*{Cholesky decomposition.}	\label{chap3:CholeskyDecompo}
+Cholesky decomposition allows to write a symmetric and definite positive matrix as the product of a lower triangular matrix with strictly positive diagonal entries by the transpose: 
+\begin{equation}
+\label{chap3:Cholesky}
+\mathbf{A} = \mathbf{L} \mathbf{L}^T,
+\end{equation}
+where $ \mathbf{L} $ is lower triangular with strictly positive diagonal entries.
+In general, Cholesky decompositions are not unique.
+Similarly to LU decomposition, a Cholesky decomposition may be used to break one set of equations into two successive but simpler systems.
+\OFF
 
@@
 Later, the \emph{biconjugate gradient method} was introduced as a generalisation of the ordinary conjugate gradient to solve systems where the matrix $ \mathbf{A} $ was not necessarily symmetric and positive-definite (see \cite{Belytschko00} for details).
+
+\ON
+\paragraph*{Preconditioners}
+Unfortunately, it is common for the matrix $ \mathbf{A} $ not to be well conditioned.
+Consequently, the convergence of the chosen iterative method is often slow.
+In order to achieve an accelerated convergence, we may use a \emph{preconditioner}.
+Although the formation and the study of preconditioners are out of the scope of this PhD thesis, preconditioners are a key element for solving systems of equations.
+To be effective, a preconditioner $ \mathbf{P} $ should be a good approximation of  $ \mathbf{A} $ and the following preconditioned system must be easier to solve than the original system \eqref{chap3:linearEquation}:
+\begin{equation}
+\label{chap3:preconditionedSystem}
+\mathbf{P}^{-1} \mathbf{A} \mathbf{x} = \mathbf{P}^{-1} \mathbf{b}.
+\end{equation}
+Two criteria should also be considered when forming a preconditioner: (1) the ease with which it may be evaluated and (2) its memory requirements.
+The easiest preconditioner would be the identity since $ \mathbf{P}^{-1} = \mathbf{I} $ in this case.
+Of course, such a conditioner leads to the resolution of the original system and is therefore totally useless.
+In constrast, the choice of $ \mathbf{P} = \mathbf{A} $ leads to the solution of the system \eqref{chap3:linearEquation} in one single iteration.
+In practice, an optimal preconditioner is to be found between these two extremes to minimise the number of iterations required by the solver while keeping the evaluation of the preconditioner simple.
+Finding a good preconditioner for all types of problems is difficult.
+A preconditioner is only useful if the overall time spent on both the formation of  $ \mathbf{P} $ and the resolution of \eqref{chap3:preconditionedSystem} is less than the time spent at solving \eqref{chap3:linearEquation} alone.
+
+One of the simplest preconditioner is the \emph{Jacobi preconditioner} where the preconditioner is chosen to be the diagonal of the matrix $ \mathbf{A} $.
+This choice is efficient in cases where $ \mathbf{A} $ is diagonally dominant.
+Sparse approximations of a LU or Cholesky decompositions may also be used as preconditioners (see sections \ref{chap3:LUdecompo} and \ref{chap3:CholeskyDecompo} for more details).
+\OFF

### 2010-09-28 Thesis/Chapter4.tex
- Update 28 September -
Major update: lot of fixes on the first part that has been reviewed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@177 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-At the beginning of computer simulation, users were immersed into environments that were merely a decor.
+%At the beginning of computer simulation, users were immersed into environments that were merely a decor.
 If it could virtually take the user somewhere else, a static scene quickly showed its limits.
@@
 
-In the field of medical simulation, the correctness of deformation is often crucial.
+\ON
+The challenges faced by the field of medical simulation were reviewed in details in the introductory chapter \label{chap1}.
+Hence, we know that at the very least, the deformation of solid organs must appear physically realist in medical simulators.
+In addition, this objective has to be achieved within heavy time constraints.
+Indeed, although each step of a simulation alone (deformation of organs, collision detection, contacts processing or visualisation) is computationally demanding, the whole simulation process must be computed in real-time or close to real-time.
+Of particular interest to us is the accurate modelling of the deformation of anatomical structures.
+The best mathematical model for predicting the deformation of a continuous structure is provided by continuum mechanics.
+If the finite element method is classicaly used in mechanical engineering for its ability to solve the equations of continuum mechanics fairly accurately, its development in medical simulation was more progressive.
+The main reason for this is that the finite element method is computationally very demanding and not easily compatible with the time constraints required in interactive simulation.
+Very often simplifications have to be made.
+Since medical simulation may find applications in various areas (such as medical training, patient-specific planning or per-operative guidance), the requirements in terms of accuracy as well as time constraints may be different.
+Consequently, a range of models has been developed over time to suit all usages.
+
+For instance, the correctness of deformation is crucial for patient-specific planning and per-operative guidance.
+Let us give an example where simulation can lead the surgeon towards its goal during surgery: the removal of a brain tumor.
+Prior surgery, images of the patient's brain were acquired and the tumor was localised.
+To access the brain during surgery, the surgeon must first drill a hole into the patient's skull.
+However, drilling a hole releases intra-cranial pressure and this change in internal pressure deforms the brain.
+\OFF This phenomenon is called \emph{brain shift}.
+The overall deformation is quite complex and the displacements depend on local elasticity of the brain, the size and the location of the tumor.
+\ON Removing the tumor relies \OFF heavily on knowing the spatial relation between the patient's brain and the images acquired prior to surgery.
+\ON And because the brain deforms during surgery, the location and the shape of the tumor will no longer match those of the images acquired.
+\OFF Therefore, a large brain shift, if not corrected, will result in inaccuracies in the surgical procedure and has the potential to cause damage to normal tissue.
+A solution is to accurately model the mechanical response of the patient's brain to predict the actual deformation and hence keep the knowledge of the tumor's location.
+\ON If the accuracy is obviously of prime importance in this situation, computation of the deformation does not necessarily need to be carried out in real-time, a reasonable delay is acceptable (a couple of minutes).
+\OFF
+
+%In the field of medical simulation, the correctness of deformation is often crucial.
 This is particulary true in the cases of per-operative guidance.
@@
 
-Conversely, when medical simulation is applied to training, such a precision in terms of distance is not required.
-The physician would not even visually notice reasonable errors as soon as it looks realistic enough.
-However, in some medical procedures, the sense of touch and the forces that the physician can feel are a fey factor for the success of the operation.
-For instance, during a colonoscopy procedure, force feedback gives precious information about what type of loop is forming and the physician can then react accordingly to prevent loop formation.
+Conversely, when medical simulation is applied to training, such a precision in terms of \ON deformation is not always \OFF required.
+\ON A visually realistic deformation is often sufficient for an inexperienced physician to learn the appropriate gestures required for a given medical procedure.
+However, in some cases, the feel of touch can be very important to master the procedure.
+\OFF For instance, during a colonoscopy procedure, force feedbacks give precious information about what type of loop is forming \ON during the colonoscope's insertion \OFF and the physician can then react accordingly to prevent loop formations.
 Consequently, a good colonoscopy simulator is required to provide accurate force feedbacks.
@@
 Obviously, the computation of a realistic force to be returned demands an appropriate mechanical modelling of all structures.
+In constrast of per-operative guidance, medical training demands real-time computations of the deformations.
+
+%However, the sense of touch and the forces that the physician can feel are a fey factor for the success of the operation. 
+%The physician would not even visually notice reasonable errors as soon as it looks realistic enough.
 
@@
 %
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=6.5cm]{chapter4/cubeFFD1.pdf}
-\includegraphics[width=6.6cm]{chapter4/cubeFFD2.pdf}
-\end{center}
-\caption[Deformation of objects by free-form deformation]{Let us imagine that the objects that we want to deform (spheres and cubes) are placed into a transparent parallelepiped \textbf{(left)}.
+\begin{figure}[ht]
+\centering 
+\subfloat[ ]{\includegraphics[width=6.5cm]{chapter4/cubeFFD1.pdf}}
+\hfill
+\subfloat[ ]{\includegraphics[width=6.6cm]{chapter4/cubeFFD2.pdf}}
+\caption[Deformation of objects by free-form deformation]{Let us imagine that the objects that we want to deform (spheres and cubes) are placed into a transparent parallelepiped (a).
 Control points are indicated by small white diamonds.
-The free-form deformation technique offers to compute the deformation of the objects from the positions of the control points \textbf{(right)}.}
+The free-form deformation technique offers to compute the deformation of the objects from the positions of the control points (b).
+Image courtesy of \cite{Sederberg86}}
 \label{chap4:fig-cubeFFD}
 \end{figure}
+
 The capability of locally applying deformations makes the technique strongly analogous to sculpting with clay, which was the sought target.
@@
 \end{center}
-\caption[Shape matching technique]{Shape matching allows to simulate hundreds of deformable objects in real-time.}
+\caption[Shape matching technique]{Shape matching allows to simulate hundreds of deformable objects in real-time. Image courtesy of \cite{Muller05}.}
 \label{chap4:fig-shoes}
@@
 \caption[Squeezing of a duck using shape matching]{A duck model is squeezed to demonstrate the stability of the shape matching technique.
-Although the flat shape of the duck is not physically plausible, this example exposes the impressive ability of the approach to recover from highly deformed or inverted configurations.}
+Although the flat shape of the duck is not physically plausible, this example exposes the impressive ability of the approach to recover from highly deformed or inverted configurations.
+Image courtesy of \cite{Muller05}}
 \label{chap4:fig-duck}
@@
 %
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=13.9cm]{chapter4/penguins.pdf}
-\end{center}
-\caption[Penguins animation with fast lattice shape matching]{\textbf{(Left)} $150$ cartoon penguins deforming dynamically using fast lattice shape matching\textbf{(Right)}.
+\begin{figure}[ht]
+\centering 
+\subfloat[ ]{\includegraphics[width=6cm]{chapter4/penguins1.pdf}}
+\hfill
+\subfloat[ ]{\includegraphics[width=6cm]{chapter4/penguins2.pdf}}
+\caption[Penguins animation with fast lattice shape matching]{(a) $150$ cartoon penguins deforming dynamically using fast lattice shape matching (b).
 Deformed lattices consisting of $150$ particles per penguin ($22\,500$ particles).
-The penguins can be deformed at $25$ frames per second on a Pentium4 $3.4 \,$GHz.}
+The penguins can be deformed at $25$ frames per second on a Pentium4 $3.4 \,$GHz.
+Image courtesy of \cite{Rivers07}.
+}
 \label{chap4:fig-penguins}
@@
 Therefore, \cite{Steinemann08} further improved the technique by allowing dynamic adaptive selection of levels of detail through the use of an octree-based hierarchical sampling.
-This adaptative sampling is illustrate with \fig{chap4:fig-flowers}.
 Their technique also handles efficient topological changes.
-%
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=13.9cm]{chapter4/flowers.pdf}
-\end{center}
-\caption[Flowers in the wind]{Complex scene with $40$ deforming flowers.
+% This adaptative sampling is illustrate with \fig{chap4:fig-flowers}
+%\begin{figure}[h]
+%\begin{center}
+%\includegraphics[width=13.9cm]{chapter4/flowers.pdf}
+%\end{center}
+%\caption[Flowers in the wind]{Complex scene with $40$ deforming flowers.
 A coarse sampling is employed when the flowers are moved by the wind, for a total of $5\,680$ nodes in the scene.
 However, the models are dynamically refined when touched by the user, as shown in the right image.}
-\label{chap4:fig-flowers}
-\end{figure}
+%\label{chap4:fig-flowers}
+%\end{figure}
 
@@
 \begin{center}
-\includegraphics[width=12cm]{chapter4/chainmail.pdf}
+\includegraphics[width=10cm]{chapter4/chainmail.pdf}
 \end{center}
-\caption[Deformation of a 2D chain]{Deformation of a 2D chain mail when the link indicated by an arrow is moved.}
+\caption[Deformation of a 2D chain]{Deformation of a 2D chain mail when the link indicated by an arrow is moved. Image courtesy of \cite{Gibson97}.}
 \label{chap4:fig-chainmail}
@@
 %
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=13cm]{chapter4/modalAnalysisLaparo.pdf}
-\end{center}
-\caption[Applications of modal analysis]{(\textbf{Left}) A jumping motion that leads to significant thigh and belly vibrations of human skin when rendered with modal analysis.
-(\textbf{Right}) The same hardware accelerated technique is applied to secondary tissues in a laparoscopic simulation to ease the CPU of computations.
-}
+\begin{figure}[ht]
+\centering 
+\subfloat[ ]{\includegraphics[height=6cm]{chapter4/modalAnalysisLaparo1.pdf}}
+\hspace{0.2cm}
+\subfloat[ ]{\includegraphics[height=6cm]{chapter4/modalAnalysisLaparo2.pdf}}
+\caption[Applications of modal analysis]{(a) A jumping motion that leads to significant thigh and belly vibrations of human skin when rendered with modal analysis.
+(b) The same hardware accelerated technique is applied to secondary tissues in a laparoscopic simulation to ease the CPU of computations.
+Image courtesy of \cite{James02}.}
 \label{chap4:fig-modalAnalysisLaparo}
@@
 In this method, the geometry is described by a network of masses connected together by springs.
-The force $ F $ exerted onto each mass by the spring may be computed with:
+The force $ \mathbf{F} $ exerted onto each mass by the spring may be computed with:
 \begin{equation}
 \label{chap4:massSpring1}
-F_{\text{spring}} = - k x,
+\mathbf{F}_{\text{spring}} = - k \mathbf{x},
 \end{equation}
-where $ k $ is the spring stiffness constant, $ x $ represents the stretch of the spring from its rest position.
+where $ k $ is the spring stiffness constant, $ \mathbf{x} $ represents the stretch of the spring from its rest position.
 From our experience, we know that a mass attached to a single spring has a tendency to oscillate.
@@
 \label{chap4:massSpring2}
-F_{\text{damper}} = - d v
+\mathbf{F}_{\text{damper}} = - d \mathbf{\dot x}
 \end{equation}
-where $ d $ is the damping constant and $ v $ the derivative of $ x $ with respect to time.
+where $ d $ is the damping constant and $ \mathbf{\dot x} $ the derivative of $ x $ with respect to time.
 The damper added between each couple of masses allows to model friction and reduce the amplitude of the oscillations.
-Applying Newton's second law $ \sum \mathbf{F} = m \mathbf{a} $ yields the standard differential equation for a mass-spring system:
+Applying Newton's second law $ \sum \mathbf{F} = m \mathbf{\ddot x} $ yields the standard differential equation for a mass-spring system:
 \begin{equation}
 \label{chap4:massSpring3}
-m \ddot x = - d \dot x - k x.
+m \mathbf{\ddot x} = - d \mathbf{\dot x} - k \mathbf{x}.
 \end{equation}
@@
 %
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=8cm]{chapter4/undesiredAnisotropy.pdf}
-\end{center}
+\begin{figure}[ht]
+\centering 
+\subfloat[ ]{\includegraphics[width=4cm]{chapter4/undesiredAnisotropy1.pdf}}
+\hspace{0.2cm}
+\subfloat[ ]{\includegraphics[width=4cm]{chapter4/undesiredAnisotropy2.pdf}}
 \caption[Undesired anisotropy with mass-spring models]{The two meshes are undergoing a downward pull at their bottom.
-While the tetrahedral mass-spring system \textbf{(left)} shows signs of anisotropy, the hexahedral mesh \textbf{(right)} with springs aligned along the gravity does not.}
+While the tetrahedral mass-spring system (a) shows signs of anisotropy, the hexahedral mesh (b) with springs aligned along the gravity does not.
+Image courtesy of \cite{Bourguignon00}.
+}
 \label{chap4:fig-undesiredAnisotropy}
@@
 \end{center}
-\caption[Ball on a deformable model]{A ball resting on a supporting elastic solid as computed by \citeauthor{Terzopoulos87}.} 
+\caption[Ball on a deformable model]{A ball resting on a supporting elastic solid as computed by \citeauthor{Terzopoulos87}. \TODO{to be deleted?}} 
 \label{chap4:fig-ball}
@@
 %
-\begin{figure}[h]
-\begin{center}
-\includegraphics[height=4.5cm]{chapter4/hybrid1.pdf}
+\begin{figure}[ht]
+\centering 
+\subfloat[ ]{\includegraphics[height=4.5cm]{chapter4/hybrid1.pdf}}
 \hspace{0.2cm}
-\includegraphics[height=4.5cm]{chapter4/hybrid2.pdf}
-\end{center}
+\subfloat[ ]{\includegraphics[height=4.5cm]{chapter4/hybrid2.pdf}}
 \caption[Simulation of a hepatectomy using a hybrid model]{Simulation of a hepatectomy using a hybrid model.
@@
 The rest of the mesh was simulated with the precomputed finite element model.
-}
+Image courtesy of \cite{Cotin00}.
+\TODO{Stéphane> c'est ptet moche comme image, mais c'est pourtant bien ça dans le papier...
+Et même dans ton HDR tu utilises cette image...
+regarde page 46 et 50...
+Donc ben c'est pas de ma faute si c'est moche ;-) }}
 \label{chap4:fig-hybridModel}
@@
 \cite{Nienhuys01} avoid the precomputations by approximating the inverse at runtime with an iterative algorithm (Conjugate Gradient).
-Moreover, instead of subdividing each tetrahedron to make the cut appear where the user performed it, they adapt the mesh locally so that there are always triangles on the scalpel path, and perform the cut along those triangles.
+Moreover, instead of subdividing each tetrahedron to make the cut appear where the user performed it, they adapt the mesh locally so that there are always triangles on the scalpel path, and perform the cut along those triangles (see \fig{chap4:fig-cutUnderScalpel}).
 If the approach is interesting, the authors reported a lag between the scalpel and the realised cut.
 In addition, repositioning the nodes within the mesh can easily generate degeneracies, which have to be eliminated.
+\TODO{add PhD of Brian Lee...
+But I don't have it, and is it published yet?}
 %
@@
 \end{center}
-\caption[The three steps in performing a cut]{The three steps in performing a cut (represented here in 2D).} 
+\caption[The three steps in performing a cut]{The three steps in performing a cut (represented here in 2D). Image courtesy of \cite{Nienhuys01}.} 
 \label{chap4:fig-cutUnderScalpel}
@@
 %
-\begin{figure}[h]
-\begin{center}
-\includegraphics[height=5.5cm]{chapter4/eyeSurgery1.pdf}
+\begin{figure}[ht]
+\centering 
+\subfloat[ ]{\includegraphics[height=5.5cm]{chapter4/eyeSurgery1.pdf}}
 \hspace{0.2cm}
-\includegraphics[height=5.5cm]{chapter4/eyeSurgery2.pdf}
-\end{center}
-\caption[Simulation of eye surgery]{(\textbf{Left}) Exterior view of the model showing the eye and the eyelashes.
-(\textbf{Right}) Surgical virtual environment including micro-tools and guidance information.}
+\subfloat[ ]{\includegraphics[height=5.5cm]{chapter4/eyeSurgery2.pdf}}
+\caption[Simulation of eye surgery]{(a) Exterior view of the model showing the eye and the eyelashes.
+(b) Surgical virtual environment including micro-tools and guidance information.
+Image courtesy of \cite{Sagar94}.}
 \label{chap4:fig-eyeSurgery}
@@
 
-\cite{Wu04} proposed the first application of GPU in medical simulation \label{chap4:GPUMedicalSimulation}.
+\cite{Wu04} proposed the first application of GPU in medical simulation\label{chap4:GPUMedicalSimulation}.
 They were able to achieve an interactive frame rate with topology changes in surgical simulation.
@@
 
+\TODO{+add paper ISBMS 2010?}
+
+\ON
+Recently, \cite{Zhu10} proposed a framework based on point-based simulation techniques to model organ deformations realistically.
+They combined a new strain-stress tensor computation scheme with smoothed particule hydrodynamics (SPH) method.
+The parameters of the tensor relating strain and stress are determined through mechanical experiments.
+
+\OFF
+
 \bigskip

### 2010-09-28 Thesis/Chapter5.tex
- Update 16 September -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@176 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 28 September -
Major update: lot of fixes on the first part that has been reviewed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@177 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 	\subsection{Computation of node displacements} \label{chap5:computeNodalDisp}
-We recall that the dynamic system of equations describing a deformable solid is the following (see section \ref{chap3:dynamic2}:
+We recall that the dynamic system of equations describing a deformable solid is the following (see section \ref{chap3:dynamic2}):
 \begin{equation}
@@
 
-Of crucial importance in explicit analyses is the restriction on time step sizes imposed by stability constraints \label{chap5:discussionTimestep}.
+Of crucial importance in explicit analyses is the restriction on time step sizes imposed by stability constraints\label{chap5:discussionTimestep}.
 As mentioned, the central difference method is only stable if $ \Delta t <  \Delta t_{cr} $.
@@
 However, if this value can be relaxed even further, significant increases in $ \Delta t_{cr}$ are possible.
-As an example, for a given $ Le $ a material with $E = 3\,000\, $Pa and $\nu = 0.45$ allows a time step of more than double that of a material with $\nu = 0.49$.
+As an example, for a given $ L_e $ a material with $E = 3\,000\, $Pa and $\nu = 0.45$ allows a time step of more than double that of a material with $\nu = 0.49$.
 Of course, lowering the Poisson ratio introduces inaccuracies, but for some applications (interactive medical simulation for instance) it may be that this is acceptable.

### 2010-09-28 Thesis/Chapter6.tex
- Update 16 September -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@176 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 28 September -
Major update: lot of fixes on the first part that has been reviewed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@177 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\section{General-purpose computation on graphics processing units}
+\section{General-purpose computation on GPU}
 
@@
 Of course, in this approach some elements will be computed several times (because being into different blocks).
-However, storing the force contributions in shared memory instead of global memory to read them later on via texture fetches widely outperforms the need of computing several elements several times.
+\ON Yet, \OFF storing the force contributions in shared memory instead of global memory to read them later on via texture fetches \ON may \OFF outperform the need of computing several elements several times.
+\ON However, the efficiency of this single kernel approach depends on the nature of the mesh.
+Because the force contributions of all elements attached to a node must be eventually added up, this approach allows to avoid the numerous (slow) accesses to global memory when the valency of the nodes is high (that is, nodes are attached to many elements).
+This type of mesh is often encountered when dealing with the complex shapes of anatomical structures.
+Therefore, where the first implementation showed its limits with complex meshes, this single kernel approach has proven to be more computationally effective.
+\OFF
 
@@
 In such a case, the critical time step for the whole simulation would be small even if there is only a single small element in the mesh.
-As an example, the liver mesh used in the section~\ref{chap6:liverSimu} had to be smoothed and remeshed properly to allow a decent simulation.
+As an example, the liver mesh used in the section~\ref{chap6:liverSimu} had to be smoothed and remeshed properly \ON with great care \OFF to allow a decent simulation.
 
@@
 Explicit time integration schemes are known to be more sensitive to contacts than implicit schemes.
-While this is beyond the scope of the present work, we note that a range of contact formulations are available in industry standard commercial explicit FE codes (e.g.
-LS-Dyna and ABAQUS).
-The methods are well tested and understood in the computational mechanics community, with two main types in common use: penalty methods and kinematic constraint methods.
+While this is beyond the scope of the present work, we note that a range of contact formulations are available, with two main types in common use: penalty methods and kinematic constraint methods.
 Penalty methods involve use of fictitious contact springs at interfaces, effectively to penalise interpenetration of opposing surfaces.
-If excessively stiff springs are used $ \Delta t_{cr} $ may indeed be reduced, however except in extreme conditions (e.g.
-from explosive or high speed impacts, or imposition of very large forces on highly constrained objects) spring stiffnesses of the same order of magnitude as the material stiffness are suitable.
-The effect on $ \Delta t_{cr} $ is therefore minimal \citep{Hallquist06}.
-Kinematic constraint methods involve (loosely) direct relocation of penetrating nodes back to the outside surface of the penetrated body, and have no effect on $ \Delta t_{cr} $.
-These may be especially useful for modelling interaction between soft tissues and stiff (effectively rigid) surgical tools.
-The computational cost of either variety has not been investigated here.
+If excessively stiff springs are used, $ \Delta t_{cr} $ may indeed be reduced.
+\ON According to LS-Dyna documentation \citep{Hallquist06},  \OFF except in extreme conditions (that is from explosive or high speed impacts, or imposition of very large forces on highly constrained objects), spring stiffnesses of the same order of magnitude as the material stiffness are suitable and the effect on $ \Delta t_{cr} $ is therefore minimal.
+\ON However, in practice, we observed that spring stiffnesses greater than the material stiffness are often required to prevent any penetration of the two objects, which results in an unstable system.
+\OFF Kinematic constraint methods involve (loosely) direct relocation of penetrating nodes back to the outside surface of the penetrated body, and have no effect on $ \Delta t_{cr} $.
+\ON In theory, \OFF these may be especially useful for modelling interaction between soft tissues and stiff (effectively rigid) surgical tools.
+\ON It is worth nothing that \OFF the computational cost of either variety has not been investigated here.
+
+\ON
+Even if a stable contact modelling could be achieved, one element of crucial importance has been omitted so far: the collision detection process.
+Indeed, before even handling the interaction between two objects, their interaction needs to be detected.
+Because the time step must be very small when using an explicit time integration scheme (often around one millisecond or less), the collision detection algorithm must also be run at the same frequency.
+The problem is that no collision detection algorithm is fast enough to execute at $ 1\,000\,$Hz or more with objects of reasonable sizes.
+Of course, one could choose to detect the collisions at a lesser frequency but the risks of penetration would substantially increases.
+One notable exception is the interaction between a soft deformable object and a fixed rigid surface where the collision detection is rapid.
+\OFF
+
+\TODO{timestep monitoring}
+
+\ON
+	\subsubsection{Conclusion}
+As we showed in the last two chapters, the TLED algorithm is a very efficient fully non-linear finite element formulation.
+We also added viscoelasticity and anisotropy features with minimal impact on its performance.
+In particular, its GPU implementation allows non-linear analyses of structures in real-time using a high number of elements as never before.
+
+However, the use of an explicit integration scheme imposes substantial constraints which limits TLED's range of applications.
+More specifically, the usage of the TLED algorithm in very interactive environments (as in the manipulation of a liver with a tool) is not recommended.
+The collision detection would either become the bottleneck of the simulation (if run at every time step) or fail to provide accurate detection of contacts (otherwise).
+In addition, the TLED algorithm is also not appropriate to simulate the deformation of rigid object as the critical time step would be too small to allow real-time computations.
+In constrast, TLED excels for computing the deformation of soft tissues when the boundary conditions are controlled.
+As an exemple, the TLED algorithm may efficiently provide brain shift predictions as the interaction of the brain with the fixed skull is determined beforehand and rapid collision detection can be achieved in this case.
+
+Hence, provided that the application is appropriate, the TLED algorithm is an efficient and accurate algorithm to model the deformation of solid soft anatomical organs.
+\OFF

### 2010-09-29 Thesis/Chapter2.tex
- Update 29 September -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@178 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 A rigid-body displacement consists of a simultaneous translation and rotation of the body without changing its shape or size.
-Deformation implies the change in shape and\slash or size of the body from an initial configuration noted $\kappa_{0}$ to a new configuration $\kappa$ called the \emph{current} or \emph{deformed configuration}.
+\ON In constrast, a change in shape and\slash or size of the body from an initial configuration $\kappa_{0}$ to a new configuration $\kappa$ is called a deformation.
+This new configuration $\kappa$ may be designated by the \emph{current} or \emph{deformed configuration}.
+\OFF
 
@@
 In the Lagrangian description, the position and physical properties of the particles are referred to a reference configuration $\kappa_{R}$, often chosen to be the undeformed configuration $\kappa_{0}$.
-Thus, the current coordinates ($\mathbf{x} \in \kappa$) are expressed in terms of the reference coordinates ($\mathbf{X} \in \kappa_{0}$):
+\ON Consequently, the current coordinates ($\mathbf{x} \in \kappa$) depend on the reference coordinates ($\mathbf{X} \in \kappa_{0}$): \OFF
 \begin{equation}
@@
 \end{equation}
-and the variation of a typical variable $\phi$ over the body is described with respect to the coordinates $\mathbf{X}$ and time $t$:
+\ON and a typical variable $\phi$ defined over the body is expressed in terms of the coordinates $\mathbf{X}$ and time $t$: \OFF
 \begin{equation}
+\label{chap2:phi}
 \phi = \phi(\textbf{X}, t).
 \end{equation}
-For a fixed value of $\mathbf{X} \in \kappa_{0}$, $\phi(\mathbf{X},t)$ gives the value of $\phi$ at time $t$ associated with the fixed particle $X$ whose position in the reference configuration is $\mathbf{X}$.
-We note that a change in time $t$ implies that the same particle $X$ has a different value $\phi$.
-Thus, the Lagrangian description focuses its attention on the particles of the continuous body and it is usually used in solid mechanics.
+\ON Let us consider a particule $X$ whose position in the reference configuration is $\mathbf{X}$.
+From \eqref{chap2:phi}, we note that the value of $\phi$ associated with this fixed particle $X$ changes over time.
+Hence, \OFF the Lagrangian description focuses its attention on the particles of the continuous body and it is usually used in solid mechanics.
 
 	\subsection{Eulerian description}
-Rather than following the particles, the Eulerian description observes the changes at fixed locations.
-	In the Eulerian description, the motion is this time referred to the current configuration $\kappa$ and $\phi$ is described with respect to the current position ($\mathbf{x} \in \kappa$):
+\ON In constrast, the Eulerian description is interested in changes at fixed locations.
+This time, the motion and $\phi$ are described with respect to the current position ($\mathbf{x} \in \kappa$): \OFF
 \begin{equation}
@@
 \end{equation}
-A \emph{displacement field} is a vector field of all displacement vectors for all particles in the body, which relates the deformed configuration $\kappa$ with the undeformed configuration $\kappa_{0}$.
-Indeed if the displacement field is known, we can construct the current configuration $\kappa$ from the undeformed configuration $\kappa_{0}$: $\chi(\mathbf{X}, t) = \mathbf{X} + \mathbf{u}(\mathbf{X})$.
+A \emph{displacement field} is a vector field of all displacement vectors for all particles in the body.
+\ON Thus, the deformed configuration $\kappa$ may be obtained from the undeformed configuration $\kappa_{0}$ by merely adding the displacement field: \OFF $\kappa_{0}$: $\chi(\mathbf{X}, t) = \mathbf{X} + \mathbf{u}(\mathbf{X})$.
 
@@
 	\subsection{Deformation gradient tensor}
-The displacement field tells us how a point displaces from the reference to the deformed configuration.
-However, we would also like to know how a piece of material is stretched and rotated as the body moves from the reference to the deformed configuration.
-Since the length of a material line $d\mathbf{X}$ can change when going to the deformed configuration as well as its orientation, we can say that $d\mathbf{X}$ deforms into $d\mathbf{x}$.
-The question then becomes how to relate $d\mathbf{x}$ in the deformed configuration with $d\mathbf{X}$ of the reference configuration.
+The displacement field tells us how a \ON particule \OFF displaces from the reference to the deformed configuration.
+\ON More interestingly, we would like to know how a material line would deform (stretch and rotation) within this displacement field.
+\OFF Since the length of a material line $d\mathbf{X}$ can change when going to the deformed configuration as well as its orientation, we can say that $d\mathbf{X}$ deforms into $d\mathbf{x}$.
+\ON We now seek the relation existing between $d\mathbf{x}$ in the deformed configuration and $d\mathbf{X}$ of the reference configuration.
+\OFF
 
@@
 Our motivation comes from the need to write global equilibrium statements that involve integrals over volumes.
-We can define volume elements in the reference and deformed configurations.
-Consider three non-coplanar line elements $d\mathbf{X}^{(1)}$, $d\mathbf{X}^{(2)}$ and $d\mathbf{X}^{(3)}$ forming a parallelepiped in the reference configuration as can be seen on \fig{chap2:fig-changeVolume}.
+\ON Let us consider a volume in the reference configuration formed by \OFF three non-coplanar line elements $d\mathbf{X}^{(1)}$, $d\mathbf{X}^{(2)}$ and $d\mathbf{X}^{(3)}$ as can be seen on \fig{chap2:fig-changeVolume}.
 % 
@@
 \end{align}
-where $\mathbf{\hat{N}}_{i}$ denote the unit vector along $d\mathbf{X}^{i}$. The corresponding volume in the deformed configuration is given by:
+where $\mathbf{\hat{N}}_{i}$ \ON is \OFF the unit vector along $d\mathbf{X}^{i}$. \ON Similarly, we can compute the \OFF corresponding volume in the deformed configuration by:
 \begin{align}
@@
 \end{align}
-The determinant of $\mathbf{F}$ is called the \emph{Jacobian} and it is denoted by $J = \det \mathbf{F}$. And we have:
+The determinant of $\mathbf{F}$ is called the \emph{Jacobian} and it is denoted by $J = \det \mathbf{F}$.
+\ON Therefore, the relation between the deformed volume $ dv $ and the undeformed volume $ dV $ may be written as: \OFF
 \begin{equation}
@@
 \end{equation}
-Thus, J has the physical meaning of being the local ratio of current to reference volume of a material volume element.
+\ON It is worth noting that the Jacobian \OFF  J has the physical meaning of being the local ratio of current to reference volume of a material volume element.
 
@@
 For similar reasons, let us find the relation between an element of area in the reference configuration $d\mathbf{A}$ which becomes $d\mathbf{a}$ in the deformed configuration.
-Considering an arbitrary material line $d\mathbf{L}$ in the reference configuration and noting $d\mathbf{l}$ the same line after deformation (see \fig{chap2:fig-changeSurface}), the corresponding reference and current volumes are given respectively by:
+Considering an arbitrary material line $d\mathbf{L}$ in the reference configuration and noting $d\mathbf{l}$ the same line after deformation (see \fig{chap2:fig-changeSurface}), the reference and current volumes are given respectively by:
 \begin{align}
@@
 \end{equation}
-The fact that this expression is valid for any vector $d\mathbf{L}$ and using the property $\mathbf{a} \cdot \mathbf{b} = \mathbf{a} ^{T} \mathbf{b}$ yields:
+\ON Since \OFF this expression is valid for any vector $d\mathbf{L}$ and by using the property $\mathbf{a} \cdot \mathbf{b} = \mathbf{a} ^{T} \mathbf{b}$, \ON the sought relation between an element of area in the reference configuration $d\mathbf{A}$ and its corresponding area $d\mathbf{a}$ in the deformed configuration may be expressed as: \OFF
 \begin{equation}
@@
 If all the curves do not change length, it is said that a rigid body displacement occurred.
-The concept of strain is used to evaluate how much a given displacement differs locally from a rigid body displacement.
+The concept of strain is used to evaluate how much a given displacement differs locally from a rigid body displacement \citep{Lubliner06}.
 Therefore, although we know how to transform vectors from the reference configuration into vectors in the current configuration using the deformation gradient, it is more useful to find a measure of the change in length of $d\mathbf{X}$.
@@
 	\subsection{Green-Lagrange strain tensor}
-Using \eqref{chap2:dS} and \eqref{chap2:ds}, the change in the squared lengths due to the body deformation between the reference and the current configuration can be expressed as:
+Using \eqref{chap2:dS} and \eqref{chap2:ds}, \ON the difference in the squared lengths between the reference and the current configuration may be evaluated as: \OFF 
 \begin{align}
@@
 By definition, the Green-Lagrange strain tensor is a symmetric second-order tensor.
-Also, the change in squared lengths is zero if and only if $\mathbf{E} = \mathbf{0}$.
+\ON Moreover, we observe that \OFF the change in squared lengths is zero if and only if \ON the Green-Lagrange strain tensor \OFF $\mathbf{E} = \mathbf{0}$.
 Using \eqref{chap2:gradient}, the Green strain tensor may be expanded as the following:
@@
 These internal forces are produced between the particles in the body as a reaction to external forces applied on the body.
-The $SI$ unit for stress is pascal (symbol $Pa$), which is equivalent to one newton (force) per square metre (unit area).
-As with strain, different measures of stress can be defined.
-For instance, stress can be measured per unit of deformed area (section \ref{chap2:CauchyStress}) or undeformed area (sections \ref{chap2:PiolaStress1} and \ref{chap2:PiolaStress2}).
+The SI unit for stress is pascal (symbol $Pa$), which is equivalent to one newton (force) per square metre (unit area).
+As with strain, stress can be measured per unit of deformed area (section \ref{chap2:CauchyStress}) or undeformed area (sections \ref{chap2:PiolaStress1} and \ref{chap2:PiolaStress2}).
 In general, the stress is not uniformly distributed over the cross section of a material body, and consequently the stress at a point on a given area is different from the average stress over this entire area.
 Therefore, it is necessary to define the stress not over a given area but at a specific point in the body.
-Stress at a point in a three-dimensional continuum can be measured in terms of nine quantities, three per plane, on three mutually perpendicular planes at the point.
-These nine quantities may be viewed as the components of a second-order tensor that we call \emph{stress tensor}.
+\ON The stress at a point in a three-dimensional continuum is fully defined by \OFF nine quantities, three per plane, on three mutually perpendicular planes at the point.
+\ON The \emph{stress tensor} is defined as the second-order tensor which components are these nine quantities.
+\OFF
 
 	\subsection{Cauchy stress}\label{chap2:CauchyStress}
-The Euler-Cauchy's stress principle states that on any surface (real or imaginary) that divides the body, the action of one part of the body on the other is equivalent to the system of distributed forces and couples on the surface dividing the body.
+The Euler-Cauchy's stress principle states that on any surface (real or imaginary) that divides the body, the action of one part of the body on the other is equivalent to the system of distributed forces and couples on the surface dividing the body \citep{Truesdell60}.
 With that consideration in mind, let us consider a plane $S$ that passes through an arbitrary internal point $P$ which has a unit normal vector $\mathbf{n}$.

### 2010-09-29 Thesis/Chapter5.tex
- Update 29 September -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@178 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 Additionally, there is no requirement for computation of tangent matrices as in implicit or quasi-static procedures, since there is no involvement of Newtonian iterations.
-While the developments above incorporated a hyperelastic constitutive formulation (neo-Hookean), thus accommodating nonlinearity of the strain-related tissue stress response, the formulation was intentionally simple nonetheless since the focus was on validation of the computational framework.
+While the developments above incorporated a hyperelastic constitutive formulation (neo-Hookean), thus accommodating non-linearity of the strain-related tissue stress response, the formulation was intentionally simple nonetheless since the focus was on validation of the computational framework.
 Two further key features of the response of most biological tissues are time- (and rate-)dependence and anisotropy \citep{Fung93}.
@@
 \end{equation}
-where $ \alpha_{\infty} $, $ \alpha_i $ and $ \tau_i $are positive real constants.
+where $ \alpha_{\infty} $, $ \alpha_i $ and $ \tau_i $ are positive real constants.
 Such forms for the relaxation functions have a physical interpretation, namely that of a generalised Maxwell model \citep{Holzapfel96}.
@@
 The instantaneous terms $ \boldsymbol \Phi^n $ may be computed directly from the (known) current deformation $ \mathbf{C}^n $.
-The main difficulty is then computation of the in- cremental rate-dependent terms $ \boldsymbol \Upsilon_i^n $ .
+The main difficulty is then computation of the incremental rate-dependent terms $ \boldsymbol \Upsilon_i^n $.
 Following \cite{Poon98}, our strategy is to maintain each $ \boldsymbol \Upsilon_i^n $ as a separate state variable to be updated at each increment also.

### 2010-09-29 Thesis/Chapter6.tex
- Update 29 September -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@178 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 where $ ^0 V $ is the initial volume of the element and $ \mathbf{\hat{S}} $ is the vector form of the stress tensor $ \mathbf{S} $.
-This integral is generally evaluated numerically, for ex- ample using Gaussian quadrature.
+This integral is generally evaluated numerically, for example using Gaussian quadrature.
 For reduced integration 8-node hexahedral elements we obtain
@@
 \begin{center}
-\includegraphics[width=10cm]{chapter6/stressCurvesCube2.pdf}
+\includegraphics[width=8cm]{chapter6/stressCurvesCube2.pdf}
 \end{center}
@@
 \begin{figure}[h]
-\begin{center}
-\includegraphics[width=13.9cm]{chapter6/cubeAnisotropic.pdf}
-\end{center}
+\centering 
+\subfloat[ ]{\includegraphics[width=5cm]{chapter6/cubeAnisotropic1.pdf}}
+\hspace{1cm}
+\subfloat[ ]{\includegraphics[width=5cm]{chapter6/cubeAnisotropic2.pdf}} \\
+\subfloat[ ]{\includegraphics[width=5cm]{chapter6/cubeAnisotropic3.pdf}}
+\hspace{1cm}
+\subfloat[ ]{\includegraphics[width=5cm]{chapter6/cubeAnisotropic4.pdf}} \\
 \caption[Deformation patterns of transversely isotropic models compared with that of an isotropic model]{Deformation patterns of transversely isotropic models compared with that of an isotropic model: (a) the undeformed cube, (b) the deformed isotropic model, (c) the deformed anisotropic model with $ \mathbf{a}_0 = \mathbf{a}_0^y $, (d) the deformed anisotropic model with $\mathbf{a}_0^{yz}$.
@@
 \begin{center}
-\includegraphics[width=10cm]{chapter6/stressDecay.pdf}
+\includegraphics[width=8cm]{chapter6/stressDecay.pdf}
 \end{center}
@@
 %
-\begin{figure}[h]
+\begin{figure}[ht]
 \begin{center}
-\includegraphics[width=13.9cm]{chapter6/timingsGPU.pdf}
+\includegraphics[width=13.5cm]{chapter6/timingsGPU.pdf}
 \end{center}
@@
 \fig{chap6:fig-timingsGPU}(b) also shows the decreasing critical time steps $ \Delta t_{cr} $ with increasing model sizes (which results from the decreasing element sizes), using liver-like and brain-like materials.
-We define liver-like materials to be those described by model TIV with material parameters quoted at the beginning of the section, whereas brain-like materials are those described by model NHE with material parameters $\mu = 1006.71\, $Pa and Poisson?s ratio $\nu = 0.45$.
+We define liver-like materials to be those described by model TIV with material parameters quoted at the beginning of the section, whereas brain-like materials are those described by model NHE with material parameters $\mu = 1006.71\, $Pa and Poisson's ratio $\nu = 0.45$.
 The intersections of these curves with the solution time curves provide an estimate of the real-time capacity of the implementation.
@@
 
-\TODO{Add comparison of different implementations}
+\bigskip
+
+\ON
+This GPU implementation of the TLED algorithm was demonstrated publically at MICCAI 2007 during the talk given by Zeike Taylor \citep{Taylor07b}.
+A non-linear finite element analysis of a cube featuring brain-like properties was solved in real-time at $1\,430\,$ Hz using $55\,566$ tetrahedral elements (see \fig{chap6:fig-demoMICCAI}).
+%
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=11cm]{chapter6/demoMICCAI.pdf}
+\end{center}
+\caption[Real-time demo at MICCAI 2007]{Non-linear finite element algorithm solved in real-time at $1\,430\,$ Hz with $55\,566$ elements.
+A strain between 0 and $40\, \%$ is periodically applied to one face whilst the opposite face is fixed.
+The material properties were the following: cube's edge = 10cm, Young modulus = $3\,000\,$ Pa, Poisson's ratio = 0.45, density = $1\,000\, kg.m^{-3}$ and the timestep was $0.7\,$ms.
+The normal for each vertex was computed in real-time for a proper lightning and a skin texture was applied.}
+\label{chap6:fig-demoMICCAI}
+\end{figure}
+\OFF
+
+\TODO{If time, add comparison of different implementations}
 
@@
 
-\begin{figure}
-\includegraphics[width=13.9cm]{chapter6/cube_and_curves.pdf}
+\begin{figure}[ht]
+\includegraphics[width=13.5cm]{chapter6/cube_and_curves.pdf}
 \caption [Computational timings within SOFA] {\textbf{Left:} FEM mesh of cube with $29\,791$ nodes deformed under a uniformed load.
-\textbf{Right:} Comparison of GPU computational timings and CPU overheads between the SOFA and standalone implementations for different mesh sizes.}
+\textbf{Right:} Comparison of GPU computational timings and CPU overheads between the SOFA and standalone implementations for different mesh sizes.
+\TODO{change the graph + nicer visual model?}}
 \label{chap6:fig-cube_and_curves}
@@
 %
-\begin{figure}[h]
+\begin{figure}[ht]
 \begin{center}
-\includegraphics[width=13.9cm]{chapter6/liver.pdf}
+\includegraphics[width=13.5cm]{chapter6/liver.pdf}
 \end{center}

### 2010-09-30 Thesis/Chapter2.tex
- Update 30 September -
Part solid organ modelling almost finished (all reviews taken care of)



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@179 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{center}
-\includegraphics[width=13cm]{chapter2/changeVolume.pdf}
+\includegraphics[width=12cm]{chapter2/changeVolume.pdf}
 \end{center}
-\caption{Transformation of a volume element under a deformation mapping. \TODO{CHANGE TO A BETTER FIGURE.}}
+\caption{Transformation of a volume element under a deformation mapping.}
 \label{chap2:fig-changeVolume}
@@
 For similar reasons, let us find the relation between an element of area in the reference configuration $d\mathbf{A}$ which becomes $d\mathbf{a}$ in the deformed configuration.
-Considering an arbitrary material line $d\mathbf{L}$ in the reference configuration and noting $d\mathbf{l}$ the same line after deformation (see \fig{chap2:fig-changeSurface}), the reference and current volumes are given respectively by:
+Considering an arbitrary material line $d\mathbf{L}$ in the reference configuration and noting $d\mathbf{l}$ the same line after deformation, the reference and current volumes are given respectively by:
 \begin{align}
@@
 % 
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=13cm]{chapter2/changeSurface.pdf}
-\end{center}
-\caption{Transformation of a surface element under a deformation mapping. \TODO{CHANGE TO A BETTER FIGURE.}}
-\label{chap2:fig-changeSurface}
-\end{figure}
+%\begin{figure}[h]
+%\begin{center}
+%\includegraphics[width=13cm]{chapter2/changeSurface.pdf}
+%\end{center}
+%\caption{Transformation of a surface element under a deformation mapping. \TODO{CHANGE TO A BETTER FIGURE.}}
+%\label{chap2:fig-changeSurface}
+%\end{figure}
 
@@
 
-To establish this relationship, let us consider an infinitesimal tetrahedron in a Cartesian coordinate system as represented \fig{chap2:fig-CauchyStress}. 
+\ON In the following we make use of a Cartesian coordinate system and we consider an infinitesimal tetrahedron as shown on \fig{chap2:fig-CauchyStress}. \OFF
 %
@@
 \begin{center}
-\includegraphics[width=13cm]{chapter2/CauchyStress.pdf}
+\includegraphics[width=6cm]{chapter2/CauchyStress.pdf}
 \end{center}
-\caption{Tetrahedral element in Cartesian coordinates. \TODO{CHANGE TO A BETTER FIGURE.}}
+\caption{Tetrahedral element in Cartesian coordinates.}
 \label{chap2:fig-CauchyStress}
@@
 This stress is basically defined as force/(unit deformed area).
-The strain measure that is appropriate to use with the Cauchy stress tensor is the infinitesimal strain tensor (see section~\ref{chap2:infinitesimalStrainTensor}).
-The problem with using the Cauchy stress tensor for analysing materials undergoing large deformation is that we generally do not know the area in the deformed configuration.
-Therefore, we need to develop alternative stress tensors that we can use in the reference configuration.
-We will introduce two new stress measures: the first Piola-Kirchhoff stress tensor and the second Piola-Kirchhoff stress tensor.
+The strain measure that is appropriate to use with the Cauchy stress tensor is \ON therefore \OFF the infinitesimal strain tensor (see section~\ref{chap2:infinitesimalStrainTensor}).
+\ON Indeed, because the Cauchy stress tensor is defined by unit of deformed area, it is not appropriate for analysing materials undergoing large deformation where the area in the deformed configuration is often unknown.
+Consequently, we require alternative stress tensors based on the reference configuration.
+\OFF We will introduce two new stress measures: the first Piola-Kirchhoff stress tensor and the second Piola-Kirchhoff stress tensor.
 
@@
 We know that the Cauchy stress tensor represents the current force per unit of deformed area: $d\mathbf{f} = \mathbf{t}(\mathbf{n}) da = \sigma \cdot d\mathbf{a}$.
-We define a stress vector $\mathbf{T}$ over the area element $dA$ with normal $\mathbf{N}$ in the underformed configuration such that it results in the same total force
+\ON We state that we can find a stress vector $\mathbf{T}$ over the area element $d\mathbf{A}$ in the underformed configuration which \OFF results in the same total force.
 \begin{equation}
@@
 
-In a similar way we defined the Cauchy stress $\sigma$ with $\mathbf{t}(\mathbf{n}) = \sigma \cdot \mathbf{n}$, we introduce a stress tensor $\mathbf{P}$ called the \emph{first Piola-Kirchhoff stress tensor} such that $\mathbf{T}(\mathbf{n}) = \mathbf{P} \cdot \mathbf{N}$.
-The first Piola-Kirchhoff stress tensor gives the current force per unit undeformed area.
+In a similar way we defined the Cauchy stress $\sigma$ with $\mathbf{t}(\mathbf{n}) = \sigma \cdot \mathbf{n}$, we introduce a stress tensor $\mathbf{P}$ such that $\mathbf{T}(\mathbf{n}) = \mathbf{P} \cdot \mathbf{N}$.
+\ON We call $ \mathbf{P} $ the \emph{first Piola-Kirchhoff stress tensor} and \OFF gives the current force per unit undeformed area.
 
@@
 	\subsection{Second Piola-Kirchhoff stress tensor}\label{chap2:PiolaStress2}
-The \emph{second Piola-Kirchhoff stress tensor} $\mathbf{S}$ is introduced as the stress tensor associated with the force $d\mathbf{F}$ in the undeformed area $d\mathbf{A}$ that corresponds to the force $d\mathbf{f}$ in the deformed area $d\mathbf{a}$:
+\ON Let us consider the stress tensor $\mathbf{S}$ resulting in the force $d\mathbf{F}$ in the undeformed area $d\mathbf{A}$ which \OFF corresponds to the force $d\mathbf{f}$ in the deformed area $d\mathbf{a}$.
+\ON We define this stress tensor $\mathbf{S}$ as the \emph{second Piola-Kirchhoff stress tensor}.
+\OFF
 \begin{equation}
@@
 \end{equation}
-Thus, the second Piola-Kirchhoff stress tensor gives the transformed current force per unit undeformed area.
-Similar to the relationship \eqref{chap2:dxdX} between $d\mathbf{x}$ and $d\mathbf{X}$, the force $d\mathbf{f}$ is related to the force $d\mathbf{F}$ by:
+\ON In other words, \OFF the second Piola-Kirchhoff stress tensor gives the transformed current force per unit undeformed area.
+Similar to the relationship \eqref{chap2:dxdX} between $d\mathbf{x}$ and $d\mathbf{X}$, the force $d\mathbf{f}$ is related to the force $d\mathbf{F}$ via the deformation gradient tensor:
 \begin{align}
@@
 So far, all the relations that we have established are valid for every continuum.
-Indeed, no mention was made of any material in the derivations.
-Therefore, these relations are not sufficient to describe the response of a material to a given loading.
-We know that the stresses result from the deformation of the material and it is now necessary to express them in terms of some measure of this deformation such as, for instance, the strain.
-These relationships, known as \emph{constitutive equations}, obviously depend on the type of material under consideration and may be dependent on time or not.
-Constitutive equations are not derived from any physical principles, although they are subject to obeying certain rules and the entropy inequality.
-In essence, constitutive equations are mathematical models of the behaviour of materials that are validated against experimental results.
+Indeed, \ON the derivations carried out so far made no mention of any material.
+\OFF Therefore, \ON we know that \OFF these relations are not sufficient to describe the response of a material to a given loading.
+\ON Moreover, because the stresses result from the deformation of the material, they may be expressed \OFF in terms of some measure of this deformation such as the strain.
+This relationship, which obviously depends on the type of material under consideration, is what we call a \emph{constitutive equation}.
+\ON Constitutives equations are an attempt to mathematically characterise the behaviour of materials observed from experimentals results.
+\OFF
 
-The relationship between stress and strain depends on the nature, or constitution of the particular material that we are studying.
-Of course, any material mechanics, and especially tissue mechanics, can be quite complex, and many assumptions must be made when deriving a constitutive equation.
-It is basically impossible to derive a particular constitutive equation that would accurately model all aspects of tissue behaviour under any type of loading.
-This is because tissues may behave very differently under small loads as compared to large loads.
-For instance, we may expect damage under large loads.
-Also, under different loading rates different tissues may behave differently.
-Therefore, whenever we develop a constitutive equation to model a tissue, we need to balance the need to accurately model tissue behaviour under the range of loadings with the need to have a constitutive equation that is simple enough to:
-\begin{enumerate}
-\item be computed in a numerical model 
-\item experimentally measure all the material constants used in the constitutive equation
-\end{enumerate}
-This section will now introduce the most common constitutive models relevant to soft-tissue modelling.
+\ON Naturally, the relationship between strain and stress is conditioned by the constitution of the material of interest.
+Because the mechanics involved in materials can be quite complex (tissue mechanics in particular), many simplifying assumptions are required to derive a constitutive equation.
+In practice, a given constitutive equation can only cover a subset of the actual mechanical behaviour and could not possibly model all aspects of tissue behaviour under any type of loading accurately.
+For instance, some tissues behave very differently under small loads as compared to large loads.
+The behaviour may also depends on loading rates and damage may even occur under large loads.
+Moreover, not only constitutive equations need to deal with the complexity of material mechanics but they must remain simple enough to measure all the included material constants experimentally.
+\OFF This section will now introduce the most common constitutive models relevant to soft-tissue modelling.
 
@@
 
-One special case of elastic materials are \emph{hyperelastic} materials. A material is said to be hyperelastic if there exists a \emph{strain energy density} function $U_0(\varepsilon)$ such that
+\ON In addition, \OFF if there exists a \emph{strain energy density} function $U_0(\varepsilon)$ such that 
 \begin{equation}
 \label{chap2:energy}
-\boldsymbol \sigma = \pd{U_0}{\varepsilon}.
+\boldsymbol \sigma = \pd{U_0}{\varepsilon},
 \end{equation}
-However, for \emph{imcompressible} materials the stress tensor is not completely determined by the deformation.
+\ON the material is said to be hyperelastic.
+\OFF However, for \emph{imcompressible} materials the stress tensor is not completely determined by the deformation.
 The hydrostatic pressure\footnote{pressure exerted by a fluid at equilibrium due to the force of gravity} affects the stress and the above equation is then written as:
@@
 It means that the sum of the displacements (and hence strains) produced by two sets of loads is equal to the displacements (and strains) produced by the sum of the two sets of loads.
-For example, the extensional strain $\varepsilon_{11}^{(1)}$ in the  coordinate direction $x_1$ due to the stress $\sigma_{11}$ in the same direction is $\sigma_{11}/E_1$, where $E_1$ denotes Young?s modulus of the material in the $x_1$ direction.
+For example, the extensional strain $\varepsilon_{11}^{(1)}$ in the  coordinate direction $x_1$ due to the stress $\sigma_{11}$ in the same direction is $\sigma_{11}/E_1$, where $E_1$ denotes Young's modulus of the material in the $x_1$ direction.
 The extensional strain $\varepsilon_{11}^{(2)}$ due to the stress $\sigma_{22}$ applied in the direction $x_2$ (as a result of the Poisson's effect) is $-\nu_{21} (\sigma_{22}/E_2)$, where $\nu_{21}$ is the Poisson's ratio:
@@
 	\subsection{Viscoelastic materials}
-Viscoelasticity is the property of materials that exhibit both viscous and elastic characteristics when undergoing deformation.
-Viscous materials, like honey, resist shear flow and strain linearly with time when a stress is applied.
-Elastic materials strain instantaneously when stretched and just as quickly return to their original state once the stress is removed.
-Viscoelastic materials have elements of both of these properties and, as such, exhibit time dependent strain.
+\ON We know that materials for which the mechanical behaviour depends solely on the state of deformation are called elastic.
+In contrast, the mechanical behaviour of \emph{viscous} materials is only a function of the current rate of deformation (like honey for instance).
+\OFF Viscoelastic materials have elements of both of these properties and, as such, exhibit time dependent strain.
 Depending on the change of strain rate versus stress inside a material the viscosity can be categorised as having a linear or non-linear response.

### 2010-09-30 Thesis/Chapter3.tex
- Update 30 September -
Part solid organ modelling almost finished (all reviews taken care of)



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@179 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 The approximation $ \psi^e_i $ can be linear or higher-order polynomial expressions and are called interpolation functions.
-%They are also commonly named \emph{shape functions} because they define the `shape' that a variable can take within an element (linear, quadratic etc.). 
 Depending on the degree of polynomial approximation used to represent the solution, additional nodes may be identified inside the element.
@@
 %
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=13cm]{chapter3/naturalCoordinatesHexa.pdf}
-\end{center}
+\begin{figure}[ht]
+\centering 
+\subfloat[ ]{\includegraphics[width=6cm]{chapter3/naturalCoordinatesHexa1.pdf}}
+\hspace{1cm}
+\subfloat[ ]{\includegraphics[width=6cm]{chapter3/naturalCoordinatesHexa2.pdf}}
 \caption[Natural coordinates of a hexaedron]{Hexahedral element as it appears in the global coordinate system (a) and its natural coordinate system (b).}
@@
 %
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=13cm]{chapter3/naturalCoordinatesTetra.pdf}
-\end{center}
+\begin{figure}[ht]
+\centering 
+\subfloat[ ]{\includegraphics[width=6cm]{chapter3/naturalCoordinatesTetra1.pdf}}
+\hspace{1cm}
+\subfloat[ ]{\includegraphics[width=6cm]{chapter3/naturalCoordinatesTetra2.pdf}}
 \caption[Natural coordinates of a tetrahedron]{Tetrahedral element as it appears in the global coordinate system (a) and its natural coordinate system (b).}
@@
 The strong form, in contrast to a \emph{weak form}, requires strong continuity on the dependent field variables, such as displacements \citep{Liu03}.
-The functions defining these field variables have to be differentiable up to the order of the partial differential equations that exist in the strong from.
-Obtaining the exact solution for a strong form is usually very difficult for practical engineering problems.
-We know that the finite element method can be used to find an approximated solution but the method usually works well only with regular geometry and boundary conditions.
+\ON Indeed, \OFF the functions defining these field variables have to be differentiable up to the order of the partial differential equations that exist in the strong from.
+\ON While the finite element method can be used to find an approximated solution for a strong form,  the method usually works well only with regular geometry and boundary conditions.
 
-A weak form is often created using energy principles or weighted residual methods.
-The weak form is often an integral form and requires a weaker continuity on the field variables.
-Due to the weaker requirement on the field variables, and the integral operation, a formulation based on a weak form usually produces a set of discretised system equations that give much more accurate results, especially for problems of complex geometry.
-Hence, the weak form is usually preferred for obtaining an approximate solution.
-Using the weak form usually leads to a set of well-behaved algebraic system equations, if the problem domain is discretised properly into elements.
+Consequently, \OFF a weak form is often created using energy principles or weighted residual methods.
+The weak form is frequently an integral form and requires a weaker continuity on the field variables.
+\ON Hence, \OFF a formulation based on a weak form usually produces a set of discretised system equations that give much more accurate results, especially for problems of complex geometry.
+\ON For this reason, \OFF the weak form is usually preferred \ON over the strong form for obtaining an approximate solution of a practical engineering problem \OFF.
 
@@
 \begin{description}
-\item[(a)] coupled formulation in which the time $ t $ is treated as an additional coordinate along with the spatial coordinate $ x $
+\item[(a)] coupled formulation in which the time $ t $ is an additional coordinate along with the spatial coordinate $ x $
 \item[(b)] decoupled formulation where time and spatial variations are assumed to be separable.
@@
 where $ \hat{\psi}^e_i(x, t) $ are time-space interpolation functions, $ \hat{u}^e_i $ are the nodal values that are independent of $ x $ and $ t $, $ \psi^e_i(x) $ are the usual interpolation functions in spatial coordinate $ x $ only and the nodal values $ u^e_i(t) $ are functions of time $ t $ only.
-Space-time coupled finite element formulations are not common and they have not been adequately studied.
+Coupled finite element formulations are not common and they have not been adequately studied.
 Hence, we consider the decoupled formulation only.
@@
 An example is given by \fig{chap3:fig-sparse} for a finite element problem in two dimensions.
-\begin{figure}
+\begin{figure}[ht]
 \begin{center}
-\includegraphics[width=13cm]{chapter3/sparse_matrix.pdf}
+\includegraphics[width=10cm]{chapter3/sparse_matrix.pdf}
 \end{center}
@@
 Furthermore, it is wasteful to reserve storage for zero elements.
-
-%The choice of a resolution method depends on the problem (linear or non-linear) and the properties of the stiffness matrix.
 
@@
 This choice is efficient in cases where $ \mathbf{A} $ is diagonally dominant.
-Sparse approximations of a LU or Cholesky decompositions may also be used as preconditioners (see sections \ref{chap3:LUdecompo} and \ref{chap3:CholeskyDecompo} for more details).
+Sparse approximations of LU or Cholesky decompositions may also be used as preconditioners (see sections \ref{chap3:LUdecompo} and \ref{chap3:CholeskyDecompo} for more details).
 \OFF

### 2010-09-30 Thesis/Chapter4.tex
- Update 30 September -
Part solid organ modelling almost finished (all reviews taken care of)



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@179 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \ON
-The challenges faced by the field of medical simulation were reviewed in details in the introductory chapter \label{chap1}.
+The challenges faced by the field of medical simulation were reviewed in details in the introductory chapter (see chapter \ref{chap1}).
 Hence, we know that at the very least, the deformation of solid organs must appear physically realist in medical simulators.
@@
 They derived a geometrically non-linear strain tensor from differential geometry theory.
-Obviously, the computations were not carried out in real-time but this technique allows \citeauthor{Terzopoulos87} to calculate the interaction between solid and deformable models in a physically realistic manner (see \fig{chap4:fig-ball}).
+Obviously, the computations were not carried out in real-time but this technique allows \citeauthor{Terzopoulos87} to calculate the interaction between solid and deformable models in a physically realistic manner.
+%(see \fig{chap4:fig-ball}).
  %
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=7cm]{chapter4/ball.pdf}
-\end{center}
-\caption[Ball on a deformable model]{A ball resting on a supporting elastic solid as computed by \citeauthor{Terzopoulos87}. \TODO{to be deleted?}} 
-\label{chap4:fig-ball}
-\end{figure}	
+%\begin{figure}[h]
+%\begin{center}
+%\includegraphics[width=7cm]{chapter4/ball.pdf}
+%\end{center}
+%\caption[Ball on a deformable model]{A ball resting on a supporting elastic solid as computed by \citeauthor{Terzopoulos87}. \TODO{to be deleted?}} 
+%\label{chap4:fig-ball}
+%\end{figure}
+
 In order to speed up the computations, linear strain measures started to be used (see section~\ref{chap2:infinitesimalStrainTensor}).
@@
 Image courtesy of \cite{Cotin00}.
-\TODO{Stéphane> c'est ptet moche comme image, mais c'est pourtant bien ça dans le papier...
+\TODO{Stéphane> tu dis que l'image est moche, mais c'est pourtant bien celle présente dans le papier...
 Et même dans ton HDR tu utilises cette image...
 regarde page 46 et 50...
-Donc ben c'est pas de ma faute si c'est moche ;-) }}
+Mine de rien ça fait plus de 10 ans ;-) }}
 \label{chap4:fig-hybridModel}
@@
 \TODO{add PhD of Brian Lee...
-But I don't have it, and is it published yet?}
+Mais je l'ai pas sa thèse...
+et tu m'as pas dit que tu étais en train de la reviewer ?
+Du coup c'est pas encore publique, c'est difficile de la citer, non ?}
 %
@@
 
-\TODO{+add paper ISBMS 2010?}
-
-\ON
-Recently, \cite{Zhu10} proposed a framework based on point-based simulation techniques to model organ deformations realistically.
+\ON Recently, \cite{Zhu10} proposed a framework based on point-based simulation techniques to model organ deformations realistically in a laparoscopic simulator.
 They combined a new strain-stress tensor computation scheme with smoothed particule hydrodynamics (SPH) method.
-The parameters of the tensor relating strain and stress are determined through mechanical experiments.
-
+The parameters of the tensor relating strain and stress were determined through mechanical experiments.
 \OFF

### 2010-09-30 Thesis/Chapter6.tex
- Update 30 September -
Part solid organ modelling almost finished (all reviews taken care of)



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@179 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\begin{figure}
+\begin{figure}[ht]
 \begin{center}
@@
 This was done foremost for simplicity, but is also a common practical feature of anisotropic models for soft tissues.
-The main reason is the clear physical interpretation of $ \bar{I}_4 $ as the square of the stretch $ \lambda $ \footnote{Strictly, $ \bar{I}_4 = J^{-2/3} \lambda^2 $} along the preferred direction $ \mathbf{a}_0 $ \cite{Holzapfel00}, whereas $ \bar{I}_5 $ has a less clear physical meaning.
-$ \bar{I}_4 $ terms are therefore more amenable to experimental investigation, and of more use in modelling structural elements (e.g.
-fibres) responsible for the anisotropy.
+The main reason is the clear physical interpretation of $ \bar{I}_4 $ as the square of the stretch $ \lambda $ \footnote{Strictly, $ \bar{I}_4 = J^{-2/3} \lambda^2 $} along the preferred direction $ \mathbf{a}_0 $ \citep{Holzapfel00}, whereas $ \bar{I}_5 $ has a less clear physical meaning.
 It must be emphasised that this model is used as an example only, and we make no claim concerning its appropriateness for any particular tissue.
+% $ \bar{I}_4 $ terms are therefore more amenable to experimental investigation, and of more use in modelling structural elements (e.g. fibres) responsible for the anisotropy.
 
@@
 
-\bigskip
+%\bigskip
 
-Material parameters were chosen based on recent results for the viscoelastic response of human liver in vivo.
+%Material parameters were chosen based on recent results for the viscoelastic response of human liver in vivo.
 \cite{Nava08} used an aspiration device to obtain an estimate of $\mu = 19\,700\,$Pa for the combined liver/liver capsule.
@@
 This GPU implementation of the TLED algorithm was demonstrated publically at MICCAI 2007 during the talk given by Zeike Taylor \citep{Taylor07b}.
-A non-linear finite element analysis of a cube featuring brain-like properties was solved in real-time at $1\,430\,$ Hz using $55\,566$ tetrahedral elements (see \fig{chap6:fig-demoMICCAI}).
+A non-linear finite element analysis of a cube featuring brain-like properties was solved in real-time at $1\,430\,$Hz using $55\,566$ tetrahedral elements (see \fig{chap6:fig-demoMICCAI}).
 %
@@
 A strain between 0 and $40\, \%$ is periodically applied to one face whilst the opposite face is fixed.
-The material properties were the following: cube's edge = 10cm, Young modulus = $3\,000\,$ Pa, Poisson's ratio = 0.45, density = $1\,000\, kg.m^{-3}$ and the timestep was $0.7\,$ms.
+The material properties were the following: cube's edge = $10\,$cm, Young modulus = $3\,000\,$ Pa, Poisson's ratio = 0.45, density = $1\,000\, kg.m^{-3}$ and the time step was $0.7\,$ms.
 The normal for each vertex was computed in real-time for a proper lightning and a skin texture was applied.}
@@
 \OFF
-
-\TODO{If time, add comparison of different implementations}
 
@@
 \end{enumerate}
-
+%
 \begin{figure}[ht]
-\includegraphics[width=13.5cm]{chapter6/cube_and_curves.pdf}
-\caption [Computational timings within SOFA] {\textbf{Left:} FEM mesh of cube with $29\,791$ nodes deformed under a uniformed load.
-\textbf{Right:} Comparison of GPU computational timings and CPU overheads between the SOFA and standalone implementations for different mesh sizes.
-\TODO{change the graph + nicer visual model?}}
+\centering 
+\subfloat[ ]{\includegraphics[width=5cm]{chapter6/cube_and_curves1.pdf}}
+\hspace{0.5cm}
+\subfloat[ ]{\includegraphics[width=8cm]{chapter6/cube_and_curves2.pdf}}
+\caption [Computational timings within SOFA] {(a) FEM mesh of cube with $29\,791$ nodes deformed under a uniformed load.
+(b) Comparison of GPU computational timings and CPU overheads between the SOFA and standalone implementations for different mesh sizes.
+\TODO{change the cube if I can get a decent visual model in SOFA}}
 \label{chap6:fig-cube_and_curves}
@@
 \begin{figure}[ht]
-\begin{center}
-\includegraphics[width=13.5cm]{chapter6/liver.pdf}
-\end{center}
-\caption[GPU solution times]{Deformation of a liver model using the TI model \textbf{(middle)} and the TIV model \textbf{(right)}.
-The undeformed mesh is shown on the left.
+\centering 
+\subfloat[ ]{\includegraphics[width=4cm]{chapter6/liver1.pdf}}
+\hfill
+\subfloat[ ]{\includegraphics[width=4cm]{chapter6/liver2.pdf}}
+\hfill
+\subfloat[ ]{\includegraphics[width=4cm]{chapter6/liver3.pdf}}
+\caption[GPU solution times]{Deformation of a liver model using the TI model (b) and the TIV model (c).
+The undeformed mesh is shown in (a).
 Colour maps indicate the relative Von Mises equivalent stress magnitude.}
@@
 The problem is that no collision detection algorithm is fast enough to execute at $ 1\,000\,$Hz or more with objects of reasonable sizes.
-Of course, one could choose to detect the collisions at a lesser frequency but the risks of penetration would substantially increases.
+Of course, one could choose to detect the collisions at a lesser frequency but the risks of penetration would substantially increase.
 One notable exception is the interaction between a soft deformable object and a fixed rigid surface where the collision detection is rapid.
-\OFF
 
-\TODO{timestep monitoring}
+One possible extension to enhance the stability is to monitor the critical time step of the simulation.
+Indeed, we know an approximation of this critical time step and we recall its expression here for convenience:
+\begin{equation}
+\Delta t_{cr} = \dfrac{L_e}{c},
+\end{equation}
+where $ L_e $ is the smallest characteristic element length (roughly interpreted as the smallest edge length) in the assembly, and $c$ is the dilatational wave speed of the material which depends on material properties (such as Young's modulus and Poisson's ratio).
+More details on the critical time step were given page \pageref{chap5:criticalTimestep}.
+If $c$ is constant for a given material, $ L_e $ varies over time during the simulation, in particular when contacts occur.
+Maintaining the time step of the simulation below the monitored critical time step would insure the stability of the simulation.
+However, by possibly reducing the time step of the simulation, real-time computations can no longer be guaranteed.
 
-\ON
 	\subsubsection{Conclusion}

### 2010-10-02 Thesis/Chapter1.tex
First table of contents + update of bibliography
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@125 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 9 August -
Added all chapters files + transition chapter2-chapter3 + summary chapter3 + removed outline



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@146 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 2 October -
A few changes in shell part + part of conclusion




git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@180 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\chapter{Introduction}
-\label{chap:intro}
-\begin{shortAbstract}
-A short abstract for the upcoming chapter
-\end{shortAbstract}
+\chapter{Medical simulation}
+\label{chap1}
 
-\section{Illustration Example}
+\section{General context}
+
+	\subsection{Historical evolution: mannequins (plain and electronic), conputer-based simulators}
+	\subsection{Ultimate goal: needs to integrate patient-specific + physiology + pathologies for medical training, patient-specific planning and per-operative guidance}
+
+\section{Challenges for computer-based simulators}					
+	\subsection{Photorealistic rendering}
+	\subsection{Accurate haptic feedback}
+	\subsection{Physically realist organ modelling in real-time}

### 2010-10-02 Thesis/Chapter11.tex
- Update 9 August -
Added all chapters files + transition chapter2-chapter3 + summary chapter3 + removed outline



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@146 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 2 October -
A few changes in shell part + part of conclusion




git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@180 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2010-08-09 Thesis/Chapter10.tex
@@
-\chapter{'Shell meshing' technique \OC{MICCAI 2010}}
-\label{chap10}
-\begin{shortAbstract}
-A short abstract for the upcoming chapter
-\end{shortAbstract}
+\chapter{Conclusion and perspectives}
+\label{chap11}
 
-\section{State of art: reconstruction/simplification}
+\section{Summary of contributions}
 
-\section{Our method}
+The objective of this PhD was to propose soft tissue models for computing the deformation of anatomical structures in real-time.
+Because the shape and the internal structure of these structures greatly differ, a unique model cannot deal with the variousness of the mechanical behaviours.
+Hence, we identified two major types of structures in human body, they can be either solid (brain, liver, prostate etc.) or hollow (colon, blood vessels, stomach etc.).
+
+Most of previous works focus on volumetric models that are able to capture the behaviour of solid organs.
+Yet, many of these models had to compromise on accuracy to achieve the real-time constraint demanded by medical simulation.
+Even though the best mathematical framework for predicting the deformation of a continuous structure is provided by continuum mechanics, solving these equations is a computationally very demanding task and simpler models are often used in practice.
+And even when the finite element method is employed to solve the equations of continuum mechanics, linear strain measure and simple constitutive laws are common and therefore does not constitute an appropriate modelling of the complex mechanical behaviours of solid anatomical structures.
+
+In contrast, much less research has been carried out for the modelling of hollow structures.
+In fact in the field of medical simulation, very few models have been proposed for simulating, in real-time, the deformation of thin anatomical structures whose volume is negligible compared to their surface area.
+Nevertheless, human body features many hollow structures and we feel that the need for modelling hollow organs in medical simulation is equally important as with solid organs.
+
+\bigskip
+
+Consequently, we strived to propose solutions to these two known problems in the field of medical simulation.
+
+\subsubsection*{Solid structure modelling.}
+We implemented the total Lagrangian explicit dynamics (TLED) algorithm on GPU using CUDA.
+This finite element formulation is fully non-linear: it makes use of a non-linear strain measure (which allows large deformations) and embodies a non-linear constitutive law between strain and stress as characterised by \cite{Fung93} for soft-tissues.
+While the first GPU implementation of the TLED was carried out by \cite{Taylor07b} in the Cg language, our re-implementation in CUDA led to a threefold increase in performance.
+Overall, we observe a maximum speed improvement for GPU solution over CPU solution of up $56.3 \times$.
+My collaboration with Zeike Taylor continued and we enhanced the formulation by adding anisotropy and viscoelasticity.
+To the best of our knowledge, this work constitutes the first GPU implementation of a non-linear, anisotropic and viscoelastic finite element procedure.
+This formulation was eventually implemented in the open source framework for medical simulation SOFA and will be released publically with the upcoming 1.0 release of SOFA.
+By providing a fast and accurate non-linear FEM with various constitutive laws for soft tissue modelling to worldwide researchers, we thus hope to assist in enhancing the realism of medical simulators.
+
+\subsubsection*{Hollow structure modelling.}
+We proposed a framework for real-time modelling of thin anatomical structures.
+We first designed a triangular shell element by combining a two-dimensional in-plane membrane energy, with an off-plane energy for describing bending and twist.
+We improved and extended a plate model first introduced by \cite{Przemieniecki85} to a co-rotational formulation to allow large displacements (but small deformations).
+Once combined with an in-plane membrane formulation we obtained an accurate, yet computationally efficient, shell finite element method featuring both membrane and bending energies.
+To the best of our knowledge, this co-rotational shell FEM is the first description of a shell finite element method in the field of medical simulation.
+However, to model the deformation of complex anatomical structures using shell elements, the first step is to describe its surface with curved patches.
+Indeed, while many flat triangles are required to describe highly curved surfaces, fewer triangular shell elements are needed to describe the given geometry with the same precision since they can be curved.
+Therefore, we suggested a simple method to mesh a surface with curved shell elements.
+The novelty of our method relies on the combination of a shell finite element formulation and a geometric surface reconstruction both based on the same polynomial interpolation function used to describe the surface of shells.
+We also showed how contacts and interactions with the curved surfaces of shells can be handled using the same function.
+
+\section{Discussion}
+
+	\subsection{Solid structures modelling}
+
+The TLED algorithm offers an accurate modelling for soft tissues and its GPU implementation using CUDA is very fast.
+Hence, real-time computations can be achieved without any trade-off on the accuracy of the deformation.
+One of the features allowing such an efficiency is the use of an explicit time integration scheme.
+If this permits significant computational advantage, the use of an explicit integration scheme imposes that the time step of the simulation remains below a critical time step.
+This critical time step depends on two kind of criterion: the size of the elements and the material?s properties (Young?s modulus and Poisson?s ratio).
+Therefore, meshing of anatomical structures to be simulated must be accomplished with great care to insure well-shaped elements.
+Even more restricting, the use of the TLED formulation is not suitable for modelling stiff objects in real-time.
+Indeed, increasing the Young's modulus translates into a lower critical time step, which necessitates an intensification of the simulation's frequency to enforce real-time.
+In some applications, the value of the Poisson?s ratio may be relaxed from 0.49 to 0.45 for instance.
+However, if it allows to double the critical time step, this relaxation introduces additional inaccuracies.
+
+One of the most common critics we received for this work concerned the poor ability of the TLED to handle interaction between organs and virtual tools.
+Indeed, interactions may impact the critical time step resulting from the use of an explicit time integration scheme and thus influence the overall stability of the simulation.
+The work carried out during my PhD was entirely concerned with the numerical solution provided by the TLED formulation.
+Yet, while this is beyond the scope of the present work, a range of contact formulations are available in commercial finite element packages.
+According to their documentation, the impact on the critical time step should be minimal.
+However, for a demonstration at an IT exhibition (CeBIT Australia), we tried to setup an interaction between a virtual tool driven by a haptic device and a cube with brain-like properties.
+If we managed to produce an interesting demo for the exhibition, its preparation brought out an unstable system.
+But even if a stable contact modelling technique could be formulated, the actual limiting factor is the collision detection process.
+Because the time step must be very small when using an explicit time integration scheme (often around one millisecond or less), the collision detection algorithm must also be run at the same frequency.
+The problem is that no collision detection algorithm is fast enough to execute at $ 1\,000\,$Hz or more with objects of reasonable sizes.
+One notable exception is the interaction with a fixed rigid surface where the collision detection is rapid.
+
+Consequently, this list of pros and cons limits the TLED's range of applications.
+If its usage in very interactive environments is not suitable, the TLED finite element algorithm is great at computing the deformation of soft tissues when the boundary conditions are known and controlled.
+Hence, provided that the application is appropriate, the TLED algorithm is an efficient and accurate algorithm to model the deformation of solid soft anatomical organs.
+
+	\subsection{Hollow structures modelling}
+
+We achieved our objectives to propose a versatile solution which can simulate, in real-time, thin objects with various shapes and material properties with good accuracy.
+The efficiency of the method was illustrated through shell-based reconstruction and real-time simulation of the deformations of various anatomical structures and other thin objects.
+Nevertheless, the use of a linear strain measure limits the range of application to small deformations only.
+Moreover, the constitutive law was chosen to be the simple (linear) Hookean law, which it is known not to describe accurately the relationship between strain and stress in soft tissues.
+However, those potential changes are substantial and their implementations are not straightforward.
+In addition, it is not clear from which method the benefits would be the greatest: whether from enhancing our model to a non-linear formulation or completely changing to another model based on true shell theory (where membrane and bending energies are not separated).
+Our shell FEM could also be improved by implementing the formulation on GPU.
+Although an optimal parallel implementation is always challenging, our method is highly parallelisable and there is no major obstacle to an efficient GPU implementation.
+
+Regarding our geometric surface reconstruction based on shells, we were able to mesh complex anatomical structures with a fairly small number of shell elements.
+However, the main drawback of our method is the lack of any assessment of the quality of the elements created.
+Indeed, if some elements are stretched from the heavy decimation, these elements are not corrected during the process of remeshing.
+Our algorithm could even worsen the situation by splitting those stretched elements into four smaller (and more stretched) elements.
+Yet, meshing a surface with curved elements is a very challenging problem and still an active area of research.
+In contrast to all geometrical techniques described in the literature, we presented a more physical approach consistent with our shell FEM formulation and obtained good results in meshing complex anatomical structures.
+
+	\subsection{Real-time constraint: use of GPU acceleration}
+
+\section{Perspectives}

### 2010-10-02 Thesis/Chapter9.tex
- Update 2 October -
A few changes in shell part + part of conclusion




git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@180 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\begin{figure}
+\begin{figure}[ht]
 \centering
@@
 %
-\begin{figure}
+\begin{figure}[ht]
 \centering
@@
 
-\begin{figure}
+\begin{figure}[ht]
 	\begin{center}
@@
 \centering 
-\subfloat[Subdivided triangles]{\includegraphics[width=6.5cm]{chapter9/sub_grid_red.png}}
-\hfill 
-\subfloat[Irregular grid]{\includegraphics[width=6.5cm]{chapter9/irregular_grid_red.png}} \\
-\subfloat[Subdivided triangles]{\includegraphics[width=6.5cm]{chapter9/sub_grid_mesh.png}}
-\hfill
-\subfloat[Irregular grid]{\includegraphics[width=6.5cm]{chapter9/irregular_grid_mesh.png}}
+\subfloat[Subdivided triangles]{\includegraphics[width=5cm]{chapter9/sub_grid_red.png}}
+\hspace{1cm}
+\subfloat[Irregular grid]{\includegraphics[width=5cm]{chapter9/irregular_grid_red.png}} \\
+\subfloat[Subdivided triangles]{\includegraphics[width=5cm]{chapter9/sub_grid_mesh.png}}
+\hspace{1cm}
+\subfloat[Irregular grid]{\includegraphics[width=5cm]{chapter9/irregular_grid_mesh.png}}
 \caption[Visual artifacts using triangle subdivisions]{In some situations, some rendering artifacts may appear when we use subdivided triangles to render the shells (a,c).
@@
 %
-\begin{figure}[h]
+\begin{figure}[ht]
 \centering 
@@
 %
-\begin{table}[h!]
+\begin{table}[ht]
 	\begin{center}
@@
 
-\begin{figure}[!h]
+\begin{figure}[ht]
 \centering
-\includegraphics[width=6cm]{chapter9/implant_folding1.pdf}
-\includegraphics[width=5cm]{chapter9/implant_folding2.pdf} \\
-\includegraphics[width=4.5cm]{chapter9/implant_folding3.pdf}
+\includegraphics[width=5cm]{chapter9/implant_folding1.pdf}
+\includegraphics[width=4.5cm]{chapter9/implant_folding2.pdf} \\
+\includegraphics[width=4cm]{chapter9/implant_folding3.pdf}
 \includegraphics[width=7cm]{chapter9/implant_folding4.pdf}
@@
 % Link where images from surgery have been found: http://asianophthalmology.org/articles/?a=articles&p=6
-\begin{figure}[!h]
+\begin{figure}[ht]
 \centering

### 2010-10-02 Thesis/TitlePage.tex
- Update 14 September -
Chapter 7, 9 and 10 finished and sent for review



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@175 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 16 September -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@176 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 2 October -
A few changes in shell part + part of conclusion




git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@180 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2010-07-16 Thesis/TitlePage.tex
@@
 \noindent \large Prepared by Olivier \textsc{Comas} at \\ 
-INRIA Lille, \textsc{Shaman} Team and CSIRO ICT Brisbane, EAHRC \\
+INRIA Lille, \textsc{Shaman} team and CSIRO ICT Brisbane, AEHRC \\
 \vspace*{0.5cm}
@@
 \begin{tabular}{llcl}
-      \textit{Reviewers:}	& Bernard \textsc{}		& - & INRIA (Shaman)\\
-				& Bernard \textsc{}		& - & INRIA (Shaman)\\
-      \textit{Advisor:}	& Stéphane \textsc{Cotin}		& - & INRIA (Shaman)\\
-      \textit{President:}	& Bernard \textsc{}		& - & INRIA (Shaman)\\
-      \textit{Examinators:}   & Bernard \textsc{}          & - & INRIA (Shaman)\\
-      				& Bernard \textsc{}			& - & INRIA (Shaman)\\
+      \textit{Advisor:}	& Stéphane \textsc{Cotin}		& - & INRIA (\textsc{Shaman} team), Lille, France\\
+      \textit{Reviewers:}	& Thomas Sangild \textsc{S\o rensen}          & - & University of Aarhus, Denmark\\
+				& Georges \textsc{Dumont}		& - & ENS Cachan, Rennes, France\\
+      \textit{President:}	& To be determined		& - & \\
+      \textit{Examinators:}   & Tamy \textsc{Boubekeur}          & - & Telecom ParisTech, Paris, France\\
+      				& Christian \textsc{Duriez}			& - & INRIA (\textsc{Shaman} team), Lille, France\\
+      \textit{Invited:}   & Hans \textsc{De Visser}          & - & CSIRO, Brisbane, Australia\\
 \end{tabular}

### 2010-10-05 Thesis/Abstract.tex
- Update 5 October -
Added conclusion, abstract, resume and beginning of introduction



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@181 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\cleardoublepage
+\chapter*{Abstract}
 
-\begin{vcenterpage}
-\noindent\rule[2pt]{\textwidth}{0.5pt}
 \begin{center}
-{\large\textbf{Real-time soft tissue modelling on GPU for medical simulation\\}}
+\large\textbf{Real-time soft tissue modelling on GPU for medical simulation\\}
 \end{center}
-{\large\textbf{Abstract:}}
-Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla. Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.
-\\
-Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla.
-\\
 
-Blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla.
-\\
+\noindent \large\textbf{Abstract:}
+Modelling the deformation of anatomical structures in real-time is a crucial problem in medical simulation.
+Because their shape and their constitution greatly differ, a unique model cannot deal with the variousness of the mechanical behaviours.
+Hence, we identified two major types of structures in human body, they can be either solid (brain, liver, prostate etc.) or hollow (colon, blood vessels, stomach etc.).
 
-Blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla.
-\\
-{\large\textbf{Keywords:}}
-Medical simulation, soft tissue modelling, finite element method, GPU
-\\
-\noindent\rule[2pt]{\textwidth}{0.5pt}
-\end{vcenterpage}
+Our answer to this problematic is twofold.
+Our first contribution is an efficient GPU implementation of a non-linear, anisotropic and viscoelastic finite element model for solid organs.
+Our second contribution is a framework for real-time modelling of thin anatomical structures via a parallelisable co-rotational shell finite element formulation and a method to mesh a complex surface with curved shell elements.
+
+Although the two soft tissue models are based on continuum mechanics for greater accuracy, they may both be employed to simulate the deformation of entire organs in real-time.
+Finally, their implementation into the open source framework SOFA will provide worldwide researchers with new models to assist in enhancing the realism of medical simulators.
+
+\bigskip
+
+\large\textbf{Keywords:}
+Medical simulation, soft tissue modelling, finite element method, GPU, TLED, shell, co-rotational méthod

### 2010-10-05 Thesis/Chapter1.tex
- Update 5 October -
Added conclusion, abstract, resume and beginning of introduction



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@181 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2010-07-12 Thesis/Chapter1.tex
@@
-\chapter{Introduction}
-\label{chap:intro}
-\begin{shortAbstract}
-A short abstract for the upcoming chapter
-\end{shortAbstract}
+\chapter{Introduction to medical simulation}
+\label{chap1}
 
-\section{Illustration Example}
+\section{General context}
+
+Virtually everyone knows this situation where a close relative has to go to the hospital, either for a simple routine check or a serious surgery.
+Hospitals are meant to be reinsuring, a refuge where people's health improves, where people get fixed.
+Obviously, patient safety is a critical concern in the medical industry.
+Yet, patients have been known to suffer injuries or even death due to errors of judgement or a lack of care and training.
+As an example, a report published in 2000 by the Institute of Medicine in the United States \citep{Kohn00} described two studies carried out in the 1990s based on large samples of hospital admissions.
+They found that the proportion of hospital admissions experiencing an adverse event, defined as causes of injuries by medical management, were 2.9 and $3.7\,$\%, respectively .
+The proportion of these events attributable to errors (that is, preventable adverse events, in other words medical errors) was 58 and $53\,$\%, respectively.
+Even when using the lower estimate, deaths due to preventable adverse events in 1997 in the US ($44\,000$) exceeded the deaths attributable to motor vehicle accidents ($43\,458$), breast cancer ($42\,297$) or AIDS ($16\,516$).
+Because many errors go unreported, detailed statistics on medical errors are fairly scarce (hospitals usually do not attempt to emphasize errors).
+Nevertheless, the problem is real.
+And this is not surprising, elaborate skills have to be mastered in the medical field.
+Moreover, achieving a high degree of personal competence is sometimes not sufficient as various members of staff must learn how to work as a team.
+In fact, one of the recommendations of this report \citep{Kohn00} was to \emph{establish interdisciplinary team training programs for providers that incorporate proven methods of training, such as simulation}.
+However, dispensing an appropriate training in such a complex and hazardous environment is very challenging.
+Because healthcare is a high risk industry (like aviation or military), training in the real world is too costly and dangerous.
+Consequently, various approaches were applied to teaching and training of medical practicians over history of healthcare.
+
+\bigskip
+
+The most basic form of medical simulators are simple models of human anatomy.
+Hundreds of years ago, representations in clay and stone were already used to demonstrate clinical features of disease states and their effects on humans.
+Models have been found from many cultures and continents.
+Nowadays, similar passive models are still used to help students learn the anatomy.
+In constrast, active models that attempt to reproduce living anatomy or physiology are recent developments.
+The first example was created in the early 1960s by Asmund Laerdal for training in mouth to mouth ventilation.
+This simulator of a dying victim not breathing and lacking a heart beat, became known as called Resusci-Anne and had been widely used for CPR training thanks to an internal spring attached to the chest wall \citep{Cooper04}.
+In the mid-1960s, Sim One became the first mannequin controlled by a computer.
+The chest was moved with breathing, the eyes blinked and the pupils could dilate for instance.
+But the computer technology was too expensive for commercialisation at the time and only one mannequin was built.
+Over time mannequins have obviously substantially improved and today such systems are essentially integrated into training centers that aim at recreating the operating room environment.
+
+%- animal and fixed models used to reproduce some of the training objectives (but imperfect substitute and expensive)
+%- animals, simulated patients, scenarios, computed-based
+
+\section{Challenges for computer-based simulators}
+%- Ultimate goal: needs to integrate patient-specific + physiology + pathologies for medical training, patient-specific planning and per-operative guidance		
+%- Demands photorealistic rendering, accurate haptic feedback and (which is of prime interest to us): physically realist organ modelling in real-time.
+%- Evolution: geometrical models => mass-springs => FEM => more elaborate constitutive laws / different type of structures
+%- Contributions + plan

### 2010-10-05 Thesis/Chapter11.tex
- Update 5 October -
Added conclusion, abstract, resume and beginning of introduction



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@181 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 The objective of this PhD was to propose soft tissue models for computing the deformation of anatomical structures in real-time.
-Because the shape and the internal structure of these structures greatly differ, a unique model cannot deal with the variousness of the mechanical behaviours.
+Because the shape and the internal nature of these structures greatly differ, a unique model cannot deal with the variousness of the mechanical behaviours.
 Hence, we identified two major types of structures in human body, they can be either solid (brain, liver, prostate etc.) or hollow (colon, blood vessels, stomach etc.).
@@
 This formulation was eventually implemented in the open source framework for medical simulation SOFA and will be released publically with the upcoming 1.0 release of SOFA.
-By providing a fast and accurate non-linear FEM with various constitutive laws for soft tissue modelling to worldwide researchers, we thus hope to assist in enhancing the realism of medical simulators.
+%By providing a fast and accurate non-linear FEM with various constitutive laws for soft tissue modelling to worldwide researchers, we thus hope to assist in enhancing the realism of medical simulators.
 
@@
 If this permits significant computational advantage, the use of an explicit integration scheme imposes that the time step of the simulation remains below a critical time step.
-This critical time step depends on two kind of criterion: the size of the elements and the material?s properties (Young?s modulus and Poisson?s ratio).
+This critical time step depends on two kind of criterion: the size of the elements and the material?s properties (Young's modulus and Poisson's ratio).
 Therefore, meshing of anatomical structures to be simulated must be accomplished with great care to insure well-shaped elements.
@@
 Indeed, increasing the Young's modulus translates into a lower critical time step, which necessitates an intensification of the simulation's frequency to enforce real-time.
-In some applications, the value of the Poisson?s ratio may be relaxed from 0.49 to 0.45 for instance.
+In some applications, the value of the Poisson's ratio may be relaxed from 0.49 to 0.45 for instance.
 However, if it allows to double the critical time step, this relaxation introduces additional inaccuracies.
@@
 
-	\subsection{Real-time constraint: use of GPU acceleration}
+	\subsection{Use of GPU acceleration}
+
+One of the strongest constraint in medical simulation is the demand for real-time computations.
+This constraint is what makes all tasks in a simulator so challenging (collision detection, contact and soft tissue modelling, haptic feedback, photorealistic rendering etc.).
+In fact, this is at the origin of the multitude of simplifications proposed in simulation for modelling the biomechanics of anatomical structures.
+The lack of accuracy induced by these simplifications is of great concern to us.
+In this PhD, a part ot the solution was to accelerate the computations by using GPU.
+While we have showed that this approach can work very well, we need to keep in mind that GPU acceleration is not the panacea.
+Firstly, a standard is yet to be imposed in GPGPU.
+While OpenCL is openly targeted as a multi-vendor and multi-platform API, its usage remains very scarce nevertheless.
+In fact, the first experiments carried out within the SOFA framework revealed that OpenCL is not as optimised as CUDA for instance.
+Secondly, GPU programming is not simple to master.
+Although we have seen many improvements in flexibility of programmation over the last 3 or 4 years, a strong understanding of the underlying GPU hardware is still mandatory to create an optimal code.
+As an example, a GPU offers different types of memory, all with distinct pros and cons (strict access patterns, latency, limited size etc.), and choosing the best type of memory for each variable may be rather challenging to say the least.
+But most importantly, GPU acceleration is only interesting if the code is highly parallelisable.
+The difficulty is to devise a model that combines (1) an excellent capability at predicting the deformation of anatomical structures and (2) great parallel properties.
+Ideally, the parallelisation of the algorithm must be thought as part of the design process rather than seen as two unconnected steps.
+Even when the code is greatly parallelisable, one could face limited GPU hardware possibilities (as with concurrent writes for instance).
+
+GPU acceleration is all but a straightforward process.
+One must often accomplish a considerable amount of work to achieve a substantial gain.
+However, a decent GPU implementation may allow improvements of up to one or two orders of magnitude.
+For instance, our CUDA implementation of the TLED algorithm was about $ 56 \times $ faster than an equivalent CPU implementation on 2008 hardware.
 
 \section{Perspectives}
+
+Earlier this year, NVIDIA released a new GPU architecture called \emph{Fermi}.
+Fermi features several major innovations for GPGPU computing like cache (faster memory access in some cases), unified address space (easier to run C and C++ programs), debugging support, faster atomic instructions, compliance with the newly revised IEEE 754 floating point standard, 3 times the amount of shared memory, etc.
+In theory, this series of improvement should boost the performance of the second and third TLED implementations (see section \ref{chap6:secondTLED}) and we should investigate and compare the efficiency of the three designs on this new hardware.
+
+In a short term perspective, our co-rotational shell FEM algorithm could also be implemented on GPU.
+We already assessed in section \ref{chap9:GPU} that there was no major obstacle which would prevent a parallel implementation.
+Nevertheless, an optimal GPU implementation will demand careful design to take full advantage of the hardware.
+
+Although the meshing of complex surfaces with curved shell elements was not the heart of this PhD, our naïve approach could be enhanced by focusing on element quality.
+One simple solution would be to add a relaxation phase where a repulsion force is applied between each vertex (constrained to remain on the surface of the object).
+This physical approach would allow a more even distribution of the vertices across the surface and would contribute to obtain better quality elements overall.
+While a more mathematical method (a derivative of advancing front techniques for instance) could probably be formulated, the problem is complex and could account for another PhD.
+
+Finally, all the contributions of this PhD could contribute to enhance the complexity of anatomical structure modelling through combining solid and hollow structures.
+While we presented our findings on the coupling between solid (tetrahedra) and thin objects (shells) via the simulation of an angioplasty procedure, these results are still preliminary and much work remains to be done.
+In particular, the validity of the interaction and the accuracy of the overall deformation of the blood vessel were not quantitatively evaluated.
+
+As a conclusion, the TLED algorithm and our co-rotational shell FEM are respectively the first non-linear FEM for solid structures and the first model featuring bending for hollow structures to be implemented into the open source framework SOFA.
+Although they are based on continuum mechanics, they may both be employed to simulate the deformation of entire organs in real-time.
+By providing efficient FEM algorithms for all types of anatomical structures to worldwide researchers, we thus hope to assist in enhancing the realism of medical simulators.

### 2010-10-05 Thesis/Resume.tex
- Update 5 October -
Added conclusion, abstract, resume and beginning of introduction



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@181 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\large\textbf{Simulation de tissus mous en temps réel sur GPU pour la simulation médicale} \\
+\begin{center}
+\large\textbf{Simulation en temps réel de tissus mous sur GPU pour la simulation médicale\\}
+\end{center}
 
-Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla. Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.
-\\
-Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla.
-\\
+\noindent \large\textbf{Résumé:}
+Modéliser la déformation de structures anatomiques en temps réel est un problème crucial en simulation médicale.
+En raison des grandes différences existantes dans leur forme et leur constitution, un modèle unique est insuffisant face à la variété des comportements mécaniques.
+Par conséquent, nous avons identifié deux principaux types de structures: les organes pleins (cerveau, foie, prostate etc.) et les organes creux (colon, vaisseaux sanguins, estomac etc.).
 
-Blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla.
-\\
+Notre réponse à cette problématique est double.
+Notre première contribution est une implémentation GPU d'un modèle éléments finis qui est non-linéaire, anisotropique et viscoélastique pour les structures pleines.
+Notre seconde contribution est un environnement pour modéliser en temps réel les structures fines via un modèle parallèlisable et co-rotationnel utilisant des éléments coques et une approche pour mailler une surface complexe avec des éléments coques courbes.
 
-Blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla.
+Bien que les deux modèles de tissus sont basés sur la mécanique continue pour une meilleure précision, ils sont tous les deux capables de simulation la déformation d'organes en temps réel.
+Enfin, leur implémentation dans l'environnement open source SOFA rendra publique ces deux modèles afin de participer à l'amélioration du réalisme des simulateurs médicaux.
+
+\bigskip
+
+\large\textbf{Mots-clés:}
+Simulation médicale, modélisation de tissus mous, méthode des éléments finis, GPU, TLED, coques, méthode co-rotationnelle

### 2010-10-06 Thesis/Chapter1.tex
- Update 6 October -
Besides acknowledgements, the draft is finished!



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@182 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 Nowadays, similar passive models are still used to help students learn the anatomy.
-In constrast, active models that attempt to reproduce living anatomy or physiology are recent developments.
+In medical schools, live animals are used for teaching physiology and surgery classes.
+Dogs, pigs, goats and sheep are commonly used in medical training.
+The exercices essentially consist in anesthetising the animal and practicing surgical techniques.
+They may also be severely injured to teach participants how to treat and manage various traumatic injuries.
+If animal training allows a better realism than passive models, this type of training is being progressively abandoned because of both the prohibitive cost and obvious ethical reasons.
+Alternatives include active models of human patients that attempt to reproduce living anatomy or physiology.
 The first example was created in the early 1960s by Asmund Laerdal for training in mouth to mouth ventilation.
@@
 But the computer technology was too expensive for commercialisation at the time and only one mannequin was built.
-Over time mannequins have obviously substantially improved and today such systems are essentially integrated into training centers that aim at recreating the operating room environment.
+Over time, computerised mannequins have substantially improved and today such systems are essentially integrated into training centers that aim at recreating the operating room environment.
+Scenarios can be created to expose medical professionals to diverse clinical situations with the goal of improving the communication between staff members.
+Rare but critical clinical situations that may require specialised training can also be simulated.
+Another approach to medical training, that may also be combined with computerised mannequins, is the use of computer-based simulators.
+Such systems provide an elegant solution for medical training since they can offer a realistic and configurable training environment.
+Computer-based simulators also appear as an ideal tool for skill assessment of students since virtually every parameter of the simulation can be measured and recorded.
+Increasing computational power and growing interest for the field over the last decade led to the rapid development of reasonably sophisticated simulation systems.
+Yet, computer-based simulators still face numerous challenges.
 
-%- animal and fixed models used to reproduce some of the training objectives (but imperfect substitute and expensive)
-%- animals, simulated patients, scenarios, computed-based
+\section{Challenges in computer-based simulation}
 
-\section{Challenges for computer-based simulators}
-%- Ultimate goal: needs to integrate patient-specific + physiology + pathologies for medical training, patient-specific planning and per-operative guidance		
-%- Demands photorealistic rendering, accurate haptic feedback and (which is of prime interest to us): physically realist organ modelling in real-time.
-%- Evolution: geometrical models => mass-springs => FEM => more elaborate constitutive laws / different type of structures
-%- Contributions + plan
+Because everything is virtual, the possibilities are almost infinite.
+Ultimately, the ideal simulator will feature a patient-specific anatomy obtained through a medical imaging technique (such as MRI or CT).
+It will also need to consider the actual properties of the anatomical structures of this given patient, to take into account both the variability across patients and the alteration due to possible diseases.
+Indeed, diseased organs are often much stiffer than healthy structures and this will affect the overall mechanical behaviour.
+The physiology relevant to the simulated procedure (motion caused by breathing, heart beat, blood flow, etc.) will have to be faithfully reproduced as well.
+Such a patient-specific simulator capable of replicating human physiology and pathologies would be a great instrument for medical training.
+Not only it would constitute an invaluable tool for teaching to medical students but it would also enable experts clinicians to practice rare and complex cases that they would otherwise never encounter before facing them.
+Moreover, a highly realistic simulator would also allow planning of complex surgical procedures and per-operative guidance.
+
+While substantial progress was accomplished during the past ten years, much remains to be done.
+In fact, the research described in this PhD fits in the context of computed-based simulation.
+Medical simulation can be broken up into different areas of research, all are very challenging.
+%
+\begin{description}
+\item[Photorealistic rendering:] first, the scene needs to look real.
+A rendered image can be understood in terms of a number of visible features that a simulated image wants to reproduce: colors, brightness, shadows, reflection, transparency, refraction, diffraction, etc.
+Therefore, a series of fairly complex calculations is required for all visible structures within each image (usually displayed at $30\,$Hz).
+\item[Collision detection] involves algorithms to check for collision between solids.
+Without them, tools could go through organs and other obstacles during the simulation.
+The collision detection algorithm has to find all intersections between the objects at each time step of the simulation.
+\item[Contact modelling:] once the collision between two objects has been detected, a variety of responses to this contact may be formulated.
+The two solids may rebound away from each other, slide (with or without friction), or settle into relative static contact.
+This step must also be carried out at each time step.
+\item[Anatomical structure modelling and haptic feedback:] the simulator must also take into account the physics involved in the deformation of anatomical structures.
+Indeed, when a pressure is exerted on a given organ, the latter must deform and react according to its mechanical properties.
+The force created by this interaction may be sent back to the user through a haptic device.
+This device provides the user with a sense of touch so he can feel the forces and vibrations due to the contact.
+The force feedback must take place at a very high frequency (often between $ 500 $ and $ 1\,000\,$Hz) for a smooth sense of touch.
+\end{description}
+%
+Although each of these steps alone is computationally demanding, the whole simulation process must be computed in real-time or close to real-time.
+Real-time means that one second of simulation corresponds to one second of real life.
+Of particular interest to us in this PhD is the accurate modelling of the deformation of anatomical structures in real-time.
+
+\bigskip
+
+Along with the increase in computational power, approaches for modelling anatomical structures became more elaborate and less approximative.
+These different approaches may be classified into three main categories: (1) geometrically based techniques, (2) approaches physically motivated and (3) methods based on the equations of continuum mechanics.
+Techniques based on geometry tolerate a degradation of realism as long as the result looks realistic and are typically used in the game industry.
+As an example, the technique of shape matching proposed by \cite{Muller05} makes use of a simple particle system without particle-particle interactions, but including response to collisions with the environment and including external forces such as gravity.
+But the most popular technique applied in simulation is the mass-spring system \citep{Terzopoulos91,Provot95,Mosegaard05b} .
+In this method, the geometry is described by a network of masses connected together by springs.
+The system of equations is solved using Newton's second law.
+Finally, the most computationally demanding techniques are the one relying on the equations of continuum mechanics.
+These equations are regarded as the most accurate mathematical description available for modelling the deformation of a continuum.
+The finite element method is a numerical procedure that allows to solve the partial differential equations provided by continuum mechanics \citep{Cotin99}.
+With the continual growth in computational power, finite element formulations applied to medical simulation become more elaborate \citep{Sagar94,Felippa00,Debunne01,Miller07}.
+For instance, they tend to employ more accurate constitutive laws to better describe the complex mechanical behaviour of organs \citep{Fung93}.
+
+\section{Our contributions}
+
+Through this PhD, we followed the trend of using more elaborate models capable of describing with great accuracy the mechanical behaviour of anatomical structures.
+Moreover, we strived to develop efficient formulations to model entire organs in real-time.
+Our work emerged from the following observation: the shape and the internal nature of the anatomical structures greatly differ.
+In the light of this statement, we believe that a unique model cannot deal with the variousness of the mechanical behaviours encountered in the human body.
+We identified two major types of structures in human body, they can be either solid (brain, liver, prostate etc.) or hollow (colon, blood vessels, stomach etc.).
+Consequently, this PhD thesis lies in two parts: solid structure modelling and hollow structure modelling.
+The first part deals with the first GPU implementation of a non-linear, anisotropic and viscoelastic finite element procedure.
+The second part introduces a framework for real-time modelling of thin anatomical structures via a parallelisable co-rotational shell finite element formulation and a method to mesh a complex surface with curved shell elements.
+In both cases, their implementation in the open source framework SOFA are presented.
+By providing these new models to the medical simulation community, we hope to assist in enhancing the realism of medical simulators.

### 2010-10-07 Thesis/Appendix1.tex
Delete most of dummy contents
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@123 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 8 September -
Started writing of state of the art for shells + bib updated



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@170 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 7 October -
Changes added in intro and conclusion + references improved + TODO removed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@183 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\chapter{Appendix Example}
-\label{chap:appendix1}
+\chapter{The weighted residual method}
+\label{appendix1}
 
-\section{Appendix example section}
+The \emph{method of weighted residuals} is an approximation technique for solving differential equations.
+A subclass of this method, the \emph{Galerkin method of weighted residuals}, is often used to derive the element equations for the finite element method.
 
-And I cite myself to show by bibtex style file (two authors) \citep{Commowick_MICCAI_2007}.
+Let us suppose that we have a linear differential operator $ D $ acting on a function $ u $ to produce a function $ p $:
+\begin{equation}
+D(u(x)) = p(x).
+\end{equation}
+We wish to approximate $ u $ by a function $ \tilde{u} $, which is a linear combination of basis functions chosen from a linearly independent set, that is: 
+\begin{equation}
+u \approx \tilde{u} = \sum_{i=1}^n a_i \phi_i .
+\end{equation}
+Now, when substituted into the differential operator $ D $, the result of the operation is not, in general, $ p(x) $. Hence, an error or \emph{residual} will exist:
+\begin{equation}
+R(x) = D(\tilde{u}(x)-p(x)) \neq 0.
+\end{equation}
+The idea behind the weighted residual method is to force the residual to zero in some average sense over the domain:
+\begin{equation}
+\int_X R(x) W_i(x) dx = 0, \quad i = 1, 2, \ldots, n
+\end{equation}
+where the number of \emph{weight functions} $ W_i $ is exactly equal to the number of unknown constants $ a_i $ in $ \tilde{u} $.
+The result is a set of $ n $ algebraic equations for the unknown constants $ a_i $.
+Depending on the choice for $ W_i $, several sub-methods can be derived: collocation, sub-domain, least squares, Ritz, Galerkin etc.
 
-This for other bibtex stye file : only one author \cite{Oakes_RStat_1999} and many authors \cite{Guimond_CVIU_2000}.
+%For instance, the weight function used by the Galerkin method \TODO{need to check this: Wi are the shape functions?} is:
+%\begin{equation}
+%W_i = \pd{\tilde{u}}{a_i}.
+%\end{equation}

### 2010-10-07 Thesis/Chapter1.tex
- Update 7 October -
Changes added in intro and conclusion + references improved + TODO removed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@183 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 As an example, a report published in 2000 by the Institute of Medicine in the United States \citep{Kohn00} described two studies carried out in the 1990s based on large samples of hospital admissions.
-They found that the proportion of hospital admissions experiencing an adverse event, defined as causes of injuries by medical management, were 2.9 and $3.7\,$\%, respectively .
+They found that the proportion of hospital admissions experiencing an adverse event, defined as causes of injuries by medical management, were 2.9 and $3.7\,$\%, respectively.
 The proportion of these events attributable to errors (that is, preventable adverse events, in other words medical errors) was 58 and $53\,$\%, respectively.
@@
 Because healthcare is a high risk industry (like aviation or military), training in the real world is too costly and dangerous.
-Consequently, various approaches were applied to teaching and training of medical practicians over history of healthcare.
+Consequently, various approaches were applied for teaching and training of medical practicians over history of healthcare.
 
@@
 Nowadays, similar passive models are still used to help students learn the anatomy.
-In medical schools, live animals are used for teaching physiology and surgery classes.
+In medical schools, live animals are also used for teaching physiology and surgery classes.
 Dogs, pigs, goats and sheep are commonly used in medical training.
@@
 Such systems provide an elegant solution for medical training since they can offer a realistic and configurable training environment.
-Computer-based simulators also appear as an ideal tool for skill assessment of students since virtually every parameter of the simulation can be measured and recorded.
+Computer-based simulators appear as an ideal tool to assess students' skills since virtually every parameter of the simulation can be measured and recorded.
 Increasing computational power and growing interest for the field over the last decade led to the rapid development of reasonably sophisticated simulation systems.
@@
 
-Because everything is virtual, the possibilities are almost infinite.
-Ultimately, the ideal simulator will feature a patient-specific anatomy obtained through a medical imaging technique (such as MRI or CT).
-It will also need to consider the actual properties of the anatomical structures of this given patient, to take into account both the variability across patients and the alteration due to possible diseases.
+Because computer-based simulators are entirely virtual, the possibilities are almost infinite.
+The ideal simulator will ultimately feature a patient-specific anatomy obtained through a medical imaging technique (such as MRI or CT).
+It will also consider the actual properties of the anatomical structures of this given patient, to take into account both the variability in mechanical properties across patients and potential alterations due to diseases.
 Indeed, diseased organs are often much stiffer than healthy structures and this will affect the overall mechanical behaviour.
@@
 Such a patient-specific simulator capable of replicating human physiology and pathologies would be a great instrument for medical training.
-Not only it would constitute an invaluable tool for teaching to medical students but it would also enable experts clinicians to practice rare and complex cases that they would otherwise never encounter before facing them.
+Not only it would constitute an invaluable tool for teaching to medical students but it would also enable experts clinicians to practice rare and complex cases that they would otherwise never encounter before facing them in real life.
 Moreover, a highly realistic simulator would also allow planning of complex surgical procedures and per-operative guidance.
@@
 While substantial progress was accomplished during the past ten years, much remains to be done.
-In fact, the research described in this PhD fits in the context of computed-based simulation.
+In fact, the research described in this PhD fits in this context of computed-based simulation.
 Medical simulation can be broken up into different areas of research, all are very challenging.
@@
 \item[Photorealistic rendering:] first, the scene needs to look real.
-A rendered image can be understood in terms of a number of visible features that a simulated image wants to reproduce: colors, brightness, shadows, reflection, transparency, refraction, diffraction, etc.
+A rendered image can be understood in terms of a number of visible features that we want to reproduce in a simulated image: colors, brightness, shadows, reflection, transparency, refraction, diffraction, etc.
 Therefore, a series of fairly complex calculations is required for all visible structures within each image (usually displayed at $30\,$Hz).
@@
 Indeed, when a pressure is exerted on a given organ, the latter must deform and react according to its mechanical properties.
-The force created by this interaction may be sent back to the user through a haptic device.
+The force created by this interaction may be sent back to the user through a so-called haptic device.
 This device provides the user with a sense of touch so he can feel the forces and vibrations due to the contact.

### 2010-10-07 Thesis/Chapter10.tex
- Update 2 October -
A few changes in shell part + part of conclusion




git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@180 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 7 October -
Changes added in intro and conclusion + references improved + TODO removed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@183 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-Our idea is the following: while many flat triangles are required to describe highly curved surfaces, fewer triangular shell elements are needed to describe the given geometry with the same precision since they can be curved.
+Our idea is the following: while many flat triangles are required to describe highly curved surfaces, fewer triangular shell elements are needed to describe a given geometry with the same precision since they can be curved.
 Our goal is to create a mesh featuring the optimal number of shell elements while staying as close as possible to our targeted geometry.
@@
 We propose a technique that can simplify the high resolution mesh of the object to be simulated and approximate the surface of anatomical structures with shell elements whose each surface is described by the shape function used in our shell formulation.
-Thus, the first step in the process of generating a shell-based mesh is an important decimation of the high resolution mesh, using quadric edge collapse technique implemented in \cite{Meshlab}.
+Thus, the first step in the process of generating a shell-based mesh is an important decimation of the high resolution mesh, using quadric edge collapse technique implemented in \citeauthor{Meshlab}.
 MeshLab is an open source program for the processing and editing of unstructured 3D triangular meshes.

### 2010-10-07 Thesis/Chapter11.tex
- Update 7 October -
Changes added in intro and conclusion + references improved + TODO removed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@183 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 However, to model the deformation of complex anatomical structures using shell elements, the first step is to describe its surface with curved patches.
-Indeed, while many flat triangles are required to describe highly curved surfaces, fewer triangular shell elements are needed to describe the given geometry with the same precision since they can be curved.
+Indeed, while many flat triangles are required to describe highly curved surfaces, fewer triangular shell elements are needed to describe a given geometry with the same precision since they can be curved.
 Therefore, we suggested a simple method to mesh a surface with curved shell elements.
 The novelty of our method relies on the combination of a shell finite element formulation and a geometric surface reconstruction both based on the same polynomial interpolation function used to describe the surface of shells.
-We also showed how contacts and interactions with the curved surfaces of shells can be handled using the same function.
+We also showed how contacts and interactions with the curved surfaces of shells can be detected and handled using the same function.
 
@@
 If this permits significant computational advantage, the use of an explicit integration scheme imposes that the time step of the simulation remains below a critical time step.
-This critical time step depends on two kind of criterion: the size of the elements and the material?s properties (Young's modulus and Poisson's ratio).
-Therefore, meshing of anatomical structures to be simulated must be accomplished with great care to insure well-shaped elements.
+This critical time step depends on two kinds of criteria: the size of the elements and the material's properties (Young's modulus and Poisson's ratio).
+Therefore, the meshing of the anatomical structures to be simulated must be accomplished with great care to insure well-shaped elements.
 Even more restricting, the use of the TLED formulation is not suitable for modelling stiff objects in real-time.
@@
 However, those potential changes are substantial and their implementations are not straightforward.
-In addition, it is not clear from which method the benefits would be the greatest: whether from enhancing our model to a non-linear formulation or completely changing to another model based on true shell theory (where membrane and bending energies are not separated).
+In addition, it is not clear from which method the benefits would be the greatest: whether from enhancing our model to a non-linear formulation or completely changing to another model based on true mathematical shell theory (where membrane and bending energies are not separated).
 Our shell FEM could also be improved by implementing the formulation on GPU.
@@
 The lack of accuracy induced by these simplifications is of great concern to us.
-In this PhD, a part ot the solution was to accelerate the computations by using GPU.
+In this PhD, a part ot the solution was to accelerate the computations by using the GPU.
 While we have showed that this approach can work very well, we need to keep in mind that GPU acceleration is not the panacea.
@@
 While OpenCL is openly targeted as a multi-vendor and multi-platform API, its usage remains very scarce nevertheless.
-In fact, the first experiments carried out within the SOFA framework revealed that OpenCL is not as optimised as CUDA for instance.
+In fact, the first experiments carried out within the SOFA framework revealed that OpenCL is not as optimised as CUDA for instance (up to twice as slow).
 Secondly, GPU programming is not simple to master.
@@
 Earlier this year, NVIDIA released a new GPU architecture called \emph{Fermi}.
-Fermi features several major innovations for GPGPU computing like cache (faster memory access in some cases), unified address space (easier to run C and C++ programs), debugging support, faster atomic instructions, compliance with the newly revised IEEE 754 floating point standard, 3 times the amount of shared memory, etc.
+Fermi features several major innovations for GPGPU computing like cache (faster memory access in some cases), unified address space (easier to run C and C++ programs), debugging support, faster atomic instructions, compliance with the newly revised IEEE 754 floating point standard, three times the amount of shared memory, etc.
 In theory, this series of improvement should boost the performance of the second and third TLED implementations (see section \ref{chap6:secondTLED}) and we should investigate and compare the efficiency of the three designs on this new hardware.
@@
 This physical approach would allow a more even distribution of the vertices across the surface and would contribute to obtain better quality elements overall.
-While a more mathematical method (a derivative of advancing front techniques for instance) could probably be formulated, the problem is complex and could account for another PhD.
+While a more mathematical method (a derivative of advancing front techniques for instance) could probably be formulated, the problem is complex and could certainly account for another PhD.
 

### 2010-10-07 Thesis/Chapter3.tex
- Update 7 October -
Changes added in intro and conclusion + references improved + TODO removed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@183 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 Algebraic relations between the undetermined coefficients are then obtained by satisfying the governing equations over each element.
-There are a few types of approach for establishing these equations but, without going into details, the mathematical foundations of all these approaches are energy principles or the \emph{weighted residual methods} (see Appendix~\ref{appendix2}), which both lead to integrals during the process.
+There are a few types of approach for establishing these equations but, without going into details, the mathematical foundations of all these approaches are energy principles or the \emph{weighted residual methods} (see Appendix~\ref{appendix1}), which both lead to integrals during the process.
 Therefore, the second stage creates two sources of error: the representation of the solution by a linear combination of functions and the evaluation of the integrals.

### 2010-10-07 Thesis/Chapter4.tex
- Update 6 October -
Besides acknowledgements, the draft is finished!



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@182 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 7 October -
Changes added in intro and conclusion + references improved + TODO removed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@183 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \ON
-The challenges faced by the field of medical simulation were reviewed in details in the introductory chapter (see chapter \ref{chap1}).
+The challenges faced by the field of medical simulation were reviewed in the introductory chapter (see chapter \ref{chap1}).
 Hence, we know that at the very least, the deformation of solid organs must appear physically realist in medical simulators.
@@
 It can produce unnatural results with deformations of large amplitude.
-\cite{Choi05} proposed to overcome this limitation by taking the rotational component of the deformation into account at each node.
+\cite{Choi05a} proposed to overcome this limitation by taking the rotational component of the deformation into account at each node.
 \cite{Barbic05} make use of modal analysis to build a deformation basis from modal derivatives.
@@
 For the first time, the whole simulation of was performed on the GPU (physics, interaction and visualisation) and they used their framework in a congenital cardiac surgery simulation.
-They compared two implementations: the first one explicitly defined the connectivity between springs using a connectivity texture while the second one determines implicitly determines at runtime that springs are connected when they are neighbours.
+They compared two implementations: the first one explicitly defines the connectivity between springs using a connectivity texture while the second one implicitly determines at runtime that springs are connected when they are neighbours.
 They were able to compute 188 iterations per second on a $42\,745$ particule model of a pig heart.
@@
 With this idea in mind, \cite{Sorensen06} reviewed the most important concepts and data structures required to realise two popular deformable models on the GPU: the spring-mass model and the finite element model.
-
-\TODO{Other GPU-based mass-spring implementations?}
+\ON In 2008, \citeauthor{Rasmusson08} investigated performance and optimisation strategies between the two GPU languages OpenGL and CUDA for their mass-spring-damper model.
+More recently, \cite{Leon10} presented a GPU implementation of a mass-spring model with a nearly $80\times$ speedup over serial execution and $20\times$ speedup over CPU-based parallel execution.
+\OFF
 

### 2010-10-07 Thesis/Chapter6.tex
- Update 5 October -
Added conclusion, abstract, resume and beginning of introduction



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@181 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 7 October -
Changes added in intro and conclusion + references improved + TODO removed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@183 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 This chapter will describe the implementation of this non-linear, viscoelastic and anisotropic finite element method algorithm in the open source framework SOFA.
-By providing an efficient and accurate non-linear FEM for soft tissue modelling to worldwide researchers, we thus hope to assist in enhancing the realism of medical simulators.
 This work has led to three publications \citep{Comas2008,Taylor2008,Taylor2009}.
@@
 
-Interface with a GPU has traditionally been via a graphic application programming interface (API) such as \citetalias{Board05} or \citetalias{DirectX10}.
+Interface with a GPU has traditionally been via a graphic application programming interface (API) such as \citetalias{OpenGL} or \citeauthor{DirectX}.
 OpenGL is an open standard API that provides a number of functions for the rendering of 2D and 3D graphics and is available on most modern operating systems including but not limited to Windows, Mac OS X and Linux.
@@
 
-	\subsection{Second implementation: a better use of shared memory}
+	\subsection{Second implementation: a better use of shared memory}	\label{chap6:secondTLED}
 
@@
 
-\TODO{If time, try this implementation on Geforce GTX 480 as it features 3 times the amount of shared memory.}
+% \TODO{If time, try this implementation on Geforce GTX 480 as it features 3 times the amount of shared memory.}
 
@@
 
-\TODO{If time, try simple implementation using atomic writes with floats on Geforce GTX 480}
+% \TODO{If time, try simple implementation using atomic writes with floats on Geforce GTX 480}
 

### 2010-10-07 Thesis/Chapter7.tex
- Update 2 October -
A few changes in shell part + part of conclusion




git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@180 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 7 October -
Changes added in intro and conclusion + references improved + TODO removed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@183 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 In this chapter, we will review diverse techniques applied for the simulation of thin objects.
-These different approaches may be grouped into three main categories: (1) mass-spring models, (2) approaches which derive a bending energy from geometrical considerations and (3) methods based on the equations of continuum mechanics.
+These different approaches may be grouped into three main categories: (1) models, (2) approaches which derive a bending energy from geometrical considerations and (3) methods based on the equations of continuum mechanics.
 Although very few models have been proposed for simulating the deformation of thin anatomical structures in the field of medical simulation, we will try to emphasize medical applications whenever possible.
@@
 \cite{Oshita01} used a sparse triangular mesh and an interpolation to generate a dense mesh.
+And \cite{Georgii05} made use of GPU acceleration and compared two different implementations for modelling cloth patches.
 
@@
 \begin{center}
-\includegraphics[width=8cm]{chapter7/hat.pdf}
+\includegraphics[width=9cm]{chapter7/hat.pdf}
 \caption[Comparison between real footage and simulation of a hat dropped on a table]{Comparison between real footage and simulation of a hat dropped on a table. Image courtesy of \cite{Grinspun03}.}
@@
 To the best of our knowledge, no real-time shell finite element method was presented in the field of medical simulation.
-\TODO{need to triple-check this...}
 

### 2010-10-07 Thesis/Chapter9.tex
- Update 7 October -
Changes added in intro and conclusion + references improved + TODO removed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@183 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-	\subsection{Inversion of matrix $ \mathbf{C} $}
-\cite{Przemieniecki85} described the condition of non-inversion for the matrix $\mathbf{C}$ as the following:
-\TODO{add condition of inversion... but need the book :)}
-
-Each triangle of a mesh obtained from a regular grid of points fits this condition of non-inversion (right triangles).
+%	\subsection{Inversion of matrix $ \mathbf{C} $}
+%\cite{Przemieniecki85} described the condition of non-inversion for the matrix $\mathbf{C}$ as the following:
+%\TODO{add condition of inversion... but need the book :)}
+%
+%Each triangle of a mesh obtained from a regular grid of points fits this condition of non-inversion (right triangles).
 We fixed this limitation by choosing a different local coordinate system that the one introduced by \citeauthor{Przemieniecki85}.
-
-\TODO{add figure with the two different coordinate system}
+%
+%\TODO{add figure with the two different coordinate system}
 

### 2010-10-07 Thesis/Resume.tex
- Update 7 October -
Changes added in intro and conclusion + references improved + TODO removed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@183 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-Bien que les deux modèles de tissus sont basés sur la mécanique continue pour une meilleure précision, ils sont tous les deux capables de simulation la déformation d'organes en temps réel.
-Enfin, leur implémentation dans l'environnement open source SOFA rendra publique ces deux modèles afin de participer à l'amélioration du réalisme des simulateurs médicaux.
+Bien que les deux modèles de tissus soient basés sur la mécanique continue pour une meilleure précision, ils sont tous les deux capables de simuler la déformation d'organes en temps réel.
+Enfin, leur implémentation dans l'environnement open source SOFA permettra la diffusion de ces deux modèles afin de participer à l'amélioration du réalisme des simulateurs médicaux.
 

### 2010-10-11 Thesis/Chapter2.tex
- Update 7 October -
Changes added in intro and conclusion + references improved + TODO removed



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@183 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 11 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@184 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 % 
-\begin{figure}[h]
+\begin{figure}[ht]
 \begin{center}
@@
 %
-\begin{figure}[h]
+\begin{figure}[ht]
 \begin{center}
@@
 The form of $U_0$ may varied for different materials.
-A material is said to be \emph{linear} if the relationship between strain and stress is also linear.
+A material is said to be \emph{linear} if the relationship between strain and stress is also linear (see \fig{chap2:fig-linear}).
 Based on this definition, $U_0$ is a cubic or higher-order function of the strains for non-linear materials whereas $U_0$ is a quadratic function of strain for linear materials.
@@
 Consequently, we will not make any distinction between the various measures of stress and strain and use $\boldsymbol \sigma$ and $\boldsymbol \varepsilon$ for stress and strain tensors respectively.
+%
+\begin{figure}[ht]
+\centering
+\includegraphics[width=7cm]{chapter2/linear.pdf}
+\caption {The relationship between strain and stress is often non-linear (blue). However, it may be approximated as a linear relationship for small deformations (red). }
+\label{chap2:fig-linear}
+\end{figure}
 
@@
 The neo-Hookean model provides a reasonable approximation of rubber-like material for moderate strains.
-\TODO{Do I keep this range of validity, which seems somewhat limited...} It is typically accurate for strains less than $20\, \%$ \citep{Gent01}.
+%  It is typically accurate for strains less than $20\, \%$ \citep{Gent01}.
 
@@
 
-Viscoelastic materials present a few characteristics: 
+Viscoelastic materials present a few characteristics (see \fig{chap2:fig-visco}): 
 \begin{enumerate}
@@
 \end{enumerate}
+%
+\begin{figure}[ht]
+\centering 
+\subfloat[Creep]{\includegraphics[width=6.5cm]{chapter2/visco1.pdf}}
+\hspace{0.5cm}
+\subfloat[Relaxation]{\includegraphics[width=6.5cm]{chapter2/visco2.pdf}} \\
+\subfloat[Rate-dependent]{\includegraphics[width=6.5cm]{chapter2/visco3.pdf}}
+\hspace{0.5cm}
+\subfloat[Hysteresis]{\includegraphics[width=6.5cm]{chapter2/visco4.pdf}}
+\caption[Characteristics of viscoelastic materials]{Characteristics of viscoelastic materials}
+\label{chap2:fig-visco}
+\end{figure}
 
@@
 A non-invasive technique called \emph{elastography} was introduced recently to measure Young's modulus.
-The first device based on this technique was commercialised by Echosens in 2003 \cite{Bastard09}.
+The first device based on this technique was commercialised by Echosens in 2003 \citep{Bastard09}.
 The techniques of elastography may be classified in two categories: static and dynamic elastography.

### 2010-10-11 Thesis/Chapter3.tex
- Update 11 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@184 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 %
-\begin{figure}[h]
+\begin{figure}[ht]
 \begin{center}
-\includegraphics[width=8cm]{chapter3/areaUnderCurve.pdf}
+\includegraphics[width=10cm]{chapter3/areaUnderCurve.pdf}
 \end{center}
@@
 \ON Hence, \OFF a formulation based on a weak form usually produces a set of discretised system equations that give much more accurate results, especially for problems of complex geometry.
-\ON For this reason, \OFF the weak form is usually preferred \ON over the strong form for obtaining an approximate solution of a practical engineering problem \OFF.
+\ON For this reason, \OFF the weak form is usually preferred \ON over the strong form for obtaining an approximate solution of a practical engineering problem.
+\OFF
 
@@
 And we suppose the element connectivity information given by Table \ref{chap3:connectivity}, which provides the global node number for each local node number within an element.
-\begin{table}
+%
+\begin{table}[ht]
 	\begin{center}
@@
 \end{table}
+%
 We also assume the following stiffness matrices for the three elements:

### 2010-10-11 Thesis/Chapter4.tex
- Update 11 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@184 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \begin{center}
@@
 In particular, the error brought by the approximated (linear) strain measure increases with the rotation of the element.
-This error in the measure creates ghost forces which overdeforms the simulated object.
+This error in the measure creates ghost forces which overdeforms the simulated object (see \fig{chap4:fig-cylinder}).
 The idea behind co-rotational methods is the decomposition of the motion into a rigid-body and deformational components.
@@
 Consequently, co-rotational formulations allow large displacements and rotations (since the rigid-body motion is not considered) as long as the strain remains small (the strain measure is still linear).
+%
+\begin{figure}[ht]
+\centering 
+\includegraphics[height=6cm]{chapter4/cylinder1.pdf}
+\hspace{0.5cm}
+\includegraphics[height=6cm]{chapter4/cylinder2.pdf}
+\caption[Deformation of cylinders]{A geometrically linear finite element model (green) is compared to a co-rotational finite element model (yellow).
+While still using a simple linear strain measure, the co-rotational formulation is not sensitive to rotations and thus does not overdeform.}
+\label{chap4:fig-cylinder}
+\end{figure}
 

### 2010-10-11 Thesis/Chapter6.tex
- Update 11 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@184 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\begin{figure}
+\begin{figure}[ht]
 \includegraphics[width=13.9cm]{chapter6/sofa.pdf}
@@
 
-% \TODO{If time, try this implementation on Geforce GTX 480 as it features 3 times the amount of shared memory.}
+% \{If time, try this implementation on Geforce GTX 480 as it features 3 times the amount of shared memory.}
 
@@
 \centering 
-\subfloat[ ]{\includegraphics[width=5cm]{chapter6/cube_and_curves1.pdf}}
-\hspace{0.5cm}
+\subfloat[ ]{\includegraphics[width=5.5cm]{chapter6/cube_and_curves1.pdf}}
+\hspace{0.2cm}
 \subfloat[ ]{\includegraphics[width=8cm]{chapter6/cube_and_curves2.pdf}}
 \caption [Computational timings within SOFA] {(a) FEM mesh of cube with $29\,791$ nodes deformed under a uniformed load.
-(b) Comparison of GPU computational timings and CPU overheads between the SOFA and standalone implementations for different mesh sizes.
-\TODO{change the cube if I can get a decent visual model in SOFA}}
+(b) Comparison of GPU computational timings and CPU overheads between the SOFA and standalone implementations for different mesh sizes.}
 \label{chap6:fig-cube_and_curves}

### 2010-10-11 Thesis/Chapter7.tex
- Update 11 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@184 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \includegraphics[width=10cm]{chapter7/dress.pdf}
-\caption[Simulation of complete garments that have large bending stiffness]{The model of \citeauthor{Volino06} allows the simulation of complete garments with material that may have fairly large bending stiffness.
+\caption[Simulation of complete garments that have large bending stiffness]{The model of Volino and Magnenat-Thalmann allows the simulation of complete garments with material that may have fairly large bending stiffness.
 Image courtesy of \cite{Volino06}.}
@@
 
-Shell finite element modelling followed the same trend than solid objects and co-rotational formulations were introduced to simplify strain expressions while allowing large rotations (but small strain) (see page \pageref{chap4:corotationalMethods} for more details).
+Shell finite element modelling followed the same trend than solid objects and co-rotational formulations were introduced to simplify strain expressions while allowing large rotations (but small strain, see page \pageref{chap4:corotationalMethods} for more details).
 For instance, \cite{Etzmuss03} used the linear Cauchy strain tensor but applied a polar decomposition on the deformation gradient to allow large deformations.

### 2010-10-11 Thesis/Chapter9.tex
- Update 11 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@184 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 The differences with the same quantities computed for the rest position during the initialisation lead to the desired displacements.
-
-\begin{lstlisting}[caption={Method used to compute the rotation of the z axis between the frames embodied by two quaternions},label=chap9:qDiffZ]
+%
+\begin{lstlisting}[caption={[Method to compute the rotation between two quaternions]Method used to compute the rotation of the z axis between the frames embodied by two quaternions},label={chap9:qDiffZ},frame=shadowbox,rulesepcolor=\color{black}]
 inline Quat qDiffZ(const Quat& Qvertex, const Quat& Qtriangle)
 {
-    // dQ is the quaternion that embodies the rotation between the z axis of the vertex and the z axis of the local triangle's frame (in local space)
+    // dQ is the quaternion that embodies the rotation between 
+    // the z axis of the vertex and the z axis of the local 
+    // triangle's frame (in local space)
     Quat dQ;
@@
 
-    // v = z axis of the vertex's frame is expressed into world space
+    // v = z axis of the vertex's frame is expressed into 
+    // world space
     Vec3d v = Qvertex.rotate(Vec3d(0.0, 0.0, 1.0));
@@
 
-    // Axis of rotation between the 2 vectors u and v lies into the plan of the 2 vectors
+    // Axis of rotation between the 2 vectors u and v lies 
+    // into the plan of the 2 vectors
     Vec3d axis = cross(u, v0);
@@
 %
-%\begin{figure}
-%	\begin{center}
-%		\begin{tabular}{cc}
-%			\includegraphics[width=6cm]{chapter9/sub_grid_red.png} & \includegraphics[width=6cm]{chapter9/irregular_grid_red.png} \\
-%			\includegraphics[width=6cm]{chapter9/sub_grid_mesh.png} & \includegraphics[width=6cm]{chapter9/irregular_grid_mesh.png} \\
-%			Subdivided triangles & Irregular grid			
-%		\end{tabular}
-%	\end{center}
-%	\label{chap9:fig-rendering}
-%	\caption{We can notice some rendering artifacts when we use subdivided triangles. Using an irregular mesh instead reduces the impact.}
-%\end{figure}
-%
 \begin{figure}[ht]
 \centering 
-\subfloat[Subdivided triangles]{\includegraphics[width=5cm]{chapter9/sub_grid_red.png}}
+\subfloat[Subdivided triangles]{\includegraphics[width=5.5cm]{chapter9/sub_grid_red.png}}
 \hspace{1cm}
-\subfloat[Irregular grid]{\includegraphics[width=5cm]{chapter9/irregular_grid_red.png}} \\
-\subfloat[Subdivided triangles]{\includegraphics[width=5cm]{chapter9/sub_grid_mesh.png}}
+\subfloat[Irregular grid]{\includegraphics[width=5.5cm]{chapter9/irregular_grid_red.png}} \\
+\subfloat[Subdivided triangles]{\includegraphics[width=5.5cm]{chapter9/sub_grid_mesh.png}}
 \hspace{1cm}
-\subfloat[Irregular grid]{\includegraphics[width=5cm]{chapter9/irregular_grid_mesh.png}}
+\subfloat[Irregular grid]{\includegraphics[width=5.5cm]{chapter9/irregular_grid_mesh.png}}
 \caption[Visual artifacts using triangle subdivisions]{In some situations, some rendering artifacts may appear when we use subdivided triangles to render the shells (a,c).

### 2010-10-12 Thesis/Chapter5.tex
- Update 11 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@184 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 12 October -
memoir class used now, fancy chapter style



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@185 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\chapter{The total Lagrangian explicit dynamics (TLED) algorithm}
+\chapter{The total Lagrangian explicit dynamics algorithm}
 \label{chap5}
@@
 \end{equation}
-where $ \mathbf{C} $ is the right Cauchy-Green deformation tensor, and we have introduced $ \boldsymbol \Phi \, {\buildrel\rm	def\over=} \, 2 \partial \Psi / \partial \mathbf{C} $ as the instantaneous hyperelastic stress response.
+where $ \mathbf{C} $ is the right Cauchy-Green deformation tensor, and we have introduced $ \boldsymbol \Phi \, {\buildrel\mathrm	def\over=} \, 2 \partial \Psi / \partial \mathbf{C} $ as the instantaneous hyperelastic stress response.
 
@@
 Transversely isotropic materials are characterised by a single preferred direction $\mathbf{a}_0$ in the reference configuration; the mechanical response is isotropic in the plane orthogonal to this direction.
-The strain energy is then a function of both $  \mathbf{C} $ and a structure tensor $\mathbf{A}_0 \, {\buildrel\rm	def\over=} \,  \mathbf{a}_0 \otimes \mathbf{a}_0$, where $\otimes$ denotes a tensor product.
+The strain energy is then a function of both $  \mathbf{C} $ and a structure tensor $\mathbf{A}_0 \, {\buildrel\mathrm	def\over=} \,  \mathbf{a}_0 \otimes \mathbf{a}_0$, where $\otimes$ denotes a tensor product.
 Analogous to the isotropic case, we consider strain energy functions of the form

### 2010-10-12 Thesis/Chapter6.tex
- Update 12 October -
memoir class used now, fancy chapter style



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@185 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 A stress-relaxation type test protocol was simulated, in which the cube was compressed by $30\, \%$ in 0.5 seconds and the compression was held for a further 4.5 seconds.
-We demonstrate the effects of anisotropy on the deformation response by comparing the deformed shapes of anisotropic models with $\mathbf{a}_0 = [0\, 1\, 0] \, {\buildrel\rm	def\over=} \, \mathbf{a}_0^y $ and $\mathbf{a}_0 = [0\, \frac{1}{\sqrt{2}} \, \frac{1}{\sqrt{2}}] \, {\buildrel\rm	def\over=} \, \mathbf{a}_0^{yz} $  with that of an isotropic model (with $\eta = 0$).
+We demonstrate the effects of anisotropy on the deformation response by comparing the deformed shapes of anisotropic models with $\mathbf{a}_0 = [0\, 1\, 0] \, {\buildrel\mathrm	def\over=} \, \mathbf{a}_0^y $ and $\mathbf{a}_0 = [0\, \frac{1}{\sqrt{2}} \, \frac{1}{\sqrt{2}}] \, {\buildrel\mathrm	def\over=} \, \mathbf{a}_0^{yz} $  with that of an isotropic model (with $\eta = 0$).
 
@@
 \begin{figure}[ht]
-\begin{center}
-\includegraphics[width=13.5cm]{chapter6/timingsGPU.pdf}
-\end{center}
+\centering 
+\subfloat[ ]{\includegraphics[width=6.5cm]{chapter6/timingsGPU1.pdf}}
+\hspace{0.5cm}
+\subfloat[ ]{\includegraphics[width=6.5cm]{chapter6/timingsGPU2.pdf}} \\
+\subfloat[ ]{\includegraphics[width=6.5cm]{chapter6/timingsGPU3.pdf}}
+\hspace{0.5cm}
+\subfloat[ ]{\includegraphics[width=6.5cm]{chapter6/timingsGPU4.pdf}} \\
 \caption[GPU solution times]{(a), (b) show CPU and GPU solution times, respectively, for a single time step for each constitutive model as a function of model size, (c) shows the ratio of CPU to GPU solution times for model TIV, and (d) shows ratios of GPU solution times for the constitutive models: TIV to TIE, TIE to NHE, and TIV to NHE.

### 2010-10-19 Thesis/Chapter2.tex
- Update 12 October -
memoir class used now, fancy chapter style



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@185 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 19 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@186 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\chapter{Background in continuum mechanics for soft-tissue modelling}
+\chapter[Background in continuum mechanics for soft-tissue modelling][Background in continuum mechanics]{Background in continuum mechanics for soft-tissue modelling}
 \label{chap2}
@@
 A rigid-body displacement consists of a simultaneous translation and rotation of the body without changing its shape or size.
-\ON In constrast, a change in shape and\slash or size of the body from an initial configuration $\kappa_{0}$ to a new configuration $\kappa$ is called a deformation.
+In constrast, a change in shape and\slash or size of the body from an initial configuration $\kappa_{0}$ to a new configuration $\kappa$ is called a deformation.
 This new configuration $\kappa$ may be designated by the \emph{current} or \emph{deformed configuration}.
@@
 In the Lagrangian description, the position and physical properties of the particles are referred to a reference configuration $\kappa_{R}$, often chosen to be the undeformed configuration $\kappa_{0}$.
-\ON Consequently, the current coordinates ($\mathbf{x} \in \kappa$) depend on the reference coordinates ($\mathbf{X} \in \kappa_{0}$): \OFF
+Consequently, the current coordinates ($\mathbf{x} \in \kappa$) depend on the reference coordinates ($\mathbf{X} \in \kappa_{0}$): \OFF
 \begin{equation}
@@
 \end{equation}
-\ON and a typical variable $\phi$ defined over the body is expressed in terms of the coordinates $\mathbf{X}$ and time $t$: \OFF
+and a typical variable $\phi$ defined over the body is expressed in terms of the coordinates $\mathbf{X}$ and time $t$: \OFF
 \begin{equation}
@@
 \end{equation}
-\ON Let us consider a particule $X$ whose position in the reference configuration is $\mathbf{X}$.
+Let us consider a particule $X$ whose position in the reference configuration is $\mathbf{X}$.
 From \eqref{chap2:phi}, we note that the value of $\phi$ associated with this fixed particle $X$ changes over time.
-Hence, \OFF the Lagrangian description focuses its attention on the particles of the continuous body and it is usually used in solid mechanics.
+Hence, the Lagrangian description focuses its attention on the particles of the continuous body and it is usually used in solid mechanics.
 
 	\subsection{Eulerian description}
-\ON In constrast, the Eulerian description is interested in changes at fixed locations.
-This time, the motion and $\phi$ are described with respect to the current position ($\mathbf{x} \in \kappa$): \OFF
+In constrast, the Eulerian description is interested in changes at fixed locations. This time, the motion and $\phi$ are described with respect to the current position ($\mathbf{x} \in \kappa$): \OFF
 \begin{equation}
@@
 A \emph{displacement field} is a vector field of all displacement vectors for all particles in the body.
-\ON Thus, the deformed configuration $\kappa$ may be obtained from the undeformed configuration $\kappa_{0}$ by merely adding the displacement field: \OFF $\kappa_{0}$: $\chi(\mathbf{X}, t) = \mathbf{X} + \mathbf{u}(\mathbf{X})$.
+Thus, the deformed configuration $\kappa$ may be obtained from the undeformed configuration $\kappa_{0}$ by merely adding the displacement field: $\kappa_{0}$: $\chi(\mathbf{X}, t) = \mathbf{X} + \mathbf{u}(\mathbf{X})$.
 
@@
 	\subsection{Deformation gradient tensor}
-The displacement field tells us how a \ON particule \OFF displaces from the reference to the deformed configuration.
-\ON More interestingly, we would like to know how a material line would deform (stretch and rotation) within this displacement field.
-\OFF Since the length of a material line $d\mathbf{X}$ can change when going to the deformed configuration as well as its orientation, we can say that $d\mathbf{X}$ deforms into $d\mathbf{x}$.
-\ON We now seek the relation existing between $d\mathbf{x}$ in the deformed configuration and $d\mathbf{X}$ of the reference configuration.
+The displacement field tells us how a particule displaces from the reference to the deformed configuration.
+More interestingly, we would like to know how a material line would deform (stretch and rotation) within this displacement field.
+Since the length of a material line $d\mathbf{X}$ can change when going to the deformed configuration as well as its orientation, we can say that $d\mathbf{X}$ deforms into $d\mathbf{x}$.
+We now seek the relation existing between $d\mathbf{x}$ in the deformed configuration and $d\mathbf{X}$ of the reference configuration.
 \OFF
@@
 Our motivation comes from the need to write global equilibrium statements that involve integrals over volumes.
-\ON Let us consider a volume in the reference configuration formed by \OFF three non-coplanar line elements $d\mathbf{X}^{(1)}$, $d\mathbf{X}^{(2)}$ and $d\mathbf{X}^{(3)}$ as can be seen on \fig{chap2:fig-changeVolume}.
+Let us consider a volume in the reference configuration formed by three non-coplanar line elements $d\mathbf{X}^{(1)}$, $d\mathbf{X}^{(2)}$ and $d\mathbf{X}^{(3)}$ as can be seen on \fig{chap2:fig-changeVolume}.
 % 
@@
 \end{align}
-where $\mathbf{\hat{N}}_{i}$ \ON is \OFF the unit vector along $d\mathbf{X}^{i}$. \ON Similarly, we can compute the \OFF corresponding volume in the deformed configuration by:
+where $\mathbf{\hat{N}}_{i}$ is the unit vector along $d\mathbf{X}^{i}$. Similarly, we can compute the corresponding volume in the deformed configuration by:
 \begin{align}
@@
 The determinant of $\mathbf{F}$ is called the \emph{Jacobian} and it is denoted by $J = \det \mathbf{F}$.
-\ON Therefore, the relation between the deformed volume $ dv $ and the undeformed volume $ dV $ may be written as: \OFF
+Therefore, the relation between the deformed volume $ dv $ and the undeformed volume $ dV $ may be written as: \OFF
 \begin{equation}
@@
 \end{equation}
-\ON It is worth noting that the Jacobian \OFF  J has the physical meaning of being the local ratio of current to reference volume of a material volume element.
+It is worth noting that the Jacobian  J has the physical meaning of being the local ratio of current to reference volume of a material volume element.
 
@@
 \end{equation}
-\ON Since \OFF this expression is valid for any vector $d\mathbf{L}$ and by using the property $\mathbf{a} \cdot \mathbf{b} = \mathbf{a} ^{T} \mathbf{b}$, \ON the sought relation between an element of area in the reference configuration $d\mathbf{A}$ and its corresponding area $d\mathbf{a}$ in the deformed configuration may be expressed as: \OFF
+Since this expression is valid for any vector $d\mathbf{L}$ and by using the property $\mathbf{a} \cdot \mathbf{b} = \mathbf{a} ^{T} \mathbf{b}$, the sought relation between an element of area in the reference configuration $d\mathbf{A}$ and its corresponding area $d\mathbf{a}$ in the deformed configuration may be expressed as: \OFF
 \begin{equation}
@@
 	\subsection{Green-Lagrange strain tensor}
-Using \eqref{chap2:dS} and \eqref{chap2:ds}, \ON the difference in the squared lengths between the reference and the current configuration may be evaluated as: \OFF 
+Using \eqref{chap2:dS} and \eqref{chap2:ds}, the difference in the squared lengths between the reference and the current configuration may be evaluated as: 
 \begin{align}
@@
 By definition, the Green-Lagrange strain tensor is a symmetric second-order tensor.
-\ON Moreover, we observe that \OFF the change in squared lengths is zero if and only if \ON the Green-Lagrange strain tensor \OFF $\mathbf{E} = \mathbf{0}$.
+Moreover, we observe that the change in squared lengths is zero if and only if the Green-Lagrange strain tensor $\mathbf{E} = \mathbf{0}$.
 Using \eqref{chap2:gradient}, the Green strain tensor may be expanded as the following:
@@
 Therefore, it is necessary to define the stress not over a given area but at a specific point in the body.
-\ON The stress at a point in a three-dimensional continuum is fully defined by \OFF nine quantities, three per plane, on three mutually perpendicular planes at the point.
-\ON The \emph{stress tensor} is defined as the second-order tensor which components are these nine quantities.
-\OFF
+The stress at a point in a three-dimensional continuum is fully defined by nine quantities, three per plane, on three mutually perpendicular planes at the point.
+The \emph{stress tensor} is defined as the second-order tensor which components are these nine quantities.
 
@@
 
-\ON In the following we make use of a Cartesian coordinate system and we consider an infinitesimal tetrahedron as shown on \fig{chap2:fig-CauchyStress}. \OFF
+In the following we make use of a Cartesian coordinate system and we consider an infinitesimal tetrahedron as shown on \fig{chap2:fig-CauchyStress}. \OFF
 %
@@
 This stress is basically defined as force/(unit deformed area).
-The strain measure that is appropriate to use with the Cauchy stress tensor is \ON therefore \OFF the infinitesimal strain tensor (see section~\ref{chap2:infinitesimalStrainTensor}).
-\ON Indeed, because the Cauchy stress tensor is defined by unit of deformed area, it is not appropriate for analysing materials undergoing large deformation where the area in the deformed configuration is often unknown.
+The strain measure that is appropriate to use with the Cauchy stress tensor is therefore the infinitesimal strain tensor (see section~\ref{chap2:infinitesimalStrainTensor}).
+Indeed, because the Cauchy stress tensor is defined by unit of deformed area, it is not appropriate for analysing materials undergoing large deformation where the area in the deformed configuration is often unknown.
 Consequently, we require alternative stress tensors based on the reference configuration.
-\OFF We will introduce two new stress measures: the first Piola-Kirchhoff stress tensor and the second Piola-Kirchhoff stress tensor.
+We will introduce two new stress measures: the first Piola-Kirchhoff stress tensor and the second Piola-Kirchhoff stress tensor.
 
@@
 We know that the Cauchy stress tensor represents the current force per unit of deformed area: $d\mathbf{f} = \mathbf{t}(\mathbf{n}) da = \sigma \cdot d\mathbf{a}$.
-\ON We state that we can find a stress vector $\mathbf{T}$ over the area element $d\mathbf{A}$ in the underformed configuration which \OFF results in the same total force.
+We state that we can find a stress vector $\mathbf{T}$ over the area element $d\mathbf{A}$ in the underformed configuration which results in the same total force.
 \begin{equation}
@@
 In a similar way we defined the Cauchy stress $\sigma$ with $\mathbf{t}(\mathbf{n}) = \sigma \cdot \mathbf{n}$, we introduce a stress tensor $\mathbf{P}$ such that $\mathbf{T}(\mathbf{n}) = \mathbf{P} \cdot \mathbf{N}$.
-\ON We call $ \mathbf{P} $ the \emph{first Piola-Kirchhoff stress tensor} and \OFF gives the current force per unit undeformed area.
+We call $ \mathbf{P} $ the \emph{first Piola-Kirchhoff stress tensor} and gives the current force per unit undeformed area.
 
@@
 	\subsection{Second Piola-Kirchhoff stress tensor}\label{chap2:PiolaStress2}
-\ON Let us consider the stress tensor $\mathbf{S}$ resulting in the force $d\mathbf{F}$ in the undeformed area $d\mathbf{A}$ which \OFF corresponds to the force $d\mathbf{f}$ in the deformed area $d\mathbf{a}$.
-\ON We define this stress tensor $\mathbf{S}$ as the \emph{second Piola-Kirchhoff stress tensor}.
+Let us consider the stress tensor $\mathbf{S}$ resulting in the force $d\mathbf{F}$ in the undeformed area $d\mathbf{A}$ which corresponds to the force $d\mathbf{f}$ in the deformed area $d\mathbf{a}$.
+We define this stress tensor $\mathbf{S}$ as the \emph{second Piola-Kirchhoff stress tensor}.
 \OFF
@@
 \end{equation}
-\ON In other words, \OFF the second Piola-Kirchhoff stress tensor gives the transformed current force per unit undeformed area.
+In other words, the second Piola-Kirchhoff stress tensor gives the transformed current force per unit undeformed area.
 Similar to the relationship \eqref{chap2:dxdX} between $d\mathbf{x}$ and $d\mathbf{X}$, the force $d\mathbf{f}$ is related to the force $d\mathbf{F}$ via the deformation gradient tensor:
@@
 So far, all the relations that we have established are valid for every continuum.
-Indeed, \ON the derivations carried out so far made no mention of any material.
-\OFF Therefore, \ON we know that \OFF these relations are not sufficient to describe the response of a material to a given loading.
-\ON Moreover, because the stresses result from the deformation of the material, they may be expressed \OFF in terms of some measure of this deformation such as the strain.
+Indeed, the derivations carried out so far made no mention of any material.
+Therefore, we know that these relations are not sufficient to describe the response of a material to a given loading.
+Moreover, because the stresses result from the deformation of the material, they may be expressed in terms of some measure of this deformation such as the strain.
 This relationship, which obviously depends on the type of material under consideration, is what we call a \emph{constitutive equation}.
-\ON Constitutives equations are an attempt to mathematically characterise the behaviour of materials observed from experimentals results.
+Constitutives equations are an attempt to mathematically characterise the behaviour of materials observed from experimentals results.
 \OFF
 
-\ON Naturally, the relationship between strain and stress is conditioned by the constitution of the material of interest.
+Naturally, the relationship between strain and stress is conditioned by the constitution of the material of interest.
 Because the mechanics involved in materials can be quite complex (tissue mechanics in particular), many simplifying assumptions are required to derive a constitutive equation.
@@
 Moreover, not only constitutive equations need to deal with the complexity of material mechanics but they must remain simple enough to measure all the included material constants experimentally.
-\OFF This section will now introduce the most common constitutive models relevant to soft-tissue modelling.
+This section will now introduce the most common constitutive models relevant to soft-tissue modelling.
 
@@
 
-\ON In addition, \OFF if there exists a \emph{strain energy density} function $U_0(\varepsilon)$ such that 
+In addition, if there exists a \emph{strain energy density} function $U_0(\varepsilon)$ such that 
 \begin{equation}
@@
 \end{equation}
-\ON the material is said to be hyperelastic.
-\OFF However, for \emph{imcompressible} materials the stress tensor is not completely determined by the deformation.
+the material is said to be hyperelastic.
+However, for \emph{imcompressible} materials the stress tensor is not completely determined by the deformation.
 The hydrostatic pressure\footnote{pressure exerted by a fluid at equilibrium due to the force of gravity} affects the stress and the above equation is then written as:
@@
 	\subsection{Viscoelastic materials}
-\ON We know that materials for which the mechanical behaviour depends solely on the state of deformation are called elastic.
+We know that materials for which the mechanical behaviour depends solely on the state of deformation are called elastic.
 In contrast, the mechanical behaviour of \emph{viscous} materials is only a function of the current rate of deformation (like honey for instance).
-\OFF Viscoelastic materials have elements of both of these properties and, as such, exhibit time dependent strain.
+Viscoelastic materials have elements of both of these properties and, as such, exhibit time dependent strain.
 Depending on the change of strain rate versus stress inside a material the viscosity can be categorised as having a linear or non-linear response.
@@
 
-\ON If continuum mechanics does teach us how to model idealised continuous structures, modelling soft tissues remains very complex.
+If continuum mechanics does teach us how to model idealised continuous structures, modelling soft tissues remains very complex.
 Indeed, in practice soft tissues are complex structures: they do not consist of a simple continuous piece of material, they are made of various constituents.
@@
 
-As we know, \OFF the constitutive equations are used to define idealised materials, which represent different aspects of the mechanical behaviour of natural materials.
-\ON However, in practice, \OFF no real material is known to behave exactly as one of these mathematical models.
+As we know, the constitutive equations are used to define idealised materials, which represent different aspects of the mechanical behaviour of natural materials.
+However, in practice, no real material is known to behave exactly as one of these mathematical models.
 In particular, few biological tissues obey Hooke's law.
@@
 These various mechanical aspects of materials make realistic modelling of anatomical structures fairly challenging.
-\ON The development of algorithms that allow the interactive computation of the deformation of soft tissues is very difficult.
-Therefore, \OFF in medical simulation, where modelling must occur in real-time or close to real-time, very often simplifications have to be made \ON and only the key characteristics of the material that are relevant for the objective of the simulation are considered and simulated.
+The development of algorithms that allow the interactive computation of the deformation of soft tissues is very difficult.
+Therefore, in medical simulation, where modelling must occur in real-time or close to real-time, very often simplifications have to be made and only the key characteristics of the material that are relevant for the objective of the simulation are considered and simulated.
 Experiments have to be performed to check the relevancy of the chosen constitutive model and to determine its parameters.
@@
 
-\OFF One of the main obstacles in developing realistic organ models is the lack of data on the material properties of live organ tissues.
+One of the main obstacles in developing realistic organ models is the lack of data on the material properties of live organ tissues.
 Measuring and characterising in vivo organ properties is a highly challenging task, but is a requirement for realistic organ modelling.
@@
 Organ models with incorrect material properties will affect training in surgical simulator systems \citep{Sedef06}.
-\ON Some devices exist for inert materials which allow the measure of Young's modulus, shear modulus and Poisson's ratio.
+Some devices exist for inert materials which allow the measure of Young's modulus, shear modulus and Poisson's ratio.
 However, the same measures are much more difficult to obtain for living tissues.
@@
 \includegraphics[width=12cm]{chapter2/elastography_liver.pdf}
-\caption {\ON Magnetic resonance elastography of the liver in a healthy volunteer and a patient with cirrhosis. Image courtesy of \cite{Talwalkar07}.}
+\caption {Magnetic resonance elastography of the liver in a healthy volunteer and a patient with cirrhosis. Image courtesy of \cite{Talwalkar07}.}
 \label{chap2:fig-elastography_liver}
@@
 The development of a simplified or optimised version of the model suitable for real-time computation is also mandatory.
-\OFF As we will see in chapter \ref{chap4} and \ref{chap7}, numerous models are available in the literature to describe physics of objects, from fairly simple and naive approaches to more complex and thorough representations.
+As we will see in chapter \ref{chap4} and \ref{chap7}, numerous models are available in the literature to describe physics of objects, from fairly simple and naive approaches to more complex and thorough representations.
 However, the most common technique to solve equations of continuum mechanics is the finite element method.

### 2010-10-19 Thesis/Chapter3.tex
- Update 19 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@186 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 The strong form, in contrast to a \emph{weak form}, requires strong continuity on the dependent field variables, such as displacements \citep{Liu03}.
-\ON Indeed, \OFF the functions defining these field variables have to be differentiable up to the order of the partial differential equations that exist in the strong from.
-\ON While the finite element method can be used to find an approximated solution for a strong form,  the method usually works well only with regular geometry and boundary conditions.
+Indeed, the functions defining these field variables have to be differentiable up to the order of the partial differential equations that exist in the strong from.
+While the finite element method can be used to find an approximated solution for a strong form,  the method usually works well only with regular geometry and boundary conditions.
 
-Consequently, \OFF a weak form is often created using energy principles or weighted residual methods.
+Consequently, a weak form is often created using energy principles or weighted residual methods.
 The weak form is frequently an integral form and requires a weaker continuity on the field variables.
-\ON Hence, \OFF a formulation based on a weak form usually produces a set of discretised system equations that give much more accurate results, especially for problems of complex geometry.
-\ON For this reason, \OFF the weak form is usually preferred \ON over the strong form for obtaining an approximate solution of a practical engineering problem.
+Hence, a formulation based on a weak form usually produces a set of discretised system equations that give much more accurate results, especially for problems of complex geometry.
+For this reason, the weak form is usually preferred over the strong form for obtaining an approximate solution of a practical engineering problem.
 \OFF

### 2010-10-19 Thesis/Chapter7.tex
- Update 19 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@186 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\chapter{Modelling the deformation of hollow objects in real-time}
+\chapter[Modelling the deformation of hollow objects in real-time][Modelling hollow objects in real-time]{Modelling the deformation of hollow objects in real-time}
 \label{chap7}
@@
 In this chapter, we will review diverse techniques applied for the simulation of thin objects.
-These different approaches may be grouped into three main categories: (1) models, (2) approaches which derive a bending energy from geometrical considerations and (3) methods based on the equations of continuum mechanics.
+These different approaches may be grouped into three main categories: (1) mass-spring models, (2) approaches which derive a non-physical bending energy and (3) methods based on the equations of continuum mechanics.
 Although very few models have been proposed for simulating the deformation of thin anatomical structures in the field of medical simulation, we will try to emphasize medical applications whenever possible.
@@
 
-\section{Introduction: the problematic}
+\section{Introduction: the problem}
 The human body is composed of various deformable anatomical structures.
@@
 The shape and the internal structure may be very different from one organ to another.
-If you take the liver for instance, it is a solid organ composed of four lobes, each one made up of units called lobules.
+For instance, the liver is a solid organ composed of four lobes, each one made up of units called lobules.
 Each lobule consists of a central vein surrounded by liver cells and constitutes the end of a dense network of blood vessels.
@@
 
-The first idea that may come to mind for modelling thin structures is to apply the same methods than with solid objects.
+An obvious modelling for thin structures is to apply the same methods than with solid objects.
 For instance, \cite{Holzapfel02} proposed a very complex 3-layer model of an aterial wall for simulating an angioplasty procedure.
@@
 Another example is given by \cite{Aloisio04} who assume that the blood flow in the vessels may be considered as incompressible and the artery consequently becomes a solid body and can be modelled as a solid deformable object.
-\TODO{why is this not appropriate?
-High-order elements could do the trick of following the curvature...} Hence, specific models for thin objects were developed.
+If this approach can work, about a dozen of elements is required within the thickness of the structure for the system to be numerically stable \TODO{reference?}.
+Consequently, many elements are needed to model thin structures and real-time performance is difficult to achieve.
+Hence, specific models for thin objects were developed.
 
@@
 Development of a satisfactory physical model that runs in real-time but produces visually convincing animation of thin objects has been a challenge in Computer Graphics, particularly in the area of cloth modelling.
-Rather than resorting to shell theory which involves the most complex formulations in continuum mechanics, most of works have relied on discrete formulations.
+Rather than resorting to shell theory which relies on continuum mechanics, most of works have relied on discrete formulations.
 We will first discuss early approaches based on mass-spring models.
@@
 Other works have considered adding bending through angular springs. 
-For instance, \cite{Taskiran05} have chosen to use a linear representation of angular springs to supply bending and torsion effects in hair modelling, which allowed them to simulate curly hair with good realism.
+%For instance, \cite{Taskiran05} have chosen to use a linear representation of angular springs to supply bending and torsion effects in hair modelling, which allowed them to simulate curly hair with good realism.
 All the forces related to springs were implemented: gravity, repulsions from collision, absorption, friction, etc.
 They could simulate up to a few hundreds of hair stripes in real-time.
-\cite{Wang07} successfully used a network of linear and angular springs to describe bending and twisting of catheters and guidewires in an interventional radiology simulator.
+For instance, \cite{Wang07} successfully used a network of linear and angular springs to describe bending and twisting of catheters and guidewires in an interventional radiology simulator.
 
@@
 Their implementation is also far from being straightforward.
-Nevertheless, we will now review a few attempts of applying shell theory to simulation.
+Nevertheless, we will now review a few examples of the application of shell theory to simulation.
 
@@
 \cite{Thomaszewski06} based their model on the Kirchhoff-Love shell theory and built a subdivision-based finite element formulation similar to \cite{Cirak00}.
-They also employed a co-rotational framework to simulate wrinkles and folds of textiles.
+They also employed a co-rotational framework to simulate wrinkles and folds of textiles (see \fig{chap7:fig-sleeve}).
 Although the rotation field is computed for each vertex, they found that using the rotation obtained for the barycentre of all vertices involved was sufficient.
@@
 \caption[Different types of folds on a garment's sleeve ]{Different types of folds on a garment's sleeve. Images courtesy of \cite{Thomaszewski06}.}
-\label{chap7:fig-sheet}
+\label{chap7:fig-sleeve}
 \end{figure}

### 2010-10-19 Thesis/Chapter9.tex
- Update 19 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@186 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 Similarly to the membrane stiffness matrix, the bending stiffness matrix in the global frame is eventually obtained using the rotation matrix of the element: $\textbf{K} = \textbf{R} \textbf{K}_e \textbf{R}^{T}$.
+While the method to numerically compute the integral will be reviewed section \ref{chap9:numericalIntegration}, we first carry on with the theory by introducing how the contacts are handled.
 
@@
 
-The co-rotational shell formulation has been successfully implemented as a Force Field into the open-source framework SOFA \citep{Allard07}.
+The co-rotational shell formulation has been successfully implemented as a Force Field component into the open-source framework SOFA \citep{Allard07}.
 A description of SOFA's architecture was given in section \ref{chap6:SOFA}.
@@
 From there, our approach to measure the rotations around local x and y axes is simple.
-At runtime and for each vertex, we compute the rotation that we need to transform the z axis of the triangle's frame into the z axis of the vertex's frame (see listing \ref{chap9:qDiffZ}).
+At runtime and for each vertex, we compute the rotation that we need to transform the z axis of the triangle's frame into the z axis of the vertex's frame.
 The differences with the same quantities computed for the rest position during the initialisation lead to the desired displacements.
+% (see listing \ref{chap9:qDiffZ})
+%\begin{lstlisting}[caption={[Method to compute the rotation between two quaternions]Method used to compute the rotation of the z axis between the frames embodied by two quaternions},label={chap9:qDiffZ},frame=shadowbox,rulesepcolor=\color{black}]
+%inline Quat qDiffZ(const Quat& Qvertex, const Quat& Qtriangle)
+%{
+%    // dQ is the quaternion that embodies the rotation between 
+%    // the z axis of the vertex and the z axis of the local 
+%    // triangle's frame (in local space)
+%    Quat dQ;
 %
-\begin{lstlisting}[caption={[Method to compute the rotation between two quaternions]Method used to compute the rotation of the z axis between the frames embodied by two quaternions},label={chap9:qDiffZ},frame=shadowbox,rulesepcolor=\color{black}]
-inline Quat qDiffZ(const Quat& Qvertex, const Quat& Qtriangle)
-{
-    // dQ is the quaternion that embodies the rotation between 
-    // the z axis of the vertex and the z axis of the local 
-    // triangle's frame (in local space)
-    Quat dQ;
-
-    // u = z axis of the triangle's frame (in local space)
-    Vec3d u(0,0,1);
-
-    // v = z axis of the vertex's frame is expressed into 
-    // world space
-    Vec3d v = Qvertex.rotate(Vec3d(0.0, 0.0, 1.0));
-
-    // v0 = v expressed into local triangle's frame
-    Vec3d v0 = Qtriangle.rotate(v);
-
-    // Axis of rotation between the 2 vectors u and v lies 
-    // into the plan of the 2 vectors
-    Vec3d axis = cross(u, v0);
-
-    // Shortest angle between the 2 vectors
-    double angle = acos(dot(u, v0));
-
-    // Quaternion associated to this axis and this angle
-    if (fabs(angle)>1e-6)
-    {
-        dQ.axisToQuat(axis,angle);
-    }
-    else
-    {
-        dQ = Quat(0,0,0,1);
-    }
-
-    return dQ;
-}
-\end{lstlisting}
+%    // u = z axis of the triangle's frame (in local space)
+%    Vec3d u(0,0,1);
+%
+%    // v = z axis of the vertex's frame is expressed into 
+%    // world space
+%    Vec3d v = Qvertex.rotate(Vec3d(0.0, 0.0, 1.0));
+%    
+%    // v0 = v expressed into local triangle's frame
+%    Vec3d v0 = Qtriangle.rotate(v);
+%
+%    // Axis of rotation between the 2 vectors u and v lies 
+%    // into the plan of the 2 vectors
+%    Vec3d axis = cross(u, v0);
+%    
+%    // Shortest angle between the 2 vectors
+%    double angle = acos(dot(u, v0));
+%
+%    // Quaternion associated to this axis and this angle
+%    if (fabs(angle)>1e-6)
+%    {
+%        dQ.axisToQuat(axis,angle);
+%    }
+%    else
+%    {
+%        dQ = Quat(0,0,0,1);
+%    }
+%
+%    return dQ;
+%}
+%\end{lstlisting}
 
@@
 
-	\subsection{Numerical integration}
+	\subsection{Numerical integration} \label{chap9:numericalIntegration}
 In practice, the integration of the integral in \eqref{chap9:Ke} must be carried out numerically using Gaussian quadrature, a numerical technique which allows to calculate the value of an integral.
@@
 Due to linearity, the same relation holds on elementary displacements $dx$.
-It also holds on accelerations, with an additional offset due to velocities when the position mapping \JNL is nonlinear.
+It also holds on accelerations, with an additional offset due to velocities when the position mapping \JNL is non-linear.
 In most cases, operators \JNL~ and \J~ are the same, but in the case of rigid bodies, \JNL~ is nonlinear with respect to $x$ and it can not be written as a matrix.
@@
 applyJ() basically embodies the same computations than apply() except that the vector of displacement $\mathbf{u}$ is different.
-Whereas the local displacement in rotation is measured from the positions to create the vector $\mathbf{u}$, this time the vector is directly obtained from the angular velocities we had at each vertex (once converted into the local frame).
+Whereas the local displacement in rotation is measured from the quaternions to create the vector $\mathbf{u}$, this time the vector is directly obtained from the angular velocities we had at each vertex (once converted into the local frame).
 We therefore have a relation between the angular velocities at each vertex and the velocity of the deflection in the direction normal to each triangle.
@@
 One potential problem however, is the assembly of the global stiffness matrix.
-Indeed, in a parallel implementation, many elements are processed concurrently.
-Inevitably, two elements sharing a node may be processed in the same time and may therefore try to both write their element contribution into the exact same location in the global stiffness matrix, and therefore the same location in GPU's memory.
+Indeed, in a parallel implementation, many elements are processed concurrently (that is, at the same time).
+Inevitably, two elements sharing a node may be processed in the same time.
+Therefore, they may try to both write their element contribution into the exact same location in the global stiffness matrix, and therefore the same location in GPU's memory.
 As with the TLED implementation described in chapter \ref{chap6}, there is two solutions.
@@
 
-\section{Validation}
+\section{Comparison with analytical results}
 
@@
 \end{table}
-
-\bigskip
-
-\TODO{maybe a bit short for a so-called validation...}
 

### 2010-10-22 Thesis/Chapter1.tex
- Update 11 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@184 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 22 October -
Almost done...



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@188 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 Virtually everyone knows this situation where a close relative has to go to the hospital, either for a simple routine check or a serious surgery.
-Hospitals are meant to be reinsuring, a refuge where people's health improves, where people get fixed.
+Hospitals are meant to be re-insuring, a refuge where people's health improves, where people get fixed.
 Obviously, patient safety is a critical concern in the medical industry.
@@
 Dogs, pigs, goats and sheep are commonly used in medical training.
-The exercices essentially consist in anesthetising the animal and practicing surgical techniques.
+The exercises essentially consist in anaesthetising the animal and practicing surgical techniques.
 They may also be severely injured to teach participants how to treat and manage various traumatic injuries.
@@
 But the computer technology was too expensive for commercialisation at the time and only one mannequin was built.
-Over time, computerised mannequins have substantially improved and today such systems are essentially integrated into training centers that aim at recreating the operating room environment.
+Over time, computerised mannequins have substantially improved and today such systems are essentially integrated into training centres that aim at recreating the operating room environment.
 Scenarios can be created to expose medical professionals to diverse clinical situations with the goal of improving the communication between staff members.
@@
 \begin{description}
-\item[Photorealistic rendering:] first, the scene needs to look real.
-A rendered image can be understood in terms of a number of visible features that we want to reproduce in a simulated image: colors, brightness, shadows, reflection, transparency, refraction, diffraction, etc.
+\item[Photo-realistic rendering:] first, the scene needs to look real.
+A rendered image can be understood in terms of a number of visible features that we want to reproduce in a simulated image: colours, brightness, shadows, reflection, transparency, refraction, diffraction, etc.
 Therefore, a series of fairly complex calculations is required for all visible structures within each image (usually displayed at $30\,$Hz).
@@
 Real-time means that one second of simulation corresponds to one second of real life.
+In that regard, a lot of work remains to be done in all these areas of research.
+Along with the increase in computational power, medical simulators endeavour to achieve the highest degree of realism possible.
 Of particular interest to us in this PhD is the accurate modelling of the deformation of anatomical structures in real-time.
 
-\bigskip
-
-Along with the increase in computational power, approaches for modelling anatomical structures became more elaborate and less approximative.
+In the area of biomechanics, approaches for modelling anatomical structures became more elaborate and less approximative.
 These different approaches may be classified into three main categories: (1) geometrically based techniques, (2) approaches physically motivated and (3) methods based on the equations of continuum mechanics.
@@
 As an example, the technique of shape matching proposed by \cite{Muller05} makes use of a simple particle system without particle-particle interactions, but including response to collisions with the environment and including external forces such as gravity.
-But the most popular technique applied in simulation is the mass-spring system \citep{Terzopoulos91,Provot95,Mosegaard05b} .
+But the most popular technique applied in simulation is the mass-spring system \citep{Terzopoulos91,Provot95,Mosegaard05b}.
 In this method, the geometry is described by a network of masses connected together by springs.
@@
 With the continual growth in computational power, finite element formulations applied to medical simulation become more elaborate \citep{Sagar94,Felippa00,Debunne01,Miller07}.
-For instance, they tend to employ more accurate constitutive laws to better describe the complex mechanical behaviour of organs \citep{Fung93}.
+For instance, they tend to employ more accurate constitutive laws to better describe the complex mechanical behaviour of organs.
+
+Yet, few biomechanical models can be computed in real-time and many approximations are still in use in medical simulation.
+Even though the mechanical behaviour of most organs is known to be non-linear, anisotropic and viscoelastic \citep{Fung93}, simple linear models are often employed in medical simulators.
+The resulting deformation of these organs is rather inaccurate and the force feedback provided to the user is not correct.
+Moreover, most of the proposed models are concerned with the mechanical behaviour of solid organs despite the fact that many thin and hollow anatomical structures may be found in the human body.
+To our knowledge, no model based on continuum mechanics was suggested in medical simulation to simulate the deformation of thin structures.
+Therefore, new algorithms are required for both type of structures.
+More elaborate models are suitable to better describe the complex behaviour of solid structures (with greater accuracy) and physics-based models are needed to model thin anatomical structures, all within the context of a real-time environment.
 

### 2010-10-22 Thesis/Chapter10.tex
- Update 11 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@184 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 19 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@186 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 22 October -
Almost done...



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@188 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 In the previous chapter, we proposed a co-rotational formulation for shell elements by extending a classical in-plane triangular finite element approach.
-This simple shell element can efficiently handle both membrane and offplane forces (bending and twist).
+This simple shell element can efficiently handle both membrane and off-plane forces (bending and twist).
 The validity of our approach has been demonstrated though comparison with theoretical results.
 It was then applied it to a rather complex application: the simulation of intra-ocular lens implant deployment in a cataract surgery simulation.
-These preliminary results are very encouraging and show the potential of such models.
+These results are very encouraging and show the potential of such models.
 Examples include the modelling of hollow anatomical structures (stomach, colon, etc.), the simulation of cardiac valve leaflets, and blood vessels.
@@
 In the next section, we propose to study how to approximate the surface of anatomical structures with (curved) shell elements.
-This work was the object of one publication \citep{Comas2010b}.
 \end{shortAbstract}
@@
 Our goal is to create a mesh featuring the optimal number of shell elements while staying as close as possible to our targeted geometry.
-Literature about volumic mesh generation algorithms is fairly dense.
+Literature about volumetric mesh generation algorithms is fairly dense.
 However, there are only a few that are concerned about the generation of meshes over curved surfaces.
@@
 Another method consists of using an a priori error estimator to build an adaptive mesh generation \citep{Baumann97}.
-However, the tolerance of this indicator should be chosen depending on the desired accuracy of the finite element solution, and therefore requires some knowledge about the problem in order to choose an effective tolerance.
+However, the tolerance of this indicator should be chosen depending on the desired accuracy of the finite element solution.
+Therefore it requires some knowledge about the problem in order to choose an effective tolerance.
 \cite{Bechet02} start from an existing triangular mesh created with a CAD (computer-aided design) software and refine and smooth the mesh based on element quality and surface curvature.
@@
 
-However, we need to keep in mind that our situation is substantially different than all previously mentioned geometrical approaches as we want to model the deformation of the structure.
+However, we need to realise that our situation is substantially different than all previously mentioned geometrical approaches as we want to model the deformation of the structure.
 In that regard, the curvature of the surface has a physical meaning: it represents the mid-surface of the shell.
@@
 A simple example is shown \fig{chap10:fig-cylinder} to illustrate the method.
-\begin{figure}
+\begin{figure}[ht]
 \centering
@@
 
-We proposed a framework for real-time modelling of thin anatomical structures.
+Although we were able to mesh complex anatomical structures with a fairly small number of shell elements, our approach could be further improved.
+In our current implementation, if the error measured with the Hausdorff distance is greater than a threshold (which is also difficult to choose), all triangular elements are subdivided into 4 new triangles and each of the new vertices is moved towards the targeted geometry based on our heuristic algorithm.
+Obviously, this is not optimal as the error between the two meshes may very well be local and therefore the use of a global mesh refinement is not very suitable.
+The second weakness is that we also do not check the quality of the elements during our mesh process.
+If some elements are stretched from the heavy decimation, these elements are not corrected during the process of re-meshing.
+Our algorithm could even worsen the situation by splitting those stretched elements into four smaller (and more stretched) elements.
+One solution for the latter issue would be to enhance our algorithm with a relaxation phase where a repulsion force is applied between each vertex (constrained to remain on the surface of the object).
+This physical approach would allow a more even distribution of the vertices across the surface and would contribute to obtain better quality elements overall.
+Yet, meshing a surface with curved elements is a very complex problem and still an active area of research.
+In contrast to all previous geometrical techniques, we suggested a more physical approach consistent with our shell FEM formulation and obtained good results in meshing complex anatomical structures.
+
+\section{Conclusion}
+
+Over the last two chapters, we proposed a complete framework for real-time modelling of thin anatomical structures.
 The novelty of our method relies on the combination of a shell finite element formulation and a geometric surface reconstruction both based on the same polynomial interpolation function used to describe the surface of shells.
@@
 We also presented preliminary results on the coupling between solid (tetrahedra) and thin objects (shells) for the advanced modelling of anatomical structures via the simulation of an angioplasty procedure.
-
-\bigskip
-
-The finite element procedure that we propose is largely developed based on physical understanding without the use of mathematical shell theories.
-Indeed, our shell FEM was obtained by simply superimposing a plane stress and bending energies.
-According to \cite{Chapelle03}, this type of approach has a limited level of accuracy.
-The effective study of shell structures clearly requires a deep physical understanding of shell structural behaviours.
-The complexity of the physical behaviours of shells requires advanced mathematical analyses from theories that are still in development.
-The interested reader may refer to the book of \cite{Chapelle03} for more details on shell theory.
-If the range of application of our shell formulation might be somewhat limited, we did not find out any inconsistencies through our tests.
-We also need to keep in mind that the accuracy required in medical simulation is not the same than the one demanded in structural analyses for instance.
-Moreover, the additional complexity of a shell finite element formulation based on true shell theory may also very well jeopardise our real-time constraints.
-Therefore, we do not think that the use of this simple shell formulation significantly diminishes our contribution to the field of medical simulation.
-Indeed, to our knowledge, our co-rotational shell FEM is the first description of a shell finite element model applied to simulating the deformation of thin anatomical structures.
-
-Nevertheless, a few improvements are conceivable.
-Large strain could be allowed by using a non-linear strain tensor and more complex material laws (non-linear, viscoelasticity, etc.) could also be added to our shell formulation.
-However, those changes are substantial and their implementations are not straightforward.
-In addition, it is not clear how the benefits would be the greatest: whether from enhancing our model to a non-linear formulation or completely changing to another model based on true shell theory (where membrane and bending energies are not separated).
-The answer may depend on the type of constraints applied on the object.
-Also (as already mentioned section \ref{chap9:GPU}), the real-time capacity of our algorithm could be further improved by implementing our shell FEM on GPU.
-Although an optimal parallel re-implementation is always challenging, we have already reviewed why there was no major obstacle to a GPU implementation.
-
-Although we were able to mesh complex anatomical structures with a fairly small number of shell elements, our approach suffers from two major drawbacks.
-In our current implementation, if the error measured with the Hausdorff distance is greater than a threshold (which is also difficult to choose), all triangular elements are subdivised into 4 new triangles and each of the new vertices is moved towards the targeted geometry based on our heuristic algorithm.
-Obviously, this is not optimal as the error between the two meshes may very well be local and therefore the use of a global mesh refinement may be very limited.
-The second drawbrack is that we also do not check the quality of the elements during our mesh process.
-If some elements are stretched from the heavy decimation, these elements are not corrected during the process of remeshing.
-Our algorithm could even worsen the situation by splitting those stretched elements into four smaller (and more stretched) elements.
-One solution for the latter issue would be to enhance our algorithm with a relaxation phase where a repulsion force is applied between each vertex (constrained to remain on the surface of the object).
-This physical approach would allow a more even distribution of the vertices across the surface and would contribute to obtain better quality elements overall.
-Yet, meshing a surface with curved elements is a very complex problem and still an active area of research.
-In contrast to all previous geometrical techniques, we suggested a more physical approach consistent with our shell FEM formulation and obtained good results in meshing complex anatomical structures.
+This work on modelling thin anatomical structures was the object of two main publications \citep{Comas2010a,Comas2010b}.

### 2010-10-22 Thesis/Chapter11.tex
- Update 22 October -
Almost done...



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@188 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 This finite element formulation is fully non-linear: it makes use of a non-linear strain measure (which allows large deformations) and embodies a non-linear constitutive law between strain and stress as characterised by \cite{Fung93} for soft-tissues.
-While the first GPU implementation of the TLED was carried out by \cite{Taylor07b} in the Cg language, our re-implementation in CUDA led to a threefold increase in performance.
-Overall, we observe a maximum speed improvement for GPU solution over CPU solution of up $56.3 \times$.
+While the first GPU implementation of the TLED was carried out by \cite{Taylor07b} in the Cg language, our efficient re-implementation in CUDA led to a threefold increase in performance.
+Overall, we observe a maximum speed improvement for GPU solution over CPU solution of up to $56.3 \times$.
 My collaboration with Zeike Taylor continued and we enhanced the formulation by adding anisotropy and viscoelasticity.
@@
 However, the main drawback of our method is the lack of any assessment of the quality of the elements created.
-Indeed, if some elements are stretched from the heavy decimation, these elements are not corrected during the process of remeshing.
+Indeed, if some elements are stretched from the heavy decimation, these elements are not corrected during the process of re-meshing.
 Our algorithm could even worsen the situation by splitting those stretched elements into four smaller (and more stretched) elements.
@@
 One of the strongest constraint in medical simulation is the demand for real-time computations.
-This constraint is what makes all tasks in a simulator so challenging (collision detection, contact and soft tissue modelling, haptic feedback, photorealistic rendering etc.).
+This constraint is what makes all tasks in a simulator so challenging (collision detection, contact and soft tissue modelling, haptic feedback, photo-realistic rendering etc.).
 In fact, this is at the origin of the multitude of simplifications proposed in simulation for modelling the biomechanics of anatomical structures.
@@
 Secondly, GPU programming is not simple to master.
-Although we have seen many improvements in flexibility of programmation over the last 3 or 4 years, a strong understanding of the underlying GPU hardware is still mandatory to create an optimal code.
+Although we have seen many improvements in flexibility of programming over the last 3 or 4 years, a strong understanding of the underlying GPU hardware is still mandatory to create an optimal code.
 As an example, a GPU offers different types of memory, all with distinct pros and cons (strict access patterns, latency, limited size etc.), and choosing the best type of memory for each variable may be rather challenging to say the least.

### 2010-10-22 Thesis/Chapter2.tex
- Update 22 October -
Almost done...



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@188 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 A rigid-body displacement consists of a simultaneous translation and rotation of the body without changing its shape or size.
-In constrast, a change in shape and\slash or size of the body from an initial configuration $\kappa_{0}$ to a new configuration $\kappa$ is called a deformation.
+In contrast, a change in shape and\slash or size of the body from an initial configuration $\kappa_{0}$ to a new configuration $\kappa$ is called a deformation.
 This new configuration $\kappa$ may be designated by the \emph{current} or \emph{deformed configuration}.
@@
 	\subsection{Deformation gradient tensor}
-The displacement field tells us how a particule displaces from the reference to the deformed configuration.
+The displacement field tells us how a particle displaces from the reference to the deformed configuration.
 More interestingly, we would like to know how a material line would deform (stretch and rotation) within this displacement field.
@@
 This relationship, which obviously depends on the type of material under consideration, is what we call a \emph{constitutive equation}.
-Constitutives equations are an attempt to mathematically characterise the behaviour of materials observed from experimentals results.
+Constitutives equations are an attempt to mathematically characterise the behaviour of materials observed from experimental results.
 \OFF
@@
 the material is said to be hyperelastic.
-However, for \emph{imcompressible} materials the stress tensor is not completely determined by the deformation.
+However, for \emph{incompressible} materials the stress tensor is not completely determined by the deformation.
 The hydrostatic pressure\footnote{pressure exerted by a fluid at equilibrium due to the force of gravity} affects the stress and the above equation is then written as:
@@
 Most often, the material properties are determined in a laboratory in terms of the engineering constants such as Young's modulus, shear modulus, and so on.
-These constants are measured using simple tests like uniaxial tension test or pure shear test.
+These constants are measured using simple tests like uni-axial tension test or pure shear test.
 Because of their direct and obvious physical meaning, engineering constants are used in place of the more abstract stiffness coefficients $C_{ijkl}$.
@@
 The strain energy density function takes different forms for different materials.
-It is often expressed as a linar combination of unknown coefficients (determined experimentally) and principal invariants of Green strain tensor $\mathbf{E}$, left Cauchy-Green strain tensor $\mathbf{B}$ and deformation gradient tensor $\mathbf{F}$ (see section \ref{chap2:strainInvariants} for more information on invariants).
+It is often expressed as a linear combination of unknown coefficients (determined experimentally) and principal invariants of Green strain tensor $\mathbf{E}$, left Cauchy-Green strain tensor $\mathbf{B}$ and deformation gradient tensor $\mathbf{F}$ (see section \ref{chap2:strainInvariants} for more information on invariants).
 The constitutive equations of  two well-known non-linear elastic materials (Mooney-Rivlin and neo-Hookean) will now be introduced.
@@
 Mechanical properties of soft tissues not only depends on the properties of these constituents, but how they are arranged relative to each other.
-This elaborate arrangement often leads to material properties for the global structure that are non-homogenous, non-linear, anisotropic and even viscoelastic.
+This elaborate arrangement often leads to material properties for the global structure that are non-homogeneous, non-linear, anisotropic and even viscoelastic.
 

### 2010-10-22 Thesis/Chapter3.tex
- Update 22 October -
Almost done...



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@188 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 Usually, the number of elements increases with the complexity of the geometry.
-For instance, if one part of the domain is thiner (see \fig{chap3:fig-errorWithSubdivision} (c)), the size of the elements must be reduced in order to tile this particular part, which increases the total number of elements, and hence the number of algebraic equations to solve.
+For instance, if one part of the domain is thinner (see \fig{chap3:fig-errorWithSubdivision} (c)), the size of the elements must be reduced in order to tile this particular part, which increases the total number of elements, and hence the number of algebraic equations to solve.
 However, adding elements is sometimes desirable.
@@
 \noindent \textbf{Remark.}
-It is worth noting that \eqref{chap3:shapeFunctions} can be easily extented to three-dimensional problems.
+It is worth noting that \eqref{chap3:shapeFunctions} can be easily extended to three-dimensional problems.
 Let us consider a Cartesian coordinate system and we denote the position of node $ i $ $ \mathbf{x}_i = [x_{i1}, x_{i2}, x_{i3}]^T$.
@@
 Note that the shape functions describe the geometry and the solution with the same degree of interpolation.
-Because isoparametric elements are very common in practice, the use of the phrase \emph{shape functions} is often extented to denote the interpolation functions employed to approximate the solution.
+Because isoparametric elements are very common in practice, the use of the phrase \emph{shape functions} is often extended to denote the interpolation functions employed to approximate the solution.
 
@@
 For some problem, this is not an issue, the motion is sufficiently slow to assume that the dynamic effects (such as damping effect) are neglectable.
-However, this assumption is not always valid, particulary in our case where we want to model organ deformation.
+However, this assumption is not always valid, particularly in our case where we want to model organ deformation.
 The deformation obviously depends on time.
@@
 We begin by introducing the simplest method, the Euler method to explain the concept of an explicit method.
-Then the central difference method will be detailled.
+Then the central difference method will be detailed.
 We then describe a few implicit methods, allowing us to highlight the relative advantages of the two types of methods.
@@
 \ON
-\paragraph*{Preconditioners}
+\paragraph*{Pre-conditioners}
 Unfortunately, it is common for the matrix $ \mathbf{A} $ not to be well conditioned.
 Consequently, the convergence of the chosen iterative method is often slow.
-In order to achieve an accelerated convergence, we may use a \emph{preconditioner}.
-Although the formation and the study of preconditioners are out of the scope of this PhD thesis, preconditioners are a key element for solving systems of equations.
+In order to achieve an accelerated convergence, we may use a \emph{pre conditioner}.
+Although the formation and the study of pre-conditioners are out of the scope of this PhD thesis, pre-conditioners are a key element for solving systems of equations.
 To be effective, a preconditioner $ \mathbf{P} $ should be a good approximation of  $ \mathbf{A} $ and the following preconditioned system must be easier to solve than the original system \eqref{chap3:linearEquation}:
@@
 Two criteria should also be considered when forming a preconditioner: (1) the ease with which it may be evaluated and (2) its memory requirements.
-The easiest preconditioner would be the identity since $ \mathbf{P}^{-1} = \mathbf{I} $ in this case.
+The easiest pre-conditioner would be the identity since $ \mathbf{P}^{-1} = \mathbf{I} $ in this case.
 Of course, such a conditioner leads to the resolution of the original system and is therefore totally useless.
 In constrast, the choice of $ \mathbf{P} = \mathbf{A} $ leads to the solution of the system \eqref{chap3:linearEquation} in one single iteration.
-In practice, an optimal preconditioner is to be found between these two extremes to minimise the number of iterations required by the solver while keeping the evaluation of the preconditioner simple.
-Finding a good preconditioner for all types of problems is difficult.
-A preconditioner is only useful if the overall time spent on both the formation of  $ \mathbf{P} $ and the resolution of \eqref{chap3:preconditionedSystem} is less than the time spent at solving \eqref{chap3:linearEquation} alone.
+In practice, an optimal pre-conditioner is to be found between these two extremes to minimise the number of iterations required by the solver while keeping the evaluation of the pre-conditioner simple.
+Finding a good pre-conditioner for all types of problems is difficult.
+A pre-conditioner is only useful if the overall time spent on both the formation of  $ \mathbf{P} $ and the resolution of \eqref{chap3:preconditionedSystem} is less than the time spent at solving \eqref{chap3:linearEquation} alone.
 
-One of the simplest preconditioner is the \emph{Jacobi preconditioner} where the preconditioner is chosen to be the diagonal of the matrix $ \mathbf{A} $.
+One of the simplest pre-conditioner is the \emph{Jacobi pre-conditioner} where the pre-conditioner is chosen to be the diagonal of the matrix $ \mathbf{A} $.
 This choice is efficient in cases where $ \mathbf{A} $ is diagonally dominant.
-Sparse approximations of LU or Cholesky decompositions may also be used as preconditioners (see sections \ref{chap3:LUdecompo} and \ref{chap3:CholeskyDecompo} for more details).
+Sparse approximations of LU or Cholesky decompositions may also be used as pre-conditioners (see sections \ref{chap3:LUdecompo} and \ref{chap3:CholeskyDecompo} for more details).
 \OFF

### 2010-10-22 Thesis/Chapter4.tex
- Update 19 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@186 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 22 October -
Almost done...



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@188 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\chapter{Modelling the deformation of solid objects in real-time}
+\chapter[Modelling the deformation of solid objects in real-time][Modelling solid objects in real-time]{Modelling the deformation of solid objects in real-time}
 \label{chap4}
@@
 
-\section{Introduction: the problematic}
+\section{Introduction: the problem}
 
@@
 
-\ON
 The challenges faced by the field of medical simulation were reviewed in the introductory chapter (see chapter \ref{chap1}).
@@
 The best mathematical model for predicting the deformation of a continuous structure is provided by continuum mechanics.
-If the finite element method is classicaly used in mechanical engineering for its ability to solve the equations of continuum mechanics fairly accurately, its development in medical simulation was more progressive.
+If the finite element method is classically used in mechanical engineering for its ability to solve the equations of continuum mechanics fairly accurately, its development in medical simulation was more progressive.
 The main reason for this is that the finite element method is computationally very demanding and not easily compatible with the time constraints required in interactive simulation.
@@
 For instance, the correctness of deformation is crucial for patient-specific planning and per-operative guidance.
-Let us give an example where simulation can lead the surgeon towards its goal during surgery: the removal of a brain tumor.
-Prior surgery, images of the patient's brain were acquired and the tumor was localised.
+Let us give an example where simulation can lead the surgeon towards its goal during surgery: the removal of a brain tumour.
+Prior surgery, images of the patient's brain were acquired and the tumour was localised.
 To access the brain during surgery, the surgeon must first drill a hole into the patient's skull.
 However, drilling a hole releases intra-cranial pressure and this change in internal pressure deforms the brain.
-\OFF This phenomenon is called \emph{brain shift}.
-The overall deformation is quite complex and the displacements depend on local elasticity of the brain, the size and the location of the tumor.
-\ON Removing the tumor relies \OFF heavily on knowing the spatial relation between the patient's brain and the images acquired prior to surgery.
-\ON And because the brain deforms during surgery, the location and the shape of the tumor will no longer match those of the images acquired.
-\OFF Therefore, a large brain shift, if not corrected, will result in inaccuracies in the surgical procedure and has the potential to cause damage to normal tissue.
-A solution is to accurately model the mechanical response of the patient's brain to predict the actual deformation and hence keep the knowledge of the tumor's location.
-\ON If the accuracy is obviously of prime importance in this situation, computation of the deformation does not necessarily need to be carried out in real-time, a reasonable delay is acceptable (a couple of minutes).
-\OFF
+This phenomenon is called \emph{brain shift}.
+The overall deformation is quite complex and the displacements depend on local elasticity of the brain, the size and the location of the tumour.
+Removing the tumour relies heavily on knowing the spatial relation between the patient's brain and the images acquired prior to surgery.
+And because the brain deforms during surgery, the location and the shape of the tumour will no longer match those of the images acquired.
+Therefore, a large brain shift, if not corrected, will result in inaccuracies in the surgical procedure and has the potential to cause damage to normal tissue.
+A solution is to accurately model the mechanical response of the patient's brain to predict the actual deformation and hence keep the knowledge of the tumour's location.
+If the accuracy is obviously of prime importance in this situation, computation of the deformation does not necessarily need to be carried out in real-time, a reasonable delay is acceptable (a couple of minutes).
 
@@
 
-Conversely, when medical simulation is applied to training, such a precision in terms of \ON deformation is not always \OFF required.
-\ON A visually realistic deformation is often sufficient for an inexperienced physician to learn the appropriate gestures required for a given medical procedure.
+Conversely, when medical simulation is applied to training, such a precision in terms of deformation is not always required.
+A visually realistic deformation is often sufficient for an inexperienced physician to learn the appropriate gestures required for a given medical procedure.
 However, in some cases, the feel of touch can be very important to master the procedure.
-\OFF For instance, during a colonoscopy procedure, force feedbacks give precious information about what type of loop is forming \ON during the colonoscope's insertion \OFF and the physician can then react accordingly to prevent loop formations.
+For instance, during a colonoscopy procedure, force feedbacks give precious information about what type of loop is forming during the colonoscope's insertion and the physician can then react accordingly to prevent loop formations.
 Consequently, a good colonoscopy simulator is required to provide accurate force feedbacks.
@@
 
-Coquillard later developed the idea of Extented free-form deformation \citeyearpar{Coquillart90} and Animated free-form technique \citeyearpar{Coquillart91}.
+Coquillard later developed the idea of extended free-form deformation \citep{Coquillart90} and animated free-form technique \citep{Coquillart91}.
 The idea of free-form deformation with lattices of arbitrary topology was introduced by \cite{MacCracken96}.
@@
 	\subsection{Shape matching}
-One field particulary interested in modelling deformable objects in a very efficient way is the game industry.
+One field particularly interested in modelling deformable objects in a very efficient way is the game industry.
 Of course, for games we are more attracted by computational efficiency and extreme stability features than accuracy to physics laws.
@@
 \cite{Muller05} developed a technique with this idea in mind called \emph{shape matching}.
-We start with a set of particules with masses $ m_i $ in an initial configuration where we denote by $ \mathbf{x}_i^0 $ the initial positions of the particules.
+We start with a set of particles with masses $ m_i $ in an initial configuration where we denote by $ \mathbf{x}_i^0 $ the initial positions of the particles.
 No connectivity information is required.
@@
 
-Later, \cite{Rivers07} extented the technique to regular lattices with embedded geometry.
+Later, \cite{Rivers07} extended the technique to regular lattices with embedded geometry.
 From a surface mesh to be deformed, they start by voxelising the model to construct a lattice of cubic cells containing the mesh.
 The embedded mesh is then deformed using trilinear interpolation of lattice vertex positions.
-A particule is placed at each vertex of the lattice and associated with a shape matching region.
+A particle is placed at each vertex of the lattice and associated with a shape matching region.
 Their technique of \emph{fast lattice shape matching} allows them to compute per-region transformations in an efficient manner (see \fig{chap4:fig-penguins}).
@@
 
-This method was extented by \cite{Schill98} to integrate inhomogeneous and anisotropic behaviour.
-This extension of the chain mail algorithm allows the authors to simulate the vitrous humor in the eye during a vitrectomy procedure.
+This method was extended by \cite{Schill98} to integrate inhomogeneous and anisotropic behaviour.
+This extension of the chain mail algorithm allows the authors to simulate the vitreous humour in the eye during a vitrectomy procedure.
 However, the main drawback of this algorithm is that it does not model volume preservation, an important characteristic of many tissues in the human body.
@@
 In particular, they use modal analysis to model human skin and secondary tissues in a laparascopic surgical simulation to allow the CPU to focus on more complex tissue models and user contact interactions (see \fig{chap4:fig-modalAnalysisLaparo}).
-The technique is fairly efficient as it allows precomputations of the vibration modes.
+The technique is fairly efficient as it allows pre-computations of the vibration modes.
 %
@@
 
-Although modal analysis significantly accelerates the simulation, it generates noticeable artifacts when applied to large deformations due to the step of linearisation.
+Although modal analysis significantly accelerates the simulation, it generates noticeable artefacts when applied to large deformations due to the step of linearisation.
 It can produce unnatural results with deformations of large amplitude.
@@
 From our experience, we know that a mass attached to a single spring has a tendency to oscillate.
-If no appropriate measure is taken, a network of springs will a forciori oscillate as well.
+If no appropriate measure is taken, a network of springs will a fortiori oscillate as well.
 To prevent the system from oscillating, a damper is also added between each couple of masses.
@@
 For instance, it was used by \cite{Miller88} to animate the motion of snakes and worms.
-It was also used by \cite{Ng-Thow-Hing97} in their muscle model where some of the springs were aligned with the muscle fibers and more recently by \cite{Magjarevic09} for modelling the mechanical behavior of the myocardial tissue.
+It was also used by \cite{Ng-Thow-Hing97} in their muscle model where some of the springs were aligned with the muscle fibers and more recently by \cite{Magjarevic09} for modelling the mechanical behaviour of the myocardial tissue.
 More details on controlling anisotropy in mass-spring systems may be found in \cite{Bourguignon00}.
@@
 Much more recently, \cite{Tamura05} introduced a method to simulate viscoelastic material using a mass-spring system.
-They use a large number of particules to create a randomly connected mesh to mimic the structure of polymeric materials and hence their viscoelascity characteristics.
+They use a large number of particles to create a randomly connected mesh to mimic the structure of polymeric materials and hence their viscoelascity characteristics.
 \cite{Basafa10} went even further and extended a mass-spring system to simulate non-linear viscoelastic deformations of soft tissue for laparoscopic surgery.
@@
 With this idea in mind, \cite{Sorensen06} reviewed the most important concepts and data structures required to realise two popular deformable models on the GPU: the spring-mass model and the finite element model.
-\ON In 2008, \citeauthor{Rasmusson08} investigated performance and optimisation strategies between the two GPU languages OpenGL and CUDA for their mass-spring-damper model.
+In 2008, \citeauthor{Rasmusson08} investigated performance and optimisation strategies between the two GPU languages OpenGL and CUDA for their mass-spring-damper model.
 More recently, \cite{Leon10} presented a GPU implementation of a mass-spring model with a nearly $80\times$ speedup over serial execution and $20\times$ speedup over CPU-based parallel execution.
-\OFF
 
@@
 The deformation is solved only at surface points while still taking the volumetric nature of the object into account.
-The second improvement concerns the fact that they explicitly precompute the inverse of the stiffness matrix and use matrix vector multiplication with this matrix to achieve a low calculation time.
+The second improvement concerns the fact that they explicitly pre-compute the inverse of the stiffness matrix and use matrix vector multiplication with this matrix to achieve a low calculation time.
 Finally, they exploit the sparse structure of the force vector.
@@
 In addition, they enhanced the formulation by adding a non-linear radial component.
-This method was applied to compute the deformation of liver in a hepatic surgery simulator.
+This method was applied to compute the deformation of liver in a hepatic surgery simulator (see \fig{chap4:fig-superposition}).
 Their approach substantially improved the computational time for a finite element model as it allowed them to compute the positions and the forces of a $1\,500$ node liver at $300\,$Hz.
 An important drawback emphasised by the authors themselves is the impossibility to simulate tissue cutting.
-Indeed, the large precomputations associated with this method are dependent of the geometry of the mesh and therefore cannot be updated in real-time.
-
-\bigskip
+Indeed, the large pre-computations associated with this method are dependent of the geometry of the mesh and therefore cannot be updated in real-time.
+%
+\begin{figure}[ht]
+\centering 
+\includegraphics[width=13.5cm]{chapter4/superposition.pdf}
+\caption[Hepatic surgery simulator]{Images of a hepatic surgery simulator.
+The liver model contains about $ 1\,500 $ nodes.
+The simulation was performed at $ 50\,$Hz for the visual feedback and $ 300\,$Hz for force feedback.
+Image courtesy of \cite{Cotin99}.}
+\label{chap4:fig-superposition}
+\end{figure}
 
@@
 The efficiency of this hybrid model was demonstrated by simulating an hepatectomy (removing of one of the eight anatomical segments of the liver).
-They reported the simulation (allowing deformation and cutting) of a mesh with more than $ 8\,000 $ tetrahedra in real-time (see \fig{chap4:fig-hybridModel}).
-%
-\begin{figure}[ht]
-\centering 
-\subfloat[ ]{\includegraphics[height=4.5cm]{chapter4/hybrid1.pdf}}
-\hspace{0.2cm}
-\subfloat[ ]{\includegraphics[height=4.5cm]{chapter4/hybrid2.pdf}}
-\caption[Simulation of a hepatectomy using a hybrid model]{Simulation of a hepatectomy using a hybrid model.
+They reported the simulation (allowing deformation and cutting) of a mesh with more than $ 8\,000 $ tetrahedra in real-time.
+% (see \fig{chap4:fig-hybridModel})
+%\begin{figure}[ht]
+%\centering 
+%\subfloat[ ]{\includegraphics[height=4.5cm]{chapter4/hybrid1.pdf}}
+%\hspace{0.2cm}
+%\subfloat[ ]{\includegraphics[height=4.5cm]{chapter4/hybrid2.pdf}}
+%\caption[Simulation of a hepatectomy using a hybrid model]{Simulation of a hepatectomy using a hybrid model.
 $ 18 $\% of the mesh was modelled with the tensor-mass model (green) and therefore could be cut out.
 The rest of the mesh was simulated with the precomputed finite element model.
-Image courtesy of \cite{Cotin00}.
-\TODO{Stéphane> tu dis que l'image est moche, mais c'est pourtant bien celle présente dans le papier...
-Et même dans ton HDR tu utilises cette image...
-regarde page 46 et 50...
-Mine de rien ça fait plus de 10 ans ;-) }}
-\label{chap4:fig-hybridModel}
-\end{figure}
+Image courtesy of \cite{Cotin00}.}
+%\label{chap4:fig-hybridModel}
+%\end{figure}
 
@@
 
-Cutting is a problem for all techniques precomputing the inverse of the stiffness matrix.
+Cutting is a problem for all techniques pre-computing the inverse of the stiffness matrix.
 Indeed, topological changes require updates of this matrix, which is too computationally heavy to be done in real-time.
@@
 In addition, repositioning the nodes within the mesh can easily generate degeneracies, which have to be eliminated.
-\TODO{add PhD of Brian Lee...
-Mais je l'ai pas sa thèse...
-et tu m'as pas dit que tu étais en train de la reviewer ?
-Du coup c'est pas encore publique, c'est difficile de la citer, non ?}
+In his PhD thesis, \cite{Lee10} presented techniques to update the precomputed inverse of the stiffness matrix rather than recomputing it, which is much more efficient.
+The proposed technique was also extended to integrate the concept of condensation, increasing computational efficiency and reducing memory requirements.
 %
 \begin{figure}[ht]
-\begin{center}
-\includegraphics[width=13cm]{chapter4/cutUnderScalpel.pdf}
-\end{center}
+\centering 
+\subfloat[Surface selection]{\includegraphics[height=3cm]{chapter4/cutUnderScalpel1.pdf}}
+\hspace{0.2cm}
+\subfloat[Node snapping]{\includegraphics[height=3cm]{chapter4/cutUnderScalpel2.pdf}}
+\hspace{0.2cm}
+\subfloat[Dissect]{\includegraphics[height=3cm]{chapter4/cutUnderScalpel3.pdf}}
 \caption[The three steps in performing a cut]{The three steps in performing a cut (represented here in 2D). Image courtesy of \cite{Nienhuys01}.} 
@@
 In particular, the error brought by the approximated (linear) strain measure increases with the rotation of the element.
-This error in the measure creates ghost forces which overdeforms the simulated object (see \fig{chap4:fig-cylinder}).
+This error in the measure creates ghost forces which over-deforms the simulated object (see \fig{chap4:fig-cylinder}).
 The idea behind co-rotational methods is the decomposition of the motion into a rigid-body and deformational components.
@@
 Although the birth of co-rotational methods is not very clear \citep{Felippa05}, \cite{Belytschko73} had a similar thinking and introduced the so-called \emph{convected coordinates} for solving dynamic equations.
-However, while the spirit is the same than co-rotational methods, convected coordinates forms a curvilinear system that fits the change of metric as the body deforms, therefore the convected metric necessarily emcompasses deformations.
+However, while the spirit is the same than co-rotational methods, convected coordinates forms a curvilinear system that fits the change of metric as the body deforms, therefore the convected metric necessarily encompasses deformations.
 In other words, the decomposition in rigid-body and deformational components is not exact.
@@
 
-In constrast of \cite{Muller02} who discretised the rotation field at each vertex, \cite{Hauth04} determined the rotation at the level of the element.
+In contrast of \cite{Muller02} who discretised the rotation field at each vertex, \cite{Hauth04} determined the rotation at the level of the element.
 If \citeauthor{Muller02}'s technique of stiffness warping yields ghost forces due to inaccuracies in the determination of the rotation field, \citeauthor{Hauth04} made use of the polar decomposition of the deformation gradient to extract the exact rotation.
@@
 The first geometrically non-linear finite element model running in real-time was reported by \cite{Debunne01}.
-They provided an adaptative method for animating viscoelastic deformable objects in a guaranteed frame rate.
+They provided an adaptive method for animating viscoelastic deformable objects in a guaranteed frame rate.
 As the object moves and deforms, the sampling is refined to concentrate the computational load into the regions that deform the most (see \fig{chap4:fig-gummyBear}).
@@
 
-In the same spirit, \cite{Wu01} applied an adaptative meshing technique to provide sufficient detail where required while minimising unnecessary computation.
+In the same spirit, \cite{Wu01} applied an adaptive meshing technique to provide sufficient detail where required while minimising unnecessary computation.
 They included both Mooney-Rivlin and Neo-Hookean material models and used the non-linear Cauchy-Green tensor.
@@
 The algorithm is based on the finite element method using the total Lagrangian formulation, where stresses and strains are measured with respect to the original configuration.
-This choice allows for precomputing of most spatial derivatives before the commencement of the time-stepping procedure.
+This choice allows for pre-computing of most spatial derivatives before the commencement of the time-stepping procedure.
 The authors reported that the average number of floating-point operations per element per time step is $35\, $\% lower than for the similar implementation of the algorithm based on updated Lagrangian formulation.
@@
 In 2007, \citeauthor{Taylor07b} introduced the first GPU implementation of a fully non-linear finite element method (more details in \cite{Taylor08}).
-They reformulated the TLED algorithm presented by \cite{Miller07} to accomodate with the limited hardware possibilities of GPUs.
+They reformulated the TLED algorithm presented by \cite{Miller07} to accommodate with the limited hardware possibilities of GPUs.
 In particular, the main restriction when using graphics-based GPU execution is the inability to scatter, which necessitated reformulation of force summation as a gather in their implementation.
@@
 A point collocation technique is used to generate the discrete equations.
-The authors qualitatively compared the method of finite spheres with the finite element technique from a commercial package and the displacement profil was similar in both cases.
+The authors qualitatively compared the method of finite spheres with the finite element technique from a commercial package and the displacement profile was similar in both cases.
 They were able to achieve a computational rate of about $ 100\, $Hz when 34 spheres where used.
@@
 
-Another example of application is the used of the technique called moving particule semi-implicit method (MPS) by \cite{Chhatkuli09}.
-They applied this method to simulate the deformation of lung and a tumor inside the left lung during inspiration.
+Another example of application is the used of the technique called moving particle semi-implicit method (MPS) by \cite{Chhatkuli09}.
+They applied this method to simulate the deformation of lung and a tumour inside the left lung during inspiration.
 The lung tissues were considered to be homogeneous, isotropic and viscoelastic.
@@
 
-\ON Recently, \cite{Zhu10} proposed a framework based on point-based simulation techniques to model organ deformations realistically in a laparoscopic simulator.
-They combined a new strain-stress tensor computation scheme with smoothed particule hydrodynamics (SPH) method.
+Recently, \cite{Zhu10} proposed a framework based on point-based simulation techniques to model organ deformations realistically in a laparoscopic simulator.
+They combined a new strain-stress tensor computation scheme with smoothed particle hydrodynamics (SPH) method.
 The parameters of the tensor relating strain and stress were determined through mechanical experiments.
-\OFF
 
@@
 
-\ON The main advantage of meshless methods is of course in the absence of a mesh.
+The main advantage of meshless methods is of course in the absence of a mesh.
 This advantage of not having to go through the complex process of meshing was often claimed by meshless methods partisans.
@@
 As a conclusion, if meshless methods are promising, the technique does not seem as mature as the finite element method for modelling soft tissues.
-\OFF

### 2010-10-22 Thesis/Chapter5.tex
- Update 22 October -
Almost done...



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@188 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 In 2007, \citeauthor{Miller07} proposed the total Lagrangian explicit dynamics (TLED) algorithm, a very efficient fully non-linear formulation.
-The authors reported that the average number of floating-point operations per element per time step is $35\, $\% lower than for the similar implementation of the algorithm based on updated Lagrangian formulation.
-Their work constituted a step towards the simulation of entire organs in real-time.
-Later in the year, \citeauthor{Taylor07b} reformulated the TLED algorithm to propose the first GPU implementation of a fully non-linear finite element method.
-However, restrictions due to using a graphics-based API necessitated reformulation of the force summation in their implementation.
-\cite{Taylor07a} also presented an efficient constitutive update procedure for viscoelastic models formulated for use in explicit dynamic finite element algorithms like the TLED algorithm.
-With the release by NVIDIA of a new graphics card architecture along with a new and more flexible non-graphics API specially designed for general-purpose GPU, we decided to investigate the re-implementation of the TLED for this new architecture with the idea of a more straightforward and efficient implementation in mind.
 We also wanted to extend the simple non-linear TLED formulation by adding the constitutive update procedure for viscoelastic models introduced by \citeauthor{Taylor07a} and an anisotropic constitutive formulation as well.
-This work eventually yields a very efficient GPU-based non-linear, viscoelastic and anisotropic formulation for modelling solid organs in medical simulation and was the object of three publications \citep{Comas2008,Taylor2008,Taylor2009}.
+A crucial advantage of this enhanced TLED algorithm is that the computations are conducted at the element level and therefore the algorithm can be easily parallelised.
+This work eventually yields a very efficient GPU-based non-linear, viscoelastic and anisotropic formulation for modelling solid organs in medical simulation.
 This chapter aims at describing the TLED algorithm in details, as well as introducing the viscoelasticity and anisotropy extensions.
@@
 
-The first key idea of the TLED algorithm is to refer all variables to the underformed configuration.
+The first key idea of the TLED algorithm is to refer all variables to the undeformed configuration.
 This type of formulation is said to be \emph{total Lagrangian}.
@@
 As we have seen in section \ref{chap2:elasticity}, hyperelastic materials are a general class of materials in which the constitutive relationship is expressed in the form of a strain energy density function $ _0^t W $.
-In such materials, the stresses are obtained by differentiating this energy density function with respect to the appropriate strain measure (that is energically conjugate).
+In such materials, the stresses are obtained by differentiating this energy density function with respect to the appropriate strain measure (that is energetically conjugate).
 In a total Lagrangian framework, the appropriate stress and strain measures are second Piola-Kirchhoff stress $ _0^t \mathbf{S} $ and Green-Lagrange strain $ _0^t \mathbf{E} $.
@@
 where $E$ is Young's modulus and $\nu$ is Poisson's ratio.
-If nonlinear analyses (either geometric or material) are conducted, a somewhat smaller time step must be employed since $c$ will change with deformation, and in particular will generally increase.
+If non-linear analyses (either geometric or material) are conducted, a somewhat smaller time step must be employed since $c$ will change with deformation, and in particular will generally increase.
 Additionally, the presence of damping necessitates a further decrease in the limit.
@@
 Transversely isotropic or orthotropic models may be produced by substituting \eqref{chap5:PhiBar} for \eqref{chap5:PhiBarTI} or \eqref{chap5:PhiBarOrtho}, respectively.
-The specification of particular forms of  $ \Psi^{vol} $ and $ \Psi^{iso} $ would be motivated by the particular tissue/material under analysis, and may stem from phenomenological or microstructural considerations.
+The specification of particular forms of  $ \Psi^{vol} $ and $ \Psi^{iso} $ would be motivated by the particular tissue/material under analysis, and may stem from phenomenological or micro-structural considerations.
 Finally we note that for separated isochoric and volumetric hyperelastic functions as used here, viscoelastic terms may be applied to either or both independently.
@@
 Assuming a constant time step size $ \Delta t $, the coefficients $ A $ and $ B $ may be precomputed.
+
+\section{Conclusion}
+
+The main advantage of the TLED is that the average number of floating-point operations per element per time step is $35\, $\% lower than for the similar implementation of the algorithm based on updated Lagrangian formulation \citep{Miller07}.
+Coupled with the use of an explicit integration scheme that eliminates the need for an iterative equation solving, the work of \cite{Miller07} constituted a step towards the simulation of entire organs in real-time.
+However, explicit analyses impose a critical time step for the simulation to remain stable (see detailed discussion about the critical time step in section~\ref{chap6:criticalTimestep} page~\pageref{chap6:criticalTimestep}).
+
+A constitutive update procedure for anisotropic and visco-hyperelastic materials has also been presented.
+The procedure allows the incorporation of these elaborate mechanical properties with only a small additional computational cost.
+
+A crucial benefit of this enhanced TLED algorithm is that the computations are conducted at the element level and therefore the algorithm can be easily parallelised.
+Hence, in 2007 \citeauthor{Taylor07b} re-formulated the TLED algorithm to propose the first GPU implementation of a fully non-linear finite element method.
+However, restrictions due to using a graphics-based API necessitated re-formulation of the force summation in their implementation.
+With the release by NVIDIA of a new graphics card architecture along with a new and more flexible non-graphics API specially designed for general-purpose GPU, we decided to investigate the re-implementation of the TLED for this new architecture with the idea of a more straightforward and efficient implementation in mind.
+We also improved the model by including the anisotropic and viscoelastic features in the GPU implementation.
+This implementation is detailed in the following chapter.

### 2010-10-22 Thesis/Chapter6.tex
- Update 22 October -
Almost done...



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@188 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 This chapter will describe the implementation of this non-linear, viscoelastic and anisotropic finite element method algorithm in the open source framework SOFA.
-This work has led to three publications \citep{Comas2008,Taylor2008,Taylor2009}.
 \end{shortAbstract}
@@
 However, let us remind the main steps of the algorithm before tackling the implementation.
-Briefly, the algorithm consists of a precomputation phase in which element shape function derivatives $ \partial \mathbf{h} $ (and other quantities) and the system mass matrix $ \mathbf{M} $ are calculated, followed by a time-loop in which incremental solutions for the node displacements $ \mathbf{U} $ are found.
+Briefly, the algorithm consists of a pre-computation phase in which element shape function derivatives $ \partial \mathbf{h} $ (and other quantities) and the system mass matrix $ \mathbf{M} $ are calculated, followed by a time-loop in which incremental solutions for the node displacements $ \mathbf{U} $ are found.
 During each step of the time-loop we:
@@
 Memory storage is allowed through the use of textures, which essentially are structured collections of floating point values arranged as arrays.
-Shaders can read textyres from any location and as many time as they like.
+Shaders can read textures from any location and as many time as they like.
 As such it is a concept very similar to regular arrays.
@@
 Indeed, in contrast to texture reads, shaders can only write once into textures (at the end of the set of instructions).
-Morevoer, as stated earlier, shaders can only write in predefined locations corresponding to their potential screen location.
+Moreover, as stated earlier, shaders can only write in predefined locations corresponding to their potential screen location.
 Therefore, graphics API does not allow writing to random memory locations (this feature is called \emph{scattering}).
-Nethertheless, shaders can have multiple render targets.
+Nevertheless, shaders can have multiple render targets.
 The write locations within each texture are still fixed though.
@@
 The first GPU implementation	of the TLED algorithm was carried out by \cite{Taylor07b,Taylor08}.
-They used the Cg language and had to reformulate the TLED algorithm presented to accomodate with the limited possibilities of this graphics API.
+They used the Cg language and had to reformulate the TLED algorithm presented to accommodate with the limited possibilities of this graphics API.
 Indeed, the major difference between this parallel implementation and a serial one is the writing of element nodal forces to memory at the end of a first kernel and their subsequent retrieval and summation during a second kernel.
@@
 Of course, in this approach some elements will be computed several times (because being into different blocks).
-\ON Yet, \OFF storing the force contributions in shared memory instead of global memory to read them later on via texture fetches \ON may \OFF outperform the need of computing several elements several times.
-\ON However, the efficiency of this single kernel approach depends on the nature of the mesh.
+Yet,  storing the force contributions in shared memory instead of global memory to read them later on via texture fetches  may  outperform the need of computing several elements several times.
+However, the efficiency of this single kernel approach depends on the nature of the mesh.
 Because the force contributions of all elements attached to a node must be eventually added up, this approach allows to avoid the numerous (slow) accesses to global memory when the valency of the nodes is high (that is, nodes are attached to many elements).
@@
 Therefore, where the first implementation showed its limits with complex meshes, this single kernel approach has proven to be more computationally effective.
-\OFF
 
@@
 
-\ON
 This GPU implementation of the TLED algorithm was demonstrated publically at MICCAI 2007 during the talk given by Zeike Taylor \citep{Taylor07b}.
@@
 \end{figure}
-\OFF
 
@@
 The tests were performed on a simple test scene featuring a cube under gravity.
-Hexahedral meshes with different resolutions from $1\,331$ to $29\,791$ nodes were used and the results are presented in figure \ref{chap6:fig-cube_and_curves}.
+Hexahedral meshes with different resolutions from $1\,331$ to $29\,791$ nodes were used and the results are presented in \fig{chap6:fig-cube_and_curves}.
 In our standalone implementation, the second kernel not only accumulates nodal forces but also adds gravity and updates positions based on the central difference integration scheme.
@@
 Finally, we performed simulations of manipulation of a liver model in order to demonstrate the feasibility of the procedure with more realistic organ geometries.
-The model geometry was based on the liver mesh available in the SOFA framework, which we smoothed and remeshed to improve uniformity of element sizes and node valencies.
+The model geometry was based on the liver mesh available in the SOFA framework, which we smoothed and re-meshed to improve uniformity of element sizes and node valencies.
 A mesh of 748 nodes and $2\,428$ tetrahedral elements was solvable in real-time.
@@
 
-\TODO{Add movies online? If possible...}
-
 \section{Discussion}
@@
 In such a case, the critical time step for the whole simulation would be small even if there is only a single small element in the mesh.
-As an example, the liver mesh used in the section~\ref{chap6:liverSimu} had to be smoothed and remeshed properly \ON with great care \OFF to allow a decent simulation.
+As an example, the liver mesh used in the section~\ref{chap6:liverSimu} had to be smoothed and re-meshed properly  with great care  to allow a decent simulation.
 
-Material's properties are also a very influent factor for the determination of a critical time step.
+Material's properties also have an influence on the determination of a critical time step.
 If the critical time step may be fairly large for very soft tissue (like brain), it is already divided by 2 when modelling liver.
@@
 If excessively stiff springs are used, $ \Delta t_{cr} $ may indeed be reduced.
-\ON According to LS-Dyna documentation \citep{Hallquist06},  \OFF except in extreme conditions (that is from explosive or high speed impacts, or imposition of very large forces on highly constrained objects), spring stiffnesses of the same order of magnitude as the material stiffness are suitable and the effect on $ \Delta t_{cr} $ is therefore minimal.
-\ON However, in practice, we observed that spring stiffnesses greater than the material stiffness are often required to prevent any penetration of the two objects, which results in an unstable system.
-\OFF Kinematic constraint methods involve (loosely) direct relocation of penetrating nodes back to the outside surface of the penetrated body, and have no effect on $ \Delta t_{cr} $.
-\ON In theory, \OFF these may be especially useful for modelling interaction between soft tissues and stiff (effectively rigid) surgical tools.
-\ON It is worth nothing that \OFF the computational cost of either variety has not been investigated here.
+According to LS-Dyna documentation \citep{Hallquist06},   except in extreme conditions (that is from explosive or high speed impacts, or imposition of very large forces on highly constrained objects), spring stiffnesses of the same order of magnitude as the material stiffness are suitable and the effect on $ \Delta t_{cr} $ is therefore minimal.
+However, in practice, we observed that spring stiffnesses greater than the material stiffness are often required to prevent any penetration of the two objects, which results in an unstable system.
+Kinematic constraint methods involve (loosely) direct relocation of penetrating nodes back to the outside surface of the penetrated body, and have no effect on $ \Delta t_{cr} $.
+In theory,  these may be especially useful for modelling interaction between soft tissues and stiff (effectively rigid) surgical tools.
+It is worth nothing that  the computational cost of either variety has not been investigated here.
 
-\ON
 Even if a stable contact modelling could be achieved, one element of crucial importance has been omitted so far: the collision detection process.
@@
 
-	\subsubsection{Conclusion}
+\section{Conclusion}
+
 As we showed in the last two chapters, the TLED algorithm is a very efficient fully non-linear finite element formulation.
@@
 In addition, the TLED algorithm is also not appropriate to simulate the deformation of rigid object as the critical time step would be too small to allow real-time computations.
-In constrast, TLED excels for computing the deformation of soft tissues when the boundary conditions are controlled.
-As an exemple, the TLED algorithm may efficiently provide brain shift predictions as the interaction of the brain with the fixed skull is determined beforehand and rapid collision detection can be achieved in this case.
+In contrast, TLED excels for computing the deformation of soft tissues when the boundary conditions are controlled.
+As an example, the TLED algorithm may efficiently provide brain shift predictions as the interaction of the brain with the fixed skull is determined beforehand and rapid collision detection can be achieved in this case.
 
 Hence, provided that the application is appropriate, the TLED algorithm is an efficient and accurate algorithm to model the deformation of solid soft anatomical organs.
-\OFF
+While the first GPU implementation of the TLED was carried out by \cite{Taylor07b} in the Cg language, my efficient re-implementation in CUDA led to a threefold increase in performance.
+Overall, we observe a maximum speed improvement for GPU solution over CPU solution of up to $56.3 \times$.
+My collaboration with Zeike Taylor continued and we enhanced the formulation by adding anisotropy and viscoelasticity.
+My contribution was once again to implement these enhancements on GPU.
+To the best of our knowledge, this work constitutes the first GPU implementation of a non-linear, anisotropic and viscoelastic finite element procedure.
+This formulation was eventually implemented in the open source framework for medical simulation SOFA and will be released publically with the upcoming 1.0 release of SOFA.
+Over the course of my PhD, this work has led to three main publications \citep{Comas2008,Taylor2008,Taylor2009}.

### 2010-10-22 Thesis/Chapter7.tex
- Update 22 October -
Almost done...



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@188 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 An obvious modelling for thin structures is to apply the same methods than with solid objects.
-For instance, \cite{Holzapfel02} proposed a very complex 3-layer model of an aterial wall for simulating an angioplasty procedure.
-Their model is non-linear and anisotropic along fibers placed according to histology.
+For instance, \cite{Holzapfel02} proposed a very complex 3-layer model of an arterial wall for simulating an angioplasty procedure.
+Their model is non-linear and anisotropic along fibres placed according to histology.
 They also model diseased part of the vessel through a hardening parameter which adds plasticity.
@@
 Then we will present techniques relying on the derivation of a bending energy from geometric considerations.
-At last, we will introduce approaches based on shell theory using the more computationaly demanding finite element method.
+At last, we will introduce approaches based on shell theory using the more computationally demanding finite element method.
 
@@
 \cite{Provot95} improved a mass-spring model to take into account the non-elastic properties of woven fabrics.
-Because of the high stiffness of textiles, mass-spring models are particulary unstable when used to simulate cloth.
+Because of the high stiffness of textiles, mass-spring models are particularly unstable when used to simulate cloth.
 The author proposed a new method inspired from dynamic inverse procedure.
@@
 A limitation of mass-spring models that we often hear is the difficulty to derive spring stiffness from elastic properties (Young's modulus and Poisson's ratio).
-With this drawback in mind, \cite{Volino97} improved a mass-spring system by allowing the modelisation of Young's modulus and Poisson's ratio (see \fig{chap7:fig-fabricSquares}).
+With this drawback in mind, \cite{Volino97} improved a mass-spring system by allowing the modelling of Young's modulus and Poisson's ratio (see \fig{chap7:fig-fabricSquares}).
 They also added a correction technique relying on the control of damping and velocities to ensure the stability of their model.
@@
 They could simulate up to a few hundreds of hair stripes in real-time.
-For instance, \cite{Wang07} successfully used a network of linear and angular springs to describe bending and twisting of catheters and guidewires in an interventional radiology simulator.
+For instance, \cite{Wang07} successfully used a network of linear and angular springs to describe bending and twisting of catheters and guide-wires in an interventional radiology simulator.
 
@@
 \cite{Volino06} suggested an alternative somewhere between mass-spring and geometrically derived approaches.
-They compute a bending vector that represents the bending of the surface through  a simple linear combination of particule positions and they distribute this vector as particule forces according to the bending stiffness.
-The authors claimed much better accuracy than mass-spring models (particulary when dealing with low curvature and high bending stiffness) with similar computation time (see \fig{chap7:fig-dress} for an illustration).
+They compute a bending vector that represents the bending of the surface through  a simple linear combination of particle positions and they distribute this vector as particule forces according to the bending stiffness.
+The authors claimed much better accuracy than mass-spring models (particularly when dealing with low curvature and high bending stiffness) with similar computation time (see \fig{chap7:fig-dress} for an illustration).
 %
@@
 
-A method to use thin shell dynamics with point sampled surfaces for efficient animation was recently proposed by \cite{Wicke05} where the curvature of the shell is measured through the use of fibers.
-The fibers are used to approximate the differential surface operators.
+A method to use thin shell dynamics with point sampled surfaces for efficient animation was recently proposed by \cite{Wicke05} where the curvature of the shell is measured through the use of fibres.
+The fibres are used to approximate the differential surface operators.
 Their method supports both elastic and plastic deformations as well as fracturing and tearing of the material.
@@
 More recently, \cite{Allard09} also used a co-rotational finite element formulation to model the membrane of the lens capsule in a cataract surgery simulator.
-The goal was to simulate the anisotropic fracture progation during capsulorhexis, the technique used to create a circular opening in the lens capsule.
+The goal was to simulate the anisotropic fracture propagation during capsulorhexis, the technique used to create a circular opening in the lens capsule.
 The formulation did not feature bending however.

### 2010-10-22 Thesis/Chapter9.tex
- Update 22 October -
Almost done...



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@188 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 We will first describe the model in details before discussing its implementation in the open source framework SOFA.
-Lastly, this co-rotational shell FEM was applied to the simulation of the lens deployment during catarct surgery.
+Lastly, this co-rotational shell FEM was applied to the simulation of the lens deployment during cataract surgery.
 This simulation also accounts for the complex contacts that take place during the injection phase.
-This work was the object of two publications \citep{Comas2010a,Comas2010b}.
 \end{shortAbstract}
@@
 Indeed, as explained section \ref{chap9:interactions}, the force applied onto the curved surface of the shell must be processed by distributing the linear force received on the bent surface between the three vertices of the triangle.
-Consequently, the mapping previously described is extented to a mechanical mapping and two additional methods need to be implemented: applyJ and applyJT.
+Consequently, the mapping previously described is extended to a mechanical mapping and two additional methods need to be implemented: applyJ and applyJT.
 %
@@
 
+Moreover, because shells are high-order (curved) elements, fewer elements are needed to describe a given geometry with the same precision (as compared to using flat triangles).
+Generally speaking, this constitutes a decisive computational advantage since fewer elements are required.
+However, GPUs are more efficient when they tackle numerous but simple computations.
+Hence, a good speedup over sequential CPU implementations is usually only achieved when the number of elements involved becomes great enough.
+In contrast, the computations for each shell element are quite heavy and the number of elements used in the whole simulation is limited.
+Therefore, the conditions are not optimal to obtain maximum speedups on GPU.
+A straightforward GPU implementation of our shell FEM formulation where each thread deals with the calculations of one element would not make use of all the GPU capacity (the occupancy would not be maximal).
+If a reasonable improvement over the CPU implementation will probably be obtained nevertheless, the gain is likely to be smaller than in the case of the TLED algorithm for instance.
+Other approaches will need to be investigated to take full advantage of the GPU.
+For instance, computations within an element could be distributed between several threads to increase the occupancy of the GPU and hence the overall efficiency.
+
 To conclude on a possible parallelisation of the algorithm, an optimal GPU implementation is always a challenge to achieve.
 One needs to choose wisely the type of memory to be used for each variable and optimise the access patterns.
-However, it seems that there is no major obstacle to a GPU implementation and our co-rotation shell FEM will eventually be ported to GPU.
+While it seems that there is no major obstacle for a simple GPU implementation, the best speedups will only be attained through more elaborate techniques of parallelisation making full usage of the GPU.
+Although it will take some time, our co-rotation shell FEM will eventually be ported to GPU.
 
-\TODO{enough details?}
-
-\section{Comparison with analytical results}
+\section{Comparison with an analytical result}
 
@@
 Top: images from a real cataract surgery, courtesy of Dr.
-Tarek Youssef, ophtalmologist in Canada.
+Tarek Youssef, ophthalmologist in Canada.
 Below: our simulation of the implant's deployment.}
@@
 \end{figure}
+
+\section{Discussion}
+
+The finite element procedure that we propose was largely developed based on physical understanding without the use of mathematical shell theories.
+Indeed, our shell FEM was obtained by simply superimposing a plane stress and bending energies.
+According to \cite{Chapelle03}, this type of approach has a limited level of accuracy.
+The effective study of shell structures clearly requires a deep physical understanding of shell structural behaviours.
+The complexity of the physical behaviours of shells requires advanced mathematical analyses from theories that are still in development.
+The interested reader may refer to the book of \cite{Chapelle03} for more details on shell theory.
+If the range of application of our shell formulation might be somewhat limited, we did not find out any inconsistencies through our tests.
+We also need to keep in mind that the accuracy required in medical simulation is not the same than the one demanded in structural analyses for instance.
+Moreover, the additional complexity of a shell finite element formulation based on true shell theory may also very well jeopardise our real-time constraints.
+Therefore, we do not think that the use of this simple shell formulation significantly diminishes our contribution to the field of medical simulation.
+Indeed, to our knowledge, our co-rotational shell FEM is the first description of a shell finite element model applied to simulating the deformation of thin anatomical structures.
+
+Nevertheless, a few improvements are conceivable.
+Large strain could be allowed by using a non-linear strain tensor and more complex material laws (non-linear, viscoelasticity, etc.) could also be added to our shell formulation.
+However, those changes are substantial and their implementations are not straightforward.
+In addition, it is not clear how the benefits would be the greatest: whether from enhancing our model to a non-linear formulation or completely changing to another model based on true shell theory (where membrane and bending energies are not separated).
+The answer may depend on the type of constraints applied on the object.
+Also (as already mentioned section \ref{chap9:GPU}), the real-time capacity of our algorithm could be further improved by implementing our shell FEM on GPU.
+Although an optimal parallel re-implementation is always challenging, we have already reviewed why there was no major obstacle to a GPU implementation.
+
+\section{Conclusion}
+
+We proposed a co-rotational formulation for shell elements by extending a classical in-plane triangular finite element approach.
+This simple shell element can efficiently handle both membrane, bending and twist forces.
+The validity of our approach has been demonstrated though comparison with theoretical results.
+It was then applied it to a rather complex application: the simulation of intra-ocular lens implant deployment in a cataract surgery simulation.
+These results are very encouraging and show the potential of such models.
+Examples include the modelling of hollow anatomical structures (colon, stomach, etc.), the simulation of cardiac valve leaflets, and blood vessels.
+This work on this co-rotational shell FEM formulation was the object of one main publications \citep{Comas2010a}.
+
+So far, the mesh of the implant used in the cataract surgery simulation was created manually.
+However, manually creating the mesh is not suitable if we want to model the deformation of complex anatomical structures using shell elements.
+Therefore, the first step is to describe the surface of the structure that we want to simulate with curved patches.
+The goal is to create a mesh featuring the optimal number of curved shell elements while staying as close as possible to our targeted geometry.
+The following chapter deals with this problem.

### 2010-10-22 Thesis/TitlePage.tex
- Update 11 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@184 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 12 October -
memoir class used now, fancy chapter style



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@185 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 22 October -
Almost done...



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@188 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\begin{titlepage}
+\thispagestyle{empty}
 
@@
 				& Georges \textsc{Dumont}		& - & ENS Cachan, Rennes, France\\
-      \textit{President:}	& To be determined		& - & \\
-      \textit{Examinators:}   & Tamy \textsc{Boubekeur}          & - & Telecom ParisTech, Paris, France\\
+      \textit{Examiners:}   & Tamy \textsc{Boubekeur}          & - & Telecom ParisTech, Paris, France\\
       				& Christian \textsc{Duriez}			& - & INRIA (\textsc{Shaman} team), Lille, France\\
+      \textit{President:}	& Laurent \textsc{Grisoni} & - & INRIA (\textsc{Mint} team), Lille, France \\
       \textit{Invited:}   & Hans \textsc{De Visser}          & - & CSIRO, Brisbane, Australia\\
@@
 \end{center}
-\end{titlepage}
 \sloppy
-
-\titlepage

### 2010-10-24 Thesis/Chapter10.tex
- Update 24 October (final) -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@189 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\chapter{Physics-based reconstruction using shell elements}
+\chapter{Conclusion and perspectives}
 \label{chap10}
-\begin{shortAbstract}
-In the previous chapter, we proposed a co-rotational formulation for shell elements by extending a classical in-plane triangular finite element approach.
-This simple shell element can efficiently handle both membrane and off-plane forces (bending and twist).
-The validity of our approach has been demonstrated though comparison with theoretical results.
-It was then applied it to a rather complex application: the simulation of intra-ocular lens implant deployment in a cataract surgery simulation.
-These results are very encouraging and show the potential of such models.
-Examples include the modelling of hollow anatomical structures (stomach, colon, etc.), the simulation of cardiac valve leaflets, and blood vessels.
+
+\section{Summary of contributions}
+
+The objective of this PhD was to propose soft tissue models for computing the deformation of anatomical structures in real-time.
+Because the shape and the internal nature of these structures greatly differ, a unique model cannot deal with the variousness of the mechanical behaviours.
+Hence, we identified two major types of structures in human body, they can be either solid (brain, liver, prostate etc.) or hollow (colon, blood vessels, stomach etc.).
+
+Most of previous works focused on volumetric models that are able to capture the behaviour of solid organs.
+Yet, many of these models had to compromise on accuracy to achieve the real-time constraint demanded by medical simulation.
+Even though the best mathematical framework for predicting the deformation of a continuous structure is provided by continuum mechanics, solving these equations is a computationally very demanding task and simpler models are often used in practice.
+And even when the finite element method is employed to solve the equations of continuum mechanics, linear strain measure and simple constitutive laws are common and therefore does not constitute an appropriate modelling of the complex mechanical behaviours of solid anatomical structures.
+
+In contrast, much less research has been carried out for the modelling of hollow structures.
+In fact in the field of medical simulation, very few models have been proposed for simulating, in real-time, the deformation of thin anatomical structures whose volume is negligible compared to their surface area.
+Nevertheless, human body features many hollow structures and we feel that the need for modelling hollow organs in medical simulation is equally important as with solid organs.
+
+\bigskip
+
+Consequently, we strived to propose solutions to these two known problems in the field of medical simulation.
+
+\subsubsection*{Solid structure modelling.}
+We implemented the total Lagrangian explicit dynamics (TLED) algorithm on GPU using CUDA.
+This finite element formulation is fully non-linear: it makes use of a non-linear strain measure (which allows large deformations) and embodies a non-linear constitutive law between strain and stress as characterised by \cite{Fung93} for soft-tissues.
+While the first GPU implementation of the TLED was carried out by \cite{Taylor07b} in the Cg language, our efficient re-implementation in CUDA led to a threefold increase in performance.
+Overall, we observe a maximum speed improvement for GPU solution over CPU solution of up to $56.3\,$x.
+My collaboration with Zeike Taylor continued and we enhanced the formulation by adding anisotropy and viscoelasticity.
+To the best of our knowledge, this work constitutes the first GPU implementation of a non-linear, anisotropic and viscoelastic finite element procedure.
+This formulation was eventually implemented in the open source framework for medical simulation SOFA and will be released publicly with the upcoming 1.0 release of SOFA.
+%By providing a fast and accurate non-linear FEM with various constitutive laws for soft tissue modelling to worldwide researchers, we thus hope to assist in enhancing the realism of medical simulators.
+
+\subsubsection*{Hollow structure modelling.}
+We proposed a framework for real-time modelling of thin anatomical structures.
+We first designed a triangular shell element by combining a two-dimensional in-plane membrane energy, with an off-plane energy for describing bending and twist.
+We improved and extended a plate model first introduced by \cite{Przemieniecki85} to a co-rotational formulation to allow large displacements (but small deformations).
+Once combined with an in-plane membrane formulation we obtained an accurate, yet computationally efficient, shell finite element method featuring both membrane and bending energies.
+To the best of our knowledge, this co-rotational shell FEM is the first description of a shell finite element method in the field of medical simulation.
 However, to model the deformation of complex anatomical structures using shell elements, the first step is to describe its surface with curved patches.
-In the next section, we propose to study how to approximate the surface of anatomical structures with (curved) shell elements.
-\end{shortAbstract}
-
-\section{Techniques of mesh simplification}
-
-\subsection{Introduction: our motivation}
-
-Our idea is the following: while many flat triangles are required to describe highly curved surfaces, fewer triangular shell elements are needed to describe a given geometry with the same precision since they can be curved.
-Our goal is to create a mesh featuring the optimal number of shell elements while staying as close as possible to our targeted geometry.
-Literature about volumetric mesh generation algorithms is fairly dense.
-However, there are only a few that are concerned about the generation of meshes over curved surfaces.
-One of the most widely used techniques for the creation of surface meshes is the plane to surface transformation method \citep{Zienkiewicz71}, mesh is first generated on a two-dimensional domain and then mapped onto the surface.
-If this method gives reasonably good meshes on smooth surfaces, the results are usually quite poor with more complex curved surfaces.
-The finite elements may be generated directly on the curved surfaces based on the advancing front technique \citep{Lo85,Lau96}.
-The main issue with this approach is that an analytical description of the geometry is needed, which is not the case in medical simulation.
-Another method consists of using an a priori error estimator to build an adaptive mesh generation \citep{Baumann97}.
-However, the tolerance of this indicator should be chosen depending on the desired accuracy of the finite element solution.
-Therefore it requires some knowledge about the problem in order to choose an effective tolerance.
-\cite{Bechet02} start from an existing triangular mesh created with a CAD (computer-aided design) software and refine and smooth the mesh based on element quality and surface curvature.
-While all those methods allow their authors to get satisfactory meshes over curved surfaces according to their needs, they all make use of flat elements to mesh geometries and do not generate actual shells.
-
-In fact, it is interesting to note that this process of meshing a curved surface is similar to the reconstruction of the surface of objects in computer vision.
-Indeed, calculating curvature maps of 3D surfaces represented as digitised data (point clouds or triangulated meshes) has been extensively studied.
-One of the most common approach is to use continuous surface patches \citep{Kolb95,Douros02}.
-The surface is locally approximated by an analytic representation called surface patch, usually chosen to be an implicit surface.
-\cite{Tang99} presented an elegant and non noise-sensitive approach that infers sign and direction of principal curvatures directly from the input and the authors used this information for coherent surface and curve extraction from 3D data.
-
-However, we need to realise that our situation is substantially different than all previously mentioned geometrical approaches as we want to model the deformation of the structure.
-In that regard, the curvature of the surface has a physical meaning: it represents the mid-surface of the shell.
-By keeping this in mind, we propose an approach to mesh a curved surface based on the same polynomial shape function \eqref{chap9:eq-deflection} used in our co-rotational shell FEM.
-
-%Simplification strategies may be grouped into two categories: local strategies that iteratively simplify the mesh by the repeated application of some local operator, and global strategies that are applied to the input mesh as a whole \citep{Cignoni97,Talton04}.
-%
-%\subsection{Local simplification strategies}
-%
-%Typically, they define some mesh operation $S$ that, when applied to a mesh $M$, acts on a small collection of its elements and produces a new mesh $\overline{M}$  with fewer elements.
-By repeated application of $S$, a mesh may be simplified arbitrarily.
-In order to determine the mesh elements to which $S$ should be applied on a given iteration, $S$ may be associated with an error function (or cost function) that measures the amount of error the operation will introduce into the approximation.
-By computing the error associated with every possible application of $S$ at a particular iteration, the algorithm can apply the one with minimal cost.
-This type of heuristic is quite reasonable for simplification problems, and in practice these methods work well.
-%
-%One nice property of local iterative algorithms is that they allow the user to specify the desired attributes of the target approximating mesh with a high degree of precision.
-For example, the user may allow the algorithm to run until the mesh contains $k$ faces, or until the error at a given vertex exceeds some threshold $\varepsilon$.
-%
-%\subsubsection*{Vertex decimation.}
-%The decimation algorithm, first proposed by \cite{Schroeder92}, is simple.
-Multiple passes are made over all vertices in the mesh.
-During a pass, each vertex is a candidate for removal and, if it meets the specified decimation criteria, the vertex and all triangles that use the vertex are deleted.
-The resulting hole in the mesh is patched by forming a local triangulation.
-The vertex removal process repeats, with possible adjustment of the decimation criteria, until some termination condition is met.
-%
-%The error associated with a particular decimation typically depends on the classification of the vertex being decimated.
-For the general case of a manifold vertex $\lbrace i \rbrace$ not near a boundary, \citeauthor{Schroeder92} consider the set of planes formed by the adjacent triangles and compute a single approximating plane based on the area-weighted average of the triangle normals $\textbf{n}_k$, centers $x_k$, and areas $A_k$:
-%%
-%\begin{equation}
-%\textbf{n} = \frac{\sum_k A_k\textbf{n}_k}{\sum_k A_k},  \hspace*{1cm} \widehat{\textbf{n}} = \frac{\textbf{n}}{\vert \textbf{n} \vert}, \hspace*{1cm}  x = \frac{\sum_k A_k\textbf{x}_k}{\sum_k A_k}
-%\end{equation}
-%
-%Then they calculate the distance from vertex $\lbrace i \rbrace$ to this plane by  $\vert \widehat{\textbf{n}} \cdot (v_i - x) \vert$ and define this quantity to be the cost associated with the decimation.
-This metric ensures that vertices in smooth regions will be decimated before vertices that define sharp features.
-%
-%In practice, simplification schemes based on vertex decimation excel at eliminating un-necessary geometry, such as is typically found in meshes generated by the marching cubes algorithm where large planar regions may be subdivided into many redundant triangles.
-%
-%\subsubsection*{Edge contraction.}
-%Edge contraction, originally proposed by \cite{Hoppe93}, is the most common simplification operation.
-An edge contraction operates on a single edge $\lbrace v_1,v_2 \rbrace$ and contracts that edge to a single vertex $\lbrace w \rbrace$, updating all edges previously incident on $\lbrace v_1 \rbrace$ and $\lbrace v_2 \rbrace$ to reference $\lbrace w \rbrace$.
-On a manifold mesh without boundary, each edge contraction will collapse precisely two faces (see \fig{chap10:fig-edgeCollapse}).
-Edge contractions can alter the topology of a mesh, since repeatedly contracting all the edges around a hole will eventually close it.
-%%
-%\begin{figure}
-%\centering
-%\includegraphics[height=3cm]{chapter10/edgeCollapse.pdf}
-%\caption[The edge contraction operation]{The edge contraction operation. Image courtesy of \cite{Hoppe93}.}
-%\label{chap10:fig-edgeCollapse}
-%\end{figure}
-%
-%For a given contraction $\lbrace v_1,v_2 \rbrace \rightarrow \lbrace w \rbrace$, it is not immediately clear what value should be assigned to $v_h$, i.e.
-where the resulting vertex should be placed.
-Obvious choices such as $v_h = v_i$, $v_h = v_j$, or $v_h = (v_i + v_j)/2$ are convenient, but can easily be shown to be non-optimal.
-Rather than placing $v_h$ arbitrarily, it is better to consider the error function associated with the contraction operation and attempt to minimise its value over the space of possible vertex placements.
-Once the optimal target position has been computed for every possible contraction, the contraction with the smallest associated error can be selected and applied.
-%
-%\subsubsection*{Adaptative subdivision.}
-%Coplanar or nearly coplanar faces are searched for in the mesh, merged into larger polygons, and then retriangulated into fewer faces than those originally required \citep{Dehaemer91,Hinker93}.
-Face merging is driven by a co-planarity test.
-%
-%The basic adaptive subdivision method as presented by \cite{Schmitt86} is sketched in \fig{chap10:fig-adaptativeSubdivision}.
-(a) shows several data points produced by the sampling process and sketch (b) adds a trial polygon that approximates the surface represented by the sample points.
-A bilinear interpolation is performed over the surface of the approximating polygon to locate the point on the surface 'directly below' a sample data point.
-The distance (or error) between the surface and the data point is then calculated, as shown in (c), and is compared to the user specified tolerance.
-If any of the errors exceeds the tolerance value then the polygon is divided into four smaller polygons, as shown in (d), and the process is repeated recursively.
-%%
-%\begin{figure}
-%\centering
-%\includegraphics[width=13.5cm]{chapter10/adaptativeSubdivision.pdf}
-%\caption[Surface-fitting polygons with adaptive subdivision]{Surface-fitting polygons with adaptive subdivision. Image courtesy of \cite{Schmitt86}.}
-%\label{chap10:fig-adaptativeSubdivision}
-%\end{figure}
-%
-%When the approximation to the data is within the specified tolerance, the recursion is terminated and the polygon is saved in a linked list.
-Upon completion of the subdivision algorithm, the resulting polygons are displayed for the user to judge the results.
-Generally, the user runs the program repeatedly while varying the maximum tolerance until the results are suitable.
-%
-%\subsubsection*{Polygon growth.}
-%Another method proposed by \cite{Dehaemer91} to systematically generate surface-fitting polygons is to use a polygon growth technique (see \fig{chap10:fig-polygonGrowth}).
-First a seed polygon is selected from the original set of data, as in (a).
-Then a neighbouring polygon is selected and the two are combined into a larger trial polygon ((b) and (c)).
-If the trial polygon passes the accuracy metric, 10 additional neighbors are selected in an attempt at further growth, as in (d) through (g).
-When all attempts at further growth fail, the polygon is added to the list of polygons describing the reduced object, as shown in (h).
-A new seed polygon is selected and the process repeated.
-When there are no more polygons from which to choose a new seed, the process is terminated.
-%%
-%\begin{figure}
-%\centering
-%\includegraphics[width=12cm]{chapter10/polygonGrowth.pdf}
-%\caption[Polygon growth]{Polygon growth. Image courtesy of \cite{Dehaemer91}.}
-%\label{chap10:fig-polygonGrowth}
-%\end{figure}
-%
-%This method of fitting polygons to the sampled data is not as useful as the adaptive subdivision methods, for several reasons.
-First, it is an essentially brute force method of surface fitting and very repetitive in nature.
-Many trial polygons are created, only to be 'thrown out' because they exceed the error tolerance.
-This causes the procedure to execute quite slowly (orders of magnitude more slowly) than the adaptive subdivision method.
-Second, the data is not significantly reduced.
-Many of the generated polygons are of complicated shapes and use a large number of vertices.
-Thus many of the original sample data points are retained.
-%
-%
-%\subsubsection*{Appearence preserving simplification.}
-%A completely different approach to determining the error associated with a given simplification operation was described by \cite{Cohen98}.
-Their metric, which is appearance-based, measures the amount of deviation caused by the operation in the screen-space (pixelised) representation of the mesh.
-Their algorithm functions by decoupling surface position from color and curvature information and storing the latter two quantities in texture and normal maps.
-A traditional geometric simplification algorithm can then be employed to filter the surface position, while a hardware-based approach is used to filter the color and normal information, resulting in simplified representations that are nearly visually indistinguishable from the original.
-However, the algorithm requires a parameterisation of the mesh in order to function.
-%
-%
-%\subsection{Global simplification strategies}
-%
-%\subsubsection*{Vertex clustering.}
-%The method of vertex clustering was originally proposed by \cite{Rossignac93} to handle meshes of arbitrary topological structure.
-In their algorithm, each vertex in the input mesh is assigned a weight based on its perceptual importance: vertices adjacent to triangles with large faces and those in areas of high curvature are weighted more heavily than vertices in smooth regions adjacent to smaller triangles.
-Next, a bounding box is placed around the mesh and subdivided into a three-dimensional grid.
-Finally, all the vertices in a given grid cell are clustered to the position of the vertex with maximum weight.
-Degenerate geometry may then be removed, although the authors proposed a novel visualisation of certain degenerate faces and edges in an effort to enhance the recognizability of their simplified meshes.
-The algorithm is extremely efficient and simple to implement, and the level of simplification may be controlled (although with some difficulty) by choosing the resolution of the overlaid grid.
-However, vertex clustering can drastically alter the topology of the input mesh in an unpredictable manner, and in practice does not produce very faithful geometric approximations at low levels of detail.
-%
-%\subsubsection*{Re-tiling.}
-%The re-tiling method described by \cite{Turk92} begins with a polygonal surface and creates a triangulation of this surface with a user-specified number of vertices.
-The points are placed uniformly over a given polygonal surface by distributing points at random over the surface and then having each point repel all of its neighbors.
-An alternative to placing the points randomly is using the variation in surface curvature.
-The features of the surface would be more accurately reflected in a re-tiling by increasing the density of vertices in regions of high curvature.
-%
-%The key notion in the re-tiling procedure is the creation of an intermediate model called the mutual tessellation of a surface that contains both the vertices from the original model and the new points that are to become vertices in the re-tiled surface.
-The new model is then created by removing each original vertex and locally re-triangulating the surface in a way that matches the local connectedness of the initial surface (see \fig{chap10:fig-retiling}).
-%
-%If this method produces nice meshes, an issue is finding measures of how closely matched a given re-tiling is to the original model. 
-%%
-%\begin{figure}
-%\centering
-%\includegraphics[width=12cm]{chapter10/retiling.pdf}
-%\caption[Retiling technique]{Upper left: original surface.
-Upper right: candidate vertices after point-repulsion.
-Lower left: mutual tesselation.
-Lower right: final tesselation.
-Image courtesy of \cite{Turk92}.}
-%\label{chap10:fig-retiling}
-%\end{figure}
-
-\section{Our method}
-
-	\subsection{A simple algorithm}
-
-In the following, we assume that we have a high resolution triangular mesh obtained from a binary segmented image of the object we want to simulate (via a Marching Cube algorithm for instance, \cite{Lorensen87}).
-We propose a technique that can simplify the high resolution mesh of the object to be simulated and approximate the surface of anatomical structures with shell elements whose each surface is described by the shape function used in our shell formulation.
-Thus, the first step in the process of generating a shell-based mesh is an important decimation of the high resolution mesh, using quadric edge collapse technique implemented in \citeauthor{Meshlab}.
-MeshLab is an open source program for the processing and editing of unstructured 3D triangular meshes.
-The algorithm tries as much as possible to preserve mesh boundaries and generates high quality triangular elements.
-We then apply a heuristic method derived from the work of \cite{Saupin07} with tetrahedral meshes based on simple geometrical rules.
-For each node of the coarse mesh, we find the three closest triangles on the high resolution mesh and we move the node to the barycentre of the three centres of mass of those triangles.
-This technique locally smoothes the surface of the mesh while converging towards the desired high resolution mesh.
-
-At each iteration of this algorithm, we want to measure the error between the curved surface of shells and the target.
-Indeed, we need to ensure that the distance between the surface of our shell-based mesh and the targeted high resolution mesh will be minimal.
-An efficient technique for measuring the error between two surfaces is the Hausdorff distance \citep{Klein96}.
-As a reminder the Hausdorff distance between two meshes is the maximum between the two so-called one-sided Hausdorff distances:
-\begin{equation}
-d_{\mathrm{H}}(X,Y) = \max \left\{ \sup_{x \in X} \inf_{y \in Y} d(x,y), \sup_{y \in Y} \inf_{x \in X} d(x,y) \right\},
-\end{equation}
-where $d()$ is the Euclidian distance between two points.
-The same technique of subdivision used for rendering allows us to sample the actual surface described by the shells to compute the Hausdorff distance with the targeted high resolution mesh.
-
-By using the Hausdorff distance between two iterations, the process may be stopped when the required precision has been reached.
-A simple example is shown \fig{chap10:fig-cylinder} to illustrate the method.
-\begin{figure}[ht]
-\centering
-\includegraphics[width=13.5cm]{chapter10/exampleCylinder.pdf}
-\caption {The target (a) is a high resolution cylinder mesh of 16,384 triangles and we start from a very coarse mesh (12 triangles), rendered with flat triangles here (b).
-In (c) the coarse mesh is rendered with shells and a one-sided Hausdorff distance colour map is applied to show the initial error with the high resolution mesh.
-(d) One-sided Hausdorff distance colour map after one iteration of our algorithm (48 shells).}
-\label{chap10:fig-cylinder}
-\end{figure}
-
-	\subsection{Results on complex anatomical structures}
-
-\subsubsection*{Meshing of anatomical structures.}
-This approach has been applied to approximate more complex anatomical geometries with curved shell elements.
-Two examples are given with the Glisson's capsule, which is the membrane surrounding the liver (\fig{chap10:fig-liver}) and an aneurysm (\fig{chap10:fig-aneurysm}).
-In each case, the error is computed through a Hausdorff distance and expressed as a percentage of the diagonal of the object's bounding box.
-%
-\begin{figure}[ht]
-\centering
-\includegraphics[width=13.5cm]{chapter10/resultsLiver.pdf}
-\caption {(a) the targeteted high resolution Glisson's capsule mesh ($8\,000$ triangles).
-(b) the one-sided Hausdorff distance error map after applying only one iteration of our algorithm to the coarse mesh ($1\,200$ shells).}
-\label{chap10:fig-liver}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=13.5cm]{chapter10/resultsAneurysm.pdf}
-\caption {(a) the targeteted high resolution aneurysm mesh ($28\,368$ triangles). (b): the one-sided Hausdorff distance error map on a mesh of 772 shells generated with our method.}
-\label{chap10:fig-aneurysm}
-\end{figure}
-
-\subsubsection*{Computation times.}
-We perform several tests on the aneurysm model at different resolutions to measure computation times.
-The shells are resisting to a uniform pressure load and solved using a Conjugate Gradient (CG) iterative solver.
-The computation times are reported in \fig{chap10:fig-computation}.
-Implicit integration allows for large time steps ($40\,ms$) and the computation is real-time for 800 shell elements and a reasonable error criterion ($5\,\%$).
-When the computation time must be bounded (critical real-time applications), one can fix the number of CG iterations to, for instance, $100$ and remains real-time for $1\,000$ shell elements.
-However, in that case the accuracy of the results is not checked.
-%
-\begin{figure}[ht]
-\centering
-\includegraphics[width=13.5cm]{chapter10/computation_times.pdf}
-\caption {Computation time on meshes of 200, 400, 600, 800 and $ 1\,000 $ elements.}.
-\label{chap10:fig-computation}
-\end{figure}
-
-\subsection{Coupling between tetrahedra and shells for advanced modelling}
-Structures in human body can be either solid (brain, liver, prostate etc.) or hollow (colon, blood vessels, stomach etc.).
-However, knowing how to model the two kind of structures is not sufficient to reach a high degree of accuracy, real life situations are more complex.
-As an example, the external surface of the liver is covered by a layer of cells called Glisson's capsule.
-Its interaction with the liver plays an important role into the overall structure's mechanical behaviour.
-Therefore considering the interaction between solid and hollow objects is as crucial as modelling the two structures separately.
-
-An example of medical procedure to illustrate this point even further is angioplasty.
-Angioplasty is the technique of mechanically widening a narrowed or obstructed blood vessel, typically as a result of atherosclerosis.
-An empty and collapsed balloon on a guide wire is passed into the narrowed locations and then inflated to a fixed size.
-The balloon crushes the fatty deposits, so opening up the blood vessel to improved flow.
-As a proof of concept, we tried to simulate an angioplasty.
-Our preliminarty results are presented on \fig{chap10:fig-stent}.
-The blood vessel is modelled using the shell FEM formulation described in this paper and the fatty deposits are simulated with a tetrahedral FEM method and are fixed to the interior wall of the blood vessel.
-When the balloon inflates it crushes the deposits and they then apply a pressure onto the curved surfaces of shells modelling the interior wall.
-The forces are then distributed onto the mechanical nodes of the blood vessel mesh as detailed in section~\ref{chap9:interactions}, which widens the blood vessel as expected.
-%
-\begin{figure}[ht]
-\centering
-\includegraphics[width=12cm]{chapter10/stenose_final.pdf}
-\caption {Simulation of an angioplasty procedure.
-(a, c): A collapsed stent is inserted into the blood vessel simulated with our shell FEM formulation.
-The fatty deposits (in yellow) are modelled with tetrahedral FEM.
-(b, d): The stent is crushing the fatty deposits which creates a pressure onto the interior wall and widens the blood vessel.}
-\label{chap10:fig-stent}
-\end{figure}
+Indeed, while many flat triangles are required to describe highly curved surfaces, fewer triangular shell elements are needed to describe a given geometry with the same precision since they can be curved.
+Therefore, we suggested a simple method to mesh a surface with curved shell elements.
+The novelty of our method relies on the combination of a shell finite element formulation and a geometric surface reconstruction both based on the same polynomial interpolation function used to describe the surface of shells.
+We also showed how contacts and interactions with the curved surfaces of shells can be detected and handled using the same function.
 
@@
 
-Although we were able to mesh complex anatomical structures with a fairly small number of shell elements, our approach could be further improved.
-In our current implementation, if the error measured with the Hausdorff distance is greater than a threshold (which is also difficult to choose), all triangular elements are subdivided into 4 new triangles and each of the new vertices is moved towards the targeted geometry based on our heuristic algorithm.
-Obviously, this is not optimal as the error between the two meshes may very well be local and therefore the use of a global mesh refinement is not very suitable.
-The second weakness is that we also do not check the quality of the elements during our mesh process.
-If some elements are stretched from the heavy decimation, these elements are not corrected during the process of re-meshing.
-Our algorithm could even worsen the situation by splitting those stretched elements into four smaller (and more stretched) elements.
-One solution for the latter issue would be to enhance our algorithm with a relaxation phase where a repulsion force is applied between each vertex (constrained to remain on the surface of the object).
-This physical approach would allow a more even distribution of the vertices across the surface and would contribute to obtain better quality elements overall.
-Yet, meshing a surface with curved elements is a very complex problem and still an active area of research.
-In contrast to all previous geometrical techniques, we suggested a more physical approach consistent with our shell FEM formulation and obtained good results in meshing complex anatomical structures.
+	\subsection{Solid structures modelling}
 
-\section{Conclusion}
+The TLED algorithm offers an accurate modelling for soft tissues and its GPU implementation using CUDA is very efficient.
+Hence, real-time computations can be achieved without any trade-off on the accuracy of the deformation.
+One of the features allowing such an efficiency is the use of an explicit time integration scheme.
+If this permits significant computational advantage, the use of an explicit integration scheme imposes that the time step of the simulation remains below a critical time step.
+This critical time step depends on two kinds of criteria: the size of the elements and the material's properties (Young's modulus and Poisson's ratio).
+Therefore, the meshing of the anatomical structures to be simulated must be accomplished with great care to insure well-shaped elements.
+Even more restricting, the use of the TLED formulation is not suitable for modelling stiff objects in real-time.
+Indeed, increasing the Young's modulus translates into a lower critical time step, which necessitates an increase of the simulation's frequency to enforce real-time.
+In some applications, the value of the Poisson's ratio may be relaxed from 0.49 to 0.45 for instance.
+However, if it allows to double the critical time step, this relaxation introduces additional inaccuracies.
 
-Over the last two chapters, we proposed a complete framework for real-time modelling of thin anatomical structures.
-The novelty of our method relies on the combination of a shell finite element formulation and a geometric surface reconstruction both based on the same polynomial interpolation function used to describe the surface of shells.
-We also show how contacts and interactions with the curved surfaces of shells can be handled using the same function.
-In summary, these polynomial shape functions are used in three different ways in our computational model:
-%
-\begin{description}
-\item[(a)] to approximate complex geometrical shapes
-\item[(b)] to compute internal forces via a co-rotational shell FEM
-\item[(c)] to compute contact forces onto a curved triangle
-\end{description}
+One of the most common critics we received for this work concerned the poor ability of the TLED to handle stiff contacts.
+Indeed, interactions may impact the critical time step resulting from the use of an explicit time integration scheme and thus influence the overall stability of the simulation.
+The work carried out during my PhD was entirely concerned with the numerical solution provided by the TLED formulation.
+Yet, while this is beyond the scope of the present work, a range of contact formulations are available in commercial finite element packages.
+According to their documentation, the impact on the critical time step should be minimal.
+However, for a demonstration at an IT exhibition (CeBIT Australia), we tried to setup an interaction between a virtual tool driven by a haptic device and a cube with brain-like properties.
+If we managed to produce an interesting demo for the exhibition, the system appeared fairly unstable.
+But even if a stable contact modelling technique could be formulated, the actual limiting factor is the collision detection process.
+Because the time step must be very small when using an explicit time integration scheme (often around one millisecond or less), the collision detection algorithm must also be run at the same frequency.
+The problem is that no collision detection algorithm is fast enough to execute at $ 1\,000\,$Hz or more with objects of reasonable sizes.
+One notable exception is the interaction with a fixed rigid surface where the collision detection is rapid.
+
+Consequently, this list of pros and cons limits the TLED's range of applications.
+If its usage in very interactive environments is not suitable, the TLED finite element algorithm is great at computing the deformation of soft tissues when the boundary conditions are known and controlled.
+The TLED is also a great algorithm for non-rigid registration problems found in image-guided therapy \citep{Modat10}.
+Indeed, intra-operative image registration must be fast enough that the work flow of the surgical procedure is not interrupted.
+Hence, provided that the application is appropriate, the TLED algorithm is an efficient and accurate algorithm to model the deformation of solid soft anatomical organs.
+
+	\subsection{Hollow structures modelling}
 
 We achieved our objectives to propose a versatile solution which can simulate, in real-time, thin objects with various shapes and material properties with good accuracy.
-The efficiency of the method was illustrated through shell-based reconstruction and real-time simulation of the deformations of various anatomical structures and other thin objects (like an intra-ocular implant).
-We also presented preliminary results on the coupling between solid (tetrahedra) and thin objects (shells) for the advanced modelling of anatomical structures via the simulation of an angioplasty procedure.
-This work on modelling thin anatomical structures was the object of two main publications \citep{Comas2010a,Comas2010b}.
+The efficiency of the method was illustrated through shell-based reconstruction and real-time simulation of the deformations of various anatomical structures and other thin objects.
+Nevertheless, the use of a linear strain measure limits the range of application to small deformations (and large displacements).
+Moreover, the constitutive law was chosen to be the simple (linear) Hookean law, which it is known not to describe accurately the relationship between strain and stress in soft tissues.
+However, those potential changes are substantial and their implementations are not straightforward.
+In addition, it is not clear from which method the benefits would be the greatest: whether from enhancing our model to a non-linear formulation or completely changing to another model based on true mathematical shell theory (where membrane and bending energies are not separated).
+Our shell FEM could also be improved by implementing the formulation on GPU.
+Although an optimal parallel implementation is always challenging, our method is highly parallelisable and there is no major obstacle to an efficient GPU implementation.
+
+Regarding our geometric surface reconstruction based on shells, we were able to mesh complex anatomical structures with a fairly small number of shell elements.
+However, the main drawback of our method is the lack of any assessment of the quality of the elements created.
+Indeed, if some elements are stretched from the heavy decimation, these elements are not corrected during the process of re-meshing.
+Our algorithm could even worsen the situation by splitting those stretched elements into four smaller (and more stretched) elements.
+Yet, meshing a surface with curved elements is a very challenging problem and still an active area of research.
+In contrast to all geometrical techniques described in the literature, we presented a more physical approach consistent with our shell FEM formulation and obtained good results in meshing complex anatomical structures.
+
+	\subsection{Use of GPU acceleration}
+
+One of the strongest constraints in medical simulation is the demand for real-time computations.
+This constraint is what makes all tasks in a simulator so challenging (collision detection, contact and soft tissue modelling, haptic feedback, photo-realistic rendering etc.).
+In fact, this is at the origin of the multitude of simplifications proposed in simulation for modelling the biomechanics of anatomical structures.
+The lack of accuracy induced by these simplifications is of great concern to us.
+In this PhD, a part ot the solution was to accelerate the computations by using the GPU.
+While we have showed that this approach can work very well, we need to keep in mind that GPU acceleration is not the ultimate solution.
+Firstly, a standard is yet to be imposed in GPGPU.
+While OpenCL is openly targeted as a multi-vendor and multi-platform API, its usage remains very scarce nevertheless.
+In fact, the first experiments carried out within the SOFA framework revealed that OpenCL is not as optimised as CUDA for instance (up to twice as slow).
+Secondly, GPU programming is not simple to master.
+Although we have seen many improvements in flexibility of programming over the last 3 or 4 years, a strong understanding of the underlying GPU hardware is still mandatory to create an optimal code.
+As an example, a GPU offers different types of memory, all with distinct pros and cons (strict access patterns, latency, limited size etc.), and choosing the best type of memory for each variable may be rather challenging to say the least.
+But most importantly, GPU acceleration is only interesting if the code is highly parallelisable.
+The difficulty is to devise a model that combines (1) an excellent capability at predicting the deformation of anatomical structures and (2) great parallel properties.
+Ideally, the parallelisation of the algorithm must be thought as part of the design process rather than seen as two unconnected steps.
+Even when the code is greatly parallelisable, one could face limited GPU hardware possibilities (as with concurrent writes for instance).
+GPU acceleration is all but a straightforward process.
+One must often accomplish a considerable amount of work to achieve a substantial gain.
+% However, a decent GPU implementation may allow improvements of up to one or two orders of magnitude.
+For instance, our CUDA implementation of the TLED algorithm was about $ 56 \times $ faster than an equivalent CPU implementation on 2008 hardware.
+
+\section{Perspectives}
+
+Earlier this year, NVIDIA released a new GPU architecture called \emph{Fermi}.
+Fermi features several major innovations for GPGPU computing like cache (faster memory access in some cases), unified address space (easier to run C and C++ programs), debugging support, faster atomic instructions, compliance with the newly revised IEEE 754 floating point standard, three times the amount of shared memory, etc.
+In theory, this series of improvement should boost the performance of the second and third TLED implementations (see section \ref{chap6:secondTLED}) and we should investigate and compare the efficiency of the three designs on this new hardware.
+
+In a short term perspective, our co-rotational shell FEM algorithm could also be implemented on GPU.
+We already assessed in section \ref{chap8:GPU} that there was no major obstacle which would prevent a parallel implementation.
+Nevertheless, an optimal GPU implementation will demand careful design to take full advantage of the hardware.
+
+Although the meshing of complex surfaces with curved shell elements was not the heart of this PhD, our naïve approach could be enhanced by focusing on element quality.
+One simple solution would be to add a relaxation phase where a repulsion force is applied between each vertex (constrained to remain on the surface of the object).
+This physical approach would allow a more even distribution of the vertices across the surface and would contribute to obtain better quality elements overall.
+While a more mathematical method (a derivative of advancing front techniques for instance) could probably be formulated, the problem is complex and could certainly account for another PhD.
+
+Finally, all the contributions of this PhD could contribute to enhance the complexity of anatomical structure modelling through combining solid and hollow structures.
+While we presented our findings on the coupling between solid (tetrahedra) and thin objects (shells) via the simulation of an angioplasty procedure, these results are still preliminary and much work remains to be done.
+In particular, the validity of the interaction and the accuracy of the overall deformation of the blood vessel were not quantitatively evaluated.
+Another interesting example of coupling would be the complete model of the liver.
+The relationship between strain and stress within the parenchyma (functional part of the liver) is known to be non-linear and anisotropic.
+Moreover, both the vascular network within the tissue and the Glisson's capsule have a strong influence on the overall mechanical behaviour of the liver.
+The contributions introduced in this PhD thesis may offer a solution to this challenging problem.
+While the parenchyma could be simulated with an algorithm such as the TLED (materially non-linear and able to account for the anisotropy), the blood vessels and the Glisson's capsule could be described using our co-rotational shell FEM.
+
+As a conclusion, the TLED algorithm and our co-rotational shell FEM are respectively the first non-linear FEM for solid structures and the first model featuring bending for hollow structures to be implemented into the open source framework SOFA.
+Although they are based on continuum mechanics, they may both be employed to simulate the deformation of entire organs in real-time.
+By providing efficient FEM algorithms for all types of anatomical structures to worldwide researchers, we thus hope to assist in enhancing the realism of medical simulators.

### 2010-10-24 Thesis/Chapter2.tex
- Update 24 October (final) -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@189 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\chapter[Background in continuum mechanics for soft-tissue modelling][Background in continuum mechanics]{Background in continuum mechanics for soft-tissue modelling}
+\chapter{Background in continuum mechanics for soft-tissue modelling}
 \label{chap2}
@@
 \section{Description of motion} 	\label{chap2:descriptionMotion}
-Let us consider a body $\textbf{\textit{B}}$ of known geometry in a three-dimensional Euclidian space $\field{R}^3$.
+Let us consider a body $\textbf{\textit{B}}$ of known geometry in a three-dimensional Euclidean space $\field{R}^3$.
 For a given geometry and loading, $\textbf{\textit{B}}$ will undergo a set of macroscopic changes which is called \emph{deformation}.
@@
 This new configuration $\kappa$ may be designated by the \emph{current} or \emph{deformed configuration}.
-\OFF
 
@@
 In the Lagrangian description, the position and physical properties of the particles are referred to a reference configuration $\kappa_{R}$, often chosen to be the undeformed configuration $\kappa_{0}$.
-Consequently, the current coordinates ($\mathbf{x} \in \kappa$) depend on the reference coordinates ($\mathbf{X} \in \kappa_{0}$): \OFF
+Consequently, the current coordinates ($\mathbf{x} \in \kappa$) depend on the reference coordinates ($\mathbf{X} \in \kappa_{0}$):
 \begin{equation}
@@
 \end{equation}
-and a typical variable $\phi$ defined over the body is expressed in terms of the coordinates $\mathbf{X}$ and time $t$: \OFF
+and a typical variable $\phi$ defined over the body is expressed in terms of the coordinates $\mathbf{X}$ and time $t$: 
 \begin{equation}
@@
 \end{equation}
-Let us consider a particule $X$ whose position in the reference configuration is $\mathbf{X}$.
+Let us consider a particle $X$ whose position in the reference configuration is $\mathbf{X}$.
 From \eqref{chap2:phi}, we note that the value of $\phi$ associated with this fixed particle $X$ changes over time.
@@
 	\subsection{Eulerian description}
-In constrast, the Eulerian description is interested in changes at fixed locations. This time, the motion and $\phi$ are described with respect to the current position ($\mathbf{x} \in \kappa$): \OFF
+In contrast, the Eulerian description is interested in changes at fixed locations. This time, the motion and $\phi$ are described with respect to the current position ($\mathbf{x} \in \kappa$): 
 \begin{equation}
@@
 We now seek the relation existing between $d\mathbf{x}$ in the deformed configuration and $d\mathbf{X}$ of the reference configuration.
-\OFF
 
@@
 The determinant of $\mathbf{F}$ is called the \emph{Jacobian} and it is denoted by $J = \det \mathbf{F}$.
-Therefore, the relation between the deformed volume $ dv $ and the undeformed volume $ dV $ may be written as: \OFF
+Therefore, the relation between the deformed volume $ dv $ and the undeformed volume $ dV $ may be written as:
 \begin{equation}
@@
 \end{equation}
-Since this expression is valid for any vector $d\mathbf{L}$ and by using the property $\mathbf{a} \cdot \mathbf{b} = \mathbf{a} ^{T} \mathbf{b}$, the sought relation between an element of area in the reference configuration $d\mathbf{A}$ and its corresponding area $d\mathbf{a}$ in the deformed configuration may be expressed as: \OFF
+Since this expression is valid for any vector $d\mathbf{L}$ and by using the property $\mathbf{a} \cdot \mathbf{b} = \mathbf{a} ^{T} \mathbf{b}$, the sought relation between an element of area in the reference configuration $d\mathbf{A}$ and its corresponding area $d\mathbf{a}$ in the deformed configuration may be expressed as:
 \begin{equation}
@@
 
-In the following we make use of a Cartesian coordinate system and we consider an infinitesimal tetrahedron as shown on \fig{chap2:fig-CauchyStress}. \OFF
+In the following we make use of a Cartesian coordinate system and we consider an infinitesimal tetrahedron as shown on \fig{chap2:fig-CauchyStress}. 
 %
@@
 We define this stress tensor $\mathbf{S}$ as the \emph{second Piola-Kirchhoff stress tensor}.
-\OFF
 \begin{equation}
@@
 This relationship, which obviously depends on the type of material under consideration, is what we call a \emph{constitutive equation}.
-Constitutives equations are an attempt to mathematically characterise the behaviour of materials observed from experimental results.
-\OFF
+Constitutive equations are an attempt to mathematically characterise the behaviour of materials observed from experimental results.
 
@@
 	\subsection{Elasticity}	\label{chap2:elasticity}
-If under applied loads a material stores but does not dissipate energy, and it returns to its orginal shape when the loads are removed, we call that material \emph{elastic}.
+If under applied loads a material stores but does not dissipate energy, and it returns to its original shape when the loads are removed, we call that material \emph{elastic}.
 The work done by the stress is generally dependent on the deformation path.
@@
 \includegraphics[width=7cm]{chapter2/linear.pdf}
-\caption {The relationship between strain and stress is often non-linear (blue). However, it may be approximated as a linear relationship for small deformations (red). }
+\caption[Relationship between strain and stress]{The relationship between strain and stress is often non-linear (blue).
+However, it may be approximated as a linear relationship for small deformations (red).
+}
 \label{chap2:fig-linear}
@@
 Note that it does mean that the material is not elastic, it may well recover all its deformation after removal of the loads.
-However, materials often exibit an elastic threshold in the non-linear range after which permanent deformation occurs.
+However, materials often exhibit an elastic threshold in the non-linear range after which permanent deformation occurs.
 This range of deformation is qualified of \emph{plastic}.
@@
 \includegraphics[width=12cm]{chapter2/elastography_liver.pdf}
-\caption {Magnetic resonance elastography of the liver in a healthy volunteer and a patient with cirrhosis. Image courtesy of \cite{Talwalkar07}.}
+\caption[Magnetic resonance elastography of a liver]{Magnetic resonance elastography of the liver in a healthy volunteer and a patient with cirrhosis. Image courtesy of \cite{Talwalkar07}.}
 \label{chap2:fig-elastography_liver}
@@
 
-Yet, precised characterisation of soft tissue is not sufficient to achieve a good simulation.
+Yet, precise characterisation of soft tissue is not sufficient to achieve a good simulation.
 The development of a simplified or optimised version of the model suitable for real-time computation is also mandatory.

### 2010-10-24 Thesis/Chapter3.tex
- Update 24 October (final) -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@189 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{shortAbstract}
-The previous chapter explained in details how physics of anatomical structures could be described by continuun mechanics.
+The previous chapter explained in details how physics of anatomical structures could be described by continuum mechanics.
 When an organ is submitted to a force, the first effect of this application is a change in the organ's shape.
@@
 The second stage is when element equations are derived.
-As mentionned earlier, the unknowns of the problem are approximated using the idea that any continuous function can be representated by a linear combination of known functions and undetermined coefficients.
+As mentioned earlier, the unknowns of the problem are approximated using the idea that any continuous function can be represented by a linear combination of known functions and undetermined coefficients.
 Algebraic relations between the undetermined coefficients are then obtained by satisfying the governing equations over each element.
@@
 The use of the natural coordinate system has several advantages \citep{Biswas76}.
-Not only the interpolation functions can be derivated only once per type of element in the mesh, but their expression is also much simpler, regardless of the actual element shape.
+Not only the interpolation functions can be derived only once per type of element in the mesh, but their expression is also much simpler, regardless of the actual element shape.
 Consequently, the element equations and the derived element matrices get to be simplified too.
@@
 \end{equation}
-where $ \mathbf{H} $ is a $ 3 \times m $ matrix built from submatrices $ \mathbf{H}_i $
+where $ \mathbf{H} $ is a $ 3 \times m $ matrix built from sub-matrices $ \mathbf{H}_i $
 \begin{equation}
@@
 For this reason, the weak form is usually preferred over the strong form for obtaining an approximate solution of a practical engineering problem.
-\OFF
 
@@
 \end{equation}
-In other words, the global matrix is merely a rewritting on the whole domain of the element stiffness matrices, which are expressed with respect to local node numbers, in terms of global node numbers.
+In other words, the global matrix is merely a rewriting on the whole domain of the element stiffness matrices, which are expressed with respect to local node numbers, in terms of global node numbers.
 In addition to constructing the global mass and damping matrices (as shown in section \ref{chap3:wordOnMatrices}), this process leads to the following global system of equations:
@@
 
-\ON
 \paragraph*{Cholesky decomposition.}	\label{chap3:CholeskyDecompo}
@@
 Similarly to LU decomposition, a Cholesky decomposition may be used to break one set of equations into two successive but simpler systems.
-\OFF
 
@@
 
-\ON
-\paragraph*{Pre-conditioners}
+\paragraph*{Pre-conditioners.}
 Unfortunately, it is common for the matrix $ \mathbf{A} $ not to be well conditioned.
@@
 Although the formation and the study of pre-conditioners are out of the scope of this PhD thesis, pre-conditioners are a key element for solving systems of equations.
-To be effective, a preconditioner $ \mathbf{P} $ should be a good approximation of  $ \mathbf{A} $ and the following preconditioned system must be easier to solve than the original system \eqref{chap3:linearEquation}:
+To be effective, a pre conditioner $ \mathbf{P} $ should be a good approximation of  $ \mathbf{A} $ and the following preconditioned system must be easier to solve than the original system \eqref{chap3:linearEquation}:
 \begin{equation}
@@
 \end{equation}
-Two criteria should also be considered when forming a preconditioner: (1) the ease with which it may be evaluated and (2) its memory requirements.
+Two criteria should also be considered when forming a pre conditioner: (1) the ease with which it may be evaluated and (2) its memory requirements.
 The easiest pre-conditioner would be the identity since $ \mathbf{P}^{-1} = \mathbf{I} $ in this case.
 Of course, such a conditioner leads to the resolution of the original system and is therefore totally useless.
-In constrast, the choice of $ \mathbf{P} = \mathbf{A} $ leads to the solution of the system \eqref{chap3:linearEquation} in one single iteration.
+In contrast, the choice of $ \mathbf{P} = \mathbf{A} $ leads to the solution of the system \eqref{chap3:linearEquation} in one single iteration.
 In practice, an optimal pre-conditioner is to be found between these two extremes to minimise the number of iterations required by the solver while keeping the evaluation of the pre-conditioner simple.
@@
 Sparse approximations of LU or Cholesky decompositions may also be used as pre-conditioners (see sections \ref{chap3:LUdecompo} and \ref{chap3:CholeskyDecompo} for more details).
-\OFF

### 2010-10-24 Thesis/Chapter4.tex
- Update 24 October (final) -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@189 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\chapter[Modelling the deformation of solid objects in real-time][Modelling solid objects in real-time]{Modelling the deformation of solid objects in real-time}
+\chapter{Modelling the deformation of solid objects in real-time}
 \label{chap4}
@@
 Modelling the deformation of solid organs is a crucial problem in medical simulation.
-If the finite element method is classicaly used in mechanical engineering for its ability to solve the equations of continuum mechanics fairly accurately, its development in medical simulation was more progressive.
+If the finite element method is classically used in mechanical engineering for its ability to solve the equations of continuum mechanics fairly accurately, its development in medical simulation was more progressive.
 The main reason for this is that the finite element method is computationally demanding and not easily compatible with time constraints required in interactive simulation.
@@
 Obviously, the computation of a realistic force to be returned demands an appropriate mechanical modelling of all structures.
-In constrast of per-operative guidance, medical training demands real-time computations of the deformations.
+In contrast of per-operative guidance, medical training demands real-time computations of the deformations.
 
@@
 For instance, it was used by \cite{Miller88} to animate the motion of snakes and worms.
-It was also used by \cite{Ng-Thow-Hing97} in their muscle model where some of the springs were aligned with the muscle fibers and more recently by \cite{Magjarevic09} for modelling the mechanical behaviour of the myocardial tissue.
+It was also used by \cite{Ng-Thow-Hing97} in their muscle model where some of the springs were aligned with the muscle fibres and more recently by \cite{Magjarevic09} for modelling the mechanical behaviour of the myocardial tissue.
 More details on controlling anisotropy in mass-spring systems may be found in \cite{Bourguignon00}.
@@
 Indeed, topological changes require updates of this matrix, which is too computationally heavy to be done in real-time.
-\cite{Nienhuys01} avoid the precomputations by approximating the inverse at runtime with an iterative algorithm (Conjugate Gradient).
+\cite{Nienhuys01} avoid the pre-computations by approximating the inverse at runtime with an iterative algorithm (Conjugate Gradient).
 Moreover, instead of subdividing each tetrahedron to make the cut appear where the user performed it, they adapt the mesh locally so that there are always triangles on the scalpel path, and perform the cut along those triangles (see \fig{chap4:fig-cutUnderScalpel}).
@@
 \caption[Deformation of cylinders]{A geometrically linear finite element model (green) is compared to a co-rotational finite element model (yellow).
-While still using a simple linear strain measure, the co-rotational formulation is not sensitive to rotations and thus does not overdeform.}
+While still using a simple linear strain measure, the co-rotational formulation is not sensitive to rotations and thus does not over-deform.}
 \label{chap4:fig-cylinder}
@@
 
-To the best of our knowledge, \cite{Rumpf01} were the firsts to leverage the power of GPUs to solve partial differential equations.
+To the best of our knowledge, \cite{Rumpf01} were the first ones to leverage the power of GPUs to solve partial differential equations.
 They established a correspondence between common mathematical operations and OpenGL operations and structures.
@@
 Nodal points are only sprinkled locally around the surgical tool tip and not over the whole domain.
-The interpolation is performed by functions that are nonzero only on spheres surrounding the nodes.
+The interpolation is performed by functions that are non-zero only on spheres surrounding the nodes.
 A point collocation technique is used to generate the discrete equations.
@@
 They used a non-linear measure of strain with a linear constitutive model.
-They reported real-time simulation of models exibiting elastic, plastic, melting and flowing effects.
+They reported real-time simulation of models exhibiting elastic, plastic, melting and flowing effects.
 
@@
 This advantage of not having to go through the complex process of meshing was often claimed by meshless methods partisans.
-It also allows much easier cutting as it saves the cost of a computationally demanding remeshing around the cut.
+It also allows much easier cutting as it saves the cost of a computationally demanding re-meshing around the cut.
 Yet, in contrast of finite element methods, meshless techniques are young and still in development.
@@
 As a conclusion, if meshless methods are promising, the technique does not seem as mature as the finite element method for modelling soft tissues.
+
+\section{Conclusion}
+
+Modelling the deformation of solid organs is a crucial problem in medical simulation.
+If the finite element method is classically used in mechanical engineering for its ability to solve the equations of continuum mechanics fairly accurately, its development in medical simulation was more progressive.
+The main reason for this is that the finite element method is computationally demanding and not easily compatible with time constraints required in interactive simulation.
+Therefore, other methods were developed to allow faster computations, at the cost of larger approximations of course.
+Following the state of the art presented in this chapter, we can highlight a few trends in medical simulation.
+
+Firstly, over time (along with the increase in computational power) models used to describe anatomical structures became more elaborate and less approximative.
+Medical simulators become more concerned with the accuracy of their deformation.
+While they used to only cover a subset of their actual mechanical behaviour, simulators now began to reproduce most aspects of tissue behaviour under various types of loading (non-linearity, anisotropy and viscoelasticity) as described by \cite{Fung93}.
+
+Secondly, we observe a tendency towards implementations on parallel architectures, and GPUs in particular.
+Despite limited hardware possibilities, GPUs allow speedups up to orders of magnitude.
+Of course, such a gain is not straightforwardly obtained and often necessitates a substantial re-write of the algorithm to achieve an optimal result.
+But in environments with real-time constraints such as medical simulators, hardware acceleration through the use of GPUs is definitely an interesting path to follow.
+
+At last, we can identify two types of anatomical structures in human body.
+They can be either solid (brain, liver, prostate etc.) or hollow (colon, blood vessels, stomach etc.).
+Despite this assertion, it is worth mentioning that all the models presented so far were applied to solid organ modelling.
+Indeed, in the medical simulation community, much more attention is allotted to solid organs and the modelling of this type of structure is better understood.
+
+We recall that one of the challenge faced by the field of medical simulation is to model the deformation of organs in a physically realistic manner.
+In addition, this objective has to be achieved within heavy time constraints.
+To answer this problem for solid organs, our idea is to combine the two recent trends observed in medical simulation: more elaborate biomechanical models and GPU acceleration.
+Our contributions towards this issue will be presented in the two following chapters.
+Since specific approaches are required for simulating the deformation of thin anatomical structures, this problem will be tackled from chapter~\ref{chap7}.

### 2010-10-24 Thesis/Chapter5.tex
- Update 24 October (final) -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@189 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 As an example, for a given $ L_e $ a material with $E = 3\,000\, $Pa and $\nu = 0.45$ allows a time step of more than double that of a material with $\nu = 0.49$.
-Of course, lowering the Poisson ratio introduces inaccuracies, but for some applications (interactive medical simulation for instance) it may be that this is acceptable.
+Of course, lowering the Poisson's ratio introduces inaccuracies, but for some applications (interactive medical simulation for instance) it may be that this is acceptable.
 Finally, it must be highlighted that $ \Delta t_{cr}$ is proportional to $ L_e $.
@@
 
-Anisotropic mechanical response may arise, for example, from the presence of a highly organised microstructure such as those of connective tissues.
+Anisotropic mechanical response may arise, for example, from the presence of a highly organised micro-structure such as those of connective tissues.
 These are predominantly composed of collagen or elastin fibres embedded in an amorphous matrix \citep{Fung93}, and may be considered as fibre reinforced composites in some cases.
@@
 We address this problem by presenting a procedure similar to that developed by \cite{Poon98} for analysis of anisotropic linear viscoelastic models.
-This was also adapted by \cite{Taylor07} for solution of their fibre composite-based microstructural model.
+This was also adapted by \cite{Taylor07c} for solution of their fibre composite-based micro-structural model.
 In these cases the rate independent responses were based on linear elasticity, rather than a hyperelastic formulation suitable for large deformations.

### 2010-10-24 Thesis/Chapter6.tex
- Update 24 October (final) -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@189 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \caption [Multi-model representation in SOFA] {Multi-model representation in SOFA.
-\textbf{Left:} a Behaviour Model controls the other representations via a series of mappings.
-\textbf{Right:} examples of representations with a liver model.}
+The different representations are connected through a series of mappings (in red).
+Examples of representations with a liver model are also shown on the right.}
 \label{chap6:fig-SOFA}
@@
 It merely comes down to creating a single C++ class and changing the position reads and force writes to integrate the algorithm into SOFA's design.
-The precomputation phase takes place in the initialisation method where relevant variables are computed and passed to an external C function that allocates memory on the GPU and binds textures to it.
+The pre-computation phase takes place in the initialisation method where relevant variables are computed and passed to an external C function that allocates memory on the GPU and binds textures to it.
 During the simulation loop, the Solver requests the computation of the forces by launching the appropriate kernels on the GPU.
@@
 
-In the precomputation phase, nodes and elements are sorted by block to leave enough space for storage in shared memory of the element force contributions initially calculated at the end of the first kernel.
+In the pre-computation phase, nodes and elements are sorted by block to leave enough space for storage in shared memory of the element force contributions initially calculated at the end of the first kernel.
 We add nodes to a block along with their unique elements until the shared memory is saturated.
@@
 We imposed a ramped loading such that the displacements $u$ of the loaded nodes at time $t$ were given by $u = rt$, where $r$ is the loading speed, and we chose the $x$-direction to be the preferred material direction ($\mathbf{a}_0 = [1\, 0\, 0 ]$).
-Since the deformation is homogenous only a single element was required to reproduce the stresses exactly.
+Since the deformation is homogeneous only a single element was required to reproduce the stresses exactly.
 Moreover an analytical solution is available with which numerical results may be compared \citep{Taylor2009}.
@@
 \end{center}
-\caption[FE computed reaction forces on the end faces of the cube models over time.]{FE computed reaction forces on the end faces of the cube models over time.
+\caption[Finite element computed reaction forces on the end faces of the cube models over time.]{Finite element computed reaction forces on the end faces of the cube models over time.
 Curves are labelled according to the constitutive model used, where \emph{Anisotropic 1} refers to the transversely isotropic viscoelastic model with $\mathbf{a}_0 = \mathbf{a}_0^y$ and \emph{Anisotropic 2} refers to the case of $\mathbf{a}_0 = \mathbf{a}_0^{yz}$.
-Note that in this figure both markers and solid lines correspond to FE solutions.}
+Note that in this figure both markers and solid lines correspond to finite element solutions.}
 \label{chap6:fig-stressDecay}
@@
 \subfloat[ ]{\includegraphics[width=6.5cm]{chapter6/timingsGPU4.pdf}} \\
-\caption[GPU solution times]{(a), (b) show CPU and GPU solution times, respectively, for a single time step for each constitutive model as a function of model size, (c) shows the ratio of CPU to GPU solution times for model TIV, and (d) shows ratios of GPU solution times for the constitutive models: TIV to TIE, TIE to NHE, and TIV to NHE.
+\caption[GPU solution times for different constitutive models]{(a), (b) show CPU and GPU solution times, respectively, for a single time step for each constitutive model as a function of model size, (c) shows the ratio of CPU to GPU solution times for model TIV, and (d) shows ratios of GPU solution times for the constitutive models: TIV to TIE, TIE to NHE, and TIV to NHE.
 (b) also shows the critical time steps $ \Delta t_{cr} $ for each model size using liver-like and brain-like properties.}
@@
 Additionally it appears that solution times are little affected by the introduction of the more complex constitutive models, and importantly by the use of the developed constitutive update scheme.
-This is borne out in \fig{chap6:fig-timingsGPU}(d), where we observe that the maximum solution time ratio for model TIE to model NHE (anisotropy vs isotropy) was $1.013$, and that of model TIV to model TIE (viscoelastic vs elastic) was $1.043$.
+This is borne out in \fig{chap6:fig-timingsGPU}(d), where we observe that the maximum solution time ratio for model TIE to model NHE (anisotropy versus isotropy) was $1.013$, and that of model TIV to model TIE (viscoelastic versus elastic) was $1.043$.
 The largest total solution time increase for an anisotropic viscoelastic model compared with an isotropic hyperelastic one was $5.1\, \%$ We conclude that the key features of anisotropy and viscoelasticity may be included in simulations at very little additional computational cost.
@@
 \subfloat[ ]{\includegraphics[width=4cm]{chapter6/liver3.pdf}}
-\caption[GPU solution times]{Deformation of a liver model using the TI model (b) and the TIV model (c).
+\caption[Deformation of a liver using TI and TIV models]{Deformation of a liver model using the TI model (b) and the TIV model (c).
 The undeformed mesh is shown in (a).
@@
 To the best of our knowledge, this work constitutes the first GPU implementation of a non-linear, anisotropic and viscoelastic finite element procedure.
-This formulation was eventually implemented in the open source framework for medical simulation SOFA and will be released publically with the upcoming 1.0 release of SOFA.
+This formulation was eventually implemented in the open source framework for medical simulation SOFA and will be released publicly with the upcoming 1.0 release of SOFA.
 Over the course of my PhD, this work has led to three main publications \citep{Comas2008,Taylor2008,Taylor2009}.

### 2010-10-24 Thesis/Chapter7.tex
- Update 24 October (final) -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@189 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\chapter[Modelling the deformation of hollow objects in real-time][Modelling hollow objects in real-time]{Modelling the deformation of hollow objects in real-time}
+\chapter{Modelling the deformation of hollow objects in real-time}
 \label{chap7}
@@
 Another example is given by \cite{Aloisio04} who assume that the blood flow in the vessels may be considered as incompressible and the artery consequently becomes a solid body and can be modelled as a solid deformable object.
-If this approach can work, about a dozen of elements is required within the thickness of the structure for the system to be numerically stable \TODO{reference?}.
+If this approach can work, about a dozen of elements is required within the thickness of the structure for the system to be numerically stable.
 Consequently, many elements are needed to model thin structures and real-time performance is difficult to achieve.
@@
 \subfloat[Corrected]{\includegraphics[width=4.5cm]{chapter7/sheet3.pdf}}
-\caption{Deformation of a sheet hanging by two adjacent corners. Images courtesy of \cite{Provot95}.}
+\caption[Deformation of a sheet hanging by two adjacent corners]{Deformation of a sheet hanging by two adjacent corners. Images courtesy of \cite{Provot95}.}
 \label{chap7:fig-sheet}
@@
 
-\section{Objectives}
+\section{Conclusion}
 
-We seek to propose a solution for simulating, in real-time, the deformation of thin anatomical structures.
+The human body is composed of various deformable anatomical structures and finding appropriate soft-tissue models for all these type of structures is challenging.
+In the previous two chapters, we described an efficient non-linear FEM formulation for simulating the deformation of solid organs.
+However, the solid models previously mentioned are not appropriate and specific approaches are required for thin anatomical structures.
+In this chapter, we reviewed diverse techniques applied for the simulation of thin objects.
+It appears that few techniques derived from continuum mechanics have been proposed for modelling in real-time hollow structures in medical simulation.
+
+Therefore, we seek to propose a solution for simulating the deformation of thin anatomical structures in a physically realistic manner.
+In addition, this approach will have to be computable in real-time.
 Since \cite{Black91} showed that formulations with only membrane energy were likely to produce significant errors in the stress distribution, this model is required to feature bending.
@@
 
-In the next section, we will first introduce our co-rotational triangular shell finite element model and show how to efficiently apply this model to simulate the implant deployment in cataract surgery.
-In a second time, we will present a technique to mesh complex anatomical shapes with curved triangles, hence allowing us to model, in real-time, any anatomical structures with our co-rotational shell FEM in an optimal way.
+In the next chapter, we will first introduce our co-rotational triangular shell finite element model and show how to efficiently apply this model to simulate the implant deployment in cataract surgery (see chapter~\ref{chap8}).
+In a second time, we will present a technique to mesh complex anatomical shapes with curved triangles, hence allowing us to model, in real-time, any anatomical structures with our co-rotational shell FEM in an optimal way (see chapter~\ref{chap9}).
 

### 2010-10-24 Thesis/Chapter8.tex
- Update 24 October (final) -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@189 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2010-08-09 Thesis/Chapter8.tex
@@
-\chapter{Mass-spring model for colon implemented on GPU \OC{ISBMS 2008b}}
+\chapter{A co-rotational triangular shell FEM model}
 \label{chap8}
 \begin{shortAbstract}
-A short abstract for the upcoming chapter
+In the field of medical simulation, very few models have been proposed for simulating, in real-time, the deformation of thin anatomical structures.
+In fact, to our knowledge, no model based on the equations of continuum mechanics was described and applied to simulating the deformation of objects.
+In this chapter, we propose to rely on continuum mechanics to accurately describe the complex deformations of the implant.
+Our approach extends the co-rotational method used in finite element analysis of in-plane deformations to incorporate a bending energy.
+We will first describe the model in details before discussing its implementation in the open source framework SOFA.
+Lastly, this co-rotational shell FEM was applied to the simulation of the lens deployment during cataract surgery.
+This simulation also accounts for the complex contacts that take place during the injection phase.
 \end{shortAbstract}
 
-\section{Colon is big, we need a fast model}
+\section{Model description}
 
-\section{Limitations (difficult to link it to actual physical properties)}
+We propose to define a triangular shell element by combining a two-dimensional in-plane membrane energy, with an off-plane energy for describing bending and twist.
+To allow for real-time simulation, a computationally efficient formulation is needed.
+We therefore propose to extend the co-rotational idea introduced by \cite{Felippa00}.
+Indeed, as we have seen in the previous chapters, co-rotational approaches have been successfully applied to real-time simulation over the last few years.
+They offer a good trade-off between computational efficiency and accuracy by allowing small deformations but large displacements.
+We propose to improve and extend a plate model first introduced by \cite{Przemieniecki85} to a co-rotational formulation.
+Once combined with an in-plane membrane formulation we obtain an accurate, yet computationally efficient, shell finite element method featuring both membrane and bending energies.
+
+\subsection{Triangular elastic membrane}
+
+The computation of the triangular elastic membrane stiffness matrix can be derived from previous works dealing with tetrahedral co-rotational elements \citep[like][for instance]{Muller04b}.
+The element stiffness matrix $\textbf{K}_e$ can be computed as follows:
+%
+\begin{equation}
+\textbf{K}_e = \int_v \textbf{J} \boldsymbol\chi \textbf{J}^{T} dV,
+\end{equation}
+where $\textbf{J}$ is the strain-displacement matrix and $\boldsymbol\chi$ embodies the material's behaviour.
+The implant is very stiff and we therefore assume that the local deformations remain limited during the deployment and a linear constitutive law is sufficient.
+Thus in the simple case of Hooke's law we have:
+%
+\begin{equation}
+\boldsymbol\chi = \frac{E}{12(1-\nu^2)}
+\begin{bmatrix}
+1 & \nu & 0 \\
+\nu & 1 & 0 \\
+0 & 0 & \frac{1}{2} (1-\nu)
+\end{bmatrix}
+\end{equation}
+
+The stiffness matrix in the global frame is eventually obtained using the rotation matrix of the element: $\textbf{K} = \textbf{R} \textbf{K}_e \textbf{R}^{T}$ where $\textbf{R}$ describes the rotation of the (triangular) element with respect to its initial configuration.
+
+\subsection{Triangular plate bending}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[height=4cm]{chapter8/bending.pdf}
+\caption[The different degrees of freedom of a triangular thin plate in bending]{The different degrees of freedom $u$ of a triangular thin plate in bending.}
+\label{chap8:fig-triangle}
+\end{figure}
+%
+To calculate the stiffness matrix for the transverse deflections and rotations shown on \fig{chap8:fig-triangle}, we start with the assumed deflection $u_z$ of the form
+\begin{equation}
+ u_z = c_1 + c_2x + c_3y + c_4x^2 + c_5xy + c_6y^2 + c_7x^3 + c_8xy^2 + c_9y^3,
+\label{chap8:eq-deflection}
+\end{equation} 
+where $c_1$, \ldots , $c_9$ are constants.
+Equation~\ref{chap8:eq-deflection} solves an issue of symmetry which was observed with the deflection function proposed by \cite{Przemieniecki85}.
+These constants can be evaluated in terms of the displacements and slopes at the three corners of the triangular plate using
+\begin{equation}
+\textbf{u} = \textbf{Cc}
+\label{chap8:eq-U}
+\end{equation} 
+where $\textbf{u} = \left\{u_1 u_2 \ldots u_9 \right\} $ and $\textbf{c} = \left\{c_1 c_2 \ldots c_9 \right\} $. Matrix $\textbf{C}$ derives from \eqref{chap8:eq-U}:
+\begin{equation}
+C = 
+	\begin{bmatrix}
+	1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+ 	0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
+	0 & -1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
+	1 & x_2 & 0 & x_2^2 & 0 & 0 & x_2^3 & 0 & 0 \\
+	0 & 0 & 1 & 0 & x_2 & 0 & 0 & 0 & 0 \\
+	0 & -1 & 0 & -2x_2 & 0 & 0 & -3x_2 & 0 & 0 \\
+	1 & x_3 & y_3 & x_3^2 & x_3y_3 & y_3^2 & x_3^2 & x_3y_3^2& y_3^3 \\
+	0 & 0 & 1 & 0 & x_3 & 2y_3 & 0 & 2x_3y_3 & 3y_3^2 \\
+	0 & -1 & 0 & -2x_3 & -y_3 & 0 & -3x_3^2 & -y_3^2 & 0
+	\end{bmatrix}
+\end{equation}
+
+We used the following notations:
+\begin{align}
+u_1 &= (u_z)_{x_1,y_1}, & u_2 &= \left(\frac{\partial u_z}{\partial y}\right)_{x_1,y_1}, & u_3 &= - \left(\frac{\partial u_z}{\partial x}\right)_{x_1,y_1} \notag \\
+u_4 &= (u_z)_{x_2,y_2}, & u_5 &= \left(\frac{\partial u_z}{\partial y}\right)_{x_2,y_2}, & u_6 &= - \left(\frac{\partial u_z}{\partial x}\right)_{x_2,y_2} \label{chap8:ui} \\
+u_7 &= (u_z)_{x_3,y_3}, & u_8 &= \left(\frac{\partial u_z}{\partial y}\right)_{x_3,y_3}, & u_9 &= - \left(\frac{\partial u_z}{\partial x}\right)_{x_3,y_3} \notag \\
+\end{align}
+
+\noindent We can calculate the strains from the flat-plate theory using:
+\begin{equation}
+e_{xx} = -z \frac{\partial^2u_z}{\partial x^2}
+\end{equation} 
+\begin{equation}
+e_{yy} = -z \frac{\partial^2u_z}{\partial y^2}
+\end{equation} 
+\begin{equation}
+e_{xy} = -2z \frac{\partial^2u_z}{\partial x \partial y}
+\end{equation}
+
+\noindent Hence, using the above equations and \eqref{chap8:eq-deflection}, we have
+\begin{equation}
+\begin{bmatrix}
+e_{xx} \\
+e_{yy} \\
+e_{xy}
+\end{bmatrix}
+= 
+-z
+\begin{bmatrix}
+0 & 0 & 0 & 2 & 0 & 0 & 6x & 0 & 0 \\
+0 & 0 & 0 & 0 & 0 & 2 & 0 & 2x & 6y \\
+0 & 0 & 0 & 0 & 2 & 0 & 0 & 4y & 0 \\
+\end{bmatrix}
+\textbf{c}
+\label{chap8:eq-strains}
+\end{equation} 
+or symbolically $\textbf{e} = \textbf{Dc}$ where $\textbf{D}$ stands for the $3\times 9$ matrix in \eqref{chap8:eq-strains}, including the pre-multiplying constant $-z$.
+Noting from \eqref{chap8:eq-U} that $\textbf{c} = \textbf{C}^{-1}\textbf{u}$, we have:
+\begin{equation}
+\textbf{e} = \textbf{DC}^{-1}\textbf{u} = \textbf{bu},
+\end{equation} 
+where $\textbf{b} = \textbf{DC}^{-1}$.
+
+\noindent Knowing that the stiffness matrix $\textbf{K}_e$ for an element is obtained from
+\begin{equation}
+\textbf{K}_e = \int_v \textbf{b}^{T} \boldsymbol\chi \textbf{b} dV,
+\end{equation} 
+where $\boldsymbol\chi$ is the material matrix, the substitution of $\textbf{b}$ into this expression yields
+\begin{equation}
+\label{chap8:Ke}
+\textbf{K}_e = (\textbf{C}^{-1})^T \int_v \textbf{D}^{T} \boldsymbol\chi \textbf{D} dV \, \textbf{C}^{-1}.
+\end{equation}
+
+Similarly to the membrane stiffness matrix, the bending stiffness matrix in the global frame is eventually obtained using the rotation matrix of the element: $\textbf{K} = \textbf{R} \textbf{K}_e \textbf{R}^{T}$.
+While the method to numerically compute the integral will be reviewed section \ref{chap8:numericalIntegration}, we first carry on with the theory by introducing how the contacts are handled.
+
+\section{Mechanical interactions with the curved surface of shells}	\label{chap8:interactions}
+
+The practical interest of modelling complex behaviours such as bending and twisting would remain fairly low for medical simulation if contacts and constraints were not handled properly.
+In our case the difficulty comes from different sources.
+First the collision detection must be carried out with the curved surface of shell elements as opposed to the classic detection on plane triangles.
+Then forces applied to a given triangle need to be distributed between linear forces and torques onto its three vertices.
+As we will see, the same polynomial interpolation function chosen to compute the bending energy in our FEM formulation is also used to capture the interactions between the curved surface and other objects.
+
+In order to detect the collision with the bent surface, we have chosen the subdivision approach.
+We first sample the flat surface of each element by recursively dividing each triangle into four smaller ones and the deflection of each new vertex is computed using \eqref{chap8:eq-deflection} according to the displacements and slopes at the three vertices of the triangular element.
+This process of subdivision may also allow us to render each shell as a curved triangle (\fig{chap8:fig-shell} (a) and (b)) and detect any collision with the curved surface of the shell using any of the classic collision detection algorithms working on flat triangles.
+
+Once a collision has been detected, it must be processed by distributing the linear force received on the bent surface between the three vertices of the triangle.
+First the linear part of the force is simply transmitted on each node using the barycentric coordinates of the contact point's projection onto the triangle.
+
+The main difficulty is to convert the normal component of the force applied to the bent surface into a torque at each of the three nodes (\fig{chap8:fig-shell} (c)).
+Our approach is the following: during force computation, we use the change in orientation measured at each node to compute the local deflection of each subvertex within the triangle.
+Differentiating the formulation twice yields a relation between the torque applied at each node and the generated force in bending.
+We therefore need to invert the latter formulation to convert a bending force into torques at each vertex.
+Thus, we are able to transmit any force coming from interactions with the curved surface of shells to the mechanical vertices used in our FEM formulation.
+%
+\begin{figure}[ht]
+\centering
+\includegraphics[width=12cm]{chapter8/shell_curvature.pdf}
+\caption[Subdivision of the shell for rendering and collision detection]{(a)  The triangle formed by the three vertices of the shell has been recursively subdivided 3 times and the deflection of each new vertex was computed according to the same deflection function used in our shell FEM.
+(b) Sampling the actual surface of the shell allows more accurate rendering and collision detection.
+(c) The shape function is used to distribute an external force $F$ onto the triangle nodes.}
+\label{chap8:fig-shell}
+\end{figure}
+
+\section{Implementation}
+
+The co-rotational shell formulation has been successfully implemented as a Force Field component into the open-source framework SOFA \citep{Allard07}.
+A description of SOFA's architecture was given in section \ref{chap6:SOFA}.
+We will now review the key points of the implementation.
+
+	\subsection{Evaluation of displacements}
+The computation of the stiffness matrix, and subsequently the internal forces, starts with the measure of the displacements $\left\{u_1 u_2 \ldots u_9 \right\} $ for every element.
+As defined by \eqref{chap8:ui}, we need to measure the displacement along the z axis and the rotations around local x and y axes for each vertex.
+For \cite{Przemieniecki85}, the deflection z is measured in the global frame.
+But in a co-rotational formulation, a local frame is attached to each triangle and used to compute displacements and forces.
+In this local frame (formed from the 3 vertices) the deflection is therefore always null (hence, $ u_1, u_4, u_7 $ are null).
+To evaluate the remaining rotations, we first associate a quaternion at each vertex which embodies its frame (or its orientation).
+In mathematics, quaternions are an extension of complex numbers and may be written as follows:
+\begin{equation}
+q = w + x i + y j + z k \quad \mbox{with } i^2 = j^2 = k^2 = -1.
+\end{equation}
+For use in geometry, quaternions are often written as a sum of a scalar and a vector:
+\begin{equation}
+q = w + (x, y, z) = \cos (\frac{\alpha}{2}) + \mathbf{u} \sin (\frac{\alpha}{2}),
+\end{equation}
+where $ \mathbf{u} $ is a unit vector and $ \alpha $ a scalar.
+If we denote by $ \mathbf{v} $ a vector in $\field{R}^3$, it can be shown that the quaternion product $ q \mathbf{v} q^{-1} $ yields a vector which is the rotation of $ \mathbf{v} $ by an angle $ \alpha $ around the axis $ \mathbf{u} $.
+Quaternions require less storage than rotation matrices and are known to be more numerically stable.
+
+From there, our approach to measure the rotations around local x and y axes is simple.
+At runtime and for each vertex, we compute the rotation that we need to transform the z axis of the triangle's frame into the z axis of the vertex's frame.
+The differences with the same quantities computed for the rest position during the initialisation lead to the desired displacements.
+% (see listing \ref{chap8:qDiffZ})
+%\begin{lstlisting}[caption={[Method to compute the rotation between two quaternions]Method used to compute the rotation of the z axis between the frames embodied by two quaternions},label={chap8:qDiffZ},frame=shadowbox,rulesepcolor=\color{black}]
+%inline Quat qDiffZ(const Quat& Qvertex, const Quat& Qtriangle)
+%{
+%    // dQ is the quaternion that embodies the rotation between 
+%    // the z axis of the vertex and the z axis of the local 
+%    // triangle's frame (in local space)
+%    Quat dQ;
+%
+%    // u = z axis of the triangle's frame (in local space)
+%    Vec3d u(0,0,1);
+%
+%    // v = z axis of the vertex's frame is expressed into 
+%    // world space
+%    Vec3d v = Qvertex.rotate(Vec3d(0.0, 0.0, 1.0));
+%    
+%    // v0 = v expressed into local triangle's frame
+%    Vec3d v0 = Qtriangle.rotate(v);
+%
+%    // Axis of rotation between the 2 vectors u and v lies 
+%    // into the plan of the 2 vectors
+%    Vec3d axis = cross(u, v0);
+%    
+%    // Shortest angle between the 2 vectors
+%    double angle = acos(dot(u, v0));
+%
+%    // Quaternion associated to this axis and this angle
+%    if (fabs(angle)>1e-6)
+%    {
+%        dQ.axisToQuat(axis,angle);
+%    }
+%    else
+%    {
+%        dQ = Quat(0,0,0,1);
+%    }
+%
+%    return dQ;
+%}
+%\end{lstlisting}
+
+%	\subsection{Inversion of matrix $ \mathbf{C} $}
+%\cite{Przemieniecki85} described the condition of non-inversion for the matrix $\mathbf{C}$ as the following:
+%\TODO{add condition of inversion... but need the book :)}
+%
+%Each triangle of a mesh obtained from a regular grid of points fits this condition of non-inversion (right triangles).
+We fixed this limitation by choosing a different local coordinate system that the one introduced by \citeauthor{Przemieniecki85}.
+%
+%\TODO{add figure with the two different coordinate system}
+
+	\subsection{Numerical integration} \label{chap8:numericalIntegration}
+In practice, the integration of the integral in \eqref{chap8:Ke} must be carried out numerically using Gaussian quadrature, a numerical technique which allows to calculate the value of an integral.
+The Gaussian quadrature states that the integral of a function $ f $ over the area $ A $ of a triangular surface may be evaluated as a weighted sum of function values at specific points within the area of integration \citep{Cowper73}.
+In other words, we have
+\begin{equation}
+\int_{A} f dA = A \sum_{i=1}^n w_i f_i,
+\end{equation}
+where $ w_i $ is a weight factor and $ f_i $ is the evaluation of $ f $ at the integration point $ i $ (also called Gauss points).
+The choice of the number and the locations of the integration points depends on the order of accuracy desired.
+We classically choose 3 integration points located at the middle of each edge of the triangle, for which the weight factors $ w_i $ are $ 1/3 $.
+The required integration over the volume is achieved by multiplying the result by the thickness $ t $ of the shell.
+Therefore, \eqref{chap8:Ke} may be numerically evaluated as
+\begin{equation}
+\textbf{K}_e = \dfrac{A t}{3} (\textbf{C}^{-1})^T \sum_{i=1}^3 \textbf{D}_i^{T} \boldsymbol\chi \textbf{D}_i  \, \textbf{C}^{-1},
+\end{equation}
+where $ D_i $ is the evaluation of the matrix $ \mathbf{D} $ defined by \eqref{chap8:eq-strains} at each Gauss point $ i $.
+
+\bigskip
+
+\subsection{Computation of stiffness matrix: summary}	\label{chap8:summary}
+In practical terms, the different computations associated with each triangular shell element can be described as follows:
+%
+\begin{enumerate}
+\item Compute the rotation matrix $\textbf{R}$ from global to triangle (local) frame
+\item Compute the local displacement vector $\textbf{u} = \{v_1, v_2, 0, u_2, u_3, v_3, v_4, 0, u_5, u_6,$ $v_5, v_6, 0, u_8, u_9 \} $ for each triangle, where $ (v_1, v_2), (v_3, v_4) and (v_5, v_6) $ are the in-plane displacements on x and y local axes and $ u_i $ were defined by \eqref{chap8:ui}.
+As we are in a co-rotational framework, we recall that the normal displacements $u_1, u_4, u_7$ at each of the three nodes are null in the local frame of the triangle.
+\item Compute matrix $\textbf{D}_i$  at each Gauss point $i$
+\item The strain-displacement matrix at each Gauss point $i$ is computed with $\textbf{J}_i = \textbf{D}_i \textbf{C}^{-1}$
+\item Compute the local stiffness matrix $\textbf{K}_e$ of the element as $\textbf{K}_e = \displaystyle{\sum^3_{i=1}} \textbf{J}_i \boldsymbol\chi \textbf{J}_i^T$
+\item Transform the local element stiffness matrix into the global frame and add it to the global stiffness matrix
+\end{enumerate}
+
+	\subsection{Concept of mappings in SOFA}	\label{chap8:mappings}
+Let us explain the concept of \emph{mappings} and \emph{mechanical mappings} in SOFA.
+In SOFA, the state of a simulated system can be described by the values and time derivatives of its independent degrees of freedom (DOF) gathered in two vectors $x_0$ and $v_0$.
+Geometry can be attached to the DOF for visualisation or contact computation.
+We call it the \emph{shape}.
+It is typically defined by points, such as triangle vertices or sphere centres, and additional data such as triangle connectivity or sphere radii.
+We call these points the vertices.
+Their positions, velocities and associated forces are stored in vectors $x_1$ , $v_1$ and $f_1$, respectively.
+They are not independent variables, since the positions and velocities are bound to the DOF using kinematic operators which we call the \emph{mappings}:
+\begin{eqnarray*}
+x_1 &=&\JNL_1(x_0)\\ 
+v_1 &=& J_1 v_0
+\end{eqnarray*}
+When the vertices and the DOF are the same, the mapping is the identity.
+Matrix $J_1 = \frac{\partial x_1}{\partial x_0}$ encodes the linear relation between the DOF velocities and the shape velocities.
+Due to linearity, the same relation holds on elementary displacements $dx$.
+It also holds on accelerations, with an additional offset due to velocities when the position mapping \JNL is non-linear.
+In most cases, operators \JNL~ and \J~ are the same, but in the case of rigid bodies, \JNL~ is non-linear with respect to $x$ and it can not be written as a matrix.
+
+When shapes collide, additional geometry can be necessary to model the contact.
+For instance, when an edge intersects another one, a contact force is applied to the intersection points.
+These points are defined by their barycentric coordinates with respect to their edge vertices.
+Other relations can be used, depending on the kind of geometrical primitives in contact.
+This additional geometry requires another geometrical layer connected to the shape by a mapping, as illustrated in \fig{chap8:fig-mappings}.
+
+\begin{figure}[ht]
+	\begin{center}
+ 		\includegraphics[width=13cm]{chapter8/mappings}
+ 	\end{center}
+ \caption[Concept of mappings in SOFA]{Mappings from the DOFs to a contact point.
+Top: two simulated objects in contact (red point).
+Bottom: hierarchy of geometrical layers.
+Positions and velocities are propagated top-down.
+The contact force $f_c$ is accumulated in the contact layers.
+Forces are then propagated bottom-up.
+Image courtesy of SOFA's documentation.
+}
+\label{chap8:fig-mappings}
+\end{figure}
+
+Positions and velocities can be propagated top-down through our layer hierarchy using the relations presented.
+In order to take the contact forces into account in the dynamics equation, we have to convert the contact forces applied to the contact points to forces applied to the DOF, where our mechanical model is applied.
+This requires an extension to the position and velocity mappings presented: we call it \emph{mechanical mapping}.
+This is a general method to propagate forces bottom-up through the layers of the geometrical hierarchy.
+Given forces $f_n$ applied to a geometry layer $n$, we derive the equivalent force applied to its parent layer $n-1$.
+Equivalent forces must have the same power.
+Thus, we have to compute $f_{n-1}$ such that:
+$$
+v_{n-1}^T f_{n-1} = v_n^T f_n
+$$
+The relation $v_{n} = J_{n}v_{n-1}$ allows us to rewrite the previous equation as
+$$
+v_{n-1}^T f_{n-1} = v_{n-1}^T J_{n}^T f_n
+$$
+Since this relation must hold for any velocity $v_{n-1}$, we simplify it and get
+\begin{equation} \label{eq:mapF}
+f_{n-1} = J_n^T f_n
+\end{equation}
+This corresponds to the principle of virtual work.
+
+\subsection{Visualisation}
+Because shells have a curved surface, a specific rendering technique is required to visualise the deformation in bending.
+Without any consideration, the shells would appear flat (as simple triangles).
+Our idea is to make use of a higher resolution mesh whose deformation would be controlled by the underlying physical mesh of shells.
+In other words, a mapping was implemented to connect the DOFs of our mechanical model to a set of vertices useful for rendering.
+In SOFA, a mapping has only 2 methods: init(), which allows computations during the initialisation phase, and apply() called at runtime and in charge of updating the positions of the high resolution mesh according to the coarser mesh constituted by the shell elements.
+
+\begin{itemize}
+\item \textbf{init()} \\
+During the initialisation phase, each vertex from the high resolution mesh is projected onto the coarse mesh.
+Thus, for each vertex we find the closest primitive (DOF, edge or triangle) on the coarse mesh and we take the list of triangles attached to it.
+Then, the barycentric coordinates of the projected point is computed within each triangle of this list.
+%
+\item \textbf{apply()} deals with the relation between the DOFs and the vertices used for rendering (shape vertices).
+The position of each vertex from the high resolution mesh must be updated according to the underlying mesh.
+After the initialisation phase, we have associated each rendering vertex to a triangle and its barycentric coordinates within this triangle are known.
+Therefore, we first start by computing the new position of the rendering vertex by weighting the coordinates of each triangle's vertex with the associated barycentric weight.
+Hence, we obtain a new position in the plane of the triangle.
+If we were to stop the procedure here, the shell elements would still be rendered as flat triangles since all rendering vertex are located in the plane of the triangles.
+Consequently, the additional step of computing the deflection is required.
+We know that the deflection at any point of local coordinates $(x, y)$ may be computed from $u_z = c_1 + c_2x + c_3y + c_4x^2 + c_5xy + c_6y^2 + c_7x^3 + c_8xy^2 + c_9y^3$ as described by \eqref{chap8:eq-deflection}.
+Hence, we calculate the local coordinates $(x, y)$ of each rendering vertex within its associated triangle.
+Moreover, we notice that the coefficients $ \mathbf{c} = \left\{c_1 c_2 \ldots c_9 \right\} $ may be evaluated using \eqref{chap8:eq-U} from
+\begin{equation}
+\mathbf{c} = \mathbf{C}^{-1} \mathbf{u},
+\end{equation}
+The deflection may therefore be computed for each vertex and added to the in-plane position previously calculated. 
+\end{itemize}
+
+It is worth noting that a high resolution mesh may be obtained by recursively subdividing each triangle.
+However, in this case the vertices are closely related to the structure of the mechanical mesh and rendering artefacts may appear.
+As we can notice on \fig{chap8:fig-rendering}, the quality of the visualisation may be improved by using a totally different mesh.
+%
+\begin{figure}[ht]
+\centering 
+\subfloat[Subdivided triangles]{\includegraphics[width=5.5cm]{chapter8/sub_grid_red.png}}
+\hspace{1cm}
+\subfloat[Irregular grid]{\includegraphics[width=5.5cm]{chapter8/irregular_grid_red.png}} \\
+\subfloat[Subdivided triangles]{\includegraphics[width=5.5cm]{chapter8/sub_grid_mesh.png}}
+\hspace{1cm}
+\subfloat[Irregular grid]{\includegraphics[width=5.5cm]{chapter8/irregular_grid_mesh.png}}
+\caption[Visual artifacts using triangle subdivisions]{In some situations, some rendering artifacts may appear when we use subdivided triangles to render the shells (a,c).
+Using an irregular mesh which is not closely related to the structure of the mechanical mesh reduces the impact (b,d).}
+\label{chap8:fig-rendering}
+\end{figure}
+
+	\subsection{Contacts with the curved surface of shells}
+If a specific rendering technique was developed to visualise the curved shape of the shells, it is also essential to interact with it.
+Indeed, as explained section \ref{chap8:interactions}, the force applied onto the curved surface of the shell must be processed by distributing the linear force received on the bent surface between the three vertices of the triangle.
+Consequently, the mapping previously described is extended to a mechanical mapping and two additional methods need to be implemented: applyJ and applyJT.
+%
+\begin{itemize}
+\item \textbf{applyJ()} encodes the linear relation between the DOF velocities and the shape velocities as described in section \ref{chap8:mappings}.
+applyJ() basically embodies the same computations than apply() except that the vector of displacement $\mathbf{u}$ is different.
+Whereas the local displacement in rotation is measured from the quaternions to create the vector $\mathbf{u}$, this time the vector is directly obtained from the angular velocities we had at each vertex (once converted into the local frame).
+We therefore have a relation between the angular velocities at each vertex and the velocity of the deflection in the direction normal to each triangle.
+\item \textbf{applyJT()} must convert the contact forces applied to the contact points to forces applied to the DOF.
+If the linear forces are easily distributed between DOFs using barycentric coordinates, it also encodes the relation between the vertical force $ \mathbf{F} $ applied onto the bent surface of the plate and the torque at each DOF.
+We start by retrieving the normal component of the applied force vector $F_{\mathrm{z}}$.
+We project the application point of the force into the triangle's plane and compute its local coordinates $(x,y)$.
+We create the polynom $P$ such as
+%
+\begin{equation}
+P = F_{\mathrm{z}}(1 \hspace{0.2cm} x \hspace{0.2cm} y \hspace{0.2cm} x^2 \hspace{0.2cm} xy \hspace{0.2cm} y^2 \hspace{0.2cm} x^3 \hspace{0.2cm} xy^2 \hspace{0.2cm} y^3)^T. 
+\end{equation}
+%
+The moments at each vertex are then obtained with 
+\begin{equation}
+\Omega = (\textbf{C}^{-1})^T P.
+\end{equation}
+\end{itemize}
+
+\subsection{Allowing a curved rest configuration}
+In \cite{Przemieniecki85}, the rest and initial configurations are both assumed to be the flat one.
+The only way to have an initial deformed shape is to apply a torque on the vertices of the shells.
+However, in this configuration we would compute a displacement vector $ \mathbf{u} $ that is not null, which leads to a bending energy.
+In order to define a curved rest configuration for our shells, we propose to compute the bending energy between the initial and rest configuration during the initialisation phase, and add the energy between the current and initial configurations at runtime.
+Consequently, by separating initial and rest configurations, it is possible to have an initial deformed shape without any initial bending energy.
+In other words, we shifted the zero energy configuration.
+
+\subsection{Possible extension: parallel implementation on GPU}	\label{chap8:GPU}
+If we look closely at all the steps required to compute the local stiffness matrix for each element (see section \ref{chap8:summary} for a summary), we can notice that the operations carried out for a given shell element are independent of the other elements.
+This feature makes our co-rotational shell FEM implementation highly parallelisable.
+One potential problem however, is the assembly of the global stiffness matrix.
+Indeed, in a parallel implementation, many elements are processed concurrently (that is, at the same time).
+Inevitably, two elements sharing a node may be processed in the same time.
+Therefore, they may try to both write their element contribution into the exact same location in the global stiffness matrix, and therefore the same location in GPU's memory.
+As with the TLED implementation described in chapter \ref{chap6}, there is two solutions.
+The first one would be to re-order the elements organisation so that no concurrent writes will occur.
+Unfortunately, this re-organisation of the element reduces the number of elements that can be processed in parallel.
+The second solution would be to take advantage of a new GPU NVIDIA architecture called Fermi, which added the capability atomic writes for floats.
+We recall that the write operation is said to be atomic in the sense that it is guaranteed to be performed without interference from other threads.
+In other words, the concurrent writes are simply serialised in hardware.
+Obviously, an additional overhead is to be expected.
+
+Moreover, because shells are high-order (curved) elements, fewer elements are needed to describe a given geometry with the same precision (as compared to using flat triangles).
+Generally speaking, this constitutes a decisive computational advantage since fewer elements are required.
+However, GPUs are more efficient when they tackle numerous but simple computations.
+Hence, a good speedup over sequential CPU implementations is usually only achieved when the number of elements involved becomes great enough.
+In contrast, the computations for each shell element are quite heavy and the number of elements used in the whole simulation is limited.
+Therefore, the conditions are not optimal to obtain maximum speedups on GPU.
+A straightforward GPU implementation of our shell FEM formulation where each thread deals with the calculations of one element would not make use of all the GPU capacity (the occupancy would not be maximal).
+If a reasonable improvement over the CPU implementation will probably be obtained nevertheless, the gain is likely to be smaller than in the case of the TLED algorithm for instance.
+Other approaches will need to be investigated to take full advantage of the GPU.
+For instance, computations within an element could be distributed between several threads to increase the occupancy of the GPU and hence the overall efficiency.
+
+To conclude on a possible parallelisation of the algorithm, an optimal GPU implementation is always a challenge to achieve.
+One needs to choose wisely the type of memory to be used for each variable and optimise the access patterns.
+While it seems that there is no major obstacle for a simple GPU implementation, the best speedups will only be attained through more elaborate techniques of parallelisation making full usage of the GPU.
+Although it will take some time, our co-rotation shell FEM will eventually be ported to GPU.
+
+\section{Comparison with an analytical result}
+
+We compared our model with some theoretical results reported by \cite{Zhongnian86} to assess its quality in modelling bending.
+The test that we carried out uses a square shape mesh clamped on all four edges.
+A uniform load is then applied on the square and the maximum deflection $z_{max}$ at the centre can be calculated.
+Several simulations are performed with increasing load values $q$ (ranging from $1$ to 5 N/m$^2$) and the following parameters were used:
+\begin{itemize}
+ \item Young Modulus $E = 1.092 \times 10^6\,$Pa
+ \item Poisson's ratio $\nu = 0.3$
+ \item edge of the square $L = 10\,$m
+ \item thickness $h = 0.1\,$m
+ \item pressure q is a uniformly distributed load per unit area
+\end{itemize}
+
+Using these particular values it can be shown that $z_{max} = 0.126\,q$.
+The maximum deflection obtained in our simulations are reported in Table~\ref{chap8:tab-results}.
+In average we found $z_{max} = 0.1248\,q $, resulting in a $0.93\,\%$ error between our model and theoretical results on that test.
+%
+\begin{table}[ht]
+	\centering
+	\begin{tabular}{p{9cm}|c|c|}
+	\cline{2-3}
+	\multirow{5}{*}{\includegraphics[height=3.5cm]{chapter8/clamped_square.pdf}} & $q$ & $z_{max}$ \tabularnewline
+	\cline{2-3}
+	& $\,1\,$ & $\, 0.1218 \,$ \tabularnewline
+	& $\,2\,$ & $\, 0.2475 \,$ \tabularnewline		
+	& $\,3\,$ & $\, 0.3747 \,$ \tabularnewline	
+	& $\,4\,$ & $\, 0.5050 \,$ \tabularnewline		
+	& $\,5\,$ & $\, 0.6374 \,$ \tabularnewline
+	\cline{2-3}
+	\end{tabular}
+	\vspace{1cm}
+	\caption[Test of our shell model with the bending of a square plate]{Comparison of our shell model with theoretical results in the case of the bending of a square plate.
+An error of less than $1\,\%$ was found between our simulation and theoretical results.}
+	\label{chap8:tab-results}
+\end{table}
+
+\section{Application to implant deployment simulation in cataract surgery}
+
+Cataract surgery consists in three main steps: capsulorhexis, phacoemulsification, and implantation of an intra-ocular lens.
+Prior to starting the surgery, a viscoelastic fluid is introduced into the capsule to facilitate capsulorhexis creation and provide protection during phacoemulsification.
+This fluid remains in the capsule for the duration of the surgery, including the injection of the implant.
+Capsulorhexis is the technique used to remove a part of the anterior lens capsule.
+Phacoemulsification consists in using a surgical device which tip vibrates at an ultrasonic frequency to emulsify the natural lens material and then aspirate the fragments of the cortical material.
+After the removal of the diseased lens, an intra-ocular lens is implanted into the eye, through a small incision (about $2\,$mm) using a foldable intra-ocular lens (see \fig{chap8:fig-surgery}).
+The foldable lens, usually made of acrylic material, is then implanted within the lens capsule through the incision used during phacoemulsification.
+In some cases the implant can be flipped or  the hooks (also called \emph{haptics}) can break.
+Therefore the simulation of the insertion and deployment of the implant is crucial for achieving a successful surgery.
+%
+\begin{figure}[ht]
+\centering 
+\subfloat[ ]{\includegraphics[height=3.2cm]{chapter8/suction.pdf}}
+\hspace{1cm} 
+\subfloat[ ]{\includegraphics[height=3.2cm]{chapter8/implant_injection_step_1.pdf}}
+\hspace{1cm} 
+\subfloat[ ]{\includegraphics[height=3.2cm]{chapter8/implant_injection_step_2.pdf}}
+\caption [Steps of cataract surgery] {(a) removal of the opacified lens by phacoemulsification.
+(b) insertion of the lens implant which is folded inside the injection device and then deploys within the lens capsule.
+(c) the lens in place in the capsule.}
+\label{chap8:fig-surgery}
+\end{figure}
+
+To simulate the insertion and deployment of an intra-ocular lens, we first created a triangulation of the lens surface.
+Particular care was given to the mesh, to ensure that areas where large stresses occur contain a higher density of elements (see \fig{chap8:fig-mesh}).
+This was done by noting the constraints applied by the surgeon to the haptics while inserting the implant within the injection device.
+During this stage, the haptics are folded onto the implant body, leading to high stresses at the junctions.
+The lens mesh contains $743$ triangles and $473$ nodes.
+Models of the injection device and the entire eye anatomy were also created.
+Physical parameters of the lens implant have been provided by the manufacturer Alcon and they are presented in Table~\ref{chap8:tab-parameters}.
+%
+\begin{table}[ht]
+	\begin{center}
+		\begin{tabular}{|p{3cm}|p{3cm}|p{3cm}|}
+		\hline
+		 \centering Young's modulus & \centering Poisson's ratio & \centering Mass density \tabularnewline
+		\hline
+		\centering $1\,$MPa & \centering $0.42$ & \centering $1.2\,$g/cm$^3$ \tabularnewline
+		\hline
+		\end{tabular}
+	\vspace{0.3cm}
+	\caption[Physical parameters of the intra-ocular implant]{Physical parameters of the intra-ocular implant (data obtained directly from Alcon at our request)}
+	\label{chap8:tab-parameters}
+	\end{center}
+\end{table}
+
+The first difficulty is to obtain the folded geometry of the lens within the injection device.
+This step is not important for the training process and does not need to be interactive.
+Indeed the surgeon does not always have to prepare the implant as some injection devices are readily available with a folded implant already in place.
+We simulate the folding process by first folding the haptics onto the implant body.
+The body was then bent while keeping the haptics inside to obtain the shape described in \fig{chap8:fig-implantFolding}.
+The whole process was carried out by applying the necessary forces and boundary conditions on the body and haptics of the implant.
+The folded implant was then placed into the injection device.
+The simulation of the injection consists in pushing the intra-ocular implant within the injection device into the lens capsule.
+During these stages of the simulation, complex contacts occur and consist of self collisions of the lens as well as collisions between the lens and the injector and later with the capsule.
+To solve the contacts we use a direct solver.
+Indeed, since the mass of the implant was small compared to its stiffness, an iterative solver (such as conjugate gradient) needed a lot of time to converge.
+As adhesions between the haptics and the body is often observed in surgery, friction is also taken into account in the contact response process.
+% the contact warping method proposed by \cite{Saupin08} as it offers an efficient way to compute physically correct contact responses in the case of co-rotational models. 
+%
+\begin{figure}[ht]
+\centering 
+\subfloat[Real implant]{\includegraphics[height=5cm]{chapter8/IOL.pdf}}
+\hspace{1cm} 
+\subfloat[Meshed model]{\includegraphics[height=5cm]{chapter8/mesh_implant.pdf}}
+\caption [Lens implant and its mesh] {An actual intra-ocular implant (a) and the triangular (b) mesh used in our simulations.
+Notice the higher density of elements in areas where large deformations will take place.
+Image of implant courtesy of Alcon.}
+\label{chap8:fig-mesh}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=5cm]{chapter8/implant_folding1.pdf}
+\includegraphics[width=4.5cm]{chapter8/implant_folding2.pdf} \\
+\includegraphics[width=4cm]{chapter8/implant_folding3.pdf}
+\includegraphics[width=7cm]{chapter8/implant_folding4.pdf}
+\caption [Folding of intra-ocular implant] {Top: intermediary steps of the intra-ocular implant folding. Bottom: fully folded implant ready to be placed into the injection device.}
+\label{chap8:fig-implantFolding}
+\end{figure}
+
+Results of our simulation are illustrated in \fig{chap8:fig-simu-results}.
+We can notice the progressive deployment of the implant when it exits the injector.
+The shape of the intra-ocular lens remains very close to that of a real one during all stages of the simulation: within the injector, during the ejection phase, and when in place within the capsule.
+Due to the high stiffness and low mass of the lens, a direct sparse solver was used at each time step (dt $= 0.01\,$s) rather than an iterative solver, resulting in a more accurate and more stable simulation, to the detriment of computation times (about 5 FPS for the complete simulation, and about 10 FPS for the deformation only, on a 2.4 GHz processor).
+
+% Link where images from surgery have been found: http://asianophthalmology.org/articles/?a=articles&p=6
+\begin{figure}[ht]
+\centering
+\includegraphics[width=4.5cm]{chapter8/surgery1.pdf}
+\hfill
+\includegraphics[width=4.5cm]{chapter8/surgery2.pdf}
+\hfill
+\includegraphics[width=4.5cm]{chapter8/surgery3.pdf} \\
+\vspace{0.1cm}
+\includegraphics[width=4.5cm]{chapter8/simu1.pdf}
+\hfill
+\includegraphics[width=4.5cm]{chapter8/simu2.pdf}
+\hfill
+\includegraphics[width=4.5cm]{chapter8/simu3.pdf}
+\caption [Lens imlant] {Three steps of the simulation of the intra-ocular lens implant injection and its deployment within the lens capsule.
+Top: images from a real cataract surgery, courtesy of Dr.
+Tarek Youssef, ophthalmologist in Canada.
+Below: our simulation of the implant's deployment.}
+\label{chap8:fig-simu-results}
+\end{figure}
+
+\section{Discussion}
+
+The finite element procedure that we propose was largely developed based on physical understanding without the use of mathematical shell theories.
+Indeed, our shell FEM was obtained by simply superimposing a plane stress and bending energies.
+According to \cite{Chapelle03}, this type of approach has a limited level of accuracy.
+The effective study of shell structures clearly requires a deep physical understanding of shell structural behaviours.
+The complexity of the physical behaviours of shells requires advanced mathematical analyses from theories that are still in development.
+The interested reader may refer to the book of \cite{Chapelle03} for more details on shell theory.
+If the range of application of our shell formulation might be somewhat limited, we did not find out any inconsistencies through our tests.
+We also need to keep in mind that the accuracy required in medical simulation is not the same than the one demanded in structural analyses for instance.
+Moreover, the additional complexity of a shell finite element formulation based on true shell theory may also very well jeopardise our real-time constraints.
+Therefore, we do not think that the use of this simple shell formulation significantly diminishes our contribution to the field of medical simulation.
+Indeed, to our knowledge, our co-rotational shell FEM is the first description of a shell finite element model applied to simulating the deformation of thin anatomical structures.
+
+Nevertheless, a few improvements are conceivable.
+Large strain could be allowed by using a non-linear strain tensor and more complex material laws (non-linear, viscoelasticity, etc.) could also be added to our shell formulation.
+However, those changes are substantial and their implementations are not straightforward.
+In addition, it is not clear how the benefits would be the greatest: whether from enhancing our model to a non-linear formulation or completely changing to another model based on true shell theory (where membrane and bending energies are not separated).
+The answer may depend on the type of constraints applied on the object.
+Also (as already mentioned section \ref{chap8:GPU}), the real-time capacity of our algorithm could be further improved by implementing our shell FEM on GPU.
+Although an optimal parallel re-implementation is always challenging, we have already reviewed why there was no major obstacle to a GPU implementation.
+
+\section{Conclusion}
+
+We proposed a co-rotational formulation for shell elements by extending a classical in-plane triangular finite element approach.
+This simple shell element can efficiently handle both membrane, bending and twist forces.
+The validity of our approach has been demonstrated though comparison with theoretical results.
+It was then applied it to a rather complex application: the simulation of intra-ocular lens implant deployment in a cataract surgery simulation.
+These results are very encouraging and show the potential of such models.
+Examples include the modelling of hollow anatomical structures (colon, stomach, etc.), the simulation of cardiac valve leaflets, and blood vessels.
+This work on this co-rotational shell FEM formulation was the object of one main publications \citep{Comas2010a}.
+
+So far, the mesh of the implant used in the cataract surgery simulation was created manually.
+However, manually creating the mesh is not suitable if we want to model the deformation of complex anatomical structures using shell elements.
+Therefore, the first step is to describe the surface of the structure that we want to simulate with curved patches.
+The goal is to create a mesh featuring the optimal number of curved shell elements while staying as close as possible to our targeted geometry.
+The following chapter deals with this problem.

### 2010-10-24 Thesis/Chapter9.tex
- Update 24 October (final) -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@189 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2010-10-11 Thesis/Chapter9.tex
@@
-\chapter{A co-rotational triangular shell FEM model}
+\chapter{Physics-based reconstruction using shell elements}
 \label{chap9}
 \begin{shortAbstract}
-In the field of medical simulation, very few models have been proposed for simulating, in real-time, the deformation of thin anatomical structures.
-In fact, to our knowledge, no model based on the equations of continuum mechanics was described and applied to simulating the deformation of objects.
-In this chapter, we propose to rely on continuum mechanics to accurately describe the complex deformations of the implant.
-Our approach extends the co-rotational method used in finite element analysis of in-plane deformations to incorporate a bending energy.
-We will first describe the model in details before discussing its implementation in the open source framework SOFA.
-Lastly, this co-rotational shell FEM was applied to the simulation of the lens deployment during catarct surgery.
-This simulation also accounts for the complex contacts that take place during the injection phase.
-This work was the object of two publications \citep{Comas2010a,Comas2010b}.
+In the previous chapter, we proposed a co-rotational formulation for shell elements by extending a classical in-plane triangular finite element approach.
+This simple shell element can efficiently handle both membrane and off-plane forces (bending and twist).
+The validity of our approach has been demonstrated though comparison with theoretical results.
+It was then applied it to a rather complex application: the simulation of intra-ocular lens implant deployment in a cataract surgery simulation.
+These results are very encouraging and show the potential of such models.
+Examples include the modelling of hollow anatomical structures (stomach, colon, etc.), the simulation of cardiac valve leaflets, and blood vessels.
+However, to model the deformation of complex anatomical structures using shell elements, the first step is to describe its surface with curved patches.
+In the next section, we propose to study how to approximate the surface of anatomical structures with (curved) shell elements.
 \end{shortAbstract}
 
-\section{Model description}
+\section{Techniques of mesh simplification}
 
-We propose to define a triangular shell element by combining a two-dimensional in-plane membrane energy, with an off-plane energy for describing bending and twist.
-To allow for real-time simulation, a computationally efficient formulation is needed.
-We therefore propose to extend the co-rotational idea introduced by \cite{Felippa00}.
-Indeed, as we have seen in the previous chapters, co-rotational approaches have been successfully applied to real-time simulation over the last few years.
-They offer a good trade-off between computational efficiency and accuracy by allowing small deformations but large displacements.
-We propose to improve and extend a plate model first introduced by \cite{Przemieniecki85} to a co-rotational formulation.
-Once combined with an in-plane membrane formulation we obtain an accurate, yet computationally efficient, shell finite element method featuring both membrane and bending energies.
+\subsection{Introduction: our motivation}
 
-\subsection{Triangular elastic membrane}
+Our idea is the following: while many flat triangles are required to describe highly curved surfaces, fewer triangular shell elements are needed to describe a given geometry with the same precision since they can be curved.
+Our goal is to create a mesh featuring the optimal number of shell elements while staying as close as possible to our targeted geometry.
+Literature about volumetric mesh generation algorithms is fairly dense.
+However, there are only a few that are concerned about the generation of meshes over curved surfaces.
+One of the most widely used techniques for the creation of surface meshes is the plane to surface transformation method \citep{Zienkiewicz71}, mesh is first generated on a two-dimensional domain and then mapped onto the surface.
+If this method gives reasonably good meshes on smooth surfaces, the results are usually quite poor with more complex curved surfaces.
+The finite elements may be generated directly on the curved surfaces based on the advancing front technique \citep{Lo85,Lau96}.
+The main issue with this approach is that an analytical description of the geometry is needed, which is not the case in medical simulation.
+Another method consists of using an a priori error estimator to build an adaptive mesh generation \citep{Baumann97}.
+However, the tolerance of this indicator should be chosen depending on the desired accuracy of the finite element solution.
+Therefore it requires some knowledge about the problem in order to choose an effective tolerance.
+\cite{Bechet02} start from an existing triangular mesh created with a CAD (computer-aided design) software and refine and smooth the mesh based on element quality and surface curvature.
+While all those methods allow their authors to get satisfactory meshes over curved surfaces according to their needs, they all make use of flat elements to mesh geometries and do not generate actual shells.
 
-The computation of the triangular elastic membrane stiffness matrix can be derived from previous works dealing with tetrahedral co-rotational elements \citep[like][for instance]{Muller04b}.
-The element stiffness matrix $\textbf{K}_e$ can be computed as follows:
+In fact, it is interesting to note that this process of meshing a curved surface is similar to the reconstruction of the surface of objects in computer vision.
+Indeed, calculating curvature maps of 3D surfaces represented as digitised data (point clouds or triangulated meshes) has been extensively studied.
+One of the most common approach is to use continuous surface patches \citep{Kolb95,Douros02}.
+The surface is locally approximated by an analytic representation called surface patch, usually chosen to be an implicit surface.
+\cite{Tang99} presented an elegant and non noise-sensitive approach that infers sign and direction of principal curvatures directly from the input and the authors used this information for coherent surface and curve extraction from 3D data.
+
+However, we need to realise that our situation is substantially different than all previously mentioned geometrical approaches as we want to model the deformation of the structure.
+In that regard, the curvature of the surface has a physical meaning: it represents the mid-surface of the shell.
+By keeping this in mind, we propose an approach to mesh a curved surface based on the same polynomial shape function \eqref{chap8:eq-deflection} used in our co-rotational shell FEM.
+
+%Simplification strategies may be grouped into two categories: local strategies that iteratively simplify the mesh by the repeated application of some local operator, and global strategies that are applied to the input mesh as a whole \citep{Cignoni97,Talton04}.
 %
+%\subsection{Local simplification strategies}
+%
+%Typically, they define some mesh operation $S$ that, when applied to a mesh $M$, acts on a small collection of its elements and produces a new mesh $\overline{M}$  with fewer elements.
+By repeated application of $S$, a mesh may be simplified arbitrarily.
+In order to determine the mesh elements to which $S$ should be applied on a given iteration, $S$ may be associated with an error function (or cost function) that measures the amount of error the operation will introduce into the approximation.
+By computing the error associated with every possible application of $S$ at a particular iteration, the algorithm can apply the one with minimal cost.
+This type of heuristic is quite reasonable for simplification problems, and in practice these methods work well.
+%
+%One nice property of local iterative algorithms is that they allow the user to specify the desired attributes of the target approximating mesh with a high degree of precision.
+For example, the user may allow the algorithm to run until the mesh contains $k$ faces, or until the error at a given vertex exceeds some threshold $\varepsilon$.
+%
+%\subsubsection*{Vertex decimation.}
+%The decimation algorithm, first proposed by \cite{Schroeder92}, is simple.
+Multiple passes are made over all vertices in the mesh.
+During a pass, each vertex is a candidate for removal and, if it meets the specified decimation criteria, the vertex and all triangles that use the vertex are deleted.
+The resulting hole in the mesh is patched by forming a local triangulation.
+The vertex removal process repeats, with possible adjustment of the decimation criteria, until some termination condition is met.
+%
+%The error associated with a particular decimation typically depends on the classification of the vertex being decimated.
+For the general case of a manifold vertex $\lbrace i \rbrace$ not near a boundary, \citeauthor{Schroeder92} consider the set of planes formed by the adjacent triangles and compute a single approximating plane based on the area-weighted average of the triangle normals $\textbf{n}_k$, centers $x_k$, and areas $A_k$:
+%%
+%\begin{equation}
+%\textbf{n} = \frac{\sum_k A_k\textbf{n}_k}{\sum_k A_k},  \hspace*{1cm} \widehat{\textbf{n}} = \frac{\textbf{n}}{\vert \textbf{n} \vert}, \hspace*{1cm}  x = \frac{\sum_k A_k\textbf{x}_k}{\sum_k A_k}
+%\end{equation}
+%
+%Then they calculate the distance from vertex $\lbrace i \rbrace$ to this plane by  $\vert \widehat{\textbf{n}} \cdot (v_i - x) \vert$ and define this quantity to be the cost associated with the decimation.
+This metric ensures that vertices in smooth regions will be decimated before vertices that define sharp features.
+%
+%In practice, simplification schemes based on vertex decimation excel at eliminating un-necessary geometry, such as is typically found in meshes generated by the marching cubes algorithm where large planar regions may be subdivided into many redundant triangles.
+%
+%\subsubsection*{Edge contraction.}
+%Edge contraction, originally proposed by \cite{Hoppe93}, is the most common simplification operation.
+An edge contraction operates on a single edge $\lbrace v_1,v_2 \rbrace$ and contracts that edge to a single vertex $\lbrace w \rbrace$, updating all edges previously incident on $\lbrace v_1 \rbrace$ and $\lbrace v_2 \rbrace$ to reference $\lbrace w \rbrace$.
+On a manifold mesh without boundary, each edge contraction will collapse precisely two faces (see \fig{chap9:fig-edgeCollapse}).
+Edge contractions can alter the topology of a mesh, since repeatedly contracting all the edges around a hole will eventually close it.
+%%
+%\begin{figure}
+%\centering
+%\includegraphics[height=3cm]{chapter9/edgeCollapse.pdf}
+%\caption[The edge contraction operation]{The edge contraction operation. Image courtesy of \cite{Hoppe93}.}
+%\label{chap9:fig-edgeCollapse}
+%\end{figure}
+%
+%For a given contraction $\lbrace v_1,v_2 \rbrace \rightarrow \lbrace w \rbrace$, it is not immediately clear what value should be assigned to $v_h$, i.e.
+where the resulting vertex should be placed.
+Obvious choices such as $v_h = v_i$, $v_h = v_j$, or $v_h = (v_i + v_j)/2$ are convenient, but can easily be shown to be non-optimal.
+Rather than placing $v_h$ arbitrarily, it is better to consider the error function associated with the contraction operation and attempt to minimise its value over the space of possible vertex placements.
+Once the optimal target position has been computed for every possible contraction, the contraction with the smallest associated error can be selected and applied.
+%
+%\subsubsection*{Adaptative subdivision.}
+%Coplanar or nearly coplanar faces are searched for in the mesh, merged into larger polygons, and then retriangulated into fewer faces than those originally required \citep{Dehaemer91,Hinker93}.
+Face merging is driven by a co-planarity test.
+%
+%The basic adaptive subdivision method as presented by \cite{Schmitt86} is sketched in \fig{chap9:fig-adaptativeSubdivision}.
+(a) shows several data points produced by the sampling process and sketch (b) adds a trial polygon that approximates the surface represented by the sample points.
+A bilinear interpolation is performed over the surface of the approximating polygon to locate the point on the surface 'directly below' a sample data point.
+The distance (or error) between the surface and the data point is then calculated, as shown in (c), and is compared to the user specified tolerance.
+If any of the errors exceeds the tolerance value then the polygon is divided into four smaller polygons, as shown in (d), and the process is repeated recursively.
+%%
+%\begin{figure}
+%\centering
+%\includegraphics[width=13.5cm]{chapter9/adaptativeSubdivision.pdf}
+%\caption[Surface-fitting polygons with adaptive subdivision]{Surface-fitting polygons with adaptive subdivision. Image courtesy of \cite{Schmitt86}.}
+%\label{chap9:fig-adaptativeSubdivision}
+%\end{figure}
+%
+%When the approximation to the data is within the specified tolerance, the recursion is terminated and the polygon is saved in a linked list.
+Upon completion of the subdivision algorithm, the resulting polygons are displayed for the user to judge the results.
+Generally, the user runs the program repeatedly while varying the maximum tolerance until the results are suitable.
+%
+%\subsubsection*{Polygon growth.}
+%Another method proposed by \cite{Dehaemer91} to systematically generate surface-fitting polygons is to use a polygon growth technique (see \fig{chap9:fig-polygonGrowth}).
+First a seed polygon is selected from the original set of data, as in (a).
+Then a neighbouring polygon is selected and the two are combined into a larger trial polygon ((b) and (c)).
+If the trial polygon passes the accuracy metric, 10 additional neighbors are selected in an attempt at further growth, as in (d) through (g).
+When all attempts at further growth fail, the polygon is added to the list of polygons describing the reduced object, as shown in (h).
+A new seed polygon is selected and the process repeated.
+When there are no more polygons from which to choose a new seed, the process is terminated.
+%%
+%\begin{figure}
+%\centering
+%\includegraphics[width=12cm]{chapter9/polygonGrowth.pdf}
+%\caption[Polygon growth]{Polygon growth. Image courtesy of \cite{Dehaemer91}.}
+%\label{chap9:fig-polygonGrowth}
+%\end{figure}
+%
+%This method of fitting polygons to the sampled data is not as useful as the adaptive subdivision methods, for several reasons.
+First, it is an essentially brute force method of surface fitting and very repetitive in nature.
+Many trial polygons are created, only to be 'thrown out' because they exceed the error tolerance.
+This causes the procedure to execute quite slowly (orders of magnitude more slowly) than the adaptive subdivision method.
+Second, the data is not significantly reduced.
+Many of the generated polygons are of complicated shapes and use a large number of vertices.
+Thus many of the original sample data points are retained.
+%
+%
+%\subsubsection*{Appearence preserving simplification.}
+%A completely different approach to determining the error associated with a given simplification operation was described by \cite{Cohen98}.
+Their metric, which is appearance-based, measures the amount of deviation caused by the operation in the screen-space (pixelised) representation of the mesh.
+Their algorithm functions by decoupling surface position from color and curvature information and storing the latter two quantities in texture and normal maps.
+A traditional geometric simplification algorithm can then be employed to filter the surface position, while a hardware-based approach is used to filter the color and normal information, resulting in simplified representations that are nearly visually indistinguishable from the original.
+However, the algorithm requires a parameterisation of the mesh in order to function.
+%
+%
+%\subsection{Global simplification strategies}
+%
+%\subsubsection*{Vertex clustering.}
+%The method of vertex clustering was originally proposed by \cite{Rossignac93} to handle meshes of arbitrary topological structure.
+In their algorithm, each vertex in the input mesh is assigned a weight based on its perceptual importance: vertices adjacent to triangles with large faces and those in areas of high curvature are weighted more heavily than vertices in smooth regions adjacent to smaller triangles.
+Next, a bounding box is placed around the mesh and subdivided into a three-dimensional grid.
+Finally, all the vertices in a given grid cell are clustered to the position of the vertex with maximum weight.
+Degenerate geometry may then be removed, although the authors proposed a novel visualisation of certain degenerate faces and edges in an effort to enhance the recognizability of their simplified meshes.
+The algorithm is extremely efficient and simple to implement, and the level of simplification may be controlled (although with some difficulty) by choosing the resolution of the overlaid grid.
+However, vertex clustering can drastically alter the topology of the input mesh in an unpredictable manner, and in practice does not produce very faithful geometric approximations at low levels of detail.
+%
+%\subsubsection*{Re-tiling.}
+%The re-tiling method described by \cite{Turk92} begins with a polygonal surface and creates a triangulation of this surface with a user-specified number of vertices.
+The points are placed uniformly over a given polygonal surface by distributing points at random over the surface and then having each point repel all of its neighbors.
+An alternative to placing the points randomly is using the variation in surface curvature.
+The features of the surface would be more accurately reflected in a re-tiling by increasing the density of vertices in regions of high curvature.
+%
+%The key notion in the re-tiling procedure is the creation of an intermediate model called the mutual tessellation of a surface that contains both the vertices from the original model and the new points that are to become vertices in the re-tiled surface.
+The new model is then created by removing each original vertex and locally re-triangulating the surface in a way that matches the local connectedness of the initial surface (see \fig{chap9:fig-retiling}).
+%
+%If this method produces nice meshes, an issue is finding measures of how closely matched a given re-tiling is to the original model. 
+%%
+%\begin{figure}
+%\centering
+%\includegraphics[width=12cm]{chapter9/retiling.pdf}
+%\caption[Retiling technique]{Upper left: original surface.
+Upper right: candidate vertices after point-repulsion.
+Lower left: mutual tesselation.
+Lower right: final tesselation.
+Image courtesy of \cite{Turk92}.}
+%\label{chap9:fig-retiling}
+%\end{figure}
+
+\section{Our method}
+
+	\subsection{A simple algorithm}
+
+In the following, we assume that we have a high resolution triangular mesh obtained from a binary segmented image of the object we want to simulate (via a Marching Cube algorithm for instance, \cite{Lorensen87}).
+We propose a technique that can simplify the high resolution mesh of the object to be simulated and approximate the surface of anatomical structures with shell elements whose each surface is described by the shape function used in our shell formulation.
+Thus, the first step in the process of generating a shell-based mesh is an important decimation of the high resolution mesh, using quadric edge collapse technique implemented in \citeauthor{Meshlab}.
+MeshLab is an open source program for the processing and editing of unstructured 3D triangular meshes.
+The algorithm tries as much as possible to preserve mesh boundaries and generates high quality triangular elements.
+We then apply a heuristic method derived from the work of \cite{Saupin07} with tetrahedral meshes based on simple geometrical rules.
+For each node of the coarse mesh, we find the three closest triangles on the high resolution mesh and we move the node to the barycentre of the three centres of mass of those triangles.
+This technique locally smooths the surface of the mesh while converging towards the desired high resolution mesh.
+
+At each iteration of this algorithm, we want to measure the error between the curved surface of shells and the target.
+Indeed, we need to ensure that the distance between the surface of our shell-based mesh and the targeted high resolution mesh will be minimal.
+An efficient technique for measuring the error between two surfaces is the Hausdorff distance \citep{Klein96}.
+As a reminder the Hausdorff distance between two meshes is the maximum between the two so-called one-sided Hausdorff distances:
 \begin{equation}
-\textbf{K}_e = \int_v \textbf{J} \boldsymbol\chi \textbf{J}^{T} dV,
+d_{\mathrm{H}}(X,Y) = \max \left\{ \sup_{x \in X} \inf_{y \in Y} d(x,y), \sup_{y \in Y} \inf_{x \in X} d(x,y) \right\},
 \end{equation}
-where $\textbf{J}$ is the strain-displacement matrix and $\boldsymbol\chi$ embodies the material's behaviour.
-The implant is very stiff and we therefore assume that the local deformations remain limited during the deployment and a linear constitutive law is sufficient.
-Thus in the simple case of Hooke's law we have:
-%
-\begin{equation}
-\boldsymbol\chi = \frac{E}{12(1-\nu^2)}
-\begin{bmatrix}
-1 & \nu & 0 \\
-\nu & 1 & 0 \\
-0 & 0 & \frac{1}{2} (1-\nu)
-\end{bmatrix}
-\end{equation}
+where $d()$ is the Euclidian distance between two points.
+The same technique of subdivision used for rendering allows us to sample the actual surface described by the shells to compute the Hausdorff distance with the targeted high resolution mesh.
 
-The stiffness matrix in the global frame is eventually obtained using the rotation matrix of the element: $\textbf{K} = \textbf{R} \textbf{K}_e \textbf{R}^{T}$ where $\textbf{R}$ describes the rotation of the (triangular) element with respect to its initial configuration.
-
-\subsection{Triangular plate bending}
-
+By using the Hausdorff distance between two iterations, the process may be stopped when the required precision has been reached.
+A simple example is shown \fig{chap9:fig-cylinder} to illustrate the method.
 \begin{figure}[ht]
 \centering
-\includegraphics[height=4cm]{chapter9/bending.pdf}
-\caption {The different degrees of freedom $u$ of a triangular thin plate in bending.}
-\label{chap9:fig-triangle}
+\includegraphics[width=13.5cm]{chapter9/exampleCylinder.pdf}
+\caption[Meshing of a cylinder with shell elements]{The target (a) is a high resolution cylinder mesh of 16,384 triangles and we start from a very coarse mesh (12 triangles), rendered with flat triangles here (b).
+In (c) the coarse mesh is rendered with shells and a one-sided Hausdorff distance colour map is applied to show the initial error with the high resolution mesh.
+(d) One-sided Hausdorff distance colour map after one iteration of our algorithm (48 shells).}
+\label{chap9:fig-cylinder}
 \end{figure}
-%
-To calculate the stiffness matrix for the transverse deflections and rotations shown on \fig{chap9:fig-triangle}, we start with the assumed deflection $u_z$ of the form
-\begin{equation}
- u_z = c_1 + c_2x + c_3y + c_4x^2 + c_5xy + c_6y^2 + c_7x^3 + c_8xy^2 + c_9y^3,
-\label{chap9:eq-deflection}
-\end{equation} 
-where $c_1$, \ldots , $c_9$ are constants.
-Equation~\ref{chap9:eq-deflection} solves an issue of symmetry which was observed with the deflection function proposed by \cite{Przemieniecki85}.
-These constants can be evaluated in terms of the displacements and slopes at the three corners of the triangular plate using
-\begin{equation}
-\textbf{u} = \textbf{Cc}
-\label{chap9:eq-U}
-\end{equation} 
-where $\textbf{u} = \left\{u_1 u_2 \ldots u_9 \right\} $ and $\textbf{c} = \left\{c_1 c_2 \ldots c_9 \right\} $. Matrix $\textbf{C}$ derives from \eqref{chap9:eq-U}:
-\begin{equation}
-C = 
-	\begin{bmatrix}
-	1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
- 	0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
-	0 & -1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
-	1 & x_2 & 0 & x_2^2 & 0 & 0 & x_2^3 & 0 & 0 \\
-	0 & 0 & 1 & 0 & x_2 & 0 & 0 & 0 & 0 \\
-	0 & -1 & 0 & -2x_2 & 0 & 0 & -3x_2 & 0 & 0 \\
-	1 & x_3 & y_3 & x_3^2 & x_3y_3 & y_3^2 & x_3^2 & x_3y_3^2& y_3^3 \\
-	0 & 0 & 1 & 0 & x_3 & 2y_3 & 0 & 2x_3y_3 & 3y_3^2 \\
-	0 & -1 & 0 & -2x_3 & -y_3 & 0 & -3x_3^2 & -y_3^2 & 0
-	\end{bmatrix}
-\end{equation}
 
-We used the following notations:
-\begin{align}
-u_1 &= (u_z)_{x_1,y_1}, & u_2 &= \left(\frac{\partial u_z}{\partial y}\right)_{x_1,y_1}, & u_3 &= - \left(\frac{\partial u_z}{\partial x}\right)_{x_1,y_1} \notag \\
-u_4 &= (u_z)_{x_2,y_2}, & u_5 &= \left(\frac{\partial u_z}{\partial y}\right)_{x_2,y_2}, & u_6 &= - \left(\frac{\partial u_z}{\partial x}\right)_{x_2,y_2} \label{chap9:ui} \\
-u_7 &= (u_z)_{x_3,y_3}, & u_8 &= \left(\frac{\partial u_z}{\partial y}\right)_{x_3,y_3}, & u_9 &= - \left(\frac{\partial u_z}{\partial x}\right)_{x_3,y_3} \notag \\
-\end{align}
+	\subsection{Results on complex anatomical structures}
 
-\noindent We can calculate the strains from the flat-plate theory using:
-\begin{equation}
-e_{xx} = -z \frac{\partial^2u_z}{\partial x^2}
-\end{equation} 
-\begin{equation}
-e_{yy} = -z \frac{\partial^2u_z}{\partial y^2}
-\end{equation} 
-\begin{equation}
-e_{xy} = -2z \frac{\partial^2u_z}{\partial x \partial y}
-\end{equation}
-
-\noindent Hence, using the above equations and \eqref{chap9:eq-deflection}, we have
-\begin{equation}
-\begin{bmatrix}
-e_{xx} \\
-e_{yy} \\
-e_{xy}
-\end{bmatrix}
-= 
--z
-\begin{bmatrix}
-0 & 0 & 0 & 2 & 0 & 0 & 6x & 0 & 0 \\
-0 & 0 & 0 & 0 & 0 & 2 & 0 & 2x & 6y \\
-0 & 0 & 0 & 0 & 2 & 0 & 0 & 4y & 0 \\
-\end{bmatrix}
-\textbf{c}
-\label{chap9:eq-strains}
-\end{equation} 
-or symbolically $\textbf{e} = \textbf{Dc}$ where $\textbf{D}$ stands for the $3\times 9$ matrix in \eqref{chap9:eq-strains}, including the pre-multiplying constant $-z$.
-Noting from \eqref{chap9:eq-U} that $\textbf{c} = \textbf{C}^{-1}\textbf{u}$, we have:
-\begin{equation}
-\textbf{e} = \textbf{DC}^{-1}\textbf{u} = \textbf{bu},
-\end{equation} 
-where $\textbf{b} = \textbf{DC}^{-1}$.
-
-\noindent Knowing that the stiffness matrix $\textbf{K}_e$ for an element is obtained from
-\begin{equation}
-\textbf{K}_e = \int_v \textbf{b}^{T} \boldsymbol\chi \textbf{b} dV,
-\end{equation} 
-where $\boldsymbol\chi$ is the material matrix, the substitution of $\textbf{b}$ into this expression yields
-\begin{equation}
-\label{chap9:Ke}
-\textbf{K}_e = (\textbf{C}^{-1})^T \int_v \textbf{D}^{T} \boldsymbol\chi \textbf{D} dV \, \textbf{C}^{-1}.
-\end{equation}
-
-Similarly to the membrane stiffness matrix, the bending stiffness matrix in the global frame is eventually obtained using the rotation matrix of the element: $\textbf{K} = \textbf{R} \textbf{K}_e \textbf{R}^{T}$.
-
-\section{Mechanical interactions with the curved surface of shells}	\label{chap9:interactions}
-
-The practical interest of modelling complex behaviours such as bending and twisting would remain fairly low for medical simulation if contacts and constraints were not handled properly.
-In our case the difficulty comes from different sources.
-First the collision detection must be carried out with the curved surface of shell elements as opposed to the classic detection on plane triangles.
-Then forces applied to a given triangle need to be distributed between linear forces and torques onto its three vertices.
-As we will see, the same polynomial interpolation function chosen to compute the bending energy in our FEM formulation is also used to capture the interactions between the curved surface and other objects.
-
-In order to detect the collision with the bent surface, we have chosen the subdivision approach.
-We first sample the flat surface of each element by recursively dividing each triangle into four smaller ones and the deflection of each new vertex is computed using \eqref{chap9:eq-deflection} according to the displacements and slopes at the three vertices of the triangular element.
-This process of subdivision may also allow us to render each shell as a curved triangle (\fig{chap9:fig-shell} (a) and (b)) and detect any collision with the curved surface of the shell using any of the classic collision detection algorithms working on flat triangles.
-
-Once a collision has been detected, it must be processed by distributing the linear force received on the bent surface between the three vertices of the triangle.
-First the linear part of the force is simply transmitted on each node using the barycentric coordinates of the contact point's projection onto the triangle.
-
-The main difficulty is to convert the normal component of the force applied to the bent surface into a torque at each of the three nodes (\fig{chap9:fig-shell} (c)).
-Our approach is the following: during force computation, we use the change in orientation measured at each node to compute the local deflection of each subvertex within the triangle.
-Differentiating the formulation twice yields a relation between the torque applied at each node and the generated force in bending.
-We therefore need to invert the latter formulation to convert a bending force into torques at each vertex.
-Thus, we are able to transmit any force coming from interactions with the curved surface of shells to the mechanical vertices used in our FEM formulation.
+\subsubsection*{Meshing of anatomical structures.}
+This approach has been applied to approximate more complex anatomical geometries with curved shell elements.
+Two examples are given with the Glisson's capsule, which is the membrane surrounding the liver (\fig{chap9:fig-liver}) and an aneurysm (\fig{chap9:fig-aneurysm}).
+In each case, the error is computed through a Hausdorff distance and expressed as a percentage of the diagonal of the object's bounding box.
 %
@@
 \centering
-\includegraphics[width=12cm]{chapter9/shell_curvature.pdf}
-\caption {(a)  The triangle formed by the three vertices of the shell has been recursively subdivided 3 times and the deflection of each new vertex was computed according to the same deflection function used in our shell FEM.
-(b) Sampling the actual surface of the shell allows more accurate rendering and collision detection.
-(c) The shape function is used to distribute an external force $F$ onto the triangle nodes.}
-\label{chap9:fig-shell}
-\end{figure}
-
-\section{Implementation}
-
-The co-rotational shell formulation has been successfully implemented as a Force Field into the open-source framework SOFA \citep{Allard07}.
-A description of SOFA's architecture was given in section \ref{chap6:SOFA}.
-We will now review the key points of the implementation.
-
-	\subsection{Evaluation of displacements}
-The computation of the stiffness matrix, and subsequently the internal forces, starts with the measure of the displacements $\left\{u_1 u_2 \ldots u_9 \right\} $ for every element.
-As defined by \eqref{chap9:ui}, we need to measure the displacement along the z axis and the rotations around local x and y axes for each vertex.
-For \cite{Przemieniecki85}, the deflection z is measured in the global frame.
-But in a co-rotational formulation, a local frame is attached to each triangle and used to compute displacements and forces.
-In this local frame (formed from the 3 vertices) the deflection is therefore always null (hence, $ u_1, u_4, u_7 $ are null).
-To evaluate the remaining rotations, we first associate a quaternion at each vertex which embodies its frame (or its orientation).
-In mathematics, quaternions are an extension of complex numbers and may be written as follows:
-\begin{equation}
-q = w + x i + y j + z k \quad \mbox{with } i^2 = j^2 = k^2 = -1.
-\end{equation}
-For use in geometry, quaternions are often written as a sum of a scalar and a vector:
-\begin{equation}
-q = w + (x, y, z) = \cos (\frac{\alpha}{2}) + \mathbf{u} \sin (\frac{\alpha}{2}),
-\end{equation}
-where $ \mathbf{u} $ is a unit vector and $ \alpha $ a scalar.
-If we denote by $ \mathbf{v} $ a vector in $\field{R}^3$, it can be shown that the quaternion product $ q \mathbf{v} q^{-1} $ yields a vector which is the rotation of $ \mathbf{v} $ by an angle $ \alpha $ around the axis $ \mathbf{u} $.
-Quaternions require less storage than rotation matrices and are known to be more numerically stable.
-
-From there, our approach to measure the rotations around local x and y axes is simple.
-At runtime and for each vertex, we compute the rotation that we need to transform the z axis of the triangle's frame into the z axis of the vertex's frame (see listing \ref{chap9:qDiffZ}).
-The differences with the same quantities computed for the rest position during the initialisation lead to the desired displacements.
-%
-\begin{lstlisting}[caption={[Method to compute the rotation between two quaternions]Method used to compute the rotation of the z axis between the frames embodied by two quaternions},label={chap9:qDiffZ},frame=shadowbox,rulesepcolor=\color{black}]
-inline Quat qDiffZ(const Quat& Qvertex, const Quat& Qtriangle)
-{
-    // dQ is the quaternion that embodies the rotation between 
-    // the z axis of the vertex and the z axis of the local 
-    // triangle's frame (in local space)
-    Quat dQ;
-
-    // u = z axis of the triangle's frame (in local space)
-    Vec3d u(0,0,1);
-
-    // v = z axis of the vertex's frame is expressed into 
-    // world space
-    Vec3d v = Qvertex.rotate(Vec3d(0.0, 0.0, 1.0));
-
-    // v0 = v expressed into local triangle's frame
-    Vec3d v0 = Qtriangle.rotate(v);
-
-    // Axis of rotation between the 2 vectors u and v lies 
-    // into the plan of the 2 vectors
-    Vec3d axis = cross(u, v0);
-
-    // Shortest angle between the 2 vectors
-    double angle = acos(dot(u, v0));
-
-    // Quaternion associated to this axis and this angle
-    if (fabs(angle)>1e-6)
-    {
-        dQ.axisToQuat(axis,angle);
-    }
-    else
-    {
-        dQ = Quat(0,0,0,1);
-    }
-
-    return dQ;
-}
-\end{lstlisting}
-
-%	\subsection{Inversion of matrix $ \mathbf{C} $}
-%\cite{Przemieniecki85} described the condition of non-inversion for the matrix $\mathbf{C}$ as the following:
-%\TODO{add condition of inversion... but need the book :)}
-%
-%Each triangle of a mesh obtained from a regular grid of points fits this condition of non-inversion (right triangles).
-We fixed this limitation by choosing a different local coordinate system that the one introduced by \citeauthor{Przemieniecki85}.
-%
-%\TODO{add figure with the two different coordinate system}
-
-	\subsection{Numerical integration}
-In practice, the integration of the integral in \eqref{chap9:Ke} must be carried out numerically using Gaussian quadrature, a numerical technique which allows to calculate the value of an integral.
-The Gaussian quadrature states that the integral of a function $ f $ over the area $ A $ of a triangular surface may be evaluated as a weighted sum of function values at specific points within the area of integration \citep{Cowper73}.
-In other words, we have
-\begin{equation}
-\int_{A} f dA = A \sum_{i=1}^n w_i f_i,
-\end{equation}
-where $ w_i $ is a weight factor and $ f_i $ is the evaluation of $ f $ at the integration point $ i $ (also called Gauss points).
-The choice of the number and the locations of the integration points depends on the order of accuracy desired.
-We classically choose 3 integration points located at the middle of each edge of the triangle, for which the weight factors $ w_i $ are $ 1/3 $.
-The required integration over the volume is achieved by multiplying the result by the thickness $ t $ of the shell.
-Therefore, \eqref{chap9:Ke} may be numerically evaluated as
-\begin{equation}
-\textbf{K}_e = \dfrac{A t}{3} (\textbf{C}^{-1})^T \sum_{i=1}^3 \textbf{D}_i^{T} \boldsymbol\chi \textbf{D}_i  \, \textbf{C}^{-1},
-\end{equation}
-where $ D_i $ is the evaluation of the matrix $ \mathbf{D} $ defined by \eqref{chap9:eq-strains} at each Gauss point $ i $.
-
-\bigskip
-
-\subsection{Computation of stiffness matrix: summary}	\label{chap9:summary}
-In practical terms, the different computations associated with each triangular shell element can be described as follows:
-%
-\begin{enumerate}
-\item Compute the rotation matrix $\textbf{R}$ from global to triangle (local) frame
-\item Compute the local displacement vector $\textbf{u} = \{v_1, v_2, 0, u_2, u_3, v_3, v_4, 0, u_5, u_6,$ $v_5, v_6, 0, u_8, u_9 \} $ for each triangle, where $ (v_1, v_2), (v_3, v_4) and (v_5, v_6) $ are the in-plane displacements on x and y local axes and $ u_i $ were defined by \eqref{chap9:ui}.
-As we are in a co-rotational framework, we recall that the normal displacements $u_1, u_4, u_7$ at each of the three nodes are null in the local frame of the triangle.
-\item Compute matrix $\textbf{D}_i$  at each Gauss point $i$
-\item The strain-displacement matrix at each Gauss point $i$ is computed with $\textbf{J}_i = \textbf{D}_i \textbf{C}^{-1}$
-\item Compute the local stiffness matrix $\textbf{K}_e$ of the element as $\textbf{K}_e = \displaystyle{\sum^3_{i=1}} \textbf{J}_i \boldsymbol\chi \textbf{J}_i^T$
-\item Transform the local element stiffness matrix into the global frame and add it to the global stiffness matrix
-\end{enumerate}
-
-	\subsection{Concept of mappings in SOFA}	\label{chap9:mappings}
-Let us explain the concept of \emph{mappings} and \emph{mechanical mappings} in SOFA.
-In SOFA, the state of a simulated system can be described by the values and time derivatives of its independent degrees of freedom (DOF) gathered in two vectors $x_0$ and $v_0$.
-Geometry can be attached to the DOF for visualisation or contact computation.
-We call it the \emph{shape}.
-It is typically defined by points, such as triangle vertices or sphere centers, and additional data such as triangle connectivity or sphere radii.
-We call these points the vertices.
-Their positions, velocities and associated forces are stored in vectors $x_1$ , $v_1$ and $f_1$, respectively.
-They are not independent variables, since the positions and velocities are bound to the DOF using kinematic operators which we call the \emph{mappings}:
-\begin{eqnarray*}
-x_1 &=&\JNL_1(x_0)\\ 
-v_1 &=& J_1 v_0
-\end{eqnarray*}
-When the vertices and the DOF are the same, the mapping is the identity.
-Matrix $J_1 = \frac{\partial x_1}{\partial x_0}$ encodes the linear relation between the DOF velocities and the shape velocities.
-Due to linearity, the same relation holds on elementary displacements $dx$.
-It also holds on accelerations, with an additional offset due to velocities when the position mapping \JNL is nonlinear.
-In most cases, operators \JNL~ and \J~ are the same, but in the case of rigid bodies, \JNL~ is nonlinear with respect to $x$ and it can not be written as a matrix.
-
-When shapes collide, additional geometry can be necessary to model the contact.
-For instance, when an edge intersects another one, a contact force is applied to the intersection points.
-These points are defined by their barycentric coordinates with respect to their edge vertices.
-Other relations can be used, depending on the kind of geometrical primitives in contact.
-This additional geometry requires another geometrical layer connected to the shape by a mapping, as illustrated in \fig{chap9:fig-mappings}.
-
-\begin{figure}[ht]
-	\begin{center}
- 		\includegraphics[width=13cm]{chapter9/mappings}
- 	\end{center}
- \caption[Concept of mappings in SOFA]{Mappings from the DOFs to a contact point.
-Top: two simulated objects in contact (red point).
-Bottom: hierarchy of geometrical layers.
-Positions and velocities are propagated top-down.
-The contact force $f_c$ is accumulated in the contact layers.
-Forces are then propagated bottom-up.
-Image courtesy of SOFA's documentation.
-}
-\label{chap9:fig-mappings}
-\end{figure}
-
-Positions and velocities can be propagated top-down through our layer hierarchy using the relations presented.
-In order to take the contact forces into account in the dynamics equation, we have to convert the contact forces applied to the contact points to forces applied to the DOF, where our mechanical model is applied.
-This requires an extension to the position and velocity mappings presented: we call it \emph{mechanical mapping}.
-This is a general method to propagate forces bottom-up through the layers of the geometrical hierarchy.
-Given forces $f_n$ applied to a geometry layer $n$, we derive the equivalent force applied to its parent layer $n-1$.
-Equivalent forces must have the same power.
-Thus, we have to compute $f_{n-1}$ such that:
-$$
-v_{n-1}^T f_{n-1} = v_n^T f_n
-$$
-The relation $v_{n} = J_{n}v_{n-1}$ allows us to rewrite the previous equation as
-$$
-v_{n-1}^T f_{n-1} = v_{n-1}^T J_{n}^T f_n
-$$
-Since this relation must hold for any velocity $v_{n-1}$, we simplify it and get
-\begin{equation} \label{eq:mapF}
-f_{n-1} = J_n^T f_n
-\end{equation}
-This corresponds to the principle of virtual work.
-
-\subsection{Visualisation}
-Because shells have a curved surface, a specific rendering technique is required to visualise the deformation in bending.
-Without any consideration, the shells would appear flat (as simple triangles).
-Our idea is to make use of a higher resolution mesh whose deformation would be controlled by the underlying physical mesh of shells.
-In other words, a mapping was implemented to connect the DOFs of our mechanical model to a set of vertices useful for rendering.
-In SOFA, a mapping has only 2 methods: init(), which allows computations during the initialisation phase, and apply() called at runtime and in charge of updating the positions of the high resolution mesh according to the coarser mesh constituted by the shell elements.
-
-\begin{itemize}
-\item \textbf{init()} \\
-During the initialisation phase, each vertex from the high resolution mesh is projected onto the coarse mesh.
-Thus, for each vertex we find the closest primitive (DOF, edge or triangle) on the coarse mesh and we take the list of triangles attached to it.
-Then, the barycentric coordinates of the projected point is computed within each triangle of this list.
-%
-\item \textbf{apply()} deals with the relation between the DOFs and the vertices used for rendering (shape vertices).
-The position of each vertex from the high resolution mesh must be updated according to the underlying mesh.
-After the initialisation phase, we have associated each rendering vertex to a triangle and its barycentric coordinates within this triangle are known.
-Therefore, we first start by computing the new position of the rendering vertex by weighting the coordinates of each triangle's vertex with the associated barycentric weight.
-Hence, we obtain a new position in the plane of the triangle.
-If we were to stop the procedure here, the shell elements would still be rendered as flat triangles since all rendering vertex are located in the plane of the triangles.
-Consequently, the additional step of computing the deflection is required.
-We know that the deflection at any point of local coordinates $(x, y)$ may be computed from $u_z = c_1 + c_2x + c_3y + c_4x^2 + c_5xy + c_6y^2 + c_7x^3 + c_8xy^2 + c_9y^3$ as described by \eqref{chap9:eq-deflection}.
-Hence, we calculate the local coordinates $(x, y)$ of each rendering vertex within its associated triangle.
-Moreover, we notice that the coefficients $ \mathbf{c} = \left\{c_1 c_2 \ldots c_9 \right\} $ may be evaluated using \eqref{chap9:eq-U} from
-\begin{equation}
-\mathbf{c} = \mathbf{C}^{-1} \mathbf{u},
-\end{equation}
-The deflection may therefore be computed for each vertex and added to the in-plane position previously calculated. 
-\end{itemize}
-
-It is worth noting that a high resolution mesh may be obtained by recursively subdividing each triangle.
-However, in this case the vertices are closely related to the structure of the mechanical mesh and rendering artifacts may appear.
-As we can notice on \fig{chap9:fig-rendering}, the quality of the visualisation may be improved by using a totally different mesh.
-%
-\begin{figure}[ht]
-\centering 
-\subfloat[Subdivided triangles]{\includegraphics[width=5.5cm]{chapter9/sub_grid_red.png}}
-\hspace{1cm}
-\subfloat[Irregular grid]{\includegraphics[width=5.5cm]{chapter9/irregular_grid_red.png}} \\
-\subfloat[Subdivided triangles]{\includegraphics[width=5.5cm]{chapter9/sub_grid_mesh.png}}
-\hspace{1cm}
-\subfloat[Irregular grid]{\includegraphics[width=5.5cm]{chapter9/irregular_grid_mesh.png}}
-\caption[Visual artifacts using triangle subdivisions]{In some situations, some rendering artifacts may appear when we use subdivided triangles to render the shells (a,c).
-Using an irregular mesh which is not closely related to the structure of the mechanical mesh reduces the impact (b,d).}
-\label{chap9:fig-rendering}
-\end{figure}
-
-	\subsection{Contacts with the curved surface of shells}
-If a specific rendering technique was developed to visualise the curved shape of the shells, it is also essential to interact with it.
-Indeed, as explained section \ref{chap9:interactions}, the force applied onto the curved surface of the shell must be processed by distributing the linear force received on the bent surface between the three vertices of the triangle.
-Consequently, the mapping previously described is extented to a mechanical mapping and two additional methods need to be implemented: applyJ and applyJT.
-%
-\begin{itemize}
-\item \textbf{applyJ()} encodes the linear relation between the DOF velocities and the shape velocities as described in section \ref{chap9:mappings}.
-applyJ() basically embodies the same computations than apply() except that the vector of displacement $\mathbf{u}$ is different.
-Whereas the local displacement in rotation is measured from the positions to create the vector $\mathbf{u}$, this time the vector is directly obtained from the angular velocities we had at each vertex (once converted into the local frame).
-We therefore have a relation between the angular velocities at each vertex and the velocity of the deflection in the direction normal to each triangle.
-\item \textbf{applyJT()} must convert the contact forces applied to the contact points to forces applied to the DOF.
-If the linear forces are easily distributed between DOFs using barycentric coordinates, it also encodes the relation between the vertical force $ \mathbf{F} $ applied onto the bent surface of the plate and the torque at each DOF.
-We start by retrieving the normal component of the applied force vector $F_{\mathrm{z}}$.
-We project the application point of the force into the triangle's plane and compute its local coordinates $(x,y)$.
-We create the polynom $P$ such as
-%
-\begin{equation}
-P = F_{\mathrm{z}}(1 \hspace{0.2cm} x \hspace{0.2cm} y \hspace{0.2cm} x^2 \hspace{0.2cm} xy \hspace{0.2cm} y^2 \hspace{0.2cm} x^3 \hspace{0.2cm} xy^2 \hspace{0.2cm} y^3)^T. 
-\end{equation}
-%
-The moments at each vertex are then obtained with 
-\begin{equation}
-\Omega = (\textbf{C}^{-1})^T P.
-\end{equation}
-\end{itemize}
-
-\subsection{Allowing a curved rest configuration}
-In \cite{Przemieniecki85}, the rest and initial configurations are both assumed to be the flat one.
-The only way to have an initial deformed shape is to apply a torque on the vertices of the shells.
-However, in this configuration we would compute a displacement vector $ \mathbf{u} $ that is not null, which leads to a bending energy.
-In order to define a curved rest configuration for our shells, we propose to compute the bending energy between the initial and rest configuration during the initialisation phase, and add the energy between the current and initial configurations at runtime.
-Consequently, by separating initial and rest configurations, it is possible to have an initial deformed shape without any initial bending energy.
-In other words, we shifted the zero energy configuration.
-
-\subsection{Possible extension: parallel implementation on GPU}	\label{chap9:GPU}
-If we look closely at all the steps required to compute the local stiffness matrix for each element (see section \ref{chap9:summary} for a summary), we can notice that the operations carried out for a given shell element are independent of the other elements.
-This feature makes our co-rotational shell FEM implementation highly parallelisable.
-One potential problem however, is the assembly of the global stiffness matrix.
-Indeed, in a parallel implementation, many elements are processed concurrently.
-Inevitably, two elements sharing a node may be processed in the same time and may therefore try to both write their element contribution into the exact same location in the global stiffness matrix, and therefore the same location in GPU's memory.
-As with the TLED implementation described in chapter \ref{chap6}, there is two solutions.
-The first one would be to re-order the elements organisation so that no concurrent writes will occur.
-Unfortunately, this re-organisation of the element reduces the number of elements that can be processed in parallel.
-The second solution would be to take advantage of a new GPU NVIDIA architecture called Fermi, which added the capability atomic writes for floats.
-We recall that the write operation is said to be atomic in the sense that it is guaranteed to be performed without interference from other threads.
-In other words, the concurrent writes are simply serialised in hardware.
-Obviously, an additional overhead is to be expected.
-
-To conclude on a possible parallelisation of the algorithm, an optimal GPU implementation is always a challenge to achieve.
-One needs to choose wisely the type of memory to be used for each variable and optimise the access patterns.
-However, it seems that there is no major obstacle to a GPU implementation and our co-rotation shell FEM will eventually be ported to GPU.
-
-\TODO{enough details?}
-
-\section{Validation}
-
-We compared our model with some theoretical results reported by \cite{Zhongnian86} to assess its quality in modelling bending.
-The test that we carried out uses a square shape mesh clamped on all four edges.
-A uniform load is then applied on the square and the maximum deflection $z_{max}$ at the centre can be calculated.
-Several simulations are performed with increasing load values $q$ (ranging from $1$ to 5 N/m$^2$) and the following parameters were used:
-\begin{itemize}
- \item Young Modulus $E = 1.092 \times 10^6\,$Pa
- \item Poisson's ratio $\nu = 0.3$
- \item edge of the square $L = 10\,$m
- \item thickness $h = 0.1\,$m
- \item pressure q is a uniformly distributed load per unit area
-\end{itemize}
-
-Using these particular values it can be shown that $z_{max} = 0.126\,q$.
-The maximum deflection obtained in our simulations are reported in Table~\ref{chap9:tab-results}.
-In average we found $z_{max} = 0.1248\,q $, resulting in a $0.93\,\%$ error between our model and theoretical results on that test.
-%
-\begin{table}[ht]
-	\centering
-	\begin{tabular}{p{9cm}|c|c|}
-	\cline{2-3}
-	\multirow{5}{*}{\includegraphics[height=3.5cm]{chapter9/clamped_square.pdf}} & $q$ & $z_{max}$ \tabularnewline
-	\cline{2-3}
-	& $\,1\,$ & $\, 0.1218 \,$ \tabularnewline
-	& $\,2\,$ & $\, 0.2475 \,$ \tabularnewline		
-	& $\,3\,$ & $\, 0.3747 \,$ \tabularnewline	
-	& $\,4\,$ & $\, 0.5050 \,$ \tabularnewline		
-	& $\,5\,$ & $\, 0.6374 \,$ \tabularnewline
-	\cline{2-3}
-	\end{tabular}
-	\vspace{1cm}
-	\caption{Comparison of our shell model with theoretical results on the bending of a square plate. An error of less than $1\,\%$ was found between our simulation and theoretical results.}
-	\label{chap9:tab-results}
-\end{table}
-
-\bigskip
-
-\TODO{maybe a bit short for a so-called validation...}
-
-\section{Application to implant deployment simulation in cataract surgery}
-
-Cataract surgery consists in three main steps: capsulorhexis, phacoemulsification, and implantation of an intra-ocular lens.
-Prior to starting the surgery, a viscoelastic fluid is introduced into the capsule to facilitate capsulorhexis creation and provide protection during phacoemulsification.
-This fluid remains in the capsule for the duration of the surgery, including the injection of the implant.
-Capsulorhexis is the technique used to remove a part of the anterior lens capsule.
-Phacoemulsification consists in using a surgical device which tip vibrates at an ultrasonic frequency to emulsify the natural lens material and then aspirate the fragments of the cortical material.
-After the removal of the diseased lens, an intra-ocular lens is implanted into the eye, through a small incision (about $2\,$mm) using a foldable intra-ocular lens (see \fig{chap9:fig-surgery}).
-The foldable lens, usually made of acrylic material, is then implanted within the lens capsule through the incision used during phacoemulsification.
-In some cases the implant can be flipped or  the hooks (also called \emph{haptics}) can break.
-Therefore the simulation of the insertion and deployment of the implant is crucial for achieving a successful surgery.
-%
-\begin{figure}[ht]
-\centering 
-\subfloat[ ]{\includegraphics[height=3.2cm]{chapter9/suction.pdf}}
-\hspace{1cm} 
-\subfloat[ ]{\includegraphics[height=3.2cm]{chapter9/implant_injection_step_1.pdf}}
-\hspace{1cm} 
-\subfloat[ ]{\includegraphics[height=3.2cm]{chapter9/implant_injection_step_2.pdf}}
-\caption [Steps of cataract surgery] {(a) removal of the opacified lens by phacoemulsification.
-(b) insertion of the lens implant which is folded inside the injection device and then deploys within the lens capsule.
-(c) the lens in place in the capsule.}
-\label{chap9:fig-surgery}
-\end{figure}
-
-To simulate the insertion and deployment of an intra-ocular lens, we first created a triangulation of the lens surface.
-Particular care was given to the mesh, to ensure that areas where large stresses occur contain a higher density of elements (see \fig{chap9:fig-mesh}).
-This was done by noting the constraints applied by the surgeon to the haptics while inserting the implant within the injection device.
-During this stage, the haptics are folded onto the implant body, leading to high stresses at the junctions.
-The lens mesh contains $743$ triangles and $473$ nodes.
-Models of the injection device and the entire eye anatomy were also created.
-Physical parameters of the lens implant have been provided by the manufacturer Alcon and they are presented in Table~\ref{chap9:tab-parameters}.
-%
-\begin{table}[ht]
-	\begin{center}
-		\begin{tabular}{|p{3cm}|p{3cm}|p{3cm}|}
-		\hline
-		 \centering Young's modulus & \centering Poisson's ratio & \centering Mass density \tabularnewline
-		\hline
-		\centering $1\,$MPa & \centering $0.42$ & \centering $1.2\,$g/cm$^3$ \tabularnewline
-		\hline
-		\end{tabular}
-	\vspace{0.3cm}
-	\caption{Physical parameters of the intra-ocular implant (source: Alcon)}
-	\label{chap9:tab-parameters}
-	\end{center}
-\end{table}
-
-The first difficulty is to obtain the folded geometry of the lens within the injection device.
-This step is not important for the training process and does not need to be interactive.
-Indeed the surgeon does not always have to prepare the implant as some injection devices are readily available with a folded implant already in place.
-We simulate the folding process by first folding the haptics onto the implant body.
-The body was then bent while keeping the haptics inside to obtain the shape described in \fig{chap9:fig-implantFolding}.
-The whole process was carried out by applying the necessary forces and boundary conditions on the body and haptics of the implant.
-The folded implant was then placed into the injection device.
-The simulation of the injection consists in pushing the intra-ocular implant within the injection device into the lens capsule.
-During these stages of the simulation, complex contacts occur and consist of self collisions of the lens as well as collisions between the lens and the injector and later with the capsule.
-To solve the contacts we use the contact warping method proposed by \cite{Saupin08} as it offers an efficient way to compute physically correct contact responses in the case of co-rotational models \TODO{is it still true?
-I'm not even sure we used this technique in the final submission to ISBMS}.
-As adhesions between the haptics and the body is often observed in surgery, friction is also taken into account in the contact response process.
-%
-\begin{figure}[ht]
-\centering 
-\subfloat[Real implant]{\includegraphics[height=5cm]{chapter9/IOL.pdf}}
-\hspace{1cm} 
-\subfloat[Meshed model]{\includegraphics[height=5cm]{chapter9/mesh_implant.pdf}}
-\caption [Lens implant and its mesh] {An actual intra-ocular implant (a) and the triangular (b) mesh used in our simulations.
-Notice the higher density of elements in areas where large deformations will take place.
-Image of implant courtesy of Alcon.}
-\label{chap9:fig-mesh}
+\includegraphics[width=13.5cm]{chapter9/resultsLiver.pdf}
+\caption[Meshing of a liver with shell elements]{(a) the targeted high resolution Glisson's capsule mesh ($8\,000$ triangles).
+(b) the one-sided Hausdorff distance error map after applying only one iteration of our algorithm to the coarse mesh ($1\,200$ shells).}
+\label{chap9:fig-liver}
 \end{figure}
@@
 \centering
-\includegraphics[width=5cm]{chapter9/implant_folding1.pdf}
-\includegraphics[width=4.5cm]{chapter9/implant_folding2.pdf} \\
-\includegraphics[width=4cm]{chapter9/implant_folding3.pdf}
-\includegraphics[width=7cm]{chapter9/implant_folding4.pdf}
-\caption [Folding of intra-ocular implant] {Top: intermediary steps of the intra-ocular implant folding. Bottom: fully folded implant ready to be placed into the injection device.}
-\label{chap9:fig-implantFolding}
+\includegraphics[width=13.5cm]{chapter9/resultsAneurysm.pdf}
+\caption[Meshing of an aneurysm with shell elements]{(a) the targeteted high resolution aneurysm mesh ($28\,368$ triangles).
+(b): the one-sided Hausdorff distance error map on a mesh of 772 shells generated with our method.}
+\label{chap9:fig-aneurysm}
 \end{figure}
 
-Results of our simulation are illustrated in \fig{chap9:fig-simu-results}.
-We can notice the progressive deployment of the implant when it exits the injector.
-The shape of the intra-ocular lens remains very close to that of a real one during all stages of the simulation: within the injector, during the ejection phase, and when in place within the capsule.
-Due to the high stiffness and low mass of the lens, a direct sparse solver was used at each time step (dt $= 0.01\,$s) rather than an iterative solver, resulting in a more accurate and more stable simulation, to the detriment of computation times (about 5 FPS for the complete simulation, and about 10 FPS for the deformation only, on a 2.4 GHz processor).
-
-% Link where images from surgery have been found: http://asianophthalmology.org/articles/?a=articles&p=6
+\subsubsection*{Computation times.}
+We perform several tests on the aneurysm model at different resolutions to measure computation times.
+The shells are resisting to a uniform pressure load and solved using a Conjugate Gradient (CG) iterative solver.
+The computation times are reported in \fig{chap9:fig-computation}.
+Implicit integration allows for large time steps ($40\,ms$) and the computation is real-time for 800 shell elements and a reasonable error criterion ($5\,\%$).
+When the computation time must be bounded (critical real-time applications), one can fix the number of CG iterations to, for instance, $100$ and remains real-time for $1\,000$ shell elements.
+However, in that case the accuracy of the results is not checked.
+%
 \begin{figure}[ht]
 \centering
-\includegraphics[width=4.5cm]{chapter9/surgery1.pdf}
-\hfill
-\includegraphics[width=4.5cm]{chapter9/surgery2.pdf}
-\hfill
-\includegraphics[width=4.5cm]{chapter9/surgery3.pdf} \\
-\vspace{0.1cm}
-\includegraphics[width=4.5cm]{chapter9/simu1.pdf}
-\hfill
-\includegraphics[width=4.5cm]{chapter9/simu2.pdf}
-\hfill
-\includegraphics[width=4.5cm]{chapter9/simu3.pdf}
-\caption [Lens imlant] {Three steps of the simulation of the intra-ocular lens implant injection and its deployment within the lens capsule.
-Top: images from a real cataract surgery, courtesy of Dr.
-Tarek Youssef, ophtalmologist in Canada.
-Below: our simulation of the implant's deployment.}
-\label{chap9:fig-simu-results}
+\includegraphics[width=13.5cm]{chapter9/computation_times.pdf}
+\caption {Computation time on meshes of 200, 400, 600, 800 and $ 1\,000 $ elements.}.
+\label{chap9:fig-computation}
 \end{figure}
+
+\subsection{Coupling between tetrahedra and shells for advanced modelling}
+Structures in human body can be either solid (brain, liver, prostate etc.) or hollow (colon, blood vessels, stomach etc.).
+However, knowing how to model the two kind of structures is not sufficient to reach a high degree of accuracy, real life situations are more complex.
+As an example, the external surface of the liver is covered by a layer of cells called Glisson's capsule.
+Its interaction with the liver plays an important role into the overall structure's mechanical behaviour.
+Therefore considering the interaction between solid and hollow objects is as crucial as modelling the two structures separately.
+
+An example of medical procedure to illustrate this point even further is angioplasty.
+Angioplasty is the technique of mechanically widening a narrowed or obstructed blood vessel, typically as a result of atherosclerosis.
+An empty and collapsed balloon on a guide wire is passed into the narrowed locations and then inflated to a fixed size.
+The balloon crushes the fatty deposits, so opening up the blood vessel to improved flow.
+As a proof of concept, we tried to simulate an angioplasty.
+Our preliminary results are presented on \fig{chap9:fig-stent}.
+The blood vessel is modelled using the shell FEM formulation described in this paper and the fatty deposits are simulated with a tetrahedral FEM method and are fixed to the interior wall of the blood vessel.
+When the balloon inflates it crushes the deposits and they then apply a pressure onto the curved surfaces of shells modelling the interior wall.
+The forces are then distributed onto the mechanical nodes of the blood vessel mesh as detailed in section~\ref{chap8:interactions}, which widens the blood vessel as expected.
+%
+\begin{figure}[ht]
+\centering
+\includegraphics[width=12cm]{chapter9/stenose_final.pdf}
+\caption[Simulation of an angioplasty procedure]{Simulation of an angioplasty procedure.
+(a, c): A collapsed stent is inserted into the blood vessel simulated with our shell FEM formulation.
+The fatty deposits (in yellow) are modelled with tetrahedral FEM.
+(b, d): The stent is crushing the fatty deposits which creates a pressure onto the interior wall and widens the blood vessel.}
+\label{chap9:fig-stent}
+\end{figure}
+
+\section{Discussion}
+
+Although we were able to mesh complex anatomical structures with a fairly small number of shell elements, our approach could be further improved.
+In our current implementation, if the error measured with the Hausdorff distance is greater than a threshold (which is also difficult to choose), all triangular elements are subdivided into 4 new triangles and each of the new vertices is moved towards the targeted geometry based on our heuristic algorithm.
+Obviously, this is not optimal as the error between the two meshes may very well be local and therefore the use of a global mesh refinement is not very suitable.
+The second weakness is that we also do not check the quality of the elements during our mesh process.
+If some elements are stretched from the heavy decimation, these elements are not corrected during the process of re-meshing.
+Our algorithm could even worsen the situation by splitting those stretched elements into four smaller (and more stretched) elements.
+One solution for the latter issue would be to enhance our algorithm with a relaxation phase where a repulsion force is applied between each vertex (constrained to remain on the surface of the object).
+This physical approach would allow a more even distribution of the vertices across the surface and would contribute to obtain better quality elements overall.
+Yet, meshing a surface with curved elements is a very complex problem and still an active area of research.
+In contrast to all previous geometrical techniques, we suggested a more physical approach consistent with our shell FEM formulation and obtained good results in meshing complex anatomical structures.
+
+\section{Conclusion}
+
+Over the last two chapters, we proposed a complete framework for real-time modelling of thin anatomical structures.
+The novelty of our method relies on the combination of a shell finite element formulation and a geometric surface reconstruction both based on the same polynomial interpolation function used to describe the surface of shells.
+We also show how contacts and interactions with the curved surfaces of shells can be handled using the same function.
+In summary, these polynomial shape functions are used in three different ways in our computational model:
+%
+\begin{description}
+\item[(a)] to approximate complex geometrical shapes
+\item[(b)] to compute internal forces via a co-rotational shell FEM
+\item[(c)] to compute contact forces onto a curved triangle
+\end{description}
+
+We achieved our objectives to propose a versatile solution which can simulate, in real-time, thin objects with various shapes and material properties with good accuracy.
+The efficiency of the method was illustrated through shell-based reconstruction and real-time simulation of the deformations of various anatomical structures and other thin objects (like an intra-ocular implant).
+We also presented preliminary results on the coupling between solid (tetrahedra) and thin objects (shells) for the advanced modelling of anatomical structures via the simulation of an angioplasty procedure.
+This work on modelling thin anatomical structures was the object of two main publications \citep{Comas2010a,Comas2010b}.

### 2012-01-04 Doc/Miccai/miccai2012.tex
add Miccai 2012 paper
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@281 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

%%%%%%%%%%%%%%%%%%%%%%% file typeinst.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is the LaTeX source for the instructions to authors using
% the LaTeX document class 'llncs.cls' for contributions to
% the Lecture Notes in Computer Sciences series.
% http://www.springer.com/lncs       Springer Heidelberg 2006/05/04
%
% It may be used as a template for your own input - copy it
% to a new file with a new name and use it as the basis
% for your article.
%
% NB: the document class 'llncs' has its own and detailed documentation, see
% ftp://ftp.springer.de/data/pubftp/pub/tex/latex/llncs/latex2e/llncsdoc.pdf
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[runningheads,a4paper]{llncs}

\usepackage[colorlinks=true,linkcolor=black,citecolor=black,pdfnewwindow=true,urlcolor=blue,filecolor=blue,baseurl=./]{hyperref}
\usepackage{url}

% \newcommand{\mycite}[2]{\url{#1}{\cite{#2}}}
% \newcommand{\mycite}[2]{\url{#1}:\cite{#2}}
% \newcommand{\mycite}[2]{\href{http://twitter.com/home}{Twitter}.}
%\newcommand{\mycite}[2]{\href{./refs/#1}{\cite{#2}}}
\newcommand{\mycite}[2]{\cite{#2}}
%\newcommand{\myciteurl}[2]{\href{#1}{\cite{#2}}}

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}

\usepackage{url}
%\urldef{\mailsa}\path|{alfred.hofmann, ursula.barth, ingrid.haas, frank.holzwarth,|
%\urldef{\mailsb}\path|anna.kramer, leonie.kunz, christine.reiss, nicole.sator,|
%\urldef{\mailsc}\path|erika.siebert-cole, peter.strasser, lncs}@springer.com|    
%\newcommand{\keywords}[1]{\par\addvspace\baselineskip
%\noindent\keywordname\enspace\ignorespaces#1}

\usepackage{color}
\usepackage{amsmath}
\usepackage[tight]{subfigure}
\usepackage{listings}
\definecolor{darkgray}{rgb}{0.2,0.2,0.2}

\lstset{language=c++,basicstyle=\scriptsize\sffamily,tabsize=2,framexleftmargin=5mm, frame=shadowbox, framesep=3pt,rulesepcolor=\color{darkgray},rulesep=.100pt,keywordstyle=\bf\color{blue},commentstyle=\color{magenta},stringstyle=\color{red},numbers=left,numberstyle=\tiny,numbersep=10pt,breaklines=true,showstringspaces=false,columns=flexible}

\newcommand{\tomas}[1]{{\color{red}{\textbf{Tom: #1}}}}
\newcommand{\stefan}[1]{{\color{magenta}{\textbf{St: #1}}}}
\newcommand{\christian}[1]{{\color{green}{\textbf{Ch: #1}}}}
\newcommand{\tobias}[1]{{\color{bleu}{\textbf{Tob: #1}}}}

\newcommand{\myvec}[1]{\ensuremath{\mathbf{#1}}}
\newcommand{\mymat}[1]{\ensuremath{\mathbf{#1}}}
\newcommand{\Vx}{\myvec{x}} % position vector
\newcommand{\Vv}{\myvec{v}} % velocity vector
\newcommand{\Va}{\myvec{a}} % acceleration vector
\newcommand{\Vdv}{\myvec{dv}} % change of velocity vector (unknown in implicit CG, and used in constraint solver

% \newcommand{\mysubsection}[1]{\noindent \textbf{#1: }}
\newcommand{\mysubsection}[1]{\textbf{#1: }}
% \newcommand{\mysubsection}[1]{\subsection{#1}}

%\usepackage{miseenpage}

\begin{document}

\mainmatter            % start of your contributions

\title{Towards planning of infantile cardiac surgery based on simulation}

\author{PaperID:}

\maketitle

\begin{abstract}
In this paper...
\end{abstract}

\section{Introduction}

\christian{We can put any comment...}
\tomas{People have different colors}

Position vector: $\Vx$

Matrices are written as $\mymat{M}$

\bibliographystyle{abbrv}
\bibliography{miccai2012}

\end{document}

### 2012-01-18 Doc/Miccai/miccai2012.tex
some initial draft of the article, nothing serious
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@287 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \newcommand{\Va}{\myvec{a}} % acceleration vector
-\newcommand{\Vdv}{\myvec{dv}} % change of velocity vector (unknown in implicit CG, and used in constraint solver
+\newcommand{\Vdv}{\myvec{dv}} % change of velocity vector (unknown in implicit CG, and used in constraint solver)
 
@@
 
-\author{PaperID:}
+\author{PaperID:} % TODO
 
@@
 \begin{abstract}
+% TODO
 In this paper...
@@
 
-\section{Introduction}
+\section{Introduction} % {{{
 
-\christian{We can put any comment...}
-\tomas{People have different colors}
+%%%%%%
+%
+%\christian{We can put any comment...}
+%\tomas{People have different colors}
+%
+%Position vector: $\Vx$
+%
+%Matrices are written as $\mymat{M}$
+%
+%%%%%%
 
-Position vector: $\Vx$
+The congenital heart diseases have become quite common \tomas{\ldots}.
+While they are very much treatable there are a lot of difficulties involved
+with the process. There is rarely only a single flaw, the lesions usually
+appear in multiples. Since most of the lesions would be fatal if untreated
+the first surgery has to be performed shortly after the birth, often
+followed with several other surgeries before the age of one year. At such
+young age the heart and blood vessels are very small which requires skills
+of the surgeon. Further more many of the critical decisions cannot be
+concluded from the CT scan or artificial model and have to be done on site
+(e.g. how exactly perform the cut and stitching). This is however in direct
+contrast with the requirement on the speed at which the operation has to be
+performed. During the surgery the blood flow has to be sustained by other
+means (\tomas{pig's heart?}) which puts overall pressure on the organism
+and thus cannot be kept for too long.
 
-Matrices are written as $\mymat{M}$
+\tomas{interesting ref. \cite{Barron2012}}
+
+\tomas{Stefan, please review this. It is just the notion that I have got
+about the problem and might not be completely correct}
+
+For these reasons there is a demand for tools helping with the
+process of pre-operative planning.
+
+\tomas{Are there other approaches to this or similar problem? My search
+did not reveal anything interesting.}
+
+Some of the most common actions in the procedure are
+
+\begin{itemize}
+    \item cutting out a piece of a blood vessel to remove an obstruction or
+    unwanted junction,
+    \item opening a narrow vessel by a cut and closing the hole by an
+    artificial material.
+    \item \tomas{more ideas?}
+\end{itemize}
+
+\tomas{add images}
+
+Both procedures are complicated from the surgeons point of view. In the
+first case the surgeon has to make an educated guess how to perform a cut
+so that after stitching the rest together the blood flow is not obstructed
+and the blood pressure at the seam is not too large so as to tear open the
+vessel. In the second case the surgeon has no prior idea about the shape
+and size of the patch he will need until he performs the cut. Therefor the
+patch has to be cut on site and cannot be prepared in advance.
+
+We are proposing new approach to pre-operative planning of cardiac surgery.
+Using shell elements to model deformations of the blood vessel we help with
+the decision making.
+
+% }}}
+
+\tomas{What to describe? -- shell elements/FEM/mechanics?}
+\tomas{shortly mention: rest shape}
+
+\section{Method Description}
+
+Outline:
+
+\begin{itemize}
+
+    \item we use shell elements to model blood vessels, dynamic system
+        (implicit integration), in Sofa framework \tomas{cite Olivier
+        \cite{Comas2010b,Comas2010c} -- shells in corotational environment}
+    \item using springs to connect the objects is not good --> generates
+        excessive amount of energy
+    \item to eliminate this we use one mesh for the simulation and
+        different mesh to describe the rest shape and we leave the dynamic
+        system to stabilize
+    \item to reduce more the generation of excessive energies in the system
+        at the beginning of the simulation we repetitively change the rest
+        shape (linear interpolation)
+\end{itemize}
+
+\section{Conclusion}
+\subsection{Verification}
+
+Verification is difficult -- limited or no knowledge of real state after
+the surgery (can't put the child under CT/MRI again). \ldots
 
@@
 \end{document}
+
+% Some Vim setup. If it interferes with anyones work feel free to remove
+% that.
+% vim: et tw=75 fdm=marker spell

### 2012-02-21 Doc/Miccai/miccai2012.tex
article: shells and method description
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@320 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \newcommand{\Vdv}{\myvec{dv}} % change of velocity vector (unknown in implicit CG, and used in constraint solver)
+\newcommand{\deriv}[2]{\frac{\partial #1}{\partial #2}} % Partial derivative
 
@@
 process of pre-operative planning.
+\tomas{Usefulness of computer aid --
+numerical models:
+3D visualisation: Hemminger2005}
 
@@
 
-\tomas{What to describe? -- shell elements/FEM/mechanics?}
-\tomas{shortly mention: rest shape}
+\section{Elastic Shell Elements}
+
+To model the blood vessels we employed shell elements in the corotational
+environment. The main advantage of shells is their physical accuracy. They
+are based on the mechanics of continuum. Also the parameters defining the
+parameters of the simulation have clear physical interpretation (unlike for
+example for mass-spring model).
+
+The common objection against using shell elements in real-time simulations
+is the numerical complexity. While this is technically valid, it is not
+good enough reason. Using low resolution mesh we can obtain results with
+good visual and physical accuracy sufficient for real-time
+simulation.\tomas{isn't this too strong?} Given the bending nature of the
+shell one can employ interpolation of the shell surface to improve the
+visual mesh presented to the user. \tomas{cite Olivier and our unpublished
+results}
+
+A shell element is a combination of two other elements: elastic membrane
+and bending plate element. While the membrane element is used to compute
+in-plane forces of the shell (like stretching or shearing), the bending
+plate elements on the other hand describes the bending properties.
+
+\tomas{add image showing the difference}
+
+\tomas{thin in one dimension}
+
+\subsection{Mechanics of Continuum Preliminaries} % {{{
+
+\tomas{recheck this. btw do we really want this?}
+
+The mathematical formulation follows the steps defined by the mechanics of
+continuum.\tomas{ref. to some text book}
+
+For every particle $\myvec{x} \in \Omega$ of the deformable body $\Omega$
+the deformation $ \phi : \Omega' \to \Omega $ is a mapping from the rest
+shape $\myvec{x}' \in \Omega'$ to it's current shape
+
+$$ \phi(x) = \myvec{x}' + u(x) $$
+
+Where $u(x)$ is the displacement for particle $x$. From the Cauchy's strain
+tensor
+
+$$
+\epsilon = \frac{1}{2}\left( \nabla^T u + \nabla u \right)
+$$
+
+\noindent%
+we can compute the stress by applying a constitutive law. In our simulation
+we are using linear Hooke's law to keep the numerical formulation simple
+enough to be able to compute it in real-time. The stress is computed by
+
+$$
+\sigma = \mymat{M}\epsilon
+$$
+
+\noindent%
+where $\mymat{M}$ is the material matrix. Further from the stress and
+strain we can compute the strain energy
+
+$$
+E = \int_{\Omega} \epsilon^T \sigma \,\mathrm{d}\Omega
+$$
+
+\tomas{now should be discretization}
+
+%}}}
+
+\subsection{shells...}
+
+In this subsection we will briefly summarize how the numerical formulation
+of the linear model is derived. For more thorough description we refer the
+reader to \tomas{add ref.} Let's define $ u_x, u_y $ to be the two
+in-plane displacements and $ u_z $ the off-plane function of deflection. We
+define the strain-displacement matrices in the following way. For the
+membrane element (using the Cauchy's strain tensor)
+
+\begin{equation}
+  \mymat{B_m} = \left[ \begin{matrix}
+    \deriv{u_x}{x} \\
+    \deriv{u_y}{y} \\
+    \deriv{u_x}{y} + \deriv{u_y}{x}
+  \end{matrix} \right]
+\end{equation}
+
+\noindent%
+and for the bending plate element (using the Kirchoff-Love theory)
+
+\begin{equation}
+  \mymat{B_b} = \left[ \begin{matrix}
+    - z \deriv{^2 u_z}{x^2} \\
+    - z \deriv{^2 u_z}{y^2} \\
+    - 2z \deriv{^2 u_z}{xy}
+  \end{matrix} \right]
+\end{equation}
+
+From these we can easily compute the stiffness matrices $K_m$ and $K_b$
+(for membrane and bending plate element respectively) by integrating over
+the volume of the shell
+
+\begin{equation}
+  \mymat{K}_i = \int_V \mymat{B}_i^T \mymat{M} \mymat{B}_i \,\mathrm{d}V
+\end{equation}
+
+\noindent%
+where $i \in \{\mathrm{b},\mathrm{m}\}$ and $\mymat{M}$ is the material
+matrix defined by Hooke's law:
+
+\begin{equation}
+  \mymat{M} = \frac{E}{1-\nu^2} \left[ \begin{matrix}
+    1   & \nu & 0 \\
+    \nu & 1   & 0 \\
+    0   & 0   & \frac{1-\nu}{2}
+  \end{matrix} \right]
+\end{equation}
+
+The only two values defining the matrix value are Poisson's ratio $\nu$ and
+the Young's modulus $E$.
+
+\subsection{Co-rotational Formulation}
+
+While the Cauchy's strain used in the formulation of the shell element is
+invariant to rigid body translation, it is known to be rotationally
+noninvariant. This produces ghost deformations just by rigid rotation. One
+way to solve the problem is using Green's strain tensor instead whose
+definition is, however, non-linear. Therefore we to overcome the problem we
+chose to use the corotational formulation instead.\tomas{add ref;
+maybe short description?}
 
 \section{Method Description}
+
+\tomas{clearly state that this is not a complete solution where user points
+and clicks -- it's just a proof-of-concept}
+
+In this chapter we will elaborate on the internal aspects of the
+simulation.
 
@@
 \begin{itemize}
-
     \item we use shell elements to model blood vessels, dynamic system
@@
         system to stabilize
+        -- we need to map the nodes between topologies
     \item to reduce more the generation of excessive energies in the system
@@
 
-\section{Conclusion}
-\subsection{Verification}
+We are aiming for non-interactive simulation without requiring the
+surgeon to go through the whole process of the surgery. This introduces several
+problems we have to deal with.
+
+The first problem is how to deform the model so that the requested parts
+can be sutured together. And second how to simulate the suture.
+The simplest solution here would be connecting the respective nodes
+together with springs. Such solution, however, has a disadvantage of adding
+excessive energies to the system.\tomas{do we need a ref.?}
+
+%\section{Conclusion}
+%\subsection{Verification}
 
 Verification is difficult -- limited or no knowledge of real state after
-the surgery (can't put the child under CT/MRI again). \ldots
+the surgery (too risky to put the child under CT/MRI again).
+
+Therefore we are interested in the geometrical evaluation of the result
+i.e. no excessive elongations, expected deformations, independence on
+physical parameters (E/nu),\ldots or something like that \ldots
 
@@
 % that.
-% vim: et tw=75 fdm=marker spell
+% vim: et sw=2 tw=75 fdm=marker spell

### 2012-02-22 Doc/Miccai/miccai2012.tex
more on article
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@326 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\subsection{shells...}
+\subsection{shells...} % {{{
 
@@
 
+% }}}
+
 \subsection{Co-rotational Formulation}
@@
 
-\section{Method Description}
+\section{Method Description} % {{{
 
@@
 
-%\section{Conclusion}
+We propose a different approach which uses a two different mesh topologies.
+The basic idea is to transform the mesh by joining the edges and nodes
+where the suturing is supposed to take place and let the dynamic system
+find a stable configuration. Such stable configuration is the result we are
+looking for. This requires us to provide a mapping to the underlying shell
+element code that would translate between nodes of the topology of the rest
+shape and the topology of the deformed mesh. Such a mapping is 1:N for
+nodes where the suturing takes place. \tomas{add image} To make it properly
+defined elements have to be taken into account also:
+
+\begin{equation}
+  \phi : N_d \times E_d -> N_r
+\end{equation}
+
+In the definition of this mapping $\phi$ the $N_d$ and $E_d$ are sets of
+nodes and elements of the deformed mesh and $N_r$ is set of nodes of the
+deformed mesh.
+
+\tomas{what more?}
+
+Another problem that a predictive simulation should deal with is filling
+the missing steps that would normally had to be performed fluently by the
+surgeon. For example when suturing two blood vessels together they have to
+be placed close to one another first. Sometimes relatively large distances
+have to be overcome. By ignoring this fact our approach with joining nodes
+and edges could generate unwanted forces at the start of the simulation. To
+alleviate large initial deformations for the elements we employed
+interpolation ...
+
+% }}}
+
+\section{Conclusion}
 %\subsection{Verification}
@@
 % that.
-% vim: et sw=2 tw=75 fdm=marker spell
+% vim: et sw=2 tw=75 fdm=marker fdc=2 spell

### 2012-02-23 Doc/Miccai/miccai2012.tex
New paper structure...
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@328 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2012-01-18 Doc/Miccai/miccai2012.tex
@@
 \begin{abstract}
-% TODO
-In this paper...
 \end{abstract}
 
-\section{Introduction} % {{{
+\section{Introduction}
+State of the Art
 
-%%%%%%
-%
-%\christian{We can put any comment...}
-%\tomas{People have different colors}
-%
-%Position vector: $\Vx$
-%
-%Matrices are written as $\mymat{M}$
-%
-%%%%%%
+Figure illustrating a surgery
 
-The congenital heart diseases have become quite common \tomas{\ldots}.
-While they are very much treatable there are a lot of difficulties involved
-with the process. There is rarely only a single flaw, the lesions usually
-appear in multiples. Since most of the lesions would be fatal if untreated
-the first surgery has to be performed shortly after the birth, often
-followed with several other surgeries before the age of one year. At such
-young age the heart and blood vessels are very small which requires skills
-of the surgeon. Further more many of the critical decisions cannot be
-concluded from the CT scan or artificial model and have to be done on site
-(e.g. how exactly perform the cut and stitching). This is however in direct
-contrast with the requirement on the speed at which the operation has to be
-performed. During the surgery the blood flow has to be sustained by other
-means (\tomas{pig's heart?}) which puts overall pressure on the organism
-and thus cannot be kept for too long.
+Requirements: Model blood vessels/wall deformation, surgical low-level tasks
 
-\tomas{interesting ref. \cite{Barron2012}}
+Plan
 
-\tomas{Stefan, please review this. It is just the notion that I have got
-about the problem and might not be completely correct}
+\section{Blood Vessel Model}
 
-For these reasons there is a demand for tools helping with the
-process of pre-operative planning.
+Thin shell model
 
-\tomas{Are there other approaches to this or similar problem? My search
-did not reveal anything interesting.}
+Eq. showing rest position stuff
 
-Some of the most common actions in the procedure are
+Fullfills requirements + collision
 
-\begin{itemize}
-    \item cutting out a piece of a blood vessel to remove an obstruction or
-    unwanted junction,
-    \item opening a narrow vessel by a cut and closing the hole by an
-    artificial material.
-    \item \tomas{more ideas?}
-\end{itemize}
+Test with tube + convergence
 
-\tomas{add images}
+\section{Simulation of Surgical Procedures}
 
-Both procedures are complicated from the surgeons point of view. In the
-first case the surgeon has to make an educated guess how to perform a cut
-so that after stitching the rest together the blood flow is not obstructed
-and the blood pressure at the seam is not too large so as to tear open the
-vessel. In the second case the surgeon has no prior idea about the shape
-and size of the patch he will need until he performs the cut. Therefor the
-patch has to be cut on site and cannot be prepared in advance.
+surgical low-level tasks done with rest position trick (joining, patch, complex cutting/suturing)
 
-We are proposing new approach to pre-operative planning of cardiac surgery.
-Using shell elements to model deformations of the blood vessel we help with
-the decision making.
+\section{Results and Discussion}
 
-% }}}
+Example with 3 diff. surgical procedures
 
-\tomas{What to describe? -- shell elements/FEM/mechanics?}
-\tomas{shortly mention: rest shape}
+Parameters
 
-\section{Method Description}
-
-Outline:
-
-\begin{itemize}
-
-    \item we use shell elements to model blood vessels, dynamic system
-        (implicit integration), in Sofa framework \tomas{cite Olivier
-        \cite{Comas2010b,Comas2010c} -- shells in corotational environment}
-    \item using springs to connect the objects is not good --> generates
-        excessive amount of energy
-    \item to eliminate this we use one mesh for the simulation and
-        different mesh to describe the rest shape and we leave the dynamic
-        system to stabilize
-    \item to reduce more the generation of excessive energies in the system
-        at the beginning of the simulation we repetitively change the rest
-        shape (linear interpolation)
-\end{itemize}
+Limitations (getting image data, meshes, dynamic remeshing, same number of nodes)
 
 \section{Conclusion}
-\subsection{Verification}
 
-Verification is difficult -- limited or no knowledge of real state after
-the surgery (can't put the child under CT/MRI again). \ldots
+%\begin{abstract}
+%% TODO
+%In this paper...
+%\end{abstract}
+%
+%\section{Introduction} % {{{
+%
+%%%%%%%
+%%
+%%\christian{We can put any comment...}
+%%\tomas{People have different colors}
+%%
+%%Position vector: $\Vx$
+%%
+%%Matrices are written as $\mymat{M}$
+%%
+%%%%%%%
+%
+%The congenital heart diseases have become quite common \tomas{\ldots}.
+%While they are very much treatable there are a lot of difficulties involved
+%with the process. There is rarely only a single flaw, the lesions usually
+%appear in multiples. Since most of the lesions would be fatal if untreated
+%the first surgery has to be performed shortly after the birth, often
+%followed with several other surgeries before the age of one year. At such
+%young age the heart and blood vessels are very small which requires skills
+%of the surgeon. Further more many of the critical decisions cannot be
+%concluded from the CT scan or artificial model and have to be done on site
+%(e.g. how exactly perform the cut and stitching). This is however in direct
+%contrast with the requirement on the speed at which the operation has to be
+%performed. During the surgery the blood flow has to be sustained by other
+%means (\tomas{pig's heart?}) which puts overall pressure on the organism
+%and thus cannot be kept for too long.
+%
+%\tomas{interesting ref. \cite{Barron2012}}
+%
+%\tomas{Stefan, please review this. It is just the notion that I have got
+%about the problem and might not be completely correct}
+%
+%For these reasons there is a demand for tools helping with the
+%process of pre-operative planning.
+%\tomas{Usefulness of computer aid --
+%numerical models:
+%3D visualisation: Hemminger2005}
+%
+%\tomas{Are there other approaches to this or similar problem? My search
+%did not reveal anything interesting.}
+%
+%Some of the most common actions in the procedure are
+%
+%\begin{itemize}
+%    \item cutting out a piece of a blood vessel to remove an obstruction or
+%    unwanted junction,
+%    \item opening a narrow vessel by a cut and closing the hole by an
+%    artificial material.
+%    \item \tomas{more ideas?}
+%\end{itemize}
+%
+%\tomas{add images}
+%
+%Both procedures are complicated from the surgeons point of view. In the
+%first case the surgeon has to make an educated guess how to perform a cut
+%so that after stitching the rest together the blood flow is not obstructed
+%and the blood pressure at the seam is not too large so as to tear open the
+%vessel. In the second case the surgeon has no prior idea about the shape
+%and size of the patch he will need until he performs the cut. Therefor the
+%patch has to be cut on site and cannot be prepared in advance.
+%
+%We are proposing new approach to pre-operative planning of cardiac surgery.
+%Using shell elements to model deformations of the blood vessel we help with
+%the decision making.
+%
+%% }}}
+%
+%\section{Elastic Shell Elements}
+%
+%To model the blood vessels we employed shell elements in the corotational
+%environment. The main advantage of shells is their physical accuracy. They
+%are based on the mechanics of continuum. Also the parameters defining the
+%parameters of the simulation have clear physical interpretation (unlike for
+%example for mass-spring model).
+%
+%
+%The common objection against using shell elements in real-time simulations
+%is the numerical complexity. While this is technically valid, it is not
+%good enough reason. Using low resolution mesh we can obtain results with
+%good visual and physical accuracy sufficient for real-time
+%simulation.\tomas{isn't this too strong?} Given the bending nature of the
+%shell one can employ interpolation of the shell surface to improve the
+%visual mesh presented to the user. \tomas{cite Olivier and our unpublished
+%results}
+%
+%A shell element is a combination of two other elements: elastic membrane
+%and bending plate element. While the membrane element is used to compute
+%in-plane forces of the shell (like stretching or shearing), the bending
+%plate elements on the other hand describes the bending properties.
+%
+%\tomas{add image showing the difference}
+%
+%\tomas{thin in one dimension}
+%
+%\subsection{Mechanics of Continuum Preliminaries} % {{{
+%
+%\tomas{recheck this. btw do we really want this?}
+%
+%The mathematical formulation follows the steps defined by the mechanics of
+%continuum.\tomas{ref. to some text book}
+%
+%For every particle $\myvec{x} \in \Omega$ of the deformable body $\Omega$
+%the deformation $ \phi : \Omega' \to \Omega $ is a mapping from the rest
+%shape $\myvec{x}' \in \Omega'$ to it's current shape
+%
+%$$ \phi(x) = \myvec{x}' + u(x) $$
+%
+%Where $u(x)$ is the displacement for particle $x$. From the Cauchy's strain
+%tensor
+%
+%$$
+%\epsilon = \frac{1}{2}\left( \nabla^T u + \nabla u \right)
+%$$
+%
+%\noindent%
+%we can compute the stress by applying a constitutive law. In our simulation
+%we are using linear Hooke's law to keep the numerical formulation simple
+%enough to be able to compute it in real-time. The stress is computed by
+%
+%$$
+%\sigma = \mymat{M}\epsilon
+%$$
+%
+%\noindent%
+%where $\mymat{M}$ is the material matrix. Further from the stress and
+%strain we can compute the strain energy
+%
+%$$
+%E = \int_{\Omega} \epsilon^T \sigma \,\mathrm{d}\Omega
+%$$
+%
+%\tomas{now should be discretization}
+%
+%%}}}
+%
+%\subsection{shells...} % {{{
+%
+%In this subsection we will briefly summarize how the numerical formulation
+%of the linear model is derived. For more thorough description we refer the
+%reader to \tomas{add ref.} Let's define $ u_x, u_y $ to be the two
+%in-plane displacements and $ u_z $ the off-plane function of deflection. We
+%define the strain-displacement matrices in the following way. For the
+%membrane element (using the Cauchy's strain tensor)
+%
+%\begin{equation}
+%  \mymat{B_m} = \left[ \begin{matrix}
+%    \deriv{u_x}{x} \\
+%    \deriv{u_y}{y} \\
+%    \deriv{u_x}{y} + \deriv{u_y}{x}
+%  \end{matrix} \right]
+%\end{equation}
+%
+%\noindent%
+%and for the bending plate element (using the Kirchoff-Love theory)
+%
+%\begin{equation}
+%  \mymat{B_b} = \left[ \begin{matrix}
+%    - z \deriv{^2 u_z}{x^2} \\
+%    - z \deriv{^2 u_z}{y^2} \\
+%    - 2z \deriv{^2 u_z}{xy}
+%  \end{matrix} \right]
+%\end{equation}
+%
+%From these we can easily compute the stiffness matrices $K_m$ and $K_b$
+%(for membrane and bending plate element respectively) by integrating over
+%the volume of the shell
+%
+%\begin{equation}
+%  \mymat{K}_i = \int_V \mymat{B}_i^T \mymat{M} \mymat{B}_i \,\mathrm{d}V
+%\end{equation}
+%
+%\noindent%
+%where $i \in \{\mathrm{b},\mathrm{m}\}$ and $\mymat{M}$ is the material
+%matrix defined by Hooke's law:
+%
+%\begin{equation}
+%  \mymat{M} = \frac{E}{1-\nu^2} \left[ \begin{matrix}
+%    1   & \nu & 0 \\
+%    \nu & 1   & 0 \\
+%    0   & 0   & \frac{1-\nu}{2}
+%  \end{matrix} \right]
+%\end{equation}
+%
+%The only two values defining the matrix value are Poisson's ratio $\nu$ and
+%the Young's modulus $E$.
+%
+%% }}}
+%
+%\subsection{Co-rotational Formulation}
+%
+%While the Cauchy's strain used in the formulation of the shell element is
+%invariant to rigid body translation, it is known to be rotationally
+%noninvariant. This produces ghost deformations just by rigid rotation. One
+%way to solve the problem is using Green's strain tensor instead whose
+%definition is, however, non-linear. Therefore we to overcome the problem we
+%chose to use the corotational formulation instead.\tomas{add ref;
+%maybe short description?}
+%
+%
+%\section{Method Description} % {{{
+%
+%\tomas{clearly state that this is not a complete solution where user points
+%and clicks -- it's just a proof-of-concept}
+%
+%In this chapter we will elaborate on the internal aspects of the
+%simulation.
+%
+%Outline:
+%
+%\begin{itemize}
+%    \item we use shell elements to model blood vessels, dynamic system
+%        (implicit integration), in Sofa framework \tomas{cite Olivier
+%        \cite{Comas2010b,Comas2010c} -- shells in corotational environment}
+%    \item using springs to connect the objects is not good --> generates
+%        excessive amount of energy
+%    \item to eliminate this we use one mesh for the simulation and
+%        different mesh to describe the rest shape and we leave the dynamic
+%        system to stabilize
+%        -- we need to map the nodes between topologies
+%    \item to reduce more the generation of excessive energies in the system
+%        at the beginning of the simulation we repetitively change the rest
+%        shape (linear interpolation)
+%\end{itemize}
+%
+%We are aiming for non-interactive simulation without requiring the
+%surgeon to go through the whole process of the surgery. This introduces several
+%problems we have to deal with.
+%
+%The first problem is how to deform the model so that the requested parts
+%can be sutured together. And second how to simulate the suture.
+%The simplest solution here would be connecting the respective nodes
+%together with springs. Such solution, however, has a disadvantage of adding
+%excessive energies to the system.\tomas{do we need a ref.?}
+%
+%We propose a different approach which uses a two different mesh topologies.
+%The basic idea is to transform the mesh by joining the edges and nodes
+%where the suturing is supposed to take place and let the dynamic system
+%find a stable configuration. Such stable configuration is the result we are
+%looking for. This requires us to provide a mapping to the underlying shell
+%element code that would translate between nodes of the topology of the rest
+%shape and the topology of the deformed mesh. Such a mapping is 1:N for
+%nodes where the suturing takes place. \tomas{add image} To make it properly
+%defined elements have to be taken into account also:
+%
+%\begin{equation}
+%  \phi : N_d \times E_d -> N_r
+%\end{equation}
+%
+%In the definition of this mapping $\phi$ the $N_d$ and $E_d$ are sets of
+%nodes and elements of the deformed mesh and $N_r$ is set of nodes of the
+%deformed mesh.
+%
+%\tomas{what more?}
+%
+%
+%Another problem that a predictive simulation should deal with is filling
+%the missing steps that would normally had to be performed fluently by the
+%surgeon. For example when suturing two blood vessels together they have to
+%be placed close to one another first. Sometimes relatively large distances
+%have to be overcome. By ignoring this fact our approach with joining nodes
+%and edges could generate unwanted forces at the start of the simulation. To
+%alleviate large initial deformations for the elements we employed
+%interpolation ...
+%
+%
+%% }}}
+%
+%\section{Conclusion}
+%%\subsection{Verification}
+%
+%Verification is difficult -- limited or no knowledge of real state after
+%the surgery (too risky to put the child under CT/MRI again). 
+%
+%Therefore we are interested in the geometrical evaluation of the result
+%i.e. no excessive elongations, expected deformations, independence on
+%physical parameters (E/nu),\ldots or something like that \ldots
 
@@
 % that.
-% vim: et tw=75 fdm=marker spell
+% vim: et sw=2 tw=75 fdm=marker fdc=2 spell

### 2012-02-24 Doc/Miccai/miccai2012.tex
extending the points we want to cover
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@329 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{abstract}
+% TODO
 \end{abstract}
@@
 
-\section{Blood Vessel Model}
+\section{Blood Vessel Model} % {{{
 
 Thin shell model
+\begin{itemize}
+  \item based on mechanics of continuum and shell theory
+  \item used in physics for deformation modelling
+  \item briefly, refer to some textbook
+\end{itemize}
 
-Eq. showing rest position stuff
+\noindent Eq. showing dependence on rest position:
+simple $\myvec{f} = K(\myvec{x} - \myvec{x_0})$
 
-Fullfills requirements + collision
+\noindent Fulfills requirements + collision
+\begin{itemize}
+  \item we need deformations: like bending and twisting
+  \item deformation should be global
+  \item should allow arbitrary models and not rely e.g. on tubular
+  structure (like \cite{Li2009})
+  \item the model has to be able to deal with collisions (e.g. between AA
+  and PA)
+\end{itemize}
 
-Test with tube + convergence
+\noindent Test with tube + convergence
+\begin{itemize}
+  \item experimented with Olivier's model \cite{Comas2010b,Comas2010c} and
+  our own model, in principle the shell implementation is not that
+  important
+  \item geometrical tool, independence on physical parameters -- E
+\end{itemize}
+
+% }}}
 
@@
 
-surgical low-level tasks done with rest position trick (joining, patch, complex cutting/suturing)
+\noindent surgical low-level tasks done with rest position trick (joining, patch, complex cutting/suturing)
+\begin{itemize}
+  \item describe the particular tasks and show that the essence is the same,
+  \item describe our method with all the tricks
+  \item include a schematic illustration
+\end{itemize}
 

### 2012-02-24 Doc/Miccai/miccai2012.tex
Added introduction to paper.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@331 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \section{Introduction}
-State of the Art
 
-Figure illustrating a surgery
+TODO: Include only a few sentences to CHD in general.
 
-Requirements: Model blood vessels/wall deformation, surgical low-level tasks
+Surgical interventions in infants with congenitally malformed great arteries and hearts can be difficult due to their disease pattern's complex and heterogenous nature.
+Today, patient-specific 3-dimensional heart models can be acquired by modern techniques of non-invasive imaging and image post-processing [REFERENCE].
+Nevertherless, cardiac surgeons tend to rely solely on subsets or 2-dimensional views of obtained 3-dimensional image data for examination.
+Possible appropriate surgical procedures may be sketched conceptually and the most promising approach is chosen just in time during the actual open-heart operation.
+This is the first time the surgeon gets a concrete idea of possible outcomes for different surgical procedures.
+However, his dicision has to be made in relatively short time and can still be based heavily on imagination capabilities and experience.
 
-Plan
+To improve planning of these kind of cardiac surgeries, optained 3-dimensional heart models could be used as input for patient-specific surgical simulation systems.
+Present simulation systems focus on low-level surgical procedures [REFERENCES], i.e.
+performing actual incisions and suturing with special input devices in real-time.
+Their purpose is either by means of educational training or planning access strategies to reach the heart's or blood vessel's malformations.
+It is more important in such \emph{reactive} simulation systems to enable simulations in real-time than to simulate tissue and blood vessels in a physically accurate manner.
 
-\section{Blood Vessel Model} % {{{
+To be able to predict the result of complex surgical procedures, a physically correct simulation model must be used.
+[REFERENCE] present a \emph{predictive} surgical simulation system for bidirectional Glenn shunts and basically similar surgeries.
+They use a hybrid model to simulate deformations of blood vessel trees by binding a surface mesh to a reference Cosserat rod elastically.
+While the mesh models surface details of blood vessels, the Cosserat rod models global deformations, i.
+e.
+stretching, bending, twisting and shearing.
+Low-level surgical procedures are done automatically by the software with only a minimum of interaction required which allows the surgeon to focus on the actual overall surgical procedure.
+Furthermore the authors show that they can attach triangular shaped patches at loose ends of dissected vessels to allow joining of vessels with different circumferences.
+Given the input of the surgeon, a typical simulation takes only seconds to produce a predictive result for further examination and evaluation of possibly undesired collisions or deformations of blood vessels.
+Although the results are promising, the author's method is limited to uniform tubular structures and operations on them, i.
+e.
+joining blood vessels end-to-end or end-to-side.
+Connection of blood vessels is done by using springs.
+This approach introduces surplus energy to the physical system and special care has to be taken during the simulation to compensate for that.
+Surgical procedures heavily changing surface details of blood vessels, i.
+e.
+arbitrary shaped patches to correct complicated coarctations and stenosis, are not possible with this hybrid model because the blood vessel's physical behavior is simulated by the Cosserat rod and its \emph{binding} to the surface mesh.
+The authors state that more sophisticated physical models including finite element methods and thin shell elements would allow this but are computationally too expensive to produce results in reasonable time.
+
+In this paper we present an alternative, universal approach using a simulation model solely based on elastic thin shell elements.
+Our method is not limited to well-formed blood vessels and captures all low-level surgical procedures necessary for aortoplasty.
+By using a relaxation process we arbitrarily join or attach blood vessels and patches anywhere without using springs that would add surplus energy to the physical system.
+We prove that simulation time can be expected to be still reasonable or even decreased by taking special care of the blood vessel's overall tubular structure during generation of thin shell elements.
+
+TODO: Structure of Paper.
+
+\section{Blood Vessel Model}
 

### 2012-02-24 Doc/Miccai/miccai2012.tex
more stuff, spell check
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@332 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-Surgical interventions in infants with congenitally malformed great arteries and hearts can be difficult due to their disease pattern's complex and heterogenous nature.
+Surgical interventions in infants with congenitally malformed great arteries and hearts can be difficult due to their disease pattern's complex and heterogeneous nature.
 Today, patient-specific 3-dimensional heart models can be acquired by modern techniques of non-invasive imaging and image post-processing [REFERENCE].
-Nevertherless, cardiac surgeons tend to rely solely on subsets or 2-dimensional views of obtained 3-dimensional image data for examination.
+Nevertheless, cardiac surgeons tend to rely solely on subsets or 2-dimensional views of obtained 3-dimensional image data for examination.
 Possible appropriate surgical procedures may be sketched conceptually and the most promising approach is chosen just in time during the actual open-heart operation.
 This is the first time the surgeon gets a concrete idea of possible outcomes for different surgical procedures.
-However, his dicision has to be made in relatively short time and can still be based heavily on imagination capabilities and experience.
+However, his decision has to be made in relatively short time and can still be based heavily on imagination capabilities and experience.
 
-To improve planning of these kind of cardiac surgeries, optained 3-dimensional heart models could be used as input for patient-specific surgical simulation systems.
+To improve planning of these kind of cardiac surgeries, obtained 3-dimensional heart models could be used as input for patient-specific surgical simulation systems.
 Present simulation systems focus on low-level surgical procedures [REFERENCES], i.e.
@@
 They use a hybrid model to simulate deformations of blood vessel trees by binding a surface mesh to a reference Cosserat rod elastically.
-While the mesh models surface details of blood vessels, the Cosserat rod models global deformations, i.
-e.
+While the mesh models surface details of blood vessels, the Cosserat rod models global deformations, i.e.
 stretching, bending, twisting and shearing.
@@
 Given the input of the surgeon, a typical simulation takes only seconds to produce a predictive result for further examination and evaluation of possibly undesired collisions or deformations of blood vessels.
-Although the results are promising, the author's method is limited to uniform tubular structures and operations on them, i.
-e.
+Although the results are promising, the author's method is limited to uniform tubular structures and operations on them, i.e.
 joining blood vessels end-to-end or end-to-side.
@@
 This approach introduces surplus energy to the physical system and special care has to be taken during the simulation to compensate for that.
-Surgical procedures heavily changing surface details of blood vessels, i.
-e.
+Surgical procedures heavily changing surface details of blood vessels, i.e.
 arbitrary shaped patches to correct complicated coarctations and stenosis, are not possible with this hybrid model because the blood vessel's physical behavior is simulated by the Cosserat rod and its \emph{binding} to the surface mesh.
@@
 
-Thin shell model
+Several different models are commonly used for modelling thin structures or
+tubular-like structures resembling blood vessels. \tomas{Add refs. or
+reformulate!}
+We choose to use thin shell elements for the predictive
+simulation of the surgery. Shell elements are often considered numerically
+too expensive, this is however not necessarily true.
+
+The formulation of the thin shell elements is based on the mechanics of
+continuum and these elements are commonly used in physics and mechanics for
+deformation modelling of thin structures.
+
+\tomas{TODO}
+\noindent
+briefly, refer to some textbook,
+Eq. showing dependence on rest position:
+simple $\myvec{f} = K(\myvec{x} - \myvec{x_0})$
+
+\ldots
+
+From our point of view, there are several requirements that the model
+should be able to fulfill. Namely:
+
 \begin{itemize}
-  \item based on mechanics of continuum and shell theory
-  \item used in physics for deformation modelling
-  \item briefly, refer to some textbook
+  \item good support for deformations by bending and twisting
+  \item the results of the deformation should be global
+  \item method should allow arbitrary models and not rely on specific
+  structure of the object (e.g. on tubular structure like \cite{Li2009})
+  \item the model has to be able to deal with collisions (e.g. between the
+  aortic arch and pulmonary artery)
 \end{itemize}
 
-\noindent Eq. showing dependence on rest position:
-simple $\myvec{f} = K(\myvec{x} - \myvec{x_0})$
+Simulation using shell model satisfies all these requirements very well.
 
-\noindent Fulfills requirements + collision
-\begin{itemize}
-  \item we need deformations: like bending and twisting
-  \item deformation should be global
-  \item should allow arbitrary models and not rely e.g. on tubular
-  structure (like \cite{Li2009})
-  \item the model has to be able to deal with collisions (e.g. between AA
-  and PA)
-\end{itemize}
-
+\tomas{TODO}
 \noindent Test with tube + convergence

### 2012-02-24 Doc/Miccai/introduction.tex
moving chapters into separate files to avoid collisions
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@333 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2012-02-24 Doc/Miccai/miccai2012.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%% file typeinst.tex %%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% This is the LaTeX source for the instructions to authors using
-% the LaTeX document class 'llncs.cls' for contributions to
-% the Lecture Notes in Computer Sciences series.
-% http://www.springer.com/lncs       Springer Heidelberg 2006/05/04
-%
-% It may be used as a template for your own input - copy it
-% to a new file with a new name and use it as the basis
-% for your article.
-%
-% NB: the document class 'llncs' has its own and detailed documentation, see
-% ftp://ftp.springer.de/data/pubftp/pub/tex/latex/llncs/latex2e/llncsdoc.pdf
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\documentclass[runningheads,a4paper]{llncs}
-
-\usepackage[colorlinks=true,linkcolor=black,citecolor=black,pdfnewwindow=true,urlcolor=blue,filecolor=blue,baseurl=./]{hyperref}
-\usepackage{url}
-
-% \newcommand{\mycite}[2]{\url{#1}{\cite{#2}}}
-% \newcommand{\mycite}[2]{\url{#1}:\cite{#2}}
-% \newcommand{\mycite}[2]{\href{http://twitter.com/home}{Twitter}.}
-%\newcommand{\mycite}[2]{\href{./refs/#1}{\cite{#2}}}
-\newcommand{\mycite}[2]{\cite{#2}}
-%\newcommand{\myciteurl}[2]{\href{#1}{\cite{#2}}}
-
-\usepackage{amssymb}
-\setcounter{tocdepth}{3}
-\usepackage{graphicx}
-
-\usepackage{url}
-%\urldef{\mailsa}\path|{alfred.hofmann, ursula.barth, ingrid.haas, frank.holzwarth,|
-%\urldef{\mailsb}\path|anna.kramer, leonie.kunz, christine.reiss, nicole.sator,|
-%\urldef{\mailsc}\path|erika.siebert-cole, peter.strasser, lncs}@springer.com|    
-%\newcommand{\keywords}[1]{\par\addvspace\baselineskip
-%\noindent\keywordname\enspace\ignorespaces#1}
-
-\usepackage{color}
-\usepackage{amsmath}
-\usepackage[tight]{subfigure}
-\usepackage{listings}
-\definecolor{darkgray}{rgb}{0.2,0.2,0.2}
-
-\lstset{language=c++,basicstyle=\scriptsize\sffamily,tabsize=2,framexleftmargin=5mm, frame=shadowbox, framesep=3pt,rulesepcolor=\color{darkgray},rulesep=.100pt,keywordstyle=\bf\color{blue},commentstyle=\color{magenta},stringstyle=\color{red},numbers=left,numberstyle=\tiny,numbersep=10pt,breaklines=true,showstringspaces=false,columns=flexible}
-
-\newcommand{\tomas}[1]{{\color{red}{\textbf{Tom: #1}}}}
-\newcommand{\stefan}[1]{{\color{magenta}{\textbf{St: #1}}}}
-\newcommand{\christian}[1]{{\color{green}{\textbf{Ch: #1}}}}
-\newcommand{\tobias}[1]{{\color{bleu}{\textbf{Tob: #1}}}}
-
-\newcommand{\myvec}[1]{\ensuremath{\mathbf{#1}}}
-\newcommand{\mymat}[1]{\ensuremath{\mathbf{#1}}}
-\newcommand{\Vx}{\myvec{x}} % position vector
-\newcommand{\Vv}{\myvec{v}} % velocity vector
-\newcommand{\Va}{\myvec{a}} % acceleration vector
-\newcommand{\Vdv}{\myvec{dv}} % change of velocity vector (unknown in implicit CG, and used in constraint solver)
-\newcommand{\deriv}[2]{\frac{\partial #1}{\partial #2}} % Partial derivative
-
-% \newcommand{\mysubsection}[1]{\noindent \textbf{#1: }}
-\newcommand{\mysubsection}[1]{\textbf{#1: }}
-% \newcommand{\mysubsection}[1]{\subsection{#1}}
-
-%\usepackage{miseenpage}
-
-\begin{document}
-
-\mainmatter            % start of your contributions
-
-\title{Towards planning of infantile cardiac surgery based on simulation}
-
-\author{PaperID:} % TODO
-
-\maketitle
-
-\begin{abstract}
-% TODO
-\end{abstract}
-
-\section{Introduction}
-
 TODO: Include only a few sentences to CHD in general.
@@
 TODO: Structure of Paper.
-
-\section{Blood Vessel Model}
-
-Several different models are commonly used for modelling thin structures or
-tubular-like structures resembling blood vessels. \tomas{Add refs. or
-reformulate!}
-We choose to use thin shell elements for the predictive
-simulation of the surgery. Shell elements are often considered numerically
-too expensive, this is however not necessarily true.
-
-The formulation of the thin shell elements is based on the mechanics of
-continuum and these elements are commonly used in physics and mechanics for
-deformation modelling of thin structures.
-
-\tomas{TODO}
-\noindent
-briefly, refer to some textbook,
-Eq. showing dependence on rest position:
-simple $\myvec{f} = K(\myvec{x} - \myvec{x_0})$
-
-\ldots
-
-From our point of view, there are several requirements that the model
-should be able to fulfill. Namely:
-
-\begin{itemize}
-  \item good support for deformations by bending and twisting
-  \item the results of the deformation should be global
-  \item method should allow arbitrary models and not rely on specific
-  structure of the object (e.g. on tubular structure like \cite{Li2009})
-  \item the model has to be able to deal with collisions (e.g. between the
-  aortic arch and pulmonary artery)
-\end{itemize}
-
-Simulation using shell model satisfies all these requirements very well.
-
-\tomas{TODO}
-\noindent Test with tube + convergence
-\begin{itemize}
-  \item experimented with Olivier's model \cite{Comas2010b,Comas2010c} and
-  our own model, in principle the shell implementation is not that
-  important
-  \item geometrical tool, independence on physical parameters -- E
-\end{itemize}
-
-% }}}
-
-\section{Simulation of Surgical Procedures}
-
-\noindent surgical low-level tasks done with rest position trick (joining, patch, complex cutting/suturing)
-\begin{itemize}
-  \item describe the particular tasks and show that the essence is the same,
-  \item describe our method with all the tricks
-  \item include a schematic illustration
-\end{itemize}
-
-\section{Results and Discussion}
-
-Example with 3 diff. surgical procedures
-
-Parameters
-
-Limitations (getting image data, meshes, dynamic remeshing, same number of nodes)
-
-\section{Conclusion}
-
-%\begin{abstract}
-%% TODO
-%In this paper...
-%\end{abstract}
-%
-%\section{Introduction} % {{{
-%
-%%%%%%%
-%%
-%%\christian{We can put any comment...}
-%%\tomas{People have different colors}
-%%
-%%Position vector: $\Vx$
-%%
-%%Matrices are written as $\mymat{M}$
-%%
-%%%%%%%
-%
-%The congenital heart diseases have become quite common \tomas{\ldots}.
-%While they are very much treatable there are a lot of difficulties involved
-%with the process. There is rarely only a single flaw, the lesions usually
-%appear in multiples. Since most of the lesions would be fatal if untreated
-%the first surgery has to be performed shortly after the birth, often
-%followed with several other surgeries before the age of one year. At such
-%young age the heart and blood vessels are very small which requires skills
-%of the surgeon. Further more many of the critical decisions cannot be
-%concluded from the CT scan or artificial model and have to be done on site
-%(e.g. how exactly perform the cut and stitching). This is however in direct
-%contrast with the requirement on the speed at which the operation has to be
-%performed. During the surgery the blood flow has to be sustained by other
-%means (\tomas{pig's heart?}) which puts overall pressure on the organism
-%and thus cannot be kept for too long.
-%
-%\tomas{interesting ref. \cite{Barron2012}}
-%
-%\tomas{Stefan, please review this. It is just the notion that I have got
-%about the problem and might not be completely correct}
-%
-%For these reasons there is a demand for tools helping with the
-%process of pre-operative planning.
-%\tomas{Usefulness of computer aid --
-%numerical models:
-%3D visualisation: Hemminger2005}
-%
-%\tomas{Are there other approaches to this or similar problem? My search
-%did not reveal anything interesting.}
-%
-%Some of the most common actions in the procedure are
-%
-%\begin{itemize}
-%    \item cutting out a piece of a blood vessel to remove an obstruction or
-%    unwanted junction,
-%    \item opening a narrow vessel by a cut and closing the hole by an
-%    artificial material.
-%    \item \tomas{more ideas?}
-%\end{itemize}
-%
-%\tomas{add images}
-%
-%Both procedures are complicated from the surgeons point of view. In the
-%first case the surgeon has to make an educated guess how to perform a cut
-%so that after stitching the rest together the blood flow is not obstructed
-%and the blood pressure at the seam is not too large so as to tear open the
-%vessel. In the second case the surgeon has no prior idea about the shape
-%and size of the patch he will need until he performs the cut. Therefor the
-%patch has to be cut on site and cannot be prepared in advance.
-%
-%We are proposing new approach to pre-operative planning of cardiac surgery.
-%Using shell elements to model deformations of the blood vessel we help with
-%the decision making.
-%
-%% }}}
-%
-%\section{Elastic Shell Elements}
-%
-%To model the blood vessels we employed shell elements in the corotational
-%environment. The main advantage of shells is their physical accuracy. They
-%are based on the mechanics of continuum. Also the parameters defining the
-%parameters of the simulation have clear physical interpretation (unlike for
-%example for mass-spring model).
-%
-%
-%The common objection against using shell elements in real-time simulations
-%is the numerical complexity. While this is technically valid, it is not
-%good enough reason. Using low resolution mesh we can obtain results with
-%good visual and physical accuracy sufficient for real-time
-%simulation.\tomas{isn't this too strong?} Given the bending nature of the
-%shell one can employ interpolation of the shell surface to improve the
-%visual mesh presented to the user. \tomas{cite Olivier and our unpublished
-%results}
-%
-%A shell element is a combination of two other elements: elastic membrane
-%and bending plate element. While the membrane element is used to compute
-%in-plane forces of the shell (like stretching or shearing), the bending
-%plate elements on the other hand describes the bending properties.
-%
-%\tomas{add image showing the difference}
-%
-%\tomas{thin in one dimension}
-%
-%\subsection{Mechanics of Continuum Preliminaries} % {{{
-%
-%\tomas{recheck this. btw do we really want this?}
-%
-%The mathematical formulation follows the steps defined by the mechanics of
-%continuum.\tomas{ref. to some text book}
-%
-%For every particle $\myvec{x} \in \Omega$ of the deformable body $\Omega$
-%the deformation $ \phi : \Omega' \to \Omega $ is a mapping from the rest
-%shape $\myvec{x}' \in \Omega'$ to it's current shape
-%
-%$$ \phi(x) = \myvec{x}' + u(x) $$
-%
-%Where $u(x)$ is the displacement for particle $x$. From the Cauchy's strain
-%tensor
-%
-%$$
-%\epsilon = \frac{1}{2}\left( \nabla^T u + \nabla u \right)
-%$$
-%
-%\noindent%
-%we can compute the stress by applying a constitutive law. In our simulation
-%we are using linear Hooke's law to keep the numerical formulation simple
-%enough to be able to compute it in real-time. The stress is computed by
-%
-%$$
-%\sigma = \mymat{M}\epsilon
-%$$
-%
-%\noindent%
-%where $\mymat{M}$ is the material matrix. Further from the stress and
-%strain we can compute the strain energy
-%
-%$$
-%E = \int_{\Omega} \epsilon^T \sigma \,\mathrm{d}\Omega
-%$$
-%
-%\tomas{now should be discretization}
-%
-%%}}}
-%
-%\subsection{shells...} % {{{
-%
-%In this subsection we will briefly summarize how the numerical formulation
-%of the linear model is derived. For more thorough description we refer the
-%reader to \tomas{add ref.} Let's define $ u_x, u_y $ to be the two
-%in-plane displacements and $ u_z $ the off-plane function of deflection. We
-%define the strain-displacement matrices in the following way. For the
-%membrane element (using the Cauchy's strain tensor)
-%
-%\begin{equation}
-%  \mymat{B_m} = \left[ \begin{matrix}
-%    \deriv{u_x}{x} \\
-%    \deriv{u_y}{y} \\
-%    \deriv{u_x}{y} + \deriv{u_y}{x}
-%  \end{matrix} \right]
-%\end{equation}
-%
-%\noindent%
-%and for the bending plate element (using the Kirchoff-Love theory)
-%
-%\begin{equation}
-%  \mymat{B_b} = \left[ \begin{matrix}
-%    - z \deriv{^2 u_z}{x^2} \\
-%    - z \deriv{^2 u_z}{y^2} \\
-%    - 2z \deriv{^2 u_z}{xy}
-%  \end{matrix} \right]
-%\end{equation}
-%
-%From these we can easily compute the stiffness matrices $K_m$ and $K_b$
-%(for membrane and bending plate element respectively) by integrating over
-%the volume of the shell
-%
-%\begin{equation}
-%  \mymat{K}_i = \int_V \mymat{B}_i^T \mymat{M} \mymat{B}_i \,\mathrm{d}V
-%\end{equation}
-%
-%\noindent%
-%where $i \in \{\mathrm{b},\mathrm{m}\}$ and $\mymat{M}$ is the material
-%matrix defined by Hooke's law:
-%
-%\begin{equation}
-%  \mymat{M} = \frac{E}{1-\nu^2} \left[ \begin{matrix}
-%    1   & \nu & 0 \\
-%    \nu & 1   & 0 \\
-%    0   & 0   & \frac{1-\nu}{2}
-%  \end{matrix} \right]
-%\end{equation}
-%
-%The only two values defining the matrix value are Poisson's ratio $\nu$ and
-%the Young's modulus $E$.
-%
-%% }}}
-%
-%\subsection{Co-rotational Formulation}
-%
-%While the Cauchy's strain used in the formulation of the shell element is
-%invariant to rigid body translation, it is known to be rotationally
-%noninvariant. This produces ghost deformations just by rigid rotation. One
-%way to solve the problem is using Green's strain tensor instead whose
-%definition is, however, non-linear. Therefore we to overcome the problem we
-%chose to use the corotational formulation instead.\tomas{add ref;
-%maybe short description?}
-%
-%
-%\section{Method Description} % {{{
-%
-%\tomas{clearly state that this is not a complete solution where user points
-%and clicks -- it's just a proof-of-concept}
-%
-%In this chapter we will elaborate on the internal aspects of the
-%simulation.
-%
-%Outline:
-%
-%\begin{itemize}
-%    \item we use shell elements to model blood vessels, dynamic system
-%        (implicit integration), in Sofa framework \tomas{cite Olivier
-%        \cite{Comas2010b,Comas2010c} -- shells in corotational environment}
-%    \item using springs to connect the objects is not good --> generates
-%        excessive amount of energy
-%    \item to eliminate this we use one mesh for the simulation and
-%        different mesh to describe the rest shape and we leave the dynamic
-%        system to stabilize
-%        -- we need to map the nodes between topologies
-%    \item to reduce more the generation of excessive energies in the system
-%        at the beginning of the simulation we repetitively change the rest
-%        shape (linear interpolation)
-%\end{itemize}
-%
-%We are aiming for non-interactive simulation without requiring the
-%surgeon to go through the whole process of the surgery. This introduces several
-%problems we have to deal with.
-%
-%The first problem is how to deform the model so that the requested parts
-%can be sutured together. And second how to simulate the suture.
-%The simplest solution here would be connecting the respective nodes
-%together with springs. Such solution, however, has a disadvantage of adding
-%excessive energies to the system.\tomas{do we need a ref.?}
-%
-%We propose a different approach which uses a two different mesh topologies.
-%The basic idea is to transform the mesh by joining the edges and nodes
-%where the suturing is supposed to take place and let the dynamic system
-%find a stable configuration. Such stable configuration is the result we are
-%looking for. This requires us to provide a mapping to the underlying shell
-%element code that would translate between nodes of the topology of the rest
-%shape and the topology of the deformed mesh. Such a mapping is 1:N for
-%nodes where the suturing takes place. \tomas{add image} To make it properly
-%defined elements have to be taken into account also:
-%
-%\begin{equation}
-%  \phi : N_d \times E_d -> N_r
-%\end{equation}
-%
-%In the definition of this mapping $\phi$ the $N_d$ and $E_d$ are sets of
-%nodes and elements of the deformed mesh and $N_r$ is set of nodes of the
-%deformed mesh.
-%
-%\tomas{what more?}
-%
-%
-%Another problem that a predictive simulation should deal with is filling
-%the missing steps that would normally had to be performed fluently by the
-%surgeon. For example when suturing two blood vessels together they have to
-%be placed close to one another first. Sometimes relatively large distances
-%have to be overcome. By ignoring this fact our approach with joining nodes
-%and edges could generate unwanted forces at the start of the simulation. To
-%alleviate large initial deformations for the elements we employed
-%interpolation ...
-%
-%
-%% }}}
-%
-%\section{Conclusion}
-%%\subsection{Verification}
-%
-%Verification is difficult -- limited or no knowledge of real state after
-%the surgery (too risky to put the child under CT/MRI again). 
-%
-%Therefore we are interested in the geometrical evaluation of the result
-%i.e. no excessive elongations, expected deformations, independence on
-%physical parameters (E/nu),\ldots or something like that \ldots
-
-\bibliographystyle{abbrv}
-\bibliography{miccai2012}
-
-\end{document}
-
-% Some Vim setup. If it interferes with anyones work feel free to remove
-% that.
-% vim: et sw=2 tw=75 fdm=marker fdc=2 spell

### 2012-02-24 Doc/Miccai/miccai2012.tex
moving chapters into separate files to avoid collisions
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@333 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2012-02-24 Doc/Miccai/miccai2012.tex
@@
 \section{Introduction}
-State of the Art
+\input{introduction.tex}
 
-Figure illustrating a surgery
-
-Requirements: Model blood vessels/wall deformation, surgical low-level tasks
-
-Plan
-
-\section{Blood Vessel Model} % {{{
-
-Thin shell model
-\begin{itemize}
-  \item based on mechanics of continuum and shell theory
-  \item used in physics for deformation modelling
-  \item briefly, refer to some textbook
-\end{itemize}
-
-\noindent Eq. showing dependence on rest position:
-simple $\myvec{f} = K(\myvec{x} - \myvec{x_0})$
-
-\noindent Fulfills requirements + collision
-\begin{itemize}
-  \item we need deformations: like bending and twisting
-  \item deformation should be global
-  \item should allow arbitrary models and not rely e.g. on tubular
-  structure (like \cite{Li2009})
-  \item the model has to be able to deal with collisions (e.g. between AA
-  and PA)
-\end{itemize}
-
-\noindent Test with tube + convergence
-\begin{itemize}
-  \item experimented with Olivier's model \cite{Comas2010b,Comas2010c} and
-  our own model, in principle the shell implementation is not that
-  important
-  \item geometrical tool, independence on physical parameters -- E
-\end{itemize}
-
-% }}}
+\section{Blood Vessel Model}
+\input{shell-model.tex}
 
 \section{Simulation of Surgical Procedures}
-
-\noindent surgical low-level tasks done with rest position trick (joining, patch, complex cutting/suturing)
-\begin{itemize}
-  \item describe the particular tasks and show that the essence is the same,
-  \item describe our method with all the tricks
-  \item include a schematic illustration
-\end{itemize}
+\input{simulation.tex}
 
 \section{Results and Discussion}
-
-Example with 3 diff. surgical procedures
-
-Parameters
-
-Limitations (getting image data, meshes, dynamic remeshing, same number of nodes)
+\input{results.tex}
 
 \section{Conclusion}
+\input{conclusion.tex}
 

### 2012-02-24 Doc/Miccai/results.tex
moving chapters into separate files to avoid collisions
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@333 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

Example with 3 diff. surgical procedures

Parameters

Limitations (getting image data, meshes, dynamic remeshing, same number of nodes)

### 2012-02-24 Doc/Miccai/shell-model.tex
moving chapters into separate files to avoid collisions
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@333 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

Several different models are commonly used for modelling thin structures or
tubular-like structures resembling blood vessels. \tomas{Add refs. or
reformulate!}
We choose to use thin shell elements for the predictive
simulation of the surgery. Shell elements are often considered numerically
too expensive, this is however not necessarily true.

The formulation of the thin shell elements is based on the mechanics of
continuum and these elements are commonly used in physics and mechanics for
deformation modelling of thin structures.

\tomas{TODO}
\noindent
briefly, refer to some textbook,
Eq. showing dependence on rest position:
simple $\myvec{f} = K(\myvec{x} - \myvec{x_0})$

\ldots

From our point of view, there are several requirements that the model
should be able to fulfill. Namely:

\begin{itemize}
  \item good support for deformations by bending and twisting
  \item the results of the deformation should be global
  \item method should allow arbitrary models and not rely on specific
  structure of the object (e.g. on tubular structure like \cite{Li2009})
  \item the model has to be able to deal with collisions (e.g. between the
  aortic arch and pulmonary artery)
\end{itemize}

Simulation using shell model satisfies all these requirements very well.

\tomas{TODO}
\noindent Test with tube + convergence
\begin{itemize}
  \item experimented with Olivier's model \cite{Comas2010b,Comas2010c} and
  our own model, in principle the shell implementation is not that
  important
  \item geometrical tool, independence on physical parameters -- E
\end{itemize}

### 2012-02-24 Doc/Miccai/simulation.tex
moving chapters into separate files to avoid collisions
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@333 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\noindent surgical low-level tasks done with rest position trick (joining, patch, complex cutting/suturing)
\begin{itemize}
  \item describe the particular tasks and show that the essence is the same,
  \item describe our method with all the tricks
  \item include a schematic illustration
\end{itemize}

### 2012-02-27 Doc/Miccai/shell-model.tex
miccai 2012: shell model and method sections
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@339 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-Several different models are commonly used for modelling thin structures or
-tubular-like structures resembling blood vessels. \tomas{Add refs. or
-reformulate!}
-We choose to use thin shell elements for the predictive
-simulation of the surgery. Shell elements are often considered numerically
-too expensive, this is however not necessarily true.
-
-The formulation of the thin shell elements is based on the mechanics of
-continuum and these elements are commonly used in physics and mechanics for
-deformation modelling of thin structures.
-
-\tomas{TODO}
-\noindent
-briefly, refer to some textbook,
-Eq. showing dependence on rest position:
-simple $\myvec{f} = K(\myvec{x} - \myvec{x_0})$
-
-\ldots
-
 From our point of view, there are several requirements that the model
@@
 
-Simulation using shell model satisfies all these requirements very well.
+Several different models are commonly used for modelling thin structures or
+tubular-like structures resembling blood vessels. \tomas{Add refs. or
+reformulate!} For the predictive simulation of the surgery we choose to use
+thin shell elements, because in our opinion it satisfies all the
+requirements very well. They are very often considered numerically too
+expensive, this is however not necessarily true.
 
-\tomas{TODO}
-\noindent Test with tube + convergence
-\begin{itemize}
-  \item experimented with Olivier's model \cite{Comas2010b,Comas2010c} and
-  our own model, in principle the shell implementation is not that
-  important
-  \item geometrical tool, independence on physical parameters -- E
-\end{itemize}
+\tomas{image: single shell}
+
+The formulation of the thin shell elements is based on the mechanics of
+continuum. These elements are commonly used in physics and mechanics for
+deformation modelling of thin structures because they are numerically more
+stable then volumetric FEM methods in this situation. The shell element is
+a 2-dimensional element whose thickness is a parameter of the simulation.
+Often the thickness is considered constant over the surface of the element.
+Shell combines forces from in-plane deformations, like stretching or
+shearing, with bending forces. Such internal forces $\myvec{f}$ are a
+function of current displacement measured as difference of current shape
+$\myvec{x}$ from the rest shape $\myvec{x_0}$.
+
+\begin{equation}
+  \myvec{f} = K(\myvec{x} - \myvec{x_0})
+\end{equation}
+
+For more thorough description we refer the reader to the textbook about
+Finite Elements for example \tomas{add ref}.
+
+In the rest of the section we want to show that the performance of a model
+based on the shell element is adequate for the simulation.
+
+\begin{figure}[tbh]
+  \begin{minipage}[b]{0.3\linewidth}
+      \framebox[1cm]{\rule{0pt}{1cm}}
+      %\includegraphics[width=\columnwidth]{img/}
+      \caption{Bending shell element}
+      \label{fig-shell}
+  \end{minipage}
+  \hspace{0.1\columnwidth}
+  \begin{minipage}[b]{0.6\linewidth}
+    \centering
+    \begin{tabular}{cc}
+     \includegraphics[width=0.5\columnwidth]{img/compare-bend.png}
+      &
+      \includegraphics[width=0.5\columnwidth]{img/compare-twist.png}
+      \\
+      \includegraphics[width=0.5\columnwidth]{img/compare-bend-other.png}
+      &
+      \includegraphics[width=0.5\columnwidth]{img/compare-twist-other.png}
+    \end{tabular}
+    \caption{Bending and twisting deformations. Upper: shell model; middle:
+    artificial tube, bottom: Cosserat rod \cite{Li2009}. (Credits for the
+    bottom images: Li et al.)}
+    \label{fig-deformations}
+  \end{minipage}
+\end{figure}
+
+Figure \ref{fig-deformations} shows different deformations that are likely to
+occur during the surgery (bending and twisting). The figure compares our
+results with the results of Li et al. \cite{Li2009}. As you can see the
+results are closer to the real deformations.
+
+The bending property of the shell allows us to use smaller number of
+elements for the simulation and use high polygonal model only for
+visualisation. The convergence of the model is a key factor here. In our experiments we
+were able to use only as few as eight nodes around the circumference of a
+tubular model while still maintaining realistic behaviour. Figure
+\tomas{add-ref} shows the results for different number of nodes along the
+circumference.
+
+\tomas{image: convergence}
+
+We have used the model of Olivier Comas \cite{Comas2010b,Comas2010c} and
+also our own model. But our method described in section \ref{sec-method}
+does not depend on any particular model. Since we use the method as a
+geometrical tool we have also certain independence on physical parameters.
+For example the variation of Young's modulus in the interval from $10^3$ to
+$10^7$ resulted to only $0.1\%$ change of the deformation.
+
+% vim: et sw=2 tw=75 fdm=marker fdc=2 spell

### 2012-02-27 Doc/Miccai/simulation.tex
add figure
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@338 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
miccai 2012: shell model and method sections
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@339 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-\noindent surgical low-level tasks done with rest position trick (joining, patch, complex cutting/suturing)
-\begin{itemize}
-  \item describe the particular tasks and show that the essence is the same,
-  \item describe our method with all the tricks
-  \item include a schematic illustration
-\end{itemize}
+There are several types of low-level task the surgeon usually employs while
+performing the surgery.
+
+One common procedure is reconnection and joining of blood vessels. Another
+is fixing the stenosis of aorta where the three most common types of fixing
+the problem are:
+
+\begin{description}
+
+  \item[End to end aortoplasty:] the surgeon completely cuts away the
+    affected part of aorta and sutures the two ends of aorta back together.
+
+  \item[Patch aortoplasty:] the surgeon performs a cut along the axis of
+    the aorta thus allowing the narrow place to expand. Then applies a
+    patch from artificial material to fill the hole in.
+
+  \item[Subclavian flap aortoplasty:] The surgeon cuts the nearest artery,
+    removing also half of the artery (along the axis) and part of the
+    aorta in the affected place. He then uses the rest of the artery to
+    patch the aorta.
+
+\end{description}
+
+From our point of view the all these procedures share the same similarity:
+parts of the circulatory system are sutured together or to a patch. These
+parts are not too far away from each other. On this we have based our
+general approach in the simulation.
+
+The simple way how to perform the suturing would be to connect the
+respective edges with springs. This approach however adds some unwanted
+energy to the system. Our method is based solely on the relaxation of the
+mesh of shell elements.
+
+We start with a mesh representing the situation after all the cuts have
+been performed but before suturing takes place. In the first step we select
+on the mesh edges that we want to suture together and join the
+corresponding nodes together. The result is topologicaly different mesh
+that has same number of elements. The Figure. \ref{fig-JoiningVessels} explains
+this operation. Then we start the simulation on the modified mesh while
+using the original mesh to define rest shapes of the shell elements. After
+the mesh relaxes and the simulation reaches energetical equilibrium the
+simulation terminates.
+
+\begin{figure}[tbh]
+\begin{center}
+\includegraphics[width=\columnwidth]{img/rest_shape_scheme.png}
+\end{center}
+
+\caption{Scheme showing the method for joining two vessels.}
+\label{fig-JoiningVessels}
+\end{figure}
+
+If the nodes that are to be sutured together are not in a close vicinity we
+may experience unwanted quick increase of internal forces at the beginning
+of the simulation. To allevieate this problem we have implemented the
+techinuqe of linear interpolation of the rest shape mesh. At the start of
+the simulation we use different rest shape mesh. It is topologicaly
+equivalent to the original mesh but nodes have positions equivalent to the
+simulated mesh (that way the initial deformations are not so big). During
+the simulation we then perform several steps of linear interpolation
+between these two rest shape meshes. At the end of interpolation the
+original rest shape is used. After the mesh relaxes the simulation again
+terminates.
+
+% vim: et sw=2 tw=75 fdm=marker fdc=2 spell

### 2012-02-27 Doc/Miccai/simulation.tex
miccai2012: few corrections
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@340 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-  \item[Subclavian flap aortoplasty:] The surgeon cuts the nearest artery,
+  \item[Subclavian flap aortoplasty:] the nearest artery is cut,
     removing also half of the artery (along the axis) and part of the
@@
 parts are not too far away from each other. On this we have based our
-general approach in the simulation.
-
-The simple way how to perform the suturing would be to connect the
-respective edges with springs. This approach however adds some unwanted
-energy to the system. Our method is based solely on the relaxation of the
-mesh of shell elements.
+general approach in the simulation. The simple way how to perform the
+suturing would be to connect the respective edges with springs. This
+approach however adds some unwanted energy to the system. Our method is
+based solely on the relaxation of the mesh of shell elements.
 

### 2012-02-27 Doc/Miccai/shell-model.tex
miccai2012: few corrections
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@340 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
miccai2012: images - convergence
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@342 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 For more thorough description we refer the reader to the textbook about
-Finite Elements for example \tomas{add ref}.
+Finite Element methods for example \cite{Reddy1993}.
 
 In the rest of the section we want to show that the performance of a model
-based on the shell element is adequate for the simulation.
+based on shell elements is adequate for the simulation.
 
@@
 elements for the simulation and use high polygonal model only for
-visualisation. The convergence of the model is a key factor here. In our experiments we
-were able to use only as few as eight nodes around the circumference of a
-tubular model while still maintaining realistic behaviour. Figure
-\tomas{add-ref} shows the results for different number of nodes along the
-circumference.
+visualisation. The convergence of the model is a key factor here. In our
+experiments we were able to use only as few as eight nodes around the
+circumference of a tubular model while still maintaining realistic
+behaviour. Figure \ref{fig-convergence} shows the results for different
+number of nodes along the circumference.
 
-\tomas{image: convergence}
+\begin{figure}[tbh]
+  \centering
+  \begin{tabular}{cccc}
+    \includegraphics[width=0.24\columnwidth]{img/twist-06-cg.png}
+    &
+    \includegraphics[width=0.24\columnwidth]{img/twist-08-cg.png}
+    &
+    \includegraphics[width=0.24\columnwidth]{img/twist-16-cg.png}
+    &
+    \includegraphics[width=0.24\columnwidth]{img/twist-31-cg.png}
+  \end{tabular}
+  \caption{Convergence on a twisted cylinder with 6, 8, 16 and 31 nodes
+  around the circumference (120, 208, 832 and 3038 elements)}
+  \label{fig-convergence}
+\end{figure}
 

### 2012-02-28 Doc/Miccai/shell-model.tex
small comments on section 2
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@343 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 Often the thickness is considered constant over the surface of the element.
-Shell combines forces from in-plane deformations, like stretching or
-shearing, with bending forces. Such internal forces $\myvec{f}$ are a
-function of current displacement measured as difference of current shape
+Shell combines forces from in-plane deformations, like stretching or shearing, with bending forces. 
+Such internal forces $\myvec{f}$ are a function of current displacement measured as difference of current shape
 $\myvec{x}$ from the rest shape $\myvec{x_0}$.
@@
   \myvec{f} = K(\myvec{x} - \myvec{x_0})
+ \label{linear_static_model} 
 \end{equation}
 
+\christian{I think that we should precise that we rely on corotational shell formulation: the model is linear in the local frame of the element, otherwise equation (\ref{linear_static_model}) is false }
 For more thorough description we refer the reader to the textbook about
@@
   \caption{Convergence on a twisted cylinder with 6, 8, 16 and 31 nodes
-  around the circumference (120, 208, 832 and 3038 elements)}
+  around the circumference (120, 208, 832 and 3038 elements).
+  \christian{Could we add wire-frame images to show the density of the meshes} } 
   \label{fig-convergence}

### 2012-02-28 Doc/Miccai/introduction.tex
comments about the intro
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@344 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 Possible appropriate surgical procedures may be sketched conceptually and the most promising approach is chosen just in time during the actual open-heart operation.
+\christian{to modify [ }
 This is the first time the surgeon gets a concrete idea of possible outcomes for different surgical procedures. 
 However, his decision has to be made in relatively short time and can still be based heavily on imagination capabilities and experience.
+\christian{  ] }
 
 To improve planning of these kind of cardiac surgeries, obtained 3-dimensional heart models could be used as input for patient-specific surgical simulation systems.
-Present simulation systems focus on low-level surgical procedures [REFERENCES], i.e.
+
+\christian{not sure this is the right place for the following paragraph (It begins the state of the art while the work is not yet introduced) : [}
+Present simulation systems \christian{what simulation systems are we talking about ?}  focus on low-level surgical procedures [REFERENCES], i.e.
 performing actual incisions and suturing with special input devices in real-time.
@@
 It is more important in such \emph{reactive} simulation systems to enable simulations in real-time than to simulate tissue and blood vessels in a physically accurate manner.
+\christian{ ] }
 
+\christian{I think we should present our work as a continuity of what was proposed with the Cosserat model. We should reduce the discussion about this paper and show what is improved by our approach.}
 To be able to predict the result of complex surgical procedures, a physically correct simulation model must be used. 
@@
 They use a hybrid model to simulate deformations of blood vessel trees by binding a surface mesh to a reference Cosserat rod elastically.
-While the mesh models surface details of blood vessels, the Cosserat rod models global deformations, i.e.
-stretching, bending, twisting and shearing.
+While the mesh models surface details of blood vessels, the Cosserat rod models global deformations, i.e. stretching, bending, twisting and shearing. 
 Low-level surgical procedures are done automatically by the software with only a minimum of interaction required which allows the surgeon to focus on the actual overall surgical procedure. 
@@
 Given the input of the surgeon, a typical simulation takes only seconds to produce a predictive result for further examination and evaluation of possibly undesired collisions or deformations of blood vessels.
-Although the results are promising, the author's method is limited to uniform tubular structures and operations on them, i.e.
-joining blood vessels end-to-end or end-to-side.
+Although the results are promising, the author's method is limited to uniform tubular structures and operations on them, i.e. joining blood vessels end-to-end or end-to-side. 
 Connection of blood vessels is done by using springs. 
@@
 
-In this paper we present an alternative, universal approach using a simulation model solely based on elastic thin shell elements.
+\christian{\sout{In this paper we present an alternative}   Precisely, I propose to place our paper as an improvement of their approach } , universal approach using a simulation model solely based on elastic thin shell elements.
 Our method is not limited to well-formed blood vessels and captures all low-level surgical procedures necessary for aortoplasty. 

### 2012-02-28 Doc/Miccai/simulation.tex
comments about the simulation section
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@345 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 performing the surgery.
-
-One common procedure is reconnection and joining of blood vessels. Another
-is fixing the stenosis of aorta where the three most common types of fixing
+%
+\christian{One common procedure is reconnection and joining of blood vessels. [should be put on the following list...]}.
+Another is fixing the stenosis of aorta where the three most common types of fixing
 the problem are:
@@
 
-From our point of view the all these procedures share the same similarity:
-parts of the circulatory system are sutured together or to a patch. These
-parts are not too far away from each other. On this we have based our
-general approach in the simulation. The simple way how to perform the
-suturing would be to connect the respective edges with springs. This
-approach however adds some unwanted energy to the system. Our method is
-based solely on the relaxation of the mesh of shell elements.
+\christian{Insist that our method is able to simulate all these surgical tasks}
+From our point of view all these procedures share the same similarity:
+Parts of the circulatory system are sutured together or to a patch. 
+\christian{These parts are not too far away from each other. (?)} 
+On this we have based our general approach in the simulation. 
+The simple way how to perform the suturing would be to connect the respective edges with springs. 
+This approach however adds some unwanted energy to the system. 
+Our method is based solely on the relaxation of the mesh of shell elements.
+%% christian: I like the "relaxation" term !!!
 
-We start with a mesh representing the situation after all the cuts have
-been performed but before suturing takes place. In the first step we select
-on the mesh edges that we want to suture together and join the
-corresponding nodes together. The result is topologicaly different mesh
-that has same number of elements. The Figure. \ref{fig-JoiningVessels} explains
-this operation. Then we start the simulation on the modified mesh while
-using the original mesh to define rest shapes of the shell elements. After
-the mesh relaxes and the simulation reaches energetical equilibrium the
-simulation terminates.
+We start with a mesh representing the situation after all the cuts have been performed but before suturing takes place. 
+\christian{I think that we should precise that we do not simulate suture : We are working on a planning system not a training simulator, and a trained surgeon knows how to perform the sutures.}
+In the first step we select on the mesh edges that we want to suture together and join the corresponding nodes together. 
+\christian{The result is topologicaly different mesh that has same number of elements. (not clear)}
+The Figure \ref{fig-JoiningVessels} explains this operation when joining vessels. 
+Then we start the simulation on the modified mesh while using the original mesh to define rest shapes of some shell elements. 
+\christian{ \sout{After the mesh relaxes and the simulation reaches energetical equilibrium the simulation terminates.}
+We progressively recontract the rest shape of the relaxed shells to their initial shapes. 
+The energetical equilibrium found by the simulation gradually forces the side edges to connect.}
 
@@
 
-\caption{Scheme showing the method for joining two vessels.}
+\caption{Scheme showing the method for joining two vessels.
+\christian{Do we need to extend the scheme for every type of aortoplasty (end to end / Patch / Subclavian): it would illustrate that the relaxation method works for every case...} }
 \label{fig-JoiningVessels}
@@
 
-If the nodes that are to be sutured together are not in a close vicinity we
-may experience unwanted quick increase of internal forces at the beginning
-of the simulation. To allevieate this problem we have implemented the
-techinuqe of linear interpolation of the rest shape mesh. At the start of
-the simulation we use different rest shape mesh. It is topologicaly
-equivalent to the original mesh but nodes have positions equivalent to the
-simulated mesh (that way the initial deformations are not so big). During
-the simulation we then perform several steps of linear interpolation
-between these two rest shape meshes. At the end of interpolation the
-original rest shape is used. After the mesh relaxes the simulation again
-terminates.
+If the nodes that are to be sutured together are not in a close vicinity we may experience unwanted quick increase of internal forces at the beginning of the simulation.
+To allevieate this problem we have implemented the technique that allows for moderate changes in the rest shape between simulation steps.
+\christian{Between the start and the end of the simulation we use a linear interpolation of the rest shape for the shells that are relaxed}
+\christian{\sout{At the start of the simulation we use different rest shape mesh.}}
+\christian{I think that it is unclear that we do the relaxation on some shells that are on the side edges of the connecting line}
+It is topologicaly equivalent to the original mesh but nodes have positions equivalent to the simulated mesh (that way the initial deformations are not so big) \christian{unclear...should be illustrated ?}.
+\christian{\sout{During the simulation we then perform several steps of linear interpolation between these two rest shape meshes. }}
+At the end of interpolation the original rest shape is used. 
+After the mesh relaxes the simulation again terminates.
 

### 2012-02-28 Doc/Miccai/shell-model.tex
adding wireframe images
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@346 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
modifs in section 2
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@347 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-From our point of view, there are several requirements that the model
-should be able to fulfill. Namely:
+Infantile cardiac surgery offers a reconfiguration of the blood vessel walls that limits or rectifies the congenital anomalies of the anatomy.
+During the surgery, the vessel walls are deformed, cut, and reconnected to other vessels.
+To do a planning of the surgery, the geometry changes of these vessel walls (the motions and deformations, their configuration when they are cut or stitched)  must be anticipated.
 
-\begin{itemize}
-  \item good support for deformations by bending and twisting
-  \item the results of the deformation should be global
-  \item method should allow arbitrary models and not rely on specific
-  structure of the object (e.g. on tubular structure like \cite{Li2009})
-  \item the model has to be able to deal with collisions (e.g. between the
-  aortic arch and pulmonary artery)
-\end{itemize}
+These requirements ask for a physics-based model for the vessel walls.
+This model should provide (I) a good support for the deformations by bending and twisting.
+(II) an accurate modeling of the wall motion when the vessel is cut along its axis (III) global deformations of stitched vessels that capture their final configuration, after surgery (IV) collision response when detected (e.g.
+between the aortic arch and pulmonary artery).
 
@@
     \includegraphics[width=0.24\columnwidth]{img/twist-31-cg.png}
+    \\
+    \includegraphics[width=0.24\columnwidth]{img/twist-06w-cg.png}
+    &
+    \includegraphics[width=0.24\columnwidth]{img/twist-08w-cg.png}
+    &
+    \includegraphics[width=0.24\columnwidth]{img/twist-16w-cg.png}
+    &
+    \includegraphics[width=0.24\columnwidth]{img/twist-31w-cg.png}
   \end{tabular}
@@
   around the circumference (120, 208, 832 and 3038 elements).
-  \christian{Could we add wire-frame images to show the density of the meshes} } 
+  \tomas{improve the wireframe images if there's time} } 
   \label{fig-convergence}

### 2012-02-29 Doc/Miccai/shell-model.tex
modifs in section 2
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@348 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
+\subsubsection{Requirements: } 
 Infantile cardiac surgery offers a reconfiguration of the blood vessel walls that limits or rectifies the congenital anomalies of the anatomy.
@@
 These requirements ask for a physics-based model for the vessel walls. 
-This model should provide (I) a good support for the deformations by bending and twisting.
-(II) an accurate modeling of the wall motion when the vessel is cut along its axis (III) global deformations of stitched vessels that capture their final configuration, after surgery (IV) collision response when detected (e.g.
-between the aortic arch and pulmonary artery).
+This model should provide 
+(I) a good support for the deformations by bending and twisting. 
+(II) an accurate modeling of the wall motion when the vessel is cut along its axis 
+(III) global deformations of stitched vessels that capture their final configuration, after surgery 
+(IV) collision response when detected (e.g. between the aortic arch and pulmonary artery).
 
-Several different models are commonly used for modelling thin structures or
-tubular-like structures resembling blood vessels. \tomas{Add refs. or
-reformulate!} For the predictive simulation of the surgery we choose to use
-thin shell elements, because in our opinion it satisfies all the
-requirements very well. They are very often considered numerically too
-expensive, this is however not necessarily true.
+In addition, the computation of this model must be fast enough to allows for interactive planning of the surgery. 
+The options on the choice of model are limited: 
+The model based on the deformation of the centerline, that can be very quickly computed, (like ??\tomas{Add refs.!} \christian{Do you have some refs that uses model based on centerline?
+}  ) must be discarded because they do not support cut of the wall along the vessel axis.
+Mass-springs approaches that are often used for cloth simulation (\christian{do we consider this option ?
+if yes add ref ??}), allow for fast computation but do not provide sufficient accuracy as they do not necessary converge to a unique solution when changing the mesh.
+In this work, we then propose to rely on Finite Element formulation of thin-shell theory. 
+These model are numerically more expensive, but provide accurate models of thin deformable structure.% that converge to unique solutions.
+Moreover, we show that we can still obtain fast computation with a reasonable number of elements.
 
+\subsubsection{Thin shell elements: } 
+The formulation of the thin shell elements is based on the mechanics of continuum. 
 \tomas{image: single shell}
-
-The formulation of the thin shell elements is based on the mechanics of
-continuum. These elements are commonly used in physics and mechanics for
-deformation modelling of thin structures because they are numerically more
-stable then volumetric FEM methods in this situation. The shell element is
-a 2-dimensional element whose thickness is a parameter of the simulation.
+For thorough description we refer the reader to the textbook about Finite Element methods for example \cite{Reddy1993}.
+These elements are commonly used in physics and mechanics for deformation modelling of thin structures because they are numerically more stable then volumetric FEM methods in this situation. 
+The shell element is a 2-dimensional element (in our case, triangle element) whose thickness is a parameter of the simulation.
 Often the thickness is considered constant over the surface of the element.
@@
 $\myvec{x}$ from the rest shape $\myvec{x_0}$.
+To accelerate the computation, we use a corotational formulation, that is based on the definition of a local frame for each element (that can be built using the geometry of the triangle, \christian{see figure "image: single shell" ??
+}).
 
+In this local frame, the computation of the internal forces is linear:
 \begin{equation}
-  \myvec{f} = K(\myvec{x} - \myvec{x_0})
+  \myvec{f} = \mathbf{K_e}(\myvec{x} - \myvec{x_0})
  \label{linear_static_model} 
 \end{equation}
+where $\mathbf{K_e}$ is a stiffness matrix that is computed thanks to a weak formulation (integral formulation) of Kirshohoff-Loves equations.
+This approach is very similar to the model presented in  \cite{Comas2010b,Comas2010c}  even if the interpolation functions we use are  different to allow for better convergence\footnote{Our shell model will be published separately}.
+We emphasis that the method described in the paper is independent of the shell model used (other non-corotational shell models can be used).
+%
+When the structure is assembled, the contributions of each element are computed in the global frame using a different rotation matrix for each element.
+\begin{equation}
+\mathbb{K} = \mathcal{A}_{e=0}^N (\mathbf{R_e}^T \mathbf{K_e} \mathbf{R_e}) \hspace{10mm}
+\mathbb{F} = \mathcal{A}_{e=0}^N ( \mathbf{R_e}^T \mathbf{K_e}(\myvec{x} - \myvec{x_0}))
+\end{equation}
+Where $\mathcal{A}$ is the assembling operator over the $N$ elements of the mesh, $\mathbb{K}$ is the global stiffness and $\mathbb{F}$ the internal forces of the entire shell mesh structure. 
+Using this approach, $\mathbb{K}$ is changing at each step of the simulation but the model is able to catch geometric non-linear deformations.
+In practice, the computation of the node motion $\mathbb{U} = \mathbb{K}^{-1} \mathbb{F}$ is performed using an optimized conjugate gradient algorithm.
 
-\christian{I think that we should precise that we rely on corotational shell formulation: the model is linear in the local frame of the element, otherwise equation (\ref{linear_static_model}) is false }
-For more thorough description we refer the reader to the textbook about
-Finite Element methods for example \cite{Reddy1993}.
-
-In the rest of the section we want to show that the performance of a model
+\subsubsection{Accuracy: } 
+In the following of the section we want to show that the performance of a model
 based on shell elements is adequate for the simulation.
@@
 results with the results of Li et al. \cite{Li2009}. As you can see the
-results are closer to the real deformations.
+results are close to the real deformations.
 
@@
 
-We have used the model of Olivier Comas \cite{Comas2010b,Comas2010c} and
-also our own model. But our method described in section \ref{sec-method}
-does not depend on any particular model. Since we use the method as a
+\christian{ I think this should be put in the result section:
+Since we use the method as a
 geometrical tool we have also certain independence on physical parameters.
@@
 $10^7$ resulted to only $0.1\%$ change of the deformation.
+}
 

### 2012-02-29 Doc/Miccai/shell-model.tex
changes to simulation and shell model sections
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@349 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 The model based on the deformation of the centerline, that can be very quickly computed, (like ??\tomas{Add refs.!} \christian{Do you have some refs that uses model based on centerline?
-}  ) must be discarded because they do not support cut of the wall along the vessel axis.
-Mass-springs approaches that are often used for cloth simulation (\christian{do we consider this option ?
-if yes add ref ??}), allow for fast computation but do not provide sufficient accuracy as they do not necessary converge to a unique solution when changing the mesh.
+}\tomas{No, but I will find something}  ) must be discarded because they do not support cut of the wall along the vessel axis.
+Mass-springs approaches that are often used for cloth simulation \cite{Volino2005b}, allow for fast computation but do not provide sufficient accuracy as they do not necessary converge to a unique solution when changing the mesh.
 In this work, we then propose to rely on Finite Element formulation of thin-shell theory. 
@@
 The formulation of the thin shell elements is based on the mechanics of continuum. 
-\tomas{image: single shell}
 For thorough description we refer the reader to the textbook about Finite Element methods for example \cite{Reddy1993}.
 These elements are commonly used in physics and mechanics for deformation modelling of thin structures because they are numerically more stable then volumetric FEM methods in this situation. 
-The shell element is a 2-dimensional element (in our case, triangle element) whose thickness is a parameter of the simulation.
-Often the thickness is considered constant over the surface of the element.
-Shell combines forces from in-plane deformations, like stretching or shearing, with bending forces. 
+The shell element is a 2-dimensional element (in our case, triangular element) whose thickness is a parameter of the simulation (often considered constant over the surface of the element).
+Shell element combines forces from in-plane deformations, like stretching or shearing, with bending forces. 
 Such internal forces $\myvec{f}$ are a function of current displacement measured as difference of current shape
@@
 \end{equation}
-where $\mathbf{K_e}$ is a stiffness matrix that is computed thanks to a weak formulation (integral formulation) of Kirshohoff-Loves equations.
+where $\mymat{K_e}$ is a stiffness matrix that is computed thanks to a weak formulation (integral formulation) of Kirshohoff-Loves equations.
 This approach is very similar to the model presented in  \cite{Comas2010b,Comas2010c}  even if the interpolation functions we use are  different to allow for better convergence\footnote{Our shell model will be published separately}.
-We emphasis that the method described in the paper is independent of the shell model used (other non-corotational shell models can be used).
+We emphasize that the method described in the paper is independent of the shell model used (other non-corotational shell models can be used as well).
 %
@@
 
-Figure \ref{fig-deformations} shows different deformations that are likely to
-occur during the surgery (bending and twisting). The figure compares our
+Figure \ref{fig-deformations} shows bending and twisting deformations.
+These are likely to occur during the surgery.
+The figure compares our
 results with the results of Li et al. \cite{Li2009}. As you can see the
@@
   around the circumference (120, 208, 832 and 3038 elements).
+  Using mesh with 3038 triangles for visualisation.
   \tomas{improve the wireframe images if there's time} } 
@@
 For example the variation of Young's modulus in the interval from $10^3$ to
-$10^7$ resulted to only $0.1\%$ change of the deformation.
+$10^7$ resulted in only $0.1\%$ change of the deformation.
 }
 
-% vim: et sw=2 tw=75 fdm=marker fdc=2 spell
+% vim: et sw=2 tw=0 spell

### 2012-02-29 Doc/Miccai/simulation.tex
changes to simulation and shell model sections
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@349 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2012-02-27 Doc/Miccai/simulation.tex
@@
 performing the surgery.
-
-One common procedure is reconnection and joining of blood vessels. Another
-is fixing the stenosis of aorta where the three most common types of fixing
-the problem are:
+The most common:
 
 \begin{description}
+  \item Reconnection and joining of blood vessels to the arteries.
 
@@
 
-From our point of view the all these procedures share the same similarity:
-parts of the circulatory system are sutured together or to a patch. These
-parts are not too far away from each other. On this we have based our
-general approach in the simulation. The simple way how to perform the
-suturing would be to connect the respective edges with springs. This
-approach however adds some unwanted energy to the system. Our method is
-based solely on the relaxation of the mesh of shell elements.
+With our method we are able to simulate all these tasks.
+We have based our general approach on the common task involved in all these operations which is suturing of two or more edges together.
+Since we are interested in predicting the result rather than creating a training simulator for suturing (trained surgeon knows well how to perform the sutures) the simplest way of performing the suture would be to connect the respective edges with springs.
+Springs however adds some unwanted energy to the system. 
+Our method is based solely on the relaxation of the mesh of shell elements.
+%% christian: I like the "relaxation" term !!!
 
-We start with a mesh representing the situation after all the cuts have
-been performed but before suturing takes place. In the first step we select
-on the mesh edges that we want to suture together and join the
-corresponding nodes together. The result is topologicaly different mesh
-that has same number of elements. The Figure. \ref{fig-JoiningVessels} explains
-this operation. Then we start the simulation on the modified mesh while
-using the original mesh to define rest shapes of the shell elements. After
-the mesh relaxes and the simulation reaches energetical equilibrium the
-simulation terminates.
+We start with a mesh representing the situation after all the cuts have been performed but before suturing takes place. 
+In the first step we perform the required topological changes to the mesh.
+We keep the number of elements in new mesh the same as in the original mesh so that we can define 1:1 mapping between elements of these meshes.
+The Figure \ref{fig-JoiningVessels} explains this operation when joining vessels. 
+Then we start the simulation on the modified mesh while using the original mesh to define rest shapes of all shell elements.
 
@@
 
-\caption{Scheme showing the method for joining two vessels.}
+\caption{Scheme showing the method for joining two vessels.
+\christian{Do we need to extend the scheme for every type of aortoplasty (end to end / Patch / Subclavian): it would illustrate that the relaxation method works for every case...}
+  \tomas{No we don't. This there will be real examples in results (if I remember correctly).} }
 \label{fig-JoiningVessels}
@@
 
-If the nodes that are to be sutured together are not in a close vicinity we
-may experience unwanted quick increase of internal forces at the beginning
-of the simulation. To allevieate this problem we have implemented the
-techinuqe of linear interpolation of the rest shape mesh. At the start of
-the simulation we use different rest shape mesh. It is topologicaly
-equivalent to the original mesh but nodes have positions equivalent to the
-simulated mesh (that way the initial deformations are not so big). During
-the simulation we then perform several steps of linear interpolation
-between these two rest shape meshes. At the end of interpolation the
-original rest shape is used. After the mesh relaxes the simulation again
-terminates.
+In case of non-trivial distance between the nodes that are sutured together the involved elements will undergo large deformation at the beginning of the simulation.
+This results in quick increase of internal forces which can cause unwanted effects.
+To allevieate the problem we have implemented the technique that allows for moderate changes in the rest shape between simulation steps.
+Between the start and the end of the simulation we use a linear interpolation of the rest shape for the shells that lie along the sutured edges. 
+\christian{I think that it is unclear that we do the relaxation on some shells that are on the side edges of the connecting line}
+\tomas{I don't understand what do you mean by that. We will have to discuss that.}
+The interpolation starts with the rest shape equal to the deformed elements and we progressively recontract the rest shape of the relaxed shells to their initial shapes. 
+The energetical equilibrium found by the simulation gradually forces the side edges to connect.
+After the mesh relaxes the simulation again terminates.
 
-% vim: et sw=2 tw=75 fdm=marker fdc=2 spell
+% vim: et sw=2 tw=0 spell

### 2012-02-29 Doc/Miccai/shell-model.tex
missing commit of changed done yesterday
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@350 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-The bending property of the shell allows us to use smaller number of
-elements for the simulation and use high polygonal model only for
-visualisation. The convergence of the model is a key factor here. In our
+The convergence of the model is a key factor here. In our
 experiments we were able to use only as few as eight nodes around the
@@
 number of nodes along the circumference.
+As the model well-converges with relatively coarse meshes, we use 
+a quite small number of elements for the simulation and use high polygonal model only for
+visualisation.
 

### 2012-02-29 Doc/Miccai/simulation.tex
modifs on section 3
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@351 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
+The primary goal of the surgery is to restore a normal heart function and circulation.
+This goal is reached by rebuilding the heart in a more or less anatomical way.
 There are several types of low-level task the surgeon usually employs while
-performing the surgery.
-The most common:
+performing the surgery that are:
 
 \begin{description}
-  \item Reconnection and joining of blood vessels to the arteries.
 
@@
 
+  \item[Reconnection and joining of blood vessels:] In addition to the modification of the aorta, the surgeon is lead to establish certain connections that are absent in pathological anatomy.
+
 \end{description}
 
-With our method we are able to simulate all these tasks.
-We have based our general approach on the common task involved in all these operations which is suturing of two or more edges together.
-Since we are interested in predicting the result rather than creating a training simulator for suturing (trained surgeon knows well how to perform the sutures) the simplest way of performing the suture would be to connect the respective edges with springs.
+With our method we are able to simulate these tasks, in order to be able to plan them.
+We have based our general approach on a common characteristic: In all these operations the surgeon sutures two or more edges of the vessel walls together.
+Since we are interested in predicting the results rather than creating a training simulator for suturing (trained surgeon knows well how to perform the sutures) the simplest way of performing the suture would be to connect the respective edges with springs.
 Springs however adds some unwanted energy to the system. 
-Our method is based solely on the relaxation of the mesh of shell elements.
-%% christian: I like the "relaxation" term !!!
+Our method is based on a topological connection of the shell elements that are neighbors of the joining edges and a controlled relaxation of these shell elements' rest shape.
 
@@
 We keep the number of elements in new mesh the same as in the original mesh so that we can define 1:1 mapping between elements of these meshes.
-The Figure \ref{fig-JoiningVessels} explains this operation when joining vessels. 
-Then we start the simulation on the modified mesh while using the original mesh to define rest shapes of all shell elements.
+The Figure \ref{fig-JoiningVessels} (a) and (b) explains this operation when joining vessels. 
+Then we start the simulation on the modified mesh while keeping the information of the original rest shapes of all shell elements.
 
@@
 
-\caption{Scheme showing the method for joining two vessels.
-\christian{Do we need to extend the scheme for every type of aortoplasty (end to end / Patch / Subclavian): it would illustrate that the relaxation method works for every case...}
-  \tomas{No we don't. This there will be real examples in results (if I remember correctly).} }
+\caption{Scheme showing the method for joining two vessels:  (a) the cut on vessel meshes is performed in a 3D geometry edition software.
+In the software, we impose that the number of segments that sample both joining edges is the same (\color{BlueGreen}{$\mathbf{\rightarrow}$}\color{black})
+Moreover, we select some shell elements that are neighbors of the edges (\color{purple}{$\mathbf{\rightarrow}$}\color{black}).
+(b) In our software, we connect the vessel topologies by performing a relaxation of the shape and the rest shape for the selected shells.  
+(c) We progressively re-contract the rest shape of the shells while computing the deformation of the whole unified mesh. 
+In the final steady state computed in the simulation, the shell elements have all their correct rest shape.}
 \label{fig-JoiningVessels}
@@
 
-In case of non-trivial distance between the nodes that are sutured together the involved elements will undergo large deformation at the beginning of the simulation.
-This results in quick increase of internal forces which can cause unwanted effects.
+We can not always step back all the shells to their original shape (passage from figure \ref{fig-JoiningVessels} (b) to (c))  in a single simulation step. 
+In case of non-small distance between the nodes that are sutured together, the involved elements will undergo large deformation that could create quick and non-smooth increase of internal forces and  cause unwanted effects.
 To allevieate the problem we have implemented the technique that allows for moderate changes in the rest shape between simulation steps.
 Between the start and the end of the simulation we use a linear interpolation of the rest shape for the shells that lie along the sutured edges. 
-\christian{I think that it is unclear that we do the relaxation on some shells that are on the side edges of the connecting line}
-\tomas{I don't understand what do you mean by that. We will have to discuss that.}
-The interpolation starts with the rest shape equal to the deformed elements and we progressively recontract the rest shape of the relaxed shells to their initial shapes. 
-The energetical equilibrium found by the simulation gradually forces the side edges to connect.
-After the mesh relaxes the simulation again terminates.
-
-% vim: et sw=2 tw=0 spell
+The interpolation starts with the rest shape equal to the deformed elements and we progressively lead the elements to recover their original rest shape.
+This way, the simulation gradually forces the side edges to connect without any abrupt jump in the behavior.
+When all elements have recover their original rest shape, the steady state found in the simulation corresponds to a physics-based energetical equilibrium of the shells deformation. 
+Thus, we take these positions as an estimation of the result of the connection of the vessel edges.

### 2012-02-29 Doc/Miccai/simulation.tex
modifs on section 3
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@352 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \caption{Scheme showing the method for joining two vessels:  (a) the cut on vessel meshes is performed in a 3D geometry edition software.
-In the software, we impose that the number of segments that sample both joining edges is the same (\color{BlueGreen}{$\mathbf{\rightarrow}$}\color{black})
-Moreover, we select some shell elements that are neighbors of the edges (\color{purple}{$\mathbf{\rightarrow}$}\color{black}).
+In the software, we impose that the number of the segments on the edges is the same on both side  (\color{BlueGreen}{$\mathbf{\rightarrow}$}\color{black})
+Moreover, we select some shell elements that lie along the sutured edges (\color{purple}{$\mathbf{\rightarrow}$}\color{black}).
 (b) In our software, we connect the vessel topologies by performing a relaxation of the shape and the rest shape for the selected shells.  
@@
 
-We can not always step back all the shells to their original shape (passage from figure \ref{fig-JoiningVessels} (b) to (c))  in a single simulation step. 
+We can not always step back all the shells to their original shape (transition from figure \ref{fig-JoiningVessels} (b) to (c))  in a single simulation step. 
 In case of non-small distance between the nodes that are sutured together, the involved elements will undergo large deformation that could create quick and non-smooth increase of internal forces and  cause unwanted effects.
@@
 When all elements have recover their original rest shape, the steady state found in the simulation corresponds to a physics-based energetical equilibrium of the shells deformation. 
-Thus, we take these positions as an estimation of the result of the connection of the vessel edges.
+Thus, we take these positions as an estimation of the result of the connection of the vessels.

### 2012-03-01 Doc/Miccai/introduction.tex
Shortened and improved introduction. Improved (not yet final) paper title.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@353 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2012-02-24 Doc/Miccai/introduction.tex
@@
-TODO: Include only a few sentences to CHD in general.
+Surgical interventions in infants with congenitally malformed great arteries and hearts are extremely challenging due to the complex and heterogeneous nature of their disease pattern.
+Cardiac surgeons rely on non-invasive imaging for patient-specific examination and possible appropriate surgical procedures may be sketched conceptually in advance, but the most promising approach is often chosen just in time during the actual open-heart operation.
+This is the first time the surgeon gets a concrete idea of possible outcomes for different surgical procedures.
+His decision has to be made in short time and is still based heavily on imagination capabilities and experience.
 
-Surgical interventions in infants with congenitally malformed great arteries and hearts can be difficult due to their disease pattern's complex and heterogeneous nature.
+To improve planning of such pediatric cardiac surgeries it is possible to create 3-dimensional heart models out of acquired image data REF and to use them as input for surgical simulation systems.
+However, most present simulation systems focus on low-level surgical procedures, i.e.
+performing actual incisions and suturing with special input devices in real-time REF.
+Their purpose is either by means of educational training or planning access strategies to reach the malformation of the heart of blood vessels.
+In such {\itshape reactive} simulation systems it is more important to enable simulations in real-time than to simulate tissue and blood vessels in a physically accurate manner.
+
+To be able to predict the result of complex surgical procedures, a physically correct simulation is mandatory.
+REF present a {\itshape predictive} surgical simulation system for bidirectional Glenn shunts and basically similar surgeries.
+They use a hybrid model to simulate deformations of blood vessel trees by binding a surface mesh to a reference Cosserat rod REF elastically.
+While the mesh models surface details of blood vessels, the Cosserat rod models global deformations, i.e.
+stretching, bending, twisting and shearing.
+Joining of blood vessels end-to-end or end-to-side is realized by attaching contracting springs.
+This approach adds surplus energy to the physical system and special care has to be taken during the simulation to compensate for that.
+The software performs low-level surgical procedures automatically with only a minimum of interaction required which allows the surgeon to focus on the actual overall surgical procedure.
+Given the input of the surgeon, a complete simulation takes only seconds to produce a predictive result for further examination and evaluation of possibly undesired collisions or deformations of blood vessels.
+Although the results are promising, the method is limited to operations on uniform tubular structures that do not change their surface arbitrary.
+Surgical procedures heavily changing surface details of blood vessels, i.e.
+arbitrary shaped patches for correction of complicated coarctations and stenosis, are not possible with this hybrid model since the physical behavior of blood vessels is simulated by the Cosserat rod and its {\itshape binding} to the surface mesh.
+
+In this paper we present an alternative, improved approach using a simulation model solely based on elastic thin shell elements.
+Our method is not limited to well-formed blood vessels and captures all low-level procedures necessary for aortoplasty.
+By using a relaxation process we can arbitrarily join or attach blood vessels and patches without using springs that would add surplus energy to the physical system.
+We prove that simulation time is still reasonable or even decreased by taking special care of the overall tubular structure of blood vessels during generation of thin shell elements.
+%TODO: Include only a few sentences to CHD in general.
+%
+%Surgical interventions in infants with congenitally malformed great arteries and hearts can be difficult due to their disease pattern's complex and heterogeneous nature.
 Today, patient-specific 3-dimensional heart models can be acquired by modern techniques of non-invasive imaging and image post-processing [REFERENCE].
@@
 Possible appropriate surgical procedures may be sketched conceptually and the most promising approach is chosen just in time during the actual open-heart operation.
-This is the first time the surgeon gets a concrete idea of possible outcomes for different surgical procedures.
-However, his decision has to be made in relatively short time and can still be based heavily on imagination capabilities and experience.
-
-To improve planning of these kind of cardiac surgeries, obtained 3-dimensional heart models could be used as input for patient-specific surgical simulation systems.
-Present simulation systems focus on low-level surgical procedures [REFERENCES], i.e.
+%\christian{to modify [ }
+%This is the first time the surgeon gets a concrete idea of possible outcomes for different surgical procedures. 
+%However, his decision has to be made in relatively short time and can still be based heavily on imagination capabilities and experience.
+%\christian{  ] }
+%
+%
+%To improve planning of these kind of cardiac surgeries, obtained 3-dimensional heart models could be used as input for patient-specific surgical simulation systems. 
+%
+%\christian{not sure this is the right place for the following paragraph (It begins the state of the art while the work is not yet introduced) : [}
+%Present simulation systems \christian{what simulation systems are we talking about ?}  focus on low-level surgical procedures [REFERENCES], i.e.
 performing actual incisions and suturing with special input devices in real-time.
-Their purpose is either by means of educational training or planning access strategies to reach the heart's or blood vessel's malformations.
-It is more important in such \emph{reactive} simulation systems to enable simulations in real-time than to simulate tissue and blood vessels in a physically accurate manner.
-
-To be able to predict the result of complex surgical procedures, a physically correct simulation model must be used.
-[REFERENCE] present a \emph{predictive} surgical simulation system for bidirectional Glenn shunts and basically similar surgeries.
-They use a hybrid model to simulate deformations of blood vessel trees by binding a surface mesh to a reference Cosserat rod elastically.
-While the mesh models surface details of blood vessels, the Cosserat rod models global deformations, i.e.
-stretching, bending, twisting and shearing.
-Low-level surgical procedures are done automatically by the software with only a minimum of interaction required which allows the surgeon to focus on the actual overall surgical procedure.
-Furthermore the authors show that they can attach triangular shaped patches at loose ends of dissected vessels to allow joining of vessels with different circumferences.
-Given the input of the surgeon, a typical simulation takes only seconds to produce a predictive result for further examination and evaluation of possibly undesired collisions or deformations of blood vessels.
-Although the results are promising, the author's method is limited to uniform tubular structures and operations on them, i.e.
-joining blood vessels end-to-end or end-to-side.
-Connection of blood vessels is done by using springs.
-This approach introduces surplus energy to the physical system and special care has to be taken during the simulation to compensate for that.
-Surgical procedures heavily changing surface details of blood vessels, i.e.
+%Their purpose is either by means of educational training or planning access strategies to reach the heart's or blood vessel's malformations. 
+%It is more important in such \emph{reactive} simulation systems to enable simulations in real-time than to simulate tissue and blood vessels in a physically accurate manner.
+%\christian{ ] }
+% 
+%
+%\christian{I think we should present our work as a continuity of what was proposed with the Cosserat model.
+We should reduce the discussion about this paper and show what is improved by our approach.}
+%To be able to predict the result of complex surgical procedures, a physically correct simulation model must be used. 
+%[REFERENCE] present a \emph{predictive} surgical simulation system for bidirectional Glenn shunts and basically similar surgeries. 
+%They use a hybrid model to simulate deformations of blood vessel trees by binding a surface mesh to a reference Cosserat rod elastically.
+%While the mesh models surface details of blood vessels, the Cosserat rod models global deformations, i.e. stretching, bending, twisting and shearing. 
+%Low-level surgical procedures are done automatically by the software with only a minimum of interaction required which allows the surgeon to focus on the actual overall surgical procedure. 
+%Furthermore the authors show that they can attach triangular shaped patches at loose ends of dissected vessels to allow joining of vessels with different circumferences. 
+%Given the input of the surgeon, a typical simulation takes only seconds to produce a predictive result for further examination and evaluation of possibly undesired collisions or deformations of blood vessels.
+%Although the results are promising, the author's method is limited to uniform tubular structures and operations on them, i.e. joining blood vessels end-to-end or end-to-side. 
+%Connection of blood vessels is done by using springs. 
+%This approach introduces surplus energy to the physical system and special care has to be taken during the simulation to compensate for that. 
+%Surgical procedures heavily changing surface details of blood vessels, i.e.
 arbitrary shaped patches to correct complicated coarctations and stenosis, are not possible with this hybrid model because the blood vessel's physical behavior is simulated by the Cosserat rod and its \emph{binding} to the surface mesh.
-The authors state that more sophisticated physical models including finite element methods and thin shell elements would allow this but are computationally too expensive to produce results in reasonable time.
-
-In this paper we present an alternative, universal approach using a simulation model solely based on elastic thin shell elements.
-Our method is not limited to well-formed blood vessels and captures all low-level surgical procedures necessary for aortoplasty.
-By using a relaxation process we arbitrarily join or attach blood vessels and patches anywhere without using springs that would add surplus energy to the physical system.
-We prove that simulation time can be expected to be still reasonable or even decreased by taking special care of the blood vessel's overall tubular structure during generation of thin shell elements.
-
-TODO: Structure of Paper.
+%The authors state that more sophisticated physical models including finite element methods and thin shell elements would allow this but are computationally too expensive to produce results in reasonable time.
+%
+%
+%\christian{\sout{In this paper we present an alternative}   Precisely, I propose to place our paper as an improvement of their approach } , universal approach using a simulation model solely based on elastic thin shell elements.
+%Our method is not limited to well-formed blood vessels and captures all low-level surgical procedures necessary for aortoplasty. 
+%By using a relaxation process we arbitrarily join or attach blood vessels and patches anywhere without using springs that would add surplus energy to the physical system. 
+%We prove that simulation time can be expected to be still reasonable or even decreased by taking special care of the blood vessel's overall tubular structure during generation of thin shell elements.
+%
+%TODO: Structure of Paper.

### 2012-03-01 Doc/Miccai/introduction.tex
Fixed typo in introduction.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@354 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
More corrections of the introduction.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@355 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 performing actual incisions and suturing with special input devices in real-time REF.
-Their purpose is either by means of educational training or planning access strategies to reach the malformation of the heart of blood vessels.
+Their purpose is either by means of educational training or planning access strategies to reach the malformations of the heart or great arteries.
 In such {\itshape reactive} simulation systems it is more important to enable simulations in real-time than to simulate tissue and blood vessels in a physically accurate manner.
@@
 REF present a {\itshape predictive} surgical simulation system for bidirectional Glenn shunts and basically similar surgeries.
-They use a hybrid model to simulate deformations of blood vessel trees by binding a surface mesh to a reference Cosserat rod REF elastically.
+They use a hybrid model to simulate deformations of blood vessels by binding a surface mesh to a reference Cosserat rod REF elastically.
 While the mesh models surface details of blood vessels, the Cosserat rod models global deformations, i.e.
@@
 Given the input of the surgeon, a complete simulation takes only seconds to produce a predictive result for further examination and evaluation of possibly undesired collisions or deformations of blood vessels.
-Although the results are promising, the method is limited to operations on uniform tubular structures that do not change their surface arbitrary.
+Although the results are promising, the method is limited to operations on uniform tubular structures that do not arbitrarily change the surface.
 Surgical procedures heavily changing surface details of blood vessels, i.e.
-arbitrary shaped patches for correction of complicated coarctations and stenosis, are not possible with this hybrid model since the physical behavior of blood vessels is simulated by the Cosserat rod and its {\itshape binding} to the surface mesh.
+patches for correction of complicated coarctations and stenosis, are not possible with this hybrid model since the physical behavior of blood vessels is simulated by the Cosserat rod and its {\itshape binding} to the surface mesh.
 

### 2012-03-01 Doc/Miccai/shell-model.tex
Proposal of an alternative start of the shell model section.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@356 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 Moreover, we show that we can still obtain fast computation with a reasonable number of elements.
+
+\stefan{Alternative start of this section: In congenital heart defect (CHD) corrective surgeries blood vessels respond to plastic interventions with a certain amount of deformation.
+Surgeons try to find an optimal approach to correct defects with a minimum of deformations and collisions that would have unnecessary negative impact on blood flow or that may lead to problems during growth of the patient.
+To be able to predict results for different surgical approaches a physical approriate model of blood vessels is mandatory.
+Furthermore, computation time of simulations must be low enough to enable interactive and usable planning software.
+Options on the choice of model are limited.
+Mass-spring models allow for fast computation but do not correlate to a physical accurate representation of reality and therefore are inappropriate for predictive simulations.
+On the contrary, simulation models based on the formulation of finite elements, either 3-dimensional elements or thin shell elements, are commonly used when high physical accuracy is required but are well known to be computationally expensive.
+The hybrid blood vessel model of REF strikes a balance but is too constrained to be used as a truly universal approach for all kinds of surgical procedures used in CHD corrective surgery.
+We propose a model based on thin shell elements that still allows for fast computation with a reasonable number of elements.}
 

### 2012-03-01 Doc/Miccai/conclusion.tex
Added authors and institutes. Began results section. Began conclusion section.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@357 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

We introduced a new approach for predictive simulation of CHD corrective surgeries. It is based on thin shell elements and a relaxation method for joining of blood vessels and patches by topologically connecting the vertices of joining edges of the simulation meshes. All necessary low-level surgical procedures can be performed using this method. We showed that we need only a small number of thin shell elemenets to achieve physically accurat simulation results which enable fast simulations. Special care has to be taken during initial mesh generation and dynamic remeshing when cutting blood vessels since the size of newly generated thin shell elements should roughly equal the size of their neighbor elements which is a common constraint for finite element simulations. To truly benefit from our simulation approach, future work has to be done in terms of remeshing, to keep a high quality of the mesh during the whole simulation.

### 2012-03-01 Doc/Miccai/miccai2012.tex
miccai 2012: shell model and method sections
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@339 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
comments about the intro
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@344 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
modifs in section 2
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@348 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
modifs on section 3
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@351 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
Shortened and improved introduction. Improved (not yet final) paper title.
git

@@
 \usepackage{color}
+\usepackage[usenames,dvipsnames]{xcolor}
 \usepackage{amsmath}
@@
 \definecolor{darkgray}{rgb}{0.2,0.2,0.2}
+\usepackage{ulem}
 
@@
 
-\title{Towards planning of infantile cardiac surgery based on simulation}
+\title{Simulation of Congenital Heart Defect Corrective Surgeries Using Thin Shell Elements}
 
-\author{PaperID:} % TODO
+\author{Stefan Kislinskiy\inst{1}
+\and Tom\'{a}\v{s} Golembiovsk\v{y}\inst{2,3}
+\and Christian Duriez\inst{2}
+\and Eug\'{e}nie Riesenkampff\inst{4}
+\and Titus Kuehne\inst{4}
+\and H.-P. Meinzer\inst{1}
+\and Tobias Heimann\inst{1}}
+
+\institute{German Cancer Research Center (DKFZ), Heidelberg, Germany
+\and Inria, Lille, France
+\and Faculty of Informatics, Masaryk University, Brno, Czech Republic
+\and German Heart Institute Berlin (DHZB), Berlin, Germany}
 
@@
 
-\section{Blood Vessel Model}
+\section{Blood Vessel Model based on shell theory}
 \input{shell-model.tex}
@@
 \section{Simulation of Surgical Procedures}
+\label{sec-method}
 \input{simulation.tex}

### 2012-03-01 Doc/Miccai/results.tex
Added authors and institutes. Began results section. Began conclusion section.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@357 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-Example with 3 diff. surgical procedures
+In order to show an improvement on existing predictive simulation systems for CHD corrective surgeries, we want to demonstrate that our blood vessel model based on thin shell elements is
+\begin{enumerate}
+\item physically accurate,
+\item converges to expected results with a low amount of elements,
+\item is able to simulate low-level surgical procedures universally.
+\end{enumerate}
+A quantitative validation by means of a comparision to real pre- and post-surgery image data is impossible at present.
+This is mainly due to ethical issues that arise during image data acquirement of infants, i.e.
+necessary heart rate lowering medications or exposure to radiation.
+There is no such data available that is obtained shortly after surgeries in infants where the growth of the patient would not interfere comparison.
+However, qualitative comparisons to a real image data set acquired months after a surgical intervention and to an elastic tube manipulated by hand, which confirm a principal suitability of predictive simulations for surgery planning, can be found in \cite{Li2009}.
 
-Parameters
+We close this section with a discussion about the limitations and constraints of our approach which have great impact on the mesh generation prior to simulation.
 
-Limitations (getting image data, meshes, dynamic remeshing, same number of nodes)
+\subsection{Accuracy}
+
+Figure \ref{fig-deformations} shows bending and twisting of an elastic tube manipulated by hand compared to the simulation results of \cite{Li2009} and our method.
+These kinds of blood vessel deformations are likely to occur during surgery.
+Our results are close to the real deformations as expected due to the physically based formulation of thin shell elements.
+
+\subsection{Convergence}
+
+It is important to minimize the number of shell elements to reduce computation time.
+This is the key factor to enable an interactive simulation system that is accepted by surgeons to simulate multiple surgical approaches in succession.
+The bending of thin shell elements allows to reduce their number by still representing the tubular structure of blood vessels while maintaining their physical behavior appropriately.
+Figure \ref{fig-convergence} shows simulation results of a strongly deformed tube represented by different numbers of thin shell elements.
+With only eight vertices along the circumference of the tube the simulation result is comparable to a simulation result with high element count.
+We map a fine polygonal mesh to the coarse shell element meshes to visualize their bending detail and therefore get comparable results that correlates with the true shape of the simulated tube.
+
+\subsection{Low-level surgical procedures}
+
+TODO: Use the remaining space we have in the paper. We should consider to shorten other sections.
+
+\subsection{Limitations and Constraints}
+
+%Example with 3 diff. surgical procedures
+%Parameters
+%Limitations (getting image data, meshes, dynamic remeshing, same number of nodes)

### 2012-03-01 Doc/Miccai/introduction.tex
Added authors and institutes. Began results section. Began conclusion section.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@357 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
comments on intro and section2
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@358 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 Their purpose is either by means of educational training or planning access strategies to reach the malformations of the heart or great arteries.
-In such {\itshape reactive} simulation systems it is more important to enable simulations in real-time than to simulate tissue and blood vessels in a physically accurate manner.
+In such {\itshape reactive} \christian{interactive ?} simulation systems it is more important to enable simulations in real-time than to simulate tissue and blood vessels in a physically accurate manner.
+\christian{We could cite some existing paper about interactive simulation for training}
 
 To be able to predict the result of complex surgical procedures, a physically correct simulation is mandatory.
-REF present a {\itshape predictive} surgical simulation system for bidirectional Glenn shunts and basically similar surgeries.
+In \cite{Li2009} is presented a {\itshape predictive} surgical simulation system for bidirectional Glenn shunts and basically similar surgeries.
 They use a hybrid model to simulate deformations of blood vessels by binding a surface mesh to a reference Cosserat rod REF elastically.
@@
 patches for correction of complicated coarctations and stenosis, are not possible with this hybrid model since the physical behavior of blood vessels is simulated by the Cosserat rod and its {\itshape binding} to the surface mesh.
+\christian{I think this paragraph is much too long and I think that our paper should not be presented in opposition to \cite{Li2009} but in the continuity of this work !!
+I put a correction of this in a commit 2 days ago, could we try to merge ?}
 

### 2012-03-01 Doc/Miccai/introduction.tex
Responded to Christian's comments.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@359 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 Their purpose is either by means of educational training or planning access strategies to reach the malformations of the heart or great arteries.
-In such {\itshape reactive} \christian{interactive ?} simulation systems it is more important to enable simulations in real-time than to simulate tissue and blood vessels in a physically accurate manner.
-\christian{We could cite some existing paper about interactive simulation for training}
+In such {\itshape reactive} \christian{interactive ?} \stefan{reactive is the term used by current papers} simulation systems it is more important to enable simulations in real-time than to simulate tissue and blood vessels in a physically accurate manner.
+\christian{We could cite some existing paper about interactive simulation for training} \stefan{We can cite two papers here I will add today. They will be put to the second REF in this paragraph.}
 
@@
 \christian{I think this paragraph is much too long and I think that our paper should not be presented in opposition to \cite{Li2009} but in the continuity of this work !!
-I put a correction of this in a commit 2 days ago, could we try to merge ?}
+I put a correction of this in a commit 2 days ago, could we try to merge ?} \stefan{I already took your comments into account and tried to present our approach as a consequent continuity of this work.
+The paragraph was shortened by some sentences also (see the comments in the bottom of the Tex-File.
+I'm not sure if we should further shorten this paragraph \emph{especially} when we present our work as continuity.
+Because that paragraph is also an introduction for our work then, isn't it?
+In the end the paragraph says what can be improved to justify our improved approach.
+Maybe when we just delete the "alternative" in the next paragpraph?}
 
-In this paper we present an alternative, improved approach using a simulation model solely based on elastic thin shell elements.
+In this paper we present an \stefan{(}alternative\stefan{)}, improved approach using a simulation model solely based on elastic thin shell elements.
 Our method is not limited to well-formed blood vessels and captures all low-level procedures necessary for aortoplasty.

### 2012-03-01 Doc/Miccai/shell-model.tex
comments on intro and section2
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@358 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
Responded to Christian's comments.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@359 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 We propose a model based on thin shell elements that still allows for fast computation with a reasonable number of elements.}
+\christian{I think we really need to do a merge because there are interesting things in both section but in both cases, we are missing references !!
+One other important feature: I think it is important to put some illustrations of the surgery either here or in the intro.} \stefan{Yes, we truly have to merge here and add references.
+My addition focused more on English language.}
 

### 2012-03-01 Doc/Miccai/simulation.tex
Added annotations to the rest shape figure.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@361 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\caption{Scheme showing the method for joining two vessels:  (a) the cut on vessel meshes is performed in a 3D geometry edition software.
+\caption{Scheme showing the method for joining two vessels \stefan{Isn't that example too complex?
+I think it is really hard to actually see what is happening here.
+At least the colors have to be changed because the different layers are colored quite equally and it should be zoomed in to the interesting area were the joining takes place.
+Remember also that the procedings are printed in black and white and that everything must be visible when printing out this figure black and white.
+There must be way more contrast.}:  (a) the cut on vessel meshes is performed in a 3D geometry edition software.
 In the software, we impose that the number of the segments on the edges is the same on both side  (\color{BlueGreen}{$\mathbf{\rightarrow}$}\color{black})

### 2012-03-01 Doc/Miccai/introduction.tex
Made some changes to introduction to shorten it.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@362 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 To be able to predict the result of complex surgical procedures, a physically correct simulation is mandatory.
-In \cite{Li2009} is presented a {\itshape predictive} surgical simulation system for bidirectional Glenn shunts and basically similar surgeries.
-They use a hybrid model to simulate deformations of blood vessels by binding a surface mesh to a reference Cosserat rod REF elastically.
+\cite{Li2009} present a {\itshape predictive} surgical simulation system for bidirectional Glenn shunts and basically similar surgeries.
+The authors use a hybrid model to simulate deformations of blood vessels by binding a surface mesh to a reference Cosserat rod REF elastically.
 While the mesh models surface details of blood vessels, the Cosserat rod models global deformations, i.e.
 stretching, bending, twisting and shearing.
-Joining of blood vessels end-to-end or end-to-side is realized by attaching contracting springs.
-This approach adds surplus energy to the physical system and special care has to be taken during the simulation to compensate for that.
-The software performs low-level surgical procedures automatically with only a minimum of interaction required which allows the surgeon to focus on the actual overall surgical procedure.
+\stefan{We can delete this: (}Joining of blood vessels end-to-end or end-to-side is realized by attaching contracting springs.
+This approach adds surplus energy to the physical system and special care has to be taken during the simulation to compensate for that.\stefan{)} The software performs low-level surgical procedures automatically with only a minimum of interaction required which allows the surgeon to focus on the actual overall surgical procedure.
 Given the input of the surgeon, a complete simulation takes only seconds to produce a predictive result for further examination and evaluation of possibly undesired collisions or deformations of blood vessels.

### 2012-03-01 Doc/Miccai/introduction.tex
fixing my name and adding comments to intro
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@365 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 To be able to predict the result of complex surgical procedures, a physically correct simulation is mandatory.
-\cite{Li2009} present a {\itshape predictive} surgical simulation system for bidirectional Glenn shunts and basically similar surgeries.
+\cite{Li2009} presents a {\itshape predictive} surgical simulation system for bidirectional Glenn shunts and basically similar surgeries.
 The authors use a hybrid model to simulate deformations of blood vessels by binding a surface mesh to a reference Cosserat rod REF elastically.
@@
 \stefan{We can delete this: (}Joining of blood vessels end-to-end or end-to-side is realized by attaching contracting springs.
-This approach adds surplus energy to the physical system and special care has to be taken during the simulation to compensate for that.\stefan{)} The software performs low-level surgical procedures automatically with only a minimum of interaction required which allows the surgeon to focus on the actual overall surgical procedure.
-Given the input of the surgeon, a complete simulation takes only seconds to produce a predictive result for further examination and evaluation of possibly undesired collisions or deformations of blood vessels.
+This approach adds surplus energy to the physical system and special care has to be taken during the simulation to compensate for that.\stefan{)}
+\tomas{We can remove this also: (} The software performs low-level surgical procedures automatically with only a minimum of interaction required which allows the surgeon to focus on the actual overall surgical procedure.
+Given the input of the surgeon, a complete simulation takes only seconds to produce a predictive result for further examination and evaluation of possibly undesired collisions or deformations of blood vessels.\tomas{) Or replace with shorter: The surgeon's input is limited to point-and-click and the software performs low-level surgical procedures automatically.}
 Although the results are promising, the method is limited to operations on uniform tubular structures that do not arbitrarily change the surface.

### 2012-03-01 Doc/Miccai/introduction.tex
Corrected Introduction and second chapter.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@370 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2012-03-01 Doc/Miccai/introduction.tex
@@
 Their purpose is either by means of educational training or planning access strategies to reach the malformations of the heart or great arteries.
-In such {\itshape reactive} \christian{interactive ?} \stefan{reactive is the term used by current papers} simulation systems it is more important to enable simulations in real-time than to simulate tissue and blood vessels in a physically accurate manner.
-\christian{We could cite some existing paper about interactive simulation for training} \stefan{We can cite two papers here I will add today. They will be put to the second REF in this paragraph.}
+In such {\itshape reactive} simulation systems it is more important to enable simulations in real-time than to simulate tissue and blood vessels in a physically accurate manner.
 
 To be able to predict the result of complex surgical procedures, a physically correct simulation is mandatory.
-\cite{Li2009} present a {\itshape predictive} surgical simulation system for bidirectional Glenn shunts and basically similar surgeries.
+\cite{Li2009} presents a {\itshape predictive} surgical simulation system for bidirectional Glenn shunts and basically similar surgeries.
 The authors use a hybrid model to simulate deformations of blood vessels by binding a surface mesh to a reference Cosserat rod REF elastically.
@@
 stretching, bending, twisting and shearing.
-\stefan{We can delete this: (}Joining of blood vessels end-to-end or end-to-side is realized by attaching contracting springs.
-This approach adds surplus energy to the physical system and special care has to be taken during the simulation to compensate for that.\stefan{)} The software performs low-level surgical procedures automatically with only a minimum of interaction required which allows the surgeon to focus on the actual overall surgical procedure.
-Given the input of the surgeon, a complete simulation takes only seconds to produce a predictive result for further examination and evaluation of possibly undesired collisions or deformations of blood vessels.
+The simulation software performs low-level surgical procedures automatically with only a minimum of interaction required which allows the surgeon to focus on the actual overall surgical procedure.
 Although the results are promising, the method is limited to operations on uniform tubular structures that do not arbitrarily change the surface.
@@
 patches for correction of complicated coarctations and stenosis, are not possible with this hybrid model since the physical behavior of blood vessels is simulated by the Cosserat rod and its {\itshape binding} to the surface mesh.
-\christian{I think this paragraph is much too long and I think that our paper should not be presented in opposition to \cite{Li2009} but in the continuity of this work !!
-I put a correction of this in a commit 2 days ago, could we try to merge ?} \stefan{I already took your comments into account and tried to present our approach as a consequent continuity of this work.
-The paragraph was shortened by some sentences also (see the comments in the bottom of the Tex-File.
-I'm not sure if we should further shorten this paragraph \emph{especially} when we present our work as continuity.
-Because that paragraph is also an introduction for our work then, isn't it?
-In the end the paragraph says what can be improved to justify our improved approach.
-Maybe when we just delete the "alternative" in the next paragpraph?}
 
-In this paper we present an \stefan{(}alternative\stefan{)}, improved approach using a simulation model solely based on elastic thin shell elements.
+In this paper we present an improved approach using a simulation model solely based on elastic thin shell elements.
 Our method is not limited to well-formed blood vessels and captures all low-level procedures necessary for aortoplasty.

### 2012-03-01 Doc/Miccai/shell-model.tex
Corrected Introduction and second chapter.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@370 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2012-03-01 Doc/Miccai/shell-model.tex
@@
-\subsubsection{Requirements: } 
-Infantile cardiac surgery offers a reconfiguration of the blood vessel walls that limits or rectifies the congenital anomalies of the anatomy.
-During the surgery, the vessel walls are deformed, cut, and reconnected to other vessels.
-To do a planning of the surgery, the geometry changes of these vessel walls (the motions and deformations, their configuration when they are cut or stitched)  must be anticipated.
-
-These requirements ask for a physics-based model for the vessel walls. 
-This model should provide 
-(I) a good support for the deformations by bending and twisting. 
-(II) an accurate modeling of the wall motion when the vessel is cut along its axis 
-(III) global deformations of stitched vessels that capture their final configuration, after surgery 
-(IV) collision response when detected (e.g. between the aortic arch and pulmonary artery).
-
-In addition, the computation of this model must be fast enough to allows for interactive planning of the surgery. 
-The options on the choice of model are limited: 
-The model based on the deformation of the centerline, that can be very quickly computed, (like ??\tomas{Add refs.!} \christian{Do you have some refs that uses model based on centerline?
-}\tomas{No, but I will find something}  ) must be discarded because they do not support cut of the wall along the vessel axis.
-Mass-springs approaches that are often used for cloth simulation \cite{Volino2005b}, allow for fast computation but do not provide sufficient accuracy as they do not necessary converge to a unique solution when changing the mesh.
-In this work, we then propose to rely on Finite Element formulation of thin-shell theory. 
-These model are numerically more expensive, but provide accurate models of thin deformable structure.% that converge to unique solutions.
-Moreover, we show that we can still obtain fast computation with a reasonable number of elements.
-
-\stefan{Alternative start of this section: In congenital heart defect (CHD) corrective surgeries blood vessels respond to plastic interventions with a certain amount of deformation.
+In congenital heart defect (CHD) corrective surgeries blood vessels respond to plastic interventions with a certain amount of deformation.
 Surgeons try to find an optimal approach to correct defects with a minimum of deformations and collisions that would have unnecessary negative impact on blood flow or that may lead to problems during growth of the patient.
@@
 On the contrary, simulation models based on the formulation of finite elements, either 3-dimensional elements or thin shell elements, are commonly used when high physical accuracy is required but are well known to be computationally expensive.
-The hybrid blood vessel model of REF strikes a balance but is too constrained to be used as a truly universal approach for all kinds of surgical procedures used in CHD corrective surgery.
-We propose a model based on thin shell elements that still allows for fast computation with a reasonable number of elements.}
+The hybrid blood vessel model of \cite{Li2009} strikes a balance but is too constrained to be used as a truly universal approach for all kinds of surgical procedures used in CHD corrective surgery.
+We propose a model based on thin shell elements that still allows for fast computation with a reasonable number of elements.
 
-\subsubsection{Thin shell elements: } 
-The formulation of the thin shell elements is based on the mechanics of continuum. 
-For thorough description we refer the reader to the textbook about Finite Element methods for example \cite{Reddy1993}.
-These elements are commonly used in physics and mechanics for deformation modelling of thin structures because they are numerically more stable then volumetric FEM methods in this situation. 
-The shell element is a 2-dimensional element (in our case, triangular element) whose thickness is a parameter of the simulation (often considered constant over the surface of the element).
-Shell element combines forces from in-plane deformations, like stretching or shearing, with bending forces. 
-Such internal forces $\myvec{f}$ are a function of current displacement measured as difference of current shape
-$\myvec{x}$ from the rest shape $\myvec{x_0}$.
-To accelerate the computation, we use a corotational formulation, that is based on the definition of a local frame for each element (that can be built using the geometry of the triangle, \christian{see figure "image: single shell" ??
-}).
+Formulation of thin shell elements is based on the machanics of continuum.
+For a thorough description we refer the reader to a standard textbook about finite element methods (FEM), i.e.
+\cite{Reddy1993}.
+Thin shell elements are commonly used in physics and mechanics to model deformation of thin structures because in that case they are numerically more stable than volumetric FEMs.
+A shell element is a 2-dimensional element whose thickness is a parameter of the simulation.
+They combine forces from in-plane deformations, i.e.
+stretching or shearing, with bending forces.
+Such internal forces $\myvec{f}$ are a function of current displacement measured as difference of the current shape $\myvec{x}$ to the rest shape $\myvec{x_0}$.
+To accelerate computation, we use a corotational formulation that is based on the definition of local frames for each element.
 
-In this local frame, the computation of the internal forces is linear:
+In these local frames computation of internal forces is linear:
 \begin{equation}
@@
 \end{equation}
-where $\mymat{K_e}$ is a stiffness matrix that is computed thanks to a weak formulation (integral formulation) of Kirshohoff-Loves equations.
-This approach is very similar to the model presented in  \cite{Comas2010b,Comas2010c}  even if the interpolation functions we use are  different to allow for better convergence\footnote{Our shell model will be published separately}.
-We emphasize that the method described in the paper is independent of the shell model used (other non-corotational shell models can be used as well).
-%
-When the structure is assembled, the contributions of each element are computed in the global frame using a different rotation matrix for each element.
+where $\mymat{K_e}$ is a stiffness matrix that is computed by Kirshohoff-Loves equations.
+This approach is similar to the model presented in  \cite{Comas2010b,Comas2010c}, although the interpolation functions we use are different to allow for better convergence\footnote{Our thin shell element model will be published separately}.
+We emphasize that the method described in this paper is independent of the concrete thin shell element model used and other non-corotational shell models can be used as well.
+When the simulation mesh is assembled, the contributions of each element are computed in a global frame using a different rotation matrix for each element:
 \begin{equation}
@@
 \end{equation}
-Where $\mathcal{A}$ is the assembling operator over the $N$ elements of the mesh, $\mathbb{K}$ is the global stiffness and $\mathbb{F}$ the internal forces of the entire shell mesh structure. 
-Using this approach, $\mathbb{K}$ is changing at each step of the simulation but the model is able to catch geometric non-linear deformations.
-In practice, the computation of the node motion $\mathbb{U} = \mathbb{K}^{-1} \mathbb{F}$ is performed using an optimized conjugate gradient algorithm.
+where $\mathcal{A}$ is the assembling operator over $N$ elements of the mesh, $\mathbb{K}$ is the global stiffness and $\mathbb{F}$ the internal forces of the entire shell element mesh structure.
+Using this approach, $\mathbb{K}$ is changing at each step of the simulation but the model is able to include geometric non-linear deformations.
+In practice, computation of node motion $\mathbb{U} = \mathbb{K}^{-1} \mathbb{F}$ is performed using a conjugate gradient algorithm.
 
-\subsubsection{Accuracy: } 
-In the following of the section we want to show that the performance of a model
-based on shell elements is adequate for the simulation.
+% \subsubsection{Requirements: } 
+% Infantile cardiac surgery offers a reconfiguration of the blood vessel walls that limits or rectifies the congenital anomalies of the anatomy.
+% During the surgery, the vessel walls are deformed, cut, and reconnected to other vessels.
+% To do a planning of the surgery, the geometry changes of these vessel walls (the motions and deformations, their configuration when they are cut or stitched)  must be anticipated.
 
-\begin{figure}[tbh]
-  \begin{minipage}[b]{0.3\linewidth}
-      \framebox[1cm]{\rule{0pt}{1cm}}
-      %\includegraphics[width=\columnwidth]{img/}
-      \caption{Bending shell element}
-      \label{fig-shell}
-  \end{minipage}
-  \hspace{0.1\columnwidth}
-  \begin{minipage}[b]{0.6\linewidth}
-    \centering
-    \begin{tabular}{cc}
-     \includegraphics[width=0.5\columnwidth]{img/compare-bend.png}
-      &
-      \includegraphics[width=0.5\columnwidth]{img/compare-twist.png}
-      \\
-      \includegraphics[width=0.5\columnwidth]{img/compare-bend-other.png}
-      &
-      \includegraphics[width=0.5\columnwidth]{img/compare-twist-other.png}
-    \end{tabular}
-    \caption{Bending and twisting deformations. Upper: shell model; middle:
-    artificial tube, bottom: Cosserat rod \cite{Li2009}. (Credits for the
-    bottom images: Li et al.)}
-    \label{fig-deformations}
-  \end{minipage}
-\end{figure}
+% These requirements ask for a physics-based model for the vessel walls. 
+% This model should provide 
+% (I) a good support for the deformations by bending and twisting. 
+% (II) an accurate modeling of the wall motion when the vessel is cut along its axis 
+% (III) global deformations of stitched vessels that capture their final configuration, after surgery 
+% (IV) collision response when detected (e.g. between the aortic arch and pulmonary artery).
 
-Figure \ref{fig-deformations} shows bending and twisting deformations.
-These are likely to occur during the surgery.
-The figure compares our
-results with the results of Li et al. \cite{Li2009}. As you can see the
-results are close to the real deformations.
+% In addition, the computation of this model must be fast enough to allows for interactive planning of the surgery. 
+% The options on the choice of model are limited: 
+% The model based on the deformation of the centerline, that can be very quickly computed, (like ??\tomas{Add refs.!} \christian{Do you have some refs that uses model based on centerline?
+}\tomas{No, but I will find something}  ) must be discarded because they do not support cut of the wall along the vessel axis.
+% Mass-springs approaches that are often used for cloth simulation \cite{Volino2005b}, allow for fast computation but do not provide sufficient accuracy as they do not necessary converge to a unique solution when changing the mesh.
+% In this work, we then propose to rely on Finite Element formulation of thin-shell theory. 
+% These model are numerically more expensive, but provide accurate models of thin deformable structure.% that converge to unique solutions.
+% Moreover, we show that we can still obtain fast computation with a reasonable number of elements.
 
-The convergence of the model is a key factor here. In our
-experiments we were able to use only as few as eight nodes around the
-circumference of a tubular model while still maintaining realistic
-behaviour. Figure \ref{fig-convergence} shows the results for different
-number of nodes along the circumference.
-As the model well-converges with relatively coarse meshes, we use 
-a quite small number of elements for the simulation and use high polygonal model only for
-visualisation.
+% \stefan{Alternative start of this section: }
+% \christian{I think we really need to do a merge because there are interesting things in both section but in both cases, we are missing references !!
+% One other important feature: I think it is important to put some illustrations of the surgery either here or in the intro.} \stefan{Yes, we truly have to merge here and add references.
+My addition focused more on English language.}
 
-\begin{figure}[tbh]
-  \centering
-  \begin{tabular}{cccc}
-    \includegraphics[width=0.24\columnwidth]{img/twist-06-cg.png}
-    &
-    \includegraphics[width=0.24\columnwidth]{img/twist-08-cg.png}
-    &
-    \includegraphics[width=0.24\columnwidth]{img/twist-16-cg.png}
-    &
-    \includegraphics[width=0.24\columnwidth]{img/twist-31-cg.png}
-    \\
-    \includegraphics[width=0.24\columnwidth]{img/twist-06w-cg.png}
-    &
-    \includegraphics[width=0.24\columnwidth]{img/twist-08w-cg.png}
-    &
-    \includegraphics[width=0.24\columnwidth]{img/twist-16w-cg.png}
-    &
-    \includegraphics[width=0.24\columnwidth]{img/twist-31w-cg.png}
-  \end{tabular}
-  \caption{Convergence on a twisted cylinder with 6, 8, 16 and 31 nodes
-  around the circumference (120, 208, 832 and 3038 elements).
-  Using mesh with 3038 triangles for visualisation.
-  \tomas{improve the wireframe images if there's time} } 
-  \label{fig-convergence}
-\end{figure}
+% \subsubsection{Thin shell elements: } 
+% The formulation of the thin shell elements is based on the mechanics of continuum. 
+% For thorough description we refer the reader to the textbook about Finite Element methods for example \cite{Reddy1993}.
+% These elements are commonly used in physics and mechanics for deformation modelling of thin structures because they are numerically more stable then volumetric FEM methods in this situation. 
+% The shell element is a 2-dimensional element (in our case, triangular element) whose thickness is a parameter of the simulation (often considered constant over the surface of the element).
+% Shell element combines forces from in-plane deformations, like stretching or shearing, with bending forces. 
+% Such internal forces $\myvec{f}$ are a function of current displacement measured as difference of current shape
+% $\myvec{x}$ from the rest shape $\myvec{x_0}$.
+% To accelerate the computation, we use a corotational formulation, that is based on the definition of a local frame for each element (that can be built using the geometry of the triangle, \christian{see figure "image: single shell" ??
+}).
 
-\christian{ I think this should be put in the result section:
-Since we use the method as a
-geometrical tool we have also certain independence on physical parameters.
-For example the variation of Young's modulus in the interval from $10^3$ to
-$10^7$ resulted in only $0.1\%$ change of the deformation.
-}
+% In this local frame, the computation of the internal forces is linear:
+% \begin{equation}
+  % \myvec{f} = \mathbf{K_e}(\myvec{x} - \myvec{x_0})
+ % \label{linear_static_model} 
+% \end{equation}
+% where $\mymat{K_e}$ is a stiffness matrix that is computed thanks to a weak formulation (integral formulation) of Kirshohoff-Loves equations.
+% This approach is very similar to the model presented in  \cite{Comas2010b,Comas2010c}  even if the interpolation functions we use are  different to allow for better convergence\footnote{Our shell model will be published separately}.
+% We emphasize that the method described in the paper is independent of the shell model used (other non-corotational shell models can be used as well).
+% %
+% When the structure is assembled, the contributions of each element are computed in the global frame using a different rotation matrix for each element.
+% \begin{equation}
+% \mathbb{K} = \mathcal{A}_{e=0}^N (\mathbf{R_e}^T \mathbf{K_e} \mathbf{R_e}) \hspace{10mm}
+% \mathbb{F} = \mathcal{A}_{e=0}^N ( \mathbf{R_e}^T \mathbf{K_e}(\myvec{x} - \myvec{x_0}))
+% \end{equation}
+% Where $\mathcal{A}$ is the assembling operator over the $N$ elements of the mesh, $\mathbb{K}$ is the global stiffness and $\mathbb{F}$ the internal forces of the entire shell mesh structure. 
+% Using this approach, $\mathbb{K}$ is changing at each step of the simulation but the model is able to catch geometric non-linear deformations.
+% In practice, the computation of the node motion $\mathbb{U} = \mathbb{K}^{-1} \mathbb{F}$ is performed using an optimized conjugate gradient algorithm.
+
+% \subsubsection{Accuracy: } 
+% In the following of the section we want to show that the performance of a model
+% based on shell elements is adequate for the simulation.
+
+% \begin{figure}[tbh]
+  % \begin{minipage}[b]{0.3\linewidth}
+      % \framebox[1cm]{\rule{0pt}{1cm}}
+      % %\includegraphics[width=\columnwidth]{img/}
+      % \caption{Bending shell element}
+      % \label{fig-shell}
+  % \end{minipage}
+  % \hspace{0.1\columnwidth}
+  % \begin{minipage}[b]{0.6\linewidth}
+    % \centering
+    % \begin{tabular}{cc}
+     % \includegraphics[width=0.5\columnwidth]{img/compare-bend.png}
+      % &
+      % \includegraphics[width=0.5\columnwidth]{img/compare-twist.png}
+      % \\
+      % \includegraphics[width=0.5\columnwidth]{img/compare-bend-other.png}
+      % &
+      % \includegraphics[width=0.5\columnwidth]{img/compare-twist-other.png}
+    % \end{tabular}
+    % \caption{Bending and twisting deformations. Upper: shell model; middle:
+    % artificial tube, bottom: Cosserat rod \cite{Li2009}. (Credits for the
+    % bottom images: Li et al.)}
+    % \label{fig-deformations}
+  % \end{minipage}
+% \end{figure}
+
+% Figure \ref{fig-deformations} shows bending and twisting deformations.
+% These are likely to occur during the surgery.
+% The figure compares our
+% results with the results of Li et al. \cite{Li2009}. As you can see the
+% results are close to the real deformations.
+
+% The convergence of the model is a key factor here. In our
+% experiments we were able to use only as few as eight nodes around the
+% circumference of a tubular model while still maintaining realistic
+% behaviour. Figure \ref{fig-convergence} shows the results for different
+% number of nodes along the circumference.
+% As the model well-converges with relatively coarse meshes, we use 
+% a quite small number of elements for the simulation and use high polygonal model only for
+% visualisation.
+
+% \begin{figure}[tbh]
+  % \centering
+  % \begin{tabular}{cccc}
+    % \includegraphics[width=0.24\columnwidth]{img/twist-06-cg.png}
+    % &
+    % \includegraphics[width=0.24\columnwidth]{img/twist-08-cg.png}
+    % &
+    % \includegraphics[width=0.24\columnwidth]{img/twist-16-cg.png}
+    % &
+    % \includegraphics[width=0.24\columnwidth]{img/twist-31-cg.png}
+    % \\
+    % \includegraphics[width=0.24\columnwidth]{img/twist-06w-cg.png}
+    % &
+    % \includegraphics[width=0.24\columnwidth]{img/twist-08w-cg.png}
+    % &
+    % \includegraphics[width=0.24\columnwidth]{img/twist-16w-cg.png}
+    % &
+    % \includegraphics[width=0.24\columnwidth]{img/twist-31w-cg.png}
+  % \end{tabular}
+  % \caption{Convergence on a twisted cylinder with 6, 8, 16 and 31 nodes
+  % around the circumference (120, 208, 832 and 3038 elements).
+  % Using mesh with 3038 triangles for visualisation.
+  % \tomas{improve the wireframe images if there's time} } 
+  % \label{fig-convergence}
+% \end{figure}
+
+% \christian{ I think this should be put in the result section:
+% Since we use the method as a
+% geometrical tool we have also certain independence on physical parameters.
+% For example the variation of Young's modulus in the interval from $10^3$ to
+% $10^7$ resulted in only $0.1\%$ change of the deformation.
+% }
 

### 2012-03-01 Doc/Miccai/introduction.tex
changes to introduction
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@372 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 In such {\itshape reactive} simulation systems it is more important to enable simulations in real-time than to simulate tissue and blood vessels in a physically accurate manner.
+In reality the surgeon is already skilled in suturing and surgical procedures necessary to access the heart.
+What is missing is a simple tool that would help the surgeon with the phase of pre-operative planing and decisions instead of focusing on how to perform low-level tasks.
 
 To be able to predict the result of complex surgical procedures, a physically correct simulation is mandatory.
-\cite{Li2009} presents a {\itshape predictive} surgical simulation system for bidirectional Glenn shunts and basically similar surgeries.
-The authors use a hybrid model to simulate deformations of blood vessels by binding a surface mesh to a reference Cosserat rod REF elastically.
-While the mesh models surface details of blood vessels, the Cosserat rod models global deformations, i.e.
-stretching, bending, twisting and shearing.
+In \cite{Li2009} a {\itshape predictive} surgical simulation system for bidirectional Glenn shunts and similar surgeries is presented.
+The authors use a hybrid model to simulate deformations of blood vessels by binding a surface mesh to a reference Cosserat rod elastically.
+%While the mesh models surface details of blood vessels, the Cosserat rod models global deformations. , i.e. stretching, bending, twisting and shearing
 The simulation software performs low-level surgical procedures automatically with only a minimum of interaction required which allows the surgeon to focus on the actual overall surgical procedure.
-Although the results are promising, the method is limited to operations on uniform tubular structures that do not arbitrarily change the surface.
-Surgical procedures heavily changing surface details of blood vessels, i.e.
+
+%Although the results are promising, the method is limited to operations on uniform tubular structures that do not arbitrarily change the surface.
+%Surgical procedures heavily changing surface details of blood vessels, i.e.
 patches for correction of complicated coarctations and stenosis, are not possible with this hybrid model since the physical behavior of blood vessels is simulated by the Cosserat rod and its {\itshape binding} to the surface mesh.
+Although the results are promising, the method is limited to operations on uniform tubular structures and doesn't allow arbitrary changes to surface mesh.
 
-In this paper we present an improved approach using a simulation model solely based on elastic thin shell elements.
+%In this paper we present an improved approach using a simulation model solely based on elastic thin shell elements.
 Our method is not limited to well-formed blood vessels and captures all low-level procedures necessary for aortoplasty.
@@
 We prove that simulation time is still reasonable or even decreased by taking special care of the overall tubular structure of blood vessels during generation of thin shell elements.
+We improve the capabilities of the simulation by using a model based on elastic thin shell elements.
+This allows us to model arbitrary blood vessels and malformations of the artery.
+It can also capture all low-level procedures necessary for aortoplasty.
+By using a relaxation process we can arbitrarily join or attach blood vessels and patches without using springs that would add surplus energy to the physical system.
+\tomas{I don't think this is correct: (}
+We prove that simulation time is still reasonable or even decreased by taking special care of the overall tubular structure of blood vessels during generation of thin shell elements.
+\tomas{) we don't make any speed comparison. I propose this: }
+By taking into consideration the overall tubular structure of blood vessels during generation of thin shell elements we can still maintain reasonable speed of the simulation.
+\tomas{)}
+
 %TODO: Include only a few sentences to CHD in general.

### 2012-03-01 Doc/Miccai/introduction.tex
shell model
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@374 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \tomas{) we don't make any speed comparison. I propose this: }
-By taking into consideration the overall tubular structure of blood vessels during generation of thin shell elements we can still maintain reasonable speed of the simulation.
+By taking into consideration the overall tubular structure of blood vessels during generation of thin shell elements we can still maintain rereal-time speed of the simulation.
 \tomas{)}

### 2012-03-01 Doc/Miccai/shell-model.tex
Added paragraph break...
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@371 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
shell model
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@374 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 Surgeons try to find an optimal approach to correct defects with a minimum of deformations and collisions that would have unnecessary negative impact on blood flow or that may lead to problems during growth of the patient.
-To be able to predict results for different surgical approaches a physical approriate model of blood vessels is mandatory.
+To be able to predict results for different surgical approaches a physicaly approriate model of blood vessels is mandatory.
 Furthermore, computation time of simulations must be low enough to enable interactive and usable planning software.
+
 Options on the choice of model are limited.
-Mass-spring models allow for fast computation but do not correlate to a physical accurate representation of reality and therefore are inappropriate for predictive simulations.
-On the contrary, simulation models based on the formulation of finite elements, either 3-dimensional elements or thin shell elements, are commonly used when high physical accuracy is required but are well known to be computationally expensive.
+Mass-spring models, often used for cloth simulation \cite{Volino2005b}, allow for fast computation but do not do not posses the required level of physical accuracy and therefore are inappropriate for predictive simulations.
+On the contrary, simulation models based on the formulation of finite elements, either 3-dimensional elements or thin shell elements, are commonly used when high physical accuracy is required but are well known to be computationally expensive \tomas{add some refs?}.
 The hybrid blood vessel model of \cite{Li2009} strikes a balance but is too constrained to be used as a truly universal approach for all kinds of surgical procedures used in CHD corrective surgery.
 We propose a model based on thin shell elements that still allows for fast computation with a reasonable number of elements.
+
+\tomas{I'm adding this back, the previous focuess only on physical correctnes.}
+Apart from physical accuracy the model should provide 
+(I) a good support for the deformations by bending and twisting,
+(II) an accurate modeling of the wall motion when the vessel is cut along its axis,
+(III) global deformations of stitched vessels that capture their final configuration, after surgery,
+(IV) collision response when detected (e.g. between the aortic arch and pulmonary artery).
+All these requirements are fullfilled by the shell element model.
 

### 2012-03-01 Doc/Miccai/introduction.tex
Included a few refs and continued writing an results.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@375 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2012-03-01 Doc/Miccai/introduction.tex
@@
 
-To improve planning of such pediatric cardiac surgeries it is possible to create 3-dimensional heart models out of acquired image data REF and to use them as input for surgical simulation systems.
+To improve planning of such pediatric cardiac surgeries it is possible to create 3-dimensional heart models out of acquired image data and to use them as input for surgical simulation systems.
 However, most present simulation systems focus on low-level surgical procedures, i.e.
-performing actual incisions and suturing with special input devices in real-time REF.
+performing actual incisions and suturing with special input devices in real-time \cite{Sorensen2006,Mosegaard2004}.
 Their purpose is either by means of educational training or planning access strategies to reach the malformations of the heart or great arteries.
@@
 To be able to predict the result of complex surgical procedures, a physically correct simulation is mandatory.
-\cite{Li2009} presents a {\itshape predictive} surgical simulation system for bidirectional Glenn shunts and basically similar surgeries.
-The authors use a hybrid model to simulate deformations of blood vessels by binding a surface mesh to a reference Cosserat rod REF elastically.
-While the mesh models surface details of blood vessels, the Cosserat rod models global deformations, i.e.
-stretching, bending, twisting and shearing.
+In \cite{Li2009}, a {\itshape predictive} surgical simulation system for bidirectional Glenn shunts and similar surgeries is presented.
+The authors use a hybrid model to simulate deformations of blood vessels by binding a surface mesh to a reference Cosserat rod elastically.
+%While the mesh models surface details of blood vessels, the Cosserat rod models global deformations, i.e. stretching, bending, twisting and shearing
 The simulation software performs low-level surgical procedures automatically with only a minimum of interaction required which allows the surgeon to focus on the actual overall surgical procedure.
-Although the results are promising, the method is limited to operations on uniform tubular structures that do not arbitrarily change the surface.
-Surgical procedures heavily changing surface details of blood vessels, i.e.
+%Although the results are promising, the method is limited to operations on uniform tubular structures that do not arbitrarily change the surface.
+%Surgical procedures heavily changing surface details of blood vessels, i.e.
 patches for correction of complicated coarctations and stenosis, are not possible with this hybrid model since the physical behavior of blood vessels is simulated by the Cosserat rod and its {\itshape binding} to the surface mesh.
+Although the results are promising, the method is limited to operations on uniform tubular structures and does not allow arbitrary manipulations of the surface mesh.
 
@@
 Our method is not limited to well-formed blood vessels and captures all low-level procedures necessary for aortoplasty.
-By using a relaxation process we can arbitrarily join or attach blood vessels and patches without using springs that would add surplus energy to the physical system.
-We prove that simulation time is still reasonable or even decreased by taking special care of the overall tubular structure of blood vessels during generation of thin shell elements.
+We use a relaxation process to arbitrarily join or attach blood vessels and patches without using springs that would add surplus energy to the physical system.
+By taking the overall tubular structure of blood vessels into consideration during generation of thin shell elements we achieve reasonable speed of the simulation.
+
+%We prove that simulation time is still reasonable or even decreased by taking special care of the overall tubular structure of blood vessels during generation of thin shell elements. 
+%We improve the capabilities of the simulation by using a model based on elastic thin shell elements.
+%This allows us to model arbitrary blood vessels and malformations of the artery.
+%It can also capture all low-level procedures necessary for aortoplasty.
+%By using a relaxation process we can arbitrarily join or attach blood vessels and patches without using springs that would add surplus energy to the physical system.
+%\tomas{I don't think this is correct: (}
+%We prove that simulation time is still reasonable or even decreased by taking special care of the overall tubular structure of blood vessels during generation of thin shell elements.
+%\tomas{) we don't make any speed comparison. I propose this: }
+%By taking into consideration the overall tubular structure of blood vessels during generation of thin shell elements we achieve reasonable speed of the simulation.
+%\tomas{)}
+
 %TODO: Include only a few sentences to CHD in general.

### 2012-03-01 Doc/Miccai/results.tex
Added surgery image.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@364 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
Included a few refs and continued writing an results.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@375 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \item physically accurate,
-\item converges to expected results with a low amount of elements,
+\item converges to expected results with a low number of elements,
 \item is able to simulate low-level surgical procedures universally.
 \end{enumerate}
+We close discussion with limitations and constraints of our simulation approach in terms of generation and dynamic remeshing of thin shell element meshes.
+
 A quantitative validation by means of a comparision to real pre- and post-surgery image data is impossible at present.
@@
 There is no such data available that is obtained shortly after surgeries in infants where the growth of the patient would not interfere comparison.
-However, qualitative comparisons to a real image data set acquired months after a surgical intervention and to an elastic tube manipulated by hand, which confirm a principal suitability of predictive simulations for surgery planning, can be found in \cite{Li2009}.
-
-We close this section with a discussion about the limitations and constraints of our approach which have great impact on the mesh generation prior to simulation.
+However, a qualitative comparison to real image data acquired months after a surgical intervention, which confirm a principal suitability of predictive simulations for surgery planning, can be found in \cite{Li2009}.
 
 \subsection{Accuracy}
+
+\begin{figure}[tbh]
+    \centering
+    \begin{tabular}{cc}
+     \includegraphics[width=0.3\columnwidth]{img/compare-bend.png}
+      &
+      \includegraphics[width=0.3\columnwidth]{img/compare-twist.png}
+      \\
+      \includegraphics[width=0.3\columnwidth]{img/compare-bend-other.png}
+      &
+      \includegraphics[width=0.3\columnwidth]{img/compare-twist-other.png}
+    \end{tabular}
+    \caption{Simulation of bending and twisting of an elestic tube.
+Top row: Thin shell element model; Bottom row: Real elastic tube manipulated by hand and Cosserat rod-based hybrid model\cite{Li2009} (Credits for the bottom images: Li et al.).}
+    \label{fig-deformations}
+\end{figure}
 
@@
 
+\begin{figure}[tbh]
+  \centering
+  \begin{tabular}{cccc}
+    \includegraphics[width=0.24\columnwidth]{img/twist-06-cg.png}
+    &
+    \includegraphics[width=0.24\columnwidth]{img/twist-08-cg.png}
+    &
+    \includegraphics[width=0.24\columnwidth]{img/twist-16-cg.png}
+    &
+    \includegraphics[width=0.24\columnwidth]{img/twist-31-cg.png}
+    \\
+    \includegraphics[width=0.24\columnwidth]{img/twist-06w-cg.png}
+    &
+    \includegraphics[width=0.24\columnwidth]{img/twist-08w-cg.png}
+    &
+    \includegraphics[width=0.24\columnwidth]{img/twist-16w-cg.png}
+    &
+    \includegraphics[width=0.24\columnwidth]{img/twist-31w-cg.png}
+  \end{tabular}
+  \caption{Convergence of a twisted, elastic tube with 6, 8, 16 and 31 vertices
+  along its circumference (120, 208, 832 and 3038 thin shell elements). Top row: High polygon count mesh mapped to simulation results of coarser shell element meshes (bottom row).}
+  \label{fig-convergence}
+\end{figure}
+
 It is important to minimize the number of shell elements to reduce computation time.
 This is the key factor to enable an interactive simulation system that is accepted by surgeons to simulate multiple surgical approaches in succession.
-The bending of thin shell elements allows to reduce their number by still representing the tubular structure of blood vessels while maintaining their physical behavior appropriately.
+The bending capabilities of thin shell elements allow to reduce their number by still representing the tubular structure of blood vessels while maintaining their physical behavior appropriately.
 Figure \ref{fig-convergence} shows simulation results of a strongly deformed tube represented by different numbers of thin shell elements.
-With only eight vertices along the circumference of the tube the simulation result is comparable to a simulation result with high element count.
-We map a fine polygonal mesh to the coarse shell element meshes to visualize their bending detail and therefore get comparable results that correlates with the true shape of the simulated tube.
+With only eight vertices along the circumference of the elastic tube the simulation result is comparable to a simulation result with high element count.
+We map a fine polygonal mesh to the coarse shell element meshes to visualize their intrinsic bending detail and therefore to get comparable results that correlates with the true shape of the simulated tube.
 
@@
 
-TODO: Use the remaining space we have in the paper. We should consider to shorten other sections.
+\begin{figure}[tbh]
+  \centering
+  \includegraphics[width=\columnwidth]{img/surgery.png}
+  \label{fig-surgery}
+  \caption{Simulation of different surgical procedures for coarctation repair of an aortic arch.
+Left image: Overview of the simulation scene consisting of an aortic arch (AO), a left subclavian artery (LSA) and a pulmonary artery (PA).
+The coarctation can be seen next to the LSA and the dashed line indicates where the aorta is to be incised for a subclavian flap aortoplasty; The vertical image pairs show the scene before and after different surgical interventions.
+From left to right: Subclavian flap aortoplasty -- the subclavian flap is used as an organic onlay patch over the area of coarctation; Patch aortoplasty -- the incision is opened by attached springs only for visualisation purposes.
+The patch is sutured in place as shown; End-to-end anastomosis -- the coarctation is resected and the loose ends of th AO sutured together.
+Due to deformation of the AO it collides with the PA, causing a slight deformation there as well.
+}
+\end{figure}
+
+We manually modeled an aortic arch based on real image data to be able to demonstrate suitability of our joining approach for different kinds of low-level surgical procedures on a concrete example.
+Therefore we chose a coarctation of an aortic arch that can be repaired by completely different surgical approaches that heavily alter the surface of the blood vessel \cite{Dodge2000}.
+Results can be examined in figure \ref{fig-surgery}.
+We prove that our topological method for joining of shell elements through a controlled relaxation to their rest shape is able to handle all different kinds of connections universally and in a correct manner.
 

### 2012-03-01 Doc/Miccai/conclusion.tex
Updated style.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@376 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 Special care has to be taken during initial mesh generation and dynamic remeshing when cutting blood vessels since the size of newly generated thin shell elements should roughly equal the size of their neighbor elements which is a common constraint for finite element simulations.
-To truly benefit from our simulation approach, future work has to be done in terms of remeshing, to keep a high quality of the mesh during the whole simulation.
+To truly benefit from our simulation approach, future work has to be done in terms of remeshing, to keep a high quality of the mesh during the simulation.

### 2012-03-01 Doc/Miccai/introduction.tex
Improved intro.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@377 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-In this paper we present an improved approach using a simulation model solely based on elastic thin shell elements.
+In this paper we present an improved approach for predictive simulation of congentital heart defect (CHD) corrective surgeries using a simulation model solely based on elastic thin shell elements.
 Our method is not limited to well-formed blood vessels and captures all low-level procedures necessary for aortoplasty.

### 2012-03-01 Doc/Miccai/shell-model.tex
Improved 2. section.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@378 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-In congenital heart defect (CHD) corrective surgeries blood vessels respond to plastic interventions with a certain amount of deformation.
+In CHD corrective surgeries blood vessels respond to plastic interventions with a certain amount of deformation.
 Surgeons try to find an optimal approach to correct defects with a minimum of deformations and collisions that would have unnecessary negative impact on blood flow or that may lead to problems during growth of the patient.
@@
 We propose a model based on thin shell elements that still allows for fast computation with a reasonable number of elements.
-
-\tomas{I'm adding this back, the previous focuess only on physical correctnes.}
-Apart from physical accuracy the model should provide 
-(I) a good support for the deformations by bending and twisting,
-(II) an accurate modeling of the wall motion when the vessel is cut along its axis,
-(III) global deformations of stitched vessels that capture their final configuration, after surgery,
-(IV) collision response when detected (e.g. between the aortic arch and pulmonary artery).
-All these requirements are fullfilled by the shell element model.
 

### 2012-03-02 Doc/Miccai/results.tex
notes on results
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@379 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{enumerate}
-\item physically accurate,
+\item physically accurate, \tomas{I suggest to use \uv{fitness for the purpose}}
 \item converges to expected results with a low number of elements,
@@
 This is the key factor to enable an interactive simulation system that is accepted by surgeons to simulate multiple surgical approaches in succession.
-The bending capabilities of thin shell elements allow to reduce their number by still representing the tubular structure of blood vessels while maintaining their physical behavior appropriately.
+The bending capabilities of thin shell elements allow to reduce their number and still representing the tubular structure of blood vessels while maintaining their physical behavior appropriately.
 Figure \ref{fig-convergence} shows simulation results of a strongly deformed tube represented by different numbers of thin shell elements.
 With only eight vertices along the circumference of the elastic tube the simulation result is comparable to a simulation result with high element count.
-We map a fine polygonal mesh to the coarse shell element meshes to visualize their intrinsic bending detail and therefore to get comparable results that correlates with the true shape of the simulated tube.
+We map a fine polygonal mesh to the coarse shell element meshes to visualize their intrinsic bending detail.
+\tomas{redundant, already in previous sentences: and therefore to get comparable results that correlates with the true shape of the simulated tube.}
 
@@
 \subsection{Limitations and Constraints}
+\tomas{rename do \uv{Discussion}}
+
+More general problems for this kind of simulation: (not limited to our method)
+
+-- getting image data -- that is something you're already working on in Germany
+
+-- hi-quality mesh with low number of polygons -- there seem to be something already, e.g. \cite{Comas2010c}.
+
+At the moment biggest limitation of our method resides in the requirement of the same number of nodes on both edges of the suture.
+Meeting this requirement may not be easy.
+Also FE methods put constrain on the quality of mesh that is necessary to maintain the convergence and stability of the model.
+These are especially important for meshes with low number of elements.
+To deal with this problems techniques for dynamic remeshig during the simulation may be required.
 

### 2012-03-02 Doc/Miccai/simulation.tex
Rewrote section 3.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@380 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2012-02-29 Doc/Miccai/simulation.tex
@@
-The primary goal of the surgery is to restore a normal heart function and circulation.
-This goal is reached by rebuilding the heart in a more or less anatomical way.
-There are several types of low-level task the surgeon usually employs while
-performing the surgery that are:
+In order to repair malformations of the great arteries surgeons perform several low-level procedures on the blood vessels.
+These include arbitrary incisions, resections of whole blood vessel parts, attaching patches and reconnection of blood vessels end-to-end or end-to-side.
+From a technical view, all cutting operations can be done prior to simulation by splitting shell elements along their edges which correlates to the actual incision.
+New shell elements can be added to a mesh or rearranged to align their edges along an incision.
+On the contrary, joining and suturing of blood vessels and patches typically deform blood vessel walls to a certain amount which has to be simulated.
+Recent simulation systems \cite{} realize suturing through attachment of contracting springs that have no physical correlation.
+Springs introduce surplus energy to a physical system and special care has to be taken during simulation to compensate for that.
 
-\begin{description}
+We present a new approach that overcomes these difficulties by connecting edges of shell elements along two opposite edges topologically (see figure \ref{fig-}.
+Since there is a distance between two joining edges in general, a topologically connection alters the shape and size of affected shell elements.
+Note that in this simulation state there is temporarily no physical correlation.
+We then start a controlled relaxation process in which the shell elements progressivly recover to their rest shape so that there is no strong increase in internal forces and the simulation remains stable.
+This is achieved by linear interpolation between the strongly deformed shell elements and their original rest shape.
+The steady state of the simulation corresponds to an energetical equilibrium of the shell element deformation and the simulation is finished.
+The effect is similar to contracting springs but there is no such problem like surplus energy after the joining process is done.
 
-  \item[End to end aortoplasty:] the surgeon completely cuts away the
-    affected part of aorta and sutures the two ends of aorta back together.
+% The primary goal of the surgery is to restore a normal heart function and circulation.
+% This goal is reached by rebuilding the heart in a more or less anatomical way.
+% There are several types of low-level task the surgeon usually employs while
+% performing the surgery that are:
 
-  \item[Patch aortoplasty:] the surgeon performs a cut along the axis of
-    the aorta thus allowing the narrow place to expand. Then applies a
-    patch from artificial material to fill the hole in.
+% \begin{description}
 
-  \item[Subclavian flap aortoplasty:] the nearest artery is cut,
-    removing also half of the artery (along the axis) and part of the
-    aorta in the affected place. He then uses the rest of the artery to
-    patch the aorta.
+  % \item[End to end aortoplasty:] the surgeon completely cuts away the
+    % affected part of aorta and sutures the two ends of aorta back together.
 
-  \item[Reconnection and joining of blood vessels:] In addition to the modification of the aorta, the surgeon is lead to establish certain connections that are absent in pathological anatomy.
+  % \item[Patch aortoplasty:] the surgeon performs a cut along the axis of
+    % the aorta thus allowing the narrow place to expand. Then applies a
+    % patch from artificial material to fill the hole in.
 
-\end{description}
+  % \item[Subclavian flap aortoplasty:] the nearest artery is cut,
+    % removing also half of the artery (along the axis) and part of the
+    % aorta in the affected place. He then uses the rest of the artery to
+    % patch the aorta.
 
-With our method we are able to simulate these tasks, in order to be able to plan them.
-We have based our general approach on a common characteristic: In all these operations the surgeon sutures two or more edges of the vessel walls together.
-Since we are interested in predicting the results rather than creating a training simulator for suturing (trained surgeon knows well how to perform the sutures) the simplest way of performing the suture would be to connect the respective edges with springs.
-Springs however adds some unwanted energy to the system. 
-Our method is based on a topological connection of the shell elements that are neighbors of the joining edges and a controlled relaxation of these shell elements' rest shape.
+  % \item[Reconnection and joining of blood vessels:] In addition to the modification of the aorta, the surgeon is lead to establish certain connections that are absent in pathological anatomy.
 
-We start with a mesh representing the situation after all the cuts have been performed but before suturing takes place. 
-In the first step we perform the required topological changes to the mesh.
+% \end{description}
+
+% With our method we are able to simulate these tasks, in order to be able to plan them.
+% We have based our general approach on a common characteristic: In all these operations the surgeon sutures two or more edges of the vessel walls together.
+% Since we are interested in predicting the results rather than creating a training simulator for suturing (trained surgeon knows well how to perform the sutures) the simplest way of performing the suture would be to connect the respective edges with springs.
+% Springs however adds some unwanted energy to the system. 
+% Our method is based on a topological connection of the shell elements that are neighbors of the joining edges and a controlled relaxation of these shell elements' rest shape.
+
+% We start with a mesh representing the situation after all the cuts have been performed but before suturing takes place. 
+% In the first step we perform the required topological changes to the mesh.
 We keep the number of elements in new mesh the same as in the original mesh so that we can define 1:1 mapping between elements of these meshes.
-The Figure \ref{fig-JoiningVessels} (a) and (b) explains this operation when joining vessels. 
-Then we start the simulation on the modified mesh while keeping the information of the original rest shapes of all shell elements.
+% The Figure \ref{fig-JoiningVessels} (a) and (b) explains this operation when joining vessels. 
+% Then we start the simulation on the modified mesh while keeping the information of the original rest shapes of all shell elements.
 
-\begin{figure}[tbh]
-\begin{center}
-\includegraphics[width=\columnwidth]{img/rest_shape_scheme.png}
-\end{center}
+% \begin{figure}[tbh]
+% \begin{center}
+% \includegraphics[width=\columnwidth]{img/rest_shape_scheme.png}
+% \end{center}
 
-\caption{Scheme showing the method for joining two vessels:  (a) the cut on vessel meshes is performed in a 3D geometry edition software.
+% \caption{Scheme showing the method for joining two vessels \stefan{Isn't that example too complex?
+I think it is really hard to actually see what is happening here.
+At least the colors have to be changed because the different layers are colored quite equally and it should be zoomed in to the interesting area were the joining takes place.
+Remember also that the procedings are printed in black and white and that everything must be visible when printing out this figure black and white.
+There must be way more contrast.}:  (a) the cut on vessel meshes is performed in a 3D geometry edition software.
 In the software, we impose that the number of the segments on the edges is the same on both side  (\color{BlueGreen}{$\mathbf{\rightarrow}$}\color{black})
-Moreover, we select some shell elements that lie along the sutured edges (\color{purple}{$\mathbf{\rightarrow}$}\color{black}).
-(b) In our software, we connect the vessel topologies by performing a relaxation of the shape and the rest shape for the selected shells.  
-(c) We progressively re-contract the rest shape of the shells while computing the deformation of the whole unified mesh. 
-In the final steady state computed in the simulation, the shell elements have all their correct rest shape.}
-\label{fig-JoiningVessels}
-\end{figure}
+% Moreover, we select some shell elements that lie along the sutured edges (\color{purple}{$\mathbf{\rightarrow}$}\color{black}).
+% (b) In our software, we connect the vessel topologies by performing a relaxation of the shape and the rest shape for the selected shells.  
+% (c) We progressively re-contract the rest shape of the shells while computing the deformation of the whole unified mesh. 
+% In the final steady state computed in the simulation, the shell elements have all their correct rest shape.}
+% \label{fig-JoiningVessels}
+% \end{figure}
 
-We can not always step back all the shells to their original shape (transition from figure \ref{fig-JoiningVessels} (b) to (c))  in a single simulation step. 
-In case of non-small distance between the nodes that are sutured together, the involved elements will undergo large deformation that could create quick and non-smooth increase of internal forces and  cause unwanted effects.
-To allevieate the problem we have implemented the technique that allows for moderate changes in the rest shape between simulation steps.
-Between the start and the end of the simulation we use a linear interpolation of the rest shape for the shells that lie along the sutured edges. 
-The interpolation starts with the rest shape equal to the deformed elements and we progressively lead the elements to recover their original rest shape.
-This way, the simulation gradually forces the side edges to connect without any abrupt jump in the behavior.
-When all elements have recover their original rest shape, the steady state found in the simulation corresponds to a physics-based energetical equilibrium of the shells deformation. 
-Thus, we take these positions as an estimation of the result of the connection of the vessels.
+% We can not always step back all the shells to their original shape (transition from figure \ref{fig-JoiningVessels} (b) to (c))  in a single simulation step. 
+% In case of non-small distance between the nodes that are sutured together, the involved elements will undergo large deformation that could create quick and non-smooth increase of internal forces and  cause unwanted effects.
+% To allevieate the problem we have implemented the technique that allows for moderate changes in the rest shape between simulation steps.
+% Between the start and the end of the simulation we use a linear interpolation of the rest shape for the shells that lie along the sutured edges. 
+% The interpolation starts with the rest shape equal to the deformed elements and we progressively lead the elements to recover their original rest shape.
+% This way, the simulation gradually forces the side edges to connect without any abrupt jump in the behavior.
+% When all elements have recover their original rest shape, the steady state found in the simulation corresponds to a physics-based energetical equilibrium of the shells deformation. 
+% Thus, we take these positions as an estimation of the result of the connection of the vessels.

### 2012-03-02 Doc/Miccai/simulation.tex
Improved simulation section.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@382 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-In order to repair malformations of the great arteries surgeons perform several low-level procedures on the blood vessels.
+In order to repair malformations of the great arteries surgeons perform several low-level procedures on blood vessels.
 These include arbitrary incisions, resections of whole blood vessel parts, attaching patches and reconnection of blood vessels end-to-end or end-to-side.
@@
 On the contrary, joining and suturing of blood vessels and patches typically deform blood vessel walls to a certain amount which has to be simulated.
-Recent simulation systems \cite{} realize suturing through attachment of contracting springs that have no physical correlation.
+Recent simulation systems \cite{Sorensen2006,Mosegaard2004,Li2009} realize suturing through attachment of contracting springs that have no physical correlation.
 Springs introduce surplus energy to a physical system and special care has to be taken during simulation to compensate for that.
 
-We present a new approach that overcomes these difficulties by connecting edges of shell elements along two opposite edges topologically (see figure \ref{fig-}.
-Since there is a distance between two joining edges in general, a topologically connection alters the shape and size of affected shell elements.
-Note that in this simulation state there is temporarily no physical correlation.
-We then start a controlled relaxation process in which the shell elements progressivly recover to their rest shape so that there is no strong increase in internal forces and the simulation remains stable.
-This is achieved by linear interpolation between the strongly deformed shell elements and their original rest shape.
-The steady state of the simulation corresponds to an energetical equilibrium of the shell element deformation and the simulation is finished.
+We present a new approach that overcomes these difficulties by performing topological changes on the mesh to connect shell elements along joining edges (see figure \ref{fig-}.
+Since there is a certain distance between two joining edges in general, a topological modification alters the shape and size of affected shell elements.
+Note that in this simulation state there is temporarily no correlation to reality.
+We then start a controlled relaxation process during which the shell elements progressivly converge to their energetical equilibrium.
+To keep the simulation stable we avoid a strong increase in internal forces by linearly interpolate the rest shape of the connected mesh to the initial rest shape before joining.
 The effect is similar to contracting springs but there is no such problem like surplus energy after the joining process is done.

### 2012-03-02 Doc/Miccai/shell-model.tex
shell element
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@383 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 Mass-spring models, often used for cloth simulation \cite{Volino2005b}, allow for fast computation but do not do not posses the required level of physical accuracy and therefore are inappropriate for predictive simulations.
-On the contrary, simulation models based on the formulation of finite elements, either 3-dimensional elements or thin shell elements, are commonly used when high physical accuracy is required but are well known to be computationally expensive \tomas{add some refs?}.
+On the contrary, simulation models based on the formulation of finite elements, either 3-dimensional elements or thin shell elements, are commonly used when high physical accuracy is required but are well known to be computationally expensive.
 The hybrid blood vessel model of \cite{Li2009} strikes a balance but is too constrained to be used as a truly universal approach for all kinds of surgical procedures used in CHD corrective surgery.
 We propose a model based on thin shell elements that still allows for fast computation with a reasonable number of elements.
+
+\begin{figure}[tbh]
+  \centering
+  \includegraphics[width=0.4\columnwidth]{img/shell.pdf}
+  \caption{Shell element combines bending with in-plane deformations}
+  \label{fig-shell}
+\end{figure}
 

### 2012-03-02 Doc/Miccai/shell-model.tex
small fixes
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@384 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 Options on the choice of model are limited.
-Mass-spring models, often used for cloth simulation \cite{Volino2005b}, allow for fast computation but do not do not posses the required level of physical accuracy and therefore are inappropriate for predictive simulations.
+Mass-spring models, often used for cloth simulation \cite{Volino2005b}, allow for fast computation but do not posses the required level of physical accuracy and therefore are inappropriate for predictive simulations.
 On the contrary, simulation models based on the formulation of finite elements, either 3-dimensional elements or thin shell elements, are commonly used when high physical accuracy is required but are well known to be computationally expensive.
@@
   \includegraphics[width=0.4\columnwidth]{img/shell.pdf}
-  \caption{Shell element combines bending with in-plane deformations}
+  \caption{Shell element combines bending with in-plane deformations.}
   \label{fig-shell}

### 2012-03-02 Doc/Miccai/conclusion.tex
Improved results and conclusion.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@385 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 We introduced a new approach for predictive simulation of CHD corrective surgeries.
-It is based on thin shell elements and a relaxation method for joining of blood vessels and patches by topologically connecting the vertices of joining edges of the simulation meshes.
-All necessary low-level surgical procedures can be performed using this method.
-We showed that we need only a small number of thin shell elemenets to achieve physically accurat simulation results which enable fast simulations.
-Special care has to be taken during initial mesh generation and dynamic remeshing when cutting blood vessels since the size of newly generated thin shell elements should roughly equal the size of their neighbor elements which is a common constraint for finite element simulations.
-To truly benefit from our simulation approach, future work has to be done in terms of remeshing, to keep a high quality of the mesh during the simulation.
+It is based on thin shell elements and a relaxation method for joining blood vessels and patches by topologically connecting affected shell elements.
+Low-level surgical procedures can be performed universally using this method.
+We showed that we only need a small number of thin shell elements to achieve predictive simulation results.
+A sophisticated remeshing algorithm is needed to allow for arbitrary incisions.

### 2012-03-02 Doc/Miccai/results.tex
fixing the quotes
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@381 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
Improved results and conclusion.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@385 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2012-03-01 Doc/Miccai/results.tex
@@
-In order to show an improvement on existing predictive simulation systems for CHD corrective surgeries, we want to demonstrate that our blood vessel model based on thin shell elements is
+In order to show an improvement on existing predictive simulation systems for CHD corrective surgeries, we want to demonstrate that our blood vessel model based on thin shell elements
 \begin{enumerate}
-\item physically accurate,
-\item converges to expected results with a low number of elements,
-\item is able to simulate low-level surgical procedures universally.
+\item physically meets the requirements for predictive simulation results,
+\item converges to these results with a low number of elements,
+\item is able to universally simulate low-level surgical procedures.
 \end{enumerate}
-We close discussion with limitations and constraints of our simulation approach in terms of generation and dynamic remeshing of thin shell element meshes.
+We close this section with a discussion about limitations and constraints of our simulation approach in terms of generation and dynamic remeshing of thin shell element meshes.
 
@@
 There is no such data available that is obtained shortly after surgeries in infants where the growth of the patient would not interfere comparison.
-However, a qualitative comparison to real image data acquired months after a surgical intervention, which confirm a principal suitability of predictive simulations for surgery planning, can be found in \cite{Li2009}.
+However, a qualitative comparison to real image data acquired months after a surgical intervention, which confirms a principal suitability of predictive simulations for surgery planning, can be found in \cite{Li2009}.
 
-\subsection{Accuracy}
+\subsection{Predictive Simulation Results}
 
@@
     \caption{Simulation of bending and twisting of an elestic tube.
-Top row: Thin shell element model; Bottom row: Real elastic tube manipulated by hand and Cosserat rod-based hybrid model\cite{Li2009} (Credits for the bottom images: Li et al.).}
+Top row: Thin shell element model; Bottom row: Real elastic tube manipulated by hand and Cosserat rod-based hybrid model \cite{Li2009} (Credits for the bottom images: Li et al.).}
     \label{fig-deformations}
@@
 This is the key factor to enable an interactive simulation system that is accepted by surgeons to simulate multiple surgical approaches in succession.
-The bending capabilities of thin shell elements allow to reduce their number by still representing the tubular structure of blood vessels while maintaining their physical behavior appropriately.
+The bending capabilities of thin shell elements allow to reduce their number and still representing the tubular structure of blood vessels while maintaining their physical behavior appropriately.
 Figure \ref{fig-convergence} shows simulation results of a strongly deformed tube represented by different numbers of thin shell elements.
 With only eight vertices along the circumference of the elastic tube the simulation result is comparable to a simulation result with high element count.
-We map a fine polygonal mesh to the coarse shell element meshes to visualize their intrinsic bending detail and therefore to get comparable results that correlates with the true shape of the simulated tube.
 
-\subsection{Low-level surgical procedures}
+\subsection{Low-Level Surgical Procedures}
 
@@
   \includegraphics[width=\columnwidth]{img/surgery.png}
-  \label{fig-surgery}
   \caption{Simulation of different surgical procedures for coarctation repair of an aortic arch.
@@
 From left to right: Subclavian flap aortoplasty -- the subclavian flap is used as an organic onlay patch over the area of coarctation; Patch aortoplasty -- the incision is opened by attached springs only for visualisation purposes.
-The patch is sutured in place as shown; End-to-end anastomosis -- the coarctation is resected and the loose ends of th AO sutured together.
-Due to deformation of the AO it collides with the PA, causing a slight deformation there as well.
+The patch is sutured in place as shown; End-to-end anastomosis -- the coarctation is resected and the loose ends of the AO sutured together.
+Due to deformation of the AO there is a collision with the PA, causing slight deformations on both blood vessels.
 }
+  \label{fig-surgery}
 \end{figure}
@@
 Results can be examined in figure \ref{fig-surgery}.
-We prove that our topological method for joining of shell elements through a controlled relaxation to their rest shape is able to handle all different kinds of connections universally and in a correct manner.
+We show that our topological method for joining of shell elements through a controlled relaxation to their rest shape is able to handle different kinds of connections universally and in a correct manner.
 
 \subsection{Limitations and Constraints}
+
+Finite element methods including thin shell elements have some constraints in common regarding the simulation mesh quality.
+A homogeneous neighborhood of equally sized elements is desirable.
+Especially when minimizing the number of elements, high mesh quality is necessary to maintain convergence and stability of the simulation.
+We showed that an elastic tubular structure can be simulated appropriately for CHD corrective surgeries with only a few elements.
+To allow for arbitrary incisions under such circumstances, the mesh must be remeshed in a way that its elements align along an incision.
+Mesh quality must be retained during remeshing and therfore the whole mesh may be affected.
+A limitation of our joining approach is that we need the same number of elements along connecting edges.
+Appropriate remeshing strategies must be developed to take full benefit of our proposed simulation approach.
+However, as a fallback, element count could be increased which will have negative impact on the simulation speed.
 

### 2012-03-02 Doc/Miccai/shell-model.tex
Improved shell-model section.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@386 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
Improved simulation chapter. Added ref to fig 1.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@387 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 The hybrid blood vessel model of \cite{Li2009} strikes a balance but is too constrained to be used as a truly universal approach for all kinds of surgical procedures used in CHD corrective surgery.
-We propose a model based on thin shell elements that still allows for fast computation with a reasonable number of elements.
+We propose to use a model based on thin shell elements.
+Such a model still allows for fast computation when using a reasonable number of elements during simulation.
 
@@
   \includegraphics[width=0.4\columnwidth]{img/shell.pdf}
-  \caption{Shell element combines bending with in-plane deformations.}
+  \caption{TODO: Thin shell elements model bended plates.}
   \label{fig-shell}
@@
 They combine forces from in-plane deformations, i.e.
-stretching or shearing, with bending forces.
+stretching or shearing, with bending forces (see figure \ref{fig-shell}).
 Such internal forces $\myvec{f}$ are a function of current displacement measured as difference of the current shape $\myvec{x}$ to the rest shape $\myvec{x_0}$.

### 2012-03-02 Doc/Miccai/simulation.tex
Improved simulation chapter. Added ref to fig 1.
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@387 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-We present a new approach that overcomes these difficulties by performing topological changes on the mesh to connect shell elements along joining edges (see figure \ref{fig-}.
+\begin{figure}[tbh]
+\begin{center}
+\includegraphics[width=\columnwidth]{img/new-joining.png}
+\end{center}
+\caption{Joining of thin shell element meshes. Left: Split meshes. Center: Shell elements along the joining edges are connected. Right: Final result after the simulation.}
+\label{fig-JoiningVessels}
+\end{figure}
+
+We present a new approach that overcomes these difficulties by performing topological changes on the mesh to connect shell elements along joining edges (see figure \ref{fig-JoiningVessels}).
 Since there is a certain distance between two joining edges in general, a topological modification alters the shape and size of affected shell elements.

### 2012-03-02 Doc/Miccai/introduction.tex
spell check
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@388 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2012-03-01 Doc/Miccai/introduction.tex
@@
 
-In this paper we present an improved approach using a simulation model solely based on elastic thin shell elements.
+In this paper we present an improved approach for predictive simulation of congenital heart defect (CHD) corrective surgeries using a simulation model solely based on elastic thin shell elements.
 Our method is not limited to well-formed blood vessels and captures all low-level procedures necessary for aortoplasty.

### 2012-03-02 Doc/Miccai/results.tex
spell check
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@388 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-A quantitative validation by means of a comparision to real pre- and post-surgery image data is impossible at present.
+A quantitative validation by means of a comparison to real pre- and post-surgery image data is impossible at present.
 This is mainly due to ethical issues that arise during image data acquirement of infants, i.e.
@@
     \end{tabular}
-    \caption{Simulation of bending and twisting of an elestic tube.
+    \caption{Simulation of bending and twisting of an elastic tube.
 Top row: Thin shell element model; Bottom row: Real elastic tube manipulated by hand and Cosserat rod-based hybrid model \cite{Li2009} (Credits for the bottom images: Li et al.).}
@@
 We manually modeled an aortic arch based on real image data to be able to demonstrate suitability of our joining approach for different kinds of low-level surgical procedures on a concrete example.
-Therefore we chose a coarctation of an aortic arch that can be repaired by completely different surgical approaches that heavily alter the surface of the blood vessel \cite{Dodge2000}.
+Therefore we chose a coarctation of an aortic arch that can be repaired by several different surgical approaches that heavily alter the surface of the blood vessel \cite{Dodge2000}.
 Results can be examined in figure \ref{fig-surgery}.
@@
 To allow for arbitrary incisions under such circumstances, the mesh must be remeshed in a way that its elements align along an incision.
-Mesh quality must be retained during remeshing and therfore the whole mesh may be affected.
+Mesh quality must be retained during remeshing and therefore the whole mesh may be affected.
 A limitation of our joining approach is that we need the same number of elements along connecting edges.
 Appropriate remeshing strategies must be developed to take full benefit of our proposed simulation approach.
-However, as a fallback, element count could be increased which will have negative impact on the simulation speed.
+However, as a fall-back, element count could be increased which will have negative impact on the simulation speed.
 

### 2012-03-02 Doc/Miccai/shell-model.tex
spell check
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@388 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 Surgeons try to find an optimal approach to correct defects with a minimum of deformations and collisions that would have unnecessary negative impact on blood flow or that may lead to problems during growth of the patient.
-To be able to predict results for different surgical approaches a physicaly approriate model of blood vessels is mandatory.
+To be able to predict results for different surgical approaches a physically appropriate model of blood vessels is mandatory.
 Furthermore, computation time of simulations must be low enough to enable interactive and usable planning software.
@@
 
-Formulation of thin shell elements is based on the machanics of continuum.
+Formulation of thin shell elements is based on the mechanics of continuum.
 For a thorough description we refer the reader to a standard textbook about finite element methods (FEM), i.e.
@@
 \end{equation}
-where $\mymat{K_e}$ is a stiffness matrix that is computed by Kirshohoff-Loves equations.
+where $\mymat{K_e}$ is a stiffness matrix that is computed by Kirchhoff-Love's equations.
 This approach is similar to the model presented in  \cite{Comas2010b,Comas2010c}, although the interpolation functions we use are different to allow for better convergence\footnote{Our thin shell element model will be published separately}.

### 2012-03-02 Doc/Miccai/introduction.tex
small modif in the intro
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@389 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2012-03-01 Doc/Miccai/introduction.tex
@@
 
-In this paper we present an improved approach using a simulation model solely based on elastic thin shell elements.
+In this paper we present an improved approach for predictive simulation of congenital heart defect (CHD) corrective surgeries using a physics based model of the vessel walls: their mechanical behavior is simulated using elastic thin shell elements.
 Our method is not limited to well-formed blood vessels and captures all low-level procedures necessary for aortoplasty.

### 2012-03-02 Doc/Miccai/abstract.tex
abstract v1
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@390 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

Pediatric cardiac surgery is particularly challenging because the final decisions on the procedure choice take place at the last minute.
The work presented in this paper aims at proposing an interactive software dedicated to preoperative planning.
This new tool would permit to evaluate several possible surgery on a virtual anatomy of the patient and anticipate the decisions.
As the surgery involves a reconfiguration of the blood vessels that are close to the heart.
During the surgery, the vessel walls are deformed, cut, and reconnected to other vessels.
We propose to anticipate the geometry changes of the vessel walls using a simulation based on thin-shell elements.
We demonstrate that our approach allows for simulating the results of low-level surgical procedures, with a unified strategy for joining the vessel models and simulate their subsequent behavior.

### 2012-03-02 Doc/Miccai/abstract.tex
changes to abstract
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@391 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 The work presented in this paper aims at proposing an interactive software dedicated to preoperative planning.
-This new tool would permit to evaluate several possible surgery on a virtual anatomy of the patient and anticipate the decisions.
-As the surgery involves a reconfiguration of the blood vessels that are close to the heart.
-During the surgery, the vessel walls are deformed, cut, and reconnected to other vessels.
+This new tool would permit to evaluate several possible surgical procedures on a virtual anatomy of the patient and anticipate the decisions.
+The surgery involves a reconfiguration of the blood vessels that are close to the heart.
+During the surgery, the vessel walls are deformed, cut, and reconnected to other vessels or closed with an artificial patch.
 We propose to anticipate the geometry changes of the vessel walls using a simulation based on thin-shell elements.

### 2012-03-02 Doc/Miccai/simulation.tex
spell check
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@388 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
modifs in simulation.tex (rest shape control)
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@394 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 Since there is a certain distance between two joining edges in general, a topological modification alters the shape and size of affected shell elements.
+Both actual shape $\myvec{x}$ and rest shape $\myvec{x_0}$ of these shells are modified.
 Note that in this simulation state there is temporarily no correlation to reality. 
-We then start a controlled relaxation process during which the shell elements progressivly converge to their energetical equilibrium.
+We then start a controlled relaxation process during which the shell elements progressively converge to their energetical equilibrium
+(i.e. we progressively re-contract the rest shape of the shells to their initial value while the simulation compute the positions of the whole mesh).
 To keep the simulation stable we avoid a strong increase in internal forces by linearly interpolate the rest shape of the connected mesh to the initial rest shape before joining. 

### 2012-03-02 Doc/Miccai/simulation.tex
fixing typos
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@396 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 We then start a controlled relaxation process during which the shell elements progressively converge to their energetical equilibrium
-(i.e. we progressively re-contract the rest shape of the shells to their initial value while the simulation compute the positions of the whole mesh).
-To keep the simulation stable we avoid a strong increase in internal forces by linearly interpolate the rest shape of the connected mesh to the initial rest shape before joining. 
+(i.e. we progressively re-contract the rest shape of the shells to their initial value while the simulation computes the deformation of the whole mesh).
+To keep the simulation stable we avoid a strong increase in internal forces by linearly interpolating the rest shape of the connected mesh to the initial rest shape before joining. 
 The effect is similar to contracting springs but there is no such problem like surplus energy after the joining process is done.

### 2012-05-22 Doc/SCA2012/sca2012.tex
paper for SCA 2012
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@407 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\documentclass{egpubl}
\usepackage{sca2012}

% --- for  Annual CONFERENCE
% \ConferenceSubmission % uncomment for Conference submission
% \ConferencePaper      % uncomment for (final) Conference Paper
% \STAR                 % uncomment for STAR contribution
% \Tutorial             % uncomment for Tutorial contribution
% \ShortPresentation    % uncomment for (final) Short Conference Presentation
%
% --- for  CGF Journal
% \JournalSubmission    % uncomment for submission to Computer Graphics Forum
% \JournalPaper         % uncomment for final version of Journal Paper
%
% --- for  EG Workshop Proceedings
\WsSubmission    % uncomment for submission to EG Workshop
% \WsPaper         % uncomment for final version of EG Workshop contribution
%
 \electronicVersion % can be used both for the printed and electronic version

% !! *please* don't change anything above
% !! unless you REALLY know what you are doing
% ------------------------------------------------------------------------

% for including postscript figures
% mind: package option 'draft' will replace PS figure by a filname within a frame
\ifpdf \usepackage[pdftex]{graphicx} \pdfcompresslevel=9
\else \usepackage[dvips]{graphicx} \fi

\PrintedOrElectronic

% prepare for electronic version of your document
\usepackage{t1enc,dfadobe}

\usepackage{egweblnk}
\usepackage{cite}

% For backwards compatibility to old LaTeX type font selection.
% Uncomment if your document adheres to LaTeX2e recommendations.
% \let\rm=\rmfamily    \let\sf=\sffamily    \let\tt=\ttfamily
% \let\it=\itshape     \let\sl=\slshape     \let\sc=\scshape
% \let\bf=\bfseries

% end of prologue

% ------------------------------------------------------------------------

\title[EG \LaTeX\ Author Guidelines]%
      {Fill in the Title} % TODO

% for anonymous conference submission please enter your SUBMISSION ID
% instead of the author's name (and leave the affiliation blank) !!
\author[...]
       {...} % TODO

% ------------------------------------------------------------------------

% if the Editors-in-Chief have given you the data, you may uncomment
% the following five lines and insert it here
%
% \volume{27}   % the volume in which the issue will be published;
% \issue{1}     % the issue number of the publication
% \pStartPage{1}      % set starting page

%-------------------------------------------------------------------------
\begin{document}

% \teaser{
%  \includegraphics[width=\linewidth]{eg_new}
%  \centering
%   \caption{New EG Logo}
% \label{fig:teaser}
% }

\maketitle

\begin{abstract} % TODO
   The ABSTRACT is to be in fully-justified italicized text, 
   between two horizontal lines,
   in one-column format, 
   below the author and affiliation information. 
   Use the word ``Abstract'' as the title, in 9-point Times, boldface type, 
   left-aligned to the text, initially capitalized. 
   The abstract is to be in 9-point, single-spaced type.
   The abstract may be up to 3 inches (7.62 cm) long. \\
   Leave one blank line after the abstract, 
   then add the subject categories according to the ACM Classification Index 
   (see http://www.acm.org/class/1998/).

% TODO
\begin{classification} % according to http://www.acm.org/class/1998/
\CCScat{Computer Graphics}{I.3.3}{Picture/Image Generation}{Line and curve generation}
\end{classification}

\end{abstract}

%-------------------------------------------------------------------------
\section{Introduction}

Long captions should be set as in Figure~\ref{fig:ex1}.

\begin{figure}[htb]
   % an empty figure just consisting of the caption lines
   \caption{\label{fig:ex1}
     'Empty' figure only to serve as an example of long caption requiring 
     more than one line. It is not typed centered but aligned on both sides.}
\end{figure}

%-------------------------------------------------------------------------
\subsection{References}

All graphics should be centered.

%%%
%%% Figure 1
%%%
%\begin{figure}[htb]
%  \centering
%  % the following command controls the width of the embedded PS file
%  % (relative to the width of the current column)
%  \includegraphics[width=.8\linewidth]{sampleFig}
%  % replacing the above command with the one below will explicitly set
%  % the bounding box of the PS figure to the rectangle (xl,yl),(xh,yh).
%  % It will also prevent LaTeX from reading the PS file to determine
%  % the bounding box (i.e., it will speed up the compilation process)
%  % \includegraphics[width=.95\linewidth, bb=39 696 126 756]{sampleFig}
%  %
%  \parbox[t]{.9\columnwidth}{\relax
%           For all figures please keep in mind that you \textbf{must not}
%           use images with transparent background! 
%           }
%  %
%  \caption{\label{fig:firstExample}
%           Here is a sample figure.}
%\end{figure}

%If your paper includes images, it is very important that they are of
%sufficient resolution to be faithfully reproduced.
%
%To determine the optimum size (width and height) of an image, measure
%the image's size as it appears in your document (in millimeters), and
%then multiply those two values by 12. The resulting values are the
%optimum $x$ and $y$ resolution, in pixels, of the image. Image quality
%will suffer if these guidelines are not followed.
%
%Example 1: 
%%
%An image measures 50\,mm by 75\,mm when placed in a document. This
%image should have a resolution of no less than 600 pixels by 900
%pixels in order to be reproduced faithfully.

%%------------------------------------------------------------------------
%\subsection{Copyright forms}
%
%You must include your signed Eurographics copyright release form
%when you submit your finished paper. We MUST have this form before
%your paper can be published in the proceedings.

%-------------------------------------------------------------------------

%\bibliographystyle{eg-alpha}
\bibliographystyle{eg-alpha-doi}

\bibliography{sca2012.bib}

\end{document}

### 2012-05-28 Doc/SCA2012/sca2012.tex
SCA 2012
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@409 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\title[EG \LaTeX\ Author Guidelines]%
+\usepackage{ucs}
+\usepackage[utf8x]{inputenc}
+
+\usepackage{amsmath}
+
+% Partial derivative
+\newcommand{\deriv}[2]{\frac{\partial #1}{\partial #2}}
+
+\newcommand{\Figure}[3]{%
+\begin{figure}[htb]
+  \centering
+  \includegraphics[width=#1]{#2}
+  \caption{\label{fig-#2}#3}
+\end{figure}}
+
+\title[Short title]%
       {Fill in the Title} % TODO
@@
 
-Long captions should be set as in Figure~\ref{fig:ex1}.
+Objects with thin structure form a very specific field of research. Because
+they have high ratio between width and thickness methods from volumetric
+deformation modeling cannot be immediately used. Such object occur in many
+places e.g. textiles, paper or leaves. They are also very common in
+anatomical structures for example in form of tubular objects like blood
+vessels or colon, in the form of thin membranes, but also many others
+(lense, skin, \ldots).
 
-\begin{figure}[htb]
-   % an empty figure just consisting of the caption lines
-   \caption{\label{fig:ex1}
-     'Empty' figure only to serve as an example of long caption requiring 
-     more than one line. It is not typed centered but aligned on both sides.}
-\end{figure}
+Methods based on continuum mechanics and namely theory of elasticity gained
+on popularity ever since Terzopoulos \cite{Terzopoulos1987} presented his
+work on elastic deformation modeling. While the methods using finite
+elements and theory of elasticity are fairly popular in volumetric modeling
+\footnote{TODO: do we need ref. here?} of deformations, they are still
+frowned uppon in the area of thin structures. Instead either the
+mass-spring models \cite{Volino2009} or bending models
+\cite{Grinspun2003,Choi2007} are preffered. Unfortunately they have the
+disadvantage that physical parameters of a material (like Young's modulus
+and Poisson's ratio) cannot be used directly.
 
-%-------------------------------------------------------------------------
-\subsection{References}
+The usual argument agains FEM based methods being that they are too slow.
+This is indeed true if lots of elements are used to discretize the object.
+However, exploiting the bending property of the shell elements allows us to
+achieve good results even with relatively small number of elements. We can
+use high-polygonal mesh only for rendering or collision detection and
+response. Similar idea has been already presented in the work of Bouthors
+et al. \cite{Bouthors2007}. They used twinned meshes for animation of mesh
+defined by implicit surface. Simple low-polygonal mesh was used for
+mechanical FEM model and the mesh was subdivided in areas with high
+curvature after every step of the simulation to produce finer mesh for
+rendering. The underlaying FEM model is however not aware of the implicit
+surface used. We build uppon the work of Comas \cite{Comas2010c} who used a
+simple cubic polynomial to describe the element geometry and derived the
+shell model from it.
 
-All graphics should be centered.
+The Comas's element has several drawbacks that we needed to address. Namely
+it produces discontinuities between the elements on the edges, it is
+missing one rotationonal degree of freedom and is not completly symetric in
+context of deformations.
 
+% TODO: cite Ubach & Oñate?
+
+In this paper we present a new shell element based on the formulation of
+Bézier triangle.
+
+\subsection{Contribution}
+
+% TODO
+The main contributions of the paper are following:
+\begin{itemize}
+    % not really novel, similar model already in Ubach2010
+    %\item shell element based on Bézier triangles
+    \item Our element solves the problems of the element presented by Comas
+        \cite{Comas2010c}.
+
+    % TODO: needs better explanation
+    % + two stage interpolation => Nodes (6DOFS) => bezier points => bezier
+    %   triangle
+    \item Two stage interpolation: 6DOF nodes -> bezier points -> bezier
+        triangle
+
+    % + separation between nodes and dofs
+    \item The model separates nodes of the mesh from degrees of freedom of the
+        control mesh. This allows mapping a surface on deformable skeleton.
+
+    % + mapping: smooth surface results with less shells + collision response
+    %   on smooth surface +  [? if I have time]  deformable on deformable:
+    %   conjonctiva 2D motion on the surface of the shells
+    \item Smooth maping of high resolution mesh on the curved surface of the
+        shell. The mapped surfce can be used for collision detection and
+        allows the propagation of proper collision response back onto the
+        (low resolution) mechanical mesh.
+
+\end{itemize}
+
+\section{Preliminaries}
+
+\subsection{Shell Element}
+
+Shell element combines two types of deformations:
+
+\begin{itemize}
+
+    \item Elastic membrane: deformation in the plane of the element,
+        described by in-plane displacements $u_x, u_y$ and rotation
+        $\theta_z$ around the axis perpendicular to the plane of the
+        element (Figure \ref{fig-membrane}).
+
+    \item Bending plate: out of the plane deformation described by rotation
+        around two in-plane axes $\theta_x, \theta_z$ and out of plane
+        displacmenent $u_z$ (Figure \ref{fig-plate}).
+
+\end{itemize}
+
+\Figure{0.8\linewidth}{membrane}
+{DOFs of elastic membrane element.}
+
+\Figure{0.8\linewidth}{plate}
+{DOFs of bending plate element.}
+
+\subsection{Bézier Triangle} % {{{
+
+In the rest of the document we define the surface over the triangle:
+\begin{equation}
+    (\xi_1,\xi_2) \in \Delta^2 = \left\{ (\xi_1,\xi_2)~|~\xi_1, \xi_2 \ge 0
+        \mathrm{~and~} \xi_1+\xi_2 \le 1 \right\}
+\end{equation}
+
+The general $n$-th order Bézier triangle is defined as:
+
+\begin{equation}
+    T(\xi_1, \xi_2) = \sum_{0 \le i + j \le n} B^n_{i,j}(\xi_1,\xi_2) P_{i,j}
+\end{equation}
+
+where $P_{i,j}$ are the control points and $B^n_{i,j}$ are the bivairate
+Bernstein basis functions defined as:
+
+\begin{equation}
+  B^n_{i,j} (\xi_1,\xi_2) =
+    \binom{n}{i~j} \xi_1^i \xi_2^j (1-\xi_1-\xi_2)^{n-i-j},
+\end{equation}
+\begin{equation}
+  \binom{n}{i~j} = \frac{n!}{i!j!(n-i-j)!}
+\end{equation}
+
+where $ 0 \le i+j \le n $.
+
+Specificaly the cubic Bézier triangle ($n=3$) used in our model is
+described by 10 control points and the surface is explicitly defined as (see
+fig. \ref{fig-bezier}):
+
+\begin{equation}\label{eq-cubicbez}
+  \begin{split}
+  T(\xi_1,\xi_2) & =
+           + \xi_1^3\ P_1
+           + \xi_2^3\ P_2
+           + \xi_3^3\ P_3 \\
+         & + 3\ \xi_1^2 \xi_2\ P_4
+           + 3\ \xi_1^2 \xi_3\ P_5
+           + 3\ \xi_2^2 \xi_3\ P_6 \\
+         & + 3\ \xi_1 \xi_2^2\ P_7
+           + 3\ \xi_1 \xi_3^2\ P_8
+           + 3\ \xi_2 \xi_3^2\ P_9 \\
+         & + 6\ \xi_1 \xi_2 \xi_3\ P_{10} \\
+  \end{split}
+\end{equation}
+
+where $ \xi_3 = 1 - \xi_1 - \xi_2 $. By naming the respective values of
+the Bernstein basis functions $n_i$ we can shortly express
+\eqref{eq-cubicbez} as:
+
+\begin{equation}\label{eq-cubicbez2}
+    T = \sum_{i=1}^{10} n_i P_i
+\end{equation}
+
+\Figure{0.8\linewidth}{bezier}
+{Cubic Bézier triangle with net of 10 control points.}
+
+% }}}
+
+\section{Bézier Shell Element} % {{{
+
+\subsection{Control Points} % {{{
+
+The Bézier triangles are defined by a net of control points, however the
+simulated object is usually described by a triangular mesh. We construct
+the Bézier mesh by taking into account the curvature of the triangular
+mesh. When constructing the mesh a special care has to be taken to maintain
+the continuity accross the nodes and edges. During the initialization phase
+we partialy employ the method described in \cite{Ubach2010} to maintain
+$C^0$ continuity on the edges. Each of the control points on the edge is
+computed as the intersection of:
+
+\begin{enumerate}
+    \item The plane perpendicular to the normal at the vertex.
+    \item The plane that contains the the curve of triangle's contour. The
+        choise is arbitrary, but necessary to maintain $C^0$ continuity. 
+        We choose the plane defined by the edge of the flat triangle and
+        average of the two normals at vertices of the edge. 
+    \item The plane perpendicular to the edge of the flat triangle placed at
+        $1/3$ of the edge length.
+\end{enumerate}
+
+Notice that the points on the edge will be same for both triangles sharing
+the edge.
+
+We do not repeat this procedure during the simulation. For faster update of
+the control points we rigidly attach the edge points to corner points. We
+remember the position and orientation of the edge point relative to the
+nearest corner point and use this information together with the rotation on
+the corner points to reconstruct new position of control points. See Figure
+\ref{fig-segments} for the correspondence between edge points and
+associated corner points.
+
+\Figure{0.8\linewidth}{segments}
+{Correspondence between edge points and associated corner points shown in
+blue.}
+
+$C^1$ continuity can also be maintained if special care is taken when
+computing the position of the central point (see \cite{Ubach2010}). For
+simplicity we compute the central point followingly:
+
+\begin{equation}\label{eq-central}
+    P_{10} = \frac{1}{3}(\sum_{i=4}^9 P_i - \sum_{i=1}^3 P_i)
+\end{equation}
+
+% TODO: explain the equation eq-central? would require an image
+
+That way the central point is slightly elevated and not in the plane of
+other nodes thus keeping the curvature of the element.
+
+%}}}
+
+% TODO - nodes + DOFs (local + global position)
+
+\subsection{Mechanics}
+
+We base the displacement field $u$ on the formulation of the Bézier
+triangle \eqref{eq-cubicbez}:
+
+\begin{equation}\label{eq-bezU}
+\begin{split}
+  u & = U_1 ( \xi_1^3 + 3 \xi_1^2 \xi_2 + 3 \xi_1^2 \xi_3
+        + 2 \xi_1 \xi_2 \xi_3 ) \\
+    & + U_2 ( \xi_2^3 + 3 \xi_1 \xi_2^2 + 3 \xi_2^2 \xi_3
+        + 2 \xi_1 \xi_2 \xi_3 ) \\
+    & + U_3 ( \xi_3^3 + 3 \xi_1 \xi_3^2 + 3 \xi_2 \xi_3^3
+        + 2 \xi_1 \xi_2 \xi_3 ) \\
+    & + 3 \xi_1^2 \xi_2 (P_{4-1} \times \theta_1) \\
+    & + 3 \xi_1^2 \xi_3 (P_{5-1} \times \theta_1) \\
+    & + 3 \xi_1 \xi_2^2 (P_{7-2} \times \theta_2) \\
+    & + 3 \xi_2^2 \xi_3 (P_{6-2} \times \theta_2) \\
+    & + 3 \xi_1 \xi_3^2 (P_{8-3} \times \theta_3) \\
+    & + 3 \xi_2 \xi_3^2 (P_{9-3} \times \theta_3) \\
+    & + 2 \xi_1 \xi_2 \xi_3 (P_{10-1} \times \theta_1 +
+        P_{10-2} \times \theta_2 +
+        P_{10-3} \times \theta_3)
+\end{split}
+\end{equation}
+
+where $U_1, U_2, U_3$ are the translational displacment vectors of the
+corner nodes of the element, $\theta_1, \theta_2, \theta_3$ are the angular
+displacements of the corner nodes and $P_{i-j}$ are the segments from
+corner nodes to other nodes of the control mesh:
+
+\begin{equation}
+    P_{i-j} = P_j-P_i, \quad i \in [4;10], j \in [1;3]
+\end{equation}
+
+Using the plate theory we can now compute the strain-displacement matrices
+from the displacement field. From the Cauchy's strain tensor for the
+elastic membrane:
+
+\begin{equation}
+    J_m = \left[ \begin{matrix}
+        \deriv{u_x}{x} \\
+        \deriv{u_y}{y} \\
+        \deriv{u_x}{y} + \deriv{u_y}{x}
+    \end{matrix} \right]
+\end{equation}
+
+and from Kirchhoff-Love theory for thin plates:
+
+\begin{equation}
+    J_b = \left[ \begin{matrix}
+        - z \deriv{^2 u_z}{x^2} \\
+        - z \deriv{^2 u_z}{y^2} \\
+        - 2z \deriv{^2 u_z}{xy}
+    \end{matrix} \right]
+\end{equation}
+
+Assuming constant thickness $t$ of the element and integrating over the volume
+of the element we compute the stiffness matrices for the elastic membrane
+and bending plate respectively:
+
+\begin{align}
+    \label{eq-Km}
+    K_m & = \iiint_V J_m^T M J_m \, \mathrm{d} V \\
+    \label{eq-Kb}
+    K_b & = \iiint_V J_b^T M J_b \, \mathrm{d} V
+\end{align}
+
+where $M$ is the material matrix. We use Hooke's law.
+
+Because the deformation field for the shell is non-linear \eqref{eq-bezU}
+the integrals \eqref{eq-Km} and \eqref{eq-Kb} have to be computed using
+numerical integration. In our implementation we employ 4-point Gaussian
+quadrature for integration over triangle area.
+% TODO: add ref
+% TODO: add abscissas/weights?
+%  \xi_1        \xi_2           w
+% 0.211324865, 0.166666667 0.197168783
+% 0.211324865, 0.622008467 0.197168783
+% 0.788675134, 0.044658198 0.052831216
+% 0.788675134, 0.166666667 0.052831216
+
+To keep the system linear during simulation we compute the forces in
+corotational frame.
+
+% }}}
+
+\section{Mechanical Mapping} % {{{
+
+% TODO: mention it is 6 DOF <-> 3 DOF
+
+Because of the bending property of the shells relatively few elements are
+necessary to simulate curved surface. To enrich the visual experience from
+the simulated object it is desirable to use more triangles in the areas
+with high curvature during rendering. To do that we can map a high
+resolution mesh onto the mechanical mesh. For every vertex of the high
+resolution mesh we first find the triangle on the mechanical mesh that is
+closest to the vertex and then assign barycentric coordinates on the
+triangle to this vertex. After every step of the simulation the high
+resolution mesh is updated using the assigned barycentric coordinates and
+the function of the surface \eqref{eq-cubicbez}.
+
+mapping velocities (necessary for collisions?)
+
+Differentiating \eqref{eq-cubicbez2} by time we get:
+
+\begin{equation}
+    \dot{T} = \sum_{i=1}^{10} n_i \dot{P_i} = \sum_{i=1}^{10} n_i V_i
+\end{equation}
+
+Which means we use the same expression to interpolate the velocities only
+by substituting the velocities at control points. For the corner nodes we
+already have the velocities $V_i$. For the internal nodes we need to compute
+them. In case of the edge nodes it is a problem of computing the velocity
+of a point attached to the moving rigid body:
+
+\begin{eqnarray}
+    V_i = V_j + \omega_j \times (P_i - P_j)
+\end{eqnarray}
+
+where $j$ is the index of the corner node this edge node is attached to and
+$\omega_j$ is the angular velocity at the node. For the central node we
+again use the interpolation as in \eqref{eq-central}.
+
+* reverse mapping forces for collision response
+
+We compute the influence of force $F$ on the corners through all the
+control points by the formula:
+
+\begin{equation}
+    F_j += F n_i
+\end{equation}
+
+and the torques applied through edge control points by:
+
+\begin{eqnarray}
+    \theta_j += (P_i - P_j) \times F
+\end{eqnarray}
+
+where $j$ is the index of the corner node associated with control point
+$P_i$.
+
+And finaly we add the torques through the central control point with:
+
+\begin{eqnarray}
+    \theta_1 += ((P_4 - P_1) + (P_5 - P_1)) \times F \\
+    \theta_2 += ((P_6 - P_2) + (P_7 - P_2)) \times F \\
+    \theta_3 += ((P_8 - P_3) + (P_9 - P_3)) \times F
+\end{eqnarray}
+
+% }}}
+
+\section{Results}
+
+proof of symmetry
+
+some other examples?
+
+computation speed
+
+\section{Discussion}
+
+% = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+%
+% Benefits:
+% + C^0 continuity on the edges
+% + Perfectly symmetric results for the element (it is not the case with
+%   "classical shells")
+% + Better conditioning (comparison between Comas model and ours in term of
+%   convergence of CG...) * Good trade-off between continuity of interpolation
+%   and spasticity: only 18 Dofs per elements for an interpolation with 3d
+%   order
+% + polynomials [Bezier triangle has theoretically 10 points x 3 dofs=> 30 dofs]
+%
+% Limitations:
+% + Continuity C1 at the nodes but C0 on the edges
+% + The segments are at fixed distance from corner nodes, may lead to
+%   uneven curvature for too stretched or shrinked elements 
+%
+% Applications:
+% + cataract surgery
+% + arterial vessel (MICCAI)
+% + vagina/uterus?
+%
+% = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+
+%Long captions should be set as in Figure~\ref{fig:ex1}.
+%
+%\begin{figure}[htb]
+%   % an empty figure just consisting of the caption lines
+%   \caption{\label{fig:ex1}
+%     'Empty' figure only to serve as an example of long caption requiring 
+%     more than one line. It is not typed centered but aligned on both sides.}
+%\end{figure}
+%
+%All graphics should be centered.
+%
 %%%
@@
 
+%-------------------------------------------------------------------------
+
 %\bibliographystyle{eg-alpha}

### 2012-07-09 Doc/VRIPHYS2012/vriphys2012.tex
changing SCA 2012 to Vriphys 2012
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@419 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2012-05-28 Doc/SCA2012/sca2012.tex
@@
 \begin{equation}
-    (\xi_1,\xi_2) \in \Delta^2 = \left\{ (\xi_1,\xi_2)~|~\xi_1, \xi_2 \ge 0
+    (\xi_1,\xi_2) \in \bigtriangleup = \left\{ (\xi_1,\xi_2)~|~\xi_1, \xi_2 \ge 0
         \mathrm{~and~} \xi_1+\xi_2 \le 1 \right\}
@@
     \item The plane perpendicular to the normal at the vertex.
-    \item The plane that contains the the curve of triangle's contour. The
+    \item The plane that contains the curve of triangle's contour. The
         choise is arbitrary, but necessary to maintain $C^0$ continuity. 
@@
 
-$C^1$ continuity can also be maintained if special care is taken when
-computing the position of the central point (see \cite{Ubach2010}). For
+$C^1$ continuity on the edges can also be maintained if special care is taken
+when computing the position of the central point (see \cite{Ubach2010}). For
 simplicity we compute the central point followingly:
@@
 
-We base the displacement field $u$ on the formulation of the Bézier
-triangle \eqref{eq-cubicbez}:
+Our definition of the displacement field $u$ is based on the formulation of the
+Bézier triangle \eqref{eq-cubicbez}:
 
@@
 \begin{equation}
-    P_{i-j} = P_j-P_i, \quad i \in [4;10], j \in [1;3]
+    P_{i-j} = P_j-P_i %, \quad i \in [4;10], j \in [1;3]
 \end{equation}
@@
 
-\begin{equation}
+\begin{equation}\label{eq-cauchy}
     J_m = \left[ \begin{matrix}
@@
 
-\begin{equation}
+\begin{equation}\label{eq-kirchhoff}
     J_b = \left[ \begin{matrix}
@@
 
-where $M$ is the material matrix. We use Hooke's law.
+where $M$ is the material matrix. To keep the system simple we use linear
+Hooke's law for isotropic materials.
 
@@
 the integrals \eqref{eq-Km} and \eqref{eq-Kb} have to be computed using
-numerical integration. In our implementation we employ 4-point Gaussian
+numerical integration. In our implementation we employ 6-point Gaussian
 quadrature for integration over triangle area.
@@
 % TODO: add abscissas/weights?
-%  \xi_1        \xi_2           w
-% 0.211324865, 0.166666667 0.197168783
-% 0.211324865, 0.622008467 0.197168783
-% 0.788675134, 0.044658198 0.052831216
-% 0.788675134, 0.166666667 0.052831216
 
-To keep the system linear during simulation we compute the forces in
-corotational frame.
+The Cauchy's strain tensor \eqref{eq-cauchy} is know to not be rotation
+invariant \footnote{add ref}. To deal with rigid body rotations we compute
+the displacements and forces in corotational frame. At each step of the
+simulation we compute the local frame for every element from the position
+of the corner nodes and we allign one axis of the frame with one edge of
+the element. \footnote{add image}
+
+The resulting internal force of deformed elemnt is then computed relative
+to this frame:
+
+\begin{equation}
+    F_e = R_e^T (K_e u) = R_e^T \left(K_e(R_e x - R_e^0 x^0)\right)
+\end{equation}
+
+\noindent
+where $R$ and $R_0$ are rotation matrices of the frame for current and rest
+mesh respectively.
 
@@
 
-\section{Mechanical Mapping} % {{{
-
-% TODO: mention it is 6 DOF <-> 3 DOF
+\section{Visual and Mechanical Mapping} % {{{
 
@@
 the simulated object it is desirable to use more triangles in the areas
-with high curvature during rendering. To do that we can map a high
-resolution mesh onto the mechanical mesh. For every vertex of the high
-resolution mesh we first find the triangle on the mechanical mesh that is
-closest to the vertex and then assign barycentric coordinates on the
-triangle to this vertex. After every step of the simulation the high
+with high curvature during rendering. Besides the visual accuracy the high
+resolution mesh correctly modeling the curved areas is needed for
+interaction and correct simulation of collisions and constraints. To handle
+these we have also to know how to map the velocities from the mechanical
+mesh onto the high resolution mesh and how to map forces acting on the high
+resolution mesh back onto the mechanical mesh.
+
+\subsection{Mapping the Surface Mesh}
+
+We use the fact that the geometry of the shell is based on the formulation
+of cubic Bézier triangle \eqref{eq-cubicbez}. For every vertex of the high
+resolution mesh we first find the corresponding triangle on the mechanical
+mesh that is closest to the vertex and assign barycentric coordinates on
+the triangle to this vertex. After every step of the simulation the high
 resolution mesh is updated using the assigned barycentric coordinates and
@@
 
-mapping velocities (necessary for collisions?)
+\subsection{Mapping Velocities}
+
+In the
 
@@
 Which means we use the same expression to interpolate the velocities only
-by substituting the velocities at control points. For the corner nodes we
+substituting the velocities at control points. For the corner nodes we
 already have the velocities $V_i$. For the internal nodes we need to compute
@@
 
-* reverse mapping forces for collision response
+\subsection{Mapping Forces}
+
+For physical interaction with the object one also requiers that the normal
+forces applied on the high resolution mesh are trasmitted back as forces
+and torques on the vertices of the mechanical mesh.
 
 We compute the influence of force $F$ on the corners through all the
-control points by the formula:
-
-\begin{equation}
-    F_j += F n_i
-\end{equation}
-
-and the torques applied through edge control points by:
+associated control points by the formula:
 
 \begin{eqnarray}
-    \theta_j += (P_i - P_j) \times F
+    F_1 & = & F \sum_{i \in \{1,4,5\}} n_i + \frac{1}{3} F n_{10} \\
+    F_2 & = & F \sum_{i \in \{2,6,7\}} n_i + \frac{1}{3} F n_{10} \\
+    F_3 & = & F \sum_{i \in \{3,8,9\}} n_i + \frac{1}{3} F n_{10} \\
 \end{eqnarray}
 
-where $j$ is the index of the corner node associated with control point
-$P_i$.
+%\begin{eqnarray}
+%    F_1 & = & F (n_1 + n_4 + n_5) + \frac{1}{3} F n_{10} \\
+%    F_2 & = & F (n_2 + n_6 + n_7) + \frac{1}{3} F n_{10} \\
+%    F_3 & = & F (n_3 + n_8 + n_9) + \frac{1}{3} F n_{10} \\
+%\end{eqnarray}
 
-And finaly we add the torques through the central control point with:
+Similarily we compute the torques applied through edge control points and
+the central control points by:
 
 \begin{eqnarray}
-    \theta_1 += ((P_4 - P_1) + (P_5 - P_1)) \times F \\
-    \theta_2 += ((P_6 - P_2) + (P_7 - P_2)) \times F \\
-    \theta_3 += ((P_8 - P_3) + (P_9 - P_3)) \times F
+    \theta_1 & = & P_{1-4} \times (n_4 F) + P_{1-5} \times (n_5 F) + \\
+        & + & \frac{1}{3} (P_{1-4} + P_{1-5}) \times (n_{10} F) \\
+    \theta_2 & = & P_{1-6} \times (n_6 F) + P_{1-7} \times (n_7 F) + \\
+        & + & \frac{1}{3} (P_{1-6} + P_{1-7}) \times (n_{10} F) \\
+    \theta_3 & = & P_{1-8} \times (n_8 F) + P_{1-9} \times (n_9 F) + \\
+        & + & \frac{1}{3} (P_{1-8} + P_{1-9}) \times (n_{10} F) \\
 \end{eqnarray}
+
+%\begin{eqnarray}
+%    \theta_1 & = & \sum_{i \in \{4,5\}} P_{1-i} \times (n_i F) \\
+%    & + &\frac{1}{3} (P_{1-4} + P_{1-5}) \times (n_{10} F) \\
+%    \theta_1 & = & \sum_{i \in \{6,7\}} P_{1-i} \times (n_i F) + \\
+%    & + &\frac{1}{3} (P_{1-6} + P_{1-7}) \times (n_{10} F) \\
+%    \theta_1 & = & \sum_{i \in \{8,9\}} P_{1-i} \times (n_i F) + \\
+%    & + &\frac{1}{3} (P_{1-8} + P_{1-9}) \times (n_{10} F) \\
+%\end{eqnarray}
+%
+%\begin{eqnarray}
+%    \theta_1 & = & \sum_{i \in \{4,5\}} P_{1-i} \times (n_i F) +
+%    \frac{1}{3} (P_{1-4} + P_{1-5}) \times (n_{10} F) \\
+%    \theta_1 & = & \sum_{i \in \{6,7\}} P_{1-i} \times (n_i F) +
+%    \frac{1}{3} (P_{1-6} + P_{1-7}) \times (n_{10} F) \\
+%    \theta_1 & = & \sum_{i \in \{8,9\}} P_{1-i} \times (n_i F) +
+%    \frac{1}{3} (P_{1-8} + P_{1-9}) \times (n_{10} F) \\
+%\end{eqnarray}
 

### 2012-07-13 Doc/VRIPHYS2012/vriphys2012.tex
changing from SCA 2012 to Vriphys 2012
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@420 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
more on the article
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@422 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \documentclass{egpubl}
-\usepackage{sca2012}
+\usepackage{vriphys2012}
 
@@
 % for including postscript figures
-% mind: package option 'draft' will replace PS figure by a filname within a frame
+% mind: package option 'draft' will replace PS figure by a filename within a frame
 \ifpdf \usepackage[pdftex]{graphicx} \pdfcompresslevel=9
@@
 
+%\Figure{width}{name}{caption}
 \newcommand{\Figure}[3]{%
@@
 %-------------------------------------------------------------------------
-\section{Introduction}
+
+\section{Introduction} % {{{
 
@@
 vessels or colon, in the form of thin membranes, but also many others
-(lense, skin, \ldots).
+(lens, skin, \ldots).
 
@@
 \footnote{TODO: do we need ref. here?} of deformations, they are still
-frowned uppon in the area of thin structures. Instead either the
+frowned upon in the area of thin structures. Instead either the
 mass-spring models \cite{Volino2009} or bending models
-\cite{Grinspun2003,Choi2007} are preffered. Unfortunately they have the
+\cite{Grinspun2003,Choi2007} are preferred. Unfortunately they have the
 disadvantage that physical parameters of a material (like Young's modulus
@@
 
-The usual argument agains FEM based methods being that they are too slow.
+The usual argument against FEM based methods being that they are too slow.
 This is indeed true if lots of elements are used to discretize the object.
@@
 rendering. The underlaying FEM model is however not aware of the implicit
-surface used. We build uppon the work of Comas \cite{Comas2010c} who used a
+surface used. We build upon the work of Comas \cite{Comas2010c} who used a
 simple cubic polynomial to describe the element geometry and derived the
@@
 it produces discontinuities between the elements on the edges, it is
-missing one rotationonal degree of freedom and is not completly symetric in
+missing one rotational degree of freedom and is not completely symmetric in
 context of deformations.
 
-% TODO: cite Ubach & Oñate?
+% TODO: cite Ubach & Oñate
+TODO: cite \cite{Ubach2010}: rotation-free shell elements, also based on
+Bézier triangle
 
@@
 
-\section{Preliminaries}
+ % }}}
+
+\section{Preliminaries} % {{{
 
@@
 
+\noindent
 where $P_{i,j}$ are the control points and $B^n_{i,j}$ are the bivairate
@@
 
+\noindent
 where $ 0 \le i+j \le n $.
@@
 
+\noindent
 where $ \xi_3 = 1 - \xi_1 - \xi_2 $. By naming the respective values of
@@
 {Cubic Bézier triangle with net of 10 control points.}
+
+% }}}
 
@@
 
-\subsection{Mechanics}
+\subsection{Mechanics} % {{{
 
@@
 
+\noindent
 where $U_1, U_2, U_3$ are the translational displacment vectors of the
@@
 
+\noindent
 and from Kirchhoff-Love theory for thin plates:
@@
 
+\noindent
 where $M$ is the material matrix. To keep the system simple we use linear
@@
 
-The Cauchy's strain tensor \eqref{eq-cauchy} is know to not be rotation
-invariant \footnote{add ref}. To deal with rigid body rotations we compute
-the displacements and forces in corotational frame. At each step of the
-simulation we compute the local frame for every element from the position
-of the corner nodes and we allign one axis of the frame with one edge of
-the element. \footnote{add image}
+% }}}
 
-The resulting internal force of deformed elemnt is then computed relative
-to this frame:
+\subsection{Corotational Formulation} % {{{
+\label{sec-corot}
+
+It is known that the Cauchy's strain tensor \eqref{eq-cauchy} is not
+rotation invariant \footnote{add ref} and produces ghost forces for rigid
+rotations. The alternative approach is to use Green's strain tensor,
+however it leads to nonlinear system. To keep the system linear and to deal
+with rigid body rotations we compute the displacements and forces in
+corotational frame. At each step of the simulation we compute the local
+frame for every element, the resulting internal force of deformed element
+is then computed relative to this frame:
 
@@
 mesh respectively.
+
+TODO: how we compute frame?
+%from the position of the corner nodes and we align one axis of the frame with one
+%edge of the element. \footnote{add image}
+
+%The displacement field defined in \eqref{eq-bezU} is clearly symmetric in
+%respect to deformations, but ...
+
+This choice of the frame aligns one axis of the frame with one of the edges
+of the triangle. Correct choice of the corotational frame is a tricky
+problem \cite{Felippa2005} and incorrect frame for the membrane element can
+have adverse effects on the simulation. In our case we have experienced
+nonnegligible difference in deformation of equilateral triangle depending
+on whether an axis was aligned to the edge favored by the deformation or to
+one of the two other edges.
+
+Performing a polar decomposition on the deformation gradient $F = \nabla u
++ I$ can be used to extract a rotation component from the deformation of
+the element \cite{Ciarlet1994}. Because we are only fixing the membrane
+element we can perform the polar decomposition in 2D with the following
+formula:
+
+\begin{equation}
+    R = F + \mathrm{sgn}(\mathrm{det}(F)) \begin{pmatrix}
+        F_{22} & - F_{21} \\
+        F_{12} & F_{11}
+    \end{pmatrix}
+\end{equation}
+
+\noindent
+and subsequently normalizing the columns. We then rotate the corotational
+frame in inverse direction. Even single step of this process improves the
+final frame but more iterations of the polar decompositions are usually
+necessary (up to 20) which severely degrades the performance. But polar
+decomposition tends to "overshoot" the ideal rotation angle and each
+subsequent step performs a rotation in the opposite direction. By properly
+scaling the rotation angle we are able to minimize the amount of needed
+iterations down to 5. The best value for the scaling factor depends on the
+simulated problem, but we have experimentally localized the best value to
+be somewhere between 0.6 and 0.65. We choose to use the value 0.61.
+
+We employ two stop conditions for the iteration process:
+\begin{itemize}
+    \item The change in rotation angle is less than $10^{-6}$,
+    \item and the maximum number of 5 iterations.
+\end{itemize}
+
+If the first condition is not met in 5 iterations it can mean one of three
+things:
+
+\begin{enumerate}
+    \item We still have very good approximation of the ideal frame.
+    \item The convergence for the element is too slow.
+    \item The process is diverging.
+\end{enumerate}
+
+\noindent
+and performing any more iterations is unnecessary.
+
+% TODO: maybe make a short table showing the convergence to the symmetric
+% solution.
+
+% }}}
 
@@
     F_2 & = & F \sum_{i \in \{2,6,7\}} n_i + \frac{1}{3} F n_{10} \\
-    F_3 & = & F \sum_{i \in \{3,8,9\}} n_i + \frac{1}{3} F n_{10} \\
+    F_3 & = & F \sum_{i \in \{3,8,9\}} n_i + \frac{1}{3} F n_{10}
 \end{eqnarray}
@@
 
-Similarily we compute the torques applied through edge control points and
+Similarly we compute the torques applied through edge control points and
 the central control points by:
 
-\begin{eqnarray}
-    \theta_1 & = & P_{1-4} \times (n_4 F) + P_{1-5} \times (n_5 F) + \\
-        & + & \frac{1}{3} (P_{1-4} + P_{1-5}) \times (n_{10} F) \\
-    \theta_2 & = & P_{1-6} \times (n_6 F) + P_{1-7} \times (n_7 F) + \\
-        & + & \frac{1}{3} (P_{1-6} + P_{1-7}) \times (n_{10} F) \\
-    \theta_3 & = & P_{1-8} \times (n_8 F) + P_{1-9} \times (n_9 F) + \\
-        & + & \frac{1}{3} (P_{1-8} + P_{1-9}) \times (n_{10} F) \\
-\end{eqnarray}
+\begin{gather}
+    \begin{split}
+        \theta_1 & = P_{1-4} \times (n_4 F) + P_{1-5} \times (n_5 F) + \\
+        & + \frac{1}{3} (P_{1-4} + P_{1-5}) \times (n_{10} F)
+    \end{split}
+    \\
+    \begin{split}
+        \theta_2 & = P_{1-6} \times (n_6 F) + P_{1-7} \times (n_7 F) + \\
+        & + \frac{1}{3} (P_{1-6} + P_{1-7}) \times (n_{10} F) \\
+    \end{split}
+    \\
+    \begin{split}
+        \theta_3 & = P_{1-8} \times (n_8 F) + P_{1-9} \times (n_9 F) + \\
+        & + \frac{1}{3} (P_{1-8} + P_{1-9}) \times (n_{10} F) \\
+    \end{split}
+\end{gather}
 
@@
 
-\section{Results}
+\section{Validation and Results} % {{{
 
-proof of symmetry
+% We cannot show results for simple plane in bending because the solution
+% seems to converge to wrong value (but it converges! :)). Thus showing
+% results for plain membrane deformations doesn't make sense (it would be
+% inconsistent).
 
-some other examples?
+In next two subsections we present tests to validate the convergence of our
+Bézier shell element. The results are compared with the shell element
+composed of DKT element and optimal ANDES element \cite{Felippa2003}. The
+last subsection presents evaluation of the computational complexity of our
+solution.
 
-computation speed
+\subsection{Roof Test}
 
-\section{Discussion}
+To validate the results of the element we have performed a modified version
+of a test known as Scordelis-Lo roof. It simulates a cylindrical roof under
+self-weight. The geometry is defined by 80° cylindrical patch of length
+$L=50$ and radius $r=25$ and thickness $t=0.25$ and is discretized into
+mesh of NxN nodes. The physical parameters are $E = 4.32 \times 10^8$ and
+$\nu = 0$. The roof is loaded with uniformly distributed load $q = 90$ per
+unit area. The curved edges are clamped at both ends to avoid rigid body
+movement. This the only difference from the original Scordelis-Lo roof test
+where the edges are free to move in longitudinal direction. The Figure
+\ref{fig-roof} shows the vertical displacement of midpoint on the free
+edge.
+
+\Figure{\linewidth}{roof}
+{Convergence for modified Scordelis-Lo roof test of DKT+ANDES element and
+our Bézier shell element.}
+
+\subsection{Hemisphere Test}
+
+Second performed test is a hemisphere with hole subjected to two opposing
+forces at the base. The geometry is described by a hemisphere with radius
+$r=10$ and thickness $t=0.04$ with 18° hole at the top and the geometry is
+discretized into the grid of NxN nodes per quadrant. The physical
+parameters are $E = 6.825 \times 10^7$ and $\nu = 0.3$. The hemisphere is
+at it's base subjected to two opposing outwards forces and two opposing
+inwards forces with magnitude $P = 4$. To avoid rigid body movement we have
+constrained two opposite nodes at the top of the hemisphere. The radial
+displacement for one point with applied load is shown in Figure
+\ref{fig-hemisphere}.
+
+\Figure{\linewidth}{hemisphere}
+{Convergence for test on hemisphere with 18° hole for the DKT+ANDES element and
+our Bézier shell element.}
+
+\subsection{Computation Speed}
+
+In figure \ref{fig-fps} we present performance of the element in terms of
+frames per second. All tests were performed on machine equipped Dual-Core
+AMD Opteron Processor 2218 and 3 GB of RAM and conjugate gradient solver
+has been to solve the system. Our implementation is without optimizations
+and isn't parallel, which means only single CPU core was used for the
+simulation. The values reported are for the raw physical simulation with
+and without the frame fixing method described in subsection
+\ref{sec-corot}. We also present results including the mapping of
+high-resolution mesh onto the physical mesh for visualisation purposes or
+collisions.
+
+\Figure{\linewidth}{fps}
+{Performance in frames per seconds for different number of elements. Shows
+values for element with frame fixing (B) method described in subsection
+\ref{sec-corot} and without it (A), and with high-resolution mesh of 2048
+(C) or 3200 (D) triangles mapped onto the physical mesh.}
+
+We can see that the frame fixing technique is really fast and we are able
+to simulate up to 1000 elements and still maintain visual refresh rate of
+25 FPS. Our implementation of mechanical mapping clearly needs improving
+because it consumes most of the computation time, but it is still able to
+provide good results for more than 600 elements.
+
+% }}}
+
+\section{Discussion and Limitations} % {{{
+
+The element has been successfully tested in SOFA framework for surgical
+simulations. The applications being for example simulation pre-operative
+planning of cardiac surgery on arterial vessels \cite{Kislinskiy2012}, or
+{\em something, something} of vagina/uterus.
 
@@
 %
-% Benefits:
-% + C^0 continuity on the edges
-% + Perfectly symmetric results for the element (it is not the case with
-%   "classical shells")
+% TODO
+% Other Applications:
+% + cataract surgery
+%
+
+Currently, due to the technical decisions the element has two limitations.
+The first limitation is that the element does not provide $C^1$ continuity
+across edges. It is $C^1$ continuous at corners but only $C^0$ on the
+edges. This may prove to be an issue for example if curvature needs to be
+computed not only inside the element but also on the edges.
+
+The second limitation is the choice of rigidly attached edge control
+points. While it greatly simplifies the formulation the rigidity may cause
+issues. In case of large compression of the element the results may be
+unpredictable. This is however unlikely to occur because the element will
+bend before reaching such configuration. It may, however, fail to maintain
+even curvature of the surface if the element is subjected to large
+stretching.
+
+On the other hand the choices provide a good trade-off between continuity
+and simplicity. Only $ 3 \times 6~\mathrm{DOFs} = 18~\mathrm{DOFs}$ are
+necessary while the Bézier triangle has theoretically $ 10 \times
+3~\mathrm{DOFs} = 30~\mathrm{DOFs}$.
+
+% }}}
+
+\subsection{Conclusion} % {{{
+
+This paper presents new shell element with geometry and displacement
+function based on the formulation of Bézier triangle. It fixes problems
+of the previous model \cite{Comas2010c}, namely:
+
+\begin{itemize}
+
+    \item It is continuous on the edges and provides $C^1$ continuity in
+    corners of the element. This is essential for good visual experience if
+    high-resolution mesh is mapped on the physical mesh.
+
+    \item Deformations of the element are symmetric. A method of improving
+    the corotational frame to keep the deformation symmetric is also
+    presented in the paper.
+
+    \item TODO: anything else \ldots? % TODO
+
+    % TODO: check this:
     % + Better conditioning (comparison between Comas model and ours in term of
-%   convergence of CG...) * Good trade-off between continuity of interpolation
-%   and spasticity: only 18 Dofs per elements for an interpolation with 3d
-%   order
-% + polynomials [Bezier triangle has theoretically 10 points x 3 dofs=> 30 dofs]
-%
-% Limitations:
-% + Continuity C1 at the nodes but C0 on the edges
-% + The segments are at fixed distance from corner nodes, may lead to
-%   uneven curvature for too stretched or shrinked elements 
-%
-% Applications:
-% + cataract surgery
-% + arterial vessel (MICCAI)
-% + vagina/uterus?
-%
-% = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
+    %   convergence of CG...)
 
-%Long captions should be set as in Figure~\ref{fig:ex1}.
-%
-%\begin{figure}[htb]
-%   % an empty figure just consisting of the caption lines
-%   \caption{\label{fig:ex1}
-%     'Empty' figure only to serve as an example of long caption requiring 
-%     more than one line. It is not typed centered but aligned on both sides.}
-%\end{figure}
-%
-%All graphics should be centered.
-%
-%%%
-%%% Figure 1
-%%%
-%\begin{figure}[htb]
-%  \centering
-%  % the following command controls the width of the embedded PS file
-%  % (relative to the width of the current column)
-%  \includegraphics[width=.8\linewidth]{sampleFig}
-%  % replacing the above command with the one below will explicitly set
-%  % the bounding box of the PS figure to the rectangle (xl,yl),(xh,yh).
-%  % It will also prevent LaTeX from reading the PS file to determine
-%  % the bounding box (i.e., it will speed up the compilation process)
-%  % \includegraphics[width=.95\linewidth, bb=39 696 126 756]{sampleFig}
-%  %
-%  \parbox[t]{.9\columnwidth}{\relax
-%           For all figures please keep in mind that you \textbf{must not}
-%           use images with transparent background! 
-%           }
-%  %
-%  \caption{\label{fig:firstExample}
-%           Here is a sample figure.}
-%\end{figure}
+\end{itemize}
 
-%If your paper includes images, it is very important that they are of
-%sufficient resolution to be faithfully reproduced.
-%
-%To determine the optimum size (width and height) of an image, measure
-%the image's size as it appears in your document (in millimeters), and
-%then multiply those two values by 12. The resulting values are the
-%optimum $x$ and $y$ resolution, in pixels, of the image. Image quality
-%will suffer if these guidelines are not followed.
-%
-%Example 1: 
-%%
-%An image measures 50\,mm by 75\,mm when placed in a document. This
-%image should have a resolution of no less than 600 pixels by 900
-%pixels in order to be reproduced faithfully.
+Because of the formulation of Bézier triangle that is intrinsically present
+there is a clean and obvious way of mapping high-resolution mesh for
+better visualisation. We also provide a mapping of forces and velocities
+between simulation mesh and physical mesh for collision detection/response
+and interaction with the mesh which are essential component for the surgical
+simulator.
 
-%%------------------------------------------------------------------------
-%\subsection{Copyright forms}
-%
-%You must include your signed Eurographics copyright release form
-%when you submit your finished paper. We MUST have this form before
-%your paper can be published in the proceedings.
+Finally examples showing the convergence to the optimal solution were also provided.
 
-%-------------------------------------------------------------------------
+% }}}
 
@@
 
-\bibliography{sca2012.bib}
+\bibliography{vriphys2012.bib}
 
 \end{document}
+
+% vim: fdm=marker tw=75

### 2012-07-13 Doc/VRIPHYS2012/fps.R

@@ -1,0 +1,23 @@
+#
+# Graph: Frames per seconds with and withouth frame fixing
+#
+
+fps <- read.csv("fps.csv");
+
+pdf("fps.pdf", width=7, height=4);
+p <- par(lwd=2, mar=c(4,4,1,1)+0.1);
+
+plot(range(fps$elements), range(c(fps$fpsBase, fps$fpsPolar, fps$fpsMap1, fps$fpsMap2)),
+    type="n", xlab="Elements", ylab="FPS", log="y");
+
+lines(fps$elements, fps$fpsBase, type="b", col="black", pch=1);
+lines(fps$elements, fps$fpsPolar, type="b", col="black", pch=4);
+lines(fps$elements, fps$fpsMap2, type="b", col="black", pch=6);
+lines(fps$elements, fps$fpsMap1, type="b", col="black", pch=5);
+
+abline(h=25, lty="dotted");
+
+legend("topright", c("A","B", "C", "D"), pch=c(1,4,5,6), col="black");
+
+dev.off()
+par(p);

### 2012-07-13 Doc/VRIPHYS2012/hemisphere.R

@@ -1,0 +1,29 @@
+#
+# Graph: modified Scordelis-Lo roof
+#
+
+n2e <- function(x) (2*4*x*x); # size to element count
+
+hem <- read.csv("hemisphere.csv");
+
+# Compute radial displacement
+d <- apply(cbind(hem$x,hem$y,hem$z), 1, function(x) (sqrt(sum(x*x)))) - 10
+hem <- data.frame(hem, d=d);
+
+# Plot to PDF
+pdf("hemisphere.pdf", width=7, height=4);
+p <- par(lwd=2, mar=c(4,4,1,1)+0.1);
+
+plot(c(1,3200), c(0,0.2), type="n", xlab="Elements", ylab="Displacement");
+abline(h=axTicks(2), lty="dotted", lwd=1); # Grid lines
+
+n <- "optdkt";
+lines(n2e(hem$size[hem$name == n]), hem$d[hem$name == n], type="b", col="black", pch=1);
+
+n<- "bezier2";
+lines(n2e(hem$size[hem$name == n]), hem$d[hem$name == n], type="b", col="black", pch=4);
+
+legend("bottomright", c("DKT","BSH"), pch=c(1,4), col=c("black","black"), bg="white");
+
+dev.off()
+par(p);

### 2012-07-13 Doc/VRIPHYS2012/roof.R

@@ -1,0 +1,28 @@
+#
+# Graph: modified Scordelis-Lo roof
+#
+
+n2e <- function(x) (2*x*x); # size to element count
+
+lo <- read.csv("roof.csv");
+
+# Compute displacement
+lo$z <- 19.1511 - lo$z;
+
+# Plot to PDF
+pdf("roof.pdf", width=7, height=4);
+p <- par(lwd=3, mar=c(4,4,1,1)+0.1);
+
+plot(c(1,3200), c(0,0.5), type="n", xlab="Elements", ylab="Displacement");
+abline(h=axTicks(2), lty="dotted", lwd=1); # Grid lines
+
+n <- "optdkt";
+lines(n2e(lo$size[lo$name == n]), lo$z[lo$name == n], type="b", col="black", pch=1);
+
+n<- "bezier2";
+lines(n2e(lo$size[lo$name == n]), lo$z[lo$name == n], type="b", col="black", pch=4);
+
+legend("topright", c("DKT","BSH"), pch=c(1,4), col=c("black","black"), bg="white");
+
+dev.off()
+par(p);

### 2012-07-13 Doc/VRIPHYS2012/fps.R
fixing a figure
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@423 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@ -12,8 +12,8 @@ plot(range(fps$elements), range(c(fps$fpsBase, fps$fpsPolar, fps$fpsMap1, fps$fp
 
 lines(fps$elements, fps$fpsBase, type="b", col="black", pch=1);
 lines(fps$elements, fps$fpsPolar, type="b", col="black", pch=4);
-lines(fps$elements, fps$fpsMap2, type="b", col="black", pch=6);
-lines(fps$elements, fps$fpsMap1, type="b", col="black", pch=5);
+lines(fps$elements, fps$fpsMap2, type="b", col="black", pch=5);
+lines(fps$elements, fps$fpsMap1, type="b", col="black", pch=6);
 
 abline(h=25, lty="dotted");
 

### 2012-07-18 Doc/VRIPHYS2012/vriphys2012.tex
more on the article
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@425 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
+% For large/wide figures
+%\FigureW{width}{name}{caption}
+\newcommand{\FigureW}[3]{%
+\begin{figure*}[htb]
+  \centering
+  \includegraphics[width=#1]{#2}
+  \caption{\label{fig-#2}#3}
+\end{figure*}}
+
 \title[Short title]%
@@
 
-Objects with thin structure form a very specific field of research. Because
-they have high ratio between width and thickness methods from volumetric
-deformation modeling cannot be immediately used. Such object occur in many
-places e.g. textiles, paper or leaves. They are also very common in
-anatomical structures for example in form of tubular objects like blood
-vessels or colon, in the form of thin membranes, but also many others
-(lens, skin, \ldots).
+Objects with thin structure occur in many forms e.g. textiles, paper,
+leaves etc. Such objects are also very common in anatomical structures for
+example in the form of tubular objects like blood vessels or colon, in the
+form of thin membranes, but also many others (lens, skin, \ldots). Because
+they have high ratio between width and thickness methods for volumetric
+deformation modeling cannot be immediately used. For this reason they form
+a very specific field of research.
 
 Methods based on continuum mechanics and namely theory of elasticity gained
-on popularity ever since Terzopoulos \cite{Terzopoulos1987} presented his
-work on elastic deformation modeling. While the methods using finite
-elements and theory of elasticity are fairly popular in volumetric modeling
-\footnote{TODO: do we need ref. here?} of deformations, they are still
-frowned upon in the area of thin structures. Instead either the
-mass-spring models \cite{Volino2009} or bending models
-\cite{Grinspun2003,Choi2007} are preferred. Unfortunately they have the
-disadvantage that physical parameters of a material (like Young's modulus
-and Poisson's ratio) cannot be used directly.
+on popularity in computer graphics and real-time simulators ever since
+Terzopoulos \cite{Terzopoulos1987} presented his work on elastic
+deformation modeling. While the methods using finite elements and theory of
+elasticity are fairly popular in volumetric modeling \footnote{TODO: do we
+need ref. here?} of deformations, they are still frowned upon in the area
+of thin structures. Instead either the mass-spring models \cite{Volino2009}
+or bending models \cite{Grinspun2003,Choi2007} are preferred. Unfortunately
+they have the disadvantage that physical parameters of a material (like
+Young's modulus and Poisson's ratio) cannot be used directly.
 
-The usual argument against FEM based methods being that they are too slow.
-This is indeed true if lots of elements are used to discretize the object.
-However, exploiting the bending property of the shell elements allows us to
-achieve good results even with relatively small number of elements. We can
-use high-polygonal mesh only for rendering or collision detection and
-response. Similar idea has been already presented in the work of Bouthors
-et al. \cite{Bouthors2007}. They used twinned meshes for animation of mesh
-defined by implicit surface. Simple low-polygonal mesh was used for
-mechanical FEM model and the mesh was subdivided in areas with high
-curvature after every step of the simulation to produce finer mesh for
-rendering. The underlaying FEM model is however not aware of the implicit
-surface used. We build upon the work of Comas \cite{Comas2010c} who used a
-simple cubic polynomial to describe the element geometry and derived the
-shell model from it.
+The usual argument against FEM based methods is that they are
+computationally too expensive. This is indeed true if lots of elements are
+used to discretize the object or non-linear models are used. However,
+exploiting the bending property of the shell elements allows us to achieve
+good results even with relatively small number of elements. We can use
+high-polygonal mesh only for rendering or collision detection and response.
+Similar idea has been already presented in the work of Bouthors et al.
+\cite{Bouthors2007}. They used twinned meshes for animation of mesh defined
+by implicit surface. Simple low-polygonal mesh was used for mechanical FEM
+model and the mesh was subdivided in areas with high curvature after every
+step of the simulation to produce finer mesh for rendering. The underlaying
+FEM model is however not aware of the implicit surface used. We build upon
+the work of Comas \cite{Comas2010c} who used a simple cubic polynomial to
+describe the element geometry and derived the shell model from it. Comas's
+element has several drawbacks that we needed to address. Namely it produces
+discontinuities between the elements on the edges, it is missing one
+rotational degree of freedom and is not completely symmetric in context of
+deformations.
 
-The Comas's element has several drawbacks that we needed to address. Namely
-it produces discontinuities between the elements on the edges, it is
-missing one rotational degree of freedom and is not completely symmetric in
-context of deformations.
+% TODO: integrate better to the rest of the section
+Ubach and Oñate \cite{Ubach2010} also present a shell element based on
+Bézier triangles. The element is however free from rotational degrees of
+freedom which is unsuitable for physical simulators. It also reduces the
+sparsity of stiffness matrix because of increased interconnection between
+nodes.
 
-% TODO: cite Ubach & Oñate
-TODO: cite \cite{Ubach2010}: rotation-free shell elements, also based on
-Bézier triangle
-
-In this paper we present a new shell element based on the formulation of
-Bézier triangle.
+% In this paper we present a new shell element based on the
+% formulation of Bézier triangle.
+%
+% TODO: paper structure
+TODO: paper structure
 
@@
     \item Our element solves the problems of the element presented by Comas
-        \cite{Comas2010c}.
+        \cite{Comas2010c}: it is continuous on the element boundary and has
+        all rotational degrees of freedom.
 
     % TODO: needs better explanation
-    % + two stage interpolation => Nodes (6DOFS) => bezier points => bezier
-    %   triangle
     \item Two stage interpolation: 6DOF nodes -> bezier points -> bezier
@@
     % + separation between nodes and dofs
-    \item The model separates nodes of the mesh from degrees of freedom of the
-        control mesh. This allows mapping a surface on deformable skeleton.
+    %\item The model separates nodes of the mesh from degrees of freedom of the
+    %    control mesh. This allows mapping a surface on deformable skeleton.
 
     % + mapping: smooth surface results with less shells + collision response
-    %   on smooth surface +  [? if I have time]  deformable on deformable:
-    %   conjonctiva 2D motion on the surface of the shells
-    \item Smooth maping of high resolution mesh on the curved surface of the
-        shell. The mapped surfce can be used for collision detection and
-        allows the propagation of proper collision response back onto the
-        (low resolution) mechanical mesh.
+    %   on smooth surface 
+    \item Mapping of high resolution mesh on the curved surface of the
+        shell. The smooth mapped surface can be used for collision
+        detection and allows the propagation of proper collision response
+        back onto the (low resolution) mechanical mesh.
 
@@
 
-Shell element combines two types of deformations:
+Shell element is a combination of two simpler elements describing two types
+of deformations:
 
@@
 
-    \item Elastic membrane: deformation in the plane of the element,
-        described by in-plane displacements $u_x, u_y$ and rotation
-        $\theta_z$ around the axis perpendicular to the plane of the
-        element (Figure \ref{fig-membrane}).
-
-    \item Bending plate: out of the plane deformation described by rotation
-        around two in-plane axes $\theta_x, \theta_z$ and out of plane
-        displacmenent $u_z$ (Figure \ref{fig-plate}).
-
-\end{itemize}
+    \item Elastic membrane defining deformations in plane of the
+    element. It encompasses deformations like stretching and shearing and
+    is described by 3 degrees of freedom: in-plane displacements $u_x, u_y$
+    and rotation $\theta_z$ around the axis perpendicular to the plane of
+    the element (Figure \ref{fig-membrane}).
 
@@
 
+    \item Bending plate defining out of plane bending deformation. It is
+    described by 3 degrees of freedom: rotation around two axes $\theta_x,
+    \theta_z$ and out of plane displacement $u_z$ (Figure \ref{fig-plate}).
+
+\end{itemize}
+
 \Figure{0.8\linewidth}{plate}
 {DOFs of bending plate element.}
+
+Combining both elements we make use of all 6 degrees of freedom available
+at each node of the element.
 
@@
 
-The general $n$-th order Bézier triangle is defined as:
+Bézier triangles were designed in sixties by French mathematician de
+Casteljau. As with other Bézier or B-spline surfaces the Bézier triangle is
+defined by a mesh of control points that do not necessarily lie on the
+surface. The general $n$-th order Bézier triangle requires $(n + 1)(n +
+2)/2$ control points and it's surface is defined as:
 
@@
 
-Specificaly the cubic Bézier triangle ($n=3$) used in our model is
-described by 10 control points and the surface is explicitly defined as (see
-fig. \ref{fig-bezier}):
+In our work we use cubic Bézier triangle ($n=3$). Triangle of smaller order
+would fail to describe all possible deformations of the element and
+triangles of higher order cannot be described unambiguously with only 18
+DOFs available to us without adding further constraints on the control
+mesh. Cubic triangle is described by 10 control points and the surface is
+explicitly defined as (see fig. \ref{fig-bezier}):
 
@@
   T(\xi_1,\xi_2) & =
-           + \xi_1^3\ P_1
+             \xi_1^3\ P_1
            + \xi_2^3\ P_2
@@
 \noindent
-where $ \xi_3 = 1 - \xi_1 - \xi_2 $. By naming the respective values of
-the Bernstein basis functions $n_i$ we can shortly express
-\eqref{eq-cubicbez} as:
+where $ \xi_3 = 1 - \xi_1 - \xi_2 $. By naming $n_i$ the respective values
+of the Bernstein basis functions linked to control point $P_i$ we can
+shortly express \eqref{eq-cubicbez} as:
 
@@
 \Figure{0.8\linewidth}{bezier}
-{Cubic Bézier triangle with net of 10 control points.}
+{Cubic Bézier triangle with mesh of 10 control points.}
 
@@
 
-The Bézier triangles are defined by a net of control points, however the
+The Bézier triangles are defined by a mesh of control points, however the
 simulated object is usually described by a triangular mesh. We construct
@@
 mesh. When constructing the mesh a special care has to be taken to maintain
-the continuity accross the nodes and edges. During the initialization phase
-we partialy employ the method described in \cite{Ubach2010} to maintain
+the continuity across the nodes and edges. During the initialization phase
+we partially employ the method described in \cite{Ubach2010} to maintain
 $C^0$ continuity on the edges. Each of the control points on the edge is
@@
     \item The plane that contains the curve of triangle's contour. The
-        choise is arbitrary, but necessary to maintain $C^0$ continuity. 
+        choice is arbitrary, but necessary to maintain $C^0$ continuity. 
         We choose the plane defined by the edge of the flat triangle and
@@
 
-That way the central point is slightly elevated and not in the plane of
-other nodes thus keeping the curvature of the element.
+That way if the element is not flat the central point is slightly elevated
+and not in the plane of other nodes thus keeping the curvature of the
+element.
 
@@
 \ref{sec-corot}. We also present results including the mapping of
-high-resolution mesh onto the physical mesh for visualisation purposes or
+high-resolution mesh onto the mechanical mesh for visualisation purposes or
 collisions.
@@
 \ref{sec-corot} and without it (A), and with high-resolution mesh of 2048
-(C) or 3200 (D) triangles mapped onto the physical mesh.}
+(C) or 3200 (D) triangles mapped onto the mechanical mesh.}
 
@@
 
+\subsection{Applications}
+
+The presented element has been successfully implemented in SOFA framework
+\cite{Faure2012} for real-time surgical simulations. It has been evaluated
+on several applications.
+
+% TODO: !!! When ANONYMISING do something with the citation !!!
+One being a tool for pre-operative planning of corrective surgery of
+congenital heart diseases in infants \cite{Kislinskiy2012}. The cardiac
+surgery is a complicated process where several arterial vessels have to be
+cut, patched and reconnected. The element has been used for to model
+deformations of blood vessels after suturing and/or applying a artificial
+patch. Figure \ref{fig-chd} presents a screenshots of the application
+prototype.
+
+\FigureW{0.75\linewidth}{chd}
+{Screenshots of the simulation system prototype for different surgical
+procedures used to repair a coarctation of an aortic arch. The screenshots
+are enriched with overlay delineations.}
+
+Another designed application is simulation of women reproductive organs
+(vagaina/uterus/\ldots). 
+The simulation is so as to evaluate
+stress on a tissue during a birth.
+
 % }}}
@@
 \section{Discussion and Limitations} % {{{
-
-The element has been successfully tested in SOFA framework for surgical
-simulations. The applications being for example simulation pre-operative
-planning of cardiac surgery on arterial vessels \cite{Kislinskiy2012}, or
-{\em something, something} of vagina/uterus.
-
-% = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = = =
-%
-% TODO
-% Other Applications:
-% + cataract surgery
-%
 
@@
 necessary while the Bézier triangle has theoretically $ 10 \times
-3~\mathrm{DOFs} = 30~\mathrm{DOFs}$.
+3~\mathrm{DOFs} = 30~\mathrm{DOFs}$. For complex elements with high number
+of nodes the interconnection between nodes increases, especially on the
+vertices where the size of nodes involved depends also on number of
+neighbouring elements. By keeping the element simple the sparsity of
+stiffness matrix is not violated.
 
@@
     corners of the element. This is essential for good visual experience if
-    high-resolution mesh is mapped on the physical mesh.
+    high-resolution mesh is mapped on the mechanical mesh.
 
@@
 better visualisation. We also provide a mapping of forces and velocities
-between simulation mesh and physical mesh for collision detection/response
+between simulation mesh and mechanical mesh for collision detection/response
 and interaction with the mesh which are essential component for the surgical

### 2012-07-18 shells2/forcefield/BezierShellForceField.h
using 6 Gauss point quadrature for bending element computing interpolation frame at center using interpolation component
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@427 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@ -171,6 +171,7 @@ public:
                 StrainDisplacement strainDisplacementMatrix4;
 
                 StrainDisplacement strainDisplacementMatrix[Gn];
+                StrainDisplacement strainDisplacementMatrixB[Gn];
 
                 // the strain-displacement matrices at each Gauss point
                 StrainDisplacementBending strainDisplacementMatrixB1;

### 2012-07-18 Doc/VRIPHYS2012/vriphys2012.tex
modifs in intro
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@428 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \title[Short title]%
-      {Fill in the Title} % TODO
+      {Bezier shell finite element for interactive medical simulation} % TODO
 
@@
 
-Objects with thin structure occur in many forms e.g. textiles, paper,
-leaves etc. Such objects are also very common in anatomical structures for
-example in the form of tubular objects like blood vessels or colon, in the
-form of thin membranes, but also many others (lens, skin, \ldots). Because
-they have high ratio between width and thickness methods for volumetric
-deformation modeling cannot be immediately used. For this reason they form
-a very specific field of research.
+Thin objects are very frequent in our every-day life. % thin wall objects ?
+%Objects with thin structure occur in many forms 
+e.g. textiles, paper, leaves etc. 
+Such objects are also very common in anatomical or pathological structures for example tubular structures (like blood vessels, colon \ldots), thin bag structures (%vessie, 
+Glisson capsule, aneurysms,  \ldots), and also many others (eyes, skin, \ldots). 
+In the field of surgical simulation, we try to reproduce or anticipate the mechanical behavior of these structures during the surgery.
+Thus, it is very important to capture accurately the physical behavior of these structure.
+However, in the same time, it is also very important to maintain fast computation rates, in order to be compatible with the targeted applications (interactive simulation for education and for planning).
 
-Methods based on continuum mechanics and namely theory of elasticity gained
-on popularity in computer graphics and real-time simulators ever since
-Terzopoulos \cite{Terzopoulos1987} presented his work on elastic
-deformation modeling. While the methods using finite elements and theory of
-elasticity are fairly popular in volumetric modeling \footnote{TODO: do we
-need ref. here?} of deformations, they are still frowned upon in the area
-of thin structures. Instead either the mass-spring models \cite{Volino2009}
-or bending models \cite{Grinspun2003,Choi2007} are preferred. Unfortunately
-they have the disadvantage that physical parameters of a material (like
-Young's modulus and Poisson's ratio) cannot be used directly.
+This problem has been faced by the computer graphics community but for different applications.
+For instance, many related work faces the problem of the simulation of cloth using mass-spring models \cite{Volino2009} or bending models \cite{Grinspun2003,Choi2007}. %% Todo verify.... 
+However, these models are discrete so their behavior depends on the mesh and their parameters are not easily measurable.
+This is completely acceptable for some computer graphics application, where the parameters can be tuned to obtain a realistic animation.
+In the case of medical simulation, more physics-based modeling is needed.
 
-The usual argument against FEM based methods is that they are
-computationally too expensive. This is indeed true if lots of elements are
-used to discretize the object or non-linear models are used. However,
-exploiting the bending property of the shell elements allows us to achieve
-good results even with relatively small number of elements. We can use
-high-polygonal mesh only for rendering or collision detection and response.
-Similar idea has been already presented in the work of Bouthors et al.
-\cite{Bouthors2007}. They used twinned meshes for animation of mesh defined
-by implicit surface. Simple low-polygonal mesh was used for mechanical FEM
-model and the mesh was subdivided in areas with high curvature after every
-step of the simulation to produce finer mesh for rendering. The underlaying
-FEM model is however not aware of the implicit surface used. We build upon
-the work of Comas \cite{Comas2010c} who used a simple cubic polynomial to
-describe the element geometry and derived the shell model from it. Comas's
-element has several drawbacks that we needed to address. Namely it produces
-discontinuities between the elements on the edges, it is missing one
-rotational degree of freedom and is not completely symmetric in context of
-deformations.
+Methods based on continuum mechanics and namely theory of elasticity gained on popularity in computer graphics and real-time simulators ever sinceTerzopoulos \cite{Terzopoulos1987} presented his work on elastic deformation modeling.
+%% Do we
+While the methods using finite elements and theory of elasticity are fairly popular in volumetric modeling \footnote{TODO: do we need ref.
+here?
+yes but for medical applications} of deformations, they are still frowned upon in the area of thin structures.
+Yet, a special field of continuum mechanics has studied the deformations of these objects, and has proposed models based on shell theory.
+This theory can be integrated using Finite Elements, but there is two drawbacks: (i) the continuum equations requires square spatial derivations that can not be handled with linear interpolation and (ii) the final system of equations that describes the mechanics is non-linear (if the deformations are not small).
 
-% TODO: integrate better to the rest of the section
-Ubach and Oñate \cite{Ubach2010} also present a shell element based on
-Bézier triangles. The element is however free from rotational degrees of
-freedom which is unsuitable for physical simulators. It also reduces the
-sparsity of stiffness matrix because of increased interconnection between
-nodes.
+The usual argument against non-linear FEM based methods is that they are computationally too expensive. 
+This is indeed true if lots of elements are used to discretize the object.
+A classical strategy, is to use high-polygonal mesh for rendering or collision detection and response, while maintaing a reasonably low number of elements for the mechanics.
+The work of Bouthors et al.\cite{Bouthors2007} is a good application example of this strategy for surface deformations.
+But this strategy is not sufficient in our case because the problem of performance is also greatly impacted when using elements that have an interpolation with high order polynoms.
+This type of interpolation is necessary because bending continuum model (based on Kirchhoff-Love theory for thin plates)  requires the computation of square spatial derivatives.
+In \cite{Comas2010c}, Comas and al.  use an hybrid interpolation: linear interpolation is used for in-plane deformations and cubic polynomials for bending deformations.
+The element is piloted by 3 nodes on which 6 degrees of freedom are defined. 
+From this point of view, our work is derived from this approach but we address several noted drawbacks: the formulation misses one rotational degree of freedom, the cubic interpolation is not symmetric and produces $C^0$  discontinuities between elements on edges.
+Ubach and al. \cite{Ubach2010}  present a shell element based on B\'ezier triangles that allows for a better behavior.
+However, the B\'ezier interpolation defined on each triangle element is based on the motion computed on adjacent triangles.
+So the stiffness forces computed on each element not only depends on nodes of the element but also on all the adjacent nodes. 
+This greatly reduces the sparsity of the global stiffness matrix and consequently increases the computation time.
 
-% In this paper we present a new shell element based on the
+hell element based on the
 % formulation of Bézier triangle.

### 2012-07-19 Doc/VRIPHYS2012/vriphys2012.tex
reorganizing sections 2 and 3, adding more stuff to sec. 2
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@429 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-Methods based on continuum mechanics and namely theory of elasticity gained on popularity in computer graphics and real-time simulators ever sinceTerzopoulos \cite{Terzopoulos1987} presented his work on elastic deformation modeling.
+Methods based on continuum mechanics and namely theory of elasticity gained on popularity in computer graphics and real-time simulators ever since Terzopoulos \cite{Terzopoulos1987} presented his work on elastic deformation modeling.
 %% Do we
@@
 Yet, a special field of continuum mechanics has studied the deformations of these objects, and has proposed models based on shell theory.
-This theory can be integrated using Finite Elements, but there is two drawbacks: (i) the continuum equations requires square spatial derivations that can not be handled with linear interpolation and (ii) the final system of equations that describes the mechanics is non-linear (if the deformations are not small).
+This theory can be integrated using Finite Elements, but there are two drawbacks: (i) the continuum equations requires second spatial derivations that can not be handled with linear interpolation and (ii) the final system of equations that describes the mechanics is non-linear (if the deformations are not small).
 
@@
 But this strategy is not sufficient in our case because the problem of performance is also greatly impacted when using elements that have an interpolation with high order polynoms.
-This type of interpolation is necessary because bending continuum model (based on Kirchhoff-Love theory for thin plates)  requires the computation of square spatial derivatives.
+This type of interpolation is necessary because bending continuum model
+(based on Kirchhoff-Love theory for thin plates)  requires the computation of second spatial derivatives.
 In \cite{Comas2010c}, Comas and al.  use an hybrid interpolation: linear interpolation is used for in-plane deformations and cubic polynomials for bending deformations.
@@
 
-hell element based on the
-% formulation of Bézier triangle.
-%
 % TODO: paper structure
@@
 
-\section{Preliminaries} % {{{
+\section{Element Geometry} % {{{
 
-\subsection{Shell Element}
-
-Shell element is a combination of two simpler elements describing two types
-of deformations:
-
-\begin{itemize}
-
-    \item Elastic membrane defining deformations in plane of the
-    element. It encompasses deformations like stretching and shearing and
-    is described by 3 degrees of freedom: in-plane displacements $u_x, u_y$
-    and rotation $\theta_z$ around the axis perpendicular to the plane of
-    the element (Figure \ref{fig-membrane}).
-
-\Figure{0.8\linewidth}{membrane}
-{DOFs of elastic membrane element.}
-
-    \item Bending plate defining out of plane bending deformation. It is
-    described by 3 degrees of freedom: rotation around two axes $\theta_x,
-    \theta_z$ and out of plane displacement $u_z$ (Figure \ref{fig-plate}).
-
-\end{itemize}
-
-\Figure{0.8\linewidth}{plate}
-{DOFs of bending plate element.}
-
-Combining both elements we make use of all 6 degrees of freedom available
-at each node of the element.
-
-\subsection{Bézier Triangle} % {{{
+\subsection{B\'ezier Triangle} % {{{
 
@@
 
-Bézier triangles were designed in sixties by French mathematician de
-Casteljau. As with other Bézier or B-spline surfaces the Bézier triangle is
+B\'ezier triangles were designed in sixties by French mathematician de
+Casteljau. As with other B\'ezier or B-spline surfaces the B\'ezier triangle is
 defined by a mesh of control points that do not necessarily lie on the
-surface. The general $n$-th order Bézier triangle requires $(n + 1)(n +
+surface. The general $n$-th order B\'ezier triangle requires $(n + 1)(n +
 2)/2$ control points and it's surface is defined as:
@@
 
-In our work we use cubic Bézier triangle ($n=3$). Triangle of smaller order
+In our work we use cubic B\'ezier triangle ($n=3$). Triangle of smaller order
 would fail to describe all possible deformations of the element and
@@
 \Figure{0.8\linewidth}{bezier}
-{Cubic Bézier triangle with mesh of 10 control points.}
+{Cubic B\'ezier triangle with mesh of 10 control points.}
 
@@
 
-% }}}
-
-\section{Bézier Shell Element} % {{{
-
 \subsection{Control Points} % {{{
 
-The Bézier triangles are defined by a mesh of control points, however the
-simulated object is usually described by a triangular mesh. We construct
-the Bézier mesh by taking into account the curvature of the triangular
-mesh. When constructing the mesh a special care has to be taken to maintain
+The B\'ezier triangles are defined by a mesh of control points, however the
+simulated object is usually described by a triangular mesh with 6 degrees of freedom per node (position \& rotation).
+We construct
+the initial B\'ezier mesh by taking into account the curvature of the triangular
+mesh. Note that special care has to be taken to maintain
 the continuity across the nodes and edges. During the initialization phase
@@
 
-We do not repeat this procedure during the simulation. For faster update of
-the control points we rigidly attach the edge points to corner points. We
-remember the position and orientation of the edge point relative to the
-nearest corner point and use this information together with the rotation on
-the corner points to reconstruct new position of control points. See Figure
+We do not repeat this procedure during the simulation. For fast update of
+the control points we rigidly attach each edge point to nearest corner point (see Figure
 \ref{fig-segments} for the correspondence between edge points and
-associated corner points.
+associated corner points).
 
@@
 
+In the initialization phase we remember the position of edge point relative to the position and orientation of the corner node:
+
+\begin{equation}
+    \overline{P_{i-j}} = (R_i^0)^T P_{i-j}^0
+\end{equation}
+
+\noindent
+where $P_{i-j}^0 = P_j^0 - P_i^0$, with $j \in [4;10]$ being index of the edge node and $i \in [1;3]$ the index of associated corner node, is the segment at the beginning of the simulation and $R_i^0$ the orientation of node $P_i$.
+
+In each step of the simulation we recompute the new position based on 
+current orientation of the corner node:
+
+\begin{equation}
+    P_j = P_i + P_{i-j} = P_i + R_i \overline{P_{i-j}}
+\end{equation}
+
+\noindent
+where $P_i$ is the current position of corner node and $R_i$ its orientation.
+
 $C^1$ continuity on the edges can also be maintained if special care is taken
 when computing the position of the central point (see \cite{Ubach2010}). For
-simplicity we compute the central point followingly:
+simplicity we compute the central point like this:
 
@@
 
-% TODO - nodes + DOFs (local + global position)
+\subsection{Deflection function} % {{{
 
-\subsection{Mechanics} % {{{
+% TODO: we need to justify how we decide to move the control points
 
-Our definition of the displacement field $u$ is based on the formulation of the
-Bézier triangle \eqref{eq-cubicbez}:
+We recall the equation for central node \eqref{eq-central} and
+considering how are the edge points of control mesh connected to corner points with
+segments $P_{i-j}$:
 
-\begin{equation}\label{eq-bezU}
+\begin{equation}
+    P_{i-j} = P_j - P_i, \quad j \in [4;9], i \in [1;3].
+\end{equation}
+
+We now add small variation to the position ($U_1, U_2, U_3$) and orientation ($\theta_1, \theta_2, \theta_3$) of corner nodes to the equation of the element \eqref{eq-cubicbez}:
+
+\begin{equation}
     \begin{split}
-  u & = U_1 ( \xi_1^3 + 3 \xi_1^2 \xi_2 + 3 \xi_1^2 \xi_3
-        + 2 \xi_1 \xi_2 \xi_3 ) \\
-    & + U_2 ( \xi_2^3 + 3 \xi_1 \xi_2^2 + 3 \xi_2^2 \xi_3
-        + 2 \xi_1 \xi_2 \xi_3 ) \\
-    & + U_3 ( \xi_3^3 + 3 \xi_1 \xi_3^2 + 3 \xi_2 \xi_3^3
-        + 2 \xi_1 \xi_2 \xi_3 ) \\
-    & + 3 \xi_1^2 \xi_2 (P_{4-1} \times \theta_1) \\
-    & + 3 \xi_1^2 \xi_3 (P_{5-1} \times \theta_1) \\
-    & + 3 \xi_1 \xi_2^2 (P_{7-2} \times \theta_2) \\
-    & + 3 \xi_2^2 \xi_3 (P_{6-2} \times \theta_2) \\
-    & + 3 \xi_1 \xi_3^2 (P_{8-3} \times \theta_3) \\
-    & + 3 \xi_2 \xi_3^2 (P_{9-3} \times \theta_3) \\
-    & + 2 \xi_1 \xi_2 \xi_3 (P_{10-1} \times \theta_1 +
-        P_{10-2} \times \theta_2 +
-        P_{10-3} \times \theta_3)
+        & n_1 (P_1 + U_1)
+        + n_2 (P_2 + U_2)
+        + n_3 (P_3 + U_3) \\
+        & + n_4 (P_1 + U_1 + P_{1-4} + P_{1-4} \times \theta_1)
+        + n_5 (P_1 + U_1 + P_{1-5} + P_{1-5} \times \theta_1) \\
+        & + n_6 (P_2 + U_2 + P_{2-6} + P_{2-6} \times \theta_2)
+        + n_7(P_2 + U_2 + P_{2-7} + P_{2-7} \times \theta_2) \\
+        & + n_8 (P_3 + U_3 + P_{3-8} + P_{3-8} \times \theta_3)
+        + n_9 (P_3 + U_3 + P_{3-9} + P_{3-9} \times \theta_3) \\
+        & + n_{10} \frac{1}{3} \left(
+        \sum_{j=4}^9 (P_i + U_i + P_{i-j} + P_{i-j} \times \theta_i) - \sum_{i=1}^3 (P_i + U_i)
+        \right) \\
+%
+        & = 
+    \sum_{i=1}^{10} n_i P_i
+      + n_1 U_1
+      + n_2 U_2
+      + n_3 U_3 \\
+      & + n_4 (U_1 + P_{1-4} \times \theta_1)
+      + n_5 (U_1 + P_{1-5} \times \theta_1)
+      + n_6 (U_2 + P_{2-6} \times \theta_2) \\
+      & + n_7 (U_2 + P_{2-7} \times \theta_2)
+      + n_8 (U_3 + P_{3-8} \times \theta_3)
+      + n_9 (U_3 + P_{3-9} \times \theta_3) \\
+      & + n_{10} \frac{1}{3} (U_1 + U_2 + U_3 + \sum_{j=4}^9 P_{i-j} \times \theta_i) \\
+%
+      & = T + U_1 (n_1 + n_4 + n_5 + \frac{1}{3} n_{10}) \\
+      & + U_2 (n_2 + n_6 + n_7 + \frac{1}{3} n_{10}) \\
+      & + U_3 (n_3 + n_8 + n_9 + \frac{1}{3} n_{10}) \\
+      & + \sum_{j=4}^9 n_j P_{i-j} \times \theta_i
+      + n_{10} \frac{1}{3} \sum_{j=4}^9 P_{i-j} \times \theta_i \\
+%
+      & = T + u
   \end{split}
@@
 \noindent
-where $U_1, U_2, U_3$ are the translational displacment vectors of the
-corner nodes of the element, $\theta_1, \theta_2, \theta_3$ are the angular
-displacements of the corner nodes and $P_{i-j}$ are the segments from
-corner nodes to other nodes of the control mesh:
+thus computing the function of deflection $u$.
 
-\begin{equation}
-    P_{i-j} = P_j-P_i %, \quad i \in [4;10], j \in [1;3]
-\end{equation}
+We have simplified the notation by leaving the variable $i$ free in the above sums involving $P_{i-j}$.
+Note however that the expressions are sound and unambiguous because there is only one possible $i$ for each $j$ depending on the connection of segments to the corner nodes (see fig.
+\ref{fig-segments}).
+
+% }}}
+
+% }}}
+
+\section{...} % {{{
+
+\subsection{Shell Element}
+
+Shell element is a combination of two simpler elements describing two types
+of deformations:
+
+\begin{itemize}
+
+    \item Elastic membrane defining deformations in plane of the
+    element. It encompasses deformations like stretching and shearing and
+    is described by 3 degrees of freedom: in-plane displacements $u_x, u_y$
+    and rotation $\theta_z$ around the axis perpendicular to the plane of
+    the element (Figure \ref{fig-membrane}).
+
+\Figure{0.8\linewidth}{membrane}
+{DOFs of elastic membrane element.}
+
+    \item Bending plate defining out of plane bending deformation. It is
+    described by 3 degrees of freedom: rotation around two axes $\theta_x,
+    \theta_z$ and out of plane displacement $u_z$ (Figure \ref{fig-plate}).
+
+\end{itemize}
+
+\Figure{0.8\linewidth}{plate}
+{DOFs of bending plate element.}
+
+Combining both elements we make use of all 6 degrees of freedom available
+at each node of the element.
+
+\subsection{Mechanics} % {{{
+
+...
 
@@
 We use the fact that the geometry of the shell is based on the formulation
-of cubic Bézier triangle \eqref{eq-cubicbez}. For every vertex of the high
+of cubic B\'ezier triangle \eqref{eq-cubicbez}. For every vertex of the high
 resolution mesh we first find the corresponding triangle on the mechanical
@@
 In next two subsections we present tests to validate the convergence of our
-Bézier shell element. The results are compared with the shell element
+B\'ezier shell element. The results are compared with the shell element
 composed of DKT element and optimal ANDES element \cite{Felippa2003}. The
@@
 {Convergence for modified Scordelis-Lo roof test of DKT+ANDES element and
-our Bézier shell element.}
+our B\'ezier shell element.}
 
@@
 {Convergence for test on hemisphere with 18° hole for the DKT+ANDES element and
-our Bézier shell element.}
+our B\'ezier shell element.}
 
@@
 and simplicity. Only $ 3 \times 6~\mathrm{DOFs} = 18~\mathrm{DOFs}$ are
-necessary while the Bézier triangle has theoretically $ 10 \times
+necessary while the B\'ezier triangle has theoretically $ 10 \times
 3~\mathrm{DOFs} = 30~\mathrm{DOFs}$. For complex elements with high number
@@
 This paper presents new shell element with geometry and displacement
-function based on the formulation of Bézier triangle. It fixes problems
+function based on the formulation of B\'ezier triangle. It fixes problems
 of the previous model \cite{Comas2010c}, namely:
@@
 
-Because of the formulation of Bézier triangle that is intrinsically present
+Because of the formulation of B\'ezier triangle that is intrinsically present
 there is a clean and obvious way of mapping high-resolution mesh for
@@
 
-% vim: fdm=marker tw=75
+% vim: fdm=marker

### 2012-07-20 Doc/VRIPHYS2012/vriphys2012.tex
modifs & comments
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@430 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
+%%% ADD COMMENTS=
+\usepackage{color} 
+\newcommand{\CD}[1]{{\color{magenta}{\textbf{C: #1}}}}
+
 \title[Short title]%
-      {Bezier shell finite element for interactive medical simulation} % TODO
+      {Bezier shell finite element for interactive surgical simulation} % TODO
 
@@
 This problem has been faced by the computer graphics community but for different applications.
-For instance, many related work faces the problem of the simulation of cloth using mass-spring models \cite{Volino2009} or bending models \cite{Grinspun2003,Choi2007}. %% Todo verify.... 
+For instance, many related work addresses the problem of the simulation of cloth using mass-spring models \cite{Volino2009} or bending models \cite{Grinspun2003,Choi2007}. %% Todo verify.... 
 However, these models are discrete so their behavior depends on the mesh and their parameters are not easily measurable.
-This is completely acceptable for some computer graphics application, where the parameters can be tuned to obtain a realistic animation.
+This is completely acceptable for some computer graphics application, where the parameters can be tuned manually to obtain a realistic animation.
 In the case of medical simulation, more physics-based modeling is needed.
@@
 Yet, a special field of continuum mechanics has studied the deformations of these objects, and has proposed models based on shell theory.
-This theory can be integrated using Finite Elements, but there are two drawbacks: (i) the continuum equations requires second spatial derivations that can not be handled with linear interpolation and (ii) the final system of equations that describes the mechanics is non-linear (if the deformations are not small).
+This theory was integrated numerically \CD{Todo citation of a shell book} using Finite Elements, but there are two issues for obtaining interactive frame rates: 
+(i) the continuum equations requires second spatial derivations that can not be handled with linear interpolation and 
+(ii) the final system of equations that describes the mechanics is non-linear (if the deformations are not small).
 
@@
 This is indeed true if lots of elements are used to discretize the object.
-A classical strategy, is to use high-polygonal mesh for rendering or collision detection and response, while maintaing a reasonably low number of elements for the mechanics.
+A classical strategy, is to use high-polygonal meshes for rendering or collision detection and response, while maintaing a reasonably low number of elements for the mechanics.
 The work of Bouthors et al.\cite{Bouthors2007} is a good application example of this strategy for surface deformations.
@@
 
-\section{Element Geometry} % {{{
+\section{Element kinematics} % {{{
 
-\subsection{B\'ezier Triangle} % {{{
+\CD{intro: we propose an interpolation with a high degree of derivability while maintaining a reasonably low number of degrees of Freedom.
+We introduce a new way of building the interpolation by using two stages: 
+(i) a B\'ezier interpolation is defined with cubic polynomial functions 
+(ii) a kinematic relation between nodes (6DOFs) defined at the vertices and the control points of the B\'ezier Triangle}
+
+\subsection{B\'ezier Triangle Interpolation} % {{{
 
@@
 
-B\'ezier triangles were designed in sixties by French mathematician de
-Casteljau. As with other B\'ezier or B-spline surfaces the B\'ezier triangle is
+B\'ezier triangles were designed in sixties by French mathematician de Casteljau.  % SURE ? What about Bezier himself ?
+As with other B\'ezier or B-spline surfaces the B\'ezier triangle is
 defined by a mesh of control points that do not necessarily lie on the
@@
 
-\subsection{Control Points} % {{{
+\subsection{Kinematic between Nodes and Control Points } % {{{
 
@@
 simulated object is usually described by a triangular mesh with 6 degrees of freedom per node (position \& rotation).
-We construct
-the initial B\'ezier mesh by taking into account the curvature of the triangular
+%
+\CD{I would not say that it is "usual"... => I would say that 10 points x 3 Dofs  = 30 Dofs for a bezier triangle $\rightarrow$ it's quite a lot ! 
+We propose to reduce the degrees of freedom by using 3 nodes with 6 dofs = 18 Dofs }
+%
+We constructthe initial B\'ezier mesh by taking into account the curvature of the triangular
 mesh. Note that special care has to be taken to maintain
@@
 $C^0$ continuity on the edges. Each of the control points on the edge is
-computed as the intersection of:
+computed as the intersection of (\CD{Need to do a figure of this}):
 
@@
 
-We do not repeat this procedure during the simulation. For fast update of
+We do not repeat this procedure during the simulation. 
+\CD{Why ?... do we argue that it is only for having fast update ?  } 
+For fast update of
 the control points we rigidly attach each edge point to nearest corner point (see Figure
@@
 {Correspondence between edge points and associated corner points shown in
-blue.}
+blue. \CD{Not clear !!}}
 
@@
 current orientation of the corner node:
+\CD{Other possible explanation: we apply the rigid transformation of frame $i$ on control point $j$ }
 
@@
 
+\CD{To be consitent with what I write in following, we could introduce a notation for the dofs of the nodes:  $\mathbf{x}_i$ for the position and $\theta_i$ for the rotation of the node $i$
+Then, the previous equation would be:
+\begin{equation}
+    P_j = \mathbf{x}_i + P_{i-j} = P_i + R(\theta_i) \overline{P_{i-j}}
+\end{equation}
+where $ R(\theta_i)$ is the rotation matrix for the angular position of node $i$...
+}
+
 \noindent
@@
 $C^1$ continuity on the edges can also be maintained if special care is taken
-when computing the position of the central point (see \cite{Ubach2010}). For
-simplicity we compute the central point like this:
+when computing the position of the central point (see \cite{Ubach2010}). 
+For simplicity we compute the central point like this: 
+\CD{Here it is not clear that with this formulation we do not maintain $C^1$ Continuity...}
 
@@
 % TODO: we need to justify how we decide to move the control points
+
+\CD{It's more about derivatives of the kinematic relation, right ?}
 
@@
 
+\CD{I would use the following notation: small variation of the node $j$ position:   $\delta \mathbf{x}_j$ for the translation and $\delta \boldsymbol{\theta}_j$ for the rotation.
+This small variation of the position creates a small variation of the control points $i$ position $\delta P_i$ that are on the edges:
+$$
+\delta P_i =  \delta \mathbf{x}_j +  P_{i-j} \times \delta \boldsymbol{\theta}_j
+$$
+ Variation of node $10$ will be:
+$$
+\delta P_{10} =  \sum_{j=1}^3  \frac{1}{3}(\delta \mathbf{x}_j +  P_{10-i} \times \delta \boldsymbol{\theta}_i)
+$$
+%
+Thus, finally, it will creates a variation of the point position on the surface $\delta T$...
+$$
+\delta T =   \sum_{i=1}^{10} n_i  \delta P_i
+$$
+which makes:
+\begin{equation}
+    \begin{split}
+\delta T =  (n_1 + n_4 + n_5 + \frac{1}{3} n_{10})  \delta \mathbf{x}_1 \\
+ + (n_2 + n_6 + n_7 + \frac{1}{3} n_{10})  \delta \mathbf{x}_2 \\
+ + (n_3 + n_8 + n_9 + \frac{1}{3} n_{10})  \delta \mathbf{x}_3 \\
++  \sum_{i=4}^9  \sum_{j=1}^3 n_i P_{i-j} \times \delta \theta_j \\
+ +  n_{10} \frac{1}{3} \sum_{j=1}^3 P_{10-j} \times \theta_j
+  \end{split}
+\end{equation}
+%
+}
+
 \noindent
-thus computing the function of deflection $u$.
+thus computing the function of deflection $u$ 
+\CD{ or $\delta T$ }.
 

### 2012-07-25 Doc/VRIPHYS2012/vriphys2012.tex
more on article
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@431 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{abstract} % TODO
-   The ABSTRACT is to be in fully-justified italicized text, 
-   between two horizontal lines,
-   in one-column format, 
-   below the author and affiliation information. 
-   Use the word ``Abstract'' as the title, in 9-point Times, boldface type, 
-   left-aligned to the text, initially capitalized. 
-   The abstract is to be in 9-point, single-spaced type.
-   The abstract may be up to 3 inches (7.62 cm) long. \\
-   Leave one blank line after the abstract, 
-   then add the subject categories according to the ACM Classification Index 
-   (see http://www.acm.org/class/1998/).
+    TODO: Abstract.
 
-% TODO
 \begin{classification} % according to http://www.acm.org/class/1998/
-\CCScat{Computer Graphics}{I.3.3}{Picture/Image Generation}{Line and curve generation}
+    \CCScat{Computer Graphics}{I.3.5}{Computational Geometry and Object Modeling}{Physically based modelling};
+    \CCScat{Computer Graphics}{I.3.7}{Three-Dimensional Graphics and Realism}{Animation}.
+    %\CCScat{Simulation and Modeling}{I.6.8}{Types of Simulation}{Animation}
+    %\CCScat{Life and Medical Sciences}{J.3}{Health}{}
 \end{classification}
@@
 %% Do we
-While the methods using finite elements and theory of elasticity are fairly popular in volumetric modeling \footnote{TODO: do we need ref.
-here?
-yes but for medical applications} of deformations, they are still frowned upon in the area of thin structures.
+While the methods using finite elements and theory of elasticity are fairly popular in volumetric modeling of deformations \cite{Cotin1999,Nesme2006,Miller2007}, they are still frowned upon in the area of thin structures.
 Yet, a special field of continuum mechanics has studied the deformations of these objects, and has proposed models based on shell theory.
@@
 
-\section{Element kinematics} % {{{
+\section{Element Kinematics} % {{{
 
-\CD{intro: we propose an interpolation with a high degree of derivability while maintaining a reasonably low number of degrees of Freedom.
+In this subsection we propose an element featuring interpolation with high degree of flexibility while still maintaining low number of degrees of freedom.
 We introduce a new way of building the interpolation by using two stages:
-(i) a B\'ezier interpolation is defined with cubic polynomial functions 
-(ii) a kinematic relation between nodes (6DOFs) defined at the vertices and the control points of the B\'ezier Triangle}
+\begin{enumerate}
+    \item A B\'ezier interpolation with cubic polynomial functions is defined on a triangle.
+    \item A kinematic relation between nodes (6DOFs) defined at the vertices and the control points of the B\'ezier triangle.
+\end{enumerate}
 
@@
 
-B\'ezier triangles were designed in sixties by French mathematician de Casteljau.  % SURE ? What about Bezier himself ?
+B\'ezier triangles were first designed in early sixties by French mathematician de Casteljau.
 As with other B\'ezier or B-spline surfaces the B\'ezier triangle is
@@
 
-\subsection{Kinematic between Nodes and Control Points } % {{{
+\subsection{Initial Construction of Control Points} % {{{
 
-The B\'ezier triangles are defined by a mesh of control points, however the
-simulated object is usually described by a triangular mesh with 6 degrees of freedom per node (position \& rotation).
-%
-\CD{I would not say that it is "usual"... => I would say that 10 points x 3 Dofs  = 30 Dofs for a bezier triangle $\rightarrow$ it's quite a lot ! 
-We propose to reduce the degrees of freedom by using 3 nodes with 6 dofs = 18 Dofs }
-%
-We constructthe initial B\'ezier mesh by taking into account the curvature of the triangular
+The cubic B\'ezier triangles are defined by a mesh of 10 control points each, however the
+simulated object is usually described by a triangular mesh possibly with normals.
+A procedure for construction of initial Bézier mesh is necessary.
+We construct the initial B\'ezier mesh by taking into account the curvature of the triangular
 mesh. Note that special care has to be taken to maintain
-the continuity across the nodes and edges. During the initialization phase
+the continuity between elements across the nodes and edges, 
 we partially employ the method described in \cite{Ubach2010} to maintain
-$C^0$ continuity on the edges. Each of the control points on the edge is
+$C^0$ continuity. Each of the control points on the edge is
 computed as the intersection of (\CD{Need to do a figure of this}):
@@
 
-We do not repeat this procedure during the simulation. 
-\CD{Why ?... do we argue that it is only for having fast update ?  } 
-For fast update of
-the control points we rigidly attach each edge point to nearest corner point (see Figure
+$C^1$ continuity between elements on the edges can also be maintained if special care is taken
+when computing the position of the central point
+(see \cite{Ubach2010} or refer to the work of Farin \cite{Farin2002} for continuity conditions). 
+For simplicity we decided to keep the edge boundary only $C^0$ continuous and we compute the central point as a function of other 9 points:
+
+\begin{equation}\label{eq-central}
+    P_{10} = \frac{1}{3}(\sum_{i=4}^9 P_i - \sum_{i=1}^3 P_i)
+\end{equation}
+
+\noindent
+where the points $P_i$ are named as in Figure \ref{fig-bezier}.
+
+% TODO: explain the equation eq-central? would require an image
+
+That way if the element is not flat the central point is slightly elevated
+and not in the plane of other points thus keeping the curvature of the
+element.
+
+% }}}
+
+\subsection{Kinematic Between Nodes and Control Points } % {{{
+
+With 10 control points and 3 DOFs per point it is a total of 30 DOFs for triangle.
+This puts ... blah blah.
+We propose a reduction of the number of degrees of freedom by using 3 nodes with 6 degrees of freedom
+by creating a kinematic link between the control points.
+Each of the edge control points is attached to nearest corner node. See Figure
 \ref{fig-segments} for the correspondence between edge points and
-associated corner points).
+associated corner nodes. 
+In the following text the index $j \in [4;10]$ refers to edge points and index $i \in [1;3]$ to the associated corner point/node.
 
@@
 
-In the initialization phase we remember the position of edge point relative to the position and orientation of the corner node:
+In the initialization phase we remember the position of edge point $j$ in the frame of corner point $i$:
 
 \begin{equation}
-    \overline{P_{i-j}} = (R_i^0)^T P_{i-j}^0
+    \overline{P_{i-j}} = R(\theta_i^0)^T P_{i-j}^0
 \end{equation}
@@
 \noindent
-where $P_{i-j}^0 = P_j^0 - P_i^0$, with $j \in [4;10]$ being index of the edge node and $i \in [1;3]$ the index of associated corner node, is the segment at the beginning of the simulation and $R_i^0$ the orientation of node $P_i$.
+where
+$\mathbf{x}_i$ and $\theta_i$ are the position and angular position of the node $i$,
+$R(\theta_i^0)$ is the rotation matrix for the angular position of node $i$ at the beginning of the simulation and
+$P_{i-j}^0 = P_j^0 - \mathbf{x}_i^0 = P_j^0 - P_i^0$ the attachment segment.
 
-In each step of the simulation we recompute the new position based on 
-current orientation of the corner node:
-\CD{Other possible explanation: we apply the rigid transformation of frame $i$ on control point $j$ }
+In each step of the simulation we
+%recompute the new position based on current orientation of the corner node:
+apply the rigid transformation of frame $i$ on control point $j$:
 
 \begin{equation}
-    P_j = P_i + P_{i-j} = P_i + R_i \overline{P_{i-j}}
+    P_j = \mathbf{x}_i + P_{i-j} = P_i + R(\theta_i) \overline{P_{i-j}}.
 \end{equation}
 
-\CD{To be consitent with what I write in following, we could introduce a notation for the dofs of the nodes:  $\mathbf{x}_i$ for the position and $\theta_i$ for the rotation of the node $i$
-Then, the previous equation would be:
-\begin{equation}
-    P_j = \mathbf{x}_i + P_{i-j} = P_i + R(\theta_i) \overline{P_{i-j}}
-\end{equation}
-where $ R(\theta_i)$ is the rotation matrix for the angular position of node $i$...
-}
-
-\noindent
-where $P_i$ is the current position of corner node and $R_i$ its orientation.
-
-$C^1$ continuity on the edges can also be maintained if special care is taken
-when computing the position of the central point (see \cite{Ubach2010}). 
-For simplicity we compute the central point like this: 
-\CD{Here it is not clear that with this formulation we do not maintain $C^1$ Continuity...}
-
-\begin{equation}\label{eq-central}
-    P_{10} = \frac{1}{3}(\sum_{i=4}^9 P_i - \sum_{i=1}^3 P_i)
-\end{equation}
-
-% TODO: explain the equation eq-central? would require an image
-
-That way if the element is not flat the central point is slightly elevated
-and not in the plane of other nodes thus keeping the curvature of the
-element.
+For the central point we again apply interpolation of \eqref{eq-central}.
 

### 2012-07-25 Doc/VRIPHYS2012/vriphys2012.tex
finishing section 2
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@432 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 where the points $P_i$ are named as in Figure \ref{fig-bezier}.
-
-% TODO: explain the equation eq-central? would require an image
+% Equation eq-central is better explained below
 
@@
 With 10 control points and 3 DOFs per point it is a total of 30 DOFs for triangle.
-This puts ... blah blah.
+Simulation of such complex system would significantly reduce the sparsity of stiffness matrix.
 We propose a reduction of the number of degrees of freedom by using 3 nodes with 6 degrees of freedom
@@
 associated corner nodes. 
-In the following text the index $j \in [4;10]$ refers to edge points and index $i \in [1;3]$ to the associated corner point/node.
+In the following text the index $j \in [4;10]$ refers to edge points and index $i \in [1;3]$ to the associated corner point/node (unless $i$ is explicitly used differently).
 
@@
 
-For the central point we again apply interpolation of \eqref{eq-central}.
+For the central point we again apply interpolation of \eqref{eq-central} which is kind of skinning function:
+
+\begin{equation}\label{eq-central}
+    \begin{split}
+        P_{10} = & \frac{1}{3}(\sum_{i=4}^9 P_i - \sum_{i=1}^3 P_i)\\
+               = & \frac{1}{3} \left[ P_1 + (P_{1-4} + P_{1-5}) \right] + \\
+                & + \frac{1}{3} \left[ P_2 + (P_{2-6} + P_{2-7}) \right] + \\
+                & + \frac{1}{3} \left[ P_3 + (P_{3-8} + P_{3-9}) \right]
+    \end{split}
+\end{equation}
 
@@
 
-\subsection{Deflection function} % {{{
+\subsection{Deflection Function} % {{{
 
-% TODO: we need to justify how we decide to move the control points
+% explicit {{{
+%We recall the equation for central node \eqref{eq-central} and
+%considering how are the edge points of control mesh connected to corner points with
+%segments $P_{i-j}$:
+%
+%\begin{equation}
+%    P_{i-j} = P_j - P_i, \quad j \in [4;9], i \in [1;3].
+%\end{equation}
+%
+%We now add small variation to the position ($U_1, U_2, U_3$) and orientation ($\theta_1, \theta_2, \theta_3$) of corner nodes to the equation of the element \eqref{eq-cubicbez}:
+%
+%\begin{equation}
+%    \begin{split}
+%        & n_1 (P_1 + U_1)
+%        + n_2 (P_2 + U_2)
+%        + n_3 (P_3 + U_3) \\
+%        & + n_4 (P_1 + U_1 + P_{1-4} + P_{1-4} \times \theta_1)
+%        + n_5 (P_1 + U_1 + P_{1-5} + P_{1-5} \times \theta_1) \\
+%        & + n_6 (P_2 + U_2 + P_{2-6} + P_{2-6} \times \theta_2)
+%        + n_7(P_2 + U_2 + P_{2-7} + P_{2-7} \times \theta_2) \\
+%        & + n_8 (P_3 + U_3 + P_{3-8} + P_{3-8} \times \theta_3)
+%        + n_9 (P_3 + U_3 + P_{3-9} + P_{3-9} \times \theta_3) \\
+%        & + n_{10} \frac{1}{3} \left(
+%        \sum_{j=4}^9 (P_i + U_i + P_{i-j} + P_{i-j} \times \theta_i) - \sum_{i=1}^3 (P_i + U_i)
+%        \right) \\
+%%
+%        & = 
+%    \sum_{i=1}^{10} n_i P_i
+%      + n_1 U_1
+%      + n_2 U_2
+%      + n_3 U_3 \\
+%      & + n_4 (U_1 + P_{1-4} \times \theta_1)
+%      + n_5 (U_1 + P_{1-5} \times \theta_1)
+%      + n_6 (U_2 + P_{2-6} \times \theta_2) \\
+%      & + n_7 (U_2 + P_{2-7} \times \theta_2)
+%      + n_8 (U_3 + P_{3-8} \times \theta_3)
+%      + n_9 (U_3 + P_{3-9} \times \theta_3) \\
+%      & + n_{10} \frac{1}{3} (U_1 + U_2 + U_3 + \sum_{j=4}^9 P_{i-j} \times \theta_i) \\
+%%
+%      & = T + U_1 (n_1 + n_4 + n_5 + \frac{1}{3} n_{10}) \\
+%      & + U_2 (n_2 + n_6 + n_7 + \frac{1}{3} n_{10}) \\
+%      & + U_3 (n_3 + n_8 + n_9 + \frac{1}{3} n_{10}) \\
+%      & + \sum_{j=4}^9 n_j P_{i-j} \times \theta_i
+%      + n_{10} \frac{1}{3} \sum_{j=4}^9 P_{i-j} \times \theta_i \\
+%%
+%      & = T + u
+%  \end{split}
+%\end{equation}
+%}}}
 
-\CD{It's more about derivatives of the kinematic relation, right ?}
+A small variation $\delta \mathbf{x}_i$ of the node $i$ position and $\delta \mathbf{\theta}_i$ for the rotation produces the following small variation of position corner points $\delta P_i$ and edge points $\delta P_j$:
 
-We recall the equation for central node \eqref{eq-central} and
-considering how are the edge points of control mesh connected to corner points with
-segments $P_{i-j}$:
+\begin{eqnarray}
+    \delta P_i & = & \delta \mathbf{x}_i \\
+    \delta P_j & = & \delta \mathbf{x}_i +  P_{i-j} \times \delta \mathbf{\theta}_i
+\end{eqnarray}
+
+\noindent
+and variation of central point $\delta P_{10}$:
 
 \begin{equation}
-    P_{i-j} = P_j - P_i, \quad j \in [4;9], i \in [1;3].
+    \delta P_{10} = \frac{1}{3} (\sum_{i=1}^3 \delta \mathbf{x}_i + \sum_{j=4}^9 P_{i-j} \times \delta \mathbf{\theta}_i)
 \end{equation}
 
-We now add small variation to the position ($U_1, U_2, U_3$) and orientation ($\theta_1, \theta_2, \theta_3$) of corner nodes to the equation of the element \eqref{eq-cubicbez}:
+Thus, finally, it creates a variation of the point position on the surface $\delta T$:
+\begin{equation}
+    \delta T = \sum_{i=1}^{10} n_i  \delta P_i
+\end{equation}
+
+\noindent
+or:
 
@@
     \begin{split}
-        & n_1 (P_1 + U_1)
-        + n_2 (P_2 + U_2)
-        + n_3 (P_3 + U_3) \\
-        & + n_4 (P_1 + U_1 + P_{1-4} + P_{1-4} \times \theta_1)
-        + n_5 (P_1 + U_1 + P_{1-5} + P_{1-5} \times \theta_1) \\
-        & + n_6 (P_2 + U_2 + P_{2-6} + P_{2-6} \times \theta_2)
-        + n_7(P_2 + U_2 + P_{2-7} + P_{2-7} \times \theta_2) \\
-        & + n_8 (P_3 + U_3 + P_{3-8} + P_{3-8} \times \theta_3)
-        + n_9 (P_3 + U_3 + P_{3-9} + P_{3-9} \times \theta_3) \\
-        & + n_{10} \frac{1}{3} \left(
-        \sum_{j=4}^9 (P_i + U_i + P_{i-j} + P_{i-j} \times \theta_i) - \sum_{i=1}^3 (P_i + U_i)
-        \right) \\
-%
-        & = 
-    \sum_{i=1}^{10} n_i P_i
-      + n_1 U_1
-      + n_2 U_2
-      + n_3 U_3 \\
-      & + n_4 (U_1 + P_{1-4} \times \theta_1)
-      + n_5 (U_1 + P_{1-5} \times \theta_1)
-      + n_6 (U_2 + P_{2-6} \times \theta_2) \\
-      & + n_7 (U_2 + P_{2-7} \times \theta_2)
-      + n_8 (U_3 + P_{3-8} \times \theta_3)
-      + n_9 (U_3 + P_{3-9} \times \theta_3) \\
-      & + n_{10} \frac{1}{3} (U_1 + U_2 + U_3 + \sum_{j=4}^9 P_{i-j} \times \theta_i) \\
-%
-      & = T + U_1 (n_1 + n_4 + n_5 + \frac{1}{3} n_{10}) \\
-      & + U_2 (n_2 + n_6 + n_7 + \frac{1}{3} n_{10}) \\
-      & + U_3 (n_3 + n_8 + n_9 + \frac{1}{3} n_{10}) \\
-      & + \sum_{j=4}^9 n_j P_{i-j} \times \theta_i
-      + n_{10} \frac{1}{3} \sum_{j=4}^9 P_{i-j} \times \theta_i \\
-%
-      & = T + u
+        \delta T = &   (n_1 + n_4 + n_5 + \frac{1}{3} n_{10})  \delta \mathbf{x}_1 + \\
+                   & + (n_2 + n_6 + n_7 + \frac{1}{3} n_{10})  \delta \mathbf{x}_2 + \\
+                   & + (n_3 + n_8 + n_9 + \frac{1}{3} n_{10})  \delta \mathbf{x}_3 + \\
+                   & + \sum_{j=4}^9 P_{i-j} \times \delta \mathbf{\theta}_j + \\
+                   & + n_{10} \frac{1}{3} \sum_{j=4}^9 P_{i-j} \times \mathbf{\theta}_i
     \end{split}
@@
 
-\CD{I would use the following notation: small variation of the node $j$ position:   $\delta \mathbf{x}_j$ for the translation and $\delta \boldsymbol{\theta}_j$ for the rotation.
-This small variation of the position creates a small variation of the control points $i$ position $\delta P_i$ that are on the edges:
-$$
-\delta P_i =  \delta \mathbf{x}_j +  P_{i-j} \times \delta \boldsymbol{\theta}_j
-$$
- Variation of node $10$ will be:
-$$
-\delta P_{10} =  \sum_{j=1}^3  \frac{1}{3}(\delta \mathbf{x}_j +  P_{10-i} \times \delta \boldsymbol{\theta}_i)
-$$
-%
-Thus, finally, it will creates a variation of the point position on the surface $\delta T$...
-$$
-\delta T =   \sum_{i=1}^{10} n_i  \delta P_i
-$$
-which makes:
-\begin{equation}
-    \begin{split}
-\delta T =  (n_1 + n_4 + n_5 + \frac{1}{3} n_{10})  \delta \mathbf{x}_1 \\
- + (n_2 + n_6 + n_7 + \frac{1}{3} n_{10})  \delta \mathbf{x}_2 \\
- + (n_3 + n_8 + n_9 + \frac{1}{3} n_{10})  \delta \mathbf{x}_3 \\
-+  \sum_{i=4}^9  \sum_{j=1}^3 n_i P_{i-j} \times \delta \theta_j \\
- +  n_{10} \frac{1}{3} \sum_{j=1}^3 P_{10-j} \times \theta_j
-  \end{split}
-\end{equation}
-%
-}
+\noindent
+thus computing the physical deflection $u$ as variation of the point position $\delta T = u$.
 
-\noindent
-thus computing the function of deflection $u$ 
-\CD{ or $\delta T$ }.
-
-We have simplified the notation by leaving the variable $i$ free in the above sums involving $P_{i-j}$.
+We have simplified the notation by leaving the variable $i$ free in the above sums ($\sum_{j=4}^9$) involving $P_{i-j}$.
 Note however that the expressions are sound and unambiguous because there is only one possible $i$ for each $j$ depending on the connection of segments to the corner nodes (see fig.
@@
 
-...
-
-Using the plate theory we can now compute the strain-displacement matrices
-from the displacement field. From the Cauchy's strain tensor for the
-elastic membrane:
+Using the plate theory we now can compute the strain-displacement matrices
+from the displacement field $u$. The Cauchy's strain tensor for the
+elastic membrane is defined as:
 
@@
 \end{equation}
+.
 
-\noindent
-and from Kirchhoff-Love theory for thin plates:
+We apply Kirchhoff-Love theory for thin plates to obtain strain-displacement matrix for bending plate element:
 

### 2012-07-27 Doc/VRIPHYS2012/vriphys2012.tex
article
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@433 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\title[Short title]%
-      {Bezier shell finite element for interactive surgical simulation} % TODO
+\title[B\'ezier Shell FE]%
+      {B\'ezier Shell Finite Element for Interactive Surgical Simulation}
 
@@
 % instead of the author's name (and leave the affiliation blank) !!
+% TODO: anonymize
 \author[...]
@@
 This problem has been faced by the computer graphics community but for different applications.
-For instance, many related work addresses the problem of the simulation of cloth using mass-spring models \cite{Volino2009} or bending models \cite{Grinspun2003,Choi2007}. %% Todo verify.... 
+For instance, many related works address the problem of the simulation of cloth using mass-spring models \cite{Volino2009} or bending models \cite{Grinspun2003}. %% Todo Choi2007?
 However, these models are discrete so their behavior depends on the mesh and their parameters are not easily measurable.
@@
 
-Methods based on continuum mechanics and namely theory of elasticity gained on popularity in computer graphics and real-time simulators ever since Terzopoulos \cite{Terzopoulos1987} presented his work on elastic deformation modeling.
-%% Do we
+Methods based on continuum mechanics and namely theory of elasticity gained on popularity in real-time simulations ever since Terzopoulos \cite{Terzopoulos1987} presented his work on elastic deformation modeling.
 While the methods using finite elements and theory of elasticity are fairly popular in volumetric modeling of deformations \cite{Cotin1999,Nesme2006,Miller2007}, they are still frowned upon in the area of thin structures.
@@
 This is indeed true if lots of elements are used to discretize the object.
-A classical strategy, is to use high-polygonal meshes for rendering or collision detection and response, while maintaing a reasonably low number of elements for the mechanics.
+A classical strategy, is to use high-polygonal meshes for rendering or collision detection and response, while maintaining a reasonably low number of elements for the mechanics.
 The work of Bouthors et al.\cite{Bouthors2007} is a good application example of this strategy for surface deformations.
-But this strategy is not sufficient in our case because the problem of performance is also greatly impacted when using elements that have an interpolation with high order polynoms.
+But this strategy is not sufficient in our case because the problem of performance is also greatly impacted when using elements that have an interpolation with high order polynomials.
 This type of interpolation is necessary because bending continuum model
 (based on Kirchhoff-Love theory for thin plates)  requires the computation of second spatial derivatives.
-In \cite{Comas2010c}, Comas and al.  use an hybrid interpolation: linear interpolation is used for in-plane deformations and cubic polynomials for bending deformations.
+On the other hand, elements common in mechanics (e.g.
+Discrete Kirchhoff Triangles) are based on shape functions that are impractical for real-time simulations or don't have explicitly defined shape functions at all.
+In \cite{Comas2010c}, Comas et al. use a hybrid interpolation: linear interpolation is used for in-plane deformations and cubic polynomials for bending deformations.
 The element is piloted by 3 nodes on which 6 degrees of freedom are defined. 
 From this point of view, our work is derived from this approach but we address several noted drawbacks: the formulation misses one rotational degree of freedom, the cubic interpolation is not symmetric and produces $C^0$  discontinuities between elements on edges.
-Ubach and al. \cite{Ubach2010}  present a shell element based on B\'ezier triangles that allows for a better behavior.
+Ubach et al. \cite{Ubach2010}  present a shell element based on B\'ezier triangles that allows for a better behavior.
 However, the B\'ezier interpolation defined on each triangle element is based on the motion computed on adjacent triangles.
@@
 \begin{equation}
-    P_j = \mathbf{x}_i + P_{i-j} = P_i + R(\theta_i) \overline{P_{i-j}}.
+    P_j = \mathbf{x}_i + P_{i-j} = P_i + R(\theta_i) \overline{P_{i-j}}
 \end{equation}
@@
 
-\begin{equation}
+\begin{equation}\label{bez-delta}
     \begin{split}
@@
 \end{equation}
-.
 
@@
 
-Because the deformation field for the shell is non-linear \eqref{eq-bezU}
-the integrals \eqref{eq-Km} and \eqref{eq-Kb} have to be computed using
+Because the deflection field for the element is non-linear \eqref{eq-delta}
+in position the integrals \eqref{eq-Km} and \eqref{eq-Kb} have to be computed using
 numerical integration. In our implementation we employ 6-point Gaussian
@@
 
-%The displacement field defined in \eqref{eq-bezU} is clearly symmetric in
+%The displacement field defined in \eqref{eq-delta} is clearly symmetric in
 %respect to deformations, but ...
@@
 In next two subsections we present tests to validate the convergence of our
-B\'ezier shell element. The results are compared with the shell element
-composed of DKT element and optimal ANDES element \cite{Felippa2003}. The
-last subsection presents evaluation of the computational complexity of our
-solution.
+B\'ezier shell element. We use a combination of two elements that are very well
+established in the area of mechanical modelling.
+Our results are compared with the shell element
+composed of Discrete Kirchoff Triangle (DKT) element \cite{Batoz1980} for bending plate and optimal ANDES element \cite{Felippa2003} for membrane.
+The last subsection presents evaluation of the computational complexity of our solution.
 
@@
 
-\subsection{Applications}
+\subsection{Application on Congenital Heart Disease Correction Planning}
 
-The presented element has been successfully implemented in SOFA framework
-\cite{Faure2012} for real-time surgical simulations. It has been evaluated
-on several applications.
+Surgical interventions in infants with congenitally malformed great arteries
+and hearts are extremely challenging due to the complex and heterogeneous
+nature of their disease patterns. At present, cardiac surgeons rely on
+non-invasive imaging for patient-specific examination and a set of preoperative
+sketches with varying approaches to the surgical procedure. However, the most
+promising approach is often chosen during the actual open-heart surgery, when
+surgeons get a more concrete idea of possible outcomes. Finalizing this
+decision during an on-pump cardiac surgery is further complicated by deflated
+blood vessels since it is difficult to imagine their exact change of shape
+caused by the intervention. The important decisions have to be made in very
+short time and are strongly dependent on the surgeon's experience.
 
-% TODO: !!! When ANONYMISING do something with the citation !!!
-One being a tool for pre-operative planning of corrective surgery of
-congenital heart diseases in infants \cite{Kislinskiy2012}. The cardiac
-surgery is a complicated process where several arterial vessels have to be
-cut, patched and reconnected. The element has been used for to model
-deformations of blood vessels after suturing and/or applying a artificial
-patch. Figure \ref{fig-chd} presents a screenshots of the application
-prototype.
+The planning process can be improved by employing surgical simulation system
+using three-dimensional heart models based on patient-specific images as input.
+Most present simulation systems focus on execution of low-level surgical
+procedures, e.g. incising and suturing. In these reactive simulation systems it
+is more important to enable simulations in real-time than accurate simulation
+of tissue and blood vessel deformation in response to simulated interventions.
+
+To be able to predict results of complex surgical procedures a physically more
+accurate approach is necessary. Using our shell finite element model we have
+developed a prototype \cite{Kislinskiy2012} of new surgical simulation system
+for preoperative planning. The element has been used to model deformations of
+blood vessels and artificial patch. By performing a set of topological changes
+we simulate the process of incising the vessel and suturing on a pre-specified
+boundary. By iteratively adjusting the underlaying mesh we search for the
+elastic equilibrium. Figure \ref{fig-chd} presents a screenshots of the
+application prototype.
+
+%%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
+%
+%% TODO: is this OK for anonymity?
+%The presented element has been successfully implemented in SOFA framework
+%\cite{Faure2012} for real-time surgical simulations. It has been evaluated
+%on several applications.
+%
+%% TODO: !!! When ANONYMISING do something with the citation !!!
+%One being a tool for pre-operative planning of corrective surgery of
+%congenital heart diseases in infants \cite{Kislinskiy2012}. The cardiac
+%surgery is a complicated process where several arterial vessels have to be
+%cut, patched and reconnected. The element has been used for to model
+%deformations of blood vessels after suturing and/or applying a artificial
+%patch. Figure \ref{fig-chd} presents a screenshots of the application
+%prototype.
 
@@
 
-Another designed application is simulation of women reproductive organs
-(vagaina/uterus/\ldots). 
-The simulation is so as to evaluate
-stress on a tissue during a birth.
+% TODO?
+%Another designed application is simulation of women reproductive organs
+%(vagaina/uterus/\ldots). 
+%The simulation is so as to evaluate
+%stress on a tissue during a birth.
 
@@
 
-This paper presents new shell element with geometry and displacement
-function based on the formulation of B\'ezier triangle. It fixes problems
-of the previous model \cite{Comas2010c}, namely:
+In this paper we presents new shell element with interpolation functions
+based on B\'ezier triangle which is more suitable for real-time simulations than elements common in mechanics with complex, obscure or unclear interpolation functions.
+It fixes problems of the previous model \cite{Comas2010c} based on cubic polynomials, namely:
 
@@
 
-    \item It is continuous on the edges and provides $C^1$ continuity in
-    corners of the element. This is essential for good visual experience if
-    high-resolution mesh is mapped on the mechanical mesh.
+    \item Element is based on all 6 DOFs, including the drilling degree of
+        freedom, not just 5 DOFs. Therefore it doesn't have the problem of
+        undefined z-rotations.
 
@@
 
-    \item TODO: anything else \ldots? % TODO
+    \item It is $C^0$ continuous on the edges and provides $C^1$ continuity in
+    corners of the element. This is essential for good visual experience if
+    high-resolution mesh is mapped on the mechanical mesh.
 
@@
 \end{itemize}
+
+Unified formulation of the deformation allows for derivation of three-dimensional 
+stress tensor. This however remains as topic for future work and
+we have kept the computation of membrane stresses separate from bending stresses.
 

### 2012-07-29 Doc/VRIPHYS2012/vriphys2012.tex
comments and some update
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@434 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 % TODO
-The main contributions of the paper are following:
+The main contributions of the paper are following \CD{- Tomas, you need to write sentences here -}
+
 \begin{itemize}
@@
 In this subsection we propose an element featuring interpolation with high degree of flexibility while still maintaining low number of degrees of freedom.
-We introduce a new way of building the interpolation by using two stages:
+We introduce a new way of building the interpolation by using two stages 
+\CD{ - Need a figure here - }:
 \begin{enumerate}
@@
 
-For the central point we again apply interpolation of \eqref{eq-central} which is kind of skinning function:
+For the central point we again apply interpolation of \eqref{eq-central} which is kind of skinning function \CD{Why so complicated ?
+You can compute  some rules to chose the first location of point $P_{10}$ (i.e. to compute $P_{i-10}$) but then use something more simple for the kinematics}:
 
@@
 
+\CD{I would have put (but that is maybe the same result...)}
+
+\begin{equation}\label{eq-central}
+    \begin{split}
+        P_{10} =  \frac{1}{3}\left( \sum_{i=1}^3 \mathbf{x}_i + R(\theta_i) \overline{P_{i-10}} \right)
+    \end{split}
+\end{equation}
+
 %}}}
+
+\CD{and it could be possible to write down (and this is helpful for the formulation of the spacial derivative):}
+\begin{equation}\label{global-transform}
+    \begin{split}
+ T(\xi_1,\xi_2, \xi_3) = \sum_{i=1}^{10} n_i(\xi_1,\xi_2, \xi_3)P_i = \\
+ 	\sum_{i=1}^{3} n_i(\xi_1,\xi_2, \xi_3) \mathbf{x}_i +  \sum_{j=4}^9 n_j (\xi_1,\xi_2, \xi_3)( \mathbf{x}_i +  R(\theta_i) \overline{P_{i-j}}) + \\  \frac{n_{10}(\xi_1,\xi_2, \xi_3)}{3}\left( \sum_{i=1}^3 \mathbf{x}_i + R(\theta_i) \overline{P_{i-10}} \right)
+    \end{split}
+\end{equation}
 
@@
 \noindent
-and variation of central point $\delta P_{10}$:
+and variation of central point $\delta P_{10}$ \CD{ this formulation would be simplified...} :
 
@@
                    & + \sum_{j=4}^9 P_{i-j} \times \delta \mathbf{\theta}_j + \\
-                   & + n_{10} \frac{1}{3} \sum_{j=4}^9 P_{i-j} \times \mathbf{\theta}_i
+                   & + n_{10} \frac{1}{3} \sum_{j=4}^9 P_{i-j} \times \delta \mathbf{\theta}_i 
     \end{split}
@@
 
-\section{...} % {{{
+%%%%%%%%%%%%%%%%%%%%%%%%% SECTION 3 %%%%%%%%%%%%%%%
+
+\section{Finite element formulation} % {{{
+
+In the previous section we present an interpolation based on cubic B\'ezier triangles and a kinematic link between the control points and the element 6DoF Nodes.
+In this section we show how this interpolation can be used to compute a finite element formulation of the shell equations.
+
+\subsection{Computation of the strain tensor}
+
+We define a local system of coordinates $(x,y,z)$ in which the axis z is oriented along the thickness of the element. (\CD{TODO: figure}) 
+For the shell formulation, there is no need to compute the derivatives along axis $z$.
+
+The 2D tensor of deformation can be defined as a $2\times2$ matrix in which the components $k=\{x,y\}$ and $l=\{x,y\}$are:
+\begin{equation}
+F_{kl} = \frac{\delta T^k}{\delta l}   
+\end{equation}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%% SECTION 4 %%%%%%%%%%%%%%%
+
 \section{Visual and Mechanical Mapping} % {{{
@@
 % inconsistent).
+  \CD{ - On this section,  we can maybe add some results from the CBM paper you did with Stephan especially  the figure 9 that illustrates the 3d contribution we list in the intro: i.e.
+Mapping of high resolution mesh on the curved surface of the shell.}
 

### 2012-07-29 Doc/VRIPHYS2012/vriphys2012.tex
begin of something in section 3.1
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@435 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-We define a local system of coordinates $(x,y,z)$ in which the axis z is oriented along the thickness of the element. (\CD{TODO: figure}) 
-For the shell formulation, there is no need to compute the derivatives along axis $z$.
+(\CD{TODO add a figure to illustrate the explanation})
 
-The 2D tensor of deformation can be defined as a $2\times2$ matrix in which the components $k=\{x,y\}$ and $l=\{x,y\}$are:
+We define a local system of coordinates $(x,y,z)$ in which the axis z is oriented along the thickness of the element\footnote{The computation of the frame of each element will be detailed in section \ref{sec-corot}.
+}.
+Given the initial position of the triangle points in this system of coordinate, $\{x_1^0, y_1^0\}$,  $\{x_2^0, y_2^0\}$, and  $\{x_3^0, y_3^0\}$, we can compute a linear relation between $\{\xi_1,\xi_2, \xi_3\}$ and $\{ x,y \}$:
+
 \begin{equation}
-F_{kl} = \frac{\delta T^k}{\delta l}   
+\left[ \begin{array}[c]{c} \xi_1(x,y) \\ \xi_2(x,y) \\ \xi_3(x,y) \end{array} \right] =
+\left[ \begin{array}[c]{ccc}
+1 & 1 &  1 \\
+x_1^0 & x_2^0 & x_3^0 \\
+y_1^0 & y_2^0 & y_3^0
+\end{array} \right] ^{-1} 
+\left[ \begin{array}[c]{c} 1 \\ x \\ y \end{array} \right] 
 \end{equation}
+
+Thus, we can locally compute some spacial derivatives, along $x$ and $y$ on the interpolation function vector $T$ (which has three components $(T^x, T^y, T^z)$ .
+The 2D tensor $F$ of the in-plane deformations can be computed as a $2\times2$ matrix where the line $k$ and the column $l$ alternatively represent $x$ and $y$\footnote{It can be noted $k=\{x,y\}$ and $l=\{x,y\}$}:
+\begin{equation}
+F_{kl} = \frac{\delta T^k}{\delta l}  =  \sum_{j=1}^3 \frac{\delta T^k}{\delta \xi_i} \frac{\delta \xi_i}{\delta l}  
+\end{equation}
+
+For the shell formulation, there is no need to compute the derivatives along direction $z$, but the second derivatives derivatives $\frac{\delta T^z}{\delta kl}$...
 
@@
 {Convergence for modified Scordelis-Lo roof test of DKT+ANDES element and
-our B\'ezier shell element.}
+our B\'ezier shell element. \CD{Can you do the same test with the model used by Comas ?}}
 
@@
 {Convergence for test on hemisphere with 18° hole for the DKT+ANDES element and
-our B\'ezier shell element.}
+our B\'ezier shell element. \CD{Same question: can you do the same test with the model we used with O.Comas ? Hopefully, it will show that our model has a better convergence ... }}
 

### 2012-07-30 Doc/VRIPHYS2012/vriphys2012.tex
trying new image for membrane/bending deformations
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@436 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
section 3.1
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@437 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
+% For large/wide figures
+%\FigureW{width}{name}{caption}
+\newcommand{\FigureWI}[3]{%
+\begin{figure*}[htb]
+  \centering
+  \input{#1.pdf_tex}
+  \caption{\label{fig-#1}#2}
+\end{figure*}}
+
 %%% ADD COMMENTS=
@@
 % TODO: anonymize
-\author[...]
-       {...} % TODO
+\author[Authors]
+       {Authors} % TODO
 
@@
 \noindent
-and variation of central point $\delta P_{10}$ \CD{ this formulation would be simplified...} :
+and variation of central point $\delta P_{10}$ \CD{ this formulation would be modified...} :
 
@@
 In this section we show how this interpolation can be used to compute a finite element formulation of the shell equations.
+The model relies on large transformation formulation: the geometrical non-linearities are handled using a corotational formulation, whereas, in the frame of the element, infinitesimal strain theory is used.
 
@@
 \begin{equation}
+\label{shape-vector}
 \left[ \begin{array}[c]{c} \xi_1(x,y) \\ \xi_2(x,y) \\ \xi_3(x,y) \end{array} \right] =
@@
 \begin{equation}
-F_{kl} = \frac{\delta T^k}{\delta l}  =  \sum_{j=1}^3 \frac{\delta T^k}{\delta \xi_i} \frac{\delta \xi_i}{\delta l}  
-\end{equation}
-
-For the shell formulation, there is no need to compute the derivatives along direction $z$, but the second derivatives derivatives $\frac{\delta T^z}{\delta kl}$...
+F_{kl} = \frac{\partial T^k}{\partial l}  =  \sum_{i=1}^3 \frac{\partial T^k}{\partial \xi_i} \frac{\partial \xi_i}{\partial l}  
+\end{equation}
+
+The term $\frac{\partial T^k}{\partial \xi_i}$ can be derived from the expression of $T(\xi_1,\xi_2, \xi_3)$ in equation (\ref{global-transform}) and 
+$\frac{\partial \xi_i}{\partial l}$ from equation (\ref{shape-vector}). 
+The in-plane displacement gradient can be obtained with tensor $F$:
+\begin{equation}
+\frac{\partial u^k}{\partial l} = F_{kl} - I_{kl} 
+\end{equation}
+where $I$ is the identity matrix ($I_{kl}  = \{1$ if $l=k$ ;  $0$ if $l \neq k \} $). 
+The in-plane displacement gradient can be linked to the degrees of freedom of the shell using the expression of the displacement of a point $\delta T$ (see equation  (\ref{bez-delta})) on the shell surface:
+$$
+\frac{\partial u^k}{\partial l} = \frac{\partial (\delta T^k)}{\partial l}
+$$
+
+In the local frame of the element, we rely on infinitesimal strain theory, thus the expression of the strain tensor (using Voigt notations) is:
+\begin{equation}
+\label{in-plane-strain}
+\left[ \begin{array}[c]{c} \epsilon_{xx} \\ \epsilon_{yy}  \\ 2\epsilon_{yx} \end{array} \right] = 
+\left[ \begin{array}[c]{c} \frac{\partial u^x}{\partial x}  \vspace{1mm}\\ \frac{\partial u^y}{\partial y}   \\ \frac{\partial u^x}{\partial y}  + \frac{\partial u^y}{\partial x}  \end{array} \right]
+\end{equation}
+
+For the bending formulation, the shell theory uses a measure of the out-of-plane-displacement $u^z(x,y)$. 
+The strain tensor has additional contributions based on Kirchhoff-Love theory for thin plates:
+\begin{equation}
+\label{out-plane-strain}
+\left[ \begin{array}[c]{c} \epsilon_{xx} \\ \epsilon_{yy}  \\ 2\epsilon_{yx} \end{array} \right] +=
+\left[ \begin{array}[c]{c} -z \frac{\partial^2 u^z}{\partial x^2}  \vspace{1mm}\\ -z \frac{\partial^2 u^z}{\partial y^2}   \vspace{1mm}\\ - 2z \frac{\partial^2 u^z}{\partial xy}    \end{array} \right]
+\end{equation}
+where $z$ is the local coordinate along $z$-axis of the point on which the strain is measured.
+\CD{This is something that is maybe wrong in our formulation...
+as the bending adds a contribution on the deformation tensor, maybe we should not separate the contribution from membrane and bending as we do now ($K_m$ and $K_b$ are computed separately) }
+
+As the term  $\frac{\partial \xi_i}{\partial l}$ provides a constant value, the second derivatives can be computed quite easily, for instance, for $\epsilon_{xx}$:
+\begin{equation}
+\frac{\partial^2 u^z}{\partial x^2}  = \sum_{i=1}^3\frac{\partial}{\partial \xi_i}\left(\sum_{j=1}^3 \frac{\partial T^z}{\partial \xi_j} \frac{\partial \xi_j}{\partial x}\right) \frac{\partial \xi_i}{\partial x}
+=  \sum_{i=1}^3  \sum_{j=1}^3 \frac{\partial^2 T^z}{\partial \xi_i\xi_j}   \frac{\partial \xi_j}{\partial x} \frac{\partial \xi_i}{\partial x}  
+\end{equation}
 
@@
 
-Shell element is a combination of two simpler elements describing two types
-of deformations:
+As previously describes, the shell element's strain is a combination of two types of deformations, as shown in figure \ref{deformations}.
+If we assume that the initial out-plane deflections are negligible and that the frame of the shell element nodes are aligned with local $z$ axis, we can distribute these two deformations into two groups of degrees of freedom at the nodes' level.
+
+\FigureWI{deformations}
+{Plate deformations:%
+a) membrane deformations like stretching or shearing, due to in-plane displacements
+b) undeformed;
+c) bending deformation, due to out-plane displacements
+\label{deformations}}
 
@@
     \item Elastic membrane defining deformations in plane of the
-    element. It encompasses deformations like stretching and shearing and
-    is described by 3 degrees of freedom: in-plane displacements $u_x, u_y$
-    and rotation $\theta_z$ around the axis perpendicular to the plane of
-    the element (Figure \ref{fig-membrane}).
-
-\Figure{0.8\linewidth}{membrane}
-{DOFs of elastic membrane element.}
-
-    \item Bending plate defining out of plane bending deformation. It is
+    element. It encompasses deformations like stretching and shearing.
+    It corresponds to the in-plane displacements $u_x, u_y$. 
+    At the nodes' level, it corresponds to in-plane displacements $\delta\mathbf{x}_x, \delta\mathbf{x}_y$ and rotation $\theta_z$.
+(Figure \ref{fig-membrane}).
+    \item Bending plate defining deformation due to out of plane bending displacement $u_z$. It is
     described by 3 degrees of freedom: rotation around two axes $\theta_x,
-    \theta_z$ and out of plane displacement $u_z$ (Figure \ref{fig-plate}).
+    \theta_z$ and out of plane displacement $\delta\mathbf{x}_z$ (Figure \ref{fig-plate}).
 
 \end{itemize}
-
-\Figure{0.8\linewidth}{plate}
-{DOFs of bending plate element.}
 
@@
 
-\subsection{Mechanics} % {{{
+\subsection{Stiffness matrix} % {{{
+
+In order to compute FEM-based stiffness for the shell element, we need to compute the jacobian matrix between strains and nodes' displacements.
+For that we can reuse equation (\ref{bez-delta}) and apply the derivatives. For instance, for the in-plane displacement:
+$$
+\frac{\delta \epsilon_{xx}} {\delta \mathbf{x}_{2}} = \sum_{i=1}^{3} (\frac{\partial n_2}{\xi_i} + \frac{\partial n_6}{\xi_i} + \frac{\partial n_9}{\xi_i} + \frac{1}{3}\frac{\partial n_{10}}{\xi_i})\frac{\partial \xi_i}{\partial x}
+$$
 
@@
 \end{align}
+
+\CD{or...
+$K =  \iiint_V \left[\begin{matrix} J_m & J_b \end{matrix} \right]^T  M \left[\begin{matrix} J_m & J_b \end{matrix} \right]$.
+Could we test that option easily in the code, or is it a nightmare ??}
 

### 2012-07-30 Doc/VRIPHYS2012/vriphys2012.tex
section 3
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@438 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \end{equation}
+%% comment
+\CD{Here, if you want to keep a separate computation for bending and membrane (see the folllowing), use $=$ instead of $+=$ !!}
+%%
 where $z$ is the local coordinate along $z$-axis of the point on which the strain is measured.
-\CD{This is something that is maybe wrong in our formulation...
-as the bending adds a contribution on the deformation tensor, maybe we should not separate the contribution from membrane and bending as we do now ($K_m$ and $K_b$ are computed separately) }
 
@@
 In order to compute FEM-based stiffness for the shell element, we need to compute the jacobian matrix between strains and nodes' displacements.
-For that we can reuse equation (\ref{bez-delta}) and apply the derivatives. For instance, for the in-plane displacement:
+As we have separated the formulations for the in-plane and the out-plane displacements, it will result with the computation of two jacobian matrices (of size $3\times9$).
+\begin{equation}
+\delta \left[ \begin{array}[c]{c} \epsilon_{xx} \\ \epsilon_{yy}  \\ 2\epsilon_{yx} \end{array} \right] = 
+\left[ \mathbf{J}_m \right] 
+\left[ \begin{array}[c]{c} \delta \mathbf{x}_{1x} \\ \delta \mathbf{x}_{1y} \\ \delta \theta_{1z} \\ \vdots    \end{array} \right]
+\end{equation}
+where $\left[ \mathbf{J}_m \right]$ maps the membrane in-plane displacements and 
+%
+\begin{equation}
+\delta \left[ \begin{array}[c]{c} \epsilon_{xx} \\ \epsilon_{yy}  \\ 2\epsilon_{yx} \end{array} \right] = 
+\left[ \mathbf{J}_b \right] 
+\left[ \begin{array}[c]{c} \delta \mathbf{x}_{1z} \\ \delta \theta_{1x} \\ \delta \theta_{1y} \\ \vdots    \end{array} \right]
+\end{equation}
+where $\left[ \mathbf{J}_b \right]$ maps the bending out-plane displacements.
+
+To compute the values in matrices $\left[ \mathbf{J}_m \right]$ and $\left[ \mathbf{J}_b \right]$,  we can reuse equation (\ref{bez-delta}) and apply the spacial derivatives.
+To illustrate, let's take one value of the matrix $\left[ \mathbf{J}_m \right]$  (first line, fourth column), we first take the kinematic between $\delta T_x$ and the concerned degree of freedom $ \delta \mathbf{x}_{2x}$:
 $$
-\frac{\delta \epsilon_{xx}} {\delta \mathbf{x}_{2}} = \sum_{i=1}^{3} (\frac{\partial n_2}{\xi_i} + \frac{\partial n_6}{\xi_i} + \frac{\partial n_9}{\xi_i} + \frac{1}{3}\frac{\partial n_{10}}{\xi_i})\frac{\partial \xi_i}{\partial x}
+   \frac{ \delta T_x}{ \delta \mathbf{x}_{2x}} = (n_2 + n_6 + n_7 + \frac{1}{3} n_{10})
 $$
-
-Using the plate theory we now can compute the strain-displacement matrices
-from the displacement field $u$. The Cauchy's strain tensor for the
-elastic membrane is defined as:
-
-\begin{equation}\label{eq-cauchy}
-    J_m = \left[ \begin{matrix}
-        \deriv{u_x}{x} \\
-        \deriv{u_y}{y} \\
-        \deriv{u_x}{y} + \deriv{u_y}{x}
-    \end{matrix} \right]
-\end{equation}
-
-We apply Kirchhoff-Love theory for thin plates to obtain strain-displacement matrix for bending plate element:
-
-\begin{equation}\label{eq-kirchhoff}
-    J_b = \left[ \begin{matrix}
-        - z \deriv{^2 u_z}{x^2} \\
-        - z \deriv{^2 u_z}{y^2} \\
-        - 2z \deriv{^2 u_z}{xy}
-    \end{matrix} \right]
-\end{equation}
+Then, we apply the spacial derivatives to compute the variation of $\epsilon_{xx}$ when 
+$$
+\delta \epsilon_{xx} += \underbrace{\sum_{i=1}^{3} (\frac{\partial n_2}{\xi_i} + \frac{\partial n_6}{\xi_i} + \frac{\partial n_9}{\xi_i} + \frac{1}{3}\frac{\partial n_{10}}{\xi_i})\frac{\partial \xi_i}{\partial x}}_{\left[ \mathbf{J}_m \right](1,4)}  {\delta \mathbf{x}_{2x}}
+$$
 
@@
     \label{eq-Km}
-    K_m & = \iiint_V J_m^T M J_m \, \mathrm{d} V \\
+    \mathbf{K}_m & = \iiint_V \mathbf{J}_m^T M \mathbf{J}_m \, \mathrm{d} V \\
     \label{eq-Kb}
-    K_b & = \iiint_V J_b^T M J_b \, \mathrm{d} V
+    \mathbf{K}_b & = \iiint_V \mathbf{J}_b^T M \mathbf{J}_b \, \mathrm{d} V
 \end{align}
-
-\CD{or...
-$K =  \iiint_V \left[\begin{matrix} J_m & J_b \end{matrix} \right]^T  M \left[\begin{matrix} J_m & J_b \end{matrix} \right]$.
+%% comment
+\CD{This is something that is maybe wrong in our formulation...
+as the bending adds a contribution on the deformation tensor, maybe we should not separate the contribution from membrane and bending as we do now ($K_m$ and $K_b$ are computed separately) }
+%% comment
+\CD{here, it would be...
+$\mathbf{K}_e =  \iiint_V \left[\begin{matrix} \mathbf{J}_m & \mathbf{J}_b \end{matrix} \right]^T  M \left[\begin{matrix} \mathbf{J}_m & \mathbf{J}_b \end{matrix} \right]$.
 Could we test that option easily in the code, or is it a nightmare ??}
+%%
 

### 2012-07-30 Doc/VRIPHYS2012/vriphys2012.tex
small changes in section 3
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@439 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-Thus, we can locally compute some spacial derivatives, along $x$ and $y$ on the interpolation function vector $T$ (which has three components $(T^x, T^y, T^z)$ .
+Thus, we can locally compute some spacial derivatives, along $x$ and $y$ on the interpolation function vector $T$ (which has three components $(T^x, T^y, T^z)$) .
 The 2D tensor $F$ of the in-plane deformations can be computed as a $2\times2$ matrix where the line $k$ and the column $l$ alternatively represent $x$ and $y$\footnote{It can be noted $k=\{x,y\}$ and $l=\{x,y\}$}:
@@
 where $I$ is the identity matrix ($I_{kl}  = \{1$ if $l=k$ ;  $0$ if $l \neq k \} $). 
-The in-plane displacement gradient can be linked to the degrees of freedom of the shell using the expression of the displacement of a point $\delta T$ (see equation  (\ref{bez-delta})) on the shell surface:
+%
+%
+% Remark: the sentence can be suppr..
+\CD{The following  sentencd I let you chose: keep or not...}The in-plane displacement gradient can be linked to the degrees of freedom of the shell using the expression of the displacement of a point $\delta T$ (see equation  (\ref{bez-delta})) on the shell surface:
 $$
@@
 $$ 
+% end of Remark
 
@@
     \item Bending plate defining deformation due to out of plane bending displacement $u_z$. It is
-    described by 3 degrees of freedom: rotation around two axes $\theta_x,
+    influenced by 3 degrees of freedom: rotation around two axes $\theta_x,
     \theta_z$ and out of plane displacement $\delta\mathbf{x}_z$ (Figure \ref{fig-plate}).
@@
 To compute the values in matrices $\left[ \mathbf{J}_m \right]$ and $\left[ \mathbf{J}_b \right]$,  we can reuse equation (\ref{bez-delta}) and apply the spacial derivatives.
-To illustrate, let's take one value of the matrix $\left[ \mathbf{J}_m \right]$  (first line, fourth column), we first take the kinematic between $\delta T_x$ and the concerned degree of freedom $ \delta \mathbf{x}_{2x}$:
+To illustrate, let's take one value of the matrix $\left[ \mathbf{J}_m \right]$  (first line, fourth column), we first take the kinematic between $\delta T_x$ and the concerned degree of freedom $ \delta \mathbf{x}_{2x}$ from equation (\ref{bez-delta}):
 $$
@@
 $$
-Then, we apply the spacial derivatives to compute the variation of $\epsilon_{xx}$ when 
+Then, we apply the spacial derivatives to compute the variation of $\epsilon_{xx}$ due to a variation of $\mathbf{x}_{2x}$.
 $$
-\delta \epsilon_{xx} += \underbrace{\sum_{i=1}^{3} (\frac{\partial n_2}{\xi_i} + \frac{\partial n_6}{\xi_i} + \frac{\partial n_9}{\xi_i} + \frac{1}{3}\frac{\partial n_{10}}{\xi_i})\frac{\partial \xi_i}{\partial x}}_{\left[ \mathbf{J}_m \right](1,4)}  {\delta \mathbf{x}_{2x}}
+\delta \epsilon_{xx} += 
+\underbrace{\left( \sum_{i=1}^{3} (\frac{\partial n_2}{\xi_i} + \frac{\partial n_6}{\xi_i} + \frac{\partial n_7}{\xi_i} + \frac{1}{3}\frac{\partial n_{10}}{\xi_i})\frac{\partial \xi_i}{\partial x} \right) }_{\left[ \mathbf{J}_m \right](1,4)}
+{\delta \mathbf{x}_{2x}}
 $$
@@
 numerical integration. In our implementation we employ 6-point Gaussian
-quadrature for integration over triangle area.
+quadrature for integration over triangle area \CD{TODO: figure + gives the formula of the quadrature...}
 % TODO: add ref

### 2012-07-31 Doc/VRIPHYS2012/vriphys2012.tex
rewritten contributon, new images for the paper
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@440 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 Yet, a special field of continuum mechanics has studied the deformations of these objects, and has proposed models based on shell theory.
-This theory was integrated numerically \CD{Todo citation of a shell book} using Finite Elements, but there are two issues for obtaining interactive frame rates: 
+This theory was integrated numerically using Finite Elements \cite{Reddy1993}, but there are two issues for obtaining interactive frame rates: 
 (i) the continuum equations requires second spatial derivations that can not be handled with linear interpolation and 
@@
 
-% TODO
-The main contributions of the paper are following \CD{- Tomas, you need to write sentences here -}
-
-\begin{itemize}
-    % not really novel, similar model already in Ubach2010
-    %\item shell element based on Bézier triangles
-    \item Our element solves the problems of the element presented by Comas
-        \cite{Comas2010c}: it is continuous on the element boundary and has
-        all rotational degrees of freedom.
-
-    % TODO: needs better explanation
-    \item Two stage interpolation: 6DOF nodes -> bezier points -> bezier
-        triangle
-
-    % + separation between nodes and dofs
-    %\item The model separates nodes of the mesh from degrees of freedom of the
-    %    control mesh. This allows mapping a surface on deformable skeleton.
-
-    % + mapping: smooth surface results with less shells + collision response
-    %   on smooth surface 
-    \item Mapping of high resolution mesh on the curved surface of the
-        shell. The smooth mapped surface can be used for collision
-        detection and allows the propagation of proper collision response
-        back onto the (low resolution) mechanical mesh.
-
-\end{itemize}
+Extending the work of Comas \cite{Comas2010c} we have designed a shell element that uses B\'ezier interpolation polynomials.
+Our element solves the problems of the element presented by Comas:
+(i) it uses all 6 degrees of freedom,
+(ii) the deformations are completly symmetric and
+(iii) the boundary between elements is continuous.
+We present a two stage interpolation:
+(i) first a kinematic link between 6 DOF nodes of element and nodes of B \'ezier control mesh is defined and
+(ii) interpolation function of cubic B\'ezier triangle is applied to interpolate on the surface of the element.
+Mapping of high resolution mesh on the curved surface of the shell.
+Finally from the kinematic link we derive formulas for
+(i) interpolating velocity on the element surface and for
+(ii) distribution of normal force applied on the element surface back to 6 DOF nodes of the element.
+This allows us to use smooth mapped surface for collision detection and allows the propagation of proper collision response back onto the mechanical mesh.
 
@@
 We introduce a new way of building the interpolation by using two stages
-\CD{ - Need a figure here - }:
+
 \begin{enumerate}
@@
 \end{enumerate}
+
+\FigureWI{stages}
+{Two stages of interpolation: First, from triangular with 6DOF nodes (a) a mesh of control points (b) is computed.
+Then based on definition of the B\'ezier triangle a surface is interpolated.}
 
@@
 
-\Figure{0.8\linewidth}{bezier}
+%\Figure{0.8\linewidth}{bezier}
+%{Cubic B\'ezier triangle with mesh of 10 control points.}
+
+\Figure{0.8\linewidth}{bezier2}
 {Cubic B\'ezier triangle with mesh of 10 control points.}
@@
 
-\Figure{0.8\linewidth}{segments}
-{Correspondence between edge points and associated corner points shown in
-blue. \CD{Not clear !!}}
+%\Figure{0.8\linewidth}{segments}
+%{Correspondence between edge points and associated corner points shown in
+%blue. \CD{Not clear !!}}
+
+\Figure{0.8\linewidth}{segments2}
+{Internal nodes of control mesh are connected to the corner nodes by kinematic link (shown in blue).}
 
@@
 \Figure{\linewidth}{roof}
-{Convergence for modified Scordelis-Lo roof test of DKT+ANDES element and
-our B\'ezier shell element. \CD{Can you do the same test with the model used by Comas ?}}
+{Convergence for modified Scordelis-Lo roof test of DKT+ANDES element, 
+our B\'ezier shell element (BSH) and elemnt of O. Comas (PSH) with polynomial shape function.}
+
+\Figure{\linewidth}{roof-img}
+{Coloured plot of vertical displacement in the roof test}
 
@@
 {Convergence for test on hemisphere with 18° hole for the DKT+ANDES element and
-our B\'ezier shell element. \CD{Same question: can you do the same test with the model we used with O.Comas ? Hopefully, it will show that our model has a better convergence ... }}
+our B\'ezier shell element (BSH) and elemnt of O. Comas (PSH) with polynomial shape function.}
+
+\Figure{\linewidth}{hemisphere-img}
+{Coloured plot of radial displacement in the hemisphere test}
 

### 2012-07-31 Doc/VRIPHYS2012/hemisphere.R

@@ -23,7 +23,10 @@ lines(n2e(hem$size[hem$name == n]), hem$d[hem$name == n], type="b", col="black",
 n<- "bezier2";
 lines(n2e(hem$size[hem$name == n]), hem$d[hem$name == n], type="b", col="black", pch=4);
 
-legend("bottomright", c("DKT","BSH"), pch=c(1,4), col=c("black","black"), bg="white");
+n<- "triangular";
+lines(n2e(hem$size[hem$name == n]), hem$d[hem$name == n], type="b", col="black", pch=5);
+
+legend("bottomright", c("DKT","BSH","PSH"), pch=c(1,4,5), col="black", bg="white");
 
 dev.off()
 par(p);

### 2012-07-31 Doc/VRIPHYS2012/roof.R

@@ -22,7 +22,10 @@ lines(n2e(lo$size[lo$name == n]), lo$z[lo$name == n], type="b", col="black", pch
 n<- "bezier2";
 lines(n2e(lo$size[lo$name == n]), lo$z[lo$name == n], type="b", col="black", pch=4);
 
-legend("topright", c("DKT","BSH"), pch=c(1,4), col=c("black","black"), bg="white");
+n<- "triangular";
+lines(n2e(lo$size[lo$name == n]), lo$z[lo$name == n], type="b", col="black", pch=5);
+
+legend("topright", c("DKT","BSH", "PSH"), pch=c(1,4,5), col=c("black","black","black"), bg="white");
 
 dev.off()
 par(p);

### 2012-07-31 Doc/VRIPHYS2012/vriphys2012.tex
article: reviewed end of section 2 and most of section 3
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@441 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 % instead of the author's name (and leave the affiliation blank) !!
-% TODO: anonymize
-\author[Authors]
-       {Authors} % TODO
+% TODO: anonymize -- Submission ID
+\author[T. Golembiovský \& C. Duriez]
+       {Tomáš Golembiovský$^{1,2}$ and Christian Duriez$^{2}$\\
+        $^1$Faculty of Informatics, Masaryk University, Czech Republic\\
+        $^2$INRIA Lille, France
+       } % TODO: ask Christian how to specify affiliation
 
@@
 {Two stages of interpolation: First, from triangular with 6DOF nodes (a) a mesh of control points (b) is computed.
-Then based on definition of the B\'ezier triangle a surface is interpolated.}
+Then based on definition of the B\'ezier triangle a surface (c) is interpolated.}
 
@@
 In the rest of the document we define the surface over the triangle:
+%
 \begin{equation}
@@
 \end{equation}
-
+%
 B\'ezier triangles were first designed in early sixties by French mathematician de Casteljau.
@@
 2)/2$ control points and it's surface is defined as:
-
+%
 \begin{equation}
@@
 \end{equation}
-
-\noindent
+%
 where $P_{i,j}$ are the control points and $B^n_{i,j}$ are the bivairate
 Bernstein basis functions defined as:
-
+%
 \begin{equation}
@@
 \end{equation}
-
-\noindent
+%
 where $ 0 \le i+j \le n $.
@@
 mesh. Cubic triangle is described by 10 control points and the surface is
-explicitly defined as (see fig. \ref{fig-bezier}):
-
+explicitly defined as (see fig. \ref{fig-bezier2}):
+%
 \begin{equation}\label{eq-cubicbez}
@@
 \end{equation}
-
-\noindent
+%
 where $ \xi_3 = 1 - \xi_1 - \xi_2 $. By naming $n_i$ the respective values
@@
 shortly express \eqref{eq-cubicbez} as:
-
+%
 \begin{equation}\label{eq-cubicbez2}
@@
 \end{equation}
-
-%\Figure{0.8\linewidth}{bezier}
-%{Cubic B\'ezier triangle with mesh of 10 control points.}
 
@@
 For simplicity we decided to keep the edge boundary only $C^0$ continuous and we compute the central point as a function of other 9 points: 
-
+%
 \begin{equation}\label{eq-central}
@@
 \end{equation}
-
-\noindent
-where the points $P_i$ are named as in Figure \ref{fig-bezier}.
-% Equation eq-central is better explained below
+%
+where the points $P_i$ are named as in Figure \ref{fig-bezier2}.
 
@@
 Each of the edge control points is attached to nearest corner node. See Figure
-\ref{fig-segments} for the correspondence between edge points and
+\ref{fig-segments2} for the correspondence between edge points and
 associated corner nodes. 
@@
 
-%\Figure{0.8\linewidth}{segments}
-%{Correspondence between edge points and associated corner points shown in
-%blue. \CD{Not clear !!}}
-
 \Figure{0.8\linewidth}{segments2}
@@
 In the initialization phase we remember the position of edge point $j$ in the frame of corner point $i$:
-
+%
 \begin{equation}
@@
 \end{equation}
-
-\noindent
+%
 where
@@
 apply the rigid transformation of frame $i$ on control point $j$:
-
+%
 \begin{equation}
@@
 
-For the central point we again apply interpolation of \eqref{eq-central} which is kind of skinning function \CD{Why so complicated ?
-You can compute  some rules to chose the first location of point $P_{10}$ (i.e. to compute $P_{i-10}$) but then use something more simple for the kinematics}:
-
-\begin{equation}\label{eq-central}
-    \begin{split}
-        P_{10} = & \frac{1}{3}(\sum_{i=4}^9 P_i - \sum_{i=1}^3 P_i)\\
-               = & \frac{1}{3} \left[ P_1 + (P_{1-4} + P_{1-5}) \right] + \\
-                & + \frac{1}{3} \left[ P_2 + (P_{2-6} + P_{2-7}) \right] + \\
-                & + \frac{1}{3} \left[ P_3 + (P_{3-8} + P_{3-9}) \right]
+Similarily for the central point we apply rigid transformation of all frames of the triangle and compute the mean:
+%
+\begin{equation}\label{eq-central2}
+    P_{10} =  \frac{1}{3}\left( \sum_{i=1}^3 \mathbf{x}_i + R(\theta_i) \overline{P_{i-10}} \right)
+\end{equation}
+
+Puting previous equations into \eqref{eq-cubicbez2} we get:
+%
+\begin{equation}\label{global-transform} \begin{split}
+    & T = \sum_{i=1}^{10} n_i P_i = \sum_{i=1}^{3} n_i \mathbf{x}_i
+    + \sum_{j=4}^9 n_j  \left(
+        \mathbf{x}_i + R(\theta_i) \overline{P_{i-j}} \right) + \\
+    &\quad + \frac{n_{10}}{3}
+        \left( \sum_{i=1}^3 \mathbf{x}_i + R(\theta_i) \overline{P_{i-10}} \right)
     \end{split}
@@
 
-\CD{I would have put (but that is maybe the same result...)}
-
-\begin{equation}\label{eq-central}
-    \begin{split}
-        P_{10} =  \frac{1}{3}\left( \sum_{i=1}^3 \mathbf{x}_i + R(\theta_i) \overline{P_{i-10}} \right)
-    \end{split}
-\end{equation}
-
-%}}}
-
-\CD{and it could be possible to write down (and this is helpful for the formulation of the spacial derivative):}
-\begin{equation}\label{global-transform}
-    \begin{split}
- T(\xi_1,\xi_2, \xi_3) = \sum_{i=1}^{10} n_i(\xi_1,\xi_2, \xi_3)P_i = \\
- 	\sum_{i=1}^{3} n_i(\xi_1,\xi_2, \xi_3) \mathbf{x}_i +  \sum_{j=4}^9 n_j (\xi_1,\xi_2, \xi_3)( \mathbf{x}_i +  R(\theta_i) \overline{P_{i-j}}) + \\  \frac{n_{10}(\xi_1,\xi_2, \xi_3)}{3}\left( \sum_{i=1}^3 \mathbf{x}_i + R(\theta_i) \overline{P_{i-10}} \right)
-    \end{split}
-\end{equation}
+% }}}
 
@@
 %        & + n_{10} \frac{1}{3} \left(
-%        \sum_{j=4}^9 (P_i + U_i + P_{i-j} + P_{i-j} \times \theta_i) - \sum_{i=1}^3 (P_i + U_i)
+%        \sum_{i=1}^3 (P_i + U_i + P_{i-10} + P_{i-10} \times \theta_i)
 %        \right) \\
@@
 %      + n_9 (U_3 + P_{3-9} \times \theta_3) \\
-%      & + n_{10} \frac{1}{3} (U_1 + U_2 + U_3 + \sum_{j=4}^9 P_{i-j} \times \theta_i) \\
+%      & + n_{10} \frac{1}{3} (U_1 + U_2 + U_3 + \sum_{i=1}^3 P_{i-10} \times \theta_i) \\
 %%
@@
 %      & + \sum_{j=4}^9 n_j P_{i-j} \times \theta_i
-%      + n_{10} \frac{1}{3} \sum_{j=4}^9 P_{i-j} \times \theta_i \\
+%      + n_{10} \frac{1}{3} \sum_{i=1}^3 P_{i-10} \times \theta_i \\
 %%
@@
 A small variation $\delta \mathbf{x}_i$ of the node $i$ position and $\delta \mathbf{\theta}_i$ for the rotation produces the following small variation of position corner points $\delta P_i$ and edge points $\delta P_j$:
-
+%
 \begin{eqnarray}
@@
 \end{eqnarray}
-
-\noindent
-and variation of central point $\delta P_{10}$ \CD{ this formulation would be modified...} :
-
-\begin{equation}
-    \delta P_{10} = \frac{1}{3} (\sum_{i=1}^3 \delta \mathbf{x}_i + \sum_{j=4}^9 P_{i-j} \times \delta \mathbf{\theta}_i)
+%
+and variation of central point $\delta P_{10}$:
+%
+\begin{equation}
+    \delta P_{10} = \frac{1}{3} \sum_{i=1}^3 \left(
+    \delta \mathbf{x}_i + P_{i-10} \times \delta \mathbf{\theta}_i \right)
 \end{equation}
@@
 Thus, finally, it creates a variation of the point position on the surface $\delta T$:
+%
 \begin{equation}
@@
 \end{equation}
-
-\noindent
+%
 or:
@@
                    & + (n_3 + n_8 + n_9 + \frac{1}{3} n_{10})  \delta \mathbf{x}_3 + \\
-                   & + \sum_{j=4}^9 P_{i-j} \times \delta \mathbf{\theta}_j + \\
-                   & + n_{10} \frac{1}{3} \sum_{j=4}^9 P_{i-j} \times \delta \mathbf{\theta}_i 
+                   & + \sum_{j=4}^9 n_j P_{i-j} \times \delta \mathbf{\theta}_i + \\
+                   & + \frac{n_{10}}{3} \sum_{i=1}^3 P_{i-10} \times \delta \mathbf{\theta}_i 
     \end{split}
 \end{equation}
-
-\noindent
-thus computing the physical deflection $u$ as variation of the point position $\delta T = u$.
 
@@
 Note however that the expressions are sound and unambiguous because there is only one possible $i$ for each $j$ depending on the connection of segments to the corner nodes (see fig.
-\ref{fig-segments}).
+\ref{fig-segments2}).
 
@@
 
-\section{Finite element formulation} % {{{
+\section{Finite Element Formulation} % {{{
 
@@
 
-\subsection{Computation of the strain tensor}
+\subsection{Computation of the Strain Tensor} % {{{
 
@@
 
-We define a local system of coordinates $(x,y,z)$ in which the axis z is oriented along the thickness of the element\footnote{The computation of the frame of each element will be detailed in section \ref{sec-corot}.
-}.
+We define a local system of coordinates $(x,y,z)$ in which the axis z is oriented along the thickness of the element (the computation of the frame of each element will be detailed in section \ref{sec-corot}).
 Given the initial position of the triangle points in this system of coordinate, $\{x_1^0, y_1^0\}$,  $\{x_2^0, y_2^0\}$, and  $\{x_3^0, y_3^0\}$, we can compute a linear relation between $\{\xi_1,\xi_2, \xi_3\}$ and $\{ x,y \}$:
-
+%
 \begin{equation}
@@
 
-Thus, we can locally compute some spacial derivatives, along $x$ and $y$ on the interpolation function vector $T$ (which has three components $(T^x, T^y, T^z)$) .
-The 2D tensor $F$ of the in-plane deformations can be computed as a $2\times2$ matrix where the line $k$ and the column $l$ alternatively represent $x$ and $y$\footnote{It can be noted $k=\{x,y\}$ and $l=\{x,y\}$}:
-\begin{equation}
-F_{kl} = \frac{\partial T^k}{\partial l}  =  \sum_{i=1}^3 \frac{\partial T^k}{\partial \xi_i} \frac{\partial \xi_i}{\partial l}  
-\end{equation}
-
-The term $\frac{\partial T^k}{\partial \xi_i}$ can be derived from the expression of $T(\xi_1,\xi_2, \xi_3)$ in equation (\ref{global-transform}) and 
-$\frac{\partial \xi_i}{\partial l}$ from equation (\ref{shape-vector}). 
+Thus, we can locally compute some spatial derivatives, along $x$ and $y$ on the interpolation function vector $T$ (which has three components $(T^x, T^y, T^z)$) .
+The 2D tensor $F$ of the in-plane deformations can be computed as a $2\times2$ matrix where the line $k$ and the column $l$ alternatively represent $x$ and $y$:
+%
+\begin{equation}
+    F_{kl} = \deriv{T^k}{l}  = 
+    \sum_{i=1}^3 \deriv{T^k}{\xi_i} \deriv{\xi_i}{l};
+    \quad k \in \{x,y\}, l \in \{x,y\}
+\end{equation}
+
+The term $\deriv{T^k}{\xi_i}$ can be derived from the expression of $T(\xi_1,\xi_2, \xi_3)$ in equation (\ref{global-transform}) and 
+$\deriv{\xi_i}{l}$ from equation (\ref{shape-vector}). 
 The in-plane displacement gradient can be obtained with tensor $F$:
-\begin{equation}
-\frac{\partial u^k}{\partial l} = F_{kl} - I_{kl} 
-\end{equation}
+%
+\begin{equation}
+\deriv{u^k}{l} = F_{kl} - I_{kl} 
+\end{equation}
+%
 where $I$ is the identity matrix ($I_{kl}  = \{1$ if $l=k$ ;  $0$ if $l \neq k \} $). 
-%
-%
-% Remark: the sentence can be suppr..
-\CD{The following  sentencd I let you chose: keep or not...}The in-plane displacement gradient can be linked to the degrees of freedom of the shell using the expression of the displacement of a point $\delta T$ (see equation  (\ref{bez-delta})) on the shell surface:
-$$
-\frac{\partial u^k}{\partial l} = \frac{\partial (\delta T^k)}{\partial l}
-$$ 
-% end of Remark
+The displacement gradient can be linked to the degrees of freedom of the shell using the expression of the displacement of a point $\delta T$ (see equation  (\ref{bez-delta})) on the shell surface:
+\begin{equation}
+    \nabla u = \nabla(\delta T)
+\end{equation}
 
 In the local frame of the element, we rely on infinitesimal strain theory, thus the expression of the strain tensor (using Voigt notations) is:
-\begin{equation}
-\label{in-plane-strain}
-\left[ \begin{array}[c]{c} \epsilon_{xx} \\ \epsilon_{yy}  \\ 2\epsilon_{yx} \end{array} \right] = 
-\left[ \begin{array}[c]{c} \frac{\partial u^x}{\partial x}  \vspace{1mm}\\ \frac{\partial u^y}{\partial y}   \\ \frac{\partial u^x}{\partial y}  + \frac{\partial u^y}{\partial x}  \end{array} \right]
-\end{equation}
-
-For the bending formulation, the shell theory uses a measure of the out-of-plane-displacement $u^z(x,y)$. 
-The strain tensor has additional contributions based on Kirchhoff-Love theory for thin plates:
-\begin{equation}
-\label{out-plane-strain}
-\left[ \begin{array}[c]{c} \epsilon_{xx} \\ \epsilon_{yy}  \\ 2\epsilon_{yx} \end{array} \right] +=
-\left[ \begin{array}[c]{c} -z \frac{\partial^2 u^z}{\partial x^2}  \vspace{1mm}\\ -z \frac{\partial^2 u^z}{\partial y^2}   \vspace{1mm}\\ - 2z \frac{\partial^2 u^z}{\partial xy}    \end{array} \right]
-\end{equation}
-%% comment
-\CD{Here, if you want to keep a separate computation for bending and membrane (see the folllowing), use $=$ instead of $+=$ !!}
-%%
+\begin{equation}\label{eq-in-plane-strain}
+    \begin{bmatrix}
+        \epsilon_{xx} \\ \epsilon_{yy} \\ 2\epsilon_{yx}
+    \end{bmatrix}
+    = 
+    \begin{bmatrix}
+        \deriv{u^x}{x}  \\[1mm]  \deriv{u^y}{y}  \\[1mm]  \deriv{u^x}{y} + \deriv{u^y}{x}
+    \end{bmatrix}
+\end{equation}
+
+For the bending formulation, the shell theory uses a measure of the out of plane-displacement $u^z(x,y)$
+The strain tensor is based on Kirchhoff-Love theory for thin plates:
+\begin{equation}\label{out-plane-strain}
+    \begin{bmatrix}
+        \epsilon_{xx} \\ \epsilon_{yy} \\ 2\epsilon_{yx}
+    \end{bmatrix}
+    =
+    \begin{bmatrix}
+        -z \deriv{^2 u^z}{x^2}  \\[1mm]  -z \deriv{^2 u^z}{y^2}  \\[1mm]  - 2z \deriv{^2 u^z}{xy}
+    \end{bmatrix}
+\end{equation}
 where $z$ is the local coordinate along $z$-axis of the point on which the strain is measured.
 
-As the term  $\frac{\partial \xi_i}{\partial l}$ provides a constant value, the second derivatives can be computed quite easily, for instance, for $\epsilon_{xx}$:
-\begin{equation}
-\frac{\partial^2 u^z}{\partial x^2}  = \sum_{i=1}^3\frac{\partial}{\partial \xi_i}\left(\sum_{j=1}^3 \frac{\partial T^z}{\partial \xi_j} \frac{\partial \xi_j}{\partial x}\right) \frac{\partial \xi_i}{\partial x}
-=  \sum_{i=1}^3  \sum_{j=1}^3 \frac{\partial^2 T^z}{\partial \xi_i\xi_j}   \frac{\partial \xi_j}{\partial x} \frac{\partial \xi_i}{\partial x}  
-\end{equation}
-
-\subsection{Shell Element}
-
-As previously describes, the shell element's strain is a combination of two types of deformations, as shown in figure \ref{deformations}.
-If we assume that the initial out-plane deflections are negligible and that the frame of the shell element nodes are aligned with local $z$ axis, we can distribute these two deformations into two groups of degrees of freedom  at the nodes' level.
+As the term $\deriv{\xi_i}{l}$ provides a constant value, the second derivatives can be computed quite easily, for instance, for $\epsilon_{xx}$:
+\begin{equation}
+    \deriv{^2 u^z}{x^2}
+    = \sum_{i=1}^3 \deriv{}{\xi_i}\left(
+        \sum_{j=1}^3 \deriv{T^z}{\xi_j} \deriv{\xi_j}{x}
+    \right) \deriv{\xi_i}{x} 
+    = \sum_{i=1}^3 \sum_{j=1}^3  \deriv{^2 T^z}{\xi_i\xi_j}  \deriv{\xi_j}{x}  \deriv{\xi_i}{x}  
+\end{equation}
+
+% }}}
+
+\subsection{Shell Element} % {{{
+
+As previously described, the shell element's strain is a combination of two types of deformations, as shown in figure \ref{deformations}.
+If we assume that the initial out of plane deflections are negligible and that the frame of the shell element nodes are aligned with local $z$ axis, we can distribute these two deformations into two groups of degrees of freedom  at the nodes' level.
 
 \FigureWI{deformations}
-{Plate deformations:%
+{Plate deformations:
 a) membrane deformations like stretching or shearing, due to in-plane displacements
 b) undeformed;
-c) bending deformation, due to out-plane displacements
+c) bending deformation, due to out of plane displacements
 \label{deformations}}
@@
     At the nodes' level, it corresponds to in-plane displacements $\delta\mathbf{x}_x, \delta\mathbf{x}_y$ and rotation $\theta_z$.
-(Figure \ref{fig-membrane}).
+
     \item Bending plate defining deformation due to out of plane bending displacement $u_z$. It is
     influenced by 3 degrees of freedom: rotation around two axes $\theta_x,
-    \theta_z$ and out of plane displacement $\delta\mathbf{x}_z$ (Figure \ref{fig-plate}).
+    \theta_z$ and out of plane displacement $\delta\mathbf{x}_z$.
 
@@
 
-\subsection{Stiffness matrix} % {{{
+% }}}
+
+\subsection{Stiffness Matrix} % {{{
 
 In order to compute FEM-based stiffness for the shell element, we need to compute the jacobian matrix between strains and nodes' displacements.
-As we have separated the formulations for the in-plane and the out-plane displacements, it will result with the computation of two jacobian matrices (of size $3\times9$).
+As we have separated the formulations for the in-plane and the out of plane displacements, it will result with the computation of two jacobian matrices (of size $3\times9$).
 \begin{equation}
@@
 \begin{eqnarray}
-    V_i = V_j + \omega_j \times (P_i - P_j)
+    V_j = V_i + \omega_i \times (P_j - P_i)
 \end{eqnarray}
@@
 $\omega_j$ is the angular velocity at the node. For the central node we
-again use the interpolation as in \eqref{eq-central}.
+the formula is again similar to \eqref{eq-central2}:
+
+\begin{eqnarray}
+    V_i = \frac{1}{3} \sum_{i=1}^3 V_i + \omega_i \times (P_{10} - P_i)
+\end{eqnarray}
 

### 2012-08-01 Doc/VRIPHYS2012/fps.R
fixing FPS for the test (only visual mapping)
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@442 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@ -7,13 +7,13 @@ fps <- read.csv("fps.csv");
 pdf("fps.pdf", width=7, height=4);
 p <- par(lwd=2, mar=c(4,4,1,1)+0.1);
 
-plot(range(fps$elements), range(c(fps$fpsBase, fps$fpsPolar, fps$fpsMap1, fps$fpsMap2)),
+plot(range(fps$elements), range(c(fps$fpsBase, fps$fpsPolar, fps$fpsMap1, fps$fpsMap2), na.rm=TRUE),
     type="n", xlab="Elements", ylab="FPS", log="y");
 
 lines(fps$elements, fps$fpsBase, type="b", col="black", pch=1);
 lines(fps$elements, fps$fpsPolar, type="b", col="black", pch=4);
-lines(fps$elements, fps$fpsMap2, type="b", col="black", pch=5);
-lines(fps$elements, fps$fpsMap1, type="b", col="black", pch=6);
+lines(fps$elements, fps$fpsMap1, type="b", col="black", pch=5);
+lines(fps$elements, fps$fpsMap2, type="b", col="black", pch=6);
 
 abline(h=25, lty="dotted");
 

### 2012-08-01 Doc/VRIPHYS2012/vriphys2012.tex
sections 3, 4 and most of 5
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@443 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \newcommand{\deriv}[2]{\frac{\partial #1}{\partial #2}}
+% Matrix
+\newcommand{\mat}[1]{\mathbf{#1}}
 
@@
 
-\section{Element Kinematics} % {{{
+\section{Element Kinematics}\label{sec-kinematics} % {{{
 
@@
 
-\begin{equation}\label{bez-delta}
+\begin{equation}\label{eq-bez-delta}
     \begin{split}
@@
 %
-where $I$ is the identity matrix ($I_{kl}  = \{1$ if $l=k$ ;  $0$ if $l \neq k \} $). 
-The displacement gradient can be linked to the degrees of freedom of the shell using the expression of the displacement of a point $\delta T$ (see equation  (\ref{bez-delta})) on the shell surface:
+where $\mat{I}$ is the identity matrix ($I_{kl}  = \{1$ if $l=k$ ;  $0$ if $l \neq k \} $). 
+The displacement gradient can be linked to the degrees of freedom of the shell using the expression of the displacement of a point $\delta T$ (see equation  (\ref{eq-bez-delta})) on the shell surface:
 \begin{equation}
@@
 \begin{equation}
-\delta \left[ \begin{array}[c]{c} \epsilon_{xx} \\ \epsilon_{yy}  \\ 2\epsilon_{yx} \end{array} \right] = 
-\left[ \mathbf{J}_m \right] 
-\left[ \begin{array}[c]{c} \delta \mathbf{x}_{1x} \\ \delta \mathbf{x}_{1y} \\ \delta \theta_{1z} \\ \vdots    \end{array} \right]
-\end{equation}
-where $\left[ \mathbf{J}_m \right]$ maps the membrane in-plane displacements and 
-%
-\begin{equation}
-\delta \left[ \begin{array}[c]{c} \epsilon_{xx} \\ \epsilon_{yy}  \\ 2\epsilon_{yx} \end{array} \right] = 
-\left[ \mathbf{J}_b \right] 
-\left[ \begin{array}[c]{c} \delta \mathbf{x}_{1z} \\ \delta \theta_{1x} \\ \delta \theta_{1y} \\ \vdots    \end{array} \right]
-\end{equation}
-where $\left[ \mathbf{J}_b \right]$ maps the bending out-plane displacements.
-
-To compute the values in matrices $\left[ \mathbf{J}_m \right]$ and $\left[ \mathbf{J}_b \right]$,  we can reuse equation (\ref{bez-delta}) and apply the spacial derivatives.
-To illustrate, let's take one value of the matrix $\left[ \mathbf{J}_m \right]$  (first line, fourth column), we first take the kinematic between $\delta T_x$ and the concerned degree of freedom $ \delta \mathbf{x}_{2x}$ from equation (\ref{bez-delta}):
-$$
-   \frac{ \delta T_x}{ \delta \mathbf{x}_{2x}} = (n_2 + n_6 + n_7 + \frac{1}{3} n_{10})
-$$
-Then, we apply the spacial derivatives to compute the variation of $\epsilon_{xx}$ due to a variation of $\mathbf{x}_{2x}$.
-$$
+    \delta 
+    \begin{bmatrix}
+        \epsilon_{xx} \\ \epsilon_{yy}  \\ 2\epsilon_{yx}
+    \end{bmatrix}
+    = \begin{bmatrix} \mat{J}_m \end{bmatrix}
+    \begin{bmatrix}
+        \delta \mat{x}_1^x \\ \delta \mat{x}_1^y \\ \delta \theta_1^z \\ \vdots
+    \end{bmatrix}
+\end{equation}
+where $\left[ \mat{J}_m \right]$ maps the membrane in-plane displacements and 
+%
+\begin{equation}
+    \delta
+    \begin{bmatrix}
+        \epsilon_{xx} \\ \epsilon_{yy}  \\ 2\epsilon_{yx}
+    \end{bmatrix}
+    = \begin{bmatrix} \mat{J}_b \end{bmatrix}
+    \begin{bmatrix}
+        \delta \mat{x}_1^z \\ \delta \theta_1^x \\ \delta \theta_1^y \\ \vdots
+    \end{bmatrix}
+\end{equation}
+where $\left[ \mat{J}_b \right]$ maps the bending displacements.
+
+To compute the values in matrices $\left[ \mat{J}_m \right]$ and $\left[ \mat{J}_b \right]$,  we can reuse equation (\ref{eq-bez-delta}) and apply the spatial derivatives.
+To illustrate, let's take one value of the matrix $\left[ \mat{J}_m \right]$  (first line, fourth column), we first take the kinematics between $\delta T_x$ and the concerned degree of freedom $ \delta \mathbf{x}_2^x$ from equation (\ref{eq-bez-delta}):
+\begin{equation}
+    \frac{ \delta T_x }{ \delta \mathbf{x}_2^x } = (n_2 + n_6 + n_7 + \frac{n_{10}}{3})
+\end{equation}
+
+Then, we apply the spatial derivatives to compute the variation of $\epsilon_{xx}$ due to a variation of $\mathbf{x}_{2x}$:
+\begin{equation}
     \delta \epsilon_{xx} += 
-\underbrace{\left( \sum_{i=1}^{3} (\frac{\partial n_2}{\xi_i} + \frac{\partial n_6}{\xi_i} + \frac{\partial n_7}{\xi_i} + \frac{1}{3}\frac{\partial n_{10}}{\xi_i})\frac{\partial \xi_i}{\partial x} \right) }_{\left[ \mathbf{J}_m \right](1,4)}
-{\delta \mathbf{x}_{2x}}
-$$
+    \underbrace{\left(
+    \sum_{i=1}^{3} (\deriv{n_2}{\xi_i} + \deriv{n_6}{\xi_i} + \deriv{n_7}{\xi_i} + \frac{1}{3}\deriv{n_{10}}{\xi_i})
+    \deriv{\xi_i}{x} \right) }_{\begin{bmatrix} \mat{J}_m \end{bmatrix}_{14}}  
+ \delta \mathbf{x}_2^x 
+\end{equation}
 
@@
     \label{eq-Km}
-    \mathbf{K}_m & = \iiint_V \mathbf{J}_m^T M \mathbf{J}_m \, \mathrm{d} V \\
+    \mat{K}_m & = \iiint_V \mat{J}_m^T \mat{M} \mat{J}_m \, \mathrm{d} V \\
     \label{eq-Kb}
-    \mathbf{K}_b & = \iiint_V \mathbf{J}_b^T M \mathbf{J}_b \, \mathrm{d} V
+    \mat{K}_b & = \iiint_V \mat{J}_b^T \mat{M} \mat{J}_b \, \mathrm{d} V
 \end{align}
-%% comment
-\CD{This is something that is maybe wrong in our formulation...
-as the bending adds a contribution on the deformation tensor, maybe we should not separate the contribution from membrane and bending as we do now ($K_m$ and $K_b$ are computed separately) }
-%% comment
-\CD{here, it would be...
-$\mathbf{K}_e =  \iiint_V \left[\begin{matrix} \mathbf{J}_m & \mathbf{J}_b \end{matrix} \right]^T  M \left[\begin{matrix} \mathbf{J}_m & \mathbf{J}_b \end{matrix} \right]$.
-Could we test that option easily in the code, or is it a nightmare ??}
-%%
 
 \noindent
-where $M$ is the material matrix. To keep the system simple we use linear
+where $\mat{M}$ is the material matrix. To keep the system simple we use linear
 Hooke's law for isotropic materials.
 
-Because the deflection field for the element is non-linear \eqref{eq-delta}
+Because the deflection field for the element is non-linear \eqref{eq-bez-delta}
 in position the integrals \eqref{eq-Km} and \eqref{eq-Kb} have to be computed using
 numerical integration. In our implementation we employ 6-point Gaussian
-quadrature for integration over triangle area \CD{TODO: figure + gives the formula of the quadrature...}
+quadrature for integration over triangle area.
+\CD{TODO: figure + gives the formula of the quadrature...}
 % TODO: add ref
@@
 
-It is known that the Cauchy's strain tensor \eqref{eq-cauchy} is not
-rotation invariant \footnote{add ref} and produces ghost forces for rigid
-rotations. The alternative approach is to use Green's strain tensor,
-however it leads to nonlinear system. To keep the system linear and to deal
+It is known that the Cauchy's strain tensor \eqref{eq-in-plane-strain} is not
+rotation invariant \cite{Hauth2004,Muller2004} and produces ghost forces for rigid
+rotations. The alternative approach is to use non-linear Green's strain tensor but the non-linearity introduced into the system is costly and introduces instabilities.
+To keep the system linear and to deal
 with rigid body rotations we compute the displacements and forces in
@@
 is then computed relative to this frame:
-
-\begin{equation}
-    F_e = R_e^T (K_e u) = R_e^T \left(K_e(R_e x - R_e^0 x^0)\right)
-\end{equation}
-
-\noindent
-where $R$ and $R_0$ are rotation matrices of the frame for current and rest
+%
+\begin{equation}
+    F_e = \mat{R}_e^T (\mat{K}_e u) = \mat{R}_e^T \left(\mat{K}_e(\mat{R}_e x - \mat{R}_e^0 x^0)\right)
+\end{equation}
+%
+where $\mat{R}$ and $\mat{R}_0$ are rotation matrices of the frame for current and rest
 mesh respectively.
 
-TODO: how we compute frame?
-%from the position of the corner nodes and we align one axis of the frame with one
-%edge of the element. \footnote{add image}
-
-%The displacement field defined in \eqref{eq-delta} is clearly symmetric in
-%respect to deformations, but ...
-
-This choice of the frame aligns one axis of the frame with one of the edges
-of the triangle. Correct choice of the corotational frame is a tricky
+The corotational frame is computed followingly:
+
+\begin{itemize}
+    \item One axis is alligned to one edge of the element:\\
+    $\mathbf{e}_1 = \mathbf{x}_2 - \mathbf{x}_1$
+    \item A normal to the plane of the element is computed:\\
+    $\mathbf{e}_3 = \mathbf{e}_1 \times (\mathbf{x}_3 - \mathbf{x}_1)$
+    \item Second axis in the plane of the element is computed:\\
+    $\mathbf{e}_2 = \mathbf{e}_1 \times \mathbf{e}_3$
+\end{itemize}
+
+Correct choice of the corotational frame is a tricky
 problem \cite{Felippa2005} and incorrect frame for the membrane element can
-have adverse effects on the simulation. In our case we have experienced
+have adverse effects on the simulation.
+This choice of the frame where we align one axis with one of the edges
+of the triangle is not a perfect one and we have experienced
 nonnegligible difference in deformation of equilateral triangle depending
@@
 
-Performing a polar decomposition on the deformation gradient $F = \nabla u
-+ I$ can be used to extract a rotation component from the deformation of
-the element \cite{Ciarlet1994}. Because we are only fixing the membrane
+A polar decomposition on the deformation gradient $F = \nabla u
++ \mat{I}$ can be used to extract a rotation component from the deformation of
+the element \cite{Ciarlet1994,Hauth2004}. Because we are only fixing the membrane
 element we can perform the polar decomposition in 2D with the following
 formula:
-
-\begin{equation}
-    R = F + \mathrm{sgn}(\mathrm{det}(F)) \begin{pmatrix}
+%
+\begin{equation}
+    \mat{R} = \mat{F} + \mathrm{sgn}(\mathrm{det}(\mat{F})) \begin{pmatrix}
         F_{22} & - F_{21} \\
@@
 \end{equation}
-
-\noindent
+%
 and subsequently normalizing the columns. We then rotate the corotational
-frame in inverse direction. Even single step of this process improves the
+frame in opposite direction. Even single step of this process improves the
 final frame but more iterations of the polar decompositions are usually
-necessary (up to 20) which severely degrades the performance. But polar
+necessary (up to 20) which severely degrades the performance. Polar
 decomposition tends to "overshoot" the ideal rotation angle and each
-subsequent step performs a rotation in the opposite direction. By properly
+subsequent step performs a rotation in the direction opposite to the rotation in previous step. By properly
 scaling the rotation angle we are able to minimize the amount of needed
@@
 We employ two stop conditions for the iteration process:
-\begin{itemize}
-    \item The change in rotation angle is less than $10^{-6}$,
-    \item and the maximum number of 5 iterations.
-\end{itemize}
+(i) the change in rotation angle is less than $10^{-6}$, and
+(ii) the maximum number of iterations is set to 5.
 
 If the first condition is not met in 5 iterations it can mean one of three
-things:
-
-\begin{enumerate}
-    \item We still have very good approximation of the ideal frame.
-    \item The convergence for the element is too slow.
-    \item The process is diverging.
-\end{enumerate}
-
-\noindent
-and performing any more iterations is unnecessary.
+things we still have very good approximation of the ideal frame. In some
+situations the process is not able to reach the ideal frame quickly because
+either the convergence for the element is too slow or the process is diverging.
+In either case performing more iterations is unnecessary.
 
@@
 
-\section{Visual and Mechanical Mapping} % {{{
+\section{Application of High Resolution Mesh} % {{{
 
@@
 interaction and correct simulation of collisions and constraints. To handle
-these we have also to know how to map the velocities from the mechanical
+these we also need to know how to map the velocities from the mechanical
 mesh onto the high resolution mesh and how to map forces acting on the high
@@
 
-\subsection{Mapping the Surface Mesh}
-
-We use the fact that the geometry of the shell is based on the formulation
+\subsection{Surface Mesh}
+
+To project vertices onto the surface of shells 
+we use the fact that the geometry of is based on the formulation
 of cubic B\'ezier triangle \eqref{eq-cubicbez}. For every vertex of the high
@@
 mesh that is closest to the vertex and assign barycentric coordinates on
-the triangle to this vertex. After every step of the simulation the high
+the triangle to this vertex. After every step of the simulation points of the
+control mesh are updated based on the kinematic link described in
+section \ref{sec-kinematics} and the high
 resolution mesh is updated using the assigned barycentric coordinates and
@@
 
-\subsection{Mapping Velocities}
-
-In the
-
+\subsection{Nodal Velocities of the Surface Mesh}
+
+For contact modeling it is often necessary to know the velocity of each point.
+The values are directly available to us for the nodes of the mechanical mesh
+but they can also be computed for any point on the surface.
 Differentiating \eqref{eq-cubicbez2} by time we get:
-
+%
 \begin{equation}
@@
 
-Which means we use the same expression to interpolate the velocities only
+Which means we use the same expression to interpolate the velocity only by
 substituting the velocities at control points. For the corner nodes we
 already have the velocities $V_i$. For the internal nodes we need to compute
-them. In case of the edge nodes it is a problem of computing the velocity
-of a point attached to the moving rigid body:
-
+them. We again make use of the kinematic link between nodes and solve the
+problem of computing the velocity of a point attached to the moving rigid body.
+For the edge points we have:
+%
 \begin{eqnarray}
-    V_j = V_i + \omega_i \times (P_j - P_i)
+    V_j = V_i + \omega_i \times P_{i-j}
 \end{eqnarray}
-
-where $j$ is the index of the corner node this edge node is attached to and
-$\omega_j$ is the angular velocity at the node. For the central node we
-the formula is again similar to \eqref{eq-central2}:
-
+%
+where $i$ is the index of the corner node the edge node is attached to and
+$\omega_i$ is the angular velocity at the node. For the central node
+the formula is based on \eqref{eq-central2}:
+%
 \begin{eqnarray}
-    V_i = \frac{1}{3} \sum_{i=1}^3 V_i + \omega_i \times (P_{10} - P_i)
+    V_i = \frac{1}{3} \sum_{i=1}^3 \left( V_i + \omega_i \times P_{i-10} \right)
 \end{eqnarray}
 
-\subsection{Mapping Forces}
+\subsection{Projection of Normal Forces}
 
 For physical interaction with the object one also requiers that the normal
-forces applied on the high resolution mesh are trasmitted back as forces
+forces applied on the object's surface are trasmitted back as forces
 and torques on the vertices of the mechanical mesh.
-
 We compute the influence of force $F$ on the corners through all the
 associated control points by the formula:
-
+%
 \begin{eqnarray}
-    F_1 & = & F \sum_{i \in \{1,4,5\}} n_i + \frac{1}{3} F n_{10} \\
-    F_2 & = & F \sum_{i \in \{2,6,7\}} n_i + \frac{1}{3} F n_{10} \\
-    F_3 & = & F \sum_{i \in \{3,8,9\}} n_i + \frac{1}{3} F n_{10}
+    F_1 & = & F \sum_{i \in \{1,4,5\}} n_i + \frac{n_{10}}{3} F \\
+    F_2 & = & F \sum_{i \in \{2,6,7\}} n_i + \frac{n_{10}}{3} F \\
+    F_3 & = & F \sum_{i \in \{3,8,9\}} n_i + \frac{n_{10}}{3} F 
 \end{eqnarray}
 
-%\begin{eqnarray}
-%    F_1 & = & F (n_1 + n_4 + n_5) + \frac{1}{3} F n_{10} \\
-%    F_2 & = & F (n_2 + n_6 + n_7) + \frac{1}{3} F n_{10} \\
-%    F_3 & = & F (n_3 + n_8 + n_9) + \frac{1}{3} F n_{10} \\
-%\end{eqnarray}
-
-Similarly we compute the torques applied through edge control points and
-the central control points by:
-
+Similarly we can compute the torques applied through edge control points and
+the central control points:
+%
 \begin{gather}
     \begin{split}
-        \theta_1 & = P_{1-4} \times (n_4 F) + P_{1-5} \times (n_5 F) + \\
-        & + \frac{1}{3} (P_{1-4} + P_{1-5}) \times (n_{10} F)
+        \theta_1 = & P_{1-4} \times (n_4 F) + P_{1-5} \times (n_5 F) + \\
+        & + \frac{1}{3} P_{1-10} \times (n_{10} F)
     \end{split}
@@
     \begin{split}
-        \theta_2 & = P_{1-6} \times (n_6 F) + P_{1-7} \times (n_7 F) + \\
-        & + \frac{1}{3} (P_{1-6} + P_{1-7}) \times (n_{10} F) \\
+        \theta_2 = & P_{2-6} \times (n_6 F) + P_{2-7} \times (n_7 F) + \\
+        & + \frac{1}{3} P_{2-10} \times (n_{10} F) \\
     \end{split}
@@
     \begin{split}
-        \theta_3 & = P_{1-8} \times (n_8 F) + P_{1-9} \times (n_9 F) + \\
-        & + \frac{1}{3} (P_{1-8} + P_{1-9}) \times (n_{10} F) \\
+        \theta_3 = & P_{3-8} \times (n_8 F) + P_{3-9} \times (n_9 F) + \\
+        & + \frac{1}{3} P_{3-10} \times (n_{10} F) \\
     \end{split}
 \end{gather}
-
-%\begin{eqnarray}
-%    \theta_1 & = & \sum_{i \in \{4,5\}} P_{1-i} \times (n_i F) \\
-%    & + &\frac{1}{3} (P_{1-4} + P_{1-5}) \times (n_{10} F) \\
-%    \theta_1 & = & \sum_{i \in \{6,7\}} P_{1-i} \times (n_i F) + \\
-%    & + &\frac{1}{3} (P_{1-6} + P_{1-7}) \times (n_{10} F) \\
-%    \theta_1 & = & \sum_{i \in \{8,9\}} P_{1-i} \times (n_i F) + \\
-%    & + &\frac{1}{3} (P_{1-8} + P_{1-9}) \times (n_{10} F) \\
-%\end{eqnarray}
-%
-%\begin{eqnarray}
-%    \theta_1 & = & \sum_{i \in \{4,5\}} P_{1-i} \times (n_i F) +
-%    \frac{1}{3} (P_{1-4} + P_{1-5}) \times (n_{10} F) \\
-%    \theta_1 & = & \sum_{i \in \{6,7\}} P_{1-i} \times (n_i F) +
-%    \frac{1}{3} (P_{1-6} + P_{1-7}) \times (n_{10} F) \\
-%    \theta_1 & = & \sum_{i \in \{8,9\}} P_{1-i} \times (n_i F) +
-%    \frac{1}{3} (P_{1-8} + P_{1-9}) \times (n_{10} F) \\
-%\end{eqnarray}
 
@@
 % inconsistent).
-  \CD{ - On this section,  we can maybe add some results from the CBM paper you did with Stephan especially  the figure 9 that illustrates the 3d contribution we list in the intro: i.e.
+
+% TODO:
+% \CD{ - On this section,  we can maybe add some results from the CBM paper you did with Stephan especially  the figure 9 that illustrates the 3d contribution we list in the intro: i.e.
 Mapping of high resolution mesh on the curved surface of the shell.}
-
-In next two subsections we present tests to validate the convergence of our
-B\'ezier shell element. We use a combination of two elements that are very well
-established in the area of mechanical modelling.
-Our results are compared with the shell element
-composed of Discrete Kirchoff Triangle (DKT) element \cite{Batoz1980} for bending plate and optimal ANDES element \cite{Felippa2003} for membrane.
-The last subsection presents evaluation of the computational complexity of our solution.
-
-\subsection{Roof Test}
+% Not enough space?
+
+In subsections \ref{sec-roof} and \ref{sec-hemisphere} we present tests to
+validate the convergence of our B\'ezier shell element. As a reference we use a
+combination of two elements that are very well established in the area of
+mechanical modelling. Our results are compared with the shell element composed
+of Discrete Kirchoff Triangle (DKT) element \cite{Batoz1980} for bending plate
+and optimal ANDES element \cite{Felippa2003} for membrane. For comparison the
+results for the model of O. Comas are also presented. Subsection \ref{sec-fps}
+presents evaluation of the computational complexity of our solution. Last
+subsections are dedicated to presentation of application of our model.
+
+\subsection{Roof Test}\label{sec-roof}
 
@@
 self-weight. The geometry is defined by 80° cylindrical patch of length
-$L=50$ and radius $r=25$ and thickness $t=0.25$ and is discretized into
-mesh of NxN nodes. The physical parameters are $E = 4.32 \times 10^8$ and
+$L=50$, radius $r=25$ and thickness $t=0.25$ and is discretized into
+mesh of NxN vertices. The physical parameters are $E = 4.32 \times 10^8$ and
 $\nu = 0$. The roof is loaded with uniformly distributed load $q = 90$ per
 unit area. The curved edges are clamped at both ends to avoid rigid body
-movement. This the only difference from the original Scordelis-Lo roof test
+movement. This is the only difference from the original Scordelis-Lo roof test
 where the edges are free to move in longitudinal direction. The Figure
@@
 
-\subsection{Hemisphere Test}
+\subsection{Hemisphere Test}\label{sec-hemisphere}
 
@@
 $r=10$ and thickness $t=0.04$ with 18° hole at the top and the geometry is
-discretized into the grid of NxN nodes per quadrant. The physical
+discretized into the grid of NxN vertices per quadrant. The physical
 parameters are $E = 6.825 \times 10^7$ and $\nu = 0.3$. The hemisphere is
@@
 
-\subsection{Computation Speed}
+\subsection{Computation Speed}\label{sec-fps} % {{{
 
 In figure \ref{fig-fps} we present performance of the element in terms of
-frames per second. All tests were performed on machine equipped Dual-Core
-AMD Opteron Processor 2218 and 3 GB of RAM and conjugate gradient solver
-has been to solve the system. Our implementation is without optimizations
+frames per second. All tests were performed on machine equipped with Dual-Core
+AMD Opteron Processor 2218 and 3~GB of RAM. We used a conjugate gradient solver
+to solve the system. Our implementation is without optimizations
 and isn't parallel, which means only single CPU core was used for the
-simulation. The values reported are for the raw physical simulation with
-and without the frame fixing method described in subsection
+simulation. The values reported are for the raw physical simulation with (b)
+and without (a) the frame fixing method described in subsection
 \ref{sec-corot}. We also present results including the mapping of
@@
 values for element with frame fixing (B) method described in subsection
-\ref{sec-corot} and without it (A), and with high-resolution mesh of 2048
-(C) or 3200 (D) triangles mapped onto the mechanical mesh.}
+\ref{sec-corot} and without it (A), and with mesh of 3\,200
+(C) or 10\,082 (D) triangles mapped onto the mechanical mesh.}
 
 We can see that the frame fixing technique is really fast and we are able
-to simulate up to 1000 elements and still maintain visual refresh rate of
-25 FPS. Our implementation of mechanical mapping clearly needs improving
-because it consumes most of the computation time, but it is still able to
-provide good results for more than 600 elements.
-
-\subsection{Application on Congenital Heart Disease Correction Planning}
+to simulate up to 1\,000 elements and still maintain visual refresh rate of
+25 FPS. While our implementation was only single-threaded and the mechanical mapping
+wasn't off-loaded to another CPU core we were still able to provide good
+results for more than 600 elements mesh with 10k triangles.
+
+% }}}
+
+\subsection{Application on Congenital Heart Disease Correction Planning} % {{{
 
@@
 %
-%% TODO: !!! When ANONYMISING do something with the citation !!!
-%One being a tool for pre-operative planning of corrective surgery of
-%congenital heart diseases in infants \cite{Kislinskiy2012}. The cardiac
-%surgery is a complicated process where several arterial vessels have to be
-%cut, patched and reconnected. The element has been used for to model
-%deformations of blood vessels after suturing and/or applying a artificial
-%patch. Figure \ref{fig-chd} presents a screenshots of the application
-%prototype.
 
@@
 are enriched with overlay delineations.}
+
+% }}}
 

### 2012-08-02 Doc/VRIPHYS2012/vriphys2012.tex
spell check
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@444 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 (i) it uses all 6 degrees of freedom,
-(ii) the deformations are completly symmetric and
+(ii) the deformations are completely symmetric and
 (iii) the boundary between elements is continuous.
@@
 %
-where $P_{i,j}$ are the control points and $B^n_{i,j}$ are the bivairate
+where $P_{i,j}$ are the control points and $B^n_{i,j}$ are the bivariate
 Bernstein basis functions defined as:
@@
 simulated object is usually described by a triangular mesh possibly with normals.
-A procedure for construction of initial Bézier mesh is necessary.
+A procedure for construction of initial B\'zier mesh is necessary.
 We construct the initial B\'ezier mesh by taking into account the curvature of the triangular
@@
 
-Similarily for the central point we apply rigid transformation of all frames of the triangle and compute the mean:
+Similarly for the central point we apply rigid transformation of all frames of the triangle and compute the mean:
 %
@@
 
-Puting previous equations into \eqref{eq-cubicbez2} we get:
+Putting previous equations into \eqref{eq-cubicbez2} we get:
 %
@@
 
-In the previous section we present an interpolation based on cubic B\'ezier triangles and a kinematic link between the control points and the element 6DoF Nodes.
+In the previous section we present an interpolation based on cubic B\'ezier triangles and a kinematic link between the control points and the element 6DOF nodes.
 In this section we show how this interpolation can be used to compute a finite element formulation of the shell equations.
@@
 
-The corotational frame is computed followingly:
+The corotational frame is computed using following steps:
 
 \begin{itemize}
-    \item One axis is alligned to one edge of the element:\\
+    \item One axis is aligned to one edge of the element:\\
     $\mathbf{e}_1 = \mathbf{x}_2 - \mathbf{x}_1$
@@
 of the triangle is not a perfect one and we have experienced
-nonnegligible difference in deformation of equilateral triangle depending
+non-negligible difference in deformation of equilateral triangle depending
 on whether an axis was aligned to the edge favored by the deformation or to
@@
 
-For physical interaction with the object one also requiers that the normal
-forces applied on the object's surface are trasmitted back as forces
+For physical interaction with the object one also requires that the normal
+forces applied on the object's surface are transmitted back as forces
 and torques on the vertices of the mechanical mesh.
@@
 {Convergence for modified Scordelis-Lo roof test of DKT+ANDES element, 
-our B\'ezier shell element (BSH) and elemnt of O. Comas (PSH) with polynomial shape function.}
+our B\'ezier shell element (BSH) and element of O. Comas (PSH) with polynomial shape function.}
 
@@
 {Convergence for test on hemisphere with 18° hole for the DKT+ANDES element and
-our B\'ezier shell element (BSH) and elemnt of O. Comas (PSH) with polynomial shape function.}
+our B\'ezier shell element (BSH) and element of O. Comas (PSH) with polynomial shape function.}
 
@@
 
-\subsection{Application on Congenital Heart Disease Correction Planning} % {{{
+\subsection{Planning of Congenital Heart Disease Correction Surgery} % {{{
 

### 2012-08-02 Doc/VRIPHYS2012/vriphys2012.tex
application on the birth simulation
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@446 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-% TODO: maybe make a short table showing the convergence to the symmetric
-% solution.
-
 % }}}
@@
 Mapping of high resolution mesh on the curved surface of the shell.}
-% Not enough space?
+% -- Not enough space?
 
@@
 
+\FigureW{0.75\linewidth}{chd}
+{Screenshots of the simulation system prototype for different surgical
+procedures used to repair a coarctation of an aortic arch. The screenshots
+are enriched with overlay delineations.}
+
 Surgical interventions in infants with congenitally malformed great arteries
@@
 for preoperative planning. The element has been used to model deformations of
-blood vessels and artificial patch. By performing a set of topological changes
+blood vessels and artificial patches. By performing a set of topological changes
 we simulate the process of incising the vessel and suturing on a pre-specified
@@
 %
-
-\FigureW{0.75\linewidth}{chd}
-{Screenshots of the simulation system prototype for different surgical
-procedures used to repair a coarctation of an aortic arch. The screenshots
-are enriched with overlay delineations.}
 
@@
 
+\subsection{Simulation of a Birth}
+
+\FigureW{\linewidth}{birth}
+{Screenshots from the simulation of child head during birth. Full visualisation (left) and the mechanical mesh (right).}
+
+The purpose of another simulator kkk 
+- Simulation of birth ! (Try to see where the tissue' stress is the highest)
+
+The trajectory of the baby-head during the birth is predefined. The collisions response drives the deformations of the uterus.
+Mechanically, the uterus is modeled with 127 shells solved by a direct solver (LDL factorization).
+Up to 60 instantaneous contact spots, simulated at 6 FPS.
+The goal is to reproduce the mechanical solicitations on the anatomical structure during the birth.
+
 % }}}
@@
 
-The second limitation is the choice of rigidly attached edge control
+The second limitation is the choice of rigidly attached control
 points. While it greatly simplifies the formulation the rigidity may cause
@@
 
-% }}}
-
-\subsection{Conclusion} % {{{
+Unified formulation of the deformation allows for derivation of three-dimensional 
+stress tensor. This however remains as topic for future work and
+we have kept the computation of membrane stresses separate from bending stresses.
+
+% }}}
+
+\section{Conclusion} % {{{
 
@@
 
-    % TODO: check this:
-    % + Better conditioning (comparison between Comas model and ours in term of
-    %   convergence of CG...)
-
 \end{itemize}
 
-Unified formulation of the deformation allows for derivation of three-dimensional 
-stress tensor. This however remains as topic for future work and
-we have kept the computation of membrane stresses separate from bending stresses.
-
 Because of the formulation of B\'ezier triangle that is intrinsically present
-there is a clean and obvious way of mapping high-resolution mesh for
-better visualisation. We also provide a mapping of forces and velocities
-between simulation mesh and mechanical mesh for collision detection/response
-and interaction with the mesh which are essential component for the surgical
-simulator.
-
-Finally examples showing the convergence to the optimal solution were also provided.
+there is a clean and obvious way of mapping high-resolution mesh for better
+visualisation. We also provide a technique of computing velocities of any point
+on the surface and formula for back-projection of normal forces applied on the
+surface onto the mechanical mesh. Those are form an essential component for the
+surgical simulator as they are necessary if contact modeling between simulation
+mesh and mechanical mesh is desired or other interaction with the mesh is
+required.
+
+At the end we present examples showing the convergence to the optimal solution
+and the application of the model for several simulation tools.
 

### 2012-08-02 Doc/VRIPHYS2012/vriphys2012.tex
abstract
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@447 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{abstract} % TODO
-    TODO: Abstract.
+There is a strong need, in surgical simulations, for physics-based deformable model of thin or hollow structures.
+The use of shell theory allows to have a well-founded formulation resulting from continuum mechanics of thin objects.
+However, this formulation asks for second order spatial derivatives so requires the use of complex elements.
+In this paper, we present a new way of building the interpolation:
+First, we use the cubic B/'ezier shell triangle to allow for a good continuity inside and between the elements and second, we build a kinematic mapping to reduce the degrees of freedom of the element from 10 control points with 3 Degrees of Freedom ($=30$ DoFs) to only 3 nodes with 6 DoFs ($=18$ DoFs).
+This reduction allows for good computation performance.
+This new shell model description is also used to map a smooth surface (for the collision detection and response) on a coarse mechanical mesh to accounts for the complex contacts that take place during surgical procedures.
+We demonstrate the convergence and the computational efficiency of our approach as well as its use in two different simulation cases: the planning of surgery for congenital heart disease correction and a preliminary simulation of childbirth.
 

### 2012-08-02 Doc/VRIPHYS2012/vriphys2012.tex
changes in sections 6 and 7
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@448 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 edges. This may prove to be an issue for example if curvature needs to be
-computed not only inside the element but also on the edges.
-
-The second limitation is the choice of rigidly attached control
-points. While it greatly simplifies the formulation the rigidity may cause
-issues. In case of large compression of the element the results may be
+computed not only inside the element but also across the edges.
+
+The second limitation is the choice of rigidly attached control points. 
+While it greatly simplifies the formulation the rigidity may cause issues.
+In case of large compression of the element the results may be
 unpredictable. This is however unlikely to occur because the element will
@@
 stretching.
+We could investigate more complex kinematic links between nodes and control points 
+and also more continuous surface descriptions.
 
@@
 
-Unified formulation of the deformation allows for derivation of three-dimensional 
-stress tensor. This however remains as topic for future work and
-we have kept the computation of membrane stresses separate from bending stresses.
-
+Our study relies on the assumption that the in-plane and the bending deformation energies can be separated.
+For complex deformation fields or material constitutive laws  this assumption is no more valid.
+However, we could keep the same way of building the element even with more complex mechanics. 
+For instance the interpolation would allow for derivation of three-dimensional non-linear strain tensors.
+We will investigate this topic in our future work.
 % }}}
@@
 In this paper we presents new shell element with interpolation functions
-based on B\'ezier triangle which is more suitable for real-time simulations than elements common in mechanics with complex, obscure or unclear interpolation functions.
+based on B\'ezier triangle which is  suitable for real-time simulations.
+Compared to  the elements commonly used in mechanics, the interpolation is less complex and ensure at least a $C^0$ continuity in the worst case. 
+(Many shell elements are not even $C^0$ when the initial shape is not flat).
+
 It fixes problems of the previous model \cite{Comas2010c} based on cubic polynomials, namely:
@@
 
-    \item It is $C^0$ continuous on the edges and provides $C^1$ continuity in
+    \item It is $C^1$ for the case of initial flat shape and, on the other cases, it is $C^0$ continuous on the edges and provides $C^1$ continuity in
     corners of the element. This is essential for good visual experience if
@@
 
-Because of the formulation of B\'ezier triangle that is intrinsically present
+Because of the formulation of B\'ezier triangle is intrinsically present
 there is a clean and obvious way of mapping high-resolution mesh for better
-visualisation. We also provide a technique of computing velocities of any point
+visualization. We also provide a technique of computing velocities of any point
 on the surface and formula for back-projection of normal forces applied on the
-surface onto the mechanical mesh. Those are form an essential component for the
-surgical simulator as they are necessary if contact modeling between simulation
-mesh and mechanical mesh is desired or other interaction with the mesh is
-required.
-
-At the end we present examples showing the convergence to the optimal solution
-and the application of the model for several simulation tools.
+surface onto the mechanical mesh. 
+Those are essential components for the surgical simulators that need accurate contact geometry modeling without an overflow of computation need
+The efficiency of the approach is demonstrated through convergence tests,  computation time benchmark and application to two different  simulation cases.
 

### 2012-08-02 Doc/VRIPHYS2012/vriphys2012.tex
proofread of section 4
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@449 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \usepackage{cite}
+\usepackage{ulem}
 
@@
 (i) interpolating velocity on the element surface and for
-(ii) distribution of normal force applied on the element surface back to 6 DOF nodes of the element.
+(ii) distribution of force applied on the element surface back to 6 DOF nodes of the element.
 This allows us to use smooth mapped surface for collision detection and allows the propagation of proper collision response back onto the mechanical mesh.
@@
 the simulated object it is desirable to use more triangles in the areas
-with high curvature during rendering. Besides the visual accuracy the high
-resolution mesh correctly modeling the curved areas is needed for
-interaction and correct simulation of collisions and constraints. To handle
-these we also need to know how to map the velocities from the mechanical
+with high curvature during rendering. 
+Besides the visual accuracy the high resolution mesh correctly modeling the curved areas is needed for
+interaction and correct simulation of collisions and constraints. 
+To handle these, we also need to know how to map the velocities from the mechanical
 mesh onto the high resolution mesh and how to map forces acting on the high
@@
 
-For contact modeling it is often necessary to know the velocity of each point.
+For contact modeling it is necessary to know the velocity of the contact points.
 The values are directly available to us for the nodes of the mechanical mesh
@@
 Which means we use the same expression to interpolate the velocity only by
-substituting the velocities at control points. For the corner nodes we
-already have the velocities $V_i$. For the internal nodes we need to compute
-them. We again make use of the kinematic link between nodes and solve the
-problem of computing the velocity of a point attached to the moving rigid body.
-For the edge points we have:
+substituting the velocities at control points. 
+For the corner \CD{\sout{nodes} $\rightarrow$ control points}  we
+already have the velocities $V_i$. For the internal \CD{\sout{nodes} $\rightarrow$ control points}  we need to compute
+them. We again make use of the kinematic link between nodes and control points.
+%solve the problem of computing the velocity of a point attached to the moving rigid body.
+For the edge points we applied to motion of a rigid body:
 %
@@
 $\omega_i$ is the angular velocity at the node. For the central node
-the formula is based on \eqref{eq-central2}:
+the formula is based on the derivate of equation \eqref{eq-central2}:
 %
 \begin{eqnarray}
-    V_i = \frac{1}{3} \sum_{i=1}^3 \left( V_i + \omega_i \times P_{i-10} \right)
+    V_{10} = \frac{1}{3} \sum_{i=1}^3 \left( V_i + \omega_i \times P_{i-10} \right)
 \end{eqnarray}
 
-\subsection{Projection of Normal Forces}
-
-For physical interaction with the object one also requires that the normal
+\subsection{Projection of surface forces}
+
+For physical interaction with the object one also requires that the
 forces applied on the object's surface are transmitted back as forces
@@
 \begin{eqnarray}
-    F_1 & = & F \sum_{i \in \{1,4,5\}} n_i + \frac{n_{10}}{3} F \\
-    F_2 & = & F \sum_{i \in \{2,6,7\}} n_i + \frac{n_{10}}{3} F \\
-    F_3 & = & F \sum_{i \in \{3,8,9\}} n_i + \frac{n_{10}}{3} F 
+    F_1 & = &  \sum_{i \in \{1,4,5\}} n_i F + \frac{n_{10}}{3} F \\
+    F_2 & = &  \sum_{i \in \{2,6,7\}} n_i F + \frac{n_{10}}{3} F \\
+    F_3 & = &  \sum_{i \in \{3,8,9\}} n_i F + \frac{n_{10}}{3} F 
 \end{eqnarray}
@@
     \begin{split}
-        \theta_1 = & P_{1-4} \times (n_4 F) + P_{1-5} \times (n_5 F) + \\
+        	\tau_1 = & P_{1-4} \times (n_4 F) + P_{1-5} \times (n_5 F) + \\
         & + \frac{1}{3} P_{1-10} \times (n_{10} F)
@@
     \begin{split}
-        \theta_2 = & P_{2-6} \times (n_6 F) + P_{2-7} \times (n_7 F) + \\
+       	\tau_2 = & P_{2-6} \times (n_6 F) + P_{2-7} \times (n_7 F) + \\
         & + \frac{1}{3} P_{2-10} \times (n_{10} F) \\
@@
     \begin{split}
-        \theta_3 = & P_{3-8} \times (n_8 F) + P_{3-9} \times (n_9 F) + \\
+        	\tau_3 = & P_{3-8} \times (n_8 F) + P_{3-9} \times (n_9 F) + \\
         & + \frac{1}{3} P_{3-10} \times (n_{10} F) \\
@@
 \end{gather}
+
+Note that these formula fulfill the physical principle of virtual power: $ P = \dot{T} \cdot F  = \sum_{i=1}^3 V_i \cdot F_i +  \omega_i  \cdot \tau_i$
 
@@
 visualization. We also provide a technique of computing velocities of any point
-on the surface and formula for back-projection of normal forces applied on the
+on the surface and formula for back-projection of forces applied on the
 surface onto the mechanical mesh. 

### 2012-08-02 Doc/VRIPHYS2012/vriphys2012.tex
proofread of section 5
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@450 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 and optimal ANDES element \cite{Felippa2003} for membrane. For comparison the
-results for the model of O. Comas are also presented. Subsection \ref{sec-fps}
+results for the model of O. Comas \cite{Comas2010c} are also presented. 
+Subsection \ref{sec-fps}
 presents evaluation of the computational complexity of our solution. Last
@@
 
-Second performed test is a hemisphere with hole subjected to two opposing
+Second performed test is an hemisphere with hole subjected to two opposing
 forces at the base. The geometry is described by a hemisphere with radius
@@
 
-\subsection{Simulation of a Birth}
+\subsection{Simulation of uterus deformation during birth}
 
 \FigureW{\linewidth}{birth}
-{Screenshots from the simulation of child head during birth. Full visualisation (left) and the mechanical mesh (right).}
-
-The purpose of another simulator kkk 
-- Simulation of birth ! (Try to see where the tissue' stress is the highest)
-
-The trajectory of the baby-head during the birth is predefined. The collisions response drives the deformations of the uterus.
-Mechanically, the uterus is modeled with 127 shells solved by a direct solver (LDL factorization).
-Up to 60 instantaneous contact spots, simulated at 6 FPS.
-The goal is to reproduce the mechanical solicitations on the anatomical structure during the birth.
-
-% }}}
+{Screenshots from the simulation of child head during birth. Full visualization (left) and the mechanical mesh (right).}
+
+The purpose of this simulator is to reproduce the tissue deformations that are encountered during a delivery.
+The goal is to evaluate the mechanical stresses which are applied on the anatomical structures (like the uterus but also like the ligaments that sustain the uterus).
+
+The data comes from a MRI of a 8 months pregnant woman. 
+The trajectory of the baby-head during the birth was predefined with gynecologists. 
+The model is limited to uterus deformations but will soon account for the vaginal tissues deformation.
+The collisions response drives the deformations of the uterus.
+Mechanically, the uterus is modeled with 127 shells that are solved by a direct solver (LDL factorization).
+Up to 60 instantaneous contact spots need to be mapped on the shell elements. 
+The simulation frame-rate is relatively low (6 FPS) due to contact response computation.
 
@@
 
+In future work, we will investigate the use of more complex constitutive laws and also the use of GPU for parallel and faster compuation.
+
 % }}}

### 2012-08-02 Doc/VRIPHYS2012/vriphys2012.tex
small corrections
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@451 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 % TODO: anonymize -- Submission ID
+%\author[paper1026]
+%       {paper1026}
 \author[T. Golembiovský \& C. Duriez]
@@
 
-\begin{abstract} % TODO
-There is a strong need, in surgical simulations, for physics-based deformable model of thin or hollow structures.
+\begin{abstract}
+There is a strong need, in surgical simulations, for physically based deformable model of thin or hollow structures.
 The use of shell theory allows to have a well-founded formulation resulting from continuum mechanics of thin objects.
@@
 In this paper, we present a new way of building the interpolation:
-First, we use the cubic B/'ezier shell triangle to allow for a good continuity inside and between the elements and second, we build a kinematic mapping to reduce the degrees of freedom of the element from 10 control points with 3 Degrees of Freedom ($=30$ DoFs) to only 3 nodes with 6 DoFs ($=18$ DoFs).
+First, we use the trianular cubic B\'ezier shell to allow for a good continuity inside and between the elements and second, we build a kinematic mapping to reduce the degrees of freedom of the element from 10 control points with 3 Degrees of Freedom ($=30$ DOFs) to only 3 nodes with 6 DOFs ($=18$ DOFs).
 This reduction allows for good computation performance.
-This new shell model description is also used to map a smooth surface (for the collision detection and response) on a coarse mechanical mesh to accounts for the complex contacts that take place during surgical procedures.
+This new shell model description is also used to map a smooth surface (for the collision detection and response) on a coarse mechanical mesh to account for the complex contacts that take place during surgical procedures.
 We demonstrate the convergence and the computational efficiency of our approach as well as its use in two different simulation cases: the planning of surgery for congenital heart disease correction and a preliminary simulation of childbirth.
@@
 \begin{classification} % according to http://www.acm.org/class/1998/
-    \CCScat{Computer Graphics}{I.3.5}{Computational Geometry and Object Modeling}{Physically based modelling};
-    \CCScat{Computer Graphics}{I.3.7}{Three-Dimensional Graphics and Realism}{Animation}.
+    \CCScat{Computer Graphics}{I.3.5}{Computational Geometry and Object Modeling}{Physically based modelling}.
+    %\CCScat{Computer Graphics}{I.3.7}{Three-Dimensional Graphics and Realism}{Animation}.
     %\CCScat{Simulation and Modeling}{I.6.8}{Types of Simulation}{Animation}
@@
 This is completely acceptable for some computer graphics application, where the parameters can be tuned manually to obtain a realistic animation.
-In the case of medical simulation, more physics-based modeling is needed.
+In the case of medical simulation, more physically based modeling is needed.
 
@@
 
-\subsection{Projection of surface forces}
+\subsection{Projection of Surface Forces}
 
@@
 
-Second performed test is an hemisphere with hole subjected to two opposing
+Second performed test is a hemisphere with hole subjected to two opposing
 forces at the base. The geometry is described by a hemisphere with radius
@@
 
-% TODO?
-%Another designed application is simulation of women reproductive organs
-%(vagaina/uterus/\ldots). 
-%The simulation is so as to evaluate
-%stress on a tissue during a birth.
-
-\subsection{Simulation of uterus deformation during birth}
+\subsection{Simulation of Uterus Deformation During Birth}
 
@@
 The simulation frame-rate is relatively low (6 FPS) due to contact response computation.
+
+% }}}
 
@@
 We will investigate this topic in our future work.
+
 % }}}
@@
 based on B\'ezier triangle which is  suitable for real-time simulations.
-Compared to  the elements commonly used in mechanics, the interpolation is less complex and ensure at least a $C^0$ continuity in the worst case. 
+Compared to  the elements commonly used in mechanics, the interpolation is less complex and ensures at least a $C^0$ continuity in the worst case. 
 (Many shell elements are not even $C^0$ when the initial shape is not flat).
@@
 surface onto the mechanical mesh. 
-Those are essential components for the surgical simulators that need accurate contact geometry modeling without an overflow of computation need
+Those are essential components for the surgical simulators that need accurate contact geometry modeling without an overflow of computation.
 The efficiency of the approach is demonstrated through convergence tests,  computation time benchmark and application to two different  simulation cases.

### 2012-08-03 Doc/VRIPHYS2012/vriphys2012.tex
changes to fit into 10 page limit
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@452 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \end{figure*}}
-
-%%% ADD COMMENTS=
-\usepackage{color} 
-\newcommand{\CD}[1]{{\color{magenta}{\textbf{C: #1}}}}
 
@@
 
-\FigureWI{stages}
+\FigureW{0.8\linewidth}{stages}
 {Two stages of interpolation: First, from triangular with 6DOF nodes (a) a mesh of control points (b) is computed.
@@
 
-\Figure{0.8\linewidth}{bezier2}
+\Figure{0.6\linewidth}{bezier2}
 {Cubic B\'ezier triangle with mesh of 10 control points.}
@@
 $C^0$ continuity. Each of the control points on the edge is
-computed as the intersection of (\CD{Need to do a figure of this}):
+computed as the intersection of:
 
@@
 
-\Figure{0.8\linewidth}{segments2}
+\Figure{0.6\linewidth}{segments2}
 {Internal nodes of control mesh are connected to the corner nodes by kinematic link (shown in blue).}
@@
 
-(\CD{TODO add a figure to illustrate the explanation})
-
 We define a local system of coordinates $(x,y,z)$ in which the axis z is oriented along the thickness of the element (the computation of the frame of each element will be detailed in section \ref{sec-corot}).
@@
 quadrature for integration over triangle area.
-\CD{TODO: figure + gives the formula of the quadrature...}
+TODO: figure + gives the formula of the quadrature...
 % TODO: add ref
@@
 substituting the velocities at control points. 
-For the corner \CD{\sout{nodes} $\rightarrow$ control points}  we
-already have the velocities $V_i$. For the internal \CD{\sout{nodes} $\rightarrow$ control points}  we need to compute
-them. We again make use of the kinematic link between nodes and control points.
+For the corner control points  we
+already have the velocities $V_i$. For the internal control points  we need to compute
+them. We again make use of the kinematic link between control points and control points.
 %solve the problem of computing the velocity of a point attached to the moving rigid body.
@@
 
-% TODO:
-% \CD{ - On this section,  we can maybe add some results from the CBM paper you did with Stephan especially  the figure 9 that illustrates the 3d contribution we list in the intro: i.e.
-Mapping of high resolution mesh on the curved surface of the shell.}
-% -- Not enough space?
-
 In subsections \ref{sec-roof} and \ref{sec-hemisphere} we present tests to
@@
 
-\FigureW{0.75\linewidth}{chd}
+\FigureW{\linewidth}{chd}
 {Screenshots of the simulation system prototype for different surgical
@@
 promising approach is often chosen during the actual open-heart surgery, when
-surgeons get a more concrete idea of possible outcomes. Finalizing this
-decision during an on-pump cardiac surgery is further complicated by deflated
-blood vessels since it is difficult to imagine their exact change of shape
-caused by the intervention. The important decisions have to be made in very
+surgeons get a more concrete idea of possible outcomes.
+The important decisions have to be made in very
 short time and are strongly dependent on the surgeon's experience.
@@
 
-%%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
-%
-%% TODO: is this OK for anonymity?
-%The presented element has been successfully implemented in SOFA framework
-%\cite{Faure2012} for real-time surgical simulations. It has been evaluated
-%on several applications.
-%
-
 % }}}
@@
 
-It fixes problems of the previous model \cite{Comas2010c} based on cubic polynomials, namely:
-
-\begin{itemize}
-
-    \item Element is based on all 6 DOFs, including the drilling degree of
-        freedom, not just 5 DOFs. Therefore it doesn't have the problem of
-        undefined z-rotations.
-
-    \item Deformations of the element are symmetric. A method of improving
-    the corotational frame to keep the deformation symmetric is also
+It fixes problems of the previous model \cite{Comas2010c} based on cubic
+polynomials, namely: element is based on all 6 DOFs, deformations of the
+element are symmetric, it is $C^1$ for the case of initial flat shape and, on
+the other cases, it is $C^0$ continuous on the edges and provides $C^1$
+continuity in corners of the element. This is essential for good visual
+experience if high-resolution mesh is mapped on the mechanical mesh. A method
+of improving the corotational frame to keep the deformation symmetric is also
 presented in the paper.
-
-    \item It is $C^1$ for the case of initial flat shape and, on the other cases, it is $C^0$ continuous on the edges and provides $C^1$ continuity in
-    corners of the element. This is essential for good visual experience if
-    high-resolution mesh is mapped on the mechanical mesh.
-
-\end{itemize}
 

### 2012-08-03 Doc/VRIPHYS2012/vriphys2012.tex
adding paper outline to introduction
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@453 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 % TODO: anonymize -- Submission ID
-%\author[paper1026]
-%       {paper1026}
-\author[T. Golembiovský \& C. Duriez]
-       {Tomáš Golembiovský$^{1,2}$ and Christian Duriez$^{2}$\\
-        $^1$Faculty of Informatics, Masaryk University, Czech Republic\\
-        $^2$INRIA Lille, France
-       } % TODO: ask Christian how to specify affiliation
+       \author[paper1026]
+              {paper1026}
+%\author[T. Golembiovský \& C. Duriez]
+%       {Tomáš Golembiovský$^{1,2}$ and Christian Duriez$^{2}$\\
+%        $^1$Faculty of Informatics, Masaryk University, Czech Republic\\
+%        $^2$INRIA Lille, France
+%       } % TODO: ask Christian how to specify affiliation
 
@@
 
-% TODO: paper structure
-TODO: paper structure
+We start by reviewing the definition of B\'ezir triangle. Then the construction
+of initial control mesh is described and  kinematic link between control points
+is defined. In Section 3 the FE model is derived and the method of computing
+the corotational frame is described. In Section 4 we specify how to map a high
+resolution surface mesh onto the mechanical mesh and give formulas for
+computing velocities of points on the surface and projecting of forces. Section
+5 presents test of convergence, speed analysis and two created applications
+that use the described shell model. We conclude with discussion of present
+limitations.
 
@@
 quadrature for integration over triangle area.
-TODO: figure + gives the formula of the quadrature...
+%TODO: figure + gives the formula of the quadrature...
 % TODO: add ref
@@
 
-Which means we use the same expression to interpolate the velocity only by
-substituting the velocities at control points. 
 For the corner control points  we
@@
 to solve the system. Our implementation is without optimizations
-and isn't parallel, which means only single CPU core was used for the
+and only single CPU core was used for the
 simulation. The values reported are for the raw physical simulation with (b)
@@
 \subsection{Planning of Congenital Heart Disease Correction Surgery} % {{{
-
-\FigureW{\linewidth}{chd}
-{Screenshots of the simulation system prototype for different surgical
-procedures used to repair a coarctation of an aortic arch. The screenshots
-are enriched with overlay delineations.}
 
@@
 
-The planning process can be improved by employing surgical simulation system
-using three-dimensional heart models based on patient-specific images as input.
-Most present simulation systems focus on execution of low-level surgical
-procedures, e.g. incising and suturing. In these reactive simulation systems it
-is more important to enable simulations in real-time than accurate simulation
-of tissue and blood vessel deformation in response to simulated interventions.
-
-To be able to predict results of complex surgical procedures a physically more
-accurate approach is necessary. Using our shell finite element model we have
-developed a prototype \cite{Kislinskiy2012} of new surgical simulation system
+To be able to predict results of complex surgical procedures a surgical simulation system can be employed.
+We have
+developed a prototype \cite{Anonymous2012} of new surgical simulation system
 for preoperative planning. The element has been used to model deformations of
@@
 boundary. By iteratively adjusting the underlaying mesh we search for the
-elastic equilibrium. Figure \ref{fig-chd} presents a screenshots of the
-application prototype.
+elastic equilibrium.
 
@@
 \subsection{Simulation of Uterus Deformation During Birth}
-
-\FigureW{\linewidth}{birth}
-{Screenshots from the simulation of child head during birth. Full visualization (left) and the mechanical mesh (right).}
 
@@
 
+\FigureW{\linewidth}{chd}
+{Screenshots of the simulation system prototype for different surgical
+procedures used to repair a coarctation of an aortic arch. The screenshots
+are enriched with overlay delineations.}
+
+\FigureW{\linewidth}{birth}
+{Screenshots from the simulation of child head during birth. Full visualization (left) and the mechanical mesh (right).}
+
 % }}}
@@
 Currently, due to the technical decisions the element has two limitations.
-The first limitation is that the element does not provide $C^1$ continuity
-across edges. It is $C^1$ continuous at corners but only $C^0$ on the
+It is $C^1$ continuous at corners but only $C^0$ on the
 edges. This may prove to be an issue for example if curvature needs to be
@@
 
-Because of the formulation of B\'ezier triangle is intrinsically present
-there is a clean and obvious way of mapping high-resolution mesh for better
+The formulation of B\'ezier triangle provides
+a clean way of mapping high-resolution mesh for better
 visualization. We also provide a technique of computing velocities of any point

### 2012-09-20 papier_afig/frame_refig.tex
Dossier papier afig
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@459 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

% Mettre un "%" devant le style à ne pas utiliser
\documentclass{refig} % style de la revue REFIG
%\documentclass{afig} % style des journées AFIG

\usepackage[latin1]{inputenc} 
\usepackage[francais]{babel} 
\usepackage{t1enc,dfadobe}
\usepackage{cite}
\usepackage[toc, page]{appendix}
%\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}

% for including postscript figures
% mind: package option 'draft' will replace PS figure by a filename within a frame
\ifpdf \usepackage[pdftex]{graphicx} \pdfcompresslevel=9
\else \usepackage[dvips]{graphicx} \fi

\title[Simulation à base de Repères]{Simulation en Lagragien Total Implicite à base de Repères}

\author[Julien Bosman]{Julien Bosman$^1$, Christian Duriez$^2$ \\ %
	$^1$LIFL/INRIA ,$^2$INRIA
}

\begin{document}

\maketitle

\begin{abstract}
   A faire

\end{abstract}
\keywords{Simulation, Méthodes sans maillage, Mécanique}
%-------------------------------------------------------------------------
\section{Introduction}
 A faire

\section{Etat de l'art}

La simulation à base de méthodes sans maillage à été introduite en informatique graphique par Cani et Desbrun dans \cite{Desbrun1996} et l'utilisation des
\emph{Smoothed Particle Hydrodynamics} (SPH) -- reposant sur l'interaction réciproque d'un ensemble de particules -- pour l'animation de modèles très
déformables. Les SPH demeurent une méthodes populaires pour l'animation de fluides et d'objets très déformables. Les travaux de \cite{Becker2009} apportent une
réponse au problème des non-linéarités géométrique en proposant un modèle corotationnel dédié. La méthode des SPH permet également de simuler des matériaux
non-newtoniens aux comportements plus complexes notamment dans \cite{Paiva2009} et \cite {hosseini06}. Cependant, Les SPH souffrent de certains problèmes qui
limitent leur utilisation notamment leur failbe précision. Le traitement des frontières est également difficile en raison de la déficience en
particule sur le frontières de l'objet. De plus, le phénomène connu sous le nom de \emph{tensile instability}, qui apparait lorsque l'objet est fortement
étiré, rend son utilisation problématique dans le cadre de la simulation de grandes déformation. Parallèlement, d'autres méthodes d'animation sans maillage ont
été développées. Des modèles basés sur la mécanique des milieux continus à l'instar de \cite{Keiser2004} et de \cite{Horton2010} où les champs de déplacement
sont approximés grâce aux \emph{Moving} Least Squares (MLS). Cependant, l'approximation MLS est impossible si la distribution des n{\oe}uds est colinéaire ou
coplanaire. En effet, la  ``moment matrix'' devient alors singulière et les fonctions de forme ne peuvent être calculées.
Des techniques d'animation, non basées sur la physique,ont également été développées. Dans \cite{Muller2005}, les objets sont déformés grâce à un système masse
ressort en utilisant la méthode dite de \emph{Shape Matching}, où les points de l'objet sont partitionnés en plusieurs groupes dont on estime une rotation
optimale et une translation par rapport à la configuration non déformée. Cette idée est reprise dans \cite{Muller2011} à la différence près que chaque possède
une orientation qui est estimée au grâce à un \emph{Shape Matching} généralisé.
\section{Simulation}

\subsection{Cinématique}
Les objets déformables sont discrétisés par un ensemble de n{\oe}uds et de points d'intégration. Les n{\oe}uds sont des repères à six degrés de liberté (DDL) :
\begin{equation}
\mathbf{q}_i = \left[ x_1 \ x_2 \ x_3 \ r_1 \ r_2 \ r_3 \right]^t = \left[ \mathbf{x}_i \ {\mathbf{r}_i} \right]^t
\end{equation}

où $\mathbf{x}_i$ sont les coordonnées du n{\oe}ud et $\mathbf{r}_i$ le vecteurs rotation correspondant à l'orientation du n{\oe}ud. Les points d'intégration ne
possèdent que trois DDL. Dans la mesure où l'on utilise une formulation en lagragien total, la position globale des points d'intégration est exprimée en
fonction de leur position dans chacun des repères dans la configuration non déformée. Les coordonnées d'un point d'intégration $s$ dans le repère $i$ dans la
configuration non déformée est donné par :
\begin{equation}
	\bar{\mathbf{l}}_s^{i} = {[\bar{\mathbf{R}}}_i]^t \cdot \left[ \bar{\mathbf{p}}_s - \bar{\mathbf{x}}_i\right]
\end{equation}
où $\bar{\mathbf{R}}_i$ et $\bar{\mathbf{x}}_i$ sont la matrice de rotation et les coordonnées du repère $i$ et $\bar{\mathbf{p}}_s$ les coordonnées du point
d'intégration $s$ dans la configuration non déformée. Ainsi, les coordonnées du point d'intégration $s$ dans la configuration déformée sont donnés par :
\begin{equation}
	\mathbf{p}_s = \sum_{i} w_s(\bar{\mathbf{x}}_i)\left(\mathbf{x}_i + [\mathbf{R}_i] \cdot \bar{\mathbf{l}}_s^{i} \right)
	\label{eq:sample_pos}
\end{equation}
où $w_s(\cdot)$ est une fonction de forme associée au point d'intégration $s$ et $\mathbf{x}_i$ et $\mathbf{R}_i$ les coordonnées et la rotation du repère $i$
dans la configuration déformée. La fonction de forme $w_s(\cdot)$ peut être à support compact (du type \emph{fonction de Wendland}) ou non (fonction de Shepard
par ex).
De la même façon, la cinématique des repères est influencée par le mouvement des points d'intégration qui lui sont associés. La variation de l'orientation du
repère $i$ dans en fonction des déplacements des points d'intégration est donné par  :
\begin{equation}
	%\mathbf{r}_i = \sum_{s}{w_s(\mathbf{x}_i)\left( \mathbf{\bar{p}}_s^{l,i} \wedge \mathbf{u}_s^{l,i} \right)}
	\frac{d\mathbf{r}_i}{d\mathbf{p}} = \sum_{s}{w_s(\mathbf{x}_i)(\mathbf{l}_s^i \wedge d\mathbf{p}_s)}
\end{equation}
où $\mathbf{l}_s^i$ sont les coordonnées du point d'intégration $s$ dans le repère $i$

\subsection{Dynamique}
\subsubsection{Expression des forces internes}
L'approche employée dans ces travaux se base sur la mécanique des milieux continus (MMC). Les matériaux étudiés sont considérés comme continus, c'est-à-dire
qu'ils sont étudiés à une échelle suffisament petite, mais aussi suffisament grande pour négliger les espaces inter-moléculaires. La MMC permet de mettre en
relation un déplacement (\textit{i.e.} les déformations subies par un point) avec une force de rappel selon le schéma suivant :
\begin{equation*}
	\underbrace{\mathbf{u}_i}_{\text{déplacements}} \rightarrow \underbrace{\mathbf{F}_i}_{\substack{\text{gradient de} \\ \text{déformation}}} \rightarrow
\underbrace{\epsilon_i}_{\substack{\text{tenseur de} \\ \text{déformation}}} \rightarrow \underbrace{\sigma_i}_{\substack{\text{tenseur de} \\
\text{contrainte}}} \rightarrow \underbrace{\mathbf{f}_i}_{\text{force}}
\end{equation*}
La méthode décrite dans cet article utilise une formulation en lagrangien total. Dans une formulation en lagrangien total les déplacements sont évalués par
rapport à la configuration non déformée de l'objet contrairement à la formulation en lagragien updatée où ils sont évalués par rapport à la
configuration de l'objet au pas de temps précédent.
Le calcul des forces internes fait appel au principe des \emph{travaux virtuels} :
\begin{equation}
	\delta W = \delta \mathbf{q} \cdot \mathbf{f} = \int_{\Omega}{ \frac{\partial \mathbf{L}^t}{\partial \mathbf{q}} \{\mathbf{S}\}d\Omega }
	\label{eq:virtual_work}
\end{equation}
où $L = \frac{1}{2}(F^t F - I)$ est le tenseur de déformation de \emph{Green-Lagrange} où $F$ est le gradient de déformation. $S$ le second tenseur de
contrainte de \emph{Piola-Kirchoff} dans la notation de \emph{Voigt}. La dérivée du tenseur de Green-Lagrange par rapport au degrés de liberté
des repères est donné par :
\begin{equation}
	\frac{\partial L}{\partial q} = \frac{1}{2}\left(\frac{\partial F^t}{\partial q} F + F^t \frac{\partial F}{\partial q} \right)
	\label{eq:green_lagrange_der}
\end{equation}

La première étape consiste à calculer le tenseur de déformation en un point d'intégration $s$ c'est-à-dire $\mathbf{F}_s = \nabla \mathbf{u}_s$. Il faut donc
calculer la dérivée spatiale de l'équation (\ref{eq:sample_pos}). On remarque alors que seul les termes $w_s(\mathbf{x}_i)$ et $\bar{\mathbf{l}}_s^i$ sont
concernés.
\begin{equation}
	F_s = \frac{\partial \mathbf{p}_s}{\partial \mathbf{X}} = \sum_{i}{\left[ \frac{\partial w_{is}}{\partial X} \left(\mathbf{x}_i + [\mathbf{R}_i] \cdot
\bar{\mathbf{l}}_s^{i} \right) + w_{is}[\mathbf{R}_i] [\bar{\mathbf{R}}_i]^t \right] } 
%& = \sum_{i}{\left[ \nabla w_s(\mathbf{x}_i) \left(\mathbf{x}_i + \mathbf{R}_i \times \bar{\mathbf{p}}_s^{l,i} \right) + w_s(\mathbf{x}_i)\mathbf{R}_i
%\mathbf{R}_i^{0t} \right] }
\end{equation}
Afin d'obtenir la dérivée du tenseur de \emph{Green-Lagrange} de l'équation (\ref{eq:green_lagrange_der}) il est nécessaire de calculer la dérivée du tenseur de
déformation :
\begin{equation}
\begin{split}
	\delta F_{IJ}^{s} =& \sum_{i}{\left(\delta \mathbf{x}_i + [-\mathbf{R}_{i}(\mathbf{\bar{p}}_s^{l,i})^{\text{\tiny{$\wedge$}}}]^{I,}\cdot \delta
\mathbf{r}_i\right) \frac{\partial w_s(\mathbf{x}_i)}{\mathbf{X}_j}} \ + \\
& [-\mathbf{R}_{i}(r^0_{,J})^{\text{\tiny{$\wedge$}}}]^{I,} \cdot \delta \mathbf{r_i} w_s(\mathbf{x}_i)
\end{split}
\label{eq:def_grad_der}
\end{equation}
avec $\mathbf{R}_{i}(\mathbf{\bar{p}}_s^{l,i})$ la rotation de $\mathbf{\bar{p}}_s^{l,i}$ par $\mathbf{R}_i$, $[\mathbf{v}^{\text{\tiny{$\wedge$}}}]$la
matrice antisymétrique $3\times 3$ associée au produit vectoriel, $[\mathbf{M}]^{i,}$ et $[\mathbf{M}]^{,j}$ correspondent respectivement à la $i^{\text{ème}}$
ligne et à la
$j^{\text{ème}}$ colonne de la matrice $\mathbf{M}$ et $r^0_{,J} = [\mathbf{R}_i^0]^{J,}$.
En réécrivant les 9 valeurs de $\delta F^{s}$ sous forme d'un vecteur colonne, l'équation (\ref{eq:def_grad_der}) peut être exprimée sous la forme d'un produit
matrice-vecteur :
\begin{equation}
	\left[
	\begin{array}{c}
		\delta F_{xx}^s \\
		\delta F_{xy}^s \\
		\vdots \\
		\delta F_{zz}^s
	\end{array}
	\right] = \sum_{i}{\mathbf{B}^{i,s} \cdot \delta \mathbf{q}_i}
	\label{eq:delta_F}
\end{equation}
avec $\delta \mathbf{q}_i^t = [\delta \mathbf{x}_i \ \delta \mathbf{r}_i]^t$. La matrice $B$ est donnée en annexe. De la même façon, il est également possible
d'écrire $\partial L / \partial q$ sous la forme d'un produit matrice-vecteur. En effet :
\begin{equation}
\begin{split}
	\delta L_{xy} &= \frac{1}{2}\left([\delta F^t]^{1,}\cdot[F]^{,2} + [F^t]^{1,}\cdot[\delta F]^{,2}\right) \\
	&= \frac{1}{2}\left([\delta F_{xx} \ \delta F_{yx} \ \delta F_{zx}]\cdot[F]^{,2} + [F^t]^{1,}\cdot[\delta F_{xy} \ \delta F_{yy} \ \delta F_{zy}]\right)
\end{split}
\end{equation}
Or, d'après l'équation (\ref{eq:delta_F}) :
\begin{equation}
	[\delta F_{xx} \ \delta F_{yx} \ \delta F_{zx}] = [B_{1,}\cdot \delta q \ B_{3,}\cdot \delta q \  B_{7,}\cdot \delta q]
\end{equation}
avec $B_{i,}$ la $i^{\text{ème}}$ ligne de la matrice $B$ d'où :
\begin{equation}
	\delta L_{xy} = \frac{1}{2}\left[F_{xy}B_{1,} +  F_{yy}B_{4,} + F_{zy}B_{7,} + F_{xx}B_{2,} + F_{yx}B_{5,} + F_{zx}B_{8,} \right] \cdot \delta q
\end{equation}
On a donc : 
\begin{equation}
	\left[
	\begin{array}{c}
		\delta L_{xx}^s \\
		\delta L_{xy}^s \\
		\vdots \\
		\delta L_{zz}^s
	\end{array}
	\right] = \sum_{i}{\mathbf{M}^{i;s} \cdot \delta \mathbf{q}_i}
	\label{eq:delta_L}	
\end{equation}
En substituant l'équation (\ref{eq:delta_L}) à l'équation (\ref{eq:virtual_work}) et en divisant par $\delta q$ on obtient :
\begin{equation}
	f = \int_{\Omega}{\mathbf{M}^t\{\mathbf{S}\}}d\Omega
\end{equation}
ou, en version discrète : 
\begin{equation}
	f_i = \sum_{s}{\mathbf{M}^{i,s}\{\mathbf{S}^s\}w_s(\mathbf{x}_i)}
	\label{eq:force_discrete}
\end{equation}
avec $S$ calculé grâce à la \emph{loi de Hooke}:
\begin{equation}
	\mathbf{S} = \lambda \ tr(\mathbf{L})\mathbf{I} + 2 \mu \mathbf{L}
\end{equation}
en notation de Voigt où $\lambda$ et $\mu$ sont les \emph{coefficients de Lamé}

\subsubsection{Formulation implicite}
La formulation précédente (équation \ref{eq:force_discrete}) est utilisable avec un solveur explicite. Cependant, bien qu'ils soient rapides, les solveurs
explicites présentent certains désavantages, notamment l'instabilité. Les oscillations et l'introduction d'énergie supplémentaire conduisent la simulation à
``exploser''. Un solution courament employée pour répondre à ce problème consiste à utiliser des pas de temps très court avec pour effet de ralentir la
simulation.
Pour ces raisons, on utilise un solveur implicte qui est inconditionnellement stable -- une proprièté intéressante dans le cadre de la simulation
d'objets déformables. Les solveurs implicites permettent de résoudre un problème non-linéaire de manière itérative selon la méthode de \emph{Newton-Raphson}
\cite{faure:hal-00681539}. Pour permettre au solveur de résoudre l'équation du mouvement, il est nécessaire de linéariser la force, c'est-à-dire dériver
l'expression de la force par rapport aux repères. D'après \cite{NFECS} on a :
\begin{equation}
	\delta f = \int_{\Omega}{\mathbf{M^t}\{\delta \mathbf{S}\}d\Omega}
	\label{eq:force_der}
\end{equation}
avec :
\begin{equation}
	\{\delta \mathbf{S}\} = \mathbf{C}\{\delta \mathbf{L}\}
\end{equation}
ou la matrice $\mathbf{C}$ donnée en annexe est la matrice des coefficients elastiques associés à la loi de Hooke. En substituant l'équation (\ref{eq:delta_L})
à l'équation (\ref{eq:force_der}) on obtient :
\begin{equation}
	\delta f = \int_{\Omega}{\mathbf{M^t}\mathbf{C}\mathbf{M}} \cdot \delta \mathbf{q} d\Omega
\end{equation}
on obtient alors, en version discrète :
\begin{equation}
	\delta f_i = \sum_{s}{\mathbf{M}^{i,s}\cdot \Delta \mathbf{S}^s}
\end{equation}
avec : 
\begin{equation}
	\Delta \mathbf{S}^s = \sum_{i}{\mathbf{C}\mathbf{M}^{i,s}\cdot \delta \mathbf{q}_i}
\end{equation}

\subsection{Figures en couleur} Etant donné qu'il s'agit d'une publication électronique, il n'y a aucune censure à avoir sur les figures en couleur (cf figure \ref{fig:ex3}.

\begin{figure*}[htb]
  \centering
  \mbox{} \hfill
  \includegraphics[width=.3\linewidth]{logoafig}
  \hfill
  \includegraphics[width=.3\linewidth]{logoafig}
  \hfill \mbox{}
  \caption{\label{fig:ex3}
           Les figures en couleurs ou en noir et blanc peuvent être placées n'importe ou dans le document. }
\end{figure*}

\section{Conclusion}
A faire

%-------------------------------------------------------------------------
\newpage

\bibliographystyle{refig-alpha}
\bibliography{biblio}

%-------------------------------------------------------------------------

\newpage
\renewcommand\arraystretch{2}
\begin{appendices}
\section{Calcul de la matrice $B$}
\end{appendices}
\begin{equation*}
\begin{split}
	& \mathbf{B}^{i,s} = \\
	&
	\left[\begin{array}{cccc}
		\frac{\partial w_{is}}{\partial x} & 0 & 0 & \left[\frac{\partial w_{is}}{\partial x}
		[-\mathbf{R}_{i}(\mathbf{\bar{p}}_s^{l,i})^{\text{\tiny{$\wedge$}}}]^{x,} - [\mathbf{R}_{i}(r^0_{,x})^{\text{\tiny{$\wedge$}}}]^{x,}\ w_{is}
		\right]\\
		\frac{\partial w_{is}}{\partial y} & 0 & 0 & \left[\frac{\partial w_{is}}{\partial y}
		[-\mathbf{R}_{i}(\mathbf{\bar{p}}_s^{l,i})^{\text{\tiny{$\wedge$}}}]^{x,} - [\mathbf{R}_{i}(r^0_{,y})^{\text{\tiny{$\wedge$}}}]^{x,}\ w_{is}
		\right] \\
		\frac{\partial w_{is}}{\partial z} & 0 & 0 & \left[\frac{\partial w_{is}}{\partial z}
		[-\mathbf{R}_{i}(\mathbf{\bar{p}}_s^{l,i})^{\text{\tiny{$\wedge$}}}]^{x,} - [\mathbf{R}_{i}(r^0_{,z})^{\text{\tiny{$\wedge$}}}]^{x,}\ w_{is}
		\right]\\		
		0 & \frac{\partial w_{is}}{\partial x} & 0 & \left[\frac{\partial w_{is}}{\partial x}
		[-\mathbf{R}_{i}(\mathbf{\bar{p}}_s^{l,i})^{\text{\tiny{$\wedge$}}}]^{y,} - [\mathbf{R}_{i}(r^0_{,x})^{\text{\tiny{$\wedge$}}}]^{y,}\ w_{is}
		\right]\\
		0 & \frac{\partial w_{is}}{\partial y} & 0 & \left[\frac{\partial w_{is}}{\partial y}
		[-\mathbf{R}_{i}(\mathbf{\bar{p}}_s^{l,i})^{\text{\tiny{$\wedge$}}}]^{y,} - [\mathbf{R}_{i}(r^0_{,y})^{\text{\tiny{$\wedge$}}}]^{y,}\ w_{is}
		\right] \\
		0 & \frac{\partial w_{is}}{\partial z} & 0 & \left[\frac{\partial w_{is}}{\partial z}
		[-\mathbf{R}_{i}(\mathbf{\bar{p}}_s^{l,i})^{\text{\tiny{$\wedge$}}}]^{y,} - [\mathbf{R}_{i}(r^0_{,z})^{\text{\tiny{$\wedge$}}}]^{y,}\ w_{is}
		\right]\\		
		0 & 0 & \frac{\partial w_{is}}{\partial x} & \left[\frac{\partial w_{is}}{\partial x}
		[-\mathbf{R}_{i}(\mathbf{\bar{p}}_s^{l,i})^{\text{\tiny{$\wedge$}}}]^{z,} - [\mathbf{R}_{i}(r^0_{,x})^{\text{\tiny{$\wedge$}}}]^{z,}\ w_{is}
		\right]\\
		0 & 0 & \frac{\partial w_{is}}{\partial y} & \left[\frac{\partial w_{is}}{\partial y}
		[-\mathbf{R}_{i}(\mathbf{\bar{p}}_s^{l,i})^{\text{\tiny{$\wedge$}}}]^{z,} - [\mathbf{R}_{i}(r^0_{,y})^{\text{\tiny{$\wedge$}}}]^{z,}\ w_{is}
		\right] \\
		0 & 0 & \frac{\partial w_{is}}{\partial z} & \left[\frac{\partial w_{is}}{\partial z}
		[-\mathbf{R}_{i}(\mathbf{\bar{p}}_s^{l,i})^{\text{\tiny{$\wedge$}}}]^{z,} - [\mathbf{R}_{i}(r^0_{,z})^{\text{\tiny{$\wedge$}}}]^{z,}\ w_{is}
		\right]\\		
	\end{array} \right]
\end{split}
\end{equation*}
\renewcommand\arraystretch{1}
avec $w_{is} = w_s(\mathbf{x}_i)$
\section{Matrice de Hooke}
\begin{equation*}
\begin{split}
	&C = \frac{E}{(1+\nu)(2-\nu)} \ \times \\
	&\left[\begin{array}{ccc}
		\begin{array}{ccc}
			1-\nu & \nu & \nu \\
			\nu & 1-\nu & \nu \\
			\nu & \nu & 1-\nu \\
		\end{array} & \text{\huge{$0$}} \\
		\text{\huge{$0$}} & 
		\begin{array}{ccc}
			\frac{(1-2\nu)}{2} & & \\
			& \frac{(1-2\nu)}{2} & \\
			& & \frac{(1-2\nu)}{2} \\
		\end{array}
	\end{array} \right]
\end{split}
\end{equation*}
où $E$ et $\nu$ sont respectivement le \emph{module de Young} et le \emph{coefficient de Poisson}.
\end{document}

### 2012-09-20 papier_afig/frame_refig.tex
Corrections + figure
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@460 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\title[Simulation à base de Repères]{Simulation en Lagragien Total Implicite à base de Repères}
+\title[Modèle pour la Simulation de Tissus Connectifs]{Modèle pour la Simulation de Tissus Connectifs}
 
-\author[Julien Bosman]{Julien Bosman$^1$, Christian Duriez$^2$ \\ %
+\author[Julien Bosman]{Julien Bosman$^1$, Christian Duriez$^2$, Stéphane Cotin$^2$ \\ %
 	$^1$LIFL/INRIA ,$^2$INRIA
@@
 non-newtoniens aux comportements plus complexes notamment dans \cite{Paiva2009} et \cite {hosseini06}. Cependant, Les SPH souffrent de certains problèmes qui
-limitent leur utilisation notamment leur failbe précision. Le traitement des frontières est également difficile en raison de la déficience en
+limitent leur utilisation notamment leur faible précision. Le traitement des frontières est également difficile en raison de la déficience en
 particule sur le frontières de l'objet. De plus, le phénomène connu sous le nom de \emph{tensile instability}, qui apparait lorsque l'objet est fortement
@@
 optimale et une translation par rapport à la configuration non déformée. Cette idée est reprise dans \cite{Muller2011} à la différence près que chaque possède
-une orientation qui est estimée au grâce à un \emph{Shape Matching} généralisé.
+une orientation qui est estimée au grâce à un \emph{Shape Matching} généralisé. Enfin, dans \cite{GilBouFauPai11}, Gilles \textit{et al.} propsent un
+modèle compatible avec la mécanique des milieux continus basé sur des repères. Cependant, leur approche concerne essentiellement des objets faiblement
+échantillonés. De fait, elle est difficile à utiliser pour modéliser des objets qui doivent se conformer étroitement à des surfaces.
 \section{Simulation}
@@
 \subsection{Cinématique}
-Les objets déformables sont discrétisés par un ensemble de n{\oe}uds et de points d'intégration. Les n{\oe}uds sont des repères à six degrés de liberté (DDL) :
+Les objets déformables sont discrétisés par un ensemble de n{\oe}uds et de points d'intégration (figure \ref{fig:discrete}). Les n{\oe}uds sont des repères à
+six degrés de liberté (DDL) :
 \begin{equation}
@@
 \end{equation}
-
 où $\mathbf{x}_i$ sont les coordonnées du n{\oe}ud et $\mathbf{r}_i$ le vecteurs rotation correspondant à l'orientation du n{\oe}ud. Les points d'intégration ne
@@
 	\bar{\mathbf{l}}_s^{i} = {[\bar{\mathbf{R}}}_i]^t \cdot \left[ \bar{\mathbf{p}}_s - \bar{\mathbf{x}}_i\right]
+	\label{eq:sample_rest_pos}
 \end{equation}
@@
 où $w_s(\cdot)$ est une fonction de forme associée au point d'intégration $s$ et $\mathbf{x}_i$ et $\mathbf{R}_i$ les coordonnées et la rotation du repère $i$
-dans la configuration déformée. La fonction de forme $w_s(\cdot)$ peut être à support compact (du type \emph{fonction de Wendland}) ou non (fonction de Shepard
-par ex).
+dans la configuration déformée. La fonction de forme $w_s(\cdot)$ peut être à support compact (du type \emph{fonction de Wendland}) ou non (\emph{fonction de
+Shepard} par ex).
+
+\begin{figure}[htb]
+	\center \includegraphics[width=6.5cm]{discrete}
+	\caption{\label{fig:discrete}Discrétisation des objets}
+\end{figure}
+
 De la même façon, la cinématique des repères est influencée par le mouvement des points d'intégration qui lui sont associés. La variation de l'orientation du
@@
 L'approche employée dans ces travaux se base sur la mécanique des milieux continus (MMC). Les matériaux étudiés sont considérés comme continus, c'est-à-dire
-qu'ils sont étudiés à une échelle suffisament petite, mais aussi suffisament grande pour négliger les espaces inter-moléculaires. La MMC permet de mettre en
+qu'ils sont étudiés à une échelle microscopique, mais une échelle suffisament grande pour négliger les espaces inter-moléculaires. La MMC permet de
+mettre en
 relation un déplacement (\textit{i.e.} les déformations subies par un point) avec une force de rappel selon le schéma suivant :
@@
 \end{equation}
-où $L = \frac{1}{2}(F^t F - I)$ est le tenseur de déformation de \emph{Green-Lagrange} où $F$ est le gradient de déformation. $S$ le second tenseur de
-contrainte de \emph{Piola-Kirchoff} dans la notation de \emph{Voigt}. La dérivée du tenseur de Green-Lagrange par rapport au degrés de liberté
-des repères est donné par :
+avec $\mathbf{L} = \frac{1}{2}(\mathbf{F}^t \mathbf{F} - \mathbf{I})$ est le tenseur de déformation de \emph{Green-Lagrange} où $F$ est le gradient de
+déformation. $\mathbf{S}$ est le second tenseur de contrainte de \emph{Piola-Kirchoff} dans la notation de \emph{Voigt}. La dérivée du tenseur de Green-Lagrange
+par rapport au degrés de liberté des repères est donné par :
 \begin{equation}
-	\frac{\partial L}{\partial q} = \frac{1}{2}\left(\frac{\partial F^t}{\partial q} F + F^t \frac{\partial F}{\partial q} \right)
+	\frac{\partial \mathbf{L}}{\partial \mathbf{q}} = \frac{1}{2}\left(\frac{\partial \mathbf{F}^t}{\partial \mathbf{q}} \mathbf{F} + \mathbf{F}^t
+\frac{\partial \mathbf{F}}{\partial \mathbf{q}} \right)
 	\label{eq:green_lagrange_der}
@@
 calculer la dérivée spatiale de l'équation (\ref{eq:sample_pos}). On remarque alors que seul les termes $w_s(\mathbf{x}_i)$ et $\bar{\mathbf{l}}_s^i$ sont
-concernés.
+concernés. Or, en reprenant l'équation (\ref{eq:sample_rest_pos}) on remarque que :
+\begin{equation}
+	\frac{\partial \bar{\mathbf{l}}_s^i}{\partial X} = \frac{\partial([\bar{\mathbf{R}}_i]^t \cdot \left[ \bar{\mathbf{p}}_s -
+\bar{\mathbf{x}}_i\right])}{\partial X} = [\bar{\mathbf{R}}_i]^t
+\end{equation}
+on obtient alors le gradient de déformation au point d'intégration $s$ :
 \begin{equation}
@@
 \end{equation}
-Afin d'obtenir la dérivée du tenseur de \emph{Green-Lagrange} de l'équation (\ref{eq:green_lagrange_der}) il est nécessaire de calculer la dérivée du tenseur de
-déformation :
+où $w_{is} = w_s(\mathbf{x}_i)$. Afin d'obtenir la dérivée du tenseur de \emph{Green-Lagrange} de l'équation (\ref{eq:green_lagrange_der}) il est nécessaire de
+calculer la dérivée du tenseur de déformation par rapport à $\delta \mathbf{q}$. On l'exprime alors en fonction d'une variation infinitésimale de déplacement
+et d'orientation, ou encore en fonction de $\delta \mathbf{x}$ et $\delta \mathbf{r}$ :
 \begin{equation}
 \begin{split}
-	\delta F_{IJ}^{s} =& \sum_{i}{\left(\delta \mathbf{x}_i + [-\mathbf{R}_{i}(\mathbf{\bar{p}}_s^{l,i})^{\text{\tiny{$\wedge$}}}]^{I,}\cdot \delta
-\mathbf{r}_i\right) \frac{\partial w_s(\mathbf{x}_i)}{\mathbf{X}_j}} \ + \\
-& [-\mathbf{R}_{i}(r^0_{,J})^{\text{\tiny{$\wedge$}}}]^{I,} \cdot \delta \mathbf{r_i} w_s(\mathbf{x}_i)
+	\delta F_{IJ}^{s} =& \sum_{i}{\left(\delta \mathbf{x}_i + [-\mathbf{R}_{i}(\mathbf{\bar{l}}_s^{i})^{\text{\tiny{$\wedge$}}}]^{I,}\cdot \delta
+\mathbf{r}_i\right) \frac{\partial w_{is}}{\mathbf{X}_j}} \ + \\
+& [-\mathbf{R}_{i}(r^0_{,J})^{\text{\tiny{$\wedge$}}}]^{I,} \cdot \delta \mathbf{r_i} w_{is}
 \end{split}
@@
 \end{equation}
-avec $\mathbf{R}_{i}(\mathbf{\bar{p}}_s^{l,i})$ la rotation de $\mathbf{\bar{p}}_s^{l,i}$ par $\mathbf{R}_i$, $[\mathbf{v}^{\text{\tiny{$\wedge$}}}]$la
+avec $\mathbf{R}_{i}(\mathbf{\bar{l}}_s^{i})$ la rotation de $\mathbf{\bar{l}}_s^{i}$ par $\mathbf{R}_i$, $[\mathbf{v}^{\text{\tiny{$\wedge$}}}]$la
 matrice antisymétrique $3\times 3$ associée au produit vectoriel, $[\mathbf{M}]^{i,}$ et $[\mathbf{M}]^{,j}$ correspondent respectivement à la $i^{\text{ème}}$
@@
 	\begin{array}{c}
-		\delta F_{xx}^s \\
-		\delta F_{xy}^s \\
+		\delta F_{11}^s \\
+		\delta F_{12}^s \\
 		\vdots \\
-		\delta F_{zz}^s
+		\delta F_{33}^s
 	\end{array}
@@
 \end{equation}
-avec $\delta \mathbf{q}_i^t = [\delta \mathbf{x}_i \ \delta \mathbf{r}_i]^t$. La matrice $B$ est donnée en annexe. De la même façon, il est également possible
+avec $\delta \mathbf{q}_i^t = [\delta \mathbf{x}_i \ \delta \mathbf{r}_i]^t$. La matrice $\mathbf{B}$ est donnée en annexe. De la même façon, il est également
+possible
 d'écrire $\partial L / \partial q$ sous la forme d'un produit matrice-vecteur. En effet :
@@
 \begin{split}
-	\delta L_{xy} &= \frac{1}{2}\left([\delta F^t]^{1,}\cdot[F]^{,2} + [F^t]^{1,}\cdot[\delta F]^{,2}\right) \\
-	&= \frac{1}{2}\left([\delta F_{xx} \ \delta F_{yx} \ \delta F_{zx}]\cdot[F]^{,2} + [F^t]^{1,}\cdot[\delta F_{xy} \ \delta F_{yy} \ \delta F_{zy}]\right)
+	\delta L_{12} &= \frac{1}{2}\left([\delta F^t]^{1,}\cdot[F]^{,2} + [F^t]^{1,}\cdot[\delta F]^{,2}\right) \\
+	&= \frac{1}{2}\left([\delta F_{11} \ \delta F_{21} \ \delta F_{31}]\cdot[F]^{,2} + [F^t]^{1,}\cdot[\delta F_{12} \ \delta F_{22} \ \delta F_{32}]\right)
 \end{split}
@@
 	\begin{array}{c}
-		\delta L_{xx}^s \\
-		\delta L_{xy}^s \\
+		\delta L_{11}^s \\
+		\delta L_{12}^s \\
 		\vdots \\
-		\delta L_{zz}^s
+		\delta L_{33}^s
 	\end{array}
@@
 \end{equation}
-En substituant l'équation (\ref{eq:delta_L}) à l'équation (\ref{eq:virtual_work}) et en divisant par $\delta q$ on obtient :
+En substituant l'équation (\ref{eq:delta_L}) à l'équation (\ref{eq:virtual_work}) et en divisant par $\delta \mathbf{q}$ on obtient :
 \begin{equation}
-	f = \int_{\Omega}{\mathbf{M}^t\{\mathbf{S}\}}d\Omega
+	\mathbf{f} = \int_{\Omega}{\mathbf{M}^t\{\mathbf{S}\}}d\Omega
+	\label{eq:force_continuous}
 \end{equation}
-ou, en version discrète : 
+Cepedant, l'équation (\ref{eq:force_continuous}) nous donne l'expression de la force dans le cas continu. Etant donné que cette intégrale ne peut être résolue
+analytiquement, hormis pour des cas simples, il est nécessaire de passer à une version discrète. L'intégrale de l'équation (\ref{eq:force_continuous}) doit être
+évaluée sur la totalité du domaine $\Omega$ occupé par l'objet. Pour ce faire, sous la forme discrète, on utilise les points d'intégration, aussi appellés
+\emph{points de Gauss} qui échantillonnent l'objet. On obtient ainsi l'expression discrète de la force en un n{\oe}ud :
 \begin{equation}
@@
 \end{equation}
-avec $S$ calculé grâce à la \emph{loi de Hooke}:
+où $\mathbf{S}$ en notation de Voigt est calculé grâce à la \emph{loi de Hooke} qui est une loi de comportement décrivant des matériaux parfaitement élastiques
+(\textit{i.e.} qui reviennent parfaitement à leur forme d'origine). Cette loi est linéaire, c'est-à-dire que la force de rappel est proportionnelle à
+l'élongation. Ainsi :
 \begin{equation}
@@
 \end{equation}
-en notation de Voigt où $\lambda$ et $\mu$ sont les \emph{coefficients de Lamé}
+avec $\lambda$ et $\mu$ qui sont les \emph{coefficients de Lamé}
 
 \subsubsection{Formulation implicite}
-La formulation précédente (équation \ref{eq:force_discrete}) est utilisable avec un solveur explicite. Cependant, bien qu'ils soient rapides, les solveurs
-explicites présentent certains désavantages, notamment l'instabilité. Les oscillations et l'introduction d'énergie supplémentaire conduisent la simulation à
+La formulation précédente (équation \ref{eq:force_discrete}) est utilisable avec un solveur explicite. En effet, en utilisant le principe fondamental de la
+dynamique (seconde loi de Newton) :
+\begin{equation}
+	m\mathbf{a} = \sum{\mathbf{f}^{int} + \mathbf{f}^{ext}}
+\end{equation}
+puis en effectuant successivement deux intégrations sur le temps, on peut obtenir les nouvelles positions et les nouvelles vitesses de chaque n{\oe}ud au pas
+de temps suivant.
+
+Cependant, bien qu'ils soient rapides, les solveurs explicites présentent certains désavantages, notamment l'instabilité. Les oscillations et l'introduction
+d'énergie supplémentaire conduisent la simulation à
 ``exploser''. Un solution courament employée pour répondre à ce problème consiste à utiliser des pas de temps très court avec pour effet de ralentir la
@@
 \cite{faure:hal-00681539}. Pour permettre au solveur de résoudre l'équation du mouvement, il est nécessaire de linéariser la force, c'est-à-dire dériver
-l'expression de la force par rapport aux repères. D'après \cite{NFECS} on a :
+l'expression de la force par rapport aux repères. Cela revient à exprimer la variation de la force en fonction d'une variation inifitésimale de la
+contrainte. En effet, d'après \cite{NFECS} on a :
 \begin{equation}
@@
 \begin{equation}
-	\delta f = \int_{\Omega}{\mathbf{M^t}\mathbf{C}\mathbf{M}} \cdot \delta \mathbf{q} d\Omega
+	\delta \mathbf{f} = \int_{\Omega}{\mathbf{M^t}\mathbf{C}\mathbf{M}} \cdot \delta \mathbf{q} d\Omega
+	\label{eq:deltaF_continuous}
 \end{equation}
-on obtient alors, en version discrète :
+De même qu'avec l'équation (\ref{eq:force_continuous}), il est nécessaire de passer à une formulation discrète afin d'être capable d'intégrer l'équation
+(\ref{eq:deltaF_continuous})
 \begin{equation}
-	\delta f_i = \sum_{s}{\mathbf{M}^{i,s}\cdot \Delta \mathbf{S}^s}
+	\delta \mathbf{f}_i = \sum_{s}{\mathbf{M}^{i,s}\cdot \Delta \mathbf{S}^s}
 \end{equation}
@@
 \end{equation}
+qui exprime la variation de la contrainte en un point d'intégration en fonction d'une variation des DDL différents n{\oe}uds.
 

### 2012-09-20 papier_afig/frame_refig.tex
Ajout intro et corrections
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@461 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \section{Introduction}
- A faire
+La simulation médicale est un domaine de plus en plus actif. Les simulateurs médicaux ont un champ d'application assez vaste allant du diagnostic à la
+formation des practiciens ou encore en aide à la chirurgie per et post opératoire. Jusqu'à maintenant, les organes simulés étaient directement en contact les
+uns avec les autres. Cependant, cette situation n'est pas conforme à la réalité. En effet, à l'intérieur du corps humains, les organes sont très rarement en
+contact direct. Ils sont généralement reliés entre eux par des fibres appelées \emph{tissus connectifs}. Ces tissus peuvent être présent en une couche très fine
+ou en un volume plus important. Ils sont difficiles à modéliser et à simuler car étant donné qu'ils occupent l'espace présent entre les organes, ils n'ont pas
+de topologie réellement définie.
 
@@
 limitent leur utilisation notamment leur faible précision. Le traitement des frontières est également difficile en raison de la déficience en
-particule sur le frontières de l'objet. De plus, le phénomène connu sous le nom de \emph{tensile instability}, qui apparait lorsque l'objet est fortement
+particule sur le frontières de l'objet. De plus, le phénomène connu sous le nom de \emph{tensile instability}, qui apparaît lorsque l'objet est fortement
 étiré, rend son utilisation problématique dans le cadre de la simulation de grandes déformation. Parallèlement, d'autres méthodes d'animation sans maillage ont
@@
 optimale et une translation par rapport à la configuration non déformée. Cette idée est reprise dans \cite{Muller2011} à la différence près que chaque possède
-une orientation qui est estimée au grâce à un \emph{Shape Matching} généralisé. Enfin, dans \cite{GilBouFauPai11}, Gilles \textit{et al.} propsent un
+une orientation qui est estimée au grâce à un \emph{Shape Matching} généralisé. Enfin, dans \cite{GilBouFauPai11}, Gilles \textit{et al.} proposent un
 modèle compatible avec la mécanique des milieux continus basé sur des repères. Cependant, leur approche concerne essentiellement des objets faiblement
-échantillonés. De fait, elle est difficile à utiliser pour modéliser des objets qui doivent se conformer étroitement à des surfaces.
+échantillonnés. De fait, elle est difficile à utiliser pour modéliser des objets qui doivent se conformer étroitement à des surfaces.
 \section{Simulation}
+
+\subsection{Nomenclature}
+\begin{tabular}{|c|p{5.5cm}|}
+	\hline
+	\raisebox{-1ex}{$\bar{\mathbf{x}}_i$, $\mathbf{x}_i$} & coordonnées du n{\oe}uds $i$ dans la configuration au repos et déformée\\
+	\hline
+	\raisebox{-1ex}{$\bar{\mathbf{r}}_i$, $\mathbf{r}_i$} & orientation du n{\oe}ud $i$ dans la configuration au repos et déformée\\
+	\hline
+	\raisebox{-1ex}{$\bar{\mathbf{p}}_s$, $\mathbf{p}_s$} & coordonnées globales du point d'intégration $s$ dans la configuration au repos et déformée\\
+	\hline
+	\raisebox{-2ex}{$\bar{\mathbf{l}}_s^i$, $\mathbf{l}_s^i$} & coordonnées du point d'intégration $s$ dans le repère $i$ dans la configuration au repos et
+déformée. \\
+	\hline
+	\raisebox{-2ex}{$[\bar{\mathbf{R}}_i]$, $[\mathbf{R}_i]$} & matrice de rotation associée à l'orientation du n{\oe}ud $i$ dans la configuration au repos et
+déformée \\
+	\hline
+	\raisebox{-1ex}{$[\mathbf{v}^{\text{\tiny{$\wedge$}}}]$} & matrice antisymétrique associée au produit vectoriel de $\mathbf{v}$\\
+	\hline
+	\raisebox{-0.1ex}{$[\mathbf{M}]^{i,}$} & $i^{\text{ème}}$ ligne de la matrice $\mathbf{M}$ \\
+	\hline
+	\raisebox{-0.1ex}{$[\mathbf{M}]^{,j}$} & $j^{\text{ème}}$ ligne de la matrice $\mathbf{M}$ \\
+	\hline
+\end{tabular}
 
@@
 	\underbrace{\mathbf{u}_i}_{\text{déplacements}} \rightarrow \underbrace{\mathbf{F}_i}_{\substack{\text{gradient de} \\ \text{déformation}}} \rightarrow
-\underbrace{\epsilon_i}_{\substack{\text{tenseur de} \\ \text{déformation}}} \rightarrow \underbrace{\sigma_i}_{\substack{\text{tenseur de} \\
+\underbrace{\L_i}_{\substack{\text{tenseur de} \\ \text{déformation}}} \rightarrow \underbrace{\S_i}_{\substack{\text{tenseur de} \\
 \text{contrainte}}} \rightarrow \underbrace{\mathbf{f}_i}_{\text{force}}
@@
 \begin{equation}
-	F_s = \frac{\partial \mathbf{p}_s}{\partial \mathbf{X}} = \sum_{i}{\left[ \frac{\partial w_{is}}{\partial X} \left(\mathbf{x}_i + [\mathbf{R}_i] \cdot
+	\mathbf{F}_s = \frac{\partial \mathbf{p}_s}{\partial \mathbf{X}} = \sum_{i}{\left[ \frac{\partial w_{is}}{\partial X} \left(\mathbf{x}_i + [\mathbf{R}_i]
+\cdot
 \bar{\mathbf{l}}_s^{i} \right) + w_{is}[\mathbf{R}_i] [\bar{\mathbf{R}}_i]^t \right] } 
@@
 où $w_{is} = w_s(\mathbf{x}_i)$. Afin d'obtenir la dérivée du tenseur de \emph{Green-Lagrange} de l'équation (\ref{eq:green_lagrange_der}) il est nécessaire de
-calculer la dérivée du tenseur de déformation par rapport à $\delta \mathbf{q}$. On l'exprime alors en fonction d'une variation infinitésimale de déplacement
-et d'orientation, ou encore en fonction de $\delta \mathbf{x}$ et $\delta \mathbf{r}$ :
+calculer la dérivée du tenseur de déformation par rapport à $\delta \mathbf{q}$ qui est une matrice $3\times 3$. On exprime alors ces neuf valeurs en fonction
+d'une variation infinitésimale de déplacement et d'orientation, c'est-à-dire en fonction de $\delta \mathbf{x}$ et $\delta \mathbf{r}$ :
 \begin{equation}
 \begin{split}
-	\delta F_{IJ}^{s} =& \sum_{i}{\left(\delta \mathbf{x}_i + [-\mathbf{R}_{i}(\mathbf{\bar{l}}_s^{i})^{\text{\tiny{$\wedge$}}}]^{I,}\cdot \delta
+	\delta \mathbf{F}_{IJ}^{s} =& \sum_{i}{\left(\delta \mathbf{x}_i + [-\mathbf{R}_{i}(\mathbf{\bar{l}}_s^{i})^{\text{\tiny{$\wedge$}}}]^{I,}\cdot \delta
 \mathbf{r}_i\right) \frac{\partial w_{is}}{\mathbf{X}_j}} \ + \\
@@
 $j^{\text{ème}}$ colonne de la matrice $\mathbf{M}$ et $r^0_{,J} = [\mathbf{R}_i^0]^{J,}$.
-En réécrivant les 9 valeurs de $\delta F^{s}$ sous forme d'un vecteur colonne, l'équation (\ref{eq:def_grad_der}) peut être exprimée sous la forme d'un produit
+En réécrivant les neuf valeurs de $\delta \mathbf{F}^{s}$ sous forme d'un vecteur colonne, l'équation (\ref{eq:def_grad_der}) peut être exprimée sous la forme
+d'un produit
 matrice-vecteur :
@@
 	\begin{array}{c}
-		\delta F_{11}^s \\
-		\delta F_{12}^s \\
+		\delta \mathbf{F}_{11}^s \\
+		\delta \mathbf{F}_{12}^s \\
 		\vdots \\
-		\delta F_{33}^s
+		\delta \mathbf{F}_{33}^s
 	\end{array}
@@
 \begin{split}
-	\delta L_{12} &= \frac{1}{2}\left([\delta F^t]^{1,}\cdot[F]^{,2} + [F^t]^{1,}\cdot[\delta F]^{,2}\right) \\
-	&= \frac{1}{2}\left([\delta F_{11} \ \delta F_{21} \ \delta F_{31}]\cdot[F]^{,2} + [F^t]^{1,}\cdot[\delta F_{12} \ \delta F_{22} \ \delta F_{32}]\right)
+	\delta L_{12} &= \frac{1}{2}\left([\delta \mathbf{F}^t]^{1,}\cdot[\mathbf{F}]^{,2} + [\mathbf{F}^t]^{1,}\cdot[\delta \mathbf{F}]^{,2}\right) \\
+	&= \frac{1}{2}\left([\delta \mathbf{F}_{11} \ \delta \mathbf{F}_{21} \ \delta \mathbf{F}_{31}]\cdot[\mathbf{F}]^{,2} + [\mathbf{F}^t]^{1,}\cdot[\delta
+\mathbf{F}_{12} \ \delta \mathbf{F}_{22} \ \delta \mathbf{F}_{32}]\right)
 \end{split}
@@
 \begin{equation}
-	[\delta F_{xx} \ \delta F_{yx} \ \delta F_{zx}] = [B_{1,}\cdot \delta q \ B_{3,}\cdot \delta q \  B_{7,}\cdot \delta q]
+	[\delta \mathbf{F}_{11} \ \delta \mathbf{F}_{21} \ \delta \mathbf{F}_{31}] = [B_{1,}\cdot \delta \mathbf{q} \ B_{3,}\cdot \delta \mathbf{q} \  B_{7,}\cdot
+\delta \mathbf{q}]
 \end{equation}
@@
 \begin{equation}
-	\delta L_{xy} = \frac{1}{2}\left[F_{xy}B_{1,} +  F_{yy}B_{4,} + F_{zy}B_{7,} + F_{xx}B_{2,} + F_{yx}B_{5,} + F_{zx}B_{8,} \right] \cdot \delta q
+\begin{split}
+	&\delta \mathbf{L}_{12} =  \\
+	&\frac{1}{2}\left[\mathbf{F}_{12}B_{1,} +  \mathbf{F}_{11}B_{4,} + \mathbf{F}_{31}B_{7,} + \mathbf{F}_{11}B_{2,} + \mathbf{F}_{21}B_{5,} +
+\mathbf{F}_{31}B_{8,} \right] \cdot \delta \mathbf{q}
+\end{split}	
 \end{equation}
@@
 	\begin{array}{c}
-		\delta L_{11}^s \\
-		\delta L_{12}^s \\
+		\delta \mathbf{L}_{11}^s \\
+		\delta \mathbf{L}_{12}^s \\
 		\vdots \\
-		\delta L_{33}^s
+		\delta \mathbf{L}_{33}^s
 	\end{array}
-	\right] = \sum_{i}{\mathbf{M}^{i;s} \cdot \delta \mathbf{q}_i}
+	\right] = \sum_{i}{\mathbf{M}^{i,s} \cdot \delta \mathbf{q}_i}
 	\label{eq:delta_L}	
@@
 \end{equation}
-Cepedant, l'équation (\ref{eq:force_continuous}) nous donne l'expression de la force dans le cas continu. Etant donné que cette intégrale ne peut être résolue
+Cependant, l'équation (\ref{eq:force_continuous}) nous donne l'expression de la force dans le cas continu. Etant donné que cette intégrale ne peut être résolue
 analytiquement, hormis pour des cas simples, il est nécessaire de passer à une version discrète. L'intégrale de l'équation (\ref{eq:force_continuous}) doit être
-évaluée sur la totalité du domaine $\Omega$ occupé par l'objet. Pour ce faire, sous la forme discrète, on utilise les points d'intégration, aussi appellés
+évaluée sur la totalité du domaine $\Omega$ occupé par l'objet. Pour ce faire, sous la forme discrète, on utilise les points d'intégration, aussi appelés
 \emph{points de Gauss} qui échantillonnent l'objet. On obtient ainsi l'expression discrète de la force en un n{\oe}ud :
 \begin{equation}
-	f_i = \sum_{s}{\mathbf{M}^{i,s}\{\mathbf{S}^s\}w_s(\mathbf{x}_i)}
+	\mathbf{f}_i = \sum_{s}{\mathbf{M}^{i,s}\{\mathbf{S}^s\}w_s(\mathbf{x}_i)}
 	\label{eq:force_discrete}
@@
 d'énergie supplémentaire conduisent la simulation à
-``exploser''. Un solution courament employée pour répondre à ce problème consiste à utiliser des pas de temps très court avec pour effet de ralentir la
+``exploser''. Un solution couramment employée pour répondre à ce problème consiste à utiliser des pas de temps très court avec pour effet de ralentir la
 simulation.
-Pour ces raisons, on utilise un solveur implicte qui est inconditionnellement stable -- une proprièté intéressante dans le cadre de la simulation
+Pour ces raisons, on utilise un solveur implicite qui est inconditionnellement stable -- une propriété intéressante dans le cadre de la simulation
 d'objets déformables. Les solveurs implicites permettent de résoudre un problème non-linéaire de manière itérative selon la méthode de \emph{Newton-Raphson}
 \cite{faure:hal-00681539}. Pour permettre au solveur de résoudre l'équation du mouvement, il est nécessaire de linéariser la force, c'est-à-dire dériver
-l'expression de la force par rapport aux repères. Cela revient à exprimer la variation de la force en fonction d'une variation inifitésimale de la
+l'expression de la force par rapport aux repères. Cela revient à exprimer la variation de la force en fonction d'une variation infinitésimale de la
 contrainte. En effet, d'après \cite{NFECS} on a :
 \begin{equation}
-	\delta f = \int_{\Omega}{\mathbf{M^t}\{\delta \mathbf{S}\}d\Omega}
+	\delta \mathbf{f} = \int_{\Omega}{\mathbf{M^t}\{\delta \mathbf{S}\}d\Omega}
 	\label{eq:force_der}
@@
 \end{equation}
-ou la matrice $\mathbf{C}$ donnée en annexe est la matrice des coefficients elastiques associés à la loi de Hooke. En substituant l'équation (\ref{eq:delta_L})
+ou la matrice $\mathbf{C}$ donnée en annexe est la matrice des coefficients élastiques associés à la loi de Hooke. En substituant l'équation (\ref{eq:delta_L})
 à l'équation (\ref{eq:force_der}) on obtient :

### 2012-09-21 papier_afig/frame_refig.tex
Papier version finale + figures
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@462 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 %\usepackage{hyperref}
+\usepackage{caption}
+\usepackage{subcaption}
 \usepackage{amsmath}
@@
 
-\author[Julien Bosman]{Julien Bosman$^1$, Christian Duriez$^2$, Stéphane Cotin$^2$ \\ %
-	$^1$LIFL/INRIA ,$^2$INRIA
+\author[Julien Bosman]{Julien Bosman, Christian Duriez, Stéphane Cotin \\ %
+	INRIA équipe SHACRA / LIFL - Université Lille 1
 }
@@
 \begin{abstract}
-   A faire
+Nous proposons un modèle de simulation de tissus organiques reposant sur des méthodes sans maillage basées sur des repères. Le réalisme des simulations
+repose sur une formulation des forces faisant appel à la mécanique des milieux continus. Notre approche permet de simuler des objets déformables aux topologies
+diverses (volumique, surfacique et linéiques). Dans le soucis de d'obtenir une simulation stable nous proposons un formulation implicite du problème.
 
 \end{abstract}
-\keywords{Simulation, Méthodes sans maillage, Mécanique}
+\keywords{Simulation médicale, Méthodes sans maillage, Mécanique, Solides déformables}
 %-------------------------------------------------------------------------
 \section{Introduction}
-La simulation médicale est un domaine de plus en plus actif. Les simulateurs médicaux ont un champ d'application assez vaste allant du diagnostic à la
-formation des practiciens ou encore en aide à la chirurgie per et post opératoire. Jusqu'à maintenant, les organes simulés étaient directement en contact les
-uns avec les autres. Cependant, cette situation n'est pas conforme à la réalité. En effet, à l'intérieur du corps humains, les organes sont très rarement en
-contact direct. Ils sont généralement reliés entre eux par des fibres appelées \emph{tissus connectifs}. Ces tissus peuvent être présent en une couche très fine
-ou en un volume plus important. Ils sont difficiles à modéliser et à simuler car étant donné qu'ils occupent l'espace présent entre les organes, ils n'ont pas
-de topologie réellement définie.
 
-\section{Etat de l'art}
+La simulation médicale est une domaine de recherche de plus en plus actif et les récents progrès ont permis d'ouvrir de nouveaux champs d'application.
+Initialement centrés sur la pratique chirurgicale ces travaux restent nécessaires car les simulateurs commerciaux actuels souffrent parfois d'un manque de
+réalisme. Cependant, les résultats obtenus en recherche permettent d'améliorer sensiblement la précision et l'efficacité des simulations tout en offrant la
+possibilité d'intégrer des données basées sur des patients réels \cite{DPP12}, \cite{LPC+12} ou encore \cite{Hostettler2010169}. Ces travaux permettent
+d'envisager de nouvelles applications, notamment la planification pré-opératoire ou encore la simulation comme assistance aux chirurgiens durant l'intervention.
 
-La simulation à base de méthodes sans maillage à été introduite en informatique graphique par Cani et Desbrun dans \cite{Desbrun1996} et l'utilisation des
-\emph{Smoothed Particle Hydrodynamics} (SPH) -- reposant sur l'interaction réciproque d'un ensemble de particules -- pour l'animation de modèles très
-déformables. Les SPH demeurent une méthodes populaires pour l'animation de fluides et d'objets très déformables. Les travaux de \cite{Becker2009} apportent une
-réponse au problème des non-linéarités géométrique en proposant un modèle corotationnel dédié. La méthode des SPH permet également de simuler des matériaux
-non-newtoniens aux comportements plus complexes notamment dans \cite{Paiva2009} et \cite {hosseini06}. Cependant, Les SPH souffrent de certains problèmes qui
-limitent leur utilisation notamment leur faible précision. Le traitement des frontières est également difficile en raison de la déficience en
-particule sur le frontières de l'objet. De plus, le phénomène connu sous le nom de \emph{tensile instability}, qui apparaît lorsque l'objet est fortement
-étiré, rend son utilisation problématique dans le cadre de la simulation de grandes déformation. Parallèlement, d'autres méthodes d'animation sans maillage ont
-été développées. Des modèles basés sur la mécanique des milieux continus à l'instar de \cite{Keiser2004} et de \cite{Horton2010} où les champs de déplacement
-sont approximés grâce aux \emph{Moving} Least Squares (MLS). Cependant, l'approximation MLS est impossible si la distribution des n{\oe}uds est colinéaire ou
-coplanaire. En effet, la  ``moment matrix'' devient alors singulière et les fonctions de forme ne peuvent être calculées.
-Des techniques d'animation, non basées sur la physique,ont également été développées. Dans \cite{Muller2005}, les objets sont déformés grâce à un système masse
-ressort en utilisant la méthode dite de \emph{Shape Matching}, où les points de l'objet sont partitionnés en plusieurs groupes dont on estime une rotation
-optimale et une translation par rapport à la configuration non déformée. Cette idée est reprise dans \cite{Muller2011} à la différence près que chaque possède
-une orientation qui est estimée au grâce à un \emph{Shape Matching} généralisé. Enfin, dans \cite{GilBouFauPai11}, Gilles \textit{et al.} proposent un
-modèle compatible avec la mécanique des milieux continus basé sur des repères. Cependant, leur approche concerne essentiellement des objets faiblement
-échantillonnés. De fait, elle est difficile à utiliser pour modéliser des objets qui doivent se conformer étroitement à des surfaces.
+Ces applications ont en commun la nécessité de continuer à fiabiliser la simulation et à améliorer l'interactivité, d'une part en améliorant les modèles
+biomécaniques et d'autre part en réduisant au maximum les temps de calcul. Parmi les travaux récents, nombreux sont ceux qui traitent d'organes spécifique
+tels que le foie \cite{marchesseau}, le cerveau \cite{Horton:2007} ou encore les organes du système pelvien \cite{LMLwww_2553}. Une des
+grande difficulté dans la chirurgie abdominale et de pouvoir fournir à ces modèles des conditions aux limites acceptables. Actuellement, peu de travaux ont
+cherché à modéliser les interactions mécaniques entre les organes. Dans \cite{CJADLC11}, ces interactions sont modélisés grâce à une loi de contact frottant
+associé à une méthode de résolution optimisée. Cependant, bien que constituant un progrès, ce modèle n'est pas totalement conforme à la réalité car les organes
+sont très rarement en contact direct.
+
+En effet, les organes sont généralement entourés et reliés entre eux par des fibres appelées \emph{tissus connectifs} ou \emph{fascias}. Ces tissus sont
+difficilement visibles par imagerie médicale, possèdent des géométries complexes et mal documentées. Pour ces raisons, ils sont très souvent négligés dans la
+plupart des simulations. Pourtant, ils jouent un rôle mécanique important étant donné qu'ils influencent les conditions aux limites des organes.
+
+Dans ce travail, nous cherche à développer une méthode numérique capable de simuler ces tissus. La principale difficulté réside dans le fait qu'ils ne possèdent
+pas une topologie uniforme dans la mesure où ils occupent l'espace présent entre les organes et qu'ils ont des géométries très variées (volumique, surfacique
+voire linéiques). Nous proposons donc une approche nouvelle, faisant appel aux techniques de simulation sans maillage, ici basées sur des repères capables
+d'être couplés avec d'autres modèles mécaniques simulant des organes.
+
+\subsection{Etat de l'art}
+
+Couramment employées en simulation mécanique, la simulation à base de méthodes sans maillage à été introduite en informatique graphique par Cani et Desbrun dans
+\cite{Desbrun1996} et l'utilisation des \emph{Smoothed Particle Hydrodynamics} (SPH) -- reposant sur l'interaction réciproque d'un ensemble de particules --
+pour l'animation de modèles très déformables. Les SPH demeurent une méthodes populaires pour l'animation de fluides et d'objets très déformables. Les travaux de
+\cite{Becker2009} apportent une réponse au problème des non-linéarités géométrique en proposant un modèle corotationnel dédié. La méthode des SPH permet
+également de simuler des matériaux non-newtoniens aux comportements plus complexes notamment dans \cite{Paiva2009} et \cite {hosseini06}. Cependant, Les SPH
+souffrent de certains problèmes qui limitent leur utilisation notamment leur faible précision. Le traitement des frontières est également difficile en raison de
+la déficience en particule sur le frontières de l'objet. De plus, le phénomène connu sous le nom de \emph{tensile instability}, qui apparaît lorsque l'objet est
+fortement étiré, rend son utilisation problématique dans le cadre de la simulation de grandes déformation. Parallèlement, d'autres méthodes d'animation sans
+maillage ont été développées. Des modèles basés sur la mécanique des milieux continus à l'instar de \cite{Keiser2004} et de \cite{Horton2010} où les champs de
+déplacement sont approximés grâce aux \emph{Moving} Least Squares (MLS), une méthode offrant une bonne précision, couramment employée en mécanique. Cependant,
+l'approximation MLS est impossible si la distribution des n{\oe}uds est colinéaire ou coplanaire. En effet, la  ``moment matrix'' devient alors singulière et
+les fonctions de forme ne peuvent être calculées. Des techniques d'animation, non basées sur la physique,ont également été développées. Dans \cite{Muller2005},
+les objets sont déformés grâce à un système masse ressort en utilisant la méthode dite de \emph{Shape Matching}, où les points de l'objet sont partitionnés en
+plusieurs groupes dont on estime une rotation optimale et une translation par rapport à la configuration non déformée. Cette idée est reprise dans
+\cite{Muller2011} à la différence près que chaque possède une orientation qui est estimée au grâce à un \emph{Shape Matching} généralisé. Cependant, malgré leur
+rapidité et leur simplicité d'utilisation, ces méthodes ne permettent pas, à proprement parler, une simulation basée sur les loi de la physique. Enfin, dans
+\cite{GilBouFauPai11}, Gilles \textit{et al.} proposent un modèle compatible avec la mécanique des milieux continus basé sur des repères. Cependant, leur
+approche vise essentiellement des objets faiblement échantillonnés. De fait, elle est difficile à utiliser pour modéliser des objets  qui doivent se
+conformer étroitement à des surfaces. Pour une nomenclature des méthodes sans maillage, le lecteur pourra se référer à \cite{Fries04}.
+
+\subsection{Contribution}
+Nos contributions sont multiples. Notre travail offre la possibilité de simuler les tissus connectifs grâce à une méthode permettant le couplage entre
+différents modèles mécaniques sans poser de problème de la discrétisation des objets dans la mesure où elle est basée sur des méthodes sans maillage. Notre
+approche permet également de simuler des structures aussi bien volumiques, surfaciques que linéiques. Enfin nous adaptons les formulation de la mécanique des
+milieux continus à des objets dont les n{\oe}uds possèdent six degrés de liberté.
+
 \section{Simulation}
@@
 \end{equation}
-où $\mathbf{x}_i$ sont les coordonnées du n{\oe}ud et $\mathbf{r}_i$ le vecteurs rotation correspondant à l'orientation du n{\oe}ud. Les points d'intégration ne
-possèdent que trois DDL. Dans la mesure où l'on utilise une formulation en lagragien total, la position globale des points d'intégration est exprimée en
-fonction de leur position dans chacun des repères dans la configuration non déformée. Les coordonnées d'un point d'intégration $s$ dans le repère $i$ dans la
-configuration non déformée est donné par :
+où $\mathbf{x}_i$ représente les coordonnées du n{\oe}ud et $\mathbf{r}_i$ le vecteurs rotation correspondant à l'orientation du n{\oe}ud. Les points
+d'intégration ne possèdent que trois DDL. Dans la mesure où l'on utilise une formulation en lagragien total, la position globale des points d'intégration est
+exprimée en fonction de leur position dans chacun des repères dans la configuration non déformée. Les coordonnées d'un point d'intégration $s$ dans le repère
+$i$ dans la configuration non déformée est donné par : 
 \begin{equation}
@@
 	%\mathbf{r}_i = \sum_{s}{w_s(\mathbf{x}_i)\left( \mathbf{\bar{p}}_s^{l,i} \wedge \mathbf{u}_s^{l,i} \right)}
-	\frac{d\mathbf{r}_i}{d\mathbf{p}} = \sum_{s}{w_s(\mathbf{x}_i)(\mathbf{l}_s^i \wedge d\mathbf{p}_s)}
+	\frac{d\mathbf{r}_i}{d\mathbf{p}} = \sum_{s}{w_s(\bar{\mathbf{x}}_i)(\mathbf{l}_s^i \wedge d\mathbf{p}_s)}
 \end{equation}
@@
 L'approche employée dans ces travaux se base sur la mécanique des milieux continus (MMC). Les matériaux étudiés sont considérés comme continus, c'est-à-dire
-qu'ils sont étudiés à une échelle microscopique, mais une échelle suffisament grande pour négliger les espaces inter-moléculaires. La MMC permet de
+qu'ils sont étudiés à une échelle microscopique, mais une échelle suffisamment grande pour négliger les espaces inter-moléculaires. La MMC permet de
 mettre en
@@
 	\underbrace{\mathbf{u}_i}_{\text{déplacements}} \rightarrow \underbrace{\mathbf{F}_i}_{\substack{\text{gradient de} \\ \text{déformation}}} \rightarrow
-\underbrace{\L_i}_{\substack{\text{tenseur de} \\ \text{déformation}}} \rightarrow \underbrace{\S_i}_{\substack{\text{tenseur de} \\
+\underbrace{\mathbf{L}_i}_{\substack{\text{tenseur de} \\ \text{déformation}}} \rightarrow \underbrace{\mathbf{S}_i}_{\substack{\text{tenseur de} \\
 \text{contrainte}}} \rightarrow \underbrace{\mathbf{f}_i}_{\text{force}}
@@
 \bar{\mathbf{l}}_s^{i} \right) + w_{is}[\mathbf{R}_i] [\bar{\mathbf{R}}_i]^t \right] }
-%& = \sum_{i}{\left[ \nabla w_s(\mathbf{x}_i) \left(\mathbf{x}_i + \mathbf{R}_i \times \bar{\mathbf{p}}_s^{l,i} \right) + w_s(\mathbf{x}_i)\mathbf{R}_i
-%\mathbf{R}_i^{0t} \right] }
 \end{equation}
-où $w_{is} = w_s(\mathbf{x}_i)$. Afin d'obtenir la dérivée du tenseur de \emph{Green-Lagrange} de l'équation (\ref{eq:green_lagrange_der}) il est nécessaire de
-calculer la dérivée du tenseur de déformation par rapport à $\delta \mathbf{q}$ qui est une matrice $3\times 3$. On exprime alors ces neuf valeurs en fonction
-d'une variation infinitésimale de déplacement et d'orientation, c'est-à-dire en fonction de $\delta \mathbf{x}$ et $\delta \mathbf{r}$ :
+où $w_{is} = w_s(\bar{\mathbf{x}}_i)$. Afin d'obtenir la dérivée du tenseur de \emph{Green-Lagrange} de l'équation (\ref{eq:green_lagrange_der}) il est
+nécessaire de calculer la dérivée du tenseur de déformation par rapport à $\delta \mathbf{q}$ qui est une matrice $3\times 3$. On exprime alors ces neuf valeurs
+en fonction d'une variation infinitésimale de déplacement et d'orientation, c'est-à-dire en fonction de $\delta \mathbf{x}$ et $\delta \mathbf{r}$ :
 \begin{equation}
@@
 \begin{split}
-	\delta L_{12} &= \frac{1}{2}\left([\delta \mathbf{F}^t]^{1,}\cdot[\mathbf{F}]^{,2} + [\mathbf{F}^t]^{1,}\cdot[\delta \mathbf{F}]^{,2}\right) \\
+	\delta \mathbf{L}_{12} &= \frac{1}{2}\left([\delta \mathbf{F}^t]^{1,}\cdot[\mathbf{F}]^{,2} + [\mathbf{F}^t]^{1,}\cdot[\delta \mathbf{F}]^{,2}\right) \\
 	&= \frac{1}{2}\left([\delta \mathbf{F}_{11} \ \delta \mathbf{F}_{21} \ \delta \mathbf{F}_{31}]\cdot[\mathbf{F}]^{,2} + [\mathbf{F}^t]^{1,}\cdot[\delta
@@
 \end{equation}
-avec $B_{i,}$ la $i^{\text{ème}}$ ligne de la matrice $B$ d'où :
+avec $B_{i,}$ la $i^{\text{ème}}$ ligne de la matrice $\mathbf{B}$ d'où :
 \begin{equation}
@@
 \begin{equation}
-	\mathbf{f}_i = \sum_{s}{\mathbf{M}^{i,s}\{\mathbf{S}^s\}w_s(\mathbf{x}_i)}
+	\mathbf{f}_i = \sum_{s}{\mathbf{M}^{i,s}\{\mathbf{S}^s\}w_s(\bar{\mathbf{x}}_i})
 	\label{eq:force_discrete}
@@
 l'expression de la force par rapport aux repères. Cela revient à exprimer la variation de la force en fonction d'une variation infinitésimale de la
-contrainte. En effet, d'après \cite{NFECS} on a :
+contrainte. En effet, d'après \cite{NFECS}(\S6.4.2) la dérivée des forces internes se décompose en deux parties :
+\begin{equation}
+	\delta \mathbf{f}^{int} = \delta \mathbf{f}^{mat} + \delta \mathbf{f}^{geo}
+\end{equation}
+où $\delta \mathbf{f}^{mat}$ est la ``material tangent stiffness'' et $\delta \mathbf{f}^{geo}$ est la ``geometric stiffness''. Dans notre cas nous nous
+limitons à l'utilisation de la première dont l'expression est :
 \begin{equation}
@@
 \begin{equation}
-	\delta \mathbf{f}_i = \sum_{s}{\mathbf{M}^{i,s}\cdot \Delta \mathbf{S}^s}
+	\delta \mathbf{f}_i = \sum_{s}{\mathbf{M}^{i,s}\cdot \Delta \mathbf{S}^s}w_s(\bar{\mathbf{x}}_i)
 \end{equation}
@@
 
-\subsection{Figures en couleur} Etant donné qu'il s'agit d'une publication électronique, il n'y a aucune censure à avoir sur les figures en couleur (cf figure \ref{fig:ex3}.
+\section{Résultats}
+Les tests de simulation de poutres modélisées avec notre méthode de la figure ({\ref{fig:beams}) montrent des résultats cohérents. Chaque test à été effectué
+avec un module de Young différent. On peut clairement observer que la rigidité de la poutre diminue avec le module de Young. La figure (\ref{fig:mod_beam})
+montre la discrétisation des poutres avec notre méthode. On peut observer que l'orientation des repères évolue conformément au fléchissement subi par la
+poutre. La figure (\ref{fig:torsion}) montre un test de torsion effectué sur une structure linéique. Le repère de gauche est fixé en rotation et translation et
+seul le repère le plus à droite à subi une rotation d'un demi tour. Initialement, tous les repères avaient la même orientation (identique à celle du repère
+fixé). On peut voir que tous les repères de la structure ont subi une torsion. La figure (\ref{fig:shells}) montre le résultat d`'un test de couplage entre
+notre méthode et d'autres modèles mécanique, en l'occurrence des éléments coque ou \emph{shells}. Ce cas test représente typiquement la situation où deux
+organes (en rouge) sont reliés par des tissus connectifs (en bleu). La surface supérieure est fixée sur deux bords opposés, les bords surface inférieure sont
+totalement libres et le centre des deux surfaces est attaché de part et d'autre de l'objet bleu. On peut observer que les conditions aux limites sont transmises
+d'un modèle à l'autre tandis que les déformations subies par les objets sont cohérentes.
 
-\begin{figure*}[htb]
+En terme de performances, notre modèle est moins rapide que la méthode des éléments finis à nombre d'éléments égal ($\approx 15$ itérations par seconde), comme
+généralement rapporté dans la littérature.
+
+\begin{figure*}[h]
         \centering
-  \mbox{} \hfill
-  \includegraphics[width=.3\linewidth]{logoafig}
-  \hfill
-  \includegraphics[width=.3\linewidth]{logoafig}
-  \hfill \mbox{}
-  \caption{\label{fig:ex3}
-           Les figures en couleurs ou en noir et blanc peuvent être placées n'importe ou dans le document. }
+        \begin{subfigure}[b]{0.3\textwidth}
+                \centering
+                \includegraphics[width=\textwidth]{beam_E20000}
+                \caption{$E=20000$}
+        \end{subfigure}%
+        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. 
+          %(or a blank line to force the subfigure onto a new line)
+        \begin{subfigure}[b]{0.3\textwidth}
+                \centering
+                \includegraphics[width=\textwidth]{beam_E10000}
+                \caption{$E=10000$}
+        \end{subfigure}
+        ~ %add desired spacing between images, e. g. ~, \quad, \qquad etc. 
+          %(or a blank line to force the subfigure onto a new line)
+        \begin{subfigure}[b]{0.3\textwidth}
+                \centering
+                \includegraphics[width=\textwidth]{beam_E3000}
+                \caption{$E=3000$}
+        \end{subfigure}
+        \caption{Poutres simulées avec différents modules de Young}\label{fig:beams}
 \end{figure*}
+
+\section{Conclusion et travaux futurs}
+La méthode présentée permet de simuler des tissus connectifs sur des topologies variées. Elle permet également l'imposition des conditions aux limites entre
+des objets simulés grâce à différentes méthodes, offrant ainsi la possibilité de simuler les interactions entre les organes. Les travaux futurs seront orientés
+selon deux axes principaux : performances et précision. En effet, dans le but d'offrir à l'utilisateur un maximum d'interactivité, il est nécessaire de se
+rapprocher de performances temps-réel. Dans cette optique, les travaux porteront sur l'implémentation de cette méthode sur GPU. Il est raisonnable de penser
+qu'il est possible d'obtenir un gain de performance non négligeable car ce type de problème est bien adapté à la parallélisation. Concernant la précision, les
+futurs travaux s'attacheront développer une expression de la ``geometric stiffness'' adaptée à notre problème et ainsi fournir une expression plus complète de
+la dérivée des forces internes. Enfin, les travaux porteront également sur l'ajout de loi de comportement, notamment hyperélastique mais également anisotrope
+car la plupart des tissus organiques possèdent ces propriétés.
+
+\begin{figure}[htb]
+	\centering
+	\includegraphics[width=\columnwidth]{beam_frames}
+	\caption{Modélisation de la poutre avec notre méthode.}\label{fig:mod_beam}
+\end{figure}
+
+\begin{figure}[htb]
+	\centering
+	\includegraphics[width=\columnwidth]{torsion}
+	\caption{Test de torsion sur une structure linéique}\label{fig:torsion}
+\end{figure}
+
+\begin{figure*}{htb}
+	\centering\includegraphics[width=.7\textwidth]{shells}
+	\caption{Couplage multi-modèle}\label{fig:shells}
+\end{figure*}
+
+% \begin{figure*}[htb]
+%   \centering
+%   \mbox{} \hfill
+%   \includegraphics[width=.3\linewidth]{logoafig}
+%   \hfill
+%   \includegraphics[width=.3\linewidth]{logoafig}
+%   \hfill \mbox{}
+%   \caption{\label{fig:ex3}
+%            Les figures en couleurs ou en noir et blanc peuvent être placées n'importe ou dans le document. }
+% \end{figure*}
 
@@
 		\frac{\partial w_{is}}{\partial x} & 0 & 0 & \left[\frac{\partial w_{is}}{\partial x}
-		[-\mathbf{R}_{i}(\mathbf{\bar{p}}_s^{l,i})^{\text{\tiny{$\wedge$}}}]^{x,} - [\mathbf{R}_{i}(r^0_{,x})^{\text{\tiny{$\wedge$}}}]^{x,}\ w_{is}
+		[-\mathbf{R}_{i}(\mathbf{\bar{l}}_s^{i})^{\text{\tiny{$\wedge$}}}]^{x,} - [\mathbf{R}_{i}(r^0_{,x})^{\text{\tiny{$\wedge$}}}]^{x,}\ w_{is}
 		\right]\\
 		\frac{\partial w_{is}}{\partial y} & 0 & 0 & \left[\frac{\partial w_{is}}{\partial y}
-		[-\mathbf{R}_{i}(\mathbf{\bar{p}}_s^{l,i})^{\text{\tiny{$\wedge$}}}]^{x,} - [\mathbf{R}_{i}(r^0_{,y})^{\text{\tiny{$\wedge$}}}]^{x,}\ w_{is}
+		[-\mathbf{R}_{i}(\mathbf{\bar{l}}_s^{i})^{\text{\tiny{$\wedge$}}}]^{x,} - [\mathbf{R}_{i}(r^0_{,y})^{\text{\tiny{$\wedge$}}}]^{x,}\ w_{is}
 		\right] \\
 		\frac{\partial w_{is}}{\partial z} & 0 & 0 & \left[\frac{\partial w_{is}}{\partial z}
-		[-\mathbf{R}_{i}(\mathbf{\bar{p}}_s^{l,i})^{\text{\tiny{$\wedge$}}}]^{x,} - [\mathbf{R}_{i}(r^0_{,z})^{\text{\tiny{$\wedge$}}}]^{x,}\ w_{is}
+		[-\mathbf{R}_{i}(\mathbf{\bar{l}}_s^{i})^{\text{\tiny{$\wedge$}}}]^{x,} - [\mathbf{R}_{i}(r^0_{,z})^{\text{\tiny{$\wedge$}}}]^{x,}\ w_{is}
 		\right]\\		
 		0 & \frac{\partial w_{is}}{\partial x} & 0 & \left[\frac{\partial w_{is}}{\partial x}
-		[-\mathbf{R}_{i}(\mathbf{\bar{p}}_s^{l,i})^{\text{\tiny{$\wedge$}}}]^{y,} - [\mathbf{R}_{i}(r^0_{,x})^{\text{\tiny{$\wedge$}}}]^{y,}\ w_{is}
+		[-\mathbf{R}_{i}(\mathbf{\bar{l}}_s^{i})^{\text{\tiny{$\wedge$}}}]^{y,} - [\mathbf{R}_{i}(r^0_{,x})^{\text{\tiny{$\wedge$}}}]^{y,}\ w_{is}
 		\right]\\
 		0 & \frac{\partial w_{is}}{\partial y} & 0 & \left[\frac{\partial w_{is}}{\partial y}
-		[-\mathbf{R}_{i}(\mathbf{\bar{p}}_s^{l,i})^{\text{\tiny{$\wedge$}}}]^{y,} - [\mathbf{R}_{i}(r^0_{,y})^{\text{\tiny{$\wedge$}}}]^{y,}\ w_{is}
+		[-\mathbf{R}_{i}(\mathbf{\bar{l}}_s^{i})^{\text{\tiny{$\wedge$}}}]^{y,} - [\mathbf{R}_{i}(r^0_{,y})^{\text{\tiny{$\wedge$}}}]^{y,}\ w_{is}
 		\right] \\
 		0 & \frac{\partial w_{is}}{\partial z} & 0 & \left[\frac{\partial w_{is}}{\partial z}
-		[-\mathbf{R}_{i}(\mathbf{\bar{p}}_s^{l,i})^{\text{\tiny{$\wedge$}}}]^{y,} - [\mathbf{R}_{i}(r^0_{,z})^{\text{\tiny{$\wedge$}}}]^{y,}\ w_{is}
+		[-\mathbf{R}_{i}(\mathbf{\bar{l}}_s^{i})^{\text{\tiny{$\wedge$}}}]^{y,} - [\mathbf{R}_{i}(r^0_{,z})^{\text{\tiny{$\wedge$}}}]^{y,}\ w_{is}
 		\right]\\		
 		0 & 0 & \frac{\partial w_{is}}{\partial x} & \left[\frac{\partial w_{is}}{\partial x}
-		[-\mathbf{R}_{i}(\mathbf{\bar{p}}_s^{l,i})^{\text{\tiny{$\wedge$}}}]^{z,} - [\mathbf{R}_{i}(r^0_{,x})^{\text{\tiny{$\wedge$}}}]^{z,}\ w_{is}
+		[-\mathbf{R}_{i}(\mathbf{\bar{l}}_s^{i})^{\text{\tiny{$\wedge$}}}]^{z,} - [\mathbf{R}_{i}(r^0_{,x})^{\text{\tiny{$\wedge$}}}]^{z,}\ w_{is}
 		\right]\\
 		0 & 0 & \frac{\partial w_{is}}{\partial y} & \left[\frac{\partial w_{is}}{\partial y}
-		[-\mathbf{R}_{i}(\mathbf{\bar{p}}_s^{l,i})^{\text{\tiny{$\wedge$}}}]^{z,} - [\mathbf{R}_{i}(r^0_{,y})^{\text{\tiny{$\wedge$}}}]^{z,}\ w_{is}
+		[-\mathbf{R}_{i}(\mathbf{\bar{l}}_s^{i})^{\text{\tiny{$\wedge$}}}]^{z,} - [\mathbf{R}_{i}(r^0_{,y})^{\text{\tiny{$\wedge$}}}]^{z,}\ w_{is}
 		\right] \\
 		0 & 0 & \frac{\partial w_{is}}{\partial z} & \left[\frac{\partial w_{is}}{\partial z}
-		[-\mathbf{R}_{i}(\mathbf{\bar{p}}_s^{l,i})^{\text{\tiny{$\wedge$}}}]^{z,} - [\mathbf{R}_{i}(r^0_{,z})^{\text{\tiny{$\wedge$}}}]^{z,}\ w_{is}
+		[-\mathbf{R}_{i}(\mathbf{\bar{l}}_s^{i})^{\text{\tiny{$\wedge$}}}]^{z,} - [\mathbf{R}_{i}(r^0_{,z})^{\text{\tiny{$\wedge$}}}]^{z,}\ w_{is}
 		\right]\\		
@@
 \renewcommand\arraystretch{1}
-avec $w_{is} = w_s(\mathbf{x}_i)$
+avec $w_{is} = w_s(\bar{\mathbf{x}}_i)$
 \section{Matrice de Hooke}

### 2012-10-08 Doc/VRIPHYS2012/vriphys2012.tex
final version of VRIPHYS paper
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@463 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 % --- for  EG Workshop Proceedings
-\WsSubmission    % uncomment for submission to EG Workshop
-% \WsPaper         % uncomment for final version of EG Workshop contribution
+%\WsSubmission    % uncomment for submission to EG Workshop
+\WsPaper         % uncomment for final version of EG Workshop contribution
 %
@@
   \caption{\label{fig-#2}#3}
+\end{figure}}
+
+%\FigureI{name}{caption}
+\newcommand{\FigureI}[2]{%
+\begin{figure}[htb]
+  \centering
+  \input{#1.pdf_tex}
+  \caption{\label{fig-#1}#2}
 \end{figure}}
@@
 
-% for anonymous conference submission please enter your SUBMISSION ID
-% instead of the author's name (and leave the affiliation blank) !!
-% TODO: anonymize -- Submission ID
-       \author[paper1026]
-              {paper1026}
-%\author[T. Golembiovský \& C. Duriez]
-%       {Tomáš Golembiovský$^{1,2}$ and Christian Duriez$^{2}$\\
-%        $^1$Faculty of Informatics, Masaryk University, Czech Republic\\
-%        $^2$INRIA Lille, France
-%       } % TODO: ask Christian how to specify affiliation
+\author[T. Golembiovský \& C. Duriez]
+       {Tomáš Golembiovský$^{1,2}$ and Christian Duriez$^{2}$\\
+        $^1$Faculty of Informatics, Masaryk University, Czech Republic\\
+        $^2$INRIA Lille -- Nord Europe and University of Lille, France
+       }
 
@@
 \begin{classification} % according to http://www.acm.org/class/1998/
-    \CCScat{Computer Graphics}{I.3.5}{Computational Geometry and Object Modeling}{Physically based modelling}.
+    \CCScat{Computer Graphics}{I.3.5}{Computational Geometry and Object Modeling}{Physically based modeling}.
     %\CCScat{Computer Graphics}{I.3.7}{Three-Dimensional Graphics and Realism}{Animation}.
@@
 In the field of surgical simulation, we try to reproduce or anticipate the mechanical behavior of these structures during the surgery.
-Thus, it is very important to capture accurately the physical behavior of these structure.
+Thus, it is very important to capture accurately their physical behavior.
 However, in the same time, it is also very important to maintain fast computation rates, in order to be compatible with the targeted applications (interactive simulation for education and for planning).
@@
 This problem has been faced by the computer graphics community but for different applications.
-For instance, many related works address the problem of the simulation of cloth using mass-spring models \cite{Volino2009} or bending models \cite{Grinspun2003}. %% Todo Choi2007?
+For instance, many related works address the problem of the simulation of cloth using for example mass-spring models or bending models \cite{Volino2009,Grinspun2003}. %% Todo Choi2007?
 However, these models are discrete so their behavior depends on the mesh and their parameters are not easily measurable.
@@
 Methods based on continuum mechanics and namely theory of elasticity gained on popularity in real-time simulations ever since Terzopoulos \cite{Terzopoulos1987} presented his work on elastic deformation modeling.
-While the methods using finite elements and theory of elasticity are fairly popular in volumetric modeling of deformations \cite{Cotin1999,Nesme2006,Miller2007}, they are still frowned upon in the area of thin structures.
+While the methods using finite elements and theory of elasticity are fairly popular in volumetric deformation modeling \cite{Cotin1999,Nesme2006,Miller2007}, they are still frowned upon in the area of thin structures.
 Yet, a special field of continuum mechanics has studied the deformations of these objects, and has proposed models based on shell theory.
-This theory was integrated numerically using Finite Elements \cite{Reddy1993}, but there are two issues for obtaining interactive frame rates: 
-(i) the continuum equations requires second spatial derivations that can not be handled with linear interpolation and 
-(ii) the final system of equations that describes the mechanics is non-linear (if the deformations are not small).
+This theory is integrated numerically using Finite Elements \cite{Reddy1993}, but there are two issues for obtaining interactive frame rates: 
+(i) the continuum equations require second spatial derivatives that can not be handled with linear interpolation and 
+(ii) the final system of equations that describes the mechanics is non-linear if large deformations are present.
 
@@
 The work of Bouthors et al.\cite{Bouthors2007} is a good application example of this strategy for surface deformations.
-But this strategy is not sufficient in our case because the problem of performance is also greatly impacted when using elements that have an interpolation with high order polynomials.
+But this strategy is not sufficient in our case because the problem of performance is also greatly impacted when using elements with interpolation based on high order polynomials.
 This type of interpolation is necessary because bending continuum model
@@
 From this point of view, our work is derived from this approach but we address several noted drawbacks: the formulation misses one rotational degree of freedom, the cubic interpolation is not symmetric and produces $C^0$  discontinuities between elements on edges.
-Ubach et al. \cite{Ubach2010}  present a shell element based on B\'ezier triangles that allows for a better behavior.
-However, the B\'ezier interpolation defined on each triangle element is based on the motion computed on adjacent triangles.
-So the stiffness forces computed on each element not only depends on nodes of the element but also on all the adjacent nodes. 
+
+B\'ezier based shell elements in their general form have been presented
+for both triangular \cite{Ainsworth2011} and quadrilateral \cite{Boumechra2006} elements. High number of simulated DOFs makes such formulations impractical for simulations.
+Ubach et al.
+\cite{Ubach2010} presented a rotation-free 3-node shell element based on B\'ezier triangles where the element geometry is estimated from all the neighbouring elements, not just those sharing an edge.
 This greatly reduces the sparsity of the global stiffness matrix and consequently increases the computation time.
+Also, the lack of rotational degrees of freedom limits the application in surgical simulations.
+
+%Ubach et al. \cite{Ubach2010}  present a shell element based on B\'ezier triangles that allows for a better behavior.
+%However, the B\'ezier interpolation defined on each triangle element is based on the motion computed on adjacent triangles.
+%So the stiffness forces computed on each element not only depends on nodes of the element but also on all the adjacent nodes. 
+%This greatly reduces the sparsity of the global stiffness matrix and consequently increases the computation time.
 
@@
 resolution surface mesh onto the mechanical mesh and give formulas for
-computing velocities of points on the surface and projecting of forces. Section
+computing velocities of points on the surface and for projection of forces. Section
 5 presents test of convergence, speed analysis and two created applications
@@
 (ii) interpolation function of cubic B\'ezier triangle is applied to interpolate on the surface of the element.
-Mapping of high resolution mesh on the curved surface of the shell.
-Finally from the kinematic link we derive formulas for
-(i) interpolating velocity on the element surface and for
-(ii) distribution of force applied on the element surface back to 6 DOF nodes of the element.
+Finally from the kinematic link we derive equations for
+(i) interpolating velocity at arbitrary position on surface and for
+(ii) distribution of forces applied on the element surface back to 6 DOF nodes of the element.
 This allows us to use smooth mapped surface for collision detection and allows the propagation of proper collision response back onto the mechanical mesh.
@@
 %
-B\'ezier triangles were first designed in early sixties by French mathematician de Casteljau.
+%B\'ezier triangles were first designed in early sixties by French mathematician de Casteljau.
 As with other B\'ezier or B-spline surfaces the B\'ezier triangle is
@@
   T(\xi_1,\xi_2) & =
-             \xi_1^3\ P_1
-           + \xi_2^3\ P_2
-           + \xi_3^3\ P_3 \\
-         & + 3\ \xi_1^2 \xi_2\ P_4
-           + 3\ \xi_1^2 \xi_3\ P_5
-           + 3\ \xi_2^2 \xi_3\ P_6 \\
-         & + 3\ \xi_1 \xi_2^2\ P_7
-           + 3\ \xi_1 \xi_3^2\ P_8
-           + 3\ \xi_2 \xi_3^2\ P_9 \\
-         & + 6\ \xi_1 \xi_2 \xi_3\ P_{10} \\
+      \xi_1^3\ \mathbf{p}_1
+      + \xi_2^3\ \mathbf{p}_2
+      + \xi_3^3\ \mathbf{p}_3 \\
+      & + 3\ \xi_1^2 \xi_2\ \mathbf{p}_4
+      + 3\ \xi_1^2 \xi_3\ \mathbf{p}_5
+      + 3\ \xi_2^2 \xi_3\ \mathbf{p}_6 \\
+      & + 3\ \xi_1 \xi_2^2\ \mathbf{p}_7
+      + 3\ \xi_1 \xi_3^2\ \mathbf{p}_8
+      + 3\ \xi_2 \xi_3^2\ \mathbf{p}_9 \\
+      & + 6\ \xi_1 \xi_2 \xi_3\ \mathbf{p}_{10} \\
   \end{split}
@@
 where $ \xi_3 = 1 - \xi_1 - \xi_2 $. By naming $n_i$ the respective values
-of the Bernstein basis functions linked to control point $P_i$ we can
+of the Bernstein basis functions linked to control point $\mathbf{p}_i$ we can
 shortly express \eqref{eq-cubicbez} as:
@@
 \begin{equation}\label{eq-cubicbez2}
-    T = \sum_{i=1}^{10} n_i P_i
+    T = \sum_{i=1}^{10} n_i \mathbf{p}_i
 \end{equation}
@@
 
-The cubic B\'ezier triangles are defined by a mesh of 10 control points each, however the
-simulated object is usually described by a triangular mesh possibly with normals.
-A procedure for construction of initial B\'zier mesh is necessary.
-We construct the initial B\'ezier mesh by taking into account the curvature of the triangular
-mesh. Note that special care has to be taken to maintain
+%The cubic B\'ezier triangles are defined by a mesh of 10 control points each, however the
+Since the
+simulated object is usually described by a triangular mesh,
+a procedure for construction of initial B\'zier mesh is necessary.
+%We construct the initial B\'ezier mesh by taking into account the curvature of the triangular
+%mesh.
+Note that special care has to be taken to maintain
 the continuity between elements across the nodes and edges, 
@@
 
-Notice that the points on the edge will be same for both triangles sharing
-the edge.
+%Notice that the points on the edge will be same for both triangles sharing
+%the edge.
 
@@
 (see \cite{Ubach2010} or refer to the work of Farin \cite{Farin2002} for continuity conditions). 
-For simplicity we decided to keep the edge boundary only $C^0$ continuous and we compute the central point as a function of other 9 points: 
+For simplicity we keep the edge boundary only $C^0$ and we compute the central point as a function of other 9 points: 
 %
 \begin{equation}\label{eq-central}
-    P_{10} = \frac{1}{3}(\sum_{i=4}^9 P_i - \sum_{i=1}^3 P_i)
-\end{equation}
-%
-where the points $P_i$ are named as in Figure \ref{fig-bezier2}.
+    \mathbf{p}_{10} = \frac{1}{3}(\sum_{i=4}^9 \mathbf{p}_i - \sum_{i=1}^3 \mathbf{p}_i)
+\end{equation}
+%
+%where the points $P_i$ are named as in Figure \ref{fig-bezier2}.
 
@@
 
-With 10 control points and 3 DOFs per point it is a total of 30 DOFs for triangle.
-Simulation of such complex system would significantly reduce the sparsity of stiffness matrix.
-We propose a reduction of the number of degrees of freedom by using 3 nodes with 6 degrees of freedom
-by creating a kinematic link between the control points.
-Each of the edge control points is attached to nearest corner node. See Figure
-\ref{fig-segments2} for the correspondence between edge points and
+With 10 control points and 3 DOFs per point it is a total of 30 DOFs for element.
+We propose a reduction of the number of DOFs by using 3 nodes with 6 degrees of freedom
+and a kinematic link between control points.
+Each of the edge control points is attached to nearest corner node. See Figure~\ref{fig-segments2} for the correspondence between edge points and
 associated corner nodes. 
-In the following text the index $j \in [4;10]$ refers to edge points and index $i \in [1;3]$ to the associated corner point/node (unless $i$ is explicitly used differently).
+In the following text the index $j \in [4;10]$ refers to internal points and index $i \in [1;3]$ to the associated corner point/node %(unless $i$ is explicitly used differently).
+if not stated otherwise.
 
@@
 \begin{equation}
-    \overline{P_{i-j}} = R(\theta_i^0)^T P_{i-j}^0
+    \overline{\mathbf{p}_{i-j}} = R(\theta_i^0)^T \mathbf{p}_{i-j}^0
 \end{equation}
@@
 $R(\theta_i^0)$ is the rotation matrix for the angular position of node $i$ at the beginning of the simulation and
-$P_{i-j}^0 = P_j^0 - \mathbf{x}_i^0 = P_j^0 - P_i^0$ the attachment segment.
+$\mathbf{p}_{i-j}^0 = \mathbf{p}_j^0 - \mathbf{x}_i^0 = \mathbf{p}_j^0 - \mathbf{p}_i^0$ the attachment segment.
+The 0 in superscript denotes value in rest state.
 
@@
 \begin{equation}
-    P_j = \mathbf{x}_i + P_{i-j} = P_i + R(\theta_i) \overline{P_{i-j}}
+    \mathbf{p}_j = \mathbf{x}_i + \mathbf{p}_{i-j} = \mathbf{p}_i + R(\theta_i) \overline{\mathbf{p}_{i-j}}
 \end{equation}
@@
 \begin{equation}\label{eq-central2}
-    P_{10} =  \frac{1}{3}\left( \sum_{i=1}^3 \mathbf{x}_i + R(\theta_i) \overline{P_{i-10}} \right)
-\end{equation}
-
-Putting previous equations into \eqref{eq-cubicbez2} we get:
+    \mathbf{p}_{10} =  \frac{1}{3} \sum_{i=1}^3 \left( \mathbf{x}_i + R(\theta_i) \overline{\mathbf{p}_{i-10}} \right)
+\end{equation}
+
+Substituting previous equations into \eqref{eq-cubicbez2} we get:
 %
 \begin{equation}\label{global-transform} \begin{split}
-    & T = \sum_{i=1}^{10} n_i P_i = \sum_{i=1}^{3} n_i \mathbf{x}_i
+    & T = \sum_{i=1}^{10} n_i \mathbf{p}_i = \sum_{i=1}^{3} n_i \mathbf{x}_i
     + \sum_{j=4}^9 n_j  \left(
-        \mathbf{x}_i + R(\theta_i) \overline{P_{i-j}} \right) + \\
+    \mathbf{x}_i + R(\theta_i) \overline{\mathbf{p}_{i-j}} \right) + \\
     &\quad + \frac{n_{10}}{3}
-        \left( \sum_{i=1}^3 \mathbf{x}_i + R(\theta_i) \overline{P_{i-10}} \right)
+    \sum_{i=1}^3 \left( \mathbf{x}_i + R(\theta_i) \overline{\mathbf{p}_{i-10}} \right)
     \end{split}
@@
 
-A small variation $\delta \mathbf{x}_i$ of the node $i$ position and $\delta \mathbf{\theta}_i$ for the rotation produces the following small variation of position corner points $\delta P_i$ and edge points $\delta P_j$:
-%
-\begin{eqnarray}
-    \delta P_i & = & \delta \mathbf{x}_i \\
-    \delta P_j & = & \delta \mathbf{x}_i +  P_{i-j} \times \delta \mathbf{\theta}_i
+A small variation $\delta \mathbf{x}_i$ of the node $i$ position and $\delta \mathbf{\theta}_i$ for the rotation produces the following small variation of position corner points $\delta \mathbf{p}_i$ and edge points $\delta \mathbf{p}_j$:
+%
+\begin{eqnarray}\label{eq-point-rel}
+    \delta \mathbf{p}_i & = & \delta \mathbf{x}_i \\
+    \delta \mathbf{p}_j & = & \delta \mathbf{x}_i +  \mathbf{p}_{i-j} \times \delta \mathbf{\theta}_i
 \end{eqnarray}
 %
-and variation of central point $\delta P_{10}$:
-%
-\begin{equation}
-    \delta P_{10} = \frac{1}{3} \sum_{i=1}^3 \left(
-    \delta \mathbf{x}_i + P_{i-10} \times \delta \mathbf{\theta}_i \right)
+and variation of central point $\delta \mathbf{p}_{10}$:
+%
+\begin{equation}\label{eq-point-rel2}
+    \delta \mathbf{p}_{10} = \frac{1}{3} \sum_{i=1}^3 \left(
+    \delta \mathbf{x}_i + \mathbf{p}_{i-10} \times \delta \mathbf{\theta}_i \right)
 \end{equation}
@@
 %
-\begin{equation}
-    \delta T = \sum_{i=1}^{10} n_i  \delta P_i
+\begin{equation}\label{eq-bez-delta}
+    \delta T = \sum_{i=1}^{10} n_i  \delta \mathbf{p}_i
 \end{equation}
@@
 
-\begin{equation}\label{eq-bez-delta}
-    \begin{split}
-        \delta T = &   (n_1 + n_4 + n_5 + \frac{1}{3} n_{10})  \delta \mathbf{x}_1 + \\
-                   & + (n_2 + n_6 + n_7 + \frac{1}{3} n_{10})  \delta \mathbf{x}_2 + \\
-                   & + (n_3 + n_8 + n_9 + \frac{1}{3} n_{10})  \delta \mathbf{x}_3 + \\
-                   & + \sum_{j=4}^9 n_j P_{i-j} \times \delta \mathbf{\theta}_i + \\
-                   & + \frac{n_{10}}{3} \sum_{i=1}^3 P_{i-10} \times \delta \mathbf{\theta}_i 
-    \end{split}
-\end{equation}
-
-We have simplified the notation by leaving the variable $i$ free in the above sums ($\sum_{j=4}^9$) involving $P_{i-j}$.
-Note however that the expressions are sound and unambiguous because there is only one possible $i$ for each $j$ depending on the connection of segments to the corner nodes (see fig.
+Substituting \eqref{eq-point-rel}-\eqref{eq-point-rel2} into \eqref{eq-bez-delta} 
+one can evaluate the equation in terms of matrix multiplication:
+
+\begin{equation}
+    \delta T = \mathbf{J}_t \left[ \delta\mathbf{x}^x_1, \hdots, \delta\mathbf{x}^z_3,
+        \delta\theta^x_1, \ldots, \delta\theta^z_3 \right]^T
+\end{equation}
+
+%\begin{equation}\label{eq-bez-delta}
+%    \begin{split}
+%        \delta T = &   (n_1 + n_4 + n_5 + \frac{1}{3} n_{10})  \delta \mathbf{x}_1 + \\
+%                   & + (n_2 + n_6 + n_7 + \frac{1}{3} n_{10})  \delta \mathbf{x}_2 + \\
+%                   & + (n_3 + n_8 + n_9 + \frac{1}{3} n_{10})  \delta \mathbf{x}_3 + \\
+%                   & + \sum_{j=4}^9 n_j P_{i-j} \times \delta \mathbf{\theta}_i + \\
+%                   & + \frac{n_{10}}{3} \sum_{i=1}^3 P_{i-10} \times \delta \mathbf{\theta}_i 
+%    \end{split}
+%\end{equation}
+%
+%We have simplified the notation by leaving the variable $i$ free in the above sums ($\sum_{j=4}^9$) involving $P_{i-j}$.
+%Note however that the expressions are sound and unambiguous because there is only one possible $i$ for each $j$ depending on the connection of segments to the corner nodes (see fig.
 \ref{fig-segments2}).
@@
 
-We define a local system of coordinates $(x,y,z)$ in which the axis z is oriented along the thickness of the element (the computation of the frame of each element will be detailed in section \ref{sec-corot}).
+We define a local system of coordinates $(x,y,z)$ in which the axis z is oriented along the thickness of the element (the computation of the element frame is covered in section \ref{sec-corot}). 
 Given the initial position of the triangle points in this system of coordinate, $\{x_1^0, y_1^0\}$,  $\{x_2^0, y_2^0\}$, and  $\{x_3^0, y_3^0\}$, we can compute a linear relation between $\{\xi_1,\xi_2, \xi_3\}$ and $\{ x,y \}$:
@@
 where $\mat{I}$ is the identity matrix ($I_{kl}  = \{1$ if $l=k$ ;  $0$ if $l \neq k \} $). 
-The displacement gradient can be linked to the degrees of freedom of the shell using the expression of the displacement of a point $\delta T$ (see equation  (\ref{eq-bez-delta})) on the shell surface:
+The displacement gradient can be linked to the degrees of freedom of the shell using the expression of the displacement of a point $\delta T$ \eqref{eq-bez-delta} on the shell surface:
 \begin{equation}
@@
 \end{equation}
-where $z$ is the local coordinate along $z$-axis of the point on which the strain is measured.
+where $z$ is the local coordinate along z-axis of the point for which the strain is measured.
 
@@
 As previously described, the shell element's strain is a combination of two types of deformations, as shown in figure \ref{deformations}.
-If we assume that the initial out of plane deflections are negligible and that the frame of the shell element nodes are aligned with local $z$ axis, we can distribute these two deformations into two groups of degrees of freedom  at the nodes' level.
+%If we assume that the initial out of plane deflections are negligible and that the frame of the shell element nodes are aligned with local $z$ axis,
+We can distribute these two deformations into two groups of degrees of freedom  at the nodes' level.
 
@@
 
-In order to compute FEM-based stiffness for the shell element, we need to compute the jacobian matrix between strains and nodes' displacements.
-As we have separated the formulations for the in-plane and the out of plane displacements, it will result with the computation of two jacobian matrices (of size $3\times9$).
-\begin{equation}
-    \delta 
+In order to compute FEM-based stiffness for the shell element, we need to compute the strain-displacement matrix between strains and nodes' displacements.
+As we have separated the formulations for the in-plane and the out of plane displacements, we obtain two strain-displacement matrices (of size $3\times9$).
+\begin{equation}
+    %\delta 
     \begin{bmatrix}
@@
     \end{bmatrix}
-    = \begin{bmatrix} \mat{J}_m \end{bmatrix}
+    = \mat{J}_m 
     \begin{bmatrix}
@@
 \end{equation}
-where $\left[ \mat{J}_m \right]$ maps the membrane in-plane displacements and 
-%
-\begin{equation}
-    \delta
+where $\mat{J}_m $ maps the membrane in-plane displacements and 
+%
+\begin{equation}
+    %\delta
     \begin{bmatrix}
@@
     \end{bmatrix}
-    = \begin{bmatrix} \mat{J}_b \end{bmatrix}
+    = \mat{J}_b 
     \begin{bmatrix}
@@
 \end{equation}
-where $\left[ \mat{J}_b \right]$ maps the bending displacements.
-
-To compute the values in matrices $\left[ \mat{J}_m \right]$ and $\left[ \mat{J}_b \right]$,  we can reuse equation (\ref{eq-bez-delta}) and apply the spatial derivatives.
-To illustrate, let's take one value of the matrix $\left[ \mat{J}_m \right]$  (first line, fourth column), we first take the kinematics between $\delta T_x$ and the concerned degree of freedom $ \delta \mathbf{x}_2^x$ from equation (\ref{eq-bez-delta}):
+where $\mat{J}_b $ maps the bending displacements.
+
+To compute the values in matrices $\mat{J}_m$ and $\mat{J}_b $,  we can reuse equation \eqref{eq-bez-delta} and apply the spatial derivatives.
+To illustrate, let's take one value of the matrix $ \mat{J}_m $  (first line, fourth column), we first take the kinematics between $\delta T_x$ and the concerned degree of freedom $ \delta \mathbf{x}_2^x$ from equation \eqref{eq-bez-delta}:
 \begin{equation}
@@
     \sum_{i=1}^{3} (\deriv{n_2}{\xi_i} + \deriv{n_6}{\xi_i} + \deriv{n_7}{\xi_i} + \frac{1}{3}\deriv{n_{10}}{\xi_i})
-    \deriv{\xi_i}{x} \right) }_{\begin{bmatrix} \mat{J}_m \end{bmatrix}_{14}}  
+    \deriv{\xi_i}{x} \right) }_{{J_m}_{14}}  
  \delta \mathbf{x}_2^x 
@@
 rotation invariant \cite{Hauth2004,Muller2004} and produces ghost forces for rigid
-rotations. The alternative approach is to use non-linear Green's strain tensor but the non-linearity introduced into the system is costly and introduces instabilities.
+rotations.% The alternative approach is to use non-linear Green's strain tensor but the non-linearity introduced into the system is costly and introduces instabilities.
 To keep the system linear and to deal
@@
 and subsequently normalizing the columns. We then rotate the corotational
-frame in opposite direction. Even single step of this process improves the
-final frame but more iterations of the polar decompositions are usually
+frame in opposite direction.
+The deformation gradient depends on frame which means multiple iterations of polar decomposition can be performed.
+Even single iteration improves the
+final frame but more are usually
 necessary (up to 20) which severely degrades the performance. Polar
@@
 
-If the first condition is not met in 5 iterations it can mean one of three
-things we still have very good approximation of the ideal frame. In some
+If the first condition is not met in 5 iterations 
+we still have very good approximation of the ideal frame. In some
 situations the process is not able to reach the ideal frame quickly because
-either the convergence for the element is too slow or the process is diverging.
+either the convergence for the element is too slow or the process is oscilating around a certain configuration.
 In either case performing more iterations is unnecessary.
@@
 resolution mesh back onto the mechanical mesh.
+This high resolution mesh may be obtained simply by refining the triangles of mechanical mesh.
 
@@
 resolution mesh is updated using the assigned barycentric coordinates and
-the function of the surface \eqref{eq-cubicbez}.
+the function of the surface \eqref{eq-cubicbez}. An alternative approach is to
+subdivide the B\'ezier triangles as necessary.
 
@@
 %
-\begin{equation}
-    \dot{T} = \sum_{i=1}^{10} n_i \dot{P_i} = \sum_{i=1}^{10} n_i V_i
+\begin{equation}\label{eq-velocity}
+    \dot{T} = \sum_{i=1}^{10} n_i \dot{\mathbf{p}_i} = \sum_{i=1}^{10} n_i \mathbf{v}_i
 \end{equation}
@@
 For the corner control points  we
-already have the velocities $V_i$. For the internal control points  we need to compute
+already have the velocities $\mathbf{v}_i$. For the internal control points  we need to compute
 them. We again make use of the kinematic link between control points and control points.
@@
 %
-\begin{eqnarray}
-    V_j = V_i + \omega_i \times P_{i-j}
+\begin{eqnarray}\label{eq-velocity-pt}
+    \mathbf{v}_j = \mathbf{v}_i + \omega_i \times \mathbf{p}_{i-j}
 \end{eqnarray}
@@
 %
-\begin{eqnarray}
-    V_{10} = \frac{1}{3} \sum_{i=1}^3 \left( V_i + \omega_i \times P_{i-10} \right)
+\begin{eqnarray}\label{eq-velocity-pt2}
+    \mathbf{v}_{10} = \frac{1}{3} \sum_{i=1}^3 \left( \mathbf{v}_i + \omega_i \times \mathbf{p}_{i-10} \right)
 \end{eqnarray}
+
+Again after substitution of \eqref{eq-velocity-pt} and \eqref{eq-velocity-pt2} into
+\eqref{eq-velocity} one can evaluate $\dot{T}$ as
+%
+$%\begin{equation}
+\dot{T} = \mathbf{J}_t \left[ \mathbf{v}^x_1, \hdots, \mathbf{v}^z_3,
+        \omega^x_1, \ldots, \omega^z_3 \right]^T
+$.%\end{equation}
 
@@
 and torques on the vertices of the mechanical mesh.
-We compute the influence of force $F$ on the corners through all the
-associated control points by the formula:
+The influence of force $\mathbf{f}$ acting on the fine mesh is transfered using the
+associated barycentric coordinates through the control points by the formula:
 %
 \begin{eqnarray}
-    F_1 & = &  \sum_{i \in \{1,4,5\}} n_i F + \frac{n_{10}}{3} F \\
-    F_2 & = &  \sum_{i \in \{2,6,7\}} n_i F + \frac{n_{10}}{3} F \\
-    F_3 & = &  \sum_{i \in \{3,8,9\}} n_i F + \frac{n_{10}}{3} F 
+    \mathbf{f}_1 & = &  \sum_{i \in \{1,4,5\}} n_i \mathbf{f} + \frac{n_{10}}{3} \mathbf{f} \\
+    \mathbf{f}_2 & = &  \sum_{i \in \{2,6,7\}} n_i \mathbf{f} + \frac{n_{10}}{3} \mathbf{f} \\
+    \mathbf{f}_3 & = &  \sum_{i \in \{3,8,9\}} n_i \mathbf{f} + \frac{n_{10}}{3} \mathbf{f} 
 \end{eqnarray}
@@
     \begin{split}
-        	\tau_1 = & P_{1-4} \times (n_4 F) + P_{1-5} \times (n_5 F) + \\
-        & + \frac{1}{3} P_{1-10} \times (n_{10} F)
+        \tau_1 = & \mathbf{p}_{1-4} \times (n_4 \mathbf{f}) + \mathbf{p}_{1-5} \times (n_5 \mathbf{f}) + \\
+                 & + \frac{1}{3} \mathbf{p}_{1-10} \times (n_{10} \mathbf{f})
     \end{split}
@@
     \begin{split}
-       	\tau_2 = & P_{2-6} \times (n_6 F) + P_{2-7} \times (n_7 F) + \\
-        & + \frac{1}{3} P_{2-10} \times (n_{10} F) \\
+        \tau_2 = & \mathbf{p}_{2-6} \times (n_6 \mathbf{f}) + \mathbf{p}_{2-7} \times (n_7 \mathbf{f}) + \\
+                 & + \frac{1}{3} \mathbf{p}_{2-10} \times (n_{10} \mathbf{f}) \\
     \end{split}
@@
     \begin{split}
-        	\tau_3 = & P_{3-8} \times (n_8 F) + P_{3-9} \times (n_9 F) + \\
-        & + \frac{1}{3} P_{3-10} \times (n_{10} F) \\
+        \tau_3 = & \mathbf{p}_{3-8} \times (n_8 \mathbf{f}) + \mathbf{p}_{3-9} \times (n_9 \mathbf{f}) + \\
+                 & + \frac{1}{3} \mathbf{p}_{3-10} \times (n_{10} \mathbf{f}) \\
     \end{split}
@@
 
-Note that these formula fulfill the physical principle of virtual power: $ P = \dot{T} \cdot F  = \sum_{i=1}^3 V_i \cdot F_i +  \omega_i  \cdot \tau_i$
+Or again as
+%
+$
+\left[ \mathbf{f}^x_1, \hdots, \mathbf{f}^z_3, \tau^x_1, \hdots, \tau^z_3 \right]^T =
+\mathbf{J}^T_i \mathbf{f}
+$.
+%
+Note that these formula fulfill the physical principle of virtual power: $ P = \dot{T} \cdot f  = \sum_{i=1}^3 V_i \cdot f_i +  \omega_i  \cdot \tau_i$
 
@@
 combination of two elements that are very well established in the area of
-mechanical modelling. Our results are compared with the shell element composed
+mechanical modeling. Our results are compared with the shell element composed
 of Discrete Kirchoff Triangle (DKT) element \cite{Batoz1980} for bending plate
@@
 results for the model of O. Comas \cite{Comas2010c} are also presented. 
+Implicit Euler integration scheme was used to solve the system. This allows us
+to use large time steps needed for real-time performance of at least 25 Hz.
 Subsection \ref{sec-fps}
@@
 movement. This is the only difference from the original Scordelis-Lo roof test
-where the edges are free to move in longitudinal direction. The Figure
-\ref{fig-roof} shows the vertical displacement of midpoint on the free
+where the edges are free to move in longitudinal direction. The
+Figure~\ref{fig-roof} shows the vertical displacement of midpoint on the free
 edge.
@@
 \Figure{\linewidth}{roof}
-{Convergence for modified Scordelis-Lo roof test of DKT+ANDES element, 
+{Convergence for modified Scordelis-Lo roof test of DKT+ANDES element (DKT), 
 our B\'ezier shell element (BSH) and element of O. Comas (PSH) with polynomial shape function.}
@@
 constrained two opposite nodes at the top of the hemisphere. The radial
-displacement for one point with applied load is shown in Figure
-\ref{fig-hemisphere}.
+displacement for one point with applied load is shown in Figure~\ref{fig-hemisphere}.
 
 \Figure{\linewidth}{hemisphere}
-{Convergence for test on hemisphere with 18° hole for the DKT+ANDES element and
+{Convergence for test on hemisphere with 18° hole for the DKT+ANDES element (DKT) and
 our B\'ezier shell element (BSH) and element of O. Comas (PSH) with polynomial shape function.}
@@
 and only single CPU core was used for the
-simulation. The values reported are for the raw physical simulation with (b)
-and without (a) the frame fixing method described in subsection
+simulation. The values reported are for the raw physical simulation with (B)
+and without (A) the frame fixing method described in subsection
 \ref{sec-corot}. We also present results including the mapping of
@@
 results for more than 600 elements mesh with 10k triangles.
+
+Both DKT+ANDES and our element are purely linear which makes their raw
+simulation time comparable. The hybrid PSH element require evaluation of
+coefficients of the polynomial during simulation which makes it inferior in
+performance.
 
@@
 We have
-developed a prototype \cite{Anonymous2012} of new surgical simulation system
+developed a prototype \cite{Kislinskiy2012} of new surgical simulation system
 for preoperative planning. The element has been used to model deformations of
@@
 The purpose of this simulator is to reproduce the tissue deformations that are encountered during a delivery.
-The goal is to evaluate the mechanical stresses which are applied on the anatomical structures (like the uterus but also like the ligaments that sustain the uterus).
+The goal is to evaluate the mechanical stresses which are applied on the anatomical structures (like the uterus but also the ligaments that sustain the uterus).
 
@@
 Currently, due to the technical decisions the element has two limitations.
-It is $C^1$ continuous at corners but only $C^0$ on the
-edges. This may prove to be an issue for example if curvature needs to be
-computed not only inside the element but also across the edges.
+It is $C^1$ continuous at corners and only $C^0$ on the
+edges. The lack of continuity may prove to be an issue for example if curvature needs to be
+computed not only inside the element but also across the corners/edges.
+In location with high initial curvature more elements may be necessary to properly model the surface.
 
@@
 Compared to  the elements commonly used in mechanics, the interpolation is less complex and ensures at least a $C^0$ continuity in the worst case. 
-(Many shell elements are not even $C^0$ when the initial shape is not flat).
+(Many shell elements are not even $C^0$ when the shape is not flat).
 
@@
 
+\renewcommand{\bibfont}{\footnotesize}
 %-------------------------------------------------------------------------

### 2013-02-18 Doc/Miccai2013/miccai2013.tex
adding first lines for MICCAI 2013 paper
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@508 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

% http://www.miccai2013.org/submission_guideline.html
\documentclass{llncs}

\usepackage{makeidx}  % allows for indexgeneration

\begin{document}
%
%\frontmatter          % for the preliminaries
%
\mainmatter              % start of the contributions
%
\title{TODO: Title} % TODO
%
\titlerunning{TODO: Title}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Anonymous}
%\author{%
%Tom\'a\v{s} Golembiovsk\'y\inst{1,2} \and%
%Igor Peterl\'ik\inst{3} \and%
%%Christian Duriez\inst{2} \and%
%Stephan\'e Cotin\inst{2,3}%
%}
%
%\authorrunning{Tom\'a\vs Golembiovsk\'y et al.} % abbreviated author list (for running head)
%
%\institute{%
%Faculty of Informatics, Masaryk University, Brno, Czech Republic\and%
%INRIA Lille -- Nord, France\and%
%IHU, Strasbourg, France%
%}
%
\maketitle

% TODO
\begin{abstract}
TODO: Abstract
\keywords{liver, Glisson's capsule, deformation modeling, membrane}
\end{abstract}

%\section{Introduction}
\section{...}

TODO

The Glisson's capsule \cite{Umale2013} has in order of three higher
stiffness than the parenchyma. As such it plays an important role in the
deformation of the liver \cite{Ahn2010,Hollenstein2006}. Evaluation of the
liver as a homogeneous material can lead to overestimation of the
mechanical properties up to the factor of 3 \cite{Hollenstein2006}.

% -- TODO: properties of the capsule

We propose to further extend the tetrahedral based model of the parenchyma
with vascularized model \cite{Peterlik2012} by the layer of membrane elements.
(Modelling the capsule with tetrahedral elements is unfeasible -- many
elements or largely deformed elements.)

Based on the small thickness it seems reasonable to assume that no bending
strains limmit the deformation of the capsule. Therefore modelling the
capsule with membrane element only should be adequate.

% -- TODO: what is membrane element
% --

\section{Experiments/Evaluation}

\section{Results}

\section{Conclusion}

%
% ---- Bibliography ----
%

\bibliographystyle{splncs03}
\bibliography{bibdata}

\end{document}
% vim:set et sw=2 tw=75 fdm=marker fdl=1 fdc=4 isk+=_,-:

### 2013-02-21 Doc/Miccai2013/miccai2013.tex
more on the miccai paper
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@513 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \usepackage{makeidx}  % allows for indexgeneration
+
+%%% Comments
+\usepackage{color}
+\newcommand{\TG}[1]{{\color{blue}\textbf{TG: #1}}}
+%%%
+
+\newcommand{\Vec}[1]{\mathbf{#1}}
+\newcommand{\Mat}[1]{\mathbf{#1}}
 
@@
 
-The Glisson's capsule \cite{Umale2013} has in order of three higher
+% -- TODO why model liver? what's interesting about it?
+% -- liver covered by capsule, thin membrane
+
+the usual approach is to model the organ as homogeneous material,
+as such complex non-linear material descriptions are inevitably necessary
+to describe the properties of non-homogeneous tissue.
+
+modeling non-homogenous volumetric object by a dense mesh is complicated by
+large number of elements. we propose a composite model and show that it
+leads to adequate results.
+
+The Glisson's capsule has in order of three higher
 stiffness than the parenchyma. As such it plays an important role in the
@@
 
-% -- TODO: what is membrane element
-% --
+GOAL: show that we are able to simulate the behaviour of non-homogeneous
+tissue covered by capsule (liver/kidney) by combined model.
+
+GOAL2: show that combination of linear tetrahedra and linear membrane is
+easy.
+
+% NOTE: the assumption is that we cannot estimate elastic modulus as a
+% combination of the two moduli (for parenchyma and capsule) because it will
+% work only for certain displacement. Is this valid?
+% -- based on the text of Hollenstein2006 it seems like it
+
+\section{Model}
+
+\subsection{parenchyma} %{{{
+
+\TG{TODO: write something like this} % (src: Peterlik2012)
+\begin{quote}
+While most studies concerning the material properties of liver parenchyma
+agree on a viscoelastic behaviour (see [5] or [7] for instance), we employ
+a simpler non-linear elastic model. This is essentially motivated by the
+fact that we are not focusing on the transient part of the deformation but
+rather the static equilibrium under some specific loading conditions.
+\end{quote}
+
+We model the parenchyma using corotational finite elmets. The corotational
+formulation allows us to deal with geometrical non-linearities but still
+rely on linear stress-strain relationship. Thus rotations and large
+deformations are possible. In the corotational formulation the deformation
+of each element is expressed in the local frame of reference. There are
+several ways of computing the corotational frame for elements, we rely on
+the geometrical method proposed in \cite{Nesme2005}.
+% NOTE: This is very simplified and could be extended.
+The stiffness matrix $\Mat{K}$ can be expressed in terms of the deformation
+$u$ and the force as $\Vec{f} = \Mat{K}(\Vec{u}) \Vec{u}$. The resulting
+system of linear equations is then solved by direct LDL solver.
+
+%}}}
+
+\subsection{Capsule} %{{{
+
+The thickness of the Glisson's capsule is relatively small (Umale et al.
+\cite{Umale2011} reported for porcine liver values in range of 10-20
+$\mu$m). This makes it very difficult to model with volumetric mesh. Such
+mesh would inevitably contain large number of elements or very irregular
+elements. Instead modeling the capsule with two-dimensional elements seems
+as a natural choice. Mechanics provides a FE tool in the form of membranes
+and shells. Based on the small thickness (compared to the area) we also
+assume negligible bending forces and propose a model based on membrane
+elements.
+
+To maintain simplicity of the composite model we choos simple triangular
+elements with constant strain.
+
+\TG{add a ref. or dive into the definition?}
+
+We again use linear elastic model and employ the  corotational formulation
+for the triangular elements.
+
+%}}}
+
+\subsection{Coupling Between Capsule and Parenchyma} %{{{
+
+High cohesion between capsule and parenchyma was reported. Based on this
+property we may assume no relative motion of the capsule to the parenchyma.
+
+While in general we could use arbitrary surface mesh for a capsule we use
+the fact that the parenchyma is modelled by tetrahedral elements which have
+triangular faces. Because the boundary of the volumetric mesh is already
+triangulated, we only extract the boundary triangles which we use for the
+model of capsule.
+
+Using directly the boundary of the tetrahedral mesh does not only solve the
+problem of obtaining the triangular mesh, but has one more advantage. Nodes
+of the triangular mesh overlap with nodes of the tetrahedral mesh and we do
+not have to project nodes of tiangular mesh onto the tetrahedra. More over,
+the stiffness matrices for capsule and parenchyma are easily assembled
+together and solved as one system.
+
+\TG{Is assembling process too obvious?}
+
+Withouth the loss of generality we can assume the tetrahedron consists of
+nodes $p_1, p_2, p_3$ and $p_4$, the boundary triangle has nodes $p_1, p_2$
+and $p_3$. We can reorder the degrees of freedom so that the stiffness
+matrix $\Mat{K}^t$ for the tetrahedron can be written as:
+
+\begin{equation}
+  \Mat{K}^t = \left[\begin{array}{c|c}
+      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
+      \hline
+      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
+  \end{array}\right]
+\end{equation}
+
+Then the assembled stiffness matrix for the element is:
+
+\begin{equation}
+  \Mat{K} = \left[\begin{array}{c|c}
+      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
+      \hline
+      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
+  \end{array}\right]
+  +
+  \left[\begin{array}{c|c}
+      \Mat{K}^m & 0 \\
+      \hline
+      0 & 0 \\
+  \end{array}\right]
+\end{equation}
+
+Where $\Mat{K}^m$ is the stiffness matrix of the triangular membrane.
+
+%}}}
+
+\subsection{Blood Vessels} %{{{
+
+We also combined our composite model with vascularization method proposed
+by Peterlik et al. \cite{Peterlik2012}. The method is based on linear beams
+with local frames of reference (in a many aspects similar to the
+corotational formulation). As such the model also handles geometric
+non-linearities in the deformation. The proposed model assumes no relative
+motion of vessels to the parenchyma. Segmented vascular network is
+discretized into the set of linear elements that are then projected into
+the tetrahedral mesh of the parenchyma. During the simulation forces and
+torques resulting from the deformations of vessels by parenchyma are
+computed and propagated back onto the tetrahedral mesh.
+
+%}}}
 
 \section{Experiments/Evaluation}
+
+setup:
+
+Test 1) Global deformation of liver under gravity
+  * show that there is a difference between simple homogeneous material
+    and material with capsule
+  * include blood vessels
+
+Test 2) Indentation test
+  * perform indentation and analyze the profile
+  * show how stiffer membrane affect the result
 
@@
 \end{document}
-% vim:set et sw=2 tw=75 fdm=marker fdl=1 fdc=4 isk+=_,-:
+% vim:set et sw=2 tw=75 fdm=marker fdl=3 fdc=4 isk+=_,-:

### 2013-02-26 Doc/Miccai2013/miccai2013.tex
abstract v1.0
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@516 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{abstract}
-TODO: Abstract
+This work provides a complete physical model for the deformations of the liver with its three constituents: parenchyma, vessels and the Glisson's capsule.
+Some recent experimental results \cite{Ahn2010} have shown that there is a significant difference of stiffness between the Glisson's capsule and the parenchyma. 
+Moreover, it has been observed that the capsule plays an important role for the deformation at the local level \cite{Hollenstein2006}.
+However, the Glisson's capsule is very thin and its influence for the global deformations of the liver was never highlighted.
+In this work, we propose to use a membrane FEM model for the capsule, that is coupled with a vascularized model of the liver.
+We show that the approach is able to reproduce the deformations observed at the local level.
+Additionally, we show, by simulating natural global deformations of the liver observed on real images (due to gravity), that the results have important differences with and without the capsule.
+
 \keywords{liver, Glisson's capsule, deformation modeling, membrane}

### 2013-02-27 Doc/Miccai2013/miccai2013.tex
more on the miccai paper
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@517 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\usepackage{makeidx}  % allows for indexgeneration
+\usepackage{makeidx}
+\usepackage[pdftex]{graphicx}
 
@@
 
+%\Figure{width}{name}{caption}
+\newcommand{\Figure}[3]{%
+\begin{figure}[htb]
+  \centering
+  \includegraphics[width=#1]{#2}
+  \caption{\label{fig-#2}#3}
+\end{figure}}
+
 \begin{document}
@@
 %
-\title{TODO: Title} % TODO
+\title{Complete Liver Model for Real-time Simulation} % TODO
+% Or: Modelling the Glisson's Gapsule in Real-time Simulation of Liver
+% or anything else ...
 %
@@
 
-% TODO
 \begin{abstract}
@@
 with vascularized model \cite{Peterlik2012} by the layer of membrane elements.
-(Modelling the capsule with tetrahedral elements is unfeasible -- many
+(Modeling the capsule with tetrahedral elements is unfeasible -- many
 elements or largely deformed elements.)
@@
 Based on the small thickness it seems reasonable to assume that no bending
-strains limmit the deformation of the capsule. Therefore modelling the
+strains limmit the deformation of the capsule. Therefore modeling the
 capsule with membrane element only should be adequate.
@@
 
+% -----
+
+The liver is usually modeled as a homogeneous isotropic (visco-)elastic
+material \TG{need refs.?}. It is however non-homogeneous and modeling it as
+such necessarily requires complex non-linear material formulation. Peterlik
+et al. \cite{Peterlik2012} showed the importance of the vascular structures
+on the liver deformation.
+
+Another overlooked component is the Glisson's capsule. Despite it's very
+small thickness it has high elastic modulus that is in three orders of
+magnitude larger than that of the parenchyma. As such it plays an important
+role in both global and local deformations of the liver.
+
+It is not possible to model such thin structure with classical tetrahedral
+elements.
+
+% -----
+
 % NOTE: the assumption is that we cannot estimate elastic modulus as a
@@
 
-\subsection{parenchyma} %{{{
+\subsection{Parenchyma} %{{{
 
-\TG{TODO: write something like this} % (src: Peterlik2012)
-\begin{quote}
-While most studies concerning the material properties of liver parenchyma
-agree on a viscoelastic behaviour (see [5] or [7] for instance), we employ
-a simpler non-linear elastic model. This is essentially motivated by the
-fact that we are not focusing on the transient part of the deformation but
-rather the static equilibrium under some specific loading conditions.
-\end{quote}
-
-We model the parenchyma using corotational finite elmets. The corotational
+It is known that the parenchyma exhibits viscoealstic behaviour \TG{Cite
+something}, we however employ simpler elastic model as we are not so much
+interested in time-dependent behaviour but rather in static equilibrium
+under certain conditions.
+%
+The parenchyma is modeled using corotational finite elmets. The corotational
 formulation allows us to deal with geometrical non-linearities but still
@@
 the geometrical method proposed in \cite{Nesme2005}.
-% NOTE: This is very simplified and could be extended.
+% NOTE: This description of corotational method is very simplified and could be extended.
 The stiffness matrix $\Mat{K}$ can be expressed in terms of the deformation
@@
 While in general we could use arbitrary surface mesh for a capsule we use
-the fact that the parenchyma is modelled by tetrahedral elements which have
+the fact that the parenchyma is modeled by tetrahedral elements which have
 triangular faces. Because the boundary of the volumetric mesh is already
@@
 
-\section{Experiments/Evaluation}
+\section{Experiments} %{{{
 
-setup:
+To evaluate the model we have implemented in the
+SOFA\footnote{www.sofa-framework.org} framework and performed a set of
+numerical simulations. We have tests the influence of the capsule for both
+local and global deformations. In the simulations material properties for a
+porcine liver were used as reported by Umale \cite{Umale2013}. The
+parenchyma was modeled with elastic modulus 1980~Pa and the capsule has
+elastic modulus 8~MPa and thickness 20~$\mu$m.
+\TG{The vessels have same parameters as Igor used in his work. Is this OK
+or should we use values reported by Sagar for consistency?}
+
+\subsection{Local deformations}
+
+Hollenstain et al. \cite{Hollenstein2006} measured the response of the
+bovine liver to the negative pressure using their aspiration test. One of
+the observations was that the Glissons capsule has non-negligible influence
+and modeling only the parenchyma leads to overestimation of the
+deformation. We reproduced the test with our model and acquired similar
+results.
+
+The aspiration device consists of a tube with 1 cm in diameter and is able
+to control internal pressure in the tube. The test is performed by
+attaching the tube to the tissue and meassuring the tissue response. We
+have set up a simulation in the spirit of this test. 
+% TODO: fix the scene! tube has 5 mm radius, not 1 cm
+The simulation is performed with a 3x3x3 cm cube of liver tissue and tube
+with 1 cm radius. The pressure of 2 kPa is simulated inside the tube.
+
+Figure \ref{fig-aspiration} shows the profiles of cuts in the middle of the
+cube for the model with and without capsule. It can be seen that the
+deformation for the model without capsule is much larger which is in
+aggreement with the data obtained by Hollenstain et al.
+
+\Figure{4in}{aspiration}{TODO: aspiration test; fix figure, add screenshot}
+
+\subsection{Global Deformations}
 
@@
 
-Test 2) Indentation test
-  * perform indentation and analyze the profile
-  * show how stiffer membrane affect the result
-
-\section{Results}
+%}}}
 
 \section{Conclusion}
+
+- we have proposed a complete liver model that employs membrane element to model
+the capsule and extends work by Peterlik et al.
+
+- we have reproduced aspiration test, results are in aggreement with Hollenstain
+  et al.
+
+- we have shown influence on global deformations
 

### 2013-02-27 Doc/Miccai2013/aspiration.R

@@ -1,0 +1,21 @@
+a1 <- data.matrix(read.csv("aspiration-simple.csv"))
+a2 <- data.matrix(read.csv("aspiration-composite.csv"))
+
+a1 <- a1[1:21,c(4,3)]; a1[,2] <- a1[,2] - 0.03
+a2 <- a2[1:21,c(4,3)]; a2[,2] <- a2[,2] - 0.03
+
+pdf("aspiration.pdf", width=4, height=2, pointsize=9); # sizes in inches
+p <- par(lwd=2, mar=c(4,4,1,1)+0.1);
+
+plot(range(c(a1[,1], a2[,1])), range(c(a1[,2], a2[,2])), #ylim=c(0, 0.004), 
+     type="n", xlab="x-coordinate", ylab="Displacement")
+
+lines(a1, col="black")
+lines(a2, col="blue")
+
+abline(h=0, lty="dotted");
+
+#legend("topright", c("A","B", "C", "D"), pch=c(1,4,5,6), col="black");
+
+dev.off()
+par(p);

### 2013-02-27 Doc/Miccai2013/miccai2013.tex
the paper
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@518 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 %Igor Peterl\'ik\inst{3} \and%
-%%Christian Duriez\inst{2} \and%
+%Christian Duriez\inst{2} \and%
 %Stephan\'e Cotin\inst{2,3}%
@@
 
-%\section{Introduction}
-\section{...}
+\section{Introduction} %{{{
 
-TODO
+Medical simulation is expected to play an integral part in many aspects of
+medical practice. From diagnosis, through planning or training to computer
+assisted intervention. A good medical model has to be simple yet efficient
+to be really usable.
+The liver is usually modeled as a homogeneous isotropic (visco-)elastic
+material \TG{refs}.% It is however non-homogeneous and modeling it as
+%homogeneous necessarily requires complex non-linear material descriptions.
+Recently
+Peterl\'{i}k et al. \cite{Peterlik2012} showed the importance of the vascular
+structures on the liver deformation.
 
-% -- TODO why model liver? what's interesting about it?
-% -- liver covered by capsule, thin membrane
+Another often overlooked component is the Glisson's capsule.
+Biomechanical measurements report that the capsule has very small
+thickness (approx. 20 $\mu$m for porcine liver \cite{Umale2011}).
+On the other hand it has in three orders of magnitude higher stiffness than
+the parenchyma.
+Because of this it plays an important role in local
+deformations of the liver \cite{Ahn2010,Hollenstein2006}.
+It was reported that evaluation of the liver as a homogeneous material
+without capsule can lead to overestimation of local deformations up to the factor of 3
+\cite{Hollenstein2006}.
 
-the usual approach is to model the organ as homogeneous material,
-as such complex non-linear material descriptions are inevitably necessary
-to describe the properties of non-homogeneous tissue.
+While the influence on local deformation was reported, no experimental
+results show whether the capsule also plays an important role on the global
+deformations of the liver. We propose a composite model that models the
+capsule with membrane FEM elements. We show that the model is able to
+simulate expected local behaviour reported in literature. Further more we
+show that the capsule plays also an important role in the deformation on
+global scale.
 
-modeling non-homogenous volumetric object by a dense mesh is complicated by
-large number of elements. we propose a composite model and show that it
-leads to adequate results.
+\TG{summary of following sections.}
 
-The Glisson's capsule has in order of three higher
-stiffness than the parenchyma. As such it plays an important role in the
-deformation of the liver \cite{Ahn2010,Hollenstein2006}. Evaluation of the
-liver as a homogeneous material can lead to overestimation of the
-mechanical properties up to the factor of 3 \cite{Hollenstein2006}.
+%}}}
 
-% -- TODO: properties of the capsule
+\section{Model} %{{{
 
-We propose to further extend the tetrahedral based model of the parenchyma
-with vascularized model \cite{Peterlik2012} by the layer of membrane elements.
-(Modeling the capsule with tetrahedral elements is unfeasible -- many
-elements or largely deformed elements.)
-
-Based on the small thickness it seems reasonable to assume that no bending
-strains limmit the deformation of the capsule. Therefore modeling the
-capsule with membrane element only should be adequate.
-
-GOAL: show that we are able to simulate the behaviour of non-homogeneous
-tissue covered by capsule (liver/kidney) by combined model.
-
-GOAL2: show that combination of linear tetrahedra and linear membrane is
-easy.
-
-% -----
-
-The liver is usually modeled as a homogeneous isotropic (visco-)elastic
-material \TG{need refs.?}. It is however non-homogeneous and modeling it as
-such necessarily requires complex non-linear material formulation. Peterlik
-et al. \cite{Peterlik2012} showed the importance of the vascular structures
-on the liver deformation.
-
-Another overlooked component is the Glisson's capsule. Despite it's very
-small thickness it has high elastic modulus that is in three orders of
-magnitude larger than that of the parenchyma. As such it plays an important
-role in both global and local deformations of the liver.
-
-It is not possible to model such thin structure with classical tetrahedral
-elements.
-
-% -----
-
-% NOTE: the assumption is that we cannot estimate elastic modulus as a
-% combination of the two moduli (for parenchyma and capsule) because it will
-% work only for certain displacement. Is this valid?
-% -- based on the text of Hollenstein2006 it seems like it
-
-\section{Model}
+In this section we will describe the construction of our composite model.
+It's three main parts are: tetrahedral model for parenchyma, membrane model
+for capsule coupled to the parenchyma and beam model of vascular system as
+proposed by Peterl\'{i}k et al. \cite{Peterlik2012}.
 
@@
 
-It is known that the parenchyma exhibits viscoealstic behaviour \TG{Cite
+It is known that the parenchyma exhibits viscoelastic behaviour \TG{Cite
 something}, we however employ simpler elastic model as we are not so much
@@
 %
-The parenchyma is modeled using corotational finite elmets. The corotational
+The parenchyma is modeled using corotational finite elements. The corotational
 formulation allows us to deal with geometrical non-linearities but still
@@
 The stiffness matrix $\Mat{K}$ can be expressed in terms of the deformation
-$u$ and the force as $\Vec{f} = \Mat{K}(\Vec{u}) \Vec{u}$. The resulting
-system of linear equations is then solved by direct LDL solver.
+$u$ and the force as $\Vec{f} = \Mat{K}(\Vec{u}) \Vec{u}$. Where the
+components of $\Mat{K}$ depend on the orientation of each element in the
+current simulation step.
 
@@
 
-\subsection{Capsule} %{{{
+\subsection{Glisson's Capsule} %{{{
 
-The thickness of the Glisson's capsule is relatively small (Umale et al.
+The thickness of the Glisson's capsule is relatively small. Umale et al.
 \cite{Umale2011} reported for porcine liver values in range of 10-20
-$\mu$m). This makes it very difficult to model with volumetric mesh. Such
-mesh would inevitably contain large number of elements or very irregular
-elements. Instead modeling the capsule with two-dimensional elements seems
-as a natural choice. Mechanics provides a FE tool in the form of membranes
-and shells. Based on the small thickness (compared to the area) we also
+$\mu$m.
+It is not possible to model such thin structure with classical tetrahedral
+elements. Modeling this non-homogeneous volumetric object would require a
+dense mesh to avoid numerical instabilities and would significantly
+violate the speed requirements imposed on medical simulators.
+Instead modeling the capsule with two-dimensional elements that abstract from the
+thickness in the third dimension seems
+as a natural choice. Mechanics provides this a FE tool in the form of membranes
+and shells. Based on the small thickness (compared to the surface area) we also
 assume negligible bending forces and propose a model based on membrane
 elements. 
-
-To maintain simplicity of the composite model we choos simple triangular
+To maintain simplicity of the composite model we choose simple triangular
 elements with constant strain.
 
-\TG{add a ref. or dive into the definition?}
+\TG{add a ref. or dive into the definition? \cite{Felippa2003}}
 
-We again use linear elastic model and employ the  corotational formulation
+We again use linear elastic material and employ the corotational formulation
 for the triangular elements.
@@
 
-High cohesion between capsule and parenchyma was reported. Based on this
+The literature reports high cohesion between capsule and parenchyma.
+Based on this
 property we may assume no relative motion of the capsule to the parenchyma.
-
 While in general we could use arbitrary surface mesh for a capsule we use
@@
 Using directly the boundary of the tetrahedral mesh does not only solve the
-problem of obtaining the triangular mesh, but has one more advantage. Nodes
+problem of obtaining the surface mesh, but has one more advantage. Nodes
 of the triangular mesh overlap with nodes of the tetrahedral mesh and we do
-not have to project nodes of tiangular mesh onto the tetrahedra. More over,
+not have to project nodes of triangular mesh onto the tetrahedra. More over,
 the stiffness matrices for capsule and parenchyma are easily assembled
@@
 
-\TG{Is assembling process too obvious?}
-
-Withouth the loss of generality we can assume the tetrahedron consists of
-nodes $p_1, p_2, p_3$ and $p_4$, the boundary triangle has nodes $p_1, p_2$
+Without the loss of generality we can assume the tetrahedron consists of
+nodes $p_1, p_2, p_3$ and $p_4$ and the boundary triangle has nodes $p_1, p_2$
 and $p_3$. We can reorder the degrees of freedom so that the stiffness
@@
 Where $\Mat{K}^m$ is the stiffness matrix of the triangular membrane.
+The resulting
+system of linear equations is then solved by direct LDL solver.
 
@@
 We also combined our composite model with vascularization method proposed
-by Peterlik et al. \cite{Peterlik2012}. The method is based on linear beams
+by Peterl\'{i}k et al. \cite{Peterlik2012}. The method is based on linear beams
 with local frames of reference (in a many aspects similar to the
@@
 
+%}}}
+
 \section{Experiments} %{{{
@@
 SOFA\footnote{www.sofa-framework.org} framework and performed a set of
-numerical simulations. We have tests the influence of the capsule for both
-local and global deformations. In the simulations material properties for a
-porcine liver were used as reported by Umale \cite{Umale2013}. The
+numerical simulations. First local deformations were compared to the
+results reported in literature to validate the method. Then the model of
+the complete liver was subjected to natural global deformations to asses
+the importance of the capsule on the deformation.
+
+Material properties used in the simulations were for a porcine liver as
+reported by Umale \cite{Umale2013}. The
 parenchyma was modeled with elastic modulus 1980~Pa and the capsule has
 elastic modulus 8~MPa and thickness 20~$\mu$m.
-\TG{The vessels have same parameters as Igor used in his work. Is this OK
-or should we use values reported by Sagar for consistency?}
+\TG{values for blood vessels}
 
-\subsection{Local deformations}
+\subsection{Local deformations} %{{{
 
-Hollenstain et al. \cite{Hollenstein2006} measured the response of the
+Hollenstein et al. \cite{Hollenstein2006} measured the response of the
 bovine liver to the negative pressure using their aspiration test. One of
@@
 deformation for the model without capsule is much larger which is in
-aggreement with the data obtained by Hollenstain et al.
+aggreement with the data obtained by Hollenstein et al.
 
@@
 
+%}}}
+
 \subsection{Global Deformations}
+
+\TG{!!!}
 
@@
 
-\section{Conclusion}
+\section{Conclusion} %{{{
 
-- we have proposed a complete liver model that employs membrane element to model
-the capsule and extends work by Peterlik et al.
+A composite model for simulation of complete liver is presented in the
+paper. The work of Peterlik et al. for vaskularized liver model was
+extended by addig the capsule modeled with membrane FEM model mechanically
+coupled with the parenchyma. To validate the model a simulation of local
+deformations was performed and was found in an aggrement with the results
+measured and published by Hollenstein et al.
 
-- we have reproduced aspiration test, results are in aggreement with Hollenstain
-  et al.
+While the influence of the capsule on local deformations was previously
+studied in the literature, it's significance on the global scale
+deformations of the liver remained unexplored. We have performed a set of
+natural global deformations of the complete liver and shown that the
+capsule, despite its small thickness, plays a significant role also in
+global context.
 
-- we have shown influence on global deformations
+%}}}
 

### 2013-02-28 Doc/Miccai2013/miccai2013.tex
few comments
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@522 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \newcommand{\TG}[1]{{\color{blue}\textbf{TG: #1}}}
+\newcommand{\CD}[1]{{\color{green}\textbf{CD: #1}}}
 %%%
@@
 However, the Glisson's capsule is very thin and its influence for the global deformations of the liver was never highlighted.
-In this work, we propose to use a membrane FEM model for the capsule, that is coupled with a vascularized model of the liver.
+In this work, we propose to use a membrane FEM model for the capsule, that is coupled with a vascularized model of the liver (\CD{based on a coupling ?})
 We show that the approach is able to reproduce the deformations observed at the local level.
@@
 to be really usable.
+\CD{develop the applications for which it has to be real-time...}
+
+\CD{First... explain the biomechanics of the liver (for instance: 
+\textit{
+The liver is a complex organ composed by three main structures: the parenchyma (soft tissue, that have a hyper-visco-elastic behavior) is coupled with a dense vascular network (tree-like tubular structure with a stiffer behavior) and is surrounded by the Glisson capsule  (a stiff collagenous membrane).
+It illustrates the need of relying on both volume, surface and line finite elements and the need of proposing coupling methods between them.}
+ }
+
+\CD{Yet, many work on liver modeling only consider the parenchyma...}
 The liver is usually modeled as a homogeneous isotropic (visco-)elastic
@@
 
+\CD{This paragraph may change if we use an other model... but maybe we should directly say that we rely on vascularized model ?}
 It is known that the parenchyma exhibits viscoelastic behaviour \TG{Cite
@@
 
-\TG{add a ref. or dive into the definition? \cite{Felippa2003}}
+\TG{add a ref. or dive into the definition? \cite{Felippa2003}} \CD{both...}
 
@@
 
+\CD{The following may appear as "trivial"... maybe we can remove this part at the end if we need space} 
 Without the loss of generality we can assume the tetrahedron consists of
@@
 
+\CD{Just a proposal: put this subsection before Glisson Capsule}
 We also combined our composite model with vascularization method proposed
@@
 
+\CD{If the experiments demonstrate that the stiffness of the material is different, the thickness is very very small...
+so we have to evaluate if it is really necessary to include the model of the capsule for both local and global deformations.}
+
 \subsection{Local deformations} %{{{
+
+\CD{Maybe explain why we call that "local deformations" and in which case it is important to capture such local deformations (i.e. insertion of needle, contacts with a sharp instrument...).}
 
@@
 \subsection{Global Deformations}
+\CD{ Experiments demonstrate that capsule plays a role at local level, but still, maybe not so important to model the capsule for more global deformation...
++ What are the applications for modeling the global deformations of the liver (training and planning simulation, physics-based registration, augmented reality,)}
 

### 2013-02-28 forcefield/CstFEMForceField.h
ADD: the effect of the force field applied gradually
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@523 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@ -170,6 +170,11 @@ public:
         Data<Real> f_young;
         Data<Real> f_thickness;
         Data<bool> f_corotated;
+        Data<int> f_startDTAppl;
+        Data<int> f_numDTAppl;
+
+        Real applyFactor;
+        int actualStep;
         //Data<sofa::helper::OptionsGroup> f_measure;
         //Data< helper::vector<Real> > f_measuredValues;
 
@@ -198,6 +203,8 @@ protected :
         virtual void applyStiffness(VecDeriv& f, const VecDeriv& dx, const Index elementIndex, const double kFactor);
 
         void convertStiffnessMatrixToGlobalSpace(StiffnessMatrixFull &K_gs, const TriangleInformation &tinfo);
+
+        void handleEvent(core::objectmodel::Event *event);
 };
 
 

### 2013-02-28 Doc/Miccai2013/miccai2013.tex
paper
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@524 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 %
-\title{Complete Liver Model for Real-time Simulation} % TODO
-% Or: Modelling the Glisson's Gapsule in Real-time Simulation of Liver
-% or anything else ...
+\title{Complete Real-Time Liver Model including Glisson Capsule, Vascular Network and Parenchyma}
 %
@@
 
-\keywords{liver, Glisson's capsule, deformation modeling, membrane}
+\keywords{liver, Glisson's capsule, vascular network, deformation modeling, membrane}
 \end{abstract}
@@
 
-\CD{This paragraph may change if we use an other model... but maybe we should directly say that we rely on vascularized model ?}
 It is known that the parenchyma exhibits viscoelastic behaviour \TG{Cite
@@
 under certain conditions.
-%
+We also rely on the vascularized model of the parenchyma proposed by
+Peterl\'{i}k et al. \cite{Peterlik2012}.
+
 The parenchyma is modeled using corotational finite elements. The corotational
@@
 current simulation step.
+
+The vascularization method is based on linear beams
+with local frames of reference (in a many aspects similar to the
+corotational formulation). As such the model also handles geometric
+non-linearities in the deformation. The proposed model assumes no relative
+motion of vessels to the parenchyma. Segmented vascular network is
+discretized into the set of linear elements that are then projected into
+the tetrahedral mesh of the parenchyma. During the simulation forces and
+torques resulting from the deformations of vessels by parenchyma are
+computed and propagated back onto the tetrahedral mesh.
 
@@
 
-\subsection{Blood Vessels} %{{{
-
-\CD{Just a proposal: put this subsection before Glisson Capsule}
-We also combined our composite model with vascularization method proposed
-by Peterl\'{i}k et al. \cite{Peterlik2012}. The method is based on linear beams
-with local frames of reference (in a many aspects similar to the
-corotational formulation). As such the model also handles geometric
-non-linearities in the deformation. The proposed model assumes no relative
-motion of vessels to the parenchyma. Segmented vascular network is
-discretized into the set of linear elements that are then projected into
-the tetrahedral mesh of the parenchyma. During the simulation forces and
-torques resulting from the deformations of vessels by parenchyma are
-computed and propagated back onto the tetrahedral mesh.
-
-%}}}
-
 %}}}
@@
 so we have to evaluate if it is really necessary to include the model of the capsule for both local and global deformations.}
+\TG{I didn't undersand that. What do you mean?}
 
-\subsection{Local deformations} %{{{
+\subsection{Local Deformations} %{{{
 
-\CD{Maybe explain why we call that "local deformations" and in which case it is important to capture such local deformations (i.e. insertion of needle, contacts with a sharp instrument...).}
+During the contact with an instrument (probe, needle insertion, sharp
+instruments, \ldots) a specific deformation occurs in the near vicinity of
+the instrument. This type of deformation may not necessarily induce the
+deformation of the object as a whole and therefore can be considered as
+local. Correct material properties are not only important for amount of the
+displacement but play also a crucial role in capturing the correct area of
+the deformation or its profile near the instrument.
+\TG{Does it make sense this way? I can add an image. :)}
 
-Hollenstein et al. \cite{Hollenstein2006} measured the response of the
-bovine liver to the negative pressure using their aspiration test. One of
-the observations was that the Glissons capsule has non-negligible influence
-and modeling only the parenchyma leads to overestimation of the
-deformation. We reproduced the test with our model and acquired similar
-results.
+One example of such local deformation occurs during the aspiration test
+when the response of liver to the negative pressure is measured.
+Hollenstein et al. \cite{Hollenstein2006} observed that the Glisson's
+capsule has non-negligible influence on response of the tissue and modeling
+only the parenchyma leads to overestimation of the deformation. We
+reproduced the test with our model and acquired similar results.
 
@@
 to control internal pressure in the tube. The test is performed by
-attaching the tube to the tissue and meassuring the tissue response. We
+attaching the tube to the tissue and measuring the tissue response. We
 have set up a simulation in the spirit of this test. 
-% TODO: fix the scene! tube has 5 mm radius, not 1 cm
-The simulation is performed with a 3x3x3 cm cube of liver tissue and tube
-with 1 cm radius. The pressure of 2 kPa is simulated inside the tube.
+The simulation is performed with a 15x15x15 mm cube of liver tissue and tube
+with 1 cm in diameter. The pressure of 3 kPa is simulated inside the tube.
 
 Figure \ref{fig-aspiration} shows the profiles of cuts in the middle of the
-cube for the model with and without capsule. It can be seen that the
-deformation for the model without capsule is much larger which is in
-aggreement with the data obtained by Hollenstein et al.
+test cube. It can be seen that the
+deformation for the model without capsule is much larger. This is in good
+agreement with the data obtained by Hollenstein et al.
 

### 2013-02-28 Doc/Miccai2013/miccai2013.tex
ADD: introduction
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@526 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
+\def\wrt{w.\,r.\,t.}
+\def\eg{e.\,g.}
+\def\ie{i.\,e.}
+\def\Dash{\nobreak\,---\penalty-500\,}
+
+\def\TODO{{\color{red}{TODO}}}
+
 %%% Comments
@@
 \newcommand{\CD}[1]{{\color{green}\textbf{CD: #1}}}
+\newcommand{\IP}[1]{{\color{cyan}\textbf{IP: #1}}}
 %%%
@@
 
-\section{Introduction} %{{{
+\section{Introduction} 
+\IP{I think we should provide some introductory motivation and broader context of the reasarch. I tried this:}
 
-Medical simulation is expected to play an integral part in many aspects of
-medical practice. From diagnosis, through planning or training to computer
-assisted intervention. A good medical model has to be simple yet efficient
-to be really usable.
-\CD{develop the applications for which it has to be real-time...}
+In 2009, nearly 100,000 European citizens died of cirrhosis or liver cancer. 
+According to the statistics, nearly 100,000 European citizens die of cirrhosis or liver cancer each year. 
+Although new methods such as radio-frequency ablation or cryo-ablation used in interventional radiology 
+represent promising improvements, surgery remains the option that offers the foremost success rate against these pathologies. 
+Nevertheless, surgery is not always performed due to several limitations, in particular the determination 
+of accurate eligibility criteria for the patient. 
+In this context, pre-operational planning becomes a crucial task having a significant impact on the treatment.
 
-\CD{First... explain the biomechanics of the liver (for instance: 
-\textit{
-The liver is a complex organ composed by three main structures: the parenchyma (soft tissue, that have a hyper-visco-elastic behavior) is coupled with a dense vascular network (tree-like tubular structure with a stiffer behavior) and is surrounded by the Glisson capsule  (a stiff collagenous membrane).
-It illustrates the need of relying on both volume, surface and line finite elements and the need of proposing coupling methods between them.}
- }
+Computer-aided physics-based medical simulation has proven to be an extremely useful technique in the area of medical training. 
+Nevertheless, whereas generic models are usually required in training simulators, accurate patient-specific modelling
+becomes necessary as soon as computer simulation is to be employed in the pre-operative planning. At the same time, 
+interactivity of such models  remains an important aspect, requiring real-time simulation which is often difficult to 
+achieve given the complexity of soft tissues.
 
-\CD{Yet, many work on liver modeling only consider the parenchyma...}
-The liver is usually modeled as a homogeneous isotropic (visco-)elastic
-material \TG{refs}.% It is however non-homogeneous and modeling it as
-%homogeneous necessarily requires complex non-linear material descriptions.
-Recently
-Peterl\'{i}k et al. \cite{Peterlik2012} showed the importance of the vascular
-structures on the liver deformation.
+In case of human liver, the task of real-time accurate modeling is challenging, mainly because of the complex structure 
+of this organ composed of three main constituents: \emph{parenchyma}, \emph{vascular networks} and \emph{Glisson's capsule}.
+The parenchyma is certainly the most studied component of the liver; today the researchers agree on hyperelastic 
+properties of the tissue, for which the mechanical parameters have been reported for example in~\cite{}. 
+Several methods have been proposed to model the hyperelastic behaviour at real-time rates, such multiplicative Jacobian decomposition
+introduced in~\cite{stephanie2008}.
 
-Another often overlooked component is the Glisson's capsule.
-Biomechanical measurements report that the capsule has very small
-thickness (approx. 20 $\mu$m for porcine liver \cite{Umale2011}).
-On the other hand it has in three orders of magnitude higher stiffness than
-the parenchyma.
-Because of this it plays an important role in local
-deformations of the liver \cite{Ahn2010,Hollenstein2006}.
-It was reported that evaluation of the liver as a homogeneous material
-without capsule can lead to overestimation of local deformations up to the factor of 3
-\cite{Hollenstein2006}.
+The mechanical importance of the vascular structures in liver is studied in~\cite{peterlik2012}. It shows that that the 
+influence of vessels on the mechanical behaviour of the liver is significant, mainly if large deformations occur. 
+Since detailed modeling of the vessels would be extremely costly (mainly because of the thickness of the vessel wall 
+being less than 250\,{$\mu$}m), the authors propose a composite model allowing for real-time simulation of entire liver
+with two vascular trees corresponding to portal and hepatic veins.
 
-While the influence on local deformation was reported, no experimental
-results show whether the capsule also plays an important role on the global
-deformations of the liver. We propose a composite model that models the
-capsule with membrane FEM elements. We show that the model is able to
-simulate expected local behaviour reported in literature. Further more we
-show that the capsule plays also an important role in the deformation on
-global scale.
+The final constituent represented by Glisson's capsule have been already studied in the literature. 
+Extensive experiments on porcine liver has been published in~\cite{Sagar}. The measurements show that although being very 
+thin (10--20$\mu m$), the Young's modulus of the capsule (reported as $8.22\pm3.42$\,MPa) exceeds the values for the parenchyma by three orders of magnitude,
+suggesting the mechanical influence of the membrane on the liver behaviour.
 
+In~\cite{Hollenstein}, a local influence of the capsule has been measured using a special aspiration device. The study was then repeated 
+in vivo on human patients during the operation, confirming the mechanical importance of the membrane~\cite{Ahn,Nava}.
+To our best knowledge, on attempt has been made to demonstrate the role of the Glisson's capsule on the global behaviour of 
+the liver, i.e. during large deformations of the organ.
+
+The main contribution of the paper is twofold: first, we present an efficient real-time model of the liver, taking into account 
+the parenchyma, vascularization and Glisson's capsule. The model is based on three different finite element representations for each constituent,
+linked together via mechanical coupling. We show that the model mimics the local experiments descibed in~\cite{Hollenstein}.
+
+Second, we use the complete model of the liver to demonstrate the influence of the Glisson's
+capsule via simulation: using a specific model of a pig liver built from CT contrast-enhanced data, we show that there is a significant 
+difference in the response of the model with and without the capsule in case when the liver undergoes large deformations.
+
+The paper is organized as follows: \TODO.
+
+%%Medical simulation is expected to play an integral part in many aspects of
+%%medical practice. From diagnosis, through planning or training to computer
+%%assisted intervention. A good medical model has to be simple yet efficient
+%%to be really usable.
+%\CD{develop the applications for which it has to be real-time...}
+%
+%\CD{First... explain the biomechanics of the liver (for instance: 
+%\textit{
+%The liver is a complex organ composed by three main structures: the parenchyma (soft tissue, that have a hyper-visco-elastic behavior) is coupled with a dense vascular network (tree-like tubular structure with a stiffer behavior) and is surrounded by the Glisson capsule  (a stiff collagenous membrane).
+%It illustrates the need of relying on both volume, surface and line finite elements and the need of proposing coupling methods between them.}
+% }
+%
+%\CD{Yet, many work on liver modeling only consider the parenchyma...}
+%The liver is usually modeled as a homogeneous isotropic (visco-)elastic
+%material \TG{refs}.% It is however non-homogeneous and modeling it as
+%%homogeneous necessarily requires complex non-linear material descriptions.
+%Recently
+%Peterl\'{i}k et al. \cite{Peterlik2012} showed the importance of the vascular
+%structures on the liver deformation.
+%
+%Another often overlooked component is the Glisson's capsule.
+%Biomechanical measurements report that the capsule has very small
+%thickness (approx. 20 $\mu$m for porcine liver \cite{Umale2011}).
+%On the other hand it has in three orders of magnitude higher stiffness than
+%the parenchyma.
+%Because of this it plays an important role in local
+%deformations of the liver \cite{Ahn2010,Hollenstein2006}.
+%It was reported that evaluation of the liver as a homogeneous material
+%without capsule can lead to overestimation of local deformations up to the factor of 3
+%\cite{Hollenstein2006}.
+%
+%While the influence on local deformation was reported, no experimental
+%results show whether the capsule also plays an important role on the global
+%deformations of the liver. We propose a composite model that models the
+%capsule with membrane FEM elements. We show that the model is able to
+%simulate expected local behaviour reported in literature. Further more we
+%show that the capsule plays also an important role in the deformation on
+%global scale.
+%
 \TG{summary of following sections.}

### 2013-02-28 Doc/Miccai2013/miccai2013.tex
paper: intro, intro to experiments
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@527 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 However, the Glisson's capsule is very thin and its influence for the global deformations of the liver was never highlighted.
-In this work, we propose to use a membrane FEM model for the capsule, that is coupled with a vascularized model of the liver (\CD{based on a coupling ?})
+In this work, we propose to use a membrane FEM model for the capsule, that is coupled with a vascularized model of the liver.
 We show that the approach is able to reproduce the deformations observed at the local level.
@@
 
-The mechanical importance of the vascular structures in liver is studied in~\cite{peterlik2012}. It shows that that the 
+The mechanical importance of the vascular structures in liver is studied in~\cite{Peterlik2012}. It shows that that the 
 influence of vessels on the mechanical behaviour of the liver is significant, mainly if large deformations occur. 
@@
 The final constituent represented by Glisson's capsule have been already studied in the literature. 
-Extensive experiments on porcine liver has been published in~\cite{Sagar}. The measurements show that although being very 
+Extensive experiments on porcine liver has been published in~\cite{Umale2013}. The measurements show that although being very 
 thin (10--20$\mu m$), the Young's modulus of the capsule (reported as $8.22\pm3.42$\,MPa) exceeds the values for the parenchyma by three orders of magnitude,
@@
 
-In~\cite{Hollenstein}, a local influence of the capsule has been measured using a special aspiration device. The study was then repeated 
-in vivo on human patients during the operation, confirming the mechanical importance of the membrane~\cite{Ahn,Nava}.
+In~\cite{Hollenstein2006}, a local influence of the capsule has been measured using a special aspiration device. The study was then repeated 
+in vivo on human patients during the operation, confirming the mechanical importance of the membrane~\cite{Ahn2010,Nava2008}.
 To our best knowledge, on attempt has been made to demonstrate the role of the Glisson's capsule on the global behaviour of 
@@
 the parenchyma, vascularization and Glisson's capsule. The model is based on three different finite element representations for each constituent,
-linked together via mechanical coupling. We show that the model mimics the local experiments descibed in~\cite{Hollenstein}.
+linked together via mechanical coupling. We show that the model mimics the local experiments descibed in~\cite{Hollenstein2006}.
 
@@
 
-%%Medical simulation is expected to play an integral part in many aspects of
-%%medical practice. From diagnosis, through planning or training to computer
-%%assisted intervention. A good medical model has to be simple yet efficient
-%%to be really usable.
-%\CD{develop the applications for which it has to be real-time...}
+%Medical simulation is expected to play an integral part in many aspects of
+%medical practice. From diagnosis, through planning or training to computer
+%assisted intervention.
+%Medical training application, like the ones for training laparoscopic
+%surgery, have to operate in real-time and a good medical model has to be
+%simple yet efficient to be really usable in this environment.
 %
-%\CD{First... explain the biomechanics of the liver (for instance: 
-%\textit{
-%The liver is a complex organ composed by three main structures: the parenchyma (soft tissue, that have a hyper-visco-elastic behavior) is coupled with a dense vascular network (tree-like tubular structure with a stiffer behavior) and is surrounded by the Glisson capsule  (a stiff collagenous membrane).
-%It illustrates the need of relying on both volume, surface and line finite elements and the need of proposing coupling methods between them.}
-% }
+%Liver is largest internal organ formed mostly by a soft tissue (parenchyma)
+%that is hyper-elastic and also visco-elastic. The parenchyma is interwoven
+%by dense vascular network, this tree-like tubular structure is stiffer then
+%the parenchyma. Finally the parenchyma is covered by very stiff collagenous
+%membrane that preserves the integrity of the organ.
 %
-%\CD{Yet, many work on liver modeling only consider the parenchyma...}
-%The liver is usually modeled as a homogeneous isotropic (visco-)elastic
-%material \TG{refs}.% It is however non-homogeneous and modeling it as
-%%homogeneous necessarily requires complex non-linear material descriptions.
-%Recently
+%Strangely, most of the work on liver modelling only considers the
+%parenchyma and model the liver as single homogeneous object \TG{Refs}.
+%Only recently
 %Peterl\'{i}k et al. \cite{Peterlik2012} showed the importance of the vascular
@@
 %
-%Another often overlooked component is the Glisson's capsule.
+%Another overlooked component is the Glisson's capsule.
 %Biomechanical measurements report that the capsule has very small
@@
 %While the influence on local deformation was reported, no experimental
-%results show whether the capsule also plays an important role on the global
+%results show whether the capsule also plays an important role in global
 %deformations of the liver. We propose a composite model that models the
@@
 %global scale.
-%
+
 \TG{summary of following sections.}
@@
 
-To evaluate the model we have implemented in the
-SOFA\footnote{www.sofa-framework.org} framework and performed a set of
-numerical simulations. First local deformations were compared to the
-results reported in literature to validate the method. Then the model of
-the complete liver was subjected to natural global deformations to asses
-the importance of the capsule on the deformation.
+To evaluate the model it was implemented in the
+SOFA\footnote{www.sofa-framework.org} framework and a set of
+numerical simulations was performed. First local deformations were compared to the
+results reported in literature to validate the method.
 
-Material properties used in the simulations were for a porcine liver as
+To show that, despite its very small thickness, the membrane cannot be
+neglected even in the context of global deformations and its overall
+stiffness plays an important role, the model of
+the complete liver was subjected to natural global deformations.
+
+In the simulations we used material properties for a porcine liver as
 reported by Umale \cite{Umale2013}. The
@@
 \TG{values for blood vessels}
-
-\CD{If the experiments demonstrate that the stiffness of the material is different, the thickness is very very small...
-so we have to evaluate if it is really necessary to include the model of the capsule for both local and global deformations.}
-\TG{I didn't undersand that. What do you mean?}
 
@@
 \end{document}
+% spell: setl spell spelllang=en spellfile=spelldict.en.utf-8.add
 % vim:set et sw=2 tw=75 fdm=marker fdl=3 fdc=4 isk+=_,-:

### 2013-02-28 Doc/Miccai2013/miccai2013.tex
ADD and CHANGE: some additions and modifications in Model -- Parenchyma and Vascularization
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@528 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\def\TODO{{\color{red}{TODO}}}
-
 %%% Comments
@@
 \newcommand{\IP}[1]{{\color{cyan}\textbf{IP: #1}}}
+\newcommand{\TODO}[1]{{\color{red}\textbf{TODO: #1}}}
 %%%
@@
 
-The paper is organized as follows: \TODO.
+The paper is organized as follows: \TODO{paper organization}.
 
@@
 
-In this section we will describe the construction of our composite model.
-It's three main parts are: tetrahedral model for parenchyma, membrane model
-for capsule coupled to the parenchyma and beam model of vascular system as
-proposed by Peterl\'{i}k et al. \cite{Peterlik2012}.
+In this section we describe the construction of our composite model
+being composed of three main components: tetrahedral FE model of the 
+parenchyma, Tymoshenko-beam model for the vascular structures and finally 
+triangular membrane elements used for the capsule.
 
-\subsection{Parenchyma} %{{{
+\subsection{Parenchyma and Vascularization} %{{{
 
-It is known that the parenchyma exhibits viscoelastic behaviour \TG{Cite
-something}, we however employ simpler elastic model as we are not so much
-interested in time-dependent behaviour but rather in static equilibrium
-under certain conditions.
-We also rely on the vascularized model of the parenchyma proposed by
-Peterl\'{i}k et al. \cite{Peterlik2012}.
+It is known that the parenchyma exhibits non-linear viscoelastic behaviour \cite{marchesseauMJED}.
+However, as we are mainly interested in the static equilibrium, we do not model the time-dependent
+phenomena related to viscosity.
 
-The parenchyma is modeled using corotational finite elements. The corotational
-formulation allows us to deal with geometrical non-linearities but still
-rely on linear stress-strain relationship. Thus rotations and large
-deformations are possible. In the corotational formulation the deformation
-of each element is expressed in the local frame of reference. There are
-several ways of computing the corotational frame for elements, we rely on
+%However, we employ simpler corotational model as we are not so much
+%interested in time-dependent behaviour but rather in static equilibrium
+%under certain conditions.
+%We also rely on the vascularized model of the parenchyma proposed by
+%Peterl\'{i}k et al. \cite{Peterlik2012}.
+
+The parenchyma is modeled using corotational finite elements~\cite{Felippa}.
+While relying on linear stress-strain relationship, it allows for large displacements including rotations. 
+While in the full non-linear formulation, the stiffness matrix relates the forces a $\Vec{f}$ and 
+displacements $\Vec{u}$ as $\Vec{f} = \Mat{K}(\Vec{u})$, the corotational model 
+requires the stiffness matrix $\Mat{K}_0$ of the system be computed only once before the simulation begins. 
+Then, in each step, the motion of each element $e$ is decomposed into rigid rotation $\Mat{R}^e$ and local deformation. 
+The rotations are then used to update each local element stiffness matrix as $\Mat{R}\Mat{K}_0\Mat{R}^{\top}$
+whereas the deformations are used to compute the linear strain in the local corotational frame.
+%The stiffness matrix $\Mat{K}$ can be expressed in terms of the deformation
+%$u$ and the force as $\Vec{f} = \Mat{K}(\Vec{u}) \Vec{u}$. 
+%Where the
+%components of $\Mat{K}$ depend on the orientation of each element in the
+%current simulation step.
+%the rigid component is  
+% The corotational
+%formulation allows us to deal with geometrical non-linearities but still
+%relies on linear stress-strain relationship. Thus rotations and large
+%deformations are modelled correctly. 
+%In the corotational formulation the deformation
+%of each element is expressed in the local frame of reference. 
+There are several ways of computing the corotational frame for elements; we rely on
 the geometrical method proposed in \cite{Nesme2005}.
 % NOTE: This description of corotational method is very simplified and could be extended.
-The stiffness matrix $\Mat{K}$ can be expressed in terms of the deformation
-$u$ and the force as $\Vec{f} = \Mat{K}(\Vec{u}) \Vec{u}$. Where the
-components of $\Mat{K}$ depend on the orientation of each element in the
-current simulation step.
 
-The vascularization method is based on linear beams
-with local frames of reference (in a many aspects similar to the
-corotational formulation). As such the model also handles geometric
-non-linearities in the deformation. The proposed model assumes no relative
-motion of vessels to the parenchyma. Segmented vascular network is
-discretized into the set of linear elements that are then projected into
-the tetrahedral mesh of the parenchyma. During the simulation forces and
-torques resulting from the deformations of vessels by parenchyma are
-computed and propagated back onto the tetrahedral mesh.
+The model of the vascularization is based on linear beams 
+with local frames of reference; in many aspects it's similar to the
+corotational formulation described above~\cite{Duriez}. As such the model also handles geometric
+non-linearities in the deformation. Through a specification of cross section and moments of inertia, 
+the model can account for the specific properties of the blood vessels.
+
+The beam-based model of vessels is mechanically coupled to the parenchyma as described in~\cite{peterlik2012}. 
+The coupling assumes that there is no relative motion between the vessels and surrounding parenchyma. 
+%The proposed model assumes no relative
+%motion of vessels to the parenchyma. Segmented vascular network is
+%discretized into the set of linear elements that are then projected into
+%the tetrahedral mesh of the parenchyma. 
+During the simulation, the nodes of linked beams are first displaced and rotated according to the actual motion of associated tetrahedra. 
+As the deformation of beams results in mechanical response represented by forces and torques, these are propagated back to 
+the tetrahedral FE model.
+
+%a displacement of tetrahedral nodes are mapped  forces and torques resulting from the deformations of vessels by parenchyma are
+%computed and propagated back onto the tetrahedral mesh.
 

### 2013-02-28 Doc/Miccai2013/miccai2013.tex
ADD: overview of "Global deformations"  section
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@529 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 elastic modulus 8~MPa and thickness 20~$\mu$m.
+\IP{OK, now we must decide what value we are going to use for the parenchyma, so it does not explode but it's somehow reasonable...)}
+
 \TG{values for blood vessels}
+
+\IP{we have to also add some information how we got the models -- either we comment briefly in each subsection:
+Local deformations: cube meshed in gmsh (graded mesh?), number of elements, etc...
+Global deformations: contrast-enhanced CT data obtained via acquisition performed on a female pig; the 3D mesh created by a
+semi-automatic segmentation using ITKSnap~\cite{itksnap}, then meshed using CGAL~\cite{cgal-meshingAlg} resulting in a mesh having 
+NNN elements. The capsule modelled using the surface triangular elements extracted from the tetrahedra. The vascularization 
+created from the segmentation of hepatic vein as described in~\cite{peterlik2012}. The structured composed of MMM elements.}
 
@@
 
-Test 1) Global deformation of liver under gravity
-  * show that there is a difference between simple homogeneous material
-    and material with capsule
-  * include blood vessels
+%Test 1) Global deformation of liver under gravity
+%  * show that there is a difference between simple homogeneous material
+%    and material with capsule
+%  * include blood vessels
+
+\IP{Short overview, I'll put it together tomorrow morning and add some fancy images. I am not sure if we are going to mess up 
+with gravity, it seems to me too much... We should rather focus only on this "registration-like" simulation (what about 
+using term "elastic warping"??? and explain properly why we think it proves the capsule plays a role. My first shot:}
+
+--we want to mimic a process takes place inside the pig such that it results in a natural deformation of the liver
+
+--such a natural deformation happens when the position of the pig is changed from the flank to a supine position
+
+--we performed a segmentation of both flank and supine and for some parts of the liver, displacements up to XXX\,cm  (I would say 3-4) 
+were observed, which is enough given the size of the entire liver (18\,cm longest dimension)
+
+--to induce the same deformation on the liver, we performed a process similar to deformation registration between the flank and supine 
+data: we chosen a set of feature in flank and supine data such that each feature could be reliably located in both supine and flank 
+data
+
+--this set of features was extracted from the vascularization, since although the vascular tree is getting deformed together with the liver, 
+the topology remains constant
+
+--then we placed a linear spring to each feature location in the supine data, so it's attracted towards its flank position
+
+--since the features were mapped to the elements via barycentric mapping, the entire model of the liver is being deformed, "elastically warped" from the 
+supine to the flank position
+
+--now the tricky part begins: we performed this process with models with and without the capsule; in each case we obtained 
+a different resulting shape of liver
+
+--additionally, we measured the forces needed to warp the liver in both cases: we obtained a difference summarized as.... ???
 

### 2013-02-28 Doc/Miccai2013/miccai2013.tex
bibliography, comments, figures for aspiration
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@530 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \newcommand{\Mat}[1]{\mathbf{#1}}
-
-%\Figure{width}{name}{caption}
-\newcommand{\Figure}[3]{%
-\begin{figure}[htb]
-  \centering
-  \includegraphics[width=#1]{#2}
-  \caption{\label{fig-#2}#3}
-\end{figure}}
 
@@
 The parenchyma is certainly the most studied component of the liver; today the researchers agree on hyperelastic 
-properties of the tissue, for which the mechanical parameters have been reported for example in~\cite{}. 
-Several methods have been proposed to model the hyperelastic behaviour at real-time rates, such multiplicative Jacobian decomposition
-introduced in~\cite{stephanie2008}.
+properties of the tissue, for which the mechanical parameters have been reported for example in~\cite{}\TODO. 
+Several methods have been proposed to model the hyperelastic behaviour at real-time rates, such as multiplicative Jacobian decomposition
+introduced in~\cite{Marchesseau2010}.
 
@@
 
-The final constituent represented by Glisson's capsule have been already studied in the literature. 
-Extensive experiments on porcine liver has been published in~\cite{Umale2013}. The measurements show that although being very 
+The final constituent represented by Glisson's capsule has been already studied in the literature. 
+Extensive experiments on porcine liver have been published in~\cite{Umale2013}. The measurements show that although being very 
 thin (10--20$\mu m$), the Young's modulus of the capsule (reported as $8.22\pm3.42$\,MPa) exceeds the values for the parenchyma by three orders of magnitude,
@@
 in vivo on human patients during the operation, confirming the mechanical importance of the membrane~\cite{Ahn2010,Nava2008}.
-To our best knowledge, on attempt has been made to demonstrate the role of the Glisson's capsule on the global behaviour of 
+To our best knowledge, no attempt has been made to demonstrate the role of the Glisson's capsule on the global behaviour of 
 the liver, i.e. during large deformations of the organ.
@@
 the parenchyma, vascularization and Glisson's capsule. The model is based on three different finite element representations for each constituent,
-linked together via mechanical coupling. We show that the model mimics the local experiments descibed in~\cite{Hollenstein2006}.
+linked together via mechanical coupling. We show that the model mimics the local experiments described in~\cite{Hollenstein2006}.
 
@@
 
-It is known that the parenchyma exhibits non-linear viscoelastic behaviour \cite{marchesseauMJED}.
+It is known that the parenchyma exhibits non-linear viscoelastic behaviour \cite{Marchesseau2010}.
 However, as we are mainly interested in the static equilibrium, we do not model the time-dependent
@@
 
-The parenchyma is modeled using corotational finite elements~\cite{Felippa}.
+The parenchyma is modeled using corotational finite elements~\cite{Felippa2005}.
 While relying on linear stress-strain relationship, it allows for large displacements including rotations. 
@@
 with local frames of reference; in many aspects it's similar to the
-corotational formulation described above~\cite{Duriez}. As such the model also handles geometric
+corotational formulation described above~\cite{Duriez2006}. As such the model also handles geometric
 non-linearities in the deformation. Through a specification of cross section and moments of inertia, 
@@
 
-The beam-based model of vessels is mechanically coupled to the parenchyma as described in~\cite{peterlik2012}. 
+The beam-based model of vessels is mechanically coupled to the parenchyma as described in~\cite{Peterlik2012}. 
 The coupling assumes that there is no relative motion between the vessels and surrounding parenchyma. 
@@
 \IP{OK, now we must decide what value we are going to use for the parenchyma, so it does not explode but it's somehow reasonable...)}
+\TG{Right now I'm using 3500 Pa for parenchyma. I can even use higher
+modulus, but not any smaller because we get large deformations and use of
+corotational model might become questionable.}
 
@@
 NNN elements. The capsule modelled using the surface triangular elements extracted from the tetrahedra. The vascularization 
-created from the segmentation of hepatic vein as described in~\cite{peterlik2012}. The structured composed of MMM elements.}
+created from the segmentation of hepatic vein as described in~\cite{Peterlik2012}. The structured composed of MMM elements.}
+\TG{Sure. Put it here or into the respective subsections? (cube: graded
+  mesh created by gmsh with 2648 tetrahedra, 518 triangles for capsule
+extracted from tetrahedral mesh.)}
 
@@
 
-\Figure{4in}{aspiration}{TODO: aspiration test; fix figure, add screenshot}
+\begin{figure}[htb]
+  \centering
+  \begin{tabular}{cc}
+    \includegraphics[width=3in]{aspiration.pdf} &
+    \includegraphics[height=1.5in]{aspiration.jpg}\\
+    a) & b)
+  \end{tabular}
+
+  \caption{\label{fig-aspiration}Aspiration test: a) displacment profile
+  with (blue) and without (black) the capsule; b) simulation scene}
+\end{figure}
 

### 2013-02-28 Doc/Miccai2013/aspiration.R

@@ -1,14 +1,14 @@
 a1 <- data.matrix(read.csv("aspiration-simple.csv"))
 a2 <- data.matrix(read.csv("aspiration-composite.csv"))
 
-a1 <- a1[1:21,c(4,3)]; a1[,2] <- a1[,2] - 0.03
-a2 <- a2[1:21,c(4,3)]; a2[,2] <- a2[,2] - 0.03
+a1 <- a1[1:31,c(4,3)]; a1[,2] <- a1[,2] - 0.015; a1[,2] <- a1[,2] * 1000
+a2 <- a2[1:33,c(4,3)]; a2[,2] <- a2[,2] - 0.015; a2[,2] <- a2[,2] * 1000
 
-pdf("aspiration.pdf", width=4, height=2, pointsize=9); # sizes in inches
+pdf("aspiration.pdf", width=3, height=1.5, pointsize=8); # sizes in inches
 p <- par(lwd=2, mar=c(4,4,1,1)+0.1);
 
 plot(range(c(a1[,1], a2[,1])), range(c(a1[,2], a2[,2])), #ylim=c(0, 0.004), 
-     type="n", xlab="x-coordinate", ylab="Displacement")
+     type="n", xlab="x-coordinate", ylab="Displacement [mm]")
 
 lines(a1, col="black")
 lines(a2, col="blue")

### 2013-03-01 Doc/Miccai2013/miccai2013.tex
TEMPORARY: aspiration.jpg => aspiration.pdf to compile the TeX. Aspiration.jpg needed!
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@531 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
FIX: small fixes in the introduction
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@534 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-In 2009, nearly 100,000 European citizens died of cirrhosis or liver cancer. 
 According to the statistics, nearly 100,000 European citizens die of cirrhosis or liver cancer each year. 
@@
 of this organ composed of three main constituents: \emph{parenchyma}, \emph{vascular networks} and \emph{Glisson's capsule}.
-The parenchyma is certainly the most studied component of the liver; today the researchers agree on hyperelastic 
+The parenchyma has certainly been the most studied component of the liver; today the researchers agree on hyperelastic 
 properties of the tissue, for which the mechanical parameters have been reported for example in~\cite{}\TODO. 
@@
 
-The mechanical importance of the vascular structures in liver is studied in~\cite{Peterlik2012}. It shows that that the 
+The mechanical importance of the vascular structures in liver is studied in~\cite{Peterlik2012}. It shows that the 
 influence of vessels on the mechanical behaviour of the liver is significant, mainly if large deformations occur. 
@@
 
-The final constituent represented by Glisson's capsule has been already studied in the literature. 
-Extensive experiments on porcine liver have been published in~\cite{Umale2013}. The measurements show that although being very 
-thin (10--20$\mu m$), the Young's modulus of the capsule (reported as $8.22\pm3.42$\,MPa) exceeds the values for the parenchyma by three orders of magnitude,
-suggesting the mechanical influence of the membrane on the liver behaviour.
+Rather a small number of studies have been conducted on the last liver constituent, the  Glisson's capsule.
+Quantitative results of experiments on a porcine liver have been published in~\cite{Umale2013}. The measurements indicate that although being very 
+thin (10--20$\mu m$), the capsule shows to be stiff in tensile tests mainly if compared to the parenchyma: the Young's modulus of the capsule (reported as $8.22\pm3.42$\,MPa) exceeds the values for the parenchyma by three orders of magnitude.
+This suggests that the mechanical influence of the membrane on the liver behaviour is significant.
 
@@
 To our best knowledge, no attempt has been made to demonstrate the role of the Glisson's capsule on the global behaviour of 
-the liver, i.e. during large deformations of the organ.
+the liver, mainly if large deformations of the organ take place.
 
@@
 the parenchyma, vascularization and Glisson's capsule. The model is based on three different finite element representations for each constituent,
-linked together via mechanical coupling. We show that the model mimics the local experiments described in~\cite{Hollenstein2006}.
+linked together via mechanical coupling. We show that the model mimics the \emph{local} experiments described in~\cite{Hollenstein2006}.
 
-Second, we use the complete model of the liver to demonstrate the influence of the Glisson's
-capsule via simulation: using a specific model of a pig liver built from CT contrast-enhanced data, we show that there is a significant 
+Second, we use the complete model of the liver to demonstrate the \emph{global} influence of the Glisson's
+capsule via simulation: using a specific model of a porcine liver built from CT contrast-enhanced data, we show that there is a significant 
 difference in the response of the model with and without the capsule in case when the liver undergoes large deformations.

### 2013-03-01 Doc/Miccai2013/aspiration.R
fixing the plot and screenshots
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@535 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@ -5,7 +5,7 @@ a1 <- a1[1:31,c(4,3)]; a1[,2] <- a1[,2] - 0.015; a1[,2] <- a1[,2] * 1000
 a2 <- a2[1:33,c(4,3)]; a2[,2] <- a2[,2] - 0.015; a2[,2] <- a2[,2] * 1000
 
 pdf("aspiration.pdf", width=3, height=1.5, pointsize=8); # sizes in inches
-p <- par(lwd=2, mar=c(4,4,1,1)+0.1);
+p <- par(lwd=0.5, mar=c(4,4,0,0)+0.1);
 
 plot(range(c(a1[,1], a2[,1])), range(c(a1[,2], a2[,2])), #ylim=c(0, 0.004), 
      type="n", xlab="x-coordinate", ylab="Displacement [mm]")

### 2013-03-01 Doc/Miccai2013/miccai2013.tex
membrane stiffness matrix
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@536 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
+\usepackage{amsmath}
 \usepackage{makeidx}
@@
 
-\newcommand{\Vec}[1]{\mathbf{#1}}
+\renewcommand{\Vec}[1]{\mathbf{#1}}
 \newcommand{\Mat}[1]{\mathbf{#1}}
@@
 To maintain simplicity of the composite model we choose simple triangular
-elements with constant strain.
+elements with constant strain (CST).
 
-\TG{add a ref. or dive into the definition? \cite{Felippa2003}} \CD{both...}
+The computation of elastic stiffness matrix follows the common derivation
+%
+\begin{eqnarray}
+  \Mat{K^m} & = \int_V \Mat{B}^T \Mat{E} \Mat{B} dV     \label{mem1} \\
+            & = h \int_A \Mat{B}^T \Mat{E} \Mat{B} dA   \label{mem2} \\
+            & = h A \Mat{B}^T \Mat{E} \Mat{B}           \label{mem3}
+\end{eqnarray}
+%
+where $\Mat{B}$ is the strain-displacement matrix, $\Mat{E}$ the material
+matrix, $h$ is the thickness and $A$ area of the element. In the previous
+\eqref{mem2} follows from the fact that we assume constant thickness of the
+element and \eqref{mem3} follows from the fact that the strain-displacement
+matrix is in our case constant. The strain-displacement matrix for the CST
+element can be expressed as:
+
+\begin{equation}
+  \Mat{B} = \frac{1}{2A} \begin{bmatrix}
+    y_{23} & 0      & y_{31} & 0      & y_{12} & 0 \\
+         0 & x_{32} & 0      & x_{13} & 0      & x_{21} \\
+    x_{32} & y_{23} & x_{13} & y_{31} & x_{21} & y_{12}
+  \end{bmatrix}
+\end{equation}
+
+The values $x_{ij} = x_i - x_j$ and $y_{ij} = y_i - y_j$ are computed from
+the $x$ or $y$ coordinates of the nodes $i,j$ of the triangular element.
+The reader can reffer to the respective literature for more thorough
+description \cite{Felippa2003}.
 

### 2013-03-01 Doc/Miccai2013/miccai2013.tex
adding references
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@537 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 %
-\titlerunning{TODO: Title}  % abbreviated title (for running head)
+\titlerunning{Complete Liver Model}  % abbreviated title (for running head)
 %                                     also used for the TOC unless
@@
 The parenchyma has certainly been the most studied component of the liver; today the researchers agree on hyperelastic 
-properties of the tissue, for which the mechanical parameters have been reported for example in~\cite{}\TODO. 
+properties of the tissue, for which the mechanical parameters have been reported for example in~\cite{Kerdok2006,Gao2009}. 
 Several methods have been proposed to model the hyperelastic behaviour at real-time rates, such as multiplicative Jacobian decomposition
@@
 The computation of elastic stiffness matrix follows the common derivation
-%
+
 \begin{eqnarray}
-  \Mat{K^m} & = \int_V \Mat{B}^T \Mat{E} \Mat{B} dV     \label{mem1} \\
+  \Mat{K}^m & = \int_V \Mat{B}^T \Mat{E} \Mat{B} dV     \label{mem1} \\
             & = h \int_A \Mat{B}^T \Mat{E} \Mat{B} dA   \label{mem2} \\
@@
 \end{eqnarray}
-%
+
 where $\Mat{B}$ is the strain-displacement matrix, $\Mat{E}$ the material
@@
 Global deformations: contrast-enhanced CT data obtained via acquisition performed on a female pig; the 3D mesh created by a
-semi-automatic segmentation using ITKSnap~\cite{itksnap}, then meshed using CGAL~\cite{cgal-meshingAlg} resulting in a mesh having 
+semi-automatic segmentation using ITKSnap\footnote{www.itksnap.org}, then meshed using CGAL\footnote{www.cgal.org} \cite{Boltcheva2009} resulting in a mesh having 
 NNN elements. The capsule modelled using the surface triangular elements extracted from the tetrahedra. The vascularization 

### 2013-03-01 Doc/Miccai2013/aspiration.R

@@ -13,7 +13,7 @@ plot(range(c(a1[,1], a2[,1])), range(c(a1[,2], a2[,2])), #ylim=c(0, 0.004),
 lines(a1, col="black")
 lines(a2, col="blue")
 
-abline(h=0, lty="dotted");
+#abline(h=0, lty="dotted");
 
 #legend("topright", c("A","B", "C", "D"), pch=c(1,4,5,6), col="black");
 

### 2013-03-01 Doc/Miccai2013/miccai2013.tex
ADD: graph showing the force evoluation when adding the membrane
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@538 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \subsection{Global Deformations}
-\CD{ Experiments demonstrate that capsule plays a role at local level, but still, maybe not so important to model the capsule for more global deformation...
-+ What are the applications for modeling the global deformations of the liver (training and planning simulation, physics-based registration, augmented reality,)}
 
-\TG{!!!}
-
-%Test 1) Global deformation of liver under gravity
-%  * show that there is a difference between simple homogeneous material
-%    and material with capsule
-%  * include blood vessels
+\begin{figure}[htb]
+  \centering
+    \includegraphics[width=.95\textwidth]{forceEvolution.pdf}
+  \caption{\label{fig-aspiration}Evolution of forces during elastic warping.}
+\end{figure}
 

### 2013-03-01 Doc/Miccai2013/miccai2013.tex
paper outline
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@539 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-The paper is organized as follows: \TODO{paper organization}.
+The paper is organized as follows: first we describe the proposed vascularized
+liver model with capsule (Sec. 2); then we show the correct behaviour of the
+model in context of local deformations and finally we analyze the influence of
+the capsule on the global deformations of the liver and show that despite the
+small thickness it plays an important role also in the global context.
 
@@
 %global scale.
-
-\TG{summary of following sections.}
 

### 2013-03-01 Doc/Miccai2013/miccai2013.tex
liver under gravity
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@540 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
small changes
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@542 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2013-03-01 Doc/Miccai2013/miccai2013.tex
@@
 
-The paper is organized as follows: \TODO{paper organization}.
+The paper is organized as follows: first we describe the proposed vascularized
+liver model with capsule (Sec. 2); then we show the correct behaviour of the
+model in context of \emph{local} deformations and finally we analyze the influence of
+the capsule on the \emph{global} deformations of the liver and show that despite the
+small thickness it plays an important role also in the \emph{global} context.
 
@@
 
-\TG{summary of following sections.}
-
 %}}}
@@
 While relying on linear stress-strain relationship, it allows for large displacements including rotations. 
-While in the full non-linear formulation, the stiffness matrix relates the forces a $\Vec{f}$ and 
+While in the full non-linear formulation, the stiffness matrix relates the forces $\Vec{f}$ and 
 displacements $\Vec{u}$ as $\Vec{f} = \Mat{K}(\Vec{u})$, the corotational model 
-requires the stiffness matrix $\Mat{K}_0$ of the system be computed only once before the simulation begins. 
+requires the stiffness matrix $\Mat{K}_0$ of the system to be computed only once before the simulation begins. 
 Then, in each step, the motion of each element $e$ is decomposed into rigid rotation $\Mat{R}^e$ and local deformation. 
@@
 thickness in the third dimension seems
-as a natural choice. Mechanics provides this a FE tool in the form of membranes
+as a natural choice. Mechanics provides this FE tool in the form of membranes
 and shells. Based on the small thickness (compared to the surface area) we also
@@
 
+\noindent%
 where $\Mat{B}$ is the strain-displacement matrix, $\Mat{E}$ the material
@@
     \includegraphics[width=.95\textwidth]{forceEvolution.pdf}
-  \caption{\label{fig-aspiration}Evolution of forces during elastic warping.}
+  \caption{\label{fig-force}Evolution of forces during elastic warping.}
+\end{figure}
+
+\begin{figure}[htb]
+  \centering
+    \includegraphics[width=.25\textwidth]{gravity.jpg}
+  \caption{\label{fig-gravity}Vascularized liver under pure gravity with (red) and without (green) the capsule.}
 \end{figure}

### 2013-03-01 Doc/Miccai2013/miccai2013.tex
ADD: part of the global deformation section plus images
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@543 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \usepackage[pdftex]{graphicx}
+\usepackage{subfigure}
 
@@
   \centering
-    \includegraphics[width=.25\textwidth]{gravity.jpg}
+    \subfigure[]{\includegraphics[width=.32\textwidth]{gravity2.png}}
+    \subfigure[]{\includegraphics[width=.32\textwidth]{registered.png}}
+    \subfigure[]{\includegraphics[width=.32\textwidth]{distanceMap.png}}
   \caption{\label{fig-gravity}Vascularized liver under pure gravity with (red) and without (green) the capsule.}
 \end{figure}
+
+In the previous section we have shown that the composite model of parenchyma and 
+capsule can be used to reproduce the \emph{local} behaviour of the tissue. In the following, 
+first we present a real-time complete model of entire liver built from image data. 
+Second, using this model we demonstrate the \emph{global} mechanical influence of the capsule when 
+the model undergoes large deformations.
+
+The complete model was constructed using constrast-enhanced CT data acquired on a female pig 
+being positioned first in supine and then flank configuration. 
+The liver and hepatic vein were segmented for both datasets  using semi-automatic methods available in ITKSnap.
+Further, an unstructured mesh was obtained from the segmented maps using algorithm~\cite{boltcheva}
+available in ITKSnap. The semiautomatic method presented in~\cite{peterlik2012} was 
+then applied to the segmented map of the hepatic vein, resulting in linked-beam model. 
+Finally, the model of the Glisson's capsule was constructed using the elements described in 
+section~\ref{} over the surface triangles extracted from the tetrahedral mesh. 
+The complete model composed of 6200 tetrahedra, 220 beams and 1820 triangles was then 
+used in simulation where external force representing the gravity was applied to the model, 
+resulting in large deformations mainly in the area of lobes. The refresh rate of 10\,FPS 
+was achieved on a desktop computer with Intel i7 CPU running at 3\,GHz and 16\,GB of memory.
 

### 2013-03-01 Doc/Miccai2013/miccai2013.tex
ADD: global deformations finished
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@544 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
     \includegraphics[width=.95\textwidth]{forceEvolution.pdf}
-  \caption{\label{fig-force}Evolution of forces during elastic warping.}
+  \caption{\label{f:forceEvol}Evolution of forces during elastic warping.}
 \end{figure}
@@
     \subfigure[]{\includegraphics[width=.32\textwidth]{distanceMap.png}}
-  \caption{\label{fig-gravity}Vascularized liver under pure gravity with (red) and without (green) the capsule.}
+  \caption{\label{f:global}Vascularized liver under pure gravity with (red) and without (green) the capsule.}
 \end{figure}
@@
 
-\IP{Short overview, I'll put it together tomorrow morning and add some fancy images. I am not sure if we are going to mess up 
-with gravity, it seems to me too much... We should rather focus only on this "registration-like" simulation (what about 
-using term "elastic warping"??? and explain properly why we think it proves the capsule plays a role. My first shot:}
+The simulation with gravity was also used as a first demonstration of the global influence of the capsule. 
+In Fig.~\ref{f:global}a, the models with capsule (red) and without (yellow) are depicted after being 
+deformed due to the gravity, showing a significant difference between the two models.
 
---we want to mimic a process takes place inside the pig such that it results in a natural deformation of the liver
+In order to demonstrate the influence of the capsule on a real deformation of the liver, we 
+performed a following protocol employing the supine and flank models built from the CT data. 
+First, we performed a rigid registration of the two organs \wrt\ the root point of the hepatic vein. 
+As shown in Fig.~\ref{f:global}b, the liver undergoes very large deformation when the body 
+is turned from the supine to the flank position, exceeding 6\,cm for some parts of the organ. 
+In order to induce similar deformation, we extracted feature pairs from the vascular system of the liver:
+we selected points such as bifurcations which can be identified easily in each supine and flank 
+model providing a set of correspondences. In the next stage, we used these correspondences as a non-homogeneous
+boundary conditions (imposed via penalty term) in order to perform a deformable registration of the supine
+model onto the flank data. During the registration, the response forces acting in the prescribed 
+points were calculated. It is important to emphasize that the goal of the simulation was not 
+to perform an accurate registration of the flank and supine data, however, the process was used 
+as an example of real deformation that can be undertaken by the liver.
 
---such a natural deformation happens when the position of the pig is changed from the flank to a supine position
-
---we performed a segmentation of both flank and supine and for some parts of the liver, displacements up to XXX\,cm  (I would say 3-4) 
-were observed, which is enough given the size of the entire liver (18\,cm longest dimension)
-
---to induce the same deformation on the liver, we performed a process similar to deformation registration between the flank and supine 
-data: we chosen a set of feature in flank and supine data such that each feature could be reliably located in both supine and flank 
-data
-
---this set of features was extracted from the vascularization, since although the vascular tree is getting deformed together with the liver, 
-the topology remains constant
-
---then we placed a linear spring to each feature location in the supine data, so it's attracted towards its flank position
-
---since the features were mapped to the elements via barycentric mapping, the entire model of the liver is being deformed, "elastically warped" from the 
-supine to the flank position
-
---now the tricky part begins: we performed this process with models with and without the capsule; in each case we obtained 
-a different resulting shape of liver
-
---additionally, we measured the forces needed to warp the liver in both cases: we obtained a difference summarized as.... ???
+The registration process was performed with the vascularized model, however, the effect of the capsule was 
+not considered. Once the static equilibrium was achieved, the effect of the capsule was gradually switched 
+on, \ie\ the stiffness of the membrane was slowly increased from 0 to 8\,MPa. The force response in the 
+prescribed points is depicted in Fig.~\ref{f:forceEvol} showing the change in the response as the effect
+of the capsule is increased. At the same time, the shape of the deformed liver with capsule was compared 
+to the model without capsule. A distance map is depicted in Fig.~\ref{f:global}c: in the red areas, the 
+maximal difference of 1\,cm. We believe that this results confirms that the Glisson's capsule 
+plays an important role in the mechanical response of the entire liver, \ie\ not only on the local, but 
+also on the global level.
 

### 2013-03-01 Doc/Miccai2013/miccai2013.tex
CHANGE: small modifications in Local Deformations section
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@545 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \section{Introduction} 
-\IP{I think we should provide some introductory motivation and broader context of the reasarch. I tried this:}
-
 According to the statistics, nearly 100,000 European citizens die of cirrhosis or liver cancer each year. 
@@
 \subsection{Glisson's Capsule} %{{{
-
+\label{ss:capsuleModel}
 The thickness of the Glisson's capsule is relatively small. Umale et al.
@@
 \section{Experiments} %{{{
-
 To evaluate the model it was implemented in the
@@
 
-In the simulations we used material properties for a porcine liver as
-reported by Umale \cite{Umale2013}. The
-parenchyma was modeled with elastic modulus 1980~Pa and the capsule has
-elastic modulus 8~MPa and thickness 20~$\mu$m.
-\IP{OK, now we must decide what value we are going to use for the parenchyma, so it does not explode but it's somehow reasonable...)}
-\TG{Right now I'm using 3500 Pa for parenchyma. I can even use higher
-modulus, but not any smaller because we get large deformations and use of
-corotational model might become questionable.}
+In the simulations we used material properties reported in literature: for the 
+Glisson's capsule we used Young's modulus $E_c$=8.22\,MPa and thickness $t_c$=20\,$\mu$m~\cite{Umale2013},
+whereas for the vessel wall we applied stiffness $E_v$=0.62\,MPa and $t_v$=200\,$\mu$m~\cite{Peterlik2012}.
+Since the values for parenchyma elasticity quite vary (usually between 2\,kPa to 5\,kPa), we perform the simulation 
+with $E_p$=3.5\,kPa.
 
-\TG{values for blood vessels}
-
-\IP{we have to also add some information how we got the models -- either we comment briefly in each subsection:
-Local deformations: cube meshed in gmsh (graded mesh?), number of elements, etc...
-Global deformations: contrast-enhanced CT data obtained via acquisition performed on a female pig; the 3D mesh created by a
-semi-automatic segmentation using ITKSnap\footnote{www.itksnap.org}, then meshed using CGAL\footnote{www.cgal.org} \cite{Boltcheva2009} resulting in a mesh having 
-NNN elements. The capsule modelled using the surface triangular elements extracted from the tetrahedra. The vascularization 
-created from the segmentation of hepatic vein as described in~\cite{Peterlik2012}. The structured composed of MMM elements.}
-\TG{Sure. Put it here or into the respective subsections? (cube: graded
-  mesh created by gmsh with 2648 tetrahedra, 518 triangles for capsule
-extracted from tetrahedral mesh.)}
+\begin{figure}[t]
+    	\subfigure[]{\includegraphics[height=3.5cm]{aspiration.jpg}}
+	\subfigure[]{\includegraphics[width=7cm]{aspiration.pdf}}
+  \caption{\label{fig-aspiration}Aspiration test: (a) the SOFA simulation scene, (b) displacment profile
+  with (blue) and without (black) the capsule.}
+\end{figure}
 
 \subsection{Local Deformations} %{{{
-
 During the contact with an instrument (probe, needle insertion, sharp
@@
 deformation of the object as a whole and therefore can be considered as
-local. Correct material properties are not only important for amount of the
-displacement but play also a crucial role in capturing the correct area of
+local. Correct material properties are not only important to quantify the
+displacement, but also play an important role in capturing the correct area of
 the deformation or its profile near the instrument.
-\TG{Does it make sense this way? I can add an image. :)}
 
-One example of such local deformation occurs during the aspiration test
+A good example of such a local deformation is the aspiration test
 when the response of liver to the negative pressure is measured.
-Hollenstein et al. \cite{Hollenstein2006} observed that the Glisson's
-capsule has non-negligible influence on response of the tissue and modeling
-only the parenchyma leads to overestimation of the deformation. We
-reproduced the test with our model and acquired similar results.
+In~\cite{Hollenstein2006} it is observed that the Glisson's
+capsule has a non-negligible influence on the response of the tissue and modeling
+only the parenchyma leads to an overestimation of the deformation.
 
-The aspiration device consists of a tube with 1 cm in diameter and is able
+%We reproduced the test with our model and acquired similar results.
+The aspiration device consists of a tube with 1\,cm in diameter and is able
 to control internal pressure in the tube. The test is performed by
 attaching the tube to the tissue and measuring the tissue response. We
-have set up a simulation in the spirit of this test. 
-The simulation is performed with a 15x15x15 mm cube of liver tissue and tube
-with 1 cm in diameter. The pressure of 3 kPa is simulated inside the tube.
+have set up a simulation to reproduce the experiment (see Fig.~\ref{fig-aspiration}a: we meshed a 15$\times$15$\times$15\,mm$^3$ 
+cube representing the tissue 
+resulting in 2648 tetrahedra. The we attached a tube having diameter d=1\,cm 
+and applied a pressure of 3\,kPa inside the tube. The interaction between the tube and 
+tissue surface was modelled with friction contact.
 
-Figure \ref{fig-aspiration} shows the profiles of cuts in the middle of the
-test cube. It can be seen that the
-deformation for the model without capsule is much larger. This is in good
-agreement with the data obtained by Hollenstein et al.
-
-\begin{figure}[htb]
-  \centering
-  \begin{tabular}{cc}
-    \includegraphics[width=3in]{aspiration.pdf} &
-    \includegraphics[height=1.5in]{aspiration.jpg}\\
-    a) & b)
-  \end{tabular}
-
-  \caption{\label{fig-aspiration}Aspiration test: a) displacment profile
-  with (blue) and without (black) the capsule; b) simulation scene}
-\end{figure}
+In Fig.~\ref{fig-aspiration}b the profiles of cuts in the middle of the
+test cube are presented showing significantly larger deformation in the model without capsule. This is in perfect 
+agreement with the results reported in~\cite{Hollenstein2006} and it can be concluded that 
+our composite model based on coupling the triangular and tetrahedral elements provides an accurate 
+model of parenchyma and capsule.
 
@@
 
-\subsection{Global Deformations}
-
-\begin{figure}[htb]
+\begin{figure}[t]
   \centering
@@
 
-\begin{figure}[htb]
+\begin{figure}[t]
   \centering
@@
 
+\subsection{Global Deformations}
 In the previous section we have shown that the composite model of parenchyma and 
@@
 being positioned first in supine and then flank configuration. 
-The liver and hepatic vein were segmented for both datasets  using semi-automatic methods available in ITKSnap.
-Further, an unstructured mesh was obtained from the segmented maps using algorithm~\cite{boltcheva}
-available in ITKSnap. The semiautomatic method presented in~\cite{peterlik2012} was 
+The liver and hepatic vein were segmented for both datasets  using semi-automatic methods available in ITKSnap~\footnote{www.itksnap.org}.
+Further, an unstructured mesh was obtained from the segmented maps using CGAL~\footnote{www.cgal.org}. 
+The semi-automatic method presented in~\cite{Peterlik2012} was 
 then applied to the segmented map of the hepatic vein, resulting in linked-beam model. 
 Finally, the model of the Glisson's capsule was constructed using the elements described in 
-section~\ref{} over the surface triangles extracted from the tetrahedral mesh. 
+section~\ref{ss:capsuleModel} over the surface triangles extracted from the tetrahedral mesh. 
 The complete model composed of 6200 tetrahedra, 220 beams and 1820 triangles was then 

### 2013-03-01 Doc/Miccai2013/miccai2013.tex
CHANGE: negligible modification of title
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@546 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
adding figure caption
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@547 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 %
-\title{Complete Real-Time Liver Model including Glisson Capsule, Vascular Network and Parenchyma}
+\title{Complete Real-Time Liver Model Including Glisson's Capsule, Vascularization and Parenchyma}
 %
@@
     \subfigure[]{\includegraphics[width=.32\textwidth]{distanceMap.png}}
-  \caption{\label{f:global}Vascularized liver under pure gravity with (red) and without (green) the capsule.}
+    \caption{\label{f:global}%
+    a) Liver under gravity with (red) and without (yellow) capsule;
+    b) liver in supine (yellow) and flank (red) position;
+    c) distance map between between registered version with and without
+    capsule}
 \end{figure}
@@
 to the model without capsule. A distance map is depicted in Fig.~\ref{f:global}c: in the red areas, the 
-maximal difference of 1\,cm. We believe that this results confirms that the Glisson's capsule 
+maximal difference of 1\,cm. We believe that this results confirm that the Glisson's capsule 
 plays an important role in the mechanical response of the entire liver, \ie\ not only on the local, but 

### 2013-03-01 Doc/Miccai2013/miccai2013.tex
removing unnecessary stuff to make it shorter; correcting insitutions
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@549 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 %Igor Peterl\'ik\inst{3} \and%
-%Christian Duriez\inst{2} \and%
-%Stephan\'e Cotin\inst{2,3}%
+%Christian Duriez\inst{2,4} \and%
+%Stephan\'e Cotin\inst{2}%
 %}
@@
 %INRIA Lille -- Nord, France\and%
-%IHU, Strasbourg, France%
+%Institut Hospitallo-Universitaire, Strasbourg, France \and%
+%University of Lille, Lille, France%
 %}
@@
 Rather a small number of studies have been conducted on the last liver constituent, the  Glisson's capsule.
-Quantitative results of experiments on a porcine liver have been published in~\cite{Umale2013}. The measurements indicate that although being very 
+Quantitative results of experiments on a porcine liver have been published in~\cite{Umale2011}. The measurements indicate that although being very 
 thin (10--20$\mu m$), the capsule shows to be stiff in tensile tests mainly if compared to the parenchyma: the Young's modulus of the capsule (reported as $8.22\pm3.42$\,MPa) exceeds the values for the parenchyma by three orders of magnitude.
@@
 The computation of elastic stiffness matrix follows the common derivation
-
+%
 \begin{eqnarray}
-  \Mat{K}^m & = \int_V \Mat{B}^T \Mat{E} \Mat{B} dV     \label{mem1} \\
-            & = h \int_A \Mat{B}^T \Mat{E} \Mat{B} dA   \label{mem2} \\
-            & = h A \Mat{B}^T \Mat{E} \Mat{B}           \label{mem3}
+  \Mat{K}^m & = & \int_V \Mat{B}^T \Mat{E} \Mat{B} dV     \label{mem1} \\
+            & = & h \int_A \Mat{B}^T \Mat{E} \Mat{B} dA   \label{mem2} \\
+            & = & h A \Mat{B}^T \Mat{E} \Mat{B}           \label{mem3}
 \end{eqnarray}
-
+%
 \noindent%
@@
 element can be expressed as:
-
+%
 \begin{equation}
@@
 \end{equation}
-
+%
 The values $x_{ij} = x_i - x_j$ and $y_{ij} = y_i - y_j$ are computed from
@@
 the stiffness matrices for capsule and parenchyma are easily assembled
-together and solved as one system.
-
-\CD{The following may appear as "trivial"... maybe we can remove this part at the end if we need space} 
-Without the loss of generality we can assume the tetrahedron consists of
-nodes $p_1, p_2, p_3$ and $p_4$ and the boundary triangle has nodes $p_1, p_2$
-and $p_3$. We can reorder the degrees of freedom so that the stiffness
-matrix $\Mat{K}^t$ for the tetrahedron can be written as:
-
-\begin{equation}
-  \Mat{K}^t = \left[\begin{array}{c|c}
-      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
-      \hline
-      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
-  \end{array}\right]
-\end{equation}
-
-Then the assembled stiffness matrix for the element is:
-
-\begin{equation}
-  \Mat{K} = \left[\begin{array}{c|c}
-      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
-      \hline
-      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
-  \end{array}\right]
-  +
-  \left[\begin{array}{c|c}
-      \Mat{K}^m & 0 \\
-      \hline
-      0 & 0 \\
-  \end{array}\right]
-\end{equation}
-
-Where $\Mat{K}^m$ is the stiffness matrix of the triangular membrane.
+together.% and solved as one system.
+%
+%\CD{The following may appear as "trivial"... maybe we can remove this part at the end if we need space} 
+%Without the loss of generality we can assume the tetrahedron consists of
+%nodes $p_1, p_2, p_3$ and $p_4$ and the boundary triangle has nodes $p_1, p_2$
+%and $p_3$. We can reorder the degrees of freedom so that the stiffness
+%matrix $\Mat{K}^t$ for the tetrahedron can be written as:
+%%
+%\begin{equation}
+%  \Mat{K}^t = \left[\begin{array}{c|c}
+%      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
+%      \hline
+%      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
+%  \end{array}\right]
+%\end{equation}
+%%
+%Then the assembled stiffness matrix for the element is:
+%%
+%\begin{equation}
+%  \Mat{K} = \left[\begin{array}{c|c}
+%      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
+%      \hline
+%      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
+%  \end{array}\right]
+%  +
+%  \left[\begin{array}{c|c}
+%      \Mat{K}^m & 0 \\
+%      \hline
+%      0 & 0 \\
+%  \end{array}\right]
+%\end{equation}
+%%
+%Where $\Mat{K}^m$ is the stiffness matrix of the triangular membrane.
 The resulting
@@
 In the simulations we used material properties reported in literature: for the 
-Glisson's capsule we used Young's modulus $E_c$=8.22\,MPa and thickness $t_c$=20\,$\mu$m~\cite{Umale2013},
+Glisson's capsule we used Young's modulus $E_c$=8.22\,MPa and thickness $t_c$=20\,$\mu$m~\cite{Umale2011},
 whereas for the vessel wall we applied stiffness $E_v$=0.62\,MPa and $t_v$=200\,$\mu$m~\cite{Peterlik2012}.

### 2013-03-02 Doc/Miccai2013/miccai2013.tex
CHANGE: new version of abstract,  small modifications in introduction
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@550 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{abstract}
-This work provides a complete physical model for the deformations of the liver with its three constituents: parenchyma, vessels and the Glisson's capsule.
-Some recent experimental results \cite{Ahn2010} have shown that there is a significant difference of stiffness between the Glisson's capsule and the parenchyma. 
-Moreover, it has been observed that the capsule plays an important role for the deformation at the local level \cite{Hollenstein2006}.
-However, the Glisson's capsule is very thin and its influence for the global deformations of the liver was never highlighted.
-In this work, we propose to use a membrane FEM model for the capsule, that is coupled with a vascularized model of the liver.
-We show that the approach is able to reproduce the deformations observed at the local level.
-Additionally, we show, by simulating natural global deformations of the liver observed on real images (due to gravity), that the results have important differences with and without the capsule.
+The accurate mechanical modeling of liver is of a paramount interest in simulation-based operation planning 
+and computer-aided per-operative guidance. However, the realistic simulation of liver behaviour is a challenging task, 
+since the organ is composed of three constituents: parenchyma, vascularization and Glisson's capsule, each having different mechanical properties. While the 
+mechanical role of vessels on both global and local behaviour of the organ has been measured and modelled previously~\cite{Peterlik2012}, only local 
+effects have been studied in the case of Glisson's capsule~\cite{Hollenstein2006,Ahn2010}.
 
-\keywords{liver, Glisson's capsule, vascular network, deformation modeling, membrane}
+In this paper we introduce what we believe is the first real-time model of liver taking into account all three constituents, each being simulated with 
+a different type of finite elements.
+Using the model, we focus on the role of the Glisson's capsule: first, we show that the model is able to reproduce correctly
+the local effects of the capsule in so called aspiration test. Second, based on CT image data of porcine abdominal cavity acquired in supine and flank 
+positions, we demonstrate 
+that the capsule has also a global impact on the mechanical behaviour of the liver in the case when the organ undergoes large deformations.
+
+%This work provides a complete physical model for the deformations of the liver with its three constituents: parenchyma, vessels and the Glisson's capsule.
+%Some recent experimental results \cite{Ahn2010} have shown that there is a significant difference of stiffness between the Glisson's capsule and the parenchyma. 
+%Moreover, it has been observed that the capsule plays an important role for the deformation at the local level \cite{Hollenstein2006}.
+%However, the Glisson's capsule is very thin and its influence for the global deformations of the liver was never highlighted.
+%In this work, we propose to use a membrane FEM model for the capsule, that is coupled with a vascularized model of the liver.
+%We show that the approach is able to reproduce the deformations observed at the local level.
+%Additionally, we show, by simulating natural global deformations of the liver observed on real images (due to gravity), that the results have important differences with and without the capsule.
+
+\keywords{deformable liver model, Glisson's capsule, vascularization, real-time soft tissue modeling}
 \end{abstract}
@@
 According to the statistics, nearly 100,000 European citizens die of cirrhosis or liver cancer each year. 
-Although new methods such as radio-frequency ablation or cryo-ablation used in interventional radiology 
-represent promising improvements, surgery remains the option that offers the foremost success rate against these pathologies. 
+Although new methods such as radio-frequency- or cryo-ablation used in the interventional radiology 
+seems to be promising, surgery remains the option that offers the foremost success rate against these pathologies. 
 Nevertheless, surgery is not always performed due to several limitations, in particular the determination 
 of accurate eligibility criteria for the patient. 
-In this context, pre-operational planning becomes a crucial task having a significant impact on the treatment.
+In this context, the pre-operational planning becomes a crucial task having a significant impact on the treatment.
 
@@
 of this organ composed of three main constituents: \emph{parenchyma}, \emph{vascular networks} and \emph{Glisson's capsule}.
-The parenchyma has certainly been the most studied component of the liver; today the researchers agree on hyperelastic 
+The parenchyma has certainly been the most studied component of the liver; today researchers an agree on hyperelastic 
 properties of the tissue, for which the mechanical parameters have been reported for example in~\cite{Kerdok2006,Gao2009}. 
@@
 The mechanical importance of the vascular structures in liver is studied in~\cite{Peterlik2012}. It shows that the 
-influence of vessels on the mechanical behaviour of the liver is significant, mainly if large deformations occur. 
-Since detailed modeling of the vessels would be extremely costly (mainly because of the thickness of the vessel wall 
-being less than 250\,{$\mu$}m), the authors propose a composite model allowing for real-time simulation of entire liver
-with two vascular trees corresponding to portal and hepatic veins.
+influence of vessels on the mechanical behaviour of the organ is significant, mainly if large deformations occur. 
+Since a detailed modeling of the vessels would be extremely costly (mainly because of the small thickness of the vessel wall, 
+the authors propose a composite model allowing for real-time simulation of entire liver with vasculature.
 
 Rather a small number of studies have been conducted on the last liver constituent, the  Glisson's capsule.
-Quantitative results of experiments on a porcine liver have been published in~\cite{Umale2011}. The measurements indicate that although being very 
-thin (10--20$\mu m$), the capsule shows to be stiff in tensile tests mainly if compared to the parenchyma: the Young's modulus of the capsule (reported as $8.22\pm3.42$\,MPa) exceeds the values for the parenchyma by three orders of magnitude.
-This suggests that the mechanical influence of the membrane on the liver behaviour is significant.
+Quantitative results of experiments on a porcine liver have been published in~\cite{Umale2013}; the measurements indicate that although being very 
+thin (10--20$\mu m$), the capsule shows to be stiff in tensile tests: the Young's modulus of the capsule (reported as $8.22\pm3.42$\,MPa) 
+exceeds the values for the parenchyma by three orders of magnitude.
+This suggests that the mechanical influence of the membrane on the liver behaviour is not negligible.
 
@@
 To our best knowledge, no attempt has been made to demonstrate the role of the Glisson's capsule on the global behaviour of 
-the liver, mainly if large deformations of the organ take place.
+the liver, mainly if the organ undergoes large deformations which is often the case during the surgery.
 
@@
 The paper is organized as follows: first we describe the proposed vascularized
-liver model with capsule (Sec. 2); then we show the correct behaviour of the
-model in context of \emph{local} deformations and finally we analyze the influence of
-the capsule on the \emph{global} deformations of the liver and show that despite the
-small thickness it plays an important role also in the \emph{global} context.
+liver model with capsule. Second, we validate our model in context of local deformations by reproducing the aspiration test 
+described in~\cite{Hollenstein2006}. Finally, we demonstrate that in spite of it's small thickness, the Glisson's 
+capsule has a global influence on the liver undergoing large deformations.
+
+%Then we show the correct behaviour of the
+%model in context of \emph{local} deformations and finally we analyze the influence of
+%the capsule on the \emph{global} deformations of the liver and show that despite the
+%small thickness it plays an important role also in the \emph{global} context.
 
@@
 
-\begin{figure}[t]
-  \centering
-    \subfigure[]{\includegraphics[width=.32\textwidth]{gravity2.png}}
-    \subfigure[]{\includegraphics[width=.32\textwidth]{registered.png}}
-    \subfigure[]{\includegraphics[width=.32\textwidth]{distanceMap.png}}
-    \caption{\label{f:global}%
-    a) Liver under gravity with (red) and without (yellow) capsule;
-    b) liver in supine (yellow) and flank (red) position;
-    c) distance map between between registered version with and without
-    capsule}
-\end{figure}
-
 \subsection{Global Deformations}
@@
 deformed due to the gravity, showing a significant difference between the two models.
+
+\begin{figure}[t]
+  \centering
+    \subfigure[]{\includegraphics[width=.32\textwidth]{gravity2.png}}
+    \subfigure[]{\includegraphics[width=.32\textwidth]{registered.png}}
+    \subfigure[]{\includegraphics[width=.32\textwidth]{distanceMap.png}}
+    \caption{\label{f:global}%
+    a) Liver under gravity with (red) and without (yellow) capsule;
+    b) liver in supine (yellow) and flank (red) position;
+    c) distance map between between registered version with and without
+    capsule}
+\end{figure}
 

### 2013-03-02 Doc/Miccai2013/miccai2013.tex
ADD: small changes in the text
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@551 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 Rather a small number of studies have been conducted on the last liver constituent, the  Glisson's capsule.
-Quantitative results of experiments on a porcine liver have been published in~\cite{Umale2013}; the measurements indicate that although being very 
+Quantitative results of experiments on a porcine liver have been published in~\cite{Umale2011}; the measurements indicate that although being very 
 thin (10--20$\mu m$), the capsule shows to be stiff in tensile tests: the Young's modulus of the capsule (reported as $8.22\pm3.42$\,MPa) 
@@
 \label{ss:capsuleModel}
-The thickness of the Glisson's capsule is relatively small. Umale et al.
-\cite{Umale2011} reported for porcine liver values in range of 10-20
-$\mu$m.
+The thickness of the Glisson's capsule is relatively small: values in range of 10-20
+$\mu$m have been reported in~\cite{Umale2011}.
 It is not possible to model such thin structure with classical tetrahedral
-elements. Modeling this non-homogeneous volumetric object would require a
-dense mesh to avoid numerical instabilities and would significantly
-violate the speed requirements imposed on medical simulators.
-Instead modeling the capsule with two-dimensional elements that abstract from the
+elements, if real-time aspect of the simulation is to be preserved.
+Furthermore, modeling both the tissue and the capsule would require an extremely 
+dense mesh to avoid numerical instabilities and thus would significantly
+violate the speed requirements required for medical simulators.
+Instead, modeling the capsule with two-dimensional elements that abstract from the
 thickness in the third dimension seems
-as a natural choice. Mechanics provides this FE tool in the form of membranes
-and shells. Based on the small thickness (compared to the surface area) we also
+to be a natural choice. In the elasticity theory, this functionality is usually provided by membrane and shell elements.
+Based on the small thickness (compared to the surface area) we also
 assume negligible bending forces and propose a model based on membrane
@@
 %
-\noindent%
 where $\Mat{B}$ is the strain-displacement matrix, $\Mat{E}$ the material
@@
 element and \eqref{mem3} follows from the fact that the strain-displacement
-matrix is in our case constant. The strain-displacement matrix for the CST
+matrix is constant in our case. The strain-displacement matrix for the CST
 element can be expressed as:
@@
 the $x$ or $y$ coordinates of the nodes $i,j$ of the triangular element.
-The reader can reffer to the respective literature for more thorough
-description \cite{Felippa2003}.
+The reader can refer to the respective literature for more thorough
+description~\cite{Felippa2003}.
 
-We again use linear elastic material and employ the corotational formulation
+We use linear elastic material and employ the corotational formulation
 for the triangular elements.
@@
 \subsection{Coupling Between Capsule and Parenchyma} %{{{
-
 The literature reports high cohesion between capsule and parenchyma.
-Based on this
-property we may assume no relative motion of the capsule to the parenchyma.
-While in general we could use arbitrary surface mesh for a capsule we use
-the fact that the parenchyma is modeled by tetrahedral elements which have
-triangular faces. Because the boundary of the volumetric mesh is already
-triangulated, we only extract the boundary triangles which we use for the
-model of capsule.
+Based on this property we assume there is no relative motion of the capsule \wrt\ the parenchyma.
+Although an arbitrary surface mesh could be used to model the capsule, we exploit 
+the fact that the parenchyma is modeled by tetrahedral elements having
+triangular faces. Thus, as the boundary of the volumetric mesh is already
+triangulated, we simply employ the triangles on the mesh surface to model the capsule.
 
 Using directly the boundary of the tetrahedral mesh does not only solve the
-problem of obtaining the surface mesh, but has one more advantage. Nodes
-of the triangular mesh overlap with nodes of the tetrahedral mesh and we do
-not have to project nodes of triangular mesh onto the tetrahedra. More over,
-the stiffness matrices for capsule and parenchyma are easily assembled
+problem of building the surface mesh, but has one more advantage: the nodes
+of the triangular mesh coincide with the nodes of the tetrahedral mesh, so no projection of one mesh onto the other is needed.
+Moreover, the stiffness matrices for capsule and parenchyma are then easily assembled
 together.% and solved as one system.
@@
 %Where $\Mat{K}^m$ is the stiffness matrix of the triangular membrane.
-The resulting
-system of linear equations is then solved by direct LDL solver.
+The resulting system of linear equations is solved by direct solver based on Cholesky decomposition.
 
@@
 \section{Conclusion} %{{{
+In the paper, a complete model of the liver was presented taking into account 
+three constituents of the organ, each being modelled with a different type 
+of finite elements: tetrahedral corotational model used the parenchyma, serially 
+linked corotational beams based on Tymoshenko formulation modelling the vascularization 
+and finally triangular elements with constant strain employed for the capsule. 
+The model was first validated for local displacements: the aspiration test presented in~\cite{Hollenstein2006} 
+was succesfully reproduced, showing a good agreement between the simulation and experiments. 
+Further, the model was used to demonstrate global influence of the capsule when the organ 
+undergoes large deformations.
 
-A composite model for simulation of complete liver is presented in the
-paper. The work of Peterlik et al. for vaskularized liver model was
-extended by addig the capsule modeled with membrane FEM model mechanically
-coupled with the parenchyma. To validate the model a simulation of local
-deformations was performed and was found in an aggrement with the results
-measured and published by Hollenstein et al.
+In the future work, we plan to employ the model to model other organs where the capsule play an important role
+(for example kidneys). We will also focus on better understanding the interaction between the organs in the abdominal 
+cavity.
 
-While the influence of the capsule on local deformations was previously
-studied in the literature, it's significance on the global scale
-deformations of the liver remained unexplored. We have performed a set of
-natural global deformations of the complete liver and shown that the
-capsule, despite its small thickness, plays a significant role also in
-global context.
-
+%While the influence of the capsule on local deformations was previously
+%studied in the literature, it's significance on the global scale
+%deformations of the liver remained unexplored. We have performed a set of
+%natural global deformations of the complete liver and shown that the
+%capsule, despite its small thickness, plays a significant role also in
+%global context.
+%
 %}}}

### 2013-03-02 Doc/Miccai2013/miccai2013.tex
small fixes
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@552 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 attaching the tube to the tissue and measuring the tissue response. We
-have set up a simulation to reproduce the experiment (see Fig.~\ref{fig-aspiration}a: we meshed a 15$\times$15$\times$15\,mm$^3$ 
+have set up a simulation to reproduce the experiment (see Fig.~\ref{fig-aspiration}a). We meshed a 15$\times$15$\times$15\,mm$^3$ 
 cube representing the tissue 
-resulting in 2648 tetrahedra. The we attached a tube having diameter d=1\,cm 
+resulting in 2648 tetrahedra. Then we attached a tube having diameter d=1\,cm 
 and applied a pressure of 3\,kPa inside the tube. The interaction between the tube and 
@@
     b) liver in supine (yellow) and flank (red) position;
-    c) distance map between between registered version with and without
-    capsule}
+    c) distance map between registered version with and without capsule}
 \end{figure}

### 2013-03-02 Doc/Miccai2013/miccai2013.tex
spell check, homogenized figure captions
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@553 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 of this organ composed of three main constituents: \emph{parenchyma}, \emph{vascular networks} and \emph{Glisson's capsule}.
-The parenchyma has certainly been the most studied component of the liver; today researchers an agree on hyperelastic 
+The parenchyma has certainly been the most studied component of the liver; today researchers agree on hyperelastic 
 properties of the tissue, for which the mechanical parameters have been reported for example in~\cite{Kerdok2006,Gao2009}. 
@@
 
-The complete model was constructed using constrast-enhanced CT data acquired on a female pig 
+The complete model was constructed using contrast-enhanced CT data acquired on a female pig 
 being positioned first in supine and then flank configuration. 
@@
     \caption{\label{f:global}%
-    a) Liver under gravity with (red) and without (yellow) capsule;
-    b) liver in supine (yellow) and flank (red) position;
-    c) distance map between registered version with and without capsule}
+      (a) Liver under gravity with (red) and without (yellow) capsule;
+      (b) liver in supine (yellow) and flank (red) position;
+      (c) distance map between registered version with and without capsule.}
 \end{figure}
@@
 The model was first validated for local displacements: the aspiration test presented in~\cite{Hollenstein2006} 
-was succesfully reproduced, showing a good agreement between the simulation and experiments. 
+was successfully reproduced, showing a good agreement between the simulation and experiments. 
 Further, the model was used to demonstrate global influence of the capsule when the organ 

### 2013-03-02 Doc/Miccai2013/miccai2013.tex
small changes
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@554 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \section{Experiments} %{{{
-To evaluate the model it was implemented in the
+The model presented in the previous section was implemented in 
 SOFA\footnote{www.sofa-framework.org} framework and a set of
@@
 results reported in literature to validate the method.
-
-To show that, despite its very small thickness, the membrane cannot be
+Second, to show that in spite of its very small thickness the membrane cannot be
 neglected even in the context of global deformations and its overall
 stiffness plays an important role, the model of
-the complete liver was subjected to natural global deformations.
+the complete liver was subjected to global deformations.
 
-In the simulations we used material properties reported in literature: for the 
+In the simulations we used material properties reported in the literature: for the 
 Glisson's capsule we used Young's modulus $E_c$=8.22\,MPa and thickness $t_c$=20\,$\mu$m~\cite{Umale2011},
-whereas for the vessel wall we applied stiffness $E_v$=0.62\,MPa and $t_v$=200\,$\mu$m~\cite{Peterlik2012}.
+whereas for the vessel wall we applied stiffness $E_v$=0.62\,MPa and thickness $t_v$=200\,$\mu$m~\cite{Peterlik2012}.
 Since the values for parenchyma elasticity quite vary (usually between 2\,kPa to 5\,kPa), we perform the simulation 
@@
 \subsection{Local Deformations} %{{{
-During the contact with an instrument (probe, needle insertion, sharp
-instruments, \ldots) a specific deformation occurs in the near vicinity of
+During the contact with an instrument such as probe, needle, scalpel and others,
+a specific deformations take place in the vicinity of
 the instrument. This type of deformation may not necessarily induce the
@@
 A good example of such a local deformation is the aspiration test
-when the response of liver to the negative pressure is measured.
-In~\cite{Hollenstein2006} it is observed that the Glisson's
+where the response of liver exposed locally to a negative pressure is measured.
+In~\cite{Hollenstein2006} it is reported that the Glisson's
 capsule has a non-negligible influence on the response of the tissue and modeling
 only the parenchyma leads to an overestimation of the deformation. 
-
 %We reproduced the test with our model and acquired similar results.
-The aspiration device consists of a tube with 1\,cm in diameter and is able
+The aspiration device consists of a tube 1\,cm in diameter and is able
 to control internal pressure in the tube. The test is performed by
 attaching the tube to the tissue and measuring the tissue response. We
-have set up a simulation to reproduce the experiment (see Fig.~\ref{fig-aspiration}a). We meshed a 15$\times$15$\times$15\,mm$^3$ 
+have set up a simulation in SOFA to reproduce the experiment (see Fig.~\ref{fig-aspiration}a: we meshed a 15$\times$15$\times$15\,mm$^3$ 
 cube representing the tissue 
-resulting in 2648 tetrahedra. Then we attached a tube having diameter d=1\,cm 
+resulting in 2648 tetrahedra. Then we attached a 1\,cm tube 
 and applied a pressure of 3\,kPa inside the tube. The interaction between the tube and 
-tissue surface was modelled with friction contact.
+tissue was modelled with friction contact.
 
@@
     \includegraphics[width=.95\textwidth]{forceEvolution.pdf}
-  \caption{\label{f:forceEvol}Evolution of forces during elastic warping.}
+  \caption{\label{f:forceEvol}Evolution of forces during elastic during the simulation, where the supine data is deformed
+towards the flank shape, and the stiffness of the capsule is increased from 0 to 8.22\,MPa after the initial equilibrium is achieved.}
 \end{figure}
@@
 was achieved on a desktop computer with Intel i7 CPU running at 3\,GHz and 16\,GB of memory. 
-
 The simulation with gravity was also used as a first demonstration of the global influence of the capsule. 
@@
 In order to induce similar deformation, we extracted feature pairs from the vascular system of the liver:
-we selected points such as bifurcations which can be identified easily in each supine and flank 
-model providing a set of correspondences. In the next stage, we used these correspondences as a non-homogeneous
+we selected points such as bifurcations which can be identified easily in both supine and flank 
+data providing a set of correspondences. In the next step, we used these correspondences as a non-homogeneous
 boundary conditions (imposed via penalty term) in order to perform a deformable registration of the supine
-model onto the flank data. During the registration, the response forces acting in the prescribed 
-points were calculated. It is important to emphasize that the goal of the simulation was not 
-to perform an accurate registration of the flank and supine data, however, the process was used 
-as an example of real deformation that can be undertaken by the liver.
+model onto the flank data. During the registration, the response forces acting in the points with prescribed displacements
+were calculated. It is important to emphasize that the goal of the simulation was not 
+to perform an accurate registration of the flank and supine data, but the simulation was used 
+as an example of real deformation of the organ.
 
-The registration process was performed with the vascularized model, however, the effect of the capsule was 
-not considered. Once the static equilibrium was achieved, the effect of the capsule was gradually switched 
-on, \ie\ the stiffness of the membrane was slowly increased from 0 to 8\,MPa. The force response in the 
-prescribed points is depicted in Fig.~\ref{f:forceEvol} showing the change in the response as the effect
-of the capsule is increased. At the same time, the shape of the deformed liver with capsule was compared 
-to the model without capsule. A distance map is depicted in Fig.~\ref{f:global}c: in the red areas, the 
-maximal difference of 1\,cm. We believe that this results confirm that the Glisson's capsule 
-plays an important role in the mechanical response of the entire liver, \ie\ not only on the local, but 
-also on the global level.
+The simulation was performed with the model composed of parenchyma,  vessels and capsule, however, zero stiffness was 
+set for the last component.
+Once the static equilibrium was achieved, the effect of the capsule was gradually switched 
+on, \ie\ the stiffness of the membrane was slowly increased from 0 to 8.22\,MPa. The evolution of the force response in the 
+points with prescribed displacements is depicted in Fig.~\ref{f:forceEvol} showing the change in the response as the effect
+of the capsule is included. At the same time, the shape of the deformed liver with capsule was compared 
+to the model without the capsule. A distance map for those cases is depicted in Fig.~\ref{f:global}c: in the red areas, the 
+maximal difference exceeding 1.5\,cm was measured. We believe that this result confirms that the Glisson's capsule 
+plays an important role in the global mechanical response of the liver.
 
@@
 
-\section{Conclusion} %{{{
+\section{Conclusions} %{{{
 In the paper, a complete model of the liver was presented taking into account 

### 2013-03-02 Doc/Miccai2013/miccai2013.tex
final fix
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@556 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
     \includegraphics[width=.95\textwidth]{forceEvolution.pdf}
-  \caption{\label{f:forceEvol}Evolution of forces during elastic during the simulation, where the supine data is deformed
+  \caption{\label{f:forceEvol}Evolution of forces during the simulation, where the supine data is deformed
 towards the flank shape, and the stiffness of the capsule is increased from 0 to 8.22\,MPa after the initial equilibrium is achieved.}

### 2013-03-02 Doc/Miccai2013/miccai2013.tex
CHANGE: two important modifs: new mesh sizes and frame rate, max. Hausdorff distance modified
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@557 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 section~\ref{ss:capsuleModel} over the surface triangles extracted from the tetrahedral mesh. 
-The complete model composed of 6200 tetrahedra, 220 beams and 1820 triangles was then 
+The complete model composed of 4266 tetrahedra, 188 beams and 1284 triangles was then 
 used in simulation where external force representing the gravity was applied to the model, 
-resulting in large deformations mainly in the area of lobes. The refresh rate of 10\,FPS 
-was achieved on a desktop computer with Intel i7 CPU running at 3\,GHz and 16\,GB of memory. 
+resulting in large deformations mainly in the area of lobes. The refresh rate of 22\,FPS 
+was achieved on a desktop computer with Intel i7 CPU running at 3.4\,GHz and 8\,GB of memory. 
 The simulation with gravity was also used as a first demonstration of the global influence of the capsule. 
@@
 of the capsule is included. At the same time, the shape of the deformed liver with capsule was compared 
-to the model without the capsule. A distance map for those cases is depicted in Fig.~\ref{f:global}c: in the red areas, the 
-maximal difference exceeding 1.5\,cm was measured. We believe that this result confirms that the Glisson's capsule 
+to the model without the capsule. A Hausdorff distance map for those cases is depicted in Fig.~\ref{f:global}c: in the red areas, the 
+maximal difference exceeding 1.25\,cm was measured. We believe that this result confirms that the Glisson's capsule 
 plays an important role in the global mechanical response of the liver.

### 2013-03-02 Doc/Miccai2013/miccai2013.tex
FIX: English and small typos
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@558 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 since the organ is composed of three constituents: parenchyma, vascularization and Glisson's capsule, each having different mechanical properties. While the 
-mechanical role of vessels on both global and local behaviour of the organ has been measured and modelled previously~\cite{Peterlik2012}, only local 
+mechanical influence of vessels on both global and local behaviour of the organ has been measured and modelled previously~\cite{Peterlik2012}, only local 
 effects have been studied in the case of Glisson's capsule~\cite{Hollenstein2006,Ahn2010}.
 
-In this paper we introduce what we believe is the first real-time model of liver taking into account all three constituents, each being simulated with 
-a different type of finite elements.
-Using the model, we focus on the role of the Glisson's capsule: first, we show that the model is able to reproduce correctly
+In this paper we introduce what we believe is the first real-time model of liver taking into account all three constituents, each being simulated 
+employing a different finite element formulation.
+Using the proposed model, we focus on the role of the Glisson's capsule: first, we show that the model is able to reproduce correctly
 the local effects of the capsule in so called aspiration test. Second, based on CT image data of porcine abdominal cavity acquired in supine and flank 
@@
 According to the statistics, nearly 100,000 European citizens die of cirrhosis or liver cancer each year. 
-Although new methods such as radio-frequency- or cryo-ablation used in the interventional radiology 
-seems to be promising, surgery remains the option that offers the foremost success rate against these pathologies. 
+Although new methods such as radio-frequency- and cryo-ablation known in the interventional radiology 
+seem to be promising, surgery remains the option that offers the foremost success rate against these pathologies. 
 Nevertheless, surgery is not always performed due to several limitations, in particular the determination 
@@
 
-In case of human liver, the task of real-time accurate modeling is challenging, mainly because of the complex structure 
+When simulating the behavior of human liver, the task of real-time accurate modeling is challenging, mainly because of the complex structure 
 of this organ composed of three main constituents: \emph{parenchyma}, \emph{vascular networks} and \emph{Glisson's capsule}.
-The parenchyma has certainly been the most studied component of the liver; today researchers agree on hyperelastic 
+The parenchyma has certainly been the most studied component of the liver; actually researchers agree on hyperelastic 
 properties of the tissue, for which the mechanical parameters have been reported for example in~\cite{Kerdok2006,Gao2009}. 
-Several methods have been proposed to model the hyperelastic behaviour at real-time rates, such as multiplicative Jacobian decomposition
+Moreover, several methods have been proposed to model the hyperelastic behaviour at real-time rates, such as multiplicative Jacobian decomposition
 introduced in~\cite{Marchesseau2010}.
@@
 Since a detailed modeling of the vessels would be extremely costly (mainly because of the small thickness of the vessel wall, 
-the authors propose a composite model allowing for real-time simulation of entire liver with vasculature.
+the authors propose a composite model allowing for real-time simulation of entire liver with vascularization.
 
-Rather a small number of studies have been conducted on the last liver constituent, the  Glisson's capsule.
+Rather a small number of studies have been conducted dealing with the third liver constituent, the  Glisson's capsule.
 Quantitative results of experiments on a porcine liver have been published in~\cite{Umale2011}; the measurements indicate that although being very 
-thin (10--20$\mu m$), the capsule shows to be stiff in tensile tests: the Young's modulus of the capsule (reported as $8.22\pm3.42$\,MPa) 
+thin (10--20$\mu m$), the capsule shows to be stiff in tensile tests: the Young's modulus of the capsule reported to be $8.22\pm3.42$\,MPa 
 exceeds the values for the parenchyma by three orders of magnitude.
@@
 
-The main contribution of the paper is twofold: first, we present an efficient real-time model of the liver, taking into account 
+The main contribution of the paper is twofold: first, we present a real-time model of the liver, including
 the parenchyma, vascularization and Glisson's capsule. The model is based on three different finite element representations for each constituent,
@@
 capsule via simulation: using a specific model of a porcine liver built from CT contrast-enhanced data, we show that there is a significant 
-difference in the response of the model with and without the capsule in case when the liver undergoes large deformations.
+difference in the response of the model with and without the capsule in the case when the liver undergoes large deformations.
 
@@
 liver model with capsule. Second, we validate our model in context of local deformations by reproducing the aspiration test 
-described in~\cite{Hollenstein2006}. Finally, we demonstrate that in spite of it's small thickness, the Glisson's 
+described in~\cite{Hollenstein2006}. Finally, we demonstrate that in spite of its small thickness, the Glisson's 
 capsule has a global influence on the liver undergoing large deformations.
@@
 
-In this section we describe the construction of our composite model
-being composed of three main components: tetrahedral FE model of the 
+In this section we describe the construction of a composite model
+based on three components: tetrahedral FE model of the 
 parenchyma, Tymoshenko-beam model for the vascular structures and finally 
@@
 The parenchyma is modeled using corotational finite elements~\cite{Felippa2005}.
-While relying on linear stress-strain relationship, it allows for large displacements including rotations. 
-While in the full non-linear formulation, the stiffness matrix relates the forces $\Vec{f}$ and 
+Although it relies on linear stress-strain relationship, large displacements including rotations are modelled correctly. 
+While in the full non-linear formulation the stiffness matrix relates the forces $\Vec{f}$ and 
 displacements $\Vec{u}$ as $\Vec{f} = \Mat{K}(\Vec{u})$, the corotational model 
@@
 The model of the vascularization is based on linear beams 
-with local frames of reference; in many aspects it's similar to the
-corotational formulation described above~\cite{Duriez2006}. As such the model also handles geometric
+with local frames of reference~\cite{Duriez2006}; in many aspects it's similar to the
+corotational formulation described above. As such the model also handles geometric
 non-linearities in the deformation. Through a specification of cross section and moments of inertia, 
@@
 The beam-based model of vessels is mechanically coupled to the parenchyma as described in~\cite{Peterlik2012}. 
-The coupling assumes that there is no relative motion between the vessels and surrounding parenchyma. 
+The coupling assumes that there is no relative motion between the vessels and the surrounding parenchyma. 
 %The proposed model assumes no relative
@@
 \label{ss:capsuleModel}
-The thickness of the Glisson's capsule is relatively small: values in range of 10-20
+The thickness of the Glisson's capsule is relatively small: the values in range of 10-20
 $\mu$m have been reported in~\cite{Umale2011}.
 It is not possible to model such thin structure with classical tetrahedral
-elements, if real-time aspect of the simulation is to be preserved.
+elements, if the real-time aspect of the simulation is to be preserved.
 Furthermore, modeling both the tissue and the capsule would require an extremely 
 dense mesh to avoid numerical instabilities and thus would significantly
-violate the speed requirements required for medical simulators.
+violate the speed requirements imposed for medical simulators.
 Instead, modeling the capsule with two-dimensional elements that abstract from the
@@
 to be a natural choice. In the elasticity theory, this functionality is usually provided by membrane and shell elements.
-Based on the small thickness (compared to the surface area) we also
+Based on the observation of its behaviour, we also
 assume negligible bending forces and propose a model based on membrane
@@
 We use linear elastic material and employ the corotational formulation
-for the triangular elements.
+for the CST elements.
 
@@
 Moreover, the stiffness matrices for capsule and parenchyma are then easily assembled
-together.% and solved as one system.
+together.
+% and solved as one system.
 %
@@
 The model presented in the previous section was implemented in 
-SOFA\footnote{www.sofa-framework.org} framework and a set of
-numerical simulations was performed. First local deformations were compared to the
+SOFA\footnote{www.sofa-framework.org} and a set of
+numerical simulations was performed. First, local deformations were compared to the
 results reported in literature to validate the method.
@@
 	\subfigure[]{\includegraphics[width=7cm]{aspiration.pdf}}
-  \caption{\label{fig-aspiration}Aspiration test: (a) the SOFA simulation scene, (b) displacment profile
-  with (blue) and without (black) the capsule.}
+  \caption{\label{fig-aspiration}Aspiration test: (a) the SOFA simulation scene, (b) cube displacement profile
+  with the capsule (in blue) and without capsule (in black).}
 \end{figure}
@@
 %We reproduced the test with our model and acquired similar results.
-The aspiration device consists of a tube 1\,cm in diameter and is able
+The aspiration device consists of a tube 1\,cm in diameter and allows
 to control internal pressure in the tube. The test is performed by
 attaching the tube to the tissue and measuring the tissue response. We
-have set up a simulation in SOFA to reproduce the experiment (see Fig.~\ref{fig-aspiration}a: we meshed a 15$\times$15$\times$15\,mm$^3$ 
+set up a simulation in SOFA to reproduce the experiment (see Fig.~\ref{fig-aspiration}a): we meshed a 15$\times$15$\times$15\,mm$^3$ 
 cube representing the tissue 
@@
 capsule can be used to reproduce the \emph{local} behaviour of the tissue. In the following, 
-first we present a real-time complete model of entire liver built from image data. 
+we first present a real-time complete model of entire liver built from image data. 
 Second, using this model we demonstrate the \emph{global} mechanical influence of the capsule when 
@@
 The complete model was constructed using contrast-enhanced CT data acquired on a female pig 
-being positioned first in supine and then flank configuration. 
+being positioned first in supine and then flank position. 
 The liver and hepatic vein were segmented for both datasets  using semi-automatic methods available in ITKSnap~\footnote{www.itksnap.org}.
@@
       (b) liver in supine (yellow) and flank (red) position;
-      (c) distance map between registered version with and without capsule.}
+      (c) Hausdorff distance map between registered version with and without capsule.}
 \end{figure}
@@
 we selected points such as bifurcations which can be identified easily in both supine and flank 
-data providing a set of correspondences. In the next step, we used these correspondences as a non-homogeneous
+data resulting in a set of correspondences. In the next step, we used these correspondences as a non-homogeneous
 boundary conditions (imposed via penalty term) in order to perform a deformable registration of the supine
@@
 to perform an accurate registration of the flank and supine data, but the simulation was used 
-as an example of real deformation of the organ.
+as an example of a real deformation of the organ.
 
@@
 maximal difference exceeding 1.25\,cm was measured. We believe that this result confirms that the Glisson's capsule 
-plays an important role in the global mechanical response of the liver.
+plays an important role in the global mechanical behaviour of the liver.
 
@@
 three constituents of the organ, each being modelled with a different type 
-of finite elements: tetrahedral corotational model used the parenchyma, serially 
-linked corotational beams based on Tymoshenko formulation modelling the vascularization 
-and finally triangular elements with constant strain employed for the capsule. 
+of finite elements: tetrahedral corotational model used for the parenchyma, serially 
+linked corotational beams based on Tymoshenko formulation employed to model the vascularization 
+and finally CST elements used for the capsule. 
 The model was first validated for local displacements: the aspiration test presented in~\cite{Hollenstein2006} 
@@
 
-In the future work, we plan to employ the model to model other organs where the capsule play an important role
-(for example kidneys). We will also focus on better understanding the interaction between the organs in the abdominal 
-cavity.
+In the future work, we plan to employ the proposed model for other organs where the capsule play an important role
+(for example kidneys). We will also focus on better description of the mechanical interaction between the organs in the abdominal 
+cavity employing a realistic modeling of constraints and contacts.
 

### 2013-05-21 Doc/Vrst2013/vrst2013.tex
Paper for VRST 2013 (reworked miccai 2013 paper)
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@573 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2013-03-02 Doc/Miccai2013/miccai2013.tex
@@
-% http://www.miccai2013.org/submission_guideline.html
-\documentclass{llncs}
+% REMEMBER: After having produced the .bbl file, and prior to final submission,
+% you need to 'insert'  your .bbl file into your source .tex file so as to provide
+% ONE 'self-contained' source file.
+
+\documentclass{acm_proc_article-sp}
 
 \usepackage{amsmath}
-\usepackage{makeidx}
-\usepackage[pdftex]{graphicx}
-\usepackage{subfigure}
+%\usepackage{makeidx}
+%\usepackage[pdftex]{graphicx} ... we have to use ps/eps
+\usepackage{subfig}
 
@@
 \begin{document}
-%
-%\frontmatter          % for the preliminaries
-%
-\mainmatter              % start of the contributions
-%
+
 \title{Complete Real-Time Liver Model Including Glisson's Capsule, Vascularization and Parenchyma}
-%
-\titlerunning{Complete Liver Model}  % abbreviated title (for running head)
-%                                     also used for the TOC unless
-%                                     \toctitle is used
-%
-\author{Anonymous}
-%\author{%
-%Tom\'a\v{s} Golembiovsk\'y\inst{1,2} \and%
-%Igor Peterl\'ik\inst{3} \and%
-%Christian Duriez\inst{2,4} \and%
-%Stephan\'e Cotin\inst{2}%
-%}
-%
-%\authorrunning{Tom\'a\vs Golembiovsk\'y et al.} % abbreviated author list (for running head)
-%
-%\institute{%
-%Faculty of Informatics, Masaryk University, Brno, Czech Republic\and%
-%INRIA Lille -- Nord, France\and%
-%Institut Hospitallo-Universitaire, Strasbourg, France \and%
-%University of Lille, Lille, France%
-%}
-%
+%\subtitle{}
+
+\numberofauthors{4}
+\author{
+% 1st. author
+\alignauthor
+Tom\'a\v{s} Golembiovsk\'y\\
+       \affaddr{Faculty of Informatics}\\
+       \affaddr{Masaryk University}\\
+       \affaddr{Brno, Czech Republic}\\
+       \email{nyoxi@ics.muni.cz}
+% 2nd. author
+\alignauthor
+Igor Peterl\'ik\\
+       \affaddr{Institut Hospitallo-Universitaire}\\
+       \affaddr{Strasbourg, France}\\
+       \email{...} % TODO
+% 3rd. author
+\alignauthor
+Christian Duriez\\
+       \affaddr{INRIA Lille}\\
+       \affaddr{Lille, France}\\
+       \email{...} % TODO
+\and
+% 4th. author
+\alignauthor
+Stephan\'e Cotin\\
+       \affaddr{Institut Hospitallo-Universitaire}\\
+       \affaddr{Strasbourg, France}\\
+       \email{...} % TODO
+}
+
+\date{30 May 2013}
+
 \maketitle
-
 \begin{abstract}
@@
 
-\keywords{deformable liver model, Glisson's capsule, vascularization, real-time soft tissue modeling}
 \end{abstract}
 
-\section{Introduction} 
+% TODO
+% A category with the (minimum) three required fields
+\category{H.4}{Information Systems Applications}{Miscellaneous}
+%A category including the fourth, optional field follows...
+\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]
+
+\terms{Theory} %TODO
+
+% NOT required for Proceedings
+\keywords{deformable liver model, Glisson's capsule, vascularization, real-time soft tissue modeling}
+
+\section{Introduction} %{{{
+
 According to the statistics, nearly 100,000 European citizens die of cirrhosis or liver cancer each year. 
@@
 
-\begin{figure}[t]
-    	\subfigure[]{\includegraphics[height=3.5cm]{aspiration.jpg}}
-	\subfigure[]{\includegraphics[width=7cm]{aspiration.pdf}}
-  \caption{\label{fig-aspiration}Aspiration test: (a) the SOFA simulation scene, (b) cube displacement profile
-  with the capsule (in blue) and without capsule (in black).}
+\begin{figure}
+  \includegraphics[height=3.5cm]{aspiration.jpg}
+  \caption{\label{fig-aspiration1} The SOFA simulation scene of the aspiration test.}
+\end{figure}
+
+\begin{figure}
+  \includegraphics[width=7cm]{aspiration.pdf}
+  \caption{\label{fig-aspiration2} Displacement profile of the cube in the
+  aspiration test.}
 \end{figure}
@@
 attaching the tube to the tissue and measuring the tissue response. We
-set up a simulation in SOFA to reproduce the experiment (see Fig.~\ref{fig-aspiration}a): we meshed a 15$\times$15$\times$15\,mm$^3$ 
+set up a simulation in SOFA to reproduce the experiment (see
+Fig.~\ref{fig-aspiration1}: we meshed a 15$\times$15$\times$15\,mm$^3$ 
 cube representing the tissue 
@@
 
-In Fig.~\ref{fig-aspiration}b the profiles of cuts in the middle of the
+In Fig.~\ref{fig-aspiration2} the profiles of cuts in the middle of the
 test cube are presented showing significantly larger deformation in the model without capsule. This is in perfect 
@@
 
-\begin{figure}[t]
+\begin{figure*}
   \centering
@@
 towards the flank shape, and the stiffness of the capsule is increased from 0 to 8.22\,MPa after the initial equilibrium is achieved.}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}[t]
+\begin{figure*}
   \centering
-    \subfigure[]{\includegraphics[width=.32\textwidth]{gravity2.png}}
-    \subfigure[]{\includegraphics[width=.32\textwidth]{registered.png}}
-    \subfigure[]{\includegraphics[width=.32\textwidth]{distanceMap.png}}
+    \subfloat[]{\includegraphics[width=.32\textwidth]{gravity2.png}}
+    \subfloat[]{\includegraphics[width=.32\textwidth]{registered.png}}
+    \subfloat[]{\includegraphics[width=.32\textwidth]{distanceMap.png}}
     \caption{\label{f:global}%
@@
       (c) Hausdorff distance map between registered version with and without capsule.}
-\end{figure}
+\end{figure*}
 
@@
 In the future work, we plan to employ the proposed model for other organs where the capsule play an important role
-(for example kidneys). We will also focus on better description of the mechanical interaction between the organs in the abdominal 
-cavity employing a realistic modeling of constraints and contacts.
+(for example kidneys). We will also focus on better description of the mechanical interactions and couplings among the organs and tissues in the abdominal 
+cavity via realistic modeling of boundary conditions using mechanical constraints and contacts.
 
@@
 
-\bibliographystyle{splncs03}
+\bibliographystyle{abbrv}
 \bibliography{bibdata}
@@
 % spell: setl spell spelllang=en spellfile=spelldict.en.utf-8.add
-% vim:set et sw=2 tw=75 fdm=marker fdl=3 fdc=4 isk+=_,-:
+% vim:set et sw=2 fdm=marker fdl=3 fdc=4 isk+=_,-:

### 2013-05-21 Doc/Vrst2013/aspiration.R

@@ -1,0 +1,21 @@
+a1 <- data.matrix(read.csv("aspiration-simple.csv"))
+a2 <- data.matrix(read.csv("aspiration-composite.csv"))
+
+a1 <- a1[1:31,c(4,3)]; a1[,2] <- a1[,2] - 0.015; a1[,2] <- a1[,2] * 1000
+a2 <- a2[1:33,c(4,3)]; a2[,2] <- a2[,2] - 0.015; a2[,2] <- a2[,2] * 1000
+
+pdf("aspiration.pdf", width=3, height=1.5, pointsize=8); # sizes in inches
+p <- par(lwd=0.5, mar=c(4,4,0,0)+0.1);
+
+plot(range(c(a1[,1], a2[,1])), range(c(a1[,2], a2[,2])), #ylim=c(0, 0.004), 
+     type="n", xlab="x-coordinate", ylab="Displacement [mm]")
+
+lines(a1, col="black")
+lines(a2, col="blue")
+
+#abline(h=0, lty="dotted");
+
+#legend("topright", c("A","B", "C", "D"), pch=c(1,4,5,6), col="black");
+
+dev.off()
+par(p);

### 2013-05-28 Doc/Vrst2013/vrst2013.tex
FIX: caption bug and affiliations
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@574 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
removing vascularization and global deformations
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@575 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
+\usepackage{caption}
+\DeclareCaptionType{copyrightbox}
 \usepackage{amsmath}
@@
 Igor Peterl\'ik\\
-       \affaddr{Institut Hospitallo-Universitaire}\\
+       \affaddr{Institut Hospitalo-Universitaire}\\
        \affaddr{Strasbourg, France}\\
@@
 Christian Duriez\\
-       \affaddr{INRIA Lille}\\
+       \affaddr{INRIA}\\
        \affaddr{Lille, France}\\
@@
 Stephan\'e Cotin\\
-       \affaddr{Institut Hospitallo-Universitaire}\\
-       \affaddr{Strasbourg, France}\\
+       \affaddr{INRIA}\\
+       \affaddr{Lille, France}\\
        \email{...} % TODO
@@
 The mechanical importance of the vascular structures in liver is studied in~\cite{Peterlik2012}. It shows that the 
-influence of vessels on the mechanical behaviour of the organ is significant, mainly if large deformations occur. 
+influence of vessels on the mechanical behaviour of the organ is significant. %, mainly if large deformations occur. 
+%Since a detailed modeling of the vessels would be extremely costly (mainly because of the small thickness of the vessel wall, 
+%the authors propose a composite model allowing for real-time simulation of entire liver with vascularization.
 Since a detailed modeling of the vessels would be extremely costly (mainly because of the small thickness of the vessel wall, 
-the authors propose a composite model allowing for real-time simulation of entire liver with vascularization.
+the authors propose a composite model allowing for real-time simulation.
 
@@
 in vivo on human patients during the operation, confirming the mechanical importance of the membrane~\cite{Ahn2010,Nava2008}.
-To our best knowledge, no attempt has been made to demonstrate the role of the Glisson's capsule on the global behaviour of 
-the liver, mainly if the organ undergoes large deformations which is often the case during the surgery.
+%To our best knowledge, no attempt has been made to demonstrate the role of the Glisson's capsule on the global behaviour of 
+%the liver, mainly if the organ undergoes large deformations which is often the case during the surgery.
 
-The main contribution of the paper is twofold: first, we present a real-time model of the liver, including
-the parenchyma, vascularization and Glisson's capsule. The model is based on three different finite element representations for each constituent,
-linked together via mechanical coupling. We show that the model mimics the \emph{local} experiments described in~\cite{Hollenstein2006}.
+In this paper we present preliminary results of our work on the complete liver
+model. We present a real-time composite model accounting for parenchyma and
+Glisson's capsule. The model is based on two different finite element
+representations for each constituent coupled together.
+%The main contribution of the paper is twofold: first, we present a real-time model of the liver, including
+%the parenchyma, vascularization and Glisson's capsule. The model is based on three different finite element representations for each constituent,
+%linked together via mechanical coupling.
+We show that the model mimics the \emph{local} experiments described in~\cite{Hollenstein2006}.
 
-Second, we use the complete model of the liver to demonstrate the \emph{global} influence of the Glisson's
-capsule via simulation: using a specific model of a porcine liver built from CT contrast-enhanced data, we show that there is a significant 
-difference in the response of the model with and without the capsule in the case when the liver undergoes large deformations.
+%Second, we use the complete model of the liver to demonstrate the \emph{global} influence of the Glisson's
+%capsule via simulation: using a specific model of a porcine liver built from CT contrast-enhanced data, we show that there is a significant 
+%difference in the response of the model with and without the capsule in the case when the liver undergoes large deformations.
 
-The paper is organized as follows: first we describe the proposed vascularized
-liver model with capsule. Second, we validate our model in context of local deformations by reproducing the aspiration test 
-described in~\cite{Hollenstein2006}. Finally, we demonstrate that in spite of its small thickness, the Glisson's 
-capsule has a global influence on the liver undergoing large deformations.
-
-%Then we show the correct behaviour of the
-%model in context of \emph{local} deformations and finally we analyze the influence of
-%the capsule on the \emph{global} deformations of the liver and show that despite the
-%small thickness it plays an important role also in the \emph{global} context.
-
-%Medical simulation is expected to play an integral part in many aspects of
-%medical practice. From diagnosis, through planning or training to computer
-%assisted intervention.
-%Medical training application, like the ones for training laparoscopic
-%surgery, have to operate in real-time and a good medical model has to be
-%simple yet efficient to be really usable in this environment.
-%
-%Liver is largest internal organ formed mostly by a soft tissue (parenchyma)
-%that is hyper-elastic and also visco-elastic. The parenchyma is interwoven
-%by dense vascular network, this tree-like tubular structure is stiffer then
-%the parenchyma. Finally the parenchyma is covered by very stiff collagenous
-%membrane that preserves the integrity of the organ.
-%
-%Strangely, most of the work on liver modelling only considers the
-%parenchyma and model the liver as single homogeneous object \TG{Refs}.
-%Only recently
-%Peterl\'{i}k et al. \cite{Peterlik2012} showed the importance of the vascular
-%structures on the liver deformation.
-%
-%Another overlooked component is the Glisson's capsule.
-%Biomechanical measurements report that the capsule has very small
-%thickness (approx. 20 $\mu$m for porcine liver \cite{Umale2011}).
-%On the other hand it has in three orders of magnitude higher stiffness than
-%the parenchyma.
-%Because of this it plays an important role in local
-%deformations of the liver \cite{Ahn2010,Hollenstein2006}.
-%It was reported that evaluation of the liver as a homogeneous material
-%without capsule can lead to overestimation of local deformations up to the factor of 3
-%\cite{Hollenstein2006}.
-%
-%While the influence on local deformation was reported, no experimental
-%results show whether the capsule also plays an important role in global
-%deformations of the liver. We propose a composite model that models the
-%capsule with membrane FEM elements. We show that the model is able to
-%simulate expected local behaviour reported in literature. Further more we
-%show that the capsule plays also an important role in the deformation on
-%global scale.
+The paper is organized as follows: first we describe the proposed model of
+liver with capsule. Second, we validate our model in context of local
+deformations by reproducing the aspiration test described
+in~\cite{Hollenstein2006}. We conclude by summarizing future steps towards the
+complete liver model.
+%Finally, we demonstrate that in spite of its small thickness, the Glisson's 
+%capsule has a global influence on the liver undergoing large deformations.
 
@@
 In this section we describe the construction of a composite model
-based on three components: tetrahedral FE model of the 
-parenchyma, Tymoshenko-beam model for the vascular structures and finally 
+based on two components: tetrahedral FE model of the 
+parenchyma and
 triangular membrane elements used for the capsule.
@@
 While in the full non-linear formulation the stiffness matrix relates the forces $\Vec{f}$ and 
-displacements $\Vec{u}$ as $\Vec{f} = \Mat{K}(\Vec{u})$, the corotational model 
+displacements $\Vec{u}$ as $\Vec{f} = K(\Vec{u})$, the corotational model 
 requires the stiffness matrix $\Mat{K}_0$ of the system to be computed only once before the simulation begins. 
@@
 whereas the deformations are used to compute the linear strain in the local corotational frame.
-%The stiffness matrix $\Mat{K}$ can be expressed in terms of the deformation
-%$u$ and the force as $\Vec{f} = \Mat{K}(\Vec{u}) \Vec{u}$. 
-%Where the
-%components of $\Mat{K}$ depend on the orientation of each element in the
-%current simulation step.
-%the rigid component is  
-% The corotational
-%formulation allows us to deal with geometrical non-linearities but still
-%relies on linear stress-strain relationship. Thus rotations and large
-%deformations are modelled correctly. 
-%In the corotational formulation the deformation
-%of each element is expressed in the local frame of reference. 
 There are several ways of computing the corotational frame for elements; we rely on
@@
 
-The model of the vascularization is based on linear beams 
-with local frames of reference~\cite{Duriez2006}; in many aspects it's similar to the
-corotational formulation described above. As such the model also handles geometric
-non-linearities in the deformation. Through a specification of cross section and moments of inertia, 
-the model can account for the specific properties of the blood vessels.
+%The model of the vascularization is based on linear beams 
+%with local frames of reference~\cite{Duriez2006}; in many aspects it's similar to the
+%corotational formulation described above. As such the model also handles geometric
+%non-linearities in the deformation. Through a specification of cross section and moments of inertia, 
+%the model can account for the specific properties of the blood vessels.
 
-The beam-based model of vessels is mechanically coupled to the parenchyma as described in~\cite{Peterlik2012}. 
-The coupling assumes that there is no relative motion between the vessels and the surrounding parenchyma. 
-%The proposed model assumes no relative
-%motion of vessels to the parenchyma. Segmented vascular network is
-%discretized into the set of linear elements that are then projected into
-%the tetrahedral mesh of the parenchyma. 
-During the simulation, the nodes of linked beams are first displaced and rotated according to the actual motion of associated tetrahedra. 
-As the deformation of beams results in mechanical response represented by forces and torques, these are propagated back to 
-the tetrahedral FE model.
-
-%a displacement of tetrahedral nodes are mapped  forces and torques resulting from the deformations of vessels by parenchyma are
-%computed and propagated back onto the tetrahedral mesh.
+%The beam-based model of vessels is mechanically coupled to the parenchyma as described in~\cite{Peterlik2012}. 
+%The coupling assumes that there is no relative motion between the vessels and the surrounding parenchyma. 
+%During the simulation, the nodes of linked beams are first displaced and rotated according to the actual motion of associated tetrahedra. 
+%As the deformation of beams results in mechanical response represented by forces and torques, these are propagated back to 
+%the tetrahedral FE model.
 
@@
 
-We use linear elastic material and employ the corotational formulation
-for the CST elements.
+Similarly as with model of parenchyma we use linear elastic material and employ
+the corotational formulation for the CST elements.
 
@@
 
+\TG{We can also mention the mechanical mapping (like the one for vessels) as an
+alternative approach. The result is same for conforming surface mesh.}
+
 %}}}
@@
 SOFA\footnote{www.sofa-framework.org} and a set of
-numerical simulations was performed. First, local deformations were compared to the
+numerical simulations was performed.
+In this section we provide comparison of local deformations with the
 results reported in literature to validate the method.
-Second, to show that in spite of its very small thickness the membrane cannot be
-neglected even in the context of global deformations and its overall
-stiffness plays an important role, the model of
-the complete liver was subjected to global deformations.
+%Second, to show that in spite of its very small thickness the membrane cannot be
+%neglected even in the context of global deformations and its overall
+%stiffness plays an important role, the model of
+%the complete liver was subjected to global deformations.
+
+\TG{Here we have to decide wether to keep the values below as well as the Fig.
+2 and specify the values used to match Hollenstein separately. Or if we take
+just one set of values and redo the figure.}
 
@@
 
-\begin{figure*}
-  \centering
-    \includegraphics[width=.95\textwidth]{forceEvolution.pdf}
-  \caption{\label{f:forceEvol}Evolution of forces during the simulation, where the supine data is deformed
-towards the flank shape, and the stiffness of the capsule is increased from 0 to 8.22\,MPa after the initial equilibrium is achieved.}
-\end{figure*}
+%\begin{figure*}
+%  \centering
+%    \includegraphics[width=.95\textwidth]{forceEvolution.pdf}
+%  \caption{\label{f:forceEvol}Evolution of forces during the simulation, where the supine data is deformed
+%towards the flank shape, and the stiffness of the capsule is increased from 0 to 8.22\,MPa after the initial equilibrium is achieved.}
+%\end{figure*}
 
-\subsection{Global Deformations}
-In the previous section we have shown that the composite model of parenchyma and 
-capsule can be used to reproduce the \emph{local} behaviour of the tissue. In the following, 
-we first present a real-time complete model of entire liver built from image data. 
-Second, using this model we demonstrate the \emph{global} mechanical influence of the capsule when 
-the model undergoes large deformations.
+%\subsection{Global Deformations}
+%In the previous section we have shown that the composite model of parenchyma and 
+%capsule can be used to reproduce the \emph{local} behaviour of the tissue. In the following, 
+%we first present a real-time complete model of entire liver built from image data. 
+%Second, using this model we demonstrate the \emph{global} mechanical influence of the capsule when 
+%the model undergoes large deformations. 
+%
+%The complete model was constructed using contrast-enhanced CT data acquired on a female pig 
+%being positioned first in supine and then flank position. 
+%The liver and hepatic vein were segmented for both datasets  using semi-automatic methods available in ITKSnap~\footnote{www.itksnap.org}.
+%Further, an unstructured mesh was obtained from the segmented maps using CGAL~\footnote{www.cgal.org}. 
+%The semi-automatic method presented in~\cite{Peterlik2012} was 
+%then applied to the segmented map of the hepatic vein, resulting in linked-beam model. 
+%Finally, the model of the Glisson's capsule was constructed using the elements described in 
+%section~\ref{ss:capsuleModel} over the surface triangles extracted from the tetrahedral mesh. 
+%The complete model composed of 4266 tetrahedra, 188 beams and 1284 triangles was then 
+%used in simulation where external force representing the gravity was applied to the model, 
+%resulting in large deformations mainly in the area of lobes. The refresh rate of 22\,FPS 
+%was achieved on a desktop computer with Intel i7 CPU running at 3.4\,GHz and 8\,GB of memory. 
+%The simulation with gravity was also used as a first demonstration of the global influence of the capsule. 
+%In Fig.~\ref{f:global}a, the models with capsule (red) and without (yellow) are depicted after being 
+%deformed due to the gravity, showing a significant difference between the two models.  
+%
+%\begin{figure*}
+%  \centering
+%    \subfloat[]{\includegraphics[width=.32\textwidth]{gravity2.png}}
+%    \subfloat[]{\includegraphics[width=.32\textwidth]{registered.png}}
+%    \subfloat[]{\includegraphics[width=.32\textwidth]{distanceMap.png}}
+%    \caption{\label{f:global}%
+%      (a) Liver under gravity with (red) and without (yellow) capsule;
+%      (b) liver in supine (yellow) and flank (red) position;
+%      (c) Hausdorff distance map between registered version with and without capsule.}
+%\end{figure*}
+%
+%In order to demonstrate the influence of the capsule on a real deformation of the liver, we 
+%performed a following protocol employing the supine and flank models built from the CT data. 
+%First, we performed a rigid registration of the two organs \wrt\ the root point of the hepatic vein. 
+%As shown in Fig.~\ref{f:global}b, the liver undergoes very large deformation when the body 
+%is turned from the supine to the flank position, exceeding 6\,cm for some parts of the organ. 
+%In order to induce similar deformation, we extracted feature pairs from the vascular system of the liver:
+%we selected points such as bifurcations which can be identified easily in both supine and flank 
+%data resulting in a set of correspondences. In the next step, we used these correspondences as a non-homogeneous
+%boundary conditions (imposed via penalty term) in order to perform a deformable registration of the supine
+%model onto the flank data. During the registration, the response forces acting in the points with prescribed displacements
+%were calculated. It is important to emphasize that the goal of the simulation was not 
+%to perform an accurate registration of the flank and supine data, but the simulation was used 
+%as an example of a real deformation of the organ.
+%
+%The simulation was performed with the model composed of parenchyma,  vessels and capsule, however, zero stiffness was 
+%set for the last component.
+%Once the static equilibrium was achieved, the effect of the capsule was gradually switched 
+%on, \ie\ the stiffness of the membrane was slowly increased from 0 to 8.22\,MPa. The evolution of the force response in the 
+%points with prescribed displacements is depicted in Fig.~\ref{f:forceEvol} showing the change in the response as the effect
+%of the capsule is included. At the same time, the shape of the deformed liver with capsule was compared 
+%to the model without the capsule. A Hausdorff distance map for those cases is depicted in Fig.~\ref{f:global}c: in the red areas, the 
+%maximal difference exceeding 1.25\,cm was measured. We believe that this result confirms that the Glisson's capsule 
+%plays an important role in the global mechanical behaviour of the liver.
 
-The complete model was constructed using contrast-enhanced CT data acquired on a female pig 
-being positioned first in supine and then flank position. 
-The liver and hepatic vein were segmented for both datasets  using semi-automatic methods available in ITKSnap~\footnote{www.itksnap.org}.
-Further, an unstructured mesh was obtained from the segmented maps using CGAL~\footnote{www.cgal.org}. 
-The semi-automatic method presented in~\cite{Peterlik2012} was 
-then applied to the segmented map of the hepatic vein, resulting in linked-beam model. 
-Finally, the model of the Glisson's capsule was constructed using the elements described in 
-section~\ref{ss:capsuleModel} over the surface triangles extracted from the tetrahedral mesh. 
-The complete model composed of 4266 tetrahedra, 188 beams and 1284 triangles was then 
-used in simulation where external force representing the gravity was applied to the model, 
-resulting in large deformations mainly in the area of lobes. The refresh rate of 22\,FPS 
-was achieved on a desktop computer with Intel i7 CPU running at 3.4\,GHz and 8\,GB of memory. 
-The simulation with gravity was also used as a first demonstration of the global influence of the capsule. 
-In Fig.~\ref{f:global}a, the models with capsule (red) and without (yellow) are depicted after being 
-deformed due to the gravity, showing a significant difference between the two models.
+%}}}
 
-\begin{figure*}
-  \centering
-    \subfloat[]{\includegraphics[width=.32\textwidth]{gravity2.png}}
-    \subfloat[]{\includegraphics[width=.32\textwidth]{registered.png}}
-    \subfloat[]{\includegraphics[width=.32\textwidth]{distanceMap.png}}
-    \caption{\label{f:global}%
-      (a) Liver under gravity with (red) and without (yellow) capsule;
-      (b) liver in supine (yellow) and flank (red) position;
-      (c) Hausdorff distance map between registered version with and without capsule.}
-\end{figure*}
+\section{Future Work} %{{{
 
-In order to demonstrate the influence of the capsule on a real deformation of the liver, we 
-performed a following protocol employing the supine and flank models built from the CT data. 
-First, we performed a rigid registration of the two organs \wrt\ the root point of the hepatic vein. 
-As shown in Fig.~\ref{f:global}b, the liver undergoes very large deformation when the body 
-is turned from the supine to the flank position, exceeding 6\,cm for some parts of the organ. 
-In order to induce similar deformation, we extracted feature pairs from the vascular system of the liver:
-we selected points such as bifurcations which can be identified easily in both supine and flank 
-data resulting in a set of correspondences. In the next step, we used these correspondences as a non-homogeneous
-boundary conditions (imposed via penalty term) in order to perform a deformable registration of the supine
-model onto the flank data. During the registration, the response forces acting in the points with prescribed displacements
-were calculated. It is important to emphasize that the goal of the simulation was not 
-to perform an accurate registration of the flank and supine data, but the simulation was used 
-as an example of a real deformation of the organ.
-
-The simulation was performed with the model composed of parenchyma,  vessels and capsule, however, zero stiffness was 
-set for the last component.
-Once the static equilibrium was achieved, the effect of the capsule was gradually switched 
-on, \ie\ the stiffness of the membrane was slowly increased from 0 to 8.22\,MPa. The evolution of the force response in the 
-points with prescribed displacements is depicted in Fig.~\ref{f:forceEvol} showing the change in the response as the effect
-of the capsule is included. At the same time, the shape of the deformed liver with capsule was compared 
-to the model without the capsule. A Hausdorff distance map for those cases is depicted in Fig.~\ref{f:global}c: in the red areas, the 
-maximal difference exceeding 1.25\,cm was measured. We believe that this result confirms that the Glisson's capsule 
-plays an important role in the global mechanical behaviour of the liver.
+\TG{including vascularization; global deformation; stress the complexity of validation of global deformations}
 
@@
 \section{Conclusions} %{{{
+\TG{TODO}
 In the paper, a complete model of the liver was presented taking into account 
@@
 
-%While the influence of the capsule on local deformations was previously
-%studied in the literature, it's significance on the global scale
-%deformations of the liver remained unexplored. We have performed a set of
-%natural global deformations of the complete liver and shown that the
-%capsule, despite its small thickness, plays a significant role also in
-%global context.
-%
 %}}}

### 2013-05-29 Doc/Vrst2013/vrst2013.tex
vrst2013: aspiration test
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@576 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
+% TODO
 % REMEMBER: After having produced the .bbl file, and prior to final submission,
@@
 
-\date{30 May 2013}
+%\date{30 May 2013}
 
@@
 % NOT required for Proceedings
-\keywords{deformable liver model, Glisson's capsule, vascularization, real-time soft tissue modeling}
+%\keywords{deformable liver model, Glisson's capsule, vascularization, real-time soft tissue modeling}
+\keywords{deformable liver model}
 
@@
 
-\TG{We can also mention the mechanical mapping (like the one for vessels) as an
-alternative approach. The result is same for conforming surface mesh.}
+Alternatively, one can use a mechanical coupling similar to the one used in
+\cite{Peterlik2012} to be able to use an arbitrary surface mesh. Nevertheless,
+for conforming triangular mesh both methods lead to the same solution.
 
@@
 
-\section{Experiments} %{{{
+\section{Aspiration Test} %{{{
+
 The model presented in the previous section was implemented in 
@@
 
-\TG{Here we have to decide wether to keep the values below as well as the Fig.
-2 and specify the values used to match Hollenstein separately. Or if we take
-just one set of values and redo the figure.}
-
-In the simulations we used material properties reported in the literature: for the 
-Glisson's capsule we used Young's modulus $E_c$=8.22\,MPa and thickness $t_c$=20\,$\mu$m~\cite{Umale2011},
-whereas for the vessel wall we applied stiffness $E_v$=0.62\,MPa and thickness $t_v$=200\,$\mu$m~\cite{Peterlik2012}.
-Since the values for parenchyma elasticity quite vary (usually between 2\,kPa to 5\,kPa), we perform the simulation 
-with $E_p$=3.5\,kPa.
-
-\begin{figure}
-  \includegraphics[height=3.5cm]{aspiration.jpg}
-  \caption{\label{fig-aspiration1} The SOFA simulation scene of the aspiration test.}
-\end{figure}
-
-\begin{figure}
-  \includegraphics[width=7cm]{aspiration.pdf}
-  \caption{\label{fig-aspiration2} Displacement profile of the cube in the
-  aspiration test.}
-\end{figure}
-
-\subsection{Local Deformations} %{{{
 During the contact with an instrument such as probe, needle, scalpel and others,
@@
 
+\begin{figure}
+  \centering
+  \includegraphics[height=3.5cm]{aspiration.jpg}
+  \caption{\label{fig-aspiration1} The SOFA simulation scene of the aspiration test.}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=7cm]{aspiration.pdf}
+  \caption{\label{fig-aspiration2} Displacement profile of the cube in the
+  aspiration test with (blue) and without (black) the capusle.}
+\end{figure}
+
 A good example of such a local deformation is the aspiration test
@@
 only the parenchyma leads to an overestimation of the deformation. 
-%We reproduced the test with our model and acquired similar results.
 The aspiration device consists of a tube 1\,cm in diameter and allows
-to control internal pressure in the tube. The test is performed by
+control of internal pressure in the tube. The test is performed by
 attaching the tube to the tissue and measuring the tissue response. We
@@
 
+\TG{Here we have to decide wether to keep the values below as well as the Fig.
+2 and specify the values used to match Hollenstein separately. Or if we take
+just one set of values and redo the figure.}
+
+In the simulations we used material properties reported in the literature: for the 
+Glisson's capsule we used Young's modulus $E_c$=8.22\,MPa and thickness $t_c$=20\,$\mu$m~\cite{Umale2011}.
+Since the values for parenchyma elasticity quite vary (usually between 2\,kPa to 5\,kPa), we perform the simulation 
+with $E_p$=3.5\,kPa.
+
 In Fig.~\ref{fig-aspiration2} the profiles of cuts in the middle of the
-test cube are presented showing significantly larger deformation in the model without capsule. This is in perfect 
-agreement with the results reported in~\cite{Hollenstein2006} and it can be concluded that 
-our composite model based on coupling the triangular and tetrahedral elements provides an accurate 
-model of parenchyma and capsule.
+test cube are presented showing significantly larger deformation in the model without capsule.
+We can observere that the deformation without the capsule is indeed overestimated approximately by the factor of 2. This is in perfect agreement with the results reported in~\cite{Hollenstein2006}
+
+Moreover, if we neglect the time-dependent behaviour we are able to reproduce
+the measurements presented in~\cite{Hollenstein2006} with less than 1\% error.
+Simulating the pressure of 3\,MPa and setting the elastic parameters to
+$E_p$=30\,kPa for parenchyma and $E_c$=3\,MPa and $t$=93\,$\mu$m for capsule.
+All the values lie within the measurements specified in the paper.
+\TG{add figure?}
+
+%and it can be concluded that 
+%our composite model based on coupling the triangular and tetrahedral elements provides an accurate 
+%model of parenchyma and capsule.
 

### 2013-05-30 Doc/Vrst2013/vrst2013.tex
vrst2013: more changes
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@577 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 % NOT required for Proceedings
-%\keywords{deformable liver model, Glisson's capsule, vascularization, real-time soft tissue modeling}
+%\keywords{deformable liver model, Glisson's capsule, real-time soft tissue modeling}
 \keywords{deformable liver model}
@@
 Moreover, if we neglect the time-dependent behaviour we are able to reproduce
-the measurements presented in~\cite{Hollenstein2006} with less than 1\% error.
-Simulating the pressure of 3\,MPa and setting the elastic parameters to
-$E_p$=30\,kPa for parenchyma and $E_c$=3\,MPa and $t$=93\,$\mu$m for capsule.
-All the values lie within the measurements specified in the paper.
-\TG{add figure?}
+the measurements presented in~\cite{Hollenstein2006,Nava2008} with less than 1\% error.
+In this case the simulation is performed with the pressure of 30\,kPa and with
+the following elastic parameters: $E_p$=30\,kPa for parenchyma and $E_c$=3\,MPa
+and $t$=93\,$\mu$m for capsule. All the values lie within the measurements
+specified in the paper. \TG{add figure?}
 
@@
 
-\section{Future Work} %{{{
+\section{Present and Future Work} %{{{
 
-\TG{including vascularization; global deformation; stress the complexity of validation of global deformations}
+Presently we are working on the validation of the model also in context of
+global deformations where we believe the capsule plays also an essential role.
+This task is however complicated by the lack of biomechanical literature
+dealing with the response of the capsule.
+%The influence of the capsule on local level has been published in literature
+%before. However, to our knowledge it's influence on global scale deformation
+%has not been studied yet. We believe that the capsule has non-negligible impact
+%also on global level.
+
+Our goal is to create complete liver model suitable for real-time medical
+simulations. For this a model accounting for all three main constituents
+(parenchyma, vascularization and capsule) is necessary. To achieve that we have
+integrated our model with the with the vasularization method proposed
+in~\cite{Peterlik2012}. In our future work we need to validate the model in the
+context of global and local deformations. This task is challenging because
+comparing in vivo measurements with the simulation is difficult due to complex
+boundary conditions. 
+\TG{ex-vivo: no information abou vessels; phantoms: cannot model capsule}
 
@@
 \section{Conclusions} %{{{
-\TG{TODO}
-In the paper, a complete model of the liver was presented taking into account 
-three constituents of the organ, each being modelled with a different type 
-of finite elements: tetrahedral corotational model used for the parenchyma, serially 
-linked corotational beams based on Tymoshenko formulation employed to model the vascularization 
-and finally CST elements used for the capsule. 
-The model was first validated for local displacements: the aspiration test presented in~\cite{Hollenstein2006} 
+
+In the paper we presented a liver model accounting for the Glisson's capsule, a component that is often overlooked. Each constituent being modelled with a different type 
+of finite elements: tetrahedral corotational model used for the parenchyma and
+and CST membrane elements used for the capsule. 
+The model was validated for local displacements based on the the aspiration test presented in~\cite{Hollenstein2006} 
 was successfully reproduced, showing a good agreement between the simulation and experiments.
-Further, the model was used to demonstrate global influence of the capsule when the organ 
-undergoes large deformations.
 
-In the future work, we plan to employ the proposed model for other organs where the capsule play an important role
-(for example kidneys). We will also focus on better description of the mechanical interactions and couplings among the organs and tissues in the abdominal 
-cavity via realistic modeling of boundary conditions using mechanical constraints and contacts.
+%Further, the model was used to demonstrate global influence of the capsule when the organ 
+%undergoes large deformations.
+
+%In the future work, we plan to employ the proposed model for other organs where the capsule play an important role
+%(for example kidneys). We will also focus on better description of the mechanical interactions and couplings among the organs and tissues in the abdominal 
+%cavity via realistic modeling of boundary conditions using mechanical constraints and contacts.
 

### 2013-05-30 Doc/Vrst2013/vrst2013.tex
more fixes in the paper
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@578 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\title{Complete Real-Time Liver Model Including Glisson's Capsule, Vascularization and Parenchyma}
+\title{Modeling Liver with Glisson's Capsule in Real-Time Medical Simulations}
 %\subtitle{}
@@
 \begin{abstract}
+  \TODO{}
 The accurate mechanical modeling of liver is of a paramount interest in simulation-based operation planning 
@@
 
+\TG{The following paragraph is optional. Unnecessary but adds to the global context}
 The mechanical importance of the vascular structures in liver is studied in~\cite{Peterlik2012}. It shows that the 
@@
 
-\subsection{Parenchyma and Vascularization} %{{{
+\subsection{Parenchyma} %{{{
 
@@
 
-\subsection{Coupling Between Capsule and Parenchyma} %{{{
+\subsection{Coupling Between Capsule and\\Parenchyma} %{{{
 The literature reports high cohesion between capsule and parenchyma.
@@
 where the response of liver exposed locally to a negative pressure is measured.
-In~\cite{Hollenstein2006} it is reported that the Glisson's
-capsule has a non-negligible influence on the response of the tissue and modeling
-only the parenchyma leads to an overestimation of the deformation. 
 The aspiration device consists of a tube 1\,cm in diameter and allows
@@
 
-In the simulations we used material properties reported in the literature: for the 
+In the simulations we used material properties for a pig liver reported in the literature: for the 
 Glisson's capsule we used Young's modulus $E_c$=8.22\,MPa and thickness $t_c$=20\,$\mu$m~\cite{Umale2011}.
@@
 
+In~\cite{Hollenstein2006} it is reported that the Glisson's
+capsule has a non-negligible influence on the response of the tissue and modeling
+only the parenchyma leads to an overestimation of the deformation up to the factor of 3. 
 In Fig.~\ref{fig-aspiration2} the profiles of cuts in the middle of the
 test cube are presented showing significantly larger deformation in the model without capsule.
-We can observere that the deformation without the capsule is indeed overestimated approximately by the factor of 2. This is in perfect agreement with the results reported in~\cite{Hollenstein2006}
+We can observere that the deformation without the capsule is indeed overestimated approximately by the factor of 2. This is in good agreement with the results reported in~\cite{Hollenstein2006}
 
@@
 the measurements presented in~\cite{Hollenstein2006,Nava2008} with less than 1\% error.
-In this case the simulation is performed with the pressure of 30\,kPa and with
+In this case the simulation was performed with the pressure of 30\,kPa and with
 the following elastic parameters: $E_p$=30\,kPa for parenchyma and $E_c$=3\,MPa
 and $t$=93\,$\mu$m for capsule. All the values lie within the measurements
-specified in the paper. \TG{add figure?}
+specified in the paper. \TG{add figure like Fig. 2?}
 
@@
 Presently we are working on the validation of the model also in context of
-global deformations where we believe the capsule plays also an essential role.
+global deformations where we believe the capsule also plays an essential role.
 This task is however complicated by the lack of biomechanical literature
@@
 (parenchyma, vascularization and capsule) is necessary. To achieve that we have
-integrated our model with the with the vasularization method proposed
+integrated our model with the vasularization method proposed
 in~\cite{Peterlik2012}. In our future work we need to validate the model in the
@@
 and CST membrane elements used for the capsule. 
-The model was validated for local displacements based on the the aspiration test presented in~\cite{Hollenstein2006} 
+The model was validated in the context for local displacements based on the the aspiration test presented in~\cite{Hollenstein2006}. The aspiration test
 was successfully reproduced, showing a good agreement between the simulation and experiments.

### 2013-05-31 Doc/Vrst2013/vrst2013.tex
reorganizing sections
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@579 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\section{Aspiration Test} %{{{
+\section{Methods and Simulation} %{{{
 
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=7cm]{aspiration.pdf}
-  \caption{\label{fig-aspiration2} Displacement profile of the cube in the
-  aspiration test with (blue) and without (black) the capusle.}
-\end{figure}
-
 A good example of such a local deformation is the aspiration test
@@
 
-\TG{Here we have to decide wether to keep the values below as well as the Fig.
-2 and specify the values used to match Hollenstein separately. Or if we take
-just one set of values and redo the figure.}
-
 In the simulations we used material properties for a pig liver reported in the literature: for the 
@@
 
+%}}}
+
+\section{Results and Discussion} %{{{
+
+\begin{figure}
+  \centering
+  \includegraphics[width=7cm]{aspiration.pdf}
+  \caption{\label{fig-aspiration2} Displacement profile of the cube in the
+  aspiration test with (blue) and without (black) the capusle.}
+\end{figure}
+
 In~\cite{Hollenstein2006} it is reported that the Glisson's
 capsule has a non-negligible influence on the response of the tissue and modeling
-only the parenchyma leads to an overestimation of the deformation up to the factor of 3. 
+only the parenchyma leads to an overestimation of the deformation by a factor of 2 to 3. 
 In Fig.~\ref{fig-aspiration2} the profiles of cuts in the middle of the
 test cube are presented showing significantly larger deformation in the model without capsule.
-We can observere that the deformation without the capsule is indeed overestimated approximately by the factor of 2. This is in good agreement with the results reported in~\cite{Hollenstein2006}
+We can indeed observere that the deformation without the capsule is indeed overestimated approximately by a factor of 2.
+This is in perfect agreement with the results reported in the paper.% ~\cite{Hollenstein2006}
 
-Moreover, if we neglect the time-dependent behaviour we are able to reproduce
+Moreover, we are able to reproduce
 the measurements presented in~\cite{Hollenstein2006,Nava2008} with less than 1\% error.
-In this case the simulation was performed with the pressure of 30\,kPa and with
-the following elastic parameters: $E_p$=30\,kPa for parenchyma and $E_c$=3\,MPa
-and $t$=93\,$\mu$m for capsule. All the values lie within the measurements
-specified in the paper. \TG{add figure like Fig. 2?}
+We do not model the time-dependent behaviour though.
+In this simulation we prescribe the the internal pressure of 30\,kPa. The other
+parameters of the model were obtained like this: First we fix the stiffness of
+parenchyma to $E_p$=30\,kPa to match the scenerio without capsule. Then,
+assuming the thickness $t$=93\,$\mu$m of the capsule (reported as average in
+the paper) we run several simulations with different capsule stiffness $E_c$.
+We have found the lowest error is when $E_c$=3\,MPa. All the stiffness
+parameters lie within the ranges measured in the paper.
 
-%and it can be concluded that 
-%our composite model based on coupling the triangular and tetrahedral elements provides an accurate 
-%model of parenchyma and capsule.
+%In this case the simulation was performed with the pressure of 30\,kPa and with
+%the following elastic parameters: $E_p$=30\,kPa for parenchyma and $E_c$=3\,MPa
+%and $t$=93\,$\mu$m for capsule.
+%All the values lie within the measurements
+%specified in the paper. \TG{add figure like Fig. 2?}
 
-%}}}
-
-%\begin{figure*}
-%  \centering
-%    \includegraphics[width=.95\textwidth]{forceEvolution.pdf}
-%  \caption{\label{f:forceEvol}Evolution of forces during the simulation, where the supine data is deformed
-%towards the flank shape, and the stiffness of the capsule is increased from 0 to 8.22\,MPa after the initial equilibrium is achieved.}
-%\end{figure*}
-
-%\subsection{Global Deformations}
-%In the previous section we have shown that the composite model of parenchyma and 
-%capsule can be used to reproduce the \emph{local} behaviour of the tissue. In the following, 
-%we first present a real-time complete model of entire liver built from image data. 
-%Second, using this model we demonstrate the \emph{global} mechanical influence of the capsule when 
-%the model undergoes large deformations. 
-%
-%The complete model was constructed using contrast-enhanced CT data acquired on a female pig 
-%being positioned first in supine and then flank position. 
-%The liver and hepatic vein were segmented for both datasets  using semi-automatic methods available in ITKSnap~\footnote{www.itksnap.org}.
-%Further, an unstructured mesh was obtained from the segmented maps using CGAL~\footnote{www.cgal.org}. 
-%The semi-automatic method presented in~\cite{Peterlik2012} was 
-%then applied to the segmented map of the hepatic vein, resulting in linked-beam model. 
-%Finally, the model of the Glisson's capsule was constructed using the elements described in 
-%section~\ref{ss:capsuleModel} over the surface triangles extracted from the tetrahedral mesh. 
-%The complete model composed of 4266 tetrahedra, 188 beams and 1284 triangles was then 
-%used in simulation where external force representing the gravity was applied to the model, 
-%resulting in large deformations mainly in the area of lobes. The refresh rate of 22\,FPS 
-%was achieved on a desktop computer with Intel i7 CPU running at 3.4\,GHz and 8\,GB of memory. 
-%The simulation with gravity was also used as a first demonstration of the global influence of the capsule. 
-%In Fig.~\ref{f:global}a, the models with capsule (red) and without (yellow) are depicted after being 
-%deformed due to the gravity, showing a significant difference between the two models.  
-%
-%\begin{figure*}
-%  \centering
-%    \subfloat[]{\includegraphics[width=.32\textwidth]{gravity2.png}}
-%    \subfloat[]{\includegraphics[width=.32\textwidth]{registered.png}}
-%    \subfloat[]{\includegraphics[width=.32\textwidth]{distanceMap.png}}
-%    \caption{\label{f:global}%
-%      (a) Liver under gravity with (red) and without (yellow) capsule;
-%      (b) liver in supine (yellow) and flank (red) position;
-%      (c) Hausdorff distance map between registered version with and without capsule.}
-%\end{figure*}
-%
-%In order to demonstrate the influence of the capsule on a real deformation of the liver, we 
-%performed a following protocol employing the supine and flank models built from the CT data. 
-%First, we performed a rigid registration of the two organs \wrt\ the root point of the hepatic vein. 
-%As shown in Fig.~\ref{f:global}b, the liver undergoes very large deformation when the body 
-%is turned from the supine to the flank position, exceeding 6\,cm for some parts of the organ. 
-%In order to induce similar deformation, we extracted feature pairs from the vascular system of the liver:
-%we selected points such as bifurcations which can be identified easily in both supine and flank 
-%data resulting in a set of correspondences. In the next step, we used these correspondences as a non-homogeneous
-%boundary conditions (imposed via penalty term) in order to perform a deformable registration of the supine
-%model onto the flank data. During the registration, the response forces acting in the points with prescribed displacements
-%were calculated. It is important to emphasize that the goal of the simulation was not 
-%to perform an accurate registration of the flank and supine data, but the simulation was used 
-%as an example of a real deformation of the organ.
-%
-%The simulation was performed with the model composed of parenchyma,  vessels and capsule, however, zero stiffness was 
-%set for the last component.
-%Once the static equilibrium was achieved, the effect of the capsule was gradually switched 
-%on, \ie\ the stiffness of the membrane was slowly increased from 0 to 8.22\,MPa. The evolution of the force response in the 
-%points with prescribed displacements is depicted in Fig.~\ref{f:forceEvol} showing the change in the response as the effect
-%of the capsule is included. At the same time, the shape of the deformed liver with capsule was compared 
-%to the model without the capsule. A Hausdorff distance map for those cases is depicted in Fig.~\ref{f:global}c: in the red areas, the 
-%maximal difference exceeding 1.25\,cm was measured. We believe that this result confirms that the Glisson's capsule 
-%plays an important role in the global mechanical behaviour of the liver.
-
-%}}}
-
-\section{Present and Future Work} %{{{
-
-Presently we are working on the validation of the model also in context of
-global deformations where we believe the capsule also plays an essential role.
-This task is however complicated by the lack of biomechanical literature
-dealing with the response of the capsule.
 %The influence of the capsule on local level has been published in literature
@@
 integrated our model with the vasularization method proposed
-in~\cite{Peterlik2012}. In our future work we need to validate the model in the
-context of global and local deformations. This task is challenging because
-comparing in vivo measurements with the simulation is difficult due to complex
-boundary conditions. 
-\TG{ex-vivo: no information abou vessels; phantoms: cannot model capsule}
+in~\cite{Peterlik2012}. Adding the capsule did not signifficantly affect the
+performance since simulation speed of \TODO{XX FPS}%
+\footnote{\TODO{Measured on...}} was achieved. This suggests
+that the proposed technique is compatible with real-time simulation of whole
+organ while modeling the local properties accurately.
+
+%In our future work we need to validate the model in the
+%context of global and local deformations. This task is challenging because
+%comparing in vivo measurements with the simulation is difficult due to complex
+%boundary conditions. 
+%\TG{ex-vivo: no information abou vessels; phantoms: cannot model capsule}
 
@@
 
-\section{Conclusions} %{{{
+\section{Conclusions and Future Work} %{{{
 
@@
 was successfully reproduced, showing a good agreement between the simulation and experiments.
+
+Presently we are focusing on the validation of the model also in context of
+global deformations of the organ where we believe the capsule also plays an
+essential role. This task is however complicated by the lack of biomechanical
+literature dealing with the response of the capsule.
 

### 2013-05-31 Doc/Vrst2013/vrst2013.tex
ADD: abstract
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@580 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
        \affaddr{Strasbourg, France}\\
-       \email{...} % TODO
+       \email{Igor.Peterlik@ihu-strasbourg.eu} % TODO
 % 3rd. author
@@
        \affaddr{Lille, France}\\
-       \email{...} % TODO
+       \email{Christian.Duriez@inria.fr} % TODO
 \and
@@
        \affaddr{Lille, France}\\
-       \email{...} % TODO
+       \email{Stephane.Cotin@inria.fr} % TODO
 }
@@
 and computer-aided per-operative guidance. However, the realistic simulation of liver behaviour is a challenging task, 
-since the organ is composed of three constituents: parenchyma, vascularization and Glisson's capsule, each having different mechanical properties. While the 
-mechanical influence of vessels on both global and local behaviour of the organ has been measured and modelled previously~\cite{Peterlik2012}, only local 
-effects have been studied in the case of Glisson's capsule~\cite{Hollenstein2006,Ahn2010}.
+since the organ is composed of three constituents: parenchyma, vascularization and Glisson's capsule, 
+each having different mechanical properties.
 
-In this paper we introduce what we believe is the first real-time model of liver taking into account all three constituents, each being simulated 
-employing a different finite element formulation.
-Using the proposed model, we focus on the role of the Glisson's capsule: first, we show that the model is able to reproduce correctly
-the local effects of the capsule in so called aspiration test. Second, based on CT image data of porcine abdominal cavity acquired in supine and flank 
-positions, we demonstrate 
-that the capsule has also a global impact on the mechanical behaviour of the liver in the case when the organ undergoes large deformations.
+While a real-time simulation of vascularized liver has been already described in~\cite{Peterlik2012}, the Glisson's 
+capsule has not been included in the modeling. In~\cite{Hollenstein2006,Ahn2010} it is shown that the capsule 
+plays an important role mainly in the vicinity of the surgical tools. Therefore, the accuracy of the liver response 
+in the vicinity of the surgical tool requires correct modeling of the capsule as a component of the liver model.
+
+The measures perfomed on the capsule show report stiffness which is significanly higher that that of the parenchyma, however, 
+the thickness of the membrane does not exceed 100$\mu$m. For this reasons, it is not possible to model the 
+capsule with standard volume elements usually employed for the parenchyma. In our approach, we rely of 
+corotational membrane elements based on constant-strain formulation, which are coupled mechanically to the underlying 
+tetrahedra. We show that using this model we are able to reproduce the experiments reported in~\cite{Hollenstein2006}, 
+while maintaining the real-time aspect of the simulation.
 

### 2013-05-31 Doc/Vrst2013/vrst2013.tex
ADD: some reformulations in section 3 and 4
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@581 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \begin{abstract}
-  \TODO{}
 The accurate mechanical modeling of liver is of a paramount interest in simulation-based operation planning 
@@
 the $x$ or $y$ coordinates of the nodes $i,j$ of the triangular element.
-The reader can refer to the respective literature for more thorough
-description~\cite{Felippa2003}.
+The reader can refer to the respective literature~\cite{Felippa2003} for more thorough
+description.
 
@@
 of the triangular mesh coincide with the nodes of the tetrahedral mesh, so no projection of one mesh onto the other is needed.
-Moreover, the stiffness matrices for capsule and parenchyma are then easily assembled
-together.
+Moreover, the stiffness matrices for capsule and parenchyma are then easily assembled together.
 % and solved as one system.
@@
 %\CD{The following may appear as "trivial"... maybe we can remove this part at the end if we need space} 
-%Without the loss of generality we can assume the tetrahedron consists of
-%nodes $p_1, p_2, p_3$ and $p_4$ and the boundary triangle has nodes $p_1, p_2$
-%and $p_3$. We can reorder the degrees of freedom so that the stiffness
-%matrix $\Mat{K}^t$ for the tetrahedron can be written as:
-%%
-%\begin{equation}
-%  \Mat{K}^t = \left[\begin{array}{c|c}
-%      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
-%      \hline
-%      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
-%  \end{array}\right]
-%\end{equation}
-%%
-%Then the assembled stiffness matrix for the element is:
-%%
-%\begin{equation}
-%  \Mat{K} = \left[\begin{array}{c|c}
-%      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
-%      \hline
-%      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
-%  \end{array}\right]
-%  +
-%  \left[\begin{array}{c|c}
-%      \Mat{K}^m & 0 \\
-%      \hline
-%      0 & 0 \\
-%  \end{array}\right]
-%\end{equation}
-%%
-%Where $\Mat{K}^m$ is the stiffness matrix of the triangular membrane.
+Without the loss of generality we can assume the tetrahedron consists of
+nodes $p_1, p_2, p_3$ and $p_4$ and the boundary triangle has nodes $p_1, p_2$
+and $p_3$. We can reorder the degrees of freedom so that the stiffness
+matrix $\Mat{K}^t$ for the tetrahedron can be written as:
+%
+\begin{equation}
+  \Mat{K}^t = \left[\begin{array}{c|c}
+      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
+      \hline
+      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
+  \end{array}\right]
+\end{equation}
+%
+Then the assembled stiffness matrix for the element is:
+%
+\begin{equation}
+  \Mat{K} = \left[\begin{array}{c|c}
+      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
+      \hline
+      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
+  \end{array}\right]
+  +
+  \left[\begin{array}{c|c}
+      \Mat{K}^m & 0 \\
+      \hline
+      0 & 0 \\
+  \end{array}\right]
+\end{equation}
+%
+Where $\Mat{K}^m$ is the stiffness matrix of the triangular membrane.
+
 The resulting system of linear equations is solved by direct solver based on Cholesky decomposition.
 
-Alternatively, one can use a mechanical coupling similar to the one used in
-\cite{Peterlik2012} to be able to use an arbitrary surface mesh. Nevertheless,
-for conforming triangular mesh both methods lead to the same solution.
+%Alternatively, one can use a mechanical coupling similar to the one used in
+%\cite{Peterlik2012} to be able to use an arbitrary surface mesh. Nevertheless,
+%for conforming triangular mesh both methods lead to the same solution.
 
@@
 \section{Methods and Simulation} %{{{
-
 The model presented in the previous section was implemented in 
@@
 where the response of liver exposed locally to a negative pressure is measured.
-The aspiration device consists of a tube 1\,cm in diameter and allows
-control of internal pressure in the tube. The test is performed by
+The aspiration device consists of a tube having 1\,cm in diameter and allows to
+control the pressure inside the tube. The test is performed by
 attaching the tube to the tissue and measuring the tissue response. We
@@
 Fig.~\ref{fig-aspiration1}: we meshed a 15$\times$15$\times$15\,mm$^3$ 
-cube representing the tissue 
-resulting in 2648 tetrahedra. Then we attached a 1\,cm tube 
-and applied a pressure of 3\,kPa inside the tube. The interaction between the tube and 
-tissue was modelled with friction contact.
+cube representing the tissue resulting in 2648 tetrahedra. Then we attached a 1\,cm tube 
+and applied a pressure of 3\,kPa inside the tube.
 
-In the simulations we used material properties for a pig liver reported in the literature: for the 
-Glisson's capsule we used Young's modulus $E_c$=8.22\,MPa and thickness $t_c$=20\,$\mu$m~\cite{Umale2011}.
-Since the values for parenchyma elasticity quite vary (usually between 2\,kPa to 5\,kPa), we perform the simulation 
-with $E_p$=3.5\,kPa.
-
-%}}}
+Since the tube is in direct contact with the tissue, uni-lateral constraints with friction were chosen 
+to model this interaction properly. We opted for a method based on \emph{non-linear complementarity problem}  (NLCP)
+which allows for solution of the Signorini's problem to avoid any interpenetration between the colliding 
+objects (see~\cite{} for details). Since NLCP requires explicitly the computation of compliance matrix which 
+is homogeneous to the inverse of the stiffness matrix, we employed a direct solver based on LDL decomposition to 
+solve the system and compute the inverse matrices.
 
 \section{Results and Discussion} %{{{
+As the first step, we evaluate the influence of the Glissons's capsule on the mechanical response of the tissue during
+the aspiration test. 
+In~\cite{Hollenstein2006} it is reported that the influence of the capsule is significant, as modeling 
+only the parenchyma without considering the membrane results in overestimation of the deformation by a factor of 2 to 3. 
+In order to verify this observation, we performed a simulation using capsule parameters measured experimentally on 
+ex-vivo pig liver~\cite{Umale2011}: Young's modulus of the membrane was set to $E_c$=8.22\,MPa and thickness was
+$t_c$=20\,$\mu$. 
+Since the values for parenchyma elasticity reported in the literature vary significantly being usually reported between 2\,kPa to 5\,kPa, 
+we employed a value $E_p$=3.5\,kPa for the Young's modulus of the parenchyma.
 
@@
 
-In~\cite{Hollenstein2006} it is reported that the Glisson's
-capsule has a non-negligible influence on the response of the tissue and modeling
-only the parenchyma leads to an overestimation of the deformation by a factor of 2 to 3. 
 In Fig.~\ref{fig-aspiration2} the profiles of cuts in the middle of the
-test cube are presented showing significantly larger deformation in the model without capsule.
-We can indeed observere that the deformation without the capsule is indeed overestimated approximately by a factor of 2.
-This is in perfect agreement with the results reported in the paper.% ~\cite{Hollenstein2006}
+test cube are presented showing significantly larger deformation in the model without capsule assuming that the same negative pressure was applied 
+on the tissue surface inside the tube. Moreover, it can be observed that the deformation without capculs is overestimated by a factor 
+close to 2, which is in perfect agreement with the results reported in~\cite{Hollenstein2006}.
 

### 2013-05-31 Doc/Vrst2013/vrst2013.tex
ADD: citations and further fixes
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@582 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\section{Methods and Simulation} %{{{
+\section{Methods} %{{{
 The model presented in the previous section was implemented in 
@@
 to model this interaction properly. We opted for a method based on \emph{non-linear complementarity problem}  (NLCP)
-which allows for solution of the Signorini's problem to avoid any interpenetration between the colliding 
-objects (see~\cite{} for details). Since NLCP requires explicitly the computation of compliance matrix which 
+where the non-linearity is introduced due to the friction. The NLCP
+allows for solution of the Signorini's problem to avoid any interpenetration between the colliding 
+objects (see~\cite{Duriez2006b} for details). Since NLCP requires explicitly the computation of compliance matrix which 
 is homogeneous to the inverse of the stiffness matrix, we employed a direct solver based on LDL decomposition to 

### 2013-05-31 Doc/Vrst2013/vrst2013.tex
ADD: reformulation in 4. chapter
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@583 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \alignauthor
-Stephan\'e Cotin\\
+St\'ephane Cotin\\
        \affaddr{INRIA}\\
@@
 
-Moreover, we are able to reproduce
-the measurements presented in~\cite{Hollenstein2006,Nava2008} with less than 1\% error.
-We do not model the time-dependent behaviour though.
-In this simulation we prescribe the the internal pressure of 30\,kPa. The other
-parameters of the model were obtained like this: First we fix the stiffness of
-parenchyma to $E_p$=30\,kPa to match the scenerio without capsule. Then,
-assuming the thickness $t$=93\,$\mu$m of the capsule (reported as average in
-the paper) we run several simulations with different capsule stiffness $E_c$.
-We have found the lowest error is when $E_c$=3\,MPa. All the stiffness
-parameters lie within the ranges measured in the paper.
+In the second step, we use our simulation to reproduce the displacements values measured experimentally in~\cite{Hollenstein2006,Nava2008}.
+Since the Young's moduli of neither parenchyma ($E_p$) nor capsule ($E_c$ were specified, we obtained both as follows: first, 
+performed a set of simulations without the capsule for different values of ($E_p$): the value that provided a good match 
+with reported displacement after applying the negative pressure of 30\,kPa inside the tube was $E_p$=30\,kPa. 
+The, we fixed the thickness of the capsule as $T$=93\,$\mu$m (reported as an average in the paper) and 
+using $E_p$=30\,kPa, we performed repeatedly the simulation with the capsule using different values of the Young's modulus
+of the capsule in order to minimize the displacement error. The minimal error (under 1\%) was obtained 
+for Young's modulus $E_c$=3\,MPa which lies in the range of capsule elasticity reported by experimental measurements.
+
+%Moreover, we are able to reproduce
+%the measurements presented in~\cite{Hollenstein2006,Nava2008} with less than 1\% error.
+%We do not model the time-dependent behaviour though.
+%In this simulation we prescribe the the internal pressure of 30\,kPa. The other
+%parameters of the model were obtained like this: First we fix the stiffness of
+%parenchyma to $E_p$=30\,kPa to match the scenerio without capsule. Then,
+%assuming the thickness $t$=93\,$\mu$m of the capsule (reported as average in
+%the paper) we run several simulations with different capsule stiffness $E_c$.
+%We have found the lowest error is when $E_c$=3\,MPa. All the stiffness
+%parameters lie within the ranges measured in the paper.
 

### 2013-05-31 Doc/Vrst2013/displacement.R
displacement evolution
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@584 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@ -1,0 +1,13 @@
+mt <- read.table("disp-tet.txt")
+mc <- read.table("disp-comp.txt")
+
+pdf("displacement.pdf", width=3, height=1.5, pointsize=8); # sizes in inches
+p <- par(lwd=0.5, mar=c(4,4,0,0)+0.1);
+
+i <- 1:500 # 5 seconds
+plot(c(0,5), range(mt[,3], mc[,3]), type="n", xlab="Time [s]", ylab="Displacement")
+lines(mt[i,1], mt[i,3], col=1)
+lines(mc[i,1], mc[i,3], col=4)
+
+dev.off()
+par(p)

### 2013-05-31 Doc/Vrst2013/vrst2013.tex
further reformulations
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@585 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-In the second step, we use our simulation to reproduce the displacements values measured experimentally in~\cite{Hollenstein2006,Nava2008}.
-Since the Young's moduli of neither parenchyma ($E_p$) nor capsule ($E_c$ were specified, we obtained both as follows: first, 
-performed a set of simulations without the capsule for different values of ($E_p$): the value that provided a good match 
-with reported displacement after applying the negative pressure of 30\,kPa inside the tube was $E_p$=30\,kPa. 
-The, we fixed the thickness of the capsule as $T$=93\,$\mu$m (reported as an average in the paper) and 
-using $E_p$=30\,kPa, we performed repeatedly the simulation with the capsule using different values of the Young's modulus
-of the capsule in order to minimize the displacement error. The minimal error (under 1\%) was obtained 
-for Young's modulus $E_c$=3\,MPa which lies in the range of capsule elasticity reported by experimental measurements.
+In the second step, we employed the simulation to reproduce the displacements values obtained via simulation in~\cite{Hollenstein2006,Nava2008}.
+Whereas the simulation in the referenced work was done on a mesh composed of tetrahedral elements for both the parenchyma 
+and the capsule, we modelled the capsule using the triangular elements as shown in section~\ref{}.
+
+Since the Young's moduli of neither parenchyma ($E_p$) nor capsule ($E_c$ were specified exactly, we obtained both as follows: first, 
+we performed a set of simulations without the capsule for different values of $E_p$: the value that provided a good match 
+with reported displacements was $E_p$=30\,kPa. 
+Then, we fixed the thickness of the capsule to be $T$=93\,$\mu$m (reported as the average value in the paper) and 
+using $E_p$=30\,kPa, we ran the simulation repeatedly using the model with the capsule. In each simulation we 
+used different value $E_c$ of the Young's modulus for the capsule in order to minimize the displacement error w.r.t. the 
+reported values. The minimal value of error (not exceeding 1\%) was obtained for $E_c$=3\,MPa. This 
+value lies in the range of capsule elasticity coefficients reported by experimental measurements.
 
@@
 
-Our goal is to create complete liver model suitable for real-time medical
-simulations. For this a model accounting for all three main constituents
-(parenchyma, vascularization and capsule) is necessary. To achieve that we have
-integrated our model with the vasularization method proposed
-in~\cite{Peterlik2012}. Adding the capsule did not signifficantly affect the
-performance since simulation speed of \TODO{XX FPS}%
-\footnote{\TODO{Measured on...}} was achieved. This suggests
-that the proposed technique is compatible with real-time simulation of whole
+In the scenario presented in this paper, we focus on local influence of the capsule.
+Nevertheless, it is possible to employ the actual model to demonstrate the global 
+impact of the capsule and to further validate the accuracy of such a complete model w.r.t. experimental measures. 
+Detailed description of such validation is however beyond the scope of this paper. 
+In spite of this we would like to emphasize that we have already integrated the
+capsule model with vascularized liver model described in~\cite{Peterlik2012}. 
+Adding the capsule did not affect the
+performance significantly: a visual refresh rate exceeding 25\,FPS was achieved on a
+PC with CPU Intel CPU i7-2630QM running at 3.00 GHz.
+This suggests that the proposed technique is compatible with real-time simulation of whole
 organ while modeling the local properties accurately.

### 2013-05-31 Doc/Vrst2013/vrst2013.tex
ADD: figure
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@586 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
+\begin{figure}
+  \centering
+  \includegraphics[height=3.5cm]{displacement.pdf}
+  \caption{\label{fig-aspiration1} The }
+\end{figure}
+
 Since the tube is in direct contact with the tissue, uni-lateral constraints with friction were chosen 
@@
 Whereas the simulation in the referenced work was done on a mesh composed of tetrahedral elements for both the parenchyma 
-and the capsule, we modelled the capsule using the triangular elements as shown in section~\ref{}.
+and the capsule, we modelled the capsule using the triangular elements as shown in section~\ref{ss:capsuleModel}.
 

### 2013-05-31 Doc/Vrst2013/displacement.R
fixing the figure
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@587 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@ -1,11 +1,15 @@
 mt <- read.table("disp-tet.txt")
 mc <- read.table("disp-comp.txt")
 
+mt[,3] <- (mt[,3] - 0.015)*1000
+mc[,3] <- (mc[,3] - 0.015)*1000
+
+
 pdf("displacement.pdf", width=3, height=1.5, pointsize=8); # sizes in inches
 p <- par(lwd=0.5, mar=c(4,4,0,0)+0.1);
 
 i <- 1:500 # 5 seconds
-plot(c(0,5), range(mt[,3], mc[,3]), type="n", xlab="Time [s]", ylab="Displacement")
+plot(c(0,5), range(mt[,3], mc[,3]), type="n", xlab="Time [s]", ylab="Displacement [mm]")
 lines(mt[i,1], mt[i,3], col=1)
 lines(mc[i,1], mc[i,3], col=4)
 

### 2013-05-31 Doc/Vrst2013/vrst2013.tex
FIX: i7 CPU
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@588 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 performance significantly: a visual refresh rate exceeding 25\,FPS was achieved on a
-PC with CPU Intel CPU i7-2630QM running at 3.00 GHz.
+PC with CPU Intel i7 running at 3.00 GHz.
 This suggests that the proposed technique is compatible with real-time simulation of whole

### 2013-05-31 Doc/Vrst2013/vrst2013.tex
FIX: conclusions
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@589 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 \section{Conclusions and Future Work} %{{{
-
-In the paper we presented a liver model accounting for the Glisson's capsule, a component that is often overlooked. Each constituent being modelled with a different type 
-of finite elements: tetrahedral corotational model used for the parenchyma and
-and CST membrane elements used for the capsule. 
-The model was validated in the context for local displacements based on the the aspiration test presented in~\cite{Hollenstein2006}. The aspiration test
-was successfully reproduced, showing a good agreement between the simulation and experiments.
+In the paper we presented a model of tissue with capsule based on two separated models for parenchyma (corotational tetrahedral elements)
+and membrane (CST triangular elements). Employing a simulation scenario known as aspiration test, 
+we compared the model with experimental data provided in the literature 
+and verified the results obtained in the simulation w.r.t.
+to a standard computationally expensive model based entirely on volume elements~\cite{Hollenstein2006}.
+We obtained values which were in good correspondence with results reported in the literature.
 
 Presently we are focusing on the validation of the model also in context of
-global deformations of the organ where we believe the capsule also plays an
-essential role. This task is however complicated by the lack of biomechanical
-literature dealing with the response of the capsule.
+global deformations of the organ where we believe the capsule plays an
+essential role as well. This task is however complicated due to the limited number of experimental studies dealing with 
+both the parenchyme and the capsule.
 

### 2013-05-31 Doc/Vrst2013/displacement.R
updating values in figure
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@590 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@ -9,7 +9,11 @@ pdf("displacement.pdf", width=3, height=1.5, pointsize=8); # sizes in inches
 p <- par(lwd=0.5, mar=c(4,4,0,0)+0.1);
 
 i <- 1:500 # 5 seconds
-plot(c(0,5), range(mt[,3], mc[,3]), type="n", xlab="Time [s]", ylab="Displacement [mm]")
+plot(c(0,5),
+     c(-0.285, 2.336), # make the range consistent
+     #range(mt[,3], mc[,3]),
+     type="n", xlab="Time [s]", ylab="Displacement [mm]")
+
 lines(mt[i,1], mt[i,3], col=1)
 lines(mc[i,1], mc[i,3], col=4)
 

### 2013-05-31 Doc/Vrst2013/vrst2013.tex
categories, term, keywords
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@591 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
final fixes
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@592 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
        \affaddr{Strasbourg, France}\\
-       \email{Igor.Peterlik@ihu-strasbourg.eu} % TODO
+       \email{Igor.Peterlik@ihu-strasbourg.eu}
 % 3rd. author
@@
        \affaddr{Lille, France}\\
-       \email{Christian.Duriez@inria.fr} % TODO
+       \email{Christian.Duriez@inria.fr}
 \and
@@
        \affaddr{Lille, France}\\
-       \email{Stephane.Cotin@inria.fr} % TODO
+       \email{Stephane.Cotin@inria.fr}
 }
@@
 
-The measures perfomed on the capsule show report stiffness which is significanly higher that that of the parenchyma, however, 
+The measures performed on the capsule report stiffness which is significantly higher that that of the parenchyma, however, 
 the thickness of the membrane does not exceed 100$\mu$m. For this reasons, it is not possible to model the 
@@
 
-% TODO
-% A category with the (minimum) three required fields
-\category{H.4}{Information Systems Applications}{Miscellaneous}
-%A category including the fourth, optional field follows...
-\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]
+% Categories
+\category{I.6.5}{Simulation and Modeling}{Model Development}
+\category{J.3}{Life and Medical Sciences}{Health}
 
-\terms{Theory} %TODO
+\terms{Modeling and Simulation}
 
 % NOT required for Proceedings
-%\keywords{deformable liver model, Glisson's capsule, real-time soft tissue modeling}
-\keywords{deformable liver model}
+\keywords{deformable liver model, Glisson's capsule, real-time soft tissue modeling}
 
@@
 Computer-aided physics-based medical simulation has proven to be an extremely useful technique in the area of medical training. 
-Nevertheless, whereas generic models are usually required in training simulators, accurate patient-specific modelling
+Nevertheless, whereas generic models are usually required in training simulators, accurate patient-specific modeling
 becomes necessary as soon as computer simulation is to be employed in the pre-operative planning. At the same time, 
@@
 
-\TG{The following paragraph is optional. Unnecessary but adds to the global context}
+%\TG{The following paragraph is optional. Unnecessary but adds to the global context}
 The mechanical importance of the vascular structures in liver is studied in~\cite{Peterlik2012}. It shows that the 
@@
 The parenchyma is modeled using corotational finite elements~\cite{Felippa2005}.
-Although it relies on linear stress-strain relationship, large displacements including rotations are modelled correctly. 
+Although it relies on linear stress-strain relationship, large displacements including rotations are modeled correctly. 
 While in the full non-linear formulation the stiffness matrix relates the forces $\Vec{f}$ and 
@@
 set up a simulation in SOFA to reproduce the experiment (see
-Fig.~\ref{fig-aspiration1}: we meshed a 15$\times$15$\times$15\,mm$^3$ 
+Fig.~\ref{fig-aspiration1}): we meshed a 15$\times$15$\times$15\,mm$^3$ 
 cube representing the tissue resulting in 2648 tetrahedra. Then we attached a 1\,cm tube 
 and applied a pressure of 3\,kPa inside the tube.
-
-\begin{figure}
-  \centering
-  \includegraphics[height=3.5cm]{displacement.pdf}
-  \caption{\label{fig-aspiration1} The }
-\end{figure}
 
@@
 \section{Results and Discussion} %{{{
-As the first step, we evaluate the influence of the Glissons's capsule on the mechanical response of the tissue during
+As the first step, we evaluate the influence of the Glisson's capsule on the mechanical response of the tissue during
 the aspiration test. 
@@
 ex-vivo pig liver~\cite{Umale2011}: Young's modulus of the membrane was set to $E_c$=8.22\,MPa and thickness was
-$t_c$=20\,$\mu$. 
+$t_c$=20\,$\mu$m. 
 Since the values for parenchyma elasticity reported in the literature vary significantly being usually reported between 2\,kPa to 5\,kPa, 
@@
   \caption{\label{fig-aspiration2} Displacement profile of the cube in the
-  aspiration test with (blue) and without (black) the capusle.}
+  aspiration test with (blue) and without (black) the capsule.}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[height=3.5cm]{displacement.pdf}
+  \caption{\label{fig-aspiration3} Evolution of the displacement at the center
+  in time for test with (blue) and without (black) the capsule.}
 \end{figure}
@@
 test cube are presented showing significantly larger deformation in the model without capsule assuming that the same negative pressure was applied 
-on the tissue surface inside the tube. Moreover, it can be observed that the deformation without capculs is overestimated by a factor 
+on the tissue surface inside the tube. Moreover, it can be observed that the deformation without capsule is overestimated by a factor 
 close to 2, which is in perfect agreement with the results reported in~\cite{Hollenstein2006}.
@@
 Whereas the simulation in the referenced work was done on a mesh composed of tetrahedral elements for both the parenchyma 
-and the capsule, we modelled the capsule using the triangular elements as shown in section~\ref{ss:capsuleModel}.
+and the capsule, we modeled the capsule using the triangular elements as shown in section~\ref{ss:capsuleModel}.
 
-Since the Young's moduli of neither parenchyma ($E_p$) nor capsule ($E_c$ were specified exactly, we obtained both as follows: first, 
+Since the Young's moduli of neither parenchyma ($E_p$) nor capsule ($E_c$) were specified exactly, we obtained both as follows: first, 
 we performed a set of simulations without the capsule for different values of $E_p$: the value that provided a good match 
 with reported displacements was $E_p$=30\,kPa. 
-Then, we fixed the thickness of the capsule to be $T$=93\,$\mu$m (reported as the average value in the paper) and 
+Then, we fixed the thickness of the capsule to be $t$=93\,$\mu$m (reported as the average value in the paper) and 
 using $E_p$=30\,kPa, we ran the simulation repeatedly using the model with the capsule. In each simulation we 
@@
 essential role as well. This task is however complicated due to the limited number of experimental studies dealing with 
-both the parenchyme and the capsule.
+both the parenchyma and the capsule.
 

### 2013-05-31 Doc/Vrst2013/displacement.R
changing the time range
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@593 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@ -8,8 +8,8 @@ mc[,3] <- (mc[,3] - 0.015)*1000
 pdf("displacement.pdf", width=3, height=1.5, pointsize=8); # sizes in inches
 p <- par(lwd=0.5, mar=c(4,4,0,0)+0.1);
 
-i <- 1:500 # 5 seconds
-plot(c(0,5),
+i <- 1:200 # 5 seconds
+plot(c(0,2),
      c(-0.285, 2.336), # make the range consistent
      #range(mt[,3], mc[,3]),
      type="n", xlab="Time [s]", ylab="Displacement [mm]")

### 2013-06-26 Doc/Media2013/media2013.tex
preparing MEDIA paper
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@597 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2013-03-02 Doc/Miccai2013/miccai2013.tex
@@
-% http://www.miccai2013.org/submission_guideline.html
-\documentclass{llncs}
+%\documentclass[preprint,12pt]{elsarticle}
+
+%% Use the option review to obtain double line spacing
+%% \documentclass[preprint,review,12pt]{elsarticle}
+
+%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
+%% for a journal layout:
+%% \documentclass[final,1p,times]{elsarticle}
+%% \documentclass[final,1p,times,twocolumn]{elsarticle}
+%% \documentclass[final,3p,times]{elsarticle}
+\documentclass[final,3p,times,twocolumn]{elsarticle}
+%% \documentclass[final,5p,times]{elsarticle}
+%% \documentclass[final,5p,times,twocolumn]{elsarticle}
 
 \usepackage{amsmath}
-\usepackage{makeidx}
-\usepackage[pdftex]{graphicx}
-\usepackage{subfigure}
+%\usepackage{amssymb}
+%\usepackage{amsthm}
+%\usepackage{makeidx}
+\usepackage{graphicx}
+%\usepackage[pdftex]{graphicx}
+\usepackage{subfig}
 
-\def\wrt{w.\,r.\,t.}
-\def\eg{e.\,g.}
-\def\ie{i.\,e.}
+%% natbib.sty is loaded by default. However, natbib options can be
+%% provided with \biboptions{...} command. Following options are
+%% valid:
+%%
+%%   round  -  round parentheses are used (default)
+%%   square -  square brackets are used   [option]
+%%   curly  -  curly braces are used      {option}
+%%   angle  -  angle brackets are used    <option>
+%%   semicolon  -  multiple citations separated by semi-colon
+%%   colon  - same as semicolon, an earlier confusion
+%%   comma  -  separated by comma
+%%   numbers-  selects numerical citations
+%%   super  -  numerical citations as superscripts
+%%   sort   -  sorts multiple citations according to order in ref. list
+%%   sort&compress   -  like sort, but also compresses numerical citations
+%%   compress - compresses without sorting
+%%
+%% \biboptions{comma,round}
+% \biboptions{}
+
+\def\wrt{w.\,r.\,t.\ }
+\def\eg{e.\,g.\ }
+\def\ie{i.\,e.\ }
 \def\Dash{\nobreak\,---\penalty-500\,}
@@
 
-\renewcommand{\Vec}[1]{\mathbf{#1}}
-\newcommand{\Mat}[1]{\mathbf{#1}}
+\renewcommand{\Vec}[1]{\boldsymbol{#1}}
+\newcommand{\Mat}[1]{\boldsymbol{#1}}
+
+\journal{Medical Image Analysis}
 
 \begin{document}
-%
-%\frontmatter          % for the preliminaries
-%
-\mainmatter              % start of the contributions
-%
-\title{Complete Real-Time Liver Model Including Glisson's Capsule, Vascularization and Parenchyma}
-%
-\titlerunning{Complete Liver Model}  % abbreviated title (for running head)
-%                                     also used for the TOC unless
-%                                     \toctitle is used
-%
-\author{Anonymous}
-%\author{%
-%Tom\'a\v{s} Golembiovsk\'y\inst{1,2} \and%
-%Igor Peterl\'ik\inst{3} \and%
-%Christian Duriez\inst{2,4} \and%
-%Stephan\'e Cotin\inst{2}%
-%}
-%
-%\authorrunning{Tom\'a\vs Golembiovsk\'y et al.} % abbreviated author list (for running head)
-%
-%\institute{%
-%Faculty of Informatics, Masaryk University, Brno, Czech Republic\and%
-%INRIA Lille -- Nord, France\and%
-%Institut Hospitallo-Universitaire, Strasbourg, France \and%
-%University of Lille, Lille, France%
-%}
-%
-\maketitle
+
+\begin{frontmatter}
+
+%% Title, authors and addresses
+
+%% use the tnoteref command within \title for footnotes;
+%% use the tnotetext command for the associated footnote;
+%% use the fnref command within \author or \address for footnotes;
+%% use the fntext command for the associated footnote;
+%% use the corref command within \author for corresponding author footnotes;
+%% use the cortext command for the associated footnote;
+%% use the ead command for the email address,
+%% and the form \ead[url] for the home page:
+%%
+%% \title{Title\tnoteref{label1}}
+%% \tnotetext[label1]{}
+%% \author{Name\corref{cor1}\fnref{label2}}
+%% \ead{email address}
+%% \ead[url]{home page}
+%% \fntext[label2]{}
+%% \cortext[cor1]{}
+%% \address{Address\fnref{label3}}
+%% \fntext[label3]{}
+
+\title{TODO: Title} % TODO
+
+%% use optional labels to link authors explicitly to addresses:
+%% \author[label1,label2]{<author name>}
+%% \address[label1]{<address>}
+%% \address[label2]{<address>}
+
+\author{}
+
+\address{}
 
 \begin{abstract}
-The accurate mechanical modeling of liver is of a paramount interest in simulation-based operation planning 
-and computer-aided per-operative guidance. However, the realistic simulation of liver behaviour is a challenging task, 
-since the organ is composed of three constituents: parenchyma, vascularization and Glisson's capsule, each having different mechanical properties. While the 
-mechanical influence of vessels on both global and local behaviour of the organ has been measured and modelled previously~\cite{Peterlik2012}, only local 
-effects have been studied in the case of Glisson's capsule~\cite{Hollenstein2006,Ahn2010}.
-
-In this paper we introduce what we believe is the first real-time model of liver taking into account all three constituents, each being simulated 
-employing a different finite element formulation.
-Using the proposed model, we focus on the role of the Glisson's capsule: first, we show that the model is able to reproduce correctly
-the local effects of the capsule in so called aspiration test. Second, based on CT image data of porcine abdominal cavity acquired in supine and flank 
-positions, we demonstrate 
-that the capsule has also a global impact on the mechanical behaviour of the liver in the case when the organ undergoes large deformations.
-
-%This work provides a complete physical model for the deformations of the liver with its three constituents: parenchyma, vessels and the Glisson's capsule.
-%Some recent experimental results \cite{Ahn2010} have shown that there is a significant difference of stiffness between the Glisson's capsule and the parenchyma. 
-%Moreover, it has been observed that the capsule plays an important role for the deformation at the local level \cite{Hollenstein2006}.
-%However, the Glisson's capsule is very thin and its influence for the global deformations of the liver was never highlighted.
-%In this work, we propose to use a membrane FEM model for the capsule, that is coupled with a vascularized model of the liver.
-%We show that the approach is able to reproduce the deformations observed at the local level.
-%Additionally, we show, by simulating natural global deformations of the liver observed on real images (due to gravity), that the results have important differences with and without the capsule.
-
-\keywords{deformable liver model, Glisson's capsule, vascularization, real-time soft tissue modeling}
+%% Text of abstract
+TODO % TODO
 \end{abstract}
 
-\section{Introduction} 
+\begin{keyword}
+deformable liver model \sep Glisson's capsule \sep vascularization \sep real-time soft tissue modeling
+
+%% MSC codes here, in the form: \MSC code \sep code
+%% or \MSC[2008] code \sep code (2000 is the default)
+
+\end{keyword}
+
+\end{frontmatter}
+
+% ---  Main text  ---
+
+\TG{mostly just copy of MICCAI paper}
+
+\section{Introduction} %{{{
+
 According to the statistics, nearly 100,000 European citizens die of cirrhosis or liver cancer each year. 
@@
 
-%Then we show the correct behaviour of the
-%model in context of \emph{local} deformations and finally we analyze the influence of
-%the capsule on the \emph{global} deformations of the liver and show that despite the
-%small thickness it plays an important role also in the \emph{global} context.
-
-%Medical simulation is expected to play an integral part in many aspects of
-%medical practice. From diagnosis, through planning or training to computer
-%assisted intervention.
-%Medical training application, like the ones for training laparoscopic
-%surgery, have to operate in real-time and a good medical model has to be
-%simple yet efficient to be really usable in this environment.
-%
-%Liver is largest internal organ formed mostly by a soft tissue (parenchyma)
-%that is hyper-elastic and also visco-elastic. The parenchyma is interwoven
-%by dense vascular network, this tree-like tubular structure is stiffer then
-%the parenchyma. Finally the parenchyma is covered by very stiff collagenous
-%membrane that preserves the integrity of the organ.
-%
-%Strangely, most of the work on liver modelling only considers the
-%parenchyma and model the liver as single homogeneous object \TG{Refs}.
-%Only recently
-%Peterl\'{i}k et al. \cite{Peterlik2012} showed the importance of the vascular
-%structures on the liver deformation.
-%
-%Another overlooked component is the Glisson's capsule.
-%Biomechanical measurements report that the capsule has very small
-%thickness (approx. 20 $\mu$m for porcine liver \cite{Umale2011}).
-%On the other hand it has in three orders of magnitude higher stiffness than
-%the parenchyma.
-%Because of this it plays an important role in local
-%deformations of the liver \cite{Ahn2010,Hollenstein2006}.
-%It was reported that evaluation of the liver as a homogeneous material
-%without capsule can lead to overestimation of local deformations up to the factor of 3
-%\cite{Hollenstein2006}.
-%
-%While the influence on local deformation was reported, no experimental
-%results show whether the capsule also plays an important role in global
-%deformations of the liver. We propose a composite model that models the
-%capsule with membrane FEM elements. We show that the model is able to
-%simulate expected local behaviour reported in literature. Further more we
-%show that the capsule plays also an important role in the deformation on
-%global scale.
-
 %}}}
@@
 
-\subsection{Parenchyma and Vascularization} %{{{
+\subsection{Parenchyma} %{{{
 
-It is known that the parenchyma exhibits non-linear viscoelastic behaviour \cite{Marchesseau2010}.
+It is known that the parenchyma exhibits non-linear viscoelastic behaviour
+\cite{Marchesseau2010,Gao2009}.
 However, as we are mainly interested in the static equilibrium, we do not model the time-dependent
@@
 
-The model of the vascularization is based on linear beams 
-with local frames of reference~\cite{Duriez2006}; in many aspects it's similar to the
-corotational formulation described above. As such the model also handles geometric
-non-linearities in the deformation. Through a specification of cross section and moments of inertia, 
-the model can account for the specific properties of the blood vessels.
+%}}}
 
-The beam-based model of vessels is mechanically coupled to the parenchyma as described in~\cite{Peterlik2012}. 
-The coupling assumes that there is no relative motion between the vessels and the surrounding parenchyma. 
-%The proposed model assumes no relative
-%motion of vessels to the parenchyma. Segmented vascular network is
-%discretized into the set of linear elements that are then projected into
-%the tetrahedral mesh of the parenchyma. 
-During the simulation, the nodes of linked beams are first displaced and rotated according to the actual motion of associated tetrahedra. 
-As the deformation of beams results in mechanical response represented by forces and torques, these are propagated back to 
-the tetrahedral FE model.
+\subsection{Vascularization} %{{{
 
-%a displacement of tetrahedral nodes are mapped  forces and torques resulting from the deformations of vessels by parenchyma are
-%computed and propagated back onto the tetrahedral mesh.
+As a first step towards a mechanical model of vessels, a continuous
+representation of vascular structures is constructed. Using segmeted data
+of a vessel tere, points along centerlines of each vessel branch are
+selected. This task can be performed either manually or automatically, \eg
+with VMTK\footnote{www.vmtk.org}. Each branch can be represented as a
+series of cubic B\'{e}zier curves fitted to the centerline points: the
+series begins in a starting point of the branch and ends either in another
+branching point or in the endpoint of the branch.
+
+A set of interpolated points is constructed using the B\'{e}zier curves;
+positions of the interpolated points can be chosen arbitrarily along each
+branch dividing it ionto intervals with a constant or varying length. For
+each interpolated point we calculate its orientation given by the tangent
+and normal of the B\'{e}zier curve in that point.
+
+The interpolated points are used as nodes of serially linked beam elements,
+in a similar way as proposed by Duriez et al. \cite{Duriez2006} for
+simulating catheters and guidewires. This model shares some similarities
+with the corotational model described above, and in particular allows for
+geometrically non-linear deformations. We introduce some modifications to
+the model to take into account particular nature of vessels, in particualr
+through specific cross section profiles and moments of inertia (see
+\cite{Przemieniecki1985} for details). The static formulation for the
+deformation of a beam is described by a system similar to the one used for
+the parenchyme, with the difference that each node is described with six
+degrees of freedom, three of which correspond to the spatial position, and
+three to the angular position of the node in a global reference frame
+\TODO{figure?}.
+
+% TODO: presunout do sekce composite model
+%The beam-based model of vessels is mechanically coupled to the parenchyma as described in~\cite{Peterlik2012}. 
+%The coupling assumes that there is no relative motion between the vessels and the surrounding parenchyma. 
+%%The proposed model assumes no relative
+%%motion of vessels to the parenchyma. Segmented vascular network is
+%%discretized into the set of linear elements that are then projected into
+%%the tetrahedral mesh of the parenchyma. 
+%During the simulation, the nodes of linked beams are first displaced and rotated according to the actual motion of associated tetrahedra. 
+%As the deformation of beams results in mechanical response represented by forces and torques, these are propagated back to 
+%the tetrahedral FE model.
+
+%%a displacement of tetrahedral nodes are mapped  forces and torques resulting from the deformations of vessels by parenchyma are
+%%computed and propagated back onto the tetrahedral mesh.
 
@@
 
+\subsection{Coupling Between Vessels and Parenchyma} %{{{
+
+We propose a mapping betwen the mesh nodes of the vessels and parenchyma to
+create the vascularized model. Since no relative motion between the vessels
+and parenchyma is observed in reality, the mapping betweeen the two can be
+modelled as a constraint. In each step of the simulation, the actual
+displacements of the parenchyma mesh nodes are mapped to the vessel nodes
+and reciprocally, the force contribution due to the deformation of the
+vessel is propagated to the parenchyma. The mapping of the forces is based
+on a principle of virtual work. Using the corotational model, the
+parenchyma is discretized by a mesh composed of a linear $P_1$ tetrahedral
+elements given by four nodes with three degrees of freedom. The vessel is
+modeled with beam elements where each beam is given by two end points each
+having 6 degreees of freedom (positions and rotations). We focus on mapping
+betwen an arbitrary beam node $b$ and tetrahedron $T$.
+
+Let us denote $\Vec{p}$ the initial position and $\Mat{R}$ initial rotation
+of the beam node $b$. Before the simulation starts, we select the
+tetrahedron $T = \{\Vec{t}_A, \Vec{t}_B, \Vec{t}_C, \Vec{t}_D \}$ which is
+closest to the beam node $b$. Then, the barycentric coordinates
+$\Vec{\beta}_T = \{\beta_A, \beta_B, \beta_C\}$ of the position $\Vec{p}$
+are computed \wrt the tetrahedron $T$ and stored. Similarly, the initial
+rotation $\Mat{R}$ is transformed to a a matrix $\Mat{O}_T$ which defines
+the orientation of the point $B$ \wrt the local basis $\Mat{L}$ defined by
+the edges of the tetrahedron, \ie $\Mat{O}_T = \Mat{L}^{-1} \Mat{R}$.
+According to the initial assumption (no relative motion between parenchyma
+and vessel), both the barycentric coordinates $\Vec{\beta}_T$ and
+$\Mat{O}_T$ are constant during the simulation. Let's suppose that in  the
+actual step, the tetrahedron is rotated and deformed as shown in
+\TODO{figure}. Denoting the nodal positions of the deformed tetrahedron $T'
+= \{\Vec{t}'_A, \Vec{t}'_B, \Vec{t}'_C, \Vec{t}'_D\}$, new actual position
+$\Vec{p}'$  of the beam node is computed using the barycentric coordinates
+$\Vec{\beta}_T$ as
+%
+\begin{equation}
+  \Vec{p}' = \Vec{t}'_A \beta_A + \Vec{t}'_B \beta_B + \Vec{t}'_C \beta_C +
+  \Vec{t}'_D (1 - \beta_A - \beta_B - \beta_C).
+\end{equation}
+%
+The updated orientation $\Mat{R}'$ of the beam node is computed using the
+basis $\Mat{L}'$ given by the actual displacement of the tetrahedron nodes
+and precomputed orientation $\Mat{O}_T$. First, $\Mat{\tilde{R}}' =
+\Mat{L}' \Mat{O}_T$ is computed. However, since besides the rotation ad
+deformation of the tetrahedron must be taken into account the matrix
+$\Mat{\tilde{R}}'$ is not necessarily orthogonal. Therefore, the updated
+orientation $\Mat{R}'$ is calculated as orthogonal component of polar
+decomposition of $\Mat{\tilde{R}}'$. As the positional mapping is applied
+in each step of the simulation, the positions of the beams are
+kinematically linked to the positions of tetrahedra.
+
+After the updated position is mapped from the tetrahedron to the beam point,
+the forces acting in the beam point must be mapped onto the vertices of the
+associated tetrahedron. For the beam point, we have linear forces $\Vec{f}
+= \{ f_x , f_y , f_z \}$
+as well as torques $\Vec{\tau} = \{ \tau_x, \tau_y, \tau_z \}$, whereas for the tetrahedron, there are only
+linear nodal forces $\Mat{F} = \{ \Vec{f}_A , \Vec{f}_B , \Vec{f}_C ,
+\Vec{f}_D \}$ . The force contribution to each node
+given by linear forces f is computed using the barycentric coordinates, required
+by the positional mapping. The torques are transformed to linear forces acting in
+the tetrahedron nodes using the equation $\Vec{\tau} = \Vec{r} \times
+\Vec{f}$ where $\Vec{r}$ is vector connecting
+the beam and tetrahedron nodes. Putting it together, the force in tetrahedron
+node $I \in \{A, B, C, D\}$ is computed as $\Vec{F}_I = \beta_I \Vec{f} −
+(\Vec{t}'_I − \Vec{p}') \times \tau$.
+
+%}}}
+
 \subsection{Coupling Between Capsule and Parenchyma} %{{{
-The literature reports high cohesion between capsule and parenchyma.
-Based on this property we assume there is no relative motion of the capsule \wrt\ the parenchyma.
+The literature reports high cohesion between capsule and parenchyma
+\TODO{need refs}.
+Based on this property we assume there is no relative motion of the capsule \wrt the parenchyma.
 Although an arbitrary surface mesh could be used to model the capsule, we exploit 
@@
 
-\begin{figure}[t]
-    	\subfigure[]{\includegraphics[height=3.5cm]{aspiration.jpg}}
-	\subfigure[]{\includegraphics[width=7cm]{aspiration.pdf}}
-  \caption{\label{fig-aspiration}Aspiration test: (a) the SOFA simulation scene, (b) cube displacement profile
-  with the capsule (in blue) and without capsule (in black).}
-\end{figure}
+%XXX
+%\begin{figure}[t]
+%    	\subfigure[]{\includegraphics[height=3.5cm]{aspiration.jpg}}
+%	\subfigure[]{\includegraphics[width=7cm]{aspiration.pdf}}
+%  \caption{\label{fig-aspiration}Aspiration test: (a) the SOFA simulation scene, (b) cube displacement profile
+%  with the capsule (in blue) and without capsule (in black).}
+%\end{figure}
 
@@
 
+\TG{
+  We are able to match the deformations presented in
+  [Hollenstein2006] with less then 1\% error using the following
+  parameters: E=30\,kPa for parenchyma; E=3\,MPa and t=93\,$\mu$mm for capsule.
+  The paper does not specify single values for elasticity modulus but specifies
+  the possible range. Our values lie in the prescribed ranges.}
+
 %}}}
 
-\begin{figure}[t]
-  \centering
-    \includegraphics[width=.95\textwidth]{forceEvolution.pdf}
-  \caption{\label{f:forceEvol}Evolution of forces during the simulation, where the supine data is deformed
-towards the flank shape, and the stiffness of the capsule is increased from 0 to 8.22\,MPa after the initial equilibrium is achieved.}
-\end{figure}
+%XXX
+%\begin{figure}[t]
+%  \centering
+%    \includegraphics[width=.95\textwidth]{forceEvolution.pdf}
+%  \caption{\label{f:forceEvol}Evolution of forces during the simulation, where the supine data is deformed
+%towards the flank shape, and the stiffness of the capsule is increased from 0 to 8.22\,MPa after the initial equilibrium is achieved.}
+%\end{figure}
 
-\subsection{Global Deformations}
+\subsection{Global Deformations} %{{{
+
 In the previous section we have shown that the composite model of parenchyma and 
@@
 
-\begin{figure}[t]
-  \centering
-    \subfigure[]{\includegraphics[width=.32\textwidth]{gravity2.png}}
-    \subfigure[]{\includegraphics[width=.32\textwidth]{registered.png}}
-    \subfigure[]{\includegraphics[width=.32\textwidth]{distanceMap.png}}
-    \caption{\label{f:global}%
-      (a) Liver under gravity with (red) and without (yellow) capsule;
-      (b) liver in supine (yellow) and flank (red) position;
-      (c) Hausdorff distance map between registered version with and without capsule.}
-\end{figure}
+%XXX
+%\begin{figure}[t]
+%  \centering
+%    \subfigure[]{\includegraphics[width=.32\textwidth]{gravity2.png}}
+%    \subfigure[]{\includegraphics[width=.32\textwidth]{registered.png}}
+%    \subfigure[]{\includegraphics[width=.32\textwidth]{distanceMap.png}}
+%    \caption{\label{f:global}%
+%      (a) Liver under gravity with (red) and without (yellow) capsule;
+%      (b) liver in supine (yellow) and flank (red) position;
+%      (c) Hausdorff distance map between registered version with and without capsule.}
+%\end{figure}
 
@@
 performed a following protocol employing the supine and flank models built from the CT data. 
-First, we performed a rigid registration of the two organs \wrt\ the root point of the hepatic vein. 
+First, we performed a rigid registration of the two organs \wrt the root point of the hepatic vein. 
 As shown in Fig.~\ref{f:global}b, the liver undergoes very large deformation when the body 
@@
 Once the static equilibrium was achieved, the effect of the capsule was gradually switched 
-on, \ie\ the stiffness of the membrane was slowly increased from 0 to 8.22\,MPa. The evolution of the force response in the 
+on, \ie the stiffness of the membrane was slowly increased from 0 to 8.22\,MPa. The evolution of the force response in the 
 points with prescribed displacements is depicted in Fig.~\ref{f:forceEvol} showing the change in the response as the effect
@@
 plays an important role in the global mechanical behaviour of the liver.
+
+%}}}
+
+\subsection{More Experiments}
+
+\TG{list of other experiments}
 
@@
 
-%
-% ---- Bibliography ----
-%
+%% The Appendices part is started with the command \appendix;
+%% appendix sections are then done as normal sections
+%% \appendix
 
-\bibliographystyle{splncs03}
+%% \section{}
+%% \label{}
+
+% ---  Bibliography  ---
+
+\bibliographystyle{model1-num-names}
 \bibliography{bibdata}
+
+%% Authors are advised to submit their bibtex database files. They are
+%% requested to list a bibtex style file in the manuscript if they do
+%% not want to use model1-num-names.bst.
 

### 2013-07-02 Doc/Media2013/media2013.tex
aspiration test
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@598 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-\TG{
-  We are able to match the deformations presented in
-  [Hollenstein2006] with less then 1\% error using the following
-  parameters: E=30\,kPa for parenchyma; E=3\,MPa and t=93\,$\mu$mm for capsule.
-  The paper does not specify single values for elasticity modulus but specifies
-  the possible range. Our values lie in the prescribed ranges.}
+---
+
+Since the tube is in direct contact with the tissue, uni-lateral constraints with friction were chosen 
+to model this interaction properly. We opted for a method based on \emph{non-linear complementarity problem}  (NLCP)
+where the non-linearity is introduced due to the friction. The NLCP
+allows for solution of the Signorini's problem to avoid any interpenetration between the colliding 
+objects (see~\cite{Duriez2006b} for details). Since NLCP requires explicitly the computation of compliance matrix which 
+is homogeneous to the inverse of the stiffness matrix, we employed a direct solver based on LDL decomposition to 
+solve the system and compute the inverse matrices.
+
+As the first step, we evaluate the influence of the Glisson's capsule on the mechanical response of the tissue during
+the aspiration test. 
+In~\cite{Hollenstein2006} it is reported that the influence of the capsule is significant, as modeling 
+only the parenchyma without considering the membrane results in overestimation of the deformation by a factor of 2 to 3. 
+In order to verify this observation, we performed a simulation using capsule parameters measured experimentally on 
+ex-vivo pig liver~\cite{Umale2011}: Young's modulus of the membrane was set to $E_c$=8.22\,MPa and thickness was
+$t_c$=20\,$\mu$m. 
+Since the values for parenchyma elasticity reported in the literature vary significantly being usually reported between 2\,kPa to 5\,kPa, 
+we employed a value $E_p$=3.5\,kPa for the Young's modulus of the parenchyma.
+
+%\begin{figure}
+%  \centering
+%  \includegraphics[width=7cm]{aspiration.pdf}
+%  \caption{\label{fig-aspiration2} Displacement profile of the cube in the
+%  aspiration test with (blue) and without (black) the capsule.}
+%\end{figure}
+%
+%\begin{figure}
+%  \centering
+%  \includegraphics[height=3.5cm]{displacement.pdf}
+%  \caption{\label{fig-aspiration3} Evolution of the displacement at the center
+%  in time for test with (blue) and without (black) the capsule.}
+%\end{figure}
+
+In Fig.~\ref{fig-aspiration2} the profiles of cuts in the middle of the
+test cube are presented showing significantly larger deformation in the model without capsule assuming that the same negative pressure was applied 
+on the tissue surface inside the tube. Moreover, it can be observed that the deformation without capsule is overestimated by a factor 
+close to 2, which is in perfect agreement with the results reported in~\cite{Hollenstein2006}.
+
+In the second step, we employed the simulation to reproduce the displacements values obtained via simulation in~\cite{Hollenstein2006,Nava2008}.
+Whereas the simulation in the referenced work was done on a mesh composed of tetrahedral elements for both the parenchyma 
+and the capsule, we modeled the capsule using the triangular elements as shown in section~\ref{ss:capsuleModel}.
+
+Since the Young's moduli of neither parenchyma ($E_p$) nor capsule ($E_c$) were specified exactly, we obtained both as follows: first, 
+we performed a set of simulations without the capsule for different values of $E_p$: the value that provided a good match 
+with reported displacements was $E_p$=30\,kPa. 
+Then, we fixed the thickness of the capsule to be $t$=93\,$\mu$m (reported as the average value in the paper) and 
+using $E_p$=30\,kPa, we ran the simulation repeatedly using the model with the capsule. In each simulation we 
+used different value $E_c$ of the Young's modulus for the capsule in order to minimize the displacement error w.r.t. the 
+reported values. The minimal value of error (not exceeding 1\%) was obtained for $E_c$=3\,MPa. This 
+value lies in the range of capsule elasticity coefficients reported by experimental measurements.
 

### 2013-07-28 Doc/Vrst2013/vrst2013.tex
using acmsiggraph class
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@600 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
-% TODO
-% REMEMBER: After having produced the .bbl file, and prior to final submission,
-% you need to 'insert'  your .bbl file into your source .tex file so as to provide
-% ONE 'self-contained' source file.
+\documentclass[sposored]{acmsiggraph}
 
-\documentclass{acm_proc_article-sp}
-
-\usepackage{caption}
-\DeclareCaptionType{copyrightbox}
 \usepackage{amsmath}
-%\usepackage{makeidx}
-%\usepackage[pdftex]{graphicx} ... we have to use ps/eps
 \usepackage{subfig}
@@
 
-%%% Comments
-\usepackage{color}
-\newcommand{\TG}[1]{{\color{blue}\textbf{TG: #1}}}
-\newcommand{\CD}[1]{{\color{green}\textbf{CD: #1}}}
-\newcommand{\IP}[1]{{\color{cyan}\textbf{IP: #1}}}
-\newcommand{\TODO}[1]{{\color{red}\textbf{TODO: #1}}}
-%%%
-
 \renewcommand{\Vec}[1]{\mathbf{#1}}
@@
 
+\title{Modeling Liver with Glisson's Capsule in Real-Time Medical Simulations}
+
+%TODO
+\author{
+\begin{tabular}[t]{cccc}%
+  Tom\'a\v{s} Golembiovsk\'y\thanks{nyoxi@ics.muni.cz} &
+  Igor Peterl\'ik &
+  Christian Duriez &
+  St\'ephane Cotin \\
+%
+  Faculty of Informatics &
+  Institut Hospitalo-Universitaire &
+  INRIA &
+  INRIA \\
+%
+  Masaryk University &
+  Strasbourg, France &
+  Lille, France &
+  Lille, France \\
+%
+  Brno, Czech Republic
+\end{tabular}
+}
+%\affiliation{foobar}
+%\numberofauthors{4}
+%\author{%
+%Tom\'a\v{s} Golembiovsk\'y\\
+%       \affaddr{Faculty of Informatics}\\
+%       \affaddr{Masaryk University}\\
+%       \affaddr{Brno, Czech Republic}\\
+%       \email{nyoxi@ics.muni.cz}
+%% 2nd. author
+%\alignauthor
+%Igor Peterl\'ik\\
+%       \affaddr{Institut Hospitalo-Universitaire}\\
+%       \affaddr{Strasbourg, France}\\
+%       \email{Igor.Peterlik@ihu-strasbourg.eu}
+%% 3rd. author
+%\alignauthor
+%Christian Duriez\\
+%       \affaddr{INRIA}\\
+%       \affaddr{Lille, France}\\
+%       \email{Christian.Duriez@inria.fr}
+%\and
+%% 4th. author
+%\alignauthor
+%St\'ephane Cotin\\
+%       \affaddr{INRIA}\\
+%       \affaddr{Lille, France}\\
+%       \email{Stephane.Cotin@inria.fr}
+%}
+\pdfauthor{Tom\'a\v{s} Golembiovsk\'y}
+
 \begin{document}
-
-\title{Modeling Liver with Glisson's Capsule in Real-Time Medical Simulations}
-%\subtitle{}
-
-\numberofauthors{4}
-\author{
-% 1st. author
-\alignauthor
-Tom\'a\v{s} Golembiovsk\'y\\
-       \affaddr{Faculty of Informatics}\\
-       \affaddr{Masaryk University}\\
-       \affaddr{Brno, Czech Republic}\\
-       \email{nyoxi@ics.muni.cz}
-% 2nd. author
-\alignauthor
-Igor Peterl\'ik\\
-       \affaddr{Institut Hospitalo-Universitaire}\\
-       \affaddr{Strasbourg, France}\\
-       \email{Igor.Peterlik@ihu-strasbourg.eu}
-% 3rd. author
-\alignauthor
-Christian Duriez\\
-       \affaddr{INRIA}\\
-       \affaddr{Lille, France}\\
-       \email{Christian.Duriez@inria.fr}
-\and
-% 4th. author
-\alignauthor
-St\'ephane Cotin\\
-       \affaddr{INRIA}\\
-       \affaddr{Lille, France}\\
-       \email{Stephane.Cotin@inria.fr}
-}
-
-%\date{30 May 2013}
 
@@
 % Categories
-\category{I.6.5}{Simulation and Modeling}{Model Development}
-\category{J.3}{Life and Medical Sciences}{Health}
+\begin{CRcatlist}
+  \CRcat{I.6.5}{Simulation and Modeling}{Model Development}
+  \CRcat{J.3}{Life and Medical Sciences}{Health};
+\end{CRcatlist}
 
-\terms{Modeling and Simulation}
+\keywords{deformable liver model, Glisson's capsule}
 
-% NOT required for Proceedings
-\keywords{deformable liver model, Glisson's capsule, real-time soft tissue modeling}
+\keywordlist
+
+\TOGlinkslist
+
+\copyrightspace
 
@@
 
-\bibliographystyle{abbrv}
+\bibliographystyle{acmsiggraph}
 \bibliography{bibdata}

### 2013-07-29 Doc/Vrst2013/vrst2013.tex
fixing some issues
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@601 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-The measures performed on the capsule report stiffness which is significantly higher that that of the parenchyma, however, 
+The measures performed on the capsule report stiffness which is significantly higher than that of the parenchyma, however, 
 the thickness of the membrane does not exceed 100$\mu$m. For this reasons, it is not possible to model the 
@@
 Then, in each step, the motion of each element $e$ is decomposed into rigid rotation $\Mat{R}^e$ and local deformation. 
-The rotations are then used to update each local element stiffness matrix as $\Mat{R}\Mat{K}_0\Mat{R}^{\top}$
+The rotations are then used to update each local element stiffness matrix as $\Mat{R}^e\Mat{K}_0{\Mat{R}^e}^{\top}$
 whereas the deformations are used to compute the linear strain in the local corotational frame.
@@
 close to 2, which is in perfect agreement with the results reported in~\cite{Hollenstein2006}.
+Figure~\ref{fig-aspiration3} presents the evolution of displacement in time at central node on the cube surface. The effect of increased combined stiffness is clearly visible.
 

### 2013-07-29 Doc/Vrst2013/vrst2013.tex
FIX: typos
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@602 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 The measures performed on the capsule report stiffness which is significantly higher than that of the parenchyma, however, 
-the thickness of the membrane does not exceed 100$\mu$m. For this reasons, it is not possible to model the 
+the thickness of the membrane does not exceed 100$\mu$m. For this reason, it is not possible to model the 
 capsule with standard volume elements usually employed for the parenchyma. In our approach, we rely of 
@@
 %the authors propose a composite model allowing for real-time simulation of entire liver with vascularization.
-Since a detailed modeling of the vessels would be extremely costly (mainly because of the small thickness of the vessel wall, 
+Since a detailed modeling of the vessels would be extremely costly (mainly because of the small thickness of the vessel wall), 
 the authors propose a composite model allowing for real-time simulation.
@@
       \hline
-      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
+      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\.
   \end{array}\right]
@@
 %
-Then the assembled stiffness matrix for the element is:
+Hence, the assembled stiffness matrix for the element is
 %
@@
 %
-Where $\Mat{K}^m$ is the stiffness matrix of the triangular membrane.
+where $\Mat{K}^m$ is the stiffness matrix of the triangular membrane.
 
@@
 During the contact with an instrument such as probe, needle, scalpel and others,
-a specific deformations take place in the vicinity of
+specific deformations take place in the vicinity of
 the instrument. This type of deformation may not necessarily induce the
@@
 close to 2, which is in perfect agreement with the results reported in~\cite{Hollenstein2006}.
-Figure~\ref{fig-aspiration3} presents the evolution of displacement in time at central node on the cube surface. The effect of increased combined stiffness is clearly visible.
+Fig.~\ref{fig-aspiration3} presents the evolution of displacement in time at central node on the cube surface. The effect of increased combined stiffness is clearly visible.
 

### 2013-07-31 Doc/MMVR2014/abstract.tex
MMVR 2014 paper - 4 pages
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@603 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\begin{frontmatter}%
\begin{abstract}%
Accurate biomechanical modeling of liver is of a paramount interest for pre-operative planning or computer-aided per-operative guidance. 
However, such simulations remain a challenging task, since the organ is composed of three constituents: parenchyma, vascularization and Glisson's capsule, 
each having different mechanical properties. 
%
%While a real-time simulation of vascularized liver has been already described in~\cite{Peterlik2012}, the Glisson's capsule has not been included in the modeling. 
%In~\cite{Hollenstein2006,Ahn2010} it is shown that the capsule plays an important role mainly in the vicinity of the surgical tools. 
%Therefore, the accuracy of the liver response in the vicinity of the surgical tool requires correct modeling of the capsule as a component of the liver model. 
%
In this paper we propose a complete liver model, where the parenchyma is modeled as a finite element volume, the vessels are modeled as series of beam elements, and the   corotational membrane elements based on constant-strain formulation, which are coupled mechanically to the underlying tetrahedra. 
We show that using this model we are able to reproduce accurately biomechanical tests, while maintaining the real-time aspect of the simulation.

%The measures performed on the capsule report stiffness which is significantly higher than that of the parenchyma, however, the thickness of the membrane does not exceed 100$\mu$m. 
%For this reason, it is not possible to model the capsule with standard volume elements usually employed for the parenchyma. 
%In our approach, we rely of corotational membrane elements based on constant-strain formulation, which are coupled mechanically to the underlying tetrahedra.

\end{abstract}%
% 
\begin{keyword}%
Simulation , modeling, liver, biomechanics
\end{keyword}%
% 
\end{frontmatter}%

### 2013-07-31 Doc/MMVR2014/figures.tex
MMVR 2014 paper - 4 pages
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@603 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2013-07-28 Doc/Vrst2013/vrst2013.tex
@@
-\documentclass[sposored]{acmsiggraph}
-
-\usepackage{amsmath}
-\usepackage{subfig}
-
-\def\wrt{w.\,r.\,t.}
-\def\eg{e.\,g.}
-\def\ie{i.\,e.}
-\def\Dash{\nobreak\,---\penalty-500\,}
-
-\renewcommand{\Vec}[1]{\mathbf{#1}}
-\newcommand{\Mat}[1]{\mathbf{#1}}
-
-\title{Modeling Liver with Glisson's Capsule in Real-Time Medical Simulations}
-
-%TODO
-\author{
-\begin{tabular}[t]{cccc}%
-  Tom\'a\v{s} Golembiovsk\'y\thanks{nyoxi@ics.muni.cz} &
-  Igor Peterl\'ik &
-  Christian Duriez &
-  St\'ephane Cotin \\
-%
-  Faculty of Informatics &
-  Institut Hospitalo-Universitaire &
-  INRIA &
-  INRIA \\
-%
-  Masaryk University &
-  Strasbourg, France &
-  Lille, France &
-  Lille, France \\
-%
-  Brno, Czech Republic
-\end{tabular}
-}
-%\affiliation{foobar}
-%\numberofauthors{4}
-%\author{%
-%Tom\'a\v{s} Golembiovsk\'y\\
-%       \affaddr{Faculty of Informatics}\\
-%       \affaddr{Masaryk University}\\
-%       \affaddr{Brno, Czech Republic}\\
-%       \email{nyoxi@ics.muni.cz}
-%% 2nd. author
-%\alignauthor
-%Igor Peterl\'ik\\
-%       \affaddr{Institut Hospitalo-Universitaire}\\
-%       \affaddr{Strasbourg, France}\\
-%       \email{Igor.Peterlik@ihu-strasbourg.eu}
-%% 3rd. author
-%\alignauthor
-%Christian Duriez\\
-%       \affaddr{INRIA}\\
-%       \affaddr{Lille, France}\\
-%       \email{Christian.Duriez@inria.fr}
-%\and
-%% 4th. author
-%\alignauthor
-%St\'ephane Cotin\\
-%       \affaddr{INRIA}\\
-%       \affaddr{Lille, France}\\
-%       \email{Stephane.Cotin@inria.fr}
-%}
-\pdfauthor{Tom\'a\v{s} Golembiovsk\'y}
-
-\begin{document}
-
-\maketitle
-\begin{abstract}
-The accurate mechanical modeling of liver is of a paramount interest in simulation-based operation planning 
-and computer-aided per-operative guidance. However, the realistic simulation of liver behaviour is a challenging task, 
-since the organ is composed of three constituents: parenchyma, vascularization and Glisson's capsule, 
-each having different mechanical properties.
-
-While a real-time simulation of vascularized liver has been already described in~\cite{Peterlik2012}, the Glisson's 
-capsule has not been included in the modeling. In~\cite{Hollenstein2006,Ahn2010} it is shown that the capsule 
-plays an important role mainly in the vicinity of the surgical tools. Therefore, the accuracy of the liver response 
-in the vicinity of the surgical tool requires correct modeling of the capsule as a component of the liver model.
-
-The measures performed on the capsule report stiffness which is significantly higher that that of the parenchyma, however, 
-the thickness of the membrane does not exceed 100$\mu$m. For this reasons, it is not possible to model the 
-capsule with standard volume elements usually employed for the parenchyma. In our approach, we rely of 
-corotational membrane elements based on constant-strain formulation, which are coupled mechanically to the underlying 
-tetrahedra. We show that using this model we are able to reproduce the experiments reported in~\cite{Hollenstein2006}, 
-while maintaining the real-time aspect of the simulation.
-
-%This work provides a complete physical model for the deformations of the liver with its three constituents: parenchyma, vessels and the Glisson's capsule.
-%Some recent experimental results \cite{Ahn2010} have shown that there is a significant difference of stiffness between the Glisson's capsule and the parenchyma. 
-%Moreover, it has been observed that the capsule plays an important role for the deformation at the local level \cite{Hollenstein2006}.
-%However, the Glisson's capsule is very thin and its influence for the global deformations of the liver was never highlighted.
-%In this work, we propose to use a membrane FEM model for the capsule, that is coupled with a vascularized model of the liver.
-%We show that the approach is able to reproduce the deformations observed at the local level.
-%Additionally, we show, by simulating natural global deformations of the liver observed on real images (due to gravity), that the results have important differences with and without the capsule.
-
-\end{abstract}
-
-% Categories
-\begin{CRcatlist}
-  \CRcat{I.6.5}{Simulation and Modeling}{Model Development}
-  \CRcat{J.3}{Life and Medical Sciences}{Health};
-\end{CRcatlist}
-
-\keywords{deformable liver model, Glisson's capsule}
-
-\keywordlist
-
-\TOGlinkslist
-
-\copyrightspace
-
-\section{Introduction} %{{{
-
-According to the statistics, nearly 100,000 European citizens die of cirrhosis or liver cancer each year. 
-Although new methods such as radio-frequency- and cryo-ablation known in the interventional radiology 
-seem to be promising, surgery remains the option that offers the foremost success rate against these pathologies. 
-Nevertheless, surgery is not always performed due to several limitations, in particular the determination 
-of accurate eligibility criteria for the patient. 
-In this context, the pre-operational planning becomes a crucial task having a significant impact on the treatment.
-
-Computer-aided physics-based medical simulation has proven to be an extremely useful technique in the area of medical training. 
-Nevertheless, whereas generic models are usually required in training simulators, accurate patient-specific modeling
-becomes necessary as soon as computer simulation is to be employed in the pre-operative planning. At the same time, 
-interactivity of such models remains an important aspect, requiring real-time simulation which is often difficult to 
-achieve given the complexity of soft tissues.
-
-When simulating the behavior of human liver, the task of real-time accurate modeling is challenging, mainly because of the complex structure 
-of this organ composed of three main constituents: \emph{parenchyma}, \emph{vascular networks} and \emph{Glisson's capsule}.
-The parenchyma has certainly been the most studied component of the liver; actually researchers agree on hyperelastic 
-properties of the tissue, for which the mechanical parameters have been reported for example in~\cite{Kerdok2006,Gao2009}. 
-Moreover, several methods have been proposed to model the hyperelastic behaviour at real-time rates, such as multiplicative Jacobian decomposition
-introduced in~\cite{Marchesseau2010}.
-
-%\TG{The following paragraph is optional. Unnecessary but adds to the global context}
-The mechanical importance of the vascular structures in liver is studied in~\cite{Peterlik2012}. It shows that the 
-influence of vessels on the mechanical behaviour of the organ is significant. %, mainly if large deformations occur. 
-%Since a detailed modeling of the vessels would be extremely costly (mainly because of the small thickness of the vessel wall, 
-%the authors propose a composite model allowing for real-time simulation of entire liver with vascularization.
-Since a detailed modeling of the vessels would be extremely costly (mainly because of the small thickness of the vessel wall, 
-the authors propose a composite model allowing for real-time simulation.
-
-Rather a small number of studies have been conducted dealing with the third liver constituent, the  Glisson's capsule.
-Quantitative results of experiments on a porcine liver have been published in~\cite{Umale2011}; the measurements indicate that although being very 
-thin (10--20$\mu m$), the capsule shows to be stiff in tensile tests: the Young's modulus of the capsule reported to be $8.22\pm3.42$\,MPa 
-exceeds the values for the parenchyma by three orders of magnitude.
-This suggests that the mechanical influence of the membrane on the liver behaviour is not negligible.
-
-In~\cite{Hollenstein2006}, a local influence of the capsule has been measured using a special aspiration device. The study was then repeated 
-in vivo on human patients during the operation, confirming the mechanical importance of the membrane~\cite{Ahn2010,Nava2008}.
-%To our best knowledge, no attempt has been made to demonstrate the role of the Glisson's capsule on the global behaviour of 
-%the liver, mainly if the organ undergoes large deformations which is often the case during the surgery.
-
-In this paper we present preliminary results of our work on the complete liver
-model. We present a real-time composite model accounting for parenchyma and
-Glisson's capsule. The model is based on two different finite element
-representations for each constituent coupled together.
-%The main contribution of the paper is twofold: first, we present a real-time model of the liver, including
-%the parenchyma, vascularization and Glisson's capsule. The model is based on three different finite element representations for each constituent,
-%linked together via mechanical coupling.
-We show that the model mimics the \emph{local} experiments described in~\cite{Hollenstein2006}.
-
-%Second, we use the complete model of the liver to demonstrate the \emph{global} influence of the Glisson's
-%capsule via simulation: using a specific model of a porcine liver built from CT contrast-enhanced data, we show that there is a significant 
-%difference in the response of the model with and without the capsule in the case when the liver undergoes large deformations.
-
-The paper is organized as follows: first we describe the proposed model of
-liver with capsule. Second, we validate our model in context of local
-deformations by reproducing the aspiration test described
-in~\cite{Hollenstein2006}. We conclude by summarizing future steps towards the
-complete liver model.
-%Finally, we demonstrate that in spite of its small thickness, the Glisson's 
-%capsule has a global influence on the liver undergoing large deformations.
-
-%}}}
-
-\section{Model} %{{{
-
-In this section we describe the construction of a composite model
-based on two components: tetrahedral FE model of the 
-parenchyma and
-triangular membrane elements used for the capsule.
-
-\subsection{Parenchyma} %{{{
-
-It is known that the parenchyma exhibits non-linear viscoelastic behaviour \cite{Marchesseau2010}.
-However, as we are mainly interested in the static equilibrium, we do not model the time-dependent
-phenomena related to viscosity.
-
-%However, we employ simpler corotational model as we are not so much
-%interested in time-dependent behaviour but rather in static equilibrium
-%under certain conditions.
-%We also rely on the vascularized model of the parenchyma proposed by
-%Peterl\'{i}k et al. \cite{Peterlik2012}.
-
-The parenchyma is modeled using corotational finite elements~\cite{Felippa2005}.
-Although it relies on linear stress-strain relationship, large displacements including rotations are modeled correctly. 
-While in the full non-linear formulation the stiffness matrix relates the forces $\Vec{f}$ and 
-displacements $\Vec{u}$ as $\Vec{f} = K(\Vec{u})$, the corotational model 
-requires the stiffness matrix $\Mat{K}_0$ of the system to be computed only once before the simulation begins. 
-Then, in each step, the motion of each element $e$ is decomposed into rigid rotation $\Mat{R}^e$ and local deformation. 
-The rotations are then used to update each local element stiffness matrix as $\Mat{R}\Mat{K}_0\Mat{R}^{\top}$
-whereas the deformations are used to compute the linear strain in the local corotational frame.
-There are several ways of computing the corotational frame for elements; we rely on
-the geometrical method proposed in \cite{Nesme2005}.
-% NOTE: This description of corotational method is very simplified and could be extended.
-
-%The model of the vascularization is based on linear beams 
-%with local frames of reference~\cite{Duriez2006}; in many aspects it's similar to the
-%corotational formulation described above. As such the model also handles geometric
-%non-linearities in the deformation. Through a specification of cross section and moments of inertia, 
-%the model can account for the specific properties of the blood vessels.
-
-%The beam-based model of vessels is mechanically coupled to the parenchyma as described in~\cite{Peterlik2012}. 
-%The coupling assumes that there is no relative motion between the vessels and the surrounding parenchyma. 
-%During the simulation, the nodes of linked beams are first displaced and rotated according to the actual motion of associated tetrahedra. 
-%As the deformation of beams results in mechanical response represented by forces and torques, these are propagated back to 
-%the tetrahedral FE model.
-
-%}}}
-
-\subsection{Glisson's Capsule} %{{{
-\label{ss:capsuleModel}
-The thickness of the Glisson's capsule is relatively small: the values in range of 10-20
-$\mu$m have been reported in~\cite{Umale2011}.
-It is not possible to model such thin structure with classical tetrahedral
-elements, if the real-time aspect of the simulation is to be preserved.
-Furthermore, modeling both the tissue and the capsule would require an extremely 
-dense mesh to avoid numerical instabilities and thus would significantly
-violate the speed requirements imposed for medical simulators.
-Instead, modeling the capsule with two-dimensional elements that abstract from the
-thickness in the third dimension seems
-to be a natural choice. In the elasticity theory, this functionality is usually provided by membrane and shell elements.
-Based on the observation of its behaviour, we also
-assume negligible bending forces and propose a model based on membrane
-elements. 
-To maintain simplicity of the composite model we choose simple triangular
-elements with constant strain (CST).
-
-The computation of elastic stiffness matrix follows the common derivation
-%
-\begin{eqnarray}
-  \Mat{K}^m & = & \int_V \Mat{B}^T \Mat{E} \Mat{B} dV     \label{mem1} \\
-            & = & h \int_A \Mat{B}^T \Mat{E} \Mat{B} dA   \label{mem2} \\
-            & = & h A \Mat{B}^T \Mat{E} \Mat{B}           \label{mem3}
-\end{eqnarray}
-%
-where $\Mat{B}$ is the strain-displacement matrix, $\Mat{E}$ the material
-matrix, $h$ is the thickness and $A$ area of the element. In the previous
-\eqref{mem2} follows from the fact that we assume constant thickness of the
-element and \eqref{mem3} follows from the fact that the strain-displacement
-matrix is constant in our case. The strain-displacement matrix for the CST
-element can be expressed as:
-%
-\begin{equation}
-  \Mat{B} = \frac{1}{2A} \begin{bmatrix}
-    y_{23} & 0      & y_{31} & 0      & y_{12} & 0 \\
-         0 & x_{32} & 0      & x_{13} & 0      & x_{21} \\
-    x_{32} & y_{23} & x_{13} & y_{31} & x_{21} & y_{12}
-  \end{bmatrix}
-\end{equation}
-%
-The values $x_{ij} = x_i - x_j$ and $y_{ij} = y_i - y_j$ are computed from
-the $x$ or $y$ coordinates of the nodes $i,j$ of the triangular element.
-The reader can refer to the respective literature~\cite{Felippa2003} for more thorough
-description.
-
-Similarly as with model of parenchyma we use linear elastic material and employ
-the corotational formulation for the CST elements.
-
-%}}}
-
-\subsection{Coupling Between Capsule and\\Parenchyma} %{{{
-The literature reports high cohesion between capsule and parenchyma.
-Based on this property we assume there is no relative motion of the capsule \wrt\ the parenchyma.
-Although an arbitrary surface mesh could be used to model the capsule, we exploit 
-the fact that the parenchyma is modeled by tetrahedral elements having
-triangular faces. Thus, as the boundary of the volumetric mesh is already
-triangulated, we simply employ the triangles on the mesh surface to model the capsule.
-
-Using directly the boundary of the tetrahedral mesh does not only solve the
-problem of building the surface mesh, but has one more advantage: the nodes
-of the triangular mesh coincide with the nodes of the tetrahedral mesh, so no projection of one mesh onto the other is needed.
-Moreover, the stiffness matrices for capsule and parenchyma are then easily assembled together.
-% and solved as one system.
-%
-%\CD{The following may appear as "trivial"... maybe we can remove this part at the end if we need space} 
-Without the loss of generality we can assume the tetrahedron consists of
-nodes $p_1, p_2, p_3$ and $p_4$ and the boundary triangle has nodes $p_1, p_2$
-and $p_3$. We can reorder the degrees of freedom so that the stiffness
-matrix $\Mat{K}^t$ for the tetrahedron can be written as:
-%
-\begin{equation}
-  \Mat{K}^t = \left[\begin{array}{c|c}
-      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
-      \hline
-      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
-  \end{array}\right]
-\end{equation}
-%
-Then the assembled stiffness matrix for the element is:
-%
-\begin{equation}
-  \Mat{K} = \left[\begin{array}{c|c}
-      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
-      \hline
-      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
-  \end{array}\right]
-  +
-  \left[\begin{array}{c|c}
-      \Mat{K}^m & 0 \\
-      \hline
-      0 & 0 \\
-  \end{array}\right]
-\end{equation}
-%
-Where $\Mat{K}^m$ is the stiffness matrix of the triangular membrane.
-
-The resulting system of linear equations is solved by direct solver based on Cholesky decomposition.
-
-%Alternatively, one can use a mechanical coupling similar to the one used in
-%\cite{Peterlik2012} to be able to use an arbitrary surface mesh. Nevertheless,
-%for conforming triangular mesh both methods lead to the same solution.
-
-%}}}
-
-%}}}
-
-\section{Methods} %{{{
-The model presented in the previous section was implemented in 
-SOFA\footnote{www.sofa-framework.org} and a set of
-numerical simulations was performed.
-In this section we provide comparison of local deformations with the
-results reported in literature to validate the method.
-%Second, to show that in spite of its very small thickness the membrane cannot be
-%neglected even in the context of global deformations and its overall
-%stiffness plays an important role, the model of
-%the complete liver was subjected to global deformations.
-
-During the contact with an instrument such as probe, needle, scalpel and others,
-a specific deformations take place in the vicinity of
-the instrument. This type of deformation may not necessarily induce the
-deformation of the object as a whole and therefore can be considered as
-local. Correct material properties are not only important to quantify the
-displacement, but also play an important role in capturing the correct area of
-the deformation or its profile near the instrument.
+\newpage
 
@@
 \end{figure}
-
-A good example of such a local deformation is the aspiration test
-where the response of liver exposed locally to a negative pressure is measured.
-The aspiration device consists of a tube having 1\,cm in diameter and allows to
-control the pressure inside the tube. The test is performed by
-attaching the tube to the tissue and measuring the tissue response. We
-set up a simulation in SOFA to reproduce the experiment (see
-Fig.~\ref{fig-aspiration1}): we meshed a 15$\times$15$\times$15\,mm$^3$ 
-cube representing the tissue resulting in 2648 tetrahedra. Then we attached a 1\,cm tube 
-and applied a pressure of 3\,kPa inside the tube.
-
-Since the tube is in direct contact with the tissue, uni-lateral constraints with friction were chosen 
-to model this interaction properly. We opted for a method based on \emph{non-linear complementarity problem}  (NLCP)
-where the non-linearity is introduced due to the friction. The NLCP
-allows for solution of the Signorini's problem to avoid any interpenetration between the colliding 
-objects (see~\cite{Duriez2006b} for details). Since NLCP requires explicitly the computation of compliance matrix which 
-is homogeneous to the inverse of the stiffness matrix, we employed a direct solver based on LDL decomposition to 
-solve the system and compute the inverse matrices.
-
-\section{Results and Discussion} %{{{
-As the first step, we evaluate the influence of the Glisson's capsule on the mechanical response of the tissue during
-the aspiration test. 
-In~\cite{Hollenstein2006} it is reported that the influence of the capsule is significant, as modeling 
-only the parenchyma without considering the membrane results in overestimation of the deformation by a factor of 2 to 3. 
-In order to verify this observation, we performed a simulation using capsule parameters measured experimentally on 
-ex-vivo pig liver~\cite{Umale2011}: Young's modulus of the membrane was set to $E_c$=8.22\,MPa and thickness was
-$t_c$=20\,$\mu$m. 
-Since the values for parenchyma elasticity reported in the literature vary significantly being usually reported between 2\,kPa to 5\,kPa, 
-we employed a value $E_p$=3.5\,kPa for the Young's modulus of the parenchyma.
 
@@
 \end{figure}
-
-In Fig.~\ref{fig-aspiration2} the profiles of cuts in the middle of the
-test cube are presented showing significantly larger deformation in the model without capsule assuming that the same negative pressure was applied 
-on the tissue surface inside the tube. Moreover, it can be observed that the deformation without capsule is overestimated by a factor 
-close to 2, which is in perfect agreement with the results reported in~\cite{Hollenstein2006}.
-
-In the second step, we employed the simulation to reproduce the displacements values obtained via simulation in~\cite{Hollenstein2006,Nava2008}.
-Whereas the simulation in the referenced work was done on a mesh composed of tetrahedral elements for both the parenchyma 
-and the capsule, we modeled the capsule using the triangular elements as shown in section~\ref{ss:capsuleModel}.
-
-Since the Young's moduli of neither parenchyma ($E_p$) nor capsule ($E_c$) were specified exactly, we obtained both as follows: first, 
-we performed a set of simulations without the capsule for different values of $E_p$: the value that provided a good match 
-with reported displacements was $E_p$=30\,kPa. 
-Then, we fixed the thickness of the capsule to be $t$=93\,$\mu$m (reported as the average value in the paper) and 
-using $E_p$=30\,kPa, we ran the simulation repeatedly using the model with the capsule. In each simulation we 
-used different value $E_c$ of the Young's modulus for the capsule in order to minimize the displacement error w.r.t. the 
-reported values. The minimal value of error (not exceeding 1\%) was obtained for $E_c$=3\,MPa. This 
-value lies in the range of capsule elasticity coefficients reported by experimental measurements.
-
-%Moreover, we are able to reproduce
-%the measurements presented in~\cite{Hollenstein2006,Nava2008} with less than 1\% error.
-%We do not model the time-dependent behaviour though.
-%In this simulation we prescribe the the internal pressure of 30\,kPa. The other
-%parameters of the model were obtained like this: First we fix the stiffness of
-%parenchyma to $E_p$=30\,kPa to match the scenerio without capsule. Then,
-%assuming the thickness $t$=93\,$\mu$m of the capsule (reported as average in
-%the paper) we run several simulations with different capsule stiffness $E_c$.
-%We have found the lowest error is when $E_c$=3\,MPa. All the stiffness
-%parameters lie within the ranges measured in the paper.
-
-%In this case the simulation was performed with the pressure of 30\,kPa and with
-%the following elastic parameters: $E_p$=30\,kPa for parenchyma and $E_c$=3\,MPa
-%and $t$=93\,$\mu$m for capsule.
-%All the values lie within the measurements
-%specified in the paper. \TG{add figure like Fig. 2?}
-
-%The influence of the capsule on local level has been published in literature
-%before. However, to our knowledge it's influence on global scale deformation
-%has not been studied yet. We believe that the capsule has non-negligible impact
-%also on global level.
-
-In the scenario presented in this paper, we focus on local influence of the capsule.
-Nevertheless, it is possible to employ the actual model to demonstrate the global 
-impact of the capsule and to further validate the accuracy of such a complete model w.r.t. experimental measures. 
-Detailed description of such validation is however beyond the scope of this paper. 
-In spite of this we would like to emphasize that we have already integrated the
-capsule model with vascularized liver model described in~\cite{Peterlik2012}. 
-Adding the capsule did not affect the
-performance significantly: a visual refresh rate exceeding 25\,FPS was achieved on a
-PC with CPU Intel i7 running at 3.00 GHz.
-This suggests that the proposed technique is compatible with real-time simulation of whole
-organ while modeling the local properties accurately.
-
-%In our future work we need to validate the model in the
-%context of global and local deformations. This task is challenging because
-%comparing in vivo measurements with the simulation is difficult due to complex
-%boundary conditions. 
-%\TG{ex-vivo: no information abou vessels; phantoms: cannot model capsule}
-
-%}}}
-
-\section{Conclusions and Future Work} %{{{
-In the paper we presented a model of tissue with capsule based on two separated models for parenchyma (corotational tetrahedral elements)
-and membrane (CST triangular elements). Employing a simulation scenario known as aspiration test, 
-we compared the model with experimental data provided in the literature 
-and verified the results obtained in the simulation w.r.t.
-to a standard computationally expensive model based entirely on volume elements~\cite{Hollenstein2006}.
-We obtained values which were in good correspondence with results reported in the literature.
-
-Presently we are focusing on the validation of the model also in context of
-global deformations of the organ where we believe the capsule plays an
-essential role as well. This task is however complicated due to the limited number of experimental studies dealing with 
-both the parenchyma and the capsule.
-
-%Further, the model was used to demonstrate global influence of the capsule when the organ 
-%undergoes large deformations.
-
-%In the future work, we plan to employ the proposed model for other organs where the capsule play an important role
-%(for example kidneys). We will also focus on better description of the mechanical interactions and couplings among the organs and tissues in the abdominal 
-%cavity via realistic modeling of boundary conditions using mechanical constraints and contacts.
-
-%}}}
-
-%
-% ---- Bibliography ----
-%
-
-\bibliographystyle{acmsiggraph}
-\bibliography{bibdata}
-
-\end{document}
-% spell: setl spell spelllang=en spellfile=spelldict.en.utf-8.add
-% vim:set et sw=2 fdm=marker fdl=3 fdc=4 isk+=_,-:

### 2013-07-31 Doc/MMVR2014/intro.tex
MMVR 2014 paper - 4 pages
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@603 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2013-07-29 Doc/Vrst2013/vrst2013.tex
@@
-\documentclass[sposored]{acmsiggraph}
-
-\usepackage{amsmath}
-\usepackage{subfig}
-
-\def\wrt{w.\,r.\,t.}
-\def\eg{e.\,g.}
-\def\ie{i.\,e.}
-\def\Dash{\nobreak\,---\penalty-500\,}
-
-\renewcommand{\Vec}[1]{\mathbf{#1}}
-\newcommand{\Mat}[1]{\mathbf{#1}}
-
-\title{Modeling Liver with Glisson's Capsule in Real-Time Medical Simulations}
-
-%TODO
-\author{
-\begin{tabular}[t]{cccc}%
-  Tom\'a\v{s} Golembiovsk\'y\thanks{nyoxi@ics.muni.cz} &
-  Igor Peterl\'ik &
-  Christian Duriez &
-  St\'ephane Cotin \\
-%
-  Faculty of Informatics &
-  Institut Hospitalo-Universitaire &
-  INRIA &
-  INRIA \\
-%
-  Masaryk University &
-  Strasbourg, France &
-  Lille, France &
-  Lille, France \\
-%
-  Brno, Czech Republic
-\end{tabular}
-}
-%\affiliation{foobar}
-%\numberofauthors{4}
-%\author{%
-%Tom\'a\v{s} Golembiovsk\'y\\
-%       \affaddr{Faculty of Informatics}\\
-%       \affaddr{Masaryk University}\\
-%       \affaddr{Brno, Czech Republic}\\
-%       \email{nyoxi@ics.muni.cz}
-%% 2nd. author
-%\alignauthor
-%Igor Peterl\'ik\\
-%       \affaddr{Institut Hospitalo-Universitaire}\\
-%       \affaddr{Strasbourg, France}\\
-%       \email{Igor.Peterlik@ihu-strasbourg.eu}
-%% 3rd. author
-%\alignauthor
-%Christian Duriez\\
-%       \affaddr{INRIA}\\
-%       \affaddr{Lille, France}\\
-%       \email{Christian.Duriez@inria.fr}
-%\and
-%% 4th. author
-%\alignauthor
-%St\'ephane Cotin\\
-%       \affaddr{INRIA}\\
-%       \affaddr{Lille, France}\\
-%       \email{Stephane.Cotin@inria.fr}
-%}
-\pdfauthor{Tom\'a\v{s} Golembiovsk\'y}
-
-\begin{document}
-
-\maketitle
-\begin{abstract}
-The accurate mechanical modeling of liver is of a paramount interest in simulation-based operation planning 
-and computer-aided per-operative guidance. However, the realistic simulation of liver behaviour is a challenging task, 
-since the organ is composed of three constituents: parenchyma, vascularization and Glisson's capsule, 
-each having different mechanical properties.
-
-While a real-time simulation of vascularized liver has been already described in~\cite{Peterlik2012}, the Glisson's 
-capsule has not been included in the modeling. In~\cite{Hollenstein2006,Ahn2010} it is shown that the capsule 
-plays an important role mainly in the vicinity of the surgical tools. Therefore, the accuracy of the liver response 
-in the vicinity of the surgical tool requires correct modeling of the capsule as a component of the liver model.
-
-The measures performed on the capsule report stiffness which is significantly higher than that of the parenchyma, however, 
-the thickness of the membrane does not exceed 100$\mu$m. For this reason, it is not possible to model the 
-capsule with standard volume elements usually employed for the parenchyma. In our approach, we rely of 
-corotational membrane elements based on constant-strain formulation, which are coupled mechanically to the underlying 
-tetrahedra. We show that using this model we are able to reproduce the experiments reported in~\cite{Hollenstein2006}, 
-while maintaining the real-time aspect of the simulation.
-
-%This work provides a complete physical model for the deformations of the liver with its three constituents: parenchyma, vessels and the Glisson's capsule.
-%Some recent experimental results \cite{Ahn2010} have shown that there is a significant difference of stiffness between the Glisson's capsule and the parenchyma. 
-%Moreover, it has been observed that the capsule plays an important role for the deformation at the local level \cite{Hollenstein2006}.
-%However, the Glisson's capsule is very thin and its influence for the global deformations of the liver was never highlighted.
-%In this work, we propose to use a membrane FEM model for the capsule, that is coupled with a vascularized model of the liver.
-%We show that the approach is able to reproduce the deformations observed at the local level.
-%Additionally, we show, by simulating natural global deformations of the liver observed on real images (due to gravity), that the results have important differences with and without the capsule.
-
-\end{abstract}
-
-% Categories
-\begin{CRcatlist}
-  \CRcat{I.6.5}{Simulation and Modeling}{Model Development}
-  \CRcat{J.3}{Life and Medical Sciences}{Health};
-\end{CRcatlist}
-
-\keywords{deformable liver model, Glisson's capsule}
-
-\keywordlist
-
-\TOGlinkslist
-
-\copyrightspace
-
-\section{Introduction} %{{{
-
+\section{Introduction}%
+\vspace{-6pt}
 According to the statistics, nearly 100,000 European citizens die of cirrhosis or liver cancer each year. 
@@
 achieve given the complexity of soft tissues. 
-
+%
 When simulating the behavior of human liver, the task of real-time accurate modeling is challenging, mainly because of the complex structure 
@@
 introduced in~\cite{Marchesseau2010}.
-
+%
 %\TG{The following paragraph is optional. Unnecessary but adds to the global context}
@@
 This suggests that the mechanical influence of the membrane on the liver behaviour is not negligible.
-
+%
 In~\cite{Hollenstein2006}, a local influence of the capsule has been measured using a special aspiration device. The study was then repeated 
@@
 
-In this paper we present preliminary results of our work on the complete liver
-model. We present a real-time composite model accounting for parenchyma and
-Glisson's capsule. The model is based on two different finite element
-representations for each constituent coupled together.
+In this paper we present preliminary results of our work on the complete liver model. 
+We present a real-time composite model accounting for parenchyma and Glisson's capsule. 
+The model is based on two different finite element representations for each constituent coupled together.
 %The main contribution of the paper is twofold: first, we present a real-time model of the liver, including
@@
 
-The paper is organized as follows: first we describe the proposed model of
-liver with capsule. Second, we validate our model in context of local
-deformations by reproducing the aspiration test described
-in~\cite{Hollenstein2006}. We conclude by summarizing future steps towards the
-complete liver model.
+%The paper is organized as follows: first we describe the proposed model of liver with capsule. 
+%Second, we validate our model in context of local deformations by reproducing the aspiration test described in~\cite{Hollenstein2006}.
+We conclude by summarizing future steps towards the complete liver model.
 %Finally, we demonstrate that in spite of its small thickness, the Glisson's 
 %capsule has a global influence on the liver undergoing large deformations.
-
-%}}}
-
-\section{Model} %{{{
-
-In this section we describe the construction of a composite model
-based on two components: tetrahedral FE model of the 
-parenchyma and
-triangular membrane elements used for the capsule.
-
-\subsection{Parenchyma} %{{{
-
-It is known that the parenchyma exhibits non-linear viscoelastic behaviour \cite{Marchesseau2010}.
-However, as we are mainly interested in the static equilibrium, we do not model the time-dependent
-phenomena related to viscosity.
-
-%However, we employ simpler corotational model as we are not so much
-%interested in time-dependent behaviour but rather in static equilibrium
-%under certain conditions.
-%We also rely on the vascularized model of the parenchyma proposed by
-%Peterl\'{i}k et al. \cite{Peterlik2012}.
-
-The parenchyma is modeled using corotational finite elements~\cite{Felippa2005}.
-Although it relies on linear stress-strain relationship, large displacements including rotations are modeled correctly. 
-While in the full non-linear formulation the stiffness matrix relates the forces $\Vec{f}$ and 
-displacements $\Vec{u}$ as $\Vec{f} = K(\Vec{u})$, the corotational model 
-requires the stiffness matrix $\Mat{K}_0$ of the system to be computed only once before the simulation begins. 
-Then, in each step, the motion of each element $e$ is decomposed into rigid rotation $\Mat{R}^e$ and local deformation. 
-The rotations are then used to update each local element stiffness matrix as $\Mat{R}^e\Mat{K}_0{\Mat{R}^e}^{\top}$
-whereas the deformations are used to compute the linear strain in the local corotational frame.
-There are several ways of computing the corotational frame for elements; we rely on
-the geometrical method proposed in \cite{Nesme2005}.
-% NOTE: This description of corotational method is very simplified and could be extended.
-
-%The model of the vascularization is based on linear beams 
-%with local frames of reference~\cite{Duriez2006}; in many aspects it's similar to the
-%corotational formulation described above. As such the model also handles geometric
-%non-linearities in the deformation. Through a specification of cross section and moments of inertia, 
-%the model can account for the specific properties of the blood vessels.
-
-%The beam-based model of vessels is mechanically coupled to the parenchyma as described in~\cite{Peterlik2012}. 
-%The coupling assumes that there is no relative motion between the vessels and the surrounding parenchyma. 
-%During the simulation, the nodes of linked beams are first displaced and rotated according to the actual motion of associated tetrahedra. 
-%As the deformation of beams results in mechanical response represented by forces and torques, these are propagated back to 
-%the tetrahedral FE model.
-
-%}}}
-
-\subsection{Glisson's Capsule} %{{{
-\label{ss:capsuleModel}
-The thickness of the Glisson's capsule is relatively small: the values in range of 10-20
-$\mu$m have been reported in~\cite{Umale2011}.
-It is not possible to model such thin structure with classical tetrahedral
-elements, if the real-time aspect of the simulation is to be preserved.
-Furthermore, modeling both the tissue and the capsule would require an extremely 
-dense mesh to avoid numerical instabilities and thus would significantly
-violate the speed requirements imposed for medical simulators.
-Instead, modeling the capsule with two-dimensional elements that abstract from the
-thickness in the third dimension seems
-to be a natural choice. In the elasticity theory, this functionality is usually provided by membrane and shell elements.
-Based on the observation of its behaviour, we also
-assume negligible bending forces and propose a model based on membrane
-elements. 
-To maintain simplicity of the composite model we choose simple triangular
-elements with constant strain (CST).
-
-The computation of elastic stiffness matrix follows the common derivation
-%
-\begin{eqnarray}
-  \Mat{K}^m & = & \int_V \Mat{B}^T \Mat{E} \Mat{B} dV     \label{mem1} \\
-            & = & h \int_A \Mat{B}^T \Mat{E} \Mat{B} dA   \label{mem2} \\
-            & = & h A \Mat{B}^T \Mat{E} \Mat{B}           \label{mem3}
-\end{eqnarray}
-%
-where $\Mat{B}$ is the strain-displacement matrix, $\Mat{E}$ the material
-matrix, $h$ is the thickness and $A$ area of the element. In the previous
-\eqref{mem2} follows from the fact that we assume constant thickness of the
-element and \eqref{mem3} follows from the fact that the strain-displacement
-matrix is constant in our case. The strain-displacement matrix for the CST
-element can be expressed as:
-%
-\begin{equation}
-  \Mat{B} = \frac{1}{2A} \begin{bmatrix}
-    y_{23} & 0      & y_{31} & 0      & y_{12} & 0 \\
-         0 & x_{32} & 0      & x_{13} & 0      & x_{21} \\
-    x_{32} & y_{23} & x_{13} & y_{31} & x_{21} & y_{12}
-  \end{bmatrix}
-\end{equation}
-%
-The values $x_{ij} = x_i - x_j$ and $y_{ij} = y_i - y_j$ are computed from
-the $x$ or $y$ coordinates of the nodes $i,j$ of the triangular element.
-The reader can refer to the respective literature~\cite{Felippa2003} for more thorough
-description.
-
-Similarly as with model of parenchyma we use linear elastic material and employ
-the corotational formulation for the CST elements.
-
-%}}}
-
-\subsection{Coupling Between Capsule and\\Parenchyma} %{{{
-The literature reports high cohesion between capsule and parenchyma.
-Based on this property we assume there is no relative motion of the capsule \wrt\ the parenchyma.
-Although an arbitrary surface mesh could be used to model the capsule, we exploit 
-the fact that the parenchyma is modeled by tetrahedral elements having
-triangular faces. Thus, as the boundary of the volumetric mesh is already
-triangulated, we simply employ the triangles on the mesh surface to model the capsule.
-
-Using directly the boundary of the tetrahedral mesh does not only solve the
-problem of building the surface mesh, but has one more advantage: the nodes
-of the triangular mesh coincide with the nodes of the tetrahedral mesh, so no projection of one mesh onto the other is needed.
-Moreover, the stiffness matrices for capsule and parenchyma are then easily assembled together.
-% and solved as one system.
-%
-%\CD{The following may appear as "trivial"... maybe we can remove this part at the end if we need space} 
-Without the loss of generality we can assume the tetrahedron consists of
-nodes $p_1, p_2, p_3$ and $p_4$ and the boundary triangle has nodes $p_1, p_2$
-and $p_3$. We can reorder the degrees of freedom so that the stiffness
-matrix $\Mat{K}^t$ for the tetrahedron can be written as:
-%
-\begin{equation}
-  \Mat{K}^t = \left[\begin{array}{c|c}
-      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
-      \hline
-      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\.
-  \end{array}\right]
-\end{equation}
-%
-Hence, the assembled stiffness matrix for the element is
-%
-\begin{equation}
-  \Mat{K} = \left[\begin{array}{c|c}
-      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
-      \hline
-      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
-  \end{array}\right]
-  +
-  \left[\begin{array}{c|c}
-      \Mat{K}^m & 0 \\
-      \hline
-      0 & 0 \\
-  \end{array}\right]
-\end{equation}
-%
-where $\Mat{K}^m$ is the stiffness matrix of the triangular membrane.
-
-The resulting system of linear equations is solved by direct solver based on Cholesky decomposition.
-
-%Alternatively, one can use a mechanical coupling similar to the one used in
-%\cite{Peterlik2012} to be able to use an arbitrary surface mesh. Nevertheless,
-%for conforming triangular mesh both methods lead to the same solution.
-
-%}}}
-
-%}}}
-
-\section{Methods} %{{{
-The model presented in the previous section was implemented in 
-SOFA\footnote{www.sofa-framework.org} and a set of
-numerical simulations was performed.
-In this section we provide comparison of local deformations with the
-results reported in literature to validate the method.
-%Second, to show that in spite of its very small thickness the membrane cannot be
-%neglected even in the context of global deformations and its overall
-%stiffness plays an important role, the model of
-%the complete liver was subjected to global deformations.
-
-During the contact with an instrument such as probe, needle, scalpel and others,
-specific deformations take place in the vicinity of
-the instrument. This type of deformation may not necessarily induce the
-deformation of the object as a whole and therefore can be considered as
-local. Correct material properties are not only important to quantify the
-displacement, but also play an important role in capturing the correct area of
-the deformation or its profile near the instrument.
-
-\begin{figure}
-  \centering
-  \includegraphics[height=3.5cm]{aspiration.jpg}
-  \caption{\label{fig-aspiration1} The SOFA simulation scene of the aspiration test.}
-\end{figure}
-
-A good example of such a local deformation is the aspiration test
-where the response of liver exposed locally to a negative pressure is measured.
-The aspiration device consists of a tube having 1\,cm in diameter and allows to
-control the pressure inside the tube. The test is performed by
-attaching the tube to the tissue and measuring the tissue response. We
-set up a simulation in SOFA to reproduce the experiment (see
-Fig.~\ref{fig-aspiration1}): we meshed a 15$\times$15$\times$15\,mm$^3$ 
-cube representing the tissue resulting in 2648 tetrahedra. Then we attached a 1\,cm tube 
-and applied a pressure of 3\,kPa inside the tube.
-
-Since the tube is in direct contact with the tissue, uni-lateral constraints with friction were chosen 
-to model this interaction properly. We opted for a method based on \emph{non-linear complementarity problem}  (NLCP)
-where the non-linearity is introduced due to the friction. The NLCP
-allows for solution of the Signorini's problem to avoid any interpenetration between the colliding 
-objects (see~\cite{Duriez2006b} for details). Since NLCP requires explicitly the computation of compliance matrix which 
-is homogeneous to the inverse of the stiffness matrix, we employed a direct solver based on LDL decomposition to 
-solve the system and compute the inverse matrices.
-
-\section{Results and Discussion} %{{{
-As the first step, we evaluate the influence of the Glisson's capsule on the mechanical response of the tissue during
-the aspiration test. 
-In~\cite{Hollenstein2006} it is reported that the influence of the capsule is significant, as modeling 
-only the parenchyma without considering the membrane results in overestimation of the deformation by a factor of 2 to 3. 
-In order to verify this observation, we performed a simulation using capsule parameters measured experimentally on 
-ex-vivo pig liver~\cite{Umale2011}: Young's modulus of the membrane was set to $E_c$=8.22\,MPa and thickness was
-$t_c$=20\,$\mu$m. 
-Since the values for parenchyma elasticity reported in the literature vary significantly being usually reported between 2\,kPa to 5\,kPa, 
-we employed a value $E_p$=3.5\,kPa for the Young's modulus of the parenchyma.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=7cm]{aspiration.pdf}
-  \caption{\label{fig-aspiration2} Displacement profile of the cube in the
-  aspiration test with (blue) and without (black) the capsule.}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[height=3.5cm]{displacement.pdf}
-  \caption{\label{fig-aspiration3} Evolution of the displacement at the center
-  in time for test with (blue) and without (black) the capsule.}
-\end{figure}
-
-In Fig.~\ref{fig-aspiration2} the profiles of cuts in the middle of the
-test cube are presented showing significantly larger deformation in the model without capsule assuming that the same negative pressure was applied 
-on the tissue surface inside the tube. Moreover, it can be observed that the deformation without capsule is overestimated by a factor 
-close to 2, which is in perfect agreement with the results reported in~\cite{Hollenstein2006}.
-Fig.~\ref{fig-aspiration3} presents the evolution of displacement in time at central node on the cube surface. The effect of increased combined stiffness is clearly visible.
-
-In the second step, we employed the simulation to reproduce the displacements values obtained via simulation in~\cite{Hollenstein2006,Nava2008}.
-Whereas the simulation in the referenced work was done on a mesh composed of tetrahedral elements for both the parenchyma 
-and the capsule, we modeled the capsule using the triangular elements as shown in section~\ref{ss:capsuleModel}.
-
-Since the Young's moduli of neither parenchyma ($E_p$) nor capsule ($E_c$) were specified exactly, we obtained both as follows: first, 
-we performed a set of simulations without the capsule for different values of $E_p$: the value that provided a good match 
-with reported displacements was $E_p$=30\,kPa. 
-Then, we fixed the thickness of the capsule to be $t$=93\,$\mu$m (reported as the average value in the paper) and 
-using $E_p$=30\,kPa, we ran the simulation repeatedly using the model with the capsule. In each simulation we 
-used different value $E_c$ of the Young's modulus for the capsule in order to minimize the displacement error w.r.t. the 
-reported values. The minimal value of error (not exceeding 1\%) was obtained for $E_c$=3\,MPa. This 
-value lies in the range of capsule elasticity coefficients reported by experimental measurements.
-
-%Moreover, we are able to reproduce
-%the measurements presented in~\cite{Hollenstein2006,Nava2008} with less than 1\% error.
-%We do not model the time-dependent behaviour though.
-%In this simulation we prescribe the the internal pressure of 30\,kPa. The other
-%parameters of the model were obtained like this: First we fix the stiffness of
-%parenchyma to $E_p$=30\,kPa to match the scenerio without capsule. Then,
-%assuming the thickness $t$=93\,$\mu$m of the capsule (reported as average in
-%the paper) we run several simulations with different capsule stiffness $E_c$.
-%We have found the lowest error is when $E_c$=3\,MPa. All the stiffness
-%parameters lie within the ranges measured in the paper.
-
-%In this case the simulation was performed with the pressure of 30\,kPa and with
-%the following elastic parameters: $E_p$=30\,kPa for parenchyma and $E_c$=3\,MPa
-%and $t$=93\,$\mu$m for capsule.
-%All the values lie within the measurements
-%specified in the paper. \TG{add figure like Fig. 2?}
-
-%The influence of the capsule on local level has been published in literature
-%before. However, to our knowledge it's influence on global scale deformation
-%has not been studied yet. We believe that the capsule has non-negligible impact
-%also on global level.
-
-In the scenario presented in this paper, we focus on local influence of the capsule.
-Nevertheless, it is possible to employ the actual model to demonstrate the global 
-impact of the capsule and to further validate the accuracy of such a complete model w.r.t. experimental measures. 
-Detailed description of such validation is however beyond the scope of this paper. 
-In spite of this we would like to emphasize that we have already integrated the
-capsule model with vascularized liver model described in~\cite{Peterlik2012}. 
-Adding the capsule did not affect the
-performance significantly: a visual refresh rate exceeding 25\,FPS was achieved on a
-PC with CPU Intel i7 running at 3.00 GHz.
-This suggests that the proposed technique is compatible with real-time simulation of whole
-organ while modeling the local properties accurately.
-
-%In our future work we need to validate the model in the
-%context of global and local deformations. This task is challenging because
-%comparing in vivo measurements with the simulation is difficult due to complex
-%boundary conditions. 
-%\TG{ex-vivo: no information abou vessels; phantoms: cannot model capsule}
-
-%}}}
-
-\section{Conclusions and Future Work} %{{{
-In the paper we presented a model of tissue with capsule based on two separated models for parenchyma (corotational tetrahedral elements)
-and membrane (CST triangular elements). Employing a simulation scenario known as aspiration test, 
-we compared the model with experimental data provided in the literature 
-and verified the results obtained in the simulation w.r.t.
-to a standard computationally expensive model based entirely on volume elements~\cite{Hollenstein2006}.
-We obtained values which were in good correspondence with results reported in the literature.
-
-Presently we are focusing on the validation of the model also in context of
-global deformations of the organ where we believe the capsule plays an
-essential role as well. This task is however complicated due to the limited number of experimental studies dealing with 
-both the parenchyma and the capsule.
-
-%Further, the model was used to demonstrate global influence of the capsule when the organ 
-%undergoes large deformations.
-
-%In the future work, we plan to employ the proposed model for other organs where the capsule play an important role
-%(for example kidneys). We will also focus on better description of the mechanical interactions and couplings among the organs and tissues in the abdominal 
-%cavity via realistic modeling of boundary conditions using mechanical constraints and contacts.
-
-%}}}
-
-%
-% ---- Bibliography ----
-%
-
-\bibliographystyle{acmsiggraph}
-\bibliography{bibdata}
-
-\end{document}
-% spell: setl spell spelllang=en spellfile=spelldict.en.utf-8.add
-% vim:set et sw=2 fdm=marker fdl=3 fdc=4 isk+=_,-:

### 2013-07-31 Doc/MMVR2014/method.tex
MMVR 2014 paper - 4 pages
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@603 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2013-07-29 Doc/Vrst2013/vrst2013.tex
@@
-\documentclass[sposored]{acmsiggraph}
+\section{Methods \& Materials}
+\label{sec:methodology}
+\vspace{-6pt}
 
-\usepackage{amsmath}
-\usepackage{subfig}
-
-\def\wrt{w.\,r.\,t.}
-\def\eg{e.\,g.}
-\def\ie{i.\,e.}
-\def\Dash{\nobreak\,---\penalty-500\,}
-
-\renewcommand{\Vec}[1]{\mathbf{#1}}
-\newcommand{\Mat}[1]{\mathbf{#1}}
-
-\title{Modeling Liver with Glisson's Capsule in Real-Time Medical Simulations}
-
-%TODO
-\author{
-\begin{tabular}[t]{cccc}%
-  Tom\'a\v{s} Golembiovsk\'y\thanks{nyoxi@ics.muni.cz} &
-  Igor Peterl\'ik &
-  Christian Duriez &
-  St\'ephane Cotin \\
-%
-  Faculty of Informatics &
-  Institut Hospitalo-Universitaire &
-  INRIA &
-  INRIA \\
-%
-  Masaryk University &
-  Strasbourg, France &
-  Lille, France &
-  Lille, France \\
-%
-  Brno, Czech Republic
-\end{tabular}
-}
-%\affiliation{foobar}
-%\numberofauthors{4}
-%\author{%
-%Tom\'a\v{s} Golembiovsk\'y\\
-%       \affaddr{Faculty of Informatics}\\
-%       \affaddr{Masaryk University}\\
-%       \affaddr{Brno, Czech Republic}\\
-%       \email{nyoxi@ics.muni.cz}
-%% 2nd. author
-%\alignauthor
-%Igor Peterl\'ik\\
-%       \affaddr{Institut Hospitalo-Universitaire}\\
-%       \affaddr{Strasbourg, France}\\
-%       \email{Igor.Peterlik@ihu-strasbourg.eu}
-%% 3rd. author
-%\alignauthor
-%Christian Duriez\\
-%       \affaddr{INRIA}\\
-%       \affaddr{Lille, France}\\
-%       \email{Christian.Duriez@inria.fr}
-%\and
-%% 4th. author
-%\alignauthor
-%St\'ephane Cotin\\
-%       \affaddr{INRIA}\\
-%       \affaddr{Lille, France}\\
-%       \email{Stephane.Cotin@inria.fr}
-%}
-\pdfauthor{Tom\'a\v{s} Golembiovsk\'y}
-
-\begin{document}
-
-\maketitle
-\begin{abstract}
-The accurate mechanical modeling of liver is of a paramount interest in simulation-based operation planning 
-and computer-aided per-operative guidance. However, the realistic simulation of liver behaviour is a challenging task, 
-since the organ is composed of three constituents: parenchyma, vascularization and Glisson's capsule, 
-each having different mechanical properties.
-
-While a real-time simulation of vascularized liver has been already described in~\cite{Peterlik2012}, the Glisson's 
-capsule has not been included in the modeling. In~\cite{Hollenstein2006,Ahn2010} it is shown that the capsule 
-plays an important role mainly in the vicinity of the surgical tools. Therefore, the accuracy of the liver response 
-in the vicinity of the surgical tool requires correct modeling of the capsule as a component of the liver model.
-
-The measures performed on the capsule report stiffness which is significantly higher than that of the parenchyma, however, 
-the thickness of the membrane does not exceed 100$\mu$m. For this reason, it is not possible to model the 
-capsule with standard volume elements usually employed for the parenchyma. In our approach, we rely of 
-corotational membrane elements based on constant-strain formulation, which are coupled mechanically to the underlying 
-tetrahedra. We show that using this model we are able to reproduce the experiments reported in~\cite{Hollenstein2006}, 
-while maintaining the real-time aspect of the simulation.
-
-%This work provides a complete physical model for the deformations of the liver with its three constituents: parenchyma, vessels and the Glisson's capsule.
-%Some recent experimental results \cite{Ahn2010} have shown that there is a significant difference of stiffness between the Glisson's capsule and the parenchyma. 
-%Moreover, it has been observed that the capsule plays an important role for the deformation at the local level \cite{Hollenstein2006}.
-%However, the Glisson's capsule is very thin and its influence for the global deformations of the liver was never highlighted.
-%In this work, we propose to use a membrane FEM model for the capsule, that is coupled with a vascularized model of the liver.
-%We show that the approach is able to reproduce the deformations observed at the local level.
-%Additionally, we show, by simulating natural global deformations of the liver observed on real images (due to gravity), that the results have important differences with and without the capsule.
-
-\end{abstract}
-
-% Categories
-\begin{CRcatlist}
-  \CRcat{I.6.5}{Simulation and Modeling}{Model Development}
-  \CRcat{J.3}{Life and Medical Sciences}{Health};
-\end{CRcatlist}
-
-\keywords{deformable liver model, Glisson's capsule}
-
-\keywordlist
-
-\TOGlinkslist
-
-\copyrightspace
-
-\section{Introduction} %{{{
-
-According to the statistics, nearly 100,000 European citizens die of cirrhosis or liver cancer each year. 
-Although new methods such as radio-frequency- and cryo-ablation known in the interventional radiology 
-seem to be promising, surgery remains the option that offers the foremost success rate against these pathologies. 
-Nevertheless, surgery is not always performed due to several limitations, in particular the determination 
-of accurate eligibility criteria for the patient. 
-In this context, the pre-operational planning becomes a crucial task having a significant impact on the treatment.
-
-Computer-aided physics-based medical simulation has proven to be an extremely useful technique in the area of medical training. 
-Nevertheless, whereas generic models are usually required in training simulators, accurate patient-specific modeling
-becomes necessary as soon as computer simulation is to be employed in the pre-operative planning. At the same time, 
-interactivity of such models remains an important aspect, requiring real-time simulation which is often difficult to 
-achieve given the complexity of soft tissues.
-
-When simulating the behavior of human liver, the task of real-time accurate modeling is challenging, mainly because of the complex structure 
-of this organ composed of three main constituents: \emph{parenchyma}, \emph{vascular networks} and \emph{Glisson's capsule}.
-The parenchyma has certainly been the most studied component of the liver; actually researchers agree on hyperelastic 
-properties of the tissue, for which the mechanical parameters have been reported for example in~\cite{Kerdok2006,Gao2009}. 
-Moreover, several methods have been proposed to model the hyperelastic behaviour at real-time rates, such as multiplicative Jacobian decomposition
-introduced in~\cite{Marchesseau2010}.
-
-%\TG{The following paragraph is optional. Unnecessary but adds to the global context}
-The mechanical importance of the vascular structures in liver is studied in~\cite{Peterlik2012}. It shows that the 
-influence of vessels on the mechanical behaviour of the organ is significant. %, mainly if large deformations occur. 
-%Since a detailed modeling of the vessels would be extremely costly (mainly because of the small thickness of the vessel wall, 
-%the authors propose a composite model allowing for real-time simulation of entire liver with vascularization.
-Since a detailed modeling of the vessels would be extremely costly (mainly because of the small thickness of the vessel wall), 
-the authors propose a composite model allowing for real-time simulation.
-
-Rather a small number of studies have been conducted dealing with the third liver constituent, the  Glisson's capsule.
-Quantitative results of experiments on a porcine liver have been published in~\cite{Umale2011}; the measurements indicate that although being very 
-thin (10--20$\mu m$), the capsule shows to be stiff in tensile tests: the Young's modulus of the capsule reported to be $8.22\pm3.42$\,MPa 
-exceeds the values for the parenchyma by three orders of magnitude.
-This suggests that the mechanical influence of the membrane on the liver behaviour is not negligible.
-
-In~\cite{Hollenstein2006}, a local influence of the capsule has been measured using a special aspiration device. The study was then repeated 
-in vivo on human patients during the operation, confirming the mechanical importance of the membrane~\cite{Ahn2010,Nava2008}.
-%To our best knowledge, no attempt has been made to demonstrate the role of the Glisson's capsule on the global behaviour of 
-%the liver, mainly if the organ undergoes large deformations which is often the case during the surgery.
-
-In this paper we present preliminary results of our work on the complete liver
-model. We present a real-time composite model accounting for parenchyma and
-Glisson's capsule. The model is based on two different finite element
-representations for each constituent coupled together.
-%The main contribution of the paper is twofold: first, we present a real-time model of the liver, including
-%the parenchyma, vascularization and Glisson's capsule. The model is based on three different finite element representations for each constituent,
-%linked together via mechanical coupling.
-We show that the model mimics the \emph{local} experiments described in~\cite{Hollenstein2006}.
-
-%Second, we use the complete model of the liver to demonstrate the \emph{global} influence of the Glisson's
-%capsule via simulation: using a specific model of a porcine liver built from CT contrast-enhanced data, we show that there is a significant 
-%difference in the response of the model with and without the capsule in the case when the liver undergoes large deformations.
-
-The paper is organized as follows: first we describe the proposed model of
-liver with capsule. Second, we validate our model in context of local
-deformations by reproducing the aspiration test described
-in~\cite{Hollenstein2006}. We conclude by summarizing future steps towards the
-complete liver model.
-%Finally, we demonstrate that in spite of its small thickness, the Glisson's 
-%capsule has a global influence on the liver undergoing large deformations.
-
-%}}}
-
-\section{Model} %{{{
-
-In this section we describe the construction of a composite model
-based on two components: tetrahedral FE model of the 
-parenchyma and
-triangular membrane elements used for the capsule.
+In this section we describe the construction of a composite model based on two components: tetrahedral FE model of the  parenchyma and triangular membrane elements used for the capsule.
 
@@
 phenomena related to viscosity.
-
+%
 %However, we employ simpler corotational model as we are not so much
@@
 %Peterl\'{i}k et al. \cite{Peterlik2012}.
-
+%
 The parenchyma is modeled using corotational finite elements~\cite{Felippa2005}.
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[height=3.5cm]{aspiration.jpg}
-  \caption{\label{fig-aspiration1} The SOFA simulation scene of the aspiration test.}
-\end{figure}
-
 A good example of such a local deformation is the aspiration test
@@
 solve the system and compute the inverse matrices.
-
-\section{Results and Discussion} %{{{
-As the first step, we evaluate the influence of the Glisson's capsule on the mechanical response of the tissue during
-the aspiration test. 
-In~\cite{Hollenstein2006} it is reported that the influence of the capsule is significant, as modeling 
-only the parenchyma without considering the membrane results in overestimation of the deformation by a factor of 2 to 3. 
-In order to verify this observation, we performed a simulation using capsule parameters measured experimentally on 
-ex-vivo pig liver~\cite{Umale2011}: Young's modulus of the membrane was set to $E_c$=8.22\,MPa and thickness was
-$t_c$=20\,$\mu$m. 
-Since the values for parenchyma elasticity reported in the literature vary significantly being usually reported between 2\,kPa to 5\,kPa, 
-we employed a value $E_p$=3.5\,kPa for the Young's modulus of the parenchyma.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=7cm]{aspiration.pdf}
-  \caption{\label{fig-aspiration2} Displacement profile of the cube in the
-  aspiration test with (blue) and without (black) the capsule.}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[height=3.5cm]{displacement.pdf}
-  \caption{\label{fig-aspiration3} Evolution of the displacement at the center
-  in time for test with (blue) and without (black) the capsule.}
-\end{figure}
-
-In Fig.~\ref{fig-aspiration2} the profiles of cuts in the middle of the
-test cube are presented showing significantly larger deformation in the model without capsule assuming that the same negative pressure was applied 
-on the tissue surface inside the tube. Moreover, it can be observed that the deformation without capsule is overestimated by a factor 
-close to 2, which is in perfect agreement with the results reported in~\cite{Hollenstein2006}.
-Fig.~\ref{fig-aspiration3} presents the evolution of displacement in time at central node on the cube surface. The effect of increased combined stiffness is clearly visible.
-
-In the second step, we employed the simulation to reproduce the displacements values obtained via simulation in~\cite{Hollenstein2006,Nava2008}.
-Whereas the simulation in the referenced work was done on a mesh composed of tetrahedral elements for both the parenchyma 
-and the capsule, we modeled the capsule using the triangular elements as shown in section~\ref{ss:capsuleModel}.
-
-Since the Young's moduli of neither parenchyma ($E_p$) nor capsule ($E_c$) were specified exactly, we obtained both as follows: first, 
-we performed a set of simulations without the capsule for different values of $E_p$: the value that provided a good match 
-with reported displacements was $E_p$=30\,kPa. 
-Then, we fixed the thickness of the capsule to be $t$=93\,$\mu$m (reported as the average value in the paper) and 
-using $E_p$=30\,kPa, we ran the simulation repeatedly using the model with the capsule. In each simulation we 
-used different value $E_c$ of the Young's modulus for the capsule in order to minimize the displacement error w.r.t. the 
-reported values. The minimal value of error (not exceeding 1\%) was obtained for $E_c$=3\,MPa. This 
-value lies in the range of capsule elasticity coefficients reported by experimental measurements.
-
-%Moreover, we are able to reproduce
-%the measurements presented in~\cite{Hollenstein2006,Nava2008} with less than 1\% error.
-%We do not model the time-dependent behaviour though.
-%In this simulation we prescribe the the internal pressure of 30\,kPa. The other
-%parameters of the model were obtained like this: First we fix the stiffness of
-%parenchyma to $E_p$=30\,kPa to match the scenerio without capsule. Then,
-%assuming the thickness $t$=93\,$\mu$m of the capsule (reported as average in
-%the paper) we run several simulations with different capsule stiffness $E_c$.
-%We have found the lowest error is when $E_c$=3\,MPa. All the stiffness
-%parameters lie within the ranges measured in the paper.
-
-%In this case the simulation was performed with the pressure of 30\,kPa and with
-%the following elastic parameters: $E_p$=30\,kPa for parenchyma and $E_c$=3\,MPa
-%and $t$=93\,$\mu$m for capsule.
-%All the values lie within the measurements
-%specified in the paper. \TG{add figure like Fig. 2?}
-
-%The influence of the capsule on local level has been published in literature
-%before. However, to our knowledge it's influence on global scale deformation
-%has not been studied yet. We believe that the capsule has non-negligible impact
-%also on global level.
-
-In the scenario presented in this paper, we focus on local influence of the capsule.
-Nevertheless, it is possible to employ the actual model to demonstrate the global 
-impact of the capsule and to further validate the accuracy of such a complete model w.r.t. experimental measures. 
-Detailed description of such validation is however beyond the scope of this paper. 
-In spite of this we would like to emphasize that we have already integrated the
-capsule model with vascularized liver model described in~\cite{Peterlik2012}. 
-Adding the capsule did not affect the
-performance significantly: a visual refresh rate exceeding 25\,FPS was achieved on a
-PC with CPU Intel i7 running at 3.00 GHz.
-This suggests that the proposed technique is compatible with real-time simulation of whole
-organ while modeling the local properties accurately.
-
-%In our future work we need to validate the model in the
-%context of global and local deformations. This task is challenging because
-%comparing in vivo measurements with the simulation is difficult due to complex
-%boundary conditions. 
-%\TG{ex-vivo: no information abou vessels; phantoms: cannot model capsule}
-
-%}}}
-
-\section{Conclusions and Future Work} %{{{
-In the paper we presented a model of tissue with capsule based on two separated models for parenchyma (corotational tetrahedral elements)
-and membrane (CST triangular elements). Employing a simulation scenario known as aspiration test, 
-we compared the model with experimental data provided in the literature 
-and verified the results obtained in the simulation w.r.t.
-to a standard computationally expensive model based entirely on volume elements~\cite{Hollenstein2006}.
-We obtained values which were in good correspondence with results reported in the literature.
-
-Presently we are focusing on the validation of the model also in context of
-global deformations of the organ where we believe the capsule plays an
-essential role as well. This task is however complicated due to the limited number of experimental studies dealing with 
-both the parenchyma and the capsule.
-
-%Further, the model was used to demonstrate global influence of the capsule when the organ 
-%undergoes large deformations.
-
-%In the future work, we plan to employ the proposed model for other organs where the capsule play an important role
-%(for example kidneys). We will also focus on better description of the mechanical interactions and couplings among the organs and tissues in the abdominal 
-%cavity via realistic modeling of boundary conditions using mechanical constraints and contacts.
-
-%}}}
-
-%
-% ---- Bibliography ----
-%
-
-\bibliographystyle{acmsiggraph}
-\bibliography{bibdata}
-
-\end{document}
-% spell: setl spell spelllang=en spellfile=spelldict.en.utf-8.add
-% vim:set et sw=2 fdm=marker fdl=3 fdc=4 isk+=_,-:

### 2013-07-31 Doc/MMVR2014/displacement.R

@@ -1,0 +1,21 @@
+mt <- read.table("disp-tet.txt")
+mc <- read.table("disp-comp.txt")
+
+mt[,3] <- (mt[,3] - 0.015)*1000
+mc[,3] <- (mc[,3] - 0.015)*1000
+
+
+pdf("displacement.pdf", width=3, height=1.5, pointsize=8); # sizes in inches
+p <- par(lwd=0.5, mar=c(4,4,0,0)+0.1);
+
+i <- 1:200 # 5 seconds
+plot(c(0,2),
+     c(-0.285, 2.336), # make the range consistent
+     #range(mt[,3], mc[,3]),
+     type="n", xlab="Time [s]", ylab="Displacement [mm]")
+
+lines(mt[i,1], mt[i,3], col=1)
+lines(mc[i,1], mc[i,3], col=4)
+
+dev.off()
+par(p)

### 2013-08-01 Doc/MMVR2014/intro.tex
UPDATE: adding the global influence, reducing the method section
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@606 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 According to the statistics, nearly 100,000 European citizens die of cirrhosis or liver cancer each year. 
-Although new methods such as radio-frequency- and cryo-ablation known in the interventional radiology 
+Although new methods 
+%such as radio-frequency- and cryo-ablation known 
+in the interventional radiology 
 seem to be promising, surgery remains the option that offers the foremost success rate against these pathologies. 
@@
 of accurate eligibility criteria for the patient. 
-In this context, the pre-operational planning becomes a crucial task having a significant impact on the treatment.
+%In this context, the pre-operational planning becomes a crucial task having a significant impact on the treatment.
 
@@
 Nevertheless, whereas generic models are usually required in training simulators, accurate patient-specific modeling
-becomes necessary as soon as computer simulation is to be employed in the pre-operative planning. At the same time, 
-interactivity of such models remains an important aspect, requiring real-time simulation which is often difficult to 
-achieve given the complexity of soft tissues. 
+becomes necessary as soon as computer simulation is to be employed in the pre-operative planning. 
+%At the same time, 
+%interactivity of such models remains an important aspect, requiring real-time simulation which is often difficult to 
+%achieve given the complexity of soft tissues. 
 %
@@
 exceeds the values for the parenchyma by three orders of magnitude.
-This suggests that the mechanical influence of the membrane on the liver behaviour is not negligible.
+This suggests that the mechanical influence of the membrane on the liver behavior is not negligible.
 %
@@
 In this paper we present preliminary results of our work on the complete liver model. 
-We present a real-time composite model accounting for parenchyma and Glisson's capsule. 
-The model is based on two different finite element representations for each constituent coupled together.
+We present a real-time composite model accounting for parenchyma and Glisson's capsule, compatible with previously proposed vascularized model~\cite{Peterlik2012}.
+%The model is based on two different finite element representations for each constituent coupled together.
 %The main contribution of the paper is twofold: first, we present a real-time model of the liver, including
@@
 %linked together via mechanical coupling.
-We show that the model mimics the \emph{local} experiments described in~\cite{Hollenstein2006}.
+%We show that the model mimics the \emph{local} experiments described in~\cite{Hollenstein2006}.
 

### 2013-08-01 Doc/MMVR2014/method.tex
UPDATE: adding the global influence, reducing the method section
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@606 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 
-In this section we describe the construction of a composite model based on two components: tetrahedral FE model of the  parenchyma and triangular membrane elements used for the capsule.
+% In this section we describe the construction of a composite model based on two components: 
+% tetrahedral FE model of the  parenchyma and triangular membrane elements used for the capsule.
 
-\subsection{Parenchyma} %{{{
+% \subsection{Biomechanical Models and Coupling}
+It is known that the parenchyma exhibits non-linear viscoelastic behavior \cite{Marchesseau2010}.
+However, as we are mainly interested in the static equilibrium, we do not model the time-dependent
+phenomena related to viscosity. In the simulation we employ corotational finite elements~\cite{Felippa2005}.
+Although it relies on linear stress-strain relationship, large displacements including rotations are modeled correctly.
+% Briefly, the motion of each element $e$ is decomposed into rigid rotation $\Mat{R}^e$ and local deformation in each step of the simulation. 
+% The rotations are then used to update each local element stiffness matrix as $\Mat{R}^e\Mat{K}_0{\Mat{R}^e}^{\top}$
+% whereas the deformations are used to compute the linear strain in the local corotational frame.
+% There are several ways of computing the corotational frame for elements; we rely on
+% the geometrical method proposed in \cite{Nesme2005}.
 
-It is known that the parenchyma exhibits non-linear viscoelastic behaviour \cite{Marchesseau2010}.
-However, as we are mainly interested in the static equilibrium, we do not model the time-dependent
-phenomena related to viscosity.
+% \subsection{Parenchyma} %{{{
+
+% It is known that the parenchyma exhibits non-linear viscoelastic behaviour \cite{Marchesseau2010}.
+% However, as we are mainly interested in the static equilibrium, we do not model the time-dependent
+% phenomena related to viscosity.
 %
-%However, we employ simpler corotational model as we are not so much
-%interested in time-dependent behaviour but rather in static equilibrium
-%under certain conditions.
-%We also rely on the vascularized model of the parenchyma proposed by
-%Peterl\'{i}k et al. \cite{Peterlik2012}.
+% % %However, we employ simpler corotational model as we are not so much
+% % %interested in time-dependent behaviour but rather in static equilibrium
+% % %under certain conditions.
+% % %We also rely on the vascularized model of the parenchyma proposed by
+% % %Peterl\'{i}k et al. \cite{Peterlik2012}.
 %
-The parenchyma is modeled using corotational finite elements~\cite{Felippa2005}.
-Although it relies on linear stress-strain relationship, large displacements including rotations are modeled correctly. 
-While in the full non-linear formulation the stiffness matrix relates the forces $\Vec{f}$ and 
-displacements $\Vec{u}$ as $\Vec{f} = K(\Vec{u})$, the corotational model 
-requires the stiffness matrix $\Mat{K}_0$ of the system to be computed only once before the simulation begins. 
-Then, in each step, the motion of each element $e$ is decomposed into rigid rotation $\Mat{R}^e$ and local deformation. 
-The rotations are then used to update each local element stiffness matrix as $\Mat{R}^e\Mat{K}_0{\Mat{R}^e}^{\top}$
-whereas the deformations are used to compute the linear strain in the local corotational frame.
-There are several ways of computing the corotational frame for elements; we rely on
-the geometrical method proposed in \cite{Nesme2005}.
-% NOTE: This description of corotational method is very simplified and could be extended.
+% The parenchyma is modeled using corotational finite elements~\cite{Felippa2005}.
+% Although it relies on linear stress-strain relationship, large displacements including rotations are modeled correctly. 
+% While in the full non-linear formulation the stiffness matrix relates the forces $\Vec{f}$ and 
+% displacements $\Vec{u}$ as $\Vec{f} = K(\Vec{u})$, the corotational model 
+% requires the stiffness matrix $\Mat{K}_0$ of the system to be computed only once before the simulation begins. 
+% Then, in each step, the motion of each element $e$ is decomposed into rigid rotation $\Mat{R}^e$ and local deformation. 
+% The rotations are then used to update each local element stiffness matrix as $\Mat{R}^e\Mat{K}_0{\Mat{R}^e}^{\top}$
+% whereas the deformations are used to compute the linear strain in the local corotational frame.
+% There are several ways of computing the corotational frame for elements; we rely on
+% the geometrical method proposed in \cite{Nesme2005}.
+% % % NOTE: This description of corotational method is very simplified and could be extended.
+% % 
+% % %The model of the vascularization is based on linear beams 
+% % %with local frames of reference~\cite{Duriez2006}; in many aspects it's similar to the
+% % %corotational formulation described above. As such the model also handles geometric
+% % %non-linearities in the deformation. Through a specification of cross section and moments of inertia, 
+% % %the model can account for the specific properties of the blood vessels. 
+% % 
+% % %The beam-based model of vessels is mechanically coupled to the parenchyma as described in~\cite{Peterlik2012}. 
+% % %The coupling assumes that there is no relative motion between the vessels and the surrounding parenchyma. 
+% % %During the simulation, the nodes of linked beams are first displaced and rotated according to the actual motion of associated tetrahedra. 
+% % %As the deformation of beams results in mechanical response represented by forces and torques, these are propagated back to 
+% % %the tetrahedral FE model. 
+% % 
+% % %}}}
 
-%The model of the vascularization is based on linear beams 
-%with local frames of reference~\cite{Duriez2006}; in many aspects it's similar to the
-%corotational formulation described above. As such the model also handles geometric
-%non-linearities in the deformation. Through a specification of cross section and moments of inertia, 
-%the model can account for the specific properties of the blood vessels.
-
-%The beam-based model of vessels is mechanically coupled to the parenchyma as described in~\cite{Peterlik2012}. 
-%The coupling assumes that there is no relative motion between the vessels and the surrounding parenchyma. 
-%During the simulation, the nodes of linked beams are first displaced and rotated according to the actual motion of associated tetrahedra. 
-%As the deformation of beams results in mechanical response represented by forces and torques, these are propagated back to 
-%the tetrahedral FE model.
-
-%}}}
-
-\subsection{Glisson's Capsule} %{{{
-\label{ss:capsuleModel}
+% \subsection{Glisson's Capsule} %{{{
+% \label{ss:capsuleModel}
 The thickness of the Glisson's capsule is relatively small: the values in range of 10-20
@@
 It is not possible to model such thin structure with classical tetrahedral
-elements, if the real-time aspect of the simulation is to be preserved.
-Furthermore, modeling both the tissue and the capsule would require an extremely 
+elements:
+%if the real-time aspect of the simulation is to be preserved.
+%Furthermore, modeling both the tissue and the capsule 
+the model would require an extremely 
 dense mesh to avoid numerical instabilities and thus would significantly
@@
 
-\subsection{Coupling Between Capsule and\\Parenchyma} %{{{
+%\subsection{Mechanical Coupling between Parenchyma and Capsule} %{{{
 The literature reports high cohesion between capsule and parenchyma.
@@
 of the triangular mesh coincide with the nodes of the tetrahedral mesh, so no projection of one mesh onto the other is needed.
-Moreover, the stiffness matrices for capsule and parenchyma are then easily assembled together.
+and the stiffness matrices for capsule and parenchyma are then easily assembled together by adding the mechanical contribution 
+of a triangle to the corresponding tetrahedron.
 % and solved as one system.
@@
 
-\section{Methods} %{{{
-The model presented in the previous section was implemented in 
-SOFA\footnote{www.sofa-framework.org} and a set of
-numerical simulations was performed.
-In this section we provide comparison of local deformations with the
-results reported in literature to validate the method.
+% \subsection{Simulation} %{{{
+% The model has been implemented SOFA\footnote{www.sofa-framework.org} and a set of
+% numerical simulations was performed.
+%In this section we provide comparison of local deformations with the
+%results reported in literature to validate the method.
 %Second, to show that in spite of its very small thickness the membrane cannot be
@@
 
-During the contact with an instrument such as probe, needle, scalpel and others,
-specific deformations take place in the vicinity of
-the instrument. This type of deformation may not necessarily induce the
-deformation of the object as a whole and therefore can be considered as
-local. Correct material properties are not only important to quantify the
-displacement, but also play an important role in capturing the correct area of
-the deformation or its profile near the instrument.
+% During the contact with an instrument such as probe, needle, scalpel and others,
+% specific deformations take place in the vicinity of
+% the instrument. This type of deformation may not necessarily induce the
+% deformation of the object as a whole and therefore can be considered as
+% local. Correct material properties are not only important to quantify the
+% displacement, but also play an important role in capturing the correct area of
+% the deformation or its profile near the instrument.
 
-A good example of such a local deformation is the aspiration test
-where the response of liver exposed locally to a negative pressure is measured.
-The aspiration device consists of a tube having 1\,cm in diameter and allows to
-control the pressure inside the tube. The test is performed by
-attaching the tube to the tissue and measuring the tissue response. We
-set up a simulation in SOFA to reproduce the experiment (see
-Fig.~\ref{fig-aspiration1}): we meshed a 15$\times$15$\times$15\,mm$^3$ 
-cube representing the tissue resulting in 2648 tetrahedra. Then we attached a 1\,cm tube 
-and applied a pressure of 3\,kPa inside the tube.
-
-Since the tube is in direct contact with the tissue, uni-lateral constraints with friction were chosen 
-to model this interaction properly. We opted for a method based on \emph{non-linear complementarity problem}  (NLCP)
-where the non-linearity is introduced due to the friction. The NLCP
-allows for solution of the Signorini's problem to avoid any interpenetration between the colliding 
-objects (see~\cite{Duriez2006b} for details). Since NLCP requires explicitly the computation of compliance matrix which 
-is homogeneous to the inverse of the stiffness matrix, we employed a direct solver based on LDL decomposition to 
-solve the system and compute the inverse matrices.
+% Since the tube is in direct contact with the tissue, uni-lateral constraints with friction were chosen 
+% to model this interaction properly. We opted for a method based on \emph{non-linear complementarity problem}  (NLCP)
+% where the non-linearity is introduced due to the friction. The NLCP
+% allows for solution of the Signorini's problem to avoid any interpenetration between the colliding 
+% objects (see~\cite{Duriez2006b} for details). Since NLCP requires explicitly the computation of compliance matrix which 
+% is homogeneous to the inverse of the stiffness matrix, we employed a direct solver based on LDL decomposition to 
+% solve the system and compute the inverse matrices.

### 2013-08-01 Doc/MMVR2014/results.tex
UPDATE: adding the global influence, reducing the method section
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@606 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\section{Results} %{{{
\subsection{Aspiration Test}
% As the first step, we evaluate the influence of the Glisson's capsule on the mechanical response of the tissue during
% the aspiration test. 
In~\cite{Hollenstein2006} an \emph{aspiration test} is presented to quantify the local influence of the Glisson's on tissue behaviour.
The aspiration device consists of a tube having 1\,cm in diameter and allows to
control the pressure inside the tube. The test is performed by
attaching the tube to the tissue and measuring the tissue response. We
set up a simulation in SOFA to reproduce the experiment (see
Fig.~\ref{fig-aspiration1}): we meshed a 15$\times$15$\times$15\,mm$^3$ 
cube representing the tissue resulting in 2648 tetrahedra. Then we attached a 1\,cm tube 
and applied a pressure of 3\,kPa inside the tube. The contacts between the tube and tissue sample were modeled using 
an approach based on non-linear complementarity problem presented in~\cite{Duriez2006b} which provides a solution for 
the Signorini's problem while taking into account friction between the objects.
% objects

According to~\cite{Hollenstein2006} the influence of the capsule is significant: modeling 
only the parenchyma without considering the membrane results in overestimation of the deformation by a factor of 2 to 3. 
In order to verify this observation, we performed a simulation using capsule parameters measured experimentally on 
ex-vivo pig liver~\cite{Umale2011}: Young's modulus of the membrane was set to $E_c$=8.22\,MPa and thickness was
$t_c$=20\,$\mu$m. 
Since the values for parenchyma elasticity reported in the literature vary significantly being usually reported between 2\,kPa to 5\,kPa, 
we employed a value $E_p$=3.5\,kPa for the Young's modulus of the parenchyma.

In Fig.~\ref{fig-aspiration2} the profiles of cuts in the middle of the
test cube are presented showing significantly larger deformation in the model without capsule assuming that the same negative pressure was applied 
on the tissue surface inside the tube. Moreover, it can be observed that the deformation without capsule is overestimated by a factor 
close to 2, which is in perfect agreement with the results reported in~\cite{Hollenstein2006}.
Fig.~\ref{fig-aspiration3} presents the evolution of displacement in time at central node on the cube surface. The effect of increased combined stiffness is clearly visible.

Further, we employed the simulation to reproduce the displacements values obtained via simulation in~\cite{Hollenstein2006,Nava2008}.
Whereas the simulation in the referenced work was done on a mesh composed of tetrahedral elements for both the parenchyma 
and the capsule, we modeled the capsule using the triangular elements as shown in section~\ref{ss:capsuleModel}.

Since the Young's moduli of neither parenchyma ($E_p$) nor capsule ($E_c$) were specified exactly, we obtained both as follows: first, 
we performed a set of simulations without the capsule for different values of $E_p$: the value that provided a good match 
with reported displacements was $E_p$=30\,kPa. 
Then, we fixed the thickness of the capsule to be $t$=93\,$\mu$m (reported as the average value in the paper) and 
using $E_p$=30\,kPa, we ran the simulation repeatedly using the model with the capsule. In each simulation we 
used different value $E_c$ of the Young's modulus for the capsule in order to minimize the displacement error w.r.t. the 
reported values. The minimal value of error (not exceeding 1\%) was obtained for $E_c$=3\,MPa. This 
value lies in the range of capsule elasticity coefficients reported by experimental measurements.

% % % %Moreover, we are able to reproduce
% % % %the measurements presented in~\cite{Hollenstein2006,Nava2008} with less than 1\% error.
% % % %We do not model the time-dependent behaviour though.
% % % %In this simulation we prescribe the the internal pressure of 30\,kPa. The other
% % % %parameters of the model were obtained like this: First we fix the stiffness of
% % % %parenchyma to $E_p$=30\,kPa to match the scenerio without capsule. Then,
% % % %assuming the thickness $t$=93\,$\mu$m of the capsule (reported as average in
% % % %the paper) we run several simulations with different capsule stiffness $E_c$.
% % % %We have found the lowest error is when $E_c$=3\,MPa. All the stiffness
% % % %parameters lie within the ranges measured in the paper.
% % % 
% % % %In this case the simulation was performed with the pressure of 30\,kPa and with
% % % %the following elastic parameters: $E_p$=30\,kPa for parenchyma and $E_c$=3\,MPa
% % % %and $t$=93\,$\mu$m for capsule.
% % % %All the values lie within the measurements
% % % %specified in the paper. \TG{add figure like Fig. 2?}
% % % 
% % % %The influence of the capsule on local level has been published in literature
% % % %before. However, to our knowledge it's influence on global scale deformation
% % % %has not been studied yet. We believe that the capsule has non-negligible impact
% % % %also on global level.

% In the scenario presented in this paper, we focus on local influence of the capsule.
% Nevertheless, it is possible to employ the actual model to demonstrate the global 
% impact of the capsule and to further validate the accuracy of such a complete model w.r.t. experimental measures. 
% Detailed description of such validation is however beyond the scope of this paper. 
% In spite of this we would like to emphasize that we have already integrated the
% capsule model with vascularized liver model described in~\cite{Peterlik2012}. 
% Adding the capsule did not affect the
% performance significantly: a visual refresh rate exceeding 25\,FPS was achieved on a
% PC with CPU Intel i7 running at 3.00 GHz.
% This suggests that the proposed technique is compatible with real-time simulation of whole
% organ while modeling the local properties accurately.

%In our future work we need to validate the model in the
%context of global and local deformations. This task is challenging because
%comparing in vivo measurements with the simulation is difficult due to complex
%boundary conditions. 
%\TG{ex-vivo: no information abou vessels; phantoms: cannot model capsule}

%}}}

\subsection{Global Deformation of Complete Model}
The complete model was constructed using contrast-enhanced CT data acquired on a female pig
being positioned first in supine and then flank position.
The liver and hepatic vein were segmented for both datasets  using semi-automatic methods available in ITKSnap~\footnote{www.itksnap.org}.
A tetrahedral mesh was obtained from the segmented maps using CGAL~\footnote{www.cgal.org}.
The semi-automatic method presented in~\cite{Peterlik2012} was
then applied to the segmented map of the hepatic vein, resulting in linked-beam model.
Finally, the model of the Glisson's capsule was constructed using the surface elements extracted from the volume mesh.
%described in section~\ref{ss:capsuleModel} over the surface triangles extracted from the tetrahedral mesh.
The complete model composed of 4266 tetrahedra, 188 beams and 1284 triangles was 
used in simulation where external force representing the gravity was applied to the model,
resulting in large deformations mainly in the area of lobes. The refresh rate of 22\,FPS
was achieved on a desktop computer with Intel i7 CPU running at 3.4\,GHz and 8\,GB of memory.
The simulation with gravity was also used as a first demonstration of the global influence of the capsule.
In Fig.~\ref{f:global}a, the models with capsule (red) and without (yellow) are depicted after being
deformed due to the gravity, showing a significant difference between the two models.

### 2013-08-01 Doc/MMVR2014/method.tex
ADD: new section about the methods showing the combination of matrices
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@607 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 % \subsection{Biomechanical Models and Coupling}
-It is known that the parenchyma exhibits non-linear viscoelastic behavior \cite{Marchesseau2010}.
-However, as we are mainly interested in the static equilibrium, we do not model the time-dependent
-phenomena related to viscosity. In the simulation we employ corotational finite elements~\cite{Felippa2005}.
-Although it relies on linear stress-strain relationship, large displacements including rotations are modeled correctly.
+%It is known that the parenchyma exhibits non-linear viscoelastic behavior \cite{Marchesseau2010}.
+Since in this paper, we are mainly interested in the static equilibrium, we do not model the time-dependent
+phenomena related to viscosity. In the simulation we employ corotational finite elements;
+There are several ways of computing the corotational frame for elements; we rely on the geometrical method proposed in~\cite{Nesme2005}.
+%although it relies on linear stress-strain relationship, large displacements including rotations are modeled correctly.
+Denoting $p$ a generic tetrahedral element of the parenchyma, the 12$\times$12 element stiffness matrix is then computed as
+\begin{equation}
+\Mat{K}_p = \int_{V_p} \Mat{B}_p^\top \Mat{D}_p \Mat{B}_p dV
+\end{equation}
+where $\Mat{B}_p$ is the strain-displacement matrix and $\Mat{D}_p$ is the stress-strain matrix.
+
+While the thickness of a vessel wall does not exceed 250$\mu$m~\cite{Forauer2003}, stiffness of 0.62$\pm$0.41\,MPa has been reported 
+in~\cite{Umale2011}, being quite important when compared to the parenchyma. Due to this property, vessels influence 
+the mechanical response of the tissue as illustrated in~\cite{Peterlik2012}. Since the vessel wall cannot be modeled 
+with tetrahedral elements if the real-time aspect of the simulation is to be preserved, 
+serially-linked beam elements based on Tymoshenko formulation are employed, taking into account the hollow 
+structure of the vessels via moments of inertia. 
+%to model the vessels which allows for correct simulation of large deformations as it is based on the corotational framework 
+%similar to the one used for the parenchyma. 
+For an arbitrary beam element representing a segment of a vessel, the corresponding local stiffness matrix $\Mat{K}_v$ is computed as shown in~\cite{Duriez2006}.
+
 % Briefly, the motion of each element $e$ is decomposed into rigid rotation $\Mat{R}^e$ and local deformation in each step of the simulation. 
@@
 % \label{ss:capsuleModel}
-The thickness of the Glisson's capsule is relatively small: the values in range of 10-20
-$\mu$m have been reported in~\cite{Umale2011}.
+The thickness of the Glisson's capsule is also very small: the values in range of 10-20 $\mu$m have been reported in~\cite{Umale2011}.
 It is not possible to model such thin structure with classical tetrahedral
@@
 thickness in the third dimension seems
-to be a natural choice. In the elasticity theory, this functionality is usually provided by membrane and shell elements.
-Based on the observation of its behaviour, we also
-assume negligible bending forces and propose a model based on membrane
-elements. 
+to be a natural choice. In the elasticity theory, this functionality is usually provided by membrane elements.
+%Based on the observation of its behaviour, we also
+%assume negligible bending forces and propose a model based on membrane
+%elements. 
 To maintain simplicity of the composite model we choose simple triangular
@@
 
-The computation of elastic stiffness matrix follows the common derivation
-%
-\begin{eqnarray}
-  \Mat{K}^m & = & \int_V \Mat{B}^T \Mat{E} \Mat{B} dV     \label{mem1} \\
-            & = & h \int_A \Mat{B}^T \Mat{E} \Mat{B} dA   \label{mem2} \\
-            & = & h A \Mat{B}^T \Mat{E} \Mat{B}           \label{mem3}
-\end{eqnarray}
-%
-where $\Mat{B}$ is the strain-displacement matrix, $\Mat{E}$ the material
-matrix, $h$ is the thickness and $A$ area of the element. In the previous
-\eqref{mem2} follows from the fact that we assume constant thickness of the
-element and \eqref{mem3} follows from the fact that the strain-displacement
-matrix is constant in our case. The strain-displacement matrix for the CST
-element can be expressed as:
+The 9$\times$9 stiffness matrix $\Mat{K}_m$ is given as 
 %
 \begin{equation}
-  \Mat{B} = \frac{1}{2A} \begin{bmatrix}
-    y_{23} & 0      & y_{31} & 0      & y_{12} & 0 \\
-         0 & x_{32} & 0      & x_{13} & 0      & x_{21} \\
-    x_{32} & y_{23} & x_{13} & y_{31} & x_{21} & y_{12}
-  \end{bmatrix}
+ \Mat{K}_c = \int_{S_c} \Mat{B}_c^\top \Mat{D}_c \Mat{B}_c dS 
 \end{equation}
+where $\Mat{B}_c$ and $\Mat{D}_c$ are strain-displacement and strain-stress matrices, respectively.
+
+The complete model of the liver takes into account all three components described above by combining the contribution 
+of parenchyma, vessels and capsule. The coupling between the vessel elements (beams) and parenchyma elements (tetrahedra)
+is described in detail in~\cite{Peterlik2012}. Briefly, each vessel node (having 6 degrees of freedom to account for torques in vessels) 
+is coupled with a tetrahedra in which it is located via barycentric coordinates. This coupling remains constant during 
+the simulation and can be described via matrix $\Mat{J}_{v\rightarrow p}$ which is a 12$\times$6 Jacobian matrix of the coupling.
+
+The coupling between the parenchyma and capsule is straightforward, since the triangles used as the 
+domain for the CST formulations are the surface faces of the volume mesh. Therefore, for given triangle 
+with vertices $v_1$, $v_2$ and $v_3$, the corresponding tetrahedron (sharing the same vertices is found) and 
+a 9$\times$12 permutation matrix $\Mat{P}_{c\rightarrow p}$ mapping the triangle vertices to tetrahedra vertices is computed.
+
+Without loss of generality, let us suppose a tetrahedral element $e$ 
+which receives a mechanical contribution from both the capsule and vessels. The composite element stiffness matrix $\Mat{K}_e$
+is then computed as:
+\begin{equation}
+\Mat{K}_e = \Mat{K}_p + \Mat{J}_{v\rightarrow p} \Mat{K}_v \Mat{J}^\top_{v\rightarrow p} + \Mat{P}_{c\rightarrow p} \Mat{K}_c \Mat{P}_{c\rightarrow p}^\top.
+\end{equation}
+
 % 
-The values $x_{ij} = x_i - x_j$ and $y_{ij} = y_i - y_j$ are computed from
-the $x$ or $y$ coordinates of the nodes $i,j$ of the triangular element.
-The reader can refer to the respective literature~\cite{Felippa2003} for more thorough
-description.
-
-Similarly as with model of parenchyma we use linear elastic material and employ
-the corotational formulation for the CST elements.
+% \begin{eqnarray}
+%   \Mat{K}^m & = & \int_V \Mat{B}^T \Mat{E} \Mat{B} dV     \label{mem1} \\
+%             & = & h \int_A \Mat{B}^T \Mat{E} \Mat{B} dA   \label{mem2} \\
+%             & = & h A \Mat{B}^T \Mat{E} \Mat{B}           \label{mem3}
+% \end{eqnarray}
+% %
+% where $\Mat{B}$ is the strain-displacement matrix, $\Mat{E}$ the material
+% matrix, $h$ is the thickness and $A$ area of the element. In the previous
+% \eqref{mem2} follows from the fact that we assume constant thickness of the
+% element and \eqref{mem3} follows from the fact that the strain-displacement
+% matrix is constant in our case. The strain-displacement matrix for the CST
+% element can be expressed as:
+% %
+% \begin{equation}
+%   \Mat{B} = \frac{1}{2A} \begin{bmatrix}
+%     y_{23} & 0      & y_{31} & 0      & y_{12} & 0 \\
+%          0 & x_{32} & 0      & x_{13} & 0      & x_{21} \\
+%     x_{32} & y_{23} & x_{13} & y_{31} & x_{21} & y_{12}
+%   \end{bmatrix}
+% \end{equation}
+% %
+% The values $x_{ij} = x_i - x_j$ and $y_{ij} = y_i - y_j$ are computed from
+% the $x$ or $y$ coordinates of the nodes $i,j$ of the triangular element.
+% The reader can refer to the respective literature~\cite{Felippa2003} for more thorough
+% description.
+% 
+% Similarly as with model of parenchyma we use linear elastic material and employ
+% the corotational formulation for the CST elements.
 
@@
 %\subsection{Mechanical Coupling between Parenchyma and Capsule} %{{{
-The literature reports high cohesion between capsule and parenchyma.
-Based on this property we assume there is no relative motion of the capsule \wrt\ the parenchyma.
-Although an arbitrary surface mesh could be used to model the capsule, we exploit 
-the fact that the parenchyma is modeled by tetrahedral elements having
-triangular faces. Thus, as the boundary of the volumetric mesh is already
-triangulated, we simply employ the triangles on the mesh surface to model the capsule.
-
-Using directly the boundary of the tetrahedral mesh does not only solve the
-problem of building the surface mesh, but has one more advantage: the nodes
-of the triangular mesh coincide with the nodes of the tetrahedral mesh, so no projection of one mesh onto the other is needed.
-and the stiffness matrices for capsule and parenchyma are then easily assembled together by adding the mechanical contribution 
-of a triangle to the corresponding tetrahedron.
-% and solved as one system.
+% The literature reports high cohesion between capsule and parenchyma.
+% Based on this property we assume there is no relative motion of the capsule \wrt\ the parenchyma.
+% Although an arbitrary surface mesh could be used to model the capsule, we exploit 
+% the fact that the parenchyma is modeled by tetrahedral elements having
+% triangular faces. Thus, as the boundary of the volumetric mesh is already
+% triangulated, we simply employ the triangles on the mesh surface to model the capsule.
 % 
-%\CD{The following may appear as "trivial"... maybe we can remove this part at the end if we need space} 
-Without the loss of generality we can assume the tetrahedron consists of
-nodes $p_1, p_2, p_3$ and $p_4$ and the boundary triangle has nodes $p_1, p_2$
-and $p_3$. We can reorder the degrees of freedom so that the stiffness
-matrix $\Mat{K}^t$ for the tetrahedron can be written as:
+% Using directly the boundary of the tetrahedral mesh does not only solve the
+% problem of building the surface mesh, but has one more advantage: the nodes
+% of the triangular mesh coincide with the nodes of the tetrahedral mesh, so no projection of one mesh onto the other is needed.
+% and the stiffness matrices for capsule and parenchyma are then easily assembled together by adding the mechanical contribution 
+% of a triangle to the corresponding tetrahedron.
+% % and solved as one system.
+% %
+% %\CD{The following may appear as "trivial"... maybe we can remove this part at the end if we need space} 
+% Without the loss of generality we can assume the tetrahedron consists of
+% nodes $p_1, p_2, p_3$ and $p_4$ and the boundary triangle has nodes $p_1, p_2$
+% and $p_3$. We can reorder the degrees of freedom so that the stiffness
+% matrix $\Mat{K}^t$ for the tetrahedron can be written as:
+% %
+% \begin{equation}
+%   \Mat{K}^t = \left[\begin{array}{c|c}
+%       \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
+%       \hline
+%       \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\.
+%   \end{array}\right]
+% \end{equation}
+% %
+% Hence, the assembled stiffness matrix for the element is
+% %
+% \begin{equation}
+%   \Mat{K} = \left[\begin{array}{c|c}
+%       \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
+%       \hline
+%       \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
+%   \end{array}\right]
+%   +
+%   \left[\begin{array}{c|c}
+%       \Mat{K}^m & 0 \\
+%       \hline
+%       0 & 0 \\
+%   \end{array}\right]
+% \end{equation}
+% %
+% where $\Mat{K}^m$ is the stiffness matrix of the triangular membrane.
 % 
-\begin{equation}
-  \Mat{K}^t = \left[\begin{array}{c|c}
-      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
-      \hline
-      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\.
-  \end{array}\right]
-\end{equation}
-%
-Hence, the assembled stiffness matrix for the element is
-%
-\begin{equation}
-  \Mat{K} = \left[\begin{array}{c|c}
-      \Mat{K}^t_{1-3,1-3} & \Mat{K}^t_{1-3,4} \\
-      \hline
-      \Mat{K}^t_{4,1-3} & \Mat{K}^t_{4,4} \\
-  \end{array}\right]
-  +
-  \left[\begin{array}{c|c}
-      \Mat{K}^m & 0 \\
-      \hline
-      0 & 0 \\
-  \end{array}\right]
-\end{equation}
-%
-where $\Mat{K}^m$ is the stiffness matrix of the triangular membrane.
-
-The resulting system of linear equations is solved by direct solver based on Cholesky decomposition.
+% The resulting system of linear equations is solved by direct solver based on Cholesky decomposition.
 

### 2013-08-02 Doc/MMVR2014/abstract.tex
FIX: finalization of the paper
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@608 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 %
-In this paper we propose a complete liver model, where the parenchyma is modeled as a finite element volume, the vessels are modeled as series of beam elements, and the   corotational membrane elements based on constant-strain formulation, which are coupled mechanically to the underlying tetrahedra.
+In this paper we propose a complete liver model, where the parenchyma is modeled as a finite element volume, the vessels are modeled as series of beam elements, 
+and the corotational membrane elements based on constant-strain formulation are used for capsule. These components are coupled mechanically as shown in the paper.
 We show that using this model we are able to reproduce accurately biomechanical tests, while maintaining the real-time aspect of the simulation.

### 2013-08-02 Doc/MMVR2014/intro.tex
FIX: finalization of the paper
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@608 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 Computer-aided physics-based medical simulation has proven to be an extremely useful technique in the area of medical training. 
-Nevertheless, whereas generic models are usually required in training simulators, accurate patient-specific modeling
+Whereas generic models are usually required in training simulators, accurate patient-specific modeling
 becomes necessary as soon as computer simulation is to be employed in the pre-operative planning. 
@@
 %achieve given the complexity of soft tissues. 
-%
 When simulating the behavior of human liver, the task of real-time accurate modeling is challenging, mainly because of the complex structure 
@@
 The parenchyma has certainly been the most studied component of the liver; actually researchers agree on hyperelastic 
-properties of the tissue, for which the mechanical parameters have been reported for example in~\cite{Kerdok2006,Gao2009}. 
+properties of the tissue, for which the mechanical parameters have been reported for example in~\cite{Kerdok2006}.%,Gao2009}. 
 Moreover, several methods have been proposed to model the hyperelastic behaviour at real-time rates, such as multiplicative Jacobian decomposition
@@
 We present a real-time composite model accounting for parenchyma and Glisson's capsule, compatible with previously proposed vascularized model~\cite{Peterlik2012}.
+We also show that the capsule has a significant impact on the mechanical response of the organ. 
 %The model is based on two different finite element representations for each constituent coupled together.

### 2013-08-02 Doc/MMVR2014/results.tex
FIX: finalization of the paper
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@608 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

@@
 close to 2, which is in perfect agreement with the results reported in~\cite{Hollenstein2006}.
-Fig.~\ref{fig-aspiration3} presents the evolution of displacement in time at central node on the cube surface. The effect of increased combined stiffness is clearly visible.
+Fig.~\ref{fig-aspiration3} presents the evolution of displacement in time at central node on the cube surface showing clear the effect of increased combined stiffness.
 
@@
 Whereas the simulation in the referenced work was done on a mesh composed of tetrahedral elements for both the parenchyma 
-and the capsule, we modeled the capsule using the triangular elements as shown in section~\ref{ss:capsuleModel}.
+and the capsule, we employ the membrane CST elements.
 
@@
 \subsection{Global Deformation of Complete Model}
-The complete model was constructed using contrast-enhanced CT data acquired on a female pig
-being positioned first in supine and then flank position.
-The liver and hepatic vein were segmented for both datasets  using semi-automatic methods available in ITKSnap~\footnote{www.itksnap.org}.
+The complete model was constructed using contrast-enhanced CT data acquired on a female pig.
+The liver and hepatic vein were segmented using semi-automatic methods available in ITKSnap~\footnote{www.itksnap.org}.
 A tetrahedral mesh was obtained from the segmented maps using CGAL~\footnote{www.cgal.org}.
@@
 The simulation with gravity was also used as a first demonstration of the global influence of the capsule.
-In Fig.~\ref{f:global}a, the models with capsule (red) and without (yellow) are depicted after being
+In Fig.~\ref{f:gravity}a, the models with capsule (red) and without (yellow) are depicted after being
 deformed due to the gravity, showing a significant difference between the two models.

### 2013-10-28 Doc/MMVR2014/IOSProceedings/conclusion.tex
ADD: proceedings version of the paper
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@610 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\section{Conclusion}
In this paper we presented a composite liver model based on volume and membrane elements. 
When compared to experimental data provided in the literature we found similar results, but with faster computation times than standard computationally expensive model based entirely on volume elements.

%Presently we are focusing on the validation of the model also in context of global deformations of the organ where we believe the capsule plays an essential role as well. 
%This task is however complicated due to the limited number of experimental studies dealing with both the parenchyma and the capsule.

%\section*{Acknowledgments} 
%The authors would like to thank the French Research Agency (ANR) for funding this study through the ACouStiC project.

### 2013-10-28 Doc/MMVR2014/IOSProceedings/results.tex
ADD: proceedings version of the paper
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@610 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

--- 2013-08-02 Doc/MMVR2014/results.tex
@@
 \section{Results} %{{{
+\begin{figure}[t]
+  \centering
+  \includegraphics[height=3.5cm]{figures/aspirationScreenshot.jpg}
+  \caption{\label{fig-aspiration1} The SOFA simulation scene of the aspiration test.}
+\end{figure}
 \subsection{Aspiration Test}
@@
 we employed a value $E_p$=3.5\,kPa for the Young's modulus of the parenchyma.
+
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=7cm]{figures/displacementProfile.pdf}
+  \caption{\label{fig-aspiration2} Displacement profile of the cube in the
+  aspiration test with (blue) and without (black) the capsule.}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[height=3.5cm]{figures/displacementEvolution.pdf}
+  \caption{\label{fig-aspiration3} Evolution of the displacement at the center
+  in time for test with (blue) and without (black) the capsule.}
+\end{figure}
 
@@
 
+\begin{figure}[b]
+  \centering
+  \includegraphics[height=7cm]{figures/modelUnderGravity.png}
+  \caption{Liver under gravity with (red) and without (yellow) capsule}
+  \label{f:gravity}
+\end{figure}
+
 \subsection{Global Deformation of Complete Model}

### 2021-01-13 initPluginShells.cpp
Set up cpp files for plugin

@@ -1,127 +1,0 @@
-/******************************************************************************
-*       SOFA, Simulation Open-Framework Architecture, version 1.0 beta 4      *
-*                (c) 2006-2009 MGH, INRIA, USTL, UJF, CNRS                    *
-*                                                                             *
-* This program is free software; you can redistribute it and/or modify it     *
-* under the terms of the GNU General Public License as published by the Free  *
-* Software Foundation; either version 2 of the License, or (at your option)   *
-* any later version.                                                          *
-*                                                                             *
-* This program is distributed in the hope that it will be useful, but WITHOUT *
-* ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or       *
-* FITNESS FOR A PARTICULAR PURPOSE. See the GNU General Public License for    *
-* more details.                                                               *
-*                                                                             *
-* You should have received a copy of the GNU General Public License along     *
-* with this program; if not, write to the Free Software Foundation, Inc., 51  *
-* Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.                   *
-*******************************************************************************
-*                            SOFA :: Applications                             *
-*                                                                             *
-* Authors: M. Adam, J. Allard, B. Andre, P-J. Bensoussan, S. Cotin, C. Duriez,*
-* H. Delingette, F. Falipou, F. Faure, S. Fonteneau, L. Heigeas, C. Mendoza,  *
-* M. Nesme, P. Neumann, J-P. de la Plata Alcade, F. Poyer and F. Roy          *
-*                                                                             *
-* Contact information: contact@sofa-framework.org                             *
-******************************************************************************/
-#include <sofa/helper/system/config.h>
-#include "initPluginShells.h"
-#include <cstring>
-#include <string>
-
-namespace sofa
-{
-
[…]

### 2021-01-13 src/SofaShells/initPluginShells.cpp

@@ -1,0 +1,75 @@
+/******************************************************************************
+*       SOFA, Simulation Open-Framework Architecture, version 1.0 beta 4      *
+*                (c) 2006-2009 MGH, INRIA, USTL, UJF, CNRS                    *
+*                                                                             *
+* This program is free software; you can redistribute it and/or modify it     *
+* under the terms of the GNU General Public License as published by the Free  *
+* Software Foundation; either version 2 of the License, or (at your option)   *
+* any later version.                                                          *
+*                                                                             *
+* This program is distributed in the hope that it will be useful, but WITHOUT *
+* ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or       *
+* FITNESS FOR A PARTICULAR PURPOSE. See the GNU General Public License for    *
+* more details.                                                               *
+*                                                                             *
+* You should have received a copy of the GNU General Public License along     *
+* with this program; if not, write to the Free Software Foundation, Inc., 51  *
+* Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.                   *
+*******************************************************************************
+*                            SOFA :: Applications                             *
+*                                                                             *
+* Authors: M. Adam, J. Allard, B. Andre, P-J. Bensoussan, S. Cotin, C. Duriez,*
+* H. Delingette, F. Falipou, F. Faure, S. Fonteneau, L. Heigeas, C. Mendoza,  *
+* M. Nesme, P. Neumann, J-P. de la Plata Alcade, F. Poyer and F. Roy          *
+*                                                                             *
+* Contact information: contact@sofa-framework.org                             *
+******************************************************************************/
+#include <SofaShells/config.h>
+#include <sofa/core/ObjectFactory.h>
+using sofa::core::ObjectFactory;
+
+
+extern "C" {
+    SOFA_SHELLS_API void initExternalModule();
+    SOFA_SHELLS_API const char* getModuleName();
+    SOFA_SHELLS_API const char* getModuleVersion();
+    SOFA_SHELLS_API const char* getModuleLicense();
+    SOFA_SHELLS_API const char* getModuleDescription();
+    SOFA_SHELLS_API const char* getModuleComponentList();
+}
+
+void initExternalModule()
+{
+    static bool first = true;
+    if (first)
+    {
+        first = false;
+    }
+}
+
+const char* getModuleName()
+{
[…]

### 2021-01-13 src/SofaShells/controller/AdaptiveCuttingController.cpp
Fix compilation, path and most warnings

@@ -1,5 +1,5 @@
-#include "../initPluginShells.h"
-#include "AdaptiveCuttingController.inl"
+#include <SofaShells/config.h>
+#include <SofaShells/cutting/AdaptiveCuttingController.inl>
 #include <sofa/core/ObjectFactory.h>
 
 

### 2021-01-13 src/SofaShells/controller/AdaptiveCuttingController.h

@@ -15,12 +15,10 @@
 #include <SofaBaseTopology/TriangleSetGeometryAlgorithms.h>
 #include <SofaBaseTopology/TopologyData.h>
 
-//#include <sofa/gui/PickHandler.h>
-
 #include <sofa/helper/map.h>
 #include <sofa/helper/vector.h>
 
-#include "Test2DAdapter.h"
+#include <SofaShells/controller/Test2DAdapter.h>
 
 namespace sofa
 {

### 2021-01-13 src/SofaShells/controller/CudaTest2DAdapter.cpp

@@ -1,7 +1,7 @@
 
-#include "../initPluginShells.h"
+#include <SofaShells/config.h>
 #include <sofa/gpu/cuda/CudaTypes.h>
-#include "CudaTest2DAdapter.inl"
+#include <SofaShells/controller/CudaTest2DAdapter.inl>
 #include <sofa/core/ObjectFactory.h>
 
 

### 2021-01-13 src/SofaShells/controller/CudaTest2DAdapter.h

@@ -2,7 +2,7 @@
 #define CUDA_TEST2DADAPTER_H
 
 #include <sofa/gpu/cuda/CudaTypes.h>
-#include "Test2DAdapter.h"
+#include <SofaShells/controller/Test2DAdapter.h>
 
 #include <sofa/helper/fixed_array.h>
 #include <sofa/helper/vector.h>

### 2021-01-13 src/SofaShells/controller/MeshChangedEvent.cpp

@@ -1,5 +1,6 @@
 
-#include "MeshChangedEvent.h"
+#include <SofaShells/controller/MeshChangedEvent.h>
+
 namespace sofa
 {
 

### 2021-01-13 src/SofaShells/controller/MeshChangedEvent.h

@@ -2,9 +2,9 @@
 #define SOFA_CORE_OBJECTMODEL_MESHCHANGEDEVENT_H
 
 #include <sofa/core/objectmodel/Event.h>
-#include <sofa/defaulttype/Vec3Types.h>
+#include <sofa/defaulttype/VecTypes.h>
 #include <sofa/defaulttype/Quat.h>
-#include <sofa/helper/system/config.h>
+#include <SofaShells/config.h>
 
 namespace sofa
 {

### 2021-01-13 src/SofaShells/controller/MeshInterpolator.cpp

@@ -1,5 +1,5 @@
 #define SOFA_COMPONENT_CONTROLLER_MESHINTERPOLATION_CPP
-#include "MeshInterpolator.inl"
+#include <SofaShells/controller/MeshInterpolator.inl>
 #include <sofa/core/ObjectFactory.h>
 
 namespace sofa

### 2021-01-13 src/SofaShells/controller/MeshInterpolator.h

@@ -1,13 +1,12 @@
 #ifndef SOFA_COMPONENT_CONTROLLER_MESHINTERPOLATION_H
 #define SOFA_COMPONENT_CONTROLLER_MESHINTERPOLATION_H
 
-//#include <sofa/component/component.h>
 #include <SofaUserInteraction/Controller.h>
 #include <sofa/defaulttype/RigidTypes.h>
 #include <sofa/defaulttype/VecTypes.h>
 #include <sofa/defaulttype/Vec.h>
 
-#include "../initPluginShells.h"
+#include <SofaShells/config.h>
 
 namespace sofa
 {
@@ -39,10 +38,10 @@ protected:
 
 public:
 
-    virtual void init();
-    virtual void reinit();
+    void init() override;
+    void reinit() override;
 
-    virtual std::string getTemplateName() const
+    std::string getTemplateName() const override
     {
         return templateName(this);
     }
@@ -64,7 +63,7 @@ public:
     Data<VecCoord>          f_position;
     Data<helper::vector<Vec3> > f_normals;
 
-    virtual void onEndAnimationStep(const double dt);
+    void onEndAnimationStep(const double dt) override;
 
     Real getInterpolationVar() { return alpha; }
 

### 2021-01-13 src/SofaShells/controller/Test2DAdapter.cpp

@@ -1,5 +1,5 @@
-#include "../initPluginShells.h"
-#include "Test2DAdapter.inl"
+#include <SofaShells/config.h>
+#include <SofaShells/controller/Test2DAdapter.inl>
 #include <sofa/core/ObjectFactory.h>
 
 

### 2021-01-13 src/SofaShells/controller/Test2DAdapter.h

@@ -18,8 +18,8 @@
 #include <sofa/helper/map.h>
 #include <sofa/helper/vector.h>
 
-#include "../misc/Optimize2DSurface.h"
-#include "../misc/SurfaceParametrization.h"
+#include <SofaShells/misc/Optimize2DSurface.h>
+#include <SofaShells/misc/SurfaceParametrization.h>
 
 namespace sofa
 {

### 2021-01-13 src/SofaShells/controller/TriangleSwitchExample.cpp

@@ -1,7 +1,7 @@
 #define SOFA_COMPONENT_CONTROLLER_TRIANGLESWITCHEXAMPLE_CPP
 
-#include "../initPluginShells.h"
-#include "TriangleSwitchExample.inl"
+#include <SofaShells/config.h>
+#include <SofaShells/controller/TriangleSwitchExample.inl>
 #include <sofa/core/ObjectFactory.h>
 
 

### 2021-01-13 src/SofaShells/controller/TriangleSwitchExample.h

@@ -1,7 +1,6 @@
 #ifndef SOFA_COMPONENT_CONTROLLER_TRIANGLESWITCHEXAMPLE_H
 #define SOFA_COMPONENT_CONTROLLER_TRIANGLESWITCHEXAMPLE_H
 
-//#include <sofa/component/component.h>
 #include <SofaUserInteraction/Controller.h>
 #include <sofa/defaulttype/RigidTypes.h>
 #include <sofa/defaulttype/VecTypes.h>
@@ -46,10 +45,10 @@ protected:
 
 public:
 
-    virtual void init();
-    virtual void reinit();
+    void init() override;
+    void reinit() override;
 
-    virtual std::string getTemplateName() const
+    std::string getTemplateName() const override
     {
         return templateName(this);
     }
@@ -61,7 +60,7 @@ public:
 
     Data<unsigned int>      f_interval;
 
-    virtual void onEndAnimationStep(const double dt);
+    void onEndAnimationStep(const double dt) override;
 
     void computeTriangleNormal(const Triangle &t, Vec3 &normal);
 

### 2021-01-13 src/SofaShells/cutting/AdaptiveCutting.cpp

@@ -1,10 +1,10 @@
-#include "../initPluginShells.h"
+#include <SofaShells/config.h>
 
 #include <sofa/core/ObjectFactory.h>
-#include <sofa/defaulttype/Vec3Types.h>
+#include <sofa/defaulttype/VecTypes.h>
 #include <sofa/helper/Factory.inl>
 #include <sofa/gui/PickHandler.h>
-#include "AdaptiveCutting.h"
+#include <SofaShells/cutting/AdaptiveCutting.h>
 
 namespace sofa
 {

### 2021-01-13 src/SofaShells/cutting/AdaptiveCutting.h

@@ -7,7 +7,7 @@
 #include <sofa/gui/MouseOperations.h>
 #include <sofa/gui/OperationFactory.h>
 
-#include "../controller/AdaptiveCuttingController.h"
+#include <SofaShells/controller/AdaptiveCuttingController.h>
 
 namespace sofa
 {

### 2021-01-13 src/SofaShells/engine/FindClosePoints.cpp

@@ -1,7 +1,7 @@
 #define SOFA_COMPONENT_ENGINE_FINDCLOSEPOINTS_CPP
-#include "FindClosePoints.inl"
+#include <SofaShells/engine/FindClosePoints.inl>
 #include <sofa/core/ObjectFactory.h>
-#include "../initPluginShells.h"
+#include <SofaShells/config.h>
 
 namespace sofa
 {

### 2021-01-13 src/SofaShells/engine/FindClosePoints.h

@@ -4,10 +4,8 @@
 #include <sofa/defaulttype/Vec.h>
 #include <sofa/core/DataEngine.h>
 #include <sofa/core/objectmodel/BaseObject.h>
-//#include <sofa/core/loader/MeshLoader.h>
 #include <sofa/defaulttype/VecTypes.h>
 #include <sofa/defaulttype/RigidTypes.h>
-//#include <sofa/component/component.h>
 
 namespace sofa
 {
@@ -42,7 +40,7 @@ protected:
 
 public:
 
-    virtual std::string getTemplateName() const
+    virtual std::string getTemplateName() const override
     {
         return templateName(this);
     }
@@ -52,9 +50,9 @@ public:
         return DataTypes::Name();
     }
 
-    void init();
-    void reinit();
-    void doUpdate();
+    void init() override;
+    void reinit() override;
+    void doUpdate() override;
 
     // Data
     Data<Real>      f_input_threshold;

### 2021-01-13 src/SofaShells/engine/JoinMeshPoints.cpp

@@ -1,5 +1,5 @@
 #define SOFA_COMPONENT_ENGINE_JOINMESHPOINTS_CPP
-#include "JoinMeshPoints.inl"
+#include <SofaShells/engine/JoinMeshPoints.inl>
 #include <sofa/core/ObjectFactory.h>
 
 namespace sofa

### 2021-01-13 src/SofaShells/engine/JoinMeshPoints.h

@@ -7,8 +7,7 @@
 #include <sofa/defaulttype/Vec.h>
 #include <sofa/defaulttype/VecTypes.h>
 #include <sofa/defaulttype/RigidTypes.h>
-//#include <sofa/component/component.h>
-#include "../initPluginShells.h"
+#include <SofaShells/config.h>
 
 namespace sofa
 {
@@ -45,7 +44,7 @@ protected:
 
 public:
 
-    virtual std::string getTemplateName() const
+    std::string getTemplateName() const override
     {
         return templateName(this);
     }
@@ -55,9 +54,9 @@ public:
         return DataTypes::Name();
     }
 
-    void init();
-    void reinit();
-    void doUpdate();
+    void init() override;
+    void reinit() override;
+    void doUpdate() override;
 
     // TODO: add methods to find out the inverse mappings
 

### 2021-01-13 src/SofaShells/forcefield/BezierTriangularBendingFEMForceField.cpp

@@ -23,12 +23,12 @@
 * Contact information: contact@sofa-framework.org                             *
 ******************************************************************************/
 #define SOFA_COMPONENT_FORCEFIELD_BEZIER_TRIANGULAR_BENDING_FEM_FORCEFIELD_CPP
-#include "BezierTriangularBendingFEMForceField.inl"
+#include <SofaShells/forcefield/BezierTriangularBendingFEMForceField.inl>
 #include <sofa/core/behavior/ForceField.inl>
 #include <sofa/defaulttype/RigidTypes.h>
 #include <SofaBaseMechanics/MechanicalObject.h>
 #include <sofa/core/ObjectFactory.h>
-#include "../initPluginShells.h"
+#include <SofaShells/config.h>
 
 
 namespace sofa

### 2021-01-13 src/SofaShells/forcefield/BezierTriangularBendingFEMForceField.h

@@ -34,13 +34,11 @@
 #include <sofa/core/behavior/MechanicalState.h>
 #include <sofa/core/objectmodel/Data.h>
 
-//#include <sofa/component/component.h>
-
 #include <sofa/core/topology/BaseMeshTopology.h>
 #include <SofaBaseTopology/TopologyData.h>
 
-#include "../controller/MeshInterpolator.h"
-#include "../engine/JoinMeshPoints.h"
+#include <SofaShells/controller/MeshInterpolator.h>
+#include <SofaShells/engine/JoinMeshPoints.h>
 
 
 // Uncomment the following to use quaternions instead of matrices for
@@ -98,7 +96,7 @@ class BezierTriangularBendingFEMForceField : public core::behavior::ForceField<D
 
         typedef Vec3Types::VecCoord VecCoordHigh;
 
-        typedef sofa::core::topology::BaseMeshTopology::index_type Index;
+        typedef sofa::Index Index;
         typedef sofa::core::topology::BaseMeshTopology::Triangle Triangle;
         typedef sofa::core::topology::BaseMeshTopology::SeqTriangles SeqTriangles;
 
@@ -221,18 +219,17 @@ public:
         BezierTriangularBendingFEMForceField();
 
         virtual ~BezierTriangularBendingFEMForceField();
-        virtual void init();
-        virtual void reinit();
-        virtual void addForce(const sofa::core::MechanicalParams* /*mparams*/, DataVecDeriv& dataF, const DataVecCoord& dataX, const DataVecDeriv& /*dataV*/ ) ;
-        virtual void addDForce(const sofa::core::MechanicalParams* /*mparams*/, DataVecDeriv& datadF, const DataVecDeriv& datadX ) ;
-        virtual void addKToMatrix(const core::MechanicalParams* mparams, const sofa::core::behavior::MultiMatrixAccessor* matrix);
-        virtual void addBToMatrix(sofa::defaulttype::BaseMatrix * /*mat*/, double /*bFact*/, unsigned int &/*offset*/);
-        //virtual double getPotentialEnergy(const VecCoord& x) const;
-        virtual void handleTopologyChange();
+        void init() override;
+        void reinit() override;
+        void addForce(const sofa::core::MechanicalParams* /*mparams*/, DataVecDeriv& dataF, const DataVecCoord& dataX, const DataVecDeriv& /*dataV*/ ) override ;
[…]

### 2021-01-13 src/SofaShells/forcefield/CstFEMForceField.cpp

@@ -23,11 +23,11 @@
 * Contact information: contact@sofa-framework.org                             *
 ******************************************************************************/
 
-#include "../initPluginShells.h"
-#include "CstFEMForceField.inl"
+#include <SofaShells/config.h>
+#include <SofaShells/forcefield/CstFEMForceField.inl>
 #include <sofa/core/behavior/ForceField.inl>
 #include <sofa/defaulttype/Vec.h>
-#include <SofaBaseMechanics/MechanicalObject.h>
+//#include <SofaBaseMechanics/MechanicalObject.h>
 #include <sofa/core/ObjectFactory.h>
 
 

### 2021-01-13 src/SofaShells/forcefield/CstFEMForceField.h

@@ -81,7 +81,7 @@ class CstFEMForceField : public core::behavior::ForceField<DataTypes>
 
         typedef Vec3Types::VecCoord VecCoordHigh;
 
-        typedef sofa::core::topology::BaseMeshTopology::index_type Index;
+        typedef sofa::Index Index;
         typedef sofa::core::topology::BaseMeshTopology::Triangle Triangle;
         typedef sofa::core::topology::BaseMeshTopology::SeqTriangles SeqTriangles;
 
@@ -156,13 +156,13 @@ public:
         CstFEMForceField();
 
         virtual ~CstFEMForceField();
-        virtual void init();
-        virtual void reinit();
-        virtual void addForce(const sofa::core::MechanicalParams* /*mparams*/, DataVecDeriv& dataF, const DataVecCoord& dataX, const DataVecDeriv& /*dataV*/ ) ;
-        virtual void addDForce(const sofa::core::MechanicalParams* /*mparams*/, DataVecDeriv& datadF, const DataVecDeriv& datadX ) ;
-        virtual void addKToMatrix(const core::MechanicalParams* mparams, const sofa::core::behavior::MultiMatrixAccessor* matrix);
+        void init() override;
+        void reinit() override;
+        void addForce(const sofa::core::MechanicalParams* /*mparams*/, DataVecDeriv& dataF, const DataVecCoord& dataX, const DataVecDeriv& /*dataV*/ ) override ;
+        void addDForce(const sofa::core::MechanicalParams* /*mparams*/, DataVecDeriv& datadF, const DataVecDeriv& datadX ) override ;
+        void addKToMatrix(const core::MechanicalParams* mparams, const sofa::core::behavior::MultiMatrixAccessor* matrix) override;
 
-        virtual SReal getPotentialEnergy(const sofa::core::MechanicalParams* /*mparams*/, const DataVecCoord& x) const { return 0; }
+        SReal getPotentialEnergy(const sofa::core::MechanicalParams* /*mparams*/, const DataVecCoord& x) const override { return 0; }
 
         sofa::core::topology::BaseMeshTopology* getTopology() {return _topology;}
 

### 2021-01-13 src/SofaShells/forcefield/TriangularBendingFEMForceField.cpp

@@ -23,12 +23,12 @@
 * Contact information: contact@sofa-framework.org                             *
 ******************************************************************************/
 #define SOFA_COMPONENT_FORCEFIELD_TRIANGULAR_BENDING_FEM_FORCEFIELD_CPP
-#include "TriangularBendingFEMForceField.inl"
+#include <SofaShells/forcefield/TriangularBendingFEMForceField.inl>
 #include <sofa/core/behavior/ForceField.inl>
 #include <sofa/defaulttype/RigidTypes.h>
 #include <SofaBaseMechanics/MechanicalObject.h>
 #include <sofa/core/ObjectFactory.h>
-#include "../initPluginShells.h"
+#include <SofaShells/config.h>
 
 
 namespace sofa

### 2021-01-13 src/SofaShells/forcefield/TriangularBendingFEMForceField.h

@@ -34,13 +34,11 @@
 #include <sofa/core/behavior/MechanicalState.h>
 #include <sofa/core/objectmodel/Data.h>
 
-//#include <sofa/component/component.h>
-
 #include <sofa/core/topology/BaseMeshTopology.h>
 #include <SofaBaseTopology/TopologyData.h>
 
-#include "../controller/MeshInterpolator.h"
-#include "../engine/JoinMeshPoints.h"
+#include <SofaShells/controller/MeshInterpolator.h>
+#include <SofaShells/engine/JoinMeshPoints.h>
 
 
 namespace sofa
@@ -88,7 +86,7 @@ class TriangularBendingFEMForceField : public core::behavior::ForceField<DataTyp
 
         typedef Vec3Types::VecCoord VecCoordHigh;
 
-        typedef sofa::core::topology::BaseMeshTopology::index_type Index;
+        typedef sofa::Index Index;
         typedef sofa::core::topology::BaseMeshTopology::Triangle Triangle;
         typedef sofa::core::topology::BaseMeshTopology::SeqTriangles SeqTriangles;
 
@@ -182,18 +180,18 @@ public:
         TriangularBendingFEMForceField();
 
         virtual ~TriangularBendingFEMForceField();
-        virtual void init();
-        virtual void reinit();
-        virtual void addForce(const sofa::core::MechanicalParams* /*mparams*/, DataVecDeriv& dataF, const DataVecCoord& dataX, const DataVecDeriv& /*dataV*/ ) ;
-        virtual void addDForce(const sofa::core::MechanicalParams* /*mparams*/, DataVecDeriv& datadF, const DataVecDeriv& datadX ) ;
-        virtual void addKToMatrix(const core::MechanicalParams* mparams, const sofa::core::behavior::MultiMatrixAccessor* matrix);
-        virtual void addBToMatrix(sofa::defaulttype::BaseMatrix * /*mat*/, double /*bFact*/, unsigned int &/*offset*/);
-        virtual double getPotentialEnergy(const VecCoord& x) const;
-        virtual void handleTopologyChange();
+        void init() override;
+        void reinit() override;
+        void addForce(const sofa::core::MechanicalParams* /*mparams*/, DataVecDeriv& dataF, const DataVecCoord& dataX, const DataVecDeriv& /*dataV*/ ) override ;
+        void addDForce(const sofa::core::MechanicalParams* /*mparams*/, DataVecDeriv& datadF, const DataVecDeriv& datadX ) override ;
[…]

### 2021-01-13 src/SofaShells/forcefield/TriangularShellForceField.cpp

@@ -23,8 +23,8 @@
 * Contact information: contact@sofa-framework.org                             *
 ******************************************************************************/
 #define SOFA_COMPONENT_FORCEFIELD_TRIANGULAR_BENDING_FEM_FORCEFIELD_CPP
-#include "../initPluginShells.h"
-#include "TriangularShellForceField.inl"
+#include <SofaShells/config.h>
+#include <SofaShells/forcefield/TriangularShellForceField.inl>
 #include <sofa/core/behavior/ForceField.inl>
 #include <sofa/defaulttype/RigidTypes.h>
 #include <SofaBaseMechanics/MechanicalObject.h>

### 2021-01-13 src/SofaShells/forcefield/TriangularShellForceField.h

@@ -35,8 +35,6 @@
 #include <sofa/core/objectmodel/Data.h>
 #include <sofa/helper/OptionsGroup.h>
 
-//#include <sofa/component/component.h>
-
 #include <sofa/core/topology/BaseMeshTopology.h>
 #include <SofaBaseTopology/TopologyData.h>
 
@@ -95,7 +93,7 @@ class TriangularShellForceField : public core::behavior::ForceField<DataTypes>
 
         typedef Vec3Types::VecCoord VecCoordHigh;
 
-        typedef sofa::core::topology::BaseMeshTopology::index_type Index;
+        typedef sofa::Index Index;
         typedef sofa::core::topology::BaseMeshTopology::Triangle Triangle;
         typedef sofa::core::topology::BaseMeshTopology::SeqTriangles SeqTriangles;
 
@@ -197,14 +195,13 @@ public:
         TriangularShellForceField();
 
         virtual ~TriangularShellForceField();
-        virtual void init();
-        virtual void reinit();
-        virtual void addForce(const sofa::core::MechanicalParams* /*mparams*/, DataVecDeriv& dataF, const DataVecCoord& dataX, const DataVecDeriv& /*dataV*/ ) ;
-        virtual void addDForce(const sofa::core::MechanicalParams* /*mparams*/, DataVecDeriv& datadF, const DataVecDeriv& datadX ) ;
-        virtual void addKToMatrix(const core::MechanicalParams* mparams, const sofa::core::behavior::MultiMatrixAccessor* matrix);
-        ////virtual void handleTopologyChange();
+        void init() override;
+        void reinit() override;
+        void addForce(const sofa::core::MechanicalParams* /*mparams*/, DataVecDeriv& dataF, const DataVecCoord& dataX, const DataVecDeriv& /*dataV*/ ) override ;
+        void addDForce(const sofa::core::MechanicalParams* /*mparams*/, DataVecDeriv& datadF, const DataVecDeriv& datadX ) override ;
+        void addKToMatrix(const core::MechanicalParams* mparams, const sofa::core::behavior::MultiMatrixAccessor* matrix) override;
 
-        virtual SReal getPotentialEnergy(const sofa::core::MechanicalParams* /*mparams*/, const DataVecCoord& x) const { return 0; }
+        SReal getPotentialEnergy(const sofa::core::MechanicalParams* /*mparams*/, const DataVecCoord& /*x*/) const override { return 0; }
 
         sofa::core::topology::BaseMeshTopology* getTopology() {return _topology;}
 

### 2021-01-13 src/SofaShells/mapping/BendingPlateMechanicalMapping.cpp

@@ -23,11 +23,11 @@
 * Contact information: contact@sofa-framework.org                             *
 ******************************************************************************/
 
-#include "BendingPlateMechanicalMapping.inl"
+#include <SofaShells/mapping/BendingPlateMechanicalMapping.inl>
 #include <sofa/defaulttype/VecTypes.h>
 #include <sofa/defaulttype/RigidTypes.h>
 #include <sofa/core/ObjectFactory.h>
-#include "../initPluginShells.h"
+#include <SofaShells/config.h>
 
 
 namespace sofa

### 2021-01-13 src/SofaShells/mapping/BendingPlateMechanicalMapping.h

@@ -33,16 +33,12 @@
 #include <sofa/core/behavior/MechanicalState.h>
 #include <sofa/helper/vector.h>
 
-
-#include <sofa/helper/gl/GLSLShader.h>
-
+#include <sofa/gl/GLSLShader.h>
 #include <SofaBaseTopology/TriangleSetTopologyContainer.h>
 
-#include "../forcefield/TriangularBendingFEMForceField.h"
-//#include <sofa/component/topology/TriangleSubdivisionTopologicalMapping.h>
+#include <SofaShells/forcefield/TriangularBendingFEMForceField.h>
 
 #include <sofa/defaulttype/VecTypes.h>
-
 #include <sofa/helper/system/thread/CTime.h>
 
 
@@ -110,15 +106,15 @@ public:
     {
     }
 
-    void init();
-    void reinit();
-    virtual void draw(const core::visual::VisualParams* vparams);
+    void init() override;
+    void reinit() override;
+    void draw(const core::visual::VisualParams* vparams) override;
 
 
-    void apply(const core::MechanicalParams *mparams, Data<OutVecCoord>& out, const Data<InVecCoord>& in);
-    void applyJ(const core::MechanicalParams *mparams, Data<OutVecDeriv>& out, const Data<InVecDeriv>& in);
-    void applyJT(const core::MechanicalParams *mparams, Data<InVecDeriv>& out, const Data<OutVecDeriv>& in);
-    void applyJT(const core::ConstraintParams *cparams, Data<InMatrixDeriv>& out, const Data<OutMatrixDeriv>& in);
+    void apply(const core::MechanicalParams *mparams, Data<OutVecCoord>& out, const Data<InVecCoord>& in) override;
+    void applyJ(const core::MechanicalParams *mparams, Data<OutVecDeriv>& out, const Data<InVecDeriv>& in) override;
+    void applyJT(const core::MechanicalParams *mparams, Data<InVecDeriv>& out, const Data<OutVecDeriv>& in) override;
+    void applyJT(const core::ConstraintParams *cparams, Data<InMatrixDeriv>& out, const Data<OutMatrixDeriv>& in) override;
 
 protected:
 
@@ -131,7 +127,7 @@ protected:
     {
     }
 
-        helper::gl::GLSLShader shader;
+        gl::GLSLShader shader;
 
         BaseMeshTopology* inputTopo;
         BaseMeshTopology* outputTopo;

### 2021-01-13 src/SofaShells/mapping/BezierTriangleMechanicalMapping.cpp

@@ -23,11 +23,11 @@
 * Contact information: contact@sofa-framework.org                             *
 ******************************************************************************/
 
-#include "BezierTriangleMechanicalMapping.inl"
+#include <SofaShells/mapping/BezierTriangleMechanicalMapping.inl>
 #include <sofa/defaulttype/VecTypes.h>
 #include <sofa/defaulttype/RigidTypes.h>
 #include <sofa/core/ObjectFactory.h>
-#include "../initPluginShells.h"
+#include <SofaShells/config.h>
 
 namespace sofa
 {

### 2021-01-13 src/SofaShells/mapping/BezierTriangleMechanicalMapping.h

@@ -28,24 +28,20 @@
 
 #include <sofa/core/Mapping.h>
 
-
-
 #include <sofa/core/behavior/MechanicalState.h>
 #include <sofa/helper/vector.h>
 
-
-#include <sofa/helper/gl/GLSLShader.h>
+#include <sofa/gl/GLSLShader.h>
 
 #include <SofaBaseLinearSolver/CompressedRowSparseMatrix.h>
 #include <SofaBaseTopology/TriangleSetTopologyContainer.h>
-//#include <sofa/component/topology/TriangleSubdivisionTopologicalMapping.h>
 #include <sofa/simulation/AnimateBeginEvent.h>
 
 #include <sofa/defaulttype/VecTypes.h>
 
 #include <sofa/helper/system/thread/CTime.h>
 
-#include "../forcefield/BezierTriangularBendingFEMForceField.h"
+#include <SofaShells/forcefield/BezierTriangularBendingFEMForceField.h>
 
 // Use quaternions for rotations
 //#define ROTQ
@@ -101,7 +97,7 @@ public:
     typedef helper::Quater<Real> Quat;
 
 
-    typedef sofa::core::topology::BaseMeshTopology::index_type Index;
+    typedef sofa::Index Index;
     //typedef BaseMeshTopology::Edge              Edge;
     typedef BaseMeshTopology::SeqEdges          SeqEdges;
     typedef BaseMeshTopology::Triangle          Triangle;
@@ -131,16 +127,16 @@ public:
     {
     }
 
-    void init();
-    void reinit();
-    virtual void draw(const core::visual::VisualParams* vparams);
+    void init() override;
+    void reinit() override;
+    void draw(const core::visual::VisualParams* vparams) override;
 
 
-    void apply(const core::MechanicalParams *mparams, Data<OutVecCoord>& out, const Data<InVecCoord>& in);
-    const sofa::defaulttype::BaseMatrix* getJ(const core::MechanicalParams * mparams);
-    void applyJ(const core::MechanicalParams *mparams, Data<OutVecDeriv>& out, const Data<InVecDeriv>& in);
-    void applyJT(const core::MechanicalParams *mparams, Data<InVecDeriv>& out, const Data<OutVecDeriv>& in);
-    void applyJT(const core::ConstraintParams *cparams, Data<InMatrixDeriv>& out, const Data<OutMatrixDeriv>& in);
+    void apply(const core::MechanicalParams *mparams, Data<OutVecCoord>& out, const Data<InVecCoord>& in) override;
[…]

### 2021-01-13 src/SofaShells/misc/Optimize2DSurface.cpp

@@ -3,8 +3,8 @@
 // triangular networks.
 //
 
-#include "../initPluginShells.h"
-#include "Optimize2DSurface.inl"
+#include <SofaShells/config.h>
+#include <SofaShells/misc/Optimize2DSurface.inl>
 
 namespace sofa
 {

### 2021-01-13 src/SofaShells/misc/Optimize2DSurface.h

@@ -13,7 +13,7 @@
 
 #include <sofa/helper/vector.h>
 
-#include "../misc/SurfaceParametrization.h"
+#include <SofaShells/misc/SurfaceParametrization.h>
 
 // Return non-zero if triangle with points (a,b,c) is defined in
 // counter-clockwise direction. (2D case)

### 2021-01-13 src/SofaShells/misc/PointProjection.cpp

@@ -2,8 +2,8 @@
 // Class for performing projection of points onto triangular surface
 //
 
-#include "../initPluginShells.h"
-#include "PointProjection.inl"
+#include <SofaShells/config.h>
+#include <SofaShells/misc/PointProjection.inl>
 
 namespace sofa
 {

### 2021-01-13 src/SofaShells/misc/SurfaceParametrization.cpp

@@ -2,8 +2,8 @@
 // Class for parametrizing 3D surface in 2D domain.
 //
 
-#include "../initPluginShells.h"
-#include "SurfaceParametrization.inl"
+#include <SofaShells/config.h>
+#include <SofaShells/misc/SurfaceParametrization.inl>
 
 namespace sofa
 {

### 2021-01-13 src/SofaShells/shells2/fem/BezierShellInterpolation.cpp

@@ -1,5 +1,5 @@
-#include "../../initPluginShells.h"
-#include "BezierShellInterpolation.inl"
+#include <SofaShells/config.h>
+#include <SofaShells/shells2/fem/BezierShellInterpolation.inl>
 #include <sofa/defaulttype/VecTypes.h>
 #include <sofa/defaulttype/RigidTypes.h>
 #include <sofa/core/ObjectFactory.h>

### 2021-01-13 src/SofaShells/shells2/fem/BezierShellInterpolation.h

@@ -8,12 +8,11 @@
 #ifndef SOFA_COMPONENT_FEM_BEZIERSHELLINTERPOLATION_H
 #define SOFA_COMPONENT_FEM_BEZIERSHELLINTERPOLATION_H
 
-#include "../../initPluginShells.h"
+#include <SofaShells/config.h>
 #include <sofa/core/behavior/MechanicalState.h>
 #include <SofaBaseMechanics/MechanicalObject.h>
 #include <SofaTopologyMapping/Mesh2PointTopologicalMapping.h>
-//#include <sofa/core/behavior/Mass.h>
-//#include <sofa/core/objectmodel/Data.h>
+
 #include <sofa/defaulttype/SolidTypes.h>
 #include <sofa/defaulttype/VecTypes.h>
 #include <sofa/defaulttype/RigidTypes.h>
@@ -52,10 +51,8 @@ class BezierShellInterpolation : public virtual sofa::core::objectmodel::BaseObj
         typedef typename DataTypes::Deriv Deriv;
         typedef typename Coord::value_type Real;
 
-        typedef sofa::core::topology::BaseMeshTopology::index_type Index;
+        typedef sofa::Index Index;
         typedef sofa::core::topology::BaseMeshTopology::TriangleID ElementID;
-        //typedef sofa::helper::vector<BaseMeshTopology::EdgeID> VecElementID;
-        //typedef sofa::helper::vector<BaseMeshTopology::Edge> VecEdges;
         typedef helper::vector<Index> VecIndex;
 
         typedef typename  sofa::defaulttype::SolidTypes<Real>::Transform Transform;
@@ -168,7 +165,7 @@ class BezierShellInterpolation : public virtual sofa::core::objectmodel::BaseObj
             if(triHandler) delete triHandler;
         }
 
-        virtual std::string getTemplateName() const
+        virtual std::string getTemplateName() const override
         {
             return templateName(this);
         }
@@ -193,14 +190,14 @@ class BezierShellInterpolation : public virtual sofa::core::objectmodel::BaseObj
         /**
          * @brief SceneGraph callback initialization method.
          */
-        void init();
+        void init() override;
 
         /**
          * @brief SceneGraph callback backward initialization method.
          */
-        void bwdInit();
+        void bwdInit() override;
 
-        void reinit() { init(); bwdInit(); }
+        void reinit() override { init(); bwdInit(); }
 
         /**
          * @brief SceneGraph callback backward reset method.
[…]

### 2021-01-13 src/SofaShells/shells2/fem/BezierShellInterpolationM.cpp

@@ -1,5 +1,5 @@
-#include "../../initPluginShells.h"
-#include "BezierShellInterpolationM.inl"
+#include <SofaShells/config.h>
+#include <SofaShells/shells2/fem/BezierShellInterpolationM.inl>
 #include <sofa/defaulttype/VecTypes.h>
 #include <sofa/defaulttype/RigidTypes.h>
 #include <sofa/core/ObjectFactory.h>

### 2021-01-13 src/SofaShells/shells2/fem/BezierShellInterpolationM.h

@@ -8,14 +8,13 @@
 #ifndef SOFA_COMPONENT_FEM_BEZIERSHELLINTERPOLATIONM_H
 #define SOFA_COMPONENT_FEM_BEZIERSHELLINTERPOLATIONM_H
 
-#include "../../initPluginShells.h"
-#include "BezierShellInterpolation.h"
+#include <SofaShells/config.h>
+#include <SofaShells/shells2/fem/BezierShellInterpolation.h>
 
 #include <sofa/core/behavior/MechanicalState.h>
 #include <SofaBaseMechanics/MechanicalObject.h>
-#include <SofaTopologyMapping/Mesh2PointTopologicalMapping.h>
-//#include <sofa/core/behavior/Mass.h>
-//#include <sofa/core/objectmodel/Data.h>
+//#include <SofaTopologyMapping/Mesh2PointTopologicalMapping.h>
+
 #include <sofa/defaulttype/SolidTypes.h>
 #include <sofa/defaulttype/VecTypes.h>
 #include <sofa/defaulttype/RigidTypes.h>
@@ -25,7 +24,6 @@
 
 #include <sofa/helper/vector.h>
 #include <sofa/defaulttype/Vec.h>
-//#include <sofa/defaulttype/Mat.h>
 
 #include <sofa/core/objectmodel/BaseObject.h>
 
@@ -82,7 +80,7 @@ class BezierShellInterpolationM : public BezierShellInterpolation<TIn>
         BezierShellInterpolationM() {}
         virtual ~BezierShellInterpolationM() {}
 
-        virtual std::string getTemplateName() const
+        virtual std::string getTemplateName() const override
         {
             return templateName(this);
         }

### 2021-01-13 src/SofaShells/shells2/forcefield/BezierShellForceField.cpp

@@ -22,12 +22,12 @@
 *                                                                             *
 * Contact information: contact@sofa-framework.org                             *
 ******************************************************************************/
-#include "BezierShellForceField.inl"
+#include <SofaShells/shells2/forcefield/BezierShellForceField.inl>
 #include <sofa/core/behavior/ForceField.inl>
 #include <sofa/defaulttype/RigidTypes.h>
 #include <SofaBaseMechanics/MechanicalObject.h>
 #include <sofa/core/ObjectFactory.h>
-#include "../../initPluginShells.h"
+#include <SofaShells/config.h>
 
 
 namespace sofa

### 2021-01-13 src/SofaShells/shells2/forcefield/BezierShellForceField.h

@@ -29,16 +29,14 @@
 #include <sofa/core/behavior/MechanicalState.h>
 #include <sofa/core/objectmodel/Data.h>
 
-//#include <sofa/component/component.h>
-
 #include <sofa/core/topology/BaseMeshTopology.h>
 #include <SofaBaseTopology/TopologyData.h>
 
 #include <sofa/helper/OptionsGroup.h>
 
-#include "../../controller/MeshInterpolator.h"
-#include "../../engine/JoinMeshPoints.h"
-#include "../fem/BezierShellInterpolation.h"
+#include <SofaShells/controller/MeshInterpolator.h>
+#include <SofaShells/engine/JoinMeshPoints.h>
+#include <SofaShells/shells2/fem/BezierShellInterpolation.h>
 
 
 // Uncomment the following to use quaternions instead of matrices for
@@ -96,7 +94,7 @@ class BezierShellForceField : public core::behavior::ForceField<DataTypes>
 
         typedef Vec3Types::VecCoord VecCoordHigh;
 
-        typedef sofa::core::topology::BaseMeshTopology::index_type Index;
+        typedef sofa::Index Index;
         typedef sofa::core::topology::BaseMeshTopology::Triangle Triangle;
         typedef sofa::core::topology::BaseMeshTopology::SeqTriangles SeqTriangles;
         typedef helper::vector<Index> VecIndex;
@@ -237,16 +235,16 @@ public:
         BezierShellForceField();
 
         virtual ~BezierShellForceField();
-        virtual void init();
-        virtual void reinit();
-        virtual void addForce(const sofa::core::MechanicalParams* /*mparams*/, DataVecDeriv& dataF, const DataVecCoord& dataX, const DataVecDeriv& /*dataV*/ ) ;
-        virtual void addDForce(const sofa::core::MechanicalParams* /*mparams*/, DataVecDeriv& datadF, const DataVecDeriv& datadX ) ;
-        virtual void addKToMatrix(const core::MechanicalParams* mparams, const sofa::core::behavior::MultiMatrixAccessor* matrix);
-        virtual void handleTopologyChange();
+        void init() override;
+        void reinit() override;
+        void addForce(const sofa::core::MechanicalParams* /*mparams*/, DataVecDeriv& dataF, const DataVecCoord& dataX, const DataVecDeriv& /*dataV*/ ) override;
[…]

### 2021-01-13 src/SofaShells/shells2/mapping/BezierShellMechanicalMapping.h

@@ -28,13 +28,10 @@
 
 #include <sofa/core/Mapping.h>
 
-
-
 #include <sofa/core/behavior/MechanicalState.h>
 #include <sofa/helper/vector.h>
 
-
-#include <sofa/helper/gl/GLSLShader.h>
+#include <sofa/gl/GLSLShader.h>
 
 #include <SofaBaseLinearSolver/CompressedRowSparseMatrix.h>
 #include <SofaBaseTopology/TriangleSetTopologyContainer.h>
@@ -42,7 +39,7 @@
 
 #include <sofa/defaulttype/VecTypes.h>
 
-#include "../fem/BezierShellInterpolationM.h"
+#include <SofaShells/shells2/fem/BezierShellInterpolationM.h>
 
 
 namespace sofa
@@ -92,8 +89,7 @@ public:
     typedef typename sofa::component::fem::BezierShellInterpolationM<TIn,TOut>::ShapeFunctions ShapeFunctions;
     typedef typename sofa::component::fem::BezierShellInterpolationM<TIn,TOut>::VecShapeFunctions VecShapeFunctions;
 
-    typedef sofa::core::topology::BaseMeshTopology::index_type Index;
-    //typedef BaseMeshTopology::Edge              Edge;
+    typedef sofa::Index Index;
     typedef BaseMeshTopology::SeqEdges          SeqEdges;
     typedef BaseMeshTopology::Triangle          Triangle;
     typedef BaseMeshTopology::SeqTriangles      SeqTriangles;
@@ -120,16 +116,15 @@ public:
     {
     }
 
-    void init();
-    void reinit();
+    void init() override;
+    void reinit() override;
     //virtual void draw(const core::visual::VisualParams* vparams);
 
 
-    void apply(const core::MechanicalParams *mparams, Data<OutVecCoord>& out, const Data<InVecCoord>& in);
-    //const sofa::defaulttype::BaseMatrix* getJ(const core::MechanicalParams * mparams);
-    void applyJ(const core::MechanicalParams *mparams, Data<OutVecDeriv>& out, const Data<InVecDeriv>& in);
-    void applyJT(const core::MechanicalParams *mparams, Data<InVecDeriv>& out, const Data<OutVecDeriv>& in);
-    void applyJT(const core::ConstraintParams *cparams, Data<InMatrixDeriv>& out, const Data<OutMatrixDeriv>& in);
[…]

### 2021-08-17 src/SofaShells/engine/FindClosePoints.h
intermediate commit for updating towards v21.06

@@ -58,7 +58,7 @@ public:
     Data<Real>      f_input_threshold;
     Data<VecCoord>  f_input_position;
 
-    Data< helper::vector< helper::fixed_array<Index,2> > > f_output_closePoints;
+    Data< type::vector< type::fixed_array<Index,2> > > f_output_closePoints;
 
 };
 

### 2021-08-17 src/SofaShells/controller/AdaptiveCuttingController.cpp
Update plugin for v21.06

@@ -20,20 +20,20 @@ SOFA_DECL_CLASS(AdaptiveCuttingController)
 int AdaptiveCuttingControllerClass = core::RegisterObject(
     "Controller that handles the cutting method based on mesh adaptivity.")
 #ifdef SOFA_FLOAT
-.add< AdaptiveCuttingController<defaulttype::Vec3fTypes> >(true) // default template
+.add< AdaptiveCuttingController<type::Vec3fTypes> >(true) // default template
 #else
-.add< AdaptiveCuttingController<defaulttype::Vec3dTypes> >(true) // default template
+.add< AdaptiveCuttingController<type::Vec3dTypes> >(true) // default template
 # ifndef SOFA_DOUBLE
-.add< AdaptiveCuttingController<defaulttype::Vec3fTypes> >()
+.add< AdaptiveCuttingController<type::Vec3fTypes> >()
 # endif
 #endif
 ;
 
 #ifndef SOFA_FLOAT
-template class SOFA_SHELLS_API AdaptiveCuttingController<defaulttype::Vec3dTypes>;
+template class SOFA_SHELLS_API AdaptiveCuttingController<type::Vec3dTypes>;
 #endif //SOFA_FLOAT
 #ifndef SOFA_DOUBLE
-template class SOFA_SHELLS_API AdaptiveCuttingController<defaulttype::Vec3fTypes>;
+template class SOFA_SHELLS_API AdaptiveCuttingController<type::Vec3fTypes>;
 #endif //SOFA_DOUBLE
 
 } // namespace controller

### 2021-08-17 src/SofaShells/controller/AdaptiveCuttingController.h

@@ -4,7 +4,7 @@
 #include <SofaUserInteraction/Controller.h>
 
 #include <sofa/defaulttype/VecTypes.h>
-#include <sofa/defaulttype/Vec.h>
+#include <sofa/type/Vec.h>
 
 #include <sofa/core/behavior/MechanicalState.h>
 #include <sofa/core/topology/BaseMeshTopology.h>
@@ -16,7 +16,7 @@
 #include <SofaBaseTopology/TopologyData.h>
 
 #include <sofa/helper/map.h>
-#include <sofa/helper/vector.h>
+#include <sofa/type/vector.h>
 
 #include <SofaShells/controller/Test2DAdapter.h>
 
@@ -50,12 +50,12 @@ public:
     //typedef typename DataTypes::VecDeriv VecDeriv;
     typedef typename Coord::value_type Real;
 
-    typedef sofa::defaulttype::Vec<2, Real> Vec2;
-    typedef sofa::defaulttype::Vec<3, Real> Vec3;
-    //typedef sofa::defaulttype::Mat<2,2,Real> Mat22;
-    //typedef sofa::defaulttype::Mat<3,3,Real> Mat33;
-    //typedef helper::vector<Vec2> VecVec2;
-    //typedef helper::vector<Vec3> VecVec3;
+    typedef sofa::type::Vec<2, Real> Vec2;
+    typedef sofa::type::Vec<3, Real> Vec3;
+    //typedef sofa::type::Mat<2,2,Real> Mat22;
+    //typedef sofa::type::Mat<3,3,Real> Mat33;
+    //typedef type::vector<Vec2> VecVec2;
+    //typedef type::vector<Vec3> VecVec3;
 
 
     typedef sofa::component::topology::EdgeSetTopologyContainer::Edge               Edge;
@@ -65,7 +65,7 @@ public:
     typedef sofa::component::topology::TriangleSetTopologyContainer::TrianglesAroundVertex  TrianglesAroundVertex;
     typedef sofa::component::topology::TriangleSetTopologyContainer::TrianglesAroundEdge    TrianglesAroundEdge;
     typedef sofa::component::topology::TriangleSetTopologyContainer::EdgesInTriangle        EdgesInTriangle;
-    typedef sofa::helper::vector<Index> VecIndex;
+    typedef sofa::type::vector<Index> VecIndex;
 
     enum { InvalidID = sofa::core::topology::Topology::InvalidID };
 

### 2021-08-17 src/SofaShells/controller/CudaTest2DAdapter.h

@@ -5,7 +5,7 @@
 #include <SofaShells/controller/Test2DAdapter.h>
 
 #include <sofa/helper/fixed_array.h>
-#include <sofa/helper/vector.h>
+#include <sofa/type/vector.h>
 
 
 namespace sofa
@@ -46,13 +46,13 @@ public:
     typedef sofa::component::topology::TriangleSetTopologyContainer::TriangleID     Index;
 
     // Graph colours (of independent sets)
-    helper::vector< sofa::gpu::cuda::CudaVector<Index> > colours;
+    type::vector< sofa::gpu::cuda::CudaVector<Index> > colours;
 
     struct TriangleData {
-        helper::fixed_array<Index,3> nodes;
+        type::fixed_array<Index,3> nodes;
         Coord normal;
         Real functional;
-        helper::fixed_array<Deriv,3> gradient;
+        type::fixed_array<Deriv,3> gradient;
     };
 
     struct PointData {

### 2021-08-17 src/SofaShells/controller/MeshChangedEvent.h

@@ -15,7 +15,7 @@ namespace core
 namespace objectmodel
 {
 
-using namespace sofa::defaulttype;
+using namespace sofa::type;
 
 // Sent by MeshInterpolator when the mesh changes
 class MeshChangedEvent : public sofa::core::objectmodel::Event

### 2021-08-17 src/SofaShells/controller/MeshInterpolator.h

@@ -4,7 +4,7 @@
 #include <SofaUserInteraction/Controller.h>
 #include <sofa/defaulttype/RigidTypes.h>
 #include <sofa/defaulttype/VecTypes.h>
-#include <sofa/defaulttype/Vec.h>
+#include <sofa/type/Vec.h>
 
 #include <SofaShells/config.h>
 
@@ -28,7 +28,7 @@ public:
     typedef typename DataTypes::Coord Coord;
     typedef typename DataTypes::Deriv Deriv;
     typedef typename Coord::value_type Real;
-    typedef sofa::defaulttype::Vec<3,Real> Vec3;
+    typedef sofa::type::Vec<3,Real> Vec3;
 
 protected:
 
@@ -56,12 +56,12 @@ public:
     Data<Real>              f_increment;
 
     Data<VecCoord>          f_startPosition;
-    Data<helper::vector<Vec3> > f_startNormals;
+    Data<sofa::type::vector<Vec3> > f_startNormals;
     Data<VecCoord>          f_endPosition;
-    Data<helper::vector<Vec3> > f_endNormals;
+    Data<sofa::type::vector<Vec3> > f_endNormals;
 
     Data<VecCoord>          f_position;
-    Data<helper::vector<Vec3> > f_normals;
+    Data<sofa::type::vector<Vec3> > f_normals;
 
     void onEndAnimationStep(const double dt) override;
 
@@ -79,16 +79,16 @@ private:
 #if defined(WIN32) && !defined(SOFA_COMPONENT_CONTROLLER_MESHINTERPOLATION_CPP)
 #pragma warning(disable : 4231)
 #ifndef SOFA_FLOAT
-//template class SOFA_SHELLS_API MeshInterpolator<defaulttype::Vec1dTypes>;
-//template class SOFA_SHELLS_API MeshInterpolator<defaulttype::Vec2dTypes>;
-//template class SOFA_SHELLS_API MeshInterpolator<defaulttype::Vec3dTypes>;
+//template class SOFA_SHELLS_API MeshInterpolator<type::Vec1dTypes>;
+//template class SOFA_SHELLS_API MeshInterpolator<type::Vec2dTypes>;
+//template class SOFA_SHELLS_API MeshInterpolator<type::Vec3dTypes>;
 //template class SOFA_SHELLS_API MeshInterpolator<defaulttype::Rigid2dTypes>;
[…]

### 2021-08-17 src/SofaShells/controller/Test2DAdapter.cpp

@@ -19,20 +19,20 @@ SOFA_DECL_CLASS(Test2DAdapter)
 // Register in the Factory
 int Test2DAdapterClass = core::RegisterObject("Adaptive mesh improvement component for 2D triangular meshes (for testing)")
 #ifdef SOFA_FLOAT
-.add< Test2DAdapter<defaulttype::Vec3fTypes> >(true) // default template
+.add< Test2DAdapter<type::Vec3fTypes> >(true) // default template
 #else
-.add< Test2DAdapter<defaulttype::Vec3dTypes> >(true) // default template
+.add< Test2DAdapter<type::Vec3dTypes> >(true) // default template
 # ifndef SOFA_DOUBLE
-.add< Test2DAdapter<defaulttype::Vec3fTypes> >()
+.add< Test2DAdapter<type::Vec3fTypes> >()
 # endif
 #endif
 ;
 
 #ifndef SOFA_FLOAT
-template class SOFA_SHELLS_API Test2DAdapter<defaulttype::Vec3dTypes>;
+template class SOFA_SHELLS_API Test2DAdapter<type::Vec3dTypes>;
 #endif //SOFA_FLOAT
 #ifndef SOFA_DOUBLE
-template class SOFA_SHELLS_API Test2DAdapter<defaulttype::Vec3fTypes>;
+template class SOFA_SHELLS_API Test2DAdapter<type::Vec3fTypes>;
 #endif //SOFA_DOUBLE
 
 } // namespace controller

### 2021-08-17 src/SofaShells/controller/Test2DAdapter.h

@@ -4,7 +4,7 @@
 #include <SofaUserInteraction/Controller.h>
 
 #include <sofa/defaulttype/VecTypes.h>
-#include <sofa/defaulttype/Vec.h>
+#include <sofa/type/Vec.h>
 
 #include <sofa/core/behavior/MechanicalState.h>
 #include <sofa/core/topology/BaseMeshTopology.h>
@@ -16,7 +16,7 @@
 #include <SofaBaseTopology/TopologyData.h>
 
 #include <sofa/helper/map.h>
-#include <sofa/helper/vector.h>
+#include <sofa/type/vector.h>
 
 #include <SofaShells/misc/Optimize2DSurface.h>
 #include <SofaShells/misc/SurfaceParametrization.h>
@@ -98,12 +98,12 @@ public:
     typedef typename DataTypes::Deriv Deriv;
     typedef typename Coord::value_type Real;
 
-    typedef sofa::defaulttype::Vec<2, Real> Vec2;
-    typedef sofa::defaulttype::Vec<3, Real> Vec3;
-    typedef sofa::defaulttype::Mat<2,2,Real> Mat22;
-    typedef sofa::defaulttype::Mat<3,3,Real> Mat33;
-    //typedef helper::vector<Vec2> VecVec2;
-    typedef helper::vector<Vec3> VecVec3;
+    typedef sofa::type::Vec<2, Real> Vec2;
+    typedef sofa::type::Vec<3, Real> Vec3;
+    typedef sofa::type::Mat<2,2,Real> Mat22;
+    typedef sofa::type::Mat<3,3,Real> Mat33;
+    //typedef type::vector<Vec2> VecVec2;
+    typedef type::vector<Vec3> VecVec3;
 
 
     typedef sofa::component::topology::EdgeSetTopologyContainer::Edge               Edge;
@@ -113,7 +113,7 @@ public:
     typedef sofa::component::topology::TriangleSetTopologyContainer::TrianglesAroundVertex  TrianglesAroundVertex;
     typedef sofa::component::topology::TriangleSetTopologyContainer::TrianglesAroundEdge    TrianglesAroundEdge;
     typedef sofa::component::topology::TriangleSetTopologyContainer::EdgesInTriangle        EdgesInTriangle;
-    typedef sofa::helper::vector<Index> VecIndex;
+    typedef sofa::type::vector<Index> VecIndex;
 
     enum { InvalidID = sofa::core::topology::Topology::InvalidID };
 
@@ -159,25 +159,25 @@ public:
[…]

### 2021-08-17 src/SofaShells/controller/TriangleSwitchExample.h

@@ -4,7 +4,7 @@
 #include <SofaUserInteraction/Controller.h>
 #include <sofa/defaulttype/RigidTypes.h>
 #include <sofa/defaulttype/VecTypes.h>
-#include <sofa/defaulttype/Vec.h>
+#include <sofa/type/Vec.h>
 
 #include <sofa/core/behavior/MechanicalState.h>
 #include <SofaBaseTopology/TriangleSetTopologyContainer.h>
@@ -30,12 +30,12 @@ public:
     typedef typename DataTypes::Coord Coord;
     typedef typename DataTypes::Deriv Deriv;
     typedef typename Coord::value_type Real;
-    typedef sofa::defaulttype::Vec<3,Real> Vec3;
+    typedef sofa::type::Vec<3,Real> Vec3;
 
     typedef sofa::component::topology::TriangleSetTopologyContainer::TriangleID     Index;
     typedef sofa::component::topology::TriangleSetTopologyContainer::Triangle       Triangle;
     typedef sofa::component::topology::TriangleSetTopologyContainer::SeqTriangles   SeqTriangles;
-    typedef sofa::helper::vector<Index> VecIndex;
+    typedef sofa::type::vector<Index> VecIndex;
 
 protected:
 

### 2021-08-17 src/SofaShells/cutting/AdaptiveCutting.cpp

@@ -33,22 +33,22 @@ namespace collision
 #ifndef SOFA_DOUBLE
 helper::Creator<
     InteractionPerformer::InteractionPerformerFactory,
-    AdaptiveCuttingPerformer<defaulttype::Vec3fTypes> >
+    AdaptiveCuttingPerformer<type::Vec3fTypes> >
         AdaptiveCuttingPerformerVec3fClass("AdaptiveCutting", true);
 #endif
 #ifndef SOFA_FLOAT
 helper::Creator<
     InteractionPerformer::InteractionPerformerFactory,
-    AdaptiveCuttingPerformer<defaulttype::Vec3dTypes> >
+    AdaptiveCuttingPerformer<type::Vec3dTypes> >
         AdaptiveCuttingPerformerVec3dClass("AdaptiveCutting", true);
 #endif
 
 #ifndef SOFA_DOUBLE
-template class SOFA_SHELLS_API  AdaptiveCuttingPerformer<defaulttype::Vec3fTypes>;
+template class SOFA_SHELLS_API  AdaptiveCuttingPerformer<type::Vec3fTypes>;
 //template class SOFA_SHELLS_API  AdaptiveCuttingPerformer<defaulttype::Rigid3fTypes>;
 #endif
 #ifndef SOFA_FLOAT
-template class SOFA_SHELLS_API  AdaptiveCuttingPerformer<defaulttype::Vec3dTypes>;
+template class SOFA_SHELLS_API  AdaptiveCuttingPerformer<type::Vec3dTypes>;
 //template class SOFA_SHELLS_API  AdaptiveCuttingPerformer<defaulttype::Rigid3dTypes>;
 #endif
 

### 2021-08-17 src/SofaShells/engine/FindClosePoints.h

@@ -1,7 +1,7 @@
 #ifndef SOFA_COMPONENT_ENGINE_FINDCLOSEPOINTS_H
 #define SOFA_COMPONENT_ENGINE_FINDCLOSEPOINTS_H
 
-#include <sofa/defaulttype/Vec.h>
+#include <sofa/type/Vec.h>
 #include <sofa/core/DataEngine.h>
 #include <sofa/core/objectmodel/BaseObject.h>
 #include <sofa/defaulttype/VecTypes.h>

### 2021-08-17 src/SofaShells/engine/JoinMeshPoints.h

@@ -1,10 +1,10 @@
 #ifndef SOFA_COMPONENT_ENGINE_JOINMESHPOINTS_H
 #define SOFA_COMPONENT_ENGINE_JOINMESHPOINTS_H
 
-#include <sofa/defaulttype/Vec.h>
+#include <sofa/type/Vec.h>
 #include <sofa/core/DataEngine.h>
 #include <sofa/core/objectmodel/BaseObject.h>
-#include <sofa/defaulttype/Vec.h>
+#include <sofa/type/Vec.h>
 #include <sofa/defaulttype/VecTypes.h>
 #include <sofa/defaulttype/RigidTypes.h>
 #include <SofaShells/config.h>
@@ -34,7 +34,7 @@ public:
     typedef typename DataTypes::Coord    Coord;
     typedef typename Coord::value_type   Real;
 
-    typedef defaulttype::Vec<3,Real> Vec3;
+    typedef type::Vec<3,Real> Vec3;
     typedef unsigned int Index;
 
 protected:
@@ -90,33 +90,33 @@ public:
 
     // Data
 
-    Data< helper::vector< helper::fixed_array<Index,2> > > f_input_joinPoints;
+    Data< type::vector< type::fixed_array<Index,2> > > f_input_joinPoints;
 
     Data<VecCoord> f_input_position;
-    Data< helper::vector<Vec3> > f_input_normals;
-    Data< helper::vector< helper::fixed_array<Index,2> > > f_input_edges;
-    Data< helper::vector< helper::fixed_array<Index,3> > > f_input_triangles;
-    Data< helper::vector< helper::fixed_array<Index,4> > > f_input_quads;
-    Data< helper::vector< helper::fixed_array<Index,4> > > f_input_tetrahedra;
-    Data< helper::vector< helper::fixed_array<Index,8> > > f_input_hexahedra;
+    Data< type::vector<Vec3> > f_input_normals;
+    Data< type::vector< type::fixed_array<Index,2> > > f_input_edges;
+    Data< type::vector< type::fixed_array<Index,3> > > f_input_triangles;
+    Data< type::vector< type::fixed_array<Index,4> > > f_input_quads;
+    Data< type::vector< type::fixed_array<Index,4> > > f_input_tetrahedra;
+    Data< type::vector< type::fixed_array<Index,8> > > f_input_hexahedra;
 
     Data<VecCoord> f_output_position;
-    Data< helper::vector<Vec3> > f_output_normals;
[…]

### 2021-08-17 src/SofaShells/forcefield/BezierTriangularBendingFEMForceField.h

@@ -56,8 +56,8 @@ namespace component
 namespace forcefield
 {
 
-using namespace sofa::defaulttype;
-using sofa::helper::vector;
+using namespace sofa::type;
+using sofa::type::vector;
 using namespace sofa::component::topology;
 using namespace sofa::core::behavior;
 
@@ -89,12 +89,12 @@ class BezierTriangularBendingFEMForceField : public core::behavior::ForceField<D
 
         typedef Mat<3,3,Real> Mat33;
 
-        typedef helper::Quater<Real> Quat;
+        typedef Quat<Real> Quat;
 
         typedef Data<VecCoord>                              DataVecCoord;
         typedef Data<VecDeriv>                              DataVecDeriv;
 
-        typedef Vec3Types::VecCoord VecCoordHigh;
+        typedef sofa::defaulttype::Vec3Types::VecCoord VecCoordHigh;
 
         typedef sofa::Index Index;
         typedef sofa::core::topology::BaseMeshTopology::Triangle Triangle;
@@ -127,11 +127,11 @@ public:
         {
             public:
 
-                helper::fixed_array <Vec3, 3> restLocalPositions;
+                type::fixed_array <Vec3, 3> restLocalPositions;
 #ifdef CRQUAT
-                helper::fixed_array <Quat, 3> restLocalOrientationsInv;
+                type::fixed_array <Quat, 3> restLocalOrientationsInv;
 #else
-                helper::fixed_array <Transformation, 3> restLocalOrientationsInv;
+                type::fixed_array <Transformation, 3> restLocalOrientationsInv;
 #endif
 
                 // Indices of each vertex
@@ -161,10 +161,10 @@ public:
                 Vec3 P2_P1_inFrame2;
 
                 // Nodes of the Bezier triangle
-                helper::fixed_array<Vec3, 10> bezierNodes;  // ... in global frame
-                helper::fixed_array<Vec3, 10> pts;          // ... in local frame
+                type::fixed_array<Vec3, 10> bezierNodes;  // ... in global frame
+                type::fixed_array<Vec3, 10> pts;          // ... in local frame
                 // ... of the rest shape
-                helper::fixed_array<Vec3, 10> bezierNodes0; // ... in global frame
+                type::fixed_array<Vec3, 10> bezierNodes0; // ... in global frame
 
                 // the strain-displacement matrices at each Gauss point
[…]

### 2021-08-17 src/SofaShells/forcefield/CstFEMForceField.cpp

@@ -26,7 +26,7 @@
 #include <SofaShells/config.h>
 #include <SofaShells/forcefield/CstFEMForceField.inl>
 #include <sofa/core/behavior/ForceField.inl>
-#include <sofa/defaulttype/Vec.h>
+#include <sofa/type/Vec.h>
 //#include <SofaBaseMechanics/MechanicalObject.h>
 #include <sofa/core/ObjectFactory.h>
 

### 2021-08-17 src/SofaShells/forcefield/CstFEMForceField.h

@@ -42,8 +42,8 @@ namespace component
 namespace forcefield
 {
 
-using namespace sofa::defaulttype;
-using sofa::helper::vector;
+using namespace sofa::type;
+using sofa::type::vector;
 using namespace sofa::component::topology;
 using namespace sofa::core::behavior;
 
@@ -74,12 +74,12 @@ class CstFEMForceField : public core::behavior::ForceField<DataTypes>
         typedef Vec<3,Real> Vec3;
         typedef Vec<9,Real> Vec9;
 
-        typedef helper::Quater<Real> Quat;
+        typedef Quat<Real> Quat;
 
         typedef Data<VecCoord>                              DataVecCoord;
         typedef Data<VecDeriv>                              DataVecDeriv;
 
-        typedef Vec3Types::VecCoord VecCoordHigh;
+        typedef sofa::defaulttype::Vec3Types::VecCoord VecCoordHigh;
 
         typedef sofa::Index Index;
         typedef sofa::core::topology::BaseMeshTopology::Triangle Triangle;
@@ -109,10 +109,10 @@ public:
                 Index a, b, c;
 
                 // Rest position in local (in-plane) coordinates
-                helper::fixed_array <Vec3, 3> restPositions;
+                type::fixed_array <Vec3, 3> restPositions;
 
                 // Deformed position in local (in-plane) coordinates
-                helper::fixed_array <Vec3, 3> deformedPositions;
+                type::fixed_array <Vec3, 3> deformedPositions;
 
                 // Frame rotation as matrix and quaternion
                 Transformation R, Rt;
@@ -124,7 +124,7 @@ public:
                 // - element area
                 Real area;
                 // - directional vectors: 1-2, 2-3, 3-1
-                helper::fixed_array <Vec2, 3> d;
+                type::fixed_array <Vec2, 3> d;
 
                 /// Output stream
                 inline friend std::ostream& operator<< ( std::ostream& os, const TriangleInformation& /*ti*/ )
@@ -139,15 +139,15 @@ public:
                 }
         };
 
-        class TRQSTriangleHandler : public TopologyDataHandler<Triangle, sofa::helper::vector<TriangleInformation> >
+        class TRQSTriangleHandler : public TopologyDataHandler<Triangle, sofa::type::vector<TriangleInformation> >
         {
             public:
[…]

### 2021-08-17 src/SofaShells/forcefield/TriangularBendingFEMForceField.h

@@ -50,8 +50,8 @@ namespace component
 namespace forcefield
 {
 
-using namespace sofa::defaulttype;
-using sofa::helper::vector;
+using namespace sofa::type;
+using sofa::type::vector;
 using namespace sofa::component::topology;
 using namespace sofa::core::behavior;
 
@@ -80,11 +80,12 @@ class TriangularBendingFEMForceField : public core::behavior::ForceField<DataTyp
 
         typedef Vec<3,Real> Vec3;
         typedef Vec<2,Real> Vec2;
+        typedef Quat<Real> Quat;
 
         typedef Data<VecCoord>                              DataVecCoord;
         typedef Data<VecDeriv>                              DataVecDeriv;
 
-        typedef Vec3Types::VecCoord VecCoordHigh;
+        typedef sofa::defaulttype::Vec3Types::VecCoord VecCoordHigh;
 
         typedef sofa::Index Index;
         typedef sofa::core::topology::BaseMeshTopology::Triangle Triangle;
@@ -115,8 +116,8 @@ public:
             public:
 
                 // position of B, C vertices in local (in-plane) coordinates (A is naturaly [0,0])
-                helper::fixed_array <Vec3, 2> restLocalPositions;
-                helper::fixed_array <Quat, 3> restLocalOrientations;
+                type::fixed_array <Vec3, 2> restLocalPositions;
+                type::fixed_array <Quat, 3> restLocalOrientations;
 
                 /// material stiffness matrices of each tetrahedron
                 MaterialStiffness materialMatrix;
@@ -163,15 +164,15 @@ public:
                 }
         };
 
-        class TRQSTriangleHandler : public TopologyDataHandler<Triangle, sofa::helper::vector<TriangleInformation> >
+        class TRQSTriangleHandler : public TopologyDataHandler<Triangle, sofa::type::vector<TriangleInformation> >
         {
             public:
-                TRQSTriangleHandler(TriangularBendingFEMForceField<DataTypes>* _ff, TriangleData<sofa::helper::vector<TriangleInformation> >* _data) : TopologyDataHandler<Triangle, sofa::helper::vector<TriangleInformation> >(_data), ff(_ff) {}
+                TRQSTriangleHandler(TriangularBendingFEMForceField<DataTypes>* _ff, TriangleData<sofa::type::vector<TriangleInformation> >* _data) : TopologyDataHandler<Triangle, sofa::type::vector<TriangleInformation> >(_data), ff(_ff) {}
 
[…]

### 2021-08-17 src/SofaShells/forcefield/TriangularShellForceField.h

@@ -54,8 +54,8 @@ namespace component
 namespace forcefield
 {
 
-using namespace sofa::defaulttype;
-using sofa::helper::vector;
+using namespace sofa::type;
+using sofa::type::vector;
 using namespace sofa::component::topology;
 using namespace sofa::core::behavior;
 
@@ -86,12 +86,12 @@ class TriangularShellForceField : public core::behavior::ForceField<DataTypes>
         typedef Vec<3,Real> Vec3;
         typedef Vec<9,Real> Vec9;
 
-        typedef helper::Quater<Real> Quat;
+        typedef Quat<Real> Quat;
 
         typedef Data<VecCoord>                              DataVecCoord;
         typedef Data<VecDeriv>                              DataVecDeriv;
 
-        typedef Vec3Types::VecCoord VecCoordHigh;
+        typedef sofa::defaulttype::Vec3Types::VecCoord VecCoordHigh;
 
         typedef sofa::Index Index;
         typedef sofa::core::topology::BaseMeshTopology::Triangle Triangle;
@@ -109,7 +109,7 @@ class TriangularShellForceField : public core::behavior::ForceField<DataTypes>
         typedef Mat<18, 18, Real> StiffnessMatrixFull;          // stiffness matrix for shell (= bending plate + membrane)
 
         typedef void (TriangularShellForceField<DataTypes>::*compstiff)(StiffnessMatrix &K, TriangleInformation &tinfo);
-        typedef helper::fixed_array<Real, 10> AndesBeta;
+        typedef type::fixed_array<Real, 10> AndesBeta;
 
         sofa::core::topology::BaseMeshTopology* _topology;
 
@@ -123,15 +123,15 @@ public:
                 Index a, b, c;
 
                 // Rest position in local (in-plane) coordinates
-                helper::fixed_array <Vec3, 3> restPositions;
+                type::fixed_array <Vec3, 3> restPositions;
 #ifdef CRQUAT
-                helper::fixed_array <Quat, 3> restOrientationsInv;
+                type::fixed_array <Quat, 3> restOrientationsInv;
 #else
-                helper::fixed_array <Transformation, 3> restOrientationsInv;
+                type::fixed_array <Transformation, 3> restOrientationsInv;
 #endif
 
                 // Deformed position in local (in-plane) coordinates
-                helper::fixed_array <Vec3, 3> deformedPositions;
+                type::fixed_array <Vec3, 3> deformedPositions;
 
[…]

### 2021-08-17 src/SofaShells/mapping/BendingPlateMechanicalMapping.h

@@ -31,7 +31,7 @@
 
 
 #include <sofa/core/behavior/MechanicalState.h>
-#include <sofa/helper/vector.h>
+#include <sofa/type/vector.h>
 
 #include <sofa/gl/GLSLShader.h>
 #include <SofaBaseTopology/TriangleSetTopologyContainer.h>
@@ -51,7 +51,7 @@ namespace component
 namespace mapping
 {
 
-using namespace sofa::defaulttype;
+using namespace sofa::type;
 using namespace sofa::component::forcefield;
 using namespace sofa::component::topology;
 using namespace sofa::helper::system::thread;
@@ -141,10 +141,10 @@ protected:
         OutVecCoord verticesTarget;
         SeqTriangles trianglesTarget;
 
-        helper::vector<Vec3> colourMapping;
-        helper::vector<Vec3> coloursPerVertex;
-        helper::vector<Real> vectorErrorCoarse;
-        helper::vector<Real> vectorErrorTarget;
+        type::vector<Vec3> colourMapping;
+        type::vector<Vec3> coloursPerVertex;
+        type::vector<Real> vectorErrorCoarse;
+        type::vector<Real> vectorErrorTarget;
 
         // Pointer on the forcefield associated with the in topology
         TriangularBendingFEMForceField<In>* triangularBendingForcefield;
@@ -155,21 +155,21 @@ protected:
 
         void HSL2RGB(Vec3 &rgb, Real h, Real sl, Real l);
         void MeasureError();
-        Real DistanceHausdorff(BaseMeshTopology *topo1, BaseMeshTopology *topo2, helper::vector<Real> &vectorError);
-        void ComputeNormals(helper::vector<Vec3> &normals);
-        void FindTriangleInNormalDirection(const InVecCoord& highResVertices, const SeqTriangles highRestriangles, const helper::vector<Vec3> &normals);
+        Real DistanceHausdorff(BaseMeshTopology *topo1, BaseMeshTopology *topo2, type::vector<Real> &vectorError);
+        void ComputeNormals(type::vector<Vec3> &normals);
+        void FindTriangleInNormalDirection(const InVecCoord& highResVertices, const SeqTriangles highRestriangles, const type::vector<Vec3> &normals);
 
         // Computes the barycentric coordinates of a vertex within a triangle
         void computeBaryCoefs(Vec3 &baryCoefs, const Vec3 &p, const Vec3 &a, const Vec3 &b, const Vec3 &c);
 
[…]

### 2021-08-17 src/SofaShells/mapping/BezierTriangleMechanicalMapping.h

@@ -29,7 +29,7 @@
 #include <sofa/core/Mapping.h>
 
 #include <sofa/core/behavior/MechanicalState.h>
-#include <sofa/helper/vector.h>
+#include <sofa/type/vector.h>
 
 #include <sofa/gl/GLSLShader.h>
 
@@ -58,7 +58,7 @@ namespace component
 namespace mapping
 {
 
-using namespace sofa::defaulttype;
+using namespace sofa::type;
 using namespace sofa::component::topology;
 using namespace sofa::helper::system::thread;
 using namespace core::topology;
@@ -94,7 +94,7 @@ public:
     typedef Vec<3, Real> Vec3;
     typedef Mat<3, 3, Real> Mat33;
 
-    typedef helper::Quater<Real> Quat;
+    typedef Quat<Real> Quat;
 
 
     typedef sofa::Index Index;
@@ -105,7 +105,7 @@ public:
 
     enum { NIn = sofa::defaulttype::DataTypeInfo<InDeriv>::Size };
     enum { NOut = sofa::defaulttype::DataTypeInfo<OutDeriv>::Size };
-    typedef defaulttype::Mat<NOut, NIn, Real> MBloc;
+    typedef type::Mat<NOut, NIn, Real> MBloc;
     typedef sofa::component::linearsolver::CompressedRowSparseMatrix<MBloc> MatrixType;
 
     BezierTriangleMechanicalMapping(core::State<In>* from, core::State<Out>* to)
@@ -199,8 +199,8 @@ protected:
     typedef struct {
 
         // Nodes of the Bézier triangle
-        sofa::helper::fixed_array<Vec3, 10> bezierNodes;
-        sofa::helper::fixed_array<Vec3, 10> bezierNodesV;
+        sofa::type::fixed_array<Vec3, 10> bezierNodes;
+        sofa::type::fixed_array<Vec3, 10> bezierNodesV;
 
         // Segments in the reference frame
         Vec3 P0_P1;
@@ -211,7 +211,7 @@ protected:
         Vec3 P2_P1;
 
         // Contains the list of poinst connected to this triangle
-        sofa::helper::vector<int> attachedPoints;
+        sofa::type::vector<int> attachedPoints;
 
     } TriangleInformation;
 
@@ -223,7 +223,7 @@ protected:
     // Pointer to the forcefield associated with the input topology
     BezierFF* bezierForcefield;
 
-    Data< helper::vector<Vec3> > normals;
+    Data< type::vector<Vec3> > normals;
[…]

### 2021-08-17 src/SofaShells/misc/Optimize2DSurface.cpp

@@ -11,11 +11,11 @@ namespace sofa
 
 #ifndef SOFA_FLOAT
 //template class SOFA_SHELLS_API Optimize2DSurface<double>;
-template class SOFA_SHELLS_API Optimize2DSurface<defaulttype::Vec3dTypes>;
+template class SOFA_SHELLS_API Optimize2DSurface<type::Vec3dTypes>;
 #endif //SOFA_FLOAT
 #ifndef SOFA_DOUBLE
 //template class SOFA_SHELLS_API Optimize2DSurface<float>;
-template class SOFA_SHELLS_API Optimize2DSurface<defaulttype::Vec3fTypes>;
+template class SOFA_SHELLS_API Optimize2DSurface<type::Vec3fTypes>;
 #endif //SOFA_DOUBLE
 
 }

### 2021-08-17 src/SofaShells/misc/Optimize2DSurface.h

@@ -9,9 +9,9 @@
 #include <SofaBaseTopology/TriangleSetTopologyContainer.h>
 
 #include <sofa/defaulttype/VecTypes.h>
-#include <sofa/defaulttype/Vec.h>
+#include <sofa/type/Vec.h>
 
-#include <sofa/helper/vector.h>
+#include <sofa/type/vector.h>
 
 #include <SofaShells/misc/SurfaceParametrization.h>
 
@@ -43,11 +43,11 @@ class Optimize2DSurface
         typedef typename Coord::value_type Real;
 
 
-        typedef sofa::defaulttype::Vec<2, Real> Vec2;
-        typedef sofa::defaulttype::Vec<3, Real> Vec3;
+        typedef sofa::type::Vec<2, Real> Vec2;
+        typedef sofa::type::Vec<3, Real> Vec3;
 
-        typedef sofa::defaulttype::Mat<2,2, Real> Mat22;
-        typedef sofa::defaulttype::Mat<3,3, Real> Mat33;
+        typedef sofa::type::Mat<2,2, Real> Mat22;
+        typedef sofa::type::Mat<3,3, Real> Mat33;
 
         typedef sofa::component::topology::EdgeSetTopologyContainer::Edge                       Edge;
         typedef sofa::component::topology::EdgeSetTopologyContainer::EdgesAroundVertex          EdgesAroundVertex;
@@ -55,10 +55,10 @@ class Optimize2DSurface
         typedef sofa::component::topology::TriangleSetTopologyContainer::Triangle               Triangle;
         typedef sofa::component::topology::TriangleSetTopologyContainer::TrianglesAroundVertex  TrianglesAroundVertex;
 
-        typedef sofa::helper::vector<Real>      VecReal;
-        typedef sofa::helper::vector<Vec2>      VecVec2;
-        typedef sofa::helper::vector<Vec3>      VecVec3;
-        //typedef sofa::helper::vector<Index>     VecIndex;
+        typedef sofa::type::vector<Real>      VecReal;
+        typedef sofa::type::vector<Vec2>      VecVec2;
+        typedef sofa::type::vector<Vec3>      VecVec3;
+        //typedef sofa::type::vector<Index>     VecIndex;
 
         enum { InvalidID = sofa::core::topology::Topology::InvalidID };
 
@@ -173,7 +173,7 @@ class Optimize2DSurface
         SurfaceParametrization<Real> &m_surf;
 
         /// Orientation of each triangle.
[…]

### 2021-08-17 src/SofaShells/misc/PointProjection.h

@@ -8,9 +8,9 @@
 #include <SofaBaseTopology/TriangleSetTopologyContainer.h>
 
 #include <sofa/defaulttype/VecTypes.h>
-#include <sofa/defaulttype/Vec.h>
+#include <sofa/type/Vec.h>
 
-#include <sofa/helper/vector.h>
+#include <sofa/type/vector.h>
 #include <sofa/defaulttype/Mat.h>
 
 
@@ -27,10 +27,10 @@ class PointProjection
 {
 
     public:
-        typedef sofa::defaulttype::Vec<2, Real> Vec2;
-        typedef sofa::defaulttype::Vec<3, Real> Vec3;
+        typedef sofa::type::Vec<2, Real> Vec2;
+        typedef sofa::type::Vec<3, Real> Vec3;
 
-        typedef sofa::defaulttype::Mat<3,3, Real> Mat33;
+        typedef sofa::type::Mat<3,3, Real> Mat33;
 
         typedef sofa::component::topology::TriangleSetTopologyContainer::Edge                   Edge;
         typedef sofa::component::topology::TriangleSetTopologyContainer::TriangleID             Index;
@@ -41,8 +41,8 @@ class PointProjection
         typedef sofa::component::topology::TriangleSetTopologyContainer::SeqEdges               SeqEdges;
         typedef sofa::component::topology::TriangleSetTopologyContainer::SeqTriangles           SeqTriangles;
 
-        typedef sofa::helper::vector<Vec3>      VecVec3;
-        typedef sofa::helper::vector<Index>     VecIndex;
+        typedef sofa::type::vector<Vec3>      VecVec3;
+        typedef sofa::type::vector<Index>     VecIndex;
 
         enum { InvalidID = sofa::core::topology::Topology::InvalidID };
 

### 2021-08-17 src/SofaShells/misc/SurfaceParametrization.h

@@ -14,9 +14,9 @@
 #include <SofaBaseTopology/TriangleSetTopologyContainer.h>
 
 #include <sofa/defaulttype/VecTypes.h>
-#include <sofa/defaulttype/Vec.h>
+#include <sofa/type/Vec.h>
 
-#include <sofa/helper/vector.h>
+#include <sofa/type/vector.h>
 
 namespace sofa
 {
@@ -34,11 +34,11 @@ class SurfaceParametrization
 {
     public:
 
-        typedef sofa::defaulttype::Vec<2, Real> Vec2;
-        typedef sofa::defaulttype::Vec<3, Real> Vec3;
+        typedef sofa::type::Vec<2, Real> Vec2;
+        typedef sofa::type::Vec<3, Real> Vec3;
 
-        typedef sofa::defaulttype::Mat<2,2, Real> Mat22;
-        typedef sofa::defaulttype::Mat<3,3, Real> Mat33;
+        typedef sofa::type::Mat<2,2, Real> Mat22;
+        typedef sofa::type::Mat<3,3, Real> Mat33;
 
         typedef sofa::component::topology::TriangleSetTopologyContainer::Edge                   Edge;
         typedef sofa::component::topology::TriangleSetTopologyContainer::TriangleID             Index;
@@ -49,10 +49,10 @@ class SurfaceParametrization
         typedef sofa::component::topology::TriangleSetTopologyContainer::SeqEdges               SeqEdges;
         typedef sofa::component::topology::TriangleSetTopologyContainer::SeqTriangles           SeqTriangles;
 
-        typedef sofa::helper::vector<Vec2>      VecVec2;
-        typedef sofa::helper::vector<Vec3>      VecVec3;
-        typedef sofa::helper::vector<Index>     VecIndex;
-        typedef sofa::helper::vector<Mat22>     VecMat22;
+        typedef sofa::type::vector<Vec2>      VecVec2;
+        typedef sofa::type::vector<Vec3>      VecVec3;
+        typedef sofa::type::vector<Index>     VecIndex;
+        typedef sofa::type::vector<Mat22>     VecMat22;
 
         enum { InvalidID = sofa::core::topology::Topology::InvalidID };
 
@@ -75,8 +75,8 @@ class SurfaceParametrization
          * @{
          */
         void pointAdd(unsigned int pointIndex, const sofa::core::topology::Point &elem,
-            const sofa::helper::vector< unsigned int > &ancestors,
-            const sofa::helper::vector< double > &coeffs);
[…]

### 2021-08-17 src/SofaShells/shells2/fem/BezierShellInterpolation.h

@@ -20,8 +20,8 @@
 #include <SofaBaseTopology/TopologyData.h>
 #include <sofa/simulation/AnimateBeginEvent.h>
 
-#include <sofa/helper/vector.h>
-#include <sofa/defaulttype/Vec.h>
+#include <sofa/type/vector.h>
+#include <sofa/type/Vec.h>
 //#include <sofa/defaulttype/Mat.h>
 
 #include <sofa/core/objectmodel/BaseObject.h>
@@ -53,23 +53,23 @@ class BezierShellInterpolation : public virtual sofa::core::objectmodel::BaseObj
 
         typedef sofa::Index Index;
         typedef sofa::core::topology::BaseMeshTopology::TriangleID ElementID;
-        typedef helper::vector<Index> VecIndex;
+        typedef type::vector<Index> VecIndex;
 
         typedef typename  sofa::defaulttype::SolidTypes<Real>::Transform Transform;
         typedef typename  sofa::defaulttype::SolidTypes<Real>::SpatialVector SpatialVector;
 
-        typedef sofa::defaulttype::Vec<2, Real> Vec2;
-        typedef sofa::defaulttype::Vec<3, Real> Vec3;
-        typedef sofa::defaulttype::Mat<3,3,Real> Mat33;
-        typedef helper::vector<Vec3> VecVec3;
+        typedef sofa::type::Vec<2, Real> Vec2;
+        typedef sofa::type::Vec<3, Real> Vec3;
+        typedef sofa::type::Mat<3,3,Real> Mat33;
+        typedef type::vector<Vec3> VecVec3;
 
         // These vector is related to Bézier nodes
-        typedef helper::vector<sofa::defaulttype::Vec<3, double> > VecVec3d;
+        typedef type::vector<sofa::type::Vec<3, double> > VecVec3d;
 
-        typedef sofa::defaulttype::Vec<10,Real>     ShapeFunctions;
-        typedef sofa::defaulttype::Vec<10,Index>    BTri;
-        typedef helper::vector<ShapeFunctions>      VecShapeFunctions;
-        typedef helper::vector<BTri>                VecBTri;
+        typedef sofa::type::Vec<10,Real>     ShapeFunctions;
+        typedef sofa::type::Vec<10,Index>    BTri;
+        typedef type::vector<ShapeFunctions>      VecShapeFunctions;
+        typedef type::vector<BTri>                VecBTri;
 
         class PointInformation
         {
[…]

### 2021-08-17 src/SofaShells/shells2/fem/BezierShellInterpolationM.h

@@ -22,8 +22,8 @@
 #include <SofaBaseTopology/TopologyData.h>
 #include <sofa/simulation/AnimateBeginEvent.h>
 
-#include <sofa/helper/vector.h>
-#include <sofa/defaulttype/Vec.h>
+#include <sofa/type/vector.h>
+#include <sofa/type/Vec.h>
 
 #include <sofa/core/objectmodel/BaseObject.h>
 

### 2021-08-17 src/SofaShells/shells2/forcefield/BezierShellForceField.h

@@ -53,8 +53,8 @@ namespace component
 namespace forcefield
 {
 
-using namespace sofa::defaulttype;
-using sofa::helper::vector;
+using namespace sofa::type;
+using sofa::type::vector;
 using namespace sofa::component::topology;
 using namespace sofa::core::behavior;
 
@@ -82,22 +82,22 @@ class BezierShellForceField : public core::behavior::ForceField<DataTypes>
 
         typedef Vec<3,Real> Vec3;
         typedef Vec<2,Real> Vec2;
-        typedef helper::vector<Vec3> VecVec3;
+        typedef type::vector<Vec3> VecVec3;
 
         typedef Mat<2,2,Real> Mat22;
         typedef Mat<3,3,Real> Mat33;
 
-        typedef helper::Quater<Real> Quat;
+        typedef Quat<Real> Quat;
 
         typedef Data<VecCoord>                              DataVecCoord;
         typedef Data<VecDeriv>                              DataVecDeriv;
 
-        typedef Vec3Types::VecCoord VecCoordHigh;
+        typedef sofa::defaulttype::Vec3Types::VecCoord VecCoordHigh;
 
         typedef sofa::Index Index;
         typedef sofa::core::topology::BaseMeshTopology::Triangle Triangle;
         typedef sofa::core::topology::BaseMeshTopology::SeqTriangles SeqTriangles;
-        typedef helper::vector<Index> VecIndex;
+        typedef type::vector<Index> VecIndex;
 
         typedef typename sofa::defaulttype::SolidTypes<Real>::Transform Transform;
         typedef typename sofa::defaulttype::SolidTypes<Real>::SpatialVector SpatialVector;
@@ -134,11 +134,11 @@ public:
         {
             public:
 
-                helper::fixed_array <Vec3, 3> restLocalPositions;
+                type::fixed_array <Vec3, 3> restLocalPositions;
 #ifdef CRQUAT
-                helper::fixed_array <Quat, 3> restLocalOrientationsInv;
+                type::fixed_array <Quat, 3> restLocalOrientationsInv;
 #else
-                helper::fixed_array <Transformation, 3> restLocalOrientationsInv;
+                type::fixed_array <Transformation, 3> restLocalOrientationsInv;
 #endif
 
                 // Index of this element
@@ -164,7 +164,7 @@ public:
                 Mat33 interpol;
 
                 // Nodes of the Bezier triangle
-                helper::fixed_array<Vec3, 10> pts;          // ... in local frame
[…]

### 2021-08-17 src/SofaShells/shells2/mapping/BezierShellMechanicalMapping.h

@@ -29,7 +29,7 @@
 #include <sofa/core/Mapping.h>
 
 #include <sofa/core/behavior/MechanicalState.h>
-#include <sofa/helper/vector.h>
+#include <sofa/type/vector.h>
 
 #include <sofa/gl/GLSLShader.h>
 
@@ -51,7 +51,7 @@ namespace component
 namespace mapping
 {
 
-using namespace sofa::defaulttype;
+using namespace sofa::type;
 using namespace sofa::component::topology;
 using namespace core::topology;
 using namespace sofa::core::behavior;
@@ -84,7 +84,7 @@ public:
     typedef Vec<3, Real> Vec3;
     typedef Mat<3, 3, Real> Mat33;
 
-    typedef helper::Quater<Real> Quat;
+    typedef Quat<Real> Quat;
 
     typedef typename sofa::component::fem::BezierShellInterpolationM<TIn,TOut>::ShapeFunctions ShapeFunctions;
     typedef typename sofa::component::fem::BezierShellInterpolationM<TIn,TOut>::VecShapeFunctions VecShapeFunctions;
@@ -96,7 +96,7 @@ public:
 
     enum { NIn = sofa::defaulttype::DataTypeInfo<InDeriv>::Size };
     enum { NOut = sofa::defaulttype::DataTypeInfo<OutDeriv>::Size };
-    typedef defaulttype::Mat<NOut, NIn, Real> MBloc;
+    typedef type::Mat<NOut, NIn, Real> MBloc;
     typedef sofa::component::linearsolver::CompressedRowSparseMatrix<MBloc> MatrixType;
 
     BezierShellMechanicalMapping(core::State<In>* from, core::State<Out>* to)
@@ -185,10 +185,10 @@ protected:
     typedef struct {
 
         // Nodes of the Bézier triangle
-        sofa::helper::fixed_array<Vec3, 10> bezierNodesV;
+        sofa::type::fixed_array<Vec3, 10> bezierNodesV;
 
         // Contains the list of poinst connected to this triangle
-        sofa::helper::vector<int> attachedPoints;
+        sofa::type::vector<int> attachedPoints;
 
     } TriangleInformation;
 
@@ -209,15 +209,15 @@ protected:
 
         TriangleSetTopologyContainer* topologyTarget;
 
-        helper::vector<Vec3> colourMapping;
-        helper::vector<Vec3> coloursPerVertex;
-        helper::vector<Real> vectorErrorCoarse;
[…]

### 2021-08-17 doc/Miccai2012-draft/abstract.tex

% Pediatric cardiac surgery of congenital heart defects is particularly challenging since final decisions on the surgical procedure take place just in time during open-heart operations.
% The work presented in this paper aims at proposing an interactive simulation software dedicated to preoperative planning.
% This new tool would permit to evaluate several possible surgical procedures on virtual anatomy of the patient and anticipate the decisions.
% The surgery involves a reconfiguration of the great arteries.
% During the surgery, the vessel walls are deformed, cut, and reconnected to other vessels or closed with an artificial patch.
% We propose to anticipate the geometry changes of the vessel walls using a simulation based on thin-shell elements.
% We demonstrate that our approach allows for simulating the results of low-level surgical procedures, with a unified strategy for joining the vessel models and simulate their subsequent behavior.

Congenital heart defect corrective surgeries are particularly challenging since final decisions on concrete surgical procedures to repair malformations take place just in time during open-heart surgeries. In this paper we propose an interactive simulation system for preoperative planning. The system allows to evaluate outcomes of different surgical procedures in advance to the actual operation. During surgery, blood vessel walls are deformed, cut, reconnected to other blood vessels and patches are attached to repair malformations. We use an approach for simulating the results of these geometric changes based on thin shell elements. We show that our method enables predictive simulation of any low-level surgical procedures by using a novel approach for joining of blood vessels without springs.
